### 2020-05-15 book/paper.tex
Starting to write a book...

\documentclass{book}
\usepackage{fullpage}
\usepackage{graphicx}
\usepackage{framed}
\usepackage{tcolorbox}
\usepackage{verbatim}

\widowpenalty 10000
\clubpenalty 10000
\tolerance=9999
\newcommand{\tm}{\raisebox{.9ex}{\tiny tm}}

% \renewcommand{\topfraction}{.99}
% \renewcommand{\bottomfraction}{.99}
% \renewcommand{\textfraction}{.01}
% \renewcommand{\floatpagefraction}{.9}
% \renewcommand{\dbltopfraction}{.99}
% \renewcommand{\dblfloatpagefraction}{.9}

\newenvironment{code}{
\tcolorbox
}{
\endtcolorbox
}

\title{Introduction Concurrent Programming with CXL}
\author{RVR and others}

\begin{document}

\maketitle

\chapter{Introduction to CXL}

CXL is a concurrent programming language.  It was designed to teach
the basics of concurrent programming, but it is also useful for
testing new concurrent algorithms or even sequential and distributed
algorithms.  CXL programs are not intended to be ``run'' like programs
in most other programming languages---instead CXL programs are
``model checked'' to verify that the program has certain desirable
properties and does not suffer from bugs.  Model checking is
limited---it can only check programs that have ``finite state
spaces.''  But even programs that have infinite state spaces can
often be constrained (for example, by specifying a finite set of
possible inputs) so that model checking is possible.  If, for
example, a sorting algorithm is shown to work for any subset of
some set of say, 10, different values, then one can hope that it
will also work for sets that are larger than 10 values.

The syntax and semantics of CXL is similar to that of Python.
Python is familiar to many programmers and is easy to learn and
use.  We will assume that the reader is familiar with the basics
of Python programming.  We also will assume that the reader
understand some basics of machine architecture and how programs
are executed.

Like Python, CXL is an imperative,
dynamically typed and garbage collected programming language.
There are also some important differences:
\begin{itemize}
\item Every statement in CXL must be terminated by a semicolon
(even \textbf{while} statements and \textbf{def} statements);
in Python semicolons are optional and rarely used;
\item Correct indentation in CXL is encouraged but optional;
\item Python is object-oriented; CXL is not;
\item CXL does not support floating point or lists natively;
\item CXL does not support I/O of any kind;
\item CXL does not support operator precedence or associativity.  Use
parentheses liberally to remove ambiguity.
\end{itemize}
There are also many unimportant ones that you will discover as
you get more familiar with programming in CXL.

\begin{figure}
\begin{code}
\verbatiminput{triangle.cxl}
\end{code}
\caption{Computing triangle numbers.}
\label{fig:triangle}
\end{figure}

Figure~\ref{fig:triangle} gives a simple example of a CXL program.
It is sequential and has a method \texttt{square} that takes
an integer number as argument.  Each method has a variable called
\texttt{result} that eventually contains the result of the
method (there is no \texttt{return} statement in CXL).  The method
also has a variable called \texttt{n} containing the value of the
argument.  The $x..y$ notation generates a set containing the numbers
from $x$ to $y$ (inclusive).  The last two lines in the program are
the most interesting.
They assign to $x$ some unspecified value in the range $0..N$, and then
verifies the \texttt{assert} statement that $\mathtt{triangle}(x)$
equals $x(x+1)/2$.
If the verification fails, the value of $x$ is reported.
(The syntax of a CXL \texttt{assert} statement is the same as in Python
except that the ``,'' clause is not optional.)

``Running'' this CXL program will try all possible executions, which
includes all possible values for $x$.  Essentially, the $\texttt{choose}(S)$
operator provides the input to the program by selecting some value from the
set $S$, while the $\texttt{assert}$ statement checks that the output is
correct.  If the program is correct, the output of CXL is the size of the
``state graph'' and its diameter (more on this later).  If not, CXL also
reports what went wrong, typically by displaying a summary of an execution in
which something went wrong.

\begin{quote}
Try it out!
\begin{itemize}
\item See what happens if, instead of initializing \texttt{result} to 0,
you initialize it to 1.
\item Write a similar program that computes squares.
\end{itemize}
\end{quote}

\chapter{The Problem of Concurrent Programming}

\begin{figure}
\begin{code}
\verbatiminput{Count.cxl}
\end{code}
\caption{Counting up and down.}
\label{fig:updown}
\end{figure}

Concurrent programming, aka multithreaded programming, involves multiple
processes running in parallel while sharing variables.
Figure~\ref{fig:updown} presents a simple example.  The program defines two
methods \texttt{up()} and \texttt{down()}.
It also initializes three shared
variables: \texttt{count}, \texttt{upDone}, and \texttt{downDone}.
The \texttt{up()} method increments \texttt{count} and sets the \texttt{upDone}
flag.  The \texttt{down()} method is similar.
Next the program spawns two processes, one that invokes \texttt{up} and the
other \texttt{down}.  Afterward there are three processes---after
all, there was already one, and it spawned two new ones.  We will call
the one that was already there the \emph{main process}.

The main process waits for the two other processes to finish by checking
the \texttt{upDone} and \texttt{downDone} flags, simply in a loop.
Finally, it verifies that the value of \texttt{count} equals 0.  If not,
it reports the actual value of \texttt{count}.

\begin{quote}
What will happen?
\begin{itemize}
\item Before you run the program, what do you think will happen?  Is the
program correct in that \texttt{count} will end up 0 after executing
\texttt{up} and \texttt{down} in parallel?
\item What are the possible values of \texttt{count} at the time of the
\texttt{assert} statement?
(You may assume that \texttt{load} and \texttt{store} instructions of the
underlying architecture are atomic---in fact they are.)
\end{itemize}
\end{quote}

What is going on is that the CXL program is compiled to machine instructions,
and it is the machine instructions that are executed by the underlying
machine.  The machine has instructions that load values from memory and store
values into memory, but it does not have instructions to atomically increment
or decrement values in memory locations.  So to increment a value in memory,
the machine must do at least three machine instructions.  Conceptually:
\begin{enumerate}
\item load the value from the memory location;
\item add 1 to the value;
\item store the value to the memory location.
\end{enumerate}

When running multiple processes, each essentially runs an instantiation of
the machine, and they do so in parallel.  As they execute, their machine
instructions are interleaved in unspecified and often random ways.
In fact, CXL will try \emph{all possible interleavings} of the processes
executing machine instructions.
And thus, the following is a possible interleaving of
the processes \texttt{up} and \texttt{down}:
\begin{enumerate}
\item \texttt{up} loads the value of \texttt{count}, which is 0;
\item \texttt{down} loads the value of \texttt{count}, which is still 0;
\item \texttt{down} subtracts one from the value that it loaded (0), and
stores $-1$ into \texttt{count};
\item \texttt{up} adds one to the value that it loaded (0), and
stores $1$ into \texttt{count};
\item \texttt{up} sets \texttt{upDone};
\item \texttt{down} sets \texttt{downDone}.
\end{enumerate}

The result in this particular interleaving is that \texttt{count} ends up
being 1.  Other interleavings can result in the value 0 or $-1$.  For the
CXL machine, those are all possible results.  When running CXL, it will
report the violations of the assertions.  It also provides an example
of an interleaving, like the one above, of how exactly the assertion
ended up failing.

If one thinks of the assertion as providing the specification of the
program, then clearly its implementation does not satisfy its specification.
Either the specification or the implementation (or both) must have a bug.
We could change the specification by changing the assertion as follows:

\begin{code}
\begin{verbatim}
    assert ((-1) <= count) and (count <= 1), count;
\end{verbatim}
\end{code}

This would fix the issue, but more likely it is the program that must
be fixed.

\begin{quote}
Can you think of a fix to the program?  You can try it out and see
if it works or not.
\end{quote}

\chapter{The CXL Machine}

Before we delve into how to solve synchronization problems, it is important
to know a bit about the underlying machine.  A CXL program is translated
into a list of machine instructions that the machine executes.
The CXL machine is not an ordinary virtual machine, but its architecture
is nonetheless representative of conventional machines such as hardware CPUs
and virtual machines such as the Java Virtual Machine.

A CXL machine has the following components:
\begin{itemize}
\item The code.  This is an immutable list of CXL machine instructions.  The
instructions will be described later.
\item The memory.  This is a tree-shaped recursive data structure.  Each node
in the tree, including the leaves and the root, are \emph{CXL values}.  CXL
values include booleans, integers, strings, sets of CXL values, and, importantly
\emph{dictionaries} that maps CXL values to other CXL values.
The root of the tree is a dictionary that maps strings to CXL values.
\item The set of processes.  Initially there is just the main process, but any process
can spawn an unbounded number of other processes and processes may terminate.
Each process has an immutable
tag, a stack of CXL values, and a mutable \emph{register}, which contains a
dictionary that maps strings to CXL values.
\end{itemize}

\end{document}

### 2020-05-15 book/paper.tex
Latest

@@
 The CXL machine is not an ordinary virtual machine, but its architecture
-is nonetheless representative of conventional machines such as hardware CPUs
-and virtual machines such as the Java Virtual Machine.
+is nonetheless representative of conventional machines such as hardware
+with memory and CPUs and virtual machines such as the Java Virtual Machine.
+
+Instead of bits and bytes, a CXL machine manipulates \emph{CXL values}.
+CXL values are recursively defined: they include booleans, integers, strings,
+sets of CXL values, and \emph{dictionaries} that map CXL values to other CXL
+values.
 
@@
 \begin{itemize}
-\item The code.  This is an immutable list of CXL machine instructions.  The
+\item Code:  This is an immutable list of CXL machine instructions.  The
 instructions will be described later.
-\item The memory.  This is a tree-shaped recursive data structure.  Each node
-in the tree, including the leaves and the root, are \emph{CXL values}.  CXL
-values include booleans, integers, strings, sets of CXL values, and, importantly
-\emph{dictionaries} that maps CXL values to other CXL values.
-The root of the tree is a dictionary that maps strings to CXL values.
-\item The set of processes.  Initially there is just the main process, but any process
+\item Shared memory: A CXL machine has just one memory location containing
+a CXL.
+\item Processes:  Initially there is just the main process, but any process
 can spawn an unbounded number of other processes and processes may terminate.
-Each process has an immutable
-tag, a stack of CXL values, and a mutable \emph{register}, which contains a
-dictionary that maps strings to CXL values.
+Each process has an immutable tag, a program counter, a stack of CXL values,
+and a single mutable general purpose \emph{register} that contains a CXL.
 \end{itemize}
 
+The state of a process is called a \emph{context}: it contains the value of
+its program counter, stack, and register.  The state of the CXL machine
+consists of the value of its memory and the multiset (or \emph{bag}) of
+contexts.  It is a multiset of contexts because two different processes can
+be in the same state.
+
+It may seem strange that there is only one memory location and that each
+process has only one register.  However, this is not a limitation because
+CXL values are unbounded trees.
+By convention, both the memory and the register of a process always contain
+a special type of CXL value in which the root is a dictionary
+that maps strings to CXL values.  It represents the state of a collection of
+variables named by the strings.
+
 \end{document}

### 2020-05-15 book/paper.tex
Latest

@@
 
+\begin{figure}
+\begin{code}
+\begin{verbatim}
+Count.cxl:1 def up():
+    0 Jump 22
+    1 Frame ('up', 'Count.cxl', 1, 5) None 10
+Count.cxl:2     count = count + 1;
+    2 Constant (1, 'Count.cxl', 2, 21)
+    3 Load ('count', 'Count.cxl', 2, 13)
+    4 Nary ('+', 'Count.cxl', 2, 19) 2
+    5 Name ('count', 'Count.cxl', 2, 5)
+    6 Store 1
+Count.cxl:3     upDone = True;
+    7 Constant (True, 'Count.cxl', 3, 14)
+    8 Name ('upDone', 'Count.cxl', 3, 5)
+    9 Store 1
+   10 Return
+   11 ...
+\end{verbatim}
+\end{code}
+\caption{The first part of the machine code corresponding to Figure~\ref{fig:updown}.}
+\label{fig:updowncode}
+\end{figure}
+
+\begin{figure}
+\begin{code}
+\begin{verbatim}
+Assertion failed ('assert', 'Count.cxl', 17, 1) 1
+
+==== Safety violation ====
+__main__ 0 [0,22-36]     37 { 'count':  0, 'upDone': False, 'downDone': False }
+up       0 [1-5]          6 { 'count':  0, 'upDone': False, 'downDone': False }
+down     0 [12-20]       21 { 'count': -1, 'upDone': False, 'downDone': True  }
+up       0 [6-9]         10 { 'count':  1, 'upDone': True,  'downDone': True  }
+__main__ 0 [37-41,43-48] 48 { 'count':  1, 'upDone': True,  'downDone': True  }
+\end{verbatim}
+\end{code}
+\caption{The output of running Figure~\ref{fig:updown}.}
+\label{fig:updownoutput}
+\end{figure}
+
 Before we delve into how to solve synchronization problems, it is important
@@
 \begin{itemize}
-\item Code:  This is an immutable list of CXL machine instructions.  The
-instructions will be described later.
+\item Code:  This is an immutable and finite list of CXL machine instructions,
+generated from a CXL program.  The types of instructions will be described later.
 \item Shared memory: A CXL machine has just one memory location containing
@@
 a special type of CXL value in which the root is a dictionary
-that maps strings to CXL values.  It represents the state of a collection of
-variables named by the strings.
+that maps strings to CXL values.  We call this a \emph{directory}.
+A directory represents the state of a collection of variables named by the strings.
+
+Directories can be organized into a tree.  Each node in the tree is then identified
+by a sequence of strings, like a path name in the file system hierarchy.  We call
+such a sequence the \emph{address} of a CXL value.
+
+Compiling the code in Figure~\ref{fig:updown} results in the CXL machine code
+listed in Figure~\ref{fig:updowncode}.
+The CXL machine is predominantly a \emph{stack machine}.
+Most instructions pop values from the stack or push values onto the stack.
+Initially there is one process that starts executing at instruction 0.
+In this case, there is a \texttt{JUMP} instruction that sets the program
+counter to 22.  At program counter 1 is the code for the \texttt{up} method.
+All methods start with a \texttt{Frame} instruction and end with a \texttt{Return}
+instruction.
+
+The \texttt{Frame} instruction has three arguments:
+\begin{enumerate}
+\item A token containing the name of the method;
+\item A token containing the argument of the method (in this case the
+method has no named argument);
+\item A program counter of the method's \texttt{Return} instruction.
+\end{enumerate}
+
+A \emph{token} is a four-tuple containing a \emph{lexeme}, a file name, a
+line number in the file, and a column number in the file.  For example,
+\texttt{(’up’, ’Count.cxl’, 1, 5)} represents the lexeme \texttt{up} in
+file \texttt{Count.xcl}, line 1, column 5.  Tokens make it easy to relate
+the CXL machine instructions to the CXL source code.
+
+The code generated from \texttt{count := count + 1} in line 2 of
+\texttt{Count.cxl} is as follows:
+
+\begin{enumerate} \setcounter{enumi}{1}
+\item the \texttt{Constant} instruction pushes the constant 1
+onto the stack of the process.
+\item The \texttt{Load} instruction loads the shared variable
+\texttt{count} from the shared memory and pushes its value onto the
+stack.
+\item \texttt{Nary} stands for \texttt{n-ary}: in this case it is a \texttt{+}
+operation with 2 arguments.  It pops two values from the stack, adds them,
+and pushes the result back onto the stack.
+\item The \texttt{Name} instruction pushes the address of the \texttt{count}
+variable onto the stack.
+\item The \texttt{Store} instruction pops the address of a variable and pops
+a CXL value, and updates the memory by replacing the CXL value oat the given
+address with the new CXL value.
+\end{enumerate}
+
+Figure~\ref{fig:updownoutput} shows the output produced by running the
+\texttt{Count.cxl} program.
+It starts by reporting that the assertion on line 17 failed and that
+\texttt{count} has value 1 instead of 0:
+\texttt{Assertion failed (’assert’, ’Count.cxl’, 17, 1) 1}.
+Next it reports an execution that failed this assertion.  The output has
+five columns:
+\begin{enumerate}
+\item The entry point of the process (\texttt{\_\_main\_\_} for the main process);
+\item The tag of the process (in this case always 0);
+\item The sequence of program counters of the CXL machine instructions that the process executed;
+\item The current program counter of the process;
+\item The contents of the shared memory.
+\end{enumerate}
+
+If we look at the middle three rows, we see that:
+\begin{enumerate}
+\item \texttt{up} executed instructions 1 through 5, loading the value of
+\texttt{count} but stopping just before storing the value 1 into \texttt{count};
+\item \texttt{down} executed instructions 12 through 20, which is all of the
+\texttt{down} method, loading the value 0 from \texttt{count} and storing $-1$
+into \texttt{count};
+\item \texttt{up} continues execution, storing the value 1 into \texttt{count}.
+\end{enumerate}
+
+This makes precise the concurrency problem that we encountered.
 

### 2020-05-15 book/paper.tex
Better example

@@
-\documentclass{book}
+\documentclass{report}
 \usepackage{fullpage}
@@
 \begin{code}
-\verbatiminput{Count.cxl}
+\verbatiminput{Up.cxl}
 \end{code}
-\caption{Counting up and down.}
-\label{fig:updown}
+\caption{Incrementing twice in parallel.}
+\label{fig:inc}
 \end{figure}
@@
 processes running in parallel while sharing variables.
-Figure~\ref{fig:updown} presents a simple example.  The program defines two
-methods \texttt{up()} and \texttt{down()}.
-It also initializes three shared
-variables: \texttt{count}, \texttt{upDone}, and \texttt{downDone}.
-The \texttt{up()} method increments \texttt{count} and sets the \texttt{upDone}
-flag.  The \texttt{down()} method is similar.
-Next the program spawns two processes, one that invokes \texttt{up} and the
-other \texttt{down}.  Afterward there are three processes---after
+Figure~\ref{fig:inc} presents a simple example.  The program defines a
+method \texttt{up} that takes a parameter called \texttt{self}.
+It also initializes two shared variables \texttt{count} and \texttt{done}.
+Variable \texttt{done} is initialized with a dictionary that maps both 1 and 2
+to \texttt{False}.
+The \texttt{up()} method increments \texttt{count} and sets \texttt{done[self]}
+to \texttt{True}.
+Next the program spawns two processes, each invoking \texttt{up}.
+Afterward there are three processes---after
 all, there was already one, and it spawned two new ones.  We will call
@@
 
-The main process waits for the two other processes to finish by checking
-the \texttt{upDone} and \texttt{downDone} flags, simply in a loop.
-Finally, it verifies that the value of \texttt{count} equals 0.  If not,
+The main process waits for the two other processes to finish by polling
+both \texttt{done} flags.
+Finally, it verifies that the value of \texttt{count} equals 2.  If not,
 it reports the actual value of \texttt{count}.
@@
 \item Before you run the program, what do you think will happen?  Is the
-program correct in that \texttt{count} will end up 0 after executing
-\texttt{up} and \texttt{down} in parallel?
+program correct in that \texttt{count} will end up being 2?
 \item What are the possible values of \texttt{count} at the time of the
@@
 and it is the machine instructions that are executed by the underlying
-machine.  The machine has instructions that load values from memory and store
-values into memory, but it does not have instructions to atomically increment
-or decrement values in memory locations.  So to increment a value in memory,
+machine.  The details of this appear in the next chapter, but suffice it to
+say that The machine has instructions that load values from memory and store
+values into memory.  Importantly, it does not have instructions to atomically
+increment or decrement values in memory locations.
+So to increment a value in memory,
 the machine must do at least three machine instructions.  Conceptually:
@@
 And thus, the following is a possible interleaving of
-the processes \texttt{up} and \texttt{down}:
+the processes 1 and 2:
 \begin{enumerate}
-\item \texttt{up} loads the value of \texttt{count}, which is 0;
-\item \texttt{down} loads the value of \texttt{count}, which is still 0;
-\item \texttt{down} subtracts one from the value that it loaded (0), and
-stores $-1$ into \texttt{count};
-\item \texttt{up} adds one to the value that it loaded (0), and
+\item Process 1 loads the value of \texttt{count}, which is 0;
+\item Process 2 loads the value of \texttt{count}, which is still 0;
+\item Process 2 adds one to the value that it loaded (0), and
 stores $1$ into \texttt{count};
-\item \texttt{up} sets \texttt{upDone};
-\item \texttt{down} sets \texttt{downDone}.
+\item Process 1 adds one to the value that it loaded (0), and
+stores $1$ into \texttt{count};
+\item Process 1 sets \texttt{upDone} to \texttt{True};
+\item Process 2 sets \texttt{upDone} to \texttt{True}.
 \end{enumerate}
@@
 The result in this particular interleaving is that \texttt{count} ends up
-being 1.  Other interleavings can result in the value 0 or $-1$.  For the
-CXL machine, those are all possible results.  When running CXL, it will
+being 1.  When running CXL, it will
 report the violations of the assertions.  It also provides an example
@@
 \begin{verbatim}
-    assert ((-1) <= count) and (count <= 1), count;
+    assert (count == 0) or (count == 1), count;
 \end{verbatim}
@@
 \begin{verbatim}
-Count.cxl:1 def up():
-    0 Jump 22
-    1 Frame ('up', 'Count.cxl', 1, 5) None 10
-Count.cxl:2     count = count + 1;
-    2 Constant (1, 'Count.cxl', 2, 21)
-    3 Load ('count', 'Count.cxl', 2, 13)
-    4 Nary ('+', 'Count.cxl', 2, 19) 2
-    5 Name ('count', 'Count.cxl', 2, 5)
+Up.cxl:1 def up(self):
+   0 Jump 12
+   1 Frame ('up', 'Up.cxl', 1, 5) ('self', 'Up.cxl', 1, 8) 11
+Up.cxl:2     count = count + 1;
+   2 Constant (1, 'Up.cxl', 2, 21)
+   3 Load ('count', 'Up.cxl', 2, 13)
+   4 Nary ('+', 'Up.cxl', 2, 19) 2
+   5 Name ('count', 'Up.cxl', 2, 5)
    6 Store 1
-Count.cxl:3     upDone = True;
-    7 Constant (True, 'Count.cxl', 3, 14)
-    8 Name ('upDone', 'Count.cxl', 3, 5)
-    9 Store 1
-   10 Return
-   11 ...
+Up.cxl:3     done[self] = True;
+   7 Constant (True, 'Up.cxl', 3, 18)
+   8 LoadVar ('self', 'Up.cxl', 3, 10)
+   9 Name ('done', 'Up.cxl', 3, 5)
+   10 Store 2
+   11 Return
+Up.cxl:5 count = 0;
+   12 Constant (0, 'Up.cxl', 5, 9)
+   13 Name ('count', 'Up.cxl', 5, 1)
+   14 Store 1
 \end{verbatim}
 \end{code}
-\caption{The first part of the machine code corresponding to Figure~\ref{fig:updown}.}
-\label{fig:updowncode}
+\caption{The first part of the machine code corresponding to Figure~\ref{fig:inc}.}
+\label{fig:inccode}
 \end{figure}
@@
 \begin{verbatim}
-Assertion failed ('assert', 'Count.cxl', 17, 1) 1
+Assertion failed ('assert', '<stdin>', 12, 1) 1
 
 ==== Safety violation ====
-__main__ 0 [0,22-36]     37 { 'count':  0, 'upDone': False, 'downDone': False }
-up       0 [1-5]          6 { 'count':  0, 'upDone': False, 'downDone': False }
-down     0 [12-20]       21 { 'count': -1, 'upDone': False, 'downDone': True  }
-up       0 [6-9]         10 { 'count':  1, 'upDone': True,  'downDone': True  }
-__main__ 0 [37-41,43-48] 48 { 'count':  1, 'upDone': True,  'downDone': True  }
+__main__ 0 [0,12-29]     30 { 'count': 0, 'done': { 1: False, 2: False } }
+up       1 [1-5]          6 { 'count': 0, 'done': { 1: False, 2: False } }
+up       2 [1-10]        11 { 'count': 1, 'done': { 1: False, 2: True  } }
+up       1 [6-10]        11 { 'count': 1, 'done': { 1: True,  2: True  } }
+__main__ 0 [30-37,39-44] 44 { 'count': 1, 'done': { 1: True,  2: True  } }
 \end{verbatim}
 \end{code}
-\caption{The output of running Figure~\ref{fig:updown}.}
-\label{fig:updownoutput}
+\caption{The output of running Figure~\ref{fig:inc}.}
+\label{fig:incoutput}
 \end{figure}
@@
 
-Compiling the code in Figure~\ref{fig:updown} results in the CXL machine code
-listed in Figure~\ref{fig:updowncode}.
+Compiling the code in Figure~\ref{fig:inc} results in the CXL machine code
+listed in Figure~\ref{fig:inccode}.
 The CXL machine is predominantly a \emph{stack machine}.
@@
 \item A token containing the name of the method;
-\item A token containing the argument of the method (in this case the
-method has no named argument);
+\item A token containing the name of the argument of the method;
 \item A program counter of the method's \texttt{Return} instruction.
@@
 line number in the file, and a column number in the file.  For example,
-\texttt{(’up’, ’Count.cxl’, 1, 5)} represents the lexeme \texttt{up} in
-file \texttt{Count.xcl}, line 1, column 5.  Tokens make it easy to relate
+\texttt{(’up’, ’Up.cxl’, 1, 5)} represents the lexeme \texttt{up} in
+file \texttt{Up.xcl}, line 1, column 5.  Tokens make it easy to relate
 the CXL machine instructions to the CXL source code.
@@
 The code generated from \texttt{count := count + 1} in line 2 of
-\texttt{Count.cxl} is as follows:
+\texttt{Up.cxl} is as follows:
 
@@
 
-Figure~\ref{fig:updownoutput} shows the output produced by running the
-\texttt{Count.cxl} program.
-It starts by reporting that the assertion on line 17 failed and that
+Figure~\ref{fig:incoutput} shows the output produced by running the
+\texttt{Up.cxl} program.
+It starts by reporting that the assertion on line 12 failed and that
 \texttt{count} has value 1 instead of 0:
-\texttt{Assertion failed (’assert’, ’Count.cxl’, 17, 1) 1}.
+\texttt{Assertion failed (’assert’, ’Up.cxl’, 12, 1) 1}.
 Next it reports an execution that failed this assertion.  The output has
@@
 \begin{enumerate}
-\item \texttt{up} executed instructions 1 through 5, loading the value of
-\texttt{count} but stopping just before storing the value 1 into \texttt{count};
-\item \texttt{down} executed instructions 12 through 20, which is all of the
-\texttt{down} method, loading the value 0 from \texttt{count} and storing $-1$
-into \texttt{count};
-\item \texttt{up} continues execution, storing the value 1 into \texttt{count}.
+\item Process 1 executed instructions 1 through 5, loading the value of
+\texttt{count} but stopping just before storing 1 into \texttt{count};
+\item Process 2 executed instructions 1 through 10, storing 1 into
+into \texttt{count} and storing \texttt{True} into \texttt{done[2]};
+\item Process 1 continues execution, storing value 1 into \texttt{count}
+and storing \texttt{True} into \texttt{done[1]}.
 \end{enumerate}
@@
 
+\chapter{Naive Solutions}
+
+Hopefully you have started thinking of how to solve the concurrency
+problem and you may already have prototyped some solutions.
+In this chapter we will go through a few reasonable but broken attempts.
+At the heart of the problem is that we would like make sure that when
+the \texttt{count} variable is being updated that no other process is
+trying to do the same thing.  We call this a \emph{critical section}: a
+set of instructions where only one process is allowed to execute at a
+time.
+
 \end{document}

### 2020-05-15 book/paper.tex
Latest

@@
 
-\title{Introduction Concurrent Programming with CXL}
+\title{Concurrent Programming with CXL}
 \author{RVR and others}
@@
 
-\chapter{Introduction to CXL}
+\chapter{Introduction to Programming with CXL}
 
@@
 
-\chapter{Naive Solutions}
+\chapter{Naive Attempts at a Solution}
+
+\begin{figure}
+\begin{code}
+\verbatiminput{UpLock.cxl}
+\end{code}
+\caption{Naive implementation of a lock.}
+\label{fig:uplock}
+\end{figure}
+
+\begin{figure}
+\begin{code}
+\verbatiminput{UpFlags.cxl}
+\end{code}
+\caption{Naive use of flags to solve mutual exclusion.}
+\label{fig:upflags}
+\end{figure}
+
+\begin{figure}
+\begin{code}
+\begin{verbatim}
+==== Livelock ====
+__main__ 0 [0,35-63]    64 {'flags': {2: False, 1: False}, 'count': 0,
+                                'done': {2: False, 1: False}}
+up       1 [1-13,21-24] 25 {'flags': {2: False, 1: True}, 'count': 0,
+                                'done': {2: False, 1: False}}
+\end{verbatim}
+\end{code}
+\caption{Error output of Figure~\ref{fig:upflags}.}
+\label{fig:upflagsoutput}
+\end{figure}
 
@@
 
+Figure~\ref{fig:uplock} presents a solution based on a \emph{lock}.
+The idea is that the lock is a baton that at most one process can own at
+a time.
+Initially the lock is not owned, indicated by lock being \texttt{False}.
+To enter the critical section, a process waits until lock is \texttt{False}
+and then sets it to \texttt{True} to indicate that the lock has been taken.
+The process then executes the critical section.  Finally the process
+releases the lock by setting it back to \texttt{False}.
+
+Unfortunately, if we run the program, we find that the assertion still
+fails.  Diagnosing the problem, we see that the \texttt{lock} variable
+suffers from the same problem as the \texttt{count} variable: operations
+on it consist of several instructions are divisible.  It is thus possible
+for both processes to believe the lock is available and obtain the lock
+at the same time.
+
+Figure~\ref{fig:upflags} presents a solution based on each process having
+a flag indicating that it is trying to enter the critical section.
+After setting the flag, the process waits until the other process is not
+trying to enter the critical section.
+If we run this program, the assertion does not fail.  In fact, this
+solution does prevent both processes being in the critical section at
+the same time.
+However, it has a different problem called livelock: if both try to
+enter the critical section at the same time, they may end up waiting for
+one another.
+
 \end{document}

### 2020-05-16 book/paper.tex
Inserting a chapter
More book stuff

@@
 ==== Safety violation ====
-__main__ 0 [0,12-29]     30 { 'count': 0, 'done': { 1: False, 2: False } }
+__main__ () [0,12-29]     30 { 'count': 0, 'done': { 1: False, 2: False } }
 up       1  [1-5]          6 { 'count': 0, 'done': { 1: False, 2: False } }
@@
 up       1  [6-10]        11 { 'count': 1, 'done': { 1: True,  2: True  } }
-__main__ 0 [30-37,39-44] 44 { 'count': 1, 'done': { 1: True,  2: True  } }
+__main__ () [30-37,39-44] 44 { 'count': 1, 'done': { 1: True,  2: True  } }
 \end{verbatim}
@@
 can spawn an unbounded number of other processes and processes may terminate.
-Each process has an immutable tag, a program counter, a stack of CXL values,
-and a single mutable general purpose \emph{register} that contains a CXL.
+Each process has an immutable \emph{name tag} (specified when spawning the
+process), a program counter, a stack of CXL values, and a single mutable
+general purpose \emph{register} that contains a CXL.
 \end{itemize}
@@
 The state of a process is called a \emph{context}: it contains the value of
-its program counter, stack, and register.  The state of the CXL machine
+its name tag, program counter, stack, and register.
+The state of the CXL machine
 consists of the value of its memory and the multiset (or \emph{bag}) of
@@
 be in the same state.
+The initial state of the CXL memory is \texttt{dict\{\}}.
+The context bag has an initial context in it with name tag
+\texttt{dict\{ .name: "\_\_main\_\_", .tag: () \}}, pc 0,
+register \texttt{dict\{\}}, and an empty stack.
+Each machine instruction updates the state in some way.
+Figure~\ref{fig:incstate} shows an example of a reachable state for 
+the program in Figure~\ref{fig:inc}.
 
@@
 
-\chapter{Naive Attempts at a Solution}
+\chapter{Critical Sections}
+
+\begin{figure}
+\begin{code}
+\verbatiminput{cs.cxl}
+\end{code}
+\caption{A barebones critical section.}
+\label{fig:cs}
+\end{figure}
+
+Hopefully you have started thinking of how to solve the concurrency
+problem and you may already have prototyped some solutions.
+In this chapter we will go through a few reasonable but broken attempts.
+At the heart of the problem is that we would like make sure that when
+the \texttt{count} variable is being updated that no other process is
+trying to do the same thing.  We call this a \emph{critical section}: a
+set of instructions where only one process is allowed to execute at a
+time.
+
+A critical section is often modeled as processes in an infinite loop
+entering and exiting the critical section.  In CXL:
+
+\begin{code}
+\begin{verbatim}
+while True:
+    # enter critical section
+    @cs: pass;  # in critical section
+    # exit critical section
+;
+\end{verbatim}
+\end{code}
+
+Here \texttt{@cs} is a \emph{label}, identifying a location in the
+CXL machine code.  The first thing we need to ensure is that there
+cannot be two processes at the critical section.  We would like to
+place an assertion at the \texttt{@cs} label that specifies that
+only the current process can be there.  CXL in fact supports this.
+It has a method \texttt{atLabel($L$)}, where $L$ is the string name
+of a label (in this case, \texttt{"cs"}).  The method returns a
+bag of name tags of processes executing at that label.  The bag is
+represented by a dictionary that maps each element in the bag to the
+number of times the element appears in the bag.
+
+Figure~\ref{fig:cs} shows the code.
+The assertion also makes use of the method \texttt{nametag()} that
+returns the name tag of the current process.
+If you run the code through CXL, the assertion should fail because
+there is no code yet for entering and exiting the critical section.
+
+\chapter{Naive Attempts at Implementing Critical Sections}
 
@@
 ==== Livelock ====
-__main__ 0 [0,35-63]    64 {'flags': {2: False, 1: False}, 'count': 0,
+__main__ () [0,35-63]    64 {'flags': {2: False, 1: False}, 'count': 0,
                                 'done': {2: False, 1: False}}
@@
 \end{figure}
-
-Hopefully you have started thinking of how to solve the concurrency
-problem and you may already have prototyped some solutions.
-In this chapter we will go through a few reasonable but broken attempts.
-At the heart of the problem is that we would like make sure that when
-the \texttt{count} variable is being updated that no other process is
-trying to do the same thing.  We call this a \emph{critical section}: a
-set of instructions where only one process is allowed to execute at a
-time.
 

### 2020-05-17 book/paper.tex
Added text about progress property

@@
 while True:
+    @ncs: pass; # in non-critical section
     # enter critical section
@@
 
-Here \texttt{@cs} is a \emph{label}, identifying a location in the
-CXL machine code.  The first thing we need to ensure is that there
-cannot be two processes at the critical section.  We would like to
-place an assertion at the \texttt{@cs} label that specifies that
-only the current process can be there.  CXL in fact supports this.
-It has a method \texttt{atLabel($L$)}, where $L$ is the string name
-of a label (in this case, \texttt{"cs"}).  The method returns a
-bag of name tags of processes executing at that label.  The bag is
-represented by a dictionary that maps each element in the bag to the
-number of times the element appears in the bag.
+Here \texttt{@ncs} and \texttt{@cs} are \emph{labels}, each identifying
+a location in the CXL machine code.  The first thing we need to
+ensure is that there cannot be two processes at the critical section.
+We would like to place an assertion at the \texttt{@cs} label that
+specifies that only the current process can be there.  CXL in fact
+supports this.  It has a method \texttt{atLabel($L$)}, where $L$
+is the string name of a label (in this case, \texttt{"cs"}).  The
+method returns a bag of name tags of processes executing at that
+label.  The bag is represented by a dictionary that maps each element
+in the bag to the number of times the element appears in the bag.
 
@@
 there is no code yet for entering and exiting the critical section.
+
+However, mutual exclusion by itself easy to ensure.
+For example, we could insert the following code to enter the
+critical section:
+\begin{code}
+\begin{verbatim}
+while True: pass;;
+\end{verbatim}
+\end{code}
+This code will surely prevent two or more processes from being
+at label \texttt{cs} at the same time.
+But it does so by preventing \emph{any} process from reaching
+the critical section.
+The point is that we need another property besides mutual exclusion.
+
+Mutual exclusion is an example of a \emph{safety property}, a
+property that ensures that \emph{nothing bad will happen}, in this case
+two processes being in the critical section. Now
+we need a \emph{liveness property}: we want to ensure that
+\emph{eventually something good will happen}.
+There are various way of specifying this,
+but here we will propose the following informally: when nobody is in the
+critical section, then any process should be able to move forward and
+enter the critical section.  We call this \emph{progress}.
 

### 2020-05-17 book/paper.tex
Slight rewriting

@@
 the critical section.
-The point is that we need another property besides mutual exclusion.
+We clearly need another property besides mutual exclusion.
 
@@
 property that ensures that \emph{nothing bad will happen}, in this case
-two processes being in the critical section. Now
-we need a \emph{liveness property}: we want to ensure that
+two processes being in the critical section.
+What we need now a \emph{liveness property}: we want to ensure that
 \emph{eventually something good will happen}.
-There are various way of specifying this,
+There are various possible liveness properties we could use,
 but here we will propose the following informally: when nobody is in the

### 2020-05-17 book/paper.tex
Working on progress property

@@
 There are various possible liveness properties we could use,
-but here we will propose the following informally: when nobody is in the
-critical section, then any process should be able to move forward and
+but here we will propose the following informally: when no process is in the
+critical section, then any process that tries should be able to
 enter the critical section.  We call this \emph{progress}.

### 2020-05-17 book/paper.tex
Refining the progress property

@@
 There are various possible liveness properties we could use,
-but here we will propose the following informally: when no process is in the
-critical section, then any process that tries should be able to
-enter the critical section.  We call this \emph{progress}.
+but here we will propose the following informally: when every process
+that is in the critical section eventually leaves it, then any process
+that tries to enter the critical section should eventually be able to do so.
+We call this \emph{progress}.
 

### 2020-05-17 cxl.py

@@ -2083,29 +2083,30 @@ def run(code, labels, invariant, pcs):
             print_shortest(visited, bad)
 
     # Check progress property
-    # First look for states where all contexts are at label "@ncs".
-    print("LABELS", labels)
+    # First look for states where all "p" contexts but one are at label "@ncs".
     ncs = labels["ncs"]
     idle = set()
     for (s, node) in visited.items():
-        all = True
+        notthere = 0
+        pick = None
         for ctx in s.ctxbag.keys():
-            if ctx.pc != ncs:
-                all = False
-                break
-        if all:
-            idle.add(s)
+            if ctx.nametag.d["name"] == "p" and ctx.pc != ncs:
+                notthere += 1
+                if notthere > 1:
+                    break
+                pick = ctx
+        if notthere == 1:
+            idle.add((s, pick))
     bad = set()
     cs = labels["cs"]
+    print("IDLE", len(idle))
     if len(idle) > 0:
-        # Make sure that from each idle state each process can reach "cs"
-        for s in idle:
-            for ctx in s.ctxbag.keys():
-                if not canReach(visited, s, ctx, cs, set()):
-                    bad.add(s)
+        for (s, pick) in idle:
+            if not canReach(visited, s, pick, cs, set()):
+                bad.add(s)
     if len(bad) > 0:
-            print("==== No Progress ====", len(bad))
-            print_shortest(visited, bad)
+        print("==== No Progress ====", len(bad))
+        print_shortest(visited, bad)
 
     return visited
 

### 2020-05-17 book/paper.tex
Further refined

--- 2020-05-17 book/paper.tex
@@
 There are various possible liveness properties we could use,
-but here we will propose the following informally: when no process is in the
-critical section, then any process that tries should be able to
-enter the critical section.  We call this \emph{progress}.
+but here we will propose the following informally: when every process
+that is in the critical section eventually leaves it and only one process
+tries to enter the critical section, then it should eventually be able to do so.
+We call this \emph{progress}.
 

### 2020-05-17 book/paper.tex
Latest version of progress property

--- 2020-05-17 book/paper.tex
@@
 There are various possible liveness properties we could use,
-but here we will propose the following informally: when no process is in the
-critical section, then any process that tries should be able to
-enter the critical section.  We call this \emph{progress}.
+but here we will propose the following informally: if some process $p$ is
+in the critical section but eventually leaves, then some process that
+tries to enter the critical section should eventually be able to do so even
+if $p$ never tries to enter the critical section again.
+We call this \emph{progress}.
 

### 2020-05-17 book/paper.tex
Latest version

--- 2020-05-17 book/paper.tex
@@
 There are various possible liveness properties we could use,
-but here we will propose the following informally: when every process
-that is in the critical section eventually leaves it, then any process
-that tries to enter the critical section should eventually be able to do so.
+but here we will propose the following informally: if some set of processes
+$S$ are trying to enter the critical section, then eventually one process in
+$S$ will enter the critical section if any process already in the
+intersection eventually leaves, even if no processes other than $S$ ever
+try to enter the critical section.
 We call this \emph{progress}.
+
+% from any state some process should be able to eventually enter the
+% critical section.  Without the help of any processes that reach ncs
+% (but don't start there).
 

### 2020-05-17 book/paper.tex
Latest and greatest

@@
 \begin{code}
+\verbatiminput{csbarebones.cxl}
+\end{code}
+\caption{A barebones critical section.}
+\label{fig:csbarebones}
+\end{figure}
+
+\begin{figure}
+\begin{code}
 \verbatiminput{cs.cxl}
 \end{code}
-\caption{A barebones critical section.}
+\caption{CXL model of a critical section.}
 \label{fig:cs}
@@
 
-Figure~\ref{fig:cs} shows the code.
+Figure~\ref{fig:csbarebones} shows the code.
 The assertion also makes use of the method \texttt{nametag()} that
@@
 
-% from any state some process should be able to eventually enter the
-% critical section.  Without the help of any processes that reach ncs
-% (but don't start there).
+In order to detect violations of progress, and other liveness problems in
+algorithms in general, CXL requires that every execution must be
+able to reach a state in which all processes have terminated.
+Clearly, even if mutual exclusion holds in Figure~\ref{fig:csbarebones},
+the spawned processes never terminate.  In order to resolve this, we
+will model processes in critical regions using the framework in
+Figure~\ref{fig:cs}: a process can \emph{choose} to enter a
+critical region more than once, but it can also choose to terminate, even
+without entering the critical region ever.
+Note also that we have dropped the \texttt{@ncs} label: it would serve no
+specific purpose.
+
+\begin{quote}
+Try it out!
+\begin{itemize}
+\item Plug \texttt{while True: pass;;} for entering the critical region
+in Figure~\ref{fig:cs} and run CXL.  It should print a trace
+to a state from which a terminating state cannot be reached.
+\end{itemize}
+\end{quote}
 

### 2020-05-17 book/paper.tex
Getting close to something I like

@@
 \begin{code}
-\begin{verbatim}
-==== Livelock ====
-__main__ () [0,35-63]    64 {'flags': {2: False, 1: False}, 'count': 0,
-                                'done': {2: False, 1: False}}
-up       1  [1-13,21-24] 25 {'flags': {2: False, 1: True}, 'count': 0,
-                                'done': {2: False, 1: False}}
-\end{verbatim}
+\verbatiminput{UpTurn.cxl}
 \end{code}
-\caption{Error output of Figure~\ref{fig:upflags}.}
-\label{fig:upflagsoutput}
+\caption{Naive use of turn variable to solve mutual exclusion.}
+\label{fig:upturn}
 \end{figure}
@@
 a flag indicating that it is trying to enter the critical section.
-After setting the flag, the process waits until the other process is not
-trying to enter the critical section.
+After setting the flag, the process waits until the other process
+(\texttt{1 - self}) is not trying to enter the critical section.
 If we run this program, the assertion does not fail.  In fact, this

### 2020-05-17 book/paper.tex
Latest

@@
 
-\chapter{Naive Attempts at Implementing Critical Sections}
+\chapter{Na\"{\i}ve Attempts at Implementing Critical Sections}
 
@@
 \end{code}
-\caption{Naive implementation of a lock.}
+\caption{Na\"{\i}ve implementation of a lock.}
 \label{fig:uplock}
@@
 \end{code}
-\caption{Naive use of flags to solve mutual exclusion.}
+\caption{Na\"{\i}ve use of flags to solve mutual exclusion.}
 \label{fig:upflags}
@@
 \end{code}
-\caption{Naive use of turn variable to solve mutual exclusion.}
+\caption{Na\"{\i}ve use of turn variable to solve mutual exclusion.}
 \label{fig:upturn}
@@
 
-Figure~\ref{fig:uplock} presents a solution based on a \emph{lock}.
+Figure~\ref{fig:uplock} presents a solution based on a na\"{\i}ve implementation
+of a \emph{lock}.
 The idea is that the lock is a baton that at most one process can own at
@@
 fails.  Diagnosing the problem, we see that the \texttt{lock} variable
-suffers from the same problem as the \texttt{count} variable: operations
+suffers from the same problem as the \texttt{count} variable
+in Figure~\ref{fig:inc}: operations
 on it consist of several instructions are divisible.  It is thus possible
@@
 a flag indicating that it is trying to enter the critical section.
-After setting the flag, the process waits until the other process
+A process can write its own flag and read the flag of its peer.
+After setting its flag, the process waits until the other process
 (\texttt{1 - self}) is not trying to enter the critical section.
@@
 
+The final na\"{\i}ve solution is based on a variable called \texttt{turn}
+that alternates between 0 and 1.  When \texttt{turn = $i$}, process~$i$ can
+enter the critical section, while process $1-i$ has to wait.  When done,
+process~$i$ sets \texttt{turn} to $1-i$ to give the other process an
+opportunity to enter.
+The solution has the nice properties that it satifies mutual exclution and,
+moreover, processes 0 and 1 alternate entering the critical section.
+The problem with the solution is
+that it violates the \emph{progress} property:
+if process~$i$ terminates instead of entering the critical section when it
+is process~$i$'s turn, process $1-i$ ends up waiting indefinitely for its
+turn.
+
+\chapter{Peterson's Algorithm}
+
+\begin{figure}
+\begin{code}
+\verbatiminput{../Peterson.cxl}
+\end{code}
+\caption{Peterson's Algorithm}
+\label{fig:peterson}
+\end{figure}
+
+\chapter{Spinlock}
+
 \end{document}

### 2020-05-18 cxl.py
got rid of "," part of assert

@@ -420,21 +420,25 @@ class ChooseOp(Op):
         context.pc += 1
 
 class AssertOp(Op):
-    def __init__(self, token):
+    def __init__(self, token, exprthere):
         self.token = token
+        self.exprthere = exprthere
 
     def __repr__(self):
         return "Assert"
 
     def eval(self, state, context):
-        expr = context.pop()
+        if self.exprthere:
+            expr = context.pop()
         cond = context.pop()
         assert isinstance(cond, bool)
         if not cond:
-            print("Assertion failed", self.token, expr)
+            if self.exprthere:
+                print("Assertion failed", self.token, expr)
+            else:
+                print("Assertion failed", self.token)
             state.failure = True
-        else:
-            context.pc += 1
+        context.pc += 1
 
 class PopOp(Op):
     def __init__(self):
@@ -1316,8 +1320,9 @@ class AssertAST(AST):
     def compile(self, scope, code):
         code.append(AtomicIncOp())
         self.cond.compile(scope, code)
-        self.expr.compile(scope, code)
-        code.append(AssertOp(self.token))
+        if self.expr != None:
+            self.expr.compile(scope, code)
+        code.append(AssertOp(self.token, self.expr != None))
         code.append(AtomicDecOp())
 
 class MethodAST(AST):
@@ -1615,8 +1620,13 @@ class StatementRule(Rule):
             assert isname(lexeme), t[1]
             return (ImportAST(t[1]), self.skip(token, t[2:]))
         if lexeme == "assert":
-            (cond, t) = NaryRule({","}).parse(t[1:])
-            (expr, t) = NaryRule({";"}).parse(t[1:])
+            (cond, t) = NaryRule({",", ";"}).parse(t[1:])
+            (lexeme, file, line, column) = t[0]
+            if lexeme == ",":
+                (expr, t) = NaryRule({";"}).parse(t[1:])
+            else:
+                assert lexeme == ";", t[0]
+                expr = None
             return (AssertAST(token, cond, expr), self.skip(token, t))
         return AssignmentRule().parse(t)
 

### 2020-05-18 book/paper.tex
Starting to add Peterson's proof
got rid of "," part of assert
Finished draft of Peterson chapter

@@
 \item Every statement in CXL must be terminated by a semicolon
-(even \textbf{while} statements and \textbf{def} statements);
+(even \texttt{while} statements and \texttt{def} statements);
 in Python semicolons are optional and rarely used;
@@
 \chapter{The CXL Machine}
+\label{ch:cxlmachine}
 
@@
 
+In 1981, Peterson came up with a beautiful solution to the mutual exclusion
+problem, now known as ``Peterson's Algorithm''~\cite{Peterson81}.
+The algorithm is an amalgam of the (incorrect) algorithms in
+Figures~\ref{fig:upflags} and~\ref{fig:upturn}, and is presented
+in Figure~\ref{fig:peterson}.
+A process first indicates its interest in entering the critical section
+by setting its flag.
+It then politely gives way to the other process should it also want to
+enter the critical section---if both do so at the same time one will
+win because writes to memory in CXL are atomic.
+The process continues to be polite, waiting in a \texttt{while} loop
+until either the other process is no where near the critical section
+or has given way.
+Running the algorithm though CXL shows that it satisfies both mutual
+exclusion and progress.
+
+Why does it work?  We will focus here on how one might go about proving
+mutual exclusion for an algorithm such as Peterson's.
+For that, we have to understand a little bit more about how the CXL
+machine works.
+In Chapter~\ref{ch:cxlmachine} we talked about the concept of \emph{state}:
+at any point of time the CXL machine is in a specific state.
+Everytime a process executes a CXL machine instruction, the
+state changes (if only because the program counter of the process is
+updated).  We call that a \emph{step}.  Steps in CXL are atomic.
+
+The CXL machine starts in an initial state in which there is only
+one process and its program counter is~0.  A \emph{trace} is a
+sequence of steps starting from the initial state.  When making a
+step, there are two sources of non-determinism in CXL.  One is when
+in some state more than one process can make a step.  The other is
+when a process executes a \texttt{choose} operation and there is
+more than one choice.
+
+A good way to prove mutual exclusion is through induction on
+the number of steps.
+In such a proof, we need a so-called
+\emph{inductive invariant} $I$, which is some predicate that holds
+over the state.  We need $I$ to satisfy the following:
+\begin{itemize}
+\item (ME1) $I$ should imply mutual exclusion.
+\item (ME2) $I$ should hold in the initial state.
+\item (ME3) If $I$ holds over some state $s$, and some process in the
+state takes a step, then $I$ should also hold in the resulting state.
+\end{itemize}
+
+One candidate for such an invariant is mutual exclusion itself.
+After all, it certainly satisfies ME1 and ME2.
+And as CXL has already determined, mutual exclusion is an invariant:
+it holds over every \emph{reachable state}.
+Unfortunately, it is not an \emph{inductive} invariant and therefore does
+not satisfy ME3.
+To see why, we need to consider an \emph{unreachable} state.
+It is easy to construct one: let process~0 be at label \texttt{@cs}
+and process~1 at the start of the \texttt{while} loop.
+Also imagine that in this particular state $\mathtt{turn} = 1$.  Now let
+process process~1 make a sequence of steps.  Because $\mathtt{turn} = 1$,
+process~1 will break out of the while loop and also enter the critical
+section, violating mutual exclusion.
+Doing a inductive proof with an invariant that is not inductive is usually
+much harder than doing one with an invariant that is.
+
+\begin{figure}
+\begin{code}
+\verbatiminput{../PetersonProof.cxl}
+\end{code}
+\caption{Peterson's Algorithm with Inductive Invariant}
+\label{fig:petersonproof}
+\end{figure}
+
+Let $C(\mathtt{self}) = \mathtt{flags}[1 - \mathtt{self}] \land
+\mathtt{turn} = 1 - \mathtt{self}$, that is, the condition on the \texttt{while} loop.
+For Peterson's Algorithm, an inductive invariant that works well is:
+if process $i$ is label \texttt{@cs} (i.e., process $i$ is in the critical section),
+then $C(i)$ does not hold or process $1-i$ is executing after setting
+$\mathtt{flags}[i]$ but still before setting \texttt{turn} to $1-i$.
+Let's call it $I_p(i)$.
+
+Figure~\ref{fig:petersonproof} formalizes $I_p(i)$ in CXL.
+The figure uses the CXL \texttt{atomic} statement for the first time.
+For now, all you need to know is that it turns the embedded statement into
+a single atomic step.  The label \texttt{@gate} thus refers exactly to
+this statement.  You can run Figure~\ref{fig:petersonproof} to determine
+that $I_p(i)$ is indeed an invariant for $i = 0, 1$.
+
+To see that $I_p(i)$ implies mutual exclusion, suppose not.  Then
+when both process 0 and process 1 are in the critical section, the
+following must hold:
+$(\lnot C(0) \lor \mathtt{process}(1)@gate) \land
+ (\lnot C(1) \lor \mathtt{process}(0)@gate)$.
+We know that both flags are set.
+We also know that neither process 1 nor process 2 is at label \texttt{@gate}
+(because they are both at label \texttt{@cs}),
+so this simplifies to $(\lnot C(0)) \land (\lnot C(1))$.
+So we conclude that $\mathtt{turn} = 0 \land \mathtt{turn} = 1$, a
+logical impossibility.  Thus $I_p(i)$ satisfies ME1.
+
+$I_p(i)$ satisfies ME2 trivially, because in the initial state no process
+is in the critical section.
+
+Finally we have to show that $I_p(i)$ satisfies ME3. Without loss of
+generality, suppose $i=0$ (a benefit from the fact that the algorithm is
+symmetric for both processes).  We have to show that if we are in a state
+in which $I_p(0)$ holds, then any step will result in a state in which
+$I_p(0)$ continues to hold.
+Since $I_p(0)$ holds, process 0 is at label \texttt{@cs}.  If process 0
+were to take a step, then in the next state process 0 would be no longer
+at that label and $I_p(0)$ would hold trivially over the next state.
+Therefore we only need to consider a step by process 1.
+
+From $I_p(0)$ we know that one of the following cases must hold before
+process 1 takes a step:
+\begin{itemize}
+\item \texttt{flags[1] = False};
+\item \texttt{turn = 0};
+\item process 1 is at label \texttt{@gate}.
+\end{itemize}
+
+Let us consider each of these cases.
+In the first case, if process 1 takes a step, there are two possibilities:
+either $flags[1]$ will still be \texttt{False} (in which case the first case
+continues to hold), or $flags[1]$ will be \texttt{True}
+and process 1 will be at label \texttt{@gate} (in which case the third case
+will hold).
+We know that process 1 never sets \texttt{turn} to 1, so
+if the second case holds before the step, it will also hold after the step.
+Finally, if process 1 is at label \texttt{@gate} before the step, then after
+the step \texttt{turn} will equal 0, and therefore the second case will hold
+after the step.
+
+We have now demonstrated mutual exclusion in Peterson's Algorithm in two
+different ways: one by letting CXL explore all possible executions, the
+other using an inductive invariant and proof by induction.  The former
+is certainly easier, but it does not provide intuition for why the
+algorithm works.  The second provides much more insight.  We therefore
+encourage to include inductive invariants in your CXL code.
+
+A cool anecdote is the following.  When the author of CXL had to teach
+Peterson's Algorithm, he refreshed his memory by looking at the Wikipedia
+page.  The page claimed that the following predicate was invariant:
+if process $i$ is in the critical section, then $\lnot C(i)$ (i.e.,
+$I_p(i)$ without the disjunct that process $1-i$ is at label \texttt{gate}).
+To demonstrate that this predicate is not invariant, you can remove the
+disjunct from Figure~\ref{fig:petersonproof} and run it to get a
+counterexample.
+
+This anecdote suggests the following.  If you need to do an induction
+proof of an algorithm, you have to come up with an inductive invariant.
+Before trying to prove the algorithm, you can check that the predicate is
+at least invariant by testing it using CXL.  Doing so could potentially
+avoid wasting your time on a proof that will not work because the
+predicate is not invariant, and therefore not an inductive invariant either.
+
+To finish the story, the author of CXL fixed the Wikipedia page, but there
+is another cool story.  Years later a colleague of the author, teaching
+the same course, asked if the first two assignments (setting \texttt{flags[self]})
+to \texttt{True} and \texttt{turn} to \texttt{1 - self}) can be reversed.
+After all, they are different variables assigned independent values---in a
+sequential program one could surely swap the two assignments.
+
+\begin{quote}
+What do you think?
+\begin{itemize}
+\item See if you can figure out for yourself if the two assignments can be
+reversed.  Then run the program in Figure~\ref{fig:peterson} after reversing
+the two assignments and see what happens.
+\end{itemize}
+\end{quote}
+
 \chapter{Spinlock}

### 2020-05-18 book/paper.tex
Latest

@@
 To see that $I_p(i)$ implies mutual exclusion, suppose not.  Then
-when both process 0 and process 1 are in the critical section, the
+when both process~0 and process~1 are in the critical section, the
 following must hold:
@@
 We know that both flags are set.
-We also know that neither process 1 nor process 2 is at label \texttt{@gate}
+We also know that neither process~0 nor process~1 is at label \texttt{@gate}
 (because they are both at label \texttt{@cs}),
@@
 $I_p(0)$ continues to hold.
-Since $I_p(0)$ holds, process 0 is at label \texttt{@cs}.  If process 0
-were to take a step, then in the next state process 0 would be no longer
+Since $I_p(0)$ holds, process~0 is at label \texttt{@cs}.  If process~0
+were to take a step, then in the next state process~0 would be no longer
 at that label and $I_p(0)$ would hold trivially over the next state.
-Therefore we only need to consider a step by process 1.
+Therefore we only need to consider a step by process~1.
 
 From $I_p(0)$ we know that one of the following cases must hold before
-process 1 takes a step:
+process~1 takes a step:
 \begin{itemize}
@@
 \item \texttt{turn = 0};
-\item process 1 is at label \texttt{@gate}.
+\item process~1 is at label \texttt{@gate}.
 \end{itemize}
@@
 Let us consider each of these cases.
-In the first case, if process 1 takes a step, there are two possibilities:
+In the first case, if process~1 takes a step, there are two possibilities:
 either $flags[1]$ will still be \texttt{False} (in which case the first case
 continues to hold), or $flags[1]$ will be \texttt{True}
-and process 1 will be at label \texttt{@gate} (in which case the third case
+and process~1 will be at label \texttt{@gate} (in which case the third case
 will hold).
-We know that process 1 never sets \texttt{turn} to 1, so
+We know that process~1 never sets \texttt{turn} to 1, so
 if the second case holds before the step, it will also hold after the step.
-Finally, if process 1 is at label \texttt{@gate} before the step, then after
+Finally, if process~1 is at label \texttt{@gate} before the step, then after
 the step \texttt{turn} will equal 0, and therefore the second case will hold
@@
 
+\begin{quote}
+Bonus question!
+\begin{itemize}
+\item Can you generalize Peterson's algorithm to more than two processes?
+\end{itemize}
+\end{quote}
+
 \chapter{Spinlock}
 
+Figure~\ref{fig:uplock} showed a faulty attempt at solving mutual exclusion.
+
 \end{document}

### 2020-05-18 book/paper.tex
Partially updated book to use process 0 and 1 rather than 1 and 2

@@
 Figure~\ref{fig:inc} presents a simple example.  The program defines a
-method \texttt{up} that takes a parameter called \texttt{self}.
+method \texttt{process} that takes a parameter called \texttt{self}.
 It also initializes two shared variables \texttt{count} and \texttt{done}.
@@
 to \texttt{False}.
-The \texttt{up()} method increments \texttt{count} and sets \texttt{done[self]}
+The \texttt{process()} method increments \texttt{count} and sets \texttt{done[self]}
 to \texttt{True}.
-Next the program spawns two processes, each invoking \texttt{up}.
+Next the program spawns two processes, each invoking \texttt{process}.
 Afterward there are three processes---after
@@
 And thus, the following is a possible interleaving of
-the processes 1 and 2:
+the processes 0 and 1:
 \begin{enumerate}
-\item Process 1 loads the value of \texttt{count}, which is 0;
-\item Process 2 loads the value of \texttt{count}, which is still 0;
-\item Process 2 adds one to the value that it loaded (0), and
-stores $1$ into \texttt{count};
+\item Process 0 loads the value of \texttt{count}, which is 0;
+\item Process 1 loads the value of \texttt{count}, which is still 0;
 \item Process 1 adds one to the value that it loaded (0), and
 stores $1$ into \texttt{count};
-\item Process 1 sets \texttt{upDone} to \texttt{True};
-\item Process 2 sets \texttt{upDone} to \texttt{True}.
+\item Process 0 adds one to the value that it loaded (0), and
+stores $1$ into \texttt{count};
+\item Process 0 sets \texttt{upDone} to \texttt{True};
+\item Process 1 sets \texttt{upDone} to \texttt{True}.
 \end{enumerate}
@@
 \begin{verbatim}
-Up.cxl:1 def up(self):
+Up.cxl:1 def process(self):
    0 Jump 12
-   1 Frame ('up', 'Up.cxl', 1, 5) ('self', 'Up.cxl', 1, 8) 11
+   1 Frame ('process', 'Up.cxl', 1, 5) ('self', 'Up.cxl', 1, 13) 11
 Up.cxl:2     count = count + 1;
@@
 \begin{verbatim}
-Assertion failed ('assert', '<stdin>', 12, 1) 1
+Assertion failed ('assert', 'Up.cxl', 12, 1) 1
 
@@
 be in the same state.
-The initial state of the CXL memory is \texttt{dict\{\}}.
+The initial state of the CXL memory is the empty tuple, \texttt{()}.
 The context bag has an initial context in it with name tag
 \texttt{dict\{ .name: "\_\_main\_\_", .tag: () \}}, pc 0,
-register \texttt{dict\{\}}, and an empty stack.
+register \texttt{()}, and an empty stack.
 Each machine instruction updates the state in some way.
@@
 \begin{enumerate}
-\item Process 1 executed instructions 1 through 5, loading the value of
+\item Process 0 executed instructions 1 through 5, loading the value of
 \texttt{count} but stopping just before storing 1 into \texttt{count};
-\item Process 2 executed instructions 1 through 10, storing 1 into
+\item Process 1 executed instructions 1 through 10, storing 1 into
 into \texttt{count} and storing \texttt{True} into \texttt{done[2]};
-\item Process 1 continues execution, storing value 1 into \texttt{count}
+\item Process 0 continues execution, storing value 1 into \texttt{count}
 and storing \texttt{True} into \texttt{done[1]}.

### 2020-05-18 cxl.py
Working on multi-argument methods

@@ -464,21 +464,21 @@ class MethodOp(Op):
         context.pc = self.endpc + 1
 
 class FrameOp(Op):
-    def __init__(self, name, arg, end):
+    def __init__(self, name, args, end):
         self.name = name
-        self.arg = arg
+        self.args = args
         self.end = end
 
     def __repr__(self):
-        return "Frame " + str(self.name) + " " + str(self.arg) + " " + str(self.end)
+        return "Frame " + str(self.name) + " " + str(self.args) + " " + str(self.end)
 
     def eval(self, state, context):
         arg = context.pop()
         context.push(context.vars)
-        if self.arg == None:
+        if self.args == []:
             context.vars = DictValue({ "result": novalue })
         else:
-            (lexeme, file, line, column) = self.arg
+            (lexeme, file, line, column) = self.args[0]
             context.vars = DictValue({ "result": novalue, lexeme: arg })
         context.pc += 1
 
@@ -1442,13 +1442,13 @@ class AssertAST(AST):
         code.append(AtomicDecOp())
 
 class MethodAST(AST):
-    def __init__(self, name, arg, stat):
+    def __init__(self, name, args, stat):
         self.name = name
-        self.arg = arg
+        self.args = args
         self.stat = stat
 
     def __repr__(self):
-        return "Method(" + str(self.name) + ", " + str(self.stat) + ")"
+        return "Method(" + str(self.name) + ", " + str(self.args) + ", " + str(self.stat) + ")"
 
     def compile(self, scope, code):
         pc = len(code)
@@ -1458,17 +1458,17 @@ class MethodAST(AST):
         scope.names[lexeme] = ("constant", (PcValue(pc + 1), file, line, column))
 
         ns = Scope(scope)
-        if self.arg == None:
+        if self.args == []:
             arg = None
         else:
-            (arg, afile, aline, acolumn) = self.arg
-            ns.names[arg] = ("variable", self.arg)
+            (arg, afile, aline, acolumn) = self.args[0]
+            ns.names[arg] = ("variable", self.args[0])
         ns.names["result"] = ("variable", ("result", file, line, column))
         self.stat.compile(ns, code)
         code.append(ReturnOp())
 
         code[pc+0] = JumpOp(len(code))
-        code[pc+1] = FrameOp(self.name, self.arg, len(code) - 1)
+        code[pc+1] = FrameOp(self.name, self.args, len(code) - 1)
 
[…]

### 2020-05-18 book/paper.tex
Latest
Working on multi-argument methods

@@
 some set of say, 10, different values, then one can hope that it
-will also work for sets that are larger than 10 values.
+will also work for sets that are larger than 10 values or sets that
+use more than 10 values.
 
@@
 \item Python is object-oriented; CXL is not;
-\item CXL does not support floating point or lists natively;
+\item CXL does not support floating point;
 \item CXL does not support I/O of any kind;
@@
 argument.  The $x..y$ notation generates a set containing the numbers
-from $x$ to $y$ (inclusive).  The last two lines in the program are
+from~$x$ to~$y$ (inclusive).  The last two lines in the program are
 the most interesting.
@@
 If the verification fails, the value of $x$ is reported.
-(The syntax of a CXL \texttt{assert} statement is the same as in Python
-except that the ``,'' clause is not optional.)
 
@@
 operator provides the input to the program by selecting some value from the
-set $S$, while the $\texttt{assert}$ statement checks that the output is
+set~$S$, while the $\texttt{assert}$ statement checks that the output is
 correct.  If the program is correct, the output of CXL is the size of the
@@
 It also initializes two shared variables \texttt{count} and \texttt{done}.
-Variable \texttt{done} is initialized with a dictionary that maps both 1 and 2
-to \texttt{False}.
 The \texttt{process()} method increments \texttt{count} and sets \texttt{done[self]}
@@
 machine.  The details of this appear in the next chapter, but suffice it to
-say that The machine has instructions that load values from memory and store
+say that the machine has instructions that load values from memory and store
 values into memory.  Importantly, it does not have instructions to atomically
@@
 executing machine instructions.
-And thus, the following is a possible interleaving of
-the processes 0 and 1:
+The following is a possible interleaving of processes~0 and~1:
 \begin{enumerate}
@@
 being 1.  When running CXL, it will
-report the violations of the assertions.  It also provides an example
-of an interleaving, like the one above, of how exactly the assertion
-ended up failing.
+report violations of assertions.  It also provides an example
+of an interleaving, like the one above, in which an assertion fails.
 
@@
 \begin{verbatim}
-    assert (count == 0) or (count == 1), count;
+    assert (count == 0) or (count == 1);
 \end{verbatim}
@@
 can spawn an unbounded number of other processes and processes may terminate.
-Each process has an immutable \emph{name tag} (specified when spawning the
-process), a program counter, a stack of CXL values, and a single mutable
-general purpose \emph{register} that contains a CXL.
+Each process has an immutable (but not necessarily unique) \emph{name tag},
+a program counter, a stack of CXL values,
+and a single mutable general purpose \emph{register} that contains a CXL.
 \end{itemize}
@@
 The context bag has an initial context in it with name tag
-\texttt{dict\{ .name: "\_\_main\_\_", .tag: () \}}, pc 0,
+\texttt{dict\{ .name: .\_\_main\_\_, .tag: () \}}, pc 0,
 register \texttt{()}, and an empty stack.
+(In CXL, dictionary values are enclosed by \texttt{dict\{} and \texttt{\}}.)
 Each machine instruction updates the state in some way.
@@
 
+\chapter{Using CXL Methods}
+
 \chapter{Spinlock}

### 2020-05-18 book/paper.tex
Better printing

@@
    0 Jump 12
-   1 Frame ('process', 'Up.cxl', 1, 5) ('self', 'Up.cxl', 1, 13) 11
+   1 Frame ('process', 'Up.cxl', 1, 5) [('self', '<stdin>', 1, 13)] 11
 Up.cxl:2     count = count + 1;
@@
 \begin{verbatim}
-Assertion failed ('assert', 'Up.cxl', 12, 1) 1
-
 ==== Safety violation ====
-__main__ () [0,12-29]     30 { 'count': 0, 'done': { 1: False, 2: False } }
-up       1  [1-5]          6 { 'count': 0, 'done': { 1: False, 2: False } }
-up       2  [1-10]        11 { 'count': 1, 'done': { 1: False, 2: True  } }
-up       1  [6-10]        11 { 'count': 1, 'done': { 1: True,  2: True  } }
-__main__ () [30-37,39-44] 44 { 'count': 1, 'done': { 1: True,  2: True  } }
+__main__/() [0,12-31]           32 dict{ .count:0, .done:dict{ 0:False, 1:False } }
+process/0   [1-5]                6 dict{ .count:0, .done:dict{ 0:False, 1:False } }
+process/1   [1-10]              11 dict{ .count:1, .done:dict{ 0:False, 1:True  } }
+process/0   [6-10]              11 dict{ .count:1, .done:dict{ 0:True,  1:True  } }
+__main__/() [32-34,37-42,44-48] 49 dict{ .count:1, .done:dict{ 0:True,  1:True  } }
 \end{verbatim}

### 2020-05-18 book/paper.tex
Working on book...

@@
 also has a variable called \texttt{n} containing the value of the
-argument.  The $x..y$ notation generates a set containing the numbers
-from~$x$ to~$y$ (inclusive).  The last two lines in the program are
+argument.  The \texttt{x..y} notation generates a set containing the numbers
+from~\texttt{x} to~\texttt{y} (inclusive).  The last two lines in the program are
 the most interesting.
-They assign to $x$ some unspecified value in the range $0..N$, and then
-verifies the \texttt{assert} statement that $\mathtt{triangle}(x)$
+They assign to \texttt{x} some unspecified value in the range \texttt{0..N},
+and then verifies the \texttt{assert} statement that $\mathtt{triangle}(x)$
 equals $x(x+1)/2$.
-If the verification fails, the value of $x$ is reported.
 
@@
    0 Jump 12
-   1 Frame ('process', 'Up.cxl', 1, 5) [('self', '<stdin>', 1, 13)] 11
+   1 Frame ('process', 'Up.cxl', 1, 5) [('self', 'Up.cxl', 1, 13)] 11
 Up.cxl:2     count = count + 1;
@@
 __main__/() [0,12-31]           32 dict{ .count:0, .done:dict{ 0:False, 1:False } }
-process/0   [1-5]                6 dict{ .count:0, .done:dict{ 0:False, 1:False } }
-process/1   [1-10]              11 dict{ .count:1, .done:dict{ 0:False, 1:True  } }
-process/0   [6-10]              11 dict{ .count:1, .done:dict{ 0:True,  1:True  } }
+process /0  [1-5]                6 dict{ .count:0, .done:dict{ 0:False, 1:False } }
+process /1  [1-10]              11 dict{ .count:1, .done:dict{ 0:False, 1:True  } }
+process /0  [6-10]              11 dict{ .count:1, .done:dict{ 0:True,  1:True  } }
 __main__/() [32-34,37-42,44-48] 49 dict{ .count:1, .done:dict{ 0:True,  1:True  } }
@@
 
+A name tag consists of the name of the main method of the process,
+along with the tag specified in the \texttt{spawn} statement.
+In Figure~\ref{fig:inc}, the created processes have name tags
+\texttt{process/0} and \texttt{process/1}.  The main process that
+starts the execution has name tag \texttt{\_\_main\_\_/()}.
+
 The state of a process is called a \emph{context}: it contains the value of
@@
 The context bag has an initial context in it with name tag
-\texttt{dict\{ .name: .\_\_main\_\_, .tag: () \}}, pc 0,
-register \texttt{()}, and an empty stack.
+\texttt{\_\_main\_\_/()}, pc 0, register \texttt{()}, and an empty stack.
 (In CXL, dictionary values are enclosed by \texttt{dict\{} and \texttt{\}}.)
@@
 
-\chapter{Using CXL Methods}
+\chapter{CXL Methods and Dictionaries}
+
+CXL methods look and feel much the same as Python methods, but there
+are some differences you should be aware of.
+First, as already noted, there is no \texttt{return} statement.
+A CXL method returns when the last statement in the method has
+completed.
+
+Like in Python, CXL methods have local variables.  You have already
+seen two types of local variables.
+First, the arguments to the method are variables, although read-only.
+Second, the variable \texttt{result} can be assigned to
+change the default return value of a method, which is \texttt{()}.
+You can declare additional local variables to hold intermediate
+results.  For example, you can
+declare a local variable $x$ with initial value 3 as follows:
+\begin{code}
+\begin{verbatim}
+var x = 3;
+\end{verbatim}
+\end{code}
+The initial value is mandatory.
+Unless a variable is declared this way, it will be considered
+a global shared variable.
+The local variables of a method cannot be accessed other than by
+the method itself.
+
+A method \texttt{m} with argument \texttt{a} is invoked in its
+most basic form as follows (assigning the result to \texttt{x}).
+\begin{code}
+\begin{verbatim}
+var x = m a;
+\end{verbatim}
+\end{code}
+That's right, no parentheses are required.  In fact, if you invoke
+\texttt{m(a)}, the argument is \texttt{(a)}, which is the same
+as \texttt{a}.
+If you invoke \texttt{m()}, the argument is \texttt{()},
+which is the empty tuple.  If you invoke
+If you invoke \texttt{m(a, b)}, the argument is \texttt{(a, b)},
+the tuple consisting of values \texttt{a} and \texttt{b}.
+You have already seen examples of this: the expression
+\texttt{atLabel.cs} invokes method \texttt{atLabel} with
+argument \texttt{.cs}.
+
+If you want to invoke a method but are not interested in its result,
+you have to use a \texttt{call} statement like so:
+\begin{code}
+\begin{verbatim}
+call m a;
+\end{verbatim}
+\end{code}
 
 \chapter{Spinlock}
-
 Figure~\ref{fig:uplock} showed a faulty attempt at solving mutual exclusion.

### 2020-05-18 book/paper.tex
Added chapter on CXL values.  Need more examples

@@
 and it is the machine instructions that are executed by the underlying
-machine.  The details of this appear in the next chapter, but suffice it to
+machine.  The details of this appear in the chapter~\ref{ch:cxlmachine},
+but suffice it to
 say that the machine has instructions that load values from memory and store
@@
 \end{quote}
+
+\chapter{CXL Values}
+
+CXL programs manipulates CXL values.
+CXL values are recursively defined:
+they include booleans (\texttt{False} and \texttt{True}),
+integers (but not floating point numbers),
+strings (enclosed by double quotes),
+sets of CXL values,
+and dictionaries that map CXL values to other CXL values.
+
+Another type of CXL value is the \emph{atom}.  You have already
+seen them: the names of constants, variables, and methods are
+all atoms.  An atom is denoted using a period followed by the
+name.  For example, the atom for the \texttt{process} method
+is denoted \texttt{.process}.
+
+CXL makes extensive use of dictionaries.  Their syntax and properties
+are a little different from Python.
+A dictionary maps keys to values.
+Unlike Python, any CXL value can be a key, including another
+dictionary.
+If \texttt{d} is a dictionary, and \texttt{k} is a key, then the
+following expression retrieves the CXL value that $k$ maps to:
+\begin{code}
+\begin{verbatim}
+d k
+\end{verbatim}
+\end{code}
+This is unfamiliar, but in CXL square brackets can be used in the
+same way as parentheses, so you can express it in the more familiar
+form:
+\begin{code}
+\begin{verbatim}
+d[k]
+\end{verbatim}
+\end{code}
+However, if the key is an atom, say \texttt{.count}, then you
+can write \texttt{d.count} instead of having to write
+\texttt{d[.count]} (both will work).
+
+Tuples are special forms of dictionaries where the keys are
+the indexes into the tuple.  For example, the tuple
+\texttt{(5, False)} is the same CXL value as
+\texttt{dict\{ 0: 5, 1: False \}}.
+The empty tuple \texttt{()} is the same value as \texttt{dict\{\}}.
+As in Python, you can create singleton tuples by adding an
+extra ",". For example, \texttt{(1,)}.
+
+Again, square brackets and parentheses work the same in CXL, so
+\texttt{[a, b, c]} (which looks like a Python list)
+is the same as \texttt{(a, b, c)} (which looks like a Python tuple),
+which in turn is the same as \texttt{dict\{ 0:a, 1:b, 2:c \}}.
+So if \texttt{x == [ False, True ]},
+then \texttt{x[0] == False} and \texttt{x[1] == True}.
+
+Dictionaries have two handy unary operators.  If \texttt{d} is a
+dictionary, then \texttt{keys d} (or equivalently \texttt{keys(d)})
+returns the set of keys and \texttt{len d} returns the size of
+this set.
 
@@
 
-Instead of bits and bytes, a CXL machine manipulates \emph{CXL values}.
-CXL values are recursively defined: they include booleans, integers, strings,
-sets of CXL values, and \emph{dictionaries} that map CXL values to other CXL
-values.
-
+Instead of bits and bytes, a CXL machine manipulates CXL values.
 A CXL machine has the following components:
@@
 \texttt{\_\_main\_\_/()}, pc 0, register \texttt{()}, and an empty stack.
-(In CXL, dictionary values are enclosed by \texttt{dict\{} and \texttt{\}}.)
 Each machine instruction updates the state in some way.
@@
 specifies that only the current process can be there.  CXL in fact
-supports this.  It has a method \texttt{atLabel($L$)}, where $L$
-is the string name of a label (in this case, \texttt{"cs"}).  The
+supports this.  It has a method \texttt{atLabel $L$}, where $L$
+is the atom of the label (in this case, \texttt{.cs}).  The
 method returns a bag of name tags of processes executing at that
@@
 
-\chapter{CXL Methods and Dictionaries}
+\chapter{CXL Methods and Pointers}
 

### 2020-05-18 book/paper.tex
Adding discussion of methods and pointers

@@
 \chapter{CXL Values}
+\label{ch:cxlvalues}
 
@@
 
+You may note that all this looks familiar.  Indeed, the syntax
+is the same as that for dictionaries (see Chapter~\ref{ch:cxlvalues}).
+Both dictionaries and methods map CXL values to CXL values,
+and their syntax is indistinguishable.
+If \texttt{f} is either a method or a
+dictionary, and \texttt{x} is an arbitrary CXL value, then
+then \texttt{f x}, \texttt{f(x)}, and \texttt{f[x]} are all
+the same expression in CXL.
+
+CXL is not an object-oriented language like Python is.  In Python
+you can pass a reference to an object to a method, and that method
+can then update the object.  In CXL, it is also sometimes convenient
+to have a method update a shared variable specified as an argument.
+For this (as well as some other uses), CXL supports \emph{pointers}
+to shared variables.
+If \texttt{x} is a shared variable, then the expression \texttt{\&(x)}
+(the parentheses are mandatory) is a pointer to \texttt{x}
+(also known as the \emph{address} of \texttt{x}.
+Conversely, if \texttt{p} is a pointer to a shared variable, then the
+expression \texttt{\^{}p} is the value of the shared variable.
+
 If you want to invoke a method but are not interested in its result,

### 2020-05-19 book/paper.tex
Fixed major bug in pointer handling
Fixed various typos

@@
 often be constrained (for example, by specifying a finite set of
-possible inputs) so that model checking is possible.  If, for
+test inputs) so that model checking is possible.  If, for
 example, a sorting algorithm is shown to work for any subset of
-some set of say, 10, different values, then one can hope that it
-will also work for sets that are larger than 10 values or sets that
-use more than 10 values.
+some set of say, 5, different values, then one can hope that it
+will also work for sets that are larger than 5 values or sets that
+use more than 5 different values.
 
@@
 understand some basics of machine architecture and how programs
-are executed.
+are executed.  For example, we assume that the reader is familiar
+with the concepts of CPU, memory, register, and stack.
 
@@
 \item Python is object-oriented; CXL is not;
-\item CXL does not support floating point;
+\item CXL does not (currently) support floating point;
 \item CXL does not support I/O of any kind;
@@
 the most interesting.
-They assign to \texttt{x} some unspecified value in the range \texttt{0..N},
-and then verifies the \texttt{assert} statement that $\mathtt{triangle}(x)$
-equals $x(x+1)/2$.
+They assign to \texttt{x} some unspecified value in the range \texttt{0..N}
+and then verifies that $\mathtt{triangle}(x)$ equals $x(x+1)/2$.
 
@@
 you initialize it to 1.
-\item Write a similar program that computes squares.
+\item Write a similar program that computes squares or factorials.
 \end{itemize}
@@
 It also initializes two shared variables \texttt{count} and \texttt{done}.
-The \texttt{process()} method increments \texttt{count} and sets \texttt{done[self]}
+The \texttt{process(self)} method increments \texttt{count} and sets \texttt{done[self]}
 to \texttt{True}.
@@
 \item Before you run the program, what do you think will happen?  Is the
-program correct in that \texttt{count} will end up being 2?
+program correct in that \texttt{count} will always end up being 2?
 \item What are the possible values of \texttt{count} at the time of the
@@
 What is going on is that the CXL program is compiled to machine instructions,
-and it is the machine instructions that are executed by the underlying
-machine.  The details of this appear in the chapter~\ref{ch:cxlmachine},
+and it is the machine instructions that are executed by the underlying CXL
+machine.  The details of this appear in Chapter~\ref{ch:cxlmachine},
 but suffice it to
@@
 instructions are interleaved in unspecified and often random ways.
-In fact, CXL will try \emph{all possible interleavings} of the processes
+A program is correct if it works for any interleaving.
+In fact, CXL will try all possible interleavings of the processes
 executing machine instructions.
+
 The following is a possible interleaving of processes~0 and~1:
@@
 \begin{verbatim}
-    assert (count == 0) or (count == 1);
+    assert (count == 1) or (count == 2);
 \end{verbatim}
@@
 \end{code}
-However, if the key is an atom, say \texttt{.count}, then you
-can write \texttt{d.count} instead of having to write
-\texttt{d[.count]} (both will work).
+However, if \texttt{d = dict\{ .count: 3 \}}, then you can write
+can write \texttt{d.count} (which has value 3) instead of having to write
+\texttt{d[.count]} (although both will work).
+The meaning of \texttt{d a b c ...} is \texttt{((((d a) b) c) ...)}.
 
@@
 The empty tuple \texttt{()} is the same value as \texttt{dict\{\}}.
+Note that this is different from the empty set, which is \texttt{\{\}}.
 As in Python, you can create singleton tuples by adding an
@@
 \texttt{[a, b, c]} (which looks like a Python list)
-is the same as \texttt{(a, b, c)} (which looks like a Python tuple),
-which in turn is the same as \texttt{dict\{ 0:a, 1:b, 2:c \}}.
+is the same CXL value as \texttt{(a, b, c)} (which looks like a Python tuple),
+which in turn is the same CXL value as \texttt{dict\{ 0:a, 1:b, 2:c \}}.
 So if \texttt{x == [ False, True ]},
-then \texttt{x[0] == False} and \texttt{x[1] == True}.
+then \texttt{x[0] == False} and \texttt{x[1] == True}, just like in Python.
 
-Dictionaries have two handy unary operators.  If \texttt{d} is a
-dictionary, then \texttt{keys d} (or equivalently \texttt{keys(d)})
-returns the set of keys and \texttt{len d} returns the size of
+CXL is not an object-oriented language, however, so objects don't have
+built-in methods.  However, CXL does have some powerful operators to
+make up for some of that.
+For example, dictionaries have two handy unary operators.
+If \texttt{d} is a
+dictionary, then \texttt{keys~d} (or equivalently \texttt{keys(d)})
+returns the set of keys and \texttt{len~d} returns the size of
 this set.
@@
 to know a bit about the underlying machine.  A CXL program is translated
-into a list of machine instructions that the machine executes.
+into a list of machine instructions that the CXL machine executes.
 The CXL machine is not an ordinary virtual machine, but its architecture
@@
 \item Shared memory: A CXL machine has just one memory location containing
-a CXL.
+an CXL value.
 \item Processes:  Initially there is just the main process, but any process
@@
 a program counter, a stack of CXL values,
-and a single mutable general purpose \emph{register} that contains a CXL.
+and a single mutable general purpose \emph{register} that contains a CXL value.
 \end{itemize}
@@
 CXL values are unbounded trees.
-By convention, both the memory and the register of a process always contain
+Both the memory and the register of a process always contain
 a special type of CXL value in which the root is a dictionary
-that maps strings to CXL values.  We call this a \emph{directory}.
-A directory represents the state of a collection of variables named by the strings.
+that maps atoms to CXL values.  We call this a \emph{directory}.
+A directory represents the state of a collection of variables named by the atoms.
 
-Directories can be organized into a tree.  Each node in the tree is then identified
-by a sequence of strings, like a path name in the file system hierarchy.  We call
+Because directories are CXL values themselves,
+directories can be organized into a tree.
+Each node in the tree is then identified
+by a sequence of atoms, like a path name in the file system hierarchy.  We call
 such a sequence the \emph{address} of a CXL value.
@@
 \item A token containing the name of the method;
-\item A token containing the name of the argument of the method;
+\item A list of tokens containing the names of the arguments of the method;
 \item A program counter of the method's \texttt{Return} instruction.
@@
 \texttt{(’up’, ’Up.cxl’, 1, 5)} represents the lexeme \texttt{up} in
-file \texttt{Up.xcl}, line 1, column 5.  Tokens make it easy to relate
+file \texttt{Up.xcl}, line~1, column~5.  Tokens make it easy to relate
 the CXL machine instructions to the CXL source code.
@@
 Next it reports an execution that failed this assertion.  The output has
-five columns:
+four columns:
 \begin{enumerate}
-\item The entry point of the process (\texttt{\_\_main\_\_} for the main process);
-\item The tag of the process (in this case always 0);
+\item The name tag of the process;
 \item The sequence of program counters of the CXL machine instructions that the process executed;
@@
 in Figure~\ref{fig:inc}: operations
-on it consist of several instructions are divisible.  It is thus possible
-for both processes to believe the lock is available and obtain the lock
+on it consist of several instructions.  It is thus possible
+for both processes to believe the lock is available and to obtain the lock
 at the same time.
@@
 the same time.
-However, it has a different problem called livelock: if both try to
+However, it has a different problem called \emph{livelock}: if both try to
 enter the critical section at the same time, they may end up waiting for
@@
 If you invoke \texttt{m()}, the argument is \texttt{()},
-which is the empty tuple.  If you invoke
+which is the empty tuple.
 If you invoke \texttt{m(a, b)}, the argument is \texttt{(a, b)},

### 2020-05-19 book/paper.tex
Example of using methods and pointers
Latest version

@@
 \usepackage{verbatim}
+\usepackage{comment}
 
@@
 
-CXL methods look and feel much the same as Python methods, but there
-are some differences you should be aware of.
-First, as already noted, there is no \texttt{return} statement.
-A CXL method returns when the last statement in the method has
-completed.
+A method \texttt{m} with argument \texttt{a} is invoked in its
+most basic form as follows (assigning the result to \texttt{r}).
+\begin{code}
+\begin{verbatim}
+r = m a;
+\end{verbatim}
+\end{code}
+That's right, no parentheses are required.  In fact, if you invoke
+\texttt{m(a)}, the argument is \texttt{(a)}, which is the same
+as \texttt{a}.
+If you invoke \texttt{m()}, the argument is \texttt{()},
+which is the empty tuple.
+If you invoke \texttt{m(a, b)}, the argument is \texttt{(a, b)},
+the tuple consisting of values \texttt{a} and \texttt{b}.
+You have already seen examples of this: the expression
+\texttt{atLabel.cs} invokes method \texttt{atLabel} with
+argument \texttt{.cs}.
+
+You may note that all this looks familiar.  Indeed, the syntax
+is the same as that for dictionaries (see Chapter~\ref{ch:cxlvalues}).
+Both dictionaries and methods map CXL values to CXL values,
+and their syntax is indistinguishable.
+If \texttt{f} is either a method or a
+dictionary, and \texttt{x} is an arbitrary CXL value, then
+\texttt{f x}, \texttt{f(x)}, and \texttt{f[x]} are all
+the same expression in CXL.
+
+\begin{figure}
+\begin{code}
+\verbatiminput{../PetersonMethod.cxl}
+\end{code}
+\caption{Peterson's Algorithm accessed through methods.}
+\label{fig:petersonmethods}
+\end{figure}
+
+CXL is not an object-oriented language like Python is.  In Python
+you can pass a reference to an object to a method, and that method
+can then update the object.  In CXL, it is also sometimes convenient
+to have a method update a shared variable specified as an argument.
+For this (as well as some other uses), CXL supports \emph{pointers}
+to shared variables.
+If \texttt{x} is a shared variable, then the expression \texttt{\&(x)}
+(the parentheses are mandatory) is a pointer to \texttt{x}
+(also known as the \emph{address} of \texttt{x}.
+Conversely, if \texttt{p} is a pointer to a shared variable, then the
+expression \texttt{\^{}p} is the value of the shared variable.
+
+It is often the case that methods that update shared variables through
+pointers do not actually have to return a result to be useful.
+In CXL, if you want to invoke a method that does not return a result
+or if you are simply not interested in it,
+you have to use a \texttt{call} statement like so:
+\begin{code}
+\begin{verbatim}
+call m a;
+\end{verbatim}
+\end{code}
+
+Figure~\ref{fig:petersonmethods} again shows Peterson's algorithm,
+but this time with methods defined to enter and exit the critical
+section.
+You can put the first three methods in its own CXL source file
+and include it using the CXL \texttt{import} statement.  This would
+make the code re-usable by other applications.  For those who are
+extra adventurous: you can add the \texttt{P\_enter} and
+\texttt{P\_exit} methods to the \texttt{P\_mutex} dictionary
+like so:
+\begin{code}
+\begin{verbatim}
+dict{ .turn: 0, .flags: [ False, False ], .enter: P_enter, .exit: P_exit }
+\end{verbatim}
+\end{code}
+That would allow you to simulate object methods.
+
+\chapter{Spinlock}
+
+\begin{figure}
+\begin{code}
+\verbatiminput{UpTAS.cxl}
+\end{code}
+\caption{Mutual Exclusion using a ``spinlock'' based on test-and-set.}
+\label{fig:tas}
+\end{figure}
+
+Figure~\ref{fig:uplock} showed a faulty attempt at solving mutual
+exclusion using a lock.  The problem with the implementation of the
+lock is that checking the lock and setting it if it is available is
+not \emph{atomic}.  Thus multiple processes contending for the lock
+can all ``grab the lock'' at the same time.  While Peterson's
+algorithm gets around the problem, it is not efficient, especially
+if generalized to multiple processes.  Instead, processors provide
+so-called \emph{interlock instructions}: special machine instructions
+that can read memory and then write it in an indivisible step.
+
+While the CXL machine does not have any built-in interlock instructions,
+it does have support for executing multiple instructions atomically.
+This feature is available in the CXL language through its \texttt{atomic}
+statement.
+You already saw an example of this in Figure~\ref{fig:petersonproof}.
+We can use \texttt{atomic} blocks to implement a wide variety of
+interlock operations.
+For example, we could fix the program in Figure~\ref{fig:inc} by
+constructing an atomic increment operation for a counter, like so:
+\begin{code}
+\begin{verbatim}
+def atomic_inc(ptr):
+    atomic:
+        ^ptr = ^ptr + 1;
+    ;
+;
+count = 0;
+call atomic_inc(&(count));
+\end{verbatim}
+\end{code}
+
+Many CPUs have an atomic ``test-and-set'' operation.
+Method \texttt{tas} in Figure~\ref{fig:tas} shows its specification.
+Here \texttt{lk} points to a shared variable holding a
+boolean.  The operation assigns \texttt{True} to the
+variable but returns the value the variable had \emph{before}
+the assignment.
+Figure~\ref{fig:tas} goes on to show how the program
+in Figure~\ref{fig:uplock} can be made correct.
+Using \texttt{tas} a process can wait until \texttt{lock} is
+\texttt{False} and then set it to \texttt{True} before another
+process is able to see its old value.
+
+Compared to Peterson's Algorithm, this approach is general for
+any number of processes.  To make sure, you may want to try it
+for three or more processes.  In fact, you may want to replace
+the last two lines by:
+\begin{code}
+\begin{verbatim}
+for i in 1..5:
+    spawn process(), ();
+;
+\end{verbatim}
+\end{code}
+This creates 5 processes with the same name tag.  CXL can
+exploit their symmetry to do a much more efficient state
+search.  Try replacing it with the following code and see
+the difference:
+\begin{code}
+\begin{verbatim}
+for i in 1..5:
+    spawn process(), i;
+;
+\end{verbatim}
+\end{code}
+
+\chapter{Local Variables}
 
@@
 
-A method \texttt{m} with argument \texttt{a} is invoked in its
-most basic form as follows (assigning the result to \texttt{x}).
-\begin{code}
-\begin{verbatim}
-var x = m a;
-\end{verbatim}
-\end{code}
-That's right, no parentheses are required.  In fact, if you invoke
-\texttt{m(a)}, the argument is \texttt{(a)}, which is the same
-as \texttt{a}.
-If you invoke \texttt{m()}, the argument is \texttt{()},
-which is the empty tuple.
-If you invoke \texttt{m(a, b)}, the argument is \texttt{(a, b)},
-the tuple consisting of values \texttt{a} and \texttt{b}.
-You have already seen examples of this: the expression
-\texttt{atLabel.cs} invokes method \texttt{atLabel} with
-argument \texttt{.cs}.
-
-You may note that all this looks familiar.  Indeed, the syntax
-is the same as that for dictionaries (see Chapter~\ref{ch:cxlvalues}).
-Both dictionaries and methods map CXL values to CXL values,
-and their syntax is indistinguishable.
-If \texttt{f} is either a method or a
-dictionary, and \texttt{x} is an arbitrary CXL value, then
-then \texttt{f x}, \texttt{f(x)}, and \texttt{f[x]} are all
-the same expression in CXL.
-
-CXL is not an object-oriented language like Python is.  In Python
-you can pass a reference to an object to a method, and that method
-can then update the object.  In CXL, it is also sometimes convenient
-to have a method update a shared variable specified as an argument.
-For this (as well as some other uses), CXL supports \emph{pointers}
-to shared variables.
-If \texttt{x} is a shared variable, then the expression \texttt{\&(x)}
-(the parentheses are mandatory) is a pointer to \texttt{x}
-(also known as the \emph{address} of \texttt{x}.
-Conversely, if \texttt{p} is a pointer to a shared variable, then the
-expression \texttt{\^{}p} is the value of the shared variable.
-
-If you want to invoke a method but are not interested in its result,
-you have to use a \texttt{call} statement like so:
-\begin{code}
-\begin{verbatim}
-call m a;
-\end{verbatim}
-\end{code}
-
-\chapter{Spinlock}
-Figure~\ref{fig:uplock} showed a faulty attempt at solving mutual exclusion.
-
 \end{document}

### 2020-05-19 book/paper.tex
Latest
Author list...
Improved formatting

@@
 \title{Concurrent Programming with CXL}
-\author{RVR and others}
+\author{RVR and possibly others}
 
@@
 count = 0;
-call atomic_inc(&(count));
+call atomic_inc &(count);
 \end{verbatim}
@@
 
+\begin{quote}
+Try doing mutual exclusion with other interlock instructions!
+\begin{itemize}
+\item Find a description for \texttt{compare-and-swap} on the internet and
+update Figure~\ref{fig:tas}, replacing \texttt{tas}.
+\item Other interesting interlock instructions include \texttt{swap} and
+\texttt{fetch-and-add}.
+\end{itemize}
+\end{quote}
+
 \chapter{Local Variables}

### 2020-05-20 book/paper.tex
Covered lock interface now

@@
 
+\begin{figure}
+\begin{code}
+\begin{verbatim}
+def tas(lk):
+    atomic:
+        result = ^lk;
+        ^lk = True;
+    ;
+;
+def lock(lk):
+    while tas(lk):
+        pass;
+    ;
+;
+def unlock(lk):
+    ^lk = False;
+;
+def Lock():
+    result = False;
+;
+\end{verbatim}
+\end{code}
+\caption{The \texttt{Lock} interface in the \texttt{synch} module.}
+\label{fig:locks}
+\end{figure}
+
+\begin{figure}
+\begin{code}
+\verbatiminput{../Lock.cxl}
+\end{code}
+\caption{Mutual Exclusion using the \texttt{Lock} interface of the \texttt{synch} module.}
+\label{fig:lockme}
+\end{figure}
+
+Locks are probably the most popular and basic form of synchronization
+in concurrent programs.  For this reason, CXL has a module called
+\texttt{synch} that includes support for locks.
+Figure~\ref{fig:locks} shows how they are implemented, and
+Figure~\ref{fig:lockme} gives an example of how they may be used.
+Notice that the module completely hides the implementation of the
+lock.
+
 \chapter{Local Variables}

### 2020-05-20 book/paper.tex
Starting semaphore chapter
Added correctness arguments for mutual exclusion in the otherwise broken solutions.

@@
 the same time.
-However, it has a different problem called \emph{livelock}: if both try to
-enter the critical section at the same time, they may end up waiting for
-one another.
+
+To see why, first note the invariant that if process $i$ is in the
+critical section, then flags[i] == True.
+Without loss of generality,
+suppose that process~0 sets \texttt{flags[0]} at time $t_0$.
+Process 0 can only reach the critical section if at some time $t_1$,
+$t_1 > t_0$, it finds that \texttt{flags[1] == False}.
+Because of the invariant, \texttt{flags[1] == False} implies that
+process~1 is not at the critical section at time $t_1$.
+Suppose that process~1 is at the critical section from $t_2$ until
+it clears its flag at time $t_3$.  Note $t_0 < t_1 < t_2 < t_3$.
+Process~1 can only reach the critical section between $(t_2, t_3)$
+if it finds \texttt{flags[0] == False} during $(t_1, t_3)$.  But
+during $(t_0, t_3)$ \texttt{flags[0] == True}.
+
+However, the solution has a different problem called \emph{livelock}:
+if both try to enter the critical section at the same time, they may
+end up waiting for one another indefinitely.  Thus the solution
+violated \emph{progress}.
 
@@
 opportunity to enter.
-The solution has the nice properties that it satifies mutual exclution and,
-moreover, processes 0 and 1 alternate entering the critical section.
+An invariant of this solution is that while process $i$ is in the critical
+section, \texttt{turn} == $i$.  Since \texttt{turn} cannot be 0 and 1 at
+the same time, mutual exclusion is satisfied.
+The solution also has the nice property that
+processes 0 and 1 alternate entering the critical section.
 The problem with the solution is
-that it violates the \emph{progress} property:
+that it also violates the \emph{progress} property:
 if process~$i$ terminates instead of entering the critical section when it
@@
 
-\chapter{Local Variables}
+\chapter{Semaphores}
 

### 2020-05-20 book/paper.tex
Better explanation of TAS

@@
 \end{code}
-\caption{Na\"{\i}ve implementation of a lock.}
+\caption{Na\"{\i}ve implementation of a shared lock.}
 \label{fig:uplock}
@@
 An invariant of this solution is that while process $i$ is in the critical
-section, \texttt{turn} == $i$.  Since \texttt{turn} cannot be 0 and 1 at
+section, \texttt{turn} == $i$.
+(If you do not believe that, replace the assertion at the critical
+section with \texttt{assert turn == self} and run CXL.)
+Since \texttt{turn} cannot be 0 and 1 at
 the same time, mutual exclusion is satisfied.
@@
 Method \texttt{tas} in Figure~\ref{fig:tas} shows its specification.
-Here \texttt{lk} points to a shared variable holding a
-boolean.  The operation assigns \texttt{True} to the
-variable but returns the value the variable had \emph{before}
-the assignment.
-Figure~\ref{fig:tas} goes on to show how the program
-in Figure~\ref{fig:uplock} can be made correct.
-Using \texttt{tas} a process can wait until \texttt{lock} is
-\texttt{False} and then set it to \texttt{True} before another
-process is able to see its old value.
-
+Here \texttt{s} points to a shared boolean variable and \texttt{p}
+to a private boolean variable, belonging to some process.
+The operation copies the value of the shared variable to the
+private variable and then sets the shared variable to \texttt{True}.
+
+Figure~\ref{fig:tas} goes on how to implement mutual exclusion for
+a set of $N$ processes.
+Variable \texttt{shared} is initialized to \texttt{False} while
+\texttt{private[$i$]} for each process $i$ is initialized to \texttt{True}.
+An important invariant, $I_1$, of the program is that at any time at most
+one of the variables is \texttt{False}.
+Another invariant, $I_2$, is that if process $i$ is in the critical section,
+then \texttt{private[$i$] == False}.
+Between the two, it is clear that only one process can be in the
+critical section at the same time.
+
+To see that invariant $I_1$ is maintained, note that
+\texttt{\^{}p == True} upon entry of \texttt{tas}.
+So there are two cases:
+\begin{enumerate}
+\item \texttt{\^{}s} is \texttt{False} upon entry to \texttt{tas}.
+Then upon exit \texttt{\^{}p == False} and \texttt{\^{}s == True}, maintaining
+the invariant.
+\item \texttt{\^{}s} is \texttt{True} upon entry to \texttt{tas}.
+Then upon exit nothing has changed, maintaining the invariant.
+\end{enumerate}
+Invariant $I_1$ is also easy to verify for exiting the critical section.
+
+\begin{figure}
+\begin{code}
+\verbatiminput{../TAS.cxl}
+\end{code}
+\caption{Fixed version of Figure~\ref{fig:uplock} using test-and-set.}
+\label{fig:tas2}
+\end{figure}
+
+Invariant $I_2$ is obvious as process $i$ only proceeds to the critical
+section if \texttt{private[$i$] == False} and no other process modified
+\texttt{private[$i$]}.
+Indeed, there is no need to keep \texttt{private} as a shared
+variable---we only did so to be able to show the invariants.
+Figure~\ref{fig:tas2} shows a more straightforward implementation.
 Compared to Peterson's Algorithm, this approach is general for
-any number of processes.  To make sure, you may want to try it
-for three or more processes.  In fact, you may want to replace
-the last two lines by:
-\begin{code}
-\begin{verbatim}
-for i in 1..5:
-    spawn process(), ();
-;
-\end{verbatim}
-\end{code}
-This creates 5 processes with the same name tag.  CXL can
-exploit their symmetry to do a much more efficient state
-search.  Try replacing it with the following code and see
-the difference:
-\begin{code}
-\begin{verbatim}
-for i in 1..5:
-    spawn process(), i;
-;
-\end{verbatim}
-\end{code}
+any number of processes.
 

### 2020-05-20 book/paper.tex
Latest
Explaining difference between atomic and locks

@@
 
+It is important to appreciate the difference between an
+\emph{atomic section} (the statements executed within an
+\texttt{atomic} statement) and a \emph{critical section}
+(protected by a lock of some sort).
+The former ensures that while the
+atomic statement is executing no other process can execute.
+The latter allows multiple processes to run concurrently,
+just not within the critical section.
+The former is rarely available to a programmer, while the latter
+is very common.
+
+In CXL, atomic statements allow you to implement your own
+low synchronization primites like test-and-set.  Atomic statements
+are not intended to replace locks or other synchonization primitives.
+By the way, CXL \texttt{assert} statements execute atomically
+as if they are nested within an atomic statement.
+
 \begin{quote}
@@
 \end{quote}
+
+\chapter{Locks and the Synch Module}
 
@@
 lock.
+The \texttt{synch} module includes a variety of other useful
+synchronization primitives, which will be discussed in later
+chapters.
 

### 2020-05-20 cxl.py
Starting on a "let" expression to replace "var" eventually

@@ -56,6 +56,7 @@ def isreserved(s):
         "choose",
         "const",
         "def",
+        "del",
         "else",
         "False",
         "for",
@@ -64,6 +65,7 @@ def isreserved(s):
         "in",
         "keys",
         "len",
+        "let",
         "nametag",
         "not",
         "or",
@@ -80,7 +82,7 @@ def isname(s):
 
 def isunaryop(s):
     return s in [ "^", "-", "atLabel", "cardinality",
-                        "nametag", "not", "keys", "len" ]
+                        "nametag", "not", "keys", "len", "processes" ]
 
 def isbinaryop(s):
     return s in [
@@ -352,6 +354,18 @@ class SplitOp(Op):
         context.push(SetValue(set(lst[1:])))
         context.pc += 1
 
+class DelVarOp(Op):
+    def __init__(self, name):
+        self.name = name
+
+    def __repr__(self):
+        return "DelVar " + str(self.name)
+
+    def eval(self, state, context):
+        (lexeme, file, line, column) = self.name
+        context.delete(lexeme)
+        context.pc += 1
+
 class LoadVarOp(Op):
     def __init__(self, name):
         self.name = name
@@ -479,9 +493,9 @@ class AssertOp(Op):
         assert isinstance(cond, bool)
         if not cond:
             if self.exprthere:
-                print("Assertion failed", self.token, expr)
+                print("CXL Assertion failed", self.token, expr)
             else:
-                print("Assertion failed", self.token)
+                print("CXL Assertion failed", self.token)
             state.failure = True
         context.pc += 1
 
@@ -666,6 +680,13 @@ class NaryOp(Op):
                 assert isinstance(e, DictValue), e
                 assert len(e) == 0
                 context.push(context.nametag)
+            elif op == "processes":
+                assert isinstance(e, DictValue), e
+                assert len(e) == 0
+                context.push(DictValue(
+                    { ctx.nametag:cnt
+                            for (ctx, cnt) in state.ctxbag.items()
+                    }))
             elif op == "len":
                 assert isinstance(e, DictValue), e
                 context.push(len(e.d))
@@ -1397,6 +1418,39 @@ class WhileAST(AST):
         code.append(JumpOp(pc1))
         code[pc2] = JumpCondOp(False, len(code))
 
+class LetAST(AST):
+    def __init__(self, var, expr, stat):
+        self.var = var
[…]

### 2020-05-21 book/paper.tex
Starting on a "let" expression to replace "var" eventually
Latest

@@
 then \texttt{x[0] == False} and \texttt{x[1] == True}, just like in Python.
+However, when creating a singleton list, make sure you include the
+comma, as in \texttt{[False,]}.
 
@@
 opportunity to enter.
-An invariant of this solution is that while process $i$ is in the critical
+An invariant of this solution is that while process~$i$ is in the critical
 section, \texttt{turn} == $i$.
@@
 a set of $N$ processes.
+It uses $N+1$ variables.
 Variable \texttt{shared} is initialized to \texttt{False} while
@@
 An important invariant, $I_1$, of the program is that at any time at most
-one of the variables is \texttt{False}.
+one of these variables is \texttt{False}.
 Another invariant, $I_2$, is that if process $i$ is in the critical section,
@@
 Invariant $I_2$ is obvious as process $i$ only proceeds to the critical
-section if \texttt{private[$i$] == False} and no other process modified
+section if \texttt{private[$i$] == False} and no other process modifies
 \texttt{private[$i$]}.
@@
 
-In CXL, atomic statements allow you to implement your own
-low synchronization primites like test-and-set.  Atomic statements
-are not intended to replace locks or other synchonization primitives.
+In CXL, atomic statements allow you to \emph{implement} your own
+low synchronization primitives like test-and-set.  Atomic statements
+are not intended to \emph{replace} locks or other synchonization primitives.
 By the way, CXL \texttt{assert} statements execute atomically

### 2020-05-21 book/paper.tex
Stuff about invariant checking

@@
 Invariant $I_1$ is also easy to verify for exiting the critical section.
-
-\begin{figure}
-\begin{code}
-\verbatiminput{../TAS.cxl}
-\end{code}
-\caption{Fixed version of Figure~\ref{fig:uplock} using test-and-set.}
-\label{fig:tas2}
-\end{figure}
-
 Invariant $I_2$ is obvious as process $i$ only proceeds to the critical
@@
 \texttt{private[$i$]}.
-Indeed, there is no need to keep \texttt{private} as a shared
-variable---we only did so to be able to show the invariants.
+
+\begin{figure}
+\begin{code}
+\verbatiminput{UpTASinv.cxl}
+\end{code}
+\caption{Checking invariants.}
+\label{fig:tasinv}
+\end{figure}
+
+CXL can check these invariants as well.  Figure~\ref{fig:tas} already
+has the code to check $I_2$.  But how would one go about checking an
+invariant like $I_1$.  Invariants must hold for every state.
+For $I_2$ we only need an assertion at label \texttt{@cs} because the
+premisse is that there is a process at that label.  However, we would
+like to check $I_1$ in \emph{every state} (after the variables have
+been initialized).
+
+We can do this by adding another process that, in a loop,
+checks the invariant.  Figure~\ref{fig:tasinv} shows the code.
+Method \texttt{checkInvariant()} checks to see if the invariant holds
+in a state.  It introduces a new feature of CXL: the ability to have
+variables local to a method.  In this case, the variable \texttt{sum}
+is used to compute the number of variables that have value
+\texttt{False}.
+The function is invoked by in a loop by a process that runs alongside
+the other processes.
+Note that this process must be started after the variables are
+initialized---before that the invariant does not hold yet.
+Because CXL tries every possible execution, the process is guaranteed
+to find violations of the invariant if it does not hold.
+
+\begin{figure}
+\begin{code}
+\verbatiminput{../TAS.cxl}
+\end{code}
+\caption{Fixed version of Figure~\ref{fig:uplock} using test-and-set.}
+\label{fig:tas2}
+\end{figure}
+
+Because each \texttt{private[i]} is only access by process~$i$,
+there is no need to keep \texttt{private} as a shared
+variable---we only did so to be able to show and check the invariants.
 Figure~\ref{fig:tas2} shows a more straightforward implementation.

### 2020-05-21 book/paper.tex
Latest

@@
 
+\begin{quote}
+Try doing mutual exclusion with other interlock instructions!
+\begin{itemize}
+\item Find a description for \texttt{compare-and-swap} on the internet and
+update Figure~\ref{fig:tas}, replacing \texttt{tas}.
+\item Other interesting interlock instructions include \texttt{swap} and
+\texttt{fetch-and-add}.
+\end{itemize}
+\end{quote}
+
+\chapter{Locks and the Synch Module}
+
 \begin{figure}
@@
 
-Because each \texttt{private[i]} is only access by process~$i$,
-there is no need to keep \texttt{private} as a shared
+In Figure~\ref{fig:tas} we have shown a solution based on a shared
+variable and a private variable for each process.   The private
+variables themselves are implemented as shared variables as well,
+but they are accessed only by their respective processes.
+There is no need to keep \texttt{private} as a shared
 variable---we only did so to be able to show and check the invariants.
 Figure~\ref{fig:tas2} shows a more straightforward implementation.
+The solution is similar to the na\"{i}ve solution of Figure~\ref{fig:uplock},
+but uses test-and-set to check and set the lock variable atomically.
 Compared to Peterson's Algorithm, this approach is general for
@@
 as if they are nested within an atomic statement.
-
-\begin{quote}
-Try doing mutual exclusion with other interlock instructions!
-\begin{itemize}
-\item Find a description for \texttt{compare-and-swap} on the internet and
-update Figure~\ref{fig:tas}, replacing \texttt{tas}.
-\item Other interesting interlock instructions include \texttt{swap} and
-\texttt{fetch-and-add}.
-\end{itemize}
-\end{quote}
-
-\chapter{Locks and the Synch Module}
 
@@
 
-Like in Python, CXL methods have local variables.  You have already
-seen two types of local variables.
-First, the arguments to the method are variables, although read-only.
-Second, the variable \texttt{result} can be assigned to
-change the default return value of a method, which is \texttt{()}.
-You can declare additional local variables to hold intermediate
-results.  For example, you can
-declare a local variable $x$ with initial value 3 as follows:
-\begin{code}
-\begin{verbatim}
-var x = 3;
-\end{verbatim}
-\end{code}
-The initial value is mandatory.
-Unless a variable is declared this way, it will be considered
-a global shared variable.
-The local variables of a method cannot be accessed other than by
-the method itself.
-
 \end{document}

### 2020-05-21 book/paper.tex
"choose" scheduling fixed

@@
 \begin{code}
-\verbatiminput{../Lock.cxl}
-\end{code}
-\caption{Mutual Exclusion using the \texttt{Lock} interface of the \texttt{synch} module.}
-\label{fig:lockme}
+\verbatiminput{UpFixed.cxl}
+\end{code}
+\caption{Program of Figure~\ref{fig:inc} fixed with a lock.}
+\label{fig:incfixed}
 \end{figure}
@@
 Figure~\ref{fig:locks} shows how they are implemented, and
-Figure~\ref{fig:lockme} gives an example of how they may be used.
+Figure~\ref{fig:figinc} gives an example of how they may be used,
+in this case to fix the program of Figure~\ref{fig:inc}.
 Notice that the module completely hides the implementation of the

### 2020-05-21 cxl.py

@@ -1921,6 +1921,7 @@ class State:
         self.ctxbag = {
             Context(DictValue({"name": "__main__", "tag": novalue}), 0, len(code)) : 1
         }
+        self.choosing = None
         self.failure = False
 
     def __repr__(self):
@@ -1939,6 +1940,8 @@ class State:
         if self.vars != other.vars:
             return False
         if self.ctxbag != other.ctxbag:
+            return False
+        if self.choosing != other.choosing:
             return False
         if self.failure != other.failure:
             return False
@@ -2139,14 +2142,19 @@ def onestep(state, ctx, choice, visited, todo, node, infloop):
         if sc.failure or cc.pc == cc.end:
             break
 
-        # if we're about to do a state change, let other processes
-        # go first assuming there are other processes and we're not
-        # in "atomic" mode
+        # See if this process is making a nondeterministic choice.
+        # If so, we break out of the microstep loop.  However, only
+        # this process is scheduled from this state.
         if isinstance(sc.code[cc.pc], ChooseOp):
             v = cc.stack[-1]
             assert isinstance(v, SetValue), v
             if len(v.s) != 1:
+                sc.choosing = cc
                 break
+
+        # if we're about to do a state change, let other processes
+        # go first assuming there are other processes and we're not
+        # in "atomic" mode
         if cc.atomic == 0 and type(sc.code[cc.pc]) in globops and len(sc.ctxbag) > 0:
             break
 
@@ -2246,14 +2254,15 @@ def run(code, labels):
             continue
         node.expanded = True
 
-        for (ctx, _) in state.ctxbag.items():
-            if ctx.pc < ctx.end and isinstance(code[ctx.pc], ChooseOp):
-                choices = ctx.stack[-1]
-                assert isinstance(choices, SetValue), choices
-                assert len(choices.s) > 0
-                for choice in choices.s:
-                    onestep(state, ctx, choice, visited, todo, node, infloop)
-            else:
+        if state.choosing != None:
+            ctx = state.choosing
+            choices = ctx.stack[-1]
+            assert isinstance(choices, SetValue), choices
+            assert len(choices.s) > 0
+            for choice in choices.s:
+                onestep(state, ctx, choice, visited, todo, node, infloop)
+        else:
+            for (ctx, _) in state.ctxbag.items():
                 onestep(state, ctx, None, visited, todo, node, infloop)
     print()
 

### 2020-05-21 book/paper.tex
Starting to work on reader/writer

@@
 \usepackage{comment}
+\usepackage{amsmath}
 
@@
 in the bag to the number of times the element appears in the bag.
+Method \texttt{atLabel} only exists for specification purposes---do not
+use it in normal code.
 
@@
 
-\chapter{Semaphores}
+\chapter{Reader/Writer Locks}
+
+\begin{figure}
+\begin{code}
+\verbatiminput{../RWbusy.cxl}
+\end{code}
+\caption{Busy-Waiting Reader/Writer Lock implementation.}
+\label{fig:rwbusy}
+\end{figure}
+
+Locks are useful when accessing a shared data structure.  By preventing
+more than one process from accessing the data structure at the same
+time, conflicting access are avoided.  However, not all concurrent
+accesses conflict, and opportunities for concurrency may be lost,
+hurting performance.  One important case is when multiple processes
+are simply reading the data structure.  There is no conflict, and
+all reads should be allowed to proceed concurrently.  In many
+applications, reads are the majority of all accesses.
+
+What we want is a special kind of lock that allows either one writer
+or one or more readers to be in the critical section.  This is called
+a \emph{reader/writer lock}.
+We will explore various ways of implementing reader/writer locks in
+this chapter and future ones.
+
+Figure~\ref{fig:rwbusy} presents a solution that uses a
+single (ordinary) lock and two counters: one that maintains the number
+of readers and one that maintains the number of writers.
+The program shows two kinds of processes: \texttt{reader()} and
+\texttt{writer()}.  The critical section is spread between two
+labels: readers access \texttt{@rcs} and writers access \texttt{@wcs}.
+The specification is that if a reader is at label \texttt{@rcs},
+no writer is allowed to be at label \texttt{@wcs}.  Vice versa, if
+a writer is at label \texttt{@wcs}, no reader is allowed to be at
+label \texttt{@rcs} \emph{and} there cannot be another writer at
+label \texttt{@wcs}.
+
+A reader process that wants to enter \texttt{@rcs} first waits
+until there are no writers: \texttt{nwriters == 0}.
+If so, it increments the number of readers.  Similarly, a writer
+process waits until there are no readers \emph{or} writers.
+If so, it increments the number of writers.
+The important invariants in this code are:
+\begin{itemize}
+\item $n$ readers at \texttt{@rcs} $\Rightarrow \mathtt{nreaders} = n$,
+\item $n$ writers at \texttt{@rcs} $\Rightarrow \mathtt{nwriters} = n$,
+\item $(\mathtt{nreaders} \ge 0 \land \mathtt{nwriters} = 0) \lor
+    (\mathtt{nreaders} = 0 \land 0 \le \mathtt{nwriters} \le 1)$.
+\end{itemize}
+It is easy to see that the invariants hold and imply the specification.
 

### 2020-05-21 book/paper.tex
More on reader/writer lock
Trying to tell why busy wait is bad...

@@
 or one or more readers to be in the critical section.  This is called
-a \emph{reader/writer lock}.
+a \emph{reader/writer lock}~\cite{courtois71}.
 We will explore various ways of implementing reader/writer locks in
@@
 It is easy to see that the invariants hold and imply the specification.
+The solution also supports progress: if no process is in the critical
+section then any process can enter.
+
+While correct, it is not considered a good solution.
+The solution is an example of what is called \emph{busy waiting}:
+processes spin in a loop until some desirable condition is met.
+The astute reader might wonder if obtaining an ordinary lock itself is an
+example of busy waiting.  After all, the CXL \texttt{synch} implementation
+of \texttt{lock()} spins in a loop until the lock is available
+(see Figure~\ref{fig:locks}).
+
+In most operating systems and programming language runtimes, however,
+when a process acquires a lock, the process is placed on a scheduling
+queue and stops using CPU cycles until the lock becomes available.
+Moreover, if there are multiple processes waiting for a lock, there
+is a scheduler that decides which of the processes gets the lock.
+Busy waiting disables all this.
+
+But even when multiple processes each run on their own core, the solution
+is inefficient.  After all, when the lock becomes available, the process
+has to check other variables to see if it can continue.  In our case,
+the process may have to read \texttt{nreaders} and/or \texttt{nwriters}
+to decide if it can go on or has to try again, involving releasing
+the lock and re-acquiring it.
+
+So it is considered ok to have processes busy wait for a lock to become
+available, but not to busy wait for application-specific conditions.
 

### 2020-05-21 book/paper.tex
Latest

@@
 While correct, it is not considered a good solution.
-The solution is an example of what is called \emph{busy waiting}:
+The solution is an example of what is called \emph{busy-waiting}:
 processes spin in a loop until some desirable condition is met.
 The astute reader might wonder if obtaining an ordinary lock itself is an
-example of busy waiting.  After all, the CXL \texttt{synch} implementation
+example of busy-waiting.  After all, the CXL \texttt{synch} implementation
 of \texttt{lock()} spins in a loop until the lock is available
@@
 
-So it is considered ok to have processes busy wait for a lock to become
-available, but not to busy wait for application-specific conditions.
+\begin{figure}
+\begin{code}
+\verbatiminput{../RWlock.cxl}
+\end{code}
+\caption{Reader/Writer with Two Locks.}
+\label{fig:rw2lock}
+\end{figure}
+
+So it is considered ok to have processes using spinlock a implement a
+lock, but not to busy-wait for application-specific conditions.
+
+Figure~\ref{fig:rw2lock} presents a reader/writer lock implementation
+that does not busy-wait.
+It uses two locks: \texttt{rwlock} is used by both readers and writers,
+while \texttt{rlock} is only used by readers.
 

### 2020-05-21 book/paper.tex
Latest

@@
 the spawned processes never terminate.  In order to resolve this, we
-will model processes in critical regions using the framework in
+will model processes in critical sections using the framework in
 Figure~\ref{fig:cs}: a process can \emph{choose} to enter a
-critical region more than once, but it can also choose to terminate, even
-without entering the critical region ever.
+critical section more than once, but it can also choose to terminate, even
+without entering the critical section ever.
 Note also that we have dropped the \texttt{@ncs} label: it would serve no
@@
 \begin{itemize}
-\item Plug \texttt{while True: pass;;} for entering the critical region
+\item Plug \texttt{while True: pass;;} for entering the critical section
 in Figure~\ref{fig:cs} and run CXL.  It should print a trace
@@
 \chapter{Peterson's Algorithm}
+\label{ch:peterson}
 
@@
 of readers and one that maintains the number of writers.
+The lock is used to protect access to those counters.
 The program shows two kinds of processes: \texttt{reader()} and
@@
 \begin{itemize}
-\item $n$ readers at \texttt{@rcs} $\Rightarrow \mathtt{nreaders} = n$,
-\item $n$ writers at \texttt{@rcs} $\Rightarrow \mathtt{nwriters} = n$,
+\item $n$ readers at \texttt{@rcs} $\Rightarrow \mathtt{nreaders} \ge n$,
+\item $n$ writers at \texttt{@rcs} $\Rightarrow \mathtt{nwriters} \ge n$,
 \item $(\mathtt{nreaders} \ge 0 \land \mathtt{nwriters} = 0) \lor
@@
 \end{itemize}
-It is easy to see that the invariants hold and imply the specification.
+It is easy to see that the invariants hold and imply the reader/writer
+specification.
 The solution also supports progress: if no process is in the critical
-section then any process can enter.
+section then any process can enter.  Better still: if any reader is in the
+critical section, any other reader is also able to enter.
 
@@
 while \texttt{rlock} is only used by readers.
+\texttt{rwlock} is held either when readers are at label \texttt{@rcs}
+or when a writer is at label \texttt{@wcs}.
+\texttt{rlock} is used to protect the \texttt{nreaders} variable that
+counts the number of readers in the critical section.
+
+The invariants that imply the reader/writer specification
+(which are again easy to verify) are as follows:
+
+\begin{itemize}
+\item $n$ readers at \texttt{@rcs} $\Rightarrow \mathtt{nreaders} \ge n$,
+\item $\exists$ writer at \texttt{@rcs} $\Rightarrow \mathtt{nreaders} = 0$,
+\end{itemize}
+
+A writer simply acquires \texttt{rwlock} to enter the critical section
+and releases it to exit.  The \emph{first} reader to enter the critical
+section acquires \texttt{rwlock} and the \emph{last} reader to exit
+the critical section releases \texttt{rwlock}.
+The implementation satisfies progress: if no process is in the critical
+section than any process enter.
+
+It is instructive to see what happens when a writer is in the critical
+section and two readers try to enter.  The first reader successfully
+acquires \texttt{rlock} and but \emph{hangs} when trying to acquire
+\texttt{rwlock}, which is held by the writer.  The second reader hangs
+trying acquire \texttt{rlock} because it is held by the first reader.
+When the writer leaves the critical section, the first reader acquires
+\texttt{rwlock}, sets \texttt{nreaders} to 1, and releases \texttt{rlock}.
+\texttt{rwlock} is still held.
+Then the second reader acquires \texttt{rlock} and, assuming the first
+reader is still in the critical section, increments \texttt{nreaders} to 2
+and enter the critical section \emph{without} acquiring \texttt{rwlock}.
+\texttt{rwlock} is essentially jointly held by both readers.
+It does not matter in which order they leave: the second will release
+\texttt{rwlock}.
+
+\chapter{Starvation}
+
+So far we have pursued two properties: emph{mutual exclusion}
+and \emph{progress}.  The former is an example of
+\texttt{safety property}---it prevents something ``bad'' from
+happening, like a reader and writer process both entering the
+critical section.  The \emph{progress} property is an example
+of a liveness property---that something good eventually happens.
+Informally (and inexactly), progress states that if no processes
+are in the critical section, then some process that wants to enter
+can.
+
+Progress is a weak form of liveness.  It says that \emph{some}
+process can enter, but it does not prevent a scenario such as
+the following.  There are three processes repeatedly trying to
+enter a critical section using a spinlock.  Two of
+the processes successfully enter, alternating, but the third
+process never gets a turn.  This is an example of
+\texttt{starvation}.  With a spinlock, this scenario could
+even happen with two processes.  Initially both processes
+try to acquire the spinlock.  One of the processes is
+successful and enter.  After the process leaves, it immediately
+tries to re-enter.  This state is identical to the initial
+state, and there is nothing that prevents the same process
+from acquiring the lock yet again.
+
+It is worth noting that Peterson's Algorithm (Chapter~\ref{ch:peterson})
+does not suffer from starvation, thanks to the \texttt{turn} variable
+that alternates between 0 and 1 when two processes are contending for
+the critical section.
+
+While spinlocks suffer from starvation, it is a uniform random
+process and each process has an equal chance of entering the critical
+section.  Thus the probability of starvation is exponentially vanishing.
+Unfortunately, such is not the case for the
+reader/writer solution of Figure~\ref{fig:rw2lock}.
+Consider this scenario: there are two readers and one writer.  One reader
+is in the critical section while the writer is waiting.  Now the
+second reader tries to enter and is able to.  The first reader leaves.
+We are now in a similar situation as the initial state with one reader
+in the critical section and the writer waiting, but it is not the same
+reader.  Unfortunately for the writer, this scenario can repeat itself
+indefinitely.  So even if neiter reader was in the critical section
+for all of the time, the writer never had a chance.
+In this section we will present a version of a reader/writer lock
+implementation that solves this type of starvation.
 

### 2020-05-21 book/paper.tex
Latest

@@
 indefinitely.  So even if neiter reader was in the critical section
-for all of the time, the writer never had a chance.
-In this section we will present a version of a reader/writer lock
-implementation that solves this type of starvation.
+for all of the time, and the second reader arrived well after the writer,
+the writer never had a chance.
+In this chapter, we will present a version of a reader/writer lock
+implementation that solves this type of starvation, but won't eliminate
+the type of starvation that comes with spinlocks.
 

### 2020-05-21 book/paper.tex
Latest
typos

@@
 
-So far we have pursued two properties: emph{mutual exclusion}
+So far we have pursued two properties: \emph{mutual exclusion}
 and \emph{progress}.  The former is an example of
@@
 reader.  Unfortunately for the writer, this scenario can repeat itself
-indefinitely.  So even if neiter reader was in the critical section
-for all of the time, and the second reader arrived well after the writer,
+indefinitely.  So even if neither reader was in the critical section
+all of the time, and the second reader arrived well after the writer,
 the writer never had a chance.

### 2020-05-21 book/paper.tex
Fixes

@@
 So far we have pursued two properties: \emph{mutual exclusion}
-and \emph{progress}.  The former is an example of
-\texttt{safety property}---it prevents something ``bad'' from
+and \emph{progress}.  The former is an example of a
+\emph{safety property}---it prevents something ``bad'' from
 happening, like a reader and writer process both entering the
 critical section.  The \emph{progress} property is an example
-of a liveness property---that something good eventually happens.
+of a \emph{liveness property}---guaranteeing that something good
+eventually happens.
 Informally (and inexactly), progress states that if no processes
@@
 try to acquire the spinlock.  One of the processes is
-successful and enter.  After the process leaves, it immediately
+successful and enters.  After the process leaves, it immediately
 tries to re-enter.  This state is identical to the initial

### 2020-05-24 book/paper.tex
Playing with producer/consumer and semaphores
Updating some cxl output

@@
 processes running in parallel while sharing variables.
-Figure~\ref{fig:inc} presents a simple example.  The program defines a
-method \texttt{process} that takes a parameter called \texttt{self}.
-It also initializes two shared variables \texttt{count} and \texttt{done}.
-The \texttt{process(self)} method increments \texttt{count} and sets \texttt{done[self]}
-to \texttt{True}.
-Next the program spawns two processes, each invoking \texttt{process}.
-Afterward there are three processes---after
-all, there was already one, and it spawned two new ones.  We will call
-the one that was already there the \emph{main process}.
-
-The main process waits for the two other processes to finish by polling
+Figure~\ref{fig:inc} presents a simple example.
+The program
+initializes two shared variables: an integer \texttt{count} and
+an array \texttt{done} with two booleans.
+Method \texttt{incrementer} takes a parameter called \texttt{self}.
+It increments \texttt{count} and sets \texttt{done[self]} to \texttt{True}.
+Method \texttt{main} waits for the processes to finish by polling
 both \texttt{done} flags.
-Finally, it verifies that the value of \texttt{count} equals 2.  If not,
+After that, it verifies that the value of \texttt{count} equals 2.  If not,
 it reports the actual value of \texttt{count}.
+The program spawns three processes.
+The first runs \texttt{incrementer(0)}, the second runs
+\texttt{incrementer(1)}, and the last runs \texttt{main()}.
 
@@
 
-The following is a possible interleaving of processes~0 and~1:
+The following is a possible interleaving of incrementers~0 and~1:
 \begin{enumerate}
-\item Process 0 loads the value of \texttt{count}, which is 0;
-\item Process 1 loads the value of \texttt{count}, which is still 0;
-\item Process 1 adds one to the value that it loaded (0), and
+\item incrementer 0 loads the value of \texttt{count}, which is 0;
+\item incrementer 1 loads the value of \texttt{count}, which is still 0;
+\item incrementer 1 adds one to the value that it loaded (0), and
 stores $1$ into \texttt{count};
-\item Process 0 adds one to the value that it loaded (0), and
+\item incrementer 0 adds one to the value that it loaded (0), and
 stores $1$ into \texttt{count};
-\item Process 0 sets \texttt{upDone} to \texttt{True};
-\item Process 1 sets \texttt{upDone} to \texttt{True}.
+\item incrementer 0 sets \texttt{upDone} to \texttt{True};
+\item incrementer 1 sets \texttt{upDone} to \texttt{True}.
 \end{enumerate}
@@
 
-Another type of CXL value is the \emph{atom}.  You have already
-seen them: the names of constants, variables, and methods are
-all atoms.  An atom is denoted using a period followed by the
-name.  For example, the atom for the \texttt{process} method
-is denoted \texttt{.process}.
+Another type of CXL value is the \emph{atom}.  It is essentially
+just a name.  An atom is denoted using a period followed by the
+name.  For example, \texttt{.main} is an atom.
 
@@
 dictionary.
+CXL dictionaries are written as
+$\mathtt{dict}\{ k_0: v_0, ~ k_1: v_1, ~ ... \}$.
 If \texttt{d} is a dictionary, and \texttt{k} is a key, then the
@@
 \begin{verbatim}
-Up.cxl:1 def process(self):
-   0 Jump 12
-   1 Frame ('process', 'Up.cxl', 1, 5) [('self', 'Up.cxl', 1, 13)] 11
+Up.cxl:1 def incrementer(self):
+   0 Jump 39
+   1 Frame ('incrementer', 'Up.cxl', 1, 5) [('self', 'Up.cxl', 1, 17)] 12
 Up.cxl:2     count = count + 1;
    2 Constant (1, 'Up.cxl', 2, 21)
-   3 Load ('count', 'Up.cxl', 2, 13)
-   4 Nary ('+', 'Up.cxl', 2, 19) 2
-   5 Name ('count', 'Up.cxl', 2, 5)
-   6 Store 1
+   3 Name ('count', 'Up.cxl', 2, 13)
+   4 Load 1
+   5 Nary ('+', 'Up.cxl', 2, 19) 2
+   6 Name ('count', 'Up.cxl', 2, 5)
+   7 Store 1
 Up.cxl:3     done[self] = True;
-   7 Constant (True, 'Up.cxl', 3, 18)
-   8 LoadVar ('self', 'Up.cxl', 3, 10)
-   9 Name ('done', 'Up.cxl', 3, 5)
-   10 Store 2
-   11 Return
-Up.cxl:5 count = 0;
-   12 Constant (0, 'Up.cxl', 5, 9)
-   13 Name ('count', 'Up.cxl', 5, 1)
-   14 Store 1
+   8 Constant (True, 'Up.cxl', 3, 18)
+   9 LoadVar ('self', 'Up.cxl', 3, 10)
+   10 Name ('done', 'Up.cxl', 3, 5)
+   11 Store 2
+   12 Return
 \end{verbatim}
@@
 ==== Safety violation ====
-__main__/() [0,12-31]           32 dict{ .count:0, .done:dict{ 0:False, 1:False } }
-process /0  [1-5]                6 dict{ .count:0, .done:dict{ 0:False, 1:False } }
-process /1  [1-10]              11 dict{ .count:1, .done:dict{ 0:False, 1:True  } }
-process /0  [6-10]              11 dict{ .count:1, .done:dict{ 0:True,  1:True  } }
-__main__/() [32-34,37-42,44-48] 49 dict{ .count:1, .done:dict{ 0:True,  1:True  } }
+incrementer/0 [1-6]          7 dict{ .count:0, .done:dict{ 0:False, 1:False } }
+incrementer/1 [1-11]        12 dict{ .count:1, .done:dict{ 0:False, 1:True  } }
+incrementer/0 [7-11]        12 dict{ .count:1, .done:dict{ 0:True,  1:True  } }
+main/() [14-19,22-27,29-36] 37 dict{ .count:1, .done:dict{ 0:True,  1:True  } }
 \end{verbatim}
@@
 an CXL value.
-\item Processes:  Initially there is just the main process, but any process
+\item Processes:  Any process
 can spawn an unbounded number of other processes and processes may terminate.
@@
 A name tag consists of the name of the main method of the process,
-along with the tag specified in the \texttt{spawn} statement.
+along with an optional tag specified in the \texttt{spawn} statement.
+The default tag is \texttt{()}.
 In Figure~\ref{fig:inc}, the created processes have name tags
-\texttt{process/0} and \texttt{process/1}.  The main process that
-starts the execution has name tag \texttt{\_\_main\_\_/()}.
+\texttt{incrementer/0}, \texttt{incrementer/1}, and \texttt{main/()}.
 
@@
 
+\chapter{Semaphores}
+
+So far we have looked at how to protect a single resource.  Sometimes
+we have multiple but a limited amount of resources that require
+protection.  That is, if there are only $n$ resources, no more than
+$n$ processes can access them at a time.
+
 \chapter{Starvation}

### 2020-05-29 book/paper.tex
Various updates

@@
 are a little different from Python.
-A dictionary maps keys to values.
+A dictionary maps CXL values, known as \emph{keys}, to CXL values.
 Unlike Python, any CXL value can be a key, including another
@@
 
-CXL is not an object-oriented language, however, so objects don't have
+CXL is not an object-oriented language, so objects don't have
 built-in methods.  However, CXL does have some powerful operators to
@@
 specifies that only the current process can be there.  CXL in fact
-supports this.  It has a method \texttt{atLabel $L$}, where $L$
+supports this.  It has an operator \texttt{atLabel $L$}, where $L$
 is the atom of the label (in this case, \texttt{.cs}).  The
-method returns a bag of name tags of processes executing at that
+operator returns a bag of name tags of processes executing at that
 label.  The bag is represented by a dictionary that maps each element
@@
 Figure~\ref{fig:csbarebones} shows the code.
-The assertion also makes use of the method \texttt{nametag()} that
+The assertion also makes use of the \texttt{nametag()} operator that
 returns the name tag of the current process.
@@
 \begin{verbatim}
-while True: pass;;
+while True:
+    pass;
+;
 \end{verbatim}
@@
 but here we will propose the following informally: if some set of processes
-$S$ are trying to enter the critical section, then eventually one process in
-$S$ will enter the critical section if any process already in the
-intersection eventually leaves, even if no processes other than $S$ ever
-try to enter the critical section.
+$S$ are trying to enter the critical section and any process already in the
+intersection eventually leaves, then eventually one process in $S$ will enter
+the critical section.
 We call this \emph{progress}.
+
+\begin{figure}
+\begin{center}
+\includegraphics[width=2in]{figures/mutex.png}
+\end{center}
+\caption{High-level state diagram specification of mutual exclusion with up to two processes.
+The first number in a state gives the number of processes; the second number is the
+number of processes in the critical section.}
+\label{fig:mutex}
+\end{figure}
 
@@
 specific purpose.
+Figure~\ref{fig:mutex} show a high-level state diagram of the code in
+Figure~\ref{fig:cs}.
 
@@
 
-Figure~\ref{fig:uplock} presents a solution based on a na\"{\i}ve implementation
-of a \emph{lock}.
-The idea is that the lock is a baton that at most one process can own at
-a time.
+In order to have at most one process in a critical section at a time, you
+probably are thinking that one might use a \emph{lock}.
+That is a good thought, but how does one implement one?
+Figure~\ref{fig:uplock} presents a mutual exclusion attempt based on a
+na\"{\i}ve implementation of a lock.
+The idea is that the lock is like a baton that at most one process can own
+(or hold) at a time.
 Initially the lock is not owned, indicated by lock being \texttt{False}.
-To enter the critical section, a process waits until lock is \texttt{False}
+To enter the critical section, a process waits until \texttt{lock} is \texttt{False}
 and then sets it to \texttt{True} to indicate that the lock has been taken.
@@
 process~1 is not at the critical section at time $t_1$.
-Suppose that process~1 is at the critical section from $t_2$ until
-it clears its flag at time $t_3$.  Note $t_0 < t_1 < t_2 < t_3$.
-Process~1 can only reach the critical section between $(t_2, t_3)$
-if it finds \texttt{flags[0] == False} during $(t_1, t_3)$.  But
-during $(t_0, t_3)$ \texttt{flags[0] == True}.
+Let $t_2$ be the time at which process~0 sets \texttt{flags[0]}
+to \texttt{False}.  Process~0 is in the critical section sometime
+between $t_1$ and $t_2$.
+It is easy to see that process~1 cannot enter the critical section
+between $t_1$ and $t_2$, because \texttt{flags[1] == False} at
+time $t_1$.  To reach the critical section between $t_1$ and $t_2$,
+it would first have to set \texttt{flags[1]} to \texttt{True} and
+then wait until \texttt{flags[0] == False}.  But that does not happen
+until time $t_2$.
 
@@
 end up waiting for one another indefinitely.  Thus the solution
-violated \emph{progress}.
+violates \emph{progress}.
 
@@
 or has given way.
-Running the algorithm though CXL shows that it satisfies both mutual
+Running the algorithm with CXL shows that it satisfies both mutual
 exclusion and progress.
@@
 section, violating mutual exclusion.
+So mutual exclusion is an invariant, but not an inductive invariant.
 Doing a inductive proof with an invariant that is not inductive is usually
@@
 Figure~\ref{fig:petersonproof} formalizes $I_p(i)$ in CXL.
-The figure uses the CXL \texttt{atomic} statement for the first time.
-For now, all you need to know is that it turns the embedded statement into
-a single atomic step.  The label \texttt{@gate} thus refers exactly to
-this statement.  You can run Figure~\ref{fig:petersonproof} to determine
+The label \texttt{@gate} refers to the instruction that sets \texttt{turn} to $1-i$.
+You can run Figure~\ref{fig:petersonproof} to determine
 that $I_p(i)$ is indeed an invariant for $i = 0, 1$.
@@
 statement.
-You already saw an example of this in Figure~\ref{fig:petersonproof}.
 We can use \texttt{atomic} blocks to implement a wide variety of
@@
 Invariant $I_1$ is also easy to verify for exiting the critical section.
-Invariant $I_2$ is obvious as process $i$ only proceeds to the critical
-section if \texttt{private[$i$] == False} and no other process modifies
+Invariant $I_2$ is obvious as (i) process $i$ only proceeds to the critical
+section if \texttt{private[$i$] == False}, and (ii) no other process modifies
 \texttt{private[$i$]}.
@@
 the other processes.
-Note that this process must be started after the variables are
-initialized---before that the invariant does not hold yet.
+In CXL, \texttt{assert} statements are executed atomically, so the
+evaluation of the assertion is not interleaved with the execution
+of other processes.
 Because CXL tries every possible execution, the process is guaranteed
@@
 are not intended to \emph{replace} locks or other synchonization primitives.
-By the way, CXL \texttt{assert} statements execute atomically
-as if they are nested within an atomic statement.
 
@@
 Figure~\ref{fig:locks} shows how they are implemented, and
-Figure~\ref{fig:figinc} gives an example of how they may be used,
+Figure~\ref{fig:incfixed} gives an example of how they may be used,
 in this case to fix the program of Figure~\ref{fig:inc}.
@@
 
+\begin{figure}
+\begin{center}
+\includegraphics[width=2in]{figures/rdwr.png}
+\end{center}
+\caption{High-level state diagram specification of reader/writer locks with
+up to two processes.
+The first number in a state gives the number of processes; the second number is the
+number of processes reading in the critical section; the third is the number of
+processes writing in the critical section.}
+\label{fig:rdwr}
+\end{figure}
+
 Locks are useful when accessing a shared data structure.  By preventing
 more than one process from accessing the data structure at the same
-time, conflicting access are avoided.  However, not all concurrent
+time, conflicting accesses are avoided.  However, not all concurrent
 accesses conflict, and opportunities for concurrency may be lost,
 hurting performance.  One important case is when multiple processes
-are simply reading the data structure.  There is no conflict, and
-all reads should be allowed to proceed concurrently.  In many
-applications, reads are the majority of all accesses.
+are simply reading the data structure.
+In many applications, reads are the majority of all accesses.
+Allowing reads to proceed concurrently can significantly improve performance.
 
@@
 The lock is used to protect access to those counters.
-The program shows two kinds of processes: \texttt{reader()} and
-\texttt{writer()}.  The critical section is spread between two
-labels: readers access \texttt{@rcs} and writers access \texttt{@wcs}.
+The program shows a process that executes in a loop.
+Each time, it decides whether to read or write.
+The critical section is spread between two labels:
+readers access \texttt{@rcs} and writers access \texttt{@wcs}.
 The specification is that if a reader is at label \texttt{@rcs},
@@
 label \texttt{@rcs} \emph{and} there cannot be another writer at
-label \texttt{@wcs}.
-
-A reader process that wants to enter \texttt{@rcs} first waits
-until there are no writers: \texttt{nwriters == 0}.
-If so, it increments the number of readers.  Similarly, a writer
-process waits until there are no readers \emph{or} writers.
+label \texttt{@wcs}.  Figure~\ref{fig:rdwr} shows a high-level
+specification for two processes.
+
+\begin{quote}
+\begin{itemize}
+\item Draw additional states and steps in Figure~\ref{fig:rdwr} to 
+for three processes.
+\end{itemize}
+\end{quote}
+
+A process that wants to read first waits until there are no writers:
+\texttt{nwriters == 0}.  If so, it increments the number of readers.
+Similarly, a
+process that wants to write waits until there are no readers \emph{or} writers.
 If so, it increments the number of writers.
@@
 \item $n$ readers at \texttt{@rcs} $\Rightarrow \mathtt{nreaders} \ge n$,
-\item $n$ writers at \texttt{@rcs} $\Rightarrow \mathtt{nwriters} \ge n$,
+\item $n$ writers at \texttt{@wcs} $\Rightarrow \mathtt{nwriters} \ge n$,
 \item $(\mathtt{nreaders} \ge 0 \land \mathtt{nwriters} = 0) \lor
@@
 \item $n$ readers at \texttt{@rcs} $\Rightarrow \mathtt{nreaders} \ge n$,
-\item $\exists$ writer at \texttt{@rcs} $\Rightarrow \mathtt{nreaders} = 0$,
+\item $\exists$ writer at \texttt{@wcs} $\Rightarrow \mathtt{nreaders} = 0$,
 \end{itemize}

### 2020-05-30 book/paper.tex
Started working on semaphores

@@
 
+Critical sections are useful when accessing a shared data
+structure, particularly when that access requires multiple underlying
+machine instructions.  A simple counter is a very simple example of
+a data structure.  A more involved one would be access to a binary tree.
+Adding a node to a binary tree or re-balancing a tree often requires
+multiple operations.  Maintaining consistency is certainly much easier
+(although not necessarily impossible) if during this time no other
+process also tries to access the binary tree.
+
 A critical section is often modeled as processes in an infinite loop
@@
 
+\begin{figure}
+\begin{code}
+\begin{verbatim}
+def Semaphore(cnt):
+    result = cnt;
+;
+def P(sema):
+    let blocked = True:
+        while blocked:
+            atomic:
+                if (^sema) > 0:
+                    ^sema = (^sema) - 1;
+                    blocked = False;
+                ;
+            ;
+        ;
+    ;
+;
+def V(sema):
+    atomic:
+        ^sema = (^sema) + 1;
+    ;
+;
+\end{verbatim}
+\end{code}
+\caption{The \texttt{Semaphore} interface in the \texttt{synch} module.}
+\label{fig:semaphore}
+\end{figure}
+
+\begin{figure}
+\begin{code}
+\verbatiminput{../PCsema.cxl}
+\end{code}
+\caption{Bounded Buffer implementation using semaphores.}
+\label{fig:boundedbuffer}
+\end{figure}
+
 So far we have looked at how to protect a single resource.  Sometimes
@@
 protection.  That is, if there are only $n$ resources, no more than
-$n$ processes can access them at a time.
+$n$ can be used at a time; if all are in use and a process come along
+that needs one of the resources, it has to wait until another process
+releases one of the resources.  Note that we cannot solve this problem
+simply using a lock per resource---allocating a resource requires access
+to all of them.
+
+A good example is the so-called \emph{bounded buffer}.  It is essentially
+a queue implemented using a circular buffer of a certain length and two pointers:
+one where items are inserted and one from where items are extracted.  If the
+buffer is full, processes that want to add more items have to wait.  In the usual
+formulation, processes that want to extract an item when the buffer is empty
+also have to wait.  It would be easy to build such a thing using a busy waiting
+approach, but again, we would like to avoid busy waiting.
+
+Introduced by the famous Dutch computer scientist Edsger Dijkstra,
+a \emph{semaphore} is a primitive that fits the bill.  A semaphore is essentially
+a counter that can be incremented and decremented but is not allowed to go
+below zero.  The semaphore counter is typically initialized to the number of
+resources available.  When allocating a resource, a process decrements the
+counter using the \texttt{procure} or simply \texttt{P} operation.  The
+\texttt{procure} operation blocks the invoking process if the counter is zero.
+To release the resource, a process increments the resource using the
+\texttt{vacate} or {V} operation.
+Figure~\ref{fig:sema} shows the \texttt{synch} module implementation of
+semaphores.
 

### 2020-06-03 book/paper.tex
Fixed the triangle program
Fixed figures a bit
Added a new introduction

@@
 
-\chapter{Introduction to Programming with CXL}
-
+\chapter{On Concurrent Programming}
+
+Programming with concurrency is hard.  On the one hand concurrency
+can make programs faster than sequential ones, but having multiple
+processes concurrently read and update shared variables and
+synchronize with one another makes programs more complicated than
+programs where only one thing happens at a time.  Why is this?
+There are, at least, two reasons:
+\begin{itemize}
+\item The execution of a sequential program is mostly \emph{deterministic}.
+If you run it twice with the same input, the same things will happen.
+Bugs are typically easily reproducible.
+\item Each statement and each function can be thought of as happening
+\emph{atomically} because there is no other activity interfering with
+their execution.
+\end{itemize}
+The lack of determinism and atomicity in concurrent programs make them
+not only hard to reason about, but also hard to test.
+Running the same test of concurrent code twice is likely to produce
+two different results.  More problematically, a test may trigger a
+bug only for certain ``lucky'' executions.  Due to the probabilistic
+nature of concurrent code, some bugs may be highly unlikely to get
+triggered even when running a test millions of times.  And even if
+a bug does get triggered, the source of the bug may be hard to find
+because it is hard to reproduce.
+
+This book is intended to help people with understanding and
+developing concurrent code.  In particular, it uses a new tool
+called CXL that help with \emph{testing} concurrent code.
+The approach is based on \emph{model checking}: instead of relying
+on luck, CXL will run \emph{all possible executions} of a particular
+test program.  So even if a bug is unlikely to occur, if the test
+\emph{can} expose the bug it \emph{will}.  Importantly, if the bug is
+found, the model checker precisely shows how to trigger it in
+the smallest number of steps.
+
+Model checking is not a replacement for formal verification.  Formal
+verification proves that a program is correct.  Model checking only
+verifies that a program is correct for some \emph{model}.  Think of
+a model as a test program.  Because model checking tries every possible
+execution, this means that the test program needs to be relatively
+simple.  In particular, it needs to have a relatively small number of
+reachable states.
+If the model is too large, it may run for longer than we may care to
+wait for or even run out of memory.
+
+So why is this useful?  Consider, for example, a sorting algorithm.
+Now suppose we create a test program, a model, that tries sorting
+\emph{all} lists of up to five numbers chosen from the set \{ 1,
+2, 3, 4, 5 \}.  Model checking proves that for those particular
+scenarios the sorting algorithm works: the output is a sorted
+permutation of the input.  In some sense it is an excellent test:
+it will have considered all corner cases, including lists where all
+numbers are the same, lists that are already sorted or reversely
+sorted, etc.  If there is a bug in the sorting algorithm, most
+likely it would be triggered and the model checker would produce a
+scenario that would make it easy to find the source of the bug.
+However, if the model checker does not find any bugs, we do not
+know necessarily that the algorithm works for lists with more than
+five numbers or for lists that have values other than the numbers
+1 through 5.  Still, we would expect that the likelihood that there
+are bugs remaining in the sorting algorithm is small.
+
+But model checking can also help with proving an algorithm correct.
+The reason is that the correctness of many programs is built around
+\emph{invariants}: predicates that must hold for every state in the
+execution of a program.  A model checker can find violations of
+those invariants when evaluating a model and provide valuable early
+feedback to somebody who is trying to construct a proof.
+
+So what is CXL?
 CXL is a concurrent programming language.  It was designed to teach
@@
 in most other programming languages---instead CXL programs are
-``model checked'' to verify that the program has certain desirable
-properties and does not suffer from bugs.  Model checking is
-limited---it can only check programs that have ``finite state
-spaces.''  But even programs that have infinite state spaces can
-often be constrained (for example, by specifying a finite set of
-test inputs) so that model checking is possible.  If, for
-example, a sorting algorithm is shown to work for any subset of
-some set of say, 5, different values, then one can hope that it
-will also work for sets that are larger than 5 values or sets that
-use more than 5 different values.
+model checked to verify that the program has certain desirable
+properties and does not suffer from bugs.
 
@@
 with the concepts of CPU, memory, register, and stack.
+
+\chapter{Introduction to Programming with CXL}
 
@@
 \begin{center}
-\includegraphics[width=2in]{figures/mutex.png}
+\includegraphics[width=4in]{figures/mutex.jpeg}
 \end{center}
@@
 \begin{center}
-\includegraphics[width=2in]{figures/rdwr.png}
+\includegraphics[width=2in]{figures/rdwr.jpeg}
 \end{center}
@@
 
+\begin{figure}
+\begin{center}
+\includegraphics[width=2.5in]{figures/pc.jpeg}
+\end{center}
+\caption{High-level state diagram specification of producers/consumers.}
+\label{fig:pc}
+\end{figure}
+
 So far we have looked at how to protect a single resource.  Sometimes
@@
 \texttt{vacate} or {V} operation.
-Figure~\ref{fig:sema} shows the \texttt{synch} module implementation of
+Figure~\ref{fig:semaphore} shows the \texttt{synch} module implementation of
 semaphores.

### 2020-06-04 book/paper.tex
Latest

@@
 feedback to somebody who is trying to construct a proof.
+Throughout this book we will include, as much as possible, examples
+of such invariants as they often provide excellent insight into
+why a particular algorithm works.
 
@@
 Like Python, CXL is an imperative,
-dynamically typed and garbage collected programming language.
+dynamically typed, and garbage collected programming language.
 There are also some important differences:
@@
 \item CXL does not support I/O of any kind;
-\item CXL does not support operator precedence or associativity.  Use
-parentheses liberally to remove ambiguity.
+\item CXL only supports basic operator precedence or associativity.
+Use parentheses liberally to remove ambiguity.
 \end{itemize}
@@
 ``Running'' this CXL program will try all possible executions, which
-includes all possible values for $x$.  Essentially, the $\texttt{choose}(S)$
+includes all possible values for $x$.  Try it out:
+
+\begin{code}
+\begin{verbatim}
+$ ./cxl triangle.cxl
+#states = 13
+no issues found
+$
+\end{verbatim}
+\end{code}
+
+Essentially, the $\texttt{choose}(S)$
 operator provides the input to the program by selecting some value from the
@@
 correct.  If the program is correct, the output of CXL is the size of the
-``state graph'' and its diameter (more on this later).  If not, CXL also
+``state graph'' (13 states in this case).  If not, CXL also
 reports what went wrong, typically by displaying a summary of an execution in
@@
 \begin{quote}
-Try it out!
+Experiment!
 \begin{itemize}
@@
     0 Jump 39
-   1 Frame ('incrementer', 'Up.cxl', 1, 5) [('self', 'Up.cxl', 1, 17)] 12
+    1 Frame incrementer ['self'] 12
 Up.cxl:2     count = count + 1;
-   2 Constant (1, 'Up.cxl', 2, 21)
-   3 Name ('count', 'Up.cxl', 2, 13)
+    2 Push 1
+    3 PushAddress count
     4 Load 1
-   5 Nary ('+', 'Up.cxl', 2, 19) 2
-   6 Name ('count', 'Up.cxl', 2, 5)
+    5 2-ary +
+    6 PushAddress count
     7 Store 1
 Up.cxl:3     done[self] = True;
-   8 Constant (True, 'Up.cxl', 3, 18)
-   9 LoadVar ('self', 'Up.cxl', 3, 10)
-   10 Name ('done', 'Up.cxl', 3, 5)
+    8 Push True
+    9 PushVar self
+   10 PushAddress done
    11 Store 2
@@
 \begin{verbatim}
+CXL Assertion failed ('assert', 'Up.cxl', 9, 5) 1
+#states = 79
 ==== Safety violation ====
+__init__/() [0,39-56]       57 dict{ .count:0, .done:dict{ 0:False, 1:False } }
 incrementer/0 [1-6]          7 dict{ .count:0, .done:dict{ 0:False, 1:False } }
@@
 \begin{enumerate}
-\item A token containing the name of the method;
-\item A list of tokens containing the names of the arguments of the method;
-\item A program counter of the method's \texttt{Return} instruction.
+\item The name of the method;
+\item A list containing the names of the arguments of the method;
+\item The program counter of the method's \texttt{Return} instruction.
 \end{enumerate}
-
-A \emph{token} is a four-tuple containing a \emph{lexeme}, a file name, a
-line number in the file, and a column number in the file.  For example,
-\texttt{(’up’, ’Up.cxl’, 1, 5)} represents the lexeme \texttt{up} in
-file \texttt{Up.xcl}, line~1, column~5.  Tokens make it easy to relate
-the CXL machine instructions to the CXL source code.
 
@@
 \begin{enumerate} \setcounter{enumi}{1}
-\item the \texttt{Constant} instruction pushes the constant 1
+\item the \texttt{Push} instruction pushes the constant 1
 onto the stack of the process.
-\item The \texttt{Load} instruction loads the shared variable
-\texttt{count} from the shared memory and pushes its value onto the
-stack.
-\item \texttt{Nary} stands for \texttt{n-ary}: in this case it is a \texttt{+}
-operation with 2 arguments.  It pops two values from the stack, adds them,
-and pushes the result back onto the stack.
-\item The \texttt{Name} instruction pushes the address of the \texttt{count}
+\item the \texttt{PushAddress} instruction pushes the address of
+the shared variable \texttt{count} onto the stack.
+\item The \texttt{Load} instruction pops the address of the
+\texttt{count} variable and then pushes the value of the
+\texttt{count} variable onto the stack.
+\item \texttt{2-ary} is a \texttt{+} operation with 2 arguments.
+It pops two values from the stack (1 and the value of \texttt{count}),
+adds them, and pushes the result back onto the stack.
+\item The \texttt{PushAddress} instruction pushes the address of the \texttt{count}
 variable onto the stack.
 \item The \texttt{Store} instruction pops the address of a variable and pops
-a CXL value, and updates the memory by replacing the CXL value oat the given
-address with the new CXL value.
+a CXL value (the sum of the \texttt{count} variable and 1), and updates the
+\texttt{count} variable.
 \end{enumerate}
@@
 \begin{enumerate}
-\item Process 0 executed instructions 1 through 5, loading the value of
+\item Process 0 executed instructions 1 through 7, loading the value of
 \texttt{count} but stopping just before storing 1 into \texttt{count};
-\item Process 1 executed instructions 1 through 10, storing 1 into
+\item Process 1 executed instructions 1 through 12, storing 1 into
 into \texttt{count} and storing \texttt{True} into \texttt{done[2]};

### 2020-06-04 book/paper.tex
Latest version

@@
 \usepackage{verbatim}
+\usepackage{fancyvrb}
 \usepackage{comment}
@@
 \begin{code}
-\verbatiminput{triangle.cxl}
+\VerbatimInput[xleftmargin=5mm,numbers=left]{triangle.cxl}
 \end{code}
@@
 \begin{code}
-\verbatiminput{Up.cxl}
+\VerbatimInput[xleftmargin=5mm,numbers=left]{Up.cxl}
 \end{code}
@@
 along with an optional tag specified in the \texttt{spawn} statement.
-The default tag is \texttt{()}.
+The default tag is the first argument to the method.
 In Figure~\ref{fig:inc}, the created processes have name tags
@@
 \texttt{Up.cxl} program.
-It starts by reporting that the assertion on line 12 failed and that
+It starts by reporting that the assertion on line 9 in column 5 failed and that
 \texttt{count} has value 1 instead of 0:
-\texttt{Assertion failed (’assert’, ’Up.cxl’, 12, 1) 1}.
+\texttt{Assertion failed (’assert’, ’Up.cxl’, 9, 5) 1}.
 Next it reports an execution that failed this assertion.  The output has
@@
 \begin{code}
-\verbatiminput{csbarebones.cxl}
+\VerbatimInput[xleftmargin=5mm,numbers=left]{csbarebones.cxl}
 \end{code}
@@
 \begin{code}
-\verbatiminput{cs.cxl}
+\VerbatimInput[xleftmargin=5mm,numbers=left]{cs.cxl}
 \end{code}
@@
 \begin{code}
-\verbatiminput{UpLock.cxl}
+\VerbatimInput[xleftmargin=5mm,numbers=left]{UpLock.cxl}
 \end{code}
@@
 \begin{code}
-\verbatiminput{UpFlags.cxl}
+\VerbatimInput[xleftmargin=5mm,numbers=left]{UpFlags.cxl}
 \end{code}
@@
 \begin{code}
-\verbatiminput{UpTurn.cxl}
+\VerbatimInput[xleftmargin=5mm,numbers=left]{UpTurn.cxl}
 \end{code}
@@
 \begin{code}
-\verbatiminput{../Peterson.cxl}
+\VerbatimInput[xleftmargin=5mm,numbers=left]{../Peterson.cxl}
 \end{code}
@@
 \begin{code}
-\verbatiminput{../PetersonProof.cxl}
+\VerbatimInput[xleftmargin=5mm,numbers=left]{../PetersonProof.cxl}
 \end{code}
@@
 \begin{code}
-\verbatiminput{../PetersonMethod.cxl}
+\VerbatimInput[xleftmargin=5mm,numbers=left]{../PetersonMethod.cxl}
 \end{code}
@@
 \begin{code}
-\verbatiminput{UpTAS.cxl}
+\VerbatimInput[xleftmargin=5mm,numbers=left]{UpTAS.cxl}
 \end{code}
@@
 \begin{code}
-\verbatiminput{UpTASinv.cxl}
+\VerbatimInput[xleftmargin=5mm,numbers=left]{UpTASinv.cxl}
 \end{code}
@@
 \begin{code}
-\verbatiminput{../TAS.cxl}
+\VerbatimInput[xleftmargin=5mm,numbers=left]{../TAS.cxl}
 \end{code}
@@
 \begin{code}
-\verbatiminput{UpFixed.cxl}
+\VerbatimInput[xleftmargin=5mm,numbers=left]{UpFixed.cxl}
 \end{code}
@@
 \begin{code}
-\verbatiminput{../RWbusy.cxl}
+\VerbatimInput[xleftmargin=5mm,numbers=left]{../RWbusy.cxl}
 \end{code}
@@
 \begin{code}
-\verbatiminput{../RWlock.cxl}
+\VerbatimInput[xleftmargin=5mm,numbers=left]{../RWlock.cxl}
 \end{code}
@@
 \begin{code}
-\verbatiminput{../PCsema.cxl}
+\VerbatimInput[xleftmargin=5mm,numbers=left]{../PCsema.cxl}
 \end{code}

### 2020-06-04 book/paper.tex
Latest

@@
 developing concurrent code.  In particular, it uses a new tool
-called CXL that help with \emph{testing} concurrent code.
+called CXL that helps with \emph{testing} concurrent code.
 The approach is based on \emph{model checking}: instead of relying
@@
 However, if the model checker does not find any bugs, we do not
-know necessarily that the algorithm works for lists with more than
+know for sure that the algorithm works for lists with more than
 five numbers or for lists that have values other than the numbers
@@
 those invariants when evaluating a model and provide valuable early
-feedback to somebody who is trying to construct a proof.
-Throughout this book we will include, as much as possible, examples
+feedback to somebody who is trying to construct a proof, even an
+informal one.
+Throughout this book we will include examples
 of such invariants as they often provide excellent insight into
@@
 in most other programming languages---instead CXL programs are
-model checked to verify that the program has certain desirable
+model checked to test that the program has certain desirable
 properties and does not suffer from bugs.
@@
 
-CXL programs manipulates CXL values.
+CXL programs manipulate CXL values.
 CXL values are recursively defined:
@@
 and dictionaries that map CXL values to other CXL values.
-
+%
 Another type of CXL value is the \emph{atom}.  It is essentially
@@
 \end{code}
-This is unfamiliar, but in CXL square brackets can be used in the
-same way as parentheses, so you can express it in the more familiar
-form:
+This is unfamiliar to Python programers, but in CXL square brackets can be used
+in the same way as parentheses, so you can express it in the more familiar form:
 \begin{code}
@@
 Note that this is different from the empty set, which is \texttt{\{\}}.
-As in Python, you can create singleton tuples by adding an
-extra ",". For example, \texttt{(1,)}.
+As in Python, you can create singleton tuples by including a comma.
+For example, \texttt{(1,)}.
 
@@
 be in the same state.
-The initial state of the CXL memory is the empty tuple, \texttt{()}.
+The initial state of the CXL memory is the empty dictionary, \texttt{()}.
 The context bag has an initial context in it with name tag
@@
 Both the memory and the register of a process always contain
-a special type of CXL value in which the root is a dictionary
-that maps atoms to CXL values.  We call this a \emph{directory}.
+a dictionary that maps atoms to CXL values.  We call this a \emph{directory}.
 A directory represents the state of a collection of variables named by the atoms.
-
+%
 Because directories are CXL values themselves,
@@
 by a sequence of atoms, like a path name in the file system hierarchy.  We call
-such a sequence the \emph{address} of a CXL value.
+such a sequence the \emph{address} of a CXL value, and it is relative to the
+``root'' of the CXL value.
 
@@
 The CXL machine is predominantly a \emph{stack machine}.
+All instructions are atomically executed.
 Most instructions pop values from the stack or push values onto the stack.
-Initially there is one process that starts executing at instruction 0.
-In this case, there is a \texttt{JUMP} instruction that sets the program
-counter to 22.  At program counter 1 is the code for the \texttt{up} method.
+Initially there is one process (with name tag \texttt{\_\_init\_\_/()}
+that starts executing at instruction 0 and keeps executing until the last
+execution in the program to initialize the state.
+In this case, the first instruction is a \texttt{JUMP} instruction that sets the
+program counter to 39.
+At program counter 1 is the code for the \texttt{incrementer} method.
 All methods start with a \texttt{Frame} instruction and end with a \texttt{Return}
@@
 It pops two values from the stack (1 and the value of \texttt{count}),
-adds them, and pushes the result back onto the stack.
+adds them and pushes the result back onto the stack.
 \item The \texttt{PushAddress} instruction pushes the address of the \texttt{count}
@@
 \item The \texttt{Store} instruction pops the address of a variable and pops
-a CXL value (the sum of the \texttt{count} variable and 1), and updates the
+a CXL value (the sum of the \texttt{count} variable and 1) and updates the
 \texttt{count} variable.
@@
 
-Figure~\ref{fig:incoutput} shows the output produced by running the
+Once initialization completes, any processes that were spawned can run.
+You can think of CXL as trying every possible interleaving of processes executing
+instructions.
+Figure~\ref{fig:incoutput} shows the output produced by running CXL on the
 \texttt{Up.cxl} program.
@@
 \item Process 1 executed instructions 1 through 12, storing 1 into
-into \texttt{count} and storing \texttt{True} into \texttt{done[2]};
+into \texttt{count} and storing \texttt{True} into \texttt{done[1]};
 \item Process 0 continues execution, storing value 1 into \texttt{count}
-and storing \texttt{True} into \texttt{done[1]}.
+and storing \texttt{True} into \texttt{done[0]}.
 \end{enumerate}
@@
 In this chapter we will go through a few reasonable but broken attempts.
-At the heart of the problem is that we would like make sure that when
-the \texttt{count} variable is being updated that no other process is
+At the heart of the problem is that we would like make sure that, when
+the \texttt{count} variable is being updated, no other process is
 trying to do the same thing.  We call this a \emph{critical section}: a
@@
 structure, particularly when that access requires multiple underlying
-machine instructions.  A simple counter is a very simple example of
+machine instructions.  A counter is a very simple example of
 a data structure.  A more involved one would be access to a binary tree.
-Adding a node to a binary tree or re-balancing a tree often requires
-multiple operations.  Maintaining consistency is certainly much easier
+Adding a node to a binary tree, or re-balancing a tree, often requires
+multiple operations.  Maintaining ``consistency'' is certainly much easier
 (although not necessarily impossible) if during this time no other
@@
 A critical section is often modeled as processes in an infinite loop
-entering and exiting the critical section.  In CXL:
-
-\begin{code}
-\begin{verbatim}
-while True:
-    @ncs: pass; # in non-critical section
-    # enter critical section
-    @cs: pass;  # in critical section
-    # exit critical section
-;
-\end{verbatim}
-\end{code}
-
+entering and exiting the critical section.
+Figure~\ref{fig:csbarebones} shows the CXL code.
 Here \texttt{@ncs} and \texttt{@cs} are \emph{labels}, each identifying
@@
 
-Figure~\ref{fig:csbarebones} shows the code.
 The assertion also makes use of the \texttt{nametag()} operator that

### 2020-06-04 book/paper.tex
Latest

@@
 section, \texttt{turn} == $i$.
-(If you do not believe that, replace the assertion at the critical
-section with \texttt{assert turn == self} and run CXL.)
 Since \texttt{turn} cannot be 0 and 1 at
@@
 The process continues to be polite, waiting in a \texttt{while} loop
-until either the other process is no where near the critical section
+until either the other process is nowhere near the critical section
 or has given way.
@@
 \emph{inductive invariant} $I$, which is some predicate that holds
-over the state.  We need $I$ to satisfy the following:
+over the state.  An inductive invariant $I$, by definition, satisfies the following:
 \begin{itemize}
-\item (ME1) $I$ should imply mutual exclusion.
-\item (ME2) $I$ should hold in the initial state.
+\item (ME1) $I$ implies mutual exclusion.
+\item (ME2) $I$ hold in the initial state.
 \item (ME3) If $I$ holds over some state $s$, and some process in the
-state takes a step, then $I$ should also hold in the resulting state.
+state takes a step, then $I$ also holds in the resulting state.
 \end{itemize}

### 2020-06-05 book/paper.tex
Improving paper

@@
 the number of steps.
-In such a proof, we need a so-called
-\emph{inductive invariant} $I$, which is some predicate that holds
-over the state.  An inductive invariant $I$, by definition, satisfies the following:
+In such a proof, we need a so-called \emph{inductive invariant} $I$ that
+implies mutual exclusion.
+An inductive invariant $I$ is a predicate over state that satisfies the following:
 \begin{itemize}
-\item (ME1) $I$ implies mutual exclusion.
-\item (ME2) $I$ hold in the initial state.
-\item (ME3) If $I$ holds over some state $s$, and some process in the
+\item $I$ hold in the initial state.
+\item For any state in which $I$ holds and any process in the
 state takes a step, then $I$ also holds in the resulting state.
@@
 One candidate for such an invariant is mutual exclusion itself.
-After all, it certainly satisfies ME1 and ME2.
+After all, it certainly holds over the initial state.
 And as CXL has already determined, mutual exclusion is an invariant:
 it holds over every \emph{reachable state}.
-Unfortunately, it is not an \emph{inductive} invariant and therefore does
-not satisfy ME3.
+Unfortunately, it is not an \emph{inductive} invariant.
 To see why, we need to consider an \emph{unreachable} state.
@@
 
-Let $C(\mathtt{self}) = \mathtt{flags}[1 - \mathtt{self}] \land
-\mathtt{turn} = 1 - \mathtt{self}$, that is, the condition on the \texttt{while} loop.
-For Peterson's Algorithm, an inductive invariant that works well is:
+Let $C(i) = \mathtt{flags}[1 - i] \land
+\mathtt{turn} = 1 - i$, that is, the condition on the \texttt{while} loop
+for process~$i$.
+Let predicate $I_p(i)$ be the following.
 if process $i$ is label \texttt{@cs} (i.e., process $i$ is in the critical section),
@@
 $\mathtt{flags}[i]$ but still before setting \texttt{turn} to $1-i$.
-Let's call it $I_p(i)$.
+More formally, $I_p(i) = \mathtt{process}(i)@cs) \Rightarrow (\lnot C(i) \lor \mathtt{process}(1-i)@gate)$.
+For Peterson's Algorithm, an inductive invariant that works well is
+$I_p(0) \land I_p(1)$.
 
@@
 
-To see that $I_p(i)$ implies mutual exclusion, suppose not.  Then
+To see that the inductive invariant implies mutual exclusion, suppose not.  Then
 when both process~0 and process~1 are in the critical section, the
@@
 So we conclude that $\mathtt{turn} = 0 \land \mathtt{turn} = 1$, a
-logical impossibility.  Thus $I_p(i)$ satisfies ME1.
-
-$I_p(i)$ satisfies ME2 trivially, because in the initial state no process
+logical impossibility.  Thus $I_p(i)$ implies mutual exclusion.
+
+To see that $I_p(0) \land I_p(1)$ is, in fact, an inductive invariant, first note that
+it certainly holds in the initial state, because in the initial state no process
 is in the critical section.
-
-Finally we have to show that $I_p(i)$ satisfies ME3. Without loss of
-generality, suppose $i=0$ (a benefit from the fact that the algorithm is
-symmetric for both processes).  We have to show that if we are in a state
+Without loss of generality, suppose $i=0$ (a benefit from the fact that the algorithm is
+symmetric for both processes).  We still have to show that if we are in a state
 in which $I_p(0)$ holds, then any step will result in a state in which
-$I_p(0)$ continues to hold.
-Since $I_p(0)$ holds, process~0 is at label \texttt{@cs}.  If process~0
+$I_p(0)$ still holds.
+If $I_p(0)$ holds, process~0 is at label \texttt{@cs}.  If process~0
 were to take a step, then in the next state process~0 would be no longer
@@
 
-From $I_p(0)$ we know that one of the following cases must hold before
+From $I_p(0)$ we know that one of the following three cases must hold before
 process~1 takes a step:
-\begin{itemize}
+\begin{enumerate}
 \item \texttt{flags[1] = False};
@@
 \item process~1 is at label \texttt{@gate}.
-\end{itemize}
+\end{enumerate}
 
@@
 the step \texttt{turn} will equal 0, and therefore the second case will hold
-after the step.
+after the step.  qed.
 
@@
 Peterson's Algorithm, he refreshed his memory by looking at the Wikipedia
-page.  The page claimed that the following predicate was invariant:
+page.  The page claimed that the following predicate is invariant:
 if process $i$ is in the critical section, then $\lnot C(i)$ (i.e.,
@@
 the tuple consisting of values \texttt{a} and \texttt{b}.
-You have already seen examples of this: the expression
-\texttt{atLabel.cs} invokes method \texttt{atLabel} with
-argument \texttt{.cs}.
 
@@
 (the parentheses are mandatory) is a pointer to \texttt{x}
-(also known as the \emph{address} of \texttt{x}.
+(also known as the \emph{address} of \texttt{x}).
 Conversely, if \texttt{p} is a pointer to a shared variable, then the
@@
 algorithm gets around the problem, it is not efficient, especially
-if generalized to multiple processes.  Instead, processors provide
+if generalized to multiple processes.  Instead, multi-core processors provide
 so-called \emph{interlock instructions}: special machine instructions
@@
 it does have support for executing multiple instructions atomically.
-This feature is available in the CXL language through its \texttt{atomic}
+This feature is available in the CXL language in two ways.
+First, any CXL statement can be made atomic by placing a label in front
+of it.  Second, a group of CXL statements can be made atomic
+through its \texttt{atomic}
 statement.
@@
 The operation copies the value of the shared variable to the
-private variable and then sets the shared variable to \texttt{True}.
+private variable (the ``test'')
+and then sets the shared variable to \texttt{True} (``set'').
 
@@
 one of these variables is \texttt{False}.
-Another invariant, $I_2$, is that if process $i$ is in the critical section,
+Another invariant, $I_2(i)$, is that if process $i$ is in the critical section,
 then \texttt{private[$i$] == False}.
-Between the two, it is clear that only one process can be in the
+Between the two (i.e., $I_1 \land \forall i: I_2(i)$),
+it is clear that only one process can be in the
 critical section at the same time.
@@
 Invariant $I_1$ is also easy to verify for exiting the critical section.
-Invariant $I_2$ is obvious as (i) process $i$ only proceeds to the critical
+Invariant $I_2(i)$ is obvious as (i) process $i$ only proceeds to the critical
 section if \texttt{private[$i$] == False}, and (ii) no other process modifies
@@
 CXL can check these invariants as well.  Figure~\ref{fig:tas} already
-has the code to check $I_2$.  But how would one go about checking an
+has the code to check $I_2(i)$.  But how would one go about checking an
 invariant like $I_1$.  Invariants must hold for every state.

### 2020-06-05 book/paper.tex
Latest

@@
 a set of $N$ processes.
-It uses $N+1$ variables.
+It is called \emph{spinlock}, because each process is ``spinning'' until
+it can acquire the lock.
+The program uses $N+1$ variables.
 Variable \texttt{shared} is initialized to \texttt{False} while
@@
 variable and a private variable for each process.   The private
-variables themselves are implemented as shared variables as well,
+variables themselves are actually implemented as shared variables,
 but they are accessed only by their respective processes.
@@
 variable---we only did so to be able to show and check the invariants.
-Figure~\ref{fig:tas2} shows a more straightforward implementation.
+Figure~\ref{fig:tas2} shows a more straightforward implementation of spinlock.
 The solution is similar to the na\"{i}ve solution of Figure~\ref{fig:uplock},
 but uses test-and-set to check and set the lock variable atomically.
-Compared to Peterson's Algorithm, this approach is general for
-any number of processes.
+This approach is general for any number of processes.
 

### 2020-06-05 book/paper.tex
Latest

@@
 in a state.  It introduces a new feature of CXL: the ability to have
-variables local to a method.  In this case, the variable \texttt{sum}
-is used to compute the number of variables that have value
+variables local to a method.  In this case, the process variable \texttt{sum}
+is used to compute the number of shared variables that have value
 \texttt{False}.
@@
 \begin{itemize}
-\item Draw additional states and steps in Figure~\ref{fig:rdwr} to 
+\item Draw additional states and steps in Figure~\ref{fig:rdwr}
 for three processes.

### 2020-06-05 book/paper.tex
Latest

@@
 \begin{verbatim}
-$ ./cxl triangle.cxl
+$ cxl triangle.cxl
 #states = 13
@@
 
+(For this to work, make sure \texttt{cxl} is in your command shell's search path.)
 Essentially, the $\texttt{choose}(S)$
@@
 listed in Figure~\ref{fig:inccode}.
+You can obtain this code by invoking \texttt{cxl} with the \texttt{-a} flag
+like so:
+\begin{code}
+\begin{verbatim}
+cxl -a Up.cxl
+\end{verbatim}
+\end{code}
 The CXL machine is predominantly a \emph{stack machine}.

### 2020-06-05 book/paper.tex
Fixed error reporting.  Said something about TLA+ and constants in book

@@
 
+CXL is heavily influenced by TLA+, TLC, and PlusCal~\cite{}.
+CXL is designed to have a lower learning curve than those
+systems, but is not as powerful.  When you finish this book
+and want to learn more, we strongly encourage checking
+those out.
+
 \chapter{Introduction to Programming with CXL}
@@
 which something went wrong.
+
+In CXL, constants have a default, but those can be overridden on the command
+line using the \texttt{-c} option.
+For example, if you want to test the code for \texttt{N = 100}, run:
+\begin{code}
+\begin{verbatim}
+$ cxl -c N=100 triangle.cxl
+#states = 103
+no issues found
+$
+\end{verbatim}
+\end{code}
 

### 2020-06-05 book/paper.tex
Working on trying to identify blocking vs busy waiting
Working on reader/writer lock

@@
 
-CXL is heavily influenced by TLA+, TLC, and PlusCal~\cite{}.
+CXL is heavily influenced by Leslie Lamport's work on
+TLA+, TLC, and PlusCal~\cite{}.
 CXL is designed to have a lower learning curve than those
@@
 those out.
+Another excellent resource is Fred Schneider's book ``On
+Concurrent Programming.''
 
@@
 
-\chapter{Reader/Writer Locks}
+\chapter{Reader/Writer Locks using Busy Waiting}
 
@@
 The solution is an example of what is called \emph{busy-waiting}:
-processes spin in a loop until some desirable condition is met.
+processes spin in a loop until some desirable application-level condition is met.
+
 The astute reader might wonder if obtaining an ordinary lock itself is an
@@
 (see Figure~\ref{fig:locks}).
-
 In most operating systems and programming language runtimes, however,
@@
 
+We call a process \emph{blocked} in a state if steps by that process
+do not change the shared memory in that state.  A process trying to
+acquire a test-and-set spinlock held by another process is a good example
+of a process being blocked: the process only reads the shared memory
+but does not change it.
+
+However, a process trying to obtain a read or write lock in
+Figure~ref{fig:rwbusy} has to actually obtain a spinlock just to check
+if the read or write lock is available.  So even when the read or write
+lock is not available, the process ends up updating the spinlock and
+therefore the shared memory.
+
+Thus, it is considered ok to have processes be blocked while waiting
+for application-specific conditions, but not for them to
+busy-wait for application-specific conditions.
+
+[TODO.  The following must be updated.]
+CXL does not complain about Figure~\ref{fig:rwbusy}, but using
+the right test, CXL can be used to check for busy-waiting.
+One approach is as follows.  First, obtain a lock in the initialization
+code by adding the following line to the end of the code:
+\begin{code}
+\begin{verbatim}
+call acquire_rlock();
+\end{verbatim}
+\end{code}
+
+Second, run CXL with the \texttt{-b} flag.  This flag tells CXL to check
+that all states are non-terminating and have all processes blocked.
+
+\chapter{Reader/Writer Lock, Part 2}
+
 \begin{figure}
@@
 
-So it is considered ok to have processes using spinlock a implement a
-lock, but not to busy-wait for application-specific conditions.
+\begin{figure}
+\begin{code}
+\VerbatimInput[xleftmargin=5mm,numbers=left]{../RWmulti.cxl}
+\end{code}
+\caption{Checking that multiple readers can acquire the read lock.}
+\label{fig:rwmulti}
+\end{figure}
 
@@
 \texttt{rwlock}.
+
+While CXL does not detect any issues, we must remember what it is checking
+for: it checks to make sure that there are never a reader and a writer
+in the critical section, and there are never two readers in the critical
+section.  Moreover, it checks progress: all executions eventually terminate.
+What it does \emph{not} check is if the code allows more than one reader
+in the critical section.  Indeed, we could have implemented the reader/writer
+lock as in Figure~\ref{fig:rwmulti} using an ordinary lock, never allowing
+more than one process in the critical section.
+
+Figure~\ref{fig:rwmulti} contains a new test that ensures that indeed more
+than one reader can enter the critical section at a time.  It does so by
+having processes sometimes wait in the critical section until all other readers
+are there.  If other readers cannot enter, then that process enters in an
+infinite loop that CXL will readily detect.  Try it out!
+
+An important lesson here is that one should not celebrate too early if CXL
+does not find any issues.  While a test may be successful, a test may not
+explore all desirable executions.  While CXL can help explore all cornercases
+in a test, it cannot find problems that the test is not looking for.
 

### 2020-06-05 book/paper.tex
Latest

@@
 
+\begin{figure}
+\begin{code}
+\VerbatimInput[xleftmargin=5mm,numbers=left]{../RWblock.cxl}
+\end{code}
+\caption{Checking for Busy Waiting.}
+\label{fig:rwblock}
+\end{figure}
+
 While correct, it is not considered a good solution.
@@
 
-[TODO.  The following must be updated.]
-CXL does not complain about Figure~\ref{fig:rwbusy}, but using
+CXL does not complain about Figure~\ref{fig:rwbusy}, but, using
 the right test, CXL can be used to check for busy-waiting.
-One approach is as follows.  First, obtain a lock in the initialization
-code by adding the following line to the end of the code:
-\begin{code}
-\begin{verbatim}
-call acquire_rlock();
-\end{verbatim}
-\end{code}
-
-Second, run CXL with the \texttt{-b} flag.  This flag tells CXL to check
+Figure~\ref{fig:rwblock} presents such a test.
+Note that the initialization code obtains a read lock, preventing
+all readers and writers from entering the critical section.
+In that case, those processes should ideally be blocked and CXL
+can test for that by running it with the \texttt{-b} flag.
+This flag tells CXL to check
 that all states are non-terminating and have all processes blocked.
@@
 in a test, it cannot find problems that the test is not looking for.
+
+\begin{quote}
+\begin{itemize}
+\item Why do you suppose the code in Figure~\ref{fig:rwmulti} uses a flag
+per process rather than a simple counter?
+\item Can you write a version of this test that uses a counter instead
+of a flag per process?
+\end{itemize}
+\end{quote}
 

### 2020-06-05 book/paper.tex
Latest

@@
 section and two readers try to enter.  The first reader successfully
-acquires \texttt{rlock} and but \emph{hangs} when trying to acquire
-\texttt{rwlock}, which is held by the writer.  The second reader hangs
+acquires \texttt{rlock} and but blocks (!) when trying to acquire
+\texttt{rwlock}, which is held by the writer.  The second reader blocks
 trying acquire \texttt{rlock} because it is held by the first reader.
@@
 of a flag per process?
+\item Write a ``library implementation'' of reader/writer locks so multiple
+reader/writer locks can be instantiated.
 \end{itemize}

### 2020-06-05 book/paper.tex
Latest
Fixed a few small things

@@
 
+\begin{quote}
+\begin{itemize}
+\item Is Peterson's Algorithm blocking or busy-waiting?
+\item Check your answer by developing a test.
+\end{itemize}
+\end{quote}
+
 \chapter{Reader/Writer Lock, Part 2}
@@
 
-So far we have looked at how to protect a single resource.  Sometimes
-we have multiple but a limited amount of resources that require
-protection.  That is, if there are only $n$ resources, no more than
-$n$ can be used at a time; if all are in use and a process come along
-that needs one of the resources, it has to wait until another process
-releases one of the resources.  Note that we cannot solve this problem
-simply using a lock per resource---allocating a resource requires access
-to all of them.
+So far we have looked at how to protect a single resource.
+Some types of resources may have multiple units or slots, each of which
+If there are only $n$ units, no more than $n$ units can be used at a time;
+if all are in use and a process comes along that needs one of the units,
+it has to wait until another process releases one of the units.
+Note that we cannot solve this problem simply using a lock per unit---allocating
+a unit requires access to the entire collection.
 
@@
 below zero.  The semaphore counter is typically initialized to the number of
-resources available.  When allocating a resource, a process decrements the
+units of a resource initially available.
+When allocating a resource, a process decrements the
 counter using the \texttt{procure} or simply \texttt{P} operation.  The

### 2020-06-05 book/paper.tex
Working on semaphore section

@@
 \end{code}
-\caption{A barebones critical section.}
+\caption{A barebones critical section with two processes.}
 \label{fig:csbarebones}
@@
 formulation, processes that want to extract an item when the buffer is empty
-also have to wait.  It would be easy to build such a thing using a busy waiting
+also have to wait.
+This problem is known as the ``Producer/Consumer Problem.''
+It would be easy to solve using a busy waiting
 approach, but again, we would like to avoid busy waiting.
@@
 semaphores.
+
+One can think of a semaphore as a generalization of a lock.  In fact, a
+lock can be implemented by a semaphore initialized to 1.  Then \texttt{P}
+acquires the lock, and \texttt{V} releases the lock.
+
+Figure~\ref{fig:boundedbuffer} presents the implementation of a bounded
+buffer using semaphores.  It features a circular bounded buffer \texttt{buf} with
+slots numbered 1 through \texttt{NSLOTS}.  There are two indexes into
+the buffer: \texttt{b\_in} specifies where the next produced item is inserted,
+while \texttt{b\_out} specifies from where the next can be consumed.
+As there may be multiple producers and multiple
+consumers, updates to the indexes, which are shared variables, must be protected.
+To this end we use two semaphores as locks: \texttt{l\_in} for \texttt{b\_in}
+and \texttt{l\_out} for \texttt{b\_out}.
+
+Without additional synchronization, the indexes may overflow and point to invalid
+entries in the circular buffer.
+To this end there is a semaphore \texttt{n\_full} that
+keeps track of the number of filled entries in the buffer and a semaphore
+\texttt{n\_empty} that keeps track of the number of empty entries in the buffer.
+
+To add an item to the bounded buffer (\texttt{produce(item)}), the producing
+process first has to wait until there is room in the bounded buffer.
+To this end, it invokes \texttt{P(n\_empty)}, which waits until
+\texttt{n\_empty > 0} and atomically decrements \texttt{n\_empty} once this
+is the case.  Next, it adds the item to the next position in the buffer.
+Since there may be more than one empty slot, multiple producers may attempt
+to do the same thing concurrently.  Hence the use of the \texttt{l\_in}
+semaphore used as a lock.  Finally, once the item is added, the
+\texttt{n\_full} semaphore is incremented to signal to consumers that
+an item is available.
+
+The consumer code is symmetric.
+
+Note the quite different usage of the \texttt{l\_} semaphores and \texttt{n\_}
+semaphores.
+People often call semaphores that are used as locks \emph{binary semaphores},
+as their counters only take on the values 0 and 1.
+They protect critical sections.
+The second type of semaphores are called \emph{counting semaphores}.
+They are essentially used to send signals between processes.
+However, binary and counting semaphores are implemented the same way.
 

### 2020-06-06 book/paper.tex
Working on producer/consumer

@@
 \begin{figure}
+\begin{center}
+\includegraphics[width=2.5in]{figures/pc.jpeg}
+\end{center}
+\caption{High-level state diagram specification of producers/consumers.}
+\label{fig:pc}
+\end{figure}
+
+\begin{figure}
 \begin{code}
@@
 
-\begin{figure}
-\begin{center}
-\includegraphics[width=2.5in]{figures/pc.jpeg}
-\end{center}
-\caption{High-level state diagram specification of producers/consumers.}
-\label{fig:pc}
-\end{figure}
-
 So far we have looked at how to protect a single resource.
@@
 This problem is known as the ``Producer/Consumer Problem.''
+
+Figure~\ref{fig:pc} gives a high-level description of what we want.  The
+two numbers in a state specify
+the number of items that the producers still can produce and
+the number of items that the consumers have consumed.  In this case
+there are three items to produce initially.
+
 It would be easy to solve using a busy waiting
 approach, but again, we would like to avoid busy waiting.
-
 Introduced by the famous Dutch computer scientist Edsger Dijkstra,
@@
 However, binary and counting semaphores are implemented the same way.
+
+We can run the code through CXL, but what does it test?  There are no
+assert statements in the code.
+We can split the functionality of this code into two pieces.  One is
+the synchronization part: producers should never run more than \texttt{NSLOTS}
+ahead of the consumers, but they should be able to produce up to \texttt{NSLOTS}
+items even if there are no consumers.  The other part is the data: we want
+to make sure that the items are sent through the buffer in order and without
+loss.  We will first focus on the synchronization.
+
+[RVR.  More here.]
 

### 2020-06-06 book/paper.tex
Added a dining philosopers chapter

@@
 
-\begin{quote}
-\begin{itemize}
-\item Is Peterson's Algorithm blocking or busy-waiting?
-\item Check your answer by developing a test.
-\end{itemize}
-\end{quote}
-
 \chapter{Reader/Writer Lock, Part 2}
@@
 reader/writer locks can be instantiated.
+\end{itemize}
+\end{quote}
+
+\chapter{Deadlock}
+
+\begin{figure}
+\begin{code}
+\VerbatimInput[xleftmargin=5mm,numbers=left]{../Diners.cxl}
+\end{code}
+\caption{Dining Philosophers.}
+\label{fig:diners}
+\end{figure}
+
+When multiple processes are synchronizing access to shared resources, they
+may end up in a \emph{deadlock} situation where one or more of the processes
+end up being blocked indefinitely because each is waiting for another to give
+up a resource.
+The famous Dutch computer scientist Edsger W.~Dijkstra illustrated this using
+a scenario he called ``Dining Philosophers.''
+
+Imagine five philosopers sitting around a table, each with a plate of food in
+front of them, and a fork between each two plates.  Each philosopher requires
+two forks to eat.  To start eating, a philosopher first picks up the fork on
+the left, then the fork on the right.  Each philospher likes to take breaks
+from eating to think for a while.  To do so, the philosopher puts down both
+forks.  The philosophers repeat this procedure.  Dijkstra had them repeating
+this for ever, but for the purposes of this book, philosphers can leave
+the table when they're not eating.
+
+Figure~\ref{fig:diners} implements the dining philosophers in CXL, using a
+process for each philosopher and a lock for each fork.  If you
+run it, CXL complains that the execution may not terminate, with all five
+processes being blocked trying to acquire the lock.
+
+\begin{quote}
+\begin{itemize}
+\item Do you see what the problem is?
+\item Does it depend on \texttt{N}, the number of philosophers?
+\item Can you come up with a solution?
+\end{itemize}
+\end{quote}
+
+We will discuss ways of preventing deadlock later, but one important
+approach to avoiding deadlock is to establish an ordering among the
+resources (in this case the forks) and, when needing more than one
+resource, always acquiring them in order.  In the case of the philosopers,
+they could prevent deadlock by always picking up the lower numbered fork
+before the higher numbered fork, like so:
+
+\begin{code}
+\begin{verbatim}
+    if left < right:
+        call lock &(forks[left]);
+        call lock &(forks[right]);
+    else:
+        call unlock &(forks[right]);
+        call unlock &(forks[left]);
+    ;
+\end{verbatim}
+\end{code}
+
+\begin{quote}
+\begin{itemize}
+\item Try it out!
 \end{itemize}
@@
 approach, but again, we would like to avoid busy waiting.
-Introduced by the famous Dutch computer scientist Edsger Dijkstra,
+Introduced by the Dijkstra,
 a \emph{semaphore} is a primitive that fits the bill.  A semaphore is essentially

### 2020-06-06 book/paper.tex
Added Italian Philosophers

@@
 \item Can you come up with a solution?
+\item Does it matter in what order the philosophers lay down their forks?
 \end{itemize}
@@
 \chapter{Semaphores}
-
-\begin{figure}
-\begin{center}
-\includegraphics[width=2.5in]{figures/pc.jpeg}
-\end{center}
-\caption{High-level state diagram specification of producers/consumers.}
-\label{fig:pc}
-\end{figure}
 
@@
 \begin{code}
-\VerbatimInput[xleftmargin=5mm,numbers=left]{../PCsema.cxl}
-\end{code}
-\caption{Bounded Buffer implementation using semaphores.}
-\label{fig:boundedbuffer}
+\VerbatimInput[xleftmargin=5mm,numbers=left]{../DinersSema.cxl}
+\end{code}
+\caption{Italian Philosophers.}
+\label{fig:dinerssema}
 \end{figure}
@@
 So far we have looked at how to protect a single resource.
-Some types of resources may have multiple units or slots, each of which
+Some types of resources may have multiple units.
 If there are only $n$ units, no more than $n$ units can be used at a time;
@@
 
-A good example is the so-called \emph{bounded buffer}.  It is essentially
-a queue implemented using a circular buffer of a certain length and two pointers:
-one where items are inserted and one from where items are extracted.  If the
-buffer is full, processes that want to add more items have to wait.  In the usual
-formulation, processes that want to extract an item when the buffer is empty
-also have to wait.
-This problem is known as the ``Producer/Consumer Problem.''
-
-Figure~\ref{fig:pc} gives a high-level description of what we want.  The
-two numbers in a state specify
-the number of items that the producers still can produce and
-the number of items that the consumers have consumed.  In this case
-there are three items to produce initially.
-
-It would be easy to solve using a busy waiting
-approach, but again, we would like to avoid busy waiting.
-Introduced by the Dijkstra,
+Consider a variant of the Dining Philosophers (which were clearly Dutch---who
+else need two forks to eat a meal?) that I call Italian Philosophers (who only
+need one fork each to eat spaghetti).  Again, there are five philosophers.
+This time, however, there is a glass sitting in the middle of the table with only
+three forks in it.  A philosopher needs one of those forks to eat, but clearly,
+no more than three philosophers can eat at a time.
+
+\begin{quote}
+\begin{itemize}
+\item See if you can come up with a deadlock-free solution using locks.  Represent
+    each fork as a lock.
+\end{itemize}
+\end{quote}
+
+It's not easy.  It would be easy to come up with a solution that uses a single lock
+and busy waiting, but we would like to avoid busy waiting.
+Introduced by Dijkstra,
 a \emph{semaphore} is a primitive that fits the bill.  A semaphore is essentially
@@
 \texttt{vacate} or {V} operation.
+
 Figure~\ref{fig:semaphore} shows the \texttt{synch} module implementation of
 semaphores.
-
+Figure~\ref{fig:dinerssema} shows a solution to the Italian Philosphers problem
+using a semaphore.
 One can think of a semaphore as a generalization of a lock.  In fact, a
@@
 acquires the lock, and \texttt{V} releases the lock.
+
+\chapter{Bounded Buffer}
+
+\begin{figure}
+\begin{center}
+\includegraphics[width=2.5in]{figures/pc.jpeg}
+\end{center}
+\caption{High-level state diagram specification of producers/consumers.}
+\label{fig:pc}
+\end{figure}
+
+\begin{figure}
+\begin{code}
+\VerbatimInput[xleftmargin=5mm,numbers=left]{../PCsema.cxl}
+\end{code}
+\caption{Bounded Buffer implementation using semaphores.}
+\label{fig:boundedbuffer}
+\end{figure}
+
+A good example of a resource with multiple units is the so-called \emph{bounded buffer}.
+It is essentially
+a queue implemented using a circular buffer of a certain length and two pointers:
+one where items are inserted and one from where items are extracted.  If the
+buffer is full, processes that want to add more items have to wait.  In the usual
+formulation, processes that want to extract an item when the buffer is empty
+also have to wait.
+This problem is known as the ``Producer/Consumer Problem.''
+
+Figure~\ref{fig:pc} gives a high-level description of what we want.  The
+two numbers in a state specify
+the number of items that the producers still can produce and
+the number of items that the consumers have consumed.  In this case
+there are three items to produce initially.
 

### 2020-06-06 book/paper.tex
fix

@@
 Consider a variant of the Dining Philosophers (which were clearly Dutch---who
-else need two forks to eat a meal?) that I call Italian Philosophers (who only
+else need two forks to eat a meal?) that we call Italian Philosophers (who only
 need one fork each to eat spaghetti).  Again, there are five philosophers.

### 2020-06-06 book/paper.tex
Added more challenges

@@
 Introduced by Dijkstra,
-a \emph{semaphore} is a primitive that fits the bill.  A semaphore is essentially
+a \emph{semaphore} is a synchronization primitive that fits the bill.
+A semaphore is essentially
 a counter that can be incremented and decremented but is not allowed to go
@@
 using a semaphore.
+
+\begin{quote}
+\begin{itemize}
+\item Can you come up with a test (an assertion) that shows that at most three
+philosophers are eating at the same time?
+\item Can you come up with a test (different code) that checks to see that
+the solution allows more than one philosopher to eat at the same time?
+\end{itemize}
+\end{quote}
+
 One can think of a semaphore as a generalization of a lock.  In fact, a
@@
 The second type of semaphores are called \emph{counting semaphores}.
-They are essentially used to send signals between processes.
+They can be used to send signals between processes.
 However, binary and counting semaphores are implemented the same way.

### 2020-06-06 book/paper.tex
Starting on new chapter
Added script

@@
 
+\begin{figure}
+\begin{code}
+\begin{verbatim}
+cxl -b -c NSLOTS=0 -c NPRODS=1 -c NCONSS=1 PCsema.cxl
+cxl -b -c NSLOTS=1 -c NPRODS=0 -c NCONSS=1 PCsema.cxl
+cxl    -c NSLOTS=1 -c NPRODS=1 -c NCONSS=0 PCsema.cxl
+cxl    -c NSLOTS=1 -c NPRODS=1 -c NCONSS=1 PCsema.cxl
+cxl -b -c NSLOTS=1 -c NPRODS=1 -c NCONSS=2 PCsema.cxl
+cxl -b -c NSLOTS=1 -c NPRODS=2 -c NCONSS=0 PCsema.cxl
+cxl    -c NSLOTS=1 -c NPRODS=2 -c NCONSS=1 PCsema.cxl
+cxl    -c NSLOTS=1 -c NPRODS=2 -c NCONSS=2 PCsema.cxl
+cxl -b -c NSLOTS=1 -c NPRODS=2 -c NCONSS=3 PCsema.cxl
+cxl    -c NSLOTS=2 -c NPRODS=1 -c NCONSS=0 PCsema.cxl
+cxl    -c NSLOTS=2 -c NPRODS=1 -c NCONSS=1 PCsema.cxl
+cxl -b -c NSLOTS=2 -c NPRODS=1 -c NCONSS=2 PCsema.cxl
+cxl    -c NSLOTS=2 -c NPRODS=2 -c NCONSS=0 PCsema.cxl
+cxl    -c NSLOTS=2 -c NPRODS=2 -c NCONSS=1 PCsema.cxl
+cxl    -c NSLOTS=2 -c NPRODS=2 -c NCONSS=2 PCsema.cxl
+cxl -b -c NSLOTS=2 -c NPRODS=2 -c NCONSS=3 PCsema.cxl
+cxl -b -c NSLOTS=2 -c NPRODS=3 -c NCONSS=0 PCsema.cxl
+cxl    -c NSLOTS=2 -c NPRODS=3 -c NCONSS=1 PCsema.cxl
+cxl    -c NSLOTS=2 -c NPRODS=3 -c NCONSS=2 PCsema.cxl
+cxl    -c NSLOTS=2 -c NPRODS=3 -c NCONSS=3 PCsema.cxl
+\end{verbatim}
+\end{code}
+\caption{Script for testing producer/consumer synchronization
+(assuming the code is in file \texttt{PCsema.cxl}).}
+\label{fig:pcscript}
+\end{figure}
+
 A good example of a resource with multiple units is the so-called \emph{bounded buffer}.
@@
 
+The first part can be tested with the code given in Figure~\ref{fig:pc}, by
+experimenting with different non-negative values for the three constants.
+In particular, any run with a positive \texttt{NSLOTS} and
+$\mathtt{NCONSS} \le \mathtt{NPRODS} \le \mathtt{NCONSS} + \mathtt{NSLOTS}$
+should terminate, while other values should lead to processes blocking.
+One could write a script like the one in Figure~\ref{fig:pcscript}.
+
 [RVR.  More here.]
+
+\chapter{Split Binary Semaphores}
 

### 2020-06-07 book/paper.tex
Finished first draft of bounded buffer chapter.

@@
 
+\begin{figure}
+\begin{code}
+\VerbatimInput[xleftmargin=5mm,numbers=left]{../PCsemadata.cxl}
+\end{code}
+\caption{Test whether the bounded buffer delivers the right data in the
+right order.}
+\label{fig:PCsemadata}
+\end{figure}
+
 Figure~\ref{fig:boundedbuffer} presents the implementation of a bounded
@@
 
-[RVR.  More here.]
+Figure~\ref{fig:PCsemadata} tests whether the correct data is sent in
+the correct order.  There are the same number of producers and consumers.
+Each producer $i$ produces two tuples: $(i, 0)$ and $(i, 1)$ in that order.
+Each consumer consumes two tuples and checks that the tuples are different
+and that, if the two tuples come from the same producer, then they have to
+be in the order sent.
+Moreover, consumers add the tuples they received to the set
+\texttt{received} (atomically---a lock could have been used here as well).)
+The \texttt{main()} process waits for all consumers to have received
+exactly the set of tuples that the producers produce.
 

### 2020-06-07 book/paper.tex
Starting on Split Binary Semaphores

@@
 a \emph{reader/writer lock}~\cite{courtois71}.
+[RVR: or was it David Parnas who first introduced them?]
 We will explore various ways of implementing reader/writer locks in
@@
 
+\begin{figure}
+\begin{code}
+\VerbatimInput[xleftmargin=5mm,numbers=left]{../RWbinsem.cxl}
+\end{code}
+\caption{Reader/Writer Lock implementation using Split Binary Semaphores.}
+\label{fig:RWsplitsema}
+\end{figure}
+
+Reader/writer locks and bounded buffers are both examples of what are
+sometimes called \emph{conditional critical regions} (CCRs).
+All critical regions are in fact conditional.  Even the basic one
+has the condition that a process can only enter if no other process
+is in the critical region.  Reader/writer locks refined that, loosening
+the restriction for improved performance.  In the case of bounded buffers,
+there were two CCRs.  First, a CCR
+to add an entry to the bounded buffer that can only be entered by one
+process and only if there is room in the bounded buffer (\texttt{produce}).
+Second, a CCR that can only be entered by one process and
+only if there is an item in the bounded buffer (\texttt{consume}).
+
+CCRs are easy to implement using busy waiting:
+\begin{code}
+\begin{verbatim}
+let blocked = True:
+    while blocked:
+        call lock();
+        if application-condition-holds:
+            blocked = False;
+        else:
+            call unlock():
+@cr: application-code;
+call unlock():
+\end{verbatim}
+\end{code}
+
+We have seen solutions to the reader/writer problem and the producer/consumer
+problem that do no busy-wait, but they are quite different.  It would be nice
+to have a general technique for all CCR problems.
+The ``Split Binary Semaphore'' is such a technique, originally proposed by
+Tony Hoare and popularized by Dijkstra.
+
 \chapter{Starvation}

### 2020-06-07 book/paper.tex
Rough draft of split binary semaphores

@@
 to have a general technique for all CCR problems.
-The ``Split Binary Semaphore'' is such a technique, originally proposed by
-Tony Hoare and popularized by Dijkstra.
+The ``Split Binary Semaphore'' (SBS) is such a technique, originally proposed by
+Tony Hoare and popularized by Edsger Dijkstra.
+SBS is an example of what one could call a ``baton-passing'' technique:
+when a process has the
+baton and is about to leave the critical region, giving up the baton, it first
+checks to see if there are any processes that are waiting and can now run.
+If so, it passes the baton directly to one of those processes.
+
+We will illustrate the technique using the reader/writer problem.
+Figure~\ref{fig:RWsplitsema} shows the
+\texttt{rlock\_acquire}, \texttt{rlock\_release},
+\texttt{wlock\_acquire}, and \texttt{wlock\_release} methods implemented using
+SBS.
+
+Step 1 is to enumerate all waiting conditions.  In the case of the reader/writer
+problem, there are two: a process that wants to read may have to wait for a
+writer to leave the critical regiion, while a process that wants to write may
+have to wait until all readers have left the critical region.  If there are $N$
+waiting conditions, the SBS technique will use $N+1$ binary semaphores: one for
+each waiting condition and an extra one that processes use when they first try
+to enter the critical region and do not yet know if they need to wait.
+We shall call this the \texttt{mutex}.
+
+An important invariant of the SBS technique is that the sum of the $N+1$ semaphores
+must always be 1 if there is no process in the CCR and 0 if there is
+a process in the CCR.  As it were, these semaphores taken together are emulating a
+single binary semaphore.
+Initially, \texttt{mutex} is 1 and the other $N$ semaphores
+are all 0.  To maintain the invariant, each process alternates \texttt{P} and
+\texttt{V} operations, starting with \texttt{P~\&(mutex)} when it tries to
+enter the CCR, and ending with a \texttt{V} operation on one of the semaphores
+(yet to be determined).
+
+Another important part of the SBS technique is to keep careful track of the
+state.  In the case of the reader/writer lock, the state consists:
+\begin{itemize}
+\item \texttt{nreadersEntered}: the number of readers in the critical region;
+\item \texttt{nwritersEntered}: the number of writers in the critical region;
+\item \texttt{nreadersWaiting}: the number of readers waiting to enter;
+\item \texttt{nwritersWaiting}: the number of writers waiting to enter.
+\end{itemize}
+Each of the
+\texttt{rlock\_acquire}, \texttt{rlock\_release},
+\texttt{wlock\_acquire}, and \texttt{wlock\_release} methods must maintain
+this state.
+To wit, check out \texttt{rlock\_acquire},  It first checks to see if there
+are any writers.  If so, it increments \texttt{nreadersWaiting}.  If not,
+it increments \texttt{nreadersEntered}.
+Either way, it calls the method \texttt{V\_one()}.
+
+Note that the four
+reader/writer methods never call \texttt{V} on a specific binary semaphore:
+\texttt{V\_one()} is the baton passing function that selects which of the
+binary semaphores to increment.
+Method \texttt{V\_one()} first checks to see if there are any readers or
+writers waiting.  If there are readers waiting and there are no writers
+in the critical section, it increments the semaphore associated with
+readers waiting.  If there are writers waiing and there are no readers
+nor writers in the critical section, it increments the semaphore associated
+with writers waiting.  If both conditions hold, it could increment either
+one---it would not matter.  However, it can only increment one of the semaphores.
+If none of the conditions hold, the \texttt{V\_one} increments \texttt{mutex}.
+
+Let us return now to \texttt{rlock\_acquire}, in the case that the process
+found there is a writer in the critical region.  After \texttt{V\_one()} returns,
+it invokes calls \texttt{P~\&(rsema)}, blocking on the semaphore associated
+with waiting readers.  When later some other process passes the baton to
+it by incrementing \texttt{rsema}), it updates the state by decrementing
+\texttt{nreadersWaiting} and incrementing \texttt{nreadersEntered}.  It
+then invokes \texttt{V\_one()} one more time.
+
+As an example, consider the case where there is a writer in the critical
+section and there are two readers waiting.  Let us see what happens when
+the writer calls \texttt{wlock\_release}:
+\begin{enumerate}
+\item After procuring \texttt{mutex},
+the sum of the three semaphores is 0.  The writer then decrements
+\texttt{nwritersEntered} and calls \texttt{V\_one()}.
+\item \texttt{V\_one()} finds that there are no writers in the critical region
+and there are two readers waiting.  It therefore vacates \texttt{rsema}.
+The sum of the semaphores is now 1, because \texttt{rsema} is 1.  This
+guarantees that only a waiting reader, procuring \texttt{rsema}, can enter
+the critical region.
+\item When it does, the reader decrements \texttt{nreadersWaiting}
+from 2 to 1, and increments \texttt{nReadersEntered} from 0 to 1.
+The reader finally calls \texttt{V\_one()}.
+\item Again, \texttt{V\_one()} finds that there are no writers and
+that there are readers waiting, so again it vacates \texttt{rsema},
+passing the baton to the one remaining waiting reader.
+\item The reader decrements \texttt{nreadersWaiting} from 1 to 0 and
+increments \texttt{nReadersEntered} from 1 to 2.
+\item Finally, the reader  calls \texttt{V\_one()}.
+\texttt{V\_one()} does not find any processes waiting,
+and so it vacates \texttt{mutex}.
+\end{enumerate}
 

### 2020-06-07 book/paper.tex
Added challenge
Updating terminology

--- 2020-06-07 book/paper.tex
@@
 the \texttt{count} variable is being updated, no other process is
-trying to do the same thing.  We call this a \emph{critical section}: a
+trying to do the same thing.  We call this a \emph{critical section}
+(aka critical region): a
 set of instructions where only one process is allowed to execute at a
@@
 However, a process trying to obtain a read or write lock in
-Figure~ref{fig:rwbusy} has to actually obtain a spinlock just to check
+Figure~\ref{fig:rwbusy} has to actually obtain a spinlock just to check
 if the read or write lock is available.  So even when the read or write
@@
 This flag tells CXL to check
-that all states are non-terminating and have all processes blocked.
+that all states are non-terminating and lead to a state in which
+all processes are blocked.
 
@@
 When allocating a resource, a process decrements the
-counter using the \texttt{procure} or simply \texttt{P} operation.  The
-\texttt{procure} operation blocks the invoking process if the counter is zero.
+counter using the \texttt{P} operation.  You can think of \texttt{P} standing
+for \emph{Procure}, as in procuring the resource associated with the semaphore.
+The \texttt{P} operation blocks the invoking process if the counter is zero.
 To release the resource, a process increments the resource using the
-\texttt{vacate} or {V} operation.
+\texttt{V} operation.  You can think of \texttt{V} standing for \emph{vacate},
+as in vacating the resource.
 
@@
 lock can be implemented by a semaphore initialized to 1.  Then \texttt{P}
-acquires the lock, and \texttt{V} releases the lock.
+procures the lock, and \texttt{V} vacates the lock.
 
@@
 \texttt{n\_empty > 0} and atomically decrements \texttt{n\_empty} once this
-is the case.  Next, it adds the item to the next position in the buffer.
+is the case.  In other words, the producer procures an empty slot.
+Next, it adds the item to the next position in the buffer.
 Since there may be more than one empty slot, multiple producers may attempt
 to do the same thing concurrently.  Hence the use of the \texttt{l\_in}
-semaphore used as a lock.  Finally, once the item is added, the
-\texttt{n\_full} semaphore is incremented to signal to consumers that
+semaphore used as a lock.  Finally, once the item is added, the process
+increments the \texttt{n\_full} semaphore to signal to consumers that
 an item is available.
@@
 
+\begin{figure}
+\begin{code}
+\VerbatimInput[xleftmargin=5mm,numbers=left]{../RWbinsem.cxl}
+\end{code}
+\caption{Reader/Writer Lock implementation using Split Binary Semaphores.}
+\label{fig:RWsplitsema}
+\end{figure}
+
+Reader/writer locks and bounded buffers are both examples of what are
+sometimes called \emph{conditional critical sections} (CCSs).
+All critical sections are in fact conditional.  Even the basic one
+has the condition that a process can only enter if no other process
+is in the critical section.  Reader/writer locks refined that, loosening
+the restriction for improved performance.
+In the case of bounded buffers, there were two CCSs.  First, a CCS
+to add an entry to the bounded buffer that can only be entered by one
+process and only if there is room in the bounded buffer (\texttt{produce}).
+Second, a CCS that can only be entered by one process and
+only if there is an item in the bounded buffer (\texttt{consume}).
+
+CCSs are easy to implement using busy waiting:
+\begin{code}
+\begin{verbatim}
+let blocked = True:
+    while blocked:
+        call lock();
+        if application-condition-holds:
+            blocked = False;
+        else:
+            call unlock():
+@cr: application-code;
+call unlock():
+\end{verbatim}
+\end{code}
+
+We have seen solutions to the reader/writer problem and the producer/consumer
+problem that do not busy-wait, but they are quite different.  It would be nice
+to have a general technique for all CCS problems.
+The ``Split Binary Semaphore'' (SBS) is such a technique, originally proposed by
+Tony Hoare and popularized by Edsger Dijkstra.
+SBS is an example of what one could call a ``baton-passing'' technique.
+A process needs the baton to access shared variables.
+When a process has the baton and does not need it any more,
+it first checks to see if there are any processes that are waiting for the
+baton and can make progress.
+If so, it passes the baton directly to one of those processes.
+If not, it simply gives up the baton.
+
+We will illustrate the technique using the reader/writer problem.
+Figure~\ref{fig:RWsplitsema} shows the
+\texttt{rlock\_acquire}, \texttt{rlock\_release},
+\texttt{wlock\_acquire}, and \texttt{wlock\_release} methods implemented using
+SBS.
+
+Step 1 is to enumerate all waiting conditions.  In the case of the reader/writer
+problem, there are two: a process that wants to read may have to wait for a
+writer to leave the critical regiion, while a process that wants to write may
+have to wait until all readers have left the critical section.  If there are $N$
+waiting conditions, the SBS technique will use $N+1$ binary semaphores: one for
+each waiting condition and an extra one that processes use when they first try
+to enter the critical section and do not yet know if they need to wait.
+We shall call this the \texttt{mutex}.
+
+An important invariant of the SBS technique is that the sum of the $N+1$ semaphores
+must always be 0 or 1, and it should be 0 when a process is accessing the shared
+variables.
+As it were, these semaphores taken together are emulating a single binary semaphore.
+Initially, \texttt{mutex} is~1 and the other $N$ semaphores
+are all~0.  To maintain the invariant, each process alternates \texttt{P} and
+\texttt{V} operations, starting with \texttt{P~\&(mutex)} when it tries to
+enter the CCS, and ending with a \texttt{V} operation on one of the semaphores
+(yet to be determined).
+
+Another important part of the SBS technique is to keep careful track of the
+state.  In the case of the reader/writer lock, the state consists of the
+following shared variables:
+\begin{itemize}
+\item \texttt{nreadersEntered}: the number of readers in the critical section;
+\item \texttt{nwritersEntered}: the number of writers in the critical section;
+\item \texttt{nreadersWaiting}: the number of readers waiting to enter;
+\item \texttt{nwritersWaiting}: the number of writers waiting to enter.
+\end{itemize}
+Each of the
+\texttt{rlock\_acquire}, \texttt{rlock\_release},
+\texttt{wlock\_acquire}, and \texttt{wlock\_release} methods must maintain
+this state.
+To wit, check out \texttt{rlock\_acquire}.  It first checks to see if there
+are any writers.  If so, it increments \texttt{nreadersWaiting}.  If not,
+it increments \texttt{nreadersEntered}.
+
+Either way, it vacates one of the semaphores next.
+\texttt{V\_one()} is the baton passing function that selects which of the
+binary semaphores to vacate.
+Method \texttt{V\_one()} first checks to see if there are any readers or
+writers waiting.  If there are readers waiting and there are no writers
+in the critical section, it vacates the semaphore associated with
+readers waiting.  If there are writers waiing and there are no readers
+nor writers in the critical section, it vacates the semaphore associated
+with writers waiting.  If both conditions hold, it could vacates either
+one---it would not matter.  However, it can only vacate one of the semaphores.
+If none of the conditions hold, \texttt{V\_one()} vacates \texttt{mutex}.
+
+Let us return now to \texttt{rlock\_acquire}, in the case that the process
+found there is a writer in the critical section, it vacates
+\texttt{mutex} and procures \texttt{rsema},
+blocking on the semaphore associated
+with waiting readers.  When later some other process passes the baton to
+it by incrementing \texttt{rsema}), it updates the state by decrementing
+\texttt{nreadersWaiting} and incrementing \texttt{nreadersEntered}.  It
+then invokes \texttt{V\_one()} one more time.
+
+As an example, consider the case where there is a writer in the critical
+section and there are two readers waiting.  Let us see what happens when
+the writer calls \texttt{wlock\_release}:
+\begin{enumerate}
+\item After procuring \texttt{mutex},
+the sum of the three semaphores is 0.  The writer then decrements
+\texttt{nwritersEntered} and calls \texttt{V\_one()}.
+\item \texttt{V\_one()} finds that there are no writers in the critical section
+and there are two readers waiting.  It therefore vacates \texttt{rsema}.
+The sum of the semaphores is now 1, because \texttt{rsema} is 1.  This
+guarantees that only a waiting reader, procuring \texttt{rsema}, can enter
+the critical section.
+\item When it does, the reader decrements \texttt{nreadersWaiting}
+from 2 to 1, and increments \texttt{nReadersEntered} from 0 to 1.
+The reader finally calls \texttt{V\_one()}.
+\item Again, \texttt{V\_one()} finds that there are no writers and
+that there are readers waiting, so again it vacates \texttt{rsema},
+passing the baton to the one remaining waiting reader.
+\item The remaining reader decrements \texttt{nreadersWaiting} from 1 to 0 and
+increments \texttt{nReadersEntered} from 1 to 2.
+\item Finally, the remaining reader  calls \texttt{V\_one()}.
+\texttt{V\_one()} does not find any processes waiting,
+and so it vacates \texttt{mutex}.
+\end{enumerate}
+
+\begin{quote}
+\begin{itemize}
+\item Implement the producer/consumer problem with SBS.
+\end{itemize}
+\end{quote}
+
 \chapter{Starvation}
+
+\begin{figure}
+\begin{code}
+\VerbatimInput[xleftmargin=5mm,numbers=left]{../RWfair.cxl}
+\end{code}
+\caption{Reader/Writer Lock implementation addressing fairness.}
+\label{fig:RWfair}
+\end{figure}
 

### 2020-06-07 book/paper.tex
Latest

@@
 Unfortunately, such is not the case for the
-reader/writer solution of Figure~\ref{fig:rw2lock}.
+reader/writer solutions that we presented thus far.
 Consider this scenario: there are two readers and one writer.  One reader
@@
 implementation that solves this type of starvation, but won't eliminate
-the type of starvation that comes with spinlocks.
+the type of starvation.
 

### 2020-06-07 book/paper.tex
Made it fit better

@@
 \begin{itemize}
-\item \texttt{nreadersEntered}: the number of readers in the critical section;
-\item \texttt{nwritersEntered}: the number of writers in the critical section;
-\item \texttt{nreadersWaiting}: the number of readers waiting to enter;
-\item \texttt{nwritersWaiting}: the number of writers waiting to enter.
+\item \texttt{r\_entered}: the number of readers in the critical section;
+\item \texttt{w\_entered}: the number of writers in the critical section;
+\item \texttt{r\_waiting}: the number of readers waiting to enter;
+\item \texttt{w\_waiting}: the number of writers waiting to enter.
 \end{itemize}
@@
 To wit, check out \texttt{rlock\_acquire}.  It first checks to see if there
-are any writers.  If so, it increments \texttt{nreadersWaiting}.  If not,
-it increments \texttt{nreadersEntered}.
+are any writers.  If so, it increments \texttt{r\_waiting}.  If not,
+it increments \texttt{r\_entered}.
 
@@
 found there is a writer in the critical section, it vacates
-\texttt{mutex} and procures \texttt{rsema},
+\texttt{mutex} and procures \texttt{r\_sema},
 blocking on the semaphore associated
 with waiting readers.  When later some other process passes the baton to
-it by incrementing \texttt{rsema}), it updates the state by decrementing
-\texttt{nreadersWaiting} and incrementing \texttt{nreadersEntered}.  It
+it by incrementing \texttt{r\_sema}), it updates the state by decrementing
+\texttt{r\_waiting} and incrementing \texttt{r\_entered}.  It
 then invokes \texttt{V\_one()} one more time.
@@
 the sum of the three semaphores is 0.  The writer then decrements
-\texttt{nwritersEntered} and calls \texttt{V\_one()}.
+\texttt{w\_entered} and calls \texttt{V\_one()}.
 \item \texttt{V\_one()} finds that there are no writers in the critical section
-and there are two readers waiting.  It therefore vacates \texttt{rsema}.
-The sum of the semaphores is now 1, because \texttt{rsema} is 1.  This
-guarantees that only a waiting reader, procuring \texttt{rsema}, can enter
+and there are two readers waiting.  It therefore vacates \texttt{r\_sema}.
+The sum of the semaphores is now 1, because \texttt{r\_sema} is 1.  This
+guarantees that only a waiting reader, procuring \texttt{r\_sema}, can enter
 the critical section.
-\item When it does, the reader decrements \texttt{nreadersWaiting}
-from 2 to 1, and increments \texttt{nReadersEntered} from 0 to 1.
+\item When it does, the reader decrements \texttt{r\_waiting}
+from 2 to 1, and increments \texttt{r\_entered} from 0 to 1.
 The reader finally calls \texttt{V\_one()}.
 \item Again, \texttt{V\_one()} finds that there are no writers and
-that there are readers waiting, so again it vacates \texttt{rsema},
+that there are readers waiting, so again it vacates \texttt{r\_sema},
 passing the baton to the one remaining waiting reader.
-\item The remaining reader decrements \texttt{nreadersWaiting} from 1 to 0 and
-increments \texttt{nReadersEntered} from 1 to 2.
+\item The remaining reader decrements \texttt{r\_waiting} from 1 to 0 and
+increments \texttt{r\_entered} from 1 to 2.
 \item Finally, the remaining reader  calls \texttt{V\_one()}.
@@
 
+\chapter{Monitors}
+
 \end{document}

### 2020-06-07 book/paper.tex
Latest

@@
 the writer never had a chance.
+
 In this chapter, we will present a version of a reader/writer lock
@@
 the type of starvation.
+In particular, the probability that either a reader or writer gets starved
+will be exponentially diminishing.
+We shall call such a solution \emph{fair} (although it does not quite
+match the usual formal nor vernacular concepts of fairness).
+
+It is difficult if not impossible to make the reader/writer lock solution of
+Figure~\ref{fig:rw2lock} fair.  However, the split binary semaphore solution of
+Figure~\ref{fig:RWsplitsema} provides much control over whom to pass baton
+to.
+Figure~\ref{fig:RWfair} contains a version of Figure~\ref{fig:RWsplitsema}
+that is a fair solution to the reader/writer problem.  There are two changes
+compared to the original version:
+
+\begin{enumerate}
+\item When a reader tries to enter the critical section, it yields not only
+if there are writers in the critical section, but also if there are writers
+waiting to enter the critical section;
+\item There are two versions of \texttt{V\_one()}.  In particular, when the
+last reader leaves the critical section, it preferentially passes the baton
+to a waiting writer rather than to a waiting reader.  Similarly,
+when a writer leaves the critical section, it preferentially passes
+the baton to a waiting reader rather than to a waiting writer.
+\end{enumerate}
+
+The net effect of this is that if there is contention between readers and
+writers, than readers and writers end up alternating entering the critical
+section.  While readers can still starve other readers and writers can still
+starve other writers, they can not starve one another.  And contention for
+a semaphore is resolved probabilistically, each reader and each writer will
+eventually have a chance to enter the critical section almost surely.
 
@@
 
+Tony Hoare, who came up with the concept of split binary semaphores, devised
+an abstraction of the concept in a programming language paradigm called
+\emph{monitors}.
+
 \end{document}

### 2020-06-07 book/paper.tex
Starting on monitors

@@
 While correct, it is not considered a good solution.
-The solution is an example of what is called \emph{busy-waiting}:
+The solution is an example of what is called \emph{busy-waiting}
+(aka spin-waiting):
 processes spin in a loop until some desirable application-level condition is met.
@@
 \emph{monitors}.
+A monitor is a special version of an object-oriented \emph{class}, comprising
+a set of variables and methods that operate on those variabled.
+There is a split binary semaphore associated with each such class.
+The \texttt{mutex} is hidden: it is automatically acquired when invoking a
+method and released upon exit.
+The other semaphores are called \emph{condition variables}.
+There are two operations on condition variables: \texttt{wait} and \texttt{signal}.
+If \texttt{c} is a pointer to a condition variable, then the semantics of the
+operations, in CXL, are as follows:
+
+\begin{code} 
+\begin{verbatim} 
+def wait(c):
+    call V &(mutex);
+    call P c;
+;
+def signal(c):
+    call V c;
+    call P &(mutex);
+;
+\end{verbatim} 
+\end{code}
+
+\texttt{wait} releases the monitor mutex and blocks while trying
+to procure the condition variable (which is a semaphore that is 0 when
+\texttt{wait} is invoked).
+\texttt{signal} vacates the condition variable, passing the baton to
+a process trying to procure it, and then procures the monitor mutex.
+The use of \texttt{wait} and \texttt{signal} ensures that \texttt{P} and
+\texttt{V} operations alternate as prescribed by the SBS technique.
+Our implementations of reader/writer locks can be easily changed to
+use \texttt{wait} and \texttt{signal} instead of using \texttt{P}
+\texttt{V}.
+
+In the late 70s, Xerox PARC developed a new programming language called
+Mesa.  Mesa introduced various important concepts to programming languages,
+including software exceptions and incremental compilation.  Mesa also
+incorporated a version of monitors.
+However, there are some subtle but important differences with Hoare-style
+monitors that make Mesa-style monitors quite unlike split binary semaphores.
 

### 2020-06-07 book/paper.tex
Mesa monitors

@@
 
-\texttt{wait} releases the monitor mutex and blocks while trying
+\texttt{wait} releases the monitor lock and blocks while trying
 to procure the condition variable (which is a semaphore that is 0 when
@@
 \texttt{signal} vacates the condition variable, passing the baton to
-a process trying to procure it, and then procures the monitor mutex.
+a process trying to procure it, and then procures the monitor lock.
 The use of \texttt{wait} and \texttt{signal} ensures that \texttt{P} and
@@
 In the late 70s, Xerox PARC developed a new programming language called
-Mesa.  Mesa introduced various important concepts to programming languages,
+Mesa~\cite{mesa}.
+Mesa introduced various important concepts to programming languages,
 including software exceptions and incremental compilation.  Mesa also
 incorporated a version of monitors.
-However, there are some subtle but important differences with Hoare-style
-monitors that make Mesa-style monitors quite unlike split binary semaphores.
+However, there are some subtle but important differences with Hoare
+monitors that make Mesa monitors quite unlike split binary semaphores.
+
+As in Hoare monitors, there is a lock associated with each monitor,
+and the lock is acquired upon entry to a method and released upon exit.
+Mesa monitors also have condition variable that a process can wait on.
+Like in Hoare monitors, the \texttt{wait} operation release the monitor
+lock.
+The most important difference is in what \texttt{signal} does.
+To make the distinction more clear, we shall call corresponding Mesa
+operation \texttt{notify} rather than \texttt{signal}.
+When a process $p$ invokes \texttt{notify}, it does not immediately pass 
+he baton to some process $q$ requiring the baton.  Instead, $p$ holds onto
+the baton until it leaves the monitor.  At that point $q$ gets the baton,
+assuming there even was such a process $q$.
+
+\begin{figure}
+\begin{center}
+\includegraphics[width=6in]{figures/monitor.jpeg}
+\end{center}
+\caption{High-level depictions of Hoare and Mesa monitors.  The hall is
+where processes wait to enter the bathroom (the critical section).  The
+bedrooms illustrate two condition variables.  The circles are processes.}
+\label{fig:pc}
+\end{figure}
+
+Figure~\ref{fig:bathroom} illustrates the difference with a drawing.
+Here a bathroom represents the critical section, allowing only one
+process at a time say.  The bedrooms represent condition variables.
+There is some condition associated with each bedroom.
+In the hall are processes waiting to enter the critical section.
+
+On the left is a Hoare monitor.  When a process leaves the monitor
+(let's call that process $p$), it must be in the bathroom.
+It can choose to either let a process
+in one of the bedrooms or a process in the hall, assuming there are any.
+To make the choice, $p$ checks for each bedroom if its condition holds
+and if there are processes in the bedroom.
+If there are such bedrooms, $p$ selects one and let's that process into
+the bathroom.  Let's call that process $q$.  Importantly, when $q$ enters
+the bathroom, the condition that $q$ was waiting for still holds.
+
+On the right is a Mesa monitor.  The only way into the bathroom is through
+the hall.  When process $p$ leaves the bathroom,
+if also checks each of the bedrooms to see if there is a process $q$ that
+can run.  But instead of letting $q$ go straight into the bathroom, $q$
+goes into the hall, joining any other processes that are also waiting to
+enter the bathroom.  Finally, when $p$ has left the bathroom, one of the
+processes in the hall is let into the bathroom.  But that may not be $q$.
+Assuming every process eventually leaves the bathroom and the system is
+fair, eventually $q$ will enter the bathroom.  However, it is no longer
+guaranteed that its condition holds.  Therefore, the first thing $q$ must
+do is check to see if the condition holds.  If not, it should go back
+into the bedroom corresponding to the condition.
+
+It turns out that with the Mesa monitor there is another important option:
+when process $p$ leaves the bathroom, it can choose to let any number of
+processes in the bedrooms into the hall.  It can do so by calling
+\texttt{notify} repeatedly, each time letting one process in one of the
+bedrooms into the hall.  Note that called \texttt{notify} on an empty
+bedroom is considered a no-op in Mesa.  Mesa also has a function called
+\texttt{notifyAll} (aka \texttt{broadcast)}) that, when applied to
+a bedroom, let's all processes in that bedroom into the hall.
+
+While Hoare monitors are conceptually simpler, the so-called
+``Mesa monitor semantics'' have become more popular, adopted by all major
+programming languages including C, Java, and Python.
+
+\chapter{Mesa Condition Variables in CXL}
+
+\begin{figure}
+\begin{code}
+\begin{verbatim}
+def Condition(lk):
+    result = dict{ .lock: lk, .waiters: {} };
+;
+def wait(c):
+    let lk = (^c).lock, blocked = True:
+        (^c).waiters = (^c).waiters + { nametag() };
+        ^lk = False;
+        while blocked:
+            atomic:
+                if (not (^lk)) and (not ((nametag()) in (^c).waiters)):
+                    ^lk = True;
+                    blocked = False;
+                ;
+            ;
+        ;
+    ;
+;
+def notify(c):
+    let waiters = (^c).waiters:
+        if waiters != {}:
+            (^c).waiters = waiters - { choose(waiters) };
+        ;
+    ;
+;
+def notifyAll(c):
+    (^c).waiters = {};
+;
+\end{verbatim}
+\end{code}
+\caption{Implementation of condition variables in the \texttt{synch} module.}
+\label{fig:cv}
+\end{figure}
 

### 2020-06-08 book/paper.tex
Fixed reference
Latest

@@
 \begin{code} 
-\begin{verbatim} 
+\begin{Verbatim}[xleftmargin=5mm,numbers=left]
 def wait(c):
@@
 ;
-\end{verbatim} 
+\end{Verbatim} 
 \end{code}
@@
 bedrooms illustrate two condition variables.  The circles are processes.}
-\label{fig:pc}
-\end{figure}
-
-Figure~\ref{fig:bathroom} illustrates the difference with a drawing.
+\label{fig:monitors}
+\end{figure}
+
+Figure~\ref{fig:monitors} illustrates the difference with a drawing.
 Here a bathroom represents the critical section, allowing only one
@@
 \begin{code}
-\begin{verbatim}
+\begin{Verbatim}[xleftmargin=5mm,numbers=left]
+import bag;
+
 def Condition(lk):
-    result = dict{ .lock: lk, .waiters: {} };
+    result = dict{ .lock: lk, .waiters: bagEmpty() };
 ;
 def wait(c):
-    let lk = (^c).lock, blocked = True:
-        (^c).waiters = (^c).waiters + { nametag() };
+    let lk = (^c).lock, blocked = True, cnt = bagCount((^c).waiters, nametag()):
+        call bagAdd(&((^c).waiters), nametag());
         ^lk = False;
@@
             atomic:
-                if (not (^lk)) and (not ((nametag()) in (^c).waiters)):
+                if (not (^lk)) and (bagCount((^c).waiters, nametag()) <= cnt):
                     ^lk = True;
@@
 def notify(c):
+    atomic:
         let waiters = (^c).waiters:
-        if waiters != {}:
-            (^c).waiters = waiters - { choose(waiters) };
+            if waiters != bagEmpty():
+                call bagRemove(&((^c).waiters), bagChoose(waiters));
+            ;
         ;
@@
 def notifyAll(c):
-    (^c).waiters = {};
-;
-\end{verbatim}
+    (^c).waiters = bagEmpty();
+;
+\end{Verbatim}
 \end{code}
@@
 
+\begin{figure}
+\begin{code}
+\VerbatimInput[xleftmargin=5mm,numbers=left]{../RWcv.cxl}
+\end{code}
+\caption{Unfair Reader/Writer Lock implemtation using Mesa-style condition variables.}
+\label{fig:RWcv}
+\end{figure}
+
 \end{document}

### 2020-06-09 book/paper.tex
Started implementation of continuations

@@
 def notify(c):
-    atomic:
     let waiters = (^c).waiters:
@@
 ;
-;
 def notifyAll(c):

### 2020-06-14 book/paper.tex
Latest

@@
 synchronize with one another makes programs more complicated than
-programs where only one thing happens at a time.  Why is this?
+programs where only one thing happens at a time.  Why are concurrent
+programs more complicated than sequential ones?
 There are, at least, two reasons:
@@
 \item The execution of a sequential program is mostly \emph{deterministic}.
-If you run it twice with the same input, the same things will happen.
-Bugs are typically easily reproducible.
-\item Each statement and each function can be thought of as happening
-\emph{atomically} because there is no other activity interfering with
-their execution.
+If you run it twice with the same input, the same output will be produced.
+Bugs are typically easily reproducible and easy to track down, for example
+by instrumenting the program.
+The output of running concurrent programs depends on how the various
+processes are interleaved.  Some bugs may occur only occasionally and
+may never occur when the program is instrumented to find them
+(so-called \emph{Heisenbugs}).
+\item In a sequential program, each statement and each function can be
+thought of as happening \emph{atomically} because there is no other activity
+interfering with their execution.  Even though a statement or function may
+be compiled into multiple machine instructions, they are executed back-to-back
+until completion.  Not so with a concurrent program, where other instructions
+may update memory locations as a statement or function is being executed.
 \end{itemize}
@@
 
-So why is this useful?  Consider, for example, a sorting algorithm.
+So if model checking does not prove a program correct, why is it
+useful?  To answer that question consider, for example, a sorting
+algorithm.
 Now suppose we create a test program, a model, that tries sorting
@@
 are bugs remaining in the sorting algorithm is small.
-
-But model checking can also help with proving an algorithm correct.
+Note, however, that it would be easy to write a program
+that sorts all lists of up to five numbers correctly but fails to
+do so for a list of 6 numbers.  (Hint: simply use an \texttt{if}
+statement.)
+
+While model checking does not in general prove an algorithm correct,
+it can help with proving an algorithm correct.
 The reason is that the correctness of many programs is built around
@@
 execution of a program.  A model checker can find violations of
-those invariants when evaluating a model and provide valuable early
+proposed invariants when evaluating a model and provide valuable early
 feedback to somebody who is trying to construct a proof, even an
@@
 are executed.  For example, we assume that the reader is familiar
-with the concepts of CPU, memory, register, and stack.
+with the concepts of CPU, memory, register, stack, and machine
+instructions.
 
@@
 \begin{itemize}
-\item Every statement in CXL must be terminated by a semicolon
-(even \texttt{while} statements and \texttt{def} statements);
-in Python semicolons are optional and rarely used;
-\item Correct indentation in CXL is encouraged but optional;
+\item Syntax: Every statement in CXL must be terminated by a semicolon
+(even \texttt{while} statements and \texttt{def} statements).
+In Python semicolons are optional and rarely used.
+There is no syntactic significance to indentation in CXL.
+Correct indentation in CXL is encouraged but not enforced.
+CXL only supports basic operator precedence or associativity.
+Use parentheses liberally to remove ambiguity.
+\item CXL does not (currently) support floating point.  CXL does not support I/O;
 \item Python is object-oriented; CXL is not;
-\item CXL does not (currently) support floating point;
-\item CXL does not support I/O of any kind;
-\item CXL only supports basic operator precedence or associativity.
-Use parentheses liberally to remove ambiguity.
 \end{itemize}
@@
 the most interesting.
-They assign to \texttt{x} some unspecified value in the range \texttt{0..N}
-and then verifies that $\mathtt{triangle}(x)$ equals $x(x+1)/2$.
+The first assigns to \texttt{x} some unspecified value in the range \texttt{0..N}
+and the second verifies that $\mathtt{triangle}(x)$ equals $x(x+1)/2$.
 
@@
 (You may assume that \texttt{load} and \texttt{store} instructions of the
-underlying architecture are atomic---in fact they are.)
+underlying machine architecture are atomic---in fact they are.)
 \end{itemize}
@@
 stores $1$ into \texttt{count};
-\item incrementer 0 sets \texttt{upDone} to \texttt{True};
-\item incrementer 1 sets \texttt{upDone} to \texttt{True}.
+\item incrementer 0 sets \texttt{done[0]} to \texttt{True};
+\item incrementer 1 sets \texttt{done[1]} to \texttt{True}.
 \end{enumerate}
@@
 
-CXL makes extensive use of dictionaries.  Their syntax and properties
-are a little different from Python.
-A dictionary maps CXL values, known as \emph{keys}, to CXL values.
+CXL makes extensive use of dictionaries.
+A dictionary maps values, known as \emph{keys}, to values.
+The CXL dictionary syntax and properties are a little different from Python.
 Unlike Python, any CXL value can be a key, including another
@@
 This is unfamiliar to Python programers, but in CXL square brackets can be used
-in the same way as parentheses, so you can express it in the more familiar form:
+in the same way as parentheses, so you can express the same thing in the form
+that is familiar to Python programmers:
 \begin{code}
@@
 \texttt{d[.count]} (although both will work).
+Thus using atoms, a dictionary can be made to look much like a Python object.
 The meaning of \texttt{d a b c ...} is \texttt{((((d a) b) c) ...)}.
@@
 The CXL machine is not an ordinary virtual machine, but its architecture
-is nonetheless representative of conventional machines such as hardware
-with memory and CPUs and virtual machines such as the Java Virtual Machine.
+is nonetheless representative of conventional computers and
+virtual machines such as the Java Virtual Machine.
 
@@
 be in the same state.
-The initial state of the CXL memory is the empty dictionary, \texttt{()}.
+The initial state of the CXL memory is the empty dictionary, \texttt{dict\{\}}.
 The context bag has an initial context in it with name tag
@@
 directories can be organized into a tree.
-Each node in the tree is then identified
+Each node in a directory tree is then identified
 by a sequence of atoms, like a path name in the file system hierarchy.  We call
 such a sequence the \emph{address} of a CXL value, and it is relative to the
-``root'' of the CXL value.
+``root'' of the directory tree.
 
@@
 Most instructions pop values from the stack or push values onto the stack.
-Initially there is one process (with name tag \texttt{\_\_init\_\_/()}
-that starts executing at instruction 0 and keeps executing until the last
-execution in the program to initialize the state.
+At first there is one process (with name tag \texttt{\_\_init\_\_/()}
+that initializes the state.
+It starts executing at instruction 0 and keeps executing until the last
+execution in the program.
 In this case, the first instruction is a \texttt{JUMP} instruction that sets the
@@
 
-The \texttt{Frame} instruction has three arguments:
-\begin{enumerate}
-\item The name of the method;
-\item A list containing the names of the arguments of the method;
-\item The program counter of the method's \texttt{Return} instruction.
-\end{enumerate}
-
+The \texttt{Frame} instruction has three arguments: the name of the method,
+a list containing the names of the arguments of the method, and
+the program counter of the method's \texttt{Return} instruction.
 The code generated from \texttt{count := count + 1} in line 2 of
@@
 machine instructions.  A counter is a very simple example of
-a data structure.  A more involved one would be access to a binary tree.
+a data structure, but as we have seen it too requires multiple instructions.
+A more involved one would be access to a binary tree.
 Adding a node to a binary tree, or re-balancing a tree, often requires
@@
 Figure~\ref{fig:csbarebones} shows the CXL code.
-Here \texttt{@ncs} and \texttt{@cs} are \emph{labels}, each identifying
+Here \texttt{@cs} is a \emph{label}, identifying
 a location in the CXL machine code.  The first thing we need to
@@
 We would like to place an assertion at the \texttt{@cs} label that
-specifies that only the current process can be there.  CXL in fact
-supports this.  It has an operator \texttt{atLabel $L$}, where $L$
-is the atom of the label (in this case, \texttt{.cs}).  The
-operator returns a bag of name tags of processes executing at that
+specifies that only the current process can be there.
+
+CXL in fact supports this.
+It has an operator \texttt{atLabel $L$}, where $L$
+is the atom containing the name of the label (in this case, \texttt{.cs}).
+The operator returns a bag (multiset) of name tags of processes executing at that
 label.  The bag is represented by a dictionary that maps each element
@@
 use it in normal code.
-
 The assertion also makes use of the \texttt{nametag()} operator that
@@
 
-\begin{figure}
-\begin{center}
-\includegraphics[width=4in]{figures/mutex.jpeg}
-\end{center}
-\caption{High-level state diagram specification of mutual exclusion with up to two processes.
-The first number in a state gives the number of processes; the second number is the
-number of processes in the critical section.}
-\label{fig:mutex}
-\end{figure}
-
 In order to detect violations of progress, and other liveness problems in
@@
 without entering the critical section ever.
-Note also that we have dropped the \texttt{@ncs} label: it would serve no
-specific purpose.
-Figure~\ref{fig:mutex} show a high-level state diagram of the code in
-Figure~\ref{fig:cs}.
 
@@
 
+\begin{figure}
+\begin{center}
+\includegraphics[width=4in]{figures/mutex.jpeg}
+\end{center}
+\caption{High-level state diagram specification of mutual exclusion with up to two processes.
+The first number in a state gives the number of processes; the second number is the
+number of processes in the critical section.}
+\label{fig:mutex}
+\end{figure}
+
+Figure~\ref{fig:mutex} show a high-level state diagram of the code in
+Figure~\ref{fig:cs}.
+The circles represent states, while the arrows represent possible state
+transitions.  The label in each circle summarizes the state;  the label
+on an arrow describes the transition.  In this case,
+the label contains two numbers: the total number of processes and the number
+of processes in the critical section.  A process that is in the critical
+section cannot terminate.
+
 \chapter{Na\"{\i}ve Attempts at Implementing Critical Sections}
@@
 
-In order to have at most one process in a critical section at a time, you
-probably are thinking that one might use a \emph{lock}.
-That is a good thought, but how does one implement one?
-Figure~\ref{fig:uplock} presents a mutual exclusion attempt based on a
-na\"{\i}ve implementation of a lock.
+You may already have heard of the concept of a \emph{lock} and have realized that
+it could be used to implement a critical section.
 The idea is that the lock is like a baton that at most one process can own
 (or hold) at a time.
+A process that wants to enter the critical section at a time must obtain the
+lock first and release it upon exiting the critical section.
+Using a lock is a good thought, but how does one implement one?
+Figure~\ref{fig:uplock} presents a mutual exclusion attempt based on a
+na\"{\i}ve (and, as it turns out, broken) implementation of a lock.
 Initially the lock is not owned, indicated by lock being \texttt{False}.
@@
 
-Unfortunately, if we run the program, we find that the assertion still
+Unfortunately, if we run the program through CXL, we find that the assertion still
 fails.  Diagnosing the problem, we see that the \texttt{lock} variable
@@
 To see why, first note the invariant that if process $i$ is in the
-critical section, then flags[i] == True.
+critical section, then \texttt{flags[$i$] == True}.
 Without loss of generality,
@@
 
-However, the solution has a different problem called \emph{livelock}:
-if both try to enter the critical section at the same time, they may
+Run the program through CXL.  Unfortunately, it turns out the solution has a
+problem: if both try to enter the critical section at the same time, they may
 end up waiting for one another indefinitely.  Thus the solution
@@
 processes 0 and 1 alternate entering the critical section.
-The problem with the solution is
-that it also violates the \emph{progress} property:
+
+Run the program through CXL.  It turns out that this solution also violates
+\emph{progress}, albeit for a different reason:
 if process~$i$ terminates instead of entering the critical section when it
@@
 turn.
+
+\begin{quote}
+Try your own solutions!
+\begin{itemize}
+\item See if you can come up with some different approaches that satisfy both
+mutual exclusion and progress.  Try them with CXL and see if they work or not.
+If they don't, try to understand why.
+Do not despair if you can't figure it out---as
+we will find out, it is possible but not easy.
+\end{itemize}
+\end{quote}
 
@@
 until either the other process is nowhere near the critical section
-or has given way.
+(\texttt{flag[1 - self] == False}) or or has given way
+(\texttt{turn == self}).
 Running the algorithm with CXL shows that it satisfies both mutual
 exclusion and progress.
+
+\begin{figure}
+\begin{center}
+\includegraphics[width=4in]{figures/states.jpeg}
+\end{center}
+\caption{Venn diagram of all states and a trace.}
+\label{fig:states}
+\end{figure}
 
@@
 at any point of time the CXL machine is in a specific state.
+A state is comprised of the values of the shared variables as well as
+the values of the process variables for each process, including its
+program counter and the contents of its stack.
 Everytime a process executes a CXL machine instruction, the
-state changes (if only because the program counter of the process is
-updated).  We call that a \emph{step}.  Steps in CXL are atomic.
+state changes (if only because the program counter of the process
+changes).  We call that a \emph{step}.  Steps in CXL are atomic.
 
@@
 one process and its program counter is~0.  A \emph{trace} is a
-sequence of steps starting from the initial state.  When making a
-step, there are two sources of non-determinism in CXL.  One is when
-in some state more than one process can make a step.  The other is
+sequence of steps starting from the initial state.
+When making a step, there are two sources of non-determinism in CXL.
+One is when
+there is more than one process that can make a step.  The other is
 when a process executes a \texttt{choose} operation and there is
 more than one choice.
-
-A good way to prove mutual exclusion is through induction on
-the number of steps.
-In such a proof, we need a so-called \emph{inductive invariant} $I$ that
-implies mutual exclusion.
-An inductive invariant $I$ is a predicate over state that satisfies the following:
+Because there is non-determinism, there are multiple possible traces.
+We call a state \emph{reachable} if it is either the initial state
+or it can be reached from the initial state through a trace.
+A state is final
+when there are no processes left to make state changes.
+
+A \emph{property} describes a subset of
+states.  In other words, a property is equivalent to a subset of the states.
+\emph{Initial}, \emph{final}, and \emph{reachable}, and \emph{unreachable}
+are all examples of properties.
+Figure~\ref{fig:states} depicts a Venn diagram of various properties
+and a trace.
+An \emph{invariant} is a property that holds over all reachable states.
+We want to show that mutual exclusion is an invariant.
+In other words, we want to show that the set of reachable states is a subset
+of the set of states where there is at most one process in the critical
+section.
+
+One way to prove that a property is an invariant is through induction
+on the number of steps.  First you prove that the property holds over
+the initial state.  Then you prove that for every reachable state,
+and for every step from that reachable state, the property also holds
+over the resulting state.
+For this you need a predicate that describes exactly which
+states are reachable.
+But we do not have such a predicate: we know how to describe the set
+of reachable states, but given an arbitrary state it is not easy to
+see whether it is reachable or not.
+
+To solve this problem, we will use what is called an
+\emph{inductive invariant}.
+An inductive invariant $I$ is a predicate over states that satisfies the following:
 \begin{itemize}
-\item $I$ hold in the initial state.
+\item $I$ holds in the initial state.
 \item For any state in which $I$ holds and any process in the
@@
 \end{itemize}
+Unlike an invariant, an inductive invariant must also hold over unreachable states.
 
@@
 
+So we are looking for an inductive invariant that \emph{implies} mutual
+exclusion.  In other words, the set of states where the inducive invariant holds
+must be a subset of the set of states where there is at most one process in
+the critical section.
 Let $C(i) = \mathtt{flags}[1 - i] \land

### 2020-06-14 book/paper.tex
Latest
Added the -m flag

@@
 
+$I_1$ is an inductive invariant.
 To see that invariant $I_1$ is maintained, note that
@@
 ;
+def Lock():
+    result = False;
+;
 def lock(lk):
@@
 ;
-def Lock():
-    result = False;
-;
 \end{verbatim}
@@
 \caption{The \texttt{Lock} interface in the \texttt{synch} module.}
-\label{fig:locks}
+\label{fig:spinlocks}
 \end{figure}
@@
 \texttt{synch} that includes support for locks.
-Figure~\ref{fig:locks} shows how they are implemented, and
+Figure~\ref{fig:spinlocks} shows how they are implemented, and
 Figure~\ref{fig:incfixed} gives an example of how they may be used,
@@
 chapters.
+
+We call a process \emph{blocked} in a state if the process is in
+an \emph{infinite loop} in which the process does not update the
+shared memory.
+A process trying to
+acquire a test-and-set spinlock held by another process is a good example
+of a process being blocked: the process only reads the shared memory
+but does not change it.
+The only way to break out of the infinite loop is if another process
+changes the shared memory.
+The \texttt{while} loop in Peterson's algorithm, in case its condition
+does not hold, is another example of a process being blocked.
+
+\begin{figure}
+\begin{code}
+\begin{verbatim}
+import list;
+
+def Lock():
+    result = dict{ .locked: False, .suspended: [] };
+;
+def lock(lk):
+    atomic:
+        if (^lk).locked:
+            call stop (^lk).suspended;
+            assert (^lk).locked;
+        else:
+            (^lk).locked = True;
+        ;
+    ;
+;
+def unlock(lk):
+    atomic:
+        if (^lk).suspended == []:
+            (^lk).locked = False;
+        else:
+            go (head((^lk).suspended)) ();
+            (^lk).suspended = tail((^lk).suspended);
+        ;
+    ;
+;
+\end{verbatim}
+\end{code}
+\caption{The \texttt{Lock} interface using suspension.}
+\label{fig:suspension}
+\end{figure}
+
+In most operating systems, processes are virtual and can be suspended
+until some condition changes.  For example, a process can be suspended
+until the lock is available to it.
+In CXL, a process can suspend itself and save its context in a
+list.  Recall that the context of a process consists of its name tag,
+its program counter, and the contents of its register and stack.
+A context is a regular CXL value.
+The syntax of the expression is as follows:
+
+\begin{code}
+\begin{verbatim}
+stop L
+\end{verbatim}
+\end{code}
+
+Here \texttt{L} is a shared list.
+Another process can revive the process using the \texttt{go} stateent:
+
+\begin{code}
+\begin{verbatim}
+go C R
+\end{verbatim}
+\end{code}
+
+Here \texttt{C} is a context and \texttt{R} is a CXL value.
+It causes a process with context \texttt{C} to be added to the state that has
+just executed the \texttt{stop} expression.  The \texttt{stop} expression
+returns the value \texttt{R}.
+
+There is a second version of the \texttt{synch} module that uses suspension
+instead of spinlocks.
+Figure~\ref{fig:suspension} shows the same \texttt{Lock} interface implemented
+using suspension.
+A \texttt{Lock} maintains both a boolean indicating whether the lock is taken,
+and a list of contexts of processes that want to acquire the lock.
+\texttt{lock} sets the lock if available, or suspends itself if not.
+Note that \texttt{stop} is called within an \texttt{atomic} statement---this is
+the only exception to an atomic statement running to completion.  While the
+process is running no other processes can run, but when it suspends itself
+it allows other processes to run.
+
+\texttt{unlock} checks to see if any processes are waiting to get the lock.
+If so, it uses the \texttt{head} and \texttt{tail}
+methods from the \texttt{list} module to resume the first process that got
+suspend and to remove its context from the list.
+Selecting the first process is a design choice.  Another implementation could
+have picked the last one, and yet another implementation could have used
+\texttt{choose} to pick an arbitrary one.  Selecting the first is a common
+choice in lock implementations as it prevents \emph{starvation}: every process
+gets a chance to obtain the lock (assuming every process eventually releases
+it).  Chapter~\ref{ch:starvation} will talk more about starvation.
+
+CXL allows you to select which version of the \texttt{synch} module you would
+like to use with the \texttt{-m} flag.  For example,
+
+\begin{code}
+\begin{verbatim}
+cxl -m synch=synchS x.cxl
+\end{verbatim}
+\end{code}
+
+runs the file \texttt{x.cxl} using the Suspension version of the \texttt{synch} module.
+You will find that using the \texttt{synchS} module often leads to searching a
+significantly larger search space than using the \texttt{synch} module.
+Part of the reason is that the \texttt{synchS} module keeps track of the order
+in which processes wait for a lock.
 
@@
 of \texttt{lock()} spins in a loop until the lock is available
-(see Figure~\ref{fig:locks}).
+(see Figure~\ref{fig:spinlocks}).
 In most operating systems and programming language runtimes, however,
@@
 
-We call a process \emph{blocked} in a state if steps by that process
-do not change the shared memory in that state.  A process trying to
-acquire a test-and-set spinlock held by another process is a good example
-of a process being blocked: the process only reads the shared memory
-but does not change it.
-
-However, a process trying to obtain a read or write lock in
-Figure~\ref{fig:rwbusy} has to actually obtain a spinlock just to check
+A process trying to obtain a read or write lock in
+Figure~\ref{fig:rwbusy} has to obtain a spinlock just to check
 if the read or write lock is available.  So even when the read or write
@@
 \chapter{Starvation}
+\label{ch:starvation}
 

### 2020-06-14 cxl.py

@@ -6,8 +6,9 @@ import collections
 import time
 
 # TODO.  This should not be global ideally
-files = {}
-constants = {}
+files = {}              # files that have been read already
+constants = {}          # constants modified with -c
+modules = {}            # modules modified with -m
 
 def load(f, filename, scope, code):
     if filename in files:
@@ -1973,6 +1974,9 @@ class ImportAST(AST):
 
     def compile(self, scope, code):
         (lexeme, file, line, column) = self.module
+        print("IMPORT", lexeme, modules)
+        if lexeme in modules:
+            lexeme = modules[lexeme]
         filename = lexeme + ".cxl"
         with open(filename) as f:
             load(f, filename, scope, code)
@@ -2726,14 +2730,24 @@ def parseConstant(c, v):
         code[ctx.pc].eval(state, ctx)
     constants[c] = ctx.pop()
 
-def doCompile(filenames, consts):
+def doCompile(filenames, consts, mods):
     for c in consts:
         try:
             i = c.index("=")
             parseConstant(c[0:i], c[i+1:])
         except IndexError:
-            print("Usage: -cC=V to define a constant")
+            print("Usage: -c C=V to define a constant")
             sys.exit(1)
+
+    global modules
+    for m in mods:
+        try:
+            i = m.index("=")
+            modules[m[0:i]] = m[i+1:]
+        except IndexError:
+            print("Usage: -m module=version to specify a module version")
+            sys.exit(1)
+
     scope = Scope(None)
     code = []
     if filenames == []:
@@ -2946,8 +2960,9 @@ def usage():
     print("  options: ")
     print("    -a: list machine code")
     print("    -b: blocking execution")
-    print("    -cname=value: define a constant")
+    print("    -c name=value: define a constant")
     print("    -h: help")
+    print("    -m module=version: select a module version")
     print("    -s: list all states")
     exit(1)
 
@@ -2957,12 +2972,13 @@ def main():
 
     # Get options.  First set default values
     consts = []
+    mods = []
     printCode = False
     blockflag = False
     printStates = False
     try:
         opts, args = getopt.getopt(sys.argv[1:],
-                        "abc:hs", ["const=", "help"])
+                        "abc:hm:s", ["const=", "help", "module="])
     except getopt.GetoptError as err:
         print(str(err))
         usage()
@@ -2973,6 +2989,8 @@ def main():
             blockflag = True
         elif o in { "-c", "--const" }:
[…]

### 2020-06-14 book/paper.tex
Latest version

@@
 a \emph{reader/writer lock}~\cite{courtois71}.
-[RVR: or was it David Parnas who first introduced them?]
 We will explore various ways of implementing reader/writer locks in
@@
 of readers and one that maintains the number of writers.
-The lock is used to protect access to those counters.
+We will call the lock the \emph{mutex} to distinguish it clearly from
+the reader/writer lock that we are implementing.
+The mutex is used to protect shared access to the counters.
 The program shows a process that executes in a loop.
@@
 processes spin in a loop until some desirable application-level condition is met.
-
-The astute reader might wonder if obtaining an ordinary lock itself is an
+The astute reader might wonder if obtaining the mutex itself is an
 example of busy-waiting.  After all, the CXL \texttt{synch} implementation
-of \texttt{lock()} spins in a loop until the lock is available
+of \texttt{lock()} spins in a loop until the mutex is available
 (see Figure~\ref{fig:spinlocks}).
-In most operating systems and programming language runtimes, however,
+
+However, there is a big difference.  As we pointed out, the processes that
+are waiting for a spinlock are \emph{blocked}: they are in an infinite loop
+in which they do not change the state.
+The processes waiting for a reader/writer lock do change the state while
+in the infinite loop: they acquire the mutex and release it.
+
+In most operating systems and programming language runtimes,
 when a process acquires a lock, the process is placed on a scheduling
 queue and stops using CPU cycles until the lock becomes available.
-Moreover, if there are multiple processes waiting for a lock, there
-is a scheduler that decides which of the processes gets the lock.
-Busy waiting disables all this.
-
-But even when multiple processes each run on their own core, the solution
-is inefficient.  After all, when the lock becomes available, the process
-has to check other variables to see if it can continue.  In our case,
-the process may have to read \texttt{nreaders} and/or \texttt{nwriters}
-to decide if it can go on or has to try again, involving releasing
-the lock and re-acquiring it.
-
+CXL supports this with the \texttt{synchS} module.
+Busy waiting disables all this: even when using the \texttt{synchS}
+module, readers and writers only get suspended temporarily in case
+there is contention for the mutex.
 A process trying to obtain a read or write lock in
-Figure~\ref{fig:rwbusy} has to obtain a spinlock just to check
-if the read or write lock is available.  So even when the read or write
-lock is not available, the process ends up updating the spinlock and
-therefore the shared memory.
+Figure~\ref{fig:rwbusy} has to obtain the mutex repeatedly
+until the read or write lock is available.
 
@@
 that does not busy-wait.
-It uses two locks: \texttt{rwlock} is used by both readers and writers,
+It uses two ordinary locks: \texttt{rwlock} is used by both readers and writers,
 while \texttt{rlock} is only used by readers.

### 2020-06-14 cxl.py

@@ -1974,7 +1974,6 @@ class ImportAST(AST):
 
     def compile(self, scope, code):
         (lexeme, file, line, column) = self.module
-        print("IMPORT", lexeme, modules)
         if lexeme in modules:
             lexeme = modules[lexeme]
         filename = lexeme + ".cxl"

### 2020-06-14 cxl.py
Updates

@@ -2945,7 +2945,7 @@ def run(code, labels, map, step, blockflag, printStates):
         bad_state = find_shortest(visited, bad)
         print_path(visited, bad_state)
         for ctx in bad_state.stopbag.keys():
-            print("  ", ctx)
+            print("stopped process:", ctx)
         issues_found = True
 
     if not issues_found:

### 2020-06-14 book/paper.tex
Updates
Fixes

@@
 \end{code}
-\caption{The \texttt{Lock} interface using suspension.}
+\caption{The \texttt{Lock} interface in the \texttt{synchS} module uses suspension.}
 \label{fig:suspension}
@@
 A monitor is a special version of an object-oriented \emph{class}, comprising
-a set of variables and methods that operate on those variabled.
+a set of variables and methods that operate on those variables.
 There is a split binary semaphore associated with each such class.
@@
 
-\texttt{wait} releases the monitor lock and blocks while trying
+\texttt{wait} releases the mutex and blocks while trying
 to procure the condition variable (which is a semaphore that is 0 when
@@
 \texttt{signal} vacates the condition variable, passing the baton to
-a process trying to procure it, and then procures the monitor lock.
+a process trying to procure it, and then procures the mutex.
 The use of \texttt{wait} and \texttt{signal} ensures that \texttt{P} and
@@
 
-As in Hoare monitors, there is a lock associated with each monitor,
-and the lock is acquired upon entry to a method and released upon exit.
+As in Hoare monitors, there is a hidden mutex associated with each monitor,
+and the mutex is automatically acquired upon entry to a method and released
+upon exit.
 Mesa monitors also have condition variable that a process can wait on.
-Like in Hoare monitors, the \texttt{wait} operation release the monitor
-lock.
+Like in Hoare monitors, the \texttt{wait} operation release the mutex.
 The most important difference is in what \texttt{signal} does.
-To make the distinction more clear, we shall call corresponding Mesa
+To make the distinction more clear, we shall call the corresponding Mesa
 operation \texttt{notify} rather than \texttt{signal}.
 When a process $p$ invokes \texttt{notify}, it does not immediately pass 
-he baton to some process $q$ requiring the baton.  Instead, $p$ holds onto
+the baton to some process $q$ requiring the baton.  Instead, $p$ holds onto
 the baton until it leaves the monitor.  At that point $q$ gets the baton,
@@
 (let's call that process $p$), it must be in the bathroom.
-It can choose to either let a process
+It can choose to either let in a process
 in one of the bedrooms or a process in the hall, assuming there are any.
@@
 and if there are processes in the bedroom.
-If there are such bedrooms, $p$ selects one and let's that process into
+If there are such bedrooms, $p$ selects one and lets that process into
 the bathroom.  Let's call that process $q$.  Importantly, when $q$ enters
@@
 guaranteed that its condition holds.  Therefore, the first thing $q$ must
-do is check to see if the condition holds.  If not, it should go back
+do is check, again, to see if the condition holds.  If not, it should go back
 into the bedroom corresponding to the condition.
 
-It turns out that with the Mesa monitor there is another important option:
+Mesa monitors provide another important option:
 when process $p$ leaves the bathroom, it can choose to let any number of
@@
 \texttt{notify} repeatedly, each time letting one process in one of the
-bedrooms into the hall.  Note that called \texttt{notify} on an empty
-bedroom is considered a no-op in Mesa.  Mesa also has a function called
+bedrooms into the hall.  Using \texttt{notify} on an empty
+bedroom is considered a no-op in Mesa.  (Hoare monitors do not allow
+using \texttt{signal} on an empty bathroom.) Mesa also has a function called
 \texttt{notifyAll} (aka \texttt{broadcast)}) that, when applied to
@@
 
+The \texttt{synch} module supports Mesa condition variables.  However, like C, Java, and Python,
+it does not have built-in language support for them but instead associates condition variables
+with explicit locks.
+Figure~\ref{fig:cv} shows the \texttt{synch} interface to condition variables.
+\texttt{Condition(lk)} creates a new condition variable associated with the given pointer
+to a lock \texttt{lk}.
+The condition variable itself is represented by a dictionary containing the pointer to the lock
+and a bag of name tags of processes waiting on the condition variable.
+(The \texttt{synchS} library instead has a list of contexts.)
+
+\texttt{wait} adds the nametag of the process to the bag.  This increments the number of processes
+in the bag with the same context.  \texttt{wait} then waits until that count is restored to the
+value that it had upon entry to \texttt{wait}.  \texttt{notify} removes an arbitrary context from
+the bag, allowing one of the processes with that context to return from \texttt{wait}.
+\texttt{notifyAll} empties out the entire bag, allowing all processes in the bag to resume.
+
 \end{document}

### 2020-06-14 book/paper.tex
Latest

@@
 \chapter{Monitors}
+\label{ch:monitors}
 
@@
 
+To illustrate how Mesa condition variables are used in practice, we demonstrate, once again,
+using an implementation of reader/writer locks.
+Figure~\ref{fig:RWcv} shows the code.  \texttt{rwlock} is the global lock or mutex for the
+reader/writer lock.  There are two condition variables: readers wait on \texttt{rcond} and
+writers wait on \texttt{wcond}.  The implementation also keeps track of the number of
+readers and writers in the critical section.
+
+When inspecting the code, one important feature to notice is that \texttt{wait} is always
+invoked within a \texttt{while} loop that checks for the condition that the process is
+waiting for.  We explained why in Chapter~\ref{ch:monitors}: when a process resumes after
+being notified, it is not guaranteed that the condition it was waiting for holds, even
+if it did at the time of invoking \texttt{notify} or \texttt{notifyAll}.
+It is \emph{imperative} that you always have a \texttt{while} loop around any invocation
+of \texttt{wait}.
+
+In \texttt{release\_rlock}, notice that \texttt{notify \&(wcond)} is invoked when there
+are no readers left, \emph{without} checking if there are writers waiting to enter.
+With Mesa monitors this is ok, because calling \texttt{notify} is a no-op if no process
+is waiting.
+
+\texttt{release\_wlock} executes \texttt{notifyAll \&(wcond)} as well as
+\texttt{notify \&(wcond)}.
+Again, because we do not keep track of the number of waiting readers or writers, we
+have to conservatively assume that all waiting readers can enter, or, alternatively,
+up to one waiting writer can enter.  So \texttt{release\_wlock} wakes up all
+potential candidates.
+There are two things to note here.  First, unlike split binary semaphores or Hoare
+monitors, where multiple waiting readers would have to be signaled one at a time in a
+baton-passing fashion (see Figure~\ref{fig:splitsema}), with Mesa monitors
+all readers are awaked in one fell swoop using \texttt{notifyAll}.
+Second, both readers and writers are awakened---this is ok because both execute
+\texttt{wait} within a \texttt{while} loop, re-checking the condition that they
+are waiting for.  So if both type of processes are waiting, either all the readers
+get to enter next or one of the writers gets to enter next.
+
 \end{document}

### 2020-06-14 book/paper.tex
Moved book code into code folder

@@
 \begin{code}
-\VerbatimInput[xleftmargin=5mm,numbers=left]{triangle.cxl}
-\end{code}
-\caption{Computing triangle numbers.}
+\VerbatimInput[xleftmargin=5mm,numbers=left]{code/triangle.cxl}
+\end{code}
+\caption{[triangle.cxl] Computing triangle numbers.}
 \label{fig:triangle}
@@
 Figure~\ref{fig:triangle} gives a simple example of a CXL program.
+(The code for examples in this book can be found in the \texttt{code} folder under
+the name listed in the caption of the example.)
 It is sequential and has a method \texttt{square} that takes
@@
 \begin{code}
-\VerbatimInput[xleftmargin=5mm,numbers=left]{Up.cxl}
-\end{code}
-\caption{Incrementing twice in parallel.}
+\VerbatimInput[xleftmargin=5mm,numbers=left]{code/Up.cxl}
+\end{code}
+\caption{[Up.cxl] Incrementing twice in parallel.}
 \label{fig:inc}
@@
 \begin{code}
-\VerbatimInput[xleftmargin=5mm,numbers=left]{csbarebones.cxl}
-\end{code}
-\caption{A barebones critical section with two processes.}
+\VerbatimInput[xleftmargin=5mm,numbers=left]{code/csbarebones.cxl}
+\end{code}
+\caption{[csbarebones.cxl] A barebones critical section with two processes.}
 \label{fig:csbarebones}
@@
 \begin{code}
-\VerbatimInput[xleftmargin=5mm,numbers=left]{cs.cxl}
-\end{code}
-\caption{CXL model of a critical section.}
+\VerbatimInput[xleftmargin=5mm,numbers=left]{code/cs.cxl}
+\end{code}
+\caption{[cx.cxl] CXL model of a critical section.}
 \label{fig:cs}
@@
 \begin{code}
-\VerbatimInput[xleftmargin=5mm,numbers=left]{UpLock.cxl}
-\end{code}
-\caption{Na\"{\i}ve implementation of a shared lock.}
+\VerbatimInput[xleftmargin=5mm,numbers=left]{code/naiveLock.cxl}
+\end{code}
+\caption{[naiveLock.cxl] Na\"{\i}ve implementation of a shared lock.}
 \label{fig:uplock}
@@
 \begin{code}
-\VerbatimInput[xleftmargin=5mm,numbers=left]{UpFlags.cxl}
-\end{code}
-\caption{Na\"{\i}ve use of flags to solve mutual exclusion.}
+\VerbatimInput[xleftmargin=5mm,numbers=left]{code/naiveFlags.cxl}
+\end{code}
+\caption{[naiveFlags.cxl] Na\"{\i}ve use of flags to solve mutual exclusion.}
 \label{fig:upflags}
@@
 \begin{code}
-\VerbatimInput[xleftmargin=5mm,numbers=left]{UpTurn.cxl}
-\end{code}
-\caption{Na\"{\i}ve use of turn variable to solve mutual exclusion.}
+\VerbatimInput[xleftmargin=5mm,numbers=left]{code/naiveTurn.cxl}
+\end{code}
+\caption{[naiveTurn.cxl] Na\"{\i}ve use of turn variable to solve mutual exclusion.}
 \label{fig:upturn}
@@
 \begin{code}
-\VerbatimInput[xleftmargin=5mm,numbers=left]{../Peterson.cxl}
-\end{code}
-\caption{Peterson's Algorithm}
+\VerbatimInput[xleftmargin=5mm,numbers=left]{code/Peterson.cxl}
+\end{code}
+\caption{[Peterson.cxl] Peterson's Algorithm}
 \label{fig:peterson}
@@
 \begin{code}
-\VerbatimInput[xleftmargin=5mm,numbers=left]{../PetersonProof.cxl}
-\end{code}
-\caption{Peterson's Algorithm with Inductive Invariant}
+\VerbatimInput[xleftmargin=5mm,numbers=left]{code/PetersonInductive.cxl}
+\end{code}
+\caption{[PetersonInductive.cxl] Peterson's Algorithm with Inductive Invariant}
 \label{fig:petersonproof}
@@
 \begin{code}
-\VerbatimInput[xleftmargin=5mm,numbers=left]{../PetersonMethod.cxl}
-\end{code}
-\caption{Peterson's Algorithm accessed through methods.}
+\VerbatimInput[xleftmargin=5mm,numbers=left]{code/PetersonMethod.cxl}
+\end{code}
+\caption{[PetersonMethod.cxl] Peterson's Algorithm accessed through methods.}
 \label{fig:petersonmethods}
@@
 \begin{code}
-\VerbatimInput[xleftmargin=5mm,numbers=left]{UpTAS.cxl}
-\end{code}
-\caption{Mutual Exclusion using a ``spinlock'' based on test-and-set.}
+\VerbatimInput[xleftmargin=5mm,numbers=left]{code/spinlock.cxl}
+\end{code}
+\caption{[spinlock.cxl] Mutual Exclusion using a ``spinlock'' based on test-and-set.}
 \label{fig:tas}
@@
 \begin{code}
-\begin{verbatim}
+\begin{Verbatim}[xleftmargin=5mm,numbers=left]
 def atomic_inc(ptr):
@@
 call atomic_inc &(count);
-\end{verbatim}
+\end{Verbatim}
 \end{code}
@@
 \begin{code}
-\VerbatimInput[xleftmargin=5mm,numbers=left]{UpTASinv.cxl}
-\end{code}
-\caption{Checking invariants.}
+\VerbatimInput[xleftmargin=5mm,numbers=left]{code/spinlockInv.cxl}
+\end{code}
+\caption{[spinlockInv.cxl] Checking invariants.}
 \label{fig:tasinv}
@@
 \begin{code}
-\VerbatimInput[xleftmargin=5mm,numbers=left]{../TAS.cxl}
-\end{code}
-\caption{Fixed version of Figure~\ref{fig:uplock} using test-and-set.}
+\VerbatimInput[xleftmargin=5mm,numbers=left]{code/csTAS.cxl}
+\end{code}
+\caption{[csTAS.cxl] Fixed version of Figure~\ref{fig:uplock} using test-and-set.}
 \label{fig:tas2}
@@
 \begin{code}
-\begin{verbatim}
+\begin{Verbatim}[xleftmargin=5mm,numbers=left]
 def tas(lk):
@@
 ;
-\end{verbatim}
+\end{Verbatim}
 \end{code}
@@
 \begin{code}
-\VerbatimInput[xleftmargin=5mm,numbers=left]{UpFixed.cxl}
-\end{code}
-\caption{Program of Figure~\ref{fig:inc} fixed with a lock.}
+\VerbatimInput[xleftmargin=5mm,numbers=left]{code/UpLock.cxl}
+\end{code}
+\caption{[UpLock.cxl] Program of Figure~\ref{fig:inc} fixed with a lock.}
 \label{fig:incfixed}
@@
 \begin{code}
-\begin{verbatim}
+\begin{Verbatim}[xleftmargin=5mm,numbers=left]
 import list;
@@
 \begin{code}
-\VerbatimInput[xleftmargin=5mm,numbers=left]{../RWbusy.cxl}
-\end{code}
-\caption{Busy-Waiting Reader/Writer Lock implementation.}
+\VerbatimInput[xleftmargin=5mm,numbers=left]{code/RWbusy.cxl}
+\end{code}
+\caption{[RWbusy.cxl] Busy-Waiting Reader/Writer Lock implementation.}
 \label{fig:rwbusy}
@@
 \begin{code}
-\VerbatimInput[xleftmargin=5mm,numbers=left]{../RWblock.cxl}
-\end{code}
-\caption{Checking for Busy Waiting.}
+\VerbatimInput[xleftmargin=5mm,numbers=left]{code/RWbusychk.cxl}
+\end{code}
+\caption{[RWbusychk.cxl] Checking for Busy Waiting.}
 \label{fig:rwblock}
@@
 \begin{code}
-\VerbatimInput[xleftmargin=5mm,numbers=left]{../RWlock.cxl}
-\end{code}
-\caption{Reader/Writer with Two Locks.}
+\VerbatimInput[xleftmargin=5mm,numbers=left]{code/RWlock.cxl}
+\end{code}
+\caption{[RWlock.cxl] Reader/Writer with Two Locks.}
 \label{fig:rw2lock}
@@
 \begin{code}
-\VerbatimInput[xleftmargin=5mm,numbers=left]{../RWmulti.cxl}
-\end{code}
-\caption{Checking that multiple readers can acquire the read lock.}
+\VerbatimInput[xleftmargin=5mm,numbers=left]{code/RWmulti.cxl}
+\end{code}
+\caption{[RWmulti.cxl] Checking that multiple readers can acquire the read lock.}
 \label{fig:rwmulti}
@@
 \begin{code}
-\VerbatimInput[xleftmargin=5mm,numbers=left]{../Diners.cxl}
-\end{code}
-\caption{Dining Philosophers.}
+\VerbatimInput[xleftmargin=5mm,numbers=left]{code/Diners.cxl}
+\end{code}
+\caption{[Diners.cxl] Dining Philosophers.}
 \label{fig:diners}
@@
 \begin{code}
-\begin{verbatim}
+\begin{Verbatim}[xleftmargin=5mm,numbers=left]
     if left < right:
@@
     ;
-\end{verbatim}
+\end{Verbatim}
 \end{code}
@@
 \begin{code}
-\begin{verbatim}
+\begin{Verbatim}[xleftmargin=5mm,numbers=left]
 def Semaphore(cnt):
@@
 \begin{code}
-\VerbatimInput[xleftmargin=5mm,numbers=left]{../DinersSema.cxl}
-\end{code}
-\caption{Italian Philosophers.}
+\VerbatimInput[xleftmargin=5mm,numbers=left]{code/DinersSema.cxl}
+\end{code}
+\caption{[DinersSema.cxl] Italian Philosophers.}
 \label{fig:dinerssema}
@@
 \begin{code}
-\VerbatimInput[xleftmargin=5mm,numbers=left]{../PCsema.cxl}
-\end{code}
-\caption{Bounded Buffer implementation using semaphores.}
+\VerbatimInput[xleftmargin=5mm,numbers=left]{code/BBsema.cxl}
+\end{code}
+\caption{[BBsema.cxl] Bounded Buffer implementation using semaphores.}
 \label{fig:boundedbuffer}
@@
 \begin{code}
-\VerbatimInput[xleftmargin=5mm,numbers=left]{../PCsemadata.cxl}
-\end{code}
-\caption{Test whether the bounded buffer delivers the right data in the
+\VerbatimInput[xleftmargin=5mm,numbers=left]{code/BBsemadata.cxl}
+\end{code}
+\caption{[BBsemadata.cxl] Test whether the bounded buffer delivers the right data in the
 right order.}
@@
 \begin{code}
-\VerbatimInput[xleftmargin=5mm,numbers=left]{../RWbinsem.cxl}
-\end{code}
-\caption{Reader/Writer Lock implementation using Split Binary Semaphores.}
+\VerbatimInput[xleftmargin=5mm,numbers=left]{code/RWsbs.cxl}
+\end{code}
+\caption{[RWsbs.cxl] Reader/Writer Lock implementation using Split Binary Semaphores.}
 \label{fig:RWsplitsema}
@@
 \begin{code}
-\begin{verbatim}
+\begin{Verbatim}[xleftmargin=5mm,numbers=left]
 let blocked = True:
@@
 call unlock():
-\end{verbatim}
+\end{Verbatim}
 \end{code}
@@
 \begin{code}
-\VerbatimInput[xleftmargin=5mm,numbers=left]{../RWfair.cxl}
-\end{code}
-\caption{Reader/Writer Lock implementation addressing fairness.}
+\VerbatimInput[xleftmargin=5mm,numbers=left]{code/RWfair.cxl}
+\end{code}
+\caption{[RWfair.cxl] Reader/Writer Lock implementation addressing fairness.}
 \label{fig:RWfair}
@@
 \begin{code}
-\VerbatimInput[xleftmargin=5mm,numbers=left]{../RWcv.cxl}
-\end{code}
-\caption{Unfair Reader/Writer Lock implemtation using Mesa-style condition variables.}
+\VerbatimInput[xleftmargin=5mm,numbers=left]{code/RWcv.cxl}
+\end{code}
+\caption{[RWcv.cxl] Reader/Writer Lock implementation using Mesa-style condition variables.}
 \label{fig:RWcv}

### 2020-06-14 book/paper.tex
Adding some appendices

@@
 \maketitle
+\tableofcontents
 
@@
 
+\appendix
+
+\chapter{Complete List of CXL Values}
+
+Chapter~\ref{ch:cxlvalues} provides an introduction to CXL values.
+Below is a complete list of CXL value types with examples:
+
+\vspace{1em}
+\begin{tabular}{|l|l|}
+\hline
+Boolean & \texttt{False}, \texttt{True} \\
+\hline
+integers & \texttt{..., -2, -1, 0, 1, 2, ...} \\
+\hline
+Atom & \texttt{.example}, \texttt{.test1} \\
+\hline
+Program Counter & (method names are program counter constants) \\
+\hline
+Dictionary & \texttt{dict\{ .account: 12345, .valid: False \} } \\
+\hline
+Set & \texttt{\{ 1, 2, 3 \}, \{ False, .id, 3 \} } \\
+\hline
+Address & \texttt{\&(lock), \&(flags[2])} \\
+\hline
+Context & (generated by \texttt{stop} expression) \\
+\hline
+\end{tabular}
+\vspace{1em}
+
+Tuples, lists, strings, and bags are all special cases of dictionaries.
+Both tuples and lists map indexes (starting at 0) to CXL values.
+Their format is either \texttt{(e, e, ..., e,)} or
+\texttt{[e, e, ..., e,]}.
+If the tuple or list has two or more elements, then the final comma
+is optional.
+A string must be enclosed by double quotes and is represented as
+a tuple of its characters.  Characters are one-character atoms.
+A bag or multiset is a dictionary that maps a value to how many
+times it occurs in the bag.
+
+All CXL values are ordered with respect to one another.  First they
+are ordered by type according to the table above.
+So, for example, \texttt{\texttt{True} < 0 < .xyz < \{ 0 \})}.
+Within types, the following rules apply:
+
+\begin{itemize}
+\item \texttt{False < True};
+\item integers are ordered in the natural way;
+\item atoms are lexicographically ordered;
+\item program counters are ordered by their integer values;
+\item dictionaries are first converted into a list of ordered (key, value)
+pairs.  Then two dictionaries are lexicographically ordered by this
+representation;
+\item a set is first converted into an ordered list, then lexicographically
+ordered;
+\item an address is a list of atoms.  Addresses are lexicographically
+ordered accordingly;
+\item a context consists of a name tag, the value of the program counter,
+the value of the register, and the contents of the stack.  Currently,
+contexts are ordered by hash.
+\end{itemize}
+
+\chapter{Complete List of CXL Operators}
+
+CXL currently supports the following operators:
+
+\vspace{1em}
+\begin{tabular}{|l|l|}
+\hline
+\texttt{e == e} & any two CXL values \\
+\hline
+\texttt{e != e} & any two CXL values \\
+\hline
+\texttt{e < e} & any two CXL values \\
+\hline
+\texttt{e <= e} & any two CXL values \\
+\hline
+\texttt{e > e} & any two CXL values \\
+\hline
+\texttt{e >= e} & any two CXL values \\
+\hline
+\texttt{e and e and ...} & two or more Booleans \\
+\hline
+\texttt{e or e or ...} & two or more Booleans \\
+\hline
+\texttt{not e} & a Boolean \\
+\hline
+\texttt{-e} & an integer \\
+\hline
+\texttt{e + e + ...} & two or more integers, sets, or lists \\
+\hline
+\texttt{e - e} & two integers or sets \\
+\hline
+\texttt{e * e * ...} & two or more integers or sets \\
+\hline
+\texttt{e / e} & two integers (integer division) \\
+\hline
+\texttt{e \% e} & two integers (division remainder) \\
+\hline
+\texttt{e~..~e} & two integers \\
+\hline
+\texttt{e in e} & first is any CXL value, second is set \\
+\hline
+\texttt{e not in e} & first is any CXL value, second is set \\
+\hline
+\texttt{\^{}e} & an address \\
+\hline
+\texttt{choose e} & a set \\
+\hline
+\texttt{min e} & a set \\
+\hline
+\texttt{max e} & a set \\
+\hline
+\texttt{cardinality e} & a set \\
+\hline
+\texttt{keys e} & a dictionary \\
+\hline
+\texttt{len e} & a dictionary (includes tuple or list) \\
+\hline
+\texttt{bagsize e} & a bag \\
+\hline
+\texttt{atLabel e} & atom (corresponding to a label) \\
+\hline
+\texttt{nametag e} & \texttt{e} must be \texttt{()} \\
+\hline
+\texttt{processes e} & \texttt{e} must be \texttt{()} \\
+\hline
+\end{tabular}
+\vspace{1em}
+
+\texttt{+} computes the union when applied to sets and
+the concatenation when applied to lists or tuples.
+\texttt{-} computes set difference when applied to two sets.
+\texttt{*} computes the intersection when applied to sets.
+\texttt{x..y} computes the set consisting of the integers \texttt{x}
+through \texttt{y} inclusive.  It returns the empty set if \texttt{x > y}.
+
+CXL also supports the following comprehensions:
+
+\vspace{1em}
+\begin{tabular}{|l|l|}
+\hline
+Set comprehension & \texttt{\{ f(e) for e in s \}} \\
+\hline
+List comprehension & \texttt{[ f(e) for e in s ]} \\
+\hline
+Dict comprehension & \texttt{dict\{ f(e) for e in s \}} \\
+\hline
+\end{tabular}
+\vspace{1em}
+
+In each case, \texttt{s} must be a set.
+For list comprehensions, the list will be sorted by \texttt{e}.
+For dictionary comprehensions, each \texttt{e} will be mapped to \texttt{f(e)}.
+
 \end{document}

### 2020-06-14 book/paper.tex
Adding appendices

@@
 
-\chapter{Complete List of CXL Values}
+\chapter{List of CXL Values}
 
@@
 
-\chapter{Complete List of CXL Operators}
+\chapter{List of CXL Operators}
 
@@
 \hline
+\texttt{e if e else e} & middle \texttt{e} must be a Boolean \\
+\hline
 \texttt{\^{}e} & an address \\
@@
 \texttt{processes e} & \texttt{e} must be \texttt{()} \\
+\hline
+\texttt{\&(lv)} & \texttt{lv} is an lvalue \\
+\hline
+\texttt{stop lv} & \texttt{lv} is an lvalue \\
 \hline
@@
 
+An \emph{lvalue} (short for left hand value of an assignment statement)
+is something that can be assigned.  This can be a
+shared variable, a process variable, or a dereferenced pointer
+variable.  It can also be indexed.  Examples include
+\texttt{var}, \texttt{var[e]}, and \texttt{\^{}p}.
+
 CXL also supports the following comprehensions:
@@
 
+\chapter{List of CXL Statements}
+
+CXL currently supports the following statements:
+
+\vspace{1em}
+\begin{tabular}{|l|l|}
+\hline
+\texttt{lv = e} & \texttt{lv} is an lvalue and \texttt{e} is an expression; \\
+\hline
+\texttt{pass} & do nothing; \\
+\hline
+\texttt{del lv} & delete; \\
+\hline
+\texttt{call e} & evaluate \texttt{e}; \\
+\hline
+\texttt{assert b [, x]} & \texttt{b} is a Boolean.  If \texttt{False}, optionally print expression \texttt{e}; \\
+\hline
+\texttt{const v = e} & \texttt{v} is an identifier, \texttt{e} is a constant expression; \\
+\hline
+\texttt{def m(v, ...): S} & \texttt{m} and \texttt{v} are identifiers, \texttt{S} a list of statements; \\
+\hline
+\texttt{let v = e, ...: S} & \texttt{e} is an expression, \texttt{S} a list of statements; \\
+\hline
+\texttt{if b: S} & \texttt{b} is a Boolean, \texttt{S} a list of statements; \\
+\hline
+\texttt{while b: S} & \texttt{b} is a Boolean, \texttt{S} a list of statements; \\
+\hline
+\texttt{for v in e: S} & \texttt{e} is a set evaluated in order,
+                            \texttt{S} a list of statements; \\
+\hline
+\texttt{atomic: S} & \texttt{S} a list of statements; \\
+\hline
+\texttt{spawn p e [, t]} & \texttt{p} is a method, \texttt{e} is an expression, \texttt{t} is an optional tag (an expression) \\
+\hline
+\texttt{go c e} & \texttt{c} is a context, \texttt{e} is an expression \\
+\hline
+\texttt{import m} & \texttt{m} identifies a module. \\
+\hline
+\end{tabular}
+\vspace{1em}
+
+To iterate through a dictionary, list, or tuple, use:
+\texttt{for v in keys(e):~S}.
+
+\chapter{List of Modules}
+
+The \texttt{bag} module has various useful methods that operate on bags or
+multisets:
+
+\vspace{1em}
+\begin{tabular}{|l|l|}
+\hline
+\texttt{bagEmpty()} & returns an empty bag; \\
+\hline
+\texttt{bagFromset(s)} & create a bag from set \texttt{s}; \\
+\hline
+\texttt{bagCount(b, e} & count how many times \texttt{e} occurs in bag \texttt{b}; \\
+\hline
+\texttt{bagChoose(b)} & like \texttt{choose(s)}, but applied to a bag; \\
+\hline
+\texttt{bagAdd(pb, e)} & add one copy of \texttt{e} to bag \texttt{\^{}pb}; \\
+\hline
+\texttt{bagRemove(pb, e)} & remove one copy of \texttt{e} from bag \texttt{\^{}pb}; \\
+\hline
+\end{tabular}
+\vspace{1em}
+
+The \texttt{list} module has various useful methods that operate on lists
+or tuples:
+
+\vspace{1em}
+\begin{tabular}{|l|l|}
+\hline
+\texttt{subseq(t, b, f)} & return a \emph{slice} of list \texttt{t} starting
+at index \texttt{b} and ending just before index \texttt{f}; \\
+\hline
+\texttt{append(t, e)} & append \texttt{e} to list \texttt{t}; \\
+\hline
+\texttt{head(t)} & return the first element of list \texttt{t}; \\
+\hline
+\texttt{tail(t)} & return all but the first element of list \texttt{t}; \\
+\hline
+\texttt{qsort(t)} & sort list \texttt{t} using quicksort; \\
+\hline
+\texttt{sum(t)} & return the sum of all elements in \texttt{t}; \\
+\hline
+\texttt{list2bag(t)} & convert list \texttt{t} into a bag. \\
+\hline
+\end{tabular}
+
 \end{document}

### 2020-06-14 book/paper.tex
Latest

--- 2020-06-14 book/paper.tex
@@
 
+\chapter{List of CXL Statements}
+
+CXL currently supports the following statements:
+
+\vspace{1em}
+\begin{tabular}{|l|l|}
+\hline
+\texttt{lv = e} & \texttt{lv} is an lvalue and \texttt{e} is an expression\\
+\hline
+\texttt{pass} & do nothing\\
+\hline
+\texttt{del lv} & delete\\
+\hline
+\texttt{call e} & evaluate \texttt{e}\\
+\hline
+\texttt{assert b [, x]} & \texttt{b} is a Boolean.  If \texttt{False}, optionally print expression \texttt{e}\\
+\hline
+\texttt{const v = e} & \texttt{v} is an identifier, \texttt{e} is a constant expression\\
+\hline
+\texttt{def m(v, ...): S} & \texttt{m} and \texttt{v} are identifiers, \texttt{S} a list of statements\\
+\hline
+\texttt{let v = e, ...: S} & \texttt{e} is an expression, \texttt{S} a list of statements\\
+\hline
+\texttt{if b: S} & \texttt{b} is a Boolean, \texttt{S} a list of statements\\
+\hline
+\texttt{while b: S} & \texttt{b} is a Boolean, \texttt{S} a list of statements\\
+\hline
+\texttt{for v in e: S} & \texttt{e} is a set evaluated in order,
+                            \texttt{S} a list of statements\\
+\hline
+\texttt{atomic: S} & \texttt{S} a list of statements\\
+\hline
+\texttt{spawn p e [, t]} & \texttt{p} is a method, \texttt{e} is an expression, \texttt{t} is an optional tag (an expression) \\
+\hline
+\texttt{go c e} & \texttt{c} is a context, \texttt{e} is an expression \\
+\hline
+\texttt{import m} & \texttt{m} identifies a module \\
+\hline
+\end{tabular}
+\vspace{1em}
+
+To iterate through a dictionary, list, or tuple, use:
+\texttt{for v in keys(e):~S}.
+
+\chapter{List of Modules}
+
+The \texttt{bag} module has various useful methods that operate on bags or
+multisets:
+
+\vspace{1em}
+\begin{tabular}{|l|l|}
+\hline
+\texttt{bagEmpty()} & returns an empty bag\\
+\hline
+\texttt{bagFromset(s)} & create a bag from set \texttt{s}\\
+\hline
+\texttt{bagCount(b, e} & count how many times \texttt{e} occurs in bag \texttt{b}\\
+\hline
+\texttt{bagChoose(b)} & like \texttt{choose(s)}, but applied to a bag\\
+\hline
+\texttt{bagAdd(pb, e)} & add one copy of \texttt{e} to bag \texttt{\^{}pb}\\
+\hline
+\texttt{bagRemove(pb, e)} & remove one copy of \texttt{e} from bag \texttt{\^{}pb}\\
+\hline
+\end{tabular}
+\vspace{1em}
+
+The \texttt{list} module has various useful methods that operate on lists
+or tuples:
+
+\vspace{1em}
+\begin{tabular}{|l|l|}
+\hline
+\texttt{subseq(t, b, f)} & return a \emph{slice} of list \texttt{t} starting
+at index \texttt{b} and ending just before index \texttt{f}\\
+\hline
+\texttt{append(t, e)} & append \texttt{e} to list \texttt{t}\\
+\hline
+\texttt{head(t)} & return the first element of list \texttt{t}\\
+\hline
+\texttt{tail(t)} & return all but the first element of list \texttt{t}\\
+\hline
+\texttt{qsort(t)} & sort list \texttt{t} using quicksort\\
+\hline
+\texttt{sum(t)} & return the sum of all elements in \texttt{t}\\
+\hline
+\texttt{list2bag(t)} & convert list \texttt{t} into a bag \\
+\hline
+\end{tabular}
+
 \end{document}

### 2020-06-14 book/paper.tex
Adding description of machine instructions

@@
 monitors, where multiple waiting readers would have to be signaled one at a time in a
-baton-passing fashion (see Figure~\ref{fig:splitsema}), with Mesa monitors
+baton-passing fashion (see Figure~\ref{fig:RWsplitsema}), with Mesa monitors
 all readers are awaked in one fell swoop using \texttt{notifyAll}.
@@
 
+\chapter{List of CXL Machine Instructions}
+
+\vspace{1em}
+\begin{tabular}{|l|l|}
+\hline
+Address & compute address \\
+\hline
+Apply & pop $m$ and $i$ and apply $i$ to $m$, pushing a value \\
+\hline
+Assert & pop $b$ and check that it is \texttt{True} \\
+\hline
+AtomicDec & decrement the atomic counter of this context \\
+\hline
+AtomicInc & increment the atomic counter of this context \\
+\hline
+Choose & choose an element from the set on top of the stack \\
+\hline
+Del & delete a shared variable \\
+\hline
+DelVar $v$ & delete process variable $v$ \\
+\hline
+Dict & pop $n$ and $n$ key/value pairs and push a dictionary \\
+\hline
+Dup & duplicate the top element of the stack \\
+\hline
+Frame $m$ $a$ $r$& start method $m$ with arguments $a$,
+initializing variables.  Last instruction is at $r$ \\
+\hline
+Go & restore context \\
+\hline
+JumpCond $e$ $p$ & pop expression and if equal to $e$ set program counter to $p$ \\
+\hline
+Jump $p$ & set program counter to $p$ \\
+\hline
+Load & load a shared variable and push it onto the stack \\
+\hline
+LoadVar $v$ & load process variable $v$ and push it onto the stack \\
+\hline
+Nary \textit{op} $n$ & apply $n$-ary operator \textit{op} to the top $n$ elements on the stack \\
+\hline
+Pop & pop a value of the stack and discard it \\
+\hline
+Push $c$ & push constant $c$ onto the stack \\
+\hline
+PushAddress $v$ & push the address of shared variable $v$ \\
+\hline
+Return & pop return address, push \texttt{result}, and restore program counter \\
+\hline
+Set & pop $n$ and $n$ elements and push a set of the elements \\
+\hline
+Spawn & pop tag, argument, and method and spawn a new context \\
+\hline
+Split & split a non-empty set in its minimum element and its remainder \\
+\hline
+Stop & save context \\
+\hline
+Store & pop a value from the stack and store it in a shared variable \\
+\hline
+StoreVar $v$ & pop a value from the stack and store it in process variable $v$ \\
+\hline
+\end{tabular}
+
 \end{document}

### 2020-06-14 book/paper.tex
Latest

@@
 
-\chapter{List of CXL Values}
+\chapter{List of Values}
 
@@
 
-\chapter{List of CXL Operators}
+\chapter{List of Operators}
 
@@
 
-\chapter{List of CXL Statements}
+\chapter{List of Statements}
 
@@
 
-\chapter{List of CXL Machine Instructions}
+\chapter{List of Machine Instructions}
 
@@
 \hline
+AtomicInc & increment the atomic counter of this context \\
+\hline
 AtomicDec & decrement the atomic counter of this context \\
-\hline
-AtomicInc & increment the atomic counter of this context \\
 \hline
@@
 \hline
-Go & restore context \\
-\hline
-JumpCond $e$ $p$ & pop expression and if equal to $e$ set program counter to $p$ \\
+Go & pop context and value, push value on context's stack, and add to context bag \\
 \hline
@@
 \hline
+JumpCond $e$ $p$ & pop expression and, if equal to $e$, set program counter to $p$ \\
+\hline
 Load & load a shared variable and push it onto the stack \\
@@
 \hline
-Split & split a non-empty set in its minimum element and its remainder \\
-\hline
-Stop & save context \\
+Split & pop set and push minimum and remainder \\
+\hline
+Stop & save context into shared variable and remove from context bag \\
 \hline
@@
 \end{tabular}
+\vspace{1em}
+
+Clarifications:
+\begin{itemize}
+\item The effect of the \texttt{Apply} instructions depends much on $m$.
+If $m$ is a dictionary, then \texttt{Apply} finds $i$ in the dictionary
+and pushes the value.
+If $m$ is a program counter, then \texttt{Apply} invokes method $m$ by
+pushing the current program counter and setting the program counter to
+$m$.  $m$ is supposed to leave the result on the stack.
+\end{itemize}
 

### 2020-06-15 book/paper.tex
Latest

@@
 Use parentheses liberally to remove ambiguity.
-\item CXL does not (currently) support floating point.  CXL does not support I/O;
+\item CXL does not (currently) support floating point, iterators, or I/O;
 \item Python is object-oriented; CXL is not;
@@
 ``Running'' this CXL program will try all possible executions, which
-includes all possible values for $x$.  Try it out:
+includes all possible values for $x$.  Try it out (here \texttt{\$}
+represents a shell prompt):
 
@@
 
+\chapter{Non-Blocking Synchronization}
+
+\chapter{Barrier Synchronization}
+
 \appendix
@@
 ordered accordingly;
-\item a context consists of a name tag, the value of the program counter,
-the value of the register, and the contents of the stack.  Currently,
-contexts are ordered by hash.
+\item contexts (Appendix~\ref{app:context} are ordered by hash.
 \end{itemize}
@@
 
+CXL does not support Python-style iterators.
 To iterate through a dictionary, list, or tuple, use:
@@
 $m$.  $m$ is supposed to leave the result on the stack.
+\item The \texttt{Frame} instruction pushes the value of the process
+register (\emph{i.e.}, the values of the process variables) onto the
+stack.  It initializes the \texttt{result} variable to the empty dictionary.
+The \texttt{Return} instruction restores the process register by popping
+its value of the stack.
+\item All method calls have exactly one argument, although it sometimes
+appears otherwise:
+\begin{itemize}
+\item \texttt{m()} invokes method \texttt{m} with the empty dictionary \texttt{()} as argument;
+\item \texttt{m(a)} invokes method \texttt{m} with argument \texttt{a};
+\item \texttt{m(a, b, c)} invokes method \texttt{m} with tuple \texttt{(a, b, c)} as argument.
 \end{itemize}
+The \texttt{Frame} instruction unpacks the argument to the method and places them into process variables by the given names.
+\end{itemize}
+
+\chapter{Contexts and Processes}
+\label{app:context}
+
+A context captures the state of a process.  Each time the process
+executes an instruction, it goes from one context to another.
+All instructions update the program counter, and so no instruction
+leaves the context the same.
+There may be multiple processes
+with the same state at the same time.
+
+A context consists of the following:
+
+\vspace{1em}
+\begin{tabular}{|l|l|}
+\hline
+name tag & a dictionary with atoms \texttt{.name} and \texttt{.tag} \\
+\hline
+program counter & a Program Counter value pointing into the code \\
+\hline
+end & a Program Counter value indicating the end of the process code \\
+\hline
+atomic & if non-zero, the process is in atomic mode \\
+\hline
+stack & a list of CXL values \\
+\hline
+register & a CXL dictionary mapping atoms (representing process variables) to CXL values \\
+\hline
+stopped & a Boolean indicating if the context is stopped \\
+\hline
+\end{tabular}
+\vspace{1em}
+
+Details:
+\begin{itemize}
+\item The name in a name tag is the name of the method that the process
+is executing;
+\item The tag in a name tag is its argument by default.  Optionally
+\texttt{spawn} allows the tag to be specified explicitly;
+\item A process terminates when it reaches the \texttt{Return} instruction
+of the method or when it hits a fault.  Faults include divide by zero,
+reading a non-existent key in a dictionary, accessing a non-existent
+variable, as well as when an assertion fails;
+\item The execution of a process in atomic mode does not get interleaved
+with that of other processes.
+\item The register of a process always contains a dictionary, mapping
+atoms to arbitrary values.  The atoms correspond to the variable names
+in a CXL program.
+\end{itemize}
+
+A context is considered \emph{blocked} if there does not exist a
+sequence instructions that the process can execute that either writes
+to a shared variable (even if that leaves the shared variable unchanged)
+or terminates the process.
+
+\chapter{The CXL Machine}
+
+The CXL machine has the following state:
+
+\vspace{1em}
+\begin{tabular}{|l|l|}
+\hline
+code & a list of CXL machine instructions \\
+\hline
+labels & a dictionary of atoms to program counters \\
+\hline
+variables & a dictionnary mapping atoms to values \\
+\hline
+ctxbag & a bag of non-stopped contexts \\
+\hline
+stopbag & a bag of stopped contexts \\
+\hline
+choosing & if not \texttt{None}, indicates a context that is choosing \\
+\hline
+failure & Boolean indicating some context has failed \\
+\hline
+initializing & Boolean indicating that the state is not fully initialized. \\
+\hline
+\end{tabular}
+\vspace{1em}
+
+There is initially a single context with nametag
+\texttt{\_\_init\_\_/()} and program counter 0.  It starts executing
+in atomic mode until it finishes executing the last instruction in
+the code.  All states until then are \texttt{initializing} states.
+Other processes, created through \texttt{spawn} statements, do not
+start executing until then.
+
+A \emph{micro step} is the execution of a single CXL machine instruction
+by a context.
+Each micro step generates a new state.
+When there are multiple contexts, the CXL machine can interleave them.
+However, trying to interleave every microstep would be needlessly expensive,
+as many micro steps involve changes to a context that are invisible to
+other contexts.
+
+A \emph{macro step} can involve multiple micro steps.  The following
+instructions start a new macro step: \texttt{Load}, \texttt{Store},
+\texttt{AtomicInc}, and \texttt{Choose}.  Also, any instruction after
+a \texttt{Stop} instruction starts a macro step.  The CXL machine
+interleaves macro steps, not micro steps.  Like micro steps, each
+macro step involves a single context.  Unlike a micro step, a macro
+step can leave the state unchanged.
+
+Executing a CXL program results in a graph where the nodes are CXL
+states and the edges are macro steps.  A terminating state is a
+state with an empty context bag or a state where all contexts are
+blocked.
 

### 2020-06-15 book/paper.tex
Latest
Added ContinueOp

@@
 The \texttt{Frame} instruction unpacks the argument to the method and places them into process variables by the given names.
+\item Every \texttt{Stop} instruction must immediately be followed by a 
+\texttt{Continue} instruction.
 \end{itemize}
@@
 instructions start a new macro step: \texttt{Load}, \texttt{Store},
-\texttt{AtomicInc}, and \texttt{Choose}.  Also, any instruction after
-a \texttt{Stop} instruction starts a macro step.  The CXL machine
+\texttt{AtomicInc}, and \texttt{Continue}.  The CXL machine
 interleaves macro steps, not micro steps.  Like micro steps, each
@@
 Executing a CXL program results in a graph where the nodes are CXL
-states and the edges are macro steps.  A terminating state is a
-state with an empty context bag or a state where all contexts are
-blocked.
+states and the edges are macro steps.
+When a state is \texttt{choosing}, the edges from that state are
+by a single context, one for each choice.  If not, the edges from
+the state are one per context.
+A terminating state is a state with an empty context bag or a state
+where all contexts are blocked.
 

### 2020-06-15 cxl.py

@@ -498,11 +498,19 @@ class StopOp(Op):
         assert isinstance(av, AddressValue), indexes
 
         # First update the context before saving it
-        context.pc += 1
         context.stopped = True
+        context.pc += 1
+        assert isinstance(state.code[context.pc], ContinueOp)
 
         # Save the context
         state.stop(av.indexes + indexes[1:], context)
+
+class ContinueOp(Op):
+    def __repr__(self):
+        return "Continue"
+
+    def eval(self, state, context):
+        context.pc += 1
 
 class AddressOp(Op):
     def __init__(self, n):
@@ -1651,6 +1659,7 @@ class StopAST(AST):
             if tv == None:
                 code.append(PushAddressOp(lv.name))
                 code.append(StopOp(n))
+                code.append(ContinueOp())
             else:
                 print("Error: Can't store state in process variable")
                 sys.exit(1)
@@ -1658,6 +1667,7 @@ class StopAST(AST):
             assert isinstance(lv, PointerAST), lv
             lv.expr.compile(scope, code)
             code.append(StopOp(n))
+            code.append(ContinueOp())
 
 class AddressAST(AST):
     def __init__(self, lv):
@@ -2522,7 +2532,7 @@ def optimize(code):
 
 # These operations cause global state changes
 globops = [
-    AtomicIncOp, LoadOp, StoreOp
+    AtomicIncOp, ContinueOp, LoadOp, StoreOp
 ]
 
 lasttime = 0

### 2020-06-15 book/paper.tex
Latest

@@
 
+\begin{figure}
+\begin{code}
+\VerbatimInput[xleftmargin=5mm,numbers=left]{code/barrier.cxl}
+\end{code}
+\caption{[barrier.cxl] Barrier synchronization with semaphores.}
+\label{fig:barrier}
+\end{figure}
+
 \appendix
@@
 \hline
-\texttt{sum(t)} & return the sum of all elements in \texttt{t}\\
-\hline
 \texttt{list2bag(t)} & convert list \texttt{t} into a bag \\
+\hline
+\texttt{list2set(t)} & convert list \texttt{t} into a set \\
+\hline
+\texttt{listMin(t)} & return the minimum of all elements in \texttt{t}\\
+\hline
+\texttt{listMax(t)} & return the maximum of all elements in \texttt{t}\\
+\hline
+\texttt{listSum(t)} & return the sum of all elements in \texttt{t}\\
 \hline
@@
 AtomicDec & decrement the atomic counter of this context \\
+\hline
+Continue & no-op (but causes a context switch) \\
 \hline

### 2020-06-15 cxl.py

@@ -2346,7 +2346,7 @@ class State:
                 v = v.d[indexes[0]]
             except KeyError:
                 print()
-                print("can't find", indexes[0], "in variable", path)
+                print("no index", indexes[0], "in variable", path)
                 sys.exit(1)
             indexes = indexes[1:]
         return v

### 2020-06-15 book/paper.tex
Latest

@@
 
+\begin{figure}
+\begin{center}
+\includegraphics[width=\textwidth]{figures/barrier.jpeg}
+\end{center}
+\caption{High-level state diagram specification of barrier synchronization
+with three processes and three rounds.  The three numbers specify the round
+number of each process.}
+\label{fig:barrierdiagram}
+\end{figure}
+
 \appendix
@@
 ordered accordingly;
-\item contexts (Appendix~\ref{app:context} are ordered by hash.
+\item contexts (Appendix~\ref{app:context}) are ordered by hash.
 \end{itemize}

### 2020-06-15 book/paper.tex
Made figures more clear
Add non-blocking queue a la Herlihy/Wing

@@
 \begin{center}
-\includegraphics[width=4in]{figures/mutex.jpeg}
+\includegraphics[width=4in]{figures/mutex-crop.pdf}
 \end{center}
@@
 \begin{center}
-\includegraphics[width=4in]{figures/states.jpeg}
+\includegraphics[width=4in]{figures/states-crop.pdf}
 \end{center}
@@
 \begin{center}
-\includegraphics[width=2in]{figures/rdwr.jpeg}
+\includegraphics[width=2in]{figures/rdwr-crop.pdf}
 \end{center}
@@
 \begin{center}
-\includegraphics[width=2.5in]{figures/pc.jpeg}
+\includegraphics[width=2.5in]{figures/pc-crop.pdf}
 \end{center}
@@
 \begin{center}
-\includegraphics[width=6in]{figures/monitor.jpeg}
+\includegraphics[width=6in]{figures/monitor-crop.pdf}
 \end{center}
@@
 \begin{center}
-\includegraphics[width=\textwidth]{figures/barrier.jpeg}
+\includegraphics[width=\textwidth]{figures/barrier-crop.pdf}
 \end{center}
@@
 \hline
-\texttt{e == e} & any two CXL values \\
-\hline
-\texttt{e != e} & any two CXL values \\
+\texttt{e == e} & two CXL values of the same type \\
+\hline
+\texttt{e != e} & two CXL values of the same type \\
 \hline

### 2020-06-15 book/paper.tex
Latest
Latest version

@@
 
+\begin{figure}
+\begin{code}
+\VerbatimInput[xleftmargin=5mm,numbers=left]{code/hw.cxl}
+\end{code}
+\caption{[hw.cxl] Non-blocking queue.}
+\label{fig:hw}
+\end{figure}
+
+So far we have concentrated on critical sections to synchronize multiple
+processes.  Certainly, preventing multiple processes from accessing
+certain code at the same time simplifies how to think about synchronization.
+However, it can lead to starvation.  Even in the absence of starvation,
+if some process is slow for some reason while being in the critical section,
+the other processes have to wait for it to finish the critical section.
+In this chapter, we will have a brief look at how one can develop concurrent
+code in which processes never have to wait for other processes.
+For this, we will revisit the producer/consumer problem, in which a
+set of processes are producing items and a set of processes are
+consuming items.
+
+The code in Figure~\ref{fig:hw} is based on code developed by Herlihy and
+Wing~\cite{HW}.  It uses two interlock instructions:
+\begin{itemize}
+\item \texttt{inc(p)}: atomically increments \texttt{\^{}p} and returns
+the old value;
+\item \texttt{swap(p)}: sets \texttt{\^{}p} to \texttt{()} and returns
+the old value.
+\end{itemize}
+
 \chapter{Barrier Synchronization}

### 2020-06-15 book/paper.tex
Latest

@@
 
-A good example of a resource with multiple units is the so-called \emph{bounded buffer}.
+A good example of a resource with multiple units is the
+so-called \emph{bounded buffer} (aka \emph{ring buffer}).
 It is essentially
@@
 The code in Figure~\ref{fig:hw} is based on code developed by Herlihy and
-Wing~\cite{HW}.  It uses two interlock instructions:
+Wing~\cite{HW}.
+The code is a ``proof of existence'' for non-blocking synchronization; it
+is not necessarily practical.
+There are two variables.  \texttt{items} is an unbounded array with each
+entry initialized to \texttt{()}.  \texttt{back} is an index into the
+array and points to the next slot where a new value is inserted.
+The code uses two interlock instructions:
 \begin{itemize}
@@
 \end{itemize}
+
+Method \texttt{produce(item)} uses \texttt{inc \&(back)} to allocate
+the next available slot in the \texttt{items} array.
+It stores the item as a singleton tuple.
+Method \texttt{consume()} repeatedly scans the array, up to the
+\texttt{back} index, trying to find an item to return.
+To check an entry, it uses \texttt{swap()}
+to atomically remove an item from a slot if there is one.
+This way, if two or more processes attempt to extract an item from
+the same slot, at most one will succeed.
+
+There is no critical section.  If one process is executing instructions
+very slowly, this does not negatively impact the other processes, as it
+would with solutions based on critical sections.
+On the contrary, it helps them because it creates less contention.
+
+The solution is not practical because the \texttt{items} array must be
+unbounded if an unbounded number of items may be produced.  Each slot
+in the array is only used once, which is inefficient.  Also, the \texttt{inc}
+and \texttt{swap} interlock instructions are not universally available
+on existing hardware.
+However, in the literature there are many examples of practical
+non-blocking (aka \emph{wait-free}) synchronization algorithms~\cite{}.
 

### 2020-06-15 book/paper.tex
Starting on bibliography

@@
 \usepackage{fullpage}
+\usepackage[nottoc,numbib]{tocbibind}
 \usepackage{graphicx}
@@
 CXL is heavily influenced by Leslie Lamport's work on
-TLA+, TLC, and PlusCal~\cite{}.
+TLA+, TLC, and PlusCal~\cite{Lamport02, Lamport09}.
 CXL is designed to have a lower learning curve than those
@@
 Another excellent resource is Fred Schneider's book ``On
-Concurrent Programming.''
+Concurrent Programming''~\cite{Schneider97}.
 
@@
 or one or more readers to be in the critical section.  This is called
-a \emph{reader/writer lock}~\cite{courtois71}.
+a \emph{reader/writer lock}~\cite{CHP71}.
 We will explore various ways of implementing reader/writer locks in
@@
 In the late 70s, Xerox PARC developed a new programming language called
-Mesa~\cite{mesa}.
+Mesa~\cite{LR80}.
 Mesa introduced various important concepts to programming languages,
@@
 The code in Figure~\ref{fig:hw} is based on code developed by Herlihy and
-Wing~\cite{HW}.
+Wing~\cite{HW87}.
 The code is a ``proof of existence'' for non-blocking synchronization; it
@@
 \end{figure}
+
+\bibliographystyle{alpha}
+\bibliography{paper}
 

### 2020-06-15 book/paper.tex
Latest
Some text for barrier synchronization

@@
 \usepackage{fullpage}
-\usepackage[nottoc,numbib]{tocbibind}
+\usepackage[nottoc,notlot,notlof]{tocbibind}
 \usepackage{graphicx}
@@
 \begin{figure}
-\begin{code}
-\VerbatimInput[xleftmargin=5mm,numbers=left]{code/barrier.cxl}
-\end{code}
-\caption{[barrier.cxl] Barrier synchronization with semaphores.}
-\label{fig:barrier}
-\end{figure}
-
-\begin{figure}
 \begin{center}
@@
 \caption{High-level state diagram specification of barrier synchronization
-with three processes and three rounds.  The three numbers specify the round
-number of each process.}
+with three processes and two rounds.  The three numbers specify how many
+rounds each process has completed.}
 \label{fig:barrierdiagram}
 \end{figure}
+
+\begin{figure}
+\begin{code}
+\VerbatimInput[xleftmargin=5mm,numbers=left]{code/barrier.cxl}
+\end{code}
+\caption{[barrier.cxl] Barrier synchronization with semaphores.}
+\label{fig:barrier}
+\end{figure}
+
+Barrier synchronization is a problem that comes up in high-performance
+parallel computing, used, among others, for scalable simulation.
+Barrier synchronization is almost the opposite of critical sections:
+the intention is to get a group of processes to run some code at the
+same time.
+More precisely, the processes run in rounds.  At the end of each round
+there is a so-called \emph{barrier} where processes wait until all
+processes have completed the round.
+For example, in an iterative matrix algorithm, the matrix may be
+cut up into fragments.  During a round, one process runs for each
+fragment concurrently.  The next round is not allowed to start
+until all processes have completed operations on their fragment.
+
+Figure~\ref{fig:barrierdiagram} shows a high-level depiction with three
+processes.  Initially all processes are in round 0.  Then each process
+can start and finish the round.  However, none of the processes can
+progress to the next round until all processes have finished the
+current round.
+
+Counting semaphores work well for implementing barrier synchronization.
+Figure~\ref{fig:barrier} shows an example.  There is a 0-initialized
+semaphore for each of the \texttt{N} processes.
+Before process $i$ enters a round, it first sends a signal to every
+other process and then waits until it receives a signal from
+every other process.  Process $i$ sends a signal to process $j$
+by incrementing \texttt{sema[$j$]} (using \texttt{V}),
+while process $i$ waits for a signal be decrementing \texttt{sema[$i$]}
+(using \texttt{P}).
+
+The \texttt{round} array is kept to check the correctness of this
+approach.  Each process increments its entry every time it enters
+a round.  If the algorithm is correct, it can never be that two processes
+are more than one round apart.
 

### 2020-06-15 book/paper.tex
Latest
Added citation to the little book of semaphores

@@
 
+You may find the free ``The Little Book of Semaphores'' by
+Allen Downey a great resource for working with semaphores.
+You can find it at
+\underline{http://greenteapress.com/semaphores/LittleBookOfSemaphores.pdf}.
+
 \chapter{Bounded Buffer}
@@
 
+Andrew Birrell's paper on Programming with Threads gives an excellent
+introduction to working with Mesa-style condition variables~\cite{Birrell89}.
+
 \chapter{Non-Blocking Synchronization}
@@
 are more than one round apart.
+
+\begin{quote}
+\begin{itemize}
+\item See if you can implement barrier synchronization for $N$ processes
+with just three semaphores.  Busy waiting is not allowed.
+\item How about just two?
+\end{itemize}
+\end{quote}
 

### 2020-06-15 book/paper.tex
Latest

@@
 You may find the free ``The Little Book of Semaphores'' by
-Allen Downey a great resource for working with semaphores.
-You can find it at
-\underline{http://greenteapress.com/semaphores/LittleBookOfSemaphores.pdf}.
+Allen Downey a great resource for working with semaphores~\cite{Downey09}.
 

### 2020-06-16 book/paper.tex
Added alternating bit protocol implementation
Added actor code

@@
 
+\chapter{Actors}
+
+\begin{figure}
+\begin{code}
+\VerbatimInput[xleftmargin=5mm,numbers=left]{code/actor.cxl}
+\end{code}
+\caption{[actor.cxl] A producer/consumer actor.}
+\label{fig:actor}
+\end{figure}
+
+\begin{figure}
+\begin{code}
+\VerbatimInput[xleftmargin=5mm,numbers=left]{code/actortest.cxl}
+\end{code}
+\caption{[actortest.cxl] Test code for producer/consumer actor.}
+\label{fig:actortest}
+\end{figure}
+
 \chapter{Non-Blocking Synchronization}
@@
 \end{quote}
+
+\chapter{Networking: Alternating Bit Protocol}
+
+\begin{figure}
+\begin{code}
+\VerbatimInput[xleftmargin=5mm,numbers=left]{code/abp.cxl}
+\end{code}
+\caption{[abp.cxl] Alternating Bit Protocol.}
+\label{fig:abp}
+\end{figure}
+
+\begin{figure}
+\begin{code}
+\VerbatimInput[xleftmargin=5mm,numbers=left]{code/abptest.cxl}
+\end{code}
+\caption{[abptest.cxl] Test code for alternating bit protocol.}
+\label{fig:abptest}
+\end{figure}
 

### 2020-06-16 book/paper.tex
Added some text for actors

@@
 Introduced by Dijkstra,
-a \emph{semaphore} is a synchronization primitive that fits the bill.
+a \emph{semaphore} is a synchronization primitive that fits the bill~\cite{EWD35}.
 A semaphore is essentially
@@
 The ``Split Binary Semaphore'' (SBS) is such a technique, originally proposed by
-Tony Hoare and popularized by Edsger Dijkstra.
+Tony Hoare and popularized by Edsger Dijkstra~\cite{EWD703}.
 SBS is an example of what one could call a ``baton-passing'' technique.
@@
 While Hoare monitors are conceptually simpler, the so-called
-``Mesa monitor semantics'' have become more popular, adopted by all major
-programming languages including C, Java, and Python.
+``Mesa monitor semantics'' or ``Mesa condition variable'' semantics
+have become more popular, adopted by all major
+programming languages.
+
+That said, few programming languages provide full support for monitors.
+In Java, each object has a hidden lock \emph{and} a hidden condition variable
+associated with it.
+Methods declared with the \texttt{synchronized} keyword automatically
+obtain the lock.  Java objects also support \texttt{wait}, \texttt{notify},
+and \texttt{notifyAll}.  Java also supports explicit allocations of locks
+and condition variables.
+In Python, locks and condition variables must be explicitly declared.
+The \texttt{with} statement makes it easy to acquire and release a lock
+for a section of code.
+In C and C++, support for locks and condition variables is through libraries.
 
@@
 of \texttt{wait}.
+Many implementation of condition variables depend on this, and optimizations of the
+code allow so-called ``spurious wakeups,'' where \texttt{wait} may sometimes return even
+if it has not been notified.
 
@@
 \end{figure}
+
+Some programming languages favor a completely different way of implementing
+synchronization using so-called \emph{actors}~\cite{HBS73}.  Actors are
+processes that have only private memory and communicate through messages.
+Given that there is no shared memory in the actor model, there is no need
+for critical sections.  Instead, some sequential process owns a particular
+piece of data and other processes access it by sending request messages
+to the process and optionally waiting for response messages.  Each process
+handles one message at a time, serializing all access to the data it owns.
+
+The actor synchronization model is popular in a variety of programming
+languages, including Erlang and Scala.  Actor support is also available
+through popular libraries such as Akka, which is available for various
+programming languages.  In Python, Java, and C/C++,
+actors can be easily emulated using threads and \emph{synchronized queues}
+for messaging.  Each thread would have one such queue for receiving messages.
+Dequeuing from an empty synchronized queue blocks the thread until
+another thread enqueues a message on the queue.
+
+The \texttt{synch} library supports a synchronized message queue,
+similar to the \texttt{Queue} object in Python.
+Its interface is as follows:
+\begin{itemize}
+\item \texttt{Queue()} returns a new message queue;
+\item \texttt{enqueue(q, item)} add \texttt{item} to the queue pointed to by \texttt{q};
+\item \texttt{dequeue(q)} waits for and returns an item on the queue pointed to by \texttt{q}.
+\end{itemize}
+
+Note that a \texttt{Queue} behaves much like a zero-initialized semaphore.
+\texttt{enqueue} is much like \texttt{V}, except that it is accompanied by data.
+\texttt{dequeue} is much like \texttt{P}, except that it also returns data.
+Thus, synchronized queues can be considered a generalization of counting semaphores.
+
+Figure~\ref{fig:actor} shows a solution to the producer/consumer problem using
+an actor.  \texttt{pc\_actor} is essentially a matchmaker: it matches
+\texttt{.produce} requests with \texttt{.consume} requests.
+Sometimes it may have buffered \texttt{.consume} requests and is waiting for
+\texttt{.produce} requests, and sometimes it is vice versa.
+The variable \texttt{balance} specifies what the situation is: if it is positive
+then it has buffered \texttt{.produce} requests;
+if it is negative
+then it has buffered \texttt{.consume} requests.
+Note that both \texttt{balance} and \texttt{requests} are variables that are
+local to the \texttt{pc\_actor} process.
+
+Each request message is a dictionary with two fields.  One of the fields is
+\texttt{.type}, which either is \texttt{.produce} or \texttt{.consume}.
+In case of \texttt{.produce}, the second field is \texttt{.item} and holds
+the data that is produced.
+In case of \texttt{.consume}, the second field is \texttt{.queue} and holds
+a pointer to the queue where the response message must be enqueued.
+\texttt{.produce} messages do not get a response.
+
+The \texttt{pc\_actor} process has two arguments.  \texttt{q} is a pointer
+to the queue on which it receives messages.  \texttt{nrequests} is the total
+number of requests that it will handle.  We need the latter because in CXL
+all processes are required to terminate.  In a more realistic implementation,
+the \texttt{pc\_actor} would be in an infinite loop awaiting requests.
+
+The \texttt{produce} method takes two arguments: a pointer to the queue
+of the \texttt{pc\_actor} and the item.  The \texttt{consume} method
+takes two queue pointer arguments.  The first is a pointer to the queue
+of the \texttt{pc\_actor}.  The second is a pointer to the queue of the
+invoker: it is the queue to which the response must be posted.
+
+Figure~\ref{fig:actortest} shows how this code may be used.  It spawns
+\texttt{NITEMS} producer and consumer processes.  Note that in that
+case the \texttt{pc\_actor} is expected to receive \texttt{2 * NITEMS}
+requests.
 

### 2020-06-16 book/paper.tex
Working on index and glossary

@@
 \usepackage{fullpage}
+\usepackage{imakeidx}\makeindex
+\usepackage{hyperref}
 \usepackage[nottoc,notlot,notlof]{tocbibind}
+\usepackage{glossaries}\makeglossaries
 \usepackage{graphicx}
@@
 \newcommand{\tm}{\raisebox{.9ex}{\tiny tm}}
+
+\newglossaryentry{macro step}
+{
+  name=macro step,
+  description={is a programmable machine that receives input,
+               stores and manipulates data, and provides
+               output in a useful format}
+}
 
@@
 \item The execution of a sequential program is mostly \emph{deterministic}.
+\index{determinism}
 If you run it twice with the same input, the same output will be produced.
@@
 (so-called \emph{Heisenbugs}).
+\index{Heisenbug}
 \item In a sequential program, each statement and each function can be
-thought of as happening \emph{atomically} because there is no other activity
+thought of as happening \emph{atomically}
+\index{atomicity}
+because there is no other activity
 interfering with their execution.  Even though a statement or function may
@@
 called CXL that helps with \emph{testing} concurrent code.
-The approach is based on \emph{model checking}: instead of relying
+The approach is based on \emph{model checking}:
+\index{model checking}
+instead of relying
 on luck, CXL will run \emph{all possible executions} of a particular
@@
 
-Model checking is not a replacement for formal verification.  Formal
-verification proves that a program is correct.  Model checking only
+Model checking is not a replacement for formal verification.
+\index{formal verification}
+Formal verification proves that a program is correct.  Model checking only
 verifies that a program is correct for some \emph{model}.  Think of
@@
 the \texttt{count} variable is being updated, no other process is
-trying to do the same thing.  We call this a \emph{critical section}
-(aka critical region): a
+trying to do the same thing.  This is called a \emph{critical section}
+(aka critical region)~\cite{EWD123}: a
 set of instructions where only one process is allowed to execute at a
@@
 process also tries to access the binary tree.
+An implementation of a data structure that can be safely accessed by multiple
+processes (or threads) is called \emph{thread-safe}.
 
@@
 \label{ch:peterson}
+\index{Peterson's Algorithm}
 
@@
 For this (as well as some other uses), CXL supports \emph{pointers}
+\index{pointer}
 to shared variables.
@@
 \chapter{Semaphores}
+\index{semaphore}
 
@@
 Reader/writer locks and bounded buffers are both examples of what are
-sometimes called \emph{conditional critical sections} (CCSs).
+sometimes called \emph{conditional critical sections} (CCSs)~\cite{Hoare73}.
 All critical sections are in fact conditional.  Even the basic one
@@
 an abstraction of the concept in a programming language paradigm called
-\emph{monitors}.
+\emph{monitors}~\cite{Hoare74}.
+\index{monitor}
+(A similar construct was independently invented by Per Brinch Hansen~\cite{BH73}.)
 A monitor is a special version of an object-oriented \emph{class}, comprising
@@
 The other semaphores are called \emph{condition variables}.
-There are two operations on condition variables: \texttt{wait} and \texttt{signal}.
+\index{condition variable}
+There are two operations on condition variables: \texttt{wait}
+\index{wait}
+and
+\texttt{signal}.
+\index{signal}
 If \texttt{c} is a pointer to a condition variable, then the semantics of the
@@
 Mesa~\cite{LR80}.
+\index{Mesa}
 Mesa introduced various important concepts to programming languages,
@@
 operation \texttt{notify} rather than \texttt{signal}.
+\index{notify}
 When a process $p$ invokes \texttt{notify}, it does not immediately pass 
@@
 using \texttt{signal} on an empty bathroom.) Mesa also has a function called
-\texttt{notifyAll} (aka \texttt{broadcast)}) that, when applied to
+\texttt{notifyAll}
+\index{notifyAll}
+(aka \texttt{broadcast)})
+\index{broadcast}
+that, when applied to
 a bedroom, let's all processes in that bedroom into the hall.
@@
 
-\chapter{Actors}
+\chapter{Actors and Synchronized Queues}
+\index{actor}
 
@@
 actors can be easily emulated using threads and \emph{synchronized queues}
-for messaging.  Each thread would have one such queue for receiving messages.
+(aka \emph{blocking queues}) for messaging.
+Each thread would have one such queue for receiving messages.
 Dequeuing from an empty synchronized queue blocks the thread until
@@
 
-A \emph{macro step} can involve multiple micro steps.  The following
+A \emph{macro step}
+\glsadd{macro step}
+can involve multiple micro steps.  The following
 instructions start a new macro step: \texttt{Load}, \texttt{Store},
@@
 
+\printindex
+
+% \glsaddall
+\printglossaries
+
 \end{document}

### 2020-06-16 book/paper.tex
More index items

@@
 The reason is that the correctness of many programs is built around
-\emph{invariants}: predicates that must hold for every state in the
+\emph{invariants}:
+\index{invariant}
+predicates that must hold for every state in the
 execution of a program.  A model checker can find violations of
@@
 What is going on is that the CXL program is compiled to machine instructions,
+\index{machine instruction}
 and it is the machine instructions that are executed by the underlying CXL
@@
 \label{ch:cxlvalues}
+\index{CXL Value}
 
@@
 \label{ch:cxlmachine}
+\index{CXL machine}
 
@@
 \chapter{Critical Sections}
+\index{critical section}
+\index{critical region}
 
@@
 ensure is that there cannot be two processes at the critical section.
+This property is called \texttt{mutual exclusion}.
+\index{mutual exclusion}
 We would like to place an assertion at the \texttt{@cs} label that
@@
 
-Mutual exclusion is an example of a \emph{safety property}, a
-property that ensures that \emph{nothing bad will happen}, in this case
+Mutual exclusion is an example of a \emph{safety property},
+\index{safety property}
+a property that ensures that \emph{nothing bad will happen}, in this case
 two processes being in the critical section.
-What we need now a \emph{liveness property}: we want to ensure that
+What we need now a \emph{liveness property}:
+\index{liveness property}
+we want to ensure that
 \emph{eventually something good will happen}.
@@
 We call this \emph{progress}.
+\index{progress}
 
@@
 
-Figure~\ref{fig:mutex} show a high-level state diagram of the code in
+Figure~\ref{fig:mutex} show a high-level state diagram
+\index{state diagram}
+of the code in
 Figure~\ref{fig:cs}.
 The circles represent states, while the arrows represent possible state
-transitions.  The label in each circle summarizes the state;  the label
-on an arrow describes the transition.  In this case,
+transitions (high-level steps).  The label in each circle summarizes the state;
+the label on an arrow describes the transition.  In this case,
 the label contains two numbers: the total number of processes and the number
@@
 
-You may already have heard of the concept of a \emph{lock} and have realized that
+You may already have heard of the concept of a \emph{lock}
+\index{lock}
+and have realized that
 it could be used to implement a critical section.
@@
 state changes (if only because the program counter of the process
-changes).  We call that a \emph{step}.  Steps in CXL are atomic.
+changes).  We call that a \emph{step}.
+\index{step}
+Steps in CXL are atomic.
 
@@
 sequence of steps starting from the initial state.
-When making a step, there are two sources of non-determinism in CXL.
+When making a step, there are two sources of non-determinism
+\index{non-determinism}
+in CXL.
 One is when
@@
 \emph{inductive invariant}.
+\index{inductive invariant}
 An inductive invariant $I$ is a predicate over states that satisfies the following:
@@
 \chapter{CXL Methods and Pointers}
+\index{CXL method}
 
@@
 \chapter{Spinlock}
+\index{spinlock}
 
@@
 through its \texttt{atomic}
+\index{atomic statement}
 statement.
@@
 
-Many CPUs have an atomic ``test-and-set'' operation.
+Many CPUs have an atomic ``test-and-set'' (TAS)
+\index{test-and-set}
+\index{TAS}
+operation.
 Method \texttt{tas} in Figure~\ref{fig:tas} shows its specification.
@@
 a set of $N$ processes.
-It is called \emph{spinlock}, because each process is ``spinning'' until
+It is called \emph{spinlock},
+\index{spinlock}
+because each process is ``spinning'' until
 it can acquire the lock.
@@
 \begin{itemize}
-\item Find a description for \texttt{compare-and-swap} on the internet and
+\item Find a description for \texttt{compare-and-swap} (CAS)
+\index{compare-and-swap}
+\index{CAS}
+on the internet and
 update Figure~\ref{fig:tas}, replacing \texttt{tas}.
@@
 \texttt{fetch-and-add}.
+\index{fetch-and-add}
 \end{itemize}
@@
 \chapter{Locks and the Synch Module}
+\index{lock}
+\index{synch module}
 
@@
 
-We call a process \emph{blocked} in a state if the process is in
+We call a process \emph{blocked}
+\index{blocked}
+in a state if the process is in
 an \emph{infinite loop} in which the process does not update the
@@
 \chapter{Reader/Writer Locks using Busy Waiting}
+\index{reader/writer lock}
+\index{busy waiting}
 
@@
 \chapter{Deadlock}
+\index{deadlock}
 
@@
 a scenario he called ``Dining Philosophers.''
+\index{dining philosopher}
 
@@
 When allocating a resource, a process decrements the
-counter using the \texttt{P} operation.  You can think of \texttt{P} standing
+counter using the \texttt{P}
+\index{P}
+\index{Procure}
+operation.  You can think of \texttt{P} standing
 for \emph{Procure}, as in procuring the resource associated with the semaphore.
@@
 To release the resource, a process increments the resource using the
-\texttt{V} operation.  You can think of \texttt{V} standing for \emph{vacate},
+\texttt{V}
+\index{V}
+\index{Vacate}
+operation.  You can think of \texttt{V} standing for \emph{vacate},
 as in vacating the resource.
@@
 \chapter{Bounded Buffer}
+\index{bounded buffer}
+\index{producer/consumer problem}
 
@@
 \chapter{Split Binary Semaphores}
+\index{split binary semaphore}
 
@@
 Reader/writer locks and bounded buffers are both examples of what are
-sometimes called \emph{conditional critical sections} (CCSs)~\cite{Hoare73}.
+sometimes called \emph{conditional critical sections}
+\index{conditional critical section}
+(CCSs)~\cite{Hoare73}.
 All critical sections are in fact conditional.  Even the basic one
@@
 \label{ch:starvation}
+\index{starvation}
 
@@
 \chapter{Non-Blocking Synchronization}
+\index{non-blocking synchronization}
 
@@
 \chapter{Barrier Synchronization}
+\index{barrier synchronization}
 
@@
 \chapter{Networking: Alternating Bit Protocol}
+\index{alternating bit protocol}
 
@@
 
-The \texttt{bag} module has various useful methods that operate on bags or
+The \texttt{bag} module
+\index{bag module}
+has various useful methods that operate on bags or
 multisets:
@@
 
+\noindent
+\index{list module}
 The \texttt{list} module has various useful methods that operate on lists
@@
 
-A \emph{micro step} is the execution of a single CXL machine instruction
+A \emph{micro step}
+\index{micro step}
+is the execution of a single CXL machine instruction
 by a context.
@@
 A \emph{macro step}
+\index{macro step}
 \glsadd{macro step}

### 2020-06-16 book/paper.tex
More index items

@@
 Each process has an immutable (but not necessarily unique) \emph{name tag},
+\index{name tag}
 a program counter, a stack of CXL values,
@@
 A name tag consists of the name of the main method of the process,
-along with an optional tag specified in the \texttt{spawn} statement.
+along with an optional tag specified in the \texttt{spawn}
+\index{spawn statement}
+\index{tag}
+statement.
 The default tag is the first argument to the method.
@@
 
-The state of a process is called a \emph{context}: it contains the value of
+The state of a process is called a \emph{context} (aka continuation):
+\index{context}
+\index{continuation}
+it contains the value of
 its name tag, program counter, stack, and register.
@@
 The CXL machine is predominantly a \emph{stack machine}.
+\index{stack machine}
 All instructions are atomically executed.
@@
 In Chapter~\ref{ch:cxlmachine} we talked about the concept of \emph{state}:
+\index{state}
 at any point of time the CXL machine is in a specific state.
@@
 Here \texttt{L} is a shared list.
-Another process can revive the process using the \texttt{go} stateent:
+Another process can revive the process using the \texttt{go}
+\index{go statement}
+statement:
 
@@
 It causes a process with context \texttt{C} to be added to the state that has
-just executed the \texttt{stop} expression.  The \texttt{stop} expression
-returns the value \texttt{R}.
+just executed the \texttt{stop}
+\index{stop expression}
+expression.  The \texttt{stop} expression returns the value \texttt{R}.
 

### 2020-06-16 book/paper.tex
Glossary now in table of contents
Still more index terms

@@
 \usepackage[nottoc,notlot,notlof]{tocbibind}
-\usepackage{glossaries}\makeglossaries
+\usepackage[toc]{glossaries}\makeglossaries
 \usepackage{graphicx}
@@
 Essentially, the $\texttt{choose}(S)$
+\index{choose operator}
 operator provides the input to the program by selecting some value from the
@@
 line using the \texttt{-c} option.
+\index{constant}
 For example, if you want to test the code for \texttt{N = 100}, run:
@@
 You can put the first three methods in its own CXL source file
-and include it using the CXL \texttt{import} statement.  This would
+and include it using the CXL \texttt{import} statement.
+\index{import statement}
+\index{module}
+This would
 make the code re-usable by other applications.  For those who are
@@
 CXL allows you to select which version of the \texttt{synch} module you would
-like to use with the \texttt{-m} flag.  For example,
+like to use with the \texttt{-m} flag.
+\index{module}
+For example,
 

### 2020-06-16 book/paper.tex
Fixed author
Latest

@@
 \title{Concurrent Programming with CXL}
-\author{RVR and possibly others}
+\author{Robbert van Renesse}
 
@@
 can make programs faster than sequential ones, but having multiple
-processes concurrently read and update shared variables and
-synchronize with one another makes programs more complicated than
+processes concurrently read and update shared variables
+\index{shared variable}
+and synchronize with one another makes programs more complicated than
 programs where only one thing happens at a time.  Why are concurrent
@@
 Concurrent programming, aka multithreaded programming, involves multiple
-processes running in parallel while sharing variables.
+processes
+\index{process}
+running in parallel while sharing variables.
 Figure~\ref{fig:inc} presents a simple example.
@@
 The program spawns three processes.
+\index{process}
 The first runs \texttt{incrementer(0)}, the second runs
@@
 sets of CXL values,
-and dictionaries that map CXL values to other CXL values.
+and dictionaries
+\index{dictionary}
+that map CXL values to other CXL values.
 %
-Another type of CXL value is the \emph{atom}.  It is essentially
+Another type of CXL value is the \emph{atom}.
+\index{atom}
+It is essentially
 just a name.  An atom is denoted using a period followed by the
@@
 \index{name tag}
-a program counter, a stack of CXL values,
-and a single mutable general purpose \emph{register} that contains a CXL value.
+a program counter,
+\index{program counter}
+a stack of CXL values,
+and a single mutable general purpose \emph{register}
+\index{register}
+that contains a CXL value.
 \end{itemize}
@@
 a dictionary that maps atoms to CXL values.  We call this a \emph{directory}.
+\index{directory}
 A directory represents the state of a collection of variables named by the atoms.
@@
 processes (or threads) is called \emph{thread-safe}.
+\index{thread safety}
+\index{thread}
 
@@
 Figure~\ref{fig:csbarebones} shows the CXL code.
-Here \texttt{@cs} is a \emph{label}, identifying
-a location in the CXL machine code.  The first thing we need to
+Here \texttt{@cs} is a \emph{label},
+\index{label}
+identifying a location in the CXL machine code.  The first thing we need to
 ensure is that there cannot be two processes at the critical section.
@@
 CXL in fact supports this.
-It has an operator \texttt{atLabel $L$}, where $L$
+It has an operator \texttt{atLabel $L$},
+\index{atLabel operator}
+where $L$
 is the atom containing the name of the label (in this case, \texttt{.cs}).
@@
 use it in normal code.
-The assertion also makes use of the \texttt{nametag()} operator that
-returns the name tag of the current process.
+The assertion also makes use of the \texttt{nametag()} operator
+\index{nametag operator}
+that returns the name tag of the current process.
 If you run the code through CXL, the assertion should fail because

### 2020-06-16 book/paper.tex
More index terms

@@
 \begin{itemize}
-\item The execution of a sequential program is mostly \emph{deterministic}.
+\item The execution of a sequential
+\index{sequential}
+program is mostly \emph{deterministic}.
 \index{determinism}
@@
 the machine, and they do so in parallel.  As they execute, their machine
-instructions are interleaved in unspecified and often random ways.
+instructions are interleaved
+\index{interleaving}
+in unspecified and often random ways.
 A program is correct if it works for any interleaving.
@@
 The state of the CXL machine
-consists of the value of its memory and the multiset (or \emph{bag}) of
-contexts.  It is a multiset of contexts because two different processes can
+consists of the value of its memory and the multiset (or \emph{bag})
+\index{bag}
+\index{multiset}
+of contexts.  It is a multiset of contexts because two different processes can
 be in the same state.
@@
 by a sequence of atoms, like a path name in the file system hierarchy.  We call
-such a sequence the \emph{address} of a CXL value, and it is relative to the
+such a sequence the \emph{address}
+\index{address}
+of a CXL value, and it is relative to the
 ``root'' of the directory tree.
@@
 \chapter{Critical Sections}
-\index{critical section}
-\index{critical region}
 
@@
 trying to do the same thing.  This is called a \emph{critical section}
-(aka critical region)~\cite{EWD123}: a
-set of instructions where only one process is allowed to execute at a
+(aka critical region)~\cite{EWD123}:
+\index{critical section}
+\index{critical region}
+a set of instructions where only one process is allowed to execute at a
 time.
@@
 The CXL machine starts in an initial state in which there is only
-one process and its program counter is~0.  A \emph{trace} is a
-sequence of steps starting from the initial state.
+one process and its program counter is~0.  A \emph{trace}
+\index{trace}
+is a sequence of steps starting from the initial state.
 When making a step, there are two sources of non-determinism
@@
 Because there is non-determinism, there are multiple possible traces.
-We call a state \emph{reachable} if it is either the initial state
+We call a state \emph{reachable}
+\index{reachable state}
+if it is either the initial state
 or it can be reached from the initial state through a trace.
@@
 
-A \emph{property} describes a subset of
+A \emph{property}
+\index{property}
+describes a subset of
 states.  In other words, a property is equivalent to a subset of the states.
@@
 and a trace.
-An \emph{invariant} is a property that holds over all reachable states.
+An \emph{invariant}
+\index{invariant}
+is a property that holds over all reachable states.
 We want to show that mutual exclusion is an invariant.
@@
 And as CXL has already determined, mutual exclusion is an invariant:
-it holds over every \emph{reachable state}.
+it holds over every reachable state.
 Unfortunately, it is not an \emph{inductive} invariant.
@@
 (the parentheses are mandatory) is a pointer to \texttt{x}
-(also known as the \emph{address} of \texttt{x}).
+(also known as the \emph{address}
+\index{address}
+of \texttt{x}).
 Conversely, if \texttt{p} is a pointer to a shared variable, then the
@@
 if generalized to multiple processes.  Instead, multi-core processors provide
-so-called \emph{interlock instructions}: special machine instructions
+so-called \emph{interlock instructions}:
+\index{interlock instruction}
+special machine instructions
 that can read memory and then write it in an indivisible step.
@@
 We call a process \emph{blocked}
-\index{blocked}
+\index{blocked process}
 in a state if the process is in
@@
 \texttt{choose} to pick an arbitrary one.  Selecting the first is a common
-choice in lock implementations as it prevents \emph{starvation}: every process
+choice in lock implementations as it prevents \emph{starvation}:
+\index{starvation}
+every process
 gets a chance to obtain the lock (assuming every process eventually releases
@@
 \index{reader/writer lock}
-\index{busy waiting}
 
@@
 The solution is an example of what is called \emph{busy-waiting}
+\index{busy waiting}
 (aka spin-waiting):
+\index{spin waiting}
 processes spin in a loop until some desirable application-level condition is met.
@@
 so-called \emph{bounded buffer} (aka \emph{ring buffer}).
+\index{ring buffer}
 It is essentially
-a queue implemented using a circular buffer of a certain length and two pointers:
+a queue implemented using a circular buffer
+\index{circular buffer}
+of a certain length and two pointers:
 one where items are inserted and one from where items are extracted.  If the
@@
 People often call semaphores that are used as locks \emph{binary semaphores},
+\index{binary semaphore}
 as their counters only take on the values 0 and 1.
@@
 The second type of semaphores are called \emph{counting semaphores}.
+\index{counting semaphore}
 They can be used to send signals between processes.
@@
 will be exponentially diminishing.
-We shall call such a solution \emph{fair} (although it does not quite
+We shall call such a solution \emph{fair}
+\index{fairness}
+(although it does not quite
 match the usual formal nor vernacular concepts of fairness).
@@
 actors can be easily emulated using threads and \emph{synchronized queues}
-(aka \emph{blocking queues}) for messaging.
+\index{synchronized queue}
+(aka \emph{blocking queues})
+\index{blocking queue}
+for messaging.
 Each thread would have one such queue for receiving messages.
@@
 However, in the literature there are many examples of practical
-non-blocking (aka \emph{wait-free}) synchronization algorithms~\cite{}.
+non-blocking (aka \emph{wait-free})
+\index{wait-free synchronization}
+synchronization algorithms~\cite{}.
 
@@
 
-An \emph{lvalue} (short for left hand value of an assignment statement)
+An \emph{lvalue}
+\index{lvalue}
+(short for left hand value of an assignment statement)
 is something that can be assigned.  This can be a
@@
 
-A context is considered \emph{blocked} if there does not exist a
+A context is considered \emph{blocked}
+\index{blocked process}
+if there does not exist a
 sequence instructions that the process can execute that either writes

### 2020-06-16 book/paper.tex
Started working on a glossary

@@
 
+\newglossaryentry{atomicity}
+{
+  name=atomicity,
+  description={describes that a certain machine instruction or sequence of machine
+    instructions by a process is executed indivisibly and cannot be interleaved with 
+    machine instructions of another process}
+}
+\newglossaryentry{concurrent execution}
+{
+  name=concurrent execution,
+  description={is when there are multiple processes executing and their machine
+    instructions are interleaved in an unpredictable manner}
+}
+\newglossaryentry{determinism}
+{
+  name=determinism,
+  description={is when the outcome of an execution is uniquely determined by the
+    initial state}
+}
+\newglossaryentry{invariant}
+{
+  name=invariant,
+  description={is a binary property that must hold for every state of a process}
+}
+\newglossaryentry{machine instruction}
+{
+  name=machine instruction,
+  description={is an atomic operation on the CXL machine, executed by a process}
+}
 \newglossaryentry{macro step}
@@
   name=macro step,
-  description={is a programmable machine that receives input,
-               stores and manipulates data, and provides
-               output in a useful format}
+  description={is a sequence of micro steps with at most one instruction
+        externally visible to other processes.  It is the unit of interleaving
+        in CXL}
+}
+\newglossaryentry{micro step}
+{
+  name=micro step,
+  description={is the execution of one machine instruction by a process, updating
+        its state and the state of the CXL machine}
+}
+\newglossaryentry{model checking}
+{
+  name=model checking,
+  description={is a formal verification method that explores all possible executions
+    of a program, which must have a finite number of states}
+}
+\newglossaryentry{process}
+{
+  name=process,
+  description={is a method in execution.  We do not make the distinction
+      between processes and threads.  A process has a program counter, a register,
+      and a stack}
+}
+\newglossaryentry{process variable}
+{
+  name=process variable,
+  description={is a variable that is private to a single process and stored in
+        its register}
+}
+\newglossaryentry{sequential execution}
+{
+  name=sequential execution,
+  description={is when there is just one process executing, as opposed to concurrent
+    execution}
+}
+\newglossaryentry{shared variable}
+{
+  name=shared variable,
+  description={is a variable that is stored in the memory of the CXL machine and
+      shared between multiple processes, as opposed to a process variable}
 }
@@
 can make programs faster than sequential ones, but having multiple
-processes concurrently read and update shared variables
+processes (aka \emph{threads}\footnote{We
+\index{thread}
+do not make a distinction between processes and threads.}) concurrently
+\glsadd{concurrent execution}
+read and update shared variables
 \index{shared variable}
+\glsadd{shared variable}
 and synchronize with one another makes programs more complicated than
@@
 \index{sequential}
+\glsadd{sequential execution}
 program is mostly \emph{deterministic}.
 \index{determinism}
+\glsadd{determinism}
 If you run it twice with the same input, the same output will be produced.
@@
 \index{atomicity}
+\glsadd{atomicity}
 because there is no other activity
@@
 \index{model checking}
+\glsadd{model checking}
 instead of relying
@@
 \index{invariant}
+\glsadd{invariant}
 predicates that must hold for every state in the
@@
 \index{process}
+\glsadd{process}
 running in parallel while sharing variables.
@@
 The program spawns three processes.
-\index{process}
 The first runs \texttt{incrementer(0)}, the second runs
@@
 \index{machine instruction}
+\glsadd{machine instruction}
 and it is the machine instructions that are executed by the underlying CXL
@@
 A state is comprised of the values of the shared variables as well as
-the values of the process variables for each process, including its
+the values of the process variables
+\index{process variable}
+\glsadd{process variable}
+for each process, including its
 program counter and the contents of its stack.
@@
 \index{micro step}
+\glsadd{micro step}
 is the execution of a single CXL machine instruction
@@
 
-% \glsaddall
 \printglossaries

### 2020-06-16 book/paper.tex
Being more precise about properties vs predicates and invariants

@@
   name=concurrent execution,
-  description={is when there are multiple processes executing and their machine
-    instructions are interleaved in an unpredictable manner}
+  description={(aka parallel execution) is when there are multiple processes executing and
+      their machine instructions are interleaved in an unpredictable manner}
+}
+\newglossaryentry{context}
+{
+  name=context,
+  description={(aka continuation) describes the state of a running process,
+    including its program counter, the values of its variables (stored in its
+    register), and the contents of its stack}
+}
+\newglossaryentry{critical section}
+{
+  name=critical section,
+  description={(aka critical region) is a set of instructions that only one process
+    is allowed to execute at a time.  The instructions are, however, not executed
+    atomically, as other processes can continue to execute and access shared
+    variables}
 }
@@
   name=invariant,
-  description={is a binary property that must hold for every state of a process}
+  description={is a binary predicate over states that must hold for every
+  reachable state of a process}
 }
@@
 }
+\newglossaryentry{property}
+{
+  name=property,
+  description={describes a set of execution traces or histories that are allowed by
+    a program.  Safety properties are properties in which ``no bad things happen,''
+    such as violating mutual exclusion in a critical section.  Liveness properties are
+    properties where ``something good eventually happens,'' like processes being
+    able to enter the critical section if they want to}
+}
 \newglossaryentry{sequential execution}
@@
       shared between multiple processes, as opposed to a process variable}
+}
+\newglossaryentry{stack machine}
+{
+  name=stack machine,
+  description={is a model of computing where the state of a process is kept on
+    a stack.  CXL uses a combination of a stack machine and a register-based machine}
+}
+\newglossaryentry{thread safety}
+{
+  name=thread safety,
+  description={is when the implementation of a data structure allows concurrent access
+    with well-defined semantics}
 }
@@
 \index{continuation}
+\glsadd{context}
 it contains the value of
@@
 \index{stack machine}
+\glsadd{stack machine}
 All instructions are atomically executed.
@@
 \index{critical section}
+\glsadd{critical section}
 \index{critical region}
@@
 \index{thread safety}
+\glsadd{thread safety}
 \index{thread}
@@
 identifying a location in the CXL machine code.  The first thing we need to
-ensure is that there cannot be two processes at the critical section.
+ensure is that there can never be two processes at the critical section.
 This property is called \texttt{mutual exclusion}.
@@
 \end{center}
-\caption{Venn diagram of all states and a trace.}
+\caption{Venn diagram classifying all states and a trace.}
 \label{fig:states}
@@
 
-A \emph{property}
-\index{property}
-describes a subset of
-states.  In other words, a property is equivalent to a subset of the states.
+It is often useful to classify states into sets.
 \emph{Initial}, \emph{final}, and \emph{reachable}, and \emph{unreachable}
-are all examples of properties.
-Figure~\ref{fig:states} depicts a Venn diagram of various properties
+are all examples of classes of states.
+Figure~\ref{fig:states} depicts a Venn diagram of various classes of states
 and a trace.
-An \emph{invariant}
+One way to classify states is to define a predicate over states.
+All states in which \texttt{x = 1}, or all states where
+there are two or more processes executing, are examples of such predicates.
+For our purposes, it is useful to define a predicate that says that at
+most one process is in the critical section.  We shall call such states
+\emph{exclusive}.
+
+An \emph{invariant} of a program
 \index{invariant}
-is a property that holds over all reachable states.
-We want to show that mutual exclusion is an invariant.
-In other words, we want to show that the set of reachable states is a subset
-of the set of states where there is at most one process in the critical
+is a predicate that holds over all states that are reachable by that program.
+We want to show that exclusivity is an invariant.
+In other words, we want to show that the set of reachable states of executing
+the program
+is a subset of the set of states where there is at most one process in the critical
 section.
 
-One way to prove that a property is an invariant is through induction
-on the number of steps.  First you prove that the property holds over
+One way to prove that a predicate is an invariant is through induction
+on the number of steps.  First you prove that the predicate holds over
 the initial state.  Then you prove that for every reachable state,
-and for every step from that reachable state, the property also holds
+and for every step from that reachable state, the predicate also holds
 over the resulting state.
@@
 
-One candidate for such an invariant is mutual exclusion itself.
+One candidate for such an invariant is exclusivity itself.
 After all, it certainly holds over the initial state.
-And as CXL has already determined, mutual exclusion is an invariant:
+And as CXL has already determined, exclusivity is an invariant:
 it holds over every reachable state.
@@
 process~1 will break out of the while loop and also enter the critical
-section, violating mutual exclusion.
-So mutual exclusion is an invariant, but not an inductive invariant.
+section, entering a state that is not exclusive.
+So exclusivity is an invariant, but not an inductive invariant.
 Doing a inductive proof with an invariant that is not inductive is usually
@@
 
-So we are looking for an inductive invariant that \emph{implies} mutual
-exclusion.  In other words, the set of states where the inducive invariant holds
+So we are looking for an inductive invariant that \emph{implies} exclusivity.
+In other words, the set of states where the inducive invariant holds
 must be a subset of the set of states where there is at most one process in
@@
 
-To see that the inductive invariant implies mutual exclusion, suppose not.  Then
+To see that the inductive invariant implies exclusivity, suppose not.  Then
 when both process~0 and process~1 are in the critical section, the
@@
 So we conclude that $\mathtt{turn} = 0 \land \mathtt{turn} = 1$, a
-logical impossibility.  Thus $I_p(i)$ implies mutual exclusion.
+logical impossibility.  Thus $I_p(i)$ implies exclusivity.
 
@@
 
+A \emph{property}
+\index{property}
+\glsadd{property}
+is a set of traces.
+If a program has a certain property, that means that the traces that that
+program allows are a subset of the traces in the property.
 So far we have pursued two properties: \emph{mutual exclusion}

### 2020-06-16 book/paper.tex
Added concept of race condition

@@
 }
+\newglossaryentry{race condition}
+{
+  name=race condition,
+  description={describes when multiple processes access shared state concurrently,
+    leading to undesirable outcomes}
+}
 \newglossaryentry{sequential execution}
@@
 The result in this particular interleaving is that \texttt{count} ends up
-being 1.  When running CXL, it will
+being 1.
+This is known as a \emph{race condition}.
+\index{race condition}
+\glsadd{race condition}
+When running CXL, it will
 report violations of assertions.  It also provides an example

### 2020-06-16 book/paper.tex
Adding a glossary

@@
 
+\newglossaryentry{actor model}
+{
+  name=actor model,
+  description={is a concurrency model where there are no shared variables, only
+      processes with private variables and exchanging messages}
+}
 \newglossaryentry{atomicity}
@@
 }
+\newglossaryentry{barrier synchronization}
+{
+  name=barrier synchronization,
+  description={is when a set of processes execute in rounds, waiting for one another
+    to complete each round}
+}
+\newglossaryentry{blocked process}
+{
+  name=blocked process,
+  description={is a process cannot make progress without the help of another process.
+      For example, a process that is waiting for a spinlock to become available}
+}
+\newglossaryentry{busy waiting}
+{
+  name=busy waiting,
+  description={(aka spin-waiting) is when a process waits in a loop for some
+      application-defined condition while changing the state, for example by
+      continually acquiring and releasing a lock}
+}
 \newglossaryentry{concurrent execution}
@@
       their machine instructions are interleaved in an unpredictable manner}
+}
+\newglossaryentry{condition variable}
+{
+  name=condition variable,
+  description={a variable that keeps track of which processes are waiting for a
+    specific application-level condition.  The variable can be waited on as well
+    as signaled or notified}
 }
@@
     register), and the contents of its stack}
+}
+\newglossaryentry{conditional critical section}
+{
+  name=conditional critical section,
+  description={is a critical section with, besides mutual exclusion, additional conditions on
+      when a process is allowed to enter the critical section}
 }
@@
 }
+\newglossaryentry{deadlock}
+{
+  name=deadlock,
+  description={is when there are two or more processes waiting indefinitely for one
+    another to release a resource}
+}
 \newglossaryentry{determinism}
@@
 }
+\newglossaryentry{fairness}
+{
+  name=fairness,
+  description={is when each process eventually can access each resource it needs to access
+    with high probability}
+}
+\newglossaryentry{interlock instruction}
+{
+  name=interlock instruction,
+  description={a machine instruction that involves multiple memory load and/or store
+    operations, executed atomically}
+}
 \newglossaryentry{invariant}
@@
 }
+\newglossaryentry{lock}
+{
+  name=lock,
+  description={an object that can be owned by at most one process at a time.  Useful
+    for implementing mutual exclusion}
+}
 \newglossaryentry{machine instruction}
@@
   description={is an atomic operation on the CXL machine, executed by a process}
-}
-\newglossaryentry{macro step}
-{
-  name=macro step,
-  description={is a sequence of micro steps with at most one instruction
-        externally visible to other processes.  It is the unit of interleaving
-        in CXL}
-}
-\newglossaryentry{micro step}
-{
-  name=micro step,
-  description={is the execution of one machine instruction by a process, updating
-        its state and the state of the CXL machine}
 }
@@
     of a program, which must have a finite number of states}
+}
+\newglossaryentry{monitor}
+{
+  name=monitor,
+  description={is a programming language paradigm that supports mutual exclusion
+    as well as waiting for resources to become available}
+}
+\newglossaryentry{mutual exclusion}
+{
+  name=mutual exclusion,
+  description={is a property that two processes never enter the same critical section}
+}
+\newglossaryentry{non-blocking synchronization}
+{
+  name=non-blocking synchronization,
+  description={(aka wait-free synchronization) is when access to a shared resource can be
+      guaranteed in a bounded number of steps even if other processes are not making progress}
 }
@@
         its register}
+}
+\newglossaryentry{producer/consumer problem}
+{
+  name=producer/consumer problem,
+  description={is a synchronization problem whereby one or more producing processes
+    submit items and one or more consuming process want to receive them.  No item
+    can get lost or forged or be delivered to more than one consumer, and producers
+    and consumers should block if resources are exhausted}
 }
@@
 }
+\newglossaryentry{reader/writer lock}
+{
+  name=reader/writer lock,
+  description={is a lock on a resource that can be held by multiple processes if they all 
+    only read the resource}
+}
+\newglossaryentry{semaphore}
+{
+  name=semaphore,
+  description={is a counter that can be atomically incremented and decremented,
+    but blocks the process until the counter is larger than zero first}
+}
 \newglossaryentry{sequential execution}
@@
 }
+\newglossaryentry{spinlock}
+{
+  name=spinlock,
+  description={is an implementation of a lock whereby a process loops until the
+    lock is available, at which point the process atomically obtains the lock}
+}
 \newglossaryentry{stack machine}
@@
 }
+\newglossaryentry{starvation}
+{
+  name=starvation,
+  description={is when a process cannot make progress because it is continuously
+    losing a competition with other processes to get access to a resource}
+}
+\newglossaryentry{state}
+{
+  name=state,
+  description={an assignment of values to variables.  In a CXL machine, this includes
+      the contents of its shared memory and the set of contexts}
+}
+\newglossaryentry{step}
+{
+  name=step,
+  description={is the execution of a machine instruction by a process, updating
+        its state.  CXL distinguishes micro steps (machine instructions) and macro
+        step (a sequence of instructions with at most one effect visible by
+        other processes)}
+}
 \newglossaryentry{thread safety}
@@
     with well-defined semantics}
+}
+\newglossaryentry{trace}
+{
+  name=trace,
+  description={is a sequence of steps, starting from an initial state.  An infinite
+    trace is also called a \emph{behavior}}
 }
@@
 \index{mutual exclusion}
+\glsadd{mutual exclusion}
 We would like to place an assertion at the \texttt{@cs} label that
@@
 \index{lock}
+\glsadd{lock}
 and have realized that
@@
 \index{state}
+\glsadd{state}
 at any point of time the CXL machine is in a specific state.
@@
 \index{step}
+\glsadd{step}
 Steps in CXL are atomic.
@@
 \index{trace}
+\glsadd{trace}
 is a sequence of steps starting from the initial state.
@@
 \index{spinlock}
+\glsadd{spinlock}
 
@@
 \index{interlock instruction}
+\glsadd{interlock instruction}
 special machine instructions
@@
 \index{blocked process}
+\glsadd{blocked process}
 in a state if the process is in
@@
 \index{reader/writer lock}
+\glsadd{reader/writer lock}
 
@@
 \index{busy waiting}
+\glsadd{busy waiting}
 (aka spin-waiting):
@@
 \index{deadlock}
+\glsadd{deadlock}
 
@@
 \index{semaphore}
+\glsadd{semaphore}
 
@@
 \index{producer/consumer problem}
+\glsadd{producer/consumer problem}
 
@@
 (CCSs)~\cite{Hoare73}.
-All critical sections are in fact conditional.  Even the basic one
+\glsadd{conditional critical section}
+All critical sections are, in a way, conditional.  Even the basic critical section
 has the condition that a process can only enter if no other process
@@
 \index{starvation}
+\glsadd{starvation}
 
@@
 In this chapter, we will present a version of a reader/writer lock
-implementation that solves this type of starvation, but won't eliminate
-the type of starvation.
+implementation that solves this type of starvation.
 In particular, the probability that either a reader or writer gets starved
@@
 \index{fairness}
+\glsadd{fairness}
 (although it does not quite
@@
 \index{monitor}
+\glsadd{monitor}
 (A similar construct was independently invented by Per Brinch Hansen~\cite{BH73}.)
@@
 \index{condition variable}
+\glsadd{condition variable}
 There are two operations on condition variables: \texttt{wait}
@@
 \chapter{Actors and Synchronized Queues}
-\index{actor}
+\index{actor model}
+\glsadd{actor model}
 
@@
 \index{non-blocking synchronization}
+\glsadd{non-blocking synchronization}
 
@@
 \index{barrier synchronization}
+\glsadd{barrier synchronization}
 
@@
 A \emph{micro step}
-\index{micro step}
-\glsadd{micro step}
 is the execution of a single CXL machine instruction
@@
 \index{macro step}
-\glsadd{macro step}
 can involve multiple micro steps.  The following

### 2020-06-16 book/paper.tex
Added a chapter outline

@@
 
+The book proceeds as follows:
+
+\begin{itemize}
+\item In Chapter~\ref{ch:cxlintro} introduces the CXL programming
+language, as it provides the language for presenting synchronization
+problems and solutions.
+\item Chapter~\ref{ch:concurrent} introduces the problem of
+concurrent programming through a simple example in which two processes
+are concurrently incrementing a counter;
+\item Chapters~\ref{ch:cxlvalues} and~\ref{ch:cxlmachine} present some more
+details of the CXL language and machine to understand the problem
+underlying concurrency better.
+\item Chapter~\ref{ch:critical} introduces the concept of a critical
+section.
+\item Chapter~\ref{ch:naive} presents various flawed implementations
+of critical sections.
+\item Chapter~\ref{ch:peterson} introduces Peterson's Algorithm, an
+elegant solution to the mutual exclusion problem.
+\item Chapter~\ref{ch:method} gives some more details on the CXL
+language.
+\item Chapter~\ref{ch:spinlock} introduces hardware-supported locks
+for mutual exclusion.  Chapter~\ref{ch:synch} presents a CXL module
+that supports locks and other synchronization primitives.
+\item Chapters~\ref{ch:rdwrbusy} and~\ref{ch:rdwr2} present the
+abstraction and implementation of reader/writer
+locks, which allow multiple processes in a critical section if they are
+all only executing read-only accesses.
+Chapter~\ref{ch:rdwrbusy} also talks about busy-waiting, which is
+an undesirable approach to synchronize processes.
+\item Chapter~\ref{ch:deadlock} introduces the notion of \emph{deadlock}
+where a set of processes are indefinitely waiting for one another to
+release a resource.
+\item Chapter~\ref{ch:semaphore} introduces a generalization of locks
+that is good not only for mutual exclusion but also for waiting for
+certain application-level conditions.
+\item Chapters~\ref{ch:bb} and~\ref{ch:sbs} introduce the bounded buffer
+problem and various solutions.
+\item Chapter~\ref{ch:starvation} talks about the problem that in some
+synchronization solutions, processes may not be able to get access to a
+resource they need, and a solution for the reader/writer problem.
+\item Chapters~\ref{ch:monitors} and~\ref{ch:mesa} present monitors and
+condition variables, another approach to process synchronication.
+\item Chapter~\ref{ch:actor} presents the actor model as an approach to
+synchronization.
+\item Chapter~\ref{ch:nonblocking} shows solutions to process synchronization
+that prevent processes waiting for one another more than a bounded number of
+steps.
+\item Chapter~\ref{ch:barrier} introduces barrier synchronization, useful in
+high-performance computing applications such as parallel simulations.
+\item Chapter~\ref{ch:abp} presents a problem and a solution to the distributed
+systems problem of having two processes communicate over an unreliable network.
+\end{itemize}
+
 \chapter{Introduction to Programming with CXL}
+\label{ch:cxlintro}
 
@@
 \chapter{The Problem of Concurrent Programming}
+\label{ch:concurrent}
 
@@
 \chapter{Critical Sections}
+\label{ch:critical}
 
@@
 \chapter{Na\"{\i}ve Attempts at Implementing Critical Sections}
+\label{ch:naive}
 
@@
 \chapter{CXL Methods and Pointers}
+\label{ch:method}
 \index{CXL method}
@@
 \chapter{Spinlock}
+\label{ch:spinlock}
 \index{spinlock}
@@
 \chapter{Locks and the Synch Module}
+\label{ch:synch}
 \index{lock}
@@
 \chapter{Reader/Writer Locks using Busy Waiting}
+\label{ch:rdwrbusy}
 \index{reader/writer lock}
@@
 \chapter{Reader/Writer Lock, Part 2}
+\label{ch:rdwr2}
 
@@
 \chapter{Deadlock}
+\label{ch:deadlock}
 \index{deadlock}
@@
 \chapter{Semaphores}
+\label{ch:semaphore}
 \index{semaphore}
@@
 \chapter{Bounded Buffer}
+\label{ch:bb}
 \index{bounded buffer}
@@
 \chapter{Split Binary Semaphores}
+\label{ch:sbs}
 \index{split binary semaphore}
@@
 \chapter{Mesa Condition Variables in CXL}
+\label{ch:mesa}
 
@@
 \chapter{Actors and Synchronized Queues}
+\label{ch:actor}
 \index{actor model}
@@
 \chapter{Non-Blocking Synchronization}
+\label{ch:nonblocking}
 \index{non-blocking synchronization}
@@
 \chapter{Barrier Synchronization}
+\label{ch:barrier}
 \index{barrier synchronization}
@@
 \chapter{Networking: Alternating Bit Protocol}
+\label{ch:abp}
 \index{alternating bit protocol}

### 2020-06-16 book/paper.tex
Starting a chapter on deadlock prevention

@@
 \item Chapter~\ref{ch:critical} introduces the concept of a critical
-section.
-\item Chapter~\ref{ch:naive} presents various flawed implementations
-of critical sections.
+section, and Chapter~\ref{ch:naive} presents various flawed implementations
+of critical sections to demonstrate that implementing a critical section
+is not trivial.
 \item Chapter~\ref{ch:peterson} introduces Peterson's Algorithm, an
-elegant solution to the mutual exclusion problem.
+elegant implementation of a critical section.
 \item Chapter~\ref{ch:method} gives some more details on the CXL
@@
 \item Chapter~\ref{ch:spinlock} introduces hardware-supported locks
-for mutual exclusion.  Chapter~\ref{ch:synch} presents a CXL module
+for implemented critical sections.  Chapter~\ref{ch:synch} presents a CXL module
 that supports locks and other synchronization primitives.
@@
 synchronization solutions, processes may not be able to get access to a
-resource they need, and a solution for the reader/writer problem.
+resource they need, and a solution applied to the reader/writer problem.
 \item Chapters~\ref{ch:monitors} and~\ref{ch:mesa} present monitors and
@@
 \item Chapter~\ref{ch:abp} presents a problem and a solution to the distributed
-systems problem of having two processes communicate over an unreliable network.
+systems problem of having two processes communicate reliably over an unreliable
+network.
 \end{itemize}
@@
 
+\chapter{Deadlock Prevention}
+\label{ch:deadlockprevention}
+
+As introduced in Chapter~\ref{ch:deadlock}, deadlock is the situation when there are
+a collection of processes waiting for one another to release one or more resources.
+There are four conditions that must hold for deadlock to occur~\cite{CES71}:
+\begin{enumerate}
+\item \emph{Mutual Exclusion}: each resource can only be used by one process at a time:
+\item \emph{Hold and Wait}: each process holds resources it already allocated while it
+waits for other resources that it needs;
+\item \emph{No Preemption}: Resources cannot be forcibly taken back from processes that
+allocated them;
+\item \emph{Circular Wait}: There exists a directed circular chain of processes, each waiting
+to allocate a resource held by the next.
+\end{enumerate}
+
+Preventing deadlock thus means preventing that one of these conditions occurs.
+However, mutual exclusion is not easily prevented (although, for some resources it is
+possible, as demonstrated in Chapter~\ref{ch:nonblocking}).
+Havender~\cite{Havender68} proposed the following techniques that avoid the remaining
+three conditions:
+
+\begin{itemize}
+\item \emph{No Hold and Wait}: a process must request all resources it is going to
+need at the same time;
+\item \emph{Preemption}: if a process is denied a request for a resource, it must
+release all resources that it has already acquired and start over;
+\item \emph{No Circular Wait}: define an ordering on all resources and allocate
+resources in a particular order.
+\end{itemize}
+
+We suggested the last technique to address the deadlock issue in the Dining
+Philosophers problem (Chapter~\ref{ch:deadlock}).
+
 \chapter{Actors and Synchronized Queues}

### 2020-06-16 book/paper.tex
Latest

@@
 
+\begin{figure}
+\begin{code}
+\VerbatimInput[xleftmargin=5mm,numbers=left]{code/DinersCV.cxl}
+\end{code}
+\caption{[DinersCV.cxl] Dining Philosophers that grab both forks at the same time.}
+\label{fig:dinerscv}
+\end{figure}
+
+\begin{figure}
+\begin{code}
+\VerbatimInput[xleftmargin=5mm,numbers=left]{code/DinersAvoid.cxl}
+\end{code}
+\caption{[DinersAvoid.cxl] Dining Philosophers that carefully avoid getting into a deadlock
+scenario.}
+\label{fig:dinersavoid}
+\end{figure}
+
 As introduced in Chapter~\ref{ch:deadlock}, deadlock is the situation when there are
@@
 Philosophers problem (Chapter~\ref{ch:deadlock}).
+
+To implement a \emph{No Hold and Wait} solution, a philosopher would need a
+way to lock both the left and right forks at the same time.  Locks do not
+have such an ability, and neither do semaphores, so we re-implement the
+Dining Philosophers using condition variables that allow one to wait for
+arbitrary application-specific conditions.
 

### 2020-06-16 book/paper.tex
Latest

@@
 condition variables, another approach to process synchronication.
+\item Chapter~\ref{ch:deadlockprevention} revisits deadlocks and how to
+avoid them.
 \item Chapter~\ref{ch:actor} presents the actor model as an approach to
@@
 \label{ch:deadlockprevention}
+\index{deadlock prevention}
 
@@
 \label{fig:dinerscv}
-\end{figure}
-
-\begin{figure}
-\begin{code}
-\VerbatimInput[xleftmargin=5mm,numbers=left]{code/DinersAvoid.cxl}
-\end{code}
-\caption{[DinersAvoid.cxl] Dining Philosophers that carefully avoid getting into a deadlock
-scenario.}
-\label{fig:dinersavoid}
 \end{figure}
@@
 way to lock both the left and right forks at the same time.  Locks do not
-have such an ability, and neither do semaphores, so we re-implement the
+have such an ability, and neither do semaphores. so we re-implement the
 Dining Philosophers using condition variables that allow one to wait for
 arbitrary application-specific conditions.
+
+Figure~\ref{fig:dinerscv} demonstrates how this might be done.
+We use a single mutex for the diners, and, for each fork, a boolean
+and a condition variable.  The boolean indicates if the fork has been
+taken.
+Each diner waits if either the left or right fork is already taken.
+But which condition variable to wait on?
+The code demonstrates an important technique to use when waiting for multiple
+conditions.
+The condition in the \texttt{while} statement is the condition that the
+diner is waiting for to be \texttt{False}.  Within the \texttt{while} statement,
+there is an \texttt{if} statement for each disjunct of that condition.
+The code waits for either or both forks if necessary.  After that, it goes
+back to the top of the \texttt{while} loop.
+
+A common mistake is to write the following code instead:
+
+\begin{code}
+\begin{Verbatim}
+while forks[left]:
+    call wait &(conds[left]);
+;
+while forks[right]:
+    call wait &(conds[right]);
+;
+\end{Verbatim}
+\end{code}
+
+\begin{quote}
+\begin{itemize}
+\item Can you see why this does not work?  What can go wrong?
+\end{itemize}
+\end{quote}
+
+The next approach suggested by Havender is to allow processes to back out.
+While this could be done, this invariably leads to a busy waiting solution
+where a process keeps obtaining locks and releasing them again until it
+finally is able to get all of them.
+The approach to order the resources was already presented in Chapter~\ref{ch:deadlock}
+
+There is, however, another approach. which is sometimes called
+\emph{deadlock avoidance}.
+\index{deadlock avoidance}.
+
+\begin{figure}
+\begin{code}
+\VerbatimInput[xleftmargin=5mm,numbers=left]{code/DinersAvoid.cxl}
+\end{code}
+\caption{[DinersAvoid.cxl] Dining Philosophers that carefully avoid getting into a deadlock
+scenario.}
+\label{fig:dinersavoid}
+\end{figure}
 

### 2020-06-17 book/paper.tex
Latest

@@
 
-We suggested the last technique to address the deadlock issue in the Dining
-Philosophers problem (Chapter~\ref{ch:deadlock}).
-
 To implement a \emph{No Hold and Wait} solution, a philosopher would need a
@@
 But which condition variable to wait on?
-The code demonstrates an important technique to use when waiting for multiple
-conditions.
+The code demonstrates an important technique to use when waiting for
+multiple conditions.
+\index{multiple conditions, waiting on}
 The condition in the \texttt{while} statement is the condition that the
@@
 \item Can you see why this does not work?  What can go wrong?
+\item Run it through CXL in case you are not sure!
 \end{itemize}
@@
 
-The next approach suggested by Havender is to allow processes to back out.
+The \texttt{Preemption} approach suggested by Havender is to allow processes to back out.
 While this could be done, this invariably leads to a busy waiting solution
@@
 finally is able to get all of them.
-The approach to order the resources was already presented in Chapter~\ref{ch:deadlock}
+The \emph{No Circular Waiting} approach to order the resources was already presented in
+Chapter~\ref{ch:deadlock}
 
@@
 \index{deadlock avoidance}.
+In the case of the Dining Philosophers, we want to avoid the situation where each
+diner picks up a fork.  So if we can prevent more than four diners from starting to
+eat at the same time, then we can avoid the conditions for deadlock from ever
+happening.
+Figure~\ref{fig:dinersavoid} demonstrates this concept.  It uses a semaphore to
+restrict the number of diners at any time to four.
+
+This concept can be generalized using something called the
+Banker's Algorithm~\cite{EWD108}, but it is outside the scope of this book.
+The problem with these kinds of schemes is that one needs to know ahead of time
+what the maximum number of resources is that each process wants to allocate,
+making them generally quite impractical.
 

### 2020-06-17 book/paper.tex
Latest
Latest version

@@
 
-\chapter{Reader/Writer Lock, Part 2}
+\chapter{Reader/Writer Locks, Part 2}
 \label{ch:rdwr2}
@@
 
-We will discuss ways of preventing deadlock later, but one important
-approach to avoiding deadlock is to establish an ordering among the
+We will discuss ways of preventing deadlock later (Chapter~\ref{ch:deadlockprevention},
+but one important
+approach to preventing deadlock is to prevent a cycle from forming, with each
+process waiting for the next process on the cycle.
+We can do this by establishing an ordering among the
 resources (in this case the forks) and, when needing more than one

### 2020-06-17 book/paper.tex
typo

@@
 
-We will discuss ways of preventing deadlock later (Chapter~\ref{ch:deadlockprevention},
+We will discuss ways of preventing deadlock later (Chapter~\ref{ch:deadlockprevention}),
 but one important

### 2020-06-17 book/paper.tex
Started working on adding exercises to chapters

@@
   name=machine instruction,
-  description={is an atomic operation on the CXL machine, executed by a process}
+  description={is an atomic operation on the CXL virtual machine, executed by a process}
 }
@@
   name=shared variable,
-  description={is a variable that is stored in the memory of the CXL machine and
+  description={is a variable that is stored in the memory of the CXL virtual machine and
       shared between multiple processes, as opposed to a process variable}
@@
   name=state,
-  description={an assignment of values to variables.  In a CXL machine, this includes
+  description={an assignment of values to variables.  In a CXL virtual machine, this includes
       the contents of its shared memory and the set of contexts}
@@
 are concurrently incrementing a counter;
-\item Chapters~\ref{ch:cxlvalues} and~\ref{ch:cxlmachine} present some more
-details of the CXL language and machine to understand the problem
+\item Chapter~\ref{ch:cxlmachine} presents the
+CXL virtual machine to understand the problem
 underlying concurrency better.
@@
 
-\begin{quote}
-Experiment!
-\begin{itemize}
+\section*{Exercises}
+\begin{enumerate}
 \item See what happens if, instead of initializing \texttt{result} to 0,
-you initialize it to 1.
-\item Write a similar program that computes squares or factorials.
-\end{itemize}
-\end{quote}
+you initialize it to 1.  (You do not need to understand the error report at this time.)
+\item Write a program that computes factorials.
+\end{enumerate}
 
@@
 \begin{quote}
-What will happen?
 \begin{itemize}
@@
 \begin{quote}
-Can you think of a fix to the program?  You can try it out and see
+\begin{itemize}
+\item Can you think of a fix to the program?  You can try it out and see
 if it works or not.
+\end{itemize}
 \end{quote}
 
-\chapter{CXL Values}
-\label{ch:cxlvalues}
-\index{CXL Value}
-
-CXL programs manipulate CXL values.
+\section*{Exercises}
+
+\begin{figure}
+\begin{code}
+\VerbatimInput[xleftmargin=5mm,numbers=left]{code/Up.py}
+\end{code}
+\caption{[Up.py] Python implementation of Figure~\ref{fig:inc}.}
+\label{fig:incpy}
+\end{figure}
+
+\begin{figure}
+\begin{code}
+\VerbatimInput[xleftmargin=5mm,numbers=left]{code/UpMany.py}
+\end{code}
+\caption{[UpMany.py] Incrementing \texttt{N} times.}
+\label{fig:incmany}
+\end{figure}
+
+\begin{enumerate}
+\item CXL programs can usually be easily translated into Python.  For example,
+Figure~\ref{fig:incpy} is a Python version of Figure~\ref{fig:inc}.
+\begin{enumerate}
+\item Run Figure~\ref{fig:incpy} using Python.  Does the assertion fail?
+\item Using a script, run Figure~\ref{fig:inc} 1000 times.  How many times does the assertion
+fail (if any).
+\end{enumerate}
+\item Figure~\ref{fig:incmany} is a version of Figure~\ref{fig:incpy} that has each
+incrementer thread increment \texttt{count} \texttt{N} times.  Run Figure~\ref{fig:incpy}
+10 times (using Python).
+Report how many times the assertion fails and what the value of \texttt{count}
+was for each of the failed runs.
+\end{enumerate}
+
+\chapter{The CXL Virtual Machine}
+\label{ch:cxlmachine}
+\index{CXL virtual machine}
+
+CXL programs are compiled to CXL bytecode
+\index{bytecode}
+(machine instructions for a virtual machine),
+which in turn is executed by the CXL virtual machine (CVM).
+\index{virtual machine}
+\index{CXL Virtual Machine}
+\index{CVM}
+To understand the problem of concurrent computing, it
+is important to have a basic understanding of the CVM.
+
+\section*{CXL Values}
+
+CXL programs, and indeed the CVM,  manipulate CXL values.
 CXL values are recursively defined:
@@
 
-\chapter{The CXL Machine}
-\label{ch:cxlmachine}
-\index{CXL machine}
+\section*{CXL Bytecode}
 
@@
 \end{code}
-\caption{The first part of the machine code corresponding to Figure~\ref{fig:inc}.}
+\caption{The first part of the CVM bytecode corresponding to Figure~\ref{fig:inc}.}
 \label{fig:inccode}
@@
 
-Before we delve into how to solve synchronization problems, it is important
-to know a bit about the underlying machine.  A CXL program is translated
-into a list of machine instructions that the CXL machine executes.
-The CXL machine is not an ordinary virtual machine, but its architecture
+A CXL program is translated
+into \emph{bytecode}, which is list of machine instructions that the
+CXL virtual machine (CVM) executes.
+The CVM is not an ordinary virtual machine, but its architecture
 is nonetheless representative of conventional computers and
@@
 
-Instead of bits and bytes, a CXL machine manipulates CXL values.
-A CXL machine has the following components:
+Instead of bits and bytes, a CVM manipulates CXL values.
+A CVM has the following components:
 \begin{itemize}
-\item Code:  This is an immutable and finite list of CXL machine instructions,
+\item Code:  This is an immutable and finite list of CVM instructions,
 generated from a CXL program.  The types of instructions will be described later.
-\item Shared memory: A CXL machine has just one memory location containing
+\item Shared memory: A CVM has just one memory location containing
 an CXL value.
@@
 its name tag, program counter, stack, and register.
-The state of the CXL machine
+The state of a CVM
 consists of the value of its memory and the multiset (or \emph{bag})
@@
 
-Compiling the code in Figure~\ref{fig:inc} results in the CXL machine code
+Compiling the code in Figure~\ref{fig:inc} results in the CVM bytecode
 listed in Figure~\ref{fig:inccode}.
@@
 \end{code}
-The CXL machine is predominantly a \emph{stack machine}.
+Each process in the CVM is predominantly a \emph{stack machine},
 \index{stack machine}
 \glsadd{stack machine}
+but it also has a register.
 All instructions are atomically executed.
@@
 \item The name tag of the process;
-\item The sequence of program counters of the CXL machine instructions that the process executed;
+\item The sequence of program counters of the CVM instructions that the process executed;
 \item The current program counter of the process;
@@
 entering and exiting the critical section.
-Figure~\ref{fig:csbarebones} shows the CXL code.
+Figure~\ref{fig:csbarebones} shows the CVM bytecode.
 Here \texttt{@cs} is a \emph{label},
 \index{label}
-identifying a location in the CXL machine code.  The first thing we need to
+identifying a location in the CVM bytecode.  The first thing we need to
 ensure is that there can never be two processes at the critical section.
@@
 \glsadd{state}
-at any point of time the CXL machine is in a specific state.
+at any point of time the CVM is in a specific state.
 A state is comprised of the values of the shared variables as well as
@@
 program counter and the contents of its stack.
-Everytime a process executes a CXL machine instruction, the
+Everytime a process executes a CVM machine instruction, the
 state changes (if only because the program counter of the process
@@
 
-The CXL machine starts in an initial state in which there is only
+The CVM starts in an initial state in which there is only
 one process and its program counter is~0.  A \emph{trace}
@@
 You may note that all this looks familiar.  Indeed, the syntax
-is the same as that for dictionaries (see Chapter~\ref{ch:cxlvalues}).
+is the same as that for dictionaries (see Chapter~\ref{ch:cxlmachine}).
 Both dictionaries and methods map CXL values to CXL values,
@@
 
-While the CXL machine does not have any built-in interlock instructions,
+While the CVM does not have any built-in interlock instructions,
 it does have support for executing multiple instructions atomically.
@@
 
-Chapter~\ref{ch:cxlvalues} provides an introduction to CXL values.
+Chapter~\ref{ch:cxlmachine} provides an introduction to CXL values.
 Below is a complete list of CXL value types with examples:
@@
 
-\chapter{The CXL Machine}
-
-The CXL machine has the following state:
+\chapter{The CXL Virtual Machine}
+
+The CXL Virtual Machine (CVM) has the following state:
 
@@
 \hline
-code & a list of CXL machine instructions \\
+code & a list of CVM machine instructions \\
 \hline
@@
 A \emph{micro step}
-is the execution of a single CXL machine instruction
+is the execution of a single CVM machine instruction
 by a context.
 Each micro step generates a new state.
-When there are multiple contexts, the CXL machine can interleave them.
+When there are multiple contexts, the CVM can interleave them.
 However, trying to interleave every microstep would be needlessly expensive,
@@
 instructions start a new macro step: \texttt{Load}, \texttt{Store},
-\texttt{AtomicInc}, and \texttt{Continue}.  The CXL machine
+\texttt{AtomicInc}, and \texttt{Continue}.  The CVM
 interleaves macro steps, not micro steps.  Like micro steps, each

### 2020-06-17 book/paper.tex
Latest

@@
 
-\begin{quote}
-\begin{itemize}
-\item Can you think of a fix to the program?  You can try it out and see
-if it works or not.
-\end{itemize}
-\end{quote}
-
 \section*{Exercises}
+
+The following exercises are intended to show you that on the one hand it
+is easy to translate CXL programs into Python, but on the other hand,
+it is easier to find concurrency bugs using CXL.
 
@@
 \item Run Figure~\ref{fig:incpy} using Python.  Does the assertion fail?
-\item Using a script, run Figure~\ref{fig:inc} 1000 times.  How many times does the assertion
+\item Using a script, run Figure~\ref{fig:incpy} 1000 times.  How many times does the assertion
 fail (if any).
@@
 was for each of the failed runs.
+\item Can you think of a fix to Figure~\ref{fig:inc}?  Try one or two different fixes
+and run them through CXL.  Do not worry about having to come up with a correct fix at this
+time---the important thing is to develop an understanding of concurrency.
 \end{enumerate}
@@
 
-\chapter{Reader/Writer Locks, Part 2}
+\chapter{Reader/Writer Locks with Blocking}
 \label{ch:rdwr2}

### 2020-06-17 book/paper.tex
Working on exercises

@@
 
-Figure~\ref{fig:triangle} gives a simple example of a CXL program.
+Figure~\ref{fig:triangle} shows a simple example of a CXL program.
 (The code for examples in this book can be found in the \texttt{code} folder under
@@
 you initialize it to 1.  (You do not need to understand the error report at this time.)
-\item Write a program that computes factorials.
+\item Write a CXL program that computes squares by repeated adding.  So the program
+should compute the square of $x$ by adding $x$ to an initial value of 0 $x$ times.
 \end{enumerate}
@@
 This makes precise the concurrency problem that we encountered.
+
+\section*{Exercises}
+
+\begin{figure}
+\begin{code}
+\VerbatimInput[xleftmargin=5mm,numbers=left]{code/UpEnter.cxl}
+\end{code}
+\caption{[UpEnter.py] Broken attempt at fixing the code of Figure~\ref{fig:inc}.}
+\label{fig:incenter}
+\end{figure}
 

### 2020-06-17 book/paper.tex
Added some exercises

@@
 
+\begin{enumerate}
+\item Figure~\ref{fig:incenter} shows an attempt at trying to fix the code of
+    Figure~\ref{fig:inc}.  Run it through CXL and see what happens.  Based on
+    the error output, Describe in English what is wrong with the code by describing,
+    in broad steps, how running the program can get into a bad state.
+\item What if we moved line~2 of Figure~\ref{fig:incenter} to after the \texttt{if}
+    statement (between lines~7 and~8)?  Do you think that would work?  Run it through
+    CXL and describe either why it works or why it does not work.
+\end{enumerate}
+
 \begin{figure}

### 2020-06-17 book/paper.tex
Latest

@@
 you initialize it to 1.  (You do not need to understand the error report at this time.)
+Explain in English what the problem is.
 \item Write a CXL program that computes squares by repeated adding.  So the program
@@
 (aka critical region)~\cite{EWD123}:
-\index{critical section}
-\glsadd{critical section}
-\index{critical region}
 a set of instructions where only one process is allowed to execute at a
 time.
+\index{critical section}
+\index{critical region}
+\glsadd{critical section}
 

### 2020-06-17 book/paper.tex
Latest

@@
 \section*{CXL Values}
+\label{ap:cxlvalues}
 
@@
 this set.
+
+Appendix~\ref{ap:cxlvalues} provides details on all the values that
+CXL currently supports.
 
@@
 
+Appendix~\ref{ap:cxlbytecode} provides a list of all CVM machine instructions,
+in case you want to read about the details.
 The \texttt{Frame} instruction has three arguments: the name of the method,
@@
 There are various possible liveness properties we could use,
-but here we will propose the following informally: if some set of processes
-$S$ are trying to enter the critical section and any process already in the
-intersection eventually leaves, then eventually one process in $S$ will enter
-the critical section.
+but here we will propose the following informally: if there exists a non-empty
+set $S$ of processes that are trying to enter the critical section and any
+process already in the intersection eventually leaves, then eventually one
+process in $S$ will enter the critical section.
 We call this \emph{progress}.
@@
 without entering the critical section ever.
-
-\begin{quote}
-Try it out!
-\begin{itemize}
-\item Plug \texttt{while True: pass;;} for entering the critical section
-in Figure~\ref{fig:cs} and run CXL.  It should print a trace
-to a state from which a terminating state cannot be reached.
-\end{itemize}
-\end{quote}
 
@@
 section cannot terminate.
+
+\section*{Exercises}
+\begin{enumerate}
+\item Run Figure~\ref{fig:cs} using CXL.  As there is no protection of the critical
+section, mutual is violated, the assertion should fail, and a trace should be reported.
+Now plug \texttt{while True: pass;;} before entering the critical section
+in Figure~\ref{fig:cs} and run CXL again.  Now mutual exclusion is guaranteed but
+progress is violated.  CXL should print a trace
+to a state from which a terminating state cannot be reached.
+Describe in English the difference in the failure reports.
+\item If we change the loop in Figure~\ref{fig:cs} into an infinite loop,
+does the \emph{progress} property guarantee that all processes that are trying
+to enter eventually enter?  Why or why not?
+\end{enumerate}
 
@@
 \chapter{List of Machine Instructions}
+\label{ap:cxlbytecode}
 
@@
 
+\newpage
 \printindex
 
+\newpage
 \printglossaries

### 2020-06-17 book/paper.tex
Latest

@@
 \item Chapter~\ref{ch:critical} introduces the concept of a critical
-section, and Chapter~\ref{ch:naive} presents various flawed implementations
+section presents various flawed implementations
 of critical sections to demonstrate that implementing a critical section
@@
 \item Chapter~\ref{ch:peterson} introduces Peterson's Algorithm, an
-elegant implementation of a critical section.
+elegant solution to implementating a critical section.
 \item Chapter~\ref{ch:method} gives some more details on the CXL
@@
 
-\section*{Exercises}
-\begin{enumerate}
-\item Run Figure~\ref{fig:cs} using CXL.  As there is no protection of the critical
-section, mutual is violated, the assertion should fail, and a trace should be reported.
-Now plug \texttt{while True: pass;;} before entering the critical section
-in Figure~\ref{fig:cs} and run CXL again.  Now mutual exclusion is guaranteed but
-progress is violated.  CXL should print a trace
-to a state from which a terminating state cannot be reached.
-Describe in English the difference in the failure reports.
-\item If we change the loop in Figure~\ref{fig:cs} into an infinite loop,
-does the \emph{progress} property guarantee that all processes that are trying
-to enter eventually enter?  Why or why not?
-\end{enumerate}
-
-\chapter{Na\"{\i}ve Attempts at Implementing Critical Sections}
-\label{ch:naive}
-
 \begin{figure}
@@
 
-\begin{quote}
-Try your own solutions!
-\begin{itemize}
+\section*{Exercises}
+\begin{enumerate}
+\item Run Figure~\ref{fig:cs} using CXL.  As there is no protection of the critical
+section, mutual is violated, the assertion should fail, and a trace should be reported.
+Now plug \texttt{while True: pass;;} before entering the critical section
+in Figure~\ref{fig:cs} and run CXL again.  Now mutual exclusion is guaranteed but
+progress is violated.  CXL should print a trace
+to a state from which a terminating state cannot be reached.
+Describe in English the difference in the failure reports.
+\item If we change the loop in Figure~\ref{fig:cs} into an infinite loop,
+does the \emph{progress} property guarantee that all processes that are trying
+to enter eventually enter?  Why or why not?
 \item See if you can come up with some different approaches that satisfy both
@@
 we will find out, it is possible but not easy.
-\end{itemize}
-\end{quote}
+\end{enumerate}
 
@@
 predicate is not invariant, and therefore not an inductive invariant either.
-
-To finish the story, the author of CXL fixed the Wikipedia page, but there
-is another cool story.  Years later a colleague of the author, teaching
-the same course, asked if the first two assignments (setting \texttt{flags[self]})
+To finish the story, the author of CXL fixed the Wikipedia page.
+
+\begin{figure}
+\begin{code}
+\VerbatimInput[xleftmargin=5mm,numbers=left]{code/csonebit.cxl}
+\end{code}
+\caption{[csonebit.cxl] Mutual exclusion using a flag per process.}
+\label{fig:csonebit}
+\end{figure}
+
+\section*{Exercises}
+\begin{enumerate}
+\item Figure~\ref{fig:csonebit} presents another correct solution to the
+mutual exclusion problem.  It is similar to the one in
+Figure~\ref{fig:upflags}, but has a process \emph{back out and try again}
+if it finds that the other process is either trying to enter the critical
+section or already has.  Compare this algorithm with Peterson's.  What are
+the advantages and disadvantages?
+\item A colleague of the author asked if the first two assignments in
+Peterson's algorithm (setting \texttt{flags[self]})
 to \texttt{True} and \texttt{turn} to \texttt{1 - self}) can be reversed.
@@
 sequential program one could surely swap the two assignments.
-
-\begin{quote}
-What do you think?
-\begin{itemize}
-\item See if you can figure out for yourself if the two assignments can be
+See if you can figure out for yourself if the two assignments can be
 reversed.  Then run the program in Figure~\ref{fig:peterson} after reversing
-the two assignments and see what happens.
-\end{itemize}
-\end{quote}
-
-\begin{quote}
-Bonus question!
-\begin{itemize}
-\item Can you generalize Peterson's algorithm to more than two processes?
-\end{itemize}
-\end{quote}
+the two assignments and describe in English what happens.
+\item Bonus question:
+Can you generalize Peterson's algorithm to more than two processes?
+\end{enumerate}
 
@@
 
-\begin{quote}
-\begin{itemize}
+\section*{Exercises}
+\begin{enumerate}
 \item Why do you suppose the code in Figure~\ref{fig:rwmulti} uses a flag
@@
 reader/writer locks can be instantiated.
-\end{itemize}
-\end{quote}
+\end{enumerate}
 

### 2020-06-18 book/paper.tex
Latest
Added an exercise

@@
 \item Bonus question:
+Can you find an inductive invariant for Figure~\ref{fig:csonebit} to prove
+it correct?
+\item Bonus question:
 Can you generalize Peterson's algorithm to more than two processes?
@@
 has the code to check $I_2(i)$.  But how would one go about checking an
-invariant like $I_1$.  Invariants must hold for every state.
+invariant like $I_1$?  Invariants must hold for every state.
 For $I_2$ we only need an assertion at label \texttt{@cs} because the
@@
 
-\begin{quote}
-Try doing mutual exclusion with other interlock instructions!
-\begin{itemize}
-\item Find a description for \texttt{compare-and-swap} (CAS)
-\index{compare-and-swap}
-\index{CAS}
-on the internet and
-update Figure~\ref{fig:tas}, replacing \texttt{tas}.
-\item Other interesting interlock instructions include \texttt{swap} and
-\texttt{fetch-and-add}.
-\index{fetch-and-add}
-\end{itemize}
-\end{quote}
+\section*{Exercises}
+\begin{enumerate}
+\item Implement an atomic swap operation.  It should take two pointer arguments
+and swap the values.
+\item Implement a spinlock using the atomic swap operation.
+\item Write out the invariants that need to hold and check them using CXL.
+\end{enumerate}
 
@@
 In CXL, atomic statements allow you to \emph{implement} your own
-low synchronization primitives like test-and-set.  Atomic statements
+synchronization primitives like test-and-set.  Other common examples
+include compare-and swap and fetch-and-add.  Atomic statements
 are not intended to \emph{replace} locks or other synchonization primitives.
+When solving synchronization problems you should not directly use
+atomic statement but use the synchronization primitives that are available
+to you.  But if you want to design a new synchronization primitive, then
+use \texttt{atomic} by all means.
 
@@
 
+\section*{Exercises}
+\begin{enumerate}
+\item Write a synchronization module using your own implementation of \texttt{Lock}
+(for example, using Peterson's algorithm or using \texttt{swap}), and run the code in
+Figure~\ref{fig:incfixed} using your own module.
+\end{enumerate}
+
 \chapter{Reader/Writer Locks using Busy Waiting}
@@
 else need two forks to eat a meal?) that we call Italian Philosophers (who only
-need one fork each to eat spaghetti).  Again, there are five philosophers.
+use one fork each to eat spaghetti---they need the other to speak\footnote{The joke
+was provided to me by my dear friend and colleague Lorenzo Alvisi.}).
+Again, there are five philosophers.
 This time, however, there is a glass sitting in the middle of the table with only

### 2020-06-18 book/paper.tex
Latest version

@@
 
-\begin{figure}
-\begin{code}
-\VerbatimInput[xleftmargin=5mm,numbers=left]{code/csonebit.cxl}
-\end{code}
-\caption{[csonebit.cxl] Mutual exclusion using a flag per process.}
-\label{fig:csonebit}
-\end{figure}
-
 \section*{Exercises}
@@
 the advantages and disadvantages?
+\item
+Can you find an inductive invariant for the algorithm in
+Figure~\ref{fig:csonebit} to prove it correct?
+Here's a pseudo-code version of the algorithm to help you.  Each line
+is an atomic action:
+\begin{code}
+\begin{verbatim}
+    initially: flagX = flagY = False
+
+    process X:                          process Y:
+        X0: flagX = True                   Y0: flagY = True
+        X1: if not flagY goto X4           Y1: if not flagX goto Y4
+        X2: flagX = False                  Y2: flagY = False
+        X3: goto X0                        Y3: goto Y0
+        X4: ...critical section...         Y4: ...critical section...
+        X5: flagX = False                  Y5: flagY = False
+\end{verbatim}
+\end{code}
 \item A colleague of the author asked if the first two assignments in
@@
 \item Bonus question:
-Can you find an inductive invariant for Figure~\ref{fig:csonebit} to prove
-it correct?
-\item Bonus question:
 Can you generalize Peterson's algorithm to more than two processes?
 \end{enumerate}
+
+\begin{figure}
+\begin{code}
+\VerbatimInput[xleftmargin=5mm,numbers=left]{code/csonebit.cxl}
+\end{code}
+\caption{[csonebit.cxl] Mutual exclusion using a flag per process.}
+\label{fig:csonebit}
+\end{figure}
 

### 2020-06-18 book/paper.tex
Latest

@@
 \begin{figure}
+\begin{center}
+\includegraphics[width=4in]{figures/actor-crop.pdf}
+\end{center}
+\caption{Depiction of three actors.  The producer does not receive messages.}
+\label{fig:actorpic}
+\end{figure}
+
+\begin{figure}
 \begin{code}
@@
 synchronization using so-called \emph{actors}~\cite{HBS73}.  Actors are
-processes that have only private memory and communicate through messages.
-Given that there is no shared memory in the actor model, there is no need
+processes that have only private memory and communicate through message queues.
+See Figure~\ref{fig:actorpic} for an illustration.
+Given that there is no shared memory in the actor model (other than the message
+queues, which have built-in synchronization), there is no need
 for critical sections.  Instead, some sequential process owns a particular
@@
 
-Figure~\ref{fig:actor} shows a solution to the producer/consumer problem using
-an actor.  \texttt{pc\_actor} is essentially a matchmaker: it matches
+Figure~\ref{fig:actor} shows a CXL solution to the producer/consumer problem using
+an actor (as illustrated in Figure~\ref{fig:actorpic}.
+\texttt{pc\_actor} is essentially a matchmaker: it matches
 \texttt{.produce} requests with \texttt{.consume} requests.

### 2020-06-19 book/paper.tex
Improved syntax

@@
 
-It is often the case that methods that update shared variables through
-pointers do not actually have to return a result to be useful.
-In CXL, if you want to invoke a method that does not return a result
-or if you are simply not interested in it,
-you have to use a \texttt{call} statement like so:
-\begin{code}
-\begin{verbatim}
-call m a;
-\end{verbatim}
-\end{code}
-
 Figure~\ref{fig:petersonmethods} again shows Peterson's algorithm,
@@
 count = 0;
-call atomic_inc &(count);
+atomic_inc &(count);
 \end{Verbatim}
@@
         if (^lk).locked:
-            call stop (^lk).suspended;
+            stop (^lk).suspended;
             assert (^lk).locked;
@@
     if left < right:
-        call lock &(forks[left]);
-        call lock &(forks[right]);
+        lock(&forks[left]);
+        lock(&(orks[right]);
     else:
-        call unlock &(forks[right]);
-        call unlock &(forks[left]);
+        unlock(&forks[right]);
+        unlock(&forks[left]);
     ;
@@
     while blocked:
-        call lock();
+        lock();
         if application-condition-holds:
@@
         else:
-            call unlock():
+            unlock():
 @cr: application-code;
-call unlock():
+unlock():
 \end{Verbatim}
@@
 def wait(c):
-    call V &(mutex);
-    call P c;
+    V(&mutex);
+    P(c);
 ;
 def signal(c):
-    call V c;
-    call P &(mutex);
+    V(c);
+    P(&mutex);
 ;
@@
     let lk = (^c).lock, blocked = True, cnt = bagCount((^c).waiters, nametag()):
-        call bagAdd(&((^c).waiters), nametag());
+        bagAdd(&(^c).waiters, nametag());
         ^lk = False;
@@
         if waiters != bagEmpty():
-            call bagRemove(&((^c).waiters), bagChoose(waiters));
+            bagRemove(&(^c).waiters, bagChoose(waiters));
         ;
@@
 while forks[left]:
-    call wait &(conds[left]);
+    wait(&conds[left]);
 ;
 while forks[right]:
-    call wait &(conds[right]);
+    wait(&conds[right]);
 ;
@@
 \texttt{del lv} & delete\\
-\hline
-\texttt{call e} & evaluate \texttt{e}\\
 \hline

### 2020-06-19 book/paper.tex
Starting to make machine code easier to read

@@
 Up.cxl:1 def incrementer(self):
-    0 Jump 39
-    1 Frame incrementer ['self'] 12
+   0 Jump 32
+   1 Frame incrementer ['self'] 9
 Up.cxl:2     count = count + 1;
    2 Push 1
-    3 PushAddress count
-    4 Load 1
-    5 2-ary +
-    6 PushAddress count
-    7 Store 1
+   3 Load count
+   4 2-ary +
+   5 Store count
 Up.cxl:3     done[self] = True;
-    8 Push True
-    9 PushVar self
-   10 PushAddress done
-   11 Store 2
-   12 Return
+   6 Push True
+   7 PushVar self
+   8 Store done 1
+   9 Return
 \end{verbatim}
@@
 \begin{verbatim}
+   ctx = __init__/() pc =   0 #states = 1 diameter = 0 queue = 0
 CXL Assertion failed ('assert', 'Up.cxl', 9, 5) 1
@@
 ==== Safety violation ====
-__init__/() [0,39-56]       57 dict{ .count:0, .done:dict{ 0:False, 1:False } }
-incrementer/0 [1-6]          7 dict{ .count:0, .done:dict{ 0:False, 1:False } }
-incrementer/1 [1-11]        12 dict{ .count:1, .done:dict{ 0:False, 1:True  } }
-incrementer/0 [7-11]        12 dict{ .count:1, .done:dict{ 0:True,  1:True  } }
-main/() [14-19,22-27,29-36] 37 dict{ .count:1, .done:dict{ 0:True,  1:True  } }
+__init__/() [0,32-47]       48 dict{ .count:0, .done:dict{ 0:False, 1:False } }
+incrementer/0 [1-4]          5 dict{ .count:0, .done:dict{ 0:False, 1:False } }
+incrementer/1 [1-8]          9 dict{ .count:1, .done:dict{ 0:False, 1:True  } }
+incrementer/0 [5-8]          9 dict{ .count:1, .done:dict{ 0:True,  1:True  } }
+main/() [11-15,18-22,24-29] 29 dict{ .count:1, .done:dict{ 0:True,  1:True  } }
 \end{verbatim}
@@
 onto the stack of the process.
-\item the \texttt{PushAddress} instruction pushes the address of
-the shared variable \texttt{count} onto the stack.
-\item The \texttt{Load} instruction pops the address of the
-\texttt{count} variable and then pushes the value of the
+\item The \texttt{Load} instruction pushes the value of the
 \texttt{count} variable onto the stack.
@@
 adds them and pushes the result back onto the stack.
-\item The \texttt{PushAddress} instruction pushes the address of the \texttt{count}
-variable onto the stack.
-\item The \texttt{Store} instruction pops the address of a variable and pops
-a CXL value (the sum of the \texttt{count} variable and 1) and updates the
-\texttt{count} variable.
+\item The \texttt{Store} instruction pops pops
+a CXL value (the sum of the \texttt{count} variable and 1) and
+stores it in the \texttt{count} variable.
 \end{enumerate}
@@
 If we look at the middle three rows, we see that:
-\begin{enumerate}
-\item Process 0 executed instructions 1 through 7, loading the value of
+\begin{itemize}
+\item Process 0 executed instructions 1 through 4, loading the value of
 \texttt{count} but stopping just before storing 1 into \texttt{count};
-\item Process 1 executed instructions 1 through 12, storing 1 into
+\item Process 1 executed instructions 1 through 8, storing 1 into
 into \texttt{count} and storing \texttt{True} into \texttt{done[1]};
@@
 and storing \texttt{True} into \texttt{done[0]}.
-\end{enumerate}
+\end{itemize}
 
@@
 \hline
-Load & load a shared variable and push it onto the stack \\
+Load $v$ & push shared variable $v$ (\texttt{\^} is indirect) onto the stack \\
 \hline
@@
 \hline
-Nary \textit{op} $n$ & apply $n$-ary operator \textit{op} to the top $n$ elements on the stack \\
+$n$-ary \textit{op} & apply $n$-ary operator \textit{op} to the top $n$ elements on the stack \\
 \hline
@@
 \hline
-Store & pop a value from the stack and store it in a shared variable \\
+Store $v$ $[n]$ & pop a value from the stack and store it in shared variable $v$ (\texttt{\^} is indirect) \\
 \hline

### 2020-06-19 cxl.py

@@ -430,42 +430,62 @@ class PushAddressOp(Op):
         context.pc += 1
 
 class LoadOp(Op):
-    def __init__(self, n):
-        self.n = n                  # number of indexes
+    def __init__(self, name):
+        self.name = name
 
     def __repr__(self):
-        return "Load " + str(self.n)
+        if self.name == None:
+            return "Load ^"
+        else:
+            (lexeme, file, line, column) = self.name
+            return "Load " + lexeme
 
     def eval(self, state, context):
-        indexes = []
-        for i in range(self.n):
-            indexes.append(context.pop())
-        av = indexes[0]
-        assert isinstance(av, AddressValue), (context.pc, indexes, state)
-        context.push(state.iget(av.indexes + indexes[1:]))
+        if self.name == None:
+            av = context.pop()
+            assert isinstance(av, AddressValue), (context.pc, indexes, state)
+            context.push(state.iget(av.indexes))
+        else:
+            (lexeme, file, line, column) = self.name
+            context.push(state.get(lexeme))
         context.pc += 1
 
 class StoreOp(Op):
-    def __init__(self, n):
-        self.n = n                  # number of indexes
+    def __init__(self, v, n):
+        self.v = v       # shared variable (None if indirect)
+        self.n = n       # number of indexes
 
     def __repr__(self):
-        return "Store " + str(self.n)
+        if self.v == None:
+            name = "^"
+        else:
+            (name, file, line, column) = self.v
+        if self.n > 0:
+            return "Store " + name + " " + str(self.n)
+        else:
+            return "Store " + name
 
     def eval(self, state, context):
         indexes = []
         for i in range(self.n):
             indexes.append(context.pop())
-        av = indexes[0]
-        # TODO.  May need to produce a nicer error message for this:
-        assert isinstance(av, AddressValue), indexes
+        if self.v == None:
+            av = indexes[0]
+            # TODO.  May need to produce a nicer error message for this:
+            assert isinstance(av, AddressValue), indexes
+            lv = av.indexes
+            name = lv[0]
+            indexes = indexes[1:]
+        else:
+            (name, file, line, column) = self.v
+            lv = [name]
 
-        if not state.initializing and (av.indexes[0] not in state.vars.d):
+        if not state.initializing and (name not in state.vars.d):
             print()
[…]

### 2020-06-19 book/paper.tex
Latest

@@
 to shared variables.
-If \texttt{x} is a shared variable, then the expression \texttt{\&(x)}
-(the parentheses are mandatory) is a pointer to \texttt{x}
-(also known as the \emph{address}
+If \texttt{x} is a shared variable, then the expression \texttt{\&x}
+ is a pointer to \texttt{x} (also known as the \emph{address}
 \index{address}
@@
 expression \texttt{\^{}p} is the value of the shared variable.
+In particular, \texttt{\^{}\&x == x}.
+This is similar to how C pointers work.
 
@@
 This would
-make the code re-usable by other applications.  For those who are
+make the code re-usable by other applications.
+
+For those who are
 extra adventurous: you can add the \texttt{P\_enter} and
@@
 count = 0;
-atomic_inc &(count);
+atomic_inc(&count);
 \end{Verbatim}

### 2020-06-19 book/paper.tex
Improved figure
Latest

@@
    0 Jump 32
-   1 Frame incrementer ['self'] 9
+   1 Frame incrementer(self) end=9
 Up.cxl:2     count = count + 1;
@@
 in case you want to read about the details.
-The \texttt{Frame} instruction has three arguments: the name of the method,
-a list containing the names of the arguments of the method, and
+The \texttt{Frame} instruction lists the name of the method,
+names of its arguments, and
 the program counter of the method's \texttt{Return} instruction.
@@
 \begin{center}
-\includegraphics[width=4in]{figures/states-crop.pdf}
+\includegraphics[width=6in]{figures/states-crop.pdf}
 \end{center}
@@
 \hline
-Address & \texttt{\&(lock), \&(flags[2])} \\
+Address & \texttt{\&lock, \&flags[2]} \\
 \hline
@@
 \hline
-\texttt{\&(lv)} & \texttt{lv} is an lvalue \\
+\texttt{\&lv} & \texttt{lv} is an lvalue \\
 \hline
@@
 \hline
-\texttt{bagCount(b, e} & count how many times \texttt{e} occurs in bag \texttt{b}\\
+\texttt{bagCount(b, e)} & count how many times \texttt{e} occurs in bag \texttt{b}\\
 \hline
@@
 \hline
-Frame $m$ $a$ $r$& start method $m$ with arguments $a$,
+Frame $m$($a$) end=$r$& start method $m$ with arguments $a$,
 initializing variables.  Last instruction is at $r$ \\

### 2020-06-19 book/paper.tex
Massaging the text

@@
 \index{thread}
-do not make a distinction between processes and threads.}) concurrently
-\glsadd{concurrent execution}
+do not make a distinction between processes and threads.})
 read and update shared variables
 \index{shared variable}
+concurrently and synchronize with one another makes programs more
+complicated than programs where only one thing happens at a time.
+\glsadd{concurrent execution}
 \glsadd{shared variable}
-and synchronize with one another makes programs more complicated than
-programs where only one thing happens at a time.  Why are concurrent
+
+Why are concurrent
 programs more complicated than sequential ones?
@@
 by instrumenting the program.
-The output of running concurrent programs depends on how the various
-processes are interleaved.  Some bugs may occur only occasionally and
+The output of running concurrent programs depends on how the
+execution of the various processes are \emph{interleaved}.
+Some bugs may occur only occasionally and
 may never occur when the program is instrumented to find them
@@
 \index{atomicity}
+because there is no other activity interfering with their execution.
 \glsadd{atomicity}
-because there is no other activity
-interfering with their execution.  Even though a statement or function may
+Even though a statement or function may
 be compiled into multiple machine instructions, they are executed back-to-back
-until completion.  Not so with a concurrent program, where other instructions
-may update memory locations as a statement or function is being executed.
+until completion.  Not so with a concurrent program, where other processes
+may update memory locations while a statement or function is being executed.
 \end{itemize}
@@
 a model as a test program.  Because model checking tries every possible
-execution, this means that the test program needs to be relatively
-simple.  In particular, it needs to have a relatively small number of
+execution, this means that the test program needs to be
+simple or it will take longer than we may care to wait for or run
+out of memory.
+In particular, the model needs to have a relatively small number of
 reachable states.
-If the model is too large, it may run for longer than we may care to
-wait for or even run out of memory.
 
@@
 informal one.
-Throughout this book we will include examples
+We will include examples
 of such invariants as they often provide excellent insight into
@@
 \begin{itemize}
-\item In Chapter~\ref{ch:cxlintro} introduces the CXL programming
+\item Chapter~\ref{ch:cxlintro} introduces the CXL programming
 language, as it provides the language for presenting synchronization
 problems and solutions.
-\item Chapter~\ref{ch:concurrent} introduces the problem of
+\item Chapter~\ref{ch:concurrent} illustrates the problem of
 concurrent programming through a simple example in which two processes
-are concurrently incrementing a counter;
+are concurrently incrementing a counter.
 \item Chapter~\ref{ch:cxlmachine} presents the
@@
 underlying concurrency better.
-\item Chapter~\ref{ch:critical} introduces the concept of a critical
-section presents various flawed implementations
+\item Chapter~\ref{ch:critical} introduces the concept of a
+\emph{critical section} and presents various flawed implementations
 of critical sections to demonstrate that implementing a critical section
@@
 an undesirable approach to synchronize processes.
-\item Chapter~\ref{ch:deadlock} introduces the notion of \emph{deadlock}
+\item Chapter~\ref{ch:deadlock} describes \emph{deadlock}
 where a set of processes are indefinitely waiting for one another to
 release a resource.
-\item Chapter~\ref{ch:semaphore} introduces a generalization of locks
+\item Chapter~\ref{ch:semaphore} introduces \emph{semaphores},
+a generalization of locks
 that is good not only for mutual exclusion but also for waiting for
 certain application-level conditions.
-\item Chapters~\ref{ch:bb} and~\ref{ch:sbs} introduce the bounded buffer
-problem and various solutions.
-\item Chapter~\ref{ch:starvation} talks about the problem that in some
-synchronization solutions, processes may not be able to get access to a
-resource they need, and a solution applied to the reader/writer problem.
-\item Chapters~\ref{ch:monitors} and~\ref{ch:mesa} present monitors and
-condition variables, another approach to process synchronication.
+\item Chapters~\ref{ch:bb} and~\ref{ch:sbs} introduce the
+\emph{bounded buffer problem} (aka \emph{producer/consumer problem})
+and various solutions.
+\item Chapter~\ref{ch:starvation} talks about \emph{starvation}:
+the problem that in some
+synchronization approaches processes may not be able to get access to a
+resource they need.
+\item Chapters~\ref{ch:monitors} and~\ref{ch:mesa} present
+\emph{monitors} and \emph{condition variables},
+another approach to process synchronication.
 \item Chapter~\ref{ch:deadlockprevention} revisits deadlocks and how to
 avoid them.
-\item Chapter~\ref{ch:actor} presents the actor model as an approach to
+\item Chapter~\ref{ch:actor} presents the \emph{actor model} as an approach to
 synchronization.
-\item Chapter~\ref{ch:nonblocking} shows solutions to process synchronization
-that prevent processes waiting for one another more than a bounded number of
+\item Chapter~\ref{ch:nonblocking} introduces \emph{non-blocking} or
+\emph{wait-free} synchronization algorithms,
+which prevent processes waiting for one another more than a bounded number of
 steps.
-\item Chapter~\ref{ch:barrier} introduces barrier synchronization, useful in
-high-performance computing applications such as parallel simulations.
+\item Chapter~\ref{ch:barrier} describes \emph{barrier synchronization},
+useful in high-performance computing applications such as parallel simulations.
 \item Chapter~\ref{ch:abp} presents a problem and a solution to the distributed
@@
 \item CXL does not (currently) support floating point, iterators, or I/O;
-\item Python is object-oriented; CXL is not;
+CXL does support \texttt{for} loops and various comprehensions over sets.
+\item Python is object-oriented, supporting classes with methods and
+inheritance; CXL has objects but does not support classes.  CXL supports
+pointers to objects.
 \end{itemize}
@@
 the name listed in the caption of the example.)
-It is sequential and has a method \texttt{square} that takes
+The example is sequential and has a method \texttt{triangle} that takes
 an integer number as argument.  Each method has a variable called
@@
 
-In CXL, constants have a default, but those can be overridden on the command
+In CXL, constants have a default value,
+but those can be overridden on the command
 line using the \texttt{-c} option.
@@
 \index{process}
+running in parallel while sharing variables.
 \glsadd{process}
-running in parallel while sharing variables.
 Figure~\ref{fig:inc} presents a simple example.
@@
 It increments \texttt{count} and sets \texttt{done[self]} to \texttt{True}.
-Method \texttt{main} waits for the processes to finish by polling
-both \texttt{done} flags.
+Method \texttt{main} waits until \texttt{done} flags are set to \texttt{True}.
 After that, it verifies that the value of \texttt{count} equals 2.  If not,
@@
 \index{race condition}
-\glsadd{race condition}
 When running CXL, it will
@@
 of an interleaving, like the one above, in which an assertion fails.
+\glsadd{race condition}
 
@@
 was for each of the failed runs.
+Also experiment with lower values of \texttt{N}.
 \item Can you think of a fix to Figure~\ref{fig:inc}?  Try one or two different fixes

### 2020-06-19 book/paper.tex
Massaging
Typo

@@
 \texttt{(5, False)} is the same CXL value as
-\texttt{dict\{ 0: 5, 1: False \}}.
+\texttt{dict\{ 0:5, 1:False \}}.
 The empty tuple \texttt{()} is the same value as \texttt{dict\{\}}.
@@
     Figure~\ref{fig:inc}.  Run it through CXL and see what happens.  Based on
-    the error output, Describe in English what is wrong with the code by describing,
+    the error output, describe in English what is wrong with the code by describing,
     in broad steps, how running the program can get into a bad state.

### 2020-06-19 book/paper.tex
Latest

@@
 \begin{verbatim}
-   ctx = __init__/() pc =   0 #states = 1 diameter = 0 queue = 0
-CXL Assertion failed ('assert', 'Up.cxl', 9, 5) 1
 #states = 79
 ==== Safety violation ====
+CXL Assertion failed ('assert', 'Up.cxl', 9, 5) 1
 __init__/() [0,32-47]       48 dict{ .count:0, .done:dict{ 0:False, 1:False } }
@@
 
-The state of a process is called a \emph{context} (aka continuation):
+The state of a process is called a \emph{context} (aka \emph{continuation}):
 \index{context}
@@
 Each machine instruction updates the state in some way.
-Figure~\ref{fig:incstate} shows an example of a reachable state for 
-the program in Figure~\ref{fig:inc}.
+% Figure~\ref{fig:incstate} shows an example of a reachable state for the program in Figure~\ref{fig:inc}.
 
@@
 Most instructions pop values from the stack or push values onto the stack.
-At first there is one process (with name tag \texttt{\_\_init\_\_/()}
+At first there is one process, with name tag \texttt{\_\_init\_\_/()},
 that initializes the state.
@@
 The code generated from \texttt{count := count + 1} in line 2 of
-\texttt{Up.cxl} is as follows:
+\texttt{Up.cxl} is as follows (see Figure~\ref{fig:inccode}):
 

### 2020-06-19 book/paper.tex
Latest version

@@
 a data structure, but as we have seen it too requires multiple instructions.
-A more involved one would be access to a binary tree.
+A more involved one would be accessing a binary tree.
 Adding a node to a binary tree, or re-balancing a tree, often requires
@@
 An implementation of a data structure that can be safely accessed by multiple
-processes (or threads) is called \emph{thread-safe}.
+processes (or threads) and free of race conditions is called \emph{thread-safe}.
 \index{thread safety}
@@
 If you run the code through CXL, the assertion should fail because
-there is no code yet for entering and exiting the critical section.
-
-However, mutual exclusion by itself easy to ensure.
+there is no code yet for safely entering and exiting the critical section.
+
+However, mutual exclusion by itself is easy to ensure.
 For example, we could insert the following code to enter the
@@
 but here we will propose the following informally: if there exists a non-empty
-set $S$ of processes that are trying to enter the critical section and any
-process already in the intersection eventually leaves, then eventually one
+set $S$ of processes that are trying to enter the critical section and
+processes in the intersection always leave eventually, then eventually one
 process in $S$ will enter the critical section.
@@
 Clearly, even if mutual exclusion holds in Figure~\ref{fig:csbarebones},
-the spawned processes never terminate.  In order to resolve this, we
+the spawned processes never terminate.  We
 will model processes in critical sections using the framework in
@@
 without entering the critical section ever.
+(Recall that CXL will try every possible execution.)
 
@@
 
-Figure~\ref{fig:mutex} show a high-level state diagram
+Figure~\ref{fig:mutex} shows a high-level state diagram
 \index{state diagram}
@@
 of processes in the critical section.  A process that is in the critical
-section cannot terminate.
+section cannot terminate until after leaving the critical section.
 

### 2020-06-20 book/paper.tex
Latest version

@@
 identifying a location in the CVM bytecode.  The first thing we need to
-ensure is that there can never be two processes at the critical section.
+ensure is that there can never be two processes in the critical section.
 This property is called \texttt{mutual exclusion}.
@@
 without entering the critical section ever.
-(Recall that CXL will try every possible execution.)
+(Recall that CXL will try every possible execution, and so it will evaluate
+both choices.)
 
@@
 Figure~\ref{fig:mutex} shows a high-level state diagram
-\index{state diagram}
-of the code in
-Figure~\ref{fig:cs}.
+\index{state diagram} of mutual exclusion and progress.
 The circles represent states, while the arrows represent possible state
@@
 section cannot terminate until after leaving the critical section.
+We will now consider various approaches toward implementing this
+specification.
 
@@
 \index{lock}
-\glsadd{lock}
 and have realized that
 it could be used to implement a critical section.
+\glsadd{lock}
 The idea is that the lock is like a baton that at most one process can own
@@
 Because of the invariant, \texttt{flags[1] == False} implies that
-process~1 is not at the critical section at time $t_1$.
+process~1 is not in the critical section at time $t_1$.
 Let $t_2$ be the time at which process~0 sets \texttt{flags[0]}
@@
 
-Run the program through CXL.  Unfortunately, it turns out the solution has a
-problem: if both try to enter the critical section at the same time, they may
-end up waiting for one another indefinitely.  Thus the solution
-violates \emph{progress}.
-
-The final na\"{\i}ve solution is based on a variable called \texttt{turn}
+However, if you run the program through CXL, it turns out the solution
+does has a problem: if both try to enter the critical section at the same
+time, they may end up waiting for one another indefinitely.  Thus the
+solution violates \emph{progress}.
+
+The final na\"{\i}ve solution that we propose
+is based on a variable called \texttt{turn}
 that alternates between 0 and 1.  When \texttt{turn = $i$}, process~$i$ can
@@
 section, mutual is violated, the assertion should fail, and a trace should be reported.
-Now plug \texttt{while True: pass;;} before entering the critical section
-in Figure~\ref{fig:cs} and run CXL again.  Now mutual exclusion is guaranteed but
-progress is violated.  CXL should print a trace
+Now plug the following code \texttt{while True: pass;;}
+just before entering the critical section
+in Figure~\ref{fig:cs} and run CXL again.
+Mutual exclusion is guaranteed but progress is violated.
+CXL should print a short trace
 to a state from which a terminating state cannot be reached.
-Describe in English the difference in the failure reports.
-\item If we change the loop in Figure~\ref{fig:cs} into an infinite loop,
-does the \emph{progress} property guarantee that all processes that are trying
-to enter eventually enter?  Why or why not?
+Describe in English the difference in the failure reports before
+and after inserting the code.
 \item See if you can come up with some different approaches that satisfy both
@@
 until either the other process is nowhere near the critical section
-(\texttt{flag[1 - self] == False}) or or has given way
+(\texttt{flag[1 - self] == False}) or has given way
 (\texttt{turn == self}).
@@
 
+\glsadd{state}
+\glsadd{process variable}
 Why does it work?  We will focus here on how one might go about proving
@@
 \index{state}
-\glsadd{state}
-at any point of time the CVM is in a specific state.
+at any point in time the CVM is in a specific state.
 A state is comprised of the values of the shared variables as well as
@@
 \index{process variable}
-\glsadd{process variable}
 for each process, including its
@@
 \index{step}
+Steps in CXL are atomic.
 \glsadd{step}
-Steps in CXL are atomic.
-
+
+\glsadd{trace}
 The CVM starts in an initial state in which there is only
@@
 \index{trace}
-\glsadd{trace}
 is a sequence of steps starting from the initial state.
@@
 Also imagine that in this particular state $\mathtt{turn} = 1$.  Now let
-process process~1 make a sequence of steps.  Because $\mathtt{turn} = 1$,
+process~1 make a sequence of steps.  Because $\mathtt{turn} = 1$,
 process~1 will break out of the while loop and also enter the critical
@@
 Let predicate $I_p(i)$ be the following.
-if process $i$ is label \texttt{@cs} (i.e., process $i$ is in the critical section),
+if process $i$ is at label \texttt{@cs} (i.e., process $i$ is in the critical section),
 then $C(i)$ does not hold or process $1-i$ is executing after setting
@@
 predicate is not invariant, and therefore not an inductive invariant either.
-To finish the story, the author of CXL fixed the Wikipedia page.
+(The author fixed the Wikipedia page.)
 

### 2020-06-20 book/paper.tex
Latest

@@
 section.
+The name \texttt{mutex} is often used to denote a variable or value
+that is used for mutual exclusion.
+\texttt{P\_mutex} is a method that returns a ``mutex,'' which, in this
+case, is a dictionary that contains Peterson's Algorithm's state:
+a turn variable and two flags.
+Both methods \texttt{P\_enter} and \texttt{P\_exit} take two arguments:
+a pointer to a mutex and the process identifier (0 or 1).
+\texttt{(\^{}pm).turn} is the value of the \texttt{.turn} key
+in the mutex dictionary.  Note that the parentheses in this expression
+are needed, as \texttt{\^{}pm.turn} would wrongly evaluate to
+\texttt{\^{}(pm.turn)}.
+
 You can put the first three methods in its own CXL source file
@@
 
-While the CVM does not have any built-in interlock instructions,
+While the CVM does not have any specific built-in interlock instructions,
 it does have support for executing multiple instructions atomically.
@@
 a set of $N$ processes.
-It is called \emph{spinlock},
+The approach is called \emph{spinlock},
 \index{spinlock}
@@
 To see that invariant $I_1$ is maintained, note that
-\texttt{\^{}p == True} upon entry of \texttt{tas}.
+\texttt{\^{}p == True} upon entry of \texttt{tas}
+(because of the condition on the \texttt{while} loop that the
+\texttt{tas} method is invoked in).
 So there are two cases:

### 2020-06-20 book/paper.tex
Fixed synchS

@@
 language.
-\item Chapter~\ref{ch:spinlock} introduces hardware-supported locks
-for implemented critical sections.  Chapter~\ref{ch:synch} presents a CXL module
+\item Chapter~\ref{ch:spinlock} introduces ``hardware'' locks
+for implemented critical sections.
+\item Chapter~\ref{ch:synch} presents CXL modules
 that supports locks and other synchronization primitives.
@@
 
-\chapter{Locks and the Synch Module}
+\chapter{Locks and Blocking}
 \label{ch:synch}
@@
 
-Locks are probably the most popular and basic form of synchronization
+Locks are probably the most prevalent and basic form of synchronization
 in concurrent programs.  For this reason, CXL has a module called
@@
 until the lock is available to it.
-In CXL, a process can suspend itself and save its context in a
+In CXL, a process can suspend itself and save its context (state) in a
 list.  Recall that the context of a process consists of its name tag,
@@
 using suspension.
-A \texttt{Lock} maintains both a boolean indicating whether the lock is taken,
-and a list of contexts of processes that want to acquire the lock.
-\texttt{lock} sets the lock if available, or suspends itself if not.
+The interface is implemented as follows:
+\begin{itemize}
+\item A \texttt{Lock} maintains both a boolean indicating whether the
+lock is taken and a list of contexts of processes that want to acquire the lock.
+\item
+\texttt{lock} sets the lock if available and suspends itself if not.
 Note that \texttt{stop} is called within an \texttt{atomic} statement---this is
@@
 it allows other processes to run.
-
+\item
 \texttt{unlock} checks to see if any processes are waiting to get the lock.
 If so, it uses the \texttt{head} and \texttt{tail}
-methods from the \texttt{list} module to resume the first process that got
-suspend and to remove its context from the list.
+methods from the \texttt{list} module (see Appendix~\ref{ap:module})
+to resume the first process that got
+suspended and to remove its context from the list.
+\end{itemize}
 Selecting the first process is a design choice.  Another implementation could
@@
 gets a chance to obtain the lock (assuming every process eventually releases
-it).  Chapter~\ref{ch:starvation} will talk more about starvation.
+it).  Chapter~\ref{ch:starvation} will talk more about starvation and how
+to prevent it.
 
@@
 
-runs the file \texttt{x.cxl} using the Suspension version of the \texttt{synch} module.
+runs the file \texttt{x.cxl} using the suspension version of the \texttt{synch} module.
 You will find that using the \texttt{synchS} module often leads to searching a
@@
 \chapter{List of Modules}
+\label{ap:module}
 

### 2020-06-20 book/paper.tex
Latest version

@@
 \item Write a synchronization module using your own implementation of \texttt{Lock}
-(for example, using Peterson's algorithm or using \texttt{swap}), and run the code in
+(for example, using Peterson's algorithm or using \texttt{swap}) and run the code in
 Figure~\ref{fig:incfixed} using your own module.

### 2020-06-20 book/paper.tex
Latest version

@@
 \item Write a synchronization module using your own implementation of \texttt{Lock}
-(for example, using Peterson's algorithm or using \texttt{swap}) and run the code in
-Figure~\ref{fig:incfixed} using your own module.
+(for example, using Peterson's algorithm or using \texttt{swap}).
+\item
+Run Figure~\ref{fig:incfixed} using (1) \texttt{synch}, (2), \texttt{synchS},
+and (3) your own module.  Report how many states were explored for each
+version of the module.
 \end{enumerate}

### 2020-06-20 book/paper.tex
Latest

@@
 Also experiment with lower values of \texttt{N}.
+How large does \texttt{N} need to be for assertions to fail?
 \item Can you think of a fix to Figure~\ref{fig:inc}?  Try one or two different fixes
@@
 
+\section*{Exercises}
+\begin{enumerate}
+\item Using busy waiting, implement a ``bound lock'' that allows
+up to \texttt{M} processes to acquire it at the same time.  A bound lock
+with \texttt{M == 1} is an ordinary lock.
+You should define a constant \texttt{M} and two methods:
+\texttt{acquire\_bound\_lock()}
+and \texttt{release\_bound\_lock()}.
+\item Write a test program for your bound lock
+that checks that no more than \texttt{M} processes can acquire the
+bound lock.
+\end{enumerate}
+
 \chapter{Reader/Writer Locks with Blocking}
@@
 reader/writer locks can be instantiated.
+\item Write a test program for ``bound locks'' (as implemented in the
+exercises of the last chapter) that checks that up to \texttt{M} processes
+can acquire the bound lock at the same time.
 \end{enumerate}

### 2020-06-20 book/paper.tex
Latest
Adding exercises

@@
 Run Figure~\ref{fig:incfixed} using (1) \texttt{synch}, (2), \texttt{synchS},
-and (3) your own module.  Report how many states were explored for each
+and (3) your own module.  Report how many states were explored by CXL for each
 version of the module.
@@
 
-\begin{quote}
-\begin{itemize}
-\item Can you come up with a test (an assertion) that shows that at most three
-philosophers are eating at the same time?
-\item Can you come up with a test (different code) that checks to see that
-the solution allows more than one philosopher to eat at the same time?
-\end{itemize}
-\end{quote}
-
 One can think of a semaphore as a generalization of a lock.  In fact, a
@@
 Allen Downey a great resource for working with semaphores~\cite{Downey09}.
+
+\section*{Exercices}
+\begin{enumerate}
+\item Add an assertion that demonstrates that at most three
+Italian philosophers are eating at the same time?
+\item Implement a test program that checks to see that
+the solution allows more than one Italian philosopher to eat at the same time?
+\end{enumerate}
 
@@
 exactly the set of tuples that the producers produce.
+
+\section*{Exercises}
+\begin{enumerate}
+\item Assume that you have only binary semaphores.  Implement counting
+semaphores using binary semaphores.
+\end{enumerate}
 

### 2020-06-20 book/paper.tex
Latest

@@
 the solution allows more than one Italian philosopher to eat at the same time?
+\item Figure~\ref{fig:qsort} shows an iterative implementation of the Qsort
+algorithm, and Figure~\ref{fig:qsorttest} an accompanying test program.
+Note that it uses a shared variable \texttt{a} containing the array to be
+sorted and a shared variable \texttt{todo} containing the ranges of the
+array that need to be sorted (initially the entire array).
+Re-using as much of this code as you can, implement a parallel version of
+this.  You should not have to change the methods \texttt{swap}, \texttt{partition},
+or \texttt{sortrange} for this.  Create \texttt{NWORKERS} ``worker processes''
+that should replace the \texttt{qsort} code.  Each worker loops until \texttt{todo}
+is empty and sorts the ranges that it finds until then.  The \texttt{main}
+process needs to wait until all workers are done.  Use a semaphore for this:
+the \texttt{main} process should use \texttt{P} to wait for each
+worker to terminate.  After that it should check if the array is sorted and a
+permutation of the original array.
 \end{enumerate}
+
+\begin{figure}
+\begin{code}
+\VerbatimInput[xleftmargin=5mm,numbers=left]{code/qsort.cxl}
+\end{code}
+\caption{[qsort.cxl] Iterative qsort() implementation.}
+\label{fig:qsort}
+\end{figure}
+
+\begin{figure}
+\begin{code}
+\VerbatimInput[xleftmargin=5mm,numbers=left]{code/qsorttest.cxl}
+\end{code}
+\caption{[qsorttest.cxl] Test program for Figure~\ref{fig:qsort}.}
+\label{fig:qsorttest}
+\end{figure}
 

### 2020-06-20 book/paper.tex
Added more exercises

@@
 
-\section*{Exercises}
-\begin{enumerate}
-\item Assume that you have only binary semaphores.  Implement counting
-semaphores using binary semaphores.
-\end{enumerate}
-
 \chapter{Split Binary Semaphores}
@@
 
-\begin{quote}
-\begin{itemize}
-\item Implement the producer/consumer problem with SBS.
-\end{itemize}
-\end{quote}
+\section*{Exercises}
+\begin{enumerate}
+\item Assume that you have only binary semaphores.  Implement counting
+semaphores using binary semaphores using split binary semaphores.
+\item Implement a solution to the producer/consumer problem
+(Figure~\ref{fig:boundedbuffer}) using split binary semaphores.
+\end{enumerate}
 

### 2020-06-20 book/paper.tex
Working on exercises

@@
 per process rather than a simple counter?
-\item Can you write a version of this test that uses a counter instead
+Can you write a version of this test that uses a counter instead
 of a flag per process?
-\item Write a ``library implementation'' of reader/writer locks so multiple
-reader/writer locks can be instantiated.
+% \item Write a ``library implementation'' of reader/writer locks so multiple
+% reader/writer locks can be instantiated.
 \item Write a test program for ``bound locks'' (as implemented in the
@@
 can acquire the bound lock at the same time.
+\item Suppose a sports club has only one dressing room.  The policy is that
+either only women or only men may be in the dressing room at any time.
+Implement a ``dressing room lock'' that allows multiple people to enter the
+dressing room while observing the policy.
 \end{enumerate}
@@
 \item Does it depend on \texttt{N}, the number of philosophers?
-\item Can you come up with a solution?
-\item Does it matter in what order the philosophers lay down their forks?
 \end{itemize}
@@
 
-\begin{quote}
-\begin{itemize}
-\item Try it out!
-\end{itemize}
-\end{quote}
+\section*{Exercises}
+\begin{enumerate}
+\item Does it matter in what order the philosophers lay down their forks?
+\end{enumerate}
 

### 2020-06-21 book/paper.tex
Added another exercise
Changed definition of blocking and added some exercises

@@
   name=blocked process,
-  description={is a process cannot make progress without the help of another process.
+  description={is a process that is waiting on a low-level synchronization primitive
+    and cannot make progress without the help of another process.
       For example, a process that is waiting for a spinlock to become available}
@@
   description={(aka spin-waiting) is when a process waits in a loop for some
-      application-defined condition while changing the state, for example by
-      continually acquiring and releasing a lock}
+      application-defined condition}
 }
@@
 statement.
-We can use \texttt{atomic} blocks to implement a wide variety of
+We can use \texttt{atomic} statement blocks to implement a wide variety of
 interlock operations.
@@
 
+\glsadd{blocked process}
 We call a process \emph{blocked}
 \index{blocked process}
-\glsadd{blocked process}
-in a state if the process is in
-an \emph{infinite loop} in which the process does not update the
-shared memory.
+in a state if the process is waiting on a low-level synchronization
+variable and cannot make progress without the help of another process.
 A process trying to
 acquire a test-and-set spinlock held by another process is a good example
-of a process being blocked: the process only reads the shared memory
-but does not change it.
-The only way to break out of the infinite loop is if another process
-changes the shared memory.
-The \texttt{while} loop in Peterson's algorithm, in case its condition
-does not hold, is another example of a process being blocked.
+of a process being blocked.
+The only way forward is if the other process releases the lock.
 
@@
 
+\glsadd{busy waiting}
 While correct, it is not considered a good solution.
@@
 \index{busy waiting}
-\glsadd{busy waiting}
 (aka spin-waiting):
@@
 processes spin in a loop until some desirable application-level condition is met.
+Figure~\ref{fig:inc} also has an example of busy-waiting: the \texttt{main}
+process waits for the other two processes to finish by checking their \texttt{done}
+flags.
 The astute reader might wonder if obtaining the mutex itself is an
@@
 However, there is a big difference.  As we pointed out, the processes that
-are waiting for a spinlock are \emph{blocked}: they are in an infinite loop
-in which they do not change the state.
-The processes waiting for a reader/writer lock do change the state while
-in the infinite loop: they acquire the mutex and release it.
-
+are waiting for a spinlock are \emph{blocked}.
 In most operating systems and programming language runtimes,
@@
 can acquire the bound lock at the same time.
-\item Suppose a sports club has only one dressing room.  The policy is that
-either only women or only men may be in the dressing room at any time.
-Implement a ``dressing room lock'' that allows multiple people to enter the
-dressing room while observing the policy.
 \end{enumerate}
@@
 \item Does it matter in what order the philosophers lay down their forks?
+\item Figure~\ref{fig:bank} shows an implementation of a bank with various
+accounts and transfers between those accounts.
+Unfortunately, running the test reveals that it sometimes leaves unterminated
+processes.  Can you fix the problem?
 \end{enumerate}
+
+\begin{figure}
+\begin{code}
+\VerbatimInput[xleftmargin=5mm,numbers=left]{code/bank.cxl}
+\end{code}
+\caption{[bank.cxl] Bank accounts.}
+\label{fig:bank}
+\end{figure}
 
@@
 exactly the set of tuples that the producers produce.
+
+\section*{Exercises}
+\begin{enumerate}
+\item Change the code in Figure~\ref{fig:boundedbuffer} so that
+\texttt{produce()} fails if the buffer is full rather than block.
+\texttt{produce()} should return \texttt{False} if it fails
+and return \texttt{True} if it succeeds.
+\item Implement a \emph{bounded stack}.  Method \texttt{push(item)}
+should block if the stack is full.  Method \texttt{pop()} should block
+until the stack is non-empty and then return an item.
+\item Implement a GPU allocator.  There are \texttt{N} GPUs identified
+by the numbers
+1 through \texttt{N}.  Method \texttt{gpuAlloc()} returns the identifier
+of an available GPU, blocking if there is currently no available GPU.
+Method \texttt{gpuRelease(gpu)} releases the given GPU.  It never needs
+to block.
+\end{enumerate}
 
@@
 (Figure~\ref{fig:boundedbuffer}) using split binary semaphores.
+\item Suppose a sports club has only one dressing room.  The policy is that
+either only women or only men may be in the dressing room together.
+Implement a ``dressing room lock'' that allows multiple people to enter the
+dressing room while observing the policy.
 \end{enumerate}

### 2020-06-21 book/paper.tex
Latest

@@
 resource they need.
-\item Chapters~\ref{ch:monitors} and~\ref{ch:mesa} present
+\item Chapter~\ref{ch:monitors} presents
 \emph{monitors} and \emph{condition variables},
@@
 \index{ring buffer}
-It is essentially
+Suppose you have a \emph{producer} (some process) wanting to
+communicate a stream of items with a \emph{consumer} (another process).
+A bounded buffer is essentially
 a queue implemented using a circular buffer
@@
 of a certain length and two pointers:
-one where items are inserted and one from where items are extracted.  If the
-buffer is full, processes that want to add more items have to wait.  In the usual
-formulation, processes that want to extract an item when the buffer is empty
-also have to wait.
-This problem is known as the ``Producer/Consumer Problem.''
+one where the producer inserts items, and one where the consumer extracts
+items.  If the
+buffer is full, the producer must wait; if the buffer is empty, the
+consumer must wait.
+This problem is known as the ``Producer/Consumer Problem'' and was
+proposed by Dijkstra~\cite{EWD329}.
+Here we generalize it to allowing multiple producers and multiple consumers
+all sharing the same bounded buffer.
 
@@
 
+\section*{Exercises}
+\begin{enumerate}
+\item TBD.  starvation-free dressing room
+\item TBD.  multiple GPU allocator
+\end{enumerate}
+
 \chapter{Monitors}
@@
 
-\chapter{Mesa Condition Variables in CXL}
-\label{ch:mesa}
+% \chapter{Mesa Condition Variables in CXL}
+% \label{ch:mesa}
 

### 2020-06-21 book/paper.tex
Typo
Massaging

@@
 \item Run Figure~\ref{fig:cs} using CXL.  As there is no protection of the critical
-section, mutual is violated, the assertion should fail, and a trace should be reported.
+section, mutual exclusion is violated,
+the assertion should fail, and a trace should be reported.
 Now plug the following code \texttt{while True: pass;;}
@@
 
-\begin{quote}
-\begin{itemize}
-\item Draw additional states and steps in Figure~\ref{fig:rdwr}
-for three processes.
-\end{itemize}
-\end{quote}
-
 A process that wants to read first waits until there are no writers:
-\texttt{nwriters == 0}.  If so, it increments the number of readers.
+\texttt{nwriters == 0}.  Then it increments the number of readers.
 Similarly, a
 process that wants to write waits until there are no readers \emph{or} writers.
-If so, it increments the number of writers.
+Then it increments the number of writers.
 The important invariants in this code are:
@@
 \index{busy waiting}
-(aka spin-waiting):
+(aka \emph{spin-waiting}):
 \index{spin waiting}
@@
 In most operating systems and programming language runtimes,
-when a process acquires a lock, the process is placed on a scheduling
-queue and stops using CPU cycles until the lock becomes available.
+when a process acquires a mutex, the process is placed on a scheduling
+queue and stops using CPU cycles until the mutex becomes available.
 CXL supports this with the \texttt{synchS} module.
@@
 \begin{enumerate}
+\item Draw additional states and steps in Figure~\ref{fig:rdwr}
+for three processes.
 \item Using busy waiting, implement a ``bound lock'' that allows
@@
 section and two readers try to enter.  The first reader successfully
-acquires \texttt{rlock} and but blocks (!) when trying to acquire
+acquires \texttt{rlock} and but blocks when trying to acquire
 \texttt{rwlock}, which is held by the writer.  The second reader blocks
@@
 reader is still in the critical section, increments \texttt{nreaders} to 2
-and enter the critical section \emph{without} acquiring \texttt{rwlock}.
+and enters the critical section \emph{without} acquiring \texttt{rwlock}.
 \texttt{rwlock} is essentially jointly held by both readers.
@@
 can acquire the bound lock at the same time.
+\item It can be useful for a process to \emph{downgrade} its write lock
+to a read lock when its done writing but not done reading.  Implement
+\texttt{downgrade}.  Note, you cannot do this simply by releasing the
+write lock and then obtaining the read lock.
 \end{enumerate}
@@
 Imagine five philosopers sitting around a table, each with a plate of food in
-front of them, and a fork between each two plates.  Each philosopher requires
+front of them and a fork between each two plates.  Each philosopher requires
 two forks to eat.  To start eating, a philosopher first picks up the fork on
@@
 from eating to think for a while.  To do so, the philosopher puts down both
-forks.  The philosophers repeat this procedure.  Dijkstra had them repeating
+forks.  Each philosopher repeat this procedure.  Dijkstra had them repeating
 this for ever, but for the purposes of this book, philosphers can leave

### 2020-06-21 book/paper.tex
Worked on Semaphores chapter

@@
 no more than three philosophers can eat at a time.
-
-\begin{quote}
-\begin{itemize}
-\item See if you can come up with a deadlock-free solution using locks.  Represent
-    each fork as a lock.
-\end{itemize}
-\end{quote}
-
-It's not easy.  It would be easy to come up with a solution that uses a single lock
+See if you can come up with a deadlock-free solution using locks.  Represent
+each fork as a lock.  It's not easy.
+It is not hard to come up with a solution that uses a single lock
 and busy waiting, but we would like to avoid busy waiting.
+
 Introduced by Dijkstra,
@@
 as in vacating the resource.
-
-Figure~\ref{fig:semaphore} shows the \texttt{synch} module implementation of
-semaphores.
-Figure~\ref{fig:dinerssema} shows a solution to the Italian Philosphers problem
-using a semaphore.
-
 One can think of a semaphore as a generalization of a lock.  In fact, a
@@
 procures the lock, and \texttt{V} vacates the lock.
+
+Figure~\ref{fig:semaphore} shows the \texttt{synch} module implementation of
+semaphores.
+It represents a semaphore as an integer.
+Figure~\ref{fig:dinerssema} shows a solution to the Italian Philosphers problem
+using a semaphore.
 
@@
 algorithm, and Figure~\ref{fig:qsorttest} an accompanying test program.
-Note that it uses a shared variable \texttt{a} containing the array to be
-sorted and a shared variable \texttt{todo} containing the ranges of the
+The array to be sorted is stored in  shared variable \texttt{a}.
+Another shared variable, \texttt{todo}, contains the ranges of the
 array that need to be sorted (initially the entire array).
@@
 the \texttt{main} process should use \texttt{P} to wait for each
-worker to terminate.  After that it should check if the array is sorted and a
-permutation of the original array.
+worker to terminate.  After that it should check that the array is sorted and
+that it is a permutation of the original array.
 \end{enumerate}

### 2020-06-22 book/paper.tex
Latest

@@
 proposed by Dijkstra~\cite{EWD329}.
-Here we generalize it to allowing multiple producers and multiple consumers
+
+The producer/consumer pattern is common.  Processes may be arranged
+in pipelines, where each upstream process is a producer and each downstream
+process is a consumer.
+Or processes may be arranged in a manager/worker pattern, with a manager
+producing jobs and workers executing them in parallel.
+
+In general, we allow multiple producers and multiple consumers
 all sharing the same bounded buffer.
-
-Figure~\ref{fig:pc} gives a high-level description of what we want.  The
+Figure~\ref{fig:pc} gives a high-level description.  The
 two numbers in a state specify
@@
 consumers, updates to the indexes, which are shared variables, must be protected.
-To this end we use two semaphores as locks: \texttt{l\_in} for \texttt{b\_in}
+To this end, we use two semaphores as locks: \texttt{l\_in} for \texttt{b\_in}
 and \texttt{l\_out} for \texttt{b\_out}.
@@
 entries in the circular buffer.
-To this end there is a semaphore \texttt{n\_full} that
+For this, there is a semaphore \texttt{n\_full} that
 keeps track of the number of filled entries in the buffer and a semaphore
@@
 an item is available.
-
 The consumer code is symmetric.
@@
 \index{binary semaphore}
-as their counters only take on the values 0 and 1.
+as they only take on the values 0 and 1.
 They protect critical sections.
@@
 
-We can run the code through CXL, but what does it test?  There are no
+We can run the code through CXL, but what would it test?  There are no
 assert statements in the code.
@@
 should terminate, while other values should lead to processes blocking.
-One could write a script like the one in Figure~\ref{fig:pcscript}.
+To this this, one can write a script like the one in Figure~\ref{fig:pcscript}.
 
@@
 until the stack is non-empty and then return an item.
-\item Implement a GPU allocator.  There are \texttt{N} GPUs identified
+\item Implement a thread-safe GPU allocator.  There are \texttt{N} GPUs identified
 by the numbers
@@
 has the condition that a process can only enter if no other process
-is in the critical section.  Reader/writer locks refined that, loosening
+is in the critical section.  Reader/writer locks refined that condition, loosening
 the restriction for improved performance.
@@
 problem, there are two: a process that wants to read may have to wait for a
-writer to leave the critical regiion, while a process that wants to write may
+writer to leave the critical region, while a process that wants to write may
 have to wait until all readers have left the critical section.  If there are $N$
@@
 must always be 0 or 1, and it should be 0 when a process is accessing the shared
-variables.
+variables and 1 when not.
 As it were, these semaphores taken together are emulating a single binary semaphore.
@@
 are all~0.  To maintain the invariant, each process alternates \texttt{P} and
-\texttt{V} operations, starting with \texttt{P~\&(mutex)} when it tries to
+\texttt{V} operations, starting with \texttt{P(\&mutex)} when it tries to
 enter the CCS, and ending with a \texttt{V} operation on one of the semaphores
@@
 nor writers in the critical section, it vacates the semaphore associated
-with writers waiting.  If both conditions hold, it could vacates either
+with writers waiting.  If both conditions hold, it could vacate either
 one---it would not matter.  However, it can only vacate one of the semaphores.
@@
 
-Let us return now to \texttt{rlock\_acquire}, in the case that the process
+Let us return now to \texttt{rlock\_acquire}. In the case that the process,
+say $p$,
 found there is a writer in the critical section, it vacates
@@
 with waiting readers.  When later some other process passes the baton to
-it by incrementing \texttt{r\_sema}), it updates the state by decrementing
-\texttt{r\_waiting} and incrementing \texttt{r\_entered}.  It
+$p$ by incrementing \texttt{r\_sema}, $p$ updates the state by decrementing
+\texttt{r\_waiting} and incrementing \texttt{r\_entered}.  $p$
 then invokes \texttt{V\_one()} one more time.

### 2020-06-22 book/paper.tex
Working on exercises
Latest

@@
 until the stack is non-empty and then return an item.
-\item Implement a thread-safe GPU allocator.  There are \texttt{N} GPUs identified
-by the numbers
+\item Implement a thread-safe GPU allocator by modifying Figure~\ref{fig:gpu}.
+There are \texttt{N} GPUs identified by the numbers
 1 through \texttt{N}.  Method \texttt{gpuAlloc()} returns the identifier
@@
 \end{enumerate}
+
+\begin{figure}
+\begin{code}
+\VerbatimInput[xleftmargin=5mm,numbers=left]{code/gpu.cxl}
+\end{code}
+\caption{[gpu.cxl] A thread-unsafe GPU allocator.}
+\label{fig:gpu}
+\end{figure}
 
@@
 enter a critical section using a spinlock.  Two of
-the processes successfully enter, alternating, but the third
+the processes successfully keep entering, alternating, but the third
 process never gets a turn.  This is an example of

### 2020-06-22 book/paper.tex
Latest
Working on starvation chapter

@@
 (Figure~\ref{fig:boundedbuffer}) using split binary semaphores.
-\item Suppose a sports club has only one dressing room.  The policy is that
-either only women or only men may be in the dressing room together.
-Implement a ``dressing room lock'' that allows multiple people to enter the
-dressing room while observing the policy.
+\item Cornell's campus features some one-lane bridges.  Cars can
+only go in one direction at a time. Consider northbound
+and southbound cars wanting to cross a one-lane bridge.
+The bridge allows arbitrary many cars, as long as they're going in the
+same direction.
+Implement a lock that observes this requirement using SBS.  Write methods
+\texttt{nb\_enter()} that a car must invoke before going northbound on
+the bridge and \texttt{nb\_leave()} that the car must invoke after leaving
+the bridge.  Similarly write \texttt{sb\_enter()} and \texttt{sb\_leave()}
+for southbound cars.
+
 \end{enumerate}
@@
 \end{code}
-\caption{[RWfair.cxl] Reader/Writer Lock implementation addressing fairness.}
+\caption{[RWfair.cxl] Reader/Writer Lock SBS implementation addressing fairness.}
 \label{fig:RWfair}
+\end{figure}
+
+\begin{figure}
+\begin{code}
+\VerbatimInput[xleftmargin=5mm,numbers=left]{code/RWqueue.cxl}
+\end{code}
+\caption{[RWqueue.cxl] Reader/Writer Lock SBS implementation using a queue.}
+\label{fig:RWqueue}
+\end{figure}
+
+\begin{figure}
+\begin{code}
+\VerbatimInput[xleftmargin=5mm,numbers=left]{code/RWqtest.cxl}
+\end{code}
+\caption{[RWqtest.cxl] Test program for Figure~\ref{fig:RWqueue}.}
+\label{fig:RWqtest}
 \end{figure}
@@
 \index{fairness}
-\glsadd{fairness}
 (although it does not quite
 match the usual formal nor vernacular concepts of fairness).
+\glsadd{fairness}
 
@@
 \begin{enumerate}
-\item TBD.  starvation-free dressing room
+\item Write a fair solution to the one-lane bridge problem of Chapter~\ref{ch:sbs}.
 \item TBD.  multiple GPU allocator

### 2020-06-22 book/paper.tex
Working on starvation solution

@@
 
-In this chapter, we will present a version of a reader/writer lock
-implementation that solves this type of starvation.
-In particular, the probability that either a reader or writer gets starved
+In this chapter, we will present two SBS-based reader/writer lock
+implementations that solves this type of starvation.
+SBS allow much control over which thread runs next and is therefore a
+good starting point for developing synchronization algorithms that do not
+suffer from starvation.
+For the first solution the probability that either a reader or writer gets starved
 will be exponentially diminishing.

### 2020-06-23 book/paper.tex
Latest

--- 2020-06-22 book/paper.tex
@@
 
-\begin{figure}
-\begin{code}
-\VerbatimInput[xleftmargin=5mm,numbers=left]{code/RWqueue.cxl}
-\end{code}
-\caption{[RWqueue.cxl] Reader/Writer Lock SBS implementation using a queue.}
-\label{fig:RWqueue}
-\end{figure}
-
-\begin{figure}
-\begin{code}
-\VerbatimInput[xleftmargin=5mm,numbers=left]{code/RWqtest.cxl}
-\end{code}
-\caption{[RWqtest.cxl] Test program for Figure~\ref{fig:RWqueue}.}
-\label{fig:RWqtest}
-\end{figure}
-
 A \emph{property}
@@
 section.  Thus the probability of starvation is exponentially vanishing.
+We shall call such a solution \emph{fair}
+\index{fairness}
+(although it does not quite
+match the usual formal nor vernacular concepts of fairness).
+\glsadd{fairness}
+
 Unfortunately, such is not the case for the
@@
 
-In this chapter, we will present a version of a reader/writer lock
-implementation that solves this type of starvation.
-In particular, the probability that either a reader or writer gets starved
-will be exponentially diminishing.
-We shall call such a solution \emph{fair}
-\index{fairness}
-(although it does not quite
-match the usual formal nor vernacular concepts of fairness).
-\glsadd{fairness}
-
-It is difficult if not impossible to make the reader/writer lock solution of
-Figure~\ref{fig:rw2lock} fair.  However, the split binary semaphore solution of
-Figure~\ref{fig:RWsplitsema} provides much control over whom to pass baton
-to.
-Figure~\ref{fig:RWfair} contains a version of Figure~\ref{fig:RWsplitsema}
-that is a fair solution to the reader/writer problem.  There are two changes
-compared to the original version:
+SBSs allow much control over which thread runs next and is therefore a
+good starting point for developing fair synchronization algorithm
+In this chapter, we will present two SBS-based fair reader/writer lock
+implementations.
+The first, in Figure~\ref{fig:RWfair}, is based on Figure~\ref{fig:RWsplitsema}.
+There are two differences:
 
@@
 section.  While readers can still starve other readers and writers can still
-starve other writers, they can not starve one another.  And contention for
+starve other writers, they cannot starve one another.  And since contention for
 a semaphore is resolved probabilistically, each reader and each writer will
 eventually have a chance to enter the critical section almost surely.
+
+Figures~\ref{fig:RWqueue} and~\ref{fig:RWqtest} present a different approach
+in which processes are allowed into the critical region in a purely First
+Come First Served manner.  When a reader arrives, it is allowed in only if
+there is no writer in the critical section or waiting to enter.  When a writer
+arrives, it is allowed in only if there is no reader or writer in the
+critical section (and therefore also not waiting to enter).  To maintain the
+arrival order, the solution uses a semaphore per process and a queue.
+The \texttt{acquire\_} methods take a pointer to
+the semaphore of the process as argument.
+Each entry in the queue is a pair consisting of a type of access
+(\texttt{.read} or \texttt{.write}) and a pointer to the semaphore of the
+processs that is waiting.  Method \texttt{V\_one} considers semaphores to vacate
+in the order in which they appear in the queue.
+
+\begin{figure}
+\begin{code}
+\VerbatimInput[xleftmargin=5mm,numbers=left]{code/RWqueue.cxl}
+\end{code}
+\caption{[RWqueue.cxl] Reader/Writer Lock SBS implementation using a queue.}
+\label{fig:RWqueue}
+\end{figure}
+
+\begin{figure}
+\begin{code}
+\VerbatimInput[xleftmargin=5mm,numbers=left]{code/RWqtest.cxl}
+\end{code}
+\caption{[RWqtest.cxl] Test program for Figure~\ref{fig:RWqueue}.}
+\label{fig:RWqtest}
+\end{figure}
 
@@
 \item Write a fair solution to the one-lane bridge problem of Chapter~\ref{ch:sbs}.
-\item TBD.  multiple GPU allocator
+If you want to use the queue method, you can change the signature of the
+methods to enter the bridge, adding an argument that contains a pointer to a
+semaphore.
 \end{enumerate}
@@
 
-In \texttt{release\_rlock}, notice that \texttt{notify \&(wcond)} is invoked when there
+In \texttt{release\_rlock}, notice that \texttt{notify(\&wcond)} is invoked when there
 are no readers left, \emph{without} checking if there are writers waiting to enter.
@@
 
-\texttt{release\_wlock} executes \texttt{notifyAll \&(wcond)} as well as
-\texttt{notify \&(wcond)}.
+\texttt{release\_wlock} executes \texttt{notifyAll(\&wcond)} as well as
+\texttt{notify(\&wcond)}.
 Again, because we do not keep track of the number of waiting readers or writers, we
@@
 introduction to working with Mesa-style condition variables~\cite{Birrell89}.
+
+\section*{Exercises}
+\begin{enumerate}
+\item Write a new version of the GPU allocator of Chapter~\ref{ch:bb}.  In this version,
+a process is allowed to allocate a set of GPUs and release a set of GPUs that it
+has allocated.  Method \texttt{gpuAllocSet(n)} should block until $n$ GPUs are
+available, but it should grant them as soon as they are available.
+Method \texttt{gpuReleaseSet(s)} takes a set of GPU identifiers as argument.
+A process does not have to return all the GPUs it allocated at once.
+\item The specification in the previous question makes the solution unfair.
+Explain why this is so.  Then change the specification and the solution so that
+it is fair.
+\end{enumerate}
 

### 2020-06-23 book/paper.tex
Better exercise numbering

@@
 \usepackage{amsmath}
+\usepackage{geometry}
+\usepackage{enumitem}
+\usepackage{etoolbox}
+
+\AtBeginEnvironment{enumerate}{\everymath{\displaystyle}}
+\newlist{problems}{enumerate}{2}
+\setlist[problems]{wide=0pt}
+\setlist[problems,1]{label=\thechapter.\arabic*, font=\bfseries, wide=0pt}
+\setlist[problems,2]{label=(\alph*), wide =0.5em, topsep=2pt, itemsep=2pt}
 
@@
 \section*{Exercises}
-\begin{enumerate}
+\begin{problems}
 \item See what happens if, instead of initializing \texttt{result} to 0,
@@
 should compute the square of $x$ by adding $x$ to an initial value of 0 $x$ times.
-\end{enumerate}
+\end{problems}
 
@@
 
-\begin{enumerate}
+\begin{problems}
 \item CXL programs can usually be easily translated into Python.  For example,
@@
 time---the important thing is to develop an understanding of concurrency.
-\end{enumerate}
+\end{problems}
 
@@
 
-\begin{enumerate}
+\begin{problems}
 \item Figure~\ref{fig:incenter} shows an attempt at trying to fix the code of
@@
     CXL and describe either why it works or why it does not work.
-\end{enumerate}
+\end{problems}
 
@@
 \section*{Exercises}
-\begin{enumerate}
+\begin{problems}
 \item Run Figure~\ref{fig:cs} using CXL.  As there is no protection of the critical
@@
 we will find out, it is possible but not easy.
-\end{enumerate}
+\end{problems}
 
@@
 \section*{Exercises}
-\begin{enumerate}
+\begin{problems}
 \item Figure~\ref{fig:csonebit} presents another correct solution to the
@@
 Can you generalize Peterson's algorithm to more than two processes?
-\end{enumerate}
+\end{problems}
 
@@
 \section*{Exercises}
-\begin{enumerate}
+\begin{problems}
 \item Implement an atomic swap operation.  It should take two pointer arguments
@@
 \item Write out the invariants that need to hold and check them using CXL.
-\end{enumerate}
+\end{problems}
 
@@
 \section*{Exercises}
-\begin{enumerate}
+\begin{problems}
 \item Write a synchronization module using your own implementation of \texttt{Lock}
@@
 version of the module.
-\end{enumerate}
+\end{problems}
 
@@
 \section*{Exercises}
-\begin{enumerate}
+\begin{problems}
 \item Draw additional states and steps in Figure~\ref{fig:rdwr}
@@
 bound lock.
-\end{enumerate}
+\end{problems}
 
@@
 \section*{Exercises}
-\begin{enumerate}
+\begin{problems}
 \item Why do you suppose the code in Figure~\ref{fig:rwmulti} uses a flag
@@
 write lock and then obtaining the read lock.
-\end{enumerate}
+\end{problems}
 
@@
 \section*{Exercises}
-\begin{enumerate}
+\begin{problems}
 \item Does it matter in what order the philosophers lay down their forks?
@@
 processes.  Can you fix the problem?
-\end{enumerate}
+\end{problems}
 
@@
 \section*{Exercices}
-\begin{enumerate}
+\begin{problems}
 \item Add an assertion that demonstrates that at most three
@@
 that it is a permutation of the original array.
-\end{enumerate}
+\end{problems}
 
@@
 \section*{Exercises}
-\begin{enumerate}
+\begin{problems}
 \item Change the code in Figure~\ref{fig:boundedbuffer} so that
@@
 until the stack is non-empty and then return an item.
-\item Implement a thread-safe GPU allocator by modifying Figure~\ref{fig:gpu}.
+\item \label{ex:gpu} Implement a thread-safe GPU allocator by modifying
+Figure~\ref{fig:gpu}.
 There are \texttt{N} GPUs identified by the numbers
@@
 to block.
-\end{enumerate}
+\end{problems}
 
@@
 \section*{Exercises}
-\begin{enumerate}
-\item Assume that you have only binary semaphores.  Implement counting
-semaphores using binary semaphores using split binary semaphores.
+\begin{problems}
+\item Assume that you have only binary semaphores.
+Implement counting semaphores using binary semaphores using split binary semaphores.
 \item Implement a solution to the producer/consumer problem
 (Figure~\ref{fig:boundedbuffer}) using split binary semaphores.
-\item Cornell's campus features some one-lane bridges.  Cars can
-only go in one direction at a time. Consider northbound
+\item \label{ex:onelane} Cornell's campus features some one-lane bridges.
+Cars can only go in one direction at a time. Consider northbound
 and southbound cars wanting to cross a one-lane bridge.
@@
 
-\end{enumerate}
+\end{problems}
 
@@
 \section*{Exercises}
-\begin{enumerate}
-\item Write a fair solution to the one-lane bridge problem of Chapter~\ref{ch:sbs}.
+\begin{problems}
+\item Write a fair solution to the one-lane bridge problem of
+Exercise~\ref{ex:onelane}.
 If you want to use the queue method, you can change the signature of the
@@
 semaphore.
-\end{enumerate}
+\end{problems}
 
@@
 \section*{Exercises}
-\begin{enumerate}
-\item Write a new version of the GPU allocator of Chapter~\ref{ch:bb}.  In this version,
+\begin{problems}
+\item Write a new version of the GPU allocator in Exercise~\ref{ex:gpu}.
+In this version,
 a process is allowed to allocate a set of GPUs and release a set of GPUs that it
@@
 it is fair.
-\end{enumerate}
+\end{problems}
 

### 2020-06-23 book/paper.tex
Latest

@@
 
-\begin{figure}
-\begin{code}
-\VerbatimInput[xleftmargin=5mm,numbers=left]{code/Up.py}
-\end{code}
-\caption{[Up.py] Python implementation of Figure~\ref{fig:inc}.}
-\label{fig:incpy}
-\end{figure}
-
-\begin{figure}
-\begin{code}
-\VerbatimInput[xleftmargin=5mm,numbers=left]{code/UpMany.py}
-\end{code}
-\caption{[UpMany.py] Incrementing \texttt{N} times.}
-\label{fig:incmany}
-\end{figure}
-
 \begin{problems}
@@
 
+\begin{figure}
+\begin{code}
+\VerbatimInput[xleftmargin=5mm,numbers=left]{code/Up.py}
+\end{code}
+\caption{[Up.py] Python implementation of Figure~\ref{fig:inc}.}
+\label{fig:incpy}
+\end{figure}
+
+\begin{figure}
+\begin{code}
+\VerbatimInput[xleftmargin=5mm,numbers=left]{code/UpMany.py}
+\end{code}
+\caption{[UpMany.py] Incrementing \texttt{N} times.}
+\label{fig:incmany}
+\end{figure}
+
 \chapter{The CXL Virtual Machine}
@@
 \end{code}
-\caption{[UpEnter.py] Broken attempt at fixing the code of Figure~\ref{fig:inc}.}
+\caption{[UpEnter.cxl] Broken attempt at fixing the code of Figure~\ref{fig:inc}.}
 \label{fig:incenter}
@@
 A method \texttt{m} with argument \texttt{a} is invoked in its
-most basic form as follows (assigning the result to \texttt{r}).
+most basic form as follows (assigning the result to~\texttt{r}).
 \begin{code}
@@
 def wait(c):
-    let lk = (^c).lock, blocked = True, cnt = bagCount((^c).waiters, nametag()):
+    let lk = (^c).lock, blocked = True,
+                cnt = bagCount((^c).waiters, nametag()):
         bagAdd(&(^c).waiters, nametag());
@@
             atomic:
-                if (not (^lk)) and (bagCount((^c).waiters, nametag()) <= cnt):
+                if (not (^lk)) and
+                        (bagCount((^c).waiters, nametag()) <= cnt):
                     ^lk = True;
@@
 Figure~\ref{fig:barrierdiagram} shows a high-level depiction with three
-processes.  Initially all processes are in round 0.  Then each process
+processes.  Initially all processes are in round~0.  Then each process
 can start and finish the round.  However, none of the processes can
@@
 
-\begin{quote}
-\begin{itemize}
+\section*{Exercises}
+\begin{problems}
 \item See if you can implement barrier synchronization for $N$ processes
-with just three semaphores.  Busy waiting is not allowed.
-\item How about just two?
-\end{itemize}
-\end{quote}
+with just three semaphores.  Busy waiting is not allowed.  How about just two?
+\end{problems}
 

### 2020-06-23 book/paper.tex
Adjusting to new formatting

@@
 an undesirable approach to synchronize processes.
-\item Chapter~\ref{ch:deadlock} describes \emph{deadlock}
-where a set of processes are indefinitely waiting for one another to
-release a resource.
 \item Chapter~\ref{ch:semaphore} introduces \emph{semaphores},
@@
 another approach to process synchronication.
-\item Chapter~\ref{ch:deadlockprevention} revisits deadlocks and how to
-avoid them.
+\item Chapter~\ref{ch:deadlock} describes \emph{deadlock}
+where a set of processes are indefinitely waiting for one another to
+release a resource.
 \item Chapter~\ref{ch:actor} presents the \emph{actor model} as an approach to
@@
 \begin{center}
-\includegraphics[width=4in]{figures/mutex-crop.pdf}
+\includegraphics[width=4.5in]{figures/mutex-crop.pdf}
 \end{center}
@@
 \begin{figure}
+\begin{code}
+\VerbatimInput[xleftmargin=5mm,numbers=left]{code/RWbusytest.cxl}
+\end{code}
+\caption{[RWbusytest.cxl] Test code for Figure~\ref{fig:rwbusy}.}
+\label{fig:rwbusytest}
+\end{figure}
+
+\begin{figure}
 \begin{center}
-\includegraphics[width=2in]{figures/rdwr-crop.pdf}
+\includegraphics[width=2.5in]{figures/rdwr-crop.pdf}
 \end{center}
@@
 of readers and one that maintains the number of writers.
+Figure~\ref{fig:rwbusytest} shows how the code may be used.
 We will call the lock the \emph{mutex} to distinguish it clearly from
@@
 flags.
+
 The astute reader might wonder if obtaining the mutex itself is an
@@
 (see Figure~\ref{fig:spinlocks}).
-
 However, there is a big difference.  As we pointed out, the processes that
@@
 \end{problems}
-
-\chapter{Deadlock}
-\label{ch:deadlock}
-\index{deadlock}
-\glsadd{deadlock}
-
-\begin{figure}
-\begin{code}
-\VerbatimInput[xleftmargin=5mm,numbers=left]{code/Diners.cxl}
-\end{code}
-\caption{[Diners.cxl] Dining Philosophers.}
-\label{fig:diners}
-\end{figure}
-
-When multiple processes are synchronizing access to shared resources, they
-may end up in a \emph{deadlock} situation where one or more of the processes
-end up being blocked indefinitely because each is waiting for another to give
-up a resource.
-The famous Dutch computer scientist Edsger W.~Dijkstra illustrated this using
-a scenario he called ``Dining Philosophers.''
-\index{dining philosopher}
-
-Imagine five philosopers sitting around a table, each with a plate of food in
-front of them and a fork between each two plates.  Each philosopher requires
-two forks to eat.  To start eating, a philosopher first picks up the fork on
-the left, then the fork on the right.  Each philospher likes to take breaks
-from eating to think for a while.  To do so, the philosopher puts down both
-forks.  Each philosopher repeat this procedure.  Dijkstra had them repeating
-this for ever, but for the purposes of this book, philosphers can leave
-the table when they're not eating.
-
-Figure~\ref{fig:diners} implements the dining philosophers in CXL, using a
-process for each philosopher and a lock for each fork.  If you
-run it, CXL complains that the execution may not terminate, with all five
-processes being blocked trying to acquire the lock.
-
-\begin{quote}
-\begin{itemize}
-\item Do you see what the problem is?
-\item Does it depend on \texttt{N}, the number of philosophers?
-\end{itemize}
-\end{quote}
-
-We will discuss ways of preventing deadlock later (Chapter~\ref{ch:deadlockprevention}),
-but one important
-approach to preventing deadlock is to prevent a cycle from forming, with each
-process waiting for the next process on the cycle.
-We can do this by establishing an ordering among the
-resources (in this case the forks) and, when needing more than one
-resource, always acquiring them in order.  In the case of the philosopers,
-they could prevent deadlock by always picking up the lower numbered fork
-before the higher numbered fork, like so:
-
-\begin{code}
-\begin{Verbatim}[xleftmargin=5mm,numbers=left]
-    if left < right:
-        lock(&forks[left]);
-        lock(&(orks[right]);
-    else:
-        unlock(&forks[right]);
-        unlock(&forks[left]);
-    ;
-\end{Verbatim}
-\end{code}
-
-\section*{Exercises}
-\begin{problems}
-\item Does it matter in what order the philosophers lay down their forks?
-\item Figure~\ref{fig:bank} shows an implementation of a bank with various
-accounts and transfers between those accounts.
-Unfortunately, running the test reveals that it sometimes leaves unterminated
-processes.  Can you fix the problem?
-\end{problems}
-
-\begin{figure}
-\begin{code}
-\VerbatimInput[xleftmargin=5mm,numbers=left]{code/bank.cxl}
-\end{code}
-\caption{[bank.cxl] Bank accounts.}
-\label{fig:bank}
-\end{figure}
 
@@
 
-\chapter{Deadlock Prevention}
-\label{ch:deadlockprevention}
-\index{deadlock prevention}
+\chapter{Deadlock}
+\label{ch:deadlock}
+\index{deadlock}
+\glsadd{deadlock}
+
+\begin{figure}
+\begin{code}
+\VerbatimInput[xleftmargin=5mm,numbers=left]{code/Diners.cxl}
+\end{code}
+\caption{[Diners.cxl] Dining Philosophers.}
+\label{fig:diners}
+\end{figure}
+
+When multiple processes are synchronizing access to shared resources, they
+may end up in a \emph{deadlock} situation where one or more of the processes
+end up being blocked indefinitely because each is waiting for another to give
+up a resource.
+The famous Dutch computer scientist Edsger W.~Dijkstra illustrated this using
+a scenario he called ``Dining Philosophers.''
+\index{dining philosopher}
+
+Imagine five philosopers sitting around a table, each with a plate of food in
+front of them and a fork between each two plates.  Each philosopher requires
+two forks to eat.  To start eating, a philosopher first picks up the fork on
+the left, then the fork on the right.  Each philospher likes to take breaks
+from eating to think for a while.  To do so, the philosopher puts down both
+forks.  Each philosopher repeat this procedure.  Dijkstra had them repeating
+this for ever, but for the purposes of this book, philosphers can leave
+the table when they're not eating.
+
+Figure~\ref{fig:diners} implements the dining philosophers in CXL, using a
+process for each philosopher and a lock for each fork.  If you
+run it, CXL complains that the execution may not terminate, with all five
+processes being blocked trying to acquire the lock.
+
+\begin{quote}
+\begin{itemize}
+\item Do you see what the problem is?
+\item Does it depend on \texttt{N}, the number of philosophers?
+\item Does it matter in what order the philosophers lay down their forks?
+\end{itemize}
+\end{quote}
+
+% \chapter{Deadlock Prevention}
+% \label{ch:deadlockprevention}
+% \index{deadlock prevention}
 
@@
 
-As introduced in Chapter~\ref{ch:deadlock}, deadlock is the situation when there are
-a collection of processes waiting for one another to release one or more resources.
+\noindent
 There are four conditions that must hold for deadlock to occur~\cite{CES71}:
@@
 
-The \texttt{Preemption} approach suggested by Havender is to allow processes to back out.
+The \emph{Preemption} approach suggested by Havender is to allow processes to back out.
 While this could be done, this invariably leads to a busy waiting solution
@@
 finally is able to get all of them.
-The \emph{No Circular Waiting} approach to order the resources was already presented in
-Chapter~\ref{ch:deadlock}
-
-There is, however, another approach. which is sometimes called
+
+The \emph{No Circular Waiting} approach
+is to prevent a cycle from forming, with each
+process waiting for the next process on the cycle.
+We can do this by establishing an ordering among the
+resources (in this case the forks) and, when needing more than one
+resource, always acquiring them in order.  In the case of the philosopers,
+they could prevent deadlock by always picking up the lower numbered fork
+before the higher numbered fork, like so:
+
+\vspace{1em}
+\begin{code}
+\begin{Verbatim}[xleftmargin=5mm,numbers=left]
+    if left < right:
+        lock(&forks[left]);
+        lock(&(orks[right]);
+    else:
+        unlock(&forks[right]);
+        unlock(&forks[left]);
+    ;
+\end{Verbatim}
+\end{code}
+\vspace{1em}
+
+This completes all the Havender methods.
+There is, however, another approach, which is sometimes called
 \emph{deadlock avoidance}.
-\index{deadlock avoidance}.
+\index{deadlock avoidance}
 In the case of the Dining Philosophers, we want to avoid the situation where each
@@
 \label{fig:dinersavoid}
+\end{figure}
+
+\section*{Exercises}
+\begin{problems}
+\item Figure~\ref{fig:bank} shows an implementation of a bank with various
+accounts and transfers between those accounts.
+Unfortunately, running the test reveals that it sometimes leaves unterminated
+processes.  Can you fix the problem?
+\end{problems}
+
+\begin{figure}
+\begin{code}
+\VerbatimInput[xleftmargin=5mm,numbers=left]{code/bank.cxl}
+\end{code}
+\caption{[bank.cxl] Bank accounts.}
+\label{fig:bank}
 \end{figure}

### 2020-06-23 book/paper.tex
Latest

@@
 \begin{code}
+{\small
 \VerbatimInput[xleftmargin=5mm,numbers=left]{code/RWfair.cxl}
+}
 \end{code}
@@
 \begin{code}
+{\small
 \VerbatimInput[xleftmargin=5mm,numbers=left]{code/RWqueue.cxl}
+}
 \end{code}
@@
 use \texttt{wait} and \texttt{signal} instead of using \texttt{P}
-\texttt{V}.
+and \texttt{V}.
 
@@
 Mesa monitors also have condition variable that a process can wait on.
-Like in Hoare monitors, the \texttt{wait} operation release the mutex.
+Like in Hoare monitors, the \texttt{wait} operation releases the mutex.
 The most important difference is in what \texttt{signal} does.
@@
 When a process $p$ invokes \texttt{notify}, it does not immediately pass 
-the baton to some process $q$ requiring the baton.  Instead, $p$ holds onto
-the baton until it leaves the monitor.  At that point $q$ gets the baton,
-assuming there even was such a process $q$.
+the baton to some process $q$ requiring the baton (assuming there even
+is such a process).  Instead, $p$ holds onto
+the baton until it leaves the monitor.  At that point, any process that
+was notified will have a chance to obtain the baton.
 
@@
 Here a bathroom represents the critical section, allowing only one
-process at a time say.  The bedrooms represent condition variables.
+process at a time.  The bedrooms represent condition variables.
 There is some condition associated with each bedroom.
@@
 processes.  Can you fix the problem?
+\item Add a method \texttt{total()} to the solution of the previous question
+that computes the total over all balances.
+It needs to obtain a lock on all accounts.  Make sure that
+it cannot cause deadlock.
+\item Use method \texttt{total()} to ensure that the sum total of the bank
+accounts at the beginning of the test is the same as at the end.
 \end{problems}

### 2020-06-23 book/paper.tex
Added more to CXL machine chapter

@@
 The default tag is the first argument to the method.
-In Figure~\ref{fig:inc}, the created processes have name tags
+A name tag is represented as a dictionary with keys \texttt{.name}
+and \texttt{.tag}, but we shall often abbreviate it using the notation
+\texttt{name/tag}.
+For example, in Figure~\ref{fig:inc}, the created processes have name tags
 \texttt{incrementer/0}, \texttt{incrementer/1}, and \texttt{main/()}.
@@
 This makes precise the concurrency problem that we encountered.
+
+CXL can report a variety of failure types:
+\begin{itemize}
+\item \texttt{Safety violation}: this means something went wrong with
+at least one of the executions of the program that it tried.  This
+can include a failing assertion, divide by zero, using an uninitialized
+or non-existent variable, dividing a set by an integer, and so on.
+CXL will print at trace of the shortest bad execution that it found.
+\item \texttt{Infinite loop}: some process ran into an infinite loop
+in which it got back to the same state without having accessed any
+shared variables.
+\item \texttt{Non-terminating States}: CXL found one or more states
+from which there does not exist an execution such that all processes
+terminate.  CXL will not only print the non-terminating state with
+the shortest trace, but also the list of processes that are running
+at that state, along with their name tags and program counters.
+\item \texttt{Stopped States}: Similar to non-terminating states,
+these are states in which some processes are left that cannot
+terminate.  Chapter~\ref{ch:synch} will explain what it means for
+a process to be stopped.
+\end{itemize}
 

### 2020-06-23 book/paper.tex
Latest

@@
   description={is a concurrency model where there are no shared variables, only
-      processes with private variables and exchanging messages}
+      processes with private variables that communicate through message passing}
 }
@@
 \item See what happens if, instead of initializing \texttt{result} to 0,
-you initialize it to 1.  (You do not need to understand the error report at this time.)
-Explain in English what the problem is.
+you initialize it to 1.  (You do not need to understand the error report at this time.  They will be explained in more detail in Chapter~\ref{ch:cxlmachine}.)
 \item Write a CXL program that computes squares by repeated adding.  So the program
@@
 the solution allows more than one Italian philosopher to eat at the same time?
-\item Figure~\ref{fig:qsort} shows an iterative implementation of the Qsort
+\item \label{ex:qsort}
+Figure~\ref{fig:qsort} shows an iterative implementation of the Qsort
 algorithm, and Figure~\ref{fig:qsorttest} an accompanying test program.
@@
 The producer/consumer pattern is common.  Processes may be arranged
-in pipelines, where each upstream process is a producer and each downstream
+in \emph{pipelines},
+\index{pipeline}
+where each upstream process is a producer and each downstream
 process is a consumer.
@@
 
-\chapter{Actors and Synchronized Queues}
+\chapter{Actors and Message Passing}
 \label{ch:actor}
 \index{actor model}
+\index{message passing}
 \glsadd{actor model}
@@
 
-Some programming languages favor a completely different way of implementing
+Some programming languages favor a different way of implementing
 synchronization using so-called \emph{actors}~\cite{HBS73}.  Actors are
-processes that have only private memory and communicate through message queues.
+processes that have only private memory and communicate through message passing.
 See Figure~\ref{fig:actorpic} for an illustration.
@@
 \item \texttt{Queue()} returns a new message queue;
-\item \texttt{enqueue(q, item)} add \texttt{item} to the queue pointed to by \texttt{q};
+\item \texttt{enqueue(q, item)} adds \texttt{item} to the queue pointed to by \texttt{q};
 \item \texttt{dequeue(q)} waits for and returns an item on the queue pointed to by \texttt{q}.
@@
 
-Figure~\ref{fig:actor} shows a CXL solution to the producer/consumer problem using
-an actor (as illustrated in Figure~\ref{fig:actorpic}.
+Figure~\ref{fig:actor} shows a CXL solution to the producer/consumer problem
+(with an unbounded buffer)
+using an actor (as illustrated in Figure~\ref{fig:actorpic}.
 \texttt{pc\_actor} is essentially a matchmaker: it matches
@@
 requests.
+
+\section*{Exercises}
+\begin{problems}
+\item Implement a version of parallel qsort (see Example~\ref{ex:qsort})
+using the actor model.  Use an actor for each worker as well as a main
+actor that waits for each worker actor to have finished.
+\item Actors and message queues are good for building pipelines.
+Develop a pipeline that computes Mersenne primes (primes that are one less
+than a factor of two).  Write four actors:
+\begin{enumerate}
+\item an actor that generates a sequence of integers 1 through \texttt{N};
+\item an actor that receives integers and forwards only those that are prime;
+\item an actor that receives integers and forwards only those that are one
+less than a factor of two;
+\item an actor that receives integers but otherwise ignores them.
+\end{enumerate}
+Configure two versions of the pipeline, one that first checks if a number
+is prime and then if it is one less than a factor of two, the other
+in the opposite order.  Which do you think is better?
+\end{problems}
 

### 2020-06-23 book/paper.tex
Latest

@@
 \begin{problems}
-\item Implement a version of parallel qsort (see Example~\ref{ex:qsort})
-using the actor model.  Use an actor for each worker as well as a main
-actor that waits for each worker actor to have finished.
 \item Actors and message queues are good for building pipelines.
@@
 in the opposite order.  Which do you think is better?
+\item Bonus question: implement a version of parallel qsort
+(see Example~\ref{ex:qsort}) using the actor model.
 \end{problems}
@@
 
-Method \texttt{produce(item)} uses \texttt{inc \&(back)} to allocate
+Method \texttt{produce(item)} uses \texttt{inc(\&back)} to allocate
 the next available slot in the \texttt{items} array.
@@
 synchronization algorithms~\cite{}.
+There are also various algorithms where read-only access does not
+require a lock but updates do require a lock---these are useful in
+situations where most accesses are read-only.
 

### 2020-06-23 book/paper.tex
Starting to add more non-blocking examples
Added something about the alloc module

@@
 
+\begin{figure}
+\begin{code}
+\VerbatimInput[xleftmargin=5mm,numbers=left]{code/lst1.cxl}
+\end{code}
+\caption{[lst1.cxl] List with non-blocking read operations, part 1.}
+\label{fig:lst1}
+\end{figure}
+
+\begin{figure}
+\begin{code}
+\VerbatimInput[xleftmargin=5mm,numbers=left]{code/lst2.cxl}
+\end{code}
+\caption{[lst2.cxl] List with non-blocking read operations, part 2.}
+\label{fig:lst2}
+\end{figure}
+
 \chapter{Barrier Synchronization}
@@
 
+The \texttt{alloc} module
+\index{alloc module}
+support thread-safe dynamic allocation of ``records.''
+Each record has two fields: \texttt{.data} and \texttt{.next} that
+can be used arbitrarily, although \texttt{.next} is intended to be
+used as a pointer to another record.
+There are just two methods:
+
+\vspace{1em}
+\begin{tabular}{|l|l|}
+\hline
+\texttt{rec\_alloc()} & allocate a record \\
+\hline
+\texttt{rec\_free(r)} & free record \texttt{r} \\
+\hline
+\end{tabular}
+\vspace{1em}
+
+CXL supports an address value \texttt{NULL} that can be
+used as a null pointer.  \texttt{rec\_alloc} returns \texttt{NULL}
+when running out of memory.
+
 The \texttt{bag} module
@@
 \hline
-\texttt{qsort(t)} & sort list \texttt{t} using quicksort\\
+\texttt{listQsort(t)} & sort list \texttt{t} using quicksort\\
 \hline

### 2020-06-24 book/paper.tex
Now supports += etc

@@
 Up.cxl:1 def incrementer(self):
-   0 Jump 32
-   1 Frame incrementer(self) end=9
+   0 Jump 34
+   1 Frame incrementer(self) end=11
 Up.cxl:2     count = count + 1;
-   2 Push 1
-   3 Load count
+   2 Load count
+   3 Push 1
    4 2-ary +
@@
 Up.cxl:3     done[self] = True;
-   6 Push True
-   7 PushVar self
-   8 Store done 1
-   9 Return
+   6 LoadVar self
+   7 PushAddress done
+   8 Address 2
+   9 Push True
+   10 Store
+   11 Return
 \end{verbatim}
@@
 ==== Safety violation ====
-CXL Assertion failed ('assert', 'Up.cxl', 9, 5) 1
-__init__/() [0,32-47]       48 dict{ .count:0, .done:dict{ 0:False, 1:False } }
-incrementer/0 [1-4]          5 dict{ .count:0, .done:dict{ 0:False, 1:False } }
-incrementer/1 [1-8]          9 dict{ .count:1, .done:dict{ 0:False, 1:True  } }
-incrementer/0 [5-8]          9 dict{ .count:1, .done:dict{ 0:True,  1:True  } }
-main/() [11-15,18-22,24-29] 29 dict{ .count:1, .done:dict{ 0:True,  1:True  } }
+CXL Assertion failed ('assert', 'Up.cxl', 9, 5): 1
+__init__/() [0,34-49]       50 dict{.count:0, .done:dict{0:False, 1:False}}
+incrementer/0 [1-4]          5 dict{.count:0, .done:dict{0:False, 1:False}}
+incrementer/1 [1-10]        11 dict{.count:1, .done:dict{0:False, 1:True }}
+incrementer/0 [5-10]        11 dict{.count:1, .done:dict{0:True,  1:True }}
+main/() [13-17,20-24,26-31] 31 dict{.count:1, .done:dict{0:True,  1:True }}
 \end{verbatim}
@@
 \begin{enumerate} \setcounter{enumi}{1}
+\item The \texttt{Load} instruction pushes the value of the
+\texttt{count} variable onto the stack.
 \item the \texttt{Push} instruction pushes the constant 1
 onto the stack of the process.
-\item The \texttt{Load} instruction pushes the value of the
-\texttt{count} variable onto the stack.
 \item \texttt{2-ary} is a \texttt{+} operation with 2 arguments.
@@
 adds them and pushes the result back onto the stack.
-\item The \texttt{Store} instruction pops pops
+\item The \texttt{Store} instruction pops
 a CXL value (the sum of the \texttt{count} variable and 1) and
@@
 \texttt{count} but stopping just before storing 1 into \texttt{count};
-\item Process 1 executed instructions 1 through 8, storing 1 into
+\item Process 1 executed instructions 1 through 10, storing 1 into
 into \texttt{count} and storing \texttt{True} into \texttt{done[1]};
@@
 \begin{center}
-\includegraphics[width=4.5in]{figures/mutex-crop.pdf}
+\includegraphics[width=4in]{figures/mutex-crop.pdf}
 \end{center}
@@
     atomic:
-        ^ptr = ^ptr + 1;
+        ^ptr += 1;
     ;
@@
                 if (^sema) > 0:
-                    ^sema = (^sema) - 1;
+                    ^sema -= 1;
                     blocked = False;
@@
     atomic:
-        ^sema = (^sema) + 1;
+        ^sema += 1;
     ;
@@
 \hline
-integers & \texttt{..., -2, -1, 0, 1, 2, ...} \\
+integers & \texttt{-inf, ..., -2, -1, 0, 1, 2, ..., inf} \\
 \hline
@@
 \hline
-Address & \texttt{\&lock, \&flags[2]} \\
+Address & \texttt{\&lock, \&flags[2], NULL} \\
 \hline
@@
 ordered;
-\item an address is a list of atoms.  Addresses are lexicographically
-ordered accordingly;
+\item an address is a list of atoms.  \texttt{NULL} is the empty list of atoms.
+Addresses are lexicographically ordered accordingly;
 \item contexts (Appendix~\ref{app:context}) are ordered by hash.
@@
 \hline
-\texttt{bagFromset(s)} & create a bag from set \texttt{s}\\
+\texttt{bagFromSet(s)} & create a bag from set \texttt{s}\\
 \hline

### 2020-06-24 book/paper.tex
Making it easier to run test code

@@
 \begin{code}
+\VerbatimInput[xleftmargin=5mm,numbers=left]{code/BBsematest.cxl}
+\end{code}
+\caption{[BBsematest.cxl] Test program for Figure~\ref{fig:boundedbuffer}.}
+\label{fig:bbtest}
+\end{figure}
+
+\begin{figure}
+\begin{code}
 \begin{verbatim}
@@
 
-We can run the code through CXL, but what would it test?  There are no
+Figure~\ref{fig:bbtest} invokes the code of Figure~\ref{fig:boundedbuffer}.
+We can run this through CXL, but what would it test?  There are no
 assert statements in the code.

### 2020-06-24 book/paper.tex
Now most code can run as is.

@@
 \begin{code}
-\VerbatimInput[xleftmargin=5mm,numbers=left]{code/RWbusy.cxl}
-\end{code}
-\caption{[RWbusy.cxl] Busy-Waiting Reader/Writer Lock implementation.}
+\VerbatimInput[xleftmargin=5mm,numbers=left]{code/RW.cxl}
+\end{code}
+\caption{[RW.cxl] Busy-Waiting Reader/Writer Lock implementation.}
 \label{fig:rwbusy}
@@
 \begin{code}
-\VerbatimInput[xleftmargin=5mm,numbers=left]{code/RWbusytest.cxl}
-\end{code}
-\caption{[RWbusytest.cxl] Test code for Figure~\ref{fig:rwbusy}.}
+\VerbatimInput[xleftmargin=5mm,numbers=left]{code/RWtest.cxl}
+\end{code}
+\caption{[RWtest.cxl] Test code for Figure~\ref{fig:rwbusy}.}
 \label{fig:rwbusytest}
@@
 that does not busy-wait.
+You can test the code by running \texttt{cxl -m RW=RWlock RWtest}.
 It uses two ordinary locks: \texttt{rwlock} is used by both readers and writers,

### 2020-06-24 book/paper.tex
Update

@@
 
+\glsadd{lock}
 You may already have heard of the concept of a \emph{lock}
@@
 it could be used to implement a critical section.
-\glsadd{lock}
 The idea is that the lock is like a baton that at most one process can own
@@
 lock first and release it upon exiting the critical section.
+Note that the word ``lock'' does not describe the concept well---it really
+is more like a baton or token that only one process can hold.
+
 Using a lock is a good thought, but how does one implement one?

### 2020-06-24 book/paper.tex
Latest

@@
 \hline
-Load $v$ & push shared variable $v$ (\texttt{\^} is indirect) onto the stack \\
-\hline
-LoadVar $v$ & load process variable $v$ and push it onto the stack \\
+Load [$v$] & push the value of a shared variable onto the stack \\
+\hline
+LoadVar [$v$] & push the value of a process variable onto the stack \\
 \hline
@@
 \hline
-Store $v$ $[n]$ & pop a value from the stack and store it in shared variable $v$ (\texttt{\^} is indirect) \\
-\hline
-StoreVar $v$ & pop a value from the stack and store it in process variable $v$ \\
+Store [$v$] & pop a value from the stack and store it in a shared variable \\
+\hline
+StoreVar [$v$] & pop a value from the stack and store it in a process variable \\
 \hline

### 2020-06-24 book/paper.tex
Worked on monitor chapter

@@
 and if there are processes in the bedroom.
-If there are such bedrooms, $p$ selects one and lets that process into
+If there are such bedrooms, $p$ selects one and one of its processes into
 the bathroom.  Let's call that process $q$.  Importantly, when $q$ enters
@@
 bedroom is considered a no-op in Mesa.  (Hoare monitors do not allow
-using \texttt{signal} on an empty bathroom.) Mesa also has a function called
-\texttt{notifyAll}
+using \texttt{signal} on an empty bathroom.) Mesa also provides an
+operation called \texttt{notifyAll}
 \index{notifyAll}
@@
 that, when applied to
-a bedroom, let's all processes in that bedroom into the hall.
+a bedroom, lets all processes in that bedroom into the hall.
+This is not possible with Hoare monitors: when needing to wake up a set
+of processes, only one can be signaled at a time.  Each of those processes
+then has to signal the next until all processes have been signaled.
 
@@
 programming languages.
-
 That said, few programming languages provide full support for monitors.
@@
 
-The \texttt{synch} module supports Mesa condition variables.  However, like C, Java, and Python,
+The CXL \texttt{synch} module supports Mesa condition variables.
+However, like C, Java, and Python,
 it does not have built-in language support for them but instead associates condition variables
 with explicit locks.
-Figure~\ref{fig:cv} shows the \texttt{synch} interface to condition variables.
+Figure~\ref{fig:cv} shows the implementation of
+condition variables in the \texttt{synch} module.
 \texttt{Condition(lk)} creates a new condition variable associated with the given pointer
@@
 and a bag of name tags of processes waiting on the condition variable.
-(The \texttt{synchS} library instead has a list of contexts.)
+(The \texttt{synchS} library instead uses a list of contexts.)
 
@@
 
-To illustrate how Mesa condition variables are used in practice, we demonstrate, once again,
-using an implementation of reader/writer locks.
+To illustrate how Mesa condition variables are used in practice, we demonstrate using,
+once again, an implementation of reader/writer locks.
 Figure~\ref{fig:RWcv} shows the code.  \texttt{rwlock} is the global lock or mutex for the
@@
 invoked within a \texttt{while} loop that checks for the condition that the process is
-waiting for.  We explained why in Chapter~\ref{ch:monitors}: when a process resumes after
+waiting for.  We explained why above: when a process resumes after
 being notified, it is not guaranteed that the condition it was waiting for holds, even
 if it did at the time of invoking \texttt{notify} or \texttt{notifyAll}.
-It is \emph{imperative} that you always have a \texttt{while} loop around any invocation
-of \texttt{wait}.
-Many implementation of condition variables depend on this, and optimizations of the
-code allow so-called ``spurious wakeups,'' where \texttt{wait} may sometimes return even
-if it has not been notified.
+It is \emph{imperative} that there is always a \texttt{while} loop around any invocation
+of \texttt{wait} containing the negation of the condition that the process is waiting for.
+Many implementation of condition variables depend on this, and optimized implementations
+of condition variables often allow so-called ``spurious wakeups,'' where \texttt{wait}
+may sometimes return even if the conditon variable has not been notified.
 
@@
 get to enter next or one of the writers gets to enter next.
+
+Much of the complexity of programming with Mesa condition variables is in figuring out
+when to invoke \texttt{notify} and when to invoke \texttt{notifyAll}.  As a rule of
+thumb: be conservative---it is better to wake up too many processes than too few.
+Waking up too many processes may lead to some inefficiency, but waking up to few
+may cause the application to get stuck.  CXL can be particularly helpful here, as
+it examines every corner case.  You can safely try to replace each \texttt{notifyAll}
+with \texttt{notify} and see if every possible execution of the the application still
+terminates.
 

### 2020-06-24 book/paper.tex
Made a pass over the deadlock chapter

@@
 \begin{problems}
-\item Write a new version of the GPU allocator in Exercise~\ref{ex:gpu}.
+\item Implement semaphores using Mesa condition variables.
+\item Write a new version of the GPU allocator in Exercise~\ref{ex:gpu}
+using Mesa condition variables.
 In this version,
@@
 possible, as demonstrated in Chapter~\ref{ch:nonblocking}).
-Havender~\cite{Havender68} proposed the following techniques that avoid the remaining
-three conditions:
+Havender proposed the following techniques that avoid the remaining
+three conditions~\cite{Havender68}:
 
@@
 \index{multiple conditions, waiting on}
-The condition in the \texttt{while} statement is the condition that the
-diner is waiting for to be \texttt{False}.  Within the \texttt{while} statement,
-there is an \texttt{if} statement for each disjunct of that condition.
+The condition in the \texttt{while} statement is the negation of the
+condition that the diner is waiting and consists of two disjuncts.
+Within the \texttt{while} statement,
+there is an \texttt{if} statement for each disjunct.
 The code waits for either or both forks if necessary.  After that, it goes
@@
 This completes all the Havender methods.
-There is, however, another approach, which is sometimes called
-\emph{deadlock avoidance}.
+There is, however, another approach, which is sometimes called deadlock
+\emph{avoidance}
 \index{deadlock avoidance}
+instead of deadlock \emph{prevention}.
 In the case of the Dining Philosophers, we want to avoid the situation where each
@@
 The problem with these kinds of schemes is that one needs to know ahead of time
-what the maximum number of resources is that each process wants to allocate,
-making them generally quite impractical.
+the set of processes and what the maximum number of resources is that each process
+wants to allocate, making them generally quite impractical.
 
@@
 it cannot cause deadlock.
-\item Use method \texttt{total()} to ensure that the sum total of the bank
-accounts at the beginning of the test is the same as at the end.
+\item Implement an extra process that checks, once, if the total of
+the balances is the same as that at the beginning (using an assertion).
+Is once enough to determine that no money gets lost or create, or should it do
+so continually?
 \end{problems}

### 2020-06-24 book/paper.tex
Made pass over actor chapter

@@
 synchronization using so-called \emph{actors}~\cite{HBS73}.  Actors are
-processes that have only private memory and communicate through message passing.
+processes that have only private memory and communicate through
+\emph{message passing}.
 See Figure~\ref{fig:actorpic} for an illustration.
@@
 handles one message at a time, serializing all access to the data it owns.
+As message queues are FCFS, starvation is prevented.
 
@@
 
-Figure~\ref{fig:actor} shows a CXL solution to the producer/consumer problem
-(with an unbounded buffer)
+Figure~\ref{fig:actor} shows a CXL solution to the
+``multiple producer / multiple consumer problem'' (with an unbounded buffer)
 using an actor (as illustrated in Figure~\ref{fig:actorpic}.
@@
 
+The shared memory consists entirely of message queues.
 The \texttt{produce} method takes two arguments: a pointer to the queue
@@
 situations where most accesses are read-only.
+
+TODO: Talk about seqlock perhaps.
 

### 2020-06-24 book/paper.tex
More detail on the machine instructions

@@
 \hline
-Address & compute address \\
+Address $n$ & compute address \\
 \hline
@@
 \begin{itemize}
+\item \texttt{Load} and \texttt{LoadVar} instructions have an optional
+variable name: if omitted the top of the stack must contain the path
+name of the variable as an AddressValue.
+\item \texttt{Store} and \texttt{StoreVar} instructions have an optional
+variable name.  In both cases the value to be assigned is on the top
+of the stack.  If the name is omitted, the path name is underneath that
+value on the stack.
 \item The effect of the \texttt{Apply} instructions depends much on $m$.
@@
 \texttt{Continue} instruction.
+\item The \texttt{Address} instruction expects $n$ values $v_1, ..., v_n$ on the stack.
+The top value must be an address value.  The remaining values are keys into dictionaries.
+The instruction then computes the address of \texttt{\^}$v_1~v_2~...~v_n$.
 \end{itemize}

### 2020-06-24 book/paper.tex
Latest

@@
 \hline
-\texttt{rec\_alloc()} & allocate a record \\
-\hline
-\texttt{rec\_free(r)} & free record \texttt{r} \\
+\texttt{recAlloc()} & allocate a record \\
+\hline
+\texttt{recFree(r)} & free record \texttt{r} \\
 \hline
@@
 
-CXL supports an address value \texttt{NULL} that can be
-used as a null pointer.  \texttt{rec\_alloc} returns \texttt{NULL}
+The usage is similar to \texttt{malloc} and \texttt{free} in C.
+CXL also supports an address value \texttt{NULL} that can be
+used as a null pointer.  \texttt{recAlloc} returns \texttt{NULL}
 when running out of memory.
@@
 \hline
-PushAddress $v$ & push the address of shared variable $v$ \\
+PushAddress $v$ & push the address of variable $v$ \\
 \hline
@@
 \begin{itemize}
+\item The \texttt{Address} instruction expects $n$ values $v_1, ..., v_n$ on the stack.
+The top value must be an address value.  The remaining values are keys into dictionaries.
+The instruction then computes the address of \texttt{\^}$v_1~v_2~...~v_n$.
+\item Even though CXL code does not allow taking addresses of process variables, both
+shared and process variables can have addresses.
 \item \texttt{Load} and \texttt{LoadVar} instructions have an optional
-variable name: if omitted the top of the stack must contain the path
-name of the variable as an AddressValue.
+variable name: if omitted the top of the stack must contain the address of
+the variable.
 \item \texttt{Store} and \texttt{StoreVar} instructions have an optional
 variable name.  In both cases the value to be assigned is on the top
-of the stack.  If the name is omitted, the path name is underneath that
+of the stack.  If the name is omitted, the address is underneath that
 value on the stack.
@@
 \texttt{Continue} instruction.
-\item The \texttt{Address} instruction expects $n$ values $v_1, ..., v_n$ on the stack.
-The top value must be an address value.  The remaining values are keys into dictionaries.
-The instruction then computes the address of \texttt{\^}$v_1~v_2~...~v_n$.
 \end{itemize}
@@
 with the same state at the same time.
-
 A context consists of the following:

### 2020-06-24 book/paper.tex
Finished a draft of the non-blocking synchronization chapter

@@
   reachable state of a process}
+}
+\newglossaryentry{linearizable}
+{
+  name=linearizable,
+  description={is a consistency condition for concurrent access to an object, requiring
+    that each access must appear to execute atomically sometime between the invocation
+    of the access and its completion}
 }
@@
 
+\glsadd{invariant}
 While model checking does not in general prove an algorithm correct,
@@
 \index{invariant}
-\glsadd{invariant}
 predicates that must hold for every state in the
@@
 \index{P}
-\index{Procure}
+\index{procure}
 operation.  You can think of \texttt{P} standing
@@
 \index{V}
-\index{Vacate}
+\index{vacate}
 operation.  You can think of \texttt{V} standing for \emph{vacate},
@@
 \begin{problems}
-\item Figure~\ref{fig:bank} shows an implementation of a bank with various
+\item \label{ex:bank} Figure~\ref{fig:bank} shows an implementation of a bank with various
 accounts and transfers between those accounts.
@@
 if some process is slow for some reason while being in the critical section,
-the other processes have to wait for it to finish the critical section.
-In this chapter, we will have a brief look at how one can develop concurrent
-code in which processes never have to wait for other processes.
-For this, we will revisit the producer/consumer problem, in which a
-set of processes are producing items and a set of processes are
-consuming items.
-
+the other processes have to wait for it to finish executing the critical section.
+In this chapter, we will have a look at how one can develop concurrent
+code in which processes do not have to wait for other processes.
+
+As a first example, we will revisit the producer/consumer problem.
 The code in Figure~\ref{fig:hw} is based on code developed by Herlihy and
@@
 On the contrary, it helps them because it creates less contention.
-
-The solution is not practical because the \texttt{items} array must be
-unbounded if an unbounded number of items may be produced.  Each slot
-in the array is only used once, which is inefficient.  Also, the \texttt{inc}
-and \texttt{swap} interlock instructions are not universally available
-on existing hardware.
+Unfortunately, the solution is not practical for the following reasons:
+\begin{itemize}
+\item The \texttt{items} array must be of infinite size if an unbounded number
+of items may be produced;
+\item Each slot in the array is only used once, which is inefficient;
+\item the \texttt{inc} and \texttt{swap} interlock instructions are not
+universally available on existing processors.
+\end{itemize}
 However, in the literature there are many examples of practical
@@
 synchronization algorithms~\cite{}.
-There are also various algorithms where read-only access does not
-require a lock but updates do require a lock---these are useful in
-situations where most accesses are read-only.
-
-TODO: Talk about seqlock perhaps.
+
+There are also various algorithms where read-only access is wait-free
+but updates do require a lock---these are useful in situations where
+most accesses are read-only and the performance of updates is less
+critical~\cite{}.
+Figures~\ref{fig:lst1} and~\ref{fig:lst2} implements an ordered linked
+list of integers without duplicates.
+There are two update operations:
+values can be added using \texttt{lst\_insert} or deleted using
+\texttt{lst\_remove}.
+There is also a read-only operation: \texttt{lst\_contains} checks if
+a particular value is in the list.
+
+\glsadd{linearizable}
+The implementation uses the \texttt{alloc} module for dynamic allocation
+\index{dynamic allocation}
+of nodes in the list using \texttt{recAlloc} and \texttt{recFree}.
+The read-only operation \texttt{lst\_contains} is wait\_free: it scans
+through the list without having to obtain a lock.
+Nonetheless, the implementation of the list is
+\emph{linearizable}~\cite{HW90},
+\index{linearizable}
+a strong notion of consistency that
+makes it appear as if each of the operations executes atomically at
+some point between their invocation and return.
+
+The list has two ``book-end'' nodes with values \texttt{-inf} and
+\texttt{inf} (like the Python \texttt{math.inf} constants).
+Each node has a lock, a value, and \texttt{next}, a pointer to the next node
+(which is \texttt{NULL} for the final \texttt{inf} node).
+The \texttt{lst\_find(lst, v)} method is a helper function for update
+methods that
+finds and locks two consecutive nodes \texttt{before} and \texttt{after}
+such that \texttt{before.value $<$ v $\le$ after.value}.
+It does so by performing something called \emph{hand-over-hand locking}.
+\index{hand-over-hand locking}
+It first locks the first node, which is the \texttt{-inf} node.
+Then, iteratively, it obtains a lock on the next node and release the
+lock on the last one, and so on, similar to climbing a tree
+hand-over-hand.
+
+Having the two adjoining nodes locked, implementing \texttt{lst\_insert}
+and \texttt{lst\_remove} is straightforward.
+The \texttt{lst\_contains} method can simply scan through the list
+because the \texttt{next} pointers are updated atomically in such a
+way that they always point to a legal suffix of the list.
 
@@
 \end{figure}
+
+\section*{Exercises}
+\begin{problems}
+\item A \emph{seqlock} consists of a lock and a version number.
+An update operation acquires the lock, increments the version number, makes the
+changes to the data structure, and then releases the lock.  A read-only operation
+does not use the lock.  Instead, it retrieves the version number,
+reads the data structure, and then checks if the
+version number has not changed.  If it has, read-only operation is retried.
+Use a seqlock to implement a bank much like Exercise~\ref{ex:bank}, with
+one seqlock for the entire bank (i.e., no locks on individual accounts).
+Method \texttt{transfer} is an update operation; method \texttt{total} is a
+read-only operation.  Explain how a \texttt{seqlock} can lead to starvation.
+\end{problems}
 

### 2020-06-24 book/paper.tex
Made pass over non-blocking synch

@@
 \label{ch:cxlmachine}
-\index{CXL virtual machine}
+\index{CXL Virtual Machine}
 
@@
 
+Determining if an implementation of a concurrent data structure
+is linearizable involves finding what are known as the
+\emph{linearization points}
+\index{linearization point}
+of the operations in an execution.  These are the unique
+points in time at which an operation appears to execute atomically.
+The linearization points for the \texttt{lst\_insert} and
+\texttt{lst\_remove} operations coincide exactly with the update
+of the \texttt{before.next} pointer.
+The linearization point of a \texttt{lst\_contains} method
+execution depends on whether the value is found or not.
+If found, it coincides with retrieving the pointer to the node
+that has the value.
+If not found, it coincides with retrieving the pointer to the
+\texttt{inf} node.
+
 \begin{figure}
@@
 \begin{problems}
+\item Add read-only methods to the data structure in Figures~\ref{fig:lst1}
+and~\ref{fig:lst2} that report the size of the list, the minimum value in the
+list, the maximum value in the list, and the sum of the values in the list.
+Are they linearizable?  If so, what are their linearization points?
 \item A \emph{seqlock} consists of a lock and a version number.

### 2020-06-24 book/paper.tex
Latest

@@
 way that they always point to a legal suffix of the list.
+An invariant of the algorithm is that at any point in time the
+list is ``complete'', starting with a \texttt{-inf} node and ending
+with a \texttt{inf} node.
 
@@
 with just three semaphores.  Busy waiting is not allowed.  How about just two?
+\item Implement barrier synchronization with Mesa condition variables.
 \end{problems}
@@
 \texttt{lv = e} & \texttt{lv} is an lvalue and \texttt{e} is an expression\\
+\hline
+\texttt{lv [op]= e} & op is one of \texttt{+}, \texttt{-}, \texttt{*}, \texttt{/}, 
+\texttt{and}, and \texttt{or}\\
 \hline

### 2020-06-24 book/paper.tex
Made a pass over barrier synchronization

@@
 parallel computing, used, among others, for scalable simulation.
-Barrier synchronization is almost the opposite of critical sections:
+A barrier is almost the opposite of a critical section:
 the intention is to get a group of processes to run some code at the
-same time.
-More precisely, the processes run in rounds.  At the end of each round
-there is a so-called \emph{barrier} where processes wait until all
-processes have completed the round.
+same time, instead of having them execute it one at a time.
+More precisely, with barrier synchronization the processes execute in rounds.
+Between each round there is a so-called \emph{barrier} where processes wait
+until all processes have completed the round and before they start the
+next one.
 For example, in an iterative matrix algorithm, the matrix may be
-cut up into fragments.  During a round, one process runs for each
-fragment concurrently.  The next round is not allowed to start
-until all processes have completed operations on their fragment.
+cut up into fragments.  During a round, the processes run concurrently
+one for each fragment.  The next round is not allowed to start
+until all processes have completed processing their fragment.
 
@@
 with just three semaphores.  Busy waiting is not allowed.  How about just two?
+(As always, the Little Book of Semaphores~\cite{Downey09} is a good resource
+for solving semaphore problems.  Look for the double turnstile solution.)
 \item Implement barrier synchronization with Mesa condition variables.
+(You may want to use a double turnstile approach here as well.)
 \end{problems}

### 2020-06-24 book/paper.tex
Reordered chapters

@@
 release a resource.
-\item Chapter~\ref{ch:actor} presents the \emph{actor model} as an approach to
-synchronization.
+\item Chapter~\ref{ch:actor} presents the \emph{actor model}
+and \emph{message passing} as an approach to synchronization.
+\item Chapter~\ref{ch:abp} presents a problem and a solution to the distributed
+systems problem of having two processes communicate reliably over an unreliable
+network.
 \item Chapter~\ref{ch:nonblocking} introduces \emph{non-blocking} or
@@
 useful in high-performance computing applications such as parallel simulations.
-\item Chapter~\ref{ch:abp} presents a problem and a solution to the distributed
-systems problem of having two processes communicate reliably over an unreliable
-network.
 \end{itemize}
@@
 
+\chapter{Networking: Alternating Bit Protocol}
+\label{ch:abp}
+\index{alternating bit protocol}
+
+\begin{figure}
+\begin{code}
+\VerbatimInput[xleftmargin=5mm,numbers=left]{code/abp.cxl}
+\end{code}
+\caption{[abp.cxl] Alternating Bit Protocol.}
+\label{fig:abp}
+\end{figure}
+
+\begin{figure}
+\begin{code}
+\VerbatimInput[xleftmargin=5mm,numbers=left]{code/abptest.cxl}
+\end{code}
+\caption{[abptest.cxl] Test code for alternating bit protocol.}
+\label{fig:abptest}
+\end{figure}
+
+A distributed system is a concurrent system in which a collection
+of processes communicate by message passing, much the same as
+in the actor model.
+The most important difference between distributed and concurrent
+systems is that the former takes \emph{failures} into account.
+In this chapter, we will consider two actors, a sender and
+a receiver, that communicate by message passing over a network
+that may lose messages.
+
 \chapter{Non-Blocking Synchronization}
@@
 \end{problems}
-
-\chapter{Networking: Alternating Bit Protocol}
-\label{ch:abp}
-\index{alternating bit protocol}
-
-\begin{figure}
-\begin{code}
-\VerbatimInput[xleftmargin=5mm,numbers=left]{code/abp.cxl}
-\end{code}
-\caption{[abp.cxl] Alternating Bit Protocol.}
-\label{fig:abp}
-\end{figure}
-
-\begin{figure}
-\begin{code}
-\VerbatimInput[xleftmargin=5mm,numbers=left]{code/abptest.cxl}
-\end{code}
-\caption{[abptest.cxl] Test code for alternating bit protocol.}
-\label{fig:abptest}
-\end{figure}
 

### 2020-06-24 book/paper.tex
Working on ABP chapter

@@
 systems is that the former takes \emph{failures} into account.
-In this chapter, we will consider two actors, a sender and
-a receiver, that communicate by message passing over a network
-that may lose messages.
+In this chapter, we will consider two actors, Alice and Bob.
+Alice wants to send a sequence of application messages to Bob,
+but the underlying network may lose messages.
+The network does not re-order messages: when sending messages
+$m_1$ and $m_2$ in that order, then if both messages are received,
+$m_1$ is received before $m_2$.
+Also, the network does not create messages out of nothing: if
+message $m$ is received, then message $m$ was sent.
+
+We need a network protocol that Alice and Bob can run.
+In particular, it has to be the case that if Alice sends application
+messages $m_1, ..., m_n$ in that order, then if Bob receives an
+application message
+$m$, then $m = m_i$ for some $i$ and Bob will already have received
+application messages $m_1, ..., m_i$ (safety).
+Also, if the network is fair, then if Alice sends application message
+$m$, then eventually Bob should deliver $m$ (liveness).
+
+We assume that there are two unreliable network channels: \texttt{s\_chan}
+from Alice to Bob, and \texttt{r\_chan} from Bob to Alice.
+Because of the network's properties, we can model each
+channel as a variable that either contains a network message or nothing
+(we use \texttt{()} in the model).
+
+The \emph{Alternating Bit Protocol} is suitable for our purposes.
+Alice and Bob each maintain a zero-initialized sequence number,
+\texttt{s\_seq} and \texttt{r\_seq} resp.
+Alice sends a network message to Bob containing an
+application message as \emph{payload} and Alice's sequence number
+as \emph{header}.
+When Bob receives such a network message, Bob returns an
+an \emph{acknowledgment} to Alice, which is a network message
+containing the sequence number in the message that Bob received.
+
+In the protocol, Alice keeps sending the same network message until she
+receives an acknowledgment with the same sequence number.
+When this happens, Alice increments her sequence number.  She is now
+ready to send the next message she wants to send to Bob.
+Bob, on the other hand, waits until he receives a message matching
+Bob's sequence number.  If so, Bob \emph{delivers} the payload in the
+message and increments his sequence number.
+Because of the network properties,  a one-bit sequence number suffices.
 

### 2020-06-25 book/paper.tex
Drafted text for ABP protocol

@@
 The most important difference between distributed and concurrent
-systems is that the former takes \emph{failures} into account.
+systems is that the former takes \emph{failures} into account,
+including failures of processes and failures of shared memory.
 In this chapter, we will consider two actors, Alice and Bob.
@@
 
-We assume that there are two unreliable network channels: \texttt{s\_chan}
-from Alice to Bob, and \texttt{r\_chan} from Bob to Alice.
-Because of the network's properties, we can model each
-channel as a variable that either contains a network message or nothing
-(we use \texttt{()} in the model).
-
 The \emph{Alternating Bit Protocol} is suitable for our purposes.
+We assume that there are two unreliable network channels: one from Alice
+to Bob and one from Bob to Alice.
 Alice and Bob each maintain a zero-initialized sequence number,
@@
 Because of the network properties,  a one-bit sequence number suffices.
+
+Because of the network's properties, we can model each
+channel as a variable that either contains a network message or nothing
+(we use \texttt{()} in the model).  Let \texttt{s\_chan} be the channel
+from Alice to Bob and \texttt{r\_chan} the channel from Bob to Alice.
+\texttt{net\_send(pchan, m, reliable)} models sending a message \texttt{m} to
+\texttt{\^{}pchan}, where \texttt{pchan} is either \texttt{\&s\_chan}
+or \texttt{\&r\_chan}.
+The method places either \texttt{m} (to model a successful send)
+or \texttt{()} (to model loss) in \texttt{\^{}pchan}.
+The use of the \texttt{reliable} flag will be explained later.
+\texttt{net\_recv(pchan)} models checking \texttt{\^{}pchan} for
+the next message.
+
+Method \texttt{app\_send(m)} tries \emph{retransmitting} \texttt{m} until
+an acknowledgment is received.
+Method \texttt{app\_recv(reliable)} returns the next successfully received
+message.
+Figure~\ref{fig:abptest} shows how the methods may be used to send and receive
+a stream of \texttt{NMSGS} messages reliably.
+It has to be bounded, because model checking requires a finite model.
+
+Only the last invocation of \texttt{app\_recv(reliable)} is invoked with
+\texttt{reliable == True}.  It causes the last acknowledgment to be sent
+reliably.  It allows the receiver (Bob) to stop, as well as the sender (Alice)
+once the last acknowledgment has been received.
+Without something like this, either the sender may be left hanging waiting
+for the last acknowledgment, or the receiver waiting for the last message.
 

### 2020-06-25 book/paper.tex
Latest

@@
 
+\section*{Exercises}
+\begin{problems}
+\item A common pattern in distributed systems is the so-called \emph{client/server}
+model, where a server process provides services to a set of clients.  The clients
+send request messages to the server, and the server sends a response to each
+request.  Develop a protocol for a single client and server using the same network
+model as for the ABP protocol.  By way of example, let a request contain an
+integer and the response consist of the square of that integer.
+\end{problems}
+
 \chapter{Non-Blocking Synchronization}

### 2020-06-25 book/paper.tex
Added some more index terms

--- 2020-06-25 book/paper.tex
@@
 producing jobs and workers executing them in parallel.
+Or, in the client/server model,
+\index{client/server model}
+some process may act as a \emph{server} that clients can send requests to
+and receive responses from.
 
@@
 \begin{problems}
+\item \label{ex:cltsvr} A popular model is the \emph{client/server} model.
+Here a single actor implements some service, like computing the square of a
+number.
+A client can send a message containing an integer to the server, and the
+server returns a response message containing the square of that message.
+Implement this.
 \item Actors and message queues are good for building pipelines.
@@
 in the opposite order.  Which do you think is better?
-\item Bonus question: implement a version of parallel qsort
-(see Example~\ref{ex:qsort}) using the actor model.
 \end{problems}
@@
 \index{alternating bit protocol}
+\index{protocol}
+\index{distributed system}
 
@@
 The most important difference between distributed and concurrent
-systems is that the former takes \emph{failures} into account,
+systems is that the former takes \emph{failures}
+\index{failure}
+into account,
 including failures of processes and failures of shared memory.
@@
 but the underlying network may lose messages.
+\index{network}
 The network does not re-order messages: when sending messages
@@
 for the last acknowledgment, or the receiver waiting for the last message.
+
+\section*{Exercises}
+\begin{problems}
+\item Exercise~\ref{ex:cltsvr} explored the \emph{client/server}.  It is popular
+in distributed systems as well.
+Develop a protocol for a single client and server using the same network
+model as for the ABP protocol.
+\end{problems}
 

### 2020-06-25 book/paper.tex
More index terms

@@
 \item
-\texttt{lock} sets the lock if available and suspends itself if not.
+\texttt{lock}
+\index{lock}
+sets the lock if available and suspends itself if not.
 Note that \texttt{stop} is called within an \texttt{atomic} statement---this is
@@
 \item
-\texttt{unlock} checks to see if any processes are waiting to get the lock.
+\texttt{unlock}
+\index{unlock}
+checks to see if any processes are waiting to get the lock.
 If so, it uses the \texttt{head} and \texttt{tail}
@@
 Are they linearizable?  If so, what are their linearization points?
-\item A \emph{seqlock} consists of a lock and a version number.
+\item A \emph{seqlock}
+\index{seqlock}
+consists of a lock and a version number.
 An update operation acquires the lock, increments the version number, makes the

### 2020-06-25 book/paper.tex
Added an exercise
Latest

@@
 model as for the ABP protocol.
+Hint: the response to a request can contain the same sequence number as the
+request.
+\item Generalize the solution in the previous exercise to multiple clients.
+Each client is uniquely identified.  You may either use separate channel pairs
+for each client, or solve the problem using a single pair of channels.
 \end{problems}
@@
 StoreVar [$v$] & pop a value from the stack and store it in a process variable \\
+\hline
+Swap & swap the two values on top of the stack \\
 \hline

### 2020-06-25 book/paper.tex
Latest

@@
 \hline
-\texttt{e == e} & two CXL values of the same type \\
-\hline
-\texttt{e != e} & two CXL values of the same type \\
-\hline
-\texttt{e < e} & any two CXL values \\
-\hline
-\texttt{e <= e} & any two CXL values \\
-\hline
-\texttt{e > e} & any two CXL values \\
-\hline
-\texttt{e >= e} & any two CXL values \\
+\texttt{e == e}, \texttt{e != e} & two CXL values of the same type \\
+\hline
+\texttt{e < e, e <= e} & any two CXL values \\
+\hline
+\texttt{e > e, e >= e} & any two CXL values \\
 \hline
@@
 \hline
-\texttt{e in e} & first is any CXL value, second is set \\
-\hline
-\texttt{e not in e} & first is any CXL value, second is set \\
+\texttt{e [not] in e} & first is any CXL value, second is set \\
 \hline
@@
 \hline
-\texttt{min e} & a set \\
-\hline
-\texttt{max e} & a set \\
+\texttt{min e, max e} & a set \\
 \hline

### 2020-06-26 book/code/pp.py
Working on pretty printing code

@@ -1,0 +1,322 @@
+import sys
+import os
+import getopt
+import traceback
+import collections
+import time
+import math
+
+# TODO.  This should not be global ideally
+files = {}              # files that have been read already
+constants = {}          # constants modified with -c
+modules = {}            # modules modified with -m
+namestack = []          # stack of module names being compiled
+node_uid = 1            # unique node identifier
+
+def load(f, filename, silent):
+    if filename in files:
+        return
+    namestack.append(filename)
+    files[filename] = []
+    all = ""
+    for line in f:
+        files[filename] += [line]
+        all += line
+    tokens = lexer(all, filename, silent)
+    namestack.pop()
+
+def islower(c):
+    return c in "abcdefghijklmnopqrstuvwxyz"
+
+def isupper(c):
+    return c in "ABCDEFGHIJKLMNOPQRSTUVWXYZ"
+
+def isletter(c):
+    return islower(c) or isupper(c)
+
+def isnumeral(c):
+    return c in "0123456789"
+
+def isalnum(c):
+    return isletter(c) or isnumeral(c)
+
+def isnamechar(c):
+    return isalnum(c) or c == "_"
+
+def isprint(c):
+    return isinstance(c, str) and len(c) == 1 and (
+        isalnum(c) or c in " ~`!@#$%^&*()-_=+[{]}\\|;:'\",<.>/?")
+
+def isnumber(s):
+    return all(isnumeral(c) for c in s)
+
+def isreserved(s):
+    return s in [
+        "and",
+        "assert",
+        "atLabel",
+        "atomic",
+        "bagsize",
+        "call",
+        "cardinality",
+        "choose",
+        "const",
+        "def",
+        "del",
+        "dict",
+        "else",
+        "False",
+        "fun",
+        "for",
+        "go",
+        "if",
+        "import",
+        "in",
+        "inf",
+        "keys",
+        "len",
+        "let",
+        "max",
+        "min",
+        "nametag",
+        "not",
+        "NULL",
+        "or",
+        "pass",
+        "spawn",
+        "stop",
+        "True",
+        "while"
+    ]
+
+def isname(s):
+    return (not isreserved(s)) and (isletter(s[0]) or s[0] == "_") and \
+                    all(isnamechar(c) for c in s)
+
+def isunaryop(s):
+    return s in [ "^", "-", "atLabel", "bagsize", "cardinality",
[…]

### 2020-06-26 book/paper.tex
Working on pretty printing code
Working on formatting CXL code

@@
 \usepackage{etoolbox}
+\usepackage{bold-extra}
 
@@
 \newcommand{\tm}{\raisebox{.9ex}{\tiny tm}}
+
+\newcommand{\cxlsource}[1]{
+\begin{tabbing}
+XX\=XXX\=XXX\kill
+    \input{code/#1.tex}
+\end{tabbing}
+}
 
@@
 \begin{code}
-\VerbatimInput[xleftmargin=5mm,numbers=left]{code/triangle.cxl}
+\cxlsource{triangle}
 \end{code}
@@
 \begin{code}
-\VerbatimInput[xleftmargin=5mm,numbers=left]{code/Up.cxl}
+\cxlsource{Up}
 \end{code}
@@
 \begin{code}
-\VerbatimInput[xleftmargin=5mm,numbers=left]{code/UpEnter.cxl}
+\cxlsource{UpEnter}
 \end{code}
@@
 \begin{code}
-\VerbatimInput[xleftmargin=5mm,numbers=left]{code/csbarebones.cxl}
+\cxlsource{csbarebones}
 \end{code}
@@
 \begin{code}
-\VerbatimInput[xleftmargin=5mm,numbers=left]{code/cs.cxl}
+\cxlsource{cs}
 \end{code}
@@
 \begin{code}
-\VerbatimInput[xleftmargin=5mm,numbers=left]{code/naiveLock.cxl}
+\cxlsource{naiveLock}
 \end{code}
@@
 \begin{code}
-\VerbatimInput[xleftmargin=5mm,numbers=left]{code/naiveFlags.cxl}
+\cxlsource{naiveFlags}
 \end{code}
@@
 \begin{code}
-\VerbatimInput[xleftmargin=5mm,numbers=left]{code/naiveTurn.cxl}
+\cxlsource{naiveTurn}
 \end{code}
@@
 \begin{code}
-\VerbatimInput[xleftmargin=5mm,numbers=left]{code/Peterson.cxl}
+\cxlsource{Peterson}
 \end{code}
@@
 \begin{code}
-\VerbatimInput[xleftmargin=5mm,numbers=left]{code/PetersonInductive.cxl}
+\cxlsource{PetersonInductive}
 \end{code}
@@
 \begin{code}
-\VerbatimInput[xleftmargin=5mm,numbers=left]{code/csonebit.cxl}
+\cxlsource{csonebit}
 \end{code}
@@
 \begin{code}
-\VerbatimInput[xleftmargin=5mm,numbers=left]{code/PetersonMethod.cxl}
+\cxlsource{PetersonMethod}
 \end{code}
@@
 \begin{code}
-\VerbatimInput[xleftmargin=5mm,numbers=left]{code/spinlock.cxl}
+\cxlsource{spinlock}
 \end{code}
@@
 \begin{code}
-\begin{Verbatim}[xleftmargin=5mm,numbers=left]
-def atomic_inc(ptr):
-    atomic:
-        ^ptr += 1;
-    ;
-;
-count = 0;
-atomic_inc(&count);
-\end{Verbatim}
+\cxlsource{atomicinc}
 \end{code}
@@
 \begin{code}
-\VerbatimInput[xleftmargin=5mm,numbers=left]{code/spinlockInv.cxl}
+\cxlsource{spinlockInv}
 \end{code}
@@
 \begin{code}
-\VerbatimInput[xleftmargin=5mm,numbers=left]{code/csTAS.cxl}
+\cxlsource{csTAS}
 \end{code}
@@
 \begin{code}
-\begin{Verbatim}[xleftmargin=5mm,numbers=left]
-def tas(lk):
-    atomic:
-        result = ^lk;
-        ^lk = True;
-    ;
-;
-def Lock():
-    result = False;
-;
-def lock(lk):
-    while tas(lk):
-        pass;
-    ;
-;
-def unlock(lk):
-    ^lk = False;
-;
-\end{Verbatim}
+\cxlsource{lockintf}
 \end{code}
@@
 \begin{code}
-\VerbatimInput[xleftmargin=5mm,numbers=left]{code/UpLock.cxl}
+\cxlsource{UpLock}
 \end{code}
@@
 \begin{code}
-\begin{Verbatim}[xleftmargin=5mm,numbers=left]
-import list;
-
-def Lock():
-    result = dict{ .locked: False, .suspended: [] };
-;
-def lock(lk):
-    atomic:
-        if (^lk).locked:
-            stop (^lk).suspended;
-            assert (^lk).locked;
-        else:
-            (^lk).locked = True;
-        ;
-    ;
-;
-def unlock(lk):
-    atomic:
-        if (^lk).suspended == []:
-            (^lk).locked = False;
-        else:
-            go (head((^lk).suspended)) ();
-            (^lk).suspended = tail((^lk).suspended);
-        ;
-    ;
-;
-\end{Verbatim}
+\cxlsource{locksusp}
 \end{code}
@@
 \begin{code}
-\VerbatimInput[xleftmargin=5mm,numbers=left]{code/RW.cxl}
+\cxlsource{RW}
 \end{code}
@@
 \begin{code}
-\VerbatimInput[xleftmargin=5mm,numbers=left]{code/RWtest.cxl}
+\cxlsource{RWtest}
 \end{code}
@@
 \begin{code}
-\VerbatimInput[xleftmargin=5mm,numbers=left]{code/RWbusychk.cxl}
+\cxlsource{RWbusychk}
 \end{code}
@@
 \begin{code}
-\VerbatimInput[xleftmargin=5mm,numbers=left]{code/RWlock.cxl}
+\cxlsource{RWlock}
 \end{code}
@@
 \begin{code}
-\VerbatimInput[xleftmargin=5mm,numbers=left]{code/RWmulti.cxl}
+\cxlsource{RWmulti}
 \end{code}
@@
 \begin{code}
-\begin{Verbatim}[xleftmargin=5mm,numbers=left]
-def Semaphore(cnt):
-    result = cnt;
-;
-def P(sema):
-    let blocked = True:
-        while blocked:
-            atomic:
-                if (^sema) > 0:
-                    ^sema -= 1;
-                    blocked = False;
-                ;
-            ;
-        ;
-    ;
-;
-def V(sema):
-    atomic:
-        ^sema += 1;
-    ;
-;
-\end{Verbatim}
+\cxlsource{semaphore}
 \end{code}
@@
 \begin{code}
-\VerbatimInput[xleftmargin=5mm,numbers=left]{code/DinersSema.cxl}
+\cxlsource{DinersSema}
 \end{code}
@@
 \begin{code}
-\VerbatimInput[xleftmargin=5mm,numbers=left]{code/qsort.cxl}
+\cxlsource{qsort}
 \end{code}
@@
 \begin{code}
-\VerbatimInput[xleftmargin=5mm,numbers=left]{code/qsorttest.cxl}
+\cxlsource{qsorttest}
 \end{code}
@@
 \begin{center}
-\includegraphics[width=2.5in]{figures/pc-crop.pdf}
+\includegraphics[width=3in]{figures/pc-crop.pdf}
 \end{center}
@@
 \begin{code}
-\VerbatimInput[xleftmargin=5mm,numbers=left]{code/BBsema.cxl}
+\cxlsource{BBsema}
 \end{code}
@@
 \begin{code}
-\VerbatimInput[xleftmargin=5mm,numbers=left]{code/BBsematest.cxl}
+\cxlsource{BBsematest}
 \end{code}
@@
 \begin{code}
-\VerbatimInput[xleftmargin=5mm,numbers=left]{code/BBsemadata.cxl}
+\cxlsource{BBsemadata}
 \end{code}
@@
 \begin{code}
-\VerbatimInput[xleftmargin=5mm,numbers=left]{code/gpu.cxl}
+\cxlsource{gpu}
 \end{code}
@@
 \begin{code}
-\VerbatimInput[xleftmargin=5mm,numbers=left]{code/RWsbs.cxl}
+\cxlsource{RWsbs}
 \end{code}
@@
 \begin{code}
-\begin{Verbatim}[xleftmargin=5mm,numbers=left]
-let blocked = True:
-    while blocked:
-        lock();
-        if application-condition-holds:
-            blocked = False;
-        else:
-            unlock():
-@cr: application-code;
-unlock():
-\end{Verbatim}
+\cxlsource{busywait}
 \end{code}
@@
 {\small
-\VerbatimInput[xleftmargin=5mm,numbers=left]{code/RWfair.cxl}
+\cxlsource{RWfair}
 }
@@
 {\small
-\VerbatimInput[xleftmargin=5mm,numbers=left]{code/RWqueue.cxl}
+\cxlsource{RWqueue}
 }
@@
 \begin{code}
-\VerbatimInput[xleftmargin=5mm,numbers=left]{code/RWqtest.cxl}
+\cxlsource{RWqtest}
 \end{code}
@@
 \begin{code}
-\begin{Verbatim}[xleftmargin=5mm,numbers=left]
-def wait(c):
-    V(&mutex);
-    P(c);
-;
-def signal(c):
-    V(c);
-    P(&mutex);
-;
-\end{Verbatim} 
+\cxlsource{hoaremon}
 \end{code}
@@
 \begin{figure}
-\begin{code}
-\begin{Verbatim}[xleftmargin=5mm,numbers=left]
-import bag;
-
-def Condition(lk):
-    result = dict{ .lock: lk, .waiters: bagEmpty() };
-;
-def wait(c):
-    let lk = (^c).lock, blocked = True,
-                cnt = bagCount((^c).waiters, nametag()):
-        bagAdd(&(^c).waiters, nametag());
-        ^lk = False;
-        while blocked:
-            atomic:
-                if (not (^lk)) and
-                        (bagCount((^c).waiters, nametag()) <= cnt):
-                    ^lk = True;
-                    blocked = False;
-                ;
-            ;
-        ;
-    ;
-;
-def notify(c):
-    let waiters = (^c).waiters:
-        if waiters != bagEmpty():
-            bagRemove(&(^c).waiters, bagChoose(waiters));
-        ;
-    ;
-;
-def notifyAll(c):
-    (^c).waiters = bagEmpty();
-;
-\end{Verbatim}
-\end{code}
+\cxlsource{mesa}
 \caption{Implementation of condition variables in the \texttt{synch} module.}
@@
 \begin{code}
-\VerbatimInput[xleftmargin=5mm,numbers=left]{code/RWcv.cxl}
+\cxlsource{RWcv}
 \end{code}
@@
 \begin{code}
-\VerbatimInput[xleftmargin=5mm,numbers=left]{code/Diners.cxl}
+\cxlsource{Diners}
 \end{code}
@@
 \begin{code}
-\VerbatimInput[xleftmargin=5mm,numbers=left]{code/DinersCV.cxl}
+\cxlsource{DinersCV}
 \end{code}
@@
 \begin{code}
-\begin{Verbatim}
-while forks[left]:
-    wait(&conds[left]);
-;
-while forks[right]:
-    wait(&conds[right]);
-;
-\end{Verbatim}
+\cxlsource{baddblwait}
 \end{code}
@@
 \begin{code}
-\begin{Verbatim}[xleftmargin=5mm,numbers=left]
-    if left < right:
-        lock(&forks[left]);
-        lock(&(orks[right]);
-    else:
-        unlock(&forks[right]);
-        unlock(&forks[left]);
-    ;
-\end{Verbatim}
+\cxlsource{dinersfix}
 \end{code}
@@
 \begin{code}
-\VerbatimInput[xleftmargin=5mm,numbers=left]{code/DinersAvoid.cxl}
+\cxlsource{DinersAvoid}
 \end{code}
@@
 \begin{code}
-\VerbatimInput[xleftmargin=5mm,numbers=left]{code/bank.cxl}
+\cxlsource{bank}
 \end{code}
@@
 \begin{code}
-\VerbatimInput[xleftmargin=5mm,numbers=left]{code/actor.cxl}
+\cxlsource{actor}
 \end{code}
@@
 \begin{code}
-\VerbatimInput[xleftmargin=5mm,numbers=left]{code/actortest.cxl}
+\cxlsource{actortest}
 \end{code}
@@
 \begin{code}
-\VerbatimInput[xleftmargin=5mm,numbers=left]{code/abp.cxl}
+\cxlsource{abp}
 \end{code}
@@
 \begin{code}
-\VerbatimInput[xleftmargin=5mm,numbers=left]{code/abptest.cxl}
+\cxlsource{abptest}
 \end{code}
@@
 \begin{code}
-\VerbatimInput[xleftmargin=5mm,numbers=left]{code/hw.cxl}
+\cxlsource{hw}
 \end{code}
@@
 \begin{code}
-\VerbatimInput[xleftmargin=5mm,numbers=left]{code/lst1.cxl}
+\cxlsource{lst1}
 \end{code}
@@
 \begin{code}
-\VerbatimInput[xleftmargin=5mm,numbers=left]{code/lst2.cxl}
+\cxlsource{lst2}
 \end{code}
@@
 \begin{code}
-\VerbatimInput[xleftmargin=5mm,numbers=left]{code/barrier.cxl}
+\cxlsource{barrier}
 \end{code}

### 2020-06-26 book/code/pp.py

@@ -8,7 +8,6 @@ import math
 
 # TODO.  This should not be global ideally
 files = {}              # files that have been read already
-constants = {}          # constants modified with -c
 modules = {}            # modules modified with -m
 namestack = []          # stack of module names being compiled
 node_uid = 1            # unique node identifier
@@ -103,12 +102,13 @@ def isbinaryop(s):
         "-", "+", "*", "/", "%", "<", "<=", ">", ">="
     ];
 
-tokens = [ "==", "!=", "<=", ">=", ".." ]
+# tokens = [ "==", "!=", "<=", ">=", ".." ]
+tokens = []
 
 def putchar(c):
     print(c, end="")
 
-constants = set()
+constants = { "lock", "unlock", "wait", "notify", "notifyAll", "signal", "P", "V" }
 
 def doimport(module, file):
     for dir in [ os.path.dirname(file), "../../modules", "." ]:
@@ -144,11 +144,18 @@ def lexer(s, file, silent):
         # skip over line comments
         if s.startswith("#"):
             s = s[1:]
-            if not silent: print("\\#\\emph{", end="")
+            if not silent: print("\\emph{\\#", end="")
             while len(s) > 0 and s[0] != '\n':
-                if not silent: putchar(s[0])
+                column += 1
+                if not silent:
+                    if s[0] in ["&", "%", "{", "}", "#", "^", "_"]:
+                        print("\\" + s[0], end="")
+                    elif s[0] in [ "<", ">" ]:
+                        print("$" + s[0] + "$", end="")
+                    else:
+                        putchar(s[0])
                 s = s[1:]
-            if not silent: print("}")
+            if not silent: print("}", end="")
             continue
 
         # skip over nested comments
@@ -191,7 +198,7 @@ def lexer(s, file, silent):
             i = 0
             while i < len(s) and isnamechar(s[i]):
                 i += 1
-            found = s[:i]
+            found = s[:i].replace("_", "\\_")
             if isreserved(found):
                 if not silent: print("\\texttt{\\textbf{%s}}"%found, end="")
                 if found in [ "def", "const", "import" ]:
@@ -262,7 +269,15 @@ def lexer(s, file, silent):
         elif s[0] == "^":
             if not silent: print("\\^{}", end="")
         else:
-            if not silent: putchar(s[0])
+            if s[0] == "@":
+                nextconst = True
+            if not silent:
+                if s[0] == "-":
+                    print("--", end="")
[…]

### 2020-06-26 book/paper.tex
Working on better fonts

@@
 complicated than programs where only one thing happens at a time.
+
 \glsadd{concurrent execution}
 \glsadd{shared variable}
+\glsadd{sequential execution}
+\glsadd{determinism}
+\glsadd{atomicity}
 
@@
 \index{sequential}
-\glsadd{sequential execution}
 program is mostly \emph{deterministic}.
 \index{determinism}
-\glsadd{determinism}
 If you run it twice with the same input, the same output will be produced.
@@
 because there is no other activity interfering with their execution.
-\glsadd{atomicity}
 Even though a statement or function may
@@
 because it is hard to reproduce.
+j
+\glsadd{model checking}
 
@@
 \index{model checking}
-\glsadd{model checking}
 instead of relying
@@
 \glsadd{invariant}
+
 While model checking does not in general prove an algorithm correct,
@@
 an integer number as argument.  Each method has a variable called
-\texttt{result} that eventually contains the result of the
+\textit{result} that eventually contains the result of the
 method (there is no \texttt{return} statement in CXL).  The method
-also has a variable called \texttt{n} containing the value of the
-argument.  The \texttt{x..y} notation generates a set containing the numbers
-from~\texttt{x} to~\texttt{y} (inclusive).  The last two lines in the program are
+also has a variable called \textit{n} containing the value of the
+argument.  The \textit{x..y} notation generates a set containing the numbers
+from~\textit{x} to~\textit{y} (inclusive).  The last two lines in the program are
 the most interesting.
-The first assigns to \texttt{x} some unspecified value in the range \texttt{0..N}
+The first assigns to \textit{x} some unspecified value in the range \texttt{0..N}
 and the second verifies that $\mathtt{triangle}(x)$ equals $x(x+1)/2$.
@@
 (For this to work, make sure \texttt{cxl} is in your command shell's search path.)
-Essentially, the $\texttt{choose}(S)$
+Essentially, the $\texttt{choose}($S$)$
 \index{choose operator}
@@
 \begin{problems}
-\item See what happens if, instead of initializing \texttt{result} to 0,
+\item See what happens if, instead of initializing \textit{result} to 0,
 you initialize it to 1.  (You do not need to understand the error report at this time.  They will be explained in more detail in Chapter~\ref{ch:cxlmachine}.)
@@
 
+\glsadd{process}
+
 Concurrent programming, aka multithreaded programming, involves multiple
@@
 running in parallel while sharing variables.
-\glsadd{process}
 Figure~\ref{fig:inc} presents a simple example.
 The program
-initializes two shared variables: an integer \texttt{count} and
-an array \texttt{done} with two booleans.
-Method \texttt{incrementer} takes a parameter called \texttt{self}.
-It increments \texttt{count} and sets \texttt{done[self]} to \texttt{True}.
-Method \texttt{main} waits until \texttt{done} flags are set to \texttt{True}.
-After that, it verifies that the value of \texttt{count} equals 2.  If not,
-it reports the actual value of \texttt{count}.
+initializes two shared variables: an integer \textit{count} and
+an array \textit{done} with two booleans.
+Method \texttt{incrementer} takes a parameter called \textit{self}.
+It increments \textit{count} and sets \textit{done}[\textit{self}] to \texttt{True}.
+Method \texttt{main} waits until \textit{done} flags are set to \texttt{True}.
+After that, it verifies that the value of \textit{count} equals 2.  If not,
+it reports the actual value of \textit{count}.
 The program spawns three processes.
@@
 \item Before you run the program, what do you think will happen?  Is the
-program correct in that \texttt{count} will always end up being 2?
-\item What are the possible values of \texttt{count} at the time of the
+program correct in that \textit{count} will always end up being 2?
+\item What are the possible values of \textit{count} at the time of the
 \texttt{assert} statement?
@@
 
+\glsadd{machine instruction}
+
 What is going on is that the CXL program is compiled to machine instructions,
 \index{machine instruction}
-\glsadd{machine instruction}
 and it is the machine instructions that are executed by the underlying CXL
@@
 \begin{enumerate}
-\item incrementer 0 loads the value of \texttt{count}, which is 0;
-\item incrementer 1 loads the value of \texttt{count}, which is still 0;
+\item incrementer 0 loads the value of \textit{count}, which is 0;
+\item incrementer 1 loads the value of \textit{count}, which is still 0;
 \item incrementer 1 adds one to the value that it loaded (0), and
-stores $1$ into \texttt{count};
+stores $1$ into \textit{count};
 \item incrementer 0 adds one to the value that it loaded (0), and
-stores $1$ into \texttt{count};
-\item incrementer 0 sets \texttt{done[0]} to \texttt{True};
-\item incrementer 1 sets \texttt{done[1]} to \texttt{True}.
+stores $1$ into \textit{count};
+\item incrementer 0 sets \textit{done}[0] to \texttt{True};
+\item incrementer 1 sets \textit{done}[1] to \texttt{True}.
 \end{enumerate}
 
-The result in this particular interleaving is that \texttt{count} ends up
+The result in this particular interleaving is that \textit{count} ends up
 being 1.
@@
 of an interleaving, like the one above, in which an assertion fails.
+
 \glsadd{race condition}
@@
 \begin{code}
-\begin{verbatim}
-    assert (count == 1) or (count == 2);
-\end{verbatim}
+\textbf{assert} (\textit{count} == 1) or (\textit{count} == 2);
 \end{code}
@@
 \item Figure~\ref{fig:incmany} is a version of Figure~\ref{fig:incpy} that has each
-incrementer thread increment \texttt{count} \texttt{N} times.  Run Figure~\ref{fig:incpy}
+incrementer thread increment \textit{count} \texttt{N} times.  Run Figure~\ref{fig:incpy}
 10 times (using Python).
-Report how many times the assertion fails and what the value of \texttt{count}
+Report how many times the assertion fails and what the value of \textit{count}
 was for each of the failed runs.
@@
 
+\glsadd{context}
+
 The state of a process is called a \emph{context} (aka \emph{continuation}):
@@
 \index{continuation}
-\glsadd{context}
 it contains the value of
@@
 
+\glsadd{stack machine}
+
 Compiling the code in Figure~\ref{fig:inc} results in the CVM bytecode
@@
 \index{stack machine}
-\glsadd{stack machine}
 but it also has a register.
@@
 the program counter of the method's \texttt{Return} instruction.
-The code generated from \texttt{count := count + 1} in line 2 of
+The code generated from $\mathit{count} := \mathit{count} + 1$ in line 2 of
 \texttt{Up.cxl} is as follows (see Figure~\ref{fig:inccode}):
@@
 \item The \texttt{Load} instruction pushes the value of the
-\texttt{count} variable onto the stack.
+\textit{count} variable onto the stack.
 \item the \texttt{Push} instruction pushes the constant 1
@@
 \item \texttt{2-ary} is a \texttt{+} operation with 2 arguments.
-It pops two values from the stack (1 and the value of \texttt{count}),
+It pops two values from the stack (1 and the value of \textit{count}),
 adds them and pushes the result back onto the stack.
 \item The \texttt{Store} instruction pops
-a CXL value (the sum of the \texttt{count} variable and 1) and
-stores it in the \texttt{count} variable.
+a CXL value (the sum of the \textit{count} variable and 1) and
+stores it in the \textit{count} variable.
 \end{enumerate}
@@
 It starts by reporting that the assertion on line 9 in column 5 failed and that
-\texttt{count} has value 1 instead of 0:
+\textit{count} has value 1 instead of 0:
 \texttt{Assertion failed (’assert’, ’Up.cxl’, 9, 5) 1}.
@@
 \item Process 0 executed instructions 1 through 4, loading the value of
-\texttt{count} but stopping just before storing 1 into \texttt{count};
+\textit{count} but stopping just before storing 1 into \textit{count};
 \item Process 1 executed instructions 1 through 10, storing 1 into
-into \texttt{count} and storing \texttt{True} into \texttt{done[1]};
-\item Process 0 continues execution, storing value 1 into \texttt{count}
-and storing \texttt{True} into \texttt{done[0]}.
+into \textit{count} and storing \texttt{True} into \textit{done}[1];
+\item Process 0 continues execution, storing value 1 into \textit{count}
+and storing \texttt{True} into \textit{done}[0].
 \end{itemize}
@@
 At the heart of the problem is that we would like make sure that, when
-the \texttt{count} variable is being updated, no other process is
+the \textit{count} variable is being updated, no other process is
 trying to do the same thing.  This is called a \emph{critical section}
@@
 \index{critical region}
+
 \glsadd{critical section}
+\glsadd{thread safety}
 
@@
 \index{thread safety}
-\glsadd{thread safety}
 \index{thread}
+
+\glsadd{mutual exclusion}
 
@@
 \index{mutual exclusion}
-\glsadd{mutual exclusion}
 We would like to place an assertion at the \texttt{@cs} label that
@@
 \glsadd{lock}
+
 You may already have heard of the concept of a \emph{lock}
@@
 Initially the lock is not owned, indicated by lock being \texttt{False}.
-To enter the critical section, a process waits until \texttt{lock} is \texttt{False}
+To enter the critical section, a process waits until \textit{lock} is \texttt{False}
 and then sets it to \texttt{True} to indicate that the lock has been taken.
@@
 Unfortunately, if we run the program through CXL, we find that the assertion still
-fails.  Diagnosing the problem, we see that the \texttt{lock} variable
-suffers from the same problem as the \texttt{count} variable
+fails.  Diagnosing the problem, we see that the \textit{lock} variable
+suffers from the same problem as the \textit{count} variable
 in Figure~\ref{fig:inc}: operations
@@
 After setting its flag, the process waits until the other process
-(\texttt{1 - self}) is not trying to enter the critical section.
+($1 - \mathit{self}$) is not trying to enter the critical section.
 If we run this program, the assertion does not fail.  In fact, this
@@
 To see why, first note the invariant that if process $i$ is in the
-critical section, then \texttt{flags[$i$] == True}.
+critical section, then \textit{flags}[$i$] == \texttt{True}.
 Without loss of generality,
-suppose that process~0 sets \texttt{flags[0]} at time $t_0$.
+suppose that process~0 sets \textit{flags}[0] at time $t_0$.
 Process 0 can only reach the critical section if at some time $t_1$,
-$t_1 > t_0$, it finds that \texttt{flags[1] == False}.
-Because of the invariant, \texttt{flags[1] == False} implies that
+$t_1 > t_0$, it finds that \textit{flags}[1] == \texttt{False}.
+Because of the invariant, \textit{flags}[1] == \texttt{False} implies that
 process~1 is not in the critical section at time $t_1$.
-Let $t_2$ be the time at which process~0 sets \texttt{flags[0]}
+Let $t_2$ be the time at which process~0 sets \textit{flags}[0]
 to \texttt{False}.  Process~0 is in the critical section sometime
@@
 It is easy to see that process~1 cannot enter the critical section
-between $t_1$ and $t_2$, because \texttt{flags[1] == False} at
+between $t_1$ and $t_2$, because \textit{flags}[1] == \texttt{False} at
 time $t_1$.  To reach the critical section between $t_1$ and $t_2$,
-it would first have to set \texttt{flags[1]} to \texttt{True} and
-then wait until \texttt{flags[0] == False}.  But that does not happen
+it would first have to set \textit{flags}[1] to \texttt{True} and
+then wait until \textit{flags}[0] == \texttt{False}.  But that does not happen
 until time $t_2$.
@@
 The final na\"{\i}ve solution that we propose
-is based on a variable called \texttt{turn}
-that alternates between 0 and 1.  When \texttt{turn = $i$}, process~$i$ can
+is based on a variable called \textit{turn}
+that alternates between 0 and 1.  When \textit{turn} = $i$, process~$i$ can
 enter the critical section, while process $1-i$ has to wait.  When done,
-process~$i$ sets \texttt{turn} to $1-i$ to give the other process an
+process~$i$ sets \textit{turn} to $1-i$ to give the other process an
 opportunity to enter.
 An invariant of this solution is that while process~$i$ is in the critical
-section, \texttt{turn} == $i$.
-Since \texttt{turn} cannot be 0 and 1 at
+section, \textit{turn} == $i$.
+Since \textit{turn} cannot be 0 and 1 at
 the same time, mutual exclusion is satisfied.
@@
 until either the other process is nowhere near the critical section
-(\texttt{flag[1 - self] == False}) or has given way
-(\texttt{turn == self}).
+($\mathit{flag}[1 - \mathit{self}] == \mathtt{False}$) or has given way
+($\mathit{turn} == \mathit{self}$).
 Running the algorithm with CXL shows that it satisfies both mutual
@@
 \glsadd{process variable}
+
 Why does it work?  We will focus here on how one might go about proving
@@
 Steps in CXL are atomic.
+
 \glsadd{step}
-
 \glsadd{trace}
+
 The CVM starts in an initial state in which there is only
@@
 One way to classify states is to define a predicate over states.
-All states in which \texttt{x = 1}, or all states where
+All states in which $x$ = 1, or all states where
 there are two or more processes executing, are examples of such predicates.
@@
 \index{inductive invariant}
-An inductive invariant $I$ is a predicate over states that satisfies the following:
+An inductive invariant $\mathcal{I}$ is a predicate over states that satisfies the following:
 \begin{itemize}
-\item $I$ holds in the initial state.
-\item For any state in which $I$ holds and any process in the
-state takes a step, then $I$ also holds in the resulting state.
+\item $\mathcal{I}$ holds in the initial state.
+\item For any state in which $\mathcal{I}$ holds and any process in the
+state takes a step, then $\mathcal{I}$ also holds in the resulting state.
 \end{itemize}
@@
 and process~1 at the start of the \texttt{while} loop.
-Also imagine that in this particular state $\mathtt{turn} = 1$.  Now let
-process~1 make a sequence of steps.  Because $\mathtt{turn} = 1$,
+Also imagine that in this particular state $\mathit{turn} = 1$.  Now let
+process~1 make a sequence of steps.  Because $\mathit{turn} = 1$,
 process~1 will break out of the while loop and also enter the critical
@@
 the critical section.
-Let $C(i) = \mathtt{flags}[1 - i] \land
-\mathtt{turn} = 1 - i$, that is, the condition on the \texttt{while} loop
+Let $\mathcal{C}(i) = \mathit{flags}[1 - i] \land
+\mathit{turn} = 1 - i$, that is, the condition on the \texttt{while} loop
 for process~$i$.
-Let predicate $I_p(i)$ be the following.
+Let predicate $\mathcal{I}_p(i)$ be the following.
 if process $i$ is at label \texttt{@cs} (i.e., process $i$ is in the critical section),
-then $C(i)$ does not hold or process $1-i$ is executing after setting
-$\mathtt{flags}[i]$ but still before setting \texttt{turn} to $1-i$.
-More formally, $I_p(i) = \mathtt{process}(i)@cs) \Rightarrow (\lnot C(i) \lor \mathtt{process}(1-i)@gate)$.
+then $\mathcal{C}(i)$ does not hold or process $1-i$ is executing after setting
+$\mathit{flags}[i]$ but still before setting \textit{turn} to $1-i$.
+More formally, $\mathcal{I}_p(i) = \mathtt{process}(i)@cs) \Rightarrow (\lnot \mathcal{C}(i) \lor \mathtt{process}(1-i)@gate)$.
 For Peterson's Algorithm, an inductive invariant that works well is
-$I_p(0) \land I_p(1)$.
-
-Figure~\ref{fig:petersonproof} formalizes $I_p(i)$ in CXL.
-The label \texttt{@gate} refers to the instruction that sets \texttt{turn} to $1-i$.
+$\mathcal{I}_p(0) \land \mathcal{I}_p(1)$.
+
+Figure~\ref{fig:petersonproof} formalizes $\mathcal{I}_p(i)$ in CXL.
+The label \texttt{@gate} refers to the instruction that sets \textit{turn} to $1-i$.
 You can run Figure~\ref{fig:petersonproof} to determine
-that $I_p(i)$ is indeed an invariant for $i = 0, 1$.
+that $\mathcal{I}_p(i)$ is indeed an invariant for $i = 0, 1$.
 
@@
 following must hold:
-$(\lnot C(0) \lor \mathtt{process}(1)@gate) \land
- (\lnot C(1) \lor \mathtt{process}(0)@gate)$.
+$(\lnot \mathcal{C}(0) \lor \mathtt{process}(1)@gate) \land
+ (\lnot \mathcal{C}(1) \lor \mathtt{process}(0)@gate)$.
 We know that both flags are set.
@@
 (because they are both at label \texttt{@cs}),
-so this simplifies to $(\lnot C(0)) \land (\lnot C(1))$.
-So we conclude that $\mathtt{turn} = 0 \land \mathtt{turn} = 1$, a
-logical impossibility.  Thus $I_p(i)$ implies exclusivity.
-
-To see that $I_p(0) \land I_p(1)$ is, in fact, an inductive invariant, first note that
+so this simplifies to $(\lnot \mathcal{C}(0)) \land (\lnot \mathcal{C}(1))$.
+So we conclude that $\mathit{turn} = 0 \land \mathit{turn} = 1$, a
+logical impossibility.  Thus $\mathcal{I}_p(i)$ implies exclusivity.
+
+To see that $\mathcal{I}_p(0) \land \mathcal{I}_p(1)$ is, in fact, an inductive invariant, first note that
 it certainly holds in the initial state, because in the initial state no process
@@
 symmetric for both processes).  We still have to show that if we are in a state
-in which $I_p(0)$ holds, then any step will result in a state in which
-$I_p(0)$ still holds.
-If $I_p(0)$ holds, process~0 is at label \texttt{@cs}.  If process~0
+in which $\mathcal{I}_p(0)$ holds, then any step will result in a state in which
+$\mathcal{I}_p(0)$ still holds.
+If $\mathcal{I}_p(0)$ holds, process~0 is at label \texttt{@cs}.  If process~0
 were to take a step, then in the next state process~0 would be no longer
-at that label and $I_p(0)$ would hold trivially over the next state.
+at that label and $\mathcal{I}_p(0)$ would hold trivially over the next state.
 Therefore we only need to consider a step by process~1.
 
-From $I_p(0)$ we know that one of the following three cases must hold before
+From $\mathcal{I}_p(0)$ we know that one of the following three cases must hold before
 process~1 takes a step:
 \begin{enumerate}
-\item \texttt{flags[1] = False};
-\item \texttt{turn = 0};
+\item \textit{flags}[1] = \texttt{False};
+\item \textit{turn} = 0;
 \item process~1 is at label \texttt{@gate}.
@@
 will hold).
-We know that process~1 never sets \texttt{turn} to 1, so
+We know that process~1 never sets \textit{turn} to 1, so
 if the second case holds before the step, it will also hold after the step.
 Finally, if process~1 is at label \texttt{@gate} before the step, then after
-the step \texttt{turn} will equal 0, and therefore the second case will hold
+the step \textit{turn} will equal 0, and therefore the second case will hold
 after the step.  qed.
@@
 page.  The page claimed that the following predicate is invariant:
-if process $i$ is in the critical section, then $\lnot C(i)$ (i.e.,
-$I_p(i)$ without the disjunct that process $1-i$ is at label \texttt{gate}).
+if process $i$ is in the critical section, then $\lnot \mathcal{C}(i)$ (i.e.,
+$\mathcal{I}_p(i)$ without the disjunct that process $1-i$ is at label \texttt{gate}).
 To demonstrate that this predicate is not invariant, you can remove the
@@
 \item A colleague of the author asked if the first two assignments in
-Peterson's algorithm (setting \texttt{flags[self]})
-to \texttt{True} and \texttt{turn} to \texttt{1 - self}) can be reversed.
+Peterson's algorithm (setting \textit{flags}[\textit{self}])
+to \texttt{True} and \textit{turn} to 1 - \textit{self}) can be reversed.
 After all, they are different variables assigned independent values---in a
@@
 \index{spinlock}
+
 \glsadd{spinlock}
@@
 \index{interlock instruction}
-\glsadd{interlock instruction}
 special machine instructions
 that can read memory and then write it in an indivisible step.
+
+\glsadd{interlock instruction}
 
@@
 The program uses $N+1$ variables.
-Variable \texttt{shared} is initialized to \texttt{False} while
-\texttt{private[$i$]} for each process $i$ is initialized to \texttt{True}.
-An important invariant, $I_1$, of the program is that at any time at most
+Variable \textit{shared} is initialized to \texttt{False} while
+\textit{private}[$i$] for each process $i$ is initialized to \texttt{True}.
+An important invariant, $\mathcal{I}_1$, of the program is that at any time at most
 one of these variables is \texttt{False}.
-Another invariant, $I_2(i)$, is that if process $i$ is in the critical section,
-then \texttt{private[$i$] == False}.
-Between the two (i.e., $I_1 \land \forall i: I_2(i)$),
+Another invariant, $\mathcal{I}_2(i)$, is that if process $i$ is in the critical section,
+then \textit{private}[$i$] == \texttt{False}.
+Between the two (i.e., $\mathcal{I}_1 \land \forall i: \mathcal{I}_2(i)$),
 it is clear that only one process can be in the
@@
 
-$I_1$ is an inductive invariant.
-To see that invariant $I_1$ is maintained, note that
-\texttt{\^{}p == True} upon entry of \texttt{tas}
+$\mathcal{I}_1$ is an inductive invariant.
+To see that invariant $\mathcal{I}_1$ is maintained, note that
+\texttt{\^{}}$p$ == \texttt{True} upon entry of \texttt{tas}
 (because of the condition on the \texttt{while} loop that the
@@
 \begin{enumerate}
-\item \texttt{\^{}s} is \texttt{False} upon entry to \texttt{tas}.
-Then upon exit \texttt{\^{}p == False} and \texttt{\^{}s == True}, maintaining
-the invariant.
-\item \texttt{\^{}s} is \texttt{True} upon entry to \texttt{tas}.
+\item \texttt{\^{}}$s$ is \texttt{False} upon entry to \texttt{tas}.
+Then upon exit \texttt{\^{}}$p$ == \texttt{False} and \texttt{\^{}}$s$ == \texttt{True},
+maintaining the invariant.
+\item \texttt{\^{}}$s$ is \texttt{True} upon entry to \texttt{tas}.
 Then upon exit nothing has changed, maintaining the invariant.
 \end{enumerate}
-Invariant $I_1$ is also easy to verify for exiting the critical section.
-Invariant $I_2(i)$ is obvious as (i) process $i$ only proceeds to the critical
-section if \texttt{private[$i$] == False}, and (ii) no other process modifies
-\texttt{private[$i$]}.
+Invariant $\mathcal{I}_1$ is also easy to verify for exiting the critical section.
+Invariant $\mathcal{I}_2(i)$ is obvious as (i) process $i$ only proceeds to the critical
+section if \textit{private}[$i$] == \texttt{False}, and (ii) no other process modifies
+\textit{private}[$i$].
 
@@
 CXL can check these invariants as well.  Figure~\ref{fig:tas} already
-has the code to check $I_2(i)$.  But how would one go about checking an
-invariant like $I_1$?  Invariants must hold for every state.
-For $I_2$ we only need an assertion at label \texttt{@cs} because the
+has the code to check $\mathcal{I}_2(i)$.  But how would one go about checking an
+invariant like $\mathcal{I}_1$?  Invariants must hold for every state.
+For $\mathcal{I}_2$ we only need an assertion at label \texttt{@cs} because the
 premisse is that there is a process at that label.  However, we would
-like to check $I_1$ in \emph{every state} (after the variables have
+like to check $\mathcal{I}_1$ in \emph{every state} (after the variables have
 been initialized).
@@
 \glsadd{blocked process}
+
 We call a process \emph{blocked}
@@
 \index{reader/writer lock}
+
 \glsadd{reader/writer lock}
@@
 \begin{itemize}
-\item $n$ readers at \texttt{@rcs} $\Rightarrow \mathtt{nreaders} \ge n$,
-\item $n$ writers at \texttt{@wcs} $\Rightarrow \mathtt{nwriters} \ge n$,
-\item $(\mathtt{nreaders} \ge 0 \land \mathtt{nwriters} = 0) \lor
-    (\mathtt{nreaders} = 0 \land 0 \le \mathtt{nwriters} \le 1)$.
+\item $n$ readers at \texttt{@rcs} $\Rightarrow \mathit{nreaders} \ge n$,
+\item $n$ writers at \texttt{@wcs} $\Rightarrow \mathit{nwriters} \ge n$,
+\item $(\mathit{nreaders} \ge 0 \land \mathit{nwriters} = 0) \lor
+    (\mathit{nreaders} = 0 \land 0 \le \mathit{nwriters} \le 1)$.
 \end{itemize}
@@
 \glsadd{busy waiting}
+
 While correct, it is not considered a good solution.
@@
 \begin{itemize}
-\item $n$ readers at \texttt{@rcs} $\Rightarrow \mathtt{nreaders} \ge n$,
-\item $\exists$ writer at \texttt{@wcs} $\Rightarrow \mathtt{nreaders} = 0$,
+\item $n$ readers at \texttt{@rcs} $\Rightarrow \mathit{nreaders} \ge n$,
+\item $\exists$ writer at \texttt{@wcs} $\Rightarrow \mathit{nreaders} = 0$,
 \end{itemize}
@@
 \index{semaphore}
+
 \glsadd{semaphore}
@@
 \index{producer/consumer problem}
+
 \glsadd{producer/consumer problem}
@@
 
+\glsadd{conditional critical section}
+
 Reader/writer locks and bounded buffers are both examples of what are
@@
 (CCSs)~\cite{Hoare73}.
-\glsadd{conditional critical section}
 All critical sections are, in a way, conditional.  Even the basic critical section
@@
 \index{starvation}
+
 \glsadd{starvation}
@@
 
+\glsadd{property}
+
 A \emph{property}
 \index{property}
-\glsadd{property}
 is a set of traces.
@@
 match the usual formal nor vernacular concepts of fairness).
+
 \glsadd{fairness}
@@
 
+\glsadd{monitor}
+\glsadd{condition variable}
+
 Tony Hoare, who came up with the concept of split binary semaphores, devised
@@
 \index{monitor}
-\glsadd{monitor}
 (A similar construct was independently invented by Per Brinch Hansen~\cite{BH73}.)
@@
 \index{condition variable}
-\glsadd{condition variable}
 There are two operations on condition variables: \texttt{wait}
@@
 \index{deadlock}
+
 \glsadd{deadlock}
@@
 \index{message passing}
+
 \glsadd{actor model}
@@
 \index{non-blocking synchronization}
+
 \glsadd{non-blocking synchronization}
@@
 \glsadd{linearizable}
+
 The implementation uses the \texttt{alloc} module for dynamic allocation
@@
 \index{barrier synchronization}
+
 \glsadd{barrier synchronization}

### 2020-06-26 book/paper.tex
Better variable name

@@
 na\"{\i}ve (and, as it turns out, broken) implementation of a lock.
-Initially the lock is not owned, indicated by lock being \texttt{False}.
-To enter the critical section, a process waits until \textit{lock} is \texttt{False}
+Initially the lock is not owned, indicated by \textit{lockTaken} being \texttt{False}.
+To enter the critical section, a process waits until \textit{lockTaken} is \texttt{False}
 and then sets it to \texttt{True} to indicate that the lock has been taken.
 The process then executes the critical section.  Finally the process
-releases the lock by setting it back to \texttt{False}.
+releases the lock by setting \textit{lockTaken} back to \texttt{False}.
 
 Unfortunately, if we run the program through CXL, we find that the assertion still
-fails.  Diagnosing the problem, we see that the \textit{lock} variable
+fails.  Diagnosing the problem, we see that the \textit{lockTaken} variable
 suffers from the same problem as the \textit{count} variable

### 2020-06-26 book/paper.tex
Renamed NULL to None

@@
 Each node has a lock, a value, and \texttt{next}, a pointer to the next node
-(which is \texttt{NULL} for the final \texttt{inf} node).
+(which is \texttt{None} for the final \texttt{inf} node).
 The \texttt{lst\_find(lst, v)} method is a helper function for update
@@
 \hline
-Address & \texttt{\&lock, \&flags[2], NULL} \\
+Address & \texttt{\&lock, \&flags[2], None} \\
 \hline
@@
 ordered;
-\item an address is a list of atoms.  \texttt{NULL} is the empty list of atoms.
+\item an address is a list of atoms.  \texttt{None} is the empty list of atoms.
 Addresses are lexicographically ordered accordingly;
@@
 The usage is similar to \texttt{malloc} and \texttt{free} in C.
-CXL also supports an address value \texttt{NULL} that can be
-used as a null pointer.  \texttt{recAlloc} returns \texttt{NULL}
-when running out of memory.
+\texttt{recAlloc} returns \texttt{None} when running out of memory.
 

### 2020-06-26 book/code/pp.py

@@ -105,7 +105,7 @@ tokens = []
 def putchar(c):
     print(c, end="")
 
-constants = { "False", "True", "NULL",
+constants = { "False", "True", "None",
     "lock", "unlock", "wait", "notify", "notifyAll", "signal", "P", "V" }
 
 def doimport(module, file):

### 2020-06-26 cxl.py

@@ -86,8 +86,8 @@ def isreserved(s):
         "max",
         "min",
         "nametag",
+        "None",
         "not",
-        "NULL",
         "or",
         "pass",
         "spawn",
@@ -334,6 +334,8 @@ class AddressValue(Value):
         self.indexes = indexes
 
     def __repr__(self):
+        if len(self.indexes) == 0:
+            return "None"
         result = "&" + self.indexes[0]
         for index in self.indexes[1:]:
             if isinstance(index, str):
@@ -1561,7 +1563,7 @@ class BasicExpressionRule(Rule):
             return (ConstantAST((False, file, line, column)), t[1:])
         if lexeme == "True":
             return (ConstantAST((True, file, line, column)), t[1:])
-        if lexeme == "NULL":
+        if lexeme == "None":
             return (ConstantAST((AddressValue([]), file, line, column)), t[1:])
         if lexeme == "inf":
             return (ConstantAST((math.inf, file, line, column)), t[1:])

### 2020-06-26 book/paper.tex
Latest

@@
 not only hard to reason about, but also hard to test.
+\index{test}
 Running the same test of concurrent code twice is likely to produce
@@
 because it is hard to reproduce.
-j
+
 \glsadd{model checking}
@@
 permutation of the input.  In some sense it is an excellent test:
-it will have considered all corner cases, including lists where all
+it will have considered all \emph{corner cases},
+\index{corner case}
+including lists where all
 numbers are the same, lists that are already sorted or reversely

### 2020-06-26 book/paper.tex
Still working on the fonts

@@
 that sorts all lists of up to five numbers correctly but fails to
-do so for a list of 6 numbers.  (Hint: simply use an \texttt{if}
+do so for a list of 6 numbers.  (Hint: simply use an \textbf{if}
 statement.)
@@
 \item Syntax: Every statement in CXL must be terminated by a semicolon
-(even \texttt{while} statements and \texttt{def} statements).
+(even \textbf{while} statements and \textbf{def} statements).
 In Python semicolons are optional and rarely used.
@@
 \item CXL does not (currently) support floating point, iterators, or I/O;
-CXL does support \texttt{for} loops and various comprehensions over sets.
+CXL does support \textbf{for} loops and various comprehensions over sets.
 \item Python is object-oriented, supporting classes with methods and
@@
 \textit{result} that eventually contains the result of the
-method (there is no \texttt{return} statement in CXL).  The method
+method (there is no \textbf{return} statement in CXL).  The method
 also has a variable called \textit{n} containing the value of the
@@
 operator provides the input to the program by selecting some value from the
-set~$S$, while the $\texttt{assert}$ statement checks that the output is
+set~$S$, while the $\textbf{assert}$ statement checks that the output is
 correct.  If the program is correct, the output of CXL is the size of the
@@
 \item What are the possible values of \textit{count} at the time of the
-\texttt{assert} statement?
+\textbf{assert} statement?
 (You may assume that \texttt{load} and \texttt{store} instructions of the
@@
 $\mathtt{dict}\{ k_0: v_0, ~ k_1: v_1, ~ ... \}$.
-If \texttt{d} is a dictionary, and \texttt{k} is a key, then the
-following expression retrieves the CXL value that $k$ maps to:
-\begin{code}
-\begin{verbatim}
-d k
-\end{verbatim}
+If $d$ is a dictionary, and $k$ is a key, then the
+following expression retrieves the CXL value that $k$ maps to in $d$:
+\begin{code}
+$d$ $k$
 \end{code}
@@
 \begin{code}
-\begin{verbatim}
-d[k]
-\end{verbatim}
-\end{code}
-However, if \texttt{d = dict\{ .count: 3 \}}, then you can write
-can write \texttt{d.count} (which has value 3) instead of having to write
-\texttt{d[.count]} (although both will work).
+$d$[$k$]
+\end{code}
+However, if $d$ = \texttt{dict\{ .count: 3 \}}, then you can write
+can write $d$.\texttt{count} (which has value 3) instead of having to write
+$d$\texttt{[.count]} (although both will work).
 Thus using atoms, a dictionary can be made to look much like a Python object.
-The meaning of \texttt{d a b c ...} is \texttt{((((d a) b) c) ...)}.
+The meaning of $d$ $a$ $b$ $...$ is ((($d$ $a$) $b$) $...$).
 
@@
 which in turn is the same CXL value as \texttt{dict\{ 0:a, 1:b, 2:c \}}.
-So if \texttt{x == [ False, True ]},
-then \texttt{x[0] == False} and \texttt{x[1] == True}, just like in Python.
+So if $x$ = \texttt{[ False, True ]},
+then $x$[0] = \texttt{False} and $x$[1] = \texttt{True}, just like in Python.
 However, when creating a singleton list, make sure you include the
@@
 For example, dictionaries have two handy unary operators.
-If \texttt{d} is a
-dictionary, then \texttt{keys~d} (or equivalently \texttt{keys(d)})
-returns the set of keys and \texttt{len~d} returns the size of
+If $d$ is a
+dictionary, then \texttt{keys}~$d$ (or equivalently \texttt{keys}($d$)
+returns the set of keys and \texttt{len}~$d$ returns the size of
 this set.
@@
     in broad steps, how running the program can get into a bad state.
-\item What if we moved line~2 of Figure~\ref{fig:incenter} to after the \texttt{if}
+\item What if we moved line~2 of Figure~\ref{fig:incenter} to after the \textbf{if}
     statement (between lines~7 and~8)?  Do you think that would work?  Run it through
@@
 ensure is that there can never be two processes in the critical section.
-This property is called \texttt{mutual exclusion}.
+This property is called \emph{mutual exclusion}.
 \index{mutual exclusion}
@@
 CXL in fact supports this.
-It has an operator \texttt{atLabel $L$},
+It has an operator \texttt{atLabel}~$L$,
 \index{atLabel operator}
@@
 This code will surely prevent two or more processes from being
-at label \texttt{cs} at the same time.
+at label \texttt{@cs} at the same time.
 But it does so by preventing \emph{any} process from reaching
@@
 the assertion should fail, and a trace should be reported.
-Now plug the following code \texttt{while True: pass;;}
+Now plug the following code \textbf{while} \texttt{True}: \textbf{pass};;
 just before entering the critical section
@@
 win because writes to memory in CXL are atomic.
-The process continues to be polite, waiting in a \texttt{while} loop
+The process continues to be polite, waiting in a \textbf{while} loop
 until either the other process is nowhere near the critical section
@@
 It is easy to construct one: let process~0 be at label \texttt{@cs}
-and process~1 at the start of the \texttt{while} loop.
+and process~1 at the start of the \textbf{while} loop.
 Also imagine that in this particular state $\mathit{turn} = 1$.  Now let
@@
 Let $\mathcal{C}(i) = \mathit{flags}[1 - i] \land
-\mathit{turn} = 1 - i$, that is, the condition on the \texttt{while} loop
+\mathit{turn} = 1 - i$, that is, the condition on the \textbf{while} loop
 for process~$i$.
@@
 if process $i$ is in the critical section, then $\lnot \mathcal{C}(i)$ (i.e.,
-$\mathcal{I}_p(i)$ without the disjunct that process $1-i$ is at label \texttt{gate}).
+$\mathcal{I}_p(i)$ without the disjunct that process $1-i$ is at label \texttt{@gate}).
 To demonstrate that this predicate is not invariant, you can remove the
@@
 
-A method \texttt{m} with argument \texttt{a} is invoked in its
+A method \texttt{m} with argument $a$ is invoked in its
 most basic form as follows (assigning the result to~\texttt{r}).
 \begin{code}
-\begin{verbatim}
-r = m a;
-\end{verbatim}
+$r$ = m $a$;
 \end{code}
 That's right, no parentheses are required.  In fact, if you invoke
-\texttt{m(a)}, the argument is \texttt{(a)}, which is the same
-as \texttt{a}.
+\texttt{m}($a$), the argument is ($a$), which is the same
+as $a$.
 If you invoke \texttt{m()}, the argument is \texttt{()},
 which is the empty tuple.
-If you invoke \texttt{m(a, b)}, the argument is \texttt{(a, b)},
-the tuple consisting of values \texttt{a} and \texttt{b}.
+If you invoke \texttt{m}($a$, $b$), the argument is ($a$, $b$),
+the tuple consisting of values $a$ and $a$.
 
@@
 If \texttt{f} is either a method or a
-dictionary, and \texttt{x} is an arbitrary CXL value, then
-\texttt{f x}, \texttt{f(x)}, and \texttt{f[x]} are all
+dictionary, and $x$ is an arbitrary CXL value, then
+\texttt{f} $x$, \texttt{f}($x$), and \texttt{f}[$x$] are all
 the same expression in CXL.
@@
 to shared variables.
-If \texttt{x} is a shared variable, then the expression \texttt{\&x}
- is a pointer to \texttt{x} (also known as the \emph{address}
+If $x$ is a shared variable, then the expression \&$x$
+ is a pointer to $x$ (also known as the \emph{address}
 \index{address}
-of \texttt{x}).
-Conversely, if \texttt{p} is a pointer to a shared variable, then the
-expression \texttt{\^{}p} is the value of the shared variable.
-In particular, \texttt{\^{}\&x == x}.
+of $x$).
+Conversely, if $p$ is a pointer to a shared variable, then the
+expression \texttt{\^{}}$p$ is the value of the shared variable.
+In particular, \texttt{\^{}\&}$x$ == $x$.
 This is similar to how C pointers work.
@@
 section.
-The name \texttt{mutex} is often used to denote a variable or value
+The name \textit{mutex} is often used to denote a variable or value
 that is used for mutual exclusion.
@@
 a pointer to a mutex and the process identifier (0 or 1).
-\texttt{(\^{}pm).turn} is the value of the \texttt{.turn} key
-in the mutex dictionary.  Note that the parentheses in this expression
-are needed, as \texttt{\^{}pm.turn} would wrongly evaluate to
-\texttt{\^{}(pm.turn)}.
+\texttt{(\^{}}\textit{pm}\texttt{.turn} is the value of the \texttt{.turn} key
+in the dictionary that \textit{pm} points to. 
+Note that the parentheses in this expression
+are needed, as \texttt{\^{}}\textit{pm}\texttt{.turn} would wrongly evaluate to
+\texttt{\^{}}(\textit{pm}\texttt{.turn}).
 
@@
 \texttt{\^{}}$p$ == \texttt{True} upon entry of \texttt{tas}
-(because of the condition on the \texttt{while} loop that the
+(because of the condition on the \textbf{while} loop that the
 \texttt{tas} method is invoked in).
@@
 the other processes.
-In CXL, \texttt{assert} statements are executed atomically, so the
+In CXL, \textbf{assert} statements are executed atomically, so the
 evaluation of the assertion is not interleaved with the execution
@@
 to enter the critical section and do not yet know if they need to wait.
-We shall call this the \texttt{mutex}.
+We shall call this the \textit{mutex}.
 
@@
 As it were, these semaphores taken together are emulating a single binary semaphore.
-Initially, \texttt{mutex} is~1 and the other $N$ semaphores
+Initially, \textit{mutex} is~1 and the other $N$ semaphores
 are all~0.  To maintain the invariant, each process alternates \texttt{P} and
@@
 one---it would not matter.  However, it can only vacate one of the semaphores.
-If none of the conditions hold, \texttt{V\_one()} vacates \texttt{mutex}.
+If none of the conditions hold, \texttt{V\_one()} vacates \textit{mutex}.
 
@@
 found there is a writer in the critical section, it vacates
-\texttt{mutex} and procures \texttt{r\_sema},
+\textit{mutex} and procures \texttt{r\_sema},
 blocking on the semaphore associated
@@
 \begin{enumerate}
-\item After procuring \texttt{mutex},
+\item After procuring \textit{mutex},
 the sum of the three semaphores is 0.  The writer then decrements
@@
 \texttt{V\_one()} does not find any processes waiting,
-and so it vacates \texttt{mutex}.
+and so it vacates \textit{mutex}.
 \end{enumerate}
@@
 There is a split binary semaphore associated with each such class.
-The \texttt{mutex} is hidden: it is automatically acquired when invoking a
+The \textit{mutex} is hidden: it is automatically acquired when invoking a
 method and released upon exit.
@@
 When inspecting the code, one important feature to notice is that \texttt{wait} is always
-invoked within a \texttt{while} loop that checks for the condition that the process is
+invoked within a \textbf{while} loop that checks for the condition that the process is
 waiting for.  We explained why above: when a process resumes after
@@
 if it did at the time of invoking \texttt{notify} or \texttt{notifyAll}.
-It is \emph{imperative} that there is always a \texttt{while} loop around any invocation
+It is \emph{imperative} that there is always a \textbf{while} loop around any invocation
 of \texttt{wait} containing the negation of the condition that the process is waiting for.
@@
 Second, both readers and writers are awakened---this is ok because both execute
-\texttt{wait} within a \texttt{while} loop, re-checking the condition that they
+\texttt{wait} within a \textbf{while} loop, re-checking the condition that they
 are waiting for.  So if both type of processes are waiting, either all the readers
@@
 \index{multiple conditions, waiting on}
-The condition in the \texttt{while} statement is the negation of the
+The condition in the \textbf{while} statement is the negation of the
 condition that the diner is waiting and consists of two disjuncts.
-Within the \texttt{while} statement,
-there is an \texttt{if} statement for each disjunct.
+Within the \textbf{while} statement,
+there is an \textbf{if} statement for each disjunct.
 The code waits for either or both forks if necessary.  After that, it goes
-back to the top of the \texttt{while} loop.
+back to the top of the \textbf{while} loop.
 

### 2020-06-26 book/code/pp.py

@@ -266,7 +266,7 @@ def lexer(s, file, silent):
         elif s[0] == "^":
             if not silent: print("\\^{}", end="")
         else:
-            if s[0] == "@":
+            if s[0] in ["@", "."]:
                 nextconst = True
             if not silent:
                 if s[0] == "-":

### 2020-06-26 book/paper.tex
Working on fonts.  Finished up to Chapter 9

@@
 CXL in fact supports this.
-It has an operator \texttt{atLabel}~$L$,
+It has an operator \textbf{atLabel}~$L$,
 \index{atLabel operator}
@@
 in the bag to the number of times the element appears in the bag.
-Method \texttt{atLabel} only exists for specification purposes---do not
+Method \textbf{atLabel} only exists for specification purposes---do not
 use it in normal code.
-The assertion also makes use of the \texttt{nametag()} operator
+The assertion also makes use of the \textbf{nametag}() operator
 \index{nametag operator}
@@
 of it.  Second, a group of CXL statements can be made atomic
-through its \texttt{atomic}
+through its \textbf{atomic}
 \index{atomic statement}
 statement.
-We can use \texttt{atomic} statement blocks to implement a wide variety of
+We can use \textbf{atomic} statement blocks to implement a wide variety of
 interlock operations.
@@
 Method \texttt{tas} in Figure~\ref{fig:tas} shows its specification.
-Here \texttt{s} points to a shared boolean variable and \texttt{p}
+Here $s$ points to a shared boolean variable and $p$
 to a private boolean variable, belonging to some process.
@@
 Figure~\ref{fig:tas} goes on how to implement mutual exclusion for
-a set of $N$ processes.
+a set of \texttt{N} processes.
 The approach is called \emph{spinlock},
@@
 it can acquire the lock.
-The program uses $N+1$ variables.
+The program uses $\mathtt{N}+1$ variables.
 Variable \textit{shared} is initialized to \texttt{False} while
@@
 in a state.  It introduces a new feature of CXL: the ability to have
-variables local to a method.  In this case, the process variable \texttt{sum}
+variables local to a method.  In this case, the process variable \textit{sum}
 is used to compute the number of shared variables that have value
@@
 but they are accessed only by their respective processes.
-There is no need to keep \texttt{private} as a shared
+There is no need to keep \textit{private} as a shared
 variable---we only did so to be able to show and check the invariants.
@@
 \emph{atomic section} (the statements executed within an
-\texttt{atomic} statement) and a \emph{critical section}
+\textbf{atomic} statement) and a \emph{critical section}
 (protected by a lock of some sort).
@@
 to you.  But if you want to design a new synchronization primitive, then
-use \texttt{atomic} by all means.
+use \textbf{atomic} by all means.
 
@@
 \begin{code}
-\begin{verbatim}
-stop L
-\end{verbatim}
-\end{code}
-
-Here \texttt{L} is a shared list.
-Another process can revive the process using the \texttt{go}
+\textbf{stop} $L$
+\end{code}
+
+Here $L$ is a shared variable containing a list.
+Another process can revive the process using the \textbf{go}
 \index{go statement}
@@
 \begin{code}
-\begin{verbatim}
-go C R
-\end{verbatim}
-\end{code}
-
-Here \texttt{C} is a context and \texttt{R} is a CXL value.
-It causes a process with context \texttt{C} to be added to the state that has
-just executed the \texttt{stop}
+\textbf{go} $C$ $R$
+\end{code}
+
+Here $C$ is a context and $R$ is a CXL value.
+It causes a process with context $C$ to be added to the state that has
+just executed the \textbf{stop}
 \index{stop expression}
-expression.  The \texttt{stop} expression returns the value \texttt{R}.
+expression.  The \textbf{stop} expression returns the value $R$.
 
@@
 sets the lock if available and suspends itself if not.
-Note that \texttt{stop} is called within an \texttt{atomic} statement---this is
+Note that \textbf{stop} is called within an \textbf{atomic} statement---this is
 the only exception to an atomic statement running to completion.  While the
@@
 \hline
-Context & (generated by \texttt{stop} expression) \\
+Context & (generated by \textbf{stop} expression) \\
 \hline

### 2020-06-27 book/paper.tex
Now fonts checked through Chapter 12

@@
 A process that wants to read first waits until there are no writers:
-\texttt{nwriters == 0}.  Then it increments the number of readers.
+\textit{nwriters} == 0.  Then it increments the number of readers.
 Similarly, a
@@
 You can test the code by running \texttt{cxl -m RW=RWlock RWtest}.
-It uses two ordinary locks: \texttt{rwlock} is used by both readers and writers,
-while \texttt{rlock} is only used by readers.
-\texttt{rwlock} is held either when readers are at label \texttt{@rcs}
+It uses two ordinary locks: \textit{rwlock} is used by both readers and writers,
+while \textit{rlock} is only used by readers.
+\textit{rwlock} is held either when readers are at label \texttt{@rcs}
 or when a writer is at label \texttt{@wcs}.
-\texttt{rlock} is used to protect the \texttt{nreaders} variable that
+\textit{rlock} is used to protect the \textit{nreaders} variable that
 counts the number of readers in the critical section.
@@
 
-A writer simply acquires \texttt{rwlock} to enter the critical section
+A writer simply acquires \textit{rwlock} to enter the critical section
 and releases it to exit.  The \emph{first} reader to enter the critical
-section acquires \texttt{rwlock} and the \emph{last} reader to exit
-the critical section releases \texttt{rwlock}.
+section acquires \textit{rwlock} and the \emph{last} reader to exit
+the critical section releases \textit{rwlock}.
 The implementation satisfies progress: if no process is in the critical
@@
 section and two readers try to enter.  The first reader successfully
-acquires \texttt{rlock} and but blocks when trying to acquire
-\texttt{rwlock}, which is held by the writer.  The second reader blocks
-trying acquire \texttt{rlock} because it is held by the first reader.
+acquires \textit{rlock} and but blocks when trying to acquire
+\textit{rwlock}, which is held by the writer.  The second reader blocks
+trying acquire \textit{rlock} because it is held by the first reader.
 When the writer leaves the critical section, the first reader acquires
-\texttt{rwlock}, sets \texttt{nreaders} to 1, and releases \texttt{rlock}.
-\texttt{rwlock} is still held.
-Then the second reader acquires \texttt{rlock} and, assuming the first
-reader is still in the critical section, increments \texttt{nreaders} to 2
-and enters the critical section \emph{without} acquiring \texttt{rwlock}.
-\texttt{rwlock} is essentially jointly held by both readers.
+\textit{rwlock}, sets \textit{nreaders} to 1, and releases \textit{rlock}.
+\textit{rwlock} is still held.
+Then the second reader acquires \textit{rlock} and, assuming the first
+reader is still in the critical section, increments \textit{nreaders} to 2
+and enters the critical section \emph{without} acquiring \textit{rwlock}.
+\textit{rwlock} is essentially jointly held by both readers.
 It does not matter in which order they leave: the second will release
-\texttt{rwlock}.
+\textit{rwlock}.
 
@@
 algorithm, and Figure~\ref{fig:qsorttest} an accompanying test program.
-The array to be sorted is stored in  shared variable \texttt{a}.
-Another shared variable, \texttt{todo}, contains the ranges of the
+The array to be sorted is stored in  shared variable $a$.
+Another shared variable, \textit{todo}, contains the ranges of the
 array that need to be sorted (initially the entire array).
@@
 or \texttt{sortrange} for this.  Create \texttt{NWORKERS} ``worker processes''
-that should replace the \texttt{qsort} code.  Each worker loops until \texttt{todo}
+that should replace the \texttt{qsort} code.  Each worker loops until \textit{todo}
 is empty and sorts the ranges that it finds until then.  The \texttt{main}
@@
 once again, an implementation of reader/writer locks.
-Figure~\ref{fig:RWcv} shows the code.  \texttt{rwlock} is the global lock or mutex for the
-reader/writer lock.  There are two condition variables: readers wait on \texttt{rcond} and
-writers wait on \texttt{wcond}.  The implementation also keeps track of the number of
+Figure~\ref{fig:RWcv} shows the code.  \textit{rwlock} is the global lock or mutex for the
+reader/writer lock.  There are two condition variables: readers wait on \textit{rcond} and
+writers wait on \textit{wcond}.  The implementation also keeps track of the number of
 readers and writers in the critical section.
@@
 \item \texttt{m()} invokes method \texttt{m} with the empty dictionary \texttt{()} as argument;
-\item \texttt{m(a)} invokes method \texttt{m} with argument \texttt{a};
-\item \texttt{m(a, b, c)} invokes method \texttt{m} with tuple \texttt{(a, b, c)} as argument.
+\item \texttt{m($a$)} invokes method \texttt{m} with argument $a$;
+\item \texttt{m($a$, $b$, $c$)} invokes method \texttt{m} with tuple \texttt{($a$, $b$, $c$)} as argument.
 \end{itemize}

### 2020-06-27 cxl.py

@@ -3291,7 +3291,7 @@ def htmlnode(s, visited, code, scope, f, verbose):
 
     print("<td valign='top'>", file=f)
     print("<table border='1'>", file=f)
-    print("<tr><th>Variable</th><th>Value</th></tr>", file=f)
+    print("<tr><th>Shared</th><th>Value</th></tr>", file=f)
     for (key, value) in s.vars.d.items():
         print("<tr>", file=f)
         print("<td>%s</td>"%strValue(key)[1:], file=f)

### 2020-06-27 book/paper.tex
Updates

@@
 \begin{verbatim}
-#states = 79
+#states = 79 diameter = 5
 ==== Safety violation ====
-CXL Assertion failed ('assert', 'Up.cxl', 9, 5): 1
 __init__/() [0,34-49]       50 dict{.count:0, .done:dict{0:False, 1:False}}
@@
 main/() [13-17,20-24,26-31] 31 dict{.count:1, .done:dict{0:True,  1:True }}
+>>> CXL Assertion failed : 1
+Open file:///Users/rvr/github/cxl/cxl.html for more information
 \end{verbatim}

### 2020-06-27 book/paper.tex
Starting to add screenshots

--- 2020-06-27 book/paper.tex
@@
 main/() [13-17,20-24,26-31] 31 dict{.count:1, .done:dict{0:True,  1:True }}
+>>> CXL Assertion failed : 1
+Open file://.../cxl.html for more information
 \end{verbatim}
 \end{code}
-\caption{The output of running Figure~\ref{fig:inc}.}
+\caption{The text output of running CXL on Figure~\ref{fig:inc}.}
 \label{fig:incoutput}
+\end{figure}
+
+\begin{figure}
+\includegraphics[width=\textwidth]{figures/Up1.png}
+\caption{The HTML output of running CXL on Figure~\ref{fig:inc}.}
+\label{fig:inchtml1}
 \end{figure}

### 2020-06-28 book/paper.tex
Latest
Now prints right variables
Added information about HTML output

@@
 
-\begin{figure}
-\begin{code}
-\begin{verbatim}
-Up.cxl:1 def incrementer(self):
-   0 Jump 34
-   1 Frame incrementer(self) end=11
-Up.cxl:2     count = count + 1;
-   2 Load count
-   3 Push 1
-   4 2-ary +
-   5 Store count
-Up.cxl:3     done[self] = True;
-   6 LoadVar self
-   7 PushAddress done
-   8 Address 2
-   9 Push True
-   10 Store
-   11 Return
-\end{verbatim}
-\end{code}
-\caption{The first part of the CVM bytecode corresponding to Figure~\ref{fig:inc}.}
-\label{fig:inccode}
-\end{figure}
-
-\begin{figure}
-\begin{code}
-\begin{verbatim}
-#states = 79 diameter = 5
-==== Safety violation ====
-__init__/() [0,34-49]       50 dict{.count:0, .done:dict{0:False, 1:False}}
-incrementer/0 [1-4]          5 dict{.count:0, .done:dict{0:False, 1:False}}
-incrementer/1 [1-10]        11 dict{.count:1, .done:dict{0:False, 1:True }}
-incrementer/0 [5-10]        11 dict{.count:1, .done:dict{0:True,  1:True }}
-main/() [13-17,20-24,26-31] 31 dict{.count:1, .done:dict{0:True,  1:True }}
->>> CXL Assertion failed : 1
-Open file://.../cxl.html for more information
-\end{verbatim}
-\end{code}
-\caption{The text output of running CXL on Figure~\ref{fig:inc}.}
-\label{fig:incoutput}
-\end{figure}
-
-\begin{figure}
-\includegraphics[width=\textwidth]{figures/Up1.png}
-\caption{The HTML output of running CXL on Figure~\ref{fig:inc}.}
-\label{fig:inchtml1}
-\end{figure}
-
 A CXL program is translated
@@
 
+\begin{figure}
+\begin{code}
+\begin{verbatim}
+Up.cxl:1 def incrementer(self):
+    0 Jump 34
+    1 Frame incrementer(self) end=11
+Up.cxl:2     count = count + 1;
+    2 Load count
+    3 Push 1
+    4 2-ary +
+    5 Store count
+Up.cxl:3     done[self] = True;
+    6 LoadVar self
+    7 PushAddress done
+    8 Address 2
+    9 Push True
+   10 Store
+   11 Return
+\end{verbatim}
+\end{code}
+\caption{The first part of the CVM bytecode corresponding to Figure~\ref{fig:inc}.}
+\label{fig:inccode}
+\end{figure}
+
 It may seem strange that there is only one memory location and that each
@@
 
+\begin{figure}
+\begin{code}
+\begin{verbatim}
+#states = 79 diameter = 5
+==== Safety violation ====
+__init__/()   [0,34-49]     50 { count=0, done=[False, False] }
+incrementer/0 [1-4]          5 { count=0, done=[False, False] }
+incrementer/1 [1-10]        11 { count=1, done=[False, True]  }
+incrementer/0 [5-10]        11 { count=1, done=[True,  True]  }
+main/() [13-17,20-24,26-31] 31 { count=1, done=[True,  True]  }
+>>> CXL Assertion failed : 1
+Open file://.../cxl.html for more information
+\end{verbatim}
+\end{code}
+\caption{The text output of running CXL on Figure~\ref{fig:inc}.}
+\label{fig:incoutput}
+\end{figure}
+
 \glsadd{stack machine}
@@
 \texttt{Up.cxl} program.
-It starts by reporting that the assertion on line 9 in column 5 failed and that
-\textit{count} has value 1 instead of 0:
-\texttt{Assertion failed (’assert’, ’Up.cxl’, 9, 5) 1}.
-Next it reports an execution that failed this assertion.  The output has
-four columns:
-\begin{enumerate}
-\item The name tag of the process;
-\item The sequence of program counters of the CVM instructions that the process executed;
-\item The current program counter of the process;
-\item The contents of the shared memory.
-\end{enumerate}
-
-If we look at the middle three rows, we see that:
-\begin{itemize}
-\item Process 0 executed instructions 1 through 4, loading the value of
-\textit{count} but stopping just before storing 1 into \textit{count};
-\item Process 1 executed instructions 1 through 10, storing 1 into
-into \textit{count} and storing \texttt{True} into \textit{done}[1];
-\item Process 0 continues execution, storing value 1 into \textit{count}
-and storing \texttt{True} into \textit{done}[0].
-\end{itemize}
-
-This makes precise the concurrency problem that we encountered.
+
+\begin{figure}
+\includegraphics[width=\textwidth]{figures/Up1.png}
+\caption{The HTML output of running CXL on Figure~\ref{fig:inc}.
+There are four sections.  The top left shows the source code and bytecode
+of the program.  The top right shows the steps in an execution.
+The middle section shows the values of the variables at a particular step.
+The bottom section shows the state of each process.
+Each row in the stack trace of a process shows the current program counter,
+the method that is being evaluated, and the line of code that is
+being evaluated.}
+\label{fig:inchtml1}
+\end{figure}
 
@@
 or non-existent variable, dividing a set by an integer, and so on.
+It also includes certain infinite loops.
 CXL will print at trace of the shortest bad execution that it found.
-\item \texttt{Infinite loop}: some process ran into an infinite loop
-in which it got back to the same state without having accessed any
-shared variables.
 \item \texttt{Non-terminating States}: CXL found one or more states
@@
 \end{itemize}
+
+In this case, CXL reports a safety violation, in particular it
+reports that the assertion on line 9 failed and that
+\textit{count} has value 1 instead of 2.
+Next it reports an execution that failed this assertion.
+The program got to the faulty execution in 5 ``steps.''
+The output has
+four columns:
+\begin{enumerate}
+\item The name tag of the process;
+\item The sequence of program counters of the CVM instructions that the process executed;
+\item The current program counter of the process;
+\item The contents of the shared memory.
+\end{enumerate}
+
+The first step is initialization.
+It sets shared variable \textit{count} to 0 and
+shared variable \textit{done} to \texttt{[False, False]}.
+If we look at the next three steps, we see that:
+\begin{itemize}
+\item Process \texttt{implementer/0}
+executed instructions 1 through 4, loading the value of
+\textit{count} but stopping just before storing 1 into \textit{count};
+\item Process \texttt{implementer/1}
+executed instructions 1 through 10, storing 1 into
+into \textit{count} and storing \texttt{True} into \textit{done}[1];
+\item Process \texttt{implementer/0}
+continues execution, storing value 1 into \textit{count}
+and storing \texttt{True} into \textit{done}[0].
+\end{itemize}
+Finally, process \texttt{main()} runs and finds the problem.
+This makes precise the concurrency issue that we encountered.
+
+CXL also generates an HTML file that allows exploring more details
+of the execution interactively.
+Open the suggested HTML file and you should see something like
+Figure~\ref{fig:inchtml1}.
 

### 2020-06-28 book/paper.tex
Latest

@@
 processes (aka \emph{threads}\footnote{We
-\index{thread}
+\index{thread}%
 do not make a distinction between processes and threads.})
 read and update shared variables
-\index{shared variable}
+\index{shared variable}%
 concurrently and synchronize with one another makes programs more
@@
 \item The execution of a sequential
-\index{sequential}
+\index{sequential}%
 program is mostly \emph{deterministic}.
-\index{determinism}
+\index{determinism}%
 If you run it twice with the same input, the same output will be produced.
@@
 (so-called \emph{Heisenbugs}).
-\index{Heisenbug}
+\index{Heisenbug}%
 \item In a sequential program, each statement and each function can be
 thought of as happening \emph{atomically}
-\index{atomicity}
+\index{atomicity}%
 because there is no other activity interfering with their execution.
@@
 not only hard to reason about, but also hard to test.
-\index{test}
+\index{test}%
 Running the same test of concurrent code twice is likely to produce
@@
 The approach is based on \emph{model checking}:
-\index{model checking}
+\index{model checking}%
 instead of relying
@@
 Model checking is not a replacement for formal verification.
-\index{formal verification}
+\index{formal verification}%
 Formal verification proves that a program is correct.  Model checking only
@@
 it will have considered all \emph{corner cases},
-\index{corner case}
+\index{corner case}%
 including lists where all
@@
 \emph{invariants}:
-\index{invariant}
+\index{invariant}%
 predicates that must hold for every state in the
@@
 Essentially, the $\texttt{choose}($S$)$
-\index{choose operator}
+\index{choose operator}%
 operator provides the input to the program by selecting some value from the
@@
 line using the \texttt{-c} option.
-\index{constant}
+\index{constant}%
 For example, if you want to test the code for \texttt{N = 100}, run:
@@
 processes
-\index{process}
+\index{process}%
 running in parallel while sharing variables.
@@
 What is going on is that the CXL program is compiled to machine instructions,
-\index{machine instruction}
+\index{machine instruction}%
 and it is the machine instructions that are executed by the underlying CXL
@@
 instructions are interleaved
-\index{interleaving}
+\index{interleaving}%
 in unspecified and often random ways.
@@
 This is known as a \emph{race condition}.
-\index{race condition}
+\index{race condition}%
 When running CXL, it will
@@
 \label{ch:cxlmachine}
-\index{CXL Virtual Machine}
+\index{CXL Virtual Machine}%
 
 CXL programs are compiled to CXL bytecode
-\index{bytecode}
+\index{bytecode}%
 (machine instructions for a virtual machine),
 which in turn is executed by the CXL virtual machine (CVM).
-\index{virtual machine}
-\index{CXL Virtual Machine}
-\index{CVM}
+\index{virtual machine}%
+\index{CXL Virtual Machine}%
+\index{CVM}%
 To understand the problem of concurrent computing, it
@@
 and dictionaries
-\index{dictionary}
+\index{dictionary}%
 that map CXL values to other CXL values.
@@
 Another type of CXL value is the \emph{atom}.
-\index{atom}
+\index{atom}%
 It is essentially
@@
 Each process has an immutable (but not necessarily unique) \emph{name tag},
-\index{name tag}
+\index{name tag}%
 a program counter,
-\index{program counter}
+\index{program counter}%
 a stack of CXL values,
 and a single mutable general purpose \emph{register}
-\index{register}
+\index{register}%
 that contains a CXL value.
@@
 along with an optional tag specified in the \texttt{spawn}
-\index{spawn statement}
-\index{tag}
+\index{spawn statement}%
+\index{tag}%
 statement.
@@
 The state of a process is called a \emph{context} (aka \emph{continuation}):
-\index{context}
-\index{continuation}
+\index{context}%
+\index{continuation}%
 it contains the value of
@@
 consists of the value of its memory and the multiset (or \emph{bag})
-\index{bag}
-\index{multiset}
+\index{bag}%
+\index{multiset}%
 of contexts.  It is a multiset of contexts because two different processes can
@@
 a dictionary that maps atoms to CXL values.  We call this a \emph{directory}.
-\index{directory}
+\index{directory}%
 A directory represents the state of a collection of variables named by the atoms.
@@
 such a sequence the \emph{address}
-\index{address}
+\index{address}%
 of a CXL value, and it is relative to the
@@
 Each process in the CVM is predominantly a \emph{stack machine},
-\index{stack machine}
+\index{stack machine}%
 but it also has a register.
@@
 time.
-\index{critical section}
-\index{critical region}
+\index{critical section}%
+\index{critical region}%
 
@@
 processes (or threads) and free of race conditions is called \emph{thread-safe}.
-\index{thread safety}
-\index{thread}
+\index{thread safety}%
+\index{thread}%
 
@@
 Here \texttt{@cs} is a \emph{label},
-\index{label}
+\index{label}%
 identifying a location in the CVM bytecode.  The first thing we need to
@@
 This property is called \emph{mutual exclusion}.
-\index{mutual exclusion}
+\index{mutual exclusion}%
 We would like to place an assertion at the \texttt{@cs} label that
@@
 It has an operator \textbf{atLabel}~$L$,
-\index{atLabel operator}
+\index{atLabel operator}%
 where $L$
@@
 The assertion also makes use of the \textbf{nametag}() operator
-\index{nametag operator}
+\index{nametag operator}%
 that returns the name tag of the current process.
@@
 Mutual exclusion is an example of a \emph{safety property},
-\index{safety property}
+\index{safety property}%
 a property that ensures that \emph{nothing bad will happen}, in this case
@@
 What we need now a \emph{liveness property}:
-\index{liveness property}
+\index{liveness property}%
 we want to ensure that
@@
 We call this \emph{progress}.
-\index{progress}
+\index{progress}%
 
@@
 \begin{center}
-\includegraphics[width=4in]{figures/mutex-crop.pdf}
+\includegraphics[width=3.5in]{figures/mutex-crop.pdf}
 \end{center}
@@
 Figure~\ref{fig:mutex} shows a high-level state diagram
-\index{state diagram} of mutual exclusion and progress.
+\index{state diagram} of mutual exclusion and progress.%
 The circles represent states, while the arrows represent possible state
@@
 You may already have heard of the concept of a \emph{lock}
-\index{lock}
+\index{lock}%
 and have realized that
@@
 \label{ch:peterson}
-\index{Peterson's Algorithm}
+\index{Peterson's Algorithm}%
 
@@
 In Chapter~\ref{ch:cxlmachine} we talked about the concept of \emph{state}:
-\index{state}
+\index{state}%
 at any point in time the CVM is in a specific state.
@@
 the values of the process variables
-\index{process variable}
+\index{process variable}%
 for each process, including its
@@
 changes).  We call that a \emph{step}.
-\index{step}
+\index{step}%
 Steps in CXL are atomic.
@@
 one process and its program counter is~0.  A \emph{trace}
-\index{trace}
+\index{trace}%
 is a sequence of steps starting from the initial state.
 When making a step, there are two sources of non-determinism
-\index{non-determinism}
+\index{non-determinism}%
 in CXL.
@@
 We call a state \emph{reachable}
-\index{reachable state}
+\index{reachable state}%
 if it is either the initial state
@@
 An \emph{invariant} of a program
-\index{invariant}
+\index{invariant}%
 is a predicate that holds over all states that are reachable by that program.
@@
 \emph{inductive invariant}.
-\index{inductive invariant}
+\index{inductive invariant}%
 An inductive invariant $\mathcal{I}$ is a predicate over states that satisfies the following:
@@
 \label{ch:method}
-\index{CXL method}
+\index{CXL method}%
 
@@
 For this (as well as some other uses), CXL supports \emph{pointers}
-\index{pointer}
+\index{pointer}%
 to shared variables.
@@
  is a pointer to $x$ (also known as the \emph{address}
-\index{address}
+\index{address}%
 of $x$).
@@
 and include it using the CXL \texttt{import} statement.
-\index{import statement}
-\index{module}
+\index{import statement}%
+\index{module}%
 This would
@@
 \label{ch:spinlock}
-\index{spinlock}
+\index{spinlock}%
 
@@
 so-called \emph{interlock instructions}:
-\index{interlock instruction}
+\index{interlock instruction}%
 special machine instructions
@@
 through its \textbf{atomic}
-\index{atomic statement}
+\index{atomic statement}%
 statement.
@@
 Many CPUs have an atomic ``test-and-set'' (TAS)
-\index{test-and-set}
-\index{TAS}
+\index{test-and-set}%
+\index{TAS}%
 operation.
@@
 The approach is called \emph{spinlock},
-\index{spinlock}
+\index{spinlock}%
 because each process is ``spinning'' until
@@
 \label{ch:synch}
-\index{lock}
-\index{synch module}
+\index{lock}%
+\index{synch module}%
 
@@
 We call a process \emph{blocked}
-\index{blocked process}
+\index{blocked process}%
 in a state if the process is waiting on a low-level synchronization
@@
 Another process can revive the process using the \textbf{go}
-\index{go statement}
+\index{go statement}%
 statement:
@@
 just executed the \textbf{stop}
-\index{stop expression}
+\index{stop expression}%
 expression.  The \textbf{stop} expression returns the value $R$.
@@
 \texttt{lock}
-\index{lock}
+\index{lock}%
 sets the lock if available and suspends itself if not.
@@
 \texttt{unlock}
-\index{unlock}
+\index{unlock}%
 checks to see if any processes are waiting to get the lock.
@@
 choice in lock implementations as it prevents \emph{starvation}:
-\index{starvation}
+\index{starvation}%
 every process
@@
 like to use with the \texttt{-m} flag.
-\index{module}
+\index{module}%
 For example,
@@
 \label{ch:rdwrbusy}
-\index{reader/writer lock}
+\index{reader/writer lock}%
 
@@
 \begin{center}
-\includegraphics[width=2.5in]{figures/rdwr-crop.pdf}
+\includegraphics[width=2.3in]{figures/rdwr-crop.pdf}
 \end{center}
@@
 The solution is an example of what is called \emph{busy-waiting}
-\index{busy waiting}
+\index{busy waiting}%
 (aka \emph{spin-waiting}):
-\index{spin waiting}
+\index{spin waiting}%
 processes spin in a loop until some desirable application-level condition is met.
@@
 \label{ch:semaphore}
-\index{semaphore}
+\index{semaphore}%
 
@@
 
-Consider a variant of the Dining Philosophers (which were clearly Dutch---who
-else need two forks to eat a meal?) that we call Italian Philosophers (who only
-use one fork each to eat spaghetti---they need the other to speak\footnote{The joke
-was provided to me by my dear friend and colleague Lorenzo Alvisi.}).
-Again, there are five philosophers.
-This time, however, there is a glass sitting in the middle of the table with only
+Consider the following synchronization problem that we call
+\emph{Italian Philosophers} (a simplification of
+the famous Dining Philosophers problem described in Chapter~\ref{ch:deadlock}).
+There are five philosophers sitting around a table, each with a plate of
+spaghetti in front of them.
+There is a glass sitting in the middle of the table with only
 three forks in it.  A philosopher needs one of those forks to eat, but clearly,
 no more than three philosophers can eat at a time.
-See if you can come up with a deadlock-free solution using locks.  Represent
-each fork as a lock.  It's not easy.
-It is not hard to come up with a solution that uses a single lock
-and busy waiting, but we would like to avoid busy waiting.
+But a lock can only count to one.
 
 Introduced by Dijkstra,
-a \emph{semaphore} is a synchronization primitive that fits the bill~\cite{EWD35}.
+a \emph{semaphore} is a synchronization primitive that
+can solve this kind of problem~\cite{EWD35}.
 A semaphore is essentially
@@
 counter using the \texttt{P}
-\index{P}
-\index{procure}
+\index{P}%
+\index{procure}%
 operation.  You can think of \texttt{P} standing
@@
 \texttt{V}
-\index{V}
-\index{vacate}
+\index{V}%
+\index{vacate}%
 operation.  You can think of \texttt{V} standing for \emph{vacate},
@@
 \label{ch:bb}
-\index{bounded buffer}
-\index{producer/consumer problem}
+\index{bounded buffer}%
+\index{producer/consumer problem}%
 
@@
 so-called \emph{bounded buffer} (aka \emph{ring buffer}).
-\index{ring buffer}
+\index{ring buffer}%
 Suppose you have a \emph{producer} (some process) wanting to
@@
 a queue implemented using a circular buffer
-\index{circular buffer}
+\index{circular buffer}%
 of a certain length and two pointers:
@@
 in \emph{pipelines},
-\index{pipeline}
+\index{pipeline}%
 where each upstream process is a producer and each downstream
@@
 Or, in the client/server model,
-\index{client/server model}
+\index{client/server model}%
 some process may act as a \emph{server} that clients can send requests to
@@
 People often call semaphores that are used as locks \emph{binary semaphores},
-\index{binary semaphore}
+\index{binary semaphore}%
 as they only take on the values 0 and 1.
@@
 The second type of semaphores are called \emph{counting semaphores}.
-\index{counting semaphore}
+\index{counting semaphore}%
 They can be used to send signals between processes.
@@
 \label{ch:sbs}
-\index{split binary semaphore}
+\index{split binary semaphore}%
 
@@
 sometimes called \emph{conditional critical sections}
-\index{conditional critical section}
+\index{conditional critical section}%
 (CCSs)~\cite{Hoare73}.
@@
 \label{ch:starvation}
-\index{starvation}
+\index{starvation}%
 
@@
 A \emph{property}
-\index{property}
+\index{property}%
 is a set of traces.
@@
 We shall call such a solution \emph{fair}
-\index{fairness}
+\index{fairness}%
 (although it does not quite
@@
 \emph{monitors}~\cite{Hoare74}.
-\index{monitor}
+\index{monitor}%
 (A similar construct was independently invented by Per Brinch Hansen~\cite{BH73}.)
@@
 The other semaphores are called \emph{condition variables}.
-\index{condition variable}
+\index{condition variable}%
 There are two operations on condition variables: \texttt{wait}
-\index{wait}
+\index{wait}%
 and
 \texttt{signal}.
-\index{signal}
+\index{signal}%
 If \texttt{c} is a pointer to a condition variable, then the semantics of the
@@
 Mesa~\cite{LR80}.
-\index{Mesa}
+\index{Mesa}%
 Mesa introduced various important concepts to programming languages,
@@
 operation \texttt{notify} rather than \texttt{signal}.
-\index{notify}
+\index{notify}%
 When a process $p$ invokes \texttt{notify}, it does not immediately pass 
@@
 operation called \texttt{notifyAll}
-\index{notifyAll}
+\index{notifyAll}%
 (aka \texttt{broadcast)})
-\index{broadcast}
+\index{broadcast}%
 that, when applied to
@@
 \label{ch:deadlock}
-\index{deadlock}
+\index{deadlock}%
 
@@
 a scenario he called ``Dining Philosophers.''
-\index{dining philosopher}
+\index{dining philosopher}%
 
@@
 multiple conditions.
-\index{multiple conditions, waiting on}
+\index{multiple conditions, waiting on}%
 The condition in the \textbf{while} statement is the negation of the
@@
 \emph{avoidance}
-\index{deadlock avoidance}
+\index{deadlock avoidance}%
 instead of deadlock \emph{prevention}.
@@
 \label{ch:actor}
-\index{actor model}
-\index{message passing}
+\index{actor model}%
+\index{message passing}%
 
@@
 actors can be easily emulated using threads and \emph{synchronized queues}
-\index{synchronized queue}
+\index{synchronized queue}%
 (aka \emph{blocking queues})
-\index{blocking queue}
+\index{blocking queue}%
 for messaging.
@@
 \label{ch:abp}
-\index{alternating bit protocol}
-\index{protocol}
-\index{distributed system}
+\index{alternating bit protocol}%
+\index{protocol}%
+\index{distributed system}%
 
@@
 systems is that the former takes \emph{failures}
-\index{failure}
+\index{failure}%
 into account,
@@
 but the underlying network may lose messages.
-\index{network}
+\index{network}%
 The network does not re-order messages: when sending messages
@@
 \label{ch:nonblocking}
-\index{non-blocking synchronization}
+\index{non-blocking synchronization}%
 
@@
 non-blocking (aka \emph{wait-free})
-\index{wait-free synchronization}
+\index{wait-free synchronization}%
 synchronization algorithms~\cite{}.
@@
 The implementation uses the \texttt{alloc} module for dynamic allocation
-\index{dynamic allocation}
+\index{dynamic allocation}%
 of nodes in the list using \texttt{recAlloc} and \texttt{recFree}.
@@
 \emph{linearizable}~\cite{HW90},
-\index{linearizable}
+\index{linearizable}%
 a strong notion of consistency that
@@
 It does so by performing something called \emph{hand-over-hand locking}.
-\index{hand-over-hand locking}
+\index{hand-over-hand locking}%
 It first locks the first node, which is the \texttt{-inf} node.
@@
 \emph{linearization points}
-\index{linearization point}
+\index{linearization point}%
 of the operations in an execution.  These are the unique
@@
 \item A \emph{seqlock}
-\index{seqlock}
+\index{seqlock}%
 consists of a lock and a version number.
@@
 \label{ch:barrier}
-\index{barrier synchronization}
+\index{barrier synchronization}%
 
@@
 An \emph{lvalue}
-\index{lvalue}
+\index{lvalue}%
 (short for left hand value of an assignment statement)
@@
 The \texttt{alloc} module
-\index{alloc module}
+\index{alloc module}%
 support thread-safe dynamic allocation of ``records.''
@@
 The \texttt{bag} module
-\index{bag module}
+\index{bag module}%
 has various useful methods that operate on bags or
@@
 \noindent
-\index{list module}
+\index{list module}%
 The \texttt{list} module has various useful methods that operate on lists
@@
 A context is considered \emph{blocked}
-\index{blocked process}
+\index{blocked process}%
 if there does not exist a
@@
 A \emph{macro step}
-\index{macro step}
+\index{macro step}%
 can involve multiple micro steps.  The following

### 2020-06-28 cxl.py

@@ -2705,9 +2705,10 @@ def print_path(visited, bad_state):
             last = (node.ctx.nametag, node.ctx.pc, node.steps, vars)
     if last != None:
         print(nametag2str(last[0]), strsteps(last[2]), last[1], strVars(last[3]))
-    (node, vars) = path[-1]
-    if node.ctx.failure != 0:
-        print(">>>", node.ctx.failure)
+    if len(path) > 0:
+        (node, vars) = path[-1]
+        if node.ctx.failure != None:
+            print(">>>", node.ctx.failure)
 
 def print_shortest(visited, bad):
     bad_state = find_shortest(visited, bad)
@@ -3567,11 +3568,9 @@ def dump(visited, code, scope, state, fulldump, verbose):
                     for (var i = 1; i < tbl.rows.length; i++) {
                         if (i == row) {
                             tbl.rows[i].style.backgroundColor = "#A5FF33";
-                            tbl.rows[i].style.color = "#FFFFFF";
                         }
                         else {
                             tbl.rows[i].style.backgroundColor = "";
-                            tbl.rows[i].style.color = "";
                         }
                     }
                   }

### 2020-06-28 book/paper.tex
Cleaned up the way processes terminate

@@
     0 Jump 34
-    1 Frame incrementer(self) end=11
+    1 Frame incrementer(self)
 Up.cxl:2     count = count + 1;
@@
 in case you want to read about the details.
-The \texttt{Frame} instruction lists the name of the method,
-names of its arguments, and
-the program counter of the method's \texttt{Return} instruction.
+The \texttt{Frame} instruction lists the name of the method and the
+names of its arguments.
 The code generated from $\mathit{count} := \mathit{count} + 1$ in line 2 of
@@
 \hline
-Frame $m$($a$) end=$r$& start method $m$ with arguments $a$,
+Frame $m$($a$) & start method $m$ with arguments $a$,
 initializing variables.  Last instruction is at $r$ \\

### 2020-06-29 book/paper.tex
Latest

@@
 
-\glsadd{concurrent execution}
-\glsadd{shared variable}
-\glsadd{sequential execution}
-\glsadd{determinism}
-\glsadd{atomicity}
+\glsadd{concurrent execution}%
+\glsadd{shared variable}%
+\glsadd{sequential execution}%
+\glsadd{determinism}%
+\glsadd{atomicity}%
 
@@
 \item In a sequential program, each statement and each function can be
-thought of as happening \emph{atomically}
+thought of as happening \emph{atomically} (indivisibly)
 \index{atomicity}%
@@
 
-\glsadd{model checking}
+\glsadd{model checking}%
 
@@
 verifies that a program is correct for some \emph{model}.  Think of
-a model as a test program.  Because model checking tries every possible
-execution, this means that the test program needs to be
-simple or it will take longer than we may care to wait for or run
-out of memory.
+a model as a test program.
+Because model checking tries every possible execution, the test
+program needs to be simple. Otherwise it may take longer than we
+care to wait for or run out of memory.
 In particular, the model needs to have a relatively small number of
@@
 So if model checking does not prove a program correct, why is it
-useful?  To answer that question consider, for example, a sorting
-algorithm.
+useful?
+To answer that question, let us consider a sorting algorithm.
 Now suppose we create a test program, a model, that tries sorting
@@
 
-\glsadd{invariant}
+\glsadd{invariant}%
 
@@
 of Python programming.  We also will assume that the reader
-understand some basics of machine architecture and how programs
+understands some basics of machine architecture and how programs
 are executed.  For example, we assume that the reader is familiar
@@
 
-\begin{figure}
-\begin{code}
-\cxlsource{Up}
-\end{code}
-\caption{[Up.cxl] Incrementing twice in parallel.}
-\label{fig:inc}
-\end{figure}
-
-\glsadd{process}
+\glsadd{process}%
 
@@
 \texttt{incrementer(1)}, and the last runs \texttt{main()}.
+Note that although the processes are \emph{spawned} one at a time,
+they will execute concurrently.  It is, for example, quite possible
+that \texttt{incrementer(1)} finishes before \texttt{incrementer(0)}
+even gets going.
+
+\begin{figure}[h]
+\begin{code}
+\cxlsource{Up}
+\end{code}
+\caption{[Up.cxl] Incrementing twice in parallel.}
+\label{fig:inc}
+\end{figure}
 
@@
 program correct in that \textit{count} will always end up being 2?
-\item What are the possible values of \textit{count} at the time of the
-\textbf{assert} statement?
 (You may assume that \texttt{load} and \texttt{store} instructions of the
-underlying machine architecture are atomic---in fact they are.)
+underlying machine architecture are atomic (indivisible)---in fact they are.)
 \end{itemize}
@@
 
-\glsadd{machine instruction}
+\glsadd{machine instruction}%
 
@@
 
-\glsadd{race condition}
+\glsadd{race condition}%
 
@@
 
-\glsadd{context}
+\glsadd{context}%
 
@@
 
-\glsadd{stack machine}
+\glsadd{stack machine}%
 
@@
 Figure~\ref{fig:inchtml1}.
+If you hover the mouse over a machine instruction, it provides a
+brief explanation.
 
@@
 
-\glsadd{critical section}
-\glsadd{thread safety}
+\glsadd{critical section}%
+\glsadd{thread safety}%
 
@@
 
-\glsadd{mutual exclusion}
+\glsadd{mutual exclusion}%
 
@@
 
-\glsadd{lock}
+\glsadd{lock}%
 
@@
 
-\glsadd{state}
-\glsadd{process variable}
+\glsadd{state}%
+\glsadd{process variable}%
 
@@
 
-\glsadd{step}
-\glsadd{trace}
+\glsadd{step}%
+\glsadd{trace}%
 
@@
 
-\glsadd{spinlock}
+\glsadd{spinlock}%
 
@@
 
-\glsadd{interlock instruction}
+\glsadd{interlock instruction}%
 
@@
 
-\glsadd{blocked process}
+\glsadd{blocked process}%
 
@@
 
-\glsadd{reader/writer lock}
+\glsadd{reader/writer lock}%
 
@@
 
-\glsadd{busy waiting}
+\glsadd{busy waiting}%
 
@@
 
-\glsadd{semaphore}
+\glsadd{semaphore}%
 
@@
 
-\glsadd{producer/consumer problem}
+\glsadd{producer/consumer problem}%
 
@@
 
-\glsadd{conditional critical section}
+\glsadd{conditional critical section}%
 
@@
 
-\glsadd{starvation}
+\glsadd{starvation}%
 
@@
 
-\glsadd{property}
+\glsadd{property}%
 
@@
 
-\glsadd{fairness}
+\glsadd{fairness}%
 
@@
 
-\glsadd{monitor}
-\glsadd{condition variable}
+\glsadd{monitor}%
+\glsadd{condition variable}%
 
@@
 
-\glsadd{deadlock}
+\glsadd{deadlock}%
 
@@
 
-\glsadd{actor model}
+\glsadd{actor model}%
 
@@
 
-\glsadd{non-blocking synchronization}
+\glsadd{non-blocking synchronization}%
 
@@
 
-\glsadd{linearizable}
+\glsadd{linearizable}%
 
@@
 
-\glsadd{barrier synchronization}
+\glsadd{barrier synchronization}%
 

### 2020-06-29 cxl.py

@@ -376,6 +376,9 @@ class SplitOp(Op):
     def __repr__(self):
         return "Split"
 
+    def explain(self):
+        return "pops a set value and pushes the smallest element and the remaining set"
+
     def eval(self, state, context):
         v = context.pop()
         assert isinstance(v, SetValue)
@@ -401,6 +404,9 @@ class DupOp(Op):
 class GoOp(Op):
     def __repr__(self):
         return "Go"
+
+    def explain(self):
+        return "pops a context and a value, restores the corresponding process, and pushes the value on its stack"
 
     def eval(self, state, context):
         ctx = context.pop()

### 2020-06-29 book/code/pp.py
Better formatting

@@ -98,8 +98,7 @@ def isbinaryop(s):
         "-", "+", "*", "/", "%", "<", "<=", ">", ">="
     ];
 
-# tokens = [ "==", "!=", "<=", ">=", ".." ]
-tokens = []
+tokens = [ "==", "!=", "<=", ">=", ".." ]
 
 def putchar(c):
     print(c, end="")

### 2020-06-29 book/paper.tex
Latest
Better formatting

@@
 Concurrent Programming''~\cite{Schneider97}.
+(This chapter is named after that book.)
 
@@
 Method \texttt{main} waits until \textit{done} flags are set to \texttt{True}.
-After that, it verifies that the value of \textit{count} equals 2.  If not,
-it reports the actual value of \textit{count}.
+After that, it verifies that the value of \textit{count} equals 2.
+Like Python, CXL supports \texttt{assert} statements with two arguments:
+the first is a Boolean condition, and the second is a value that is
+reported if the condition fails, in this case
+the actual value of \textit{count}.
 The program spawns three processes.
@@
 \item Run Figure~\ref{fig:incpy} using Python.  Does the assertion fail?
-\item Using a script, run Figure~\ref{fig:incpy} 1000 times.  How many times does the assertion
+\item Using a script, run Figure~\ref{fig:incpy} 1000 times.
+For example, if you are using the bash shell, you can do the following:
+\begin{code}
+\begin{verbatim}
+for i in {1..1000}
+do
+    python Up.py
+done
+\end{verbatim}
+\end{code}
+How many times does the assertion
 fail (if any).
@@
 Method \texttt{tas} in Figure~\ref{fig:tas} shows its specification.
-Here $s$ points to a shared boolean variable and $p$
-to a private boolean variable, belonging to some process.
+Here $s$ points to a shared Boolean variable and $p$
+to a private Boolean variable, belonging to some process.
 The operation copies the value of the shared variable to the
@@
 \hline
-integers & \texttt{-inf, ..., -2, -1, 0, 1, 2, ..., inf} \\
+Integer & \texttt{-inf, ..., -2, -1, 0, 1, 2, ..., inf} \\
 \hline
@@
 \hline
-\texttt{e and e and ...} & two or more Booleans \\
-\hline
-\texttt{e or e or ...} & two or more Booleans \\
-\hline
-\texttt{not e} & a Boolean \\
+\texttt{e and e and ...} & two or more booleans \\
+\hline
+\texttt{e or e or ...} & two or more booleans \\
+\hline
+\texttt{not e} & a boolean \\
 \hline
@@
 \hline
-\texttt{e if e else e} & middle \texttt{e} must be a Boolean \\
+\texttt{e if e else e} & middle \texttt{e} must be a boolean \\
 \hline
@@
 \hline
-\texttt{assert b [, x]} & \texttt{b} is a Boolean.  If \texttt{False}, optionally print expression \texttt{e}\\
+\texttt{assert b [, x]} & \texttt{b} is a boolean.  If \texttt{False}, optionally print expression \texttt{e}\\
 \hline
@@
 \hline
-\texttt{if b: S} & \texttt{b} is a Boolean, \texttt{S} a list of statements\\
-\hline
-\texttt{while b: S} & \texttt{b} is a Boolean, \texttt{S} a list of statements\\
+\texttt{if b: S} & \texttt{b} is a boolean, \texttt{S} a list of statements\\
+\hline
+\texttt{while b: S} & \texttt{b} is a boolean, \texttt{S} a list of statements\\
 \hline
@@
 \hline
-stopped & a Boolean indicating if the context is stopped \\
+stopped & a boolean indicating if the context is stopped \\
 \hline
@@
 variable, as well as when an assertion fails;
-\item The execution of a process in atomic mode does not get interleaved
+\item The execution of a process in \emph{atomic mode} does not get interleaved
 with that of other processes.
@@
 \hline
-failure & Boolean indicating some context has failed \\
-\hline
-initializing & Boolean indicating that the state is not fully initialized. \\
+failure & boolean indicating some context has failed \\
+\hline
+initializing & boolean indicating that the state is not fully initialized. \\
 \hline

### 2020-06-29 cxl.py

@@ -562,46 +562,92 @@ class StoreOp(Op):
             context.pc += 1
 
 class DelOp(Op):
-    def __init__(self, n):
-        self.n = n                  # number of indexes
+    def __init__(self, name):
+        self.name = name
 
     def __repr__(self):
-        return "Del " + str(self.n)
+        if self.name != None:
+            (lexeme, file, line, column) = self.name
+            return "Del " + lexeme
+        else:
+            return "Del"
+
+    def explain(self):
+        if self.name == None:
+            return "pop an address and delete the shared variable at the address"
+        else:
+            return "delete the shared variable " + self.name[0]
 
     def eval(self, state, context):
-        indexes = []
-        for i in range(self.n):
-            indexes.append(context.pop())
-        av = indexes[0]
-        assert isinstance(av, AddressValue), indexes
-        state.delete(av.indexes + indexes[1:])
-        context.pc += 1
+        if self.name == None:
+            av = context.pop()
+            if not isinstance(av, AddressValue):
+                context.failure = "Error: not an address " + \
+                                    str(self.token) + " -> " + str(av)
+                return
+            lv = av.indexes
+            name = lv[0]
+        else:
+            (lexeme, file, line, column) = self.name
+            lv = [lexeme]
+            name = lexeme
+
+        if not state.initializing and (name not in state.vars.d):
+            context.failure = "Error: deleting an uninitialized shared variable " \
+                    + name + ": " + str(self.token)
+        else:
+            state.delete(lv)
+            context.pc += 1
 
 class StopOp(Op):
-    def __init__(self, n):
-        self.n = n                  # number of indexes
+    def __init__(self, name):
+        self.name = name
 
     def __repr__(self):
-        return "Stop " + str(self.n)
+        if self.name != None:
+            (lexeme, file, line, column) = self.name
+            return "Stop " + lexeme
+        else:
+            return "Stop"
+
+    def explain(self):
+        if self.name == None:
+            return "pop an address and store context at that address"
+        else:
+            return "store context at " + self.name[0]
 
     def eval(self, state, context):
-        indexes = []
-        for i in range(self.n):
-            indexes.append(context.pop())
-        av = indexes[0]
-        assert isinstance(av, AddressValue), indexes
[…]

### 2020-06-29 book/paper.tex
Fixed some things that were outdated

@@
 \hline
-Address $n$ & compute address \\
+Address $n$ & compute address from $n$ components \\
 \hline
@@
 \hline
-Del & delete a shared variable \\
-\hline
-DelVar $v$ & delete process variable $v$ \\
+Del [$v$] & delete shared variable $v$ \\
+\hline
+DelVar [$v$] & delete process variable $v$ \\
 \hline
@@
 \hline
-Stop & save context into shared variable and remove from context bag \\
+Stop [$v$] & save context into shared variable $v$ and remove from context bag \\
 \hline
@@
 shared and process variables can have addresses.
-\item \texttt{Load} and \texttt{LoadVar} instructions have an optional
+\item The \texttt{Load}, \texttt{LoadVar}, \texttt{Del}, \texttt{DelVar},
+and \texttt{Stop} instructions have an optional
 variable name: if omitted the top of the stack must contain the address of
@@
 \hline
-program counter & a Program Counter value pointing into the code \\
-\hline
-end & a Program Counter value indicating the end of the process code \\
+program counter & an integer value pointing into the code \\
+\hline
+frame pointer & an integer value pointing into the stack \\
 \hline
@@
 \texttt{spawn} allows the tag to be specified explicitly;
+\item The frame pointer points to the current \emph{stack frame},
+which consists of the caller's frame and variables, the argument to
+the method, and the return address.
 \item A process terminates when it reaches the \texttt{Return} instruction
@@
 if there does not exist a
-sequence instructions that the process can execute that either writes
-to a shared variable (even if that leaves the shared variable unchanged)
-or terminates the process.
+sequence instructions that the process can execute by itself
+that leads to its termination.
 

### 2020-06-29 book/paper.tex
Latest

@@
 \begin{verbatim}
-cxl -b -c NSLOTS=0 -c NPRODS=1 -c NCONSS=1 PCsema.cxl
-cxl -b -c NSLOTS=1 -c NPRODS=0 -c NCONSS=1 PCsema.cxl
-cxl    -c NSLOTS=1 -c NPRODS=1 -c NCONSS=0 PCsema.cxl
-cxl    -c NSLOTS=1 -c NPRODS=1 -c NCONSS=1 PCsema.cxl
-cxl -b -c NSLOTS=1 -c NPRODS=1 -c NCONSS=2 PCsema.cxl
-cxl -b -c NSLOTS=1 -c NPRODS=2 -c NCONSS=0 PCsema.cxl
-cxl    -c NSLOTS=1 -c NPRODS=2 -c NCONSS=1 PCsema.cxl
-cxl    -c NSLOTS=1 -c NPRODS=2 -c NCONSS=2 PCsema.cxl
-cxl -b -c NSLOTS=1 -c NPRODS=2 -c NCONSS=3 PCsema.cxl
-cxl    -c NSLOTS=2 -c NPRODS=1 -c NCONSS=0 PCsema.cxl
-cxl    -c NSLOTS=2 -c NPRODS=1 -c NCONSS=1 PCsema.cxl
-cxl -b -c NSLOTS=2 -c NPRODS=1 -c NCONSS=2 PCsema.cxl
-cxl    -c NSLOTS=2 -c NPRODS=2 -c NCONSS=0 PCsema.cxl
-cxl    -c NSLOTS=2 -c NPRODS=2 -c NCONSS=1 PCsema.cxl
-cxl    -c NSLOTS=2 -c NPRODS=2 -c NCONSS=2 PCsema.cxl
-cxl -b -c NSLOTS=2 -c NPRODS=2 -c NCONSS=3 PCsema.cxl
-cxl -b -c NSLOTS=2 -c NPRODS=3 -c NCONSS=0 PCsema.cxl
-cxl    -c NSLOTS=2 -c NPRODS=3 -c NCONSS=1 PCsema.cxl
-cxl    -c NSLOTS=2 -c NPRODS=3 -c NCONSS=2 PCsema.cxl
-cxl    -c NSLOTS=2 -c NPRODS=3 -c NCONSS=3 PCsema.cxl
+cxl -b -c NSLOTS=0 -c NPRODS=1 -c NCONSS=1 BBsematest.cxl
+cxl -b -c NSLOTS=1 -c NPRODS=0 -c NCONSS=1 BBsematest.cxl
+cxl    -c NSLOTS=1 -c NPRODS=1 -c NCONSS=0 BBsematest.cxl
+cxl    -c NSLOTS=1 -c NPRODS=1 -c NCONSS=1 BBsematest.cxl
+cxl -b -c NSLOTS=1 -c NPRODS=1 -c NCONSS=2 BBsematest.cxl
+cxl -b -c NSLOTS=1 -c NPRODS=2 -c NCONSS=0 BBsematest.cxl
+cxl    -c NSLOTS=1 -c NPRODS=2 -c NCONSS=1 BBsematest.cxl
+cxl    -c NSLOTS=1 -c NPRODS=2 -c NCONSS=2 BBsematest.cxl
+cxl -b -c NSLOTS=1 -c NPRODS=2 -c NCONSS=3 BBsematest.cxl
+cxl    -c NSLOTS=2 -c NPRODS=1 -c NCONSS=0 BBsematest.cxl
+cxl    -c NSLOTS=2 -c NPRODS=1 -c NCONSS=1 BBsematest.cxl
+cxl -b -c NSLOTS=2 -c NPRODS=1 -c NCONSS=2 BBsematest.cxl
+cxl    -c NSLOTS=2 -c NPRODS=2 -c NCONSS=0 BBsematest.cxl
+cxl    -c NSLOTS=2 -c NPRODS=2 -c NCONSS=1 BBsematest.cxl
+cxl    -c NSLOTS=2 -c NPRODS=2 -c NCONSS=2 BBsematest.cxl
+cxl -b -c NSLOTS=2 -c NPRODS=2 -c NCONSS=3 BBsematest.cxl
+cxl -b -c NSLOTS=2 -c NPRODS=3 -c NCONSS=0 BBsematest.cxl
+cxl    -c NSLOTS=2 -c NPRODS=3 -c NCONSS=1 BBsematest.cxl
+cxl    -c NSLOTS=2 -c NPRODS=3 -c NCONSS=2 BBsematest.cxl
+cxl    -c NSLOTS=2 -c NPRODS=3 -c NCONSS=3 BBsematest.cxl
 \end{verbatim}
@@
 \caption{Script for testing producer/consumer synchronization
-(assuming the code is in file \texttt{PCsema.cxl}).}
+(assuming the code is in file \texttt{BBsematest.cxl}).}
 \label{fig:pcscript}
@@
 right order.}
-\label{fig:PCsemadata}
+\label{fig:BBsemadata}
 \end{figure}
@@
 
-Figure~\ref{fig:PCsemadata} tests whether the correct data is sent in
+Figure~\ref{fig:BBsemadata} tests whether the correct data is sent in
 the correct order.  There are the same number of producers and consumers.
@@
 \hline
+failure & if not None, string that describes how the process failed \\
+\hline
 \end{tabular}
@@
 \hline
-failure & boolean indicating some context has failed \\
-\hline
 initializing & boolean indicating that the state is not fully initialized. \\

### 2020-06-29 cxl.py

@@ -3072,26 +3072,6 @@ def onestep(state, ctx, choice, visited, todo, node):
     if cc.failure != None:
         next.issues.add("process failure")
 
-def explore(s, visited, mapping, reach):
-    reach[s] = None         # prevent infinite loops
-    hs = mapping[s]
-    n = visited[s]
-    result = set()
-    assert not s.choosing
-    for (ctx, edge) in n.edges.items():
-        assert isinstance(ctx, ContextValue)
-        (nextState, nextContext, nextSteps) = edge
-        next = mapping[nextState]
-        if next == hs:
-            if nextState not in reach:
-                explore(nextState, visited, mapping, reach)
-            r = reach[nextState]
-            if r != None:
-                result = result.union(r)
-        else:
-            result.add(next)
-    reach[s] = result
-
 def parseConstant(c, v):
     tokens = lexer(v, "<constant argument>")
     try:
@@ -3104,7 +3084,7 @@ def parseConstant(c, v):
     code = []
     ast.compile(scope, code)
     state = State(code, scope.labels)
-    ctx = ContextValue(DictValue({"name": "__arg__", "tag": novalue}), 0, len(code))
+    ctx = ContextValue(DictValue({"name": "__arg__", "tag": novalue}), 0)
     ctx.atomic = 1
     while ctx.pc != len(code):
         code[ctx.pc].eval(state, ctx)
@@ -3140,6 +3120,27 @@ def doCompile(filenames, consts, mods):
     code.append(ReturnOp())     # to terminate "__init__" process
     optimize(code)
     return (code, scope)
+
+# No longer using this code
+def explore(s, visited, mapping, reach):
+    reach[s] = None         # prevent infinite loops
+    hs = mapping[s]
+    n = visited[s]
+    result = set()
+    assert not s.choosing
+    for (ctx, edge) in n.edges.items():
+        assert isinstance(ctx, ContextValue)
+        (nextState, nextContext, nextSteps) = edge
+        next = mapping[nextState]
+        if next == hs:
+            if nextState not in reach:
+                explore(nextState, visited, mapping, reach)
+            r = reach[nextState]
+            if r != None:
+                result = result.union(r)
+        else:
+            result.add(next)
+    reach[s] = result
 
 # No longer using this code
 def mapcheck():

### 2020-06-29 book/paper.tex
Changing definition of "blocked"

@@
   description={is a process that is waiting on a low-level synchronization primitive
-    and cannot make progress without the help of another process.
+    and cannot terminate without the help of another process.
       For example, a process that is waiting for a spinlock to become available}
@@
 in a state if the process is waiting on a low-level synchronization
-variable and cannot make progress without the help of another process.
+variable and cannot terminate without the help of another process.
 A process trying to

### 2020-06-29 book/paper.tex
Fixed formatting for Ch 13

@@
 \begin{center}
-\includegraphics[width=3in]{figures/pc-crop.pdf}
+\includegraphics[width=2.5in]{figures/pc-crop.pdf}
 \end{center}
@@
 Figure~\ref{fig:boundedbuffer} presents the implementation of a bounded
-buffer using semaphores.  It features a circular bounded buffer \texttt{buf} with
+buffer using semaphores.  It features a circular bounded buffer \textit{buf} with
 slots numbered 1 through \texttt{NSLOTS}.  There are two indexes into
-the buffer: \texttt{b\_in} specifies where the next produced item is inserted,
-while \texttt{b\_out} specifies from where the next can be consumed.
+the buffer: \textit{b\_in} specifies where the next produced item is inserted,
+while \textit{b\_out} specifies from where the next can be consumed.
 As there may be multiple producers and multiple
 consumers, updates to the indexes, which are shared variables, must be protected.
-To this end, we use two semaphores as locks: \texttt{l\_in} for \texttt{b\_in}
-and \texttt{l\_out} for \texttt{b\_out}.
+To this end, we use two semaphores as locks: \textit{l\_in} for \textit{b\_in}
+and \textit{l\_out} for \textit{b\_out}.
 
@@
 entries in the circular buffer.
-For this, there is a semaphore \texttt{n\_full} that
+For this, there is a semaphore \textit{n\_full} that
 keeps track of the number of filled entries in the buffer and a semaphore
-\texttt{n\_empty} that keeps track of the number of empty entries in the buffer.
-
-To add an item to the bounded buffer (\texttt{produce(item)}), the producing
+\textit{n\_empty} that keeps track of the number of empty entries in the buffer.
+
+To add an item to the bounded buffer (\texttt{produce}(\textit{item})), the producing
 process first has to wait until there is room in the bounded buffer.
-To this end, it invokes \texttt{P(n\_empty)}, which waits until
-\texttt{n\_empty > 0} and atomically decrements \texttt{n\_empty} once this
+To this end, it invokes \texttt{P}(\textit{n\_empty}), which waits until
+$\mathit{n\_empty} > 0$ and atomically decrements \textit{n\_empty} once this
 is the case.  In other words, the producer procures an empty slot.
@@
 Since there may be more than one empty slot, multiple producers may attempt
-to do the same thing concurrently.  Hence the use of the \texttt{l\_in}
+to do the same thing concurrently.  Hence the use of the \textit{l\_in}
 semaphore used as a lock.  Finally, once the item is added, the process
-increments the \texttt{n\_full} semaphore to signal to consumers that
+increments the \textit{n\_full} semaphore to signal to consumers that
 an item is available.
@@
 
-Note the quite different usage of the \texttt{l\_} semaphores and \texttt{n\_}
+Note the quite different usage of the \textit{l\_} semaphores and \textit{n\_}
 semaphores.
@@
 Moreover, consumers add the tuples they received to the set
-\texttt{received} (atomically---a lock could have been used here as well).)
+\textit{received} (atomically---a lock could have been used here as well).)
 The \texttt{main()} process waits for all consumers to have received
@@
 and return \texttt{True} if it succeeds.
-\item Implement a \emph{bounded stack}.  Method \texttt{push(item)}
+\item Implement a \emph{bounded stack}.  Method \texttt{push}(\textit{item})
 should block if the stack is full.  Method \texttt{pop()} should block
@@
 of an available GPU, blocking if there is currently no available GPU.
-Method \texttt{gpuRelease(gpu)} releases the given GPU.  It never needs
+Method \texttt{gpuRelease}(\textit{gpu}) releases the given GPU.  It never needs
 to block.

### 2020-06-29 book/paper.tex
Formatting up to monitors

@@
 are all~0.  To maintain the invariant, each process alternates \texttt{P} and
-\texttt{V} operations, starting with \texttt{P(\&mutex)} when it tries to
+\texttt{V} operations, starting with \texttt{P}(\&\textit{mutex}) when it tries to
 enter the CCS, and ending with a \texttt{V} operation on one of the semaphores
@@
 \begin{itemize}
-\item \texttt{r\_entered}: the number of readers in the critical section;
-\item \texttt{w\_entered}: the number of writers in the critical section;
-\item \texttt{r\_waiting}: the number of readers waiting to enter;
-\item \texttt{w\_waiting}: the number of writers waiting to enter.
+\item \textit{r\_entered}: the number of readers in the critical section;
+\item \textit{w\_entered}: the number of writers in the critical section;
+\item \textit{r\_waiting}: the number of readers waiting to enter;
+\item \textit{w\_waiting}: the number of writers waiting to enter.
 \end{itemize}
@@
 To wit, check out \texttt{rlock\_acquire}.  It first checks to see if there
-are any writers.  If so, it increments \texttt{r\_waiting}.  If not,
-it increments \texttt{r\_entered}.
+are any writers.  If so, it increments \textit{r\_waiting}.  If not,
+it increments \textit{r\_entered}.
 
@@
 found there is a writer in the critical section, it vacates
-\textit{mutex} and procures \texttt{r\_sema},
+\textit{mutex} and procures \textit{r\_sema},
 blocking on the semaphore associated
 with waiting readers.  When later some other process passes the baton to
-$p$ by incrementing \texttt{r\_sema}, $p$ updates the state by decrementing
-\texttt{r\_waiting} and incrementing \texttt{r\_entered}.  $p$
+$p$ by incrementing \textit{r\_sema}, $p$ updates the state by decrementing
+\textit{r\_waiting} and incrementing \textit{r\_entered}.  $p$
 then invokes \texttt{V\_one()} one more time.
@@
 the sum of the three semaphores is 0.  The writer then decrements
-\texttt{w\_entered} and calls \texttt{V\_one()}.
+\textit{w\_entered} and calls \texttt{V\_one()}.
 \item \texttt{V\_one()} finds that there are no writers in the critical section
-and there are two readers waiting.  It therefore vacates \texttt{r\_sema}.
-The sum of the semaphores is now 1, because \texttt{r\_sema} is 1.  This
-guarantees that only a waiting reader, procuring \texttt{r\_sema}, can enter
+and there are two readers waiting.  It therefore vacates \textit{r\_sema}.
+The sum of the semaphores is now 1, because \textit{r\_sema} is 1.  This
+guarantees that only a waiting reader, procuring \textit{r\_sema}, can enter
 the critical section.
-\item When it does, the reader decrements \texttt{r\_waiting}
-from 2 to 1, and increments \texttt{r\_entered} from 0 to 1.
+\item When it does, the reader decrements \textit{r\_waiting}
+from 2 to 1, and increments \textit{r\_entered} from 0 to 1.
 The reader finally calls \texttt{V\_one()}.
 \item Again, \texttt{V\_one()} finds that there are no writers and
-that there are readers waiting, so again it vacates \texttt{r\_sema},
+that there are readers waiting, so again it vacates \textit{r\_sema},
 passing the baton to the one remaining waiting reader.
-\item The remaining reader decrements \texttt{r\_waiting} from 1 to 0 and
-increments \texttt{r\_entered} from 1 to 2.
+\item The remaining reader decrements \textit{r\_waiting} from 1 to 0 and
+increments \textit{r\_entered} from 1 to 2.
 \item Finally, the remaining reader  calls \texttt{V\_one()}.
@@
 Alice and Bob each maintain a zero-initialized sequence number,
-\texttt{s\_seq} and \texttt{r\_seq} resp.
+\texttt{s\_seq} and \textit{r\_seq} resp.
 Alice sends a network message to Bob containing an
@@
 (we use \texttt{()} in the model).  Let \texttt{s\_chan} be the channel
-from Alice to Bob and \texttt{r\_chan} the channel from Bob to Alice.
+from Alice to Bob and \textit{r\_chan} the channel from Bob to Alice.
 \texttt{net\_send(pchan, m, reliable)} models sending a message \texttt{m} to

### 2020-06-29 book/paper.tex
Formatting up to Chapter 18

@@
 \index{signal}%
-If \texttt{c} is a pointer to a condition variable, then the semantics of the
+If $c$ is a pointer to a condition variable, then the semantics of the
 operations, in CXL, are as follows:
@@
 condition variables in the \texttt{synch} module.
-\texttt{Condition(lk)} creates a new condition variable associated with the given pointer
-to a lock \texttt{lk}.
+\texttt{Condition}(\textit{lk}) creates a new condition variable associated with the given pointer
+to a lock \textit{lk}.
 The condition variable itself is represented by a dictionary containing the pointer to the lock
@@
 
-In \texttt{release\_rlock}, notice that \texttt{notify(\&wcond)} is invoked when there
+In \texttt{release\_rlock}, notice that \texttt{notify}(\&\textit{wcond}) is invoked when there
 are no readers left, \emph{without} checking if there are writers waiting to enter.
@@
 
-\texttt{release\_wlock} executes \texttt{notifyAll(\&wcond)} as well as
-\texttt{notify(\&wcond)}.
+\texttt{release\_wlock} executes \texttt{notifyAll}(\&\textit{wcond}) as well as
+\texttt{notify}(\&\textit{wcond)}.
 Again, because we do not keep track of the number of waiting readers or writers, we

### 2020-06-29 book/paper.tex
Formatting up to Chapter 19

@@
 \item \texttt{Queue()} returns a new message queue;
-\item \texttt{enqueue(q, item)} adds \texttt{item} to the queue pointed to by \texttt{q};
-\item \texttt{dequeue(q)} waits for and returns an item on the queue pointed to by \texttt{q}.
+\item \texttt{enqueue}($q$, \textit{item}) adds \textit{item} to the queue pointed to by $q$;
+\item \texttt{dequeue}($q$) waits for and returns an item on the queue pointed to by $q$.
 \end{itemize}
@@
 \texttt{.produce} requests, and sometimes it is vice versa.
-The variable \texttt{balance} specifies what the situation is: if it is positive
+The variable \textit{balance} specifies what the situation is: if it is positive
 then it has buffered \texttt{.produce} requests;
@@
 then it has buffered \texttt{.consume} requests.
-Note that both \texttt{balance} and \texttt{requests} are variables that are
+Note that both \textit{balance} and \textit{requests} are variables that are
 local to the \texttt{pc\_actor} process.
@@
 
-The \texttt{pc\_actor} process has two arguments.  \texttt{q} is a pointer
-to the queue on which it receives messages.  \texttt{nrequests} is the total
+The \texttt{pc\_actor} process has two arguments.  $q$ is a pointer
+to the queue on which it receives messages.  \textit{nrequests} is the total
 number of requests that it will handle.  We need the latter because in CXL
@@
 Alice and Bob each maintain a zero-initialized sequence number,
-\texttt{s\_seq} and \textit{r\_seq} resp.
+\textit{s\_seq} and \textit{r\_seq} resp.
 Alice sends a network message to Bob containing an
@@
 channel as a variable that either contains a network message or nothing
-(we use \texttt{()} in the model).  Let \texttt{s\_chan} be the channel
+(we use \texttt{()} in the model).  Let \textit{s\_chan} be the channel
 from Alice to Bob and \textit{r\_chan} the channel from Bob to Alice.
-\texttt{net\_send(pchan, m, reliable)} models sending a message \texttt{m} to
-\texttt{\^{}pchan}, where \texttt{pchan} is either \texttt{\&s\_chan}
-or \texttt{\&r\_chan}.
-The method places either \texttt{m} (to model a successful send)
-or \texttt{()} (to model loss) in \texttt{\^{}pchan}.
-The use of the \texttt{reliable} flag will be explained later.
-\texttt{net\_recv(pchan)} models checking \texttt{\^{}pchan} for
+\texttt{net\_send}(\textit{pchan}, $m$, \textit{reliable}) models sending
+a message $m$ to
+\texttt{\^{}}\textit{pchan}, where \textit{pchan} is either \&\textit{s\_chan}
+or \&\textit{r\_chan}.
+The method places either $m$ (to model a successful send)
+or \texttt{()} (to model loss) in \texttt{\^{}}\textit{pchan}.
+The use of the \textit{reliable} flag will be explained later.
+\texttt{net\_recv}(\textit{pchan}) models checking \texttt{\^{}pchan} for
 the next message.
 
-Method \texttt{app\_send(m)} tries \emph{retransmitting} \texttt{m} until
+Method \texttt{app\_send}($m$) tries \emph{retransmitting} $m$ until
 an acknowledgment is received.
-Method \texttt{app\_recv(reliable)} returns the next successfully received
+Method \texttt{app\_recv}(\textit{reliable}) returns the next successfully received
 message.
@@
 
-Only the last invocation of \texttt{app\_recv(reliable)} is invoked with
-\texttt{reliable == True}.  It causes the last acknowledgment to be sent
+Only the last invocation of \texttt{app\_recv}(\textit{reliable}) is invoked with
+\textit{reliable} == \texttt{True}.  It causes the last acknowledgment to be sent
 reliably.  It allows the receiver (Bob) to stop, as well as the sender (Alice)

### 2020-06-29 book/paper.tex
Made a formatting pass over all chapters

@@
 is not necessarily practical.
-There are two variables.  \texttt{items} is an unbounded array with each
-entry initialized to \texttt{()}.  \texttt{back} is an index into the
+There are two variables.  \textit{items} is an unbounded array with each
+entry initialized to \texttt{()}.  \textit{back} is an index into the
 array and points to the next slot where a new value is inserted.
@@
 \begin{itemize}
-\item \texttt{inc(p)}: atomically increments \texttt{\^{}p} and returns
+\item \texttt{inc}($p$): atomically increments \texttt{\^{}}$p$ and returns
 the old value;
-\item \texttt{swap(p)}: sets \texttt{\^{}p} to \texttt{()} and returns
+\item \texttt{swap}($p$): sets \texttt{\^{}}$p$ to \texttt{()} and returns
 the old value.
@@
 
-Method \texttt{produce(item)} uses \texttt{inc(\&back)} to allocate
-the next available slot in the \texttt{items} array.
+Method \texttt{produce}(\textit{item}) uses \texttt{inc}(\&\textit{back})
+to allocate
+the next available slot in the \textit{items} array.
 It stores the item as a singleton tuple.
 Method \texttt{consume()} repeatedly scans the array, up to the
-\texttt{back} index, trying to find an item to return.
+\textit{back} index, trying to find an item to return.
 To check an entry, it uses \texttt{swap()}
@@
 \begin{itemize}
-\item The \texttt{items} array must be of infinite size if an unbounded number
+\item The \textit{items} array must be of infinite size if an unbounded number
 of items may be produced;
@@
 \texttt{inf} (like the Python \texttt{math.inf} constants).
-Each node has a lock, a value, and \texttt{next}, a pointer to the next node
+Each node has a lock, a value, and \textit{next}, a pointer to the next node
 (which is \texttt{None} for the final \texttt{inf} node).
-The \texttt{lst\_find(lst, v)} method is a helper function for update
+The \texttt{lst\_find}(\textit{lst}, $v$) method is a helper function for update
 methods that
-finds and locks two consecutive nodes \texttt{before} and \texttt{after}
-such that \texttt{before.value $<$ v $\le$ after.value}.
+finds and locks two consecutive nodes \textit{before} and \textit{after}
+such that $\mathit{before}.\mathtt{data.value} < v \le \mathit{after}.\mathtt{data.value}$.
 It does so by performing something called \emph{hand-over-hand locking}.
@@
 The \texttt{lst\_contains} method can simply scan through the list
-because the \texttt{next} pointers are updated atomically in such a
+because the \textit{next} pointers are updated atomically in such a
 way that they always point to a legal suffix of the list.
@@
 \texttt{lst\_remove} operations coincide exactly with the update
-of the \texttt{before.next} pointer.
+of the \textit{before}.\texttt{next} pointer.
 The linearization point of a \texttt{lst\_contains} method
@@
 Method \texttt{transfer} is an update operation; method \texttt{total} is a
-read-only operation.  Explain how a \texttt{seqlock} can lead to starvation.
+read-only operation.  Explain how a seqlock can lead to starvation.
 \end{problems}
@@
 
-The \texttt{round} array is kept to check the correctness of this
+The \textit{round} array is kept to check the correctness of this
 approach.  Each process increments its entry every time it enters
@@
 \begin{problems}
-\item See if you can implement barrier synchronization for $N$ processes
+\item See if you can implement barrier synchronization for \texttt{N} processes
 with just three semaphores.  Busy waiting is not allowed.  How about just two?

### 2020-06-29 book/paper.tex
Made formatting pass over appendices

@@
 Both tuples and lists map indexes (starting at 0) to CXL values.
-Their format is either \texttt{(e, e, ..., e,)} or
-\texttt{[e, e, ..., e,]}.
+Their format is either \texttt{($e$, $e$, ..., $e$,)} or
+\texttt{[$e$, $e$, ..., $e$,]}.
 If the tuple or list has two or more elements, then the final comma
@@
 Addresses are lexicographically ordered accordingly;
-\item contexts (Appendix~\ref{app:context}) are ordered by hash.
+\item contexts (Appendix~\ref{app:context}) are ordered first by name tag, then by program counter, then by hash.
 \end{itemize}
@@
 \hline
-\texttt{e == e}, \texttt{e != e} & two CXL values of the same type \\
-\hline
-\texttt{e < e, e <= e} & any two CXL values \\
-\hline
-\texttt{e > e, e >= e} & any two CXL values \\
-\hline
-\texttt{e and e and ...} & two or more booleans \\
-\hline
-\texttt{e or e or ...} & two or more booleans \\
-\hline
-\texttt{not e} & a boolean \\
-\hline
-\texttt{-e} & an integer \\
-\hline
-\texttt{e + e + ...} & two or more integers, sets, or lists \\
-\hline
-\texttt{e - e} & two integers or sets \\
-\hline
-\texttt{e * e * ...} & two or more integers or sets \\
-\hline
-\texttt{e / e} & two integers (integer division) \\
-\hline
-\texttt{e \% e} & two integers (division remainder) \\
-\hline
-\texttt{e~..~e} & two integers \\
-\hline
-\texttt{e [not] in e} & first is any CXL value, second is set \\
-\hline
-\texttt{e if e else e} & middle \texttt{e} must be a boolean \\
-\hline
-\texttt{\^{}e} & an address \\
-\hline
-\texttt{choose e} & a set \\
-\hline
-\texttt{min e, max e} & a set \\
-\hline
-\texttt{cardinality e} & a set \\
-\hline
-\texttt{keys e} & a dictionary \\
-\hline
-\texttt{len e} & a dictionary (includes tuple or list) \\
-\hline
-\texttt{bagsize e} & a bag \\
-\hline
-\texttt{atLabel e} & atom (corresponding to a label) \\
-\hline
-\texttt{nametag e} & \texttt{e} must be \texttt{()} \\
-\hline
-\texttt{processes e} & \texttt{e} must be \texttt{()} \\
-\hline
-\texttt{\&lv} & \texttt{lv} is an lvalue \\
-\hline
-\texttt{stop lv} & \texttt{lv} is an lvalue \\
+\texttt{$e$ == $e$}, \texttt{$e$ != $e$} & two CXL values of the same type \\
+\hline
+\texttt{$e$ < $e$, $e$ <= $e$} & any two CXL values \\
+\hline
+\texttt{$e$ > $e$, $e$ >= $e$} & any two CXL values \\
+\hline
+\texttt{$e$ and $e$ and ...} & two or more booleans \\
+\hline
+\texttt{$e$ or $e$ or ...} & two or more booleans \\
+\hline
+\texttt{not $e$} & a boolean \\
+\hline
+\texttt{-$e$} & an integer \\
+\hline
+\texttt{$e$ + $e$ + ...} & two or more integers, sets, or lists \\
+\hline
+\texttt{$e$ - $e$} & two integers or sets \\
+\hline
+\texttt{$e$ * $e$ * ...} & two or more integers or sets \\
+\hline
+\texttt{$e$ / $e$} & two integers (integer division) \\
+\hline
+\texttt{$e$ \% $e$} & two integers (division remainder) \\
+\hline
+\texttt{$e$~..~$e$} & two integers \\
+\hline
+\texttt{$e$ [not] in $e$} & first is any CXL value, second is set \\
+\hline
+\texttt{$e$ if $e$ else $e$} & middle $e$ must be a boolean \\
+\hline
+\texttt{\^{}$e$} & an address \\
+\hline
+\texttt{choose $e$} & a set \\
+\hline
+\texttt{min $e$, max $e$} & a set \\
+\hline
+\texttt{cardinality $e$} & a set \\
+\hline
+\texttt{keys $e$, len $e$} & a dictionary (includes tuple or list) \\
+\hline
+\texttt{bagsize $e$} & a bag \\
+\hline
+\texttt{atLabel $e$} & atom (corresponding to a label) \\
+\hline
+\texttt{nametag $e$} & $e$ must be \texttt{()} \\
+\hline
+\texttt{processes $e$} & $e$ must be \texttt{()} \\
+\hline
+\texttt{\&}\textit{lv} & \textit{lv} is an lvalue \\
+\hline
+\texttt{stop} \textit{lv} & \textit{lv} is an lvalue \\
 \hline
@@
 \texttt{*} computes the intersection when applied to sets.
-\texttt{x..y} computes the set consisting of the integers \texttt{x}
-through \texttt{y} inclusive.  It returns the empty set if \texttt{x > y}.
+\texttt{$x$..$y$} computes the set consisting of the integers $x$
+through $y$ inclusive.  It returns the empty set if $x > y$.
 
@@
 variable.  It can also be indexed.  Examples include
-\texttt{var}, \texttt{var[e]}, and \texttt{\^{}p}.
+\textit{var}, \textit{var}[$e$], and \texttt{\^{}}\textit{p}.
 
@@
 \hline
-Set comprehension & \texttt{\{ f(e) for e in s \}} \\
-\hline
-List comprehension & \texttt{[ f(e) for e in s ]} \\
-\hline
-Dict comprehension & \texttt{dict\{ f(e) for e in s \}} \\
+Set comprehension & \texttt{\{ f($e$) for $e$ in $s$ \}} \\
+\hline
+List comprehension & \texttt{[ f($e$) for $e$ in $s$ ]} \\
+\hline
+Dict comprehension & \texttt{dict\{ f($e$) for $e$ in $s$ \}} \\
 \hline
@@
 
-In each case, \texttt{s} must be a set.
-For list comprehensions, the list will be sorted by \texttt{e}.
-For dictionary comprehensions, each \texttt{e} will be mapped to \texttt{f(e)}.
+In each case, $s$ must be a set.
+For list comprehensions, the list will be sorted by $e$.
+For dictionary comprehensions, each $e$ will be mapped to \texttt{f}($e$).
 
@@
 \hline
-\texttt{lv = e} & \texttt{lv} is an lvalue and \texttt{e} is an expression\\
-\hline
-\texttt{lv [op]= e} & op is one of \texttt{+}, \texttt{-}, \texttt{*}, \texttt{/}, 
+\textit{lv} = $e$ & \textit{lv} is an lvalue and $e$ is an expression\\
+\hline
+\textit{lv} \texttt{[op]=} $e$ & \texttt{op} is one of \texttt{+}, \texttt{-}, \texttt{*}, \texttt{/}, 
 \texttt{and}, and \texttt{or}\\
@@
 \hline
-\texttt{del lv} & delete\\
-\hline
-\texttt{assert b [, x]} & \texttt{b} is a boolean.  If \texttt{False}, optionally print expression \texttt{e}\\
-\hline
-\texttt{const v = e} & \texttt{v} is an identifier, \texttt{e} is a constant expression\\
-\hline
-\texttt{def m(v, ...): S} & \texttt{m} and \texttt{v} are identifiers, \texttt{S} a list of statements\\
-\hline
-\texttt{let v = e, ...: S} & \texttt{e} is an expression, \texttt{S} a list of statements\\
-\hline
-\texttt{if b: S} & \texttt{b} is a boolean, \texttt{S} a list of statements\\
-\hline
-\texttt{while b: S} & \texttt{b} is a boolean, \texttt{S} a list of statements\\
-\hline
-\texttt{for v in e: S} & \texttt{e} is a set evaluated in order,
+\texttt{del} \textit{lv} & delete\\
+\hline
+\texttt{assert $b$ [, $x$]} & $b$ is a Boolean expression.  Optionally report value of expression $e$\\
+\hline
+\texttt{const v = $e$} & \texttt{v} is an identifier, $e$ is a constant expression\\
+\hline
+\texttt{def m($v$, ...): S} & \texttt{m} and \texttt{v} are identifiers, \texttt{S} a list of statements\\
+\hline
+\texttt{let $v$ = $e$, ...: S} & $v$ is an identifier, $e$ is an expression, \texttt{S} a list of statements\\
+\hline
+\texttt{if $b$: S} & $b$ is a Boolean expression, \texttt{S} a list of statements\\
+\hline
+\texttt{while $b$: S} & $b$ is a Boolean expression, \texttt{S} a list of statements\\
+\hline
+\texttt{for $v$ in $e$: S} & $v$ is an identifier, $e$ is a set evaluated in order,
                             \texttt{S} a list of statements\\
@@
 \hline
-\texttt{spawn p e [, t]} & \texttt{p} is a method, \texttt{e} is an expression, \texttt{t} is an optional tag (an expression) \\
-\hline
-\texttt{go c e} & \texttt{c} is a context, \texttt{e} is an expression \\
+\texttt{spawn m $e$ [, $t$]} & \texttt{m} is a method,
+$e$ is an expression, $t$ is a tag (an expression) \\
+\hline
+\texttt{go $c$ $e$} & $c$ is a context, $e$ is an expression \\
 \hline
@@
 To iterate through a dictionary, list, or tuple, use:
-\texttt{for v in keys(e):~S}.
+\texttt{for $v$ in keys($e$):~S}.
 
@@
 \label{ap:module}
+
+\section{The \texttt{alloc} module}
 
@@
 \hline
-\texttt{recFree(r)} & free record \texttt{r} \\
+\texttt{recFree($r$)} & free record $r$ \\
 \hline
@@
 \texttt{recAlloc} returns \texttt{None} when running out of memory.
+
+\section{The \texttt{bag} module}
 
@@
 \hline
-\texttt{bagFromSet(s)} & create a bag from set \texttt{s}\\
-\hline
-\texttt{bagCount(b, e)} & count how many times \texttt{e} occurs in bag \texttt{b}\\
-\hline
-\texttt{bagChoose(b)} & like \texttt{choose(s)}, but applied to a bag\\
-\hline
-\texttt{bagAdd(pb, e)} & add one copy of \texttt{e} to bag \texttt{\^{}pb}\\
-\hline
-\texttt{bagRemove(pb, e)} & remove one copy of \texttt{e} from bag \texttt{\^{}pb}\\
+\texttt{bagFromSet($s$)} & create a bag from set $s$\\
+\hline
+\texttt{bagCount($b$, $e$)} & count how many times $e$ occurs in bag $b$\\
+\hline
+\texttt{bagChoose($b$)} & like \texttt{choose(s)}, but applied to a bag\\
+\hline
+\texttt{bagAdd}(\textit{pb}, $e$) & add one copy of $e$ to bag \texttt{\^{}}\textit{pb}\\
+\hline
+\texttt{bagRemove}(\textit{pb}, $e$) & remove one copy of $e$ from bag \texttt{\^{}}\textit{pb}\\
 \hline
 \end{tabular}
-\vspace{1em}
-
-\noindent
+
+\section{The \texttt{list} module}
+
 \index{list module}%
@@
 \hline
-\texttt{subseq(t, b, f)} & return a \emph{slice} of list \texttt{t} starting
-at index \texttt{b} and ending just before index \texttt{f}\\
-\hline
-\texttt{append(t, e)} & append \texttt{e} to list \texttt{t}\\
-\hline
-\texttt{head(t)} & return the first element of list \texttt{t}\\
-\hline
-\texttt{tail(t)} & return all but the first element of list \texttt{t}\\
-\hline
-\texttt{listQsort(t)} & sort list \texttt{t} using quicksort\\
-\hline
-\texttt{list2bag(t)} & convert list \texttt{t} into a bag \\
-\hline
-\texttt{list2set(t)} & convert list \texttt{t} into a set \\
-\hline
-\texttt{listMin(t)} & return the minimum of all elements in \texttt{t}\\
-\hline
-\texttt{listMax(t)} & return the maximum of all elements in \texttt{t}\\
-\hline
-\texttt{listSum(t)} & return the sum of all elements in \texttt{t}\\
+\texttt{subseq($t$, $b$, $b$)} & return a \emph{slice} of list $t$ starting
+at index $b$ and ending just before index $f$\\
+\hline
+\texttt{append($t$, $e$)} & append $e$ to list $t$\\
+\hline
+\texttt{head($t$)} & return the first element of list $t$\\
+\hline
+\texttt{tail($t$)} & return all but the first element of list $t$\\
+\hline
+\texttt{listQsort($t$)} & sort list $t$ using quicksort\\
+\hline
+\texttt{list2bag($t$)} & convert list $t$ into a bag \\
+\hline
+\texttt{list2set($t$)} & convert list $t$ into a set \\
+\hline
+\texttt{listMin($t$)} & return the minimum of all elements in $t$\\
+\hline
+\texttt{listMax($t$)} & return the maximum of all elements in $t$\\
+\hline
+\texttt{listSum($t$)} & return the sum of all elements in $t$\\
+\hline
+\end{tabular}
+
+\section{The \texttt{synch} module}
+
+\index{synch module}%
+The \texttt{synch} module provides the following methods:
+
+\vspace{1em}
+\begin{tabular}{|l|l|}
+\hline
+tas(\textit{lk}) & test-and-set on \texttt{\^{}}\textit{lk} \\
+\hline
+Lock() & return a lock object \\
+\hline
+lock(\textit{lk}) & acquire \texttt{\^{}}\textit{lk} \\
+\hline
+unlock(\textit{lk}) & release \texttt{\^{}}\textit{lk} \\
+\hline
+Condition(\textit{lk}) & return a condition variable on \texttt{\^{}}\textit{lk} \\
+\hline
+wait($c$) & wait on condition variable \texttt{\^{}}$c$ \\
+\hline
+notify($c$) & notify a process waiting on condition variable \texttt{\^{}}$c$ \\
+\hline
+notifyAll($c$) & notify all processes waiting on condition variable \texttt{\^{}}$c$ \\
+\hline
+Semaphore(\textit{cnt}) & return a semaphore object initialized to \textit{cnt} \\
+\hline
+P(\textit{sema}) & procure \texttt{\^{}}\textit{sema}  \\
+\hline
+V(\textit{sema}) & vacate \texttt{\^{}}\textit{sema}  \\
+\hline
+Queue() & return a synchronized queue object \\
+\hline
+dequeue($q$) & return next element of $q$, blocking if empty \\
+\hline
+enqueue($q$, \textit{item}) & add \textit{item} to $a$ \\
 \hline
@@
 \hline
-Frame $m$($a$) & start method $m$ with arguments $a$,
-initializing variables.  Last instruction is at $r$ \\
+Frame \texttt{m}($a$) & start method \texttt{m} with arguments $a$,
+initializing variables.  \\
 \hline

### 2020-06-29 book/paper.tex
Added acknowledgments

@@
 
-\newpage
+\cleardoublepage
+\phantomsection
+\addcontentsline{toc}{chapter}{Acknowledgments}
+\chapter*{Acknowledgments}
+
+I received considerable help and inspiration from various people
+while writing this book.
+
+First and foremost I would like to think my student Haobin Ni with
+whom I've had numerous discussions about the design of CXL.
+Haobin even contributed some code to the CXL compiler.
+
+Most of what I know about concurrent programming I learned from
+my colleague Fred Schneider.  He suggested I write this book after
+demonstrating CXL to him.
+
+Leslie Lamport introduced me to using model checking to test properties
+of a concurrent system.  My experimentation with using TLC on Peterson's
+Algorithm became an aha momment for me.
+
+I first demonstrated CXL to the students in my CS6480 class on systems
+and formal verification and
+received valuable feedback from them.
+
+I would also like to thank my proofreaders Anneke van Renesse and
+Zhuoyu Xu.
+
+Finally, I would like to thank my family who had to suffer as I obsessed
+over writing the code and the book.
+
+% I wrote the CXL compiler and this book in May and June of 2020,
+% during the pandemic and the George Floyd demonstrations.
+
+\cleardoublepage
+\phantomsection
 \printindex
 
-\newpage
+\cleardoublepage
+\phantomsection
 \printglossaries

### 2020-06-29 book/paper.tex
Implemented hash()

@@
 There are various possible liveness properties we could use,
-but here we will propose the following informally: if there exists a non-empty
+but here we will propose the following informally: if
+(1) there exists a non-empty
 set $S$ of processes that are trying to enter the critical section and
-processes in the intersection always leave eventually, then eventually one
-process in $S$ will enter the critical section.
+(2) processes in the critical section always leave eventually, then
+eventually one process in $S$ will enter the critical section.
 We call this \emph{progress}.
@@
 Mutual exclusion is guaranteed but progress is violated.
-CXL should print a short trace
+CXL should print a trace
 to a state from which a terminating state cannot be reached.
@@
 \texttt{\^{}$e$} & an address \\
+\hline
+\texttt{hash $e$} & any CXL value \\
 \hline

### 2020-07-02 book/paper.tex
Latest
Improved layout

@@
 \caption{The HTML output of running CXL on Figure~\ref{fig:inc}.
-There are four sections.  The top left shows the source code and bytecode
-of the program.  The top right shows the steps in an execution.
-The middle section shows the values of the variables at a particular step.
-The bottom section shows the state of each process.
+There are three sections.
+The top shows the steps to the problematic state.
+The bottom left shows the source code and bytecode
+of the program.
+The bottom right shows the state of each process after the selected step.
 Each row in the stack trace of a process shows the current program counter,
@@
 \end{code}
-\caption{[csbarebones.cxl] A barebones critical section with two processes.}
+\caption{[csbarebones.cxl] A bare bones critical section with two processes.}
 \label{fig:csbarebones}
@@
 entering and exiting the critical section.
-Figure~\ref{fig:csbarebones} shows the CVM bytecode.
+Figure~\ref{fig:csbarebones} shows the CXL code.
 Here \texttt{@cs} is a \emph{label},

### 2020-07-02 cxl.py

@@ -2859,10 +2859,10 @@ def find_shortest(visited, bad):
 
 def print_path(visited, bad_state):
     path = genpath(bad_state, visited)
-    for (ctx, steps, states) in path:
+    for (ctx, steps, states, vars) in path:
         print(nametag2str(ctx.nametag), strsteps(steps))
     if len(path) > 0:
-        (ctx, steps, states) = path[-1]
+        (ctx, steps, states, vars) = path[-1]
         if ctx.failure != None:
             print(">>>", ctx.failure)
 
@@ -3388,17 +3388,19 @@ def genpath(s, visited):
     lastctx = None
     laststeps = []
     laststates = []
+    lastvars = DictValue({})
     for (s, n) in path:
         if lastctx == None or lastctx == n.before:
             laststeps += n.steps
             lastctx = n.after
             laststates.append(n.uid)
+            lastvars = s.vars
             continue
-        path2.append((lastctx, laststeps, laststates))
+        path2.append((lastctx, laststeps, laststates, lastvars))
         lastctx = n.after
         laststeps = n.steps.copy()
         laststates = []
-    path2.append((lastctx, laststeps, laststates))
+    path2.append((lastctx, laststeps, laststates, lastvars))
     return path2
 
 def htmlpath(s, visited, color, f):
@@ -3412,11 +3414,16 @@ def htmlpath(s, visited, color, f):
             label += ", "
         label += issue
     label = "Issue: " + label
+    keys = sorted(s.vars.d.keys(), key=lambda x: keyValue(x))
     path = genpath(s, visited)
-    print("<table id='issuestbl' border='1' width='100%%' style='color: %s'><tr><th colspan='3' align='center'>%s</th>"%(color, label), file=f)
+    print("<table id='issuestbl' border='1' width='100%%'><tr><th colspan='2' align='left' style='color: %s'>%s</th><th colspan='%d'>Shared Variables</th>"%(color, label, len(keys)), file=f)
+    print("<col style='width:15%'>", file=f)
     print("<tr><th>Process</th><th>Steps</th>", file=f)
+    for v in keys:
+        print("<th>%s</th>"%v, file=f)
+    print("</tr>", file=f)
     row = 1
-    for (ctx, steps, states) in path:
+    for (ctx, steps, states, vars) in path:
         row += 1
         if len(states) > 0:
             sid = states[-1]
@@ -3424,6 +3431,9 @@ def htmlpath(s, visited, color, f):
             sid = visited[s].uid
[…]

### 2020-07-02 book/paper.tex
Added another picture

@@
 \begin{figure}
+\includegraphics[width=\textwidth]{figures/naiveLock.png}
+\caption{The HTML output of running CXL on Figure~\ref{fig:uplock}.}
+\label{fig:naiveLockhtml}
+\end{figure}
+
+\begin{figure}
 \begin{code}
@@
 
-Unfortunately, if we run the program through CXL, we find that the assertion still
-fails.  Diagnosing the problem, we see that the \textit{lockTaken} variable
+Unfortunately, if we run the program through CXL,
+we find that the assertion still fails.
+Figure~\ref{fig:naiveLockhtml} shows the CXL output.
+Process~0 finds that the lock is available, but just before it stores
+\texttt{True} in \textit{lockTaken} in instruction~9, process~1 gets
+to run.  Because \textit{lockTaken} is still \texttt{False}, it too
+believes it can acquire the lock, and stores \texttt{True} in
+\texttt{lockTaken} and moves on to the critical section.
+Finally, process~0 moves on, also stores \texttt{True} and also moves
+into the critical section.  Process~0 is the one that detects the
+problem.
+The \textit{lockTaken} variable
 suffers from the same problem as the \textit{count} variable

### 2020-07-02 book/paper.tex
Latest

@@
 \begin{figure}
+\includegraphics[width=\textwidth]{figures/naiveFlags.png}
+\caption{The HTML output of running CXL on Figure~\ref{fig:upflags}.}
+\label{fig:naiveflagshtml}
+\end{figure}
+
+\begin{figure}
 \begin{code}
@@
 
-However, if you run the program through CXL, it turns out the solution
+However, if you run the program through CXL (Figure~\ref{fig:naiveflagshtml}),
+it turns out the solution
 does has a problem: if both try to enter the critical section at the same

### 2020-07-02 book/paper.tex
Latest

@@
 \texttt{True} in \textit{lockTaken} in instruction~9, process~1 gets
-to run.  Because \textit{lockTaken} is still \texttt{False}, it too
+to run.
+(Recall that you can hover your mouse over a machine instruction in order
+to see what it does.)
+Because \textit{lockTaken} is still \texttt{False}, it too
 believes it can acquire the lock, and stores \texttt{True} in

### 2020-07-03 book/paper.tex
Latest

@@
 process also tries to access the binary tree.
+Typically, you want some invariant property of the data structure to hold
+at the beginning and at the end of the critical section, but in the middle
+the invariant may be temporarily broken---this is no issue as no other
+process will be able to see it.
 An implementation of a data structure that can be safely accessed by multiple
@@
 \item Write out the invariants that need to hold and check them using CXL.
+\item People who use an ATM often first check their balance and then withdraw
+a certain amount of money not exceeding their balance.  A negative balance
+is not allowed.  Figure~\ref{fig:atm} shows two operations on bank accounts:
+one to check the balance and one to withdraw money.
+Note that all operations on accounts are carefully protected by a lock.
+The \texttt{customer}
+method models going to a particular ATM and withdrawing money not exceeding
+the balance.  Method \texttt{checker} looks for negative balances.
+Running the code through CXL reveals that there is a bug.
+It is a common type of concurrency bug known as \emph{Time Of Check Time Of
+Execution} (TOCTOE).
+\index{Time Of Check Time Of Execution}
+\index{TOCTOE}
+In this case, by the time the withdraw operation is performed,
+the balance can have changed.
+\begin{enumerate}
+\item Fix the code in Figure~\ref{fig:atm}.
+\item Is it necessary to obtain a lock in \texttt{atm\_check\_balance()}?
+\end{enumerate}
 \end{problems}
+
+\begin{figure}
+\begin{code}
+\cxlsource{atm}
+\end{code}
+\caption{[atm.cxl] Withdrawing money from an ATM.}
+\label{fig:atm}
+\end{figure}
 
@@
 Locks are probably the most prevalent and basic form of synchronization
-in concurrent programs.  For this reason, CXL has a module called
+in concurrent programs.  Typically, whenever you have a shared data
+structure, you want to protect the data structure with a lock and
+acquire the lock before access and release it immediately afterward.
+In other words, you want the access to the data structure to be a
+critical section.
+When there is a bug in a program because some code omitted obtaining
+a lock before accessing the data structure, that is known as a
+\emph{data race}.
+\index{data race}
+
+Because locks are so common, CXL has a module called
 \texttt{synch} that includes support for locks.

### 2020-07-03 book/code/pp.py

@@ -184,7 +184,11 @@ def lexer(s, file, silent):
                 found = t
         if found != "":
             result += [ (found, file, line, column) ]
-            if not silent: print(found, end="")
+            if not silent:
+                if found in [ "<=", ">=" ]:
+                    print("$%s$"%found, end="")
+                else:
+                    print(found, end="")
             s = s[len(found):]
             column += len(found)
             continue

### 2020-07-03 book/paper.tex
Added links to web site

@@
 \includegraphics[width=\textwidth]{figures/Up1.png}
-\caption{The HTML output of running CXL on Figure~\ref{fig:inc}.
+\caption{The
+\href{https://www.cs.cornell.edu/home/rvr/cxl/output/Up.html}{HTML output of running CXL on Figure~\ref{fig:inc}}.
 There are three sections.
@@
 \includegraphics[width=\textwidth]{figures/naiveLock.png}
-\caption{The HTML output of running CXL on Figure~\ref{fig:uplock}.}
+\caption{The \href{https://www.cs.cornell.edu/home/rvr/cxl/output/naiveLock.html}{HTML output of running CXL on Figure~\ref{fig:uplock}}.}
 \label{fig:naiveLockhtml}
@@
 \includegraphics[width=\textwidth]{figures/naiveFlags.png}
-\caption{The HTML output of running CXL on Figure~\ref{fig:upflags}.}
+\caption{The \href{https://www.cs.cornell.edu/home/rvr/cxl/output/naiveFlags.html}{HTML output of running CXL on Figure~\ref{fig:upflags}}.}
 \label{fig:naiveflagshtml}

### 2020-07-03 book/paper.tex
Added links to sources

@@
 \end{tabbing}
+}
+
+\newcommand{\cxllink}[1]{
+    [\href{https://www.cs.cornell.edu/home/rvr/cxl/code/#1}{#1}]
 }
@@
 \end{code}
-\caption{[triangle.cxl] Computing triangle numbers.}
+\caption{\cxllink{triangle.cxl} Computing triangle numbers.}
 \label{fig:triangle}
@@
 \end{code}
-\caption{[Up.cxl] Incrementing twice in parallel.}
+\caption{\cxllink{Up.cxl} Incrementing twice in parallel.}
 \label{fig:inc}
@@
 \end{code}
-\caption{[UpEnter.cxl] Broken attempt at fixing the code of Figure~\ref{fig:inc}.}
+\caption{\cxllink{UpEnter.cxl} Broken attempt at fixing the code of Figure~\ref{fig:inc}.}
 \label{fig:incenter}
@@
 \end{code}
-\caption{[csbarebones.cxl] A bare bones critical section with two processes.}
+\caption{\cxllink{csbarebones.cxl} A bare bones critical section with two processes.}
 \label{fig:csbarebones}
@@
 \end{code}
-\caption{[cx.cxl] CXL model of a critical section.}
+\caption{\cxllink{cx.cxl} CXL model of a critical section.}
 \label{fig:cs}
@@
 \end{code}
-\caption{[naiveLock.cxl] Na\"{\i}ve implementation of a shared lock.}
+\caption{\cxllink{naiveLock.cxl} Na\"{\i}ve implementation of a shared lock.}
 \label{fig:uplock}
@@
 \end{code}
-\caption{[naiveFlags.cxl] Na\"{\i}ve use of flags to solve mutual exclusion.}
+\caption{\cxllink{naiveFlags.cxl} Na\"{\i}ve use of flags to solve mutual exclusion.}
 \label{fig:upflags}
@@
 \end{code}
-\caption{[naiveTurn.cxl] Na\"{\i}ve use of turn variable to solve mutual exclusion.}
+\caption{\cxllink{naiveTurn.cxl} Na\"{\i}ve use of turn variable to solve mutual exclusion.}
 \label{fig:upturn}
@@
 \end{code}
-\caption{[Peterson.cxl] Peterson's Algorithm}
+\caption{\cxllink{Peterson.cxl} Peterson's Algorithm}
 \label{fig:peterson}
@@
 \end{code}
-\caption{[PetersonInductive.cxl] Peterson's Algorithm with Inductive Invariant}
+\caption{\cxllink{PetersonInductive.cxl} Peterson's Algorithm with Inductive Invariant}
 \label{fig:petersonproof}
@@
 \end{code}
-\caption{[csonebit.cxl] Mutual exclusion using a flag per process.}
+\caption{\cxllink{csonebit.cxl} Mutual exclusion using a flag per process.}
 \label{fig:csonebit}
@@
 \end{code}
-\caption{[PetersonMethod.cxl] Peterson's Algorithm accessed through methods.}
+\caption{\cxllink{PetersonMethod.cxl} Peterson's Algorithm accessed through methods.}
 \label{fig:petersonmethods}
@@
 \end{code}
-\caption{[spinlock.cxl] Mutual Exclusion using a ``spinlock'' based on test-and-set.}
+\caption{\cxllink{spinlock.cxl} Mutual Exclusion using a ``spinlock'' based on test-and-set.}
 \label{fig:tas}
@@
 \end{code}
-\caption{[spinlockInv.cxl] Checking invariants.}
+\caption{\cxllink{spinlockInv.cxl} Checking invariants.}
 \label{fig:tasinv}
@@
 \end{code}
-\caption{[atm.cxl] Withdrawing money from an ATM.}
+\caption{\cxllink{atm.cxl} Withdrawing money from an ATM.}
 \label{fig:atm}
@@
 \end{code}
-\caption{[csTAS.cxl] Fixed version of Figure~\ref{fig:uplock} using test-and-set.}
+\caption{\cxllink{csTAS.cxl} Fixed version of Figure~\ref{fig:uplock} using test-and-set.}
 \label{fig:tas2}
@@
 \end{code}
-\caption{[UpLock.cxl] Program of Figure~\ref{fig:inc} fixed with a lock.}
+\caption{\cxllink{UpLock.cxl} Program of Figure~\ref{fig:inc} fixed with a lock.}
 \label{fig:incfixed}
@@
 \end{code}
-\caption{[RW.cxl] Busy-Waiting Reader/Writer Lock implementation.}
+\caption{\cxllink{RW.cxl} Busy-Waiting Reader/Writer Lock implementation.}
 \label{fig:rwbusy}
@@
 \end{code}
-\caption{[RWtest.cxl] Test code for Figure~\ref{fig:rwbusy}.}
+\caption{\cxllink{RWtest.cxl} Test code for Figure~\ref{fig:rwbusy}.}
 \label{fig:rwbusytest}
@@
 \end{code}
-\caption{[RWbusychk.cxl] Checking for Busy Waiting.}
+\caption{\cxllink{RWbusychk.cxl} Checking for Busy Waiting.}
 \label{fig:rwblock}
@@
 \end{code}
-\caption{[RWlock.cxl] Reader/Writer with Two Locks.}
+\caption{\cxllink{RWlock.cxl} Reader/Writer with Two Locks.}
 \label{fig:rw2lock}
@@
 \end{code}
-\caption{[RWmulti.cxl] Checking that multiple readers can acquire the read lock.}
+\caption{\cxllink{RWmulti.cxl} Checking that multiple readers can acquire the read lock.}
 \label{fig:rwmulti}
@@
 \end{code}
-\caption{[DinersSema.cxl] Italian Philosophers.}
+\caption{\cxllink{DinersSema.cxl} Italian Philosophers.}
 \label{fig:dinerssema}
@@
 \end{code}
-\caption{[qsort.cxl] Iterative qsort() implementation.}
+\caption{\cxllink{qsort.cxl} Iterative qsort() implementation.}
 \label{fig:qsort}
@@
 \end{code}
-\caption{[qsorttest.cxl] Test program for Figure~\ref{fig:qsort}.}
+\caption{\cxllink{qsorttest.cxl} Test program for Figure~\ref{fig:qsort}.}
 \label{fig:qsorttest}
@@
 \end{code}
-\caption{[BBsema.cxl] Bounded Buffer implementation using semaphores.}
+\caption{\cxllink{BBsema.cxl} Bounded Buffer implementation using semaphores.}
 \label{fig:boundedbuffer}
@@
 \end{code}
-\caption{[BBsematest.cxl] Test program for Figure~\ref{fig:boundedbuffer}.}
+\caption{\cxllink{BBsematest.cxl} Test program for Figure~\ref{fig:boundedbuffer}.}
 \label{fig:bbtest}
@@
 \end{code}
-\caption{[BBsemadata.cxl] Test whether the bounded buffer delivers the right data in the
+\caption{\cxllink{BBsemadata.cxl} Test whether the bounded buffer delivers the right data in the
 right order.}
@@
 \end{code}
-\caption{[gpu.cxl] A thread-unsafe GPU allocator.}
+\caption{\cxllink{gpu.cxl} A thread-unsafe GPU allocator.}
 \label{fig:gpu}
@@
 \end{code}
-\caption{[RWsbs.cxl] Reader/Writer Lock implementation using Split Binary Semaphores.}
+\caption{\cxllink{RWsbs.cxl} Reader/Writer Lock implementation using Split Binary Semaphores.}
 \label{fig:RWsplitsema}
@@
 \end{code}
-\caption{[RWfair.cxl] Reader/Writer Lock SBS implementation addressing fairness.}
+\caption{\cxllink{RWfair.cxl} Reader/Writer Lock SBS implementation addressing fairness.}
 \label{fig:RWfair}
@@
 \end{code}
-\caption{[RWqueue.cxl] Reader/Writer Lock SBS implementation using a queue.}
+\caption{\cxllink{RWqueue.cxl} Reader/Writer Lock SBS implementation using a queue.}
 \label{fig:RWqueue}
@@
 \end{code}
-\caption{[RWqtest.cxl] Test program for Figure~\ref{fig:RWqueue}.}
+\caption{\cxllink{RWqtest.cxl} Test program for Figure~\ref{fig:RWqueue}.}
 \label{fig:RWqtest}
@@
 \end{code}
-\caption{[RWcv.cxl] Reader/Writer Lock implementation using Mesa-style condition variables.}
+\caption{\cxllink{RWcv.cxl} Reader/Writer Lock implementation using Mesa-style condition variables.}
 \label{fig:RWcv}
@@
 \end{code}
-\caption{[Diners.cxl] Dining Philosophers.}
+\caption{\cxllink{Diners.cxl} Dining Philosophers.}
 \label{fig:diners}
@@
 \end{code}
-\caption{[DinersCV.cxl] Dining Philosophers that grab both forks at the same time.}
+\caption{\cxllink{DinersCV.cxl} Dining Philosophers that grab both forks at the same time.}
 \label{fig:dinerscv}
@@
 \end{code}
-\caption{[DinersAvoid.cxl] Dining Philosophers that carefully avoid getting into a deadlock
+\caption{\cxllink{DinersAvoid.cxl} Dining Philosophers that carefully avoid getting into a deadlock
 scenario.}
@@
 \end{code}
-\caption{[bank.cxl] Bank accounts.}
+\caption{\cxllink{bank.cxl} Bank accounts.}
 \label{fig:bank}
@@
 \end{code}
-\caption{[actor.cxl] A producer/consumer actor.}
+\caption{\cxllink{actor.cxl} A producer/consumer actor.}
 \label{fig:actor}
@@
 \end{code}
-\caption{[actortest.cxl] Test code for producer/consumer actor.}
+\caption{\cxllink{actortest.cxl} Test code for producer/consumer actor.}
 \label{fig:actortest}
@@
 \end{code}
-\caption{[abp.cxl] Alternating Bit Protocol.}
+\caption{\cxllink{abp.cxl} Alternating Bit Protocol.}
 \label{fig:abp}
@@
 \end{code}
-\caption{[abptest.cxl] Test code for alternating bit protocol.}
+\caption{\cxllink{abptest.cxl} Test code for alternating bit protocol.}
 \label{fig:abptest}
@@
 \end{code}
-\caption{[hw.cxl] Non-blocking queue.}
+\caption{\cxllink{hw.cxl} Non-blocking queue.}
 \label{fig:hw}
@@
 \end{code}
-\caption{[lst1.cxl] List with non-blocking read operations, part 1.}
+\caption{\cxllink{lst1.cxl} List with non-blocking read operations, part 1.}
 \label{fig:lst1}
@@
 \end{code}
-\caption{[lst2.cxl] List with non-blocking read operations, part 2.}
+\caption{\cxllink{lst2.cxl} List with non-blocking read operations, part 2.}
 \label{fig:lst2}
@@
 \end{code}
-\caption{[barrier.cxl] Barrier synchronization with semaphores.}
+\caption{\cxllink{barrier.cxl} Barrier synchronization with semaphores.}
 \label{fig:barrier}

### 2020-07-03 book/paper.tex
Latest

@@
 \usepackage{imakeidx}\makeindex
-\usepackage{hyperref}
 \usepackage[nottoc,notlot,notlof]{tocbibind}
@@
 \usepackage{bold-extra}
+\usepackage{xcolor}
+\usepackage{hyperref}
+\hypersetup{linkbordercolor=blue, pdfborderstyle={/S/U/W 1}}
 
@@
 \newcommand{\cxllink}[1]{
-    [\href{https://www.cs.cornell.edu/home/rvr/cxl/code/#1}{#1}]
+    [\href[pdfnewwindow=true]{https://www.cs.cornell.edu/home/rvr/cxl/code/#1}{#1}]
 }
@@
 \caption{The
-\href{https://www.cs.cornell.edu/home/rvr/cxl/output/Up.html}{HTML output of running CXL on Figure~\ref{fig:inc}}.
+\href[pdfnewwindow=true]{https://www.cs.cornell.edu/home/rvr/cxl/output/Up.html}{HTML output of running CXL on Figure~\ref{fig:inc}}.
 There are three sections.
@@
 \includegraphics[width=\textwidth]{figures/naiveLock.png}
-\caption{The \href{https://www.cs.cornell.edu/home/rvr/cxl/output/naiveLock.html}{HTML output of running CXL on Figure~\ref{fig:uplock}}.}
+\caption{The \href[pdfnewwindow=true]{https://www.cs.cornell.edu/home/rvr/cxl/output/naiveLock.html}{HTML output of running CXL on Figure~\ref{fig:uplock}}.}
 \label{fig:naiveLockhtml}
@@
 \includegraphics[width=\textwidth]{figures/naiveFlags.png}
-\caption{The \href{https://www.cs.cornell.edu/home/rvr/cxl/output/naiveFlags.html}{HTML output of running CXL on Figure~\ref{fig:upflags}}.}
+\caption{The \href[pdfnewwindow=true]{https://www.cs.cornell.edu/home/rvr/cxl/output/naiveFlags.html}{HTML output of running CXL on Figure~\ref{fig:upflags}}.}
 \label{fig:naiveflagshtml}

### 2020-07-03 book/paper.tex
Latest

--- 2020-07-03 book/paper.tex
@@
 
-\newcommand{\cxllink}[1]{
-    [\href{https://www.cs.cornell.edu/home/rvr/cxl/code/#1}{#1}]
+\newcommand{\cxllink}[1]{%
+[\href{https://www.cs.cornell.edu/home/rvr/cxl/code/#1}{\underline{#1}}]%
+}
+
+\newcommand{\cxlref}[2]{%
+\href{https://www.cs.cornell.edu/home/rvr/cxl/output/#1}{\underline{#2}}%
 }
@@
 \caption{The
-\href{https://www.cs.cornell.edu/home/rvr/cxl/output/Up.html}{HTML output of running CXL on Figure~\ref{fig:inc}}.
+\cxlref{Up.html}{HTML output of running CXL on Figure~\ref{fig:inc}}.
 There are three sections.

### 2020-07-03 book/paper.tex
Latest

@@
 \includegraphics[width=\textwidth]{figures/naiveLock.png}
-\caption{The \href{https://www.cs.cornell.edu/home/rvr/cxl/output/naiveLock.html}{HTML output of running CXL on Figure~\ref{fig:uplock}}.}
+\caption{The \cxlref{naiveLock.html}{HTML output of running CXL on Figure~\ref{fig:uplock}}.}
 \label{fig:naiveLockhtml}
@@
 \includegraphics[width=\textwidth]{figures/naiveFlags.png}
-\caption{The \href{https://www.cs.cornell.edu/home/rvr/cxl/output/naiveFlags.html}{HTML output of running CXL on Figure~\ref{fig:upflags}}.}
+\caption{The \cxlref{naiveFlags.html}{HTML output of running CXL on Figure~\ref{fig:upflags}}.}
 \label{fig:naiveflagshtml}

### 2020-07-03 cxl.py
Improved formatting

@@ -3641,6 +3641,7 @@ def htmlnode(s, visited, code, scope, f, verbose):
         print("<th>Steps</th><th>Next State</th></tr>", file=f)
     else:
         print("</tr>", file=f)
+        print("<tr><td></td><td></td><td></td><td></td></tr>", file=f)
     for ctx in sorted(s.ctxbag.keys(), key=lambda x: nametag2str(x.nametag)):
         htmlrow(ctx, s.ctxbag, s, n, code, scope, f, verbose)
     for ctx in sorted(s.stopbag.keys(), key=lambda x: nametag2str(x.nametag)):

### 2020-07-03 book/paper.tex
Improved formatting
Latest

@@
 XX\=XXX\=XXX\kill
-    \input{code/#1.tex}
+    \input{sources/#1.tex}
 \end{tabbing}
@@
 \newcommand{\cxllink}[1]{%
-[\href{https://www.cs.cornell.edu/home/rvr/cxl/code/#1}{\underline{#1}}]%
+[\href{https://www.cs.cornell.edu/home/rvr/cxl/sources/#1}{\underline{#1}}]%
 }
@@
 \begin{code}
-\VerbatimInput[xleftmargin=5mm,numbers=left]{code/Up.py}
+\VerbatimInput[xleftmargin=5mm,numbers=left]{../code/Up.py}
 \end{code}
@@
 \begin{code}
-\VerbatimInput[xleftmargin=5mm,numbers=left]{code/UpMany.py}
+\VerbatimInput[xleftmargin=5mm,numbers=left]{../code/UpMany.py}
 \end{code}
@@
 one to check the balance and one to withdraw money.
-Note that all operations on accounts are carefully protected by a lock.
+Note that all operations on accounts are carefully protected by a lock
+(i.e., there are no data races).
 The \texttt{customer}

### 2020-07-04 book/paper.tex

@@
 \newcommand{\cxllink}[1]{%
-[\href{https://www.cs.cornell.edu/home/rvr/cxl/sources/#1}{\underline{#1}}]%
+[\href{https://www.cs.cornell.edu/home/rvr/cxl/code/#1}{\underline{#1}}]%
 }
@@
 \end{code}
-\caption{[Up.py] Python implementation of Figure~\ref{fig:inc}.}
+\caption{\cxllink{Up.py} Python implementation of Figure~\ref{fig:inc}.}
 \label{fig:incpy}
@@
 \end{code}
-\caption{[UpMany.py] Incrementing \texttt{N} times.}
+\caption{\cxllink{UpMany.py} Incrementing \texttt{N} times.}
 \label{fig:incmany}

### 2020-07-04 book/paper.tex
Moved python source files for better web experience

@@
 \end{code}
-\caption{\cxllink{triangle.cxl} Computing triangle numbers.}
+\caption{\cxllink{code/triangle.cxl} Computing triangle numbers.}
 \label{fig:triangle}
@@
 \end{code}
-\caption{\cxllink{Up.cxl} Incrementing twice in parallel.}
+\caption{\cxllink{code/Up.cxl} Incrementing twice in parallel.}
 \label{fig:inc}
@@
 \begin{code}
-\VerbatimInput[xleftmargin=5mm,numbers=left]{../code/Up.py}
-\end{code}
-\caption{\cxllink{Up.py} Python implementation of Figure~\ref{fig:inc}.}
+\VerbatimInput[xleftmargin=5mm,numbers=left]{../python/Up.py}
+\end{code}
+\caption{\cxllink{python/Up.py} Python implementation of Figure~\ref{fig:inc}.}
 \label{fig:incpy}
@@
 \begin{code}
-\VerbatimInput[xleftmargin=5mm,numbers=left]{../code/UpMany.py}
-\end{code}
-\caption{\cxllink{UpMany.py} Incrementing \texttt{N} times.}
+\VerbatimInput[xleftmargin=5mm,numbers=left]{../python/UpMany.py}
+\end{code}
+\caption{\cxllink{python/UpMany.py} Incrementing \texttt{N} times.}
 \label{fig:incmany}
@@
 \end{code}
-\caption{\cxllink{UpEnter.cxl} Broken attempt at fixing the code of Figure~\ref{fig:inc}.}
+\caption{\cxllink{code/UpEnter.cxl} Broken attempt at fixing the code of Figure~\ref{fig:inc}.}
 \label{fig:incenter}
@@
 \end{code}
-\caption{\cxllink{csbarebones.cxl} A bare bones critical section with two processes.}
+\caption{\cxllink{code/csbarebones.cxl} A bare bones critical section with two processes.}
 \label{fig:csbarebones}
@@
 \end{code}
-\caption{\cxllink{cx.cxl} CXL model of a critical section.}
+\caption{\cxllink{code/cx.cxl} CXL model of a critical section.}
 \label{fig:cs}
@@
 \end{code}
-\caption{\cxllink{naiveLock.cxl} Na\"{\i}ve implementation of a shared lock.}
+\caption{\cxllink{code/naiveLock.cxl} Na\"{\i}ve implementation of a shared lock.}
 \label{fig:uplock}
@@
 \end{code}
-\caption{\cxllink{naiveFlags.cxl} Na\"{\i}ve use of flags to solve mutual exclusion.}
+\caption{\cxllink{code/naiveFlags.cxl} Na\"{\i}ve use of flags to solve mutual exclusion.}
 \label{fig:upflags}
@@
 \end{code}
-\caption{\cxllink{naiveTurn.cxl} Na\"{\i}ve use of turn variable to solve mutual exclusion.}
+\caption{\cxllink{code/naiveTurn.cxl} Na\"{\i}ve use of turn variable to solve mutual exclusion.}
 \label{fig:upturn}
@@
 \end{code}
-\caption{\cxllink{Peterson.cxl} Peterson's Algorithm}
+\caption{\cxllink{code/Peterson.cxl} Peterson's Algorithm}
 \label{fig:peterson}
@@
 \end{code}
-\caption{\cxllink{PetersonInductive.cxl} Peterson's Algorithm with Inductive Invariant}
+\caption{\cxllink{code/PetersonInductive.cxl} Peterson's Algorithm with Inductive Invariant}
 \label{fig:petersonproof}
@@
 \end{code}
-\caption{\cxllink{csonebit.cxl} Mutual exclusion using a flag per process.}
+\caption{\cxllink{code/csonebit.cxl} Mutual exclusion using a flag per process.}
 \label{fig:csonebit}
@@
 \end{code}
-\caption{\cxllink{PetersonMethod.cxl} Peterson's Algorithm accessed through methods.}
+\caption{\cxllink{code/PetersonMethod.cxl} Peterson's Algorithm accessed through methods.}
 \label{fig:petersonmethods}
@@
 \end{code}
-\caption{\cxllink{spinlock.cxl} Mutual Exclusion using a ``spinlock'' based on test-and-set.}
+\caption{\cxllink{code/spinlock.cxl} Mutual Exclusion using a ``spinlock'' based on test-and-set.}
 \label{fig:tas}
@@
 \end{code}
-\caption{\cxllink{spinlockInv.cxl} Checking invariants.}
+\caption{\cxllink{code/spinlockInv.cxl} Checking invariants.}
 \label{fig:tasinv}
@@
 \end{code}
-\caption{\cxllink{atm.cxl} Withdrawing money from an ATM.}
+\caption{\cxllink{code/atm.cxl} Withdrawing money from an ATM.}
 \label{fig:atm}
@@
 \end{code}
-\caption{\cxllink{csTAS.cxl} Fixed version of Figure~\ref{fig:uplock} using test-and-set.}
+\caption{\cxllink{code/csTAS.cxl} Fixed version of Figure~\ref{fig:uplock} using test-and-set.}
 \label{fig:tas2}
@@
 \end{code}
-\caption{\cxllink{UpLock.cxl} Program of Figure~\ref{fig:inc} fixed with a lock.}
+\caption{\cxllink{code/UpLock.cxl} Program of Figure~\ref{fig:inc} fixed with a lock.}
 \label{fig:incfixed}
@@
 \end{code}
-\caption{\cxllink{RW.cxl} Busy-Waiting Reader/Writer Lock implementation.}
+\caption{\cxllink{code/RW.cxl} Busy-Waiting Reader/Writer Lock implementation.}
 \label{fig:rwbusy}
@@
 \end{code}
-\caption{\cxllink{RWtest.cxl} Test code for Figure~\ref{fig:rwbusy}.}
+\caption{\cxllink{code/RWtest.cxl} Test code for Figure~\ref{fig:rwbusy}.}
 \label{fig:rwbusytest}
@@
 \end{code}
-\caption{\cxllink{RWbusychk.cxl} Checking for Busy Waiting.}
+\caption{\cxllink{code/RWbusychk.cxl} Checking for Busy Waiting.}
 \label{fig:rwblock}
@@
 \end{code}
-\caption{\cxllink{RWlock.cxl} Reader/Writer with Two Locks.}
+\caption{\cxllink{code/RWlock.cxl} Reader/Writer with Two Locks.}
 \label{fig:rw2lock}
@@
 \end{code}
-\caption{\cxllink{RWmulti.cxl} Checking that multiple readers can acquire the read lock.}
+\caption{\cxllink{code/RWmulti.cxl} Checking that multiple readers can acquire the read lock.}
 \label{fig:rwmulti}
@@
 \end{code}
-\caption{\cxllink{DinersSema.cxl} Italian Philosophers.}
+\caption{\cxllink{code/DinersSema.cxl} Italian Philosophers.}
 \label{fig:dinerssema}
@@
 \end{code}
-\caption{\cxllink{qsort.cxl} Iterative qsort() implementation.}
+\caption{\cxllink{code/qsort.cxl} Iterative qsort() implementation.}
 \label{fig:qsort}
@@
 \end{code}
-\caption{\cxllink{qsorttest.cxl} Test program for Figure~\ref{fig:qsort}.}
+\caption{\cxllink{code/qsorttest.cxl} Test program for Figure~\ref{fig:qsort}.}
 \label{fig:qsorttest}
@@
 \end{code}
-\caption{\cxllink{BBsema.cxl} Bounded Buffer implementation using semaphores.}
+\caption{\cxllink{code/BBsema.cxl} Bounded Buffer implementation using semaphores.}
 \label{fig:boundedbuffer}
@@
 \end{code}
-\caption{\cxllink{BBsematest.cxl} Test program for Figure~\ref{fig:boundedbuffer}.}
+\caption{\cxllink{code/BBsematest.cxl} Test program for Figure~\ref{fig:boundedbuffer}.}
 \label{fig:bbtest}
@@
 \end{code}
-\caption{\cxllink{BBsemadata.cxl} Test whether the bounded buffer delivers the right data in the
+\caption{\cxllink{code/BBsemadata.cxl} Test whether the bounded buffer delivers the right data in the
 right order.}
@@
 \end{code}
-\caption{\cxllink{gpu.cxl} A thread-unsafe GPU allocator.}
+\caption{\cxllink{code/gpu.cxl} A thread-unsafe GPU allocator.}
 \label{fig:gpu}
@@
 \end{code}
-\caption{\cxllink{RWsbs.cxl} Reader/Writer Lock implementation using Split Binary Semaphores.}
+\caption{\cxllink{code/RWsbs.cxl} Reader/Writer Lock implementation using Split Binary Semaphores.}
 \label{fig:RWsplitsema}
@@
 \end{code}
-\caption{\cxllink{RWfair.cxl} Reader/Writer Lock SBS implementation addressing fairness.}
+\caption{\cxllink{code/RWfair.cxl} Reader/Writer Lock SBS implementation addressing fairness.}
 \label{fig:RWfair}
@@
 \end{code}
-\caption{\cxllink{RWqueue.cxl} Reader/Writer Lock SBS implementation using a queue.}
+\caption{\cxllink{code/RWqueue.cxl} Reader/Writer Lock SBS implementation using a queue.}
 \label{fig:RWqueue}
@@
 \end{code}
-\caption{\cxllink{RWqtest.cxl} Test program for Figure~\ref{fig:RWqueue}.}
+\caption{\cxllink{code/RWqtest.cxl} Test program for Figure~\ref{fig:RWqueue}.}
 \label{fig:RWqtest}
@@
 \end{code}
-\caption{\cxllink{RWcv.cxl} Reader/Writer Lock implementation using Mesa-style condition variables.}
+\caption{\cxllink{code/RWcv.cxl} Reader/Writer Lock implementation using Mesa-style condition variables.}
 \label{fig:RWcv}
@@
 \end{code}
-\caption{\cxllink{Diners.cxl} Dining Philosophers.}
+\caption{\cxllink{code/Diners.cxl} Dining Philosophers.}
 \label{fig:diners}
@@
 \end{code}
-\caption{\cxllink{DinersCV.cxl} Dining Philosophers that grab both forks at the same time.}
+\caption{\cxllink{code/DinersCV.cxl} Dining Philosophers that grab both forks at the same time.}
 \label{fig:dinerscv}
@@
 \end{code}
-\caption{\cxllink{DinersAvoid.cxl} Dining Philosophers that carefully avoid getting into a deadlock
+\caption{\cxllink{code/DinersAvoid.cxl} Dining Philosophers that carefully avoid getting into a deadlock
 scenario.}
@@
 \end{code}
-\caption{\cxllink{bank.cxl} Bank accounts.}
+\caption{\cxllink{code/bank.cxl} Bank accounts.}
 \label{fig:bank}
@@
 \end{code}
-\caption{\cxllink{actor.cxl} A producer/consumer actor.}
+\caption{\cxllink{code/actor.cxl} A producer/consumer actor.}
 \label{fig:actor}
@@
 \end{code}
-\caption{\cxllink{actortest.cxl} Test code for producer/consumer actor.}
+\caption{\cxllink{code/actortest.cxl} Test code for producer/consumer actor.}
 \label{fig:actortest}
@@
 \end{code}
-\caption{\cxllink{abp.cxl} Alternating Bit Protocol.}
+\caption{\cxllink{code/abp.cxl} Alternating Bit Protocol.}
 \label{fig:abp}
@@
 \end{code}
-\caption{\cxllink{abptest.cxl} Test code for alternating bit protocol.}
+\caption{\cxllink{code/abptest.cxl} Test code for alternating bit protocol.}
 \label{fig:abptest}
@@
 \end{code}
-\caption{\cxllink{hw.cxl} Non-blocking queue.}
+\caption{\cxllink{code/hw.cxl} Non-blocking queue.}
 \label{fig:hw}
@@
 \end{code}
-\caption{\cxllink{lst1.cxl} List with non-blocking read operations, part 1.}
+\caption{\cxllink{code/lst1.cxl} List with non-blocking read operations, part 1.}
 \label{fig:lst1}
@@
 \end{code}
-\caption{\cxllink{lst2.cxl} List with non-blocking read operations, part 2.}
+\caption{\cxllink{code/lst2.cxl} List with non-blocking read operations, part 2.}
 \label{fig:lst2}
@@
 \end{code}
-\caption{\cxllink{barrier.cxl} Barrier synchronization with semaphores.}
+\caption{\cxllink{code/barrier.cxl} Barrier synchronization with semaphores.}
 \label{fig:barrier}

### 2020-07-04 book/paper.tex
Latest

@@
 \newcommand{\cxllink}[1]{%
-[\href{https://www.cs.cornell.edu/home/rvr/cxl/code/#1}{\underline{#1}}]%
+[\href{https://www.cs.cornell.edu/home/rvr/cxl/#1}{\underline{#1}}]%
 }
@@
 \end{code}
-\caption{\cxllink{code/cx.cxl} CXL model of a critical section.}
+\caption{\cxllink{code/csbarebones.cxl} CXL model of a critical section.}
 \label{fig:cs}

### 2020-07-05 book/paper.tex
Made cxl more self-reliant

@@
 \usepackage{xcolor}
-\usepackage{hyperref}
-\hypersetup{linkbordercolor=blue, pdfborderstyle={/S/U/W 1}}
+\usepackage[colorlinks=true, linkcolor=blue, urlcolor=blue]{hyperref}
+\def\itemautorefname{Exercise}%
+\def\chapterautorefname{Chapter}%
+\def\figureautorefname{Figure}%
 
@@
 \begin{document}
-
 \maketitle
+
+Permission is granted to copy, distribute and/or modify this
+document under the terms of the
+Creative Commons AttributionNonCommercial-ShareAlike 4.0 International
+(CC BY-NC-SA 4.0) at
+\url{http://creativecommons.org/licenses/by-nc-sa/4.0}.
+
 \tableofcontents
@@
 \begin{itemize}
-\item Chapter~\ref{ch:cxlintro} introduces the CXL programming
+\item \autoref{ch:cxlintro} introduces the CXL programming
 language, as it provides the language for presenting synchronization
 problems and solutions.
-\item Chapter~\ref{ch:concurrent} illustrates the problem of
+\item \autoref{ch:concurrent} illustrates the problem of
 concurrent programming through a simple example in which two processes
 are concurrently incrementing a counter.
-\item Chapter~\ref{ch:cxlmachine} presents the
+\item \autoref{ch:cxlmachine} presents the
 CXL virtual machine to understand the problem
 underlying concurrency better.
-\item Chapter~\ref{ch:critical} introduces the concept of a
+\item \autoref{ch:critical} introduces the concept of a
 \emph{critical section} and presents various flawed implementations
@@
 is not trivial.
-\item Chapter~\ref{ch:peterson} introduces Peterson's Algorithm, an
+\item \autoref{ch:peterson} introduces Peterson's Algorithm, an
 elegant solution to implementating a critical section.
-\item Chapter~\ref{ch:method} gives some more details on the CXL
+\item \autoref{ch:method} gives some more details on the CXL
 language.
-\item Chapter~\ref{ch:spinlock} introduces ``hardware'' locks
+\item \autoref{ch:spinlock} introduces ``hardware'' locks
 for implemented critical sections.
-\item Chapter~\ref{ch:synch} presents CXL modules
+\item \autoref{ch:synch} presents CXL modules
 that supports locks and other synchronization primitives.
-\item Chapters~\ref{ch:rdwrbusy} and~\ref{ch:rdwr2} present the
+\item \autoref{ch:rdwrbusy} and \autoref{ch:rdwr2} present the
 abstraction and implementation of reader/writer
@@
 all only executing read-only accesses.
-Chapter~\ref{ch:rdwrbusy} also talks about busy-waiting, which is
+\autoref{ch:rdwrbusy} also talks about busy-waiting, which is
 an undesirable approach to synchronize processes.
-\item Chapter~\ref{ch:semaphore} introduces \emph{semaphores},
+\item \autoref{ch:semaphore} introduces \emph{semaphores},
 a generalization of locks
@@
 certain application-level conditions.
-\item Chapters~\ref{ch:bb} and~\ref{ch:sbs} introduce the
+\item \autoref{ch:bb} and \autoref{ch:sbs} introduce the
 \emph{bounded buffer problem} (aka \emph{producer/consumer problem})
 and various solutions.
-\item Chapter~\ref{ch:starvation} talks about \emph{starvation}:
+\item \autoref{ch:starvation} talks about \emph{starvation}:
 the problem that in some
@@
 resource they need.
-\item Chapter~\ref{ch:monitors} presents
+\item \autoref{ch:monitors} presents
 \emph{monitors} and \emph{condition variables},
 another approach to process synchronication.
-\item Chapter~\ref{ch:deadlock} describes \emph{deadlock}
+\item \autoref{ch:deadlock} describes \emph{deadlock}
 where a set of processes are indefinitely waiting for one another to
 release a resource.
-\item Chapter~\ref{ch:actor} presents the \emph{actor model}
+\item \autoref{ch:actor} presents the \emph{actor model}
 and \emph{message passing} as an approach to synchronization.
-\item Chapter~\ref{ch:abp} presents a problem and a solution to the distributed
+\item \autoref{ch:abp} presents a problem and a solution to the distributed
 systems problem of having two processes communicate reliably over an unreliable
 network.
-\item Chapter~\ref{ch:nonblocking} introduces \emph{non-blocking} or
+\item \autoref{ch:nonblocking} introduces \emph{non-blocking} or
 \emph{wait-free} synchronization algorithms,
@@
 steps.
-\item Chapter~\ref{ch:barrier} describes \emph{barrier synchronization},
+\item \autoref{ch:barrier} describes \emph{barrier synchronization},
 useful in high-performance computing applications such as parallel simulations.
@@
 
-Figure~\ref{fig:triangle} shows a simple example of a CXL program.
+\autoref{fig:triangle} shows a simple example of a CXL program.
 (The code for examples in this book can be found in the \texttt{code} folder under
@@
 \item See what happens if, instead of initializing \textit{result} to 0,
-you initialize it to 1.  (You do not need to understand the error report at this time.  They will be explained in more detail in Chapter~\ref{ch:cxlmachine}.)
+you initialize it to 1.  (You do not need to understand the error report at this time.  They will be explained in more detail in \autoref{ch:cxlmachine}.)
 \item Write a CXL program that computes squares by repeated adding.  So the program
@@
 running in parallel while sharing variables.
-Figure~\ref{fig:inc} presents a simple example.
+\autoref{fig:inc} presents a simple example.
 The program
@@
 and it is the machine instructions that are executed by the underlying CXL
-machine.  The details of this appear in Chapter~\ref{ch:cxlmachine},
+machine.  The details of this appear in \autoref{ch:cxlmachine},
 but suffice it to
@@
 \item CXL programs can usually be easily translated into Python.  For example,
-Figure~\ref{fig:incpy} is a Python version of Figure~\ref{fig:inc}.
+\autoref{fig:incpy} is a Python version of \autoref{fig:inc}.
 \begin{enumerate}
-\item Run Figure~\ref{fig:incpy} using Python.  Does the assertion fail?
-\item Using a script, run Figure~\ref{fig:incpy} 1000 times.
+\item Run \autoref{fig:incpy} using Python.  Does the assertion fail?
+\item Using a script, run \autoref{fig:incpy} 1000 times.
 For example, if you are using the bash shell, you can do the following:
@@
 \end{enumerate}
-\item Figure~\ref{fig:incmany} is a version of Figure~\ref{fig:incpy} that has each
-incrementer thread increment \textit{count} \texttt{N} times.  Run Figure~\ref{fig:incpy}
+\item \autoref{fig:incmany} is a version of \autoref{fig:incpy} that has each
+incrementer thread increment \textit{count} \texttt{N} times.  Run \autoref{fig:incpy}
 10 times (using Python).
@@
 How large does \texttt{N} need to be for assertions to fail?
-\item Can you think of a fix to Figure~\ref{fig:inc}?  Try one or two different fixes
+\item Can you think of a fix to \autoref{fig:inc}?  Try one or two different fixes
 and run them through CXL.  Do not worry about having to come up with a correct fix at this
@@
 \end{code}
-\caption{\cxllink{python/Up.py} Python implementation of Figure~\ref{fig:inc}.}
+\caption{\cxllink{python/Up.py} Python implementation of \autoref{fig:inc}.}
 \label{fig:incpy}
@@
 \texttt{name/tag}.
-For example, in Figure~\ref{fig:inc}, the created processes have name tags
+For example, in \autoref{fig:inc}, the created processes have name tags
 \texttt{incrementer/0}, \texttt{incrementer/1}, and \texttt{main/()}.
@@
 Each machine instruction updates the state in some way.
-% Figure~\ref{fig:incstate} shows an example of a reachable state for the program in Figure~\ref{fig:inc}.
+% \autoref{fig:incstate} shows an example of a reachable state for the program in \autoref{fig:inc}.
 
@@
 \end{code}
-\caption{The first part of the CVM bytecode corresponding to Figure~\ref{fig:inc}.}
+\caption{The first part of the CVM bytecode corresponding to \autoref{fig:inc}.}
 \label{fig:inccode}
@@
 \end{code}
-\caption{The text output of running CXL on Figure~\ref{fig:inc}.}
+\caption{The text output of running CXL on \autoref{fig:inc}.}
 \label{fig:incoutput}
@@
 
-Compiling the code in Figure~\ref{fig:inc} results in the CVM bytecode
-listed in Figure~\ref{fig:inccode}.
+Compiling the code in \autoref{fig:inc} results in the CVM bytecode
+listed in \autoref{fig:inccode}.
 You can obtain this code by invoking \texttt{cxl} with the \texttt{-a} flag
@@
 The code generated from $\mathit{count} := \mathit{count} + 1$ in line 2 of
-\texttt{Up.cxl} is as follows (see Figure~\ref{fig:inccode}):
+\texttt{Up.cxl} is as follows (see \autoref{fig:inccode}):
 
@@
 instructions.
-Figure~\ref{fig:incoutput} shows the output produced by running CXL on the
+\autoref{fig:incoutput} shows the output produced by running CXL on the
 \texttt{Up.cxl} program.
@@
 \caption{The
-\cxlref{Up.html}{HTML output of running CXL on Figure~\ref{fig:inc}}.
+\cxlref{Up.html}{HTML output of running CXL on \autoref{fig:inc}}.
 There are three sections.
@@
 these are states in which some processes are left that cannot
-terminate.  Chapter~\ref{ch:synch} will explain what it means for
+terminate.  \autoref{ch:synch} will explain what it means for
 a process to be stopped.
@@
 Open the suggested HTML file and you should see something like
-Figure~\ref{fig:inchtml1}.
+\autoref{fig:inchtml1}.
 If you hover the mouse over a machine instruction, it provides a
@@
 \begin{problems}
-\item Figure~\ref{fig:incenter} shows an attempt at trying to fix the code of
-    Figure~\ref{fig:inc}.  Run it through CXL and see what happens.  Based on
+\item \autoref{fig:incenter} shows an attempt at trying to fix the code of
+    \autoref{fig:inc}.  Run it through CXL and see what happens.  Based on
     the error output, describe in English what is wrong with the code by describing,
     in broad steps, how running the program can get into a bad state.
-\item What if we moved line~2 of Figure~\ref{fig:incenter} to after the \textbf{if}
+\item What if we moved line~2 of \autoref{fig:incenter} to after the \textbf{if}
     statement (between lines~7 and~8)?  Do you think that would work?  Run it through
@@
 \end{code}
-\caption{\cxllink{code/UpEnter.cxl} Broken attempt at fixing the code of Figure~\ref{fig:inc}.}
+\caption{\cxllink{code/UpEnter.cxl} Broken attempt at fixing the code of \autoref{fig:inc}.}
 \label{fig:incenter}
@@
 entering and exiting the critical section.
-Figure~\ref{fig:csbarebones} shows the CXL code.
+\autoref{fig:csbarebones} shows the CXL code.
 Here \texttt{@cs} is a \emph{label},
@@
 able to reach a state in which all processes have terminated.
-Clearly, even if mutual exclusion holds in Figure~\ref{fig:csbarebones},
+Clearly, even if mutual exclusion holds in \autoref{fig:csbarebones},
 the spawned processes never terminate.  We
 will model processes in critical sections using the framework in
-Figure~\ref{fig:cs}: a process can \emph{choose} to enter a
+\autoref{fig:cs}: a process can \emph{choose} to enter a
 critical section more than once, but it can also choose to terminate, even
@@
 
-Figure~\ref{fig:mutex} shows a high-level state diagram
+\autoref{fig:mutex} shows a high-level state diagram
 \index{state diagram} of mutual exclusion and progress.%
@@
 \includegraphics[width=\textwidth]{figures/naiveLock.png}
-\caption{The \cxlref{naiveLock.html}{HTML output of running CXL on Figure~\ref{fig:uplock}}.}
+\caption{The \cxlref{naiveLock.html}{HTML output of running CXL on \autoref{fig:uplock}}.}
 \label{fig:naiveLockhtml}
@@
 \includegraphics[width=\textwidth]{figures/naiveFlags.png}
-\caption{The \cxlref{naiveFlags.html}{HTML output of running CXL on Figure~\ref{fig:upflags}}.}
+\caption{The \cxlref{naiveFlags.html}{HTML output of running CXL on \autoref{fig:upflags}}.}
 \label{fig:naiveflagshtml}
@@
 Using a lock is a good thought, but how does one implement one?
-Figure~\ref{fig:uplock} presents a mutual exclusion attempt based on a
+\autoref{fig:uplock} presents a mutual exclusion attempt based on a
 na\"{\i}ve (and, as it turns out, broken) implementation of a lock.
@@
 we find that the assertion still fails.
-Figure~\ref{fig:naiveLockhtml} shows the CXL output.
+\autoref{fig:naiveLockhtml} shows the CXL output.
 Process~0 finds that the lock is available, but just before it stores
@@
 suffers from the same problem as the \textit{count} variable
-in Figure~\ref{fig:inc}: operations
+in \autoref{fig:inc}: operations
 on it consist of several instructions.  It is thus possible
@@
 
-Figure~\ref{fig:upflags} presents a solution based on each process having
+\autoref{fig:upflags} presents a solution based on each process having
 a flag indicating that it is trying to enter the critical section.
@@
 
-However, if you run the program through CXL (Figure~\ref{fig:naiveflagshtml}),
+However, if you run the program through CXL (\autoref{fig:naiveflagshtml}),
 it turns out the solution
@@
 \begin{problems}
-\item Run Figure~\ref{fig:cs} using CXL.  As there is no protection of the critical
+\item Run \autoref{fig:cs} using CXL.  As there is no protection of the critical
 section, mutual exclusion is violated,
@@
 just before entering the critical section
-in Figure~\ref{fig:cs} and run CXL again.
+in \autoref{fig:cs} and run CXL again.
 Mutual exclusion is guaranteed but progress is violated.
@@
 The algorithm is an amalgam of the (incorrect) algorithms in
-Figures~\ref{fig:upflags} and~\ref{fig:upturn}, and is presented
-in Figure~\ref{fig:peterson}.
+\autoref{fig:upflags} and \autoref{fig:upturn}, and is presented
+in \autoref{fig:peterson}.
 A process first indicates its interest in entering the critical section
@@
 machine works.
-In Chapter~\ref{ch:cxlmachine} we talked about the concept of \emph{state}:
+In \autoref{ch:cxlmachine} we talked about the concept of \emph{state}:
 \index{state}%
@@
 are all examples of classes of states.
-Figure~\ref{fig:states} depicts a Venn diagram of various classes of states
+\autoref{fig:states} depicts a Venn diagram of various classes of states
 and a trace.
@@
 
-Figure~\ref{fig:petersonproof} formalizes $\mathcal{I}_p(i)$ in CXL.
+\autoref{fig:petersonproof} formalizes $\mathcal{I}_p(i)$ in CXL.
 The label \texttt{@gate} refers to the instruction that sets \textit{turn} to $1-i$.
-You can run Figure~\ref{fig:petersonproof} to determine
+You can run \autoref{fig:petersonproof} to determine
 that $\mathcal{I}_p(i)$ is indeed an invariant for $i = 0, 1$.
@@
 To demonstrate that this predicate is not invariant, you can remove the
-disjunct from Figure~\ref{fig:petersonproof} and run it to get a
+disjunct from \autoref{fig:petersonproof} and run it to get a
 counterexample.
@@
 \begin{problems}
-\item Figure~\ref{fig:csonebit} presents another correct solution to the
+\item \autoref{fig:csonebit} presents another correct solution to the
 mutual exclusion problem.  It is similar to the one in
-Figure~\ref{fig:upflags}, but has a process \emph{back out and try again}
+\autoref{fig:upflags}, but has a process \emph{back out and try again}
 if it finds that the other process is either trying to enter the critical
@@
 Can you find an inductive invariant for the algorithm in
-Figure~\ref{fig:csonebit} to prove it correct?
+\autoref{fig:csonebit} to prove it correct?
 Here's a pseudo-code version of the algorithm to help you.  Each line
@@
 See if you can figure out for yourself if the two assignments can be
-reversed.  Then run the program in Figure~\ref{fig:peterson} after reversing
+reversed.  Then run the program in \autoref{fig:peterson} after reversing
 the two assignments and describe in English what happens.
@@
 You may note that all this looks familiar.  Indeed, the syntax
-is the same as that for dictionaries (see Chapter~\ref{ch:cxlmachine}).
+is the same as that for dictionaries (see \autoref{ch:cxlmachine}).
 Both dictionaries and methods map CXL values to CXL values,
@@
 
-Figure~\ref{fig:petersonmethods} again shows Peterson's algorithm,
+\autoref{fig:petersonmethods} again shows Peterson's algorithm,
 but this time with methods defined to enter and exit the critical
@@
 
-Figure~\ref{fig:uplock} showed a faulty attempt at solving mutual
+\autoref{fig:uplock} showed a faulty attempt at solving mutual
 exclusion using a lock.  The problem with the implementation of the
@@
 interlock operations.
-For example, we could fix the program in Figure~\ref{fig:inc} by
+For example, we could fix the program in \autoref{fig:inc} by
 constructing an atomic increment operation for a counter, like so:
@@
 operation.
-Method \texttt{tas} in Figure~\ref{fig:tas} shows its specification.
+Method \texttt{tas} in \autoref{fig:tas} shows its specification.
 Here $s$ points to a shared Boolean variable and $p$
@@
 
-Figure~\ref{fig:tas} goes on how to implement mutual exclusion for
+\autoref{fig:tas} goes on how to implement mutual exclusion for
 a set of \texttt{N} processes.
@@
 
-CXL can check these invariants as well.  Figure~\ref{fig:tas} already
+CXL can check these invariants as well.  \autoref{fig:tas} already
 has the code to check $\mathcal{I}_2(i)$.  But how would one go about checking an
@@
 We can do this by adding another process that, in a loop,
-checks the invariant.  Figure~\ref{fig:tasinv} shows the code.
+checks the invariant.  \autoref{fig:tasinv} shows the code.
 Method \texttt{checkInvariant()} checks to see if the invariant holds
@@
 a certain amount of money not exceeding their balance.  A negative balance
-is not allowed.  Figure~\ref{fig:atm} shows two operations on bank accounts:
+is not allowed.  \autoref{fig:atm} shows two operations on bank accounts:
 one to check the balance and one to withdraw money.
@@
 \begin{enumerate}
-\item Fix the code in Figure~\ref{fig:atm}.
+\item Fix the code in \autoref{fig:atm}.
 \item Is it necessary to obtain a lock in \texttt{atm\_check\_balance()}?
@@
 \end{code}
-\caption{\cxllink{code/csTAS.cxl} Fixed version of Figure~\ref{fig:uplock} using test-and-set.}
+\caption{\cxllink{code/csTAS.cxl} Fixed version of \autoref{fig:uplock} using test-and-set.}
 \label{fig:tas2}
@@
 
-In Figure~\ref{fig:tas} we have shown a solution based on a shared
+In \autoref{fig:tas} we have shown a solution based on a shared
 variable and a private variable for each process.   The private
@@
 variable---we only did so to be able to show and check the invariants.
-Figure~\ref{fig:tas2} shows a more straightforward implementation of spinlock.
-The solution is similar to the na\"{i}ve solution of Figure~\ref{fig:uplock},
+\autoref{fig:tas2} shows a more straightforward implementation of spinlock.
+The solution is similar to the na\"{i}ve solution of \autoref{fig:uplock},
 but uses test-and-set to check and set the lock variable atomically.
@@
 \end{code}
-\caption{\cxllink{code/UpLock.cxl} Program of Figure~\ref{fig:inc} fixed with a lock.}
+\caption{\cxllink{code/UpLock.cxl} Program of \autoref{fig:inc} fixed with a lock.}
 \label{fig:incfixed}
@@
 \texttt{synch} that includes support for locks.
-Figure~\ref{fig:spinlocks} shows how they are implemented, and
-Figure~\ref{fig:incfixed} gives an example of how they may be used,
-in this case to fix the program of Figure~\ref{fig:inc}.
+\autoref{fig:spinlocks} shows how they are implemented, and
+\autoref{fig:incfixed} gives an example of how they may be used,
+in this case to fix the program of \autoref{fig:inc}.
 Notice that the module completely hides the implementation of the
@@
 instead of spinlocks.
-Figure~\ref{fig:suspension} shows the same \texttt{Lock} interface implemented
+\autoref{fig:suspension} shows the same \texttt{Lock} interface implemented
 using suspension.
@@
 gets a chance to obtain the lock (assuming every process eventually releases
-it).  Chapter~\ref{ch:starvation} will talk more about starvation and how
+it).  \autoref{ch:starvation} will talk more about starvation and how
 to prevent it.
@@
 \item
-Run Figure~\ref{fig:incfixed} using (1) \texttt{synch}, (2), \texttt{synchS},
+Run \autoref{fig:incfixed} using (1) \texttt{synch}, (2), \texttt{synchS},
 and (3) your own module.  Report how many states were explored by CXL for each
@@
 \end{code}
-\caption{\cxllink{code/RWtest.cxl} Test code for Figure~\ref{fig:rwbusy}.}
+\caption{\cxllink{code/RWtest.cxl} Test code for \autoref{fig:rwbusy}.}
 \label{fig:rwbusytest}
@@
 
-Figure~\ref{fig:rwbusy} presents a solution that uses a
+\autoref{fig:rwbusy} presents a solution that uses a
 single (ordinary) lock and two counters: one that maintains the number
 of readers and one that maintains the number of writers.
-Figure~\ref{fig:rwbusytest} shows how the code may be used.
+\autoref{fig:rwbusytest} shows how the code may be used.
 We will call the lock the \emph{mutex} to distinguish it clearly from
@@
 label \texttt{@rcs} \emph{and} there cannot be another writer at
-label \texttt{@wcs}.  Figure~\ref{fig:rdwr} shows a high-level
+label \texttt{@wcs}.  \autoref{fig:rdwr} shows a high-level
 specification for two processes.
@@
 processes spin in a loop until some desirable application-level condition is met.
-Figure~\ref{fig:inc} also has an example of busy-waiting: the \texttt{main}
+\autoref{fig:inc} also has an example of busy-waiting: the \texttt{main}
 process waits for the other two processes to finish by checking their \texttt{done}
@@
 of \texttt{lock()} spins in a loop until the mutex is available
-(see Figure~\ref{fig:spinlocks}).
+(see \autoref{fig:spinlocks}).
 However, there is a big difference.  As we pointed out, the processes that
@@
 A process trying to obtain a read or write lock in
-Figure~\ref{fig:rwbusy} has to obtain the mutex repeatedly
+\autoref{fig:rwbusy} has to obtain the mutex repeatedly
 until the read or write lock is available.
@@
 
-CXL does not complain about Figure~\ref{fig:rwbusy}, but, using
+CXL does not complain about \autoref{fig:rwbusy}, but, using
 the right test, CXL can be used to check for busy-waiting.
-Figure~\ref{fig:rwblock} presents such a test.
+\autoref{fig:rwblock} presents such a test.
 Note that the initialization code obtains a read lock, preventing
@@
 \begin{problems}
-\item Draw additional states and steps in Figure~\ref{fig:rdwr}
+\item Draw additional states and steps in \autoref{fig:rdwr}
 for three processes.
@@
 
-Figure~\ref{fig:rw2lock} presents a reader/writer lock implementation
+\autoref{fig:rw2lock} presents a reader/writer lock implementation
 that does not busy-wait.
@@
 in the critical section.  Indeed, we could have implemented the reader/writer
-lock as in Figure~\ref{fig:rwmulti} using an ordinary lock, never allowing
+lock as in \autoref{fig:rwmulti} using an ordinary lock, never allowing
 more than one process in the critical section.
 
-Figure~\ref{fig:rwmulti} contains a new test that ensures that indeed more
+\autoref{fig:rwmulti} contains a new test that ensures that indeed more
 than one reader can enter the critical section at a time.  It does so by
@@
 \begin{problems}
-\item Why do you suppose the code in Figure~\ref{fig:rwmulti} uses a flag
+\item Why do you suppose the code in \autoref{fig:rwmulti} uses a flag
 per process rather than a simple counter?
@@
 \emph{Italian Philosophers} (a simplification of
-the famous Dining Philosophers problem described in Chapter~\ref{ch:deadlock}).
+the famous Dining Philosophers problem described in \autoref{ch:deadlock}).
 There are five philosophers sitting around a table, each with a plate of
@@
 
-Figure~\ref{fig:semaphore} shows the \texttt{synch} module implementation of
+\autoref{fig:semaphore} shows the \texttt{synch} module implementation of
 semaphores.
 It represents a semaphore as an integer.
-Figure~\ref{fig:dinerssema} shows a solution to the Italian Philosphers problem
+\autoref{fig:dinerssema} shows a solution to the Italian Philosphers problem
 using a semaphore.
@@
 \item \label{ex:qsort}
-Figure~\ref{fig:qsort} shows an iterative implementation of the Qsort
-algorithm, and Figure~\ref{fig:qsorttest} an accompanying test program.
+\autoref{fig:qsort} shows an iterative implementation of the Qsort
+algorithm, and \autoref{fig:qsorttest} an accompanying test program.
 The array to be sorted is stored in  shared variable $a$.
@@
 \end{code}
-\caption{\cxllink{code/qsorttest.cxl} Test program for Figure~\ref{fig:qsort}.}
+\caption{\cxllink{code/qsorttest.cxl} Test program for \autoref{fig:qsort}.}
 \label{fig:qsorttest}
@@
 \end{code}
-\caption{\cxllink{code/BBsematest.cxl} Test program for Figure~\ref{fig:boundedbuffer}.}
+\caption{\cxllink{code/BBsematest.cxl} Test program for \autoref{fig:boundedbuffer}.}
 \label{fig:bbtest}
@@
 all sharing the same bounded buffer.
-Figure~\ref{fig:pc} gives a high-level description.  The
+\autoref{fig:pc} gives a high-level description.  The
 two numbers in a state specify
@@
 
-Figure~\ref{fig:boundedbuffer} presents the implementation of a bounded
+\autoref{fig:boundedbuffer} presents the implementation of a bounded
 buffer using semaphores.  It features a circular bounded buffer \textit{buf} with
@@
 
-Figure~\ref{fig:bbtest} invokes the code of Figure~\ref{fig:boundedbuffer}.
+\autoref{fig:bbtest} invokes the code of \autoref{fig:boundedbuffer}.
 We can run this through CXL, but what would it test?  There are no
@@
 
-The first part can be tested with the code given in Figure~\ref{fig:pc}, by
+The first part can be tested with the code given in \autoref{fig:pc}, by
 experimenting with different non-negative values for the three constants.
@@
 should terminate, while other values should lead to processes blocking.
-To this this, one can write a script like the one in Figure~\ref{fig:pcscript}.
-
-Figure~\ref{fig:BBsemadata} tests whether the correct data is sent in
+To this this, one can write a script like the one in \autoref{fig:pcscript}.
+
+\autoref{fig:BBsemadata} tests whether the correct data is sent in
 the correct order.  There are the same number of producers and consumers.
@@
 \begin{problems}
-\item Change the code in Figure~\ref{fig:boundedbuffer} so that
+\item Change the code in \autoref{fig:boundedbuffer} so that
 \texttt{produce()} fails if the buffer is full rather than block.
@@
 \item \label{ex:gpu} Implement a thread-safe GPU allocator by modifying
-Figure~\ref{fig:gpu}.
+\autoref{fig:gpu}.
 There are \texttt{N} GPUs identified by the numbers
@@
 We will illustrate the technique using the reader/writer problem.
-Figure~\ref{fig:RWsplitsema} shows the
+\autoref{fig:RWsplitsema} shows the
 \texttt{rlock\_acquire}, \texttt{rlock\_release},
@@
 \item Implement a solution to the producer/consumer problem
-(Figure~\ref{fig:boundedbuffer}) using split binary semaphores.
+(\autoref{fig:boundedbuffer}) using split binary semaphores.
 \item \label{ex:onelane} Cornell's campus features some one-lane bridges.
@@
 
-It is worth noting that Peterson's Algorithm (Chapter~\ref{ch:peterson})
+It is worth noting that Peterson's Algorithm (\autoref{ch:peterson})
 does not suffer from starvation, thanks to the \texttt{turn} variable
@@
 implementations.
-The first, in Figure~\ref{fig:RWfair}, is based on Figure~\ref{fig:RWsplitsema}.
+The first, in \autoref{fig:RWfair}, is based on \autoref{fig:RWsplitsema}.
 There are two differences:
@@
 
-Figures~\ref{fig:RWqueue} and~\ref{fig:RWqtest} present a different approach
+\autoref{fig:RWqueue} and \autoref{fig:RWqtest} present a different approach
 in which processes are allowed into the critical region in a purely First
@@
 \end{code}
-\caption{\cxllink{code/RWqtest.cxl} Test program for Figure~\ref{fig:RWqueue}.}
+\caption{\cxllink{code/RWqtest.cxl} Test program for \autoref{fig:RWqueue}.}
 \label{fig:RWqtest}
@@
 \item Write a fair solution to the one-lane bridge problem of
-Exercise~\ref{ex:onelane}.
+\autoref{ex:onelane}.
 If you want to use the queue method, you can change the signature of the
@@
 
-Figure~\ref{fig:monitors} illustrates the difference with a drawing.
+\autoref{fig:monitors} illustrates the difference with a drawing.
 Here a bathroom represents the critical section, allowing only one
@@
 with explicit locks.
-Figure~\ref{fig:cv} shows the implementation of
+\autoref{fig:cv} shows the implementation of
 condition variables in the \texttt{synch} module.
@@
 once again, an implementation of reader/writer locks.
-Figure~\ref{fig:RWcv} shows the code.  \textit{rwlock} is the global lock or mutex for the
+\autoref{fig:RWcv} shows the code.  \textit{rwlock} is the global lock or mutex for the
 reader/writer lock.  There are two condition variables: readers wait on \textit{rcond} and
@@
 monitors, where multiple waiting readers would have to be signaled one at a time in a
-baton-passing fashion (see Figure~\ref{fig:RWsplitsema}), with Mesa monitors
+baton-passing fashion (see \autoref{fig:RWsplitsema}), with Mesa monitors
 all readers are awaked in one fell swoop using \texttt{notifyAll}.
@@
 \item Implement semaphores using Mesa condition variables.
-\item Write a new version of the GPU allocator in Exercise~\ref{ex:gpu}
+\item Write a new version of the GPU allocator in \autoref{ex:gpu}
 using Mesa condition variables.
@@
 
-Figure~\ref{fig:diners} implements the dining philosophers in CXL, using a
+\autoref{fig:diners} implements the dining philosophers in CXL, using a
 process for each philosopher and a lock for each fork.  If you
@@
 However, mutual exclusion is not easily prevented (although, for some resources it is
-possible, as demonstrated in Chapter~\ref{ch:nonblocking}).
+possible, as demonstrated in \autoref{ch:nonblocking}).
 Havender proposed the following techniques that avoid the remaining
@@
 
-Figure~\ref{fig:dinerscv} demonstrates how this might be done.
+\autoref{fig:dinerscv} demonstrates how this might be done.
 We use a single mutex for the diners, and, for each fork, a boolean
@@
 happening.
-Figure~\ref{fig:dinersavoid} demonstrates this concept.  It uses a semaphore to
+\autoref{fig:dinersavoid} demonstrates this concept.  It uses a semaphore to
 restrict the number of diners at any time to four.
@@
 \begin{problems}
-\item \label{ex:bank} Figure~\ref{fig:bank} shows an implementation of a bank with various
+\item \label{ex:bank} \autoref{fig:bank} shows an implementation of a bank with various
 accounts and transfers between those accounts.
@@
 \emph{message passing}.
-See Figure~\ref{fig:actorpic} for an illustration.
+See \autoref{fig:actorpic} for an illustration.
 Given that there is no shared memory in the actor model (other than the message
@@
 
-Figure~\ref{fig:actor} shows a CXL solution to the
+\autoref{fig:actor} shows a CXL solution to the
 ``multiple producer / multiple consumer problem'' (with an unbounded buffer)
-using an actor (as illustrated in Figure~\ref{fig:actorpic}.
+using an actor (as illustrated in \autoref{fig:actorpic}.
 \texttt{pc\_actor} is essentially a matchmaker: it matches
@@
 
-Figure~\ref{fig:actortest} shows how this code may be used.  It spawns
+\autoref{fig:actortest} shows how this code may be used.  It spawns
 \texttt{NITEMS} producer and consumer processes.  Note that in that
@@
 message.
-Figure~\ref{fig:abptest} shows how the methods may be used to send and receive
+\autoref{fig:abptest} shows how the methods may be used to send and receive
 a stream of \texttt{NMSGS} messages reliably.
@@
 \begin{problems}
-\item Exercise~\ref{ex:cltsvr} explored the \emph{client/server}.  It is popular
+\item \autoref{ex:cltsvr} explored the \emph{client/server}.  It is popular
 in distributed systems as well.
@@
 As a first example, we will revisit the producer/consumer problem.
-The code in Figure~\ref{fig:hw} is based on code developed by Herlihy and
+The code in \autoref{fig:hw} is based on code developed by Herlihy and
 Wing~\cite{HW87}.
@@
 critical~\cite{}.
-Figures~\ref{fig:lst1} and~\ref{fig:lst2} implements an ordered linked
+\autoref{fig:lst1} and \autoref{fig:lst2} implements an ordered linked
 list of integers without duplicates.
@@
 \begin{problems}
-\item Add read-only methods to the data structure in Figures~\ref{fig:lst1}
-and~\ref{fig:lst2} that report the size of the list, the minimum value in the
+\item Add read-only methods to the data structure in \autoref{fig:lst1}
+and \autoref{fig:lst2} that report the size of the list, the minimum value in the
 list, the maximum value in the list, and the sum of the values in the list.
@@
 version number has not changed.  If it has, read-only operation is retried.
-Use a seqlock to implement a bank much like Exercise~\ref{ex:bank}, with
+Use a seqlock to implement a bank much like \autoref{ex:bank}, with
 one seqlock for the entire bank (i.e., no locks on individual accounts).
@@
 
-Figure~\ref{fig:barrierdiagram} shows a high-level depiction with three
+\autoref{fig:barrierdiagram} shows a high-level depiction with three
 processes.  Initially all processes are in round~0.  Then each process
@@
 Counting semaphores work well for implementing barrier synchronization.
-Figure~\ref{fig:barrier} shows an example.  There is a 0-initialized
+\autoref{fig:barrier} shows an example.  There is a 0-initialized
 semaphore for each of the \texttt{N} processes.
@@
 
-Chapter~\ref{ch:cxlmachine} provides an introduction to CXL values.
+\autoref{ch:cxlmachine} provides an introduction to CXL values.
 Below is a complete list of CXL value types with examples:

### 2020-07-06 book/paper.tex
Rename CXL --> Harmony

@@
 
-\newcommand{\cxlsource}[1]{
+\newcommand{\harmonysource}[1]{
 \begin{tabbing}
@@
 
-\newcommand{\cxllink}[1]{%
-[\href{https://www.cs.cornell.edu/home/rvr/cxl/#1}{\underline{#1}}]%
+\newcommand{\harmonylink}[1]{%
+[\href{https://www.cs.cornell.edu/home/rvr/harmony/#1}{\underline{#1}}]%
 }
 
-\newcommand{\cxlref}[2]{%
-\href{https://www.cs.cornell.edu/home/rvr/cxl/output/#1}{\underline{#2}}%
+\newcommand{\harmonyref}[2]{%
+\href{https://www.cs.cornell.edu/home/rvr/harmony/output/#1}{\underline{#2}}%
 }
@@
   name=machine instruction,
-  description={is an atomic operation on the CXL virtual machine, executed by a process}
+  description={is an atomic operation on the Harmony virtual machine, executed by a process}
 }
@@
   name=shared variable,
-  description={is a variable that is stored in the memory of the CXL virtual machine and
+  description={is a variable that is stored in the memory of the Harmony virtual machine and
       shared between multiple processes, as opposed to a process variable}
@@
   description={is a model of computing where the state of a process is kept on
-    a stack.  CXL uses a combination of a stack machine and a register-based machine}
+    a stack.  Harmony uses a combination of a stack machine and a register-based machine}
 }
@@
   name=state,
-  description={an assignment of values to variables.  In a CXL virtual machine, this includes
+  description={an assignment of values to variables.  In a Harmony virtual machine, this includes
       the contents of its shared memory and the set of contexts}
@@
   description={is the execution of a machine instruction by a process, updating
-        its state.  CXL distinguishes micro steps (machine instructions) and macro
+        its state.  Harmony distinguishes micro steps (machine instructions) and macro
         step (a sequence of instructions with at most one effect visible by
@@
 
-\title{Concurrent Programming with CXL}
+\title{Concurrent Programming with Harmony}
 \author{Robbert van Renesse}
@@
 developing concurrent code.  In particular, it uses a new tool
-called CXL that helps with \emph{testing} concurrent code.
+called Harmony that helps with \emph{testing} concurrent code.
 The approach is based on \emph{model checking}:
@@
 instead of relying
-on luck, CXL will run \emph{all possible executions} of a particular
+on luck, Harmony will run \emph{all possible executions} of a particular
 test program.  So even if a bug is unlikely to occur, if the test
@@
 
-So what is CXL?
-CXL is a concurrent programming language.  It was designed to teach
+So what is Harmony?
+Harmony is a concurrent programming language.  It was designed to teach
 the basics of concurrent programming, but it is also useful for
 testing new concurrent algorithms or even sequential and distributed
-algorithms.  CXL programs are not intended to be ``run'' like programs
-in most other programming languages---instead CXL programs are
+algorithms.  Harmony programs are not intended to be ``run'' like programs
+in most other programming languages---instead Harmony programs are
 model checked to test that the program has certain desirable
@@
 
-The syntax and semantics of CXL is similar to that of Python.
+The syntax and semantics of Harmony is similar to that of Python.
 Python is familiar to many programmers and is easy to learn and
@@
 
-CXL is heavily influenced by Leslie Lamport's work on
+Harmony is heavily influenced by Leslie Lamport's work on
 TLA+, TLC, and PlusCal~\cite{Lamport02, Lamport09}.
-CXL is designed to have a lower learning curve than those
+Harmony is designed to have a lower learning curve than those
 systems, but is not as powerful.  When you finish this book
@@
 \begin{itemize}
-\item \autoref{ch:cxlintro} introduces the CXL programming
+\item \autoref{ch:harmonyintro} introduces the Harmony programming
 language, as it provides the language for presenting synchronization
@@
 are concurrently incrementing a counter.
-\item \autoref{ch:cxlmachine} presents the
-CXL virtual machine to understand the problem
+\item \autoref{ch:harmonymachine} presents the
+Harmony virtual machine to understand the problem
 underlying concurrency better.
@@
 elegant solution to implementating a critical section.
-\item \autoref{ch:method} gives some more details on the CXL
+\item \autoref{ch:method} gives some more details on the Harmony
 language.
@@
 for implemented critical sections.
-\item \autoref{ch:synch} presents CXL modules
+\item \autoref{ch:synch} presents Harmony modules
 that supports locks and other synchronization primitives.
@@
 
-\chapter{Introduction to Programming with CXL}
-\label{ch:cxlintro}
-
-Like Python, CXL is an imperative,
+\chapter{Introduction to Programming with Harmony}
+\label{ch:harmonyintro}
+
+Like Python, Harmony is an imperative,
 dynamically typed, and garbage collected programming language.
@@
 \begin{itemize}
-\item Syntax: Every statement in CXL must be terminated by a semicolon
+\item Syntax: Every statement in Harmony must be terminated by a semicolon
 (even \textbf{while} statements and \textbf{def} statements).
 In Python semicolons are optional and rarely used.
-There is no syntactic significance to indentation in CXL.
-Correct indentation in CXL is encouraged but not enforced.
-CXL only supports basic operator precedence or associativity.
+There is no syntactic significance to indentation in Harmony.
+Correct indentation in Harmony is encouraged but not enforced.
+Harmony only supports basic operator precedence or associativity.
 Use parentheses liberally to remove ambiguity.
-\item CXL does not (currently) support floating point, iterators, or I/O;
-CXL does support \textbf{for} loops and various comprehensions over sets.
+\item Harmony does not (currently) support floating point, iterators, or I/O;
+Harmony does support \textbf{for} loops and various comprehensions over sets.
 \item Python is object-oriented, supporting classes with methods and
-inheritance; CXL has objects but does not support classes.  CXL supports
+inheritance; Harmony has objects but does not support classes.  Harmony supports
 pointers to objects.
@@
 There are also many unimportant ones that you will discover as
-you get more familiar with programming in CXL.
-
-\begin{figure}
-\begin{code}
-\cxlsource{triangle}
-\end{code}
-\caption{\cxllink{code/triangle.cxl} Computing triangle numbers.}
+you get more familiar with programming in Harmony.
+
+\begin{figure}
+\begin{code}
+\harmonysource{triangle}
+\end{code}
+\caption{\harmonylink{code/triangle.hny} Computing triangle numbers.}
 \label{fig:triangle}
@@
 
-\autoref{fig:triangle} shows a simple example of a CXL program.
+\autoref{fig:triangle} shows a simple example of a Harmony program.
 (The code for examples in this book can be found in the \texttt{code} folder under
@@
 \textit{result} that eventually contains the result of the
-method (there is no \textbf{return} statement in CXL).  The method
+method (there is no \textbf{return} statement in Harmony).  The method
 also has a variable called \textit{n} containing the value of the
@@
 
-``Running'' this CXL program will try all possible executions, which
+``Running'' this Harmony program will try all possible executions, which
 includes all possible values for $x$.  Try it out (here \texttt{\$}
@@
 \begin{verbatim}
-$ cxl triangle.cxl
+$ harmony triangle.hny
 #states = 13
@@
 
-(For this to work, make sure \texttt{cxl} is in your command shell's search path.)
+(For this to work, make sure \texttt{harmony} is in your command shell's search path.)
 Essentially, the $\texttt{choose}($S$)$
@@
 set~$S$, while the $\textbf{assert}$ statement checks that the output is
-correct.  If the program is correct, the output of CXL is the size of the
-``state graph'' (13 states in this case).  If not, CXL also
+correct.  If the program is correct, the output of Harmony is the size of the
+``state graph'' (13 states in this case).  If not, Harmony also
 reports what went wrong, typically by displaying a summary of an execution in
@@
 
-In CXL, constants have a default value,
+In Harmony, constants have a default value,
 but those can be overridden on the command
@@
 \begin{verbatim}
-$ cxl -c N=100 triangle.cxl
+$ harmony -c N=100 triangle.hny
 #states = 103
@@
 \item See what happens if, instead of initializing \textit{result} to 0,
-you initialize it to 1.  (You do not need to understand the error report at this time.  They will be explained in more detail in \autoref{ch:cxlmachine}.)
-\item Write a CXL program that computes squares by repeated adding.  So the program
+you initialize it to 1.  (You do not need to understand the error report at this time.  They will be explained in more detail in \autoref{ch:harmonymachine}.)
+\item Write a Harmony program that computes squares by repeated adding.  So the program
 should compute the square of $x$ by adding $x$ to an initial value of 0 $x$ times.
@@
 After that, it verifies that the value of \textit{count} equals 2.
-Like Python, CXL supports \texttt{assert} statements with two arguments:
+Like Python, Harmony supports \texttt{assert} statements with two arguments:
 the first is a Boolean condition, and the second is a value that is
@@
 \begin{code}
-\cxlsource{Up}
-\end{code}
-\caption{\cxllink{code/Up.cxl} Incrementing twice in parallel.}
+\harmonysource{Up}
+\end{code}
+\caption{\harmonylink{code/Up.hny} Incrementing twice in parallel.}
 \label{fig:inc}
@@
 
-What is going on is that the CXL program is compiled to machine instructions,
+What is going on is that the Harmony program is compiled to machine instructions,
 \index{machine instruction}%
-and it is the machine instructions that are executed by the underlying CXL
-machine.  The details of this appear in \autoref{ch:cxlmachine},
+and it is the machine instructions that are executed by the underlying Harmony
+machine.  The details of this appear in \autoref{ch:harmonymachine},
 but suffice it to
@@
 A program is correct if it works for any interleaving.
-In fact, CXL will try all possible interleavings of the processes
+In fact, Harmony will try all possible interleavings of the processes
 executing machine instructions.
@@
 \index{race condition}%
-When running CXL, it will
+When running Harmony, it will
 report violations of assertions.  It also provides an example
@@
 The following exercises are intended to show you that on the one hand it
-is easy to translate CXL programs into Python, but on the other hand,
-it is easier to find concurrency bugs using CXL.
+is easy to translate Harmony programs into Python, but on the other hand,
+it is easier to find concurrency bugs using Harmony.
 
 \begin{problems}
-\item CXL programs can usually be easily translated into Python.  For example,
+\item Harmony programs can usually be easily translated into Python.  For example,
 \autoref{fig:incpy} is a Python version of \autoref{fig:inc}.
@@
 \item Can you think of a fix to \autoref{fig:inc}?  Try one or two different fixes
-and run them through CXL.  Do not worry about having to come up with a correct fix at this
+and run them through Harmony.  Do not worry about having to come up with a correct fix at this
 time---the important thing is to develop an understanding of concurrency.
@@
 \end{code}
-\caption{\cxllink{python/Up.py} Python implementation of \autoref{fig:inc}.}
+\caption{\harmonylink{python/Up.py} Python implementation of \autoref{fig:inc}.}
 \label{fig:incpy}
@@
 \end{code}
-\caption{\cxllink{python/UpMany.py} Incrementing \texttt{N} times.}
+\caption{\harmonylink{python/UpMany.py} Incrementing \texttt{N} times.}
 \label{fig:incmany}
@@
 
-\chapter{The CXL Virtual Machine}
-\label{ch:cxlmachine}
-\index{CXL Virtual Machine}%
-
-CXL programs are compiled to CXL bytecode
+\chapter{The Harmony Virtual Machine}
+\label{ch:harmonymachine}
+\index{Harmony Virtual Machine}%
+
+Harmony programs are compiled to Harmony bytecode
 \index{bytecode}%
 (machine instructions for a virtual machine),
-which in turn is executed by the CXL virtual machine (CVM).
+which in turn is executed by the Harmony virtual machine (HVM).
 \index{virtual machine}%
-\index{CXL Virtual Machine}%
-\index{CVM}%
+\index{Harmony Virtual Machine}%
+\index{HVM}%
 To understand the problem of concurrent computing, it
-is important to have a basic understanding of the CVM.
-
-\section*{CXL Values}
-\label{ap:cxlvalues}
-
-CXL programs, and indeed the CVM,  manipulate CXL values.
-CXL values are recursively defined:
+is important to have a basic understanding of the HVM.
+
+\section*{Harmony Values}
+\label{ap:harmonyvalues}
+
+Harmony programs, and indeed the HVM,  manipulate Harmony values.
+Harmony values are recursively defined:
 they include booleans (\texttt{False} and \texttt{True}),
@@
 strings (enclosed by double quotes),
-sets of CXL values,
+sets of Harmony values,
 and dictionaries
 \index{dictionary}%
-that map CXL values to other CXL values.
+that map Harmony values to other Harmony values.
 %
-Another type of CXL value is the \emph{atom}.
+Another type of Harmony value is the \emph{atom}.
 \index{atom}%
@@
 
-CXL makes extensive use of dictionaries.
+Harmony makes extensive use of dictionaries.
 A dictionary maps values, known as \emph{keys}, to values.
-The CXL dictionary syntax and properties are a little different from Python.
-Unlike Python, any CXL value can be a key, including another
+The Harmony dictionary syntax and properties are a little different from Python.
+Unlike Python, any Harmony value can be a key, including another
 dictionary.
-CXL dictionaries are written as
+Harmony dictionaries are written as
 $\mathtt{dict}\{ k_0: v_0, ~ k_1: v_1, ~ ... \}$.
 If $d$ is a dictionary, and $k$ is a key, then the
-following expression retrieves the CXL value that $k$ maps to in $d$:
+following expression retrieves the Harmony value that $k$ maps to in $d$:
 \begin{code}
@@
 \end{code}
-This is unfamiliar to Python programers, but in CXL square brackets can be used
+This is unfamiliar to Python programers, but in Harmony square brackets can be used
 in the same way as parentheses, so you can express the same thing in the form
@@
 the indexes into the tuple.  For example, the tuple
-\texttt{(5, False)} is the same CXL value as
+\texttt{(5, False)} is the same Harmony value as
 \texttt{dict\{ 0:5, 1:False \}}.
@@
 
-Again, square brackets and parentheses work the same in CXL, so
+Again, square brackets and parentheses work the same in Harmony, so
 \texttt{[a, b, c]} (which looks like a Python list)
-is the same CXL value as \texttt{(a, b, c)} (which looks like a Python tuple),
-which in turn is the same CXL value as \texttt{dict\{ 0:a, 1:b, 2:c \}}.
+is the same Harmony value as \texttt{(a, b, c)} (which looks like a Python tuple),
+which in turn is the same Harmony value as \texttt{dict\{ 0:a, 1:b, 2:c \}}.
 So if $x$ = \texttt{[ False, True ]},
@@
 
-CXL is not an object-oriented language, so objects don't have
-built-in methods.  However, CXL does have some powerful operators to
+Harmony is not an object-oriented language, so objects don't have
+built-in methods.  However, Harmony does have some powerful operators to
 make up for some of that.
@@
 
-Appendix~\ref{ap:cxlvalues} provides details on all the values that
-CXL currently supports.
-
-\section*{CXL Bytecode}
-
-A CXL program is translated
+Appendix~\ref{ap:harmonyvalues} provides details on all the values that
+Harmony currently supports.
+
+\section*{Harmony Bytecode}
+
+A Harmony program is translated
 into \emph{bytecode}, which is list of machine instructions that the
-CXL virtual machine (CVM) executes.
-The CVM is not an ordinary virtual machine, but its architecture
+Harmony virtual machine (HVM) executes.
+The HVM is not an ordinary virtual machine, but its architecture
 is nonetheless representative of conventional computers and
@@
 
-Instead of bits and bytes, a CVM manipulates CXL values.
-A CVM has the following components:
+Instead of bits and bytes, a HVM manipulates Harmony values.
+A HVM has the following components:
 \begin{itemize}
-\item Code:  This is an immutable and finite list of CVM instructions,
-generated from a CXL program.  The types of instructions will be described later.
-\item Shared memory: A CVM has just one memory location containing
-an CXL value.
+\item Code:  This is an immutable and finite list of HVM instructions,
+generated from a Harmony program.  The types of instructions will be described later.
+\item Shared memory: A HVM has just one memory location containing
+an Harmony value.
 \item Processes:  Any process
@@
 \index{program counter}%
-a stack of CXL values,
+a stack of Harmony values,
 and a single mutable general purpose \emph{register}
 \index{register}%
-that contains a CXL value.
+that contains a Harmony value.
 \end{itemize}
@@
 its name tag, program counter, stack, and register.
-The state of a CVM
+The state of a HVM
 consists of the value of its memory and the multiset (or \emph{bag})
@@
 be in the same state.
-The initial state of the CXL memory is the empty dictionary, \texttt{dict\{\}}.
+The initial state of the Harmony memory is the empty dictionary, \texttt{dict\{\}}.
 The context bag has an initial context in it with name tag
@@
 \begin{verbatim}
-Up.cxl:1 def incrementer(self):
+Up.hny:1 def incrementer(self):
     0 Jump 34
     1 Frame incrementer(self)
-Up.cxl:2     count = count + 1;
+Up.hny:2     count = count + 1;
     2 Load count
@@
     5 Store count
-Up.cxl:3     done[self] = True;
+Up.hny:3     done[self] = True;
     6 LoadVar self
@@
 \end{code}
-\caption{The first part of the CVM bytecode corresponding to \autoref{fig:inc}.}
+\caption{The first part of the HVM bytecode corresponding to \autoref{fig:inc}.}
 \label{fig:inccode}
@@
 process has only one register.  However, this is not a limitation because
-CXL values are unbounded trees.
+Harmony values are unbounded trees.
 Both the memory and the register of a process always contain
-a dictionary that maps atoms to CXL values.  We call this a \emph{directory}.
+a dictionary that maps atoms to Harmony values.  We call this a \emph{directory}.
 \index{directory}%
@@
 %
-Because directories are CXL values themselves,
+Because directories are Harmony values themselves,
 directories can be organized into a tree.
@@
 \index{address}%
-of a CXL value, and it is relative to the
+of a Harmony value, and it is relative to the
 ``root'' of the directory tree.
@@
 main/() [13-17,20-24,26-31] 31 { count=1, done=[True,  True]  }
->>> CXL Assertion failed : 1
-Open file://.../cxl.html for more information
+>>> Harmony Assertion failed : 1
+Open file://.../harmony.html for more information
 \end{verbatim}
 \end{code}
-\caption{The text output of running CXL on \autoref{fig:inc}.}
+\caption{The text output of running Harmony on \autoref{fig:inc}.}
 \label{fig:incoutput}
@@
 
-Compiling the code in \autoref{fig:inc} results in the CVM bytecode
+Compiling the code in \autoref{fig:inc} results in the HVM bytecode
 listed in \autoref{fig:inccode}.
-You can obtain this code by invoking \texttt{cxl} with the \texttt{-a} flag
+You can obtain this code by invoking \texttt{harmony} with the \texttt{-a} flag
 like so:
@@
 \begin{verbatim}
-cxl -a Up.cxl
+harmony -a Up.hny
 \end{verbatim}
 \end{code}
-Each process in the CVM is predominantly a \emph{stack machine},
+Each process in the HVM is predominantly a \emph{stack machine},
 \index{stack machine}%
@@
 
-Appendix~\ref{ap:cxlbytecode} provides a list of all CVM machine instructions,
+Appendix~\ref{ap:harmonybytecode} provides a list of all HVM machine instructions,
 in case you want to read about the details.
@@
 The code generated from $\mathit{count} := \mathit{count} + 1$ in line 2 of
-\texttt{Up.cxl} is as follows (see \autoref{fig:inccode}):
+\texttt{Up.hny} is as follows (see \autoref{fig:inccode}):
 
@@
 \item The \texttt{Store} instruction pops
-a CXL value (the sum of the \textit{count} variable and 1) and
+a Harmony value (the sum of the \textit{count} variable and 1) and
 stores it in the \textit{count} variable.
@@
 Once initialization completes, any processes that were spawned can run.
-You can think of CXL as trying every possible interleaving of processes executing
+You can think of Harmony as trying every possible interleaving of processes executing
 instructions.
-\autoref{fig:incoutput} shows the output produced by running CXL on the
-\texttt{Up.cxl} program.
+\autoref{fig:incoutput} shows the output produced by running Harmony on the
+\texttt{Up.hny} program.
 
@@
 \caption{The
-\cxlref{Up.html}{HTML output of running CXL on \autoref{fig:inc}}.
+\harmonyref{Up.html}{HTML output of running Harmony on \autoref{fig:inc}}.
 There are three sections.
@@
 
-CXL can report a variety of failure types:
+Harmony can report a variety of failure types:
 \begin{itemize}
@@
 It also includes certain infinite loops.
-CXL will print at trace of the shortest bad execution that it found.
-\item \texttt{Non-terminating States}: CXL found one or more states
+Harmony will print at trace of the shortest bad execution that it found.
+\item \texttt{Non-terminating States}: Harmony found one or more states
 from which there does not exist an execution such that all processes
-terminate.  CXL will not only print the non-terminating state with
+terminate.  Harmony will not only print the non-terminating state with
 the shortest trace, but also the list of processes that are running
@@
 
-In this case, CXL reports a safety violation, in particular it
+In this case, Harmony reports a safety violation, in particular it
 reports that the assertion on line 9 failed and that
@@
 \item The name tag of the process;
-\item The sequence of program counters of the CVM instructions that the process executed;
+\item The sequence of program counters of the HVM instructions that the process executed;
 \item The current program counter of the process;
@@
 
-CXL also generates an HTML file that allows exploring more details
+Harmony also generates an HTML file that allows exploring more details
 of the execution interactively.
@@
 \item \autoref{fig:incenter} shows an attempt at trying to fix the code of
-    \autoref{fig:inc}.  Run it through CXL and see what happens.  Based on
+    \autoref{fig:inc}.  Run it through Harmony and see what happens.  Based on
     the error output, describe in English what is wrong with the code by describing,
@@
     statement (between lines~7 and~8)?  Do you think that would work?  Run it through
-    CXL and describe either why it works or why it does not work.
+    Harmony and describe either why it works or why it does not work.
 \end{problems}
@@
 \begin{code}
-\cxlsource{UpEnter}
-\end{code}
-\caption{\cxllink{code/UpEnter.cxl} Broken attempt at fixing the code of \autoref{fig:inc}.}
+\harmonysource{UpEnter}
+\end{code}
+\caption{\harmonylink{code/UpEnter.hny} Broken attempt at fixing the code of \autoref{fig:inc}.}
 \label{fig:incenter}
@@
 \begin{code}
-\cxlsource{csbarebones}
-\end{code}
-\caption{\cxllink{code/csbarebones.cxl} A bare bones critical section with two processes.}
+\harmonysource{csbarebones}
+\end{code}
+\caption{\harmonylink{code/csbarebones.hny} A bare bones critical section with two processes.}
 \label{fig:csbarebones}
@@
 \begin{code}
-\cxlsource{cs}
-\end{code}
-\caption{\cxllink{code/csbarebones.cxl} CXL model of a critical section.}
+\harmonysource{cs}
+\end{code}
+\caption{\harmonylink{code/csbarebones.hny} Harmony model of a critical section.}
 \label{fig:cs}
@@
 entering and exiting the critical section.
-\autoref{fig:csbarebones} shows the CXL code.
+\autoref{fig:csbarebones} shows the Harmony code.
 Here \texttt{@cs} is a \emph{label},
 \index{label}%
-identifying a location in the CVM bytecode.  The first thing we need to
+identifying a location in the HVM bytecode.  The first thing we need to
 ensure is that there can never be two processes in the critical section.
@@
 
-CXL in fact supports this.
+Harmony in fact supports this.
 It has an operator \textbf{atLabel}~$L$,
@@
 that returns the name tag of the current process.
-If you run the code through CXL, the assertion should fail because
+If you run the code through Harmony, the assertion should fail because
 there is no code yet for safely entering and exiting the critical section.
@@
 In order to detect violations of progress, and other liveness problems in
-algorithms in general, CXL requires that every execution must be
+algorithms in general, Harmony requires that every execution must be
 able to reach a state in which all processes have terminated.
@@
 without entering the critical section ever.
-(Recall that CXL will try every possible execution, and so it will evaluate
+(Recall that Harmony will try every possible execution, and so it will evaluate
 both choices.)
@@
 \begin{code}
-\cxlsource{naiveLock}
-\end{code}
-\caption{\cxllink{code/naiveLock.cxl} Na\"{\i}ve implementation of a shared lock.}
+\harmonysource{naiveLock}
+\end{code}
+\caption{\harmonylink{code/naiveLock.hny} Na\"{\i}ve implementation of a shared lock.}
 \label{fig:uplock}
@@
 \includegraphics[width=\textwidth]{figures/naiveLock.png}
-\caption{The \cxlref{naiveLock.html}{HTML output of running CXL on \autoref{fig:uplock}}.}
+\caption{The \harmonyref{naiveLock.html}{HTML output of running Harmony on \autoref{fig:uplock}}.}
 \label{fig:naiveLockhtml}
@@
 \begin{code}
-\cxlsource{naiveFlags}
-\end{code}
-\caption{\cxllink{code/naiveFlags.cxl} Na\"{\i}ve use of flags to solve mutual exclusion.}
+\harmonysource{naiveFlags}
+\end{code}
+\caption{\harmonylink{code/naiveFlags.hny} Na\"{\i}ve use of flags to solve mutual exclusion.}
 \label{fig:upflags}
@@
 \includegraphics[width=\textwidth]{figures/naiveFlags.png}
-\caption{The \cxlref{naiveFlags.html}{HTML output of running CXL on \autoref{fig:upflags}}.}
+\caption{The \harmonyref{naiveFlags.html}{HTML output of running Harmony on \autoref{fig:upflags}}.}
 \label{fig:naiveflagshtml}
@@
 \begin{code}
-\cxlsource{naiveTurn}
-\end{code}
-\caption{\cxllink{code/naiveTurn.cxl} Na\"{\i}ve use of turn variable to solve mutual exclusion.}
+\harmonysource{naiveTurn}
+\end{code}
+\caption{\harmonylink{code/naiveTurn.hny} Na\"{\i}ve use of turn variable to solve mutual exclusion.}
 \label{fig:upturn}
@@
 
-Unfortunately, if we run the program through CXL,
+Unfortunately, if we run the program through Harmony,
 we find that the assertion still fails.
-\autoref{fig:naiveLockhtml} shows the CXL output.
+\autoref{fig:naiveLockhtml} shows the Harmony output.
 Process~0 finds that the lock is available, but just before it stores
@@
 
-However, if you run the program through CXL (\autoref{fig:naiveflagshtml}),
+However, if you run the program through Harmony (\autoref{fig:naiveflagshtml}),
 it turns out the solution
@@
 
-Run the program through CXL.  It turns out that this solution also violates
+Run the program through Harmony.  It turns out that this solution also violates
 \emph{progress}, albeit for a different reason:
@@
 \begin{problems}
-\item Run \autoref{fig:cs} using CXL.  As there is no protection of the critical
+\item Run \autoref{fig:cs} using Harmony.  As there is no protection of the critical
 section, mutual exclusion is violated,
@@
 just before entering the critical section
-in \autoref{fig:cs} and run CXL again.
+in \autoref{fig:cs} and run Harmony again.
 Mutual exclusion is guaranteed but progress is violated.
-CXL should print a trace
+Harmony should print a trace
 to a state from which a terminating state cannot be reached.
@@
 \item See if you can come up with some different approaches that satisfy both
-mutual exclusion and progress.  Try them with CXL and see if they work or not.
+mutual exclusion and progress.  Try them with Harmony and see if they work or not.
 If they don't, try to understand why.
@@
 \begin{code}
-\cxlsource{Peterson}
-\end{code}
-\caption{\cxllink{code/Peterson.cxl} Peterson's Algorithm}
+\harmonysource{Peterson}
+\end{code}
+\caption{\harmonylink{code/Peterson.hny} Peterson's Algorithm}
 \label{fig:peterson}
@@
 enter the critical section---if both do so at the same time one will
-win because writes to memory in CXL are atomic.
+win because writes to memory in Harmony are atomic.
 The process continues to be polite, waiting in a \textbf{while} loop
@@
 ($\mathit{turn} == \mathit{self}$).
-Running the algorithm with CXL shows that it satisfies both mutual
+Running the algorithm with Harmony shows that it satisfies both mutual
 exclusion and progress.
@@
 mutual exclusion for an algorithm such as Peterson's.
-For that, we have to understand a little bit more about how the CXL
+For that, we have to understand a little bit more about how the Harmony
 machine works.
-In \autoref{ch:cxlmachine} we talked about the concept of \emph{state}:
+In \autoref{ch:harmonymachine} we talked about the concept of \emph{state}:
 \index{state}%
-at any point in time the CVM is in a specific state.
+at any point in time the HVM is in a specific state.
 A state is comprised of the values of the shared variables as well as
@@
 program counter and the contents of its stack.
-Everytime a process executes a CVM machine instruction, the
+Everytime a process executes a HVM machine instruction, the
 state changes (if only because the program counter of the process
@@
 \index{step}%
-Steps in CXL are atomic.
+Steps in Harmony are atomic.
 
@@
 
-The CVM starts in an initial state in which there is only
+The HVM starts in an initial state in which there is only
 one process and its program counter is~0.  A \emph{trace}
@@
 \index{non-determinism}%
-in CXL.
+in Harmony.
 One is when
@@
 After all, it certainly holds over the initial state.
-And as CXL has already determined, exclusivity is an invariant:
+And as Harmony has already determined, exclusivity is an invariant:
 it holds over every reachable state.
@@
 \begin{code}
-\cxlsource{PetersonInductive}
-\end{code}
-\caption{\cxllink{code/PetersonInductive.cxl} Peterson's Algorithm with Inductive Invariant}
+\harmonysource{PetersonInductive}
+\end{code}
+\caption{\harmonylink{code/PetersonInductive.hny} Peterson's Algorithm with Inductive Invariant}
 \label{fig:petersonproof}
@@
 
-\autoref{fig:petersonproof} formalizes $\mathcal{I}_p(i)$ in CXL.
+\autoref{fig:petersonproof} formalizes $\mathcal{I}_p(i)$ in Harmony.
 The label \texttt{@gate} refers to the instruction that sets \textit{turn} to $1-i$.
@@
 We have now demonstrated mutual exclusion in Peterson's Algorithm in two
-different ways: one by letting CXL explore all possible executions, the
+different ways: one by letting Harmony explore all possible executions, the
 other using an inductive invariant and proof by induction.  The former
@@
 algorithm works.  The second provides much more insight.  We therefore
-encourage to include inductive invariants in your CXL code.
-
-A cool anecdote is the following.  When the author of CXL had to teach
+encourage to include inductive invariants in your Harmony code.
+
+A cool anecdote is the following.  When the author of Harmony had to teach
 Peterson's Algorithm, he refreshed his memory by looking at the Wikipedia
@@
 Before trying to prove the algorithm, you can check that the predicate is
-at least invariant by testing it using CXL.  Doing so could potentially
+at least invariant by testing it using Harmony.  Doing so could potentially
 avoid wasting your time on a proof that will not work because the
@@
 \begin{code}
-\cxlsource{csonebit}
-\end{code}
-\caption{\cxllink{code/csonebit.cxl} Mutual exclusion using a flag per process.}
+\harmonysource{csonebit}
+\end{code}
+\caption{\harmonylink{code/csonebit.hny} Mutual exclusion using a flag per process.}
 \label{fig:csonebit}
@@
 
-\chapter{CXL Methods and Pointers}
+\chapter{Harmony Methods and Pointers}
 \label{ch:method}
-\index{CXL method}%
+\index{Harmony method}%
 
@@
 You may note that all this looks familiar.  Indeed, the syntax
-is the same as that for dictionaries (see \autoref{ch:cxlmachine}).
-Both dictionaries and methods map CXL values to CXL values,
+is the same as that for dictionaries (see \autoref{ch:harmonymachine}).
+Both dictionaries and methods map Harmony values to Harmony values,
 and their syntax is indistinguishable.
 If \texttt{f} is either a method or a
-dictionary, and $x$ is an arbitrary CXL value, then
+dictionary, and $x$ is an arbitrary Harmony value, then
 \texttt{f} $x$, \texttt{f}($x$), and \texttt{f}[$x$] are all
-the same expression in CXL.
-
-\begin{figure}
-\begin{code}
-\cxlsource{PetersonMethod}
-\end{code}
-\caption{\cxllink{code/PetersonMethod.cxl} Peterson's Algorithm accessed through methods.}
+the same expression in Harmony.
+
+\begin{figure}
+\begin{code}
+\harmonysource{PetersonMethod}
+\end{code}
+\caption{\harmonylink{code/PetersonMethod.hny} Peterson's Algorithm accessed through methods.}
 \label{fig:petersonmethods}
@@
 
-CXL is not an object-oriented language like Python is.  In Python
+Harmony is not an object-oriented language like Python is.  In Python
 you can pass a reference to an object to a method, and that method
-can then update the object.  In CXL, it is also sometimes convenient
+can then update the object.  In Harmony, it is also sometimes convenient
 to have a method update a shared variable specified as an argument.
-For this (as well as some other uses), CXL supports \emph{pointers}
+For this (as well as some other uses), Harmony supports \emph{pointers}
 \index{pointer}%
@@
 
-You can put the first three methods in its own CXL source file
-and include it using the CXL \texttt{import} statement.
+You can put the first three methods in its own Harmony source file
+and include it using the Harmony \texttt{import} statement.
 \index{import statement}%
@@
 \begin{code}
-\cxlsource{spinlock}
-\end{code}
-\caption{\cxllink{code/spinlock.cxl} Mutual Exclusion using a ``spinlock'' based on test-and-set.}
+\harmonysource{spinlock}
+\end{code}
+\caption{\harmonylink{code/spinlock.hny} Mutual Exclusion using a ``spinlock'' based on test-and-set.}
 \label{fig:tas}
@@
 
-While the CVM does not have any specific built-in interlock instructions,
+While the HVM does not have any specific built-in interlock instructions,
 it does have support for executing multiple instructions atomically.
-This feature is available in the CXL language in two ways.
-First, any CXL statement can be made atomic by placing a label in front
-of it.  Second, a group of CXL statements can be made atomic
+This feature is available in the Harmony language in two ways.
+First, any Harmony statement can be made atomic by placing a label in front
+of it.  Second, a group of Harmony statements can be made atomic
 through its \textbf{atomic}
@@
 \begin{code}
-\cxlsource{atomicinc}
+\harmonysource{atomicinc}
 \end{code}
@@
 \begin{code}
-\cxlsource{spinlockInv}
-\end{code}
-\caption{\cxllink{code/spinlockInv.cxl} Checking invariants.}
+\harmonysource{spinlockInv}
+\end{code}
+\caption{\harmonylink{code/spinlockInv.hny} Checking invariants.}
 \label{fig:tasinv}
@@
 
-CXL can check these invariants as well.  \autoref{fig:tas} already
+Harmony can check these invariants as well.  \autoref{fig:tas} already
 has the code to check $\mathcal{I}_2(i)$.  But how would one go about checking an
@@
 Method \texttt{checkInvariant()} checks to see if the invariant holds
-in a state.  It introduces a new feature of CXL: the ability to have
+in a state.  It introduces a new feature of Harmony: the ability to have
 variables local to a method.  In this case, the process variable \textit{sum}
@@
 the other processes.
-In CXL, \textbf{assert} statements are executed atomically, so the
+In Harmony, \textbf{assert} statements are executed atomically, so the
 evaluation of the assertion is not interleaved with the execution
 of other processes.
-Because CXL tries every possible execution, the process is guaranteed
+Because Harmony tries every possible execution, the process is guaranteed
 to find violations of the invariant if it does not hold.
@@
 \item Implement a spinlock using the atomic swap operation.
-\item Write out the invariants that need to hold and check them using CXL.
+\item Write out the invariants that need to hold and check them using Harmony.
 \item People who use an ATM often first check their balance and then withdraw
@@
 the balance.  Method \texttt{checker} looks for negative balances.
-Running the code through CXL reveals that there is a bug.
+Running the code through Harmony reveals that there is a bug.
 It is a common type of concurrency bug known as \emph{Time Of Check Time Of
@@
 \begin{code}
-\cxlsource{atm}
-\end{code}
-\caption{\cxllink{code/atm.cxl} Withdrawing money from an ATM.}
+\harmonysource{atm}
+\end{code}
+\caption{\harmonylink{code/atm.hny} Withdrawing money from an ATM.}
 \label{fig:atm}
@@
 \begin{code}
-\cxlsource{csTAS}
-\end{code}
-\caption{\cxllink{code/csTAS.cxl} Fixed version of \autoref{fig:uplock} using test-and-set.}
+\harmonysource{csTAS}
+\end{code}
+\caption{\harmonylink{code/csTAS.hny} Fixed version of \autoref{fig:uplock} using test-and-set.}
 \label{fig:tas2}
@@
 
-In CXL, atomic statements allow you to \emph{implement} your own
+In Harmony, atomic statements allow you to \emph{implement} your own
 synchronization primitives like test-and-set.  Other common examples
@@
 \begin{code}
-\cxlsource{lockintf}
+\harmonysource{lockintf}
 \end{code}
@@
 \begin{code}
-\cxlsource{UpLock}
-\end{code}
-\caption{\cxllink{code/UpLock.cxl} Program of \autoref{fig:inc} fixed with a lock.}
+\harmonysource{UpLock}
+\end{code}
+\caption{\harmonylink{code/UpLock.hny} Program of \autoref{fig:inc} fixed with a lock.}
 \label{fig:incfixed}
@@
 
-Because locks are so common, CXL has a module called
+Because locks are so common, Harmony has a module called
 \texttt{synch} that includes support for locks.
@@
 \begin{code}
-\cxlsource{locksusp}
+\harmonysource{locksusp}
 \end{code}
@@
 until the lock is available to it.
-In CXL, a process can suspend itself and save its context (state) in a
+In Harmony, a process can suspend itself and save its context (state) in a
 list.  Recall that the context of a process consists of its name tag,
 its program counter, and the contents of its register and stack.
-A context is a regular CXL value.
+A context is a regular Harmony value.
 The syntax of the expression is as follows:
@@
 
-Here $C$ is a context and $R$ is a CXL value.
+Here $C$ is a context and $R$ is a Harmony value.
 It causes a process with context $C$ to be added to the state that has
@@
 
-CXL allows you to select which version of the \texttt{synch} module you would
+Harmony allows you to select which version of the \texttt{synch} module you would
 like to use with the \texttt{-m} flag.
@@
 \begin{verbatim}
-cxl -m synch=synchS x.cxl
+harmony -m synch=synchS x.hny
 \end{verbatim}
@@
 
-runs the file \texttt{x.cxl} using the suspension version of the \texttt{synch} module.
+runs the file \texttt{x.hny} using the suspension version of the \texttt{synch} module.
 You will find that using the \texttt{synchS} module often leads to searching a
@@
 Run \autoref{fig:incfixed} using (1) \texttt{synch}, (2), \texttt{synchS},
-and (3) your own module.  Report how many states were explored by CXL for each
+and (3) your own module.  Report how many states were explored by Harmony for each
 version of the module.
@@
 \begin{code}
-\cxlsource{RW}
-\end{code}
-\caption{\cxllink{code/RW.cxl} Busy-Waiting Reader/Writer Lock implementation.}
+\harmonysource{RW}
+\end{code}
+\caption{\harmonylink{code/RW.hny} Busy-Waiting Reader/Writer Lock implementation.}
 \label{fig:rwbusy}
@@
 \begin{code}
-\cxlsource{RWtest}
-\end{code}
-\caption{\cxllink{code/RWtest.cxl} Test code for \autoref{fig:rwbusy}.}
+\harmonysource{RWtest}
+\end{code}
+\caption{\harmonylink{code/RWtest.hny} Test code for \autoref{fig:rwbusy}.}
 \label{fig:rwbusytest}
@@
 \begin{code}
-\cxlsource{RWbusychk}
-\end{code}
-\caption{\cxllink{code/RWbusychk.cxl} Checking for Busy Waiting.}
+\harmonysource{RWbusychk}
+\end{code}
+\caption{\harmonylink{code/RWbusychk.hny} Checking for Busy Waiting.}
 \label{fig:rwblock}
@@
 The astute reader might wonder if obtaining the mutex itself is an
-example of busy-waiting.  After all, the CXL \texttt{synch} implementation
+example of busy-waiting.  After all, the Harmony \texttt{synch} implementation
 of \texttt{lock()} spins in a loop until the mutex is available
@@
 queue and stops using CPU cycles until the mutex becomes available.
-CXL supports this with the \texttt{synchS} module.
+Harmony supports this with the \texttt{synchS} module.
 Busy waiting disables all this: even when using the \texttt{synchS}
@@
 
-CXL does not complain about \autoref{fig:rwbusy}, but, using
-the right test, CXL can be used to check for busy-waiting.
+Harmony does not complain about \autoref{fig:rwbusy}, but, using
+the right test, Harmony can be used to check for busy-waiting.
 \autoref{fig:rwblock} presents such a test.
@@
 all readers and writers from entering the critical section.
-In that case, those processes should ideally be blocked and CXL
+In that case, those processes should ideally be blocked and Harmony
 can test for that by running it with the \texttt{-b} flag.
-This flag tells CXL to check
+This flag tells Harmony to check
 that all states are non-terminating and lead to a state in which
@@
 \begin{code}
-\cxlsource{RWlock}
-\end{code}
-\caption{\cxllink{code/RWlock.cxl} Reader/Writer with Two Locks.}
+\harmonysource{RWlock}
+\end{code}
+\caption{\harmonylink{code/RWlock.hny} Reader/Writer with Two Locks.}
 \label{fig:rw2lock}
@@
 \begin{code}
-\cxlsource{RWmulti}
-\end{code}
-\caption{\cxllink{code/RWmulti.cxl} Checking that multiple readers can acquire the read lock.}
+\harmonysource{RWmulti}
+\end{code}
+\caption{\harmonylink{code/RWmulti.hny} Checking that multiple readers can acquire the read lock.}
 \label{fig:rwmulti}
@@
 that does not busy-wait.
-You can test the code by running \texttt{cxl -m RW=RWlock RWtest}.
+You can test the code by running \texttt{harmony -m RW=RWlock RWtest}.
 It uses two ordinary locks: \textit{rwlock} is used by both readers and writers,
@@
 
-While CXL does not detect any issues, we must remember what it is checking
+While Harmony does not detect any issues, we must remember what it is checking
 for: it checks to make sure that there are never a reader and a writer
@@
 are there.  If other readers cannot enter, then that process enters in an
-infinite loop that CXL will readily detect.  Try it out!
-
-An important lesson here is that one should not celebrate too early if CXL
+infinite loop that Harmony will readily detect.  Try it out!
+
+An important lesson here is that one should not celebrate too early if Harmony
 does not find any issues.  While a test may be successful, a test may not
-explore all desirable executions.  While CXL can help explore all cornercases
+explore all desirable executions.  While Harmony can help explore all cornercases
 in a test, it cannot find problems that the test is not looking for.
@@
 \begin{code}
-\cxlsource{semaphore}
+\harmonysource{semaphore}
 \end{code}
@@
 \begin{code}
-\cxlsource{DinersSema}
-\end{code}
-\caption{\cxllink{code/DinersSema.cxl} Italian Philosophers.}
+\harmonysource{DinersSema}
+\end{code}
+\caption{\harmonylink{code/DinersSema.hny} Italian Philosophers.}
 \label{fig:dinerssema}
@@
 \begin{code}
-\cxlsource{qsort}
-\end{code}
-\caption{\cxllink{code/qsort.cxl} Iterative qsort() implementation.}
+\harmonysource{qsort}
+\end{code}
+\caption{\harmonylink{code/qsort.hny} Iterative qsort() implementation.}
 \label{fig:qsort}
@@
 \begin{code}
-\cxlsource{qsorttest}
-\end{code}
-\caption{\cxllink{code/qsorttest.cxl} Test program for \autoref{fig:qsort}.}
+\harmonysource{qsorttest}
+\end{code}
+\caption{\harmonylink{code/qsorttest.hny} Test program for \autoref{fig:qsort}.}
 \label{fig:qsorttest}
@@
 \begin{code}
-\cxlsource{BBsema}
-\end{code}
-\caption{\cxllink{code/BBsema.cxl} Bounded Buffer implementation using semaphores.}
+\harmonysource{BBsema}
+\end{code}
+\caption{\harmonylink{code/BBsema.hny} Bounded Buffer implementation using semaphores.}
 \label{fig:boundedbuffer}
@@
 \begin{code}
-\cxlsource{BBsematest}
-\end{code}
-\caption{\cxllink{code/BBsematest.cxl} Test program for \autoref{fig:boundedbuffer}.}
+\harmonysource{BBsematest}
+\end{code}
+\caption{\harmonylink{code/BBsematest.hny} Test program for \autoref{fig:boundedbuffer}.}
 \label{fig:bbtest}
@@
 \begin{verbatim}
-cxl -b -c NSLOTS=0 -c NPRODS=1 -c NCONSS=1 BBsematest.cxl
-cxl -b -c NSLOTS=1 -c NPRODS=0 -c NCONSS=1 BBsematest.cxl
-cxl    -c NSLOTS=1 -c NPRODS=1 -c NCONSS=0 BBsematest.cxl
-cxl    -c NSLOTS=1 -c NPRODS=1 -c NCONSS=1 BBsematest.cxl
-cxl -b -c NSLOTS=1 -c NPRODS=1 -c NCONSS=2 BBsematest.cxl
-cxl -b -c NSLOTS=1 -c NPRODS=2 -c NCONSS=0 BBsematest.cxl
-cxl    -c NSLOTS=1 -c NPRODS=2 -c NCONSS=1 BBsematest.cxl
-cxl    -c NSLOTS=1 -c NPRODS=2 -c NCONSS=2 BBsematest.cxl
-cxl -b -c NSLOTS=1 -c NPRODS=2 -c NCONSS=3 BBsematest.cxl
-cxl    -c NSLOTS=2 -c NPRODS=1 -c NCONSS=0 BBsematest.cxl
-cxl    -c NSLOTS=2 -c NPRODS=1 -c NCONSS=1 BBsematest.cxl
-cxl -b -c NSLOTS=2 -c NPRODS=1 -c NCONSS=2 BBsematest.cxl
-cxl    -c NSLOTS=2 -c NPRODS=2 -c NCONSS=0 BBsematest.cxl
-cxl    -c NSLOTS=2 -c NPRODS=2 -c NCONSS=1 BBsematest.cxl
-cxl    -c NSLOTS=2 -c NPRODS=2 -c NCONSS=2 BBsematest.cxl
-cxl -b -c NSLOTS=2 -c NPRODS=2 -c NCONSS=3 BBsematest.cxl
-cxl -b -c NSLOTS=2 -c NPRODS=3 -c NCONSS=0 BBsematest.cxl
-cxl    -c NSLOTS=2 -c NPRODS=3 -c NCONSS=1 BBsematest.cxl
-cxl    -c NSLOTS=2 -c NPRODS=3 -c NCONSS=2 BBsematest.cxl
-cxl    -c NSLOTS=2 -c NPRODS=3 -c NCONSS=3 BBsematest.cxl
+harmony -b -c NSLOTS=0 -c NPRODS=1 -c NCONSS=1 BBsematest.hny
+harmony -b -c NSLOTS=1 -c NPRODS=0 -c NCONSS=1 BBsematest.hny
+harmony    -c NSLOTS=1 -c NPRODS=1 -c NCONSS=0 BBsematest.hny
+harmony    -c NSLOTS=1 -c NPRODS=1 -c NCONSS=1 BBsematest.hny
+harmony -b -c NSLOTS=1 -c NPRODS=1 -c NCONSS=2 BBsematest.hny
+harmony -b -c NSLOTS=1 -c NPRODS=2 -c NCONSS=0 BBsematest.hny
+harmony    -c NSLOTS=1 -c NPRODS=2 -c NCONSS=1 BBsematest.hny
+harmony    -c NSLOTS=1 -c NPRODS=2 -c NCONSS=2 BBsematest.hny
+harmony -b -c NSLOTS=1 -c NPRODS=2 -c NCONSS=3 BBsematest.hny
+harmony    -c NSLOTS=2 -c NPRODS=1 -c NCONSS=0 BBsematest.hny
+harmony    -c NSLOTS=2 -c NPRODS=1 -c NCONSS=1 BBsematest.hny
+harmony -b -c NSLOTS=2 -c NPRODS=1 -c NCONSS=2 BBsematest.hny
+harmony    -c NSLOTS=2 -c NPRODS=2 -c NCONSS=0 BBsematest.hny
+harmony    -c NSLOTS=2 -c NPRODS=2 -c NCONSS=1 BBsematest.hny
+harmony    -c NSLOTS=2 -c NPRODS=2 -c NCONSS=2 BBsematest.hny
+harmony -b -c NSLOTS=2 -c NPRODS=2 -c NCONSS=3 BBsematest.hny
+harmony -b -c NSLOTS=2 -c NPRODS=3 -c NCONSS=0 BBsematest.hny
+harmony    -c NSLOTS=2 -c NPRODS=3 -c NCONSS=1 BBsematest.hny
+harmony    -c NSLOTS=2 -c NPRODS=3 -c NCONSS=2 BBsematest.hny
+harmony    -c NSLOTS=2 -c NPRODS=3 -c NCONSS=3 BBsematest.hny
 \end{verbatim}
@@
 \caption{Script for testing producer/consumer synchronization
-(assuming the code is in file \texttt{BBsematest.cxl}).}
+(assuming the code is in file \texttt{BBsematest.hny}).}
 \label{fig:pcscript}
@@
 \begin{code}
-\cxlsource{BBsemadata}
-\end{code}
-\caption{\cxllink{code/BBsemadata.cxl} Test whether the bounded buffer delivers the right data in the
+\harmonysource{BBsemadata}
+\end{code}
+\caption{\harmonylink{code/BBsemadata.hny} Test whether the bounded buffer delivers the right data in the
 right order.}
@@
 \autoref{fig:bbtest} invokes the code of \autoref{fig:boundedbuffer}.
-We can run this through CXL, but what would it test?  There are no
+We can run this through Harmony, but what would it test?  There are no
 assert statements in the code.
@@
 \begin{code}
-\cxlsource{gpu}
-\end{code}
-\caption{\cxllink{code/gpu.cxl} A thread-unsafe GPU allocator.}
+\harmonysource{gpu}
+\end{code}
+\caption{\harmonylink{code/gpu.hny} A thread-unsafe GPU allocator.}
 \label{fig:gpu}
@@
 \begin{code}
-\cxlsource{RWsbs}
-\end{code}
-\caption{\cxllink{code/RWsbs.cxl} Reader/Writer Lock implementation using Split Binary Semaphores.}
+\harmonysource{RWsbs}
+\end{code}
+\caption{\harmonylink{code/RWsbs.hny} Reader/Writer Lock implementation using Split Binary Semaphores.}
 \label{fig:RWsplitsema}
@@
 \begin{code}
-\cxlsource{busywait}
+\harmonysource{busywait}
 \end{code}
@@
 {\small
-\cxlsource{RWfair}
+\harmonysource{RWfair}
 }
 \end{code}
-\caption{\cxllink{code/RWfair.cxl} Reader/Writer Lock SBS implementation addressing fairness.}
+\caption{\harmonylink{code/RWfair.hny} Reader/Writer Lock SBS implementation addressing fairness.}
 \label{fig:RWfair}
@@
 {\small
-\cxlsource{RWqueue}
+\harmonysource{RWqueue}
 }
 \end{code}
-\caption{\cxllink{code/RWqueue.cxl} Reader/Writer Lock SBS implementation using a queue.}
+\caption{\harmonylink{code/RWqueue.hny} Reader/Writer Lock SBS implementation using a queue.}
 \label{fig:RWqueue}
@@
 \begin{code}
-\cxlsource{RWqtest}
-\end{code}
-\caption{\cxllink{code/RWqtest.cxl} Test program for \autoref{fig:RWqueue}.}
+\harmonysource{RWqtest}
+\end{code}
+\caption{\harmonylink{code/RWqtest.hny} Test program for \autoref{fig:RWqueue}.}
 \label{fig:RWqtest}
@@
 If $c$ is a pointer to a condition variable, then the semantics of the
-operations, in CXL, are as follows:
-
-\begin{code}
-\cxlsource{hoaremon}
+operations, in Harmony, are as follows:
+
+\begin{code}
+\harmonysource{hoaremon}
 \end{code}
@@
 
-% \chapter{Mesa Condition Variables in CXL}
+% \chapter{Mesa Condition Variables in Harmony}
 % \label{ch:mesa}
@@
 \begin{figure}
-\cxlsource{mesa}
+\harmonysource{mesa}
 \caption{Implementation of condition variables in the \texttt{synch} module.}
@@
 \begin{code}
-\cxlsource{RWcv}
-\end{code}
-\caption{\cxllink{code/RWcv.cxl} Reader/Writer Lock implementation using Mesa-style condition variables.}
+\harmonysource{RWcv}
+\end{code}
+\caption{\harmonylink{code/RWcv.hny} Reader/Writer Lock implementation using Mesa-style condition variables.}
 \label{fig:RWcv}
@@
 
-The CXL \texttt{synch} module supports Mesa condition variables.
+The Harmony \texttt{synch} module supports Mesa condition variables.
 However, like C, Java, and Python,
@@
 Waking up too many processes may lead to some inefficiency, but waking up to few
-may cause the application to get stuck.  CXL can be particularly helpful here, as
+may cause the application to get stuck.  Harmony can be particularly helpful here, as
 it examines every corner case.  You can safely try to replace each \texttt{notifyAll}
@@
 \begin{code}
-\cxlsource{Diners}
-\end{code}
-\caption{\cxllink{code/Diners.cxl} Dining Philosophers.}
+\harmonysource{Diners}
+\end{code}
+\caption{\harmonylink{code/Diners.hny} Dining Philosophers.}
 \label{fig:diners}
@@
 
-\autoref{fig:diners} implements the dining philosophers in CXL, using a
+\autoref{fig:diners} implements the dining philosophers in Harmony, using a
 process for each philosopher and a lock for each fork.  If you
-run it, CXL complains that the execution may not terminate, with all five
+run it, Harmony complains that the execution may not terminate, with all five
 processes being blocked trying to acquire the lock.
@@
 \begin{code}
-\cxlsource{DinersCV}
-\end{code}
-\caption{\cxllink{code/DinersCV.cxl} Dining Philosophers that grab both forks at the same time.}
+\harmonysource{DinersCV}
+\end{code}
+\caption{\harmonylink{code/DinersCV.hny} Dining Philosophers that grab both forks at the same time.}
 \label{fig:dinerscv}
@@
 \begin{code}
-\cxlsource{baddblwait}
+\harmonysource{baddblwait}
 \end{code}
@@
 \item Can you see why this does not work?  What can go wrong?
-\item Run it through CXL in case you are not sure!
+\item Run it through Harmony in case you are not sure!
 \end{itemize}
@@
 \begin{code}
-\cxlsource{dinersfix}
+\harmonysource{dinersfix}
 \end{code}
@@
 \begin{code}
-\cxlsource{DinersAvoid}
-\end{code}
-\caption{\cxllink{code/DinersAvoid.cxl} Dining Philosophers that carefully avoid getting into a deadlock
+\harmonysource{DinersAvoid}
+\end{code}
+\caption{\harmonylink{code/DinersAvoid.hny} Dining Philosophers that carefully avoid getting into a deadlock
 scenario.}
@@
 \begin{code}
-\cxlsource{bank}
-\end{code}
-\caption{\cxllink{code/bank.cxl} Bank accounts.}
+\harmonysource{bank}
+\end{code}
+\caption{\harmonylink{code/bank.hny} Bank accounts.}
 \label{fig:bank}
@@
 \begin{code}
-\cxlsource{actor}
-\end{code}
-\caption{\cxllink{code/actor.cxl} A producer/consumer actor.}
+\harmonysource{actor}
+\end{code}
+\caption{\harmonylink{code/actor.hny} A producer/consumer actor.}
 \label{fig:actor}
@@
 \begin{code}
-\cxlsource{actortest}
-\end{code}
-\caption{\cxllink{code/actortest.cxl} Test code for producer/consumer actor.}
+\harmonysource{actortest}
+\end{code}
+\caption{\harmonylink{code/actortest.hny} Test code for producer/consumer actor.}
 \label{fig:actortest}
@@
 
-\autoref{fig:actor} shows a CXL solution to the
+\autoref{fig:actor} shows a Harmony solution to the
 ``multiple producer / multiple consumer problem'' (with an unbounded buffer)
@@
 to the queue on which it receives messages.  \textit{nrequests} is the total
-number of requests that it will handle.  We need the latter because in CXL
+number of requests that it will handle.  We need the latter because in Harmony
 all processes are required to terminate.  In a more realistic implementation,
@@
 \begin{code}
-\cxlsource{abp}
-\end{code}
-\caption{\cxllink{code/abp.cxl} Alternating Bit Protocol.}
+\harmonysource{abp}
+\end{code}
+\caption{\harmonylink{code/abp.hny} Alternating Bit Protocol.}
 \label{fig:abp}
@@
 \begin{code}
-\cxlsource{abptest}
-\end{code}
-\caption{\cxllink{code/abptest.cxl} Test code for alternating bit protocol.}
+\harmonysource{abptest}
+\end{code}
+\caption{\harmonylink{code/abptest.hny} Test code for alternating bit protocol.}
 \label{fig:abptest}
@@
 \begin{code}
-\cxlsource{hw}
-\end{code}
-\caption{\cxllink{code/hw.cxl} Non-blocking queue.}
+\harmonysource{hw}
+\end{code}
+\caption{\harmonylink{code/hw.hny} Non-blocking queue.}
 \label{fig:hw}
@@
 \begin{code}
-\cxlsource{lst1}
-\end{code}
-\caption{\cxllink{code/lst1.cxl} List with non-blocking read operations, part 1.}
+\harmonysource{lst1}
+\end{code}
+\caption{\harmonylink{code/lst1.hny} List with non-blocking read operations, part 1.}
 \label{fig:lst1}
@@
 \begin{code}
-\cxlsource{lst2}
-\end{code}
-\caption{\cxllink{code/lst2.cxl} List with non-blocking read operations, part 2.}
+\harmonysource{lst2}
+\end{code}
+\caption{\harmonylink{code/lst2.hny} List with non-blocking read operations, part 2.}
 \label{fig:lst2}
@@
 \begin{code}
-\cxlsource{barrier}
-\end{code}
-\caption{\cxllink{code/barrier.cxl} Barrier synchronization with semaphores.}
+\harmonysource{barrier}
+\end{code}
+\caption{\harmonylink{code/barrier.hny} Barrier synchronization with semaphores.}
 \label{fig:barrier}
@@
 
-\autoref{ch:cxlmachine} provides an introduction to CXL values.
-Below is a complete list of CXL value types with examples:
+\autoref{ch:harmonymachine} provides an introduction to Harmony values.
+Below is a complete list of Harmony value types with examples:
 
@@
 Tuples, lists, strings, and bags are all special cases of dictionaries.
-Both tuples and lists map indexes (starting at 0) to CXL values.
+Both tuples and lists map indexes (starting at 0) to Harmony values.
 Their format is either \texttt{($e$, $e$, ..., $e$,)} or
@@
 
-All CXL values are ordered with respect to one another.  First they
+All Harmony values are ordered with respect to one another.  First they
 are ordered by type according to the table above.
@@
 
-CXL currently supports the following operators:
+Harmony currently supports the following operators:
 
@@
 \hline
-\texttt{$e$ == $e$}, \texttt{$e$ != $e$} & two CXL values of the same type \\
-\hline
-\texttt{$e$ < $e$, $e$ <= $e$} & any two CXL values \\
-\hline
-\texttt{$e$ > $e$, $e$ >= $e$} & any two CXL values \\
+\texttt{$e$ == $e$}, \texttt{$e$ != $e$} & two Harmony values of the same type \\
+\hline
+\texttt{$e$ < $e$, $e$ <= $e$} & any two Harmony values \\
+\hline
+\texttt{$e$ > $e$, $e$ >= $e$} & any two Harmony values \\
 \hline
@@
 \hline
-\texttt{$e$ [not] in $e$} & first is any CXL value, second is set \\
+\texttt{$e$ [not] in $e$} & first is any Harmony value, second is set \\
 \hline
@@
 \hline
-\texttt{hash $e$} & any CXL value \\
+\texttt{hash $e$} & any Harmony value \\
 \hline
@@
 
-CXL also supports the following comprehensions:
+Harmony also supports the following comprehensions:
 
@@
 
-CXL currently supports the following statements:
+Harmony currently supports the following statements:
 
@@
 
-CXL does not support Python-style iterators.
+Harmony does not support Python-style iterators.
 To iterate through a dictionary, list, or tuple, use:
@@
 \chapter{List of Machine Instructions}
-\label{ap:cxlbytecode}
+\label{ap:harmonybytecode}
 
@@
 The instruction then computes the address of \texttt{\^}$v_1~v_2~...~v_n$.
-\item Even though CXL code does not allow taking addresses of process variables, both
+\item Even though Harmony code does not allow taking addresses of process variables, both
 shared and process variables can have addresses.
@@
 \hline
-stack & a list of CXL values \\
-\hline
-register & a CXL dictionary mapping atoms (representing process variables) to CXL values \\
+stack & a list of Harmony values \\
+\hline
+register & a Harmony dictionary mapping atoms (representing process variables) to Harmony values \\
 \hline
@@
 atoms to arbitrary values.  The atoms correspond to the variable names
-in a CXL program.
+in a Harmony program.
 \end{itemize}
@@
 
-\chapter{The CXL Virtual Machine}
-
-The CXL Virtual Machine (CVM) has the following state:
+\chapter{The Harmony Virtual Machine}
+
+The Harmony Virtual Machine (HVM) has the following state:
 
@@
 \hline
-code & a list of CVM machine instructions \\
+code & a list of HVM machine instructions \\
 \hline
@@
 A \emph{micro step}
-is the execution of a single CVM machine instruction
+is the execution of a single HVM machine instruction
 by a context.
 Each micro step generates a new state.
-When there are multiple contexts, the CVM can interleave them.
+When there are multiple contexts, the HVM can interleave them.
 However, trying to interleave every microstep would be needlessly expensive,
@@
 instructions start a new macro step: \texttt{Load}, \texttt{Store},
-\texttt{AtomicInc}, and \texttt{Continue}.  The CVM
+\texttt{AtomicInc}, and \texttt{Continue}.  The HVM
 interleaves macro steps, not micro steps.  Like micro steps, each
@@
 
-Executing a CXL program results in a graph where the nodes are CXL
+Executing a Harmony program results in a graph where the nodes are Harmony
 states and the edges are macro steps.
@@
 First and foremost I would like to think my student Haobin Ni with
-whom I've had numerous discussions about the design of CXL.
-Haobin even contributed some code to the CXL compiler.
+whom I've had numerous discussions about the design of Harmony.
+Haobin even contributed some code to the Harmony compiler.
 
@@
 my colleague Fred Schneider.  He suggested I write this book after
-demonstrating CXL to him.
+demonstrating Harmony to him.
 
@@
 
-I first demonstrated CXL to the students in my CS6480 class on systems
+I first demonstrated Harmony to the students in my CS6480 class on systems
 and formal verification and
@@
 
-% I wrote the CXL compiler and this book in May and June of 2020,
+% I wrote the Harmony compiler and this book in May and June of 2020,
 % during the pandemic and the George Floyd demonstrations.

### 2020-07-06 book/paper.tex
Latest

@@
 version of the module.
+\item \label{ex:trylock} Implement \texttt{trylock}(\textit{lk}) as an additional
+interface to the \texttt{synch} and \texttt{synchS} modules.
+This interface is like \texttt{lock}(\textit{lk}) but never blocks.  It
+returns \texttt{True} if the lock was available (and now acquired) or \texttt{False}
+if the lock was already taken.
 \end{problems}
@@
 for three processes.
-\item Using busy waiting, implement a ``bound lock'' that allows
+\item \label{ex:boundlock} Using busy waiting, implement a ``bound lock'' that allows
 up to \texttt{M} processes to acquire it at the same time.  A bound lock
@@
 % reader/writer locks can be instantiated.
-\item Write a test program for ``bound locks'' (as implemented in the
-exercises of the last chapter) that checks that up to \texttt{M} processes
+\item Write a test program for ``bound locks'' (Exercise~\ref{ex:boundlock})
+that checks that up to \texttt{M} processes
 can acquire the bound lock at the same time.
@@
 \begin{problems}
+\item Implement a ``try lock'' using Mesa condition variables
+(see also Exercise~\ref{ex:trylock}).  It should
+have the following API:
+\begin{enumerate}
+\item \textit{tl} = \texttt{TryLock}() \# \emph{allocate a try lock}
+\item \texttt{tlLock}(\&\textit{tl}) \# \emph{acquire a try lock}
+\item \texttt{tlTrylock}(\&\textit{tl}) \# \emph{attempt to acquire a try lock}
+\item \texttt{tlUnlock}(\&\textit{tl}) \# \emph{release a try lock}
+\end{enumerate}
+\noindent
+\texttt{tlTrylock} should not wait.
+Instead it should return \texttt{True} if the lock was successfully
+acquired and \texttt{False} if the lock was not available.
 \item Implement semaphores using Mesa condition variables.

### 2020-07-13 book/paper.tex
Latest
Improved question

@@
 Can you generalize Peterson's algorithm to more than two processes?
+\item Bonus question:
+Implement the
+\href{https://en.wikipedia.org/wiki/Lamport%27s_bakery_algorithm}{Bakery Algorithm}
+or \href{https://en.wikipedia.org/wiki/Dekker%27s_algorithm}{Dekker's Algorithm}.
 \end{problems}
@@
 and swap the values.
-\item Implement a spinlock using the atomic swap operation.
-\item Write out the invariants that need to hold and check them using Harmony.
+\item \label{ex:swap} Implement a spinlock using the atomic swap operation.
+\item For the solution to \autoref{ex:swap},
+write out the invariants that need to hold and check them using Harmony.
 \item People who use an ATM often first check their balance and then withdraw

### 2020-07-15 book/paper.tex
Clarification
Typo

@@
 If you invoke \texttt{m}($a$, $b$), the argument is ($a$, $b$),
-the tuple consisting of values $a$ and $a$.
+the tuple consisting of values $a$ and $b$.
 
@@
 \begin{enumerate}
-\item Fix the code in \autoref{fig:atm}.
+\item Fix the code in \autoref{fig:atm}.  Note, you should leave the
+customer code the same. You are only allowed to change the
+\texttt{atm\_} methods, and you cannot use the \textbf{atomic}
+statement.
 \item Is it necessary to obtain a lock in \texttt{atm\_check\_balance()}?

### 2020-07-19 book/paper.tex
Latest

@@
 and \emph{message passing} as an approach to synchronization.
+\item \autoref{ch:interrupts} discusses how to handle interrupts to a process,
+a problem closely related to synchronization among multiple processes.
 \item \autoref{ch:abp} presents a problem and a solution to the distributed
@@
 
+\chapter{Interrupts}
+\label{ch:interrupts}
+\index{interrupt}
+
+\begin{figure}
+\begin{code}
+\harmonysource{trap}
+\end{code}
+\caption{\harmonylink{code/trap.hny} How to use \textbf{trap}.}
+\label{fig:trap}
+\end{figure}
+
+\begin{figure}
+\begin{code}
+\harmonysource{trap2}
+\end{code}
+\caption{\harmonylink{code/trap2.hny} A race condition with interrupts.}
+\label{fig:trap}
+\end{figure}
+
+\begin{figure}
+\begin{code}
+\harmonysource{trap3}
+\end{code}
+\caption{\harmonylink{code/trap3.hny} Locks do not work with interrupts.}
+\label{fig:trap}
+\end{figure}
+
+\begin{figure}
+\begin{code}
+\harmonysource{trap4}
+\end{code}
+\caption{\harmonylink{code/trap4.hny} Disabling and enabling interrupts.}
+\label{fig:trap}
+\end{figure}
+
+Processes can be \emph{interrupted}.  An interrupt is a notification of some event
+such as a keystroke, a timer expiring, the reception of a network packet, the completion
+of a disk operation, and so on.  We distinguish \emph{interrupts} and \emph{exceptions}.
+An exception is caused by the process executing an invalid machine instruction such as
+divide-by-zero.  An interrupt is caused by some peripheral device.
+
+Harmony allows modeling interrupts using operator \texttt{trap}:
+\begin{code}
+\textbf{trap} \textit{handler} \textit{argument}
+\end{code}
+invokes \textit{handler argument} at some later, unspecified time.
+Only one of these asynchronous events can be outstanding at a time; a new call to
+\textbf{trap} overwrites any outstanding one.
+Figure~\ref{fig:trap} gives an example of how \textbf{trap} might be used.
+Here, the \textit{main}() process loops until the interrupt has occurred and
+the \textit{done} flag has been set.
+
 \chapter{Networking: Alternating Bit Protocol}
@@
 \hline
+Move $i$ & move stack element at offset $i$ to top of the stack \\
+\hline
 $n$-ary \textit{op} & apply $n$-ary operator \textit{op} to the top $n$ elements on the stack \\
@@
 StoreVar [$v$] & pop a value from the stack and store it in a process variable \\
-\hline
-Swap & swap the two values on top of the stack \\
 \hline

### 2020-07-19 book/paper.tex
Latest

@@
 \usepackage{etoolbox}
-\usepackage{bold-extra}
+% \usepackage{bold-extra}
 \usepackage{xcolor}
@@
 \caption{\harmonylink{code/trap2.hny} A race condition with interrupts.}
-\label{fig:trap}
+\label{fig:trap2}
 \end{figure}
@@
 \caption{\harmonylink{code/trap3.hny} Locks do not work with interrupts.}
-\label{fig:trap}
+\label{fig:trap3}
 \end{figure}
@@
 \caption{\harmonylink{code/trap4.hny} Disabling and enabling interrupts.}
-\label{fig:trap}
+\label{fig:trap4}
 \end{figure}
@@
 divide-by-zero.  An interrupt is caused by some peripheral device.
+\index{exception}
 
@@
 \textbf{trap} overwrites any outstanding one.
-Figure~\ref{fig:trap} gives an example of how \textbf{trap} might be used.
+\autoref{fig:trap} gives an example of how \textbf{trap} might be used.
 Here, the \textit{main}() process loops until the interrupt has occurred and
 the \textit{done} flag has been set.
+
+But now consider \autoref{fig:trap2}.  The difference with \autoref{fig:trap} is
+that both the \textit{main}() and \textit{handler}() methods increment \textit{count}.
+This is not unlike the example we gave in \autoref{fig:inc}, except that only a single
+process is involved now.  And, indeed, it suffers from a similar race condition; run
+it through Harmony to see for yourself.  If the interrupt occurs after \textit{main}()
+read \textit{count} (and thus still has value 0) but before \textit{main()} writes the
+updated value~1, then the interrupt handler will also read value~0 and write value~1.
+We say that the code in \autoref{fig:trap2} is not \textit{interrupt-safe} (as opposed
+to not being \textit{thread-safe}).
+\index{interrupt-safety}
+
+You would be excused if you wanted to solve the problem using locks, similar to
+\autoref{fig:incfixed}.  \autoref{fig:trap3} shows how one might go about this.
+But locks are intended to solve synchronization issues between multiple processes.
+If you run the code through Harmony, you will find that the code may not terminate.
+The issue is that a process can only acquire a lock once.  If the interrupt happens
+after \textit{main}() acquires the lock but before \textit{main}() releases it, the
+\textit{handler}() method will block trying to acquire the lock, even though
+it is executed the same process that holds the lock.
+
+Instead, the way one fixes interrupt-safety issues is through disabling interrupts
+temporarily.  In Harmony, this can be done by setting the \textit{interrupt level}
+of a process to \texttt{True} using the \textbf{setintlevel} interface.
+\autoref{fig:trap4} illustrates how this is done.
+Note that it is not necessary to change the interrupt level during servicing an
+interrupt, because it is automatically set to \texttt{True} upon entry to the interrupt
+handler and restored to \texttt{False} upon exit.
+It is important that the \textit{main}() code re-enables interrupts after incrementing
+\textit{count}.  What would happen if \textit{main}() left interrupt disabled?
+
+\textbf{setintlevel}(\textit{il}) sets the interrupt level to \textit{il} and returns
+the prior interrupt level.  Returning the old level is handy when writing interrupt-safe
+methods that can be called from ordinary code as well as from an interrupt handler.
+\autoref{fig:trap5} shows how one might write a interrupt-safe method
+to increment the counter.
+
+\begin{figure}
+\begin{code}
+\harmonysource{trap5}
+\end{code}
+\caption{\harmonylink{code/trap5.hny} Example of an interrupt-safe method.}
+\label{fig:trap5}
+\end{figure}
+
+It will often be necessary to write code that is both interrupt-safe \emph{and}
+thread-safe.  As you might expect, this involves both managing locks and
+interrupt levels.
 

### 2020-07-19 book/paper.tex
Latest
Updated some appendices
Added an exercise to the interrupts chapter

@@
 
+\begin{figure}
+\begin{code}
+\harmonysource{trap6}
+\end{code}
+\caption{\harmonylink{code/trap6.hny} Code that is both interrupt-safe and thread-safe.}
+\label{fig:trap6}
+\end{figure}
+
 It will often be necessary to write code that is both interrupt-safe \emph{and}
@@
 interrupt levels.
+To increment \textit{count}, the interrupt level must be \textit{True} and
+\textit{countlock} must be held.
+\autoref{fig:trap6} gives an example of how this might be done.
+The important rule to remember is for a process to disable interrupts \emph{before}
+attempting to acquire a lock.
+
+Try moving \texttt{lock}() to the beginning of \texttt{increment}()
+and \texttt{unlock}() to the end of \texttt{increment}() and see what happens.
+While Harmony will only report one faulty run, this incorrect code can lead to
+the assertion failing as well as processes getting blocked indefinitely.
+
+\section*{Exercises}
+\begin{problems}
+\item Make methods \texttt{produce}() and \texttt{consume}() in \autoref{fig:boundedbuffer}
+both thread-safe and interrupt-safe.
+\end{problems}
 
@@
 \hline
-\texttt{hash $e$} & any Harmony value \\
-\hline
 \texttt{choose $e$} & a set \\
@@
 \hline
+\texttt{setintlevel} \textit{e} & \textit{e} is a boolean \\
+\hline
+\texttt{stop} \textit{lv} & \textit{lv} is an lvalue \\
+\hline
 \texttt{\&}\textit{lv} & \textit{lv} is an lvalue \\
-\hline
-\texttt{stop} \textit{lv} & \textit{lv} is an lvalue \\
 \hline
@@
 
-\vspace{1em}
+{\small
 \begin{tabular}{|l|l|}
@@
 \hline
+SetIntLevel & pop $e$, set interrupt level to $e$, and push old interrupt level \\
+\hline
 Spawn & pop tag, argument, and method and spawn a new context \\
@@
 \hline
+Trap & pop interrupt argument and method \\
+\hline
 \end{tabular}
-\vspace{1em}
-
+}
+
+\newpage
 Clarifications:

### 2020-07-20 book/paper.tex
Added some stuff
Improved interrupt chapter

@@
 An exception is caused by the process executing an invalid machine instruction such as
-divide-by-zero.  An interrupt is caused by some peripheral device.
+divide-by-zero.  An interrupt is caused by some peripheral device and can be handled in
+Harmony; an exception is an error.
 \index{exception}
 
-Harmony allows modeling interrupts using operator \texttt{trap}:
+Harmony allows modeling interrupts using the \texttt{trap} statement:
 \begin{code}
@@
 it through Harmony to see for yourself.  If the interrupt occurs after \textit{main}()
-read \textit{count} (and thus still has value 0) but before \textit{main()} writes the
+reads \textit{count} (and thus still has value 0) but before \textit{main()} writes the
 updated value~1, then the interrupt handler will also read value~0 and write value~1.
@@
 \textit{handler}() method will block trying to acquire the lock, even though
-it is executed the same process that holds the lock.
+it is being acquired by the same process that already holds the lock.
 
@@
 \autoref{fig:trap6} gives an example of how this might be done.
-The important rule to remember is for a process to disable interrupts \emph{before}
+One important rule to remember is for a process to disable interrupts \emph{before}
 attempting to acquire a lock.
 
-Try moving \texttt{lock}() to the beginning of \texttt{increment}()
-and \texttt{unlock}() to the end of \texttt{increment}() and see what happens.
+Try moving \texttt{lock}() to the beginning of the \texttt{increment} method
+and \texttt{unlock}() to the end of \texttt{increment} and see what happens.
 While Harmony will only report one faulty run, this incorrect code can lead to
@@
 
+There is another important rule to keep in mind.  Just locks should never be held
+for long, interrupts should never be disabled for long.  With locks the issue is to
+maximize concurrent performance.  The interrupts the issue is fast response to
+asynchronous events.  Because interrupts may be disabled only briefly, interrupt
+handlers must run quickly and cannot wait for other events.  So it is ok to invoke
+synchronization calls such as \texttt{V} and \texttt{notify}, but calls such as
+\texttt{P} and \texttt{wait} should only be used if it's known that they will not
+block for long.  Informally, interrupt handlers can be
+\emph{producers} but not \emph{consumers} of synchronization events.
+
 \section*{Exercises}
 \begin{problems}
-\item Make methods \texttt{produce}() and \texttt{consume}() in \autoref{fig:boundedbuffer}
-both thread-safe and interrupt-safe.
+\item The \texttt{produce} method in \autoref{fig:boundedbuffer} cannot be used
+from interrupt handlers for two reasons: (1) it is not interrupt-safe, and (2)
+it may block for a long time if the buffer is full.  Yet, it would be useful if,
+say, a keyboard interrupt handler could place an event on a shared queue.  Implement
+a new method \texttt{iproduce}(\textit{item}) that is interrupt-safe and does not
+block.  Instead, it should return \texttt{False} if the buffer is full and \texttt{True}
+if the item was successfully enqueued.
 \end{problems}
@@
 $e$ is an expression, $t$ is a tag (an expression) \\
+\hline
+\texttt{trap m $e$} & \texttt{m} is a method and $e$ is an expression \\
 \hline

### 2020-07-20 book/paper.tex
Latest

@@
 \item The \texttt{produce} method in \autoref{fig:boundedbuffer} cannot be used
-from interrupt handlers for two reasons: (1) it is not interrupt-safe, and (2)
+in interrupt handlers for two reasons: (1) it is not interrupt-safe, and (2)
 it may block for a long time if the buffer is full.  Yet, it would be useful if,

### 2020-07-23 book/paper.tex
Updated syntax
Added bitwise operators

@@
 $ harmony triangle.hny
-#states = 13
+#states = 13 diameter = 1
 no issues found
@@
 $ harmony -c N=100 triangle.hny
-#states = 103
+#states = 103 diameter = 1
 no issues found
@@
 \end{code}
-\caption{\harmonylink{code/csbarebones.hny} Harmony model of a critical section.}
+\caption{\harmonylink{code/cs.hny} Harmony model of a critical section.}
 \label{fig:cs}
@@
 to shared variables.
-If $x$ is a shared variable, then the expression \&$x$
+If $x$ is a shared variable, then the expression ?$x$
  is a pointer to $x$ (also known as the \emph{address}
@@
 Conversely, if $p$ is a pointer to a shared variable, then the
-expression \texttt{\^{}}$p$ is the value of the shared variable.
-In particular, \texttt{\^{}\&}$x$ == $x$.
+expression \texttt{!{}}$p$ is the value of the shared variable.
+In particular, \texttt{!{}?}$x$ == $x$.
 This is similar to how C pointers work.
@@
 a pointer to a mutex and the process identifier (0 or 1).
-\texttt{(\^{}}\textit{pm}\texttt{.turn} is the value of the \texttt{.turn} key
+\texttt{(!{}}\textit{pm}\texttt{.turn} is the value of the \texttt{.turn} key
 in the dictionary that \textit{pm} points to. 
 Note that the parentheses in this expression
-are needed, as \texttt{\^{}}\textit{pm}\texttt{.turn} would wrongly evaluate to
-\texttt{\^{}}(\textit{pm}\texttt{.turn}).
+are needed, as \texttt{!{}}\textit{pm}\texttt{.turn} would wrongly evaluate to
+\texttt{!{}}(\textit{pm}\texttt{.turn}).
 
@@
 To see that invariant $\mathcal{I}_1$ is maintained, note that
-\texttt{\^{}}$p$ == \texttt{True} upon entry of \texttt{tas}
+\texttt{!{}}$p$ == \texttt{True} upon entry of \texttt{tas}
 (because of the condition on the \textbf{while} loop that the
@@
 \begin{enumerate}
-\item \texttt{\^{}}$s$ is \texttt{False} upon entry to \texttt{tas}.
-Then upon exit \texttt{\^{}}$p$ == \texttt{False} and \texttt{\^{}}$s$ == \texttt{True},
+\item \texttt{!{}}$s$ is \texttt{False} upon entry to \texttt{tas}.
+Then upon exit \texttt{!{}}$p$ == \texttt{False} and \texttt{!{}}$s$ == \texttt{True},
 maintaining the invariant.
-\item \texttt{\^{}}$s$ is \texttt{True} upon entry to \texttt{tas}.
+\item \texttt{!{}}$s$ is \texttt{True} upon entry to \texttt{tas}.
 Then upon exit nothing has changed, maintaining the invariant.
@@
 are all~0.  To maintain the invariant, each process alternates \texttt{P} and
-\texttt{V} operations, starting with \texttt{P}(\&\textit{mutex}) when it tries to
+\texttt{V} operations, starting with \texttt{P}(?\textit{mutex}) when it tries to
 enter the CCS, and ending with a \texttt{V} operation on one of the semaphores
@@
 \begin{figure}
+\begin{code}
 \harmonysource{mesa}
+\end{code}
 \caption{Implementation of condition variables in the \texttt{synch} module.}
@@
 
-In \texttt{release\_rlock}, notice that \texttt{notify}(\&\textit{wcond}) is invoked when there
+In \texttt{release\_rlock}, notice that \texttt{notify}(?\textit{wcond}) is invoked when there
 are no readers left, \emph{without} checking if there are writers waiting to enter.
@@
 
-\texttt{release\_wlock} executes \texttt{notifyAll}(\&\textit{wcond}) as well as
-\texttt{notify}(\&\textit{wcond)}.
+\texttt{release\_wlock} executes \texttt{notifyAll}(?\textit{wcond}) as well as
+\texttt{notify}(?\textit{wcond)}.
 Again, because we do not keep track of the number of waiting readers or writers, we
@@
 \item \textit{tl} = \texttt{TryLock}() \# \emph{allocate a try lock}
-\item \texttt{tlLock}(\&\textit{tl}) \# \emph{acquire a try lock}
-\item \texttt{tlTrylock}(\&\textit{tl}) \# \emph{attempt to acquire a try lock}
-\item \texttt{tlUnlock}(\&\textit{tl}) \# \emph{release a try lock}
+\item \texttt{tlLock}(?\textit{tl}) \# \emph{acquire a try lock}
+\item \texttt{tlTrylock}(?\textit{tl}) \# \emph{attempt to acquire a try lock}
+\item \texttt{tlUnlock}(?\textit{tl}) \# \emph{release a try lock}
 \end{enumerate}
@@
 a message $m$ to
-\texttt{\^{}}\textit{pchan}, where \textit{pchan} is either \&\textit{s\_chan}
-or \&\textit{r\_chan}.
+\texttt{!{}}\textit{pchan}, where \textit{pchan} is either ?\textit{s\_chan}
+or ?\textit{r\_chan}.
 The method places either $m$ (to model a successful send)
-or \texttt{()} (to model loss) in \texttt{\^{}}\textit{pchan}.
+or \texttt{()} (to model loss) in \texttt{!{}}\textit{pchan}.
 The use of the \textit{reliable} flag will be explained later.
-\texttt{net\_recv}(\textit{pchan}) models checking \texttt{\^{}pchan} for
+\texttt{net\_recv}(\textit{pchan}) models checking \texttt{!{}pchan} for
 the next message.
@@
 \begin{itemize}
-\item \texttt{inc}($p$): atomically increments \texttt{\^{}}$p$ and returns
+\item \texttt{inc}($p$): atomically increments \texttt{!{}}$p$ and returns
 the old value;
-\item \texttt{swap}($p$): sets \texttt{\^{}}$p$ to \texttt{()} and returns
+\item \texttt{swap}($p$): sets \texttt{!{}}$p$ to \texttt{()} and returns
 the old value.
@@
 
-Method \texttt{produce}(\textit{item}) uses \texttt{inc}(\&\textit{back})
+Method \texttt{produce}(\textit{item}) uses \texttt{inc}(?\textit{back})
 to allocate
@@
 \hline
-Address & \texttt{\&lock, \&flags[2], None} \\
+Address & \texttt{?lock, ?flags[2], None} \\
 \hline
@@
 
+{\small
+\begin{tabular}{|l|l|}
+\hline
+\texttt{$e$ == $e$}, \texttt{$e$ != $e$} & two Harmony values of the same type \\
+\hline
+\texttt{$e$ < $e$, $e$ <= $e$} & any two Harmony values \\
+\hline
+\texttt{$e$ > $e$, $e$ >= $e$} & any two Harmony values \\
+\hline
+\texttt{$e$ and $e$ and ...} & two or more booleans \\
+\hline
+\texttt{$e$ or $e$ or ...} & two or more booleans \\
+\hline
+\texttt{not $e$} & a boolean \\
+\hline
+\texttt{-$e$, \string~$e$} & an integer \\
+\hline
+\texttt{$e$ + $e$ + ...} & two or more integers, tuples, or lists \\
+\hline
+\texttt{$e$ - $e$} & two integers or sets \\
+\hline
+\texttt{$e$ * $e$ * ...} & two or more integers \\
+\hline
+\texttt{$e$ \& $e$ \& * ...} & two or more integers or sets \\
+\hline
+\texttt{$e$ | $e$ | * ...} & two or more integers or sets \\
+\hline
+\texttt{$e$ \string^ $e$ \string^ * ...} & two or more integers or sets \\
+\hline
+\texttt{$e$ / $e$} & two integers (integer division) \\
+\hline
+\texttt{$e$ \% $e$} & two integers (division remainder) \\
+\hline
+\texttt{$\{ e..e \}$} & two integers \\
+\hline
+\texttt{$e$ [not] in $e$} & first is any Harmony value, second is set \\
+\hline
+\texttt{$e$ if $e$ else $e$} & middle $e$ must be a boolean \\
+\hline
+\texttt{!{}$e$} & an address \\
+\hline
+\texttt{choose $e$} & a set \\
+\hline
+\texttt{min $e$, max $e$} & a set \\
+\hline
+\texttt{cardinality $e$} & a set \\
+\hline
+\texttt{keys $e$, len $e$} & a dictionary (includes tuple or list) \\
+\hline
+\texttt{bagsize $e$} & a bag \\
+\hline
+\texttt{atLabel $e$} & atom (corresponding to a label) \\
+\hline
+\texttt{nametag $e$} & $e$ must be \texttt{()} \\
+\hline
+\texttt{processes $e$} & $e$ must be \texttt{()} \\
+\hline
+\texttt{setintlevel} \textit{e} & \textit{e} is a boolean \\
+\hline
+\texttt{stop} \textit{lv} & \textit{lv} is an lvalue \\
+\hline
+\texttt{?}\textit{lv} & \textit{lv} is an lvalue \\
+\hline
+\end{tabular}
+}
 \vspace{1em}
-\begin{tabular}{|l|l|}
-\hline
-\texttt{$e$ == $e$}, \texttt{$e$ != $e$} & two Harmony values of the same type \\
-\hline
-\texttt{$e$ < $e$, $e$ <= $e$} & any two Harmony values \\
-\hline
-\texttt{$e$ > $e$, $e$ >= $e$} & any two Harmony values \\
-\hline
-\texttt{$e$ and $e$ and ...} & two or more booleans \\
-\hline
-\texttt{$e$ or $e$ or ...} & two or more booleans \\
-\hline
-\texttt{not $e$} & a boolean \\
-\hline
-\texttt{-$e$} & an integer \\
-\hline
-\texttt{$e$ + $e$ + ...} & two or more integers, sets, or lists \\
-\hline
-\texttt{$e$ - $e$} & two integers or sets \\
-\hline
-\texttt{$e$ * $e$ * ...} & two or more integers or sets \\
-\hline
-\texttt{$e$ / $e$} & two integers (integer division) \\
-\hline
-\texttt{$e$ \% $e$} & two integers (division remainder) \\
-\hline
-\texttt{$e$~..~$e$} & two integers \\
-\hline
-\texttt{$e$ [not] in $e$} & first is any Harmony value, second is set \\
-\hline
-\texttt{$e$ if $e$ else $e$} & middle $e$ must be a boolean \\
-\hline
-\texttt{\^{}$e$} & an address \\
-\hline
-\texttt{choose $e$} & a set \\
-\hline
-\texttt{min $e$, max $e$} & a set \\
-\hline
-\texttt{cardinality $e$} & a set \\
-\hline
-\texttt{keys $e$, len $e$} & a dictionary (includes tuple or list) \\
-\hline
-\texttt{bagsize $e$} & a bag \\
-\hline
-\texttt{atLabel $e$} & atom (corresponding to a label) \\
-\hline
-\texttt{nametag $e$} & $e$ must be \texttt{()} \\
-\hline
-\texttt{processes $e$} & $e$ must be \texttt{()} \\
-\hline
-\texttt{setintlevel} \textit{e} & \textit{e} is a boolean \\
-\hline
-\texttt{stop} \textit{lv} & \textit{lv} is an lvalue \\
-\hline
-\texttt{\&}\textit{lv} & \textit{lv} is an lvalue \\
-\hline
-\end{tabular}
-\vspace{1em}
-
-\texttt{+} computes the union when applied to sets and
-the concatenation when applied to lists or tuples.
+
+\texttt{+} concatenates when applied to lists or tuples.
 \texttt{-} computes set difference when applied to two sets.
-\texttt{*} computes the intersection when applied to sets.
-\texttt{$x$..$y$} computes the set consisting of the integers $x$
+\texttt{\&} computes the intersection when applied to sets.
+\texttt{|} computes the union when applied to sets.
+\texttt{\^} computes $(s \cup t) \backslash (s \cap t)$ when applied to sets $s$ and $t$.
+\texttt{$\{ x .. y \}$} computes the set consisting of the integers $x$
 through $y$ inclusive.  It returns the empty set if $x > y$.
@@
 variable.  It can also be indexed.  Examples include
-\textit{var}, \textit{var}[$e$], and \texttt{\^{}}\textit{p}.
+\textit{var}, \textit{var}[$e$], and \texttt{!{}}\textit{p}.
 
@@
 \hline
-Set comprehension & \texttt{\{ f($e$) for $e$ in $s$ \}} \\
-\hline
-List comprehension & \texttt{[ f($e$) for $e$ in $s$ ]} \\
-\hline
-Dict comprehension & \texttt{dict\{ f($e$) for $e$ in $s$ \}} \\
+Set comprehension & \texttt{\{ f($v$) for $v$ in $s$ \}} \\
+\hline
+List comprehension & \texttt{[ f($v$) for $v$ in $s$ ]} \\
+\hline
+Dict comprehension & \texttt{dict\{ f($v$) for $v$ in $s$ \}} \\
 \hline
@@
 \hline
-\texttt{bagAdd}(\textit{pb}, $e$) & add one copy of $e$ to bag \texttt{\^{}}\textit{pb}\\
-\hline
-\texttt{bagRemove}(\textit{pb}, $e$) & remove one copy of $e$ from bag \texttt{\^{}}\textit{pb}\\
+\texttt{bagAdd}(\textit{pb}, $e$) & add one copy of $e$ to bag \texttt{!{}}\textit{pb}\\
+\hline
+\texttt{bagRemove}(\textit{pb}, $e$) & remove one copy of $e$ from bag \texttt{!{}}\textit{pb}\\
 \hline
@@
 \hline
-tas(\textit{lk}) & test-and-set on \texttt{\^{}}\textit{lk} \\
+tas(\textit{lk}) & test-and-set on \texttt{!{}}\textit{lk} \\
 \hline
@@
 \hline
-lock(\textit{lk}) & acquire \texttt{\^{}}\textit{lk} \\
-\hline
-unlock(\textit{lk}) & release \texttt{\^{}}\textit{lk} \\
-\hline
-Condition(\textit{lk}) & return a condition variable on \texttt{\^{}}\textit{lk} \\
-\hline
-wait($c$) & wait on condition variable \texttt{\^{}}$c$ \\
-\hline
-notify($c$) & notify a process waiting on condition variable \texttt{\^{}}$c$ \\
-\hline
-notifyAll($c$) & notify all processes waiting on condition variable \texttt{\^{}}$c$ \\
+lock(\textit{lk}) & acquire \texttt{!{}}\textit{lk} \\
+\hline
+unlock(\textit{lk}) & release \texttt{!{}}\textit{lk} \\
+\hline
+Condition(\textit{lk}) & return a condition variable on \texttt{!{}}\textit{lk} \\
+\hline
+wait($c$) & wait on condition variable \texttt{!{}}$c$ \\
+\hline
+notify($c$) & notify a process waiting on condition variable \texttt{!{}}$c$ \\
+\hline
+notifyAll($c$) & notify all processes waiting on condition variable \texttt{!{}}$c$ \\
 \hline
@@
 \hline
-P(\textit{sema}) & procure \texttt{\^{}}\textit{sema}  \\
-\hline
-V(\textit{sema}) & vacate \texttt{\^{}}\textit{sema}  \\
+P(\textit{sema}) & procure \texttt{!{}}\textit{sema}  \\
+\hline
+V(\textit{sema}) & vacate \texttt{!{}}\textit{sema}  \\
 \hline
@@
 The top value must be an address value.  The remaining values are keys into dictionaries.
-The instruction then computes the address of \texttt{\^}$v_1~v_2~...~v_n$.
+The instruction then computes the address of \texttt{!}$v_1~v_2~...~v_n$.
 \item Even though Harmony code does not allow taking addresses of process variables, both

### 2020-07-23 harmony.py
Added mod

@@ -509,8 +509,8 @@ def isunaryop(s):
 def isbinaryop(s):
     return s in [
         "==", "!=", "in", "and", "or", "&", "|", "^",
-        "-", "+", "*", "/", "%", "<", "<=", ">", ">=",
-        "//", "**", "<<", ">>"
+        "-", "+", "*", "/", "//", "%", "mod", "<", "<=", ">", ">=",
+        "**", "<<", ">>"
     ];
 
 tokens = [ "dict{", "==", "!=", "<=", ">=", "//", "**", "<<", ">>", ".." ]
@@ -1655,7 +1655,7 @@ class NaryOp(Op):
                     context.push(0)
                 else:
                     context.push(e1 // e2)
-            elif op == "%":
+            elif op in { "%", "mod" }:
                 if not self.checktype(state, context, sa, isinstance(e1, int)):
                     return
                 if not self.checktype(state, context, sa, isinstance(e2, int)):
@@ -3324,7 +3324,7 @@ class StatementRule(Rule):
 
         (lexeme, file, line, column) = op = t[0]
         self.expect("assignment statement",
-            lexeme in { "=", "+", "-", "*", "/", "//", "**", "<<", ">>",
+            lexeme in { "=", "+", "-", "*", "/", "//", "%", "mod", "**", "<<", ">>",
                             "and", "or", "&", "|", "^" }, op, "expected '='")
         if lexeme != "=":
             (eq, file, line, column) = t[1];

### 2020-07-23 book/paper.tex
Added mod
Latest

@@
 Harmony currently supports the following operators:
+\vspace{1em}
 
@@
 \hline
-\texttt{$e$ / $e$} & two integers (integer division) \\
-\hline
-\texttt{$e$ \% $e$} & two integers (division remainder) \\
+\texttt{$e$ / $e$, $e$ // $e$} & two integers (integer division) \\
+\hline
+\texttt{$e$ \% $e$, $e$ mod $e$} & two integers (division remainder) \\
+\hline
+\texttt{$e$ ** $e$, $e << e$, $e >> e$} & two integers \\
 \hline
@@
 \hline
-\texttt{assert $b$ [, $x$]} & $b$ is a Boolean expression.  Optionally report value of expression $e$\\
+\texttt{assert $b$ [, $e$]} & $b$ is a boolean.  Optionally report value of expression $e$\\
 \hline
@@
 \hline
-\texttt{let $v$ = $e$, ...: S} & $v$ is an identifier, $e$ is an expression, \texttt{S} a list of statements\\
-\hline
-\texttt{if $b$: S} & $b$ is a Boolean expression, \texttt{S} a list of statements\\
-\hline
-\texttt{while $b$: S} & $b$ is a Boolean expression, \texttt{S} a list of statements\\
-\hline
-\texttt{for $v$ in $e$: S} & $v$ is an identifier, $e$ is a set evaluated in order,
+\texttt{let $v$ = $e$ [let ...]: S} & $v$ is an identifier, $e$ is an expression, \texttt{S} a list of statements\\
+\hline
+\texttt{if $b$: S} & $b$ is a boolean, \texttt{S} a list of statements\\
+\hline
+\texttt{while $b$: S} & $b$ is a boolean, \texttt{S} a list of statements\\
+\hline
+\texttt{for $v$ in $e$: S} & $v$ is an identifier, $e$ is a set,
                             \texttt{S} a list of statements\\

### 2020-07-24 book/paper.tex
Fixes bug

@@
 \hline
+\texttt{$e$ * $e$ * ...} & two or more integers \\
+\hline
 \texttt{$e$ - $e$} & two integers or sets \\
 \hline
-\texttt{$e$ * $e$ * ...} & two or more integers \\
-\hline
-\texttt{$e$ \& $e$ \& * ...} & two or more integers or sets \\
-\hline
-\texttt{$e$ | $e$ | * ...} & two or more integers or sets \\
-\hline
-\texttt{$e$ \string^ $e$ \string^ * ...} & two or more integers or sets \\
+\texttt{$e$ \& $e$ \& ...} & two or more integers or sets \\
+\hline
+\texttt{$e$ | $e$ | ...} & two or more integers or sets \\
+\hline
+\texttt{$e$ \string^ $e$ \string^ ...} & two or more integers or sets \\
 \hline

### 2020-07-25 book/paper.tex
Fixed question
Updating book

@@
 \begin{problems}
-\item Write a synchronization module using your own implementation of \texttt{Lock}
-(for example, using Peterson's algorithm or using \texttt{swap}).
+\item Write a new synchronization module, similar to \autoref{fig:spinlocks}, but using
+atomic swap.
 \item
@@
 \hline
+\texttt{lambda $b$: $e$ end} & $b$ is a bounded variable, $e$ an expression\\
+\hline
 \end{tabular}
@@
 \hline
-\texttt{const v = $e$} & \texttt{v} is an identifier, $e$ is a constant expression\\
-\hline
-\texttt{def m($v$, ...): S} & \texttt{m} and \texttt{v} are identifiers, \texttt{S} a list of statements\\
-\hline
-\texttt{let $v$ = $e$ [let ...]: S} & $v$ is an identifier, $e$ is an expression, \texttt{S} a list of statements\\
+\texttt{const b = $e$} & \texttt{b} is a bounded variable, $e$ is a constant expression\\
+\hline
+\texttt{def m $b$: S} & \texttt{m} is an identifier, $b$ a bounded variable, \texttt{S} a list of statements\\
+\hline
+\texttt{let $b$ = $e$ [let ...]: S} & $b$ is a bounded variable, $e$ is an expression, \texttt{S} a list of statements\\
 \hline
@@
 \hline
-\texttt{for $v$ in $e$: S} & $v$ is an identifier, $e$ is a set,
+\texttt{for $b$ in $e$: S} & $b$ is a bounded variable, $e$ is a set,
                             \texttt{S} a list of statements\\

### 2020-07-25 book/paper.tex
Renamed bounded variable to a

@@
 \hline
-\texttt{$e$ < $e$, $e$ <= $e$} & any two Harmony values \\
-\hline
-\texttt{$e$ > $e$, $e$ >= $e$} & any two Harmony values \\
+\texttt{$e$ < $e$, $e$ <= $e$, $e$ > $e$, $e$ >= $e$} & any two Harmony values \\
 \hline
@@
 \hline
-\texttt{min $e$, max $e$} & a set \\
-\hline
-\texttt{cardinality $e$} & a set \\
-\hline
-\texttt{keys $e$, len $e$} & a dictionary (includes tuple or list) \\
+\texttt{min $e$, max $e$, cardinality $e$} & a set \\
+\hline
+\texttt{keys $e$, len $e$} & a dictionary (includes tuple and list) \\
 \hline
@@
 \hline
-\texttt{lambda $b$: $e$ end} & $b$ is a bounded variable, $e$ an expression\\
+\texttt{lambda $a$: $e$ end} & $a$ is a bounded variable, $e$ an expression\\
 \hline
@@
 \hline
-\texttt{const b = $e$} & \texttt{b} is a bounded variable, $e$ is a constant expression\\
-\hline
-\texttt{def m $b$: S} & \texttt{m} is an identifier, $b$ a bounded variable, \texttt{S} a list of statements\\
-\hline
-\texttt{let $b$ = $e$ [let ...]: S} & $b$ is a bounded variable, $e$ is an expression, \texttt{S} a list of statements\\
+\texttt{const a = $e$} & \texttt{a} is a bounded variable, $e$ is a constant expression\\
+\hline
+\texttt{def m $a$: S} & \texttt{m} is an identifier, $a$ a bounded variable, \texttt{S} a list of statements\\
+\hline
+\texttt{let $a$ = $e$ [let ...]: S} & $a$ is a bounded variable, $e$ is an expression, \texttt{S} a list of statements\\
 \hline
@@
 \hline
-\texttt{for $b$ in $e$: S} & $b$ is a bounded variable, $e$ is a set,
+\texttt{for $a$ in $e$: S} & $a$ is a bounded variable, $e$ is a set,
                             \texttt{S} a list of statements\\

### 2020-07-27 book/paper.tex
Now for loops also work over lists and dictionaries in general

@@
 Now suppose we create a test program, a model, that tries sorting
-\emph{all} lists of up to five numbers chosen from the set \{ 1,
-2, 3, 4, 5 \}.  Model checking proves that for those particular
+\emph{all} lists of up to five numbers chosen from the set
+\{ 1, 2, 3, 4, 5 \}.  Model checking proves that for those particular
 scenarios the sorting algorithm works: the output is a sorted
@@
 \item Harmony does not (currently) support floating point, iterators, or I/O;
-Harmony does support \textbf{for} loops and various comprehensions over sets.
+Harmony does support \textbf{for} loops and various comprehensions.
 \item Python is object-oriented, supporting classes with methods and
@@
 \hline
-\texttt{min $e$, max $e$, cardinality $e$} & a set \\
-\hline
-\texttt{keys $e$, len $e$} & a dictionary (includes tuple and list) \\
+\texttt{min $e$, max $e$, len $e$} & a set or a dict \\
+\hline
+\texttt{keys $e$} & a dictionary (includes tuple and list) \\
 \hline

### 2020-07-27 harmony.py
Implemented int * list

@@ -1515,11 +1515,26 @@ class NaryOp(Op):
             return False
         return True
 
+    def checkdmult(self, state, context, args, d, e):
+        if not self.checktype(state, context, args, isinstance(e, int)):
+            return False
+        keys = set(range(len(d.d)))
+        if d.d.keys() != keys:
+            context.failure = "Error: one operand in " + str(self.op) + \
+                        " must be a list: " + str(list(reversed(args)))
+            return False
+        return True
+
+    def dmult(self, d, e):
+        n = len(d.d)
+        lst = { i:d.d[i % n] for i in range(e * n) }
+        return DictValue(lst)
+
     def eval(self, state, context):
         (op, file, line, column) = self.op
         assert len(context.stack) >= self.n
         sa = context.stack[-self.n:]
-        if op in { "+", "*", "&", "|", "^" }:
+        if op in { "+", "&", "|", "^" }:
             assert self.n > 1
             e2 = context.pop()
             for i in range(1, self.n):
@@ -1535,12 +1550,6 @@ class NaryOp(Op):
                         if not self.checktype(state, context, sa, isinstance(e2, DictValue)):
                             return
                         e2 = self.concat(e1, e2)
-                elif op == "*":
-                    if not self.checktype(state, context, sa, isinstance(e1, int)):
-                        return
-                    if not self.checktype(state, context, sa, isinstance(e2, int)):
-                        return
-                    e2 *= e1
                 elif op == "&":
                     if isinstance(e1, int):
                         if not self.checktype(state, context, sa, isinstance(e2, int)):
@@ -1736,6 +1745,22 @@ class NaryOp(Op):
                 if not self.checktype(state, context, sa, isinstance(e2, SetValue)):
                     return
                 context.push(e1 in e2.s)
+            elif op == "*":
+                if isinstance(e1, DictValue) or isinstance(e2, DictValue):
+                    if isinstance(e1, DictValue) and not self.checkdmult(state, context, sa, e1, e2):
+                        return
+                    if isinstance(e2, DictValue) and not self.checkdmult(state, context, sa, e2, e1):
+                        return
+                    if isinstance(e1, DictValue):
+                        context.push(self.dmult(e1, e2))
+                    else:
+                        context.push(self.dmult(e2, e1))
+                else:
+                    if not self.checktype(state, context, sa, isinstance(e1, int)):
+                        return
[…]

### 2020-07-27 book/paper.tex
Implemented int * list
Made a pass over all book examples

@@
 Up.hny:3     done[self] = True;
-    6 LoadVar self
-    7 PushAddress done
+    6 PushAddress done
+    7 LoadVar self
     8 Address 2
@@
 \hline
-\texttt{$e$ * $e$ * ...} & two or more integers \\
-\hline
 \texttt{$e$ - $e$} & two integers or sets \\
@@
 \texttt{$e$ \string^ $e$ \string^ ...} & two or more integers or sets \\
+\hline
+\texttt{$e$ * $e$} & two integers or an integer and a list \\
 \hline

### 2020-07-27 harmony.py

@@ -1592,7 +1592,7 @@ class NaryOp(Op):
                 if not self.checktype(state, context, sa, isinstance(e, int) or isinstance(e, float)):
                     return
                 context.push(-e)
-            if op == "~":
+            elif op == "~":
                 if not self.checktype(state, context, sa, isinstance(e, int)):
                     return
                 context.push(~e)
@@ -2044,7 +2044,7 @@ class NaryAST(AST):
     def __init__(self, op, args):
         self.op = op
         self.args = args
-        assert all(isinstance(x, AST) for x in args)
+        assert all(isinstance(x, AST) for x in args), args
 
     def __repr__(self):
         return "NaryOp(" + str(self.op) + ", " + str(self.args) + ")"
@@ -2115,6 +2115,8 @@ class NaryRule(Rule):
 
     def parse(self, t):
         (ast, t) = ExpressionRule().parse(t)
+        if ast == False:
+            return (ast, t)
         (lexeme, file, line, column) = t[0]
         if lexeme in self.closers:
             return (ast, t)

### 2020-07-27 book/paper.tex
Added more python-like operators

@@
 \hline
+\texttt{reversed($t$)} & reverse a list \\
+\hline
+\texttt{sorted($t$)} & sorted set or list \\
+\hline
+\texttt{set($t$)} & convert values of a dict or list into a set \\
+\hline
+\texttt{list($t$)} & convert set into a list \\
+\hline
+\texttt{values($t$)} & convert values of a dict into a list sorted by key \\
+\hline
+\texttt{enumerate($t$)} & like Python enumerate \\
+\hline
+\texttt{sum($t$)} & return the sum of all elements in $t$\\
+\hline
+\texttt{list2bag($t$)} & convert list $t$ into a bag \\
+\hline
 \texttt{listQsort($t$)} & sort list $t$ using quicksort\\
-\hline
-\texttt{list2bag($t$)} & convert list $t$ into a bag \\
-\hline
-\texttt{list2set($t$)} & convert list $t$ into a set \\
-\hline
-\texttt{listMin($t$)} & return the minimum of all elements in $t$\\
-\hline
-\texttt{listMax($t$)} & return the maximum of all elements in $t$\\
-\hline
-\texttt{listSum($t$)} & return the sum of all elements in $t$\\
 \hline

### 2020-07-27 book/paper.tex
Fixed formatting

@@
 \hline
-tas(\textit{lk}) & test-and-set on \texttt{!{}}\textit{lk} \\
-\hline
-Lock() & return a lock object \\
-\hline
-lock(\textit{lk}) & acquire \texttt{!{}}\textit{lk} \\
-\hline
-unlock(\textit{lk}) & release \texttt{!{}}\textit{lk} \\
-\hline
-Condition(\textit{lk}) & return a condition variable on \texttt{!{}}\textit{lk} \\
-\hline
-wait($c$) & wait on condition variable \texttt{!{}}$c$ \\
-\hline
-notify($c$) & notify a process waiting on condition variable \texttt{!{}}$c$ \\
-\hline
-notifyAll($c$) & notify all processes waiting on condition variable \texttt{!{}}$c$ \\
-\hline
-Semaphore(\textit{cnt}) & return a semaphore object initialized to \textit{cnt} \\
-\hline
-P(\textit{sema}) & procure \texttt{!{}}\textit{sema}  \\
-\hline
-V(\textit{sema}) & vacate \texttt{!{}}\textit{sema}  \\
-\hline
-Queue() & return a synchronized queue object \\
-\hline
-dequeue($q$) & return next element of $q$, blocking if empty \\
-\hline
-enqueue($q$, \textit{item}) & add \textit{item} to $a$ \\
+\texttt{tas}(\textit{lk}) & test-and-set on \texttt{!{}}\textit{lk} \\
+\hline
+\texttt{Lock}() & return a lock object \\
+\hline
+\texttt{lock}(\textit{lk}) & acquire \texttt{!{}}\textit{lk} \\
+\hline
+\texttt{unlock}(\textit{lk}) & release \texttt{!{}}\textit{lk} \\
+\hline
+\texttt{Condition}(\textit{lk}) & return a condition variable on \texttt{!{}}\textit{lk} \\
+\hline
+\texttt{wait}($c$) & wait on condition variable \texttt{!{}}$c$ \\
+\hline
+\texttt{notify}($c$) & notify a process waiting on condition variable \texttt{!{}}$c$ \\
+\hline
+\texttt{notifyAll}($c$) & notify all processes waiting on condition variable \texttt{!{}}$c$ \\
+\hline
+\texttt{Semaphore}(\textit{cnt}) & return a semaphore object initialized to \textit{cnt} \\
+\hline
+\texttt{P}(\textit{sema}) & procure \texttt{!{}}\textit{sema}  \\
+\hline
+\texttt{V}(\textit{sema}) & vacate \texttt{!{}}\textit{sema}  \\
+\hline
+\texttt{Queue}() & return a synchronized queue object \\
+\hline
+\texttt{dequeue}($q$) & return next element of $q$, blocking if empty \\
+\hline
+\texttt{enqueue}($q$, \textit{item}) & add \textit{item} to $a$ \\
 \hline

### 2020-07-27 book/paper.tex
Fixed formatting

@@
 \hline
+Cut & pop set or dict and push minimum and remainder \\
+\hline
 Del [$v$] & delete shared variable $v$ \\
@@
 \hline
-Split & pop set and push minimum and remainder \\
+Split & pop tuple and push its elements \\
 \hline
@@
 \hline
-register & a Harmony dictionary mapping atoms (representing process variables) to Harmony values \\
+register & a dictionary mapping atoms (names of variables) to values \\
 \hline

### 2020-07-27 book/paper.tex
Adding a chapter on Harmony language details

@@
 \def\figureautorefname{Figure}%
+\def\sectionautorefname{Section}%
 
@@
 to have a method update a shared variable specified as an argument.
-For this (as well as some other uses), Harmony supports \emph{pointers}
+For this, each shared variable has an \emph{address}.
+\index{address}%
+If $x$ is a shared variable, then the expression ?$x$ is the address of $x$.
+If a variable contains an address, we call that variable a \emph{pointer}.
 \index{pointer}%
-to shared variables.
-If $x$ is a shared variable, then the expression ?$x$
- is a pointer to $x$ (also known as the \emph{address}
-\index{address}%
-of $x$).
-Conversely, if $p$ is a pointer to a shared variable, then the
+If $p$ is a pointer to a shared variable, then the
 expression \texttt{!{}}$p$ is the value of the shared variable.
 In particular, \texttt{!{}?}$x$ == $x$.
-This is similar to how C pointers work.
+This is similar to how C pointers work (\texttt{*\string&}$x$ == $x$).
+
+Because there are no objects in Harmony,
+there are no object methods either.
+However, methods in Harmony are program counter constants.
 
@@
 make the code re-usable by other applications.
-
-For those who are
-extra adventurous: you can add the \texttt{P\_enter} and
-\texttt{P\_exit} methods to the \texttt{P\_mutex} dictionary
-like so:
-\begin{code}
-\begin{verbatim}
-dict{ .turn: 0, .flags: [ False, False ], .enter: P_enter, .exit: P_exit }
-\end{verbatim}
-\end{code}
-That would allow you to simulate object methods.
 
@@
 \chapter{List of Values}
+\label{ap:values}
 
@@
 \hline
-\textit{lv} = $e$ & \textit{lv} is an lvalue and $e$ is an expression\\
-\hline
-\textit{lv} \texttt{[op]=} $e$ & \texttt{op} is one of \texttt{+}, \texttt{-}, \texttt{*}, \texttt{/}, 
+\textit{lv} = [lv =]... $e$ & \textit{lv} is an lvalue and $e$ is an expression\\
+\hline
+\textit{lv} \texttt{[op]=} $e$ & \texttt{op} is one of \texttt{+}, \texttt{-},
+\texttt{*}, \texttt{/}, \texttt{//}, \texttt{\%},
+\texttt{\string\&}, \texttt{|}, \texttt{\string\^},
 \texttt{and}, and \texttt{or}\\
@@
 \section{The \texttt{list} module}
+\label{ap:list}
 
@@
 \texttt{values($t$)} & convert values of a dict into a list sorted by key \\
+\hline
+\texttt{items($t$)} & convert dict into (key, value) list sorted by key \\
 \hline
@@
 
+\chapter{Harmony Language Details}
+
+The Harmony language borrows heavily from Python.  However, there are
+some important differences that we will describe in this chapter.
+
+\section{Harmony is not object-oriented}
+
+Python is object-oriented, but Harmony is not.  This can lead to some
+unexpected differences.  For example, consider the following code:
+
+\begin{code}
+\harmonysource{oo}
+\end{code}
+
+In Python, lists are objects.  Thus $x$ and $y$ point to the same list,
+and the assertion will fail.
+In Harmony, lists are values.  So when $x$ is updated in Line~3, it does
+not affect the value of $y$.  The assertion succeeds.
+Harmony does support references to values (\autoref{ch:method}),
+allowing programs to implement shared objects.
+
+Because Harmony does not have objects, it also does not have object methods.
+However, Harmony methods and lambdas are program counter constants.
+These constants can be added to dictionaries.
+For example, in Figure~\ref{fig:petersonmethods}
+you can add the \texttt{P\_enter} and
+\texttt{P\_exit} methods to the \texttt{P\_mutex} dictionary
+like so:
+\begin{code}
+\begin{verbatim}
+dict{ .turn: 0, .flags: [ False, False ], .enter: P_enter, .exit: P_exit }
+\end{verbatim}
+\end{code}
+That would allow you to simulate object methods.
+
+\section{Constants, Global and Local Variables}
+
+Each (non-reserved) identifier in a Harmony program refers to either
+a constant, a global variable, or a local variable.
+Constants are declared using \texttt{const} statements.
+Those constants are computed at compile-time.
+
+Local variables all declared.  They can be declared in
+\texttt{def} statements (i.e., arguments),
+\texttt{let} statements, and in \texttt{for} loops.
+Also, each method has an implicitly declared
+\textit{result} variable.
+Local variables are tightly scoped and cannot be shared
+between processes.
+While in theory one method can be declared within another,
+they cannot share variables either.
+All other variables are global and must be initialized
+before any processes are spawned.
+
+While arguments to a method and variables in \texttt{for} loops
+can be modified, we discourage it for improved code readability.
+
+\section{Tuples, Lists, and Pattern Matching}
+
+Harmony's tuples and, equivalently, lists, are just special cases of dictionaries.
+They can be bracketed either by '(' and ')' or by '[' and ']', and
+even the brackets are often optional.  Importantly, with a singleton list, the
+one element must be followed by a comma.
+So the statement \texttt{$x$ = 1,;} assigns a singleton tuple (or list) to $x$.
+
+Because tuples and lists are dictionaries, the \texttt{del} statement is
+different Python.  For example, if $x$ = [.a, .b, .c], then
+\texttt{del} $x$[1] results in $x$ = \texttt{dict}\{ 0:.a, 2:.c \}, \emph{not}
+$x$ = [.a, .c].
+
+Harmony allows pattern matching against nested tuples in various
+language constructs.  The following are the same in Python and Harmony:
+\begin{itemize}
+\item \texttt{$x,$ = 1,}: assigns 1 to $x$;
+\item \texttt{$x$, $y$ = 1, (2, 3)}: assigns 1 to $x$ and (2, 3) to $y$;
+\item \texttt{$x$, ($y$, $z$) = 1, (2, 3)}: assigns 1 to $x$, 2 to $y$, and
+3 to $z$;
+\item \texttt{$x$, ($y$, $z$) = 1, 2;} generates an runtime error because 2 cannot
+be matched with ($y$, $z$);
+\item \texttt{$x[0]$, $x[1]$ = $x[1]$, $x[0]$;} swaps the first two elements of list $x$.
+\end{itemize}
+
+As in Python, pattern matching can also be used in \texttt{for} statements.
+For example:
+\begin{code}
+\texttt{for} key, value in [ (1, 2), (3, 4) ]:
+    ...
+\end{code}
+
+Harmony also allows pattern matching in defining and invoking methods.
+For example, you can write:
+\begin{code}
+\texttt{def} f[$a$, ($b$, $c$)]: ...
+\end{code}
+and then call \texttt{f[1, (2, 3)]}.
+Note that the more familiar: \texttt{def} g($a$) defines a method $g$ with
+a single argument $a$.  Invoking g(1, 2) would assign the tuple (1, 2) to
+$a$.  This is not consistent with Python syntax.  For single argument methods,
+you may want to declare as follows: \texttt{def} g($a,$).
+Calling g(1,) assigns 1 to $a$, while calling g(1, 2) would result in a
+runtime error as (1, 2) cannot be matched with ($a$,).
+
+Pattern matching can also be used in \texttt{const} and \texttt{let}
+statements.
+
+\section{For Loops and Comprehensions}
+
+While Harmony does not support general iterators such as Python does,
+Harmony allows iterating over sets and dictionaries (and thus lists
+and tuples).  The details are a little different from Python:
+
+\begin{itemize}
+\item When iterating over a set, the set is always traversed in order
+(see \autoref{ap:values} for a description how Harmony values are ordered);
+\item In case of a dictionary, the iteration is over the values of the
+dictionary, but in the order of the keys.  In the case of lists, this
+works much the same as in Python;
+\item If you want to iterate over the keys of a dictionary $d$, use
+\texttt{for} $k$ \texttt{in} \texttt{keys} $d$;
+\item The \texttt{list} module (\autoref{ap:list}) provides methods \texttt{values}(),
+\texttt{items}(), \texttt{enumerate}(), and \texttt{reversed}() for
+other types of iteration supported by Python.
+\end{itemize}
+
+Harmony supports filtering in \texttt{for loops}.
+For example:
+\begin{code}
+\texttt{for} $i$ in \{ 1..10 \} \texttt{such that} ($i$ \% 2) == 0: ...
+\end{code}
+\noindent iterates over the even numbers between 1 and 10.
+
+Harmony also supports set, list, and dictionary comprehensions.
+Set and list comprehensions are similar to Python, except that filtering uses
+the keywords \texttt{such that} instead of \texttt{if}.
+Dictionary comprehensions are a little different.
+The dictionary comprehension \texttt{dict}\{ f($i$) \texttt{for} $i$ \texttt{in} $I$ \}
+creates a dictionary of $i$:f($i$) entries.
+
 \cleardoublepage

### 2020-07-27 book/paper.tex
Latest

@@
 \texttt{subseq($t$, $b$, $b$)} & return a \emph{slice} of list $t$ starting
-at index $b$ and ending just before index $f$\\
+at index $b$ and ending just before $f$\\
 \hline
@@
 Harmony allows pattern matching against nested tuples in various
-language constructs.  The following are the same in Python and Harmony:
+language constructs.
+\index{pattern matching}%
+The following are the same in Python and Harmony:
 \begin{itemize}
@@
 Harmony also supports set, list, and dictionary comprehensions.
+\index{comprehensions}%
 Set and list comprehensions are similar to Python, except that filtering uses

### 2020-07-27 book/paper.tex
Latest version of book

@@
 not affect the value of $y$.  The assertion succeeds.
-Harmony does support references to values (\autoref{ch:method}),
+Harmony supports references to values (\autoref{ch:method}),
 allowing programs to implement shared objects.
@@
 Harmony's tuples and, equivalently, lists, are just special cases of dictionaries.
-They can be bracketed either by '(' and ')' or by '[' and ']', and
-even the brackets are often optional.  Importantly, with a singleton list, the
+They can be bracketed either by '(' and ')' or by '[' and ']', but
+the brackets are often optional.  Importantly, with a singleton list, the
 one element must be followed by a comma.
@@
 $x$ = [.a, .c].
+Harmony also does not support special slicing syntax like Python.
+To modify lists, use the \texttt{subseq} method in the \texttt{list} module
+(\autoref{ap:list}).
 
@@
 
-Harmony also allows pattern matching in defining and invoking methods.
+Harmony (but not Python)
+also allows pattern matching in defining and invoking methods.
 For example, you can write:

### 2020-07-28 book/paper.tex
Latest
Massaging the book

@@
 by instrumenting the program.
-The output of running concurrent programs depends on how the
+On the other hand,
+the output of running concurrent programs depends on how the
 execution of the various processes are \emph{interleaved}.
@@
 \emph{can} expose the bug it \emph{will}.  Importantly, if the bug is
-found, the model checker precisely shows how to trigger it in
+found, the model checker precisely shows how to trigger the bug in
 the smallest number of steps.
@@
 is not trivial.
-\item \autoref{ch:peterson} introduces Peterson's Algorithm, an
+\item \autoref{ch:peterson} introduces \emph{Peterson's Algorithm}, an
 elegant solution to implementating a critical section.
 \item \autoref{ch:method} gives some more details on the Harmony
-language.
-\item \autoref{ch:spinlock} introduces ``hardware'' locks
+language needed for the rest of the book.
+\item \autoref{ch:spinlock} introduces ``hardware'' \emph{locks}
 for implemented critical sections.
@@
 \item \autoref{ch:rdwrbusy} and \autoref{ch:rdwr2} present the
-abstraction and implementation of reader/writer
-locks, which allow multiple processes in a critical section if they are
+abstraction and implementation of \emph{reader/writer locks},
+which allow multiple processes in a critical section if they are
 all only executing read-only accesses.
-\autoref{ch:rdwrbusy} also talks about busy-waiting, which is
-an undesirable approach to synchronize processes.
+\autoref{ch:rdwrbusy} also discusses \emph{busy-waiting}, which is
+an easy but undesirable approach to synchronize processes.
 \item \autoref{ch:semaphore} introduces \emph{semaphores},
@@
 
-\chapter{Introduction to Programming with Harmony}
+\chapter{Introduction to Harmony}
 \label{ch:harmonyintro}
@@
 \begin{itemize}
-\item Syntax: Every statement in Harmony must be terminated by a semicolon
-(even \textbf{while} statements and \textbf{def} statements).
+\item Every statement in Harmony must be terminated by a semicolon
+(even \textbf{def} statements).
 In Python semicolons are optional and rarely used.
@@
 \item Harmony does not (currently) support floating point, iterators, or I/O;
-Harmony does support \textbf{for} loops and various comprehensions.
+Harmony does support \textbf{for} loops and various ``comprehensions.''
 \item Python is object-oriented, supporting classes with methods and
 inheritance; Harmony has objects but does not support classes.  Harmony supports
-pointers to objects.
+pointers to objects and methods.
 \end{itemize}
@@
 also has a variable called \textit{n} containing the value of the
-argument.  The \textit{x..y} notation generates a set containing the numbers
+argument.  The \{ \textit{x..y} \} notation generates a set containing the numbers
 from~\textit{x} to~\textit{y} (inclusive).  The last two lines in the program are
@@
 even gets going.
+And because Harmony tries every possible execution, it will consider
+such executions as well.
 
@@
 A program is correct if it works for any interleaving.
-In fact, Harmony will try all possible interleavings of the processes
+Harmony will try all possible interleavings of the processes
 executing machine instructions.
@@
 
-The following exercises are intended to show you that on the one hand it
-is easy to translate Harmony programs into Python, but on the other hand,
-it is easier to find concurrency bugs using Harmony.
+The following exercises are intended to show you that while it is just
+as easy to write concurrent programs in Python, it is much easier to
+find concurrency bugs using Harmony.
 
@@
 \end{code}
-How many times does the assertion
-fail (if any).
+How many times does the assertion fail (if any)?
 \end{enumerate}
@@
 to have a method update a shared variable specified as an argument.
-For this, each shared variable has an \emph{address}.
+For this, as mentioned in \autoref{ch:harmonymachine},
+each shared variable has an \emph{address}.
 \index{address}%
@@
 This is similar to how C pointers work (\texttt{*\string&}$x$ == $x$).
-
-Because there are no objects in Harmony,
-there are no object methods either.
-However, methods in Harmony are program counter constants.
 
@@
 that does not busy-wait.
-You can test the code by running \texttt{harmony -m RW=RWlock RWtest}.
+You can test the code by running \texttt{harmony -m RW=RWlock RWtest.hny}.
 It uses two ordinary locks: \textit{rwlock} is used by both readers and writers,
@@
 The implementation satisfies progress: if no process is in the critical
-section than any process enter.
+section then any process can enter.
 

### 2020-07-28 book/paper.tex
Made a pass over the first four chapters

@@
 
-Harmony programs are compiled to Harmony bytecode
+Harmony programs are compiled to Harmony \emph{bytecode}
 \index{bytecode}%
-(machine instructions for a virtual machine),
+(a list of machine instructions for a virtual machine),
 which in turn is executed by the Harmony virtual machine (HVM).
@@
 However, when creating a singleton list, make sure you include the
-comma, as in \texttt{[False,]}.
+comma, as in \texttt{[False,]}.  The expression \texttt{[False]} just means
+\texttt{False}.
 
@@
 If $d$ is a
-dictionary, then \texttt{keys}~$d$ (or equivalently \texttt{keys}($d$)
+dictionary, then \texttt{keys}~$d$ (or equivalently \texttt{keys}($d$))
 returns the set of keys and \texttt{len}~$d$ returns the size of
@@
 
-A Harmony program is translated
-into \emph{bytecode}, which is list of machine instructions that the
-Harmony virtual machine (HVM) executes.
+A Harmony program is translated into HVM bytecode.
 The HVM is not an ordinary virtual machine, but its architecture
@@
 The context bag has an initial context in it with name tag
-\texttt{\_\_main\_\_/()}, pc 0, register \texttt{()}, and an empty stack.
+\texttt{\_\_init\_\_/()}, pc 0, register \texttt{()}, and an empty stack.
 Each machine instruction updates the state in some way.
@@
 ``root'' of the directory tree.
+For example, in \autoref{fig:inc} the memory is a dictionary with two
+entries: \texttt{.count} and \texttt{.done}.  And the value of entry
+\texttt{.done} is a dictionary with keys 0 and 1.
 
@@
 In this case, the first instruction is a \texttt{JUMP} instruction that sets the
-program counter to 39.
+program counter to 34 (skipping over the code for the methods).
 At program counter 1 is the code for the \texttt{incrementer} method.
@@
 \item \texttt{2-ary} is a \texttt{+} operation with 2 arguments.
-It pops two values from the stack (1 and the value of \textit{count}),
-adds them and pushes the result back onto the stack.
+It pops two values from the stack (the value of \textit{count} and 1),
+adds them, and pushes the result back onto the stack.
 \item The \texttt{Store} instruction pops

### 2020-07-28 book/paper.tex
Working on massaging section 5

@@
 at the beginning and at the end of the critical section, but in the middle
-the invariant may be temporarily broken---this is no issue as no other
+the invariant may be temporarily broken---this is no issue as critical
+sections guarantee that no other
 process will be able to see it.
@@
 two processes being in the critical section.
-What we need now a \emph{liveness property}:
+What we need now is a \emph{liveness property}:
 \index{liveness property}%
@@
 the spawned processes never terminate.  We
-will model processes in critical sections using the framework in
+will instead model processes in critical sections using the framework in
 \autoref{fig:cs}: a process can \emph{choose} to enter a
@@
 \autoref{fig:mutex} shows a high-level state diagram
-\index{state diagram} of mutual exclusion and progress.%
+\index{state diagram}%
+of mutual exclusion and progress.
 The circles represent states, while the arrows represent possible state
@@
 lock first and release it upon exiting the critical section.
-Note that the word ``lock'' does not describe the concept well---it really
-is more like a baton or token that only one process can hold.
+% Note that the word ``lock'' does not describe the concept well---it really
+% is more like a baton or token that only one process can hold.
 

### 2020-07-28 book/paper.tex
Finished pass over Chapter 5
Latest

@@
 
-To see why, first note the invariant that if process $i$ is in the
+To see why, first note the following invariant: if process $i$ is in the
 critical section, then \textit{flags}[$i$] == \texttt{True}.
@@
 
-In 1981, Peterson came up with a beautiful solution to the mutual exclusion
+In 1981, Gary L.~Peterson came up with a beautiful solution to the mutual exclusion
 problem, now known as ``Peterson's Algorithm''~\cite{Peterson81}.
@@
 For that, we have to understand a little bit more about how the Harmony
-machine works.
+virtual machine (HVM) works.
 In \autoref{ch:harmonymachine} we talked about the concept of \emph{state}:
@@
 \index{process variable}%
-for each process, including its
+of each process, including its
 program counter and the contents of its stack.
-Everytime a process executes a HVM machine instruction, the
+Each time a process executes a HVM machine instruction, the
 state changes (if only because the program counter of the process
@@
 
-It is often useful to classify states into sets.
+It is often useful to classify states.
 \emph{Initial}, \emph{final}, and \emph{reachable}, and \emph{unreachable}
@@
 over the resulting state.
-For this you need a predicate that describes exactly which
+For this to work you would need a predicate that describes exactly which
 states are reachable.
@@
 \item $\mathcal{I}$ holds in the initial state.
-\item For any state in which $\mathcal{I}$ holds and any process in the
-state takes a step, then $\mathcal{I}$ also holds in the resulting state.
+\item For any state in which $\mathcal{I}$ holds and for any process in the
+state, if the process takes a step, then $\mathcal{I}$ also holds in the resulting state.
 \end{itemize}
-Unlike an invariant, an inductive invariant must also hold over unreachable states.
+Unlike an ordinary invariant, an inductive invariant must also hold over unreachable states.
 
@@
 It is easy to construct one: let process~0 be at label \texttt{@cs}
-and process~1 at the start of the \textbf{while} loop.
+and process~1 be at the start of the \textbf{while} loop.
 Also imagine that in this particular state $\mathit{turn} = 1$.  Now let
@@
 $\mathit{flags}[i]$ but still before setting \textit{turn} to $1-i$.
-More formally, $\mathcal{I}_p(i) = \mathtt{process}(i)@cs) \Rightarrow (\lnot \mathcal{C}(i) \lor \mathtt{process}(1-i)@gate)$.
+More formally, $\mathcal{I}_p(i) = \mathtt{process}(i)@cs \Rightarrow (\lnot \mathcal{C}(i) \lor \mathtt{process}(1-i)@gate)$.
 For Peterson's Algorithm, an inductive invariant that works well is
@@
 To see that $\mathcal{I}_p(0) \land \mathcal{I}_p(1)$ is, in fact, an inductive invariant, first note that
-it certainly holds in the initial state, because in the initial state no process
+it holds in the initial state, because in the initial state no process
 is in the critical section.
@@
 Let us consider each of these cases.
+We have to show that if process~1 takes a step, then one of those
+cases must hold after the step.
 In the first case, if process~1 takes a step, there are two possibilities:
@@
 algorithm works.  The second provides much more insight.  We therefore
-encourage to include inductive invariants in your Harmony code.
+encourage to include invariants, particularly inductive ones, in your
+Harmony code.
 
@@
 
-This anecdote suggests the following.  If you need to do an induction
-proof of an algorithm, you have to come up with an inductive invariant.
+This anecdote suggests the following.  If you need to do a proof by induction
+of an algorithm, you have to come up with an inductive invariant.
 Before trying to prove the algorithm, you can check that the predicate is
@@
 predicate is not invariant, and therefore not an inductive invariant either.
-(The author fixed the Wikipedia page.)
+(The author fixed the Wikipedia page with the help of Fred B.~Schneider.)
 

### 2020-07-28 book/paper.tex
Finished pass over chapter 7

@@
 If $p$ is a pointer to a shared variable, then the
-expression \texttt{!{}}$p$ is the value of the shared variable.
-In particular, \texttt{!{}?}$x$ == $x$.
+expression \texttt{!}$p$ is the value of the shared variable.
+In particular, \texttt{!?}$x$ == $x$.
 This is similar to how C pointers work (\texttt{*\string&}$x$ == $x$).
@@
 a pointer to a mutex and the process identifier (0 or 1).
-\texttt{(!{}}\textit{pm}\texttt{.turn} is the value of the \texttt{.turn} key
+(\texttt{!}\textit{pm})\texttt{.turn} is the value of the \texttt{.turn} key
 in the dictionary that \textit{pm} points to. 
 Note that the parentheses in this expression
-are needed, as \texttt{!{}}\textit{pm}\texttt{.turn} would wrongly evaluate to
-\texttt{!{}}(\textit{pm}\texttt{.turn}).
+are needed, as \texttt{!}\textit{pm}\texttt{.turn} would wrongly evaluate
+\texttt{!}(\textit{pm}\texttt{.turn}).
 
@@
 To see that invariant $\mathcal{I}_1$ is maintained, note that
-\texttt{!{}}$p$ == \texttt{True} upon entry of \texttt{tas}
+\texttt{!}$p$ == \texttt{True} upon entry of \texttt{tas}
 (because of the condition on the \textbf{while} loop that the
@@
 \begin{enumerate}
-\item \texttt{!{}}$s$ is \texttt{False} upon entry to \texttt{tas}.
-Then upon exit \texttt{!{}}$p$ == \texttt{False} and \texttt{!{}}$s$ == \texttt{True},
+\item \texttt{!}$s$ is \texttt{False} upon entry to \texttt{tas}.
+Then upon exit \texttt{!}$p$ == \texttt{False} and \texttt{!}$s$ == \texttt{True},
 maintaining the invariant.
-\item \texttt{!{}}$s$ is \texttt{True} upon entry to \texttt{tas}.
+\item \texttt{!}$s$ is \texttt{True} upon entry to \texttt{tas}.
 Then upon exit nothing has changed, maintaining the invariant.
@@
 a message $m$ to
-\texttt{!{}}\textit{pchan}, where \textit{pchan} is either ?\textit{s\_chan}
+\texttt{!}\textit{pchan}, where \textit{pchan} is either ?\textit{s\_chan}
 or ?\textit{r\_chan}.
 The method places either $m$ (to model a successful send)
-or \texttt{()} (to model loss) in \texttt{!{}}\textit{pchan}.
+or \texttt{()} (to model loss) in \texttt{!}\textit{pchan}.
 The use of the \textit{reliable} flag will be explained later.
-\texttt{net\_recv}(\textit{pchan}) models checking \texttt{!{}pchan} for
+\texttt{net\_recv}(\textit{pchan}) models checking \texttt{!pchan} for
 the next message.
@@
 \begin{itemize}
-\item \texttt{inc}($p$): atomically increments \texttt{!{}}$p$ and returns
+\item \texttt{inc}($p$): atomically increments \texttt{!}$p$ and returns
 the old value;
-\item \texttt{swap}($p$): sets \texttt{!{}}$p$ to \texttt{()} and returns
+\item \texttt{swap}($p$): sets \texttt{!}$p$ to \texttt{()} and returns
 the old value.
@@
 \hline
-\texttt{!{}$e$} & an address \\
+\texttt{!$e$} & an address \\
 \hline
@@
 variable.  It can also be indexed.  Examples include
-\textit{var}, \textit{var}[$e$], and \texttt{!{}}\textit{p}.
+\textit{var}, \textit{var}[$e$], and \texttt{!}\textit{p}.
 
@@
 \hline
-\texttt{bagAdd}(\textit{pb}, $e$) & add one copy of $e$ to bag \texttt{!{}}\textit{pb}\\
-\hline
-\texttt{bagRemove}(\textit{pb}, $e$) & remove one copy of $e$ from bag \texttt{!{}}\textit{pb}\\
+\texttt{bagAdd}(\textit{pb}, $e$) & add one copy of $e$ to bag \texttt{!}\textit{pb}\\
+\hline
+\texttt{bagRemove}(\textit{pb}, $e$) & remove one copy of $e$ from bag \texttt{!}\textit{pb}\\
 \hline
@@
 \hline
-\texttt{tas}(\textit{lk}) & test-and-set on \texttt{!{}}\textit{lk} \\
+\texttt{tas}(\textit{lk}) & test-and-set on \texttt{!}\textit{lk} \\
 \hline
@@
 \hline
-\texttt{lock}(\textit{lk}) & acquire \texttt{!{}}\textit{lk} \\
-\hline
-\texttt{unlock}(\textit{lk}) & release \texttt{!{}}\textit{lk} \\
-\hline
-\texttt{Condition}(\textit{lk}) & return a condition variable on \texttt{!{}}\textit{lk} \\
-\hline
-\texttt{wait}($c$) & wait on condition variable \texttt{!{}}$c$ \\
-\hline
-\texttt{notify}($c$) & notify a process waiting on condition variable \texttt{!{}}$c$ \\
-\hline
-\texttt{notifyAll}($c$) & notify all processes waiting on condition variable \texttt{!{}}$c$ \\
+\texttt{lock}(\textit{lk}) & acquire \texttt{!}\textit{lk} \\
+\hline
+\texttt{unlock}(\textit{lk}) & release \texttt{!}\textit{lk} \\
+\hline
+\texttt{Condition}(\textit{lk}) & return a condition variable on \texttt{!}\textit{lk} \\
+\hline
+\texttt{wait}($c$) & wait on condition variable \texttt{!}$c$ \\
+\hline
+\texttt{notify}($c$) & notify a process waiting on condition variable \texttt{!}$c$ \\
+\hline
+\texttt{notifyAll}($c$) & notify all processes waiting on condition variable \texttt{!}$c$ \\
 \hline
@@
 \hline
-\texttt{P}(\textit{sema}) & procure \texttt{!{}}\textit{sema}  \\
-\hline
-\texttt{V}(\textit{sema}) & vacate \texttt{!{}}\textit{sema}  \\
+\texttt{P}(\textit{sema}) & procure \texttt{!}\textit{sema}  \\
+\hline
+\texttt{V}(\textit{sema}) & vacate \texttt{!}\textit{sema}  \\
 \hline

### 2020-07-28 book/paper.tex
Working on Section 8

@@
 
-\autoref{fig:tas} goes on how to implement mutual exclusion for
+\autoref{fig:tas} goes on to implement mutual exclusion for
 a set of \texttt{N} processes.
@@
 \index{spinlock}%
-because each process is ``spinning'' until
+because each process is ``spinning'' (executing a tight loop) until
 it can acquire the lock.

### 2020-07-28 book/paper.tex
Fixed invariant checking

@@
 
-We can do this by adding another process that, in a loop,
+We can do this by adding another process that
 checks the invariant.  \autoref{fig:tasinv} shows the code.
@@
 \texttt{False}.
-The function is invoked by in a loop by a process that runs alongside
+The method is invoked by a process that runs alongside
 the other processes.

### 2020-07-28 book/paper.tex
Made pass over Chapter 9

@@
 The former ensures that while the
-atomic statement is executing no other process can execute.
+\texttt{atomic} statement is executing no other process can execute.
 The latter allows multiple processes to run concurrently,
 just not within the critical section.
-The former is rarely available to a programmer, while the latter
+The former is rarely available to a programmer (e.g., none of
+Python, C, or Java support it), while the latter
 is very common.
 
-In Harmony, atomic statements allow you to \emph{implement} your own
+In Harmony, \texttt{atomic} statements allow you to \emph{implement} your own
 synchronization primitives like test-and-set.  Other common examples
-include compare-and swap and fetch-and-add.  Atomic statements
+include compare-and swap and fetch-and-add.  \texttt{atomic} statements
 are not intended to \emph{replace} locks or other synchonization primitives.
 When solving synchronization problems you should not directly use
-atomic statement but use the synchronization primitives that are available
+\emph{atomic} statements but use the synchronization primitives that are available
 to you.  But if you want to design a new synchronization primitive, then
@@
 \end{code}
-\caption{The \texttt{Lock} interface in the \texttt{synch} module.}
+\caption{The \texttt{Lock} interface and implementation in the \texttt{synch} module.}
 \label{fig:spinlocks}
@@
 You will find that using the \texttt{synchS} module often leads to searching a
-significantly larger search space than using the \texttt{synch} module.
+significantly larger state space than using the \texttt{synch} module.
 Part of the reason is that the \texttt{synchS} module keeps track of the order
-in which processes wait for a lock.
+in which processes wait for a lock, while the \texttt{synch} module does not.
 
@@
 \item Write a new synchronization module, similar to \autoref{fig:spinlocks}, but using
-atomic swap.
+atomic swap (similar to \autoref{ex:swap}).
 \item
@@
 \item \label{ex:trylock} Implement \texttt{trylock}(\textit{lk}) as an additional
-interface to the \texttt{synch} and \texttt{synchS} modules.
+interface for both the \texttt{synch} and \texttt{synchS} modules.
 This interface is like \texttt{lock}(\textit{lk}) but never blocks.  It

### 2020-07-28 book/paper.tex
Working on Chapter 10

@@
 
-What we want is a special kind of lock that allows either one writer
-or one or more readers to be in the critical section.  This is called
+What we want is a special kind of lock that allows either (i) one writer
+or (ii) one or more readers to be in the critical section.  This is called
 a \emph{reader/writer lock}~\cite{CHP71}.
@@
 \index{spin waiting}%
-processes spin in a loop until some desirable application-level condition is met.
+processes spin in a loop until some desirable application-level condition is met,
+wasting CPU cycles.
 \autoref{fig:inc} also has an example of busy-waiting: the \texttt{main}
@@
 A process trying to obtain a read or write lock in
-\autoref{fig:rwbusy} has to obtain the mutex repeatedly
-until the read or write lock is available.
+\autoref{fig:rwbusy} tries to obtain the mutex repeatedly.
 

### 2020-07-28 book/paper.tex
Latest

@@
   name=blocked process,
-  description={is a process that is waiting on a low-level synchronization primitive
-    and cannot terminate without the help of another process.
-      For example, a process that is waiting for a spinlock to become available}
+  description={is a process that cannot change the state unless another
+    process changes the state first.  For example, a process that is waiting
+    for a lock to become available.}
 }
@@
 terminate.  Harmony will not only print the non-terminating state with
-the shortest trace, but also the list of processes that are running
+the shortest trace, but also the list of processes
 at that state, along with their name tags and program counters.
@@
 \index{blocked process}%
-in a state if the process is waiting on a low-level synchronization
-variable and cannot terminate without the help of another process.
+if the process can only change the state if another process changes
+the state first.
 A process trying to
@@
 
-A context is considered \emph{blocked}
-\index{blocked process}%
-if there does not exist a
-sequence instructions that the process can execute by itself
-that leads to its termination.
-
 \chapter{The Harmony Virtual Machine}
@@
 the state are one per context.
-A terminating state is a state with an empty context bag or a state
-where all contexts are blocked.
 

### 2020-07-29 book/paper.tex
Latest

@@
 section and two readers try to enter.  The first reader successfully
-acquires \textit{rlock} and but blocks when trying to acquire
+acquires \textit{rlock} but blocks when trying to acquire
 \textit{rwlock}, which is held by the writer.  The second reader blocks
@@
 for: it checks to make sure that there are never a reader and a writer
-in the critical section, and there are never two readers in the critical
+in the critical section, and there are never two writers in the critical
 section.  Moreover, it checks progress: all executions eventually terminate.
@@
 in the critical section.  Indeed, we could have implemented the reader/writer
-lock as in \autoref{fig:rwmulti} using an ordinary lock, never allowing
-more than one process in the critical section.
+lock using an ordinary lock, never allowing
+more than one process in the critical section, and Harmony would not have
+complained.
 
@@
 can acquire the bound lock at the same time.
-\item It can be useful for a process to \emph{downgrade} its write lock
-to a read lock when its done writing but not done reading.  Implement
-\texttt{downgrade}.  Note, you cannot do this simply by releasing the
-write lock and then obtaining the read lock.
 \end{problems}
@@
 \item Assume that you have only binary semaphores.
-Implement counting semaphores using binary semaphores using split binary semaphores.
+Implement counting semaphores using split binary semaphores.
+\item With reader/writer locks,
+it can be useful for a process to \emph{downgrade} its write lock
+to a read lock when its done writing but not done reading.  Implement
+a \texttt{downgrade} method.  Note, you cannot do this simply by releasing the
+write lock and then obtaining the read lock, because the process needs
+to remain in the critical section.
 \item Implement a solution to the producer/consumer problem

### 2020-07-29 book/paper.tex
Updated documentation
Latest

@@
 \begin{problems}
-\item Write a new synchronization module, similar to \autoref{fig:spinlocks}, but using
-atomic swap (similar to \autoref{ex:swap}).
 \item
-Run \autoref{fig:incfixed} using (1) \texttt{synch}, (2), \texttt{synchS},
-and (3) your own module.  Report how many states were explored by Harmony for each
-version of the module.
+Run \autoref{fig:incfixed} using (i) \texttt{synch}, and (ii), \texttt{synchS}.
+Report how many states were explored by Harmony for each module.
 \item \label{ex:trylock} Implement \texttt{trylock}(\textit{lk}) as an additional
@@
 if the lock was already taken.
+Hint: you do not have to change the existing code.
 \end{problems}
@@
 
-Harmony supports filtering in \texttt{for loops}.
+Harmony supports nesting and filtering in \texttt{for loops}.
 For example:
 \begin{code}
-\texttt{for} $i$ in \{ 1..10 \} \texttt{such that} ($i$ \% 2) == 0: ...
-\end{code}
-\noindent iterates over the even numbers between 1 and 10.
+\texttt{for} $i$ in \{ 1..10 \}
+\texttt{for} $j$ in \{ 1..10 \}
+\texttt{such that} $i < j$: ...
+\end{code}
 

### 2020-07-30 book/paper.tex
Latest
Made pass over Chapter 12

@@
 \item
-Run \autoref{fig:incfixed} using (i) \texttt{synch}, and (ii), \texttt{synchS}.
+Run \autoref{fig:incfixed} using (i) \texttt{synch} and (ii) \texttt{synchS}.
 Report how many states were explored by Harmony for each module.
@@
 no more than three philosophers can eat at a time.
-But a lock can only count to one.
+Unfortunately, a lock can only count to one.
 
@@
 \begin{problems}
-\item Add an assertion that demonstrates that at most three
-Italian philosophers are eating at the same time?
+\item How might you check with Harmony that at most three
+Italian philosophers can eat at the same time?
 \item Implement a test program that checks to see that

### 2020-07-30 book/paper.tex
Latest

@@
 use \textbf{atomic} by all means.
+You can also use \textbf{atomic} statements in your test code.
+In fact, \texttt{assert} statements are executed atomically.
 
@@
 Or processes may be arranged in a manager/worker pattern, with a manager
-producing jobs and workers executing them in parallel.
+producing jobs and workers consuming and executing them in parallel.
 Or, in the client/server model,
@@
 some process may act as a \emph{server} that clients can send requests to
-and receive responses from.
+and receive responses from.  In that case, there is a bounded buffer
+for each client/server pair. Clients produce requests and
+consume responses, while the server consumes requests and produces responses.
 
@@
 the number of items that the producers still can produce and
-the number of items that the consumers have consumed.  In this case
-there are three items to produce initially.
+the number of items that the consumers have consumed.  In this
+description there are three items to produce initially.
 
@@
 \autoref{fig:bbtest} invokes the code of \autoref{fig:boundedbuffer}.
-We can run this through Harmony, but what would it test?  There are no
-assert statements in the code.
-We can split the functionality of this code into two pieces.  One is
-the synchronization part: producers should never run more than \texttt{NSLOTS}
-ahead of the consumers, but they should be able to produce up to \texttt{NSLOTS}
-items even if there are no consumers.  The other part is the data: we want
-to make sure that the items are sent through the buffer in order and without
-loss.  We will first focus on the synchronization.
+We can run this through Harmony, but what would it test?  There are
+no assert statements in the code.  We can split the functionality
+of this code into two and test each separately.  One is the
+synchronization part: producers should never run more than
+\texttt{NSLOTS} ahead of the consumers, but they should be able to
+produce up to \texttt{NSLOTS} items even if there are no consumers.
+The other part is the data: we want to make sure that the items are
+sent through the buffer in order and without loss.  We will first
+focus on the synchronization.
 

### 2020-07-30 book/paper.tex
Latest

@@
 \end{code}
-\caption{Script for testing producer/consumer synchronization
-(assuming the code is in file \texttt{BBsematest.hny}).}
+\caption{Script for testing producer/consumer synchronization.}
 \label{fig:pcscript}
@@
 Moreover, consumers add the tuples they received to the set
-\textit{received} (atomically---a lock could have been used here as well).)
+\textit{received} (atomically---a lock could have been used here as well).
 The \texttt{main()} process waits for all consumers to have received
@@
 until the stack is non-empty and then return an item.
-\item \label{ex:gpu} Implement a thread-safe GPU allocator by modifying
+Also implement test programs for your solution.
+\item \label{ex:gpu} Implement a thread-safe \emph{GPU allocator} by modifying
 \autoref{fig:gpu}.

### 2020-07-30 book/paper.tex
Latest

@@
 \end{code}
-\caption{\harmonylink{code/RWsbs.hny} Reader/Writer Lock implementation using Split Binary Semaphores.}
+\caption{\harmonylink{code/RWsbs.hny} Reader/Writer Lock using Split Binary Semaphores.}
 \label{fig:RWsplitsema}
@@
 If so, it passes the baton directly to one of those processes.
-If not, it simply gives up the baton.
+If not, it releases the baton for some other process to pick up at a
+later time.
 
@@
 writer to leave the critical region, while a process that wants to write may
-have to wait until all readers have left the critical section.  If there are $N$
+have to wait until all readers have left the critical section or until a writer has left.  If there are $N$
 waiting conditions, the SBS technique will use $N+1$ binary semaphores: one for
@@
 to enter the critical section and do not yet know if they need to wait.
-We shall call this the \textit{mutex}.
+We shall call this last semaphore the \textit{mutex}.
 
@@
 this state.
-To wit, check out \texttt{rlock\_acquire}.  It first checks to see if there
-are any writers.  If so, it increments \textit{r\_waiting}.  If not,
-it increments \textit{r\_entered}.
+For example, check out \texttt{rlock\_acquire}.
+It first checks to see if there
+are any writers.  If so, it increments \textit{r\_waiting}.
+If not, it increments \textit{r\_entered}.
 

### 2020-07-30 book/sources/pp.py

@@ -63,6 +63,7 @@ def isreserved(s):
         "def",
         "del",
         "dict",
+        "elif",
         "else",
         "fun",
         "for",

### 2020-07-30 harmony.py

@@ -468,6 +468,7 @@ def isreserved(s):
         "const",
         "def",
         "del",
+        "elif",
         "else",
         "end",
         "False",

### 2020-07-30 book/paper.tex
Latest

@@
 in the critical section, it vacates the semaphore associated with
-readers waiting.  If there are writers waiing and there are no readers
+readers waiting, passing the baton to the next reader.
+If there are writers waiting and there are no readers
 nor writers in the critical section, it vacates the semaphore associated
-with writers waiting.  If both conditions hold, it could vacate either
-one---it would not matter.  However, it can only vacate one of the semaphores.
-If none of the conditions hold, \texttt{V\_one()} vacates \textit{mutex}.
+with writers waiting, passing the baton to the next writer.
+If both conditions hold, it could vacate either
+one---it would not matter.
+However, it can only vacate one of the semaphores.
+In this implementation, it prefers to pass the baton to another reader.
+If none of the conditions hold, \texttt{V\_one()} vacates \textit{mutex},
+leaving the batom for a future process to pick up.
 
@@
 \textit{r\_waiting} and incrementing \textit{r\_entered}.  $p$
-then invokes \texttt{V\_one()} one more time.
+then invokes \texttt{V\_one()} to pass the baton to another waiting
+reader, if any, or to leave it for another process to pick up.
 
@@
 and there are two readers waiting.  It therefore vacates \textit{r\_sema}.
-The sum of the semaphores is now 1, because \textit{r\_sema} is 1.  This
+The sum of the semaphores is now~1, because \textit{r\_sema} is~1.  This
 guarantees that only a waiting reader, procuring \textit{r\_sema}, can enter
@@
 \item When it does, the reader decrements \textit{r\_waiting}
-from 2 to 1, and increments \textit{r\_entered} from 0 to 1.
+from 2 to 1, and increments \textit{r\_entered} from~0 to~1.
 The reader finally calls \texttt{V\_one()}.
@@
 \item The remaining reader decrements \textit{r\_waiting} from 1 to 0 and
-increments \textit{r\_entered} from 1 to 2.
+increments \textit{r\_entered} from~1 to~2.
 \item Finally, the remaining reader  calls \texttt{V\_one()}.

### 2020-07-30 book/paper.tex
Added an exercise

@@
 \item Assume that you have only binary semaphores.
-Implement counting semaphores using split binary semaphores.
+Implement counting semaphores using the split binary semaphores technique.
 \item With reader/writer locks,
 it can be useful for a process to \emph{downgrade} its write lock
-to a read lock when its done writing but not done reading.  Implement
-a \texttt{downgrade} method.  Note, you cannot do this simply by releasing the
-write lock and then obtaining the read lock, because the process needs
+to a read lock when its done writing but not done reading.  Add
+a \texttt{downgrade} method to the code in \autoref{fig:RWsplitsema}.
+How easy or hard would it be to do the same for the code in
+\autoref{fig:rw2lock}?
+Note, you cannot simply release the
+write lock and then obtain the read lock, because the process needs
 to remain in the critical section.
-\item Implement a solution to the producer/consumer problem
+\item \label{ex:onelane} Implement a solution to the producer/consumer problem
 (\autoref{fig:boundedbuffer}) using split binary semaphores.
 \item \label{ex:onelane} Cornell's campus features some one-lane bridges.
-Cars can only go in one direction at a time. Consider northbound
+On a one-lane bridge,
+cars can only go in one direction at a time. Consider northbound
 and southbound cars wanting to cross a one-lane bridge.
@@
 for southbound cars.
+\item Extend the solution to \autoref{ex:onelane} by implementing the
+requirement that at most 3 cars are allowed on the bridge.
 

### 2020-07-30 book/paper.tex
Latest

@@
 SBSs allow much control over which thread runs next and is therefore a
-good starting point for developing fair synchronization algorithm
+good starting point for developing fair synchronization algorithms.
 In this chapter, we will present two SBS-based fair reader/writer lock
@@
 
-In the late 70s, Xerox PARC developed a new programming language called
+In the late 70s, researchers at Xerox PARC,
+where among others the desktop and Ethernet
+were invented, developed a new programming language called
 Mesa~\cite{LR80}.
@@
 upon exit.
-Mesa monitors also have condition variable that a process can wait on.
+Mesa monitors also have condition variables that a process can wait on.
 Like in Hoare monitors, the \texttt{wait} operation releases the mutex.

### 2020-07-30 book/paper.tex
Improved monitor chapter

@@
 the baton until it leaves the monitor.  At that point, any process that
-was notified will have a chance to obtain the baton.
+was notified will have a chance to obtain the baton, but they compete
+with other processes also trying to obtain the baton.
 
@@
 
-On the left is a Hoare monitor.  When a process leaves the monitor
-(let's call that process $p$), it must be in the bathroom.
-It can choose to either let in a process
-in one of the bedrooms or a process in the hall, assuming there are any.
-To make the choice, $p$ checks for each bedroom if its condition holds
-and if there are processes in the bedroom.
-If there are such bedrooms, $p$ selects one and one of its processes into
-the bathroom.  Let's call that process $q$.  Importantly, when $q$ enters
-the bathroom, the condition that $q$ was waiting for still holds.
+On the left is a Hoare monitor.
+When some process $p$ is in the bathroom, it can invoke \textbf{signal}
+on some non-empty bedroom associated with a condition that holds.
+This select a process $q$ in that bedroom.
+As a result of invoking \textbf{signal}, $p$ goes back into the hall
+and $q$ enters the bathroom.
+Importantly, when $q$ enters the bathroom, the condition that $q$ was
+waiting for is guaranteed to hold.
 
 On the right is a Mesa monitor.  The only way into the bathroom is through
-the hall.  When process $p$ leaves the bathroom,
-if also checks each of the bedrooms to see if there is a process $q$ that
-can run.  But instead of letting $q$ go straight into the bathroom, $q$
-goes into the hall, joining any other processes that are also waiting to
-enter the bathroom.  Finally, when $p$ has left the bathroom, one of the
-processes in the hall is let into the bathroom.  But that may not be $q$.
+the hall.  When process $p$ invokes \textbf{notify} on a bedroom, it
+selects a process $q$ in that bedroom, if any.
+Process $q$ goes into the hall,
+joining any other processes that are also waiting to
+enter the bathroom.
+If there are no processes in that bedroom, \textbf{notify} is a no-op.
+Importantly, in either case $p$ remains in the bathroom.
+Finally, when $p$ leaves the bathroom, one of the
+processes in the hall can enter the bathroom.  But that may not be $q$.
 Assuming every process eventually leaves the bathroom and the system is
 fair, eventually $q$ will enter the bathroom.  However, it is no longer
-guaranteed that its condition holds.  Therefore, the first thing $q$ must
+guaranteed that the condition it was waiting for still holds.
+Therefore, the first thing $q$ must
 do is check, again, to see if the condition holds.  If not, it should go back
@@
 
-Mesa monitors provide another important option:
-when process $p$ leaves the bathroom, it can choose to let any number of
-processes in the bedrooms into the hall.  It can do so by calling
-\texttt{notify} repeatedly, each time letting one process in one of the
-bedrooms into the hall.  Using \texttt{notify} on an empty
-bedroom is considered a no-op in Mesa.  (Hoare monitors do not allow
-using \texttt{signal} on an empty bathroom.) Mesa also provides an
+Besides notifying processes, Mesa monitors provide another important option:
+it can choose to let any number of processes in the bedrooms into the hall.
+It can do so by calling
+\texttt{notify} repeatedly, each time letting one process in the specified
+bedroom into the hall.  Mesa also provides an
 operation called \texttt{notifyAll}
@@
 of processes, only one can be signaled at a time.  Each of those processes
-then has to signal the next until all processes have been signaled.
+then has to signal the next in a baton-passing style
+until all processes have been signaled.
 
@@
 obtain the lock.  Java objects also support \texttt{wait}, \texttt{notify},
-and \texttt{notifyAll}.  Java also supports explicit allocations of locks
+and \texttt{notifyAll}.
+In addition, Java supports explicit allocations of locks
 and condition variables.
@@
 \end{code}
-\caption{\harmonylink{code/RWcv.hny} Reader/Writer Lock implementation using Mesa-style condition variables.}
+\caption{\harmonylink{code/RWcv.hny} Reader/Writer Lock using Mesa-style condition variables.}
 \label{fig:RWcv}
@@
 
-\texttt{wait} adds the nametag of the process to the bag.  This increments the number of processes
-in the bag with the same context.  \texttt{wait} then waits until that count is restored to the
+\texttt{wait} adds the nametag of the process to the bag.
+This increments the number of processes in the bag with the same context.
+\texttt{wait} then waits until that count is restored to the
 value that it had upon entry to \texttt{wait}.  \texttt{notify} removes an arbitrary context from
@@
 terminates.
+
+While Hoare moniters embody the split binary semaphore approach,
+Mesa monitors are much closer to the busy-waiting approach.
+When comparing the busy-waiting version of reader/writer locks in \autoref{fig:rwbusy}
+and the Mesa version in \autoref{fig:RWcv}, you will see important similarities.
+Busy-waiting is ``fixed'' by placing a \textbf{wait} statement in the waiting loop.
+The Mesa code then has to be careful to invoke \textbf{notify} or \textbf{notifyAll}
+in the right places.
 

### 2020-07-30 book/paper.tex
Added some info

@@
 As a result of invoking \textbf{signal}, $p$ goes back into the hall
-and $q$ enters the bathroom.
+and $q$ enters the bathroom.\footnote{Per Brinch Hansen' monitors have
+a \textbf{continue} operation instead of \textbf{signal}.  The difference
+is that processes that invoke \textbf{continue} leave the monitor rather
+than returning to the hall.}
 Importantly, when $q$ enters the bathroom, the condition that $q$ was

### 2020-07-30 book/paper.tex
Latest

@@
 two forks to eat.  To start eating, a philosopher first picks up the fork on
-the left, then the fork on the right.  Each philospher likes to take breaks
+the left, then the fork on the right.  Each philosopher likes to take breaks
 from eating to think for a while.  To do so, the philosopher puts down both
 forks.  Each philosopher repeat this procedure.  Dijkstra had them repeating
-this for ever, but for the purposes of this book, philosphers can leave
+this for ever, but for the purposes of this book, philosophers can leave
 the table when they're not eating.
@@
 waits for other resources that it needs;
-\item \emph{No Preemption}: Resources cannot be forcibly taken back from processes that
+\item \emph{No Preemption}: Resources cannot be forcibly taken away from processes that
 allocated them;
@@
 arbitrary application-specific conditions.
-
+%
 \autoref{fig:dinerscv} demonstrates how this might be done.

### 2020-07-30 book/paper.tex
Fixed typo

@@
 handles one message at a time, serializing all access to the data it owns.
-As message queues are FCFS, starvation is prevented.
+As message queues are FCFS (First-Come-First-Served), starvation is prevented.
 
@@
 ``multiple producer / multiple consumer problem'' (with an unbounded buffer)
-using an actor (as illustrated in \autoref{fig:actorpic}.
+using an actor (as illustrated in \autoref{fig:actorpic}).
 \texttt{pc\_actor} is essentially a matchmaker: it matches
@@
 A client can send a message containing an integer to the server, and the
-server returns a response message containing the square of that message.
+server returns a response message containing the square of that integer.
 Implement this.
@@
 Develop a pipeline that computes Mersenne primes (primes that are one less
-than a factor of two).  Write four actors:
+than a power of two).  Write four actors:
 \begin{enumerate}
@@
 \item an actor that receives integers and forwards only those that are one
-less than a factor of two;
+less than a power of two;
 \item an actor that receives integers but otherwise ignores them.
@@
 Configure two versions of the pipeline, one that first checks if a number
-is prime and then if it is one less than a factor of two, the other
+is prime and then if it is one less than a power of two, the other
 in the opposite order.  Which do you think is better?

### 2020-07-30 book/paper.tex
Latest

@@
 \begin{problems}
-\item Change the code in \autoref{fig:boundedbuffer} so that
+\item \label{ex:nbproduce} Change the code in \autoref{fig:boundedbuffer} so that
 \texttt{produce()} fails if the buffer is full rather than block.
@@
 divide-by-zero.  An interrupt is caused by some peripheral device and can be handled in
-Harmony; an exception is an error.
+Harmony.
+In other words: an interrupt is a notification, while an exception is an error.
 \index{exception}
@@
 It is important that the \textit{main}() code re-enables interrupts after incrementing
-\textit{count}.  What would happen if \textit{main}() left interrupt disabled?
+\textit{count}.  What would happen if \textit{main}() left interrupts disabled?
 
@@
 \autoref{fig:trap6} gives an example of how this might be done.
-One important rule to remember is for a process to disable interrupts \emph{before}
+One important rule to remember is that a process should disable interrupts \emph{before}
 attempting to acquire a lock.
@@
 
-There is another important rule to keep in mind.  Just locks should never be held
+There is another important rule to keep in mind.  Just like locks should never be held
 for long, interrupts should never be disabled for long.  With locks the issue is to
-maximize concurrent performance.  The interrupts the issue is fast response to
+maximize concurrent performance.  For interrupts the issue is fast response to
 asynchronous events.  Because interrupts may be disabled only briefly, interrupt
@@
 synchronization calls such as \texttt{V} and \texttt{notify}, but calls such as
-\texttt{P} and \texttt{wait} should only be used if it's known that they will not
+\texttt{P} and \texttt{wait} should only be used if it is certain that they will not
 block for long.  Informally, interrupt handlers can be
@@
 it may block for a long time if the buffer is full.  Yet, it would be useful if,
-say, a keyboard interrupt handler could place an event on a shared queue.  Implement
-a new method \texttt{iproduce}(\textit{item}) that is interrupt-safe and does not
+say, a keyboard interrupt handler could place an event on a shared queue.
+Similar to \autoref{ex:nbproduce}, implement
+a new method \texttt{iproduce}(\textit{item}) that does not
 block.  Instead, it should return \texttt{False} if the buffer is full and \texttt{True}
 if the item was successfully enqueued.
+The method also needs to be interrupt-safe.
 \end{problems}

### 2020-07-30 book/paper.tex
Pass over Chapter 20

@@
 
-\chapter{Networking: Alternating Bit Protocol}
+\chapter{Alternating Bit Protocol}
 \label{ch:abp}
@@
 
-A distributed system is a concurrent system in which a collection
+A
+\emph{distributed system}
+\index{distributed system}%
+is a concurrent system in which a collection
 of processes communicate by message passing, much the same as
@@
 
-We need a network protocol that Alice and Bob can run.
+It is useful to create an abstract network that reliably sends messages
+between processes, much like the FIFO queue in the \texttt{synch} module.
+For this, we need a network protocol that Alice and Bob can run.
 In particular, it has to be the case that if Alice sends application
@@
 application messages $m_1, ..., m_i$ (safety).
-Also, if the network is fair, then if Alice sends application message
+Also, if the network is fair and Alice sends application message
 $m$, then eventually Bob should deliver $m$ (liveness).
@@
 to Bob and one from Bob to Alice.
-Alice and Bob each maintain a zero-initialized sequence number,
+Alice and Bob each maintain a zero-initialized
+\emph{sequence number},
+\index{sequence number}%
 \textit{s\_seq} and \textit{r\_seq} resp.
@@
 When Bob receives such a network message, Bob returns an
-an \emph{acknowledgment} to Alice, which is a network message
-containing the sequence number in the message that Bob received.
+\emph{acknowledgment}
+\index{acknowledgment}%
+to Alice, which is a network message
+containing the same sequence number as in the message that Bob received.
 
@@
 receives an acknowledgment with the same sequence number.
-When this happens, Alice increments her sequence number.  She is now
-ready to send the next message she wants to send to Bob.
+This is called \emph{retransmission}.
+\index{retransmission}%
+When she receives the desired sequence number, Alice increments her sequence number.
+She is now ready to send the next message she wants to send to Bob.
 Bob, on the other hand, waits until he receives a message matching
@@
 
-Because of the network's properties, we can model each
+We can model each
 channel as a variable that either contains a network message or nothing
@@
 
-Method \texttt{app\_send}($m$) tries \emph{retransmitting} $m$ until
+Method \texttt{app\_send}($m$) retransmits $m$ until
 an acknowledgment is received.
@@
 \begin{problems}
-\item \autoref{ex:cltsvr} explored the \emph{client/server}.  It is popular
+\item \autoref{ex:cltsvr} explored the \emph{client/server model}.  It is popular
 in distributed systems as well.

### 2020-07-30 book/paper.tex
Latest

@@
 the old value;
-\item \texttt{swap}($p$): sets \texttt{!}$p$ to \texttt{()} and returns
+\item \texttt{exch}($p$): sets \texttt{!}$p$ to \texttt{()} and returns
 the old value.
@@
 \textit{back} index, trying to find an item to return.
-To check an entry, it uses \texttt{swap()}
+To check an entry, it uses \texttt{exch()}
 to atomically remove an item from a slot if there is one.
@@
 \item Each slot in the array is only used once, which is inefficient;
-\item the \texttt{inc} and \texttt{swap} interlock instructions are not
+\item the \texttt{inc} and \texttt{exch} interlock instructions are not
 universally available on existing processors.

### 2020-07-30 book/paper.tex
Latest

@@
 synchronization algorithms~\cite{}.
+% TODO
 
@@
 a particular value is in the list.
-
 \glsadd{linearizable}%
@@
 The list has two ``book-end'' nodes with values \texttt{-inf} and
-\texttt{inf} (like the Python \texttt{math.inf} constants).
+\texttt{inf} (similar to the Python \texttt{math.inf} constant).
 Each node has a lock, a value, and \textit{next}, a pointer to the next node
@@
 An invariant of the algorithm is that at any point in time the
-list is ``complete'', starting with a \texttt{-inf} node and ending
+list is ``valid,'' starting with a \texttt{-inf} node and ending
 with a \texttt{inf} node.
@@
 reads the data structure, and then checks if the
-version number has not changed.  If it has, read-only operation is retried.
+version number has changed.  If so, the read-only operation is retried.
 Use a seqlock to implement a bank much like \autoref{ex:bank}, with

### 2020-07-30 book/paper.tex
Updated barrier

@@
 Between each round there is a so-called \emph{barrier} where processes wait
-until all processes have completed the round and before they start the
+until all processes have completed the previous round, before they start the
 next one.
 For example, in an iterative matrix algorithm, the matrix may be
-cut up into fragments.  During a round, the processes run concurrently
+cut up into fragments.  During a round, the processes run concurrently,
 one for each fragment.  The next round is not allowed to start
@@
 
-\autoref{fig:barrierdiagram} shows a high-level depiction with three
-processes.  Initially all processes are in round~0.  Then each process
+\autoref{fig:barrierdiagram} shows a high-level depiction of barrier synchronization
+with three and processes.  Initially all processes are in round~0.  Then each process
 can start and finish the round.  However, none of the processes can

### 2020-07-30 book/paper.tex
Fixed typo

@@
 \index{wait-free synchronization}%
-synchronization algorithms~\cite{}.
-% TODO
+synchronization algorithms.
+% TODO~\cite{}.
 
@@
 (As always, the Little Book of Semaphores~\cite{Downey09} is a good resource
-for solving semaphore problems.  Look for the double turnstile solution.)
+for solving synchronization problems with semaphores.
+Look for the \emph{double turnstile}
+\index{double turnstile}%
+solution.)
 \item Implement barrier synchronization with Mesa condition variables.
@@
 
-In each case, $s$ must be a set.
-For list comprehensions, the list will be sorted by $e$.
-For dictionary comprehensions, each $e$ will be mapped to \texttt{f}($e$).
+In each case, $s$ must be a set or a dictionary.
+\autoref{ap:details} provides more information on comprehensions.
 
@@
 \end{tabular}
-\vspace{1em}
-
-Harmony does not support Python-style iterators.
-To iterate through a dictionary, list, or tuple, use:
-\texttt{for $v$ in keys($e$):~S}.
 
@@
 \hline
-\texttt{subseq($t$, $b$, $b$)} & return a \emph{slice} of list $t$ starting
+\texttt{subseq($t$, $b$, $f$)} & return a \emph{slice} of list $t$ starting
 at index $b$ and ending just before $f$\\
@@
 \chapter{Harmony Language Details}
+\label{ap:details}
 

### 2020-07-30 book/sources/pp.py

@@ -84,6 +84,8 @@ def isreserved(s):
         "spawn",
         "setintlevel",
         "stop",
+        "such",
+        "that",
         "trap",
         "while"
     ]

### 2020-07-30 book/paper.tex
Latest

@@
 \hline
-Frame \texttt{m}($a$) & start method \texttt{m} with arguments $a$,
+Frame \texttt{m} $a$ & start method \texttt{m} with arguments $a$,
 initializing variables.  \\
@@
 executes an instruction, it goes from one context to another.
-All instructions update the program counter, and so no instruction
+All instructions update the program counter (\texttt{Jump} instructions
+are not allowed to jump to their own locations), and so no instruction
 leaves the context the same.

### 2020-07-30 book/paper.tex
Latest

@@
 is executing;
-\item The tag in a name tag is its argument by default.  Optionally
+\item The tag in a name tag the argument to the method by default.  Optionally
 \texttt{spawn} allows the tag to be specified explicitly;
@@
 \item A process terminates when it reaches the \texttt{Return} instruction
-of the method or when it hits a fault.  Faults include divide by zero,
+of the method or when it hits an exception.  Exceptions include divide by zero,
 reading a non-existent key in a dictionary, accessing a non-existent

### 2020-07-30 book/paper.tex
Latest

@@
 \hline
-variables & a dictionnary mapping atoms to values \\
+variables & a dictionary mapping atoms to values \\
 \hline
@@
 In Python, lists are objects.  Thus $x$ and $y$ point to the same list,
-and the assertion will fail.
+and the assertion would fail if executed by Python.
 In Harmony, lists are values.  So when $x$ is updated in Line~3, it does
@@
 \item When iterating over a set, the set is always traversed in order
-(see \autoref{ap:values} for a description how Harmony values are ordered);
+(see \autoref{ap:values} for how Harmony values are ordered);
 \item In case of a dictionary, the iteration is over the values of the
@@
 of a concurrent system.  My experimentation with using TLC on Peterson's
-Algorithm became an aha momment for me.
+Algorithm became an aha moment for me.
 
@@
 Finally, I would like to thank my family who had to suffer as I obsessed
-over writing the code and the book.
-
-% I wrote the Harmony compiler and this book in May and June of 2020,
+over writing the code and the book during the turbulent months of
+May and June 2020.
+
+% I wrote the Harmony compiler and this book
 % during the pandemic and the George Floyd demonstrations.

### 2020-07-30 book/paper.tex
Latest

@@
     process changes the state first.  For example, a process that is waiting
-    for a lock to become available.}
+    for a lock to become available}
 }
@@
   description={(aka spin-waiting) is when a process waits in a loop for some
-      application-defined condition}
+      application-defined condition instead of blocking}
 }
@@
   name=mutual exclusion,
-  description={is a property that two processes never enter the same critical section}
+  description={is the property that two processes never enter the same critical section}
 }
@@
   description={is a method in execution.  We do not make the distinction
-      between processes and threads.  A process has a program counter, a register,
-      and a stack}
+      between processes and threads.  A process has a current context and
+      updates its context every time it executes a machine instruction}
 }
@@
   description={is a synchronization problem whereby one or more producing processes
-    submit items and one or more consuming process want to receive them.  No item
+    submit items and one or more consuming processes want to receive them.  No item
     can get lost or forged or be delivered to more than one consumer, and producers

### 2020-08-03 book/paper.tex
Latest

@@
   name=blocked process,
-  description={is a process that cannot change the state unless another
+  description={is a process that cannot change the state or terminate unless another
     process changes the state first.  For example, a process that is waiting
@@
 \begin{figure}
-\includegraphics[width=\textwidth]{figures/Up1.png}
+\includegraphics[width=\textwidth]{figures/Up1-crop.pdf}
 \caption{The
@@
 \index{blocked process}%
-if the process can only change the state if another process changes
-the state first.
+if the process can only change the state or terminate if another process
+changes the state first.
 A process trying to

### 2020-08-03 harmony.py

@@ -4318,8 +4318,7 @@ def htmlrow(ctx, bag, node, code, scope, f, verbose):
             print("<td>choosing</td>", file=f)
         else:
             if ctx in node.edges:
-                (nxtnode, nxtctx, steps) = node.edges[ctx]
-                if nxtnode == node:
+                if node.isblocked(ctx):
                     print("<td>blocked</td>", file=f)
                 else:
                     print("<td>running</td>", file=f)

### 2020-08-05 book/paper.tex
Latest

--- 2020-07-30 book/paper.tex
@@
   name=blocked process,
-  description={is a process that cannot change the state unless another
-    process changes the state first.  For example, a process that is waiting
-    for a lock to become available}
+  description={is a process that cannot change the state or terminate or
+      can only do so after another process changes the state first.
+      For example, a process that is waiting for a lock to become available}
 }
@@
 \index{blocked process}%
-if the process can only change the state if another process changes
-the state first.
+if a process cannot change the state or terminate or can only do so if
+another process changes the state first.
 A process trying to
@@
 The only way forward is if the other process releases the lock.
+A process that is in an infinite loop is also considered blocked,
+although Harmony can often detect infinite loops specifically.
 

### 2020-08-09 book/paper.tex
New solutions

@@
   name=process,
-  description={is a method in execution.  We do not make the distinction
+  description={is code in execution.  We do not make the distinction
       between processes and threads.  A process has a current context and
@@
 \item Using a script, run \autoref{fig:incpy} 1000 times.
-For example, if you are using the bash shell, you can do the following:
+For example, if you are using the bash shell (in Linux, say), you can do the following:
 \begin{code}
@@
 the assertion should fail, and a trace should be reported.
-Now plug the following code \textbf{while} \texttt{True}: \textbf{pass};;
+Now insert the following code:
+\begin{code}
+\begin{tabbing}
+XX\=XXX\=XXX\kill
+\textbf{while} \texttt{True}: \\
+\> \textbf{pass}; \\
+;
+\end{tabbing}
+\end{code}
 just before entering the critical section

### 2020-08-09 book/paper.tex
Added set module

@@
 \item Bonus question:
-Implement the
-\href{https://en.wikipedia.org/wiki/Lamport%27s_bakery_algorithm}{Bakery Algorithm}
-or \href{https://en.wikipedia.org/wiki/Dekker%27s_algorithm}{Dekker's Algorithm}.
+Implement
+\href{https://en.wikipedia.org/wiki/Dekker%27s_algorithm}{Dekker's Algorithm},
+\href{https://en.wikipedia.org/wiki/Szymanski%27s_algorithm}{Szymański's Algorithm}, or the
+\href{https://en.wikipedia.org/wiki/Lamport%27s_bakery_algorithm}{Bakery Algorithm}.
+Note that the last one uses unbounded state, so you should modify the processes so they
+only try to enter the critical section a bounded number of times.
 \end{problems}
@@
 \end{tabular}
+
+\section{The \texttt{set} module}
+
+The \texttt{set} module
+\index{set module}%
+implements the following method:
+
+\vspace{1em}
+\begin{tabular}{|l|l|}
+\hline
+\texttt{issubset($s$, $t$)} & returns whether $s$ is a subset of $t$ \\
+\hline
+\end{tabular}
+
+\vspace{1em}
+For Python users: note that $s <= t$ does not check if $s$ is a subset of
+$t$ when $s$ and $t$ are sets, as ``$<=$'' implements a total order on all Harmony values
+including sets.
 

### 2020-08-10 book/paper.tex
Implemented any/all
Now also using 'all' in Python code

@@
 It increments \textit{count} and sets \textit{done}[\textit{self}] to \texttt{True}.
-Method \texttt{main} waits until \textit{done} flags are set to \texttt{True}.
+Method \texttt{main} waits until all \textit{done} flags are set to \texttt{True}.
+(\textbf{await} $c$ is semantically the same as \textbf{while not} $c$: \textbf{pass}.)
 After that, it verifies that the value of \textit{count} equals 2.
@@
 \hline
-\texttt{$e$ [not] in $e$} & first is any Harmony value, second is set \\
+\texttt{$e$ [not] in $e$} & first is any Harmony value, second is a set or dict \\
 \hline
@@
 \hline
-\texttt{min $e$, max $e$, len $e$} & a set or a dict \\
+\texttt{min $e$, max $e$, any $e$, all $e$, len $e$} & a set or a dict \\
 \hline
@@
 through $y$ inclusive.  It returns the empty set if $x > y$.
+If $d$ is a dictionary, $e$ \texttt{in} $d$ tests if $e$ is in the set of
+\emph{values} of $d$, not the set of \emph{keys} (this is different from Python).
 
@@
 (see \autoref{ap:values} for how Harmony values are ordered);
-\item In case of a dictionary, the iteration is over the values of the
+\item In case of a dictionary, the iteration is over the \emph{values} of the
 dictionary, but in the order of the keys.  In the case of lists, this
-works much the same as in Python;
+works much the same as in Python, but in the case of general dictionaries,
+Python iterates over the keys rather than the values;
 \item If you want to iterate over the keys of a dictionary $d$, use

### 2020-08-10 python/UpMany.py

@@ -9,7 +9,7 @@ def incrementer(self):
     done[self] = True
 
 def main():
-    while not (done[0] and done[1]):
+    while not all(done):
         pass
     assert count == 2*N, count
     print("Done")

### 2020-08-10 book/paper.tex
Latest

@@
 \begin{code}
-\begin{verbatim}
-while True:
-    pass;
-;
-\end{verbatim}
+\texttt{await False;}
 \end{code}
@@
 the assertion should fail, and a trace should be reported.
-Now insert the following code:
-\begin{code}
-\begin{tabbing}
-XX\=XXX\=XXX\kill
-\textbf{while} \texttt{True}: \\
-\> \textbf{pass}; \\
-;
-\end{tabbing}
+Now insert
+\begin{code}
+\texttt{await False;}
 \end{code}
@@
 win because writes to memory in Harmony are atomic.
-The process continues to be polite, waiting in a \textbf{while} loop
+The process continues to be polite, waiting
 until either the other process is nowhere near the critical section
@@
 It is easy to construct one: let process~0 be at label \texttt{@cs}
-and process~1 be at the start of the \textbf{while} loop.
+and process~1 be at the start of the \textbf{await} statement.
 Also imagine that in this particular state $\mathit{turn} = 1$.  Now let
 process~1 make a sequence of steps.  Because $\mathit{turn} = 1$,
-process~1 will break out of the while loop and also enter the critical
+process~1 will stop waiting and also enter the critical
 section, entering a state that is not exclusive.
@@
 Let $\mathcal{C}(i) = \mathit{flags}[1 - i] \land
-\mathit{turn} = 1 - i$, that is, the condition on the \textbf{while} loop
-for process~$i$.
+\mathit{turn} = 1 - i$, that is, the negation of the condition on the \textbf{await}
+statement for process~$i$.
 Let predicate $\mathcal{I}_p(i)$ be the following.

### 2020-08-10 book/paper.tex
Added await to list of statements
Latest

@@
 the critical section.
-Let $\mathcal{C}(i) = \mathit{flags}[1 - i] \land
-\mathit{turn} = 1 - i$, that is, the negation of the condition on the \textbf{await}
+Let $\mathcal{C}(i) = \lnot\mathit{flags}[1 - i] \lor
+\mathit{turn} = i$, that is, the condition on the \textbf{await}
 statement for process~$i$.
-Let predicate $\mathcal{I}_p(i)$ be the following.
+Let predicate $\mathcal{I}_p(i)$ be the following:
 if process $i$ is at label \texttt{@cs} (i.e., process $i$ is in the critical section),
-then $\mathcal{C}(i)$ does not hold or process $1-i$ is executing after setting
+then $\mathcal{C}(i)$ holds or process $1-i$ is executing after setting
 $\mathit{flags}[i]$ but still before setting \textit{turn} to $1-i$.
-More formally, $\mathcal{I}_p(i) = \mathtt{process}(i)@cs \Rightarrow (\lnot \mathcal{C}(i) \lor \mathtt{process}(1-i)@gate)$.
+More formally, $\mathcal{I}_p(i) = \mathtt{process}(i)@cs \Rightarrow \mathcal{C}(i) \lor \mathtt{process}(1-i)@gate$.
 For Peterson's Algorithm, an inductive invariant that works well is
@@
 following must hold:
-$(\lnot \mathcal{C}(0) \lor \mathtt{process}(1)@gate) \land
- (\lnot \mathcal{C}(1) \lor \mathtt{process}(0)@gate)$.
+$(\mathcal{C}(0) \lor \mathtt{process}(1)@gate) \land
+ (\mathcal{C}(1) \lor \mathtt{process}(0)@gate)$.
 We know that both flags are set.
@@
 (because they are both at label \texttt{@cs}),
-so this simplifies to $(\lnot \mathcal{C}(0)) \land (\lnot \mathcal{C}(1))$.
+so this simplifies to $\mathcal{C}(0) \land \mathcal{C}(1)$.
 So we conclude that $\mathit{turn} = 0 \land \mathit{turn} = 1$, a
@@
 page.  The page claimed that the following predicate is invariant:
-if process $i$ is in the critical section, then $\lnot \mathcal{C}(i)$ (i.e.,
+if process $i$ is in the critical section, then $\mathcal{C}(i)$ (i.e.,
 $\mathcal{I}_p(i)$ without the disjunct that process $1-i$ is at label \texttt{@gate}).
@@
 \hline
-\texttt{if $b$: S} & $b$ is a boolean, \texttt{S} a list of statements\\
+\texttt{if $b$: S else: S} & $b$ is a boolean, \texttt{S} a list of statements\\
 \hline
 \texttt{while $b$: S} & $b$ is a boolean, \texttt{S} a list of statements\\
+\hline
+\texttt{await $b$} & $b$ is a boolean\\
 \hline

### 2020-08-10 book/paper.tex
Latest

@@
 \hline
-Atom & \texttt{.example}, \texttt{.test1} \\
+Atom & \texttt{.example}, \texttt{.test1}, \texttt{.0x4A} \\
 \hline
@@
 is optional.
-A string must be enclosed by double quotes and is represented as
-a tuple of its characters.  Characters are one-character atoms.
+A string is represented as a tuple of its characters.
+Characters are one-character atoms, which can be expressed
+in hexadecimal unicode using the syntax \texttt{.0xXX}.
 A bag or multiset is a dictionary that maps a value to how many
@@
 \index{set module}%
-implements the following method:
+implements the following methods:
 
@@
 \texttt{issubset($s$, $t$)} & returns whether $s$ is a subset of $t$ \\
+\hline
+\texttt{issuperset($s$, $t$)} & returns whether $s$ is a superset of $t$ \\
 \hline
@@
 
+\section{Comments}
+
+Harmony supports the same commenting conventions as Python.
+In addition, Harmony supports nested multi-line comments
+of the form \texttt{(* comment *)}.
+
 \cleardoublepage

### 2020-08-10 book/paper.tex
Updated example with graphics

@@
 #states = 13 diameter = 1
+#components: 13
 no issues found
+Open file:///.../harmony.html for more information
 $
@@
 #states = 103 diameter = 1
+#components: 103
 no issues found
+Open file:///.../harmony.html for more information
 $
@@
 \begin{verbatim}
-Up.hny:1 def incrementer(self):
-    0 Jump 34
-    1 Frame incrementer(self)
-Up.hny:2     count = count + 1;
+code/Up.hny:1 def incrementer(self):
+   0 Jump 25
+   1 Frame incrementer self
+code/Up.hny:2     count = count + 1;
    2 Load count
@@
    5 Store count
-Up.hny:3     done[self] = True;
+code/Up.hny:3     done[self] = True;
    6 PushAddress done
@@
    11 Return
+code/Up.hny:5 def main():
+   12 Jump 25
+   13 Frame main ()
+code/Up.hny:6     await all(done);
+   14 Load done
+   15 1-ary all
+   16 JumpCond False 14
+code/Up.hny:7     assert count == 2, count;
+   17 AtomicInc
+   18 Load count
+   19 Push 2
+   20 2-ary ==
+   21 Load count
+   22 Assert2
+   23 AtomicDec
+   24 Return
+code/Up.hny:9 count = 0;
+   25 Push 0
+   26 Store count
+code/Up.hny:10 done = [ False, False ];
+   27 Push [False, False]
+   28 Store done
+code/Up.hny:11 spawn incrementer(0);
+   29 Push 0
+   30 Dup
+   31 Push PC(1)
+   32 Spawn
 \end{verbatim}
@@
 \begin{verbatim}
-#states = 79 diameter = 5
+#states = 71 diameter = 5
 ==== Safety violation ====
-__init__/()   [0,34-49]     50 { count=0, done=[False, False] }
-incrementer/0 [1-4]          5 { count=0, done=[False, False] }
-incrementer/1 [1-10]        11 { count=1, done=[False, True]  }
-incrementer/0 [5-10]        11 { count=1, done=[True,  True]  }
-main/() [13-17,20-24,26-31] 31 { count=1, done=[True,  True]  }
+__init__/()   [0,25-41] 41 { count: 0, done: [False, False] }
+incrementer/0 [1-4]      5 { count: 0, done: [False, False] }
+incrementer/1 [1-11]    11 { count: 1, done: [False, True ] }
+incrementer/0 [5-11]    11 { count: 1, done: [True,  True ] }
+main/()       [13-22]   22 { count: 1, done: [True,  True ] }
 >>> Harmony Assertion failed : 1
@@
 In this case, the first instruction is a \texttt{JUMP} instruction that sets the
-program counter to 34 (skipping over the code for the methods).
+program counter to 25 (skipping over the code for the methods).
 At program counter 1 is the code for the \texttt{incrementer} method.
@@
 In this case, Harmony reports a safety violation, in particular it
-reports that the assertion on line 9 failed and that
+reports that the assertion and that
 \textit{count} has value 1 instead of 2.
@@
 \item Process \texttt{implementer/1}
-executed instructions 1 through 10, storing 1 into
+executed instructions 1 through 11, storing 1 into
 into \textit{count} and storing \texttt{True} into \textit{done}[1];
 \item Process \texttt{implementer/0}
-continues execution, storing value 1 into \textit{count}
+continues execution, executing instructions 5 through 11
+storing value 1 into \textit{count}
 and storing \texttt{True} into \textit{done}[0].

### 2020-08-10 harmony.py

@@ -3790,10 +3790,18 @@ def find_shortest(bad):
             best_len = node.len
     return best_node
 
+def strvars(vars):
+    result = ""
+    for k in sorted(vars.keys()):
+        if result != "":
+            result += ", "
+        result += "%s: %s"%(k, strValue(vars[k]))
+    return "{ " + result + " }"
+
 def print_path(bad_node):
     path = genpath(bad_node)
     for (ctx, steps, states, vars) in path:
-        print(nametag2str(ctx.nametag), strsteps(steps))
+        print(nametag2str(ctx.nametag), strsteps(steps), ctx.pc, strvars(vars.d))
     if len(path) > 0:
         (ctx, steps, states, vars) = path[-1]
         if ctx.failure != None:

### 2020-08-11 book/paper.tex
Added some more details on the virtual machine

@@
 \item The frame pointer points to the current \emph{stack frame},
-which consists of the caller's frame and variables, the argument to
-the method, and the return address.
+which consists of the caller's frame pointer and variables, the argument to
+the method, an ``invocation type atom'' (\texttt{normal}, \texttt{interrupt}, or
+\texttt{process}), and the return address (in case of \texttt{normal}).
 \item A process terminates when it reaches the \texttt{Return} instruction
-of the method or when it hits an exception.  Exceptions include divide by zero,
+of the top-level method (when the stack frame is of type \texttt{process})
+or when it hits an exception.  Exceptions include divide by zero,
 reading a non-existent key in a dictionary, accessing a non-existent
-variable, as well as when an assertion fails;
+variable, infinite loops of microsteps, as well as when an assertion fails;
 \item The execution of a process in \emph{atomic mode} does not get interleaved
@@
 the state are one per context.
+
+Consecutive macro steps by the same process are called \emph{mega steps}.
+Each state maintains the shortest path to it from the initial state in terms
+of mega steps.
+The diameter of the graph is the length of the longest path found.
+
+If some states have a problem, the state with the shortest path is reported.
+Problematic states include states that experienced exceptions.
+If there are no exceptions, Harmony computes the strongly connected components (SCCs)
+of the graph (the number of such components are printed as part of the output).
+The sink SCCs should each consist of a terminal state without any processes.
+If not, again the state with the shortest path is reported.
+
+If there are no problematic states, Harmony reports ``no issues found'' and outputs
+in the HTML file the state with the longest path.
 
@@
 Because tuples and lists are dictionaries, the \texttt{del} statement is
-different Python.  For example, if $x$ = [.a, .b, .c], then
+different than in Python.  For example, if $x$ = [.a, .b, .c], then
 \texttt{del} $x$[1] results in $x$ = \texttt{dict}\{ 0:.a, 2:.c \}, \emph{not}

### 2020-08-11 book/paper.tex
Latest

@@
 \vspace{1em}
-For Python users: note that $s <= t$ does not check if $s$ is a subset of
+For Python programmers: note that $s <= t$ does not check if $s$ is a subset of
 $t$ when $s$ and $t$ are sets, as ``$<=$'' implements a total order on all Harmony values

### 2020-08-11 book/paper.tex
Latest

@@
 \index{dynamic allocation}%
-of nodes in the list using \texttt{recAlloc} and \texttt{recFree}.
+of nodes in the list using \texttt{malloc} and \texttt{free}.
 The read-only operation \texttt{lst\_contains} is wait\_free: it scans
@@
 \index{alloc module}%
-support thread-safe dynamic allocation of ``records.''
-Each record has two fields: \texttt{.data} and \texttt{.next} that
-can be used arbitrarily, although \texttt{.next} is intended to be
-used as a pointer to another record.
-There are just two methods:
+support thread-safe (but not interrupt-safe) dynamic allocation of
+dictionaries.  There are just two methods:
 
@@
 \hline
-\texttt{recAlloc()} & allocate a record \\
-\hline
-\texttt{recFree($r$)} & free record $r$ \\
+\texttt{malloc()} & allocate a dictionary \\
+\hline
+\texttt{free($r$)} & free an allocated dictionary $r$ \\
 \hline
@@
 The usage is similar to \texttt{malloc} and \texttt{free} in C.
-\texttt{recAlloc} returns \texttt{None} when running out of memory.
+\texttt{malloc}() returns \texttt{None} when running out of memory.
 

### 2020-08-11 book/paper.tex
Latest

@@
 In this chapter, we will have a look at how one can develop concurrent
-code in which processes do not have to wait for other processes.
+code in which processes do not have to wait for other processes to complete their
+ongoing operations.
 
@@
 most accesses are read-only and the performance of updates is less
-critical~\cite{}.
+critical. %~\cite{}.
 \autoref{fig:lst1} and \autoref{fig:lst2} implements an ordered linked
@@
 \end{code}
-\caption{\harmonylink{code/lst1.hny} List with non-blocking read operations, part 1.}
+\caption{\harmonylink{code/lst1.hny} List with non-blocking read operation, part 1.}
 \label{fig:lst1}
@@
 \end{code}
-\caption{\harmonylink{code/lst2.hny} List with non-blocking read operations, part 2.}
+\caption{\harmonylink{code/lst2.hny} List with non-blocking read operation, part 2.}
 \label{fig:lst2}

### 2020-08-11 book/sources/pp.py

@@ -64,6 +64,7 @@ def isreserved(s):
         "const",
         "def",
         "del",
+        "dict",
         "elif",
         "else",
         "end",

### 2020-08-12 book/paper.tex
Updated alloc module

@@
 \index{alloc module}%
-support thread-safe (but not interrupt-safe) dynamic allocation of
-dictionaries.  There are just two methods:
+maintains a pool of ``memory locations'' and
+supports thread-safe (but not interrupt-safe) dynamic allocation of
+memory locations.  There are just two methods:
 
@@
 \hline
-\texttt{malloc()} & allocate a dictionary \\
-\hline
-\texttt{free($r$)} & free an allocated dictionary $r$ \\
+\texttt{malloc()} & allocate and return a pointer to a memory location \\
+\hline
+\texttt{free($p$)} & free an allocated memory location $p$ \\
 \hline

### 2020-08-12 harmony.py

@@ -391,13 +391,13 @@ def malloc():
     lock(?alloc_lock);
     result = alloc_flist;
     if result != None:
-        alloc_flist = (!result).nxt;
+        alloc_flist = !result;
     ;
     unlock(?alloc_lock);
 ;
 def free(r):
     lock(?alloc_lock);
-    (!r).nxt = alloc_flist;
+    !r = alloc_flist;
     alloc_flist = r;
     unlock(?alloc_lock);
 ;

### 2020-08-12 book/paper.tex
Latest

@@
 Report how many states were explored by Harmony for each module.
+\item \label{ex:xy} \autoref{fig:xy} shows a Harmony program with two variables $x$
+(initially 0) and $y$ (initially 100) that can be accessed through methods
+\texttt{setX} and \texttt{getXY}.  An application invariant is that \texttt{getXY}
+should return a pair that sums to 100.  Add the necessary synchronization code.
 \item \label{ex:trylock} Implement \texttt{trylock}(\textit{lk}) as an additional
@@
 \end{problems}
+
+\begin{figure}
+\begin{code}
+\harmonysource{xy}
+\end{code}
+\caption{$x + y = 100$.}
+\label{fig:xy}
+\end{figure}
 
@@
 for three processes.
+\item Optimize your solution to \autoref{ex:xy} to use reader/writer locks.
 \item \label{ex:boundlock} Using busy waiting, implement a ``bound lock'' that allows

### 2020-08-13 book/paper.tex
Latest

@@
 \end{code}
-\caption{$x + y = 100$.}
+\caption{Imcomplete code with desired invariant $x + y = 100$.}
 \label{fig:xy}
@@
 \index{alloc module}%
-maintains a pool of ``memory locations'' and
 supports thread-safe (but not interrupt-safe) dynamic allocation of
@@
 The usage is similar to \texttt{malloc} and \texttt{free} in C.
-\texttt{malloc}() returns \texttt{None} when running out of memory.
+\texttt{malloc}() is specified to \texttt{None} when running out of
+memory, although this is an impossible outcome in the current
+implementation of the module.
 

### 2020-08-13 book/paper.tex
Latest

@@
 
+\section{Dynamic Allocation}
+
+\begin{figure}
+\begin{code}
+\harmonysource{stacktest}
+\end{code}
+\caption{\harmonylink{code/stacktest.hny} Testing a stack implementation.}
+\label{fig:stacktest}
+\end{figure}
+
+\begin{figure}
+\begin{code}
+\harmonysource{stack1}
+\end{code}
+\caption{\harmonylink{code/stack1.hny} Stack implemented using a dynamically updated list.}
+\label{fig:stack1}
+\end{figure}
+
+\begin{figure}
+\begin{code}
+\harmonysource{stack2}
+\end{code}
+\caption{\harmonylink{code/stack2.hny} Stack implemented using static lists.}
+\label{fig:stack2}
+\end{figure}
+
+\begin{figure}
+\begin{code}
+\harmonysource{stack3}
+\end{code}
+\caption{\harmonylink{code/stack3.hny} Stack implemented using a recursive tuple data structure.}
+\label{fig:stack3}
+\end{figure}
+
+\begin{figure}
+\begin{code}
+\harmonysource{stack4}
+\end{code}
+\caption{\harmonylink{code/stack4.hny} Stack implemented using a linked list.}
+\label{fig:stack4}
+\end{figure}
+
+Harmony supports various options for dynamic allocation.
+By way of example, consider a stack.
+\autoref{fig:stacktest} presents a test program for a stack.
+We present four different stack implementations to illustrate
+options for dynamic allocation:
+\begin{enumerate}
+\item[] \autoref{fig:stack1} uses a single list to represent the
+stack.  It is updated to perform \texttt{push} and \texttt{pop} operations;
+\item[] \autoref{fig:stack2} also uses a list but, instead of updating
+the list, it replaces the list with a new one for each operation;
+\item[] \autoref{fig:stack3} represents a stack as a recursively nested tuple
+$(v, f)$, where $v$ is the element on top of the stack and $r$ is a stack
+that is the remainder;
+\item[] \autoref{fig:stack4} implements a stack as a linked list with nodes
+allocated using the \texttt{alloc} module.
+\end{enumerate}
+
+While the last option is the most versatile (it allows cyclic
+data structures), Harmony does not support garbage collection
+for memory allocated this way and so allocated memory that is no
+longer in use must be explicitly released using \texttt{free}.
+
 \section{Comments}

### 2020-08-13 harmony.py

@@ -236,10 +236,7 @@ def enqueue(q, item):
 """,
     "list": """# return s[b:e]
 def subseq(s, b, e):
-    result = [];
-    for x in {b..e-1}:
-        result[x - b] = s[x];
-    ;
+    result = [ s[x] for x in {b..e-1} ];
 ;
 
 def append(s, e):
@@ -267,26 +264,17 @@ def set(a):
 
 # convert set into a (sorted) list
 def list(a):
-    result = [];
-    for v in a:
-        result += [v,];
-    ;
+    result = [ v for v in a ];
 ;
 
 # convert values of a dictionary into a list
 def values(d):
-    result = [];
-    for k in keys d:
-        result += [ d[k], ];
-    ;
+    result = [ d[k] for k in keys d ];
 ;
 
 # like Python .items()
 def items(d):
-    result = [];
-    for k in keys d:
-        result += [ (k, d[k]), ];
-    ;
+    result = [ (k, d[k]) for k in keys d ];
 ;
 
 # like Python enumerate()
@@ -326,10 +314,7 @@ def sorted(d):
 # like Python reversed()
 def reversed(d):
     n = len(d);
-    result = [];
-    for i in { 1..n }:
-        result += [ d[n - i], ];
-    ;
+    result = [ d[n-i] for i in { 1..n } ];
 ;
 
 # turn a list into a bag (multiset)
@@ -385,7 +370,9 @@ def issuperset(s, t):
     result = (s & t) == t;
 ;
 """,
-    "alloc": """def malloc():
+    "alloc": """import synch;
+
+def malloc():
     lock(?alloc_lock);
     if alloc_flist == None:
         let i = len(alloc_pool):

### 2020-08-14 book/paper.tex
Added section on operator precedence

@@
 
+\section{Operator Precedence}
+
+In Harmony, there is no syntactic difference between applying an argument to a function
+or an index to a dictionary.  Both use the syntax \textit{a b c ...}.
+We call this \emph{application}, and application is left-associative.
+So \textit{a b c} is interpreted as (\textit{a b}) $c$: $b$ is applied to $a$,
+and then $c$ is applied to the result.
+For readability, it sometimes helps to write $a(b)$ for function application and
+$a[b]$ for indexing.  In case $b$ is an atom, you can also write $a.b$ for indexing.
+
+There are three classes of precedence.
+Application has the highest precedence.  So \texttt{!}\textit{a b} is interpreted as
+\texttt{!}(\textit{a b}) and \textit{a b} \texttt{+} \textit{c d} is interpreted as
+(\textit{a b}) \texttt{+} (\textit{c d}).
+Unary operators are have the next highest precedence,
+and the remaining operators have the lowest precedence.
+So $-2 + 3$ evaluates to 1, not $-5$.
+
+Associative operators ($+$, $*$, $|$, $\string&$, $\string^$, \textbf{and}, \textbf{or})
+are interpreted as general $n$-ary operators, and you are allowed to write
+$a + b + c$.  However, $a - b - c$ is illegal, as is any combination of operators such
+as $a + b < c$.  In such cases you have to add parentheses or brackets to indicate what
+the intended evaluation order is, such as $(a + b) < c$.
+
+In almost all expressions, subexpressions are evaluated left to right.  So $a[b] + c$
+first evaluates $a$, then $b$ (and then applies $b$ to $a$), and then $c$.  The one
+exception is the expression $a$ \textbf{if} $c$ \textbf{else} $b$, where $b$ is evaluated
+first.  In that expression, only $a$ or $b$ is evaluated depending on the value of $c$.
+In the expression $a$ \textbf{and} $b$ \textbf{and} $...$, evaluation is left
+to right but stops once one of the subexpressions evaluates to \textbf{False}.
+Similarly for \textbf{or}, where evaluation stops once one of the subexpressions
+evaluates to \texttt{True}.
+
 \section{Comments}

### 2020-08-14 book/paper.tex
Latest

@@
 
+\section{Operator Precedence}
+
+In Harmony, there is no syntactic difference between applying an argument to a function
+or an index to a dictionary.  Both use the syntax \textit{a b c ...}.
+We call this \emph{application}, and application is left-associative.
+So \textit{a b c} is interpreted as (\textit{a b}) $c$: $b$ is applied to $a$,
+and then $c$ is applied to the result.
+For readability, it may help to write $a(b)$ for function application and
+$a[b]$ for indexing.  In case $b$ is an atom, you can also write $a.b$ for indexing.
+
+There are three classes of precedence.
+Application has the highest precedence.  So \texttt{!}\textit{a b} is interpreted as
+\texttt{!}(\textit{a b}) and \textit{a b} \texttt{+} \textit{c d} is interpreted as
+(\textit{a b}) \texttt{+} (\textit{c d}).
+Unary operators are have the next highest precedence,
+and the remaining operators have the lowest precedence.
+So $-2 + 3$ evaluates to 1, not $-5$.
+
+Associative operators ($+$, $*$, $|$, $\string&$, $\string^$, \textbf{and}, \textbf{or})
+are interpreted as general $n$-ary operators, and you are allowed to write
+$a + b + c$.  However, $a - b - c$ is illegal, as is any combination of operators with an
+arity larger than one, such as $a + b < c$.
+In such cases you have to add parentheses or brackets to indicate what
+the intended evaluation order is, such as $(a + b) < c$.
+
+In almost all expressions, subexpressions are evaluated left to right.  So $a[b] + c$
+first evaluates $a$, then $b$ (and then applies $b$ to $a$), and then $c$.  The one
+exception is the expression $a$ \textbf{if} $c$ \textbf{else} $b$, where $c$ is evaluated
+first.  In that expression, only $a$ or $b$ is evaluated depending on the value of $c$.
+In the expression $a$ \textbf{and} $b$ \textbf{and} $...$, evaluation is left
+to right but stops once one of the subexpressions evaluates to \textbf{False}.
+Similarly for \textbf{or}, where evaluation stops once one of the subexpressions
+evaluates to \texttt{True}.
+
+As an aside:
+the expression $a$ \textbf{not in} $b$ is equivalent to \textbf{not} ($a$ \textbf{in} $b$).
+Harmony generalizes this construct for any unary operator besides \textbf{not}.
+So $a$ \textbf{not and} $b$ is the same as \textbf{not} ($a$ \textbf{and} $b$).  
+For those familiar with logic gates, \textbf{not and} is the equivalent of \texttt{NAND}.
+
 \section{Tuples, Lists, and Pattern Matching}
@@
 
-\section{Operator Precedence}
-
-In Harmony, there is no syntactic difference between applying an argument to a function
-or an index to a dictionary.  Both use the syntax \textit{a b c ...}.
-We call this \emph{application}, and application is left-associative.
-So \textit{a b c} is interpreted as (\textit{a b}) $c$: $b$ is applied to $a$,
-and then $c$ is applied to the result.
-For readability, it sometimes helps to write $a(b)$ for function application and
-$a[b]$ for indexing.  In case $b$ is an atom, you can also write $a.b$ for indexing.
-
-There are three classes of precedence.
-Application has the highest precedence.  So \texttt{!}\textit{a b} is interpreted as
-\texttt{!}(\textit{a b}) and \textit{a b} \texttt{+} \textit{c d} is interpreted as
-(\textit{a b}) \texttt{+} (\textit{c d}).
-Unary operators are have the next highest precedence,
-and the remaining operators have the lowest precedence.
-So $-2 + 3$ evaluates to 1, not $-5$.
-
-Associative operators ($+$, $*$, $|$, $\string&$, $\string^$, \textbf{and}, \textbf{or})
-are interpreted as general $n$-ary operators, and you are allowed to write
-$a + b + c$.  However, $a - b - c$ is illegal, as is any combination of operators such
-as $a + b < c$.  In such cases you have to add parentheses or brackets to indicate what
-the intended evaluation order is, such as $(a + b) < c$.
-
-In almost all expressions, subexpressions are evaluated left to right.  So $a[b] + c$
-first evaluates $a$, then $b$ (and then applies $b$ to $a$), and then $c$.  The one
-exception is the expression $a$ \textbf{if} $c$ \textbf{else} $b$, where $b$ is evaluated
-first.  In that expression, only $a$ or $b$ is evaluated depending on the value of $c$.
-In the expression $a$ \textbf{and} $b$ \textbf{and} $...$, evaluation is left
-to right but stops once one of the subexpressions evaluates to \textbf{False}.
-Similarly for \textbf{or}, where evaluation stops once one of the subexpressions
-evaluates to \texttt{True}.
-
 \section{Comments}

### 2020-08-16 book/paper.tex
Latest

@@
 the expression $a$ \textbf{not in} $b$ is equivalent to \textbf{not} ($a$ \textbf{in} $b$).
-Harmony generalizes this construct for any unary operator besides \textbf{not}.
-So $a$ \textbf{not and} $b$ is the same as \textbf{not} ($a$ \textbf{and} $b$).  
+Harmony generalizes this construct for any pair of a unary (except '$-$') and a binary operator.
+In particular, $a$ \textbf{not and} $b$ is the same as \textbf{not} ($a$ \textbf{and} $b$).  
 For those familiar with logic gates, \textbf{not and} is the equivalent of \texttt{NAND}.

### 2020-08-16 harmony.py

@@ -2258,7 +2258,7 @@ class NaryRule(Rule):
         args = [ast]
         op = t[0]
         invert = None
-        if isunaryop(op[0]):      # mostly for use in "a not in s", but more general
+        if isunaryop(op[0]) and op[0] != '-':  # mostly for use in "a not in s", but more general
             invert = op
             t = t[1:]
             op = t[0]

### 2020-08-18 book/paper.tex
Latest

@@
 \href{https://en.wikipedia.org/wiki/Dekker%27s_algorithm}{Dekker's Algorithm},
+\href{https://en.wikipedia.org/wiki/Eisenberg_%26_McGuire_algorithm}{Eisenstein and McGuire's Algorithm},
 \href{https://en.wikipedia.org/wiki/Szymanski%27s_algorithm}{Szymański's Algorithm}, or the
-\href{https://en.wikipedia.org/wiki/Lamport%27s_bakery_algorithm}{Bakery Algorithm}.
+\href{https://en.wikipedia.org/wiki/Lamport%27s_bakery_algorithm}{Lamport's Bakery Algorithm}.
 Note that the last one uses unbounded state, so you should modify the processes so they
@@
 \end{figure}
+
+Harmony does not have a \textbf{return} statement.  (You can assign a return value
+to a method by setting the \textit{result} variable.)  Neither does it support
+\textbf{break} or \textbf{continue} statements in loops.  One reason for their absence is
+that, particularly in concurrent programming, such control flow directions are highly
+error-prone.  It's easy to forget to, say, release a lock when returning a value in the
+middle of a method, and a major source of bugs in practice.
 

### 2020-08-18 book/paper.tex
Added a new chapter on concurrent data structures

@@
 that supports locks and other synchronization primitives.
+\item \autoref{ch:cds} provides an introduction to building concurrent
+data structures.
 \item \autoref{ch:rdwrbusy} and \autoref{ch:rdwr2} present the
@@
 
+\chapter{Concurrent Data Structures}
+\label{ch:cds}
+
+The most common use for locks is in building concurrent data structures.
+By way of example, we will demonstrate how to build a concurrent queue.
+The queue will have the following API:
+\begin{itemize}
+\item $q = \mathtt{Qnew}()$: allocate a new queue;
+\item $Qenqueue(q, v)$: add $v$ to the tail of the queue;
+\item $r = Qdequeue(q)$: returns $r = ()$ if $q$ is empty or $r = (v,)$ (a singleton tuple)
+if $v$ was at the head of the queue.
+\end{itemize}
+See \autoref{fig:queuetest} for a simple test program that uses the queue.
+
+We will first implement the queue as a linked list.
+The implementation in \autoref{fig:queue}
+uses the \texttt{alloc} module for dynamic allocation
+\index{dynamic allocation}%
+of nodes in the list using \texttt{malloc}() and \texttt{free}().
+\texttt{malloc}() returns a new memory location, which should be released
+with \texttt{free}() when it is no longer in use.
+The queue maintains a pointer to the first and last elements in the list,
+which are \textbf{None} if the queue is empty.
+
+\texttt{Qnew()} returns a new queue object consisting of a \textbf{None} head
+and tail pointer and a lock.
+\texttt{Qenqueue(q, v)} and \texttt{Qdequeue(q)} both take a pointer $q$ to the
+queue object because both may modify the queue.
+Before they access the value of the head or tail of the queue they first obtain
+the lock.
+When they are done, they release the lock.
+
+The implementation in \autoref{fig:queueMS}~\cite{MS98} is similar but uses
+separate locks for the head and tail, allowing an enqueue and dequeue operation
+to proceed concurrently.  To avoid contention between the head and the tail, the
+queue uses a dummy item on the linked list.
+
+\begin{figure}
+\begin{code}
+\harmonysource{queuetest}
+\end{code}
+\caption{Test program for a concurrent queue.}
+\label{fig:queuetest}
+\end{figure}
+
+\begin{figure}
+\begin{code}
+\harmonysource{queue}
+\end{code}
+\caption{A basic concurrent queue data structure.}
+\label{fig:queue}
+\end{figure}
+
+\begin{figure}
+\begin{code}
+\harmonysource{queueMS}
+\end{code}
+\caption{A queue with separate locks for enqueuing and dequeuing items.}
+\label{fig:queueMS}
+\end{figure}
+
 \chapter{Reader/Writer Locks using Busy Waiting}
@@
 
-The implementation uses the \texttt{alloc} module for dynamic allocation
-\index{dynamic allocation}%
-of nodes in the list using \texttt{malloc} and \texttt{free}.
 The read-only operation \texttt{lst\_contains} is wait\_free: it scans

### 2020-08-18 harmony.py

@@ -1337,7 +1337,8 @@ class AssertOp(Op):
                         " must be a boolean: " + strValue(cond)
             return
         if not cond:
-            context.failure = "Harmony Assertion failed "
+            (lexeme, file, line, column) = self.token
+            context.failure = "Harmony Assertion (file=%s, line=%d) failed"%(file, line)
             if self.exprthere:
                 context.failure += ": " + strValue(expr)
             return

### 2020-08-19 book/paper.tex
Now supports ->

@@
 error-prone.  It's easy to forget to, say, release a lock when returning a value in the
-middle of a method, and a major source of bugs in practice.
+middle of a method, a major source of bugs in practice.
 
@@
 
+Often, pointers point to dictionaries, and so if $p$ is such a pointer,
+then $!(p).\mathtt{field}$ would evaluate to the specified field in the dictionary.
+Note that the parentheses in this expression
+are needed, as \texttt{!}\textit{p}\texttt{.field} would wrongly evaluate
+\texttt{!}(\textit{p}\texttt{.field}).
+$!(p).\mathtt{field}$ is such a common expression that, like C, Harmony supports the
+shorthand $p-$$>$\texttt{field}, which greatly improves readability.
+
 \autoref{fig:petersonmethods} again shows Peterson's algorithm,
@@
 a pointer to a mutex and the process identifier (0 or 1).
-(\texttt{!}\textit{pm})\texttt{.turn} is the value of the \texttt{.turn} key
+$\mathit{pm}\rightarrow\mathtt{turn}$ is the value of the \texttt{.turn} key
 in the dictionary that \textit{pm} points to.
-Note that the parentheses in this expression
-are needed, as \texttt{!}\textit{pm}\texttt{.turn} would wrongly evaluate
-\texttt{!}(\textit{pm}\texttt{.turn}).
 
@@
 \end{code}
-\caption{The \texttt{Lock} interface and implementation in the \texttt{synch} module.}
+\caption{\harmonylink{modules/synch.hny} The \texttt{Lock} interface and implementation in the \texttt{synch} module.}
 \label{fig:spinlocks}
@@
 \end{code}
-\caption{The \texttt{Lock} interface in the \texttt{synchS} module uses suspension.}
+\caption{\harmonylink{modules/syncS.hny} The \texttt{Lock} interface in the \texttt{synchS} module uses suspension.}
 \label{fig:suspension}
@@
 \end{code}
-\caption{Imcomplete code with desired invariant $x + y = 100$.}
+\caption{\harmonylink{code/xy.hny} Incomplete code with desired invariant $x + y = 100$.}
 \label{fig:xy}
@@
 \end{code}
-\caption{Test program for a concurrent queue.}
+\caption{\harmonylink{code/queuetest.hny} Test program for a concurrent queue.}
 \label{fig:queuetest}
@@
 \end{code}
-\caption{A basic concurrent queue data structure.}
+\caption{\harmonylink{code/queue.hny}A basic concurrent queue data structure.}
 \label{fig:queue}
@@
 \end{code}
-\caption{A queue with separate locks for enqueuing and dequeuing items.}
+\caption{\harmonylink{code/queueMS.hny} A queue with separate locks for enqueuing and dequeuing items.}
 \label{fig:queueMS}
@@
 supports thread-safe (but not interrupt-safe) dynamic allocation of
-memory locations.  There are just two methods:
+shared memory locations.  There are just two methods:
 
@@
 The usage is similar to \texttt{malloc} and \texttt{free} in C.
-\texttt{malloc}() is specified to \texttt{None} when running out of
+\texttt{malloc}() is specified to return \texttt{None} when running out of
 memory, although this is an impossible outcome in the current

### 2020-08-19 book/sources/pp.py

@@ -113,7 +113,7 @@ def isbinaryop(s):
         "-", "+", "*", "/", "%", "<", "<=", ">", ">="
     ];
 
-tokens = [ "==", "!=", "<=", ">=", ".." ]
+tokens = [ "==", "!=", "<=", ">=", "..", "->" ]
 
 def putchar(c):
     print(c, end="")
@@ -202,6 +202,8 @@ def lexer(s, file, silent):
             if not silent:
                 if found in [ "<=", ">=" ]:
                     print("$%s$"%found, end="")
+                elif found == "->":
+                    print("$\\rightarrow$", end="")
                 else:
                     print(found, end="")
             s = s[len(found):]

### 2020-08-19 book/paper.tex
Latest

@@
 of nodes in the list using \texttt{malloc}() and \texttt{free}().
-\texttt{malloc}() returns a new memory location, which should be released
-with \texttt{free}() when it is no longer in use.
-The queue maintains a pointer to the first and last elements in the list,
-which are \textbf{None} if the queue is empty.
+\texttt{malloc}($v$) returns a new memory location initialized to $v$,
+which should be released with \texttt{free}() when it is no longer in use.
+The queue maintains a \texttt{head} pointer to the first element in the list
+and a \texttt{tail} pointer to the last element in the list.
+The \texttt{head} pointer is \textbf{None} if and only if the queue is empty.
 
@@
 and tail pointer and a lock.
-\texttt{Qenqueue(q, v)} and \texttt{Qdequeue(q)} both take a pointer $q$ to the
+$\mathtt{Qenqueue}(q, v)$ and $\mathtt{Qdequeue}(q)$ both take a pointer $q$ to the
 queue object because both may modify the queue.
@@
 The implementation in \autoref{fig:queueMS}~\cite{MS98} is similar but uses
-separate locks for the head and tail, allowing an enqueue and dequeue operation
-to proceed concurrently.  To avoid contention between the head and the tail, the
-queue uses a dummy item on the linked list.
+separate locks for the head and tail, allowing an enqueue and a dequeue operation
+to proceed concurrently.  To avoid contention between the head and the tail that
+requires both locks, the queue uses a dummy node at the head of the linked list.
+Except of the initial state of the queue, the dummy node is the last node that
+was dequeued.  Note that neither the \texttt{head} nor \texttt{tail} pointer are
+ever \textbf{None}.
 
@@
 \end{figure}
+
+\section*{Exercises}
+\begin{problems}
+\item \label{ex:qcontains} Add a method $\mathtt{Qcontains}(q, v)$ to both \autoref{fig:queue} and \autoref{fig:queueMS}
+that checks to see if $v$ is in queue $q$.
+\item Add a method $\mathtt{Qremove}(q, v)$ to both \autoref{fig:queue} and \autoref{fig:queueMS}
+that removes $v$, if any, from queue $q$.
+\item The test program in \autoref{fig:queuetest} is not very thorough.  Design
+and implement a better one.
+\item Create a thread-safe sorted binary tree.  Implement methods
+$\mathtt{BTnew}()$ to create a new binary tree, $\mathtt{BTinsert}(t, v)$ that inserts $v$ into
+tree $t$, and $\mathtt{BTcontains}(t, v)$ that checks if $v$ is in tree $t$.  Use a single
+lock per binary tree.
+\item Bonus question: create a binary tree that uses, instead of a single lock per tree,
+a lock for each node in the tree.
+\end{problems}
 
@@
 for three processes.
-\item Optimize your solution to \autoref{ex:xy} to use reader/writer locks.
+\item Optimize your solutions to \autoref{ex:qcontains} to use reader/writer locks.
 \item \label{ex:boundlock} Using busy waiting, implement a ``bound lock'' that allows
@@
 \end{code}
-\caption{The \texttt{Semaphore} interface in the \texttt{synch} module.}
+\caption{\harmonylink{modules/synch.hny} The \texttt{Semaphore} interface in the \texttt{synch} module.}
 \label{fig:semaphore}
@@
 \end{code}
-\caption{Implementation of condition variables in the \texttt{synch} module.}
+\caption{\harmonylink{modules/synch.hny}
+Implementation of condition variables in the \texttt{synch} module.}
 \label{fig:cv}
@@
 \hline
-\texttt{malloc()} & allocate and return a pointer to a memory location \\
+\texttt{malloc($v$)} & return a pointer to a memory location initialized to $v$ \\
 \hline

### 2020-08-21 book/paper.tex
Latest

@@
 and \texttt{release\_bound\_lock()}.
+(Bound locks are useful for situations where too many processes working
+at the same time might exhaust some resource such as a cache.)
 \item Write a test program for your bound lock
@@
 and return \texttt{True} if it succeeds.
+\item Optimize the code in \autoref{fig:boundedbuffer} for the case
+where there is only one consumer and one producer.
 \item Implement a \emph{bounded stack}.  Method \texttt{push}(\textit{item})
@@
 
+(Another option is to use synchronization techniques that do not use locks.
+See \autoref{ch:nonblocking} for more information.)
+
 There is another important rule to keep in mind.  Just like locks should never be held
@@
 the other processes have to wait for it to finish executing the critical section.
+Also, using synchronization primitives in interrupt handlers is tricky
+to get right (\autoref{ch:interrupts}) and might run too long.
 In this chapter, we will have a look at how one can develop concurrent

### 2020-08-22 book/paper.tex
Better lvalue parsing

@@
    7 LoadVar self
-   8 Address 2
+   8 Address
    9 Push True
@@
 \hline
-Address $n$ & compute address from $n$ components \\
+Address & compute address from two components \\
 \hline
@@
 \begin{itemize}
-\item The \texttt{Address} instruction expects $n$ values $v_1, ..., v_n$ on the stack.
-The top value must be an address value.  The remaining values are keys into dictionaries.
-The instruction then computes the address of \texttt{!}$v_1~v_2~...~v_n$.
+\item The \texttt{Address} instruction expects two values on the stack.
+The top value must be an address value, representing a dictionary 
+The other value must be a key into the dictionary.
+The instruction then computes the address of the given key.
 \item Even though Harmony code does not allow taking addresses of process variables, both

### 2020-08-27 book/paper.tex
Latest

@@
 \item $q = \mathtt{Qnew}()$: allocate a new queue;
-\item $Qenqueue(q, v)$: add $v$ to the tail of the queue;
-\item $r = Qdequeue(q)$: returns $r = ()$ if $q$ is empty or $r = (v,)$ (a singleton tuple)
+\item $\mathtt{Qenqueue}(q, v)$: add $v$ to the tail of the queue;
+\item $r = \mathtt{Qdequeue}(q)$: returns $r = ()$ if $q$ is empty or $r = (v,)$ (a singleton tuple)
 if $v$ was at the head of the queue.
@@
 requires both locks, the queue uses a dummy node at the head of the linked list.
-Except of the initial state of the queue, the dummy node is the last node that
+Except for the initial state of the queue, the dummy node is the last node that
 was dequeued.  Note that neither the \texttt{head} nor \texttt{tail} pointer are
@@
 \item With reader/writer locks,
-it can be useful for a process to \emph{downgrade} its write lock
+concurrency can be improved if a process \emph{downgrades} its write lock
 to a read lock when its done writing but not done reading.  Add

### 2020-08-28 book/paper.tex
Last

@@
 
-\texttt{release\_wlock} executes \texttt{notifyAll}(?\textit{wcond}) as well as
+\texttt{release\_wlock} executes \texttt{notifyAll}(?\textit{rcond}) as well as
 \texttt{notify}(?\textit{wcond)}.
@@
 thumb: be conservative---it is better to wake up too many processes than too few.
-Waking up too many processes may lead to some inefficiency, but waking up to few
+Waking up too many processes may lead to some inefficiency, but waking up too few
 may cause the application to get stuck.  Harmony can be particularly helpful here, as

### 2020-08-28 harmony.py

@@ -2507,6 +2507,8 @@ class ArrowExpressionRule(Rule):
         (ast, t) = BasicExpressionRule().parse(t)
         if ast == False:
             return (False, t)
+        if t == []:
+            return (ast, t)
         (lexeme, file, line, column) = token = t[0]
         while lexeme == "->":
             (lexeme, file, line, column) = t[1]

### 2020-08-30 book/paper.tex
Got rid of PushAddress

@@
 code/Up.hny:3     done[self] = True;
-   6 PushAddress done
+   6 Push ?done
    7 LoadVar self
@@
 Push $c$ & push constant $c$ onto the stack \\
-\hline
-PushAddress $v$ & push the address of variable $v$ \\
 \hline

### 2020-08-30 harmony.py

@@ -1043,22 +1043,6 @@ class PushOp(Op):
         context.push(lexeme)
         context.pc += 1
 
-class PushAddressOp(Op):
-    def __init__(self, name):
-        self.name = name 
-
-    def __repr__(self):
-        (lexeme, file, line, column) = self.name
-        return "PushAddress " + str(lexeme)
-
-    def explain(self):
-        return "push the address of shared variable " + self.name[0]
-
-    def eval(self, state, context):
-        (lexeme, file, line, column) = self.name
-        context.push(AddressValue([lexeme]))
-        context.pc += 1
-
 class LoadOp(Op):
     def __init__(self, name, token):
         self.name = name
@@ -2067,11 +2051,12 @@ class NameAST(AST):
     def ph1(self, scope, code):
         tv = scope.lookup(self.name)
         if tv == None:
-            code.append(PushAddressOp(self.name))
+            (lexeme, file, line, column) = self.name
         else:
             (t, v) = tv
             assert t == "variable", (self, tv)
-            code.append(PushAddressOp(v))
+            (lexeme, file, line, column) = v
+        code.append(PushOp((AddressValue([lexeme]), file, line, column)))
 
     def ph2(self, scope, code, skip):
         if skip > 0:

### 2020-09-06 book/paper.tex
Fixed typo
Latest

@@
 \item Using a script, run \autoref{fig:incpy} 1000 times.
-For example, if you are using the bash shell (in Linux, say), you can do the following:
+For example, if you are using the bash shell (in Linux or Mac OS X, say), you can do the following:
 \begin{code}
@@
 \end{code}
+If you're using Windows, the following batch script does the trick:
+\begin{code}
+\begin{verbatim}
+FOR /L %%i IN (1, 1, 1000) DO python Up.py
+PAUSE
+\end{verbatim}
+\end{code}
 How many times does the assertion fail (if any)?
@@
 \item \autoref{fig:incmany} is a version of \autoref{fig:incpy} that has each
-incrementer thread increment \textit{count} \texttt{N} times.  Run \autoref{fig:incpy}
+incrementer thread increment \textit{count} \texttt{N} times.  Run \autoref{fig:incmany}
 10 times (using Python).

### 2020-09-06 book/paper.tex
Latest

@@
 However, if $d$ = \texttt{dict\{ .count: 3 \}}, then you can write
-can write $d$.\texttt{count} (which has value 3) instead of having to write
+$d$.\texttt{count} (which has value 3) instead of having to write
 $d$\texttt{[.count]} (although both will work).
@@
 \begin{itemize}
-\item Process \texttt{implementer/0}
+\item Process \texttt{incrementer/0}
 executed instructions 1 through 4, loading the value of
 \textit{count} but stopping just before storing 1 into \textit{count};
-\item Process \texttt{implementer/1}
+\item Process \texttt{incrementer/1}
 executed instructions 1 through 11, storing 1 into
 into \textit{count} and storing \texttt{True} into \textit{done}[1];
-\item Process \texttt{implementer/0}
+\item Process \texttt{incrementer/0}
 continues execution, executing instructions 5 through 11

### 2020-09-20 book/paper.tex
Title change
Latest
Fixed typos

@@
 
-\title{Concurrent Programming with Harmony}
+\title{Concurrent Programming in Harmony}
 \author{Robbert van Renesse}
@@
 How large does \texttt{N} need to be for assertions to fail?
+(Try powers of 10 for \texttt{N}.)
 \item Can you think of a fix to \autoref{fig:inc}?  Try one or two different fixes
@@
 \item Shared memory: A HVM has just one memory location containing
-an Harmony value.
+a Harmony value.
 \item Processes:  Any process
@@
 \item What if we moved line~2 of \autoref{fig:incenter} to after the \textbf{if}
-    statement (between lines~7 and~8)?  Do you think that would work?  Run it through
+    statement (between lines~5 and~6)?  Do you think that would work?  Run it through
     Harmony and describe either why it works or why it does not work.
@@
 from eating to think for a while.  To do so, the philosopher puts down both
-forks.  Each philosopher repeat this procedure.  Dijkstra had them repeating
+forks.  Each philosopher repeats this procedure.  Dijkstra had them repeating
 this for ever, but for the purposes of this book, philosophers can leave

### 2020-09-26 book/paper.tex
Latest

@@
 it turns out the solution
-does has a problem: if both try to enter the critical section at the same
+does have a problem: if both try to enter the critical section at the same
 time, they may end up waiting for one another indefinitely.  Thus the
@@
 So we are looking for an inductive invariant that \emph{implies} exclusivity.
-In other words, the set of states where the inducive invariant holds
+In other words, the set of states where the inductive invariant holds
 must be a subset of the set of states where there is at most one process in
@@
 \begin{center}
-\includegraphics[width=2.3in]{figures/rdwr-crop.pdf}
+\includegraphics[width=2.3in]{figures/rdwr.png}
 \end{center}
@@
 \item \label{ex:boundlock} Using busy waiting, implement a ``bound lock'' that allows
-up to \texttt{M} processes to acquire it at the same time.  A bound lock
+up to \texttt{M} processes to acquire it at the same time.\footnote{A bound lock is a restricted version of a counting semaphore, as we will learn later in this book.}
+A bound lock
 with \texttt{M == 1} is an ordinary lock.
@@
 \begin{itemize}
-\item $n$ readers at \texttt{@rcs} $\Rightarrow \mathit{nreaders} \ge n$,
-\item $\exists$ writer at \texttt{@wcs} $\Rightarrow \mathit{nreaders} = 0$,
+\item $n$ readers at \texttt{@rcs} $\Rightarrow \mathit{nreaders} \ge n$
+\item $\exists$ writer at \texttt{@wcs} $\Rightarrow \mathit{nreaders} = 0$
 \end{itemize}
 
-A writer simply acquires \textit{rwlock} to enter the critical section
+A writer just acquires \textit{rwlock} to enter the critical section
 and releases it to exit.  The \emph{first} reader to enter the critical

### 2020-09-27 book/paper.tex
Fixed Peterson's proof

@@
 $\mathcal{I}_p(0)$ still holds.
-If $\mathcal{I}_p(0)$ holds, process~0 is at label \texttt{@cs}.  If process~0
+
+First consider the case that process~0 is at label \texttt{@cs}.  If process~0
 were to take a step, then in the next state process~0 would be no longer
@@
 Therefore we only need to consider a step by process~1.
-
+%
 From $\mathcal{I}_p(0)$ we know that one of the following three cases must hold before
@@
 the step \textit{turn} will equal 0, and therefore the second case will hold
-after the step.  qed.
+after the step.
+
+Now consider the case where process~0 is not in the critical section,
+and therefore $\mathcal{I}_p(0)$ holds trivially because false implies
+anything.
+There are three cases to consider:
+\begin{enumerate}
+\item Process~1 takes a step.  But then process~0 is still not in the critical
+section and $\mathcal{I}_p(0)$ continues to hold;
+\item Process~0 takes a step but still is not in the critical section.
+Then again $\mathcal{I}_p(0)$ continues to hold.
+\item Process~0 takes a step and ends up in the critical section.
+Because process~0 entered the critical section, we know that
+\textit{flags}[1] = \texttt{False} or \textit{turn} == 0 because
+of the \texttt{await} condition.
+And hence $\mathcal{I}_p(0)$ continues to hold in that case as well.
+\end{enumerate}
 

### 2020-09-28 book/paper.tex
Improved

@@
 the critical section.
+For simplicity, we will suppose as a given that 
+$\mathtt{process}(i)@cs \Rightarrow \mathit{flags][i]$, the truth of which is
+easy to see from the code.
 Let $\mathcal{C}(i) = \lnot\mathit{flags}[1 - i] \lor
@@
 \item $n$ readers at \texttt{@rcs} $\Rightarrow \mathit{nreaders} \ge n$
-\item $\exists$ writer at \texttt{@wcs} $\Rightarrow \mathit{nreaders} = 0$
+\item $\exists$ writer at \texttt{@wcs} $\Rightarrow \mathit{nreaders} = 0$ and writer holds \textit{rwlock}
+\item at most one writer can hold \textit{rwlock}
 \end{itemize}
@@
 
-I would also like to thank my proofreaders Anneke van Renesse and
-Zhuoyu Xu.
+The following people contributed by making comments on or finding bugs in
+early drafts of the book:
+Anneke van Renesse,
+Brendon Nguyen,
+Sunwook Kim,
+Trishita Tiwari,
+Yidan Wang,
+Zhuoyu Xu,
+and
+Zoltan Csaki.
 

### 2020-09-28 book/paper.tex
Fixed typo

@@
 For simplicity, we will suppose as a given that 
-$\mathtt{process}(i)@cs \Rightarrow \mathit{flags][i]$, the truth of which is
+$\mathtt{process}(i)@cs \Rightarrow \mathit{flags}[i]$, the truth of which is
 easy to see from the code.

### 2020-09-30 book/paper.tex
Latest
Fixed Hoare monitor semantics

@@
 the critical section.
-For simplicity, we will suppose as a given that 
-$\mathtt{process}(i)@cs \Rightarrow \mathit{flags}[i]$, the truth of which is
-easy to see from the code.
 Let $\mathcal{C}(i) = \lnot\mathit{flags}[1 - i] \lor
@@
 
+\begin{figure}
+\begin{code}
+\harmonysource{hoare}
+\end{code}
+\caption{\harmonylink{code/hoare.hny} Implementation of Hoare monitors.}
+\label{fig:hoare}
+\end{figure}
+
+\begin{figure}
+\begin{code}
+\harmonysource{hoaretest}
+\end{code}
+\caption{\harmonylink{code/hoaretest.hny} Example of Hoare monitors.}
+\label{fig:hoaretest}
+\end{figure}
+
 Tony Hoare, who came up with the concept of split binary semaphores, devised
@@
 method and released upon exit.
-The other semaphores are called \emph{condition variables}.
+The other semaphores are encapsulated in \emph{condition variables}.
 \index{condition variable}%
+Each condition variable also keeps track of the number of processes waiting
+on the condition variable.
 There are two operations on condition variables: \texttt{wait}
@@
 \index{signal}%
-If $c$ is a pointer to a condition variable, then the semantics of the
-operations, in Harmony, are as follows:
-
-\begin{code}
-\harmonysource{hoaremon}
-\end{code}
-
-\texttt{wait} releases the mutex and blocks while trying
-to procure the condition variable (which is a semaphore that is 0 when
-\texttt{wait} is invoked).
-\texttt{signal} vacates the condition variable, passing the baton to
-a process trying to procure it, and then procures the mutex.
+
+\autoref{fig:hoare} presents the semantics by giving an implementation of
+\texttt{wait} and \texttt{signal} as well as the code that is used to enter
+and exit monitor methods.
+\autoref{fig:hoaretest} presents a simple example lifted directly from
+the original paper~\cite{Hoare74}.
+In the design proposed by Hoare, the monitor methods automatically acquire
+and release the monitor mutex.
+
+\texttt{wait} increments the condition's counter, releases the monitor mutex,
+blocks while trying to procure the condition's semaphore, and finally decrements
+the counter---in much the same way as we have seen for SBS.
+\texttt{signal} checks to see if the condition's count is non-zero, and if so
+vacates the condition's semaphore---passing the baton to
+a process trying to procure it---and then blocks by procuring the mutex again.
 The use of \texttt{wait} and \texttt{signal} ensures that \texttt{P} and

### 2020-09-30 book/paper.tex
Latest

@@
 on the condition variable.
+
 There are two operations on condition variables: \texttt{wait}
@@
 \index{signal}%
-
 \autoref{fig:hoare} presents the semantics by giving an implementation of
@@
 the original paper~\cite{Hoare74}.
-In the design proposed by Hoare, the monitor methods automatically acquire
-and release the monitor mutex.
 
 \texttt{wait} increments the condition's counter, releases the monitor mutex,
-blocks while trying to procure the condition's semaphore, and finally decrements
-the counter---in much the same way as we have seen for SBS.
+blocks while trying to procure the condition's semaphore, and after resuming
+decrements the counter---in much the same way as we have seen for split binary
+semaphores (SBS).
 \texttt{signal} checks to see if the condition's count is non-zero, and if so

### 2020-10-09 book/paper.tex
Fixed bug in pointer checking

@@
 \autoref{fig:rwblock} presents such a test.
-Note that the initialization code obtains a read lock, preventing
+Note that the initialization code obtains a write lock, preventing
 all readers and writers from entering the critical section.
@@
 Brendon Nguyen,
+Saleh Hassen,
 Sunwook Kim,

### 2020-10-09 harmony.py

@@ -1104,6 +1104,9 @@ class StoreOp(Op):
                                     str(self.token) + " -> " + str(av)
                 return
             lv = av.indexes
+            if len(lv) == 0:
+                context.failure = "Error: bad address " + str(self.token)
+                return
             name = lv[0]
         else:
             (lexeme, file, line, column) = self.name

### 2020-10-09 book/paper.tex
Latest

@@
 Brendon Nguyen,
+Heather Zheng,
 Saleh Hassen,

### 2020-10-18 book/paper.tex
Latest

--- 2020-10-09 book/paper.tex
@@
 the balances is the same as that at the beginning (using an assertion).
-Is once enough to determine that no money gets lost or create, or should it do
+Is once enough to determine that no money gets lost or created, or should it do
 so continually?
@@
 early drafts of the book:
+Alex Chang,
 Anneke van Renesse,
 Brendon Nguyen,
+Heather Zheng,
 Saleh Hassen,

### 2020-11-26 book/paper.tex
Latest

@@
   description={is a concurrency model where there are no shared variables, only
-      processes with private variables that communicate through message passing}
+      threads with private variables that communicate through message passing}
 }
@@
   description={describes that a certain machine instruction or sequence of machine
-    instructions by a process is executed indivisibly and cannot be interleaved with 
-    machine instructions of another process}
+    instructions is executed indivisibly by a thread and cannot be interleaved with 
+    machine instructions of another thread}
 }
@@
   name=barrier synchronization,
-  description={is when a set of processes execute in rounds, waiting for one another
+  description={is when a set of threads execute in rounds, waiting for one another
     to complete each round}
 }
-\newglossaryentry{blocked process}
+\newglossaryentry{blocked thread}
 {
-  name=blocked process,
-  description={is a process that cannot change the state or terminate or
-      can only do so after another process changes the state first.
-      For example, a process that is waiting for a lock to become available}
+  name=blocked thread,
+  description={is a thread that cannot change the state or terminate or
+      can only do so after another thread changes the state first.
+      For example, a thread that is waiting for a lock to become available}
 }
@@
   name=busy waiting,
-  description={(aka spin-waiting) is when a process waits in a loop for some
+  description={(aka spin-waiting) is when a thread waits in a loop for some
       application-defined condition instead of blocking}
@@
   name=concurrent execution,
-  description={(aka parallel execution) is when there are multiple processes executing and
+  description={(aka parallel execution) is when there are multiple threads executing and
       their machine instructions are interleaved in an unpredictable manner}
@@
   name=condition variable,
-  description={a variable that keeps track of which processes are waiting for a
+  description={a variable that keeps track of which threads are waiting for a
     specific application-level condition.  The variable can be waited on as well
@@
   name=context,
-  description={(aka continuation) describes the state of a running process,
+  description={(aka continuation) describes the state of a running thread,
     including its program counter, the values of its variables (stored in its
@@
   description={is a critical section with, besides mutual exclusion, additional conditions on
-      when a process is allowed to enter the critical section}
+      when a thread is allowed to enter the critical section}
 }
@@
   name=critical section,
-  description={(aka critical region) is a set of instructions that only one process
+  description={(aka critical region) is a set of instructions that only one thread
     is allowed to execute at a time.  The instructions are, however, not executed
-    atomically, as other processes can continue to execute and access shared
+    atomically, as other threads can continue to execute and access shared
     variables}
@@
   name=deadlock,
-  description={is when there are two or more processes waiting indefinitely for one
+  description={is when there are two or more threads waiting indefinitely for one
     another to release a resource}
@@
   name=fairness,
-  description={is when each process eventually can access each resource it needs to access
+  description={is when each thread eventually can access each resource it needs to access
     with high probability}
@@
   description={is a binary predicate over states that must hold for every
-  reachable state of a process}
+  reachable state of a thread}
 }
@@
   name=lock,
-  description={an object that can be owned by at most one process at a time.  Useful
+  description={an object that can be owned by at most one thread at a time.  Useful
     for implementing mutual exclusion}
@@
   name=machine instruction,
-  description={is an atomic operation on the Harmony virtual machine, executed by a process}
+  description={is an atomic operation on the Harmony virtual machine, executed by a thread}
 }
@@
   name=mutual exclusion,
-  description={is the property that two processes never enter the same critical section}
+  description={is the property that two threads never enter the same critical section}
 }
@@
   description={(aka wait-free synchronization) is when access to a shared resource can be
-      guaranteed in a bounded number of steps even if other processes are not making progress}
+      guaranteed in a bounded number of steps even if other threads are not making progress}
 }
-\newglossaryentry{process}
+\newglossaryentry{thread}
 {
-  name=process,
+  name=thread,
   description={is code in execution.  We do not make the distinction
-      between processes and threads.  A process has a current context and
+      between threads and threads.  A thread has a current context and
       updates its context every time it executes a machine instruction}
 }
-\newglossaryentry{process variable}
+\newglossaryentry{thread variable}
 {
-  name=process variable,
-  description={is a variable that is private to a single process and stored in
+  name=thread variable,
+  description={is a variable that is private to a single thread and stored in
         its register}
@@
   name=producer/consumer problem,
-  description={is a synchronization problem whereby one or more producing processes
-    submit items and one or more consuming processes want to receive them.  No item
+  description={is a synchronization problem whereby one or more producing threads
+    submit items and one or more consuming threads want to receive them.  No item
     can get lost or forged or be delivered to more than one consumer, and producers
@@
     such as violating mutual exclusion in a critical section.  Liveness properties are
-    properties where ``something good eventually happens,'' like processes being
+    properties where ``something good eventually happens,'' like threads being
     able to enter the critical section if they want to}
@@
   name=race condition,
-  description={describes when multiple processes access shared state concurrently,
+  description={describes when multiple threads access shared state concurrently,
     leading to undesirable outcomes}
@@
   name=reader/writer lock,
-  description={is a lock on a resource that can be held by multiple processes if they all 
+  description={is a lock on a resource that can be held by multiple threads if they all 
     only read the resource}
@@
   description={is a counter that can be atomically incremented and decremented,
-    but blocks the process until the counter is larger than zero first}
+    but blocks the thread until the counter is larger than zero first}
 }
@@
   name=sequential execution,
-  description={is when there is just one process executing, as opposed to concurrent
+  description={is when there is just one thread executing, as opposed to concurrent
     execution}
@@
   description={is a variable that is stored in the memory of the Harmony virtual machine and
-      shared between multiple processes, as opposed to a process variable}
+      shared between multiple threads, as opposed to a thread variable}
 }
@@
   name=spinlock,
-  description={is an implementation of a lock whereby a process loops until the
-    lock is available, at which point the process atomically obtains the lock}
+  description={is an implementation of a lock whereby a thread loops until the
+    lock is available, at which point the thread atomically obtains the lock}
 }
@@
   name=stack machine,
-  description={is a model of computing where the state of a process is kept on
+  description={is a model of computing where the state of a thread is kept on
     a stack.  Harmony uses a combination of a stack machine and a register-based machine}
@@
   name=starvation,
-  description={is when a process cannot make progress because it is continuously
-    losing a competition with other processes to get access to a resource}
+  description={is when a thread cannot make progress because it is continuously
+    losing a competition with other threads to get access to a resource}
 }
@@
   name=step,
-  description={is the execution of a machine instruction by a process, updating
+  description={is the execution of a machine instruction by a thread, updating
         its state.  Harmony distinguishes micro steps (machine instructions) and macro
         step (a sequence of instructions with at most one effect visible by
-        other processes)}
+        other threads)}
 }
@@
 can make programs faster than sequential ones, but having multiple
-processes (aka \emph{threads}\footnote{We
 \index{thread}%
-do not make a distinction between processes and threads.})
-read and update shared variables
+threads read and update shared variables
 \index{shared variable}%
@@
 complicated than programs where only one thing happens at a time.
-
+%
 \glsadd{concurrent execution}%
@@
 \glsadd{atomicity}%
-
+%
 Why are concurrent
@@
 the output of running concurrent programs depends on how the
-execution of the various processes are \emph{interleaved}.
+execution of the various threads are \emph{interleaved}.
 Some bugs may occur only occasionally and
 may never occur when the program is instrumented to find them
-(so-called \emph{Heisenbugs}).
+(so-called \emph{Heisenbugs}---overhead caused by instrumentation
+leads to timing changes that makes such bugs less likely to occur).
 \index{Heisenbug}%
@@
 be compiled into multiple machine instructions, they are executed back-to-back
-until completion.  Not so with a concurrent program, where other processes
+until completion.  Not so with a concurrent program, where other threads
 may update memory locations while a statement or function is being executed.
@@
 test program.  So even if a bug is unlikely to occur, if the test
-\emph{can} expose the bug it \emph{will}.  Importantly, if the bug is
+\emph{can} expose the bug it \emph{will}.  Moreover, if the bug is
 found, the model checker precisely shows how to trigger the bug in
@@
 To answer that question, let us consider a sorting algorithm.
-Now suppose we create a test program, a model, that tries sorting
+Suppose we create a test program, a model, that tries sorting
 \emph{all} lists of up to five numbers chosen from the set
@@
 scenario that would make it easy to find the source of the bug.
+
 However, if the model checker does not find any bugs, we do not
@@
 are bugs remaining in the sorting algorithm is small.
-Note, however, that it would be easy to write a program
+That said, it would be easy to write a program
 that sorts all lists of up to five numbers correctly but fails to
@@
 it can help with proving an algorithm correct.
-The reason is that the correctness of many programs is built around
+The reason is that many correctness properties can be proved using
 \emph{invariants}:
@@
 \item \autoref{ch:concurrent} illustrates the problem of
-concurrent programming through a simple example in which two processes
+concurrent programming through a simple example in which two threads
 are concurrently incrementing a counter.
@@
 abstraction and implementation of \emph{reader/writer locks},
-which allow multiple processes in a critical section if they are
+which allow multiple threads in a critical section if they are
 all only executing read-only accesses.
 \autoref{ch:rdwrbusy} also discusses \emph{busy-waiting}, which is
-an easy but undesirable approach to synchronize processes.
+an easy but undesirable approach to synchronize threads.
 \item \autoref{ch:semaphore} introduces \emph{semaphores},
@@
 the problem that in some
-synchronization approaches processes may not be able to get access to a
+synchronization approaches threads may not be able to get access to a
 resource they need.
@@
 \emph{monitors} and \emph{condition variables},
-another approach to process synchronication.
+another approach to thread synchronication.
 \item \autoref{ch:deadlock} describes \emph{deadlock}
-where a set of processes are indefinitely waiting for one another to
+where a set of threads are indefinitely waiting for one another to
 release a resource.
@@
 and \emph{message passing} as an approach to synchronization.
-\item \autoref{ch:interrupts} discusses how to handle interrupts to a process,
-a problem closely related to synchronization among multiple processes.
+\item \autoref{ch:interrupts} discusses how to handle interrupts,
+a problem closely related to synchronization among multiple threads.
 \item \autoref{ch:abp} presents a problem and a solution to the distributed
-systems problem of having two processes communicate reliably over an unreliable
+systems problem of having two threads communicate reliably over an unreliable
 network.
@@
 \emph{wait-free} synchronization algorithms,
-which prevent processes waiting for one another more than a bounded number of
+which prevent threads waiting for one another more than a bounded number of
 steps.
@@
 Correct indentation in Harmony is encouraged but not enforced.
-Harmony only supports basic operator precedence or associativity.
+\item Harmony only supports basic operator precedence or associativity.
 Use parentheses liberally to remove ambiguity.
@@
 \end{itemize}
-There are also many unimportant ones that you will discover as
+There are also less important ones that you will discover as
 you get more familiar with programming in Harmony.
@@
 no issues found
-Open file:///.../harmony.html for more information
 $
@@
 
-\glsadd{process}%
+\glsadd{thread}%
 
 Concurrent programming, aka multithreaded programming, involves multiple
-processes
-\index{process}%
+threads
+\index{thread}%
 running in parallel while sharing variables.
@@
 (\textbf{await} $c$ is semantically the same as \textbf{while not} $c$: \textbf{pass}.)
-After that, it verifies that the value of \textit{count} equals 2.
+After that, method \texttt{incrementer}
+verifies that the value of \textit{count} equals 2.
 Like Python, Harmony supports \texttt{assert} statements with two arguments:
@@
 the actual value of \textit{count}.
-The program spawns three processes.
+
+The program spawns three threads.
 The first runs \texttt{incrementer(0)}, the second runs
 \texttt{incrementer(1)}, and the last runs \texttt{main()}.
-Note that although the processes are \emph{spawned} one at a time,
+Note that although the threads are \emph{spawned} one at a time,
 they will execute concurrently.  It is, for example, quite possible
@@
 
-When running multiple processes, each essentially runs an instantiation of
+When running multiple threads, each essentially runs an instantiation of
 the machine, and they do so in parallel.  As they execute, their machine
@@
 in unspecified and often random ways.
-A program is correct if it works for any interleaving.
-Harmony will try all possible interleavings of the processes
+A program is correct if it works for any interleaving of threads.
+Harmony will try all possible interleavings of the threads
 executing machine instructions.
 
-The following is a possible interleaving of incrementers~0 and~1:
+If the threads run one at a time, then \textit{count} will be incremented
+twice and ends up being 2.  However, the following is also
+a possible interleaving of incrementers~0 and~1:
 \begin{enumerate}
@@
 This would fix the issue, but more likely it is the program that must
-be fixed.
+be fixed, not the specification.
 
@@
 which in turn is the same Harmony value as \texttt{dict\{ 0:a, 1:b, 2:c \}}.
-So if $x$ = \texttt{[ False, True ]},
+So if $x$ = \texttt{[False, True]},
 then $x$[0] = \texttt{False} and $x$[1] = \texttt{True}, just like in Python.
@@
 a Harmony value.
-\item Processes:  Any process
-can spawn an unbounded number of other processes and processes may terminate.
-Each process has an immutable (but not necessarily unique) \emph{name tag},
+\item Threads:  Any thread
+can spawn an unbounded number of other threads and threads may terminate.
+Each thread has an immutable (but not necessarily unique) \emph{name tag},
 \index{name tag}%
@@
 
-A name tag consists of the name of the main method of the process,
+A name tag consists of the name of the main method of the thread,
 along with an optional tag specified in the \texttt{spawn}
@@
 statement.
-The default tag is the first argument to the method.
+The default tag is the argument to the method.
 A name tag is represented as a dictionary with keys \texttt{.name}
@@
 \texttt{name/tag}.
-For example, in \autoref{fig:inc}, the created processes have name tags
+For example, in \autoref{fig:inc}, the created threads have name tags
 \texttt{incrementer/0}, \texttt{incrementer/1}, and \texttt{main/()}.
@@
 
-The state of a process is called a \emph{context} (aka \emph{continuation}):
+The state of a thread is called a \emph{context} (aka \emph{continuation}):
 \index{context}%
 \index{continuation}%
-it contains the value of
+it contains the values of
 its name tag, program counter, stack, and register.
@@
 \index{multiset}%
-of contexts.  It is a multiset of contexts because two different processes can
-be in the same state.
+of contexts.  It is a multiset of contexts because two different threads can
+have the same context.
 The initial state of the Harmony memory is the empty dictionary, \texttt{dict\{\}}.
@@
 \begin{verbatim}
-code/Up.hny:1 def incrementer(self):
-   0 Jump 25
-   1 Frame incrementer self
-code/Up.hny:2     count = count + 1;
-   2 Load count
-   3 Push 1
-   4 2-ary +
-   5 Store count
-code/Up.hny:3     done[self] = True;
-   6 Push ?done
-   7 LoadVar self
-   8 Address
-   9 Push True
-   10 Store
-   11 Return
-code/Up.hny:5 def main():
-   12 Jump 25
-   13 Frame main ()
-code/Up.hny:6     await all(done);
-   14 Load done
-   15 1-ary all
-   16 JumpCond False 14
-code/Up.hny:7     assert count == 2, count;
-   17 AtomicInc
-   18 Load count
-   19 Push 2
-   20 2-ary ==
-   21 Load count
-   22 Assert2
-   23 AtomicDec
-   24 Return
-code/Up.hny:9 count = 0;
-   25 Push 0
-   26 Store count
-code/Up.hny:10 done = [ False, False ];
-   27 Push [False, False]
-   28 Store done
-code/Up.hny:11 spawn incrementer(0);
-   29 Push 0
-   30 Dup
-   31 Push PC(1)
-   32 Spawn
+code/Up.hny:1 count = 0;
+   0 Push 0
+   1 Store count
+code/Up.hny:2 done = [ False, False ];
+   2 Push [False, False]
+   3 Store done
+code/Up.hny:4 def incrementer(self):
+   4 Jump 31
+   5 Frame incrementer self
+code/Up.hny:5     count = count + 1;
+   6 Load count
+   7 Push 1
+   8 2-ary +
+   9 Store count
+code/Up.hny:6     done[self] = True;
+   10 Push ?done
+   11 LoadVar self
+   12 Address
+   13 Push True
+   14 Store
+   15 Return
+code/Up.hny:8 def main():
+   16 Jump 31
+   17 Frame main ()
+code/Up.hny:9     await all(done);
+   18 Load done
+   19 1-ary all
+   20 JumpCond False 18
+code/Up.hny:10     assert count == 2, count;
+   21 ReadonlyInc
+   22 AtomicInc
+   23 Load count
+   24 Push 2
+   25 2-ary ==
+   26 Load count
+   27 Assert2
+   28 AtomicDec
+   29 ReadonlyDec
+   30 Return
+code/Up.hny:13 spawn incrementer(0);
+   31 Push 0
+   32 Dup
+   33 Push PC(5)
+   34 Spawn
 \end{verbatim}
@@
 It may seem strange that there is only one memory location and that each
-process has only one register.  However, this is not a limitation because
+thread has only one register.  However, this is not a limitation because
 Harmony values are unbounded trees.
-Both the memory and the register of a process always contain
+Both the memory and the register of a thread always contain
 a dictionary that maps atoms to Harmony values.  We call this a \emph{directory}.
@@
 \begin{verbatim}
-#states = 71 diameter = 5
-==== Safety violation ====
-__init__/()   [0,25-41] 41 { count: 0, done: [False, False] }
-incrementer/0 [1-4]      5 { count: 0, done: [False, False] }
-incrementer/1 [1-11]    11 { count: 1, done: [False, True ] }
-incrementer/0 [5-11]    11 { count: 1, done: [True,  True ] }
-main/()       [13-22]   22 { count: 1, done: [True,  True ] }
->>> Harmony Assertion failed : 1
-Open file://.../harmony.html for more information
+#states = 71 diameter = 5                                                                              ==== Safety violation ====
+P0: __init__/()   [0-4,31-43] 43 { .count: 0, .done: [False, False] }
+P1: incrementer/0 [5-8]        9 { .count: 0, .done: [False, False] }
+P2: incrementer/1 [5-15]      15 { .count: 1, .done: [False, True ] }
+P1: incrementer/0 [9-15]      15 { .count: 1, .done: [True,  True ] }
+P3: main/()       [17-27]     27 { .count: 1, .done: [True,  True ] }
+>>> Harmony Assertion (file=code/Up.hny, line=10) failed: 1
+Open file:///Users/rvr/github/harmony.new/harmony.html for more information
 \end{verbatim}
@@
 \end{code}
-Each process in the HVM is predominantly a \emph{stack machine},
+Each thread in the HVM is predominantly a \emph{stack machine},
 \index{stack machine}%
@@
 Most instructions pop values from the stack or push values onto the stack.
-At first there is one process, with name tag \texttt{\_\_init\_\_/()},
+At first there is one thread, with name tag \texttt{\_\_init\_\_/()},
 that initializes the state.
@@
 execution in the program.
-In this case, the first instruction is a \texttt{JUMP} instruction that sets the
-program counter to 25 (skipping over the code for the methods).
-At program counter 1 is the code for the \texttt{incrementer} method.
+In this case, it executes instructions 0 through 4 first.
+The last instruction is a \texttt{JUMP} instruction that sets the
+program counter to 31 (skipping over the code for the methods).
+At program counter 5 is the code for the \texttt{incrementer} method.
 All methods start with a \texttt{Frame} instruction and end with a \texttt{Return}
@@
 names of its arguments.
-The code generated from $\mathit{count} := \mathit{count} + 1$ in line 2 of
+The code generated from $\mathit{count} := \mathit{count} + 1$ in line~6 of
 \texttt{Up.hny} is as follows (see \autoref{fig:inccode}):
 
-\begin{enumerate} \setcounter{enumi}{1}
+\begin{enumerate} \setcounter{enumi}{5}
 \item The \texttt{Load} instruction pushes the value of the
@@
 \item the \texttt{Push} instruction pushes the constant 1
-onto the stack of the process.
+onto the stack of the thread.
 \item \texttt{2-ary} is a \texttt{+} operation with 2 arguments.
@@
 
-Once initialization completes, any processes that were spawned can run.
-You can think of Harmony as trying every possible interleaving of processes executing
+Once initialization completes, any threads that were spawned can run.
+You can think of Harmony as trying every possible interleaving of threads executing
 instructions.
@@
 \begin{figure}
-\includegraphics[width=\textwidth]{figures/Up1.png}
+\includegraphics[width=\textwidth]{figures/Up1.pdf}
 \caption{The
@@
 of the program.
-The bottom right shows the state of each process after the selected step.
-Each row in the stack trace of a process shows the current program counter,
+The bottom right shows the state of each thread after the selected step.
+Each row in the stack trace of a thread shows the current program counter,
 the method that is being evaluated, and the line of code that is
@@
 \item \texttt{Non-terminating States}: Harmony found one or more states
-from which there does not exist an execution such that all processes
+from which there does not exist an execution such that all threads
 terminate.  Harmony will not only print the non-terminating state with
-the shortest trace, but also the list of processes
+the shortest trace, but also the list of threads
 at that state, along with their name tags and program counters.
 \item \texttt{Stopped States}: Similar to non-terminating states,
-these are states in which some processes are left that cannot
+these are states in which some threads are left that cannot
 terminate.  \autoref{ch:synch} will explain what it means for
-a process to be stopped.
+a thread to be stopped.
 \end{itemize}
 
-In this case, Harmony reports a safety violation, in particular it
+In our case, Harmony reports a safety violation, in particular it
 reports that the assertion and that
@@
 The program got to the faulty execution in 5 ``steps.''
-The output has
-four columns:
+The output has four columns:
 \begin{enumerate}
-\item The name tag of the process;
-\item The sequence of program counters of the HVM instructions that the process executed;
-\item The current program counter of the process;
+\item The name tag of the thread;
+\item The sequence of program counters of the HVM instructions that the thread executed;
+\item The current program counter of the thread;
 \item The contents of the shared memory.
@@
 \begin{itemize}
-\item Process \texttt{incrementer/0}
-executed instructions 1 through 4, loading the value of
+\item Thread \texttt{incrementer/0}
+executed instructions 5 through 8, loading the value of
 \textit{count} but stopping just before storing 1 into \textit{count};
-\item Process \texttt{incrementer/1}
-executed instructions 1 through 11, storing 1 into
+\item Thread \texttt{incrementer/1}
+executed instructions 5 through 15, storing 1 into
 into \textit{count} and storing \texttt{True} into \textit{done}[1];
-\item Process \texttt{incrementer/0}
-continues execution, executing instructions 5 through 11
+\item Thread \texttt{incrementer/0}
+continues execution, executing instructions 9 through 15
 storing value 1 into \textit{count}
@@
 \end{itemize}
-Finally, process \texttt{main()} runs and finds the problem.
+Finally, thread \texttt{main()} runs and finds the problem.
 This makes precise the concurrency issue that we encountered.
@@
     in broad steps, how running the program can get into a bad state.
-\item What if we moved line~2 of \autoref{fig:incenter} to after the \textbf{if}
-    statement (between lines~5 and~6)?  Do you think that would work?  Run it through
+\item What if we moved line~6 of \autoref{fig:incenter} to after the \textbf{if}
+    statement (between lines~9 and~10)?  Do you think that would work?  Run it through
     Harmony and describe either why it works or why it does not work.
@@
 \end{code}
-\caption{\harmonylink{code/csbarebones.hny} A bare bones critical section with two processes.}
+\caption{\harmonylink{code/csbarebones.hny} A bare bones critical section with two threads.}
 \label{fig:csbarebones}
@@
 At the heart of the problem is that we would like make sure that, when
-the \textit{count} variable is being updated, no other process is
+the \textit{count} variable is being updated, no other thread is
 trying to do the same thing.  This is called a \emph{critical section}
 (aka critical region)~\cite{EWD123}:
-a set of instructions where only one process is allowed to execute at a
+a set of instructions where only one thread is allowed to execute at a
 time.
@@
 (although not necessarily impossible) if during this time no other
-process also tries to access the binary tree.
+thread also tries to access the binary tree.
 Typically, you want some invariant property of the data structure to hold
@@
 sections guarantee that no other
-process will be able to see it.
+thread will be able to see it.
 An implementation of a data structure that can be safely accessed by multiple
-processes (or threads) and free of race conditions is called \emph{thread-safe}.
+threads (or threads) and free of race conditions is called \emph{thread-safe}.
 \index{thread safety}%
@@
 
-A critical section is often modeled as processes in an infinite loop
+A critical section is often modeled as threads in an infinite loop
 entering and exiting the critical section.
@@
 identifying a location in the HVM bytecode.  The first thing we need to
-ensure is that there can never be two processes in the critical section.
+ensure is that there can never be two threads in the critical section.
 This property is called \emph{mutual exclusion}.
@@
 We would like to place an assertion at the \texttt{@cs} label that
-specifies that only the current process can be there.
+specifies that only the current thread can be there.
 
@@
 is the atom containing the name of the label (in this case, \texttt{.cs}).
-The operator returns a bag (multiset) of name tags of processes executing at that
+The operator returns a bag (multiset) of name tags of threads executing at that
 label.  The bag is represented by a dictionary that maps each element
@@
 \index{nametag operator}%
-that returns the name tag of the current process.
+that returns the name tag of the current thread.
 If you run the code through Harmony, the assertion should fail because
@@
 \end{code}
-This code will surely prevent two or more processes from being
+This code will surely prevent two or more threads from being
 at label \texttt{@cs} at the same time.
-But it does so by preventing \emph{any} process from reaching
+But it does so by preventing \emph{any} thread from reaching
 the critical section.
@@
 a property that ensures that \emph{nothing bad will happen}, in this case
-two processes being in the critical section.
+two threads being in the critical section.
 What we need now is a \emph{liveness property}:
@@
 (1) there exists a non-empty
-set $S$ of processes that are trying to enter the critical section and
-(2) processes in the critical section always leave eventually, then
-eventually one process in $S$ will enter the critical section.
+set $S$ of threads that are trying to enter the critical section and
+(2) threads in the critical section always leave eventually, then
+eventually one thread in $S$ will enter the critical section.
 We call this \emph{progress}.
@@
 algorithms in general, Harmony requires that every execution must be
-able to reach a state in which all processes have terminated.
+able to reach a state in which all threads have terminated.
 Clearly, even if mutual exclusion holds in \autoref{fig:csbarebones},
-the spawned processes never terminate.  We
-will instead model processes in critical sections using the framework in
-\autoref{fig:cs}: a process can \emph{choose} to enter a
+the spawned threads never terminate.  We
+will instead model threads in critical sections using the framework in
+\autoref{fig:cs}: a thread can \emph{choose} to enter a
 critical section more than once, but it can also choose to terminate, even
@@
 
+\begin{comment}
 \begin{figure}
@@
 \end{center}
-\caption{High-level state diagram specification of mutual exclusion with up to two processes.
-The first number in a state gives the number of processes; the second number is the
-number of processes in the critical section.}
+\caption{High-level state diagram specification of mutual exclusion with up to two threads.
+The first number in a state gives the number of threads; the second number is the
+number of threads in the critical section.}
 \label{fig:mutex}
@@
 the label on an arrow describes the transition.  In this case,
-the label contains two numbers: the total number of processes and the number
-of processes in the critical section.  A process that is in the critical
+the label contains two numbers: the total number of threads and the number
+of threads in the critical section.
+\end{comment}
+
+A thread that is in the critical
 section cannot terminate until after leaving the critical section.
@@
 it could be used to implement a critical section.
-The idea is that the lock is like a baton that at most one process can own
+The idea is that the lock is like a baton that at most one thread can own
 (or hold) at a time.
-A process that wants to enter the critical section at a time must obtain the
+A thread that wants to enter the critical section at a time must obtain the
 lock first and release it upon exiting the critical section.
 % Note that the word ``lock'' does not describe the concept well---it really
-% is more like a baton or token that only one process can hold.
+% is more like a baton or token that only one thread can hold.
 
@@
 Initially the lock is not owned, indicated by \textit{lockTaken} being \texttt{False}.
-To enter the critical section, a process waits until \textit{lockTaken} is \texttt{False}
+To enter the critical section, a thread waits until \textit{lockTaken} is \texttt{False}
 and then sets it to \texttt{True} to indicate that the lock has been taken.
-The process then executes the critical section.  Finally the process
+The thread then executes the critical section.  Finally the thread
 releases the lock by setting \textit{lockTaken} back to \texttt{False}.
@@
 \autoref{fig:naiveLockhtml} shows the Harmony output.
-Process~0 finds that the lock is available, but just before it stores
-\texttt{True} in \textit{lockTaken} in instruction~9, process~1 gets
+Thread~0 finds that the lock is available, but just before it stores
+\texttt{True} in \textit{lockTaken} in instruction~9, thread~1 gets
 to run.
@@
 \texttt{lockTaken} and moves on to the critical section.
-Finally, process~0 moves on, also stores \texttt{True} and also moves
-into the critical section.  Process~0 is the one that detects the
+Finally, thread~0 moves on, also stores \texttt{True} and also moves
+into the critical section.  Thread~0 is the one that detects the
 problem.
 The \textit{lockTaken} variable
-suffers from the same problem as the \textit{count} variable
-in \autoref{fig:inc}: operations
-on it consist of several instructions.  It is thus possible
-for both processes to believe the lock is available and to obtain the lock
+suffers from the same sort of race condition as the \textit{count} variable
+in \autoref{fig:inc}: testing and setting the lock
+consists of several instructions.  It is thus possible
+for both threads to believe the lock is available and to obtain the lock
 at the same time.
 
-\autoref{fig:upflags} presents a solution based on each process having
+Preventing multiple threads from updating the same variable,
+\autoref{fig:upflags} presents a solution based on each thread having
 a flag indicating that it is trying to enter the critical section.
-A process can write its own flag and read the flag of its peer.
-After setting its flag, the process waits until the other process
+A thread can write its own flag and read the flag of its peer.
+After setting its flag, the thread waits until the other thread
 ($1 - \mathit{self}$) is not trying to enter the critical section.
 If we run this program, the assertion does not fail.  In fact, this
-solution does prevent both processes being in the critical section at
+solution does prevent both threads being in the critical section at
 the same time.
 
-To see why, first note the following invariant: if process $i$ is in the
+To see why, first note the following invariant: if thread $i$ is in the
 critical section, then \textit{flags}[$i$] == \texttt{True}.
 Without loss of generality,
-suppose that process~0 sets \textit{flags}[0] at time $t_0$.
-Process 0 can only reach the critical section if at some time $t_1$,
+suppose that thread~0 sets \textit{flags}[0] at time $t_0$.
+Thread 0 can only reach the critical section if at some time $t_1$,
 $t_1 > t_0$, it finds that \textit{flags}[1] == \texttt{False}.
 Because of the invariant, \textit{flags}[1] == \texttt{False} implies that
-process~1 is not in the critical section at time $t_1$.
-Let $t_2$ be the time at which process~0 sets \textit{flags}[0]
-to \texttt{False}.  Process~0 is in the critical section sometime
+thread~1 is not in the critical section at time $t_1$.
+Let $t_2$ be the time at which thread~0 sets \textit{flags}[0]
+to \texttt{False}.  Thread~0 is in the critical section sometime
 between $t_1$ and $t_2$.
-It is easy to see that process~1 cannot enter the critical section
+It is easy to see that thread~1 cannot enter the critical section
 between $t_1$ and $t_2$, because \textit{flags}[1] == \texttt{False} at
@@
 The final na\"{\i}ve solution that we propose
-is based on a variable called \textit{turn}
-that alternates between 0 and 1.  When \textit{turn} = $i$, process~$i$ can
-enter the critical section, while process $1-i$ has to wait.  When done,
-process~$i$ sets \textit{turn} to $1-i$ to give the other process an
-opportunity to enter.
-An invariant of this solution is that while process~$i$ is in the critical
+is based on a variable called \textit{turn}.
+Each thread politely lets the other thread have a turn first.
+When \textit{turn} = $i$, thread~$i$ can
+enter the critical section, while thread $1-i$ has to wait.
+An invariant of this solution is that while thread~$i$ is in the critical
 section, \textit{turn} == $i$.
@@
 the same time, mutual exclusion is satisfied.
-The solution also has the nice property that
-processes 0 and 1 alternate entering the critical section.
+The solution also has the nice property that the thread that has been waiting the
+longest to enter the critical section can go next.
 
@@
 \emph{progress}, albeit for a different reason:
-if process~$i$ terminates instead of entering the critical section when it
-is process~$i$'s turn, process $1-i$ ends up waiting indefinitely for its
-turn.
+if thread~$i$ terminates instead of entering the critical section,
+thread $1-i$, politely, ends up waiting indefinitely for its turn.
+Too bad, it would have been a great solution if both threads try to enter the
+critical section continually.
 
@@
 If they don't, try to understand why.
-Do not despair if you can't figure it out---as
+Do not despair if you can't figure out how to develop a solution that satisfies both
+mutual exclusion and progress---as
 we will find out, it is possible but not easy.
@@
 in \autoref{fig:peterson}.
-A process first indicates its interest in entering the critical section
+A thread first indicates its interest in entering the critical section
 by setting its flag.
-It then politely gives way to the other process should it also want to
+It then politely gives way to the other thread should it also want to
 enter the critical section---if both do so at the same time one will
 win because writes to memory in Harmony are atomic.
-The process continues to be polite, waiting
-until either the other process is nowhere near the critical section
+The thread continues to be polite, waiting
+until either the other thread is nowhere near the critical section
 ($\mathit{flag}[1 - \mathit{self}] == \mathtt{False}$) or has given way
@@
 \glsadd{state}%
-\glsadd{process variable}%
+\glsadd{thread variable}%
 
@@
 A state is comprised of the values of the shared variables as well as
-the values of the process variables
-\index{process variable}%
-of each process, including its
+the values of the thread variables
+\index{thread variable}%
+of each thread, including its
 program counter and the contents of its stack.
-Each time a process executes a HVM machine instruction, the
-state changes (if only because the program counter of the process
+Each time a thread executes a HVM machine instruction, the
+state changes (if only because the program counter of the thread
 changes).  We call that a \emph{step}.
@@
 The HVM starts in an initial state in which there is only
-one process and its program counter is~0.  A \emph{trace}
+one thread and its program counter is~0.  A \emph{trace}
 \index{trace}%
@@
 One is when
-there is more than one process that can make a step.  The other is
-when a process executes a \texttt{choose} operation and there is
+there is more than one thread that can make a step.  The other is
+when a thread executes a \texttt{choose} operation and there is
 more than one choice.
@@
 A state is final
-when there are no processes left to make state changes.
+when there are no threads left to make state changes.
 
@@
 All states in which $x$ = 1, or all states where
-there are two or more processes executing, are examples of such predicates.
+there are two or more threads executing, are examples of such predicates.
 For our purposes, it is useful to define a predicate that says that at
-most one process is in the critical section.  We shall call such states
+most one thread is in the critical section.  We shall call such states
 \emph{exclusive}.
@@
 is a predicate that holds over all states that are reachable by that program.
-We want to show that exclusivity is an invariant.
+We want to show that exclusivity is an invariant because mutual exclusion means
+that all reachable states are exclusive.
 In other words, we want to show that the set of reachable states of executing
 the program
-is a subset of the set of states where there is at most one process in the critical
+is a subset of the set of states where there is at most one thread in the critical
 section.
@@
 states are reachable.
-But we do not have such a predicate: we know how to describe the set
+But we do not have such a predicate: we know how to define the set
 of reachable states, but given an arbitrary state it is not easy to
@@
 \item $\mathcal{I}$ holds in the initial state.
-\item For any state in which $\mathcal{I}$ holds and for any process in the
-state, if the process takes a step, then $\mathcal{I}$ also holds in the resulting state.
+\item For any state in which $\mathcal{I}$ holds (including unreachable ones) and for any
+thread in the state, if the thread takes a step, then $\mathcal{I}$ also holds in the
+resulting state.
 \end{itemize}
-Unlike an ordinary invariant, an inductive invariant must also hold over unreachable states.
-
-One candidate for such an invariant is exclusivity itself.
+
+One candidate for such a predicate is exclusivity itself.
 After all, it certainly holds over the initial state.
@@
 it holds over every reachable state.
-Unfortunately, it is not an \emph{inductive} invariant.
-To see why, we need to consider an \emph{unreachable} state.
-It is easy to construct one: let process~0 be at label \texttt{@cs}
-and process~1 be at the start of the \textbf{await} statement.
-Also imagine that in this particular state $\mathit{turn} = 1$.  Now let
-process~1 make a sequence of steps.  Because $\mathit{turn} = 1$,
-process~1 will stop waiting and also enter the critical
+Unfortunately, exclusivity is not an \emph{inductive} invariant.
+To see why, consider the following state $s$: let thread~0 be at label \texttt{@cs}
+and thread~1 be at the start of the \textbf{await} statement.
+Also, in state $s$, $\mathit{turn} = 1$.  Now let
+thread~1 make a sequence of steps.  Because $\mathit{turn} = 1$,
+thread~1 will stop waiting and also enter the critical
 section, entering a state that is not exclusive.
-So exclusivity is an invariant, but not an inductive invariant.
-Doing a inductive proof with an invariant that is not inductive is usually
-much harder than doing one with an invariant that is.
+So exclusivity is an invariant (holds over every reachable state, as demonstrated
+by Harmony),
+but not an inductive invariant (it will turn out that $s$ is not reachable).
+
+So we are looking for an inductive invariant that \emph{implies} exclusivity.
+In other words, the set of states where the inductive invariant holds
+must be a subset of the set of states where there is at most one thread in
+the critical section.
+
+Let us begin with considering the following important property:
+$\mathcal{F}(i) = \mathtt{thread}(i)@[8 \cdots 15] \Rightarrow \mathit{flags}[i]$,
+that is, if thread~$i$ is executing in lines 8 through 15, then $\mathit{flags}[i]$
+is set.
+Although it does not, by itself, imply exclusivity, we can show that
+$\mathcal{F}(i)$ is an inductive invariant (for both threads 1 and 2).
+To wit, it holds in the initial state, because in the initial state thread $i$ does
+not even exist yet.
+Now we have to show that if $\mathcal{F}(i)$ holds in some state, then
+$\mathcal{F}(i)$ also holds in a next state.
+Since only thread~$i$ ever changes $\mathit{flags}[i]$, we only need to
+consider steps by thread~$i$.
+Since $\mathcal{F}(i)$ holds, there are two cases to consider:
+\begin{enumerate}
+\item states in which $\mathit{flags}[i] = \texttt{true}$
+\item states in which $\lnot \mathtt{thread}(i)@[8 \cdots 15]$
+\end{enumerate}
+In each case we need to show that if thread $i$ takes a step, 
+$\mathcal{F}(i)$ still holds.
+In the first case, there is only one step that thread $i$ can take that would
+set $\mathit{flags}[i]$ to false: the step in line 15.  But executing the line
+would also take the thread out of lines $8 \cdots 15$, so $\mathcal{F}(i)$ continues to hold.
+In the second case (thread $i$ is not executing in lines $8 \cdots 15$), the only step
+that would cause thread $i$ to execute in lines $8 \cdots 15$ would be the step in line 7.
+But in that case $\mathit{flags}[i]$ would end up being true, so
+$\mathcal{F}(i)$ continues to hold as well.
+So $\mathcal{F}(i)$ is an inductive invariant (for both threads~0 and~1).
+
+While $\mathcal{F}(i)$ does not imply mutual exclusion, it does imply the following useful
+invariant: $\mathtt{thread}(i)@cs \Rightarrow \mathit{flags}[i]$: when thread $i$ is
+at the critical section, $\mathit{flags}[i]$ is set.  This seems obvious from the code,
+but now you know how to prove it.
+
+We need a stronger inductive invariant than $\mathcal{F}(i)$ to prove mutual exclusion.
+What else do we know when thread $i$ is in the critical section?
+Let $\mathcal{C}(i) = \lnot\mathit{flags}[1 - i] \lor \mathit{turn} = i$, that is,
+the condition on the \textbf{await} statement for thread~$i$.
+In a sequential program, $\mathcal{C}(i)$ would clearly hold if thread $i$ is in
+the critical section: $\mathtt{thread}(i)@cs \Rightarrow \mathcal{C}(i)$.
+However, because thread $1-i$ is executing concurrently, this property does not
+hold.  In particular, suppose thread 0 is at the critical section, $\mathit{flags}[0]$ = true,
+$\mathit{turn} = 1$, and thread 1 just finished the step in line 7,
+setting $\mathit{flags}[1]$ to true.  Notice that $C(0)$ is violated.
 
@@
 
-So we are looking for an inductive invariant that \emph{implies} exclusivity.
-In other words, the set of states where the inductive invariant holds
-must be a subset of the set of states where there is at most one process in
-the critical section.
-Let $\mathcal{C}(i) = \lnot\mathit{flags}[1 - i] \lor
-\mathit{turn} = i$, that is, the condition on the \textbf{await}
-statement for process~$i$.
-Let predicate $\mathcal{I}_p(i)$ be the following:
-if process $i$ is at label \texttt{@cs} (i.e., process $i$ is in the critical section),
-then $\mathcal{C}(i)$ holds or process $1-i$ is executing after setting
-$\mathit{flags}[i]$ but still before setting \textit{turn} to $1-i$.
-More formally, $\mathcal{I}_p(i) = \mathtt{process}(i)@cs \Rightarrow \mathcal{C}(i) \lor \mathtt{process}(1-i)@gate$.
-For Peterson's Algorithm, an inductive invariant that works well is
-$\mathcal{I}_p(0) \land \mathcal{I}_p(1)$.
-
-\autoref{fig:petersonproof} formalizes $\mathcal{I}_p(i)$ in Harmony.
-The label \texttt{@gate} refers to the instruction that sets \textit{turn} to $1-i$.
+Instead, we will use the following property: $\mathcal{G}(i) =
+\mathtt{thread}(i)@cs \Rightarrow \mathcal{C}(i) \lor \mathtt{thread}(1-i)@[8]$.
+That is, if thread $i$ is at the critical section, then
+either $\mathcal{C}(i)$ holds or thread $1-i$ is about to execute line 8.
+\autoref{fig:petersonproof} formalizes $\mathcal{G}(i)$ in Harmony.
+The label \texttt{@gate} refers to the step that sets \textit{turn} to $1-i$.
 You can run \autoref{fig:petersonproof} to determine
-that $\mathcal{I}_p(i)$ is indeed an invariant for $i = 0, 1$.
-
-To see that the inductive invariant implies exclusivity, suppose not.  Then
-when both process~0 and process~1 are in the critical section, the
-following must hold:
-$(\mathcal{C}(0) \lor \mathtt{process}(1)@gate) \land
- (\mathcal{C}(1) \lor \mathtt{process}(0)@gate)$.
-We know that both flags are set.
-We also know that neither process~0 nor process~1 is at label \texttt{@gate}
-(because they are both at label \texttt{@cs}),
-so this simplifies to $\mathcal{C}(0) \land \mathcal{C}(1)$.
-So we conclude that $\mathit{turn} = 0 \land \mathit{turn} = 1$, a
-logical impossibility.  Thus $\mathcal{I}_p(i)$ implies exclusivity.
-
-To see that $\mathcal{I}_p(0) \land \mathcal{I}_p(1)$ is, in fact, an inductive invariant, first note that
-it holds in the initial state, because in the initial state no process
-is in the critical section.
+that $\mathcal{G}(i)$ is an invariant for $i = 0, 1$.
+Moreover, if $\mathcal{F}(i)$ and $\mathcal{G}(i)$ both hold for $i = 0, 1$,
+then mutual exclusion holds.  We can show this using proof by
+contradiction.  Suppose mutual exclusion is violated and thus both threads are in
+the critical section.  By $\mathcal{F}$ it must be the case that both
+\textit{flags} are true.  By $\mathcal{G}$ and the fact that neither thread
+is at label \texttt{@gate}, we know that both $C(0)$ and $C(1)$ must hold.
+This then implies that $\mathit{turn} = 0 \land \mathit{turn} = 1$, providing
+the desired contradiction.
+
+We claim that $\mathcal{G}(i)$ is an inductive invariant.
+First, since neither thread exists in the initial state, it is clear that
+$\mathcal{G}(i)$ holds in the initial state.
 Without loss of generality, suppose $i=0$ (a benefit from the fact that the algorithm is
-symmetric for both processes).  We still have to show that if we are in a state
-in which $\mathcal{I}_p(0)$ holds, then any step will result in a state in which
-$\mathcal{I}_p(0)$ still holds.
-
-First consider the case that process~0 is at label \texttt{@cs}.  If process~0
-were to take a step, then in the next state process~0 would be no longer
-at that label and $\mathcal{I}_p(0)$ would hold trivially over the next state.
-Therefore we only need to consider a step by process~1.
+symmetric for both threads).  We still have to show that if we are in a state
+in which $\mathcal{G}(0)$ holds, then any step will result in a
+state in which $\mathcal{G}(0)$ still holds.
+
+First consider the case that thread~0 is at label \texttt{@cs}.  If thread~0
+were to take a step, then in the next state thread~0 would be no longer
+at that label and $\mathcal{G}(0)$ would hold trivially over the next state.
+Therefore we only need to consider a step by thread~1.
 %
-From $\mathcal{I}_p(0)$ we know that one of the following three cases must hold before
-process~1 takes a step:
+From $\mathcal{G}$ we know that one of the following three cases must hold before
+thread~1 takes a step:
 \begin{enumerate}
@@
 \item \textit{turn} = 0;
-\item process~1 is at label \texttt{@gate}.
+\item thread~1 is at label \texttt{@gate}.
 \end{enumerate}
@@
 Let us consider each of these cases.
-We have to show that if process~1 takes a step, then one of those
+We have to show that if thread~1 takes a step, then one of those
 cases must hold after the step.
-In the first case, if process~1 takes a step, there are two possibilities:
+In the first case, if thread~1 takes a step, there are two possibilities:
 either $flags[1]$ will still be \texttt{False} (in which case the first case
 continues to hold), or $flags[1]$ will be \texttt{True}
-and process~1 will be at label \texttt{@gate} (in which case the third case
+and thread~1 will be at label \texttt{@gate} (in which case the third case
 will hold).
-We know that process~1 never sets \textit{turn} to 1, so
+We know that thread~1 never sets \textit{turn} to 1, so
 if the second case holds before the step, it will also hold after the step.
-Finally, if process~1 is at label \texttt{@gate} before the step, then after
+Finally, if thread~1 is at label \texttt{@gate} before the step, then after
 the step \textit{turn} will equal 0, and therefore the second case will hold
@@
 
-Now consider the case where process~0 is not in the critical section,
-and therefore $\mathcal{I}_p(0)$ holds trivially because false implies
+Now consider the case where thread~0 is not in the critical section,
+and therefore $\mathcal{G}(0)$ holds trivially because false implies
 anything.
@@
 \begin{enumerate}
-\item Process~1 takes a step.  But then process~0 is still not in the critical
-section and $\mathcal{I}_p(0)$ continues to hold;
-\item Process~0 takes a step but still is not in the critical section.
-Then again $\mathcal{I}_p(0)$ continues to hold.
-\item Process~0 takes a step and ends up in the critical section.
-Because process~0 entered the critical section, we know that
+\item Thread~1 takes a step.  But then thread~0 is still not in the critical
+section and $\mathcal{G}(0)$ continues to hold;
+\item Thread~0 takes a step but still is not in the critical section.
+Then again $\mathcal{G}(0)$ continues to hold.
+\item Thread~0 takes a step and ends up in the critical section.
+Because thread~0 entered the critical section, we know that
 \textit{flags}[1] = \texttt{False} or \textit{turn} == 0 because
 of the \texttt{await} condition.
-And hence $\mathcal{I}_p(0)$ continues to hold in that case as well.
+And hence $\mathcal{G}(0)$ continues to hold in that case as well.
 \end{enumerate}
@@
 different ways: one by letting Harmony explore all possible executions, the
-other using an inductive invariant and proof by induction.  The former
+other using inductive invariants and proof by induction.  The former
 is certainly easier, but it does not provide intuition for why the
-algorithm works.  The second provides much more insight.  We therefore
-encourage to include invariants, particularly inductive ones, in your
-Harmony code.
+algorithm works.  The second provides much more insight.
+
+Even though they are not strictly necessary, we encourage to include invariants
+in your Harmony code. They can provide important insights into why the code works.
 
@@
 page.  The page claimed that the following predicate is invariant:
-if process $i$ is in the critical section, then $\mathcal{C}(i)$ (i.e.,
-$\mathcal{I}_p(i)$ without the disjunct that process $1-i$ is at label \texttt{@gate}).
+if thread $i$ is in the critical section, then $\mathcal{C}(i)$ (i.e.,
+$\mathcal{G}$ without the disjunct that thread $1-i$ is at label \texttt{@gate}).
 To demonstrate that this predicate is not invariant, you can remove the
@@
 mutual exclusion problem.  It is similar to the one in
-\autoref{fig:upflags}, but has a process \emph{back out and try again}
-if it finds that the other process is either trying to enter the critical
+\autoref{fig:upflags}, but has a thread \emph{back out and try again}
+if it finds that the other thread is either trying to enter the critical
 section or already has.  Compare this algorithm with Peterson's.  What are
@@
 \item
-Can you find an inductive invariant for the algorithm in
+Can you find one or more inductive invariants for the algorithm in
 \autoref{fig:csonebit} to prove it correct?
@@
 
-    process X:                          process Y:
+    thread X:                          thread Y:
         X0: flagX = True                   Y0: flagY = True
@@
 Peterson's algorithm (setting \textit{flags}[\textit{self}])
-to \texttt{True} and \textit{turn} to 1 - \textit{self}) can be reversed.
+to \texttt{True} and \textit{turn} to $1 - \mathit{self}$) can be reversed.
 After all, they are different variables assigned independent values---in a
@@
 \item Bonus question:
-Can you generalize Peterson's algorithm to more than two processes?
+Can you generalize Peterson's algorithm to more than two threads?
 \item Bonus question:
@@
 \href{https://en.wikipedia.org/wiki/Lamport%27s_bakery_algorithm}{Lamport's Bakery Algorithm}.
-Note that the last one uses unbounded state, so you should modify the processes so they
+Note that the last one uses unbounded state, so you should modify the threads so they
 only try to enter the critical section a bounded number of times.
@@
 \end{code}
-\caption{\harmonylink{code/csonebit.hny} Mutual exclusion using a flag per process.}
+\caption{\harmonylink{code/csonebit.hny} Mutual exclusion using a flag per thread.}
 \label{fig:csonebit}
@@
 Both methods \texttt{P\_enter} and \texttt{P\_exit} take two arguments:
-a pointer to a mutex and the process identifier (0 or 1).
+a pointer to a mutex and the thread identifier (0 or 1).
 $\mathit{pm}\rightarrow\mathtt{turn}$ is the value of the \texttt{.turn} key
@@
 lock is that checking the lock and setting it if it is available is
-not \emph{atomic}.  Thus multiple processes contending for the lock
+not \emph{atomic}.  Thus multiple threads contending for the lock
 can all ``grab the lock'' at the same time.  While Peterson's
 algorithm gets around the problem, it is not efficient, especially
-if generalized to multiple processes.  Instead, multi-core processors provide
+if generalized to multiple threads.  Instead, multi-core processors provide
 so-called \emph{interlock instructions}:
@@
 of it.  Second, a group of Harmony statements can be made atomic
-through its \textbf{atomic}
+through the \textbf{atomic}
 \index{atomic statement}%
@@
 Here $s$ points to a shared Boolean variable and $p$
-to a private Boolean variable, belonging to some process.
+to a private Boolean variable, belonging to some thread.
 The operation copies the value of the shared variable to the
@@
 \autoref{fig:tas} goes on to implement mutual exclusion for
-a set of \texttt{N} processes.
+a set of \texttt{N} threads.
 The approach is called \emph{spinlock},
 \index{spinlock}%
-because each process is ``spinning'' (executing a tight loop) until
+because each thread is ``spinning'' (executing a tight loop) until
 it can acquire the lock.
@@
 Variable \textit{shared} is initialized to \texttt{False} while
-\textit{private}[$i$] for each process $i$ is initialized to \texttt{True}.
+\textit{private}[$i$] for each thread $i$ is initialized to \texttt{True}.
 An important invariant, $\mathcal{I}_1$, of the program is that at any time at most
 one of these variables is \texttt{False}.
-Another invariant, $\mathcal{I}_2(i)$, is that if process $i$ is in the critical section,
+Another invariant, $\mathcal{I}_2(i)$, is that if thread $i$ is executing between
+lines $18 \cdots 25$ (which includes the critical section),
 then \textit{private}[$i$] == \texttt{False}.
 Between the two (i.e., $\mathcal{I}_1 \land \forall i: \mathcal{I}_2(i)$),
-it is clear that only one process can be in the
+it is clear that only one thread can be in the
 critical section at the same time.
 
-$\mathcal{I}_1$ is an inductive invariant.
+$\mathcal{I}_1 \land \forall i: \mathcal{I}_2(i)$ is an inductive invariant.
 To see that invariant $\mathcal{I}_1$ is maintained, note that
@@
 \end{enumerate}
-Invariant $\mathcal{I}_1$ is also easy to verify for exiting the critical section.
-Invariant $\mathcal{I}_2(i)$ is obvious as (i) process $i$ only proceeds to the critical
-section if \textit{private}[$i$] == \texttt{False}, and (ii) no other process modifies
+Invariant $\mathcal{I}_1$ is also easy to verify for exiting the critical section
+because we can assume, by the induction hypothesis, that $\textit{private}[i]$ is
+\texttt{True} just before exiting the critical section.
+Invariant $\mathcal{I}_2(i)$ is obvious as (i) thread $i$ only proceeds to the critical
+section if \textit{private}[$i$] is \texttt{False}, and (ii) no other thread modifies
 \textit{private}[$i$].
@@
 For $\mathcal{I}_2$ we only need an assertion at label \texttt{@cs} because the
-premisse is that there is a process at that label.  However, we would
+premisse is that there is a thread at that label.  However, we would
 like to check $\mathcal{I}_1$ in \emph{every state} (after the variables have
@@
 
-We can do this by adding another process that
+We can do this by adding another thread that
 checks the invariant.  \autoref{fig:tasinv} shows the code.
@@
 in a state.  It introduces a new feature of Harmony: the ability to have
-variables local to a method.  In this case, the process variable \textit{sum}
+variables local to a method.  In this case, the thread variable \textit{sum}
 is used to compute the number of shared variables that have value
 \texttt{False}.
-The method is invoked by a process that runs alongside
-the other processes.
+The method is invoked by a thread that runs alongside
+the other threads.
 In Harmony, \textbf{assert} statements are executed atomically, so the
 evaluation of the assertion is not interleaved with the execution
-of other processes.
-Because Harmony tries every possible execution, the process is guaranteed
+of other threads.
+Because Harmony tries every possible execution, the thread is guaranteed
 to find violations of the invariant if it does not hold.
@@
 
-\chapter{Locks and Blocking}
+\chapter{Binary Semaphores and Blocking}
 \label{ch:synch}
@@
 In \autoref{fig:tas} we have shown a solution based on a shared
-variable and a private variable for each process.   The private
+variable and a private variable for each thread.   The private
 variables themselves are actually implemented as shared variables,
-but they are accessed only by their respective processes.
-There is no need to keep \textit{private} as a shared
+but they are accessed only by their respective threads.
+A thread usually does not keep explicit track of whether it has a lock
+or not because it is usually implied by the control flow of the program.
+So there is no need to keep \textit{private} as a shared
 variable---we only did so to be able to show and check the invariants.
@@
 but uses test-and-set to check and set the lock variable atomically.
-This approach is general for any number of processes.
+This approach is general for any number of threads.
 
@@
 The former ensures that while the
-\texttt{atomic} statement is executing no other process can execute.
-The latter allows multiple processes to run concurrently,
+\texttt{atomic} statement is executing no other thread can execute.
+The latter allows multiple threads to run concurrently,
 just not within the critical section.
@@
 When solving synchronization problems you should not directly use
-\emph{atomic} statements but use the synchronization primitives that are available
+\texttt{atomic} statements but use the synchronization primitives that are available
 to you.  But if you want to design a new synchronization primitive, then
@@
 You can also use \textbf{atomic} statements in your test code.
-In fact, \texttt{assert} statements are executed atomically.
+In fact, as mentioned before, \texttt{assert} statements are executed atomically.
 
@@
 \end{code}
-\caption{\harmonylink{modules/synch.hny} The \texttt{Lock} interface and implementation in the \texttt{synch} module.}
+\caption{\harmonylink{modules/synch.hny} The binary semaphore interface and implementation in the \texttt{synch} module.}
 \label{fig:spinlocks}
@@
 critical section.
+That way, when a thread makes modifications to the data structure that take
+multiple steps, other threads will not see the intermediate inconsistent
+states of the data structure.
 When there is a bug in a program because some code omitted obtaining
@@
 
-Because locks are so common, Harmony has a module called
-\texttt{synch} that includes support for locks.
+\index{binary semaphore}
+Locks are a special case of \emph{binary semaphores}.
+Like locks, binary semaphores are either in an \emph{acquired} state
+or a \emph{released} state.
+Processes trying to acquire an already acquired binary semaphore
+have to wait until the binary semaphore is released.
+It is illegal to release a binary semaphore that is in a released
+state.
+A lock is a binary semaphore that, by convention, is initialized to a
+released state, and is always released by the same thread that acquired it.
+A binary semaphore can be initialized to an acquired state and can be
+released by threads other than the thread that last acquired the
+binary semaphore.
+
+Harmony has a module called \texttt{synch} that includes support for
+binary semaphores and locks.
 \autoref{fig:spinlocks} shows how they are implemented, and
@@
 in this case to fix the program of \autoref{fig:inc}.
-Notice that the module completely hides the implementation of the
-lock.
+Notice that the module hides the implementation of
+binary semaphores.
 The \texttt{synch} module includes a variety of other useful
@@
 
-\glsadd{blocked process}%
-
-We call a process \emph{blocked}
-\index{blocked process}%
-if a process cannot change the state or terminate or can only do so if
-another process changes the state first.
-A process trying to
-acquire a test-and-set spinlock held by another process is a good example
-of a process being blocked.
-The only way forward is if the other process releases the lock.
-A process that is in an infinite loop is also considered blocked,
+\glsadd{blocked thread}%
+
+We call a thread \emph{blocked}
+\index{blocked thread}%
+if a thread cannot change the state or terminate unless
+another thread changes the state first.
+A thread trying to
+acquire a test-and-set spinlock held by another thread is a good example
+of a thread being blocked.
+The only way forward is if the other thread releases the spinlock.
+A thread that is in an infinite loop is also considered blocked,
 although Harmony can often detect infinite loops specifically.
@@
 \end{code}
-\caption{\harmonylink{modules/syncS.hny} The \texttt{Lock} interface in the \texttt{synchS} module uses suspension.}
+\caption{\harmonylink{modules/syncS.hny} The binary semaphore interface in the \texttt{synchS} module uses suspension.}
 \label{fig:suspension}
@@
 
-In most operating systems, processes are virtual and can be suspended
-until some condition changes.  For example, a process can be suspended
-until the lock is available to it.
-In Harmony, a process can suspend itself and save its context (state) in a
-list.  Recall that the context of a process consists of its name tag,
+In most operating systems, threads are virtual (as opposed to ``raw CPU cores'')
+and can be suspended until some condition changes.
+For example, a thread that is trying to acquire a lock can be suspended until the lock is
+available.
+In Harmony, a thread can suspend itself and save its context (state) in a
+list.  Recall that the context of a thread consists of its name tag,
 its program counter, and the contents of its register and stack.
@@
 Here $L$ is a shared variable containing a list.
-Another process can revive the process using the \textbf{go}
+Another thread can revive the thread using the \textbf{go}
 \index{go statement}%
@@
 Here $C$ is a context and $R$ is a Harmony value.
-It causes a process with context $C$ to be added to the state that has
+It causes a thread with context $C$ to be added to the state that has
 just executed the \textbf{stop}
@@
 
-There is a second version of the \texttt{synch} module that uses suspension
-instead of spinlocks.
-\autoref{fig:suspension} shows the same \texttt{Lock} interface implemented
+There is a second version of the \texttt{synch} module, called \texttt{synchS},
+that uses suspension instead of spinlocks.
+\autoref{fig:suspension} shows the same interface implemented
 using suspension.
@@
 \begin{itemize}
-\item A \texttt{Lock} maintains both a boolean indicating whether the
-lock is taken and a list of contexts of processes that want to acquire the lock.
+\item A binary semaphore maintains both a boolean indicating whether the
+binary semaphore is currently acquired and a list of contexts of threads that want to
+acquire the binary semaphore.
 \item
-\texttt{lock}
-\index{lock}%
-sets the lock if available and suspends itself if not.
+\texttt{acquire}()
+\index{acquire}%
+acquires the binary semaphore if available and suspends the invoking thread if not.
 Note that \textbf{stop} is called within an \textbf{atomic} statement---this is
 the only exception to an atomic statement running to completion.  While the
-process is running no other processes can run, but when it suspends itself
-it allows other processes to run.
+thread is running no other threads can run, but when the thread suspends itself
+other threads can run.
 \item
-\texttt{unlock}
-\index{unlock}%
-checks to see if any processes are waiting to get the lock.
+\texttt{release}()
+\index{release}%
+checks to see if any threads are waiting to acquire the binary semaphore.
 If so, it uses the \texttt{head} and \texttt{tail}
 methods from the \texttt{list} module (see Appendix~\ref{ap:module})
-to resume the first process that got
+to resume the first thread that got
 suspended and to remove its context from the list.
 \end{itemize}
-Selecting the first process is a design choice.  Another implementation could
+Selecting the first thread is a design choice.  Another implementation could
 have picked the last one, and yet another implementation could have used
 \texttt{choose} to pick an arbitrary one.  Selecting the first is a common
-choice in lock implementations as it prevents \emph{starvation}:
+choice in binary semaphore implementations as it prevents \emph{starvation}:
 \index{starvation}%
-every process
-gets a chance to obtain the lock (assuming every process eventually releases
+every thread
+gets a chance to acquire the binary semaphore (assuming every thread eventually releases
 it).  \autoref{ch:starvation} will talk more about starvation and how
@@
 Part of the reason is that the \texttt{synchS} module keeps track of the order
-in which processes wait for a lock, while the \texttt{synch} module does not.
+in which threads wait to acquire a binary semaphore, while the \texttt{synch} module does not.
 
@@
 should return a pair that sums to 100.  Add the necessary synchronization code.
-\item \label{ex:trylock} Implement \texttt{trylock}(\textit{lk}) as an additional
+\item \label{ex:trylock} Implement \texttt{tryAcquire}(\textit{b}) as an additional
 interface for both the \texttt{synch} and \texttt{synchS} modules.
-This interface is like \texttt{lock}(\textit{lk}) but never blocks.  It
-returns \texttt{True} if the lock was available (and now acquired) or \texttt{False}
-if the lock was already taken.
+This interface is like \texttt{acquire}(\textit{b}) but never blocks.  It
+returns \texttt{True} if the binary semaphore was available (and now acquired) or \texttt{False}
+if the binary semaphore was already acquired.
 Hint: you do not have to change the existing code.
@@
 By way of example, we will demonstrate how to build a concurrent queue.
-The queue will have the following API:
+The \texttt{queue} module will have the following API:
 \begin{itemize}
-\item $q = \mathtt{Qnew}()$: allocate a new queue;
-\item $\mathtt{Qenqueue}(q, v)$: add $v$ to the tail of the queue;
-\item $r = \mathtt{Qdequeue}(q)$: returns $r = ()$ if $q$ is empty or $r = (v,)$ (a singleton tuple)
+\item $q = \mathtt{Queue}()$: allocate a new queue;
+\item $\mathtt{enqueue}(q, v)$: add $v$ to the tail of the queue;
+\item $r = \mathtt{dequeue}(q)$: returns $r = ()$ if $q$ is empty or $r = (v,)$ (a singleton tuple)
 if $v$ was at the head of the queue.
@@
 The \texttt{head} pointer is \textbf{None} if and only if the queue is empty.
-
-\texttt{Qnew()} returns a new queue object consisting of a \textbf{None} head
+(\textbf{None} is a special address value that is not the address of any
+memory location.)
+
+\texttt{queue.Queue()} returns a new queue object consisting of a \textbf{None} head
 and tail pointer and a lock.
-$\mathtt{Qenqueue}(q, v)$ and $\mathtt{Qdequeue}(q)$ both take a pointer $q$ to the
+$\mathtt{queue.enqueue}(q, v)$ and $\mathtt{queue.dequeue}(q)$ both take a pointer $q$ to the
 queue object because both may modify the queue.
@@
 separate locks for the head and tail, allowing an enqueue and a dequeue operation
-to proceed concurrently.  To avoid contention between the head and the tail that
-requires both locks, the queue uses a dummy node at the head of the linked list.
-Except for the initial state of the queue, the dummy node is the last node that
-was dequeued.  Note that neither the \texttt{head} nor \texttt{tail} pointer are
+to proceed concurrently.  To avoid contention between the head and the tail,
+the queue uses a dummy node at the head of the linked list.
+Except initially, the dummy node is the last node that was dequeued.
+Note that neither the \texttt{head} nor \texttt{tail} pointer are
 ever \textbf{None}.
@@
 \begin{problems}
-\item \label{ex:qcontains} Add a method $\mathtt{Qcontains}(q, v)$ to both \autoref{fig:queue} and \autoref{fig:queueMS}
+\item \label{ex:qcontains} Add a method $\mathtt{contains}(q, v)$ to both \autoref{fig:queue} and \autoref{fig:queueMS}
 that checks to see if $v$ is in queue $q$.
-\item Add a method $\mathtt{Qremove}(q, v)$ to both \autoref{fig:queue} and \autoref{fig:queueMS}
-that removes $v$, if any, from queue $q$.
+\item Add a method $\mathtt{remove}(q, v)$ to both \autoref{fig:queue} and \autoref{fig:queueMS}
+that removes all occurrences of $v$, if any, from queue $q$.
 \item The test program in \autoref{fig:queuetest} is not very thorough.  Design
 and implement a better one.
-\item Create a thread-safe sorted binary tree.  Implement methods
-$\mathtt{BTnew}()$ to create a new binary tree, $\mathtt{BTinsert}(t, v)$ that inserts $v$ into
-tree $t$, and $\mathtt{BTcontains}(t, v)$ that checks if $v$ is in tree $t$.  Use a single
+\item Create a thread-safe sorted binary tree.  Implement a module
+\texttt{bintree} with methods
+$\mathtt{BinTree}()$ to create a new binary tree, $\mathtt{insert}(t, v)$ that inserts $v$ into
+tree $t$, and $\mathtt{contains}(t, v)$ that checks if $v$ is in tree $t$.  Use a single
 lock per binary tree.
-\item Bonus question: create a binary tree that uses, instead of a single lock per tree,
+\item Bonus problem: create a binary tree that uses, instead of a single lock per tree,
 a lock for each node in the tree.
+\item \label{ex:qsort}
+Bonus problem: \autoref{fig:qsort} shows an iterative implementation of the Qsort
+algorithm, and \autoref{fig:qsorttest} an accompanying test program.
+The array to be sorted is stored in  shared variable $a$.
+Another shared variable, \textit{todo}, contains the ranges of the
+array that need to be sorted (initially the entire array).
+Re-using as much of this code as you can, implement a parallel version of
+this.  You should not have to change the methods \texttt{swap}, \texttt{partition},
+or \texttt{sortrange} for this.  Create \texttt{NWORKERS} ``worker threads''
+that should replace the \texttt{qsort} code.  Each worker loops until \textit{todo}
+is empty and sorts the ranges that it finds until then.  The \texttt{main}
+thread needs to wait until all workers are done.  Use a semaphore for this:
+the \texttt{main} thread should use \texttt{P} to wait for each
+worker to terminate.  After that it should check that the array is sorted and
+that it is a permutation of the original array.
 \end{problems}
 
-\chapter{Reader/Writer Locks using Busy Waiting}
-\label{ch:rdwrbusy}
+\begin{figure}
+\begin{code}
+\harmonysource{qsort}
+\end{code}
+\caption{\harmonylink{code/qsort.hny} Iterative qsort() implementation.}
+\label{fig:qsort}
+\end{figure}
+
+\begin{figure}
+\begin{code}
+\harmonysource{qsorttest}
+\end{code}
+\caption{\harmonylink{code/qsorttest.hny} Test program for \autoref{fig:qsort}.}
+\label{fig:qsorttest}
+\end{figure}
+
+\chapter{Conditional Waiting}
+\label{ch:condwait}
+
+Critical sections enable multiple threads
+to easily share data structures whose modification
+requires multiple steps.
+A critical section allows only one thread to execute its code at a time.
+When a thread arrives at a critical section,
+the thread blocks until there is no other thread in the critical section.
+
+\glsadd{busy waiting}%
+\index{busy waiting}%
+Sometimes it is useful for a thread to block waiting for additional
+conditions.
+For example, when dequeuing from an empty shared queue,
+it may be useful for the thread to block until the queue is non-empty
+instead of returning an error.
+The alternative would be \emph{busy waiting} (aka \emph{spin-waiting}),
+where the thread repeatedly tries to dequeue an item until it is successful.
+Doing so wastes CPU cycles and adds contention to queue access.
+%
+A thread that is busy waiting until the queue is non-empty cannot
+make progress until another thread enqueues an item.
+However, the thread is not considered \emph{blocked} because it is
+changing the shared state by repeatedly acquiring and releasing the
+lock.
+We would like to find a solution to \emph{conditional waiting}
+so that a thread blocks until the condition holds---or at least most
+of the time.
+
+Before we do so, we will give two classic examples of synchronization
+problems that involve conditional waiting: \emph{reader/writer locks}
+and \emph{bounded buffers}.
+
+\begin{figure}
+\begin{code}
+\harmonysource{RWtest}
+\end{code}
+\caption{\harmonylink{code/RWtest.hny} Test code for reader/writer locks.}
+\label{fig:rwtest}
+\end{figure}
+
+\section{Reader/Writer Locks}
 \index{reader/writer lock}%
-
 \glsadd{reader/writer lock}%
 
+\begin{comment}
 \begin{figure}
@@
 \label{fig:rwbusy}
-\end{figure}
-
-\begin{figure}
-\begin{code}
-\harmonysource{RWtest}
-\end{code}
-\caption{\harmonylink{code/RWtest.hny} Test code for \autoref{fig:rwbusy}.}
-\label{fig:rwbusytest}
 \end{figure}
@@
 \caption{High-level state diagram specification of reader/writer locks with
-up to two processes.
-The first number in a state gives the number of processes; the second number is the
-number of processes reading in the critical section; the third is the number of
-processes writing in the critical section.}
+up to two threads.
+The first number in a state gives the number of threads; the second number is the
+number of threads reading in the critical section; the third is the number of
+threads writing in the critical section.}
 \label{fig:rdwr}
 \end{figure}
+\end{comment}
 
 Locks are useful when accessing a shared data structure.  By preventing
-more than one process from accessing the data structure at the same
+more than one thread from accessing the data structure at the same
 time, conflicting accesses are avoided.  However, not all concurrent
 accesses conflict, and opportunities for concurrency may be lost,
-hurting performance.  One important case is when multiple processes
+hurting performance.  One important case is when multiple threads
 are simply reading the data structure.
@@
 What we want is a special kind of lock that allows either (i) one writer
-or (ii) one or more readers to be in the critical section.  This is called
+or (ii) one or more readers to acquire the lock.  This is called
 a \emph{reader/writer lock}~\cite{CHP71}.
+%
+A reader/writer lock is an object whose abstract (and opaque)
+state contains two integer counters:
+\begin{enumerate}
+\item \textit{nreaders}: the number of readers
+\item \textit{nwriters}: the number of writers
+\end{enumerate}
+satisfying the following invariant:
+\begin{itemize}
+\item [] $(\mathit{nreaders} \ge 0 \land \mathit{nwriters} = 0) \lor
+    (\mathit{nreaders} = 0 \land 0 \le \mathit{nwriters} \le 1)$
+\end{itemize}
+
+There are four operations on a reader/writer lock \textit{rw}:
+\begin{itemize}
+\item \texttt{read\_acquire}(\textit{rw}): waits until $\mathit{nwriters} = 0$
+and then increments \textit{nreaders};
+\item \texttt{read\_release}(\textit{rw}): decrements $\mathit{nreaders}$;
+\item \texttt{write\_acquire}(\textit{rw}): waits until
+$\mathit{nreaders} = \mathit{nwriters} = 0$
+and then sets \textit{nwriters} to 1;
+\item \texttt{write\_release}(\textit{rw}): sets $\mathit{nwriters}$ to 0.
+\end{itemize}
+
+\autoref{fig:rwtest} shows how reader/writer locks operations
+may be tested.
+Similar to ordinary locks, a thread is restricted in how it is allowed to
+invoke these operations.
+In particular, a thread can only release a reader/writer lock for reading
+if it acquired it for reading and the same for writing.
+Moreover, a thread is only allowed the acquire a reader/writer lock once.
+
+\begin{comment}
 We will explore various ways of implementing reader/writer locks in
@@
 of readers and one that maintains the number of writers.
-\autoref{fig:rwbusytest} shows how the code may be used.
+\autoref{fig:rwtest} shows how the code may be used.
 We will call the lock the \emph{mutex} to distinguish it clearly from
@@
 The mutex is used to protect shared access to the counters.
-The program shows a process that executes in a loop.
+The program shows a thread that executes in a loop.
 Each time, it decides whether to read or write.
@@
 label \texttt{@rcs} \emph{and} there cannot be another writer at
-label \texttt{@wcs}.  \autoref{fig:rdwr} shows a high-level
-specification for two processes.
-
-A process that wants to read first waits until there are no writers:
+label \texttt{@wcs}.
+
+\autoref{fig:rdwr} shows a high-level
+specification for two threads.
+
+A thread that wants to read first waits until there are no writers:
 \textit{nwriters} == 0.  Then it increments the number of readers.
 Similarly, a
-process that wants to write waits until there are no readers \emph{or} writers.
+thread that wants to write waits until there are no readers \emph{or} writers.
 Then it increments the number of writers.
@@
 specification.
-The solution also supports progress: if no process is in the critical
-section then any process can enter.  Better still: if any reader is in the
+The solution also supports progress: if no thread is in the critical
+section then any thread can enter.  Better still: if any reader is in the
 critical section, any other reader is also able to enter.
@@
 \index{spin waiting}%
-processes spin in a loop until some desirable application-level condition is met,
+threads spin in a loop until some desirable application-level condition is met,
 wasting CPU cycles.
 \autoref{fig:inc} also has an example of busy-waiting: the \texttt{main}
-process waits for the other two processes to finish by checking their \texttt{done}
+thread waits for the other two threads to finish by checking their \texttt{done}
 flags.
@@
 (see \autoref{fig:spinlocks}).
-However, there is a big difference.  As we pointed out, the processes that
+However, there is a big difference.  As we pointed out, the threads that
 are waiting for a spinlock are \emph{blocked}.
 In most operating systems and programming language runtimes,
-when a process acquires a mutex, the process is placed on a scheduling
+when a thread acquires a mutex, the thread is placed on a scheduling
 queue and stops using CPU cycles until the mutex becomes available.
@@
 there is contention for the mutex.
-A process trying to obtain a read or write lock in
+A thread trying to obtain a read or write lock in
 \autoref{fig:rwbusy} tries to obtain the mutex repeatedly.
 
-Thus, it is considered ok to have processes be blocked while waiting
+Thus, it is considered ok to have threads be blocked while waiting
 for application-specific conditions, but not for them to
@@
 all readers and writers from entering the critical section.
-In that case, those processes should ideally be blocked and Harmony
+In that case, those threads should ideally be blocked and Harmony
 can test for that by running it with the \texttt{-b} flag.
@@
 that all states are non-terminating and lead to a state in which
-all processes are blocked.
+all threads are blocked.
 
@@
 \item Draw additional states and steps in \autoref{fig:rdwr}
-for three processes.
-\item Optimize your solutions to \autoref{ex:qcontains} to use reader/writer locks.
-\item \label{ex:boundlock} Using busy waiting, implement a ``bound lock'' that allows
-up to \texttt{M} processes to acquire it at the same time.\footnote{A bound lock is a restricted version of a counting semaphore, as we will learn later in this book.}
-A bound lock
-with \texttt{M == 1} is an ordinary lock.
-You should define a constant \texttt{M} and two methods:
-\texttt{acquire\_bound\_lock()}
-and \texttt{release\_bound\_lock()}.
-(Bound locks are useful for situations where too many processes working
-at the same time might exhaust some resource such as a cache.)
-\item Write a test program for your bound lock
-that checks that no more than \texttt{M} processes can acquire the
-bound lock.
+for three threads.
 \end{problems}
@@
 the critical section releases \textit{rwlock}.
-The implementation satisfies progress: if no process is in the critical
-section then any process can enter.
+The implementation satisfies progress: if no thread is in the critical
+section then any thread can enter.
 
@@
 lock using an ordinary lock, never allowing
-more than one process in the critical section, and Harmony would not have
+more than one thread in the critical section, and Harmony would not have
 complained.
@@
 than one reader can enter the critical section at a time.  It does so by
-having processes sometimes wait in the critical section until all other readers
-are there.  If other readers cannot enter, then that process enters in an
+having threads sometimes wait in the critical section until all other readers
+are there.  If other readers cannot enter, then that thread enters in an
 infinite loop that Harmony will readily detect.  Try it out!
@@
 \item Why do you suppose the code in \autoref{fig:rwmulti} uses a flag
-per process rather than a simple counter?
+per thread rather than a simple counter?
 Can you write a version of this test that uses a counter instead
-of a flag per process?
+of a flag per thread?
 % \item Write a ``library implementation'' of reader/writer locks so multiple
 % reader/writer locks can be instantiated.
-\item Write a test program for ``bound locks'' (Exercise~\ref{ex:boundlock})
-that checks that up to \texttt{M} processes
-can acquire the bound lock at the same time.
 \end{problems}
@@
 If there are only $n$ units, no more than $n$ units can be used at a time;
-if all are in use and a process comes along that needs one of the units,
-it has to wait until another process releases one of the units.
+if all are in use and a thread comes along that needs one of the units,
+it has to wait until another thread releases one of the units.
 Note that we cannot solve this problem simply using a lock per unit---allocating
@@
 units of a resource initially available.
-When allocating a resource, a process decrements the
+When allocating a resource, a thread decrements the
 counter using the \texttt{P}
@@
 for \emph{Procure}, as in procuring the resource associated with the semaphore.
-The \texttt{P} operation blocks the invoking process if the counter is zero.
-To release the resource, a process increments the resource using the
+The \texttt{P} operation blocks the invoking thread if the counter is zero.
+To release the resource, a thread increments the resource using the
 \texttt{V}
@@
 
-\section*{Exercices}
+\section*{Exercises}
 \begin{problems}
@@
 the solution allows more than one Italian philosopher to eat at the same time?
-\item \label{ex:qsort}
-\autoref{fig:qsort} shows an iterative implementation of the Qsort
-algorithm, and \autoref{fig:qsorttest} an accompanying test program.
-The array to be sorted is stored in  shared variable $a$.
-Another shared variable, \textit{todo}, contains the ranges of the
-array that need to be sorted (initially the entire array).
-Re-using as much of this code as you can, implement a parallel version of
-this.  You should not have to change the methods \texttt{swap}, \texttt{partition},
-or \texttt{sortrange} for this.  Create \texttt{NWORKERS} ``worker processes''
-that should replace the \texttt{qsort} code.  Each worker loops until \textit{todo}
-is empty and sorts the ranges that it finds until then.  The \texttt{main}
-process needs to wait until all workers are done.  Use a semaphore for this:
-the \texttt{main} process should use \texttt{P} to wait for each
-worker to terminate.  After that it should check that the array is sorted and
-that it is a permutation of the original array.
 \end{problems}
-
-\begin{figure}
-\begin{code}
-\harmonysource{qsort}
-\end{code}
-\caption{\harmonylink{code/qsort.hny} Iterative qsort() implementation.}
-\label{fig:qsort}
-\end{figure}
-
-\begin{figure}
-\begin{code}
-\harmonysource{qsorttest}
-\end{code}
-\caption{\harmonylink{code/qsorttest.hny} Test program for \autoref{fig:qsort}.}
-\label{fig:qsorttest}
-\end{figure}
-
-\chapter{Bounded Buffer}
-\label{ch:bb}
+\end{comment}
+
+\section{Bounded Buffer}
 \index{bounded buffer}%
 \index{producer/consumer problem}%
-
 \glsadd{producer/consumer problem}%
 
+\begin{comment}
 \begin{figure}
@@
 \end{figure}
-
-A good example of a resource with multiple units is the
-so-called \emph{bounded buffer} (aka \emph{ring buffer}).
+\end{comment}
+
 \index{ring buffer}%
-Suppose you have a \emph{producer} (some process) wanting to
-communicate a stream of items with a \emph{consumer} (another process).
-A bounded buffer is essentially
-a queue implemented using a circular buffer
+A \emph{bounded buffer} (aka \emph{ring buffer}) is
+a queue with the usual enqueue/dequeue interface,
+but implemented using a circular buffer
 \index{circular buffer}%
 of a certain length and two pointers:
-one where the producer inserts items, and one where the consumer extracts
-items.  If the
-buffer is full, the producer must wait; if the buffer is empty, the
-consumer must wait.
+the \emph{tail} points where new items are enqueued and the \emph{head}
+points where items are dequeued.
+If the buffer is full, the enqueur must wait; if the buffer is empty, the
+dequeuer must wait.
 This problem is known as the ``Producer/Consumer Problem'' and was
@@
 
-The producer/consumer pattern is common.  Processes may be arranged
+The producer/consumer pattern is common.  Threads may be arranged
 in \emph{pipelines},
 \index{pipeline}%
-where each upstream process is a producer and each downstream
-process is a consumer.
-Or processes may be arranged in a manager/worker pattern, with a manager
+where each upstream thread is a producer and each downstream
+thread is a consumer.
+Or threads may be arranged in a manager/worker pattern, with a manager
 producing jobs and workers consuming and executing them in parallel.
@@
 \index{client/server model}%
-some process may act as a \emph{server} that clients can send requests to
+some thread may act as a \emph{server} that clients can send requests to
 and receive responses from.  In that case, there is a bounded buffer
@@
 consume responses, while the server consumes requests and produces responses.
-
-In general, we allow multiple producers and multiple consumers
-all sharing the same bounded buffer.
+%
+Multiple producers and multiple consumers may all share
+the same bounded buffer.
+
+\begin{comment}
 \autoref{fig:pc} gives a high-level description.  The
@@
 To add an item to the bounded buffer (\texttt{produce}(\textit{item})), the producing
-process first has to wait until there is room in the bounded buffer.
+thread first has to wait until there is room in the bounded buffer.
 To this end, it invokes \texttt{P}(\textit{n\_empty}), which waits until
@@
 to do the same thing concurrently.  Hence the use of the \textit{l\_in}
-semaphore used as a lock.  Finally, once the item is added, the process
+semaphore used as a lock.  Finally, once the item is added, the thread
 increments the \textit{n\_full} semaphore to signal to consumers that
@@
 \index{counting semaphore}%
-They can be used to send signals between processes.
+They can be used to send signals between threads.
 However, binary and counting semaphores are implemented the same way.
@@
 $\mathtt{NCONSS} \le \mathtt{NPRODS} \le \mathtt{NCONSS} + \mathtt{NSLOTS}$
-should terminate, while other values should lead to processes blocking.
+should terminate, while other values should lead to threads blocking.
 To this this, one can write a script like the one in \autoref{fig:pcscript}.
@@
 \textit{received} (atomically---a lock could have been used here as well).
-The \texttt{main()} process waits for all consumers to have received
+The \texttt{main()} thread waits for all consumers to have received
 exactly the set of tuples that the producers produce.
@@
 \begin{problems}
-\item \label{ex:nbproduce} Change the code in \autoref{fig:boundedbuffer} so that
-\texttt{produce()} fails if the buffer is full rather than block.
-\texttt{produce()} should return \texttt{False} if it fails
-and return \texttt{True} if it succeeds.
-\item Optimize the code in \autoref{fig:boundedbuffer} for the case
-where there is only one consumer and one producer.
+\end{problems}
+\end{comment}
+
+\chapter{Split Binary Semaphores}
+\label{ch:sbs}
+\index{split binary semaphore}%
+
+\begin{figure}
+\begin{code}
+\harmonysource{RWsbs}
+\end{code}
+\caption{\harmonylink{code/RWsbs.hny} Reader/Writer Lock using Split Binary Semaphores.}
+\label{fig:RWsplitsema}
+\end{figure}
+
+\glsadd{conditional critical section}%
+
+Split Binary Semaphores (SBS) is a general technique to implement conditional
+waiting.  It was originally proposed by
+Tony Hoare and popularized by Edsger Dijkstra~\cite{EWD703}.
+SBS is an extension of a critical section that is protected by a single
+binary semaphore.
+If there are $n$ \emph{waiting conditions},
+then SBS uses $n+1$ binary semaphores to protect the critical section.
+An ordinary critical section has no waiting conditions.
+Reader/writer locks have two waiting conditions:
+\begin{enumerate}
+\item readers waiting for a writer to release the lock;
+\item writers waiting for another writer or all readers to release the lock.
+\end{enumerate}
+Similarly, a bounded buffer has two waiting conditions:
+\begin{enumerate}
+\item consumers waiting for the buffer to be non-empty;
+\item producers waiting for an empty slot in the buffer.
+\end{enumerate}
+So each will require 3 binary semaphores if the SBS technique is applied.
+
+Think of each of these semaphores as a gate that a thread must go
+through in order to enter the critical section.  A gate is either open
+or closed.  Initially, exactly one gate, the main gate, is open.
+Each of the other gates, the \emph{waiting gates}, is associated with a
+waiting condition.
+When a gate is open, one thread can enter the critical section,
+closing the gate behind it.
+
+When leaving the critical section, the thread must open exactly one
+of the gates, but it does not have to be the gate that it used to enter
+the critical section.
+In particular, when a thread leaves the critical section it should
+check for each waiting gate if its waiting condition hold and if there are
+processes trying to get through the gate.  If there is such a gate,
+it must select one and open that gate.  If there is no such gate,
+it must open the main gate.
+
+Finally, if a thread is executing in the critical section and needs to
+wait for a particular condition, it leaves the critical section and waits
+for the gate associated with that condition to open.
+
+Note that the following invariants holds:
+\begin{itemize}
+\item At any time, at most one gate is open;
+\item If some gate is open, then no thread is in the critical section;
+\item If some thread is in the critical section, all gates are closed;
+\item At any time, at most one thread is in the critical section.
+\end{itemize}
+
+The main gate is implemented by a binary semaphore, initialized in the
+released state (signifying that the gate is open).
+The waiting gates each consist of a pair: a counter that counts how many
+threads are waiting behind the gate, and a binary semaphore initialized
+in the acquired state (signifying that the gate is closed).
+
+We will illustrate the technique using the reader/writer problem.
+\autoref{fig:RWsplitsema} shows code.
+The first step is to enumerate all waiting conditions.
+In the case of the reader/writer
+problem, there are two: a thread that wants to read may have to wait for a
+writer to leave the critical section, while a thread that wants to write may
+have to wait until all readers have left the critical section or until a writer has left.
+
+The state of a reader/writer lock thus consists of the following:
+\begin{itemize}
+\item \textit{nreaders}: the number of readers in the critical section;
+\item \textit{nwriters}: the number of writers in the critical section;
+\item \textit{mutex}: the ``main gate'' binary semaphore;
+\item \textit{r\_gate}: the ``waiting gate'' used by readers, consisting of a binary semaphore and the number of readers waiting to enter;
+\item \textit{w\_gate}: the ``waiting gate'' used by writers, similar to the readers' gate.
+\end{itemize}
+
+Each of the
+\texttt{read\_acquire}, \texttt{read\_release},
+\texttt{write\_acquire}, and \texttt{write\_release} methods must maintain
+this state.
+Each of these methods first has to acquire the \textit{mutex}
+(i.e., enter the main gate).
+\texttt{read\_acquire} and \texttt{write\_acquire} must check to see
+if it has to wait.
+If so, it increments the count associated with its respective gate,
+opens a gate (using \texttt{release\_one}, and then blocks until
+its waiting gate opens up.
+
+\texttt{release\_one()} is the function that a thread uses when leaving
+the critical section.  It must check to see if there is a waiting gate
+that has threads waiting behind it and whose condition is met.
+If so, it selects one and opens that gate.  In the given code,
+\texttt{release\_one()} first checks the readers' gate and then the
+writers' gate, but the other way around would have worked as well.
+If neither waiting gate qualifies, then \texttt{release\_one()} 
+has to open the main gate (i.e., release \textit{mutex}).
+
+Let us examine \texttt{read\_acquire} more carefully.
+First the method acquires the \textit{mutex}.
+Then, in the case that the thread, say $p$,
+finds that there is a writer in the critical section
+($\mathit{nwriters > 0}$), it increments the counter associated with
+the readers' gate, leaves the critical section (\texttt{release\_one}),
+and then tries to acquire the semaphore associated with the gate.
+This will cause it to block until some thread opens that gate.
+
+Now consider the case where there is a writer in the critical
+section and there are two readers waiting.  Let us see what happens when
+the writer calls \texttt{write\_release}:
+\begin{enumerate}
+\item After acquiring \textit{mutex}, the writer decrements
+\textit{nwriters}, which must be 1 at this time, and thus becomes 0.
+\item It then calls \texttt{release\_one()}.
+\texttt{release\_one()} finds that there are no writers in the critical section
+and there are two readers waiting.  It therefore releases not
+\textit{mutex} but the readers' gate's binary semaphore.
+\item One of the waiting reader can now re-enter the critical section.
+When it does, the reader decrements the gate's counter (from 2 to 1),
+and increments \textit{nreaders} (from~0 to~1).
+The reader finally calls \texttt{release\_one()}.
+\item Again, \texttt{release\_one()} finds that there are no writers and
+that there are readers waiting, so again it releases the readers' semaphore.
+\item The second reader can now enter the critical section.
+It decrements the gate's count from 1 to 0 and increments \textit{nreaders}
+from~1 to~2.
+\item Finally, the second reader  calls \texttt{release\_one()}.
+This time \texttt{release\_one()} does not find any threads waiting,
+and so it releases \textit{mutex}.
+There are now two threads that are holding the reader/writer lock.
+\end{enumerate}
+
+\section*{Exercises}
+\begin{problems}
+% \item Assume that you have only binary semaphores.
+% Implement counting semaphores using the split binary semaphores technique.
+\item Several of the instantiations of \texttt{release\_one()} in
+\autoref{fig:RWsplitsema} can be replaced by simply releasing \textit{mutex}.
+Which ones?
+\item Optimize your solutions to \autoref{ex:qcontains} to use reader/writer locks.
+\item \label{ex:bbsbs} Implement a solution to the producer/consumer problem
+using split binary semaphores.
+\item \label{ex:boundlock} Using busy waiting, implement a ``bound lock'' that allows
+up to \texttt{M} threads to acquire it at the same time.\footnote{A bound lock is a restricted version of a \emph{counting} semaphore.}
+A bound lock
+with \texttt{M == 1} is an ordinary lock.
+You should define a constant \texttt{M} and two methods:
+\texttt{acquire\_bound\_lock()}
+and \texttt{release\_bound\_lock()}.
+(Bound locks are useful for situations where too many threads working
+at the same time might exhaust some resource such as a cache.)
+\item Write a test program for your bound lock
+that checks that no more than \texttt{M} threads can acquire the
+bound lock.
+\item Write a test program for bound locks
+that checks that up to \texttt{M} threads
+can acquire the bound lock at the same time.
 \item Implement a \emph{bounded stack}.  Method \texttt{push}(\textit{item})
@@
 to block.
-\end{problems}
-
-\begin{figure}
-\begin{code}
-\harmonysource{gpu}
-\end{code}
-\caption{\harmonylink{code/gpu.hny} A thread-unsafe GPU allocator.}
-\label{fig:gpu}
-\end{figure}
-
-\chapter{Split Binary Semaphores}
-\label{ch:sbs}
-\index{split binary semaphore}%
-
-\begin{figure}
-\begin{code}
-\harmonysource{RWsbs}
-\end{code}
-\caption{\harmonylink{code/RWsbs.hny} Reader/Writer Lock using Split Binary Semaphores.}
-\label{fig:RWsplitsema}
-\end{figure}
-
-\glsadd{conditional critical section}%
-
-Reader/writer locks and bounded buffers are both examples of what are
-sometimes called \emph{conditional critical sections}
-\index{conditional critical section}%
-(CCSs)~\cite{Hoare73}.
-All critical sections are, in a way, conditional.  Even the basic critical section
-has the condition that a process can only enter if no other process
-is in the critical section.  Reader/writer locks refined that condition, loosening
-the restriction for improved performance.
-In the case of bounded buffers, there were two CCSs.  First, a CCS
-to add an entry to the bounded buffer that can only be entered by one
-process and only if there is room in the bounded buffer (\texttt{produce}).
-Second, a CCS that can only be entered by one process and
-only if there is an item in the bounded buffer (\texttt{consume}).
-
-CCSs are easy to implement using busy waiting:
-\begin{code}
-\harmonysource{busywait}
-\end{code}
-
-We have seen solutions to the reader/writer problem and the producer/consumer
-problem that do not busy-wait, but they are quite different.  It would be nice
-to have a general technique for all CCS problems.
-The ``Split Binary Semaphore'' (SBS) is such a technique, originally proposed by
-Tony Hoare and popularized by Edsger Dijkstra~\cite{EWD703}.
-SBS is an example of what one could call a ``baton-passing'' technique.
-A process needs the baton to access shared variables.
-When a process has the baton and does not need it any more,
-it first checks to see if there are any processes that are waiting for the
-baton and can make progress.
-If so, it passes the baton directly to one of those processes.
-If not, it releases the baton for some other process to pick up at a
-later time.
-
-We will illustrate the technique using the reader/writer problem.
-\autoref{fig:RWsplitsema} shows the
-\texttt{rlock\_acquire}, \texttt{rlock\_release},
-\texttt{wlock\_acquire}, and \texttt{wlock\_release} methods implemented using
-SBS.
-
-Step 1 is to enumerate all waiting conditions.  In the case of the reader/writer
-problem, there are two: a process that wants to read may have to wait for a
-writer to leave the critical region, while a process that wants to write may
-have to wait until all readers have left the critical section or until a writer has left.  If there are $N$
-waiting conditions, the SBS technique will use $N+1$ binary semaphores: one for
-each waiting condition and an extra one that processes use when they first try
-to enter the critical section and do not yet know if they need to wait.
-We shall call this last semaphore the \textit{mutex}.
-
-An important invariant of the SBS technique is that the sum of the $N+1$ semaphores
-must always be 0 or 1, and it should be 0 when a process is accessing the shared
-variables and 1 when not.
-As it were, these semaphores taken together are emulating a single binary semaphore.
-Initially, \textit{mutex} is~1 and the other $N$ semaphores
-are all~0.  To maintain the invariant, each process alternates \texttt{P} and
-\texttt{V} operations, starting with \texttt{P}(?\textit{mutex}) when it tries to
-enter the CCS, and ending with a \texttt{V} operation on one of the semaphores
-(yet to be determined).
-
-Another important part of the SBS technique is to keep careful track of the
-state.  In the case of the reader/writer lock, the state consists of the
-following shared variables:
-\begin{itemize}
-\item \textit{r\_entered}: the number of readers in the critical section;
-\item \textit{w\_entered}: the number of writers in the critical section;
-\item \textit{r\_waiting}: the number of readers waiting to enter;
-\item \textit{w\_waiting}: the number of writers waiting to enter.
-\end{itemize}
-Each of the
-\texttt{rlock\_acquire}, \texttt{rlock\_release},
-\texttt{wlock\_acquire}, and \texttt{wlock\_release} methods must maintain
-this state.
-For example, check out \texttt{rlock\_acquire}.
-It first checks to see if there
-are any writers.  If so, it increments \textit{r\_waiting}.
-If not, it increments \textit{r\_entered}.
-
-Either way, it vacates one of the semaphores next.
-\texttt{V\_one()} is the baton passing function that selects which of the
-binary semaphores to vacate.
-Method \texttt{V\_one()} first checks to see if there are any readers or
-writers waiting.  If there are readers waiting and there are no writers
-in the critical section, it vacates the semaphore associated with
-readers waiting, passing the baton to the next reader.
-If there are writers waiting and there are no readers
-nor writers in the critical section, it vacates the semaphore associated
-with writers waiting, passing the baton to the next writer.
-If both conditions hold, it could vacate either
-one---it would not matter.
-However, it can only vacate one of the semaphores.
-In this implementation, it prefers to pass the baton to another reader.
-If none of the conditions hold, \texttt{V\_one()} vacates \textit{mutex},
-leaving the batom for a future process to pick up.
-
-Let us return now to \texttt{rlock\_acquire}. In the case that the process,
-say $p$,
-found there is a writer in the critical section, it vacates
-\textit{mutex} and procures \textit{r\_sema},
-blocking on the semaphore associated
-with waiting readers.  When later some other process passes the baton to
-$p$ by incrementing \textit{r\_sema}, $p$ updates the state by decrementing
-\textit{r\_waiting} and incrementing \textit{r\_entered}.  $p$
-then invokes \texttt{V\_one()} to pass the baton to another waiting
-reader, if any, or to leave it for another process to pick up.
-
-As an example, consider the case where there is a writer in the critical
-section and there are two readers waiting.  Let us see what happens when
-the writer calls \texttt{wlock\_release}:
-\begin{enumerate}
-\item After procuring \textit{mutex},
-the sum of the three semaphores is 0.  The writer then decrements
-\textit{w\_entered} and calls \texttt{V\_one()}.
-\item \texttt{V\_one()} finds that there are no writers in the critical section
-and there are two readers waiting.  It therefore vacates \textit{r\_sema}.
-The sum of the semaphores is now~1, because \textit{r\_sema} is~1.  This
-guarantees that only a waiting reader, procuring \textit{r\_sema}, can enter
-the critical section.
-\item When it does, the reader decrements \textit{r\_waiting}
-from 2 to 1, and increments \textit{r\_entered} from~0 to~1.
-The reader finally calls \texttt{V\_one()}.
-\item Again, \texttt{V\_one()} finds that there are no writers and
-that there are readers waiting, so again it vacates \textit{r\_sema},
-passing the baton to the one remaining waiting reader.
-\item The remaining reader decrements \textit{r\_waiting} from 1 to 0 and
-increments \textit{r\_entered} from~1 to~2.
-\item Finally, the remaining reader  calls \texttt{V\_one()}.
-\texttt{V\_one()} does not find any processes waiting,
-and so it vacates \textit{mutex}.
-\end{enumerate}
-
-\section*{Exercises}
-\begin{problems}
-\item Assume that you have only binary semaphores.
-Implement counting semaphores using the split binary semaphores technique.
 \item With reader/writer locks,
-concurrency can be improved if a process \emph{downgrades} its write lock
+concurrency can be improved if a thread \emph{downgrades} its write lock
 to a read lock when its done writing but not done reading.  Add
 a \texttt{downgrade} method to the code in \autoref{fig:RWsplitsema}.
-How easy or hard would it be to do the same for the code in
-\autoref{fig:rw2lock}?
-Note, you cannot simply release the
-write lock and then obtain the read lock, because the process needs
-to remain in the critical section.
-\item \label{ex:onelane} Implement a solution to the producer/consumer problem
-(\autoref{fig:boundedbuffer}) using split binary semaphores.
+% How easy or hard would it be to do the same for the code in
+% \autoref{fig:rw2lock}?
+% Note, you cannot simply release the
+% write lock and then obtain the read lock, because the thread needs
+% to remain in the critical section.
 \item \label{ex:onelane} Cornell's campus features some one-lane bridges.
@@
 same direction.
-Implement a lock that observes this requirement using SBS.  Write methods
+Implement a lock that observes this requirement using SBS.
+Write methods \texttt{OLBlock()} to create a new ``one lane bridge'' lock,
 \texttt{nb\_enter()} that a car must invoke before going northbound on
@@
 \item Extend the solution to \autoref{ex:onelane} by implementing the
-requirement that at most 3 cars are allowed on the bridge.
+requirement that at most $n$ cars are allowed on the bridge.  Add $n$
+as an argument to \texttt{OLBlock}.
 
 \end{problems}
+
+\begin{figure}
+\begin{code}
+\harmonysource{gpu}
+\end{code}
+\caption{\harmonylink{code/gpu.hny} A thread-unsafe GPU allocator.}
+\label{fig:gpu}
+\end{figure}
 
@@
 \emph{safety property}---it prevents something ``bad'' from
-happening, like a reader and writer process both entering the
+happening, like a reader and writer thread both entering the
 critical section.  The \emph{progress} property is an example
@@
 eventually happens.
-Informally (and inexactly), progress states that if no processes
-are in the critical section, then some process that wants to enter
+Informally (and inexactly), progress states that if no threads
+are in the critical section, then some thread that wants to enter
 can.
@@
 Progress is a weak form of liveness.  It says that \emph{some}
-process can enter, but it does not prevent a scenario such as
-the following.  There are three processes repeatedly trying to
+thread can enter, but it does not prevent a scenario such as
+the following.  There are three threads repeatedly trying to
 enter a critical section using a spinlock.  Two of
-the processes successfully keep entering, alternating, but the third
-process never gets a turn.  This is an example of
+the threads successfully keep entering, alternating, but the third
+thread never gets a turn.  This is an example of
 \texttt{starvation}.  With a spinlock, this scenario could
-even happen with two processes.  Initially both processes
-try to acquire the spinlock.  One of the processes is
-successful and enters.  After the process leaves, it immediately
+even happen with two threads.  Initially both threads
+try to acquire the spinlock.  One of the threads is
+successful and enters.  After the thread leaves, it immediately
 tries to re-enter.  This state is identical to the initial
-state, and there is nothing that prevents the same process
+state, and there is nothing that prevents the same thread
 from acquiring the lock yet again.
@@
 does not suffer from starvation, thanks to the \texttt{turn} variable
-that alternates between 0 and 1 when two processes are contending for
+that alternates between 0 and 1 when two threads are contending for
 the critical section.
@@
 While spinlocks suffer from starvation, it is a uniform random
-process and each process has an equal chance of entering the critical
+process and each thread has an equal chance of entering the critical
 section.  Thus the probability of starvation is exponentially vanishing.
@@
 Unfortunately, such is not the case for the
-reader/writer solutions that we presented thus far.
+reader/writer solution that we presented in~\autoref{ch:sbs}.
 Consider this scenario: there are two readers and one writer.  One reader
@@
 
-SBSs allow much control over which thread runs next and is therefore a
-good starting point for developing fair synchronization algorithms.
-In this chapter, we will present two SBS-based fair reader/writer lock
-implementations.
-The first, in \autoref{fig:RWfair}, is based on \autoref{fig:RWsplitsema}.
-There are two differences:
+SBSs allow much control over which type of thread runs next and is therefore
+a good starting point for developing fair synchronization algorithms.
+\autoref{fig:RWfair} is based on \autoref{fig:RWsplitsema}, but there
+are two important differences:
 
@@
 waiting to enter the critical section;
-\item There are two versions of \texttt{V\_one()}.  In particular, when the
-last reader leaves the critical section, it preferentially passes the baton
-to a waiting writer rather than to a waiting reader.  Similarly,
-when a writer leaves the critical section, it preferentially passes
-the baton to a waiting reader rather than to a waiting writer.
+\item Instead of a one-size-fits-all \texttt{release\_one} method, each
+method has a custom way of selecting which gate to open.  In particular,
+\texttt{read\_release} prefers the write gate, while \texttt{write\_release}
+prefers the read gate.
 \end{enumerate}
@@
 The net effect of this is that if there is contention between readers and
-writers, than readers and writers end up alternating entering the critical
+writers, then readers and writers end up alternating entering the critical
 section.  While readers can still starve other readers and writers can still
-starve other writers, they cannot starve one another.  And since contention for
-a semaphore is resolved probabilistically, each reader and each writer will
-eventually have a chance to enter the critical section almost surely.
-
+starve other writers, readers can no longer starve writers nor vice versa.
+Other fairness is based on the fairness of semaphores themselves.
+
+\begin{comment}
 \autoref{fig:RWqueue} and \autoref{fig:RWqtest} present a different approach
-in which processes are allowed into the critical region in a purely First
+in which threads are allowed into the critical section in a purely First
 Come First Served manner.  When a reader arrives, it is allowed in only if
@@
 critical section (and therefore also not waiting to enter).  To maintain the
-arrival order, the solution uses a semaphore per process and a queue.
+arrival order, the solution uses a semaphore per thread and a queue.
 The \texttt{acquire\_} methods take a pointer to
-the semaphore of the process as argument.
+the semaphore of the thread as argument.
 Each entry in the queue is a pair consisting of a type of access
 (\texttt{.read} or \texttt{.write}) and a pointer to the semaphore of the
-processs that is waiting.  Method \texttt{V\_one} considers semaphores to vacate
-in the order in which they appear in the queue.
+thread that is waiting.  Method \texttt{release\_one} considers semaphores
+to release in the order in which they appear in the queue.
 
@@
 \end{figure}
+\end{comment}
 
@@
 \begin{code}
-\harmonysource{hoaretest}
-\end{code}
-\caption{\harmonylink{code/hoaretest.hny} Example of Hoare monitors.}
+\harmonysource{BBhoare}
+\end{code}
+\caption{\harmonylink{code/BBhoare.hny} Bounded Buffer implemented using
+a Hoare monitor.}
 \label{fig:hoaretest}
@@
 \index{condition variable}%
-Each condition variable also keeps track of the number of processes waiting
+Each condition variable also keeps track of the number of threads waiting
 on the condition variable.
@@
 and exit monitor methods.
-\autoref{fig:hoaretest} presents a simple example lifted directly from
-the original paper~\cite{Hoare74}.
-
 \texttt{wait} increments the condition's counter, releases the monitor mutex,
-blocks while trying to procure the condition's semaphore, and after resuming
+blocks while trying to acquire the condition's semaphore, and after resuming
 decrements the counter---in much the same way as we have seen for split binary
@@
 \texttt{signal} checks to see if the condition's count is non-zero, and if so
-vacates the condition's semaphore---passing the baton to
-a process trying to procure it---and then blocks by procuring the mutex again.
-The use of \texttt{wait} and \texttt{signal} ensures that \texttt{P} and
-\texttt{V} operations alternate as prescribed by the SBS technique.
-Our implementations of reader/writer locks can be easily changed to
-use \texttt{wait} and \texttt{signal} instead of using \texttt{P}
-and \texttt{V}.
+releases the condition's semaphore,
+and then blocks by trying to acquire the mutex again.
+
+\autoref{fig:hoaretest} presents a bounded buffer implemented using
+Hoare monitors.  It is written in much the same way you would if using
+the SBS technique (see \autoref{ex:bbsbs}).
+There is no \texttt{release\_one} method.  Instead, one can conclude
+that \texttt{enqueue} causes the queue to be non-empty, and
+\texttt{signal} will check if there are any threads waiting for
+this event.  If so, \texttt{signal} will pass control to one such thread
+and, unlike \texttt{release\_one}, re-enter the critical
+section afterwards by acquiring the \textit{mutex}.
+
+Implementing a reader/writer lock with Hoare monitors is not quite so
+straightforward, unfortunately.  When a writer releases the lock, it
+has to choose whether to signal a reader or another writer.  For that
+it needs to know if there is a reader or writer waiting.  The simplest
+solution would be to peek at the counters inside the respective condition
+variable, but that breaks the abstraction.  The alternative is for the
+reader/writer implementation to keep track of that state explicitly,
+which complicates the code.  Also, it requires a deep understanding of
+the SBS method to remember to place a call \texttt{signal} in the
+\texttt{read\_acquire} method that releases additional readers that
+may be waiting to acquire the lock.
 
@@
 However, there are some subtle but important differences with Hoare
-monitors that make Mesa monitors quite unlike split binary semaphores.
+monitors that make Mesa monitors quite unlike split binary semaphores
+and mostly easier to use in practice.
 
@@
 upon exit.
-Mesa monitors also have condition variables that a process can wait on.
+Mesa monitors also have condition variables that a thread can wait on.
 Like in Hoare monitors, the \texttt{wait} operation releases the mutex.
@@
 \index{notify}%
-When a process $p$ invokes \texttt{notify}, it does not immediately pass 
-the baton to some process $q$ requiring the baton (assuming there even
-is such a process).  Instead, $p$ holds onto
-the baton until it leaves the monitor.  At that point, any process that
-was notified will have a chance to obtain the baton, but they compete
-with other processes also trying to obtain the baton.
-
+When a thread $p$ invokes \texttt{notify}, it does not immediately pass 
+control to a thread that is waiting on the corresponding condition (if there
+is such a thread).  Instead, $p$ remains in the critical section
+until it leaves the monitor explicitly.  At that point, any thread that
+was notified will have a chance to enter the critical section, but they compete
+with other threads trying to enter the critical section.
+Basically, there is just one gate to enter the critical section, instead
+of a main gate and a gate per waiting condition.
+
+\begin{comment}
 \begin{figure}
@@
 \caption{High-level depictions of Hoare and Mesa monitors.  The hall is
-where processes wait to enter the bathroom (the critical section).  The
-bedrooms illustrate two condition variables.  The circles are processes.}
+where threads wait to enter the bathroom (the critical section).  The
+bedrooms illustrate two condition variables.  The circles are threads.}
 \label{fig:monitors}
@@
 Here a bathroom represents the critical section, allowing only one
-process at a time.  The bedrooms represent condition variables.
+thread at a time.  The bedrooms represent condition variables.
 There is some condition associated with each bedroom.
-In the hall are processes waiting to enter the critical section.
+In the hall are threads waiting to enter the critical section.
 
 On the left is a Hoare monitor.
-When some process $p$ is in the bathroom, it can invoke \textbf{signal}
+When some thread $p$ is in the bathroom, it can invoke \textbf{signal}
 on some non-empty bedroom associated with a condition that holds.
-This select a process $q$ in that bedroom.
+This select a thread $q$ in that bedroom.
 As a result of invoking \textbf{signal}, $p$ goes back into the hall
@@
 a \textbf{continue} operation instead of \textbf{signal}.  The difference
-is that processes that invoke \textbf{continue} leave the monitor rather
+is that threads that invoke \textbf{continue} leave the monitor rather
 than returning to the hall.}
@@
 On the right is a Mesa monitor.  The only way into the bathroom is through
-the hall.  When process $p$ invokes \textbf{notify} on a bedroom, it
-selects a process $q$ in that bedroom, if any.
-Process $q$ goes into the hall,
-joining any other processes that are also waiting to
+the hall.  When thread $p$ invokes \textbf{notify} on a bedroom, it
+selects a thread $q$ in that bedroom, if any.
+Thread $q$ goes into the hall,
+joining any other threads that are also waiting to
 enter the bathroom.
-If there are no processes in that bedroom, \textbf{notify} is a no-op.
+If there are no threads in that bedroom, \textbf{notify} is a no-op.
 Importantly, in either case $p$ remains in the bathroom.
 Finally, when $p$ leaves the bathroom, one of the
-processes in the hall can enter the bathroom.  But that may not be $q$.
-Assuming every process eventually leaves the bathroom and the system is
+threads in the hall can enter the bathroom.  But that may not be $q$.
+Assuming every thread eventually leaves the bathroom and the system is
 fair, eventually $q$ will enter the bathroom.  However, it is no longer
@@
 into the bedroom corresponding to the condition.
-
-Besides notifying processes, Mesa monitors provide another important option:
-it can choose to let any number of processes in the bedrooms into the hall.
-It can do so by calling
-\texttt{notify} repeatedly, each time letting one process in the specified
-bedroom into the hall.  Mesa also provides an
-operation called \texttt{notifyAll}
+\end{comment}
+
+This is a very important difference.  In Hoare monitors, when a thread
+enters through a waiting gate, it can assume that the condition associated
+with the waiting gate still holds because no other thread can run in between.
+Not so with Mesa monitors: by the time a thread that was notified enters
+through the main gate, other threads may have entered first and falsified
+the condition.  So in Mesa, threads always have to check the condition
+again after resuming from the \texttt{wait} operation.  This is accomplished
+by wrapping each \texttt{wait} operation in a \textbf{while} statement that
+loops until the condition of interest becomes valid.
+A Mesa monitor therefore is more closely related to \emph{busy waiting}
+than to split binary semaphores.
+
+Mesa monitors provide another important and convenient option:
+operation \texttt{notifyAll}
+notifies \emph{all} threads that are waiting on a condition
+instead of just one.
+Having such an operation is not possible with Hoare monitors because using
+Hoare monitors control
+must be passed immediately to a thread that has been signaled, and that
+can only be done if there is just one such thread.
 \index{notifyAll}%
@@
 \index{broadcast}%
-that, when applied to
-a bedroom, lets all processes in that bedroom into the hall.
-This is not possible with Hoare monitors: when needing to wake up a set
-of processes, only one can be signaled at a time.  Each of those processes
-then has to signal the next in a baton-passing style
-until all processes have been signaled.
-
-While Hoare monitors are conceptually simpler, the so-called
+lets all threads in that bedroom into the hall.
+
+The so-called
 ``Mesa monitor semantics'' or ``Mesa condition variable'' semantics
-have become more popular, adopted by all major
-programming languages.
+have become more popular than Hoare monitor semantics, and have been
+adopted by all major programming languages.
 That said, few programming languages provide full support for monitors.
@@
 
-The Harmony \texttt{synch} module supports Mesa condition variables.
-However, like C, Java, and Python,
-it does not have built-in language support for them but instead associates condition variables
-with explicit locks.
+Like C, Java, and Python, Harmony does not have built-in language
+support for monitors, but provides library support through the
+Harmony \texttt{synch} module.
 \autoref{fig:cv} shows the implementation of
 condition variables in the \texttt{synch} module.
-\texttt{Condition}(\textit{lk}) creates a new condition variable associated with the given pointer
-to a lock \textit{lk}.
-The condition variable itself is represented by a dictionary containing the pointer to the lock
-and a bag of name tags of processes waiting on the condition variable.
+\texttt{Condition}() creates a new condition variable.
+It is represented by a dictionary containing
+a bag of name tags of threads waiting on the condition variable.
 (The \texttt{synchS} library instead uses a list of contexts.)
 
-\texttt{wait} adds the nametag of the process to the bag.
-This increments the number of processes in the bag with the same context.
+\texttt{wait} adds the nametag of the thread to the bag.
+This increments the number of threads in the bag with the same context.
 \texttt{wait} then waits until that count is restored to the
-value that it had upon entry to \texttt{wait}.  \texttt{notify} removes an arbitrary context from
-the bag, allowing one of the processes with that context to return from \texttt{wait}.
-\texttt{notifyAll} empties out the entire bag, allowing all processes in the bag to resume.
-
-To illustrate how Mesa condition variables are used in practice, we demonstrate using,
-once again, an implementation of reader/writer locks.
-\autoref{fig:RWcv} shows the code.  \textit{rwlock} is the global lock or mutex for the
-reader/writer lock.  There are two condition variables: readers wait on \textit{rcond} and
-writers wait on \textit{wcond}.  The implementation also keeps track of the number of
+value that it had upon entry to \texttt{wait}.
+\texttt{notify} removes an arbitrary context from
+the bag, allowing one of the threads with that context to
+resume and re-acquire the lock associated with the monitor.
+\texttt{notifyAll} empties out the entire bag, allowing all threads
+in the bag to resume.
+
+To illustrate how Mesa condition variables are used in practice, we demonstrate using an implementation of reader/writer locks.
+\autoref{fig:RWcv} shows the code.  \textit{mutex} is the shared lock
+that protects the critical region.
+(Again, there is only one ``gate.'')
+There are two condition variables: readers wait on \textit{r\_cond} and
+writers wait on \textit{w\_cond}.
+The implementation also keeps track of the number of
 readers and writers in the critical section.
 
-When inspecting the code, one important feature to notice is that \texttt{wait} is always
-invoked within a \textbf{while} loop that checks for the condition that the process is
-waiting for.  We explained why above: when a process resumes after
-being notified, it is not guaranteed that the condition it was waiting for holds, even
-if it did at the time of invoking \texttt{notify} or \texttt{notifyAll}.
-It is \emph{imperative} that there is always a \textbf{while} loop around any invocation
-of \texttt{wait} containing the negation of the condition that the process is waiting for.
-Many implementation of condition variables depend on this, and optimized implementations
-of condition variables often allow so-called ``spurious wakeups,'' where \texttt{wait}
-may sometimes return even if the conditon variable has not been notified.
-
-In \texttt{release\_rlock}, notice that \texttt{notify}(?\textit{wcond}) is invoked when there
-are no readers left, \emph{without} checking if there are writers waiting to enter.
-With Mesa monitors this is ok, because calling \texttt{notify} is a no-op if no process
-is waiting.
-
-\texttt{release\_wlock} executes \texttt{notifyAll}(?\textit{rcond}) as well as
-\texttt{notify}(?\textit{wcond)}.
-Again, because we do not keep track of the number of waiting readers or writers, we
-have to conservatively assume that all waiting readers can enter, or, alternatively,
-up to one waiting writer can enter.  So \texttt{release\_wlock} wakes up all
-potential candidates.
-There are two things to note here.  First, unlike split binary semaphores or Hoare
-monitors, where multiple waiting readers would have to be signaled one at a time in a
-baton-passing fashion (see \autoref{fig:RWsplitsema}), with Mesa monitors
-all readers are awaked in one fell swoop using \texttt{notifyAll}.
-Second, both readers and writers are awakened---this is ok because both execute
-\texttt{wait} within a \textbf{while} loop, re-checking the condition that they
-are waiting for.  So if both type of processes are waiting, either all the readers
-get to enter next or one of the writers gets to enter next.
-
-Much of the complexity of programming with Mesa condition variables is in figuring out
-when to invoke \texttt{notify} and when to invoke \texttt{notifyAll}.  As a rule of
-thumb: be conservative---it is better to wake up too many processes than too few.
-Waking up too many processes may lead to some inefficiency, but waking up too few
-may cause the application to get stuck.  Harmony can be particularly helpful here, as
-it examines every corner case.  You can safely try to replace each \texttt{notifyAll}
-with \texttt{notify} and see if every possible execution of the the application still
-terminates.
+Note that \texttt{wait} is always invoked within a \textbf{while}
+loop that checks for the condition that the thread is waiting for.
+It is \emph{imperative} that there is always a \textbf{while} loop
+around any invocation of \texttt{wait} containing the negation of
+the condition that the thread is waiting for.  Many implementation
+of condition variables depend on this, and optimized implementations
+of condition variables often allow so-called ``spurious wakeups,''
+where \texttt{wait} may sometimes return even if the conditon
+variable has not been notified.
+In particular, one should always be able to replace \texttt{wait}
+by \texttt{unlock} followed by \texttt{lock}.
+This turns the solution into a busy-waiting one, inefficient but still
+correct.
+
+In \texttt{release\_rlock}, notice that \texttt{notify}(?\textit{w\_cond})
+is invoked when there are no readers left, \emph{without} checking
+if there are writers waiting to enter.  This is
+ok, because calling \texttt{notify} is a no-op if no thread is
+waiting.
+
+\texttt{release\_wlock} executes \texttt{notifyAll}(?\textit{r\_cond})
+as well as \texttt{notify}(?\textit{w\_cond)}.  Again, because we
+do not keep track of the number of waiting readers or writers, we
+have to conservatively assume that all waiting readers can enter,
+or, alternatively, up to one waiting writer can enter.  So
+\texttt{release\_wlock} wakes up all potential candidates.  There
+are two things to note here.  First, unlike split binary semaphores
+or Hoare monitors, where multiple waiting readers would have to be
+signaled one at a time in a baton-passing fashion (see
+\autoref{fig:RWsplitsema}), with Mesa monitors all readers are
+awakened in one fell swoop using \texttt{notifyAll}.  Second, both
+readers and writers are awakened---this is ok because both execute
+\texttt{wait} within a \textbf{while} loop, re-checking the condition
+that they are waiting for.  So if both type of threads are waiting,
+either all the readers get to enter next or one of the writers gets
+to enter next.
 
@@
 Mesa monitors are much closer to the busy-waiting approach.
-When comparing the busy-waiting version of reader/writer locks in \autoref{fig:rwbusy}
-and the Mesa version in \autoref{fig:RWcv}, you will see important similarities.
-Busy-waiting is ``fixed'' by placing a \textbf{wait} statement in the waiting loop.
-The Mesa code then has to be careful to invoke \textbf{notify} or \textbf{notifyAll}
-in the right places.
+The Mesa code then has to be careful to invoke \textbf{notify} or
+\textbf{notifyAll} in the right places.
+Much of the complexity of programming with Mesa condition variables
+is in figuring out when to invoke \texttt{notify} and when to invoke
+\texttt{notifyAll}.  As a rule of thumb: be conservative---it is
+better to wake up too many threads than too few.  Waking up too
+many threads may lead to some inefficiency, but waking up too few
+may cause the application to get stuck.  Harmony can be particularly
+helpful here, as it examines every corner case.  You can safely try
+to replace each \texttt{notifyAll} with \texttt{notify} and see if
+every possible execution of the the application still terminates.
 
@@
 \begin{problems}
-\item Implement a ``try lock'' using Mesa condition variables
+\item Implement a ``try lock'' module using Mesa condition variables
 (see also Exercise~\ref{ex:trylock}).  It should
@@
 \begin{enumerate}
-\item \textit{tl} = \texttt{TryLock}() \# \emph{allocate a try lock}
-\item \texttt{tlLock}(?\textit{tl}) \# \emph{acquire a try lock}
-\item \texttt{tlTrylock}(?\textit{tl}) \# \emph{attempt to acquire a try lock}
-\item \texttt{tlUnlock}(?\textit{tl}) \# \emph{release a try lock}
+\item \textit{tl} = \texttt{TryLock}() \# \emph{create a try lock}
+\item \texttt{acquire}(?\textit{tl}) \# \emph{acquire a try lock}
+\item \texttt{tryAcquire}(?\textit{tl}) \# \emph{attempt to acquire a try lock}
+\item \texttt{release}(?\textit{tl}) \# \emph{release a try lock}
 \end{enumerate}
 \noindent
-\texttt{tlTrylock} should not wait.
+\texttt{tryAcquire} should not wait.
 Instead it should return \texttt{True} if the lock was successfully
 acquired and \texttt{False} if the lock was not available.
-\item Implement semaphores using Mesa condition variables.
+% \item Implement semaphores using Mesa condition variables.
 \item Write a new version of the GPU allocator in \autoref{ex:gpu}
@@
 In this version,
-a process is allowed to allocate a set of GPUs and release a set of GPUs that it
+a thread is allowed to allocate a set of GPUs and release a set of GPUs that it
 has allocated.  Method \texttt{gpuAllocSet(n)} should block until $n$ GPUs are
@@
 Method \texttt{gpuReleaseSet(s)} takes a set of GPU identifiers as argument.
-A process does not have to return all the GPUs it allocated at once.
+A thread does not have to return all the GPUs it allocated at once.
 \item The specification in the previous question makes the solution unfair.
@@
 
-When multiple processes are synchronizing access to shared resources, they
-may end up in a \emph{deadlock} situation where one or more of the processes
+When multiple threads are synchronizing access to shared resources, they
+may end up in a \emph{deadlock} situation where one or more of the threads
 end up being blocked indefinitely because each is waiting for another to give
@@
 \autoref{fig:diners} implements the dining philosophers in Harmony, using a
-process for each philosopher and a lock for each fork.  If you
+thread for each philosopher and a lock for each fork.  If you
 run it, Harmony complains that the execution may not terminate, with all five
-processes being blocked trying to acquire the lock.
+threads being blocked trying to acquire the lock.
 
@@
 \begin{enumerate}
-\item \emph{Mutual Exclusion}: each resource can only be used by one process at a time:
-\item \emph{Hold and Wait}: each process holds resources it already allocated while it
+\item \emph{Mutual Exclusion}: each resource can only be used by one thread at a time:
+\item \emph{Hold and Wait}: each thread holds resources it already allocated while it
 waits for other resources that it needs;
-\item \emph{No Preemption}: Resources cannot be forcibly taken away from processes that
+\item \emph{No Preemption}: Resources cannot be forcibly taken away from threads that
 allocated them;
-\item \emph{Circular Wait}: There exists a directed circular chain of processes, each waiting
+\item \emph{Circular Wait}: There exists a directed circular chain of threads, each waiting
 to allocate a resource held by the next.
@@
 \begin{itemize}
-\item \emph{No Hold and Wait}: a process must request all resources it is going to
+\item \emph{No Hold and Wait}: a thread must request all resources it is going to
 need at the same time;
-\item \emph{Preemption}: if a process is denied a request for a resource, it must
+\item \emph{Preemption}: if a thread is denied a request for a resource, it must
 release all resources that it has already acquired and start over;
@@
 
-The \emph{Preemption} approach suggested by Havender is to allow processes to back out.
+The \emph{Preemption} approach suggested by Havender is to allow threads to back out.
 While this could be done, this invariably leads to a busy waiting solution
-where a process keeps obtaining locks and releasing them again until it
+where a thread keeps obtaining locks and releasing them again until it
 finally is able to get all of them.
@@
 is to prevent a cycle from forming, with each
-process waiting for the next process on the cycle.
+thread waiting for the next thread on the cycle.
 We can do this by establishing an ordering among the
@@
 The problem with these kinds of schemes is that one needs to know ahead of time
-the set of processes and what the maximum number of resources is that each process
+the set of threads and what the maximum number of resources is that each thread
 wants to allocate, making them generally quite impractical.
@@
 Unfortunately, running the test reveals that it sometimes leaves unterminated
-processes.  Can you fix the problem?
+threads.  Can you fix the problem?
 \item Add a method \texttt{total()} to the solution of the previous question
@@
 it cannot cause deadlock.
-\item Implement an extra process that checks, once, if the total of
+\item Implement an extra thread that checks, once, if the total of
 the balances is the same as that at the beginning (using an assertion).
@@
 synchronization using so-called \emph{actors}~\cite{HBS73}.  Actors are
-processes that have only private memory and communicate through
+threads that have only private memory and communicate through
 \emph{message passing}.
@@
 queues, which have built-in synchronization), there is no need
-for critical sections.  Instead, some sequential process owns a particular
-piece of data and other processes access it by sending request messages
-to the process and optionally waiting for response messages.  Each process
+for critical sections.  Instead, some sequential thread owns a particular
+piece of data and other threads access it by sending request messages
+to the thread and optionally waiting for response messages.  Each thread
 handles one message at a time, serializing all access to the data it owns.
@@
 Note that both \textit{balance} and \textit{requests} are variables that are
-local to the \texttt{pc\_actor} process.
+local to the \texttt{pc\_actor} thread.
 
@@
 
-The \texttt{pc\_actor} process has two arguments.  $q$ is a pointer
+The \texttt{pc\_actor} thread has two arguments.  $q$ is a pointer
 to the queue on which it receives messages.  \textit{nrequests} is the total
 number of requests that it will handle.  We need the latter because in Harmony
-all processes are required to terminate.  In a more realistic implementation,
+all threads are required to terminate.  In a more realistic implementation,
 the \texttt{pc\_actor} would be in an infinite loop awaiting requests.
@@
 \autoref{fig:actortest} shows how this code may be used.  It spawns
-\texttt{NITEMS} producer and consumer processes.  Note that in that
+\texttt{NITEMS} producer and consumer threads.  Note that in that
 case the \texttt{pc\_actor} is expected to receive \texttt{2 * NITEMS}
@@
 
-Processes can be \emph{interrupted}.  An interrupt is a notification of some event
+Threads can be \emph{interrupted}.  An interrupt is a notification of some event
 such as a keystroke, a timer expiring, the reception of a network packet, the completion
 of a disk operation, and so on.  We distinguish \emph{interrupts} and \emph{exceptions}.
-An exception is caused by the process executing an invalid machine instruction such as
+An exception is caused by the thread executing an invalid machine instruction such as
 divide-by-zero.  An interrupt is caused by some peripheral device and can be handled in
@@
 \autoref{fig:trap} gives an example of how \textbf{trap} might be used.
-Here, the \textit{main}() process loops until the interrupt has occurred and
+Here, the \textit{main}() thread loops until the interrupt has occurred and
 the \textit{done} flag has been set.
@@
 This is not unlike the example we gave in \autoref{fig:inc}, except that only a single
-process is involved now.  And, indeed, it suffers from a similar race condition; run
+thread is involved now.  And, indeed, it suffers from a similar race condition; run
 it through Harmony to see for yourself.  If the interrupt occurs after \textit{main}()
@@
 \autoref{fig:incfixed}.  \autoref{fig:trap3} shows how one might go about this.
-But locks are intended to solve synchronization issues between multiple processes.
+But locks are intended to solve synchronization issues between multiple threads.
 If you run the code through Harmony, you will find that the code may not terminate.
-The issue is that a process can only acquire a lock once.  If the interrupt happens
+The issue is that a thread can only acquire a lock once.  If the interrupt happens
 after \textit{main}() acquires the lock but before \textit{main}() releases it, the
 \textit{handler}() method will block trying to acquire the lock, even though
-it is being acquired by the same process that already holds the lock.
+it is being acquired by the same thread that already holds the lock.
 
@@
 temporarily.  In Harmony, this can be done by setting the \textit{interrupt level}
-of a process to \texttt{True} using the \textbf{setintlevel} interface.
+of a thread to \texttt{True} using the \textbf{setintlevel} interface.
 \autoref{fig:trap4} illustrates how this is done.
@@
 \autoref{fig:trap6} gives an example of how this might be done.
-One important rule to remember is that a process should disable interrupts \emph{before}
+One important rule to remember is that a thread should disable interrupts \emph{before}
 attempting to acquire a lock.
@@
 While Harmony will only report one faulty run, this incorrect code can lead to
-the assertion failing as well as processes getting blocked indefinitely.
+the assertion failing as well as threads getting blocked indefinitely.
 
@@
 is a concurrent system in which a collection
-of processes communicate by message passing, much the same as
+of threads communicate by message passing, much the same as
 in the actor model.
@@
 into account,
-including failures of processes and failures of shared memory.
+including failures of threads and failures of shared memory.
 In this chapter, we will consider two actors, Alice and Bob.
@@
 It is useful to create an abstract network that reliably sends messages
-between processes, much like the FIFO queue in the \texttt{synch} module.
+between threads, much like the FIFO queue in the \texttt{synch} module.
 For this, we need a network protocol that Alice and Bob can run.
@@
 So far we have concentrated on critical sections to synchronize multiple
-processes.  Certainly, preventing multiple processes from accessing
+threads.  Certainly, preventing multiple threads from accessing
 certain code at the same time simplifies how to think about synchronization.
 However, it can lead to starvation.  Even in the absence of starvation,
-if some process is slow for some reason while being in the critical section,
-the other processes have to wait for it to finish executing the critical section.
+if some thread is slow for some reason while being in the critical section,
+the other threads have to wait for it to finish executing the critical section.
 Also, using synchronization primitives in interrupt handlers is tricky
@@
 In this chapter, we will have a look at how one can develop concurrent
-code in which processes do not have to wait for other processes to complete their
+code in which threads do not have to wait for other threads to complete their
 ongoing operations.
@@
 to atomically remove an item from a slot if there is one.
-This way, if two or more processes attempt to extract an item from
+This way, if two or more threads attempt to extract an item from
 the same slot, at most one will succeed.
 
-There is no critical section.  If one process is executing instructions
-very slowly, this does not negatively impact the other processes, as it
+There is no critical section.  If one thread is executing instructions
+very slowly, this does not negatively impact the other threads, as it
 would with solutions based on critical sections.
@@
 \caption{High-level state diagram specification of barrier synchronization
-with three processes and two rounds.  The three numbers specify how many
-rounds each process has completed.}
+with three threads and two rounds.  The three numbers specify how many
+rounds each thread has completed.}
 \label{fig:barrierdiagram}
@@
 A barrier is almost the opposite of a critical section:
-the intention is to get a group of processes to run some code at the
+the intention is to get a group of threads to run some code at the
 same time, instead of having them execute it one at a time.
-More precisely, with barrier synchronization the processes execute in rounds.
-Between each round there is a so-called \emph{barrier} where processes wait
-until all processes have completed the previous round, before they start the
+More precisely, with barrier synchronization the threads execute in rounds.
+Between each round there is a so-called \emph{barrier} where threads wait
+until all threads have completed the previous round, before they start the
 next one.
 For example, in an iterative matrix algorithm, the matrix may be
-cut up into fragments.  During a round, the processes run concurrently,
+cut up into fragments.  During a round, the threads run concurrently,
 one for each fragment.  The next round is not allowed to start
-until all processes have completed processing their fragment.
+until all threads have completed processing their fragment.
 
 \autoref{fig:barrierdiagram} shows a high-level depiction of barrier synchronization
-with three and processes.  Initially all processes are in round~0.  Then each process
-can start and finish the round.  However, none of the processes can
-progress to the next round until all processes have finished the
+with three and threads.  Initially all threads are in round~0.  Then each thread
+can start and finish the round.  However, none of the threads can
+progress to the next round until all threads have finished the
 current round.
@@
 \autoref{fig:barrier} shows an example.  There is a 0-initialized
-semaphore for each of the \texttt{N} processes.
-Before process $i$ enters a round, it first sends a signal to every
-other process and then waits until it receives a signal from
-every other process.  Process $i$ sends a signal to process $j$
+semaphore for each of the \texttt{N} threads.
+Before thread $i$ enters a round, it first sends a signal to every
+other thread and then waits until it receives a signal from
+every other thread.  Thread $i$ sends a signal to thread $j$
 by incrementing \texttt{sema[$j$]} (using \texttt{V}),
-while process $i$ waits for a signal be decrementing \texttt{sema[$i$]}
+while thread $i$ waits for a signal be decrementing \texttt{sema[$i$]}
 (using \texttt{P}).
@@
 The \textit{round} array is kept to check the correctness of this
-approach.  Each process increments its entry every time it enters
-a round.  If the algorithm is correct, it can never be that two processes
+approach.  Each thread increments its entry every time it enters
+a round.  If the algorithm is correct, it can never be that two threads
 are more than one round apart.
@@
 \begin{problems}
-\item See if you can implement barrier synchronization for \texttt{N} processes
+\item See if you can implement barrier synchronization for \texttt{N} threads
 with just three semaphores.  Busy waiting is not allowed.  How about just two?
@@
 \hline
-\texttt{processes $e$} & $e$ must be \texttt{()} \\
+\texttt{threads $e$} & $e$ must be \texttt{()} \\
 \hline
@@
 is something that can be assigned.  This can be a
-shared variable, a process variable, or a dereferenced pointer
+shared variable, a thread variable, or a dereferenced pointer
 variable.  It can also be indexed.  Examples include
@@
 \hline
-\texttt{notify}($c$) & notify a process waiting on condition variable \texttt{!}$c$ \\
-\hline
-\texttt{notifyAll}($c$) & notify all processes waiting on condition variable \texttt{!}$c$ \\
+\texttt{notify}($c$) & notify a thread waiting on condition variable \texttt{!}$c$ \\
+\hline
+\texttt{notifyAll}($c$) & notify all threads waiting on condition variable \texttt{!}$c$ \\
 \hline
@@
 \hline
-DelVar [$v$] & delete process variable $v$ \\
+DelVar [$v$] & delete thread variable $v$ \\
 \hline
@@
 \hline
-LoadVar [$v$] & push the value of a process variable onto the stack \\
+LoadVar [$v$] & push the value of a thread variable onto the stack \\
 \hline
@@
 \hline
+ReadonlyInc & increment the read-only counter of this context \\
+\hline
+ReadonlyDec & decrement the read-only counter of this context \\
+\hline
 Return & pop return address, push \texttt{result}, and restore program counter \\
@@
 \hline
-StoreVar [$v$] & pop a value from the stack and store it in a process variable \\
+StoreVar [$v$] & pop a value from the stack and store it in a thread variable \\
 \hline
@@
 The instruction then computes the address of the given key.
-\item Even though Harmony code does not allow taking addresses of process variables, both
-shared and process variables can have addresses.
+\item Even though Harmony code does not allow taking addresses of thread variables, both
+shared and thread variables can have addresses.
 \item The \texttt{Load}, \texttt{LoadVar}, \texttt{Del}, \texttt{DelVar},
@@
 $m$.  $m$ is supposed to leave the result on the stack.
-\item The \texttt{Frame} instruction pushes the value of the process
-register (\emph{i.e.}, the values of the process variables) onto the
+\item The \texttt{Frame} instruction pushes the value of the thread
+register (\emph{i.e.}, the values of the thread variables) onto the
 stack.  It initializes the \texttt{result} variable to the empty dictionary.
-The \texttt{Return} instruction restores the process register by popping
+The \texttt{Return} instruction restores the thread register by popping
 its value of the stack.
@@
 \end{itemize}
-The \texttt{Frame} instruction unpacks the argument to the method and places them into process variables by the given names.
+The \texttt{Frame} instruction unpacks the argument to the method and places them into thread variables by the given names.
 \item Every \texttt{Stop} instruction must immediately be followed by a 
@@
 
-\chapter{Contexts and Processes}
+\chapter{Contexts and Threads}
 \label{app:context}
 
-A context captures the state of a process.  Each time the process
+A context captures the state of a thread.  Each time the thread
 executes an instruction, it goes from one context to another.
@@
 leaves the context the same.
-There may be multiple processes
+There may be multiple threads
 with the same state at the same time.
@@
 \hline
-atomic & if non-zero, the process is in atomic mode \\
+atomic & if non-zero, the thread is in atomic mode \\
 \hline
@@
 \hline
-failure & if not None, string that describes how the process failed \\
+failure & if not None, string that describes how the thread failed \\
 \hline
@@
 \begin{itemize}
-\item The name in a name tag is the name of the method that the process
+\item The name in a name tag is the name of the method that the thread
 is executing;
@@
 the method, an ``invocation type atom'' (\texttt{normal}, \texttt{interrupt}, or
-\texttt{process}), and the return address (in case of \texttt{normal}).
-\item A process terminates when it reaches the \texttt{Return} instruction
-of the top-level method (when the stack frame is of type \texttt{process})
+\texttt{thread}), and the return address (in case of \texttt{normal}).
+\item A thread terminates when it reaches the \texttt{Return} instruction
+of the top-level method (when the stack frame is of type \texttt{thread})
 or when it hits an exception.  Exceptions include divide by zero,
@@
 variable, infinite loops of microsteps, as well as when an assertion fails;
-\item The execution of a process in \emph{atomic mode} does not get interleaved
-with that of other processes.
-\item The register of a process always contains a dictionary, mapping
+\item The execution of a thread in \emph{atomic mode} does not get interleaved
+with that of other threads.
+\item The register of a thread always contains a dictionary, mapping
 atoms to arbitrary values.  The atoms correspond to the variable names
@@
 the code.  All states until then are \texttt{initializing} states.
-Other processes, created through \texttt{spawn} statements, do not
+Other threads, created through \texttt{spawn} statements, do not
 start executing until then.
@@
 
-Consecutive macro steps by the same process are called \emph{mega steps}.
+Consecutive macro steps by the same thread are called \emph{mega steps}.
 Each state maintains the shortest path to it from the initial state in terms
@@
 of the graph (the number of such components are printed as part of the output).
-The sink SCCs should each consist of a terminal state without any processes.
+The sink SCCs should each consist of a terminal state without any threads.
 If not, again the state with the shortest path is reported.
@@
 Local variables are tightly scoped and cannot be shared
-between processes.
+between threads.
 While in theory one method can be declared within another,
@@
 All other variables are global and must be initialized
-before any processes are spawned.
+before any threads are spawned.
 

### 2020-11-26 book/sources/pp.py

@@ -54,6 +54,7 @@ def isreserved(s):
         "all",
         "and",
         "any",
+        "as",
         "assert",
         "atLabel",
         "atomic",
@@ -68,6 +69,7 @@ def isreserved(s):
         "elif",
         "else",
         "end",
+        "from",
         "False",
         "fun",
         "for",
@@ -220,9 +222,9 @@ def lexer(s, file, silent):
             found = s[:i].replace("_", "\\_")
             if isreserved(found):
                 if not silent: print("\\texttt{\\textbf{%s}}"%found, end="")
-                if found in [ "def", "const", "import" ]:
+                if found in [ "def", "const", "import", "from" ]:
                     nextconst = True
-                    if found == "import":
+                    if found in { "import", "from" }:
                         nextimport = True
             elif isletter(found[0]):
                 if nextconst:

### 2020-11-26 python/UpMany.py

@@ -1,6 +1,8 @@
 import threading
 
 N = 10000000
+count = 0
+done = [ False, False ]
 
 def incrementer(self):
     global count
@@ -17,8 +19,6 @@ def main():
 def spawn(f, a):
     threading.Thread(target=f, args=a).start()
 
-count = 0
-done = [ False, False ]
 spawn(incrementer, (0,))
 spawn(incrementer, (1,))
 spawn(main, ())

### 2020-11-26 book/paper.tex
Various updates

@@
 that supports locks and other synchronization primitives.
-\item \autoref{ch:cds} provides an introduction to building concurrent
+\item \autoref{ch:cds} gives an introduction to building concurrent
 data structures.
-\item \autoref{ch:rdwrbusy} and \autoref{ch:rdwr2} present the
-abstraction and implementation of \emph{reader/writer locks},
-which allow multiple threads in a critical section if they are
-all only executing read-only accesses.
+\item \autoref{ch:condwait} talks about threads having to wait for
+certain conditions.  As examples, it presents the reader/writer lock
+problem and the bounded buffer problem.
+\item \autoref{ch:sbs} presents \emph{Split Binary Semaphores}, a
+general technique for solving synchronization problems.
+\begin{comment}
 \autoref{ch:rdwrbusy} also discusses \emph{busy-waiting}, which is
@@
 and various solutions.
+\end{comment}
 \item \autoref{ch:starvation} talks about \emph{starvation}:
@@
 
-\chapter{Binary Semaphores and Blocking}
+\chapter{Blocking}
 \label{ch:synch}
@@
 In Harmony, \texttt{atomic} statements allow you to \emph{implement} your own
-synchronization primitives like test-and-set.  Other common examples
+interlock primitives such as test-and-set.  Other common examples
 include compare-and swap and fetch-and-add.  \texttt{atomic} statements
@@
 its program counter, and the contents of its register and stack.
-A context is a regular Harmony value.
+A context is a regular (if complex) Harmony value.
 The syntax of the expression is as follows:
@@
 \begin{code}
-\textbf{go} $C$ $R$
+\textbf{go} $C$ $R$;
 \end{code}
@@
 runs the file \texttt{x.hny} using the suspension version of the \texttt{synch} module.
-You will find that using the \texttt{synchS} module often leads to searching a
+You will find that using the \texttt{synchS} module often leads to
+the model checker searching a
 significantly larger state space than using the \texttt{synch} module.
@@
 \chapter{Reader/Writer Locks with Blocking}
-\label{ch:rdwr2}
+\label{ch:rdwr}
 
@@
 \hline
-\texttt{import m} & \texttt{m} identifies a module \\
+\texttt{import m, ...} & \texttt{m} identifies a module \\
+\hline
+\texttt{from m import ...} & \texttt{m} identifies a module \\
 \hline
@@
 \hline
-\texttt{bagEmpty()} & returns an empty bag\\
-\hline
-\texttt{bagFromSet($s$)} & create a bag from set $s$\\
-\hline
-\texttt{bagCount($b$, $e$)} & count how many times $e$ occurs in bag $b$\\
-\hline
-\texttt{bagChoose($b$)} & like \texttt{choose(s)}, but applied to a bag\\
-\hline
-\texttt{bagAdd}(\textit{pb}, $e$) & add one copy of $e$ to bag \texttt{!}\textit{pb}\\
-\hline
-\texttt{bagRemove}(\textit{pb}, $e$) & remove one copy of $e$ from bag \texttt{!}\textit{pb}\\
+\texttt{empty()} & returns an empty bag\\
+\hline
+\texttt{fromSet($s$)} & create a bag from set $s$\\
+\hline
+\texttt{fromList($t$)} & convert list $t$ into a bag \\
+\hline
+\texttt{count($b$, $e$)} & count how many times $e$ occurs in bag $b$\\
+\hline
+\texttt{bchoose($b$)} & like \texttt{choose(s)}, but applied to a bag\\
+\hline
+\texttt{add}(\textit{pb}, $e$) & add one copy of $e$ to bag \texttt{!}\textit{pb}\\
+\hline
+\texttt{remove}(\textit{pb}, $e$) & remove one copy of $e$ from bag \texttt{!}\textit{pb}\\
 \hline
@@
 \hline
-\texttt{list2bag($t$)} & convert list $t$ into a bag \\
-\hline
-\texttt{listQsort($t$)} & sort list $t$ using quicksort\\
+\texttt{qsort($t$)} & sort list $t$ using quicksort\\
 \hline
@@
 \hline
-\texttt{Lock}() & return a lock object \\
-\hline
-\texttt{lock}(\textit{lk}) & acquire \texttt{!}\textit{lk} \\
-\hline
-\texttt{unlock}(\textit{lk}) & release \texttt{!}\textit{lk} \\
-\hline
-\texttt{Condition}(\textit{lk}) & return a condition variable on \texttt{!}\textit{lk} \\
-\hline
-\texttt{wait}($c$) & wait on condition variable \texttt{!}$c$ \\
-\hline
+\texttt{BinSem}($v$) & return a binary semaphore initialized to $v$ \\
+\hline
+\texttt{Lock}() & return a binary semaphore initialized to \texttt{False} \\
+\hline
+\texttt{acquire}(\textit{bs}) & acquire binary semaphore \texttt{!}\textit{bs} \\
+\hline
+\texttt{release}(\textit{bs}) & release binary semaphore \texttt{!}\textit{bs} \\
+\hline
+\texttt{Condition}() & return a condition variable \\
+\hline
+\texttt{wait}($c$, \textit{lk}) & wait on condition variable \texttt{!}$c$
+and lock \textit{lk} \\ \hline
 \texttt{notify}($c$) & notify a thread waiting on condition variable \texttt{!}$c$ \\
@@
 \hline
-\texttt{Semaphore}(\textit{cnt}) & return a semaphore object initialized to \textit{cnt} \\
+\texttt{Semaphore}(\textit{cnt}) & return a counting semaphore initialized to \textit{cnt} \\
 \hline
@@
 \hline
+readonly & if non-zero, the thread is in read-only mode \\
+\hline
 stack & a list of Harmony values \\
@@
 with that of other threads.
+\item The execution of a thread in \emph{read-only mode} is not allowed
+to update shared variables of spawn threads.
 \item The register of a thread always contains a dictionary, mapping

### 2020-11-26 harmony.py

@@ -1428,6 +1428,9 @@ class SpawnOp(Op):
         return "pop a pc, argument, and tag and spawn a new process"
 
     def eval(self, state, context):
+        if context.readonly > 0:
+            context.failure = "Error: no spawn allowed in assert"
+            return
         method = context.pop()
         assert isinstance(method, PcValue)
         arg = context.pop()

### 2020-11-27 book/paper.tex
Latest

@@
 requires multiple steps.
-A critical section allows only one thread to execute its code at a time.
-When a thread arrives at a critical section,
+A critical section only allows one thread to execute the code
+of the critical section at a time.
+Therefore, when a thread arrives at a critical section,
 the thread blocks until there is no other thread in the critical section.
@@
 \end{code}
-\caption{\harmonylink{code/hoare.hny} Implementation of Hoare monitors.}
+\caption{\harmonylink{modules/hoare.hny} Implementation of Hoare monitors.}
 \label{fig:hoare}
@@
 happening.
-\autoref{fig:dinersavoid} demonstrates this concept.  It uses a semaphore to
-restrict the number of diners at any time to four.
+\autoref{fig:dinersavoid} demonstrates this concept.  It uses a
+\emph{counting semaphore} to restrict the number of diners at any time to
+four.  A counting semaphore is like a binary semaphore, but can be
+acquired a given number of times.  It is supported by the \texttt{synch}
+module.
 
@@
 \texttt{remove}(\textit{pb}, $e$) & remove one copy of $e$ from bag \texttt{!}\textit{pb}\\
+\hline
+\end{tabular}
+
+\section{The \texttt{Hoare} module}
+\label{ap:hoare}
+
+\index{list module}%
+The \texttt{hoare} module implements support for Hoare-style monitors
+and condition variables.
+
+\vspace{1em}
+\begin{tabular}{|l|l|}
+\hline
+\texttt{Monitor()} & return a monitor mutex\\
+\hline
+\texttt{enter($m$)} & enter a monitor.  $m$ points to a monitor mutex\\
+\hline
+\texttt{exit($m$)} & exit a monitor.\\
+\hline
+\texttt{Condition()} & return a condition variable\\
+\hline
+\texttt{wait($c$, $m$)} & wait on condition variable pointed to be $c$ in monitor pointed to by $m$\\
+\hline
+\texttt{signal($,$. $m$)} & signal a condition variable\\
 \hline

### 2020-11-27 code/x.c

@@ -1,0 +1,23 @@
+#include <stdio.h>
+#include <signal.h>
+#include <string.h>
+
+int done = 0;
+
+void handler(int signum){
+	done = 1;
+}
+
+int main(){
+    struct sigaction sa;
+
+    memset(&sa, 0, sizeof(sa));
+    sa.sa_handler = handler;
+    sa.sa_flags = SA_RESTART; /* Restart syscalls if interrupted */
+    sigaction(SIGINT, &sa, NULL);
+
+    while (!done)
+        ;
+    printf("DONE\n");
+    return 0;
+}

### 2020-11-27 book/paper.tex
Latest

@@
 a lock for each node in the tree.
-\item \label{ex:qsort}
-Bonus problem: \autoref{fig:qsort} shows an iterative implementation of the Qsort
-algorithm, and \autoref{fig:qsorttest} an accompanying test program.
-The array to be sorted is stored in  shared variable $a$.
-Another shared variable, \textit{todo}, contains the ranges of the
-array that need to be sorted (initially the entire array).
-Re-using as much of this code as you can, implement a parallel version of
-this.  You should not have to change the methods \texttt{swap}, \texttt{partition},
-or \texttt{sortrange} for this.  Create \texttt{NWORKERS} ``worker threads''
-that should replace the \texttt{qsort} code.  Each worker loops until \textit{todo}
-is empty and sorts the ranges that it finds until then.  The \texttt{main}
-thread needs to wait until all workers are done.  Use a semaphore for this:
-the \texttt{main} thread should use \texttt{P} to wait for each
-worker to terminate.  After that it should check that the array is sorted and
-that it is a permutation of the original array.
 \end{problems}
-
-\begin{figure}
-\begin{code}
-\harmonysource{qsort}
-\end{code}
-\caption{\harmonylink{code/qsort.hny} Iterative qsort() implementation.}
-\label{fig:qsort}
-\end{figure}
-
-\begin{figure}
-\begin{code}
-\harmonysource{qsorttest}
-\end{code}
-\caption{\harmonylink{code/qsorttest.hny} Test program for \autoref{fig:qsort}.}
-\label{fig:qsorttest}
-\end{figure}
 
@@
 it is fair.
+\item \label{ex:qsort}
+Bonus problem: \autoref{fig:qsort} shows an iterative implementation of the Qsort
+algorithm, and \autoref{fig:qsorttest} an accompanying test program.
+The array to be sorted is stored in shared variable \textit{textqs.arr}.
+Another shared variable, \textit{testqs.todo}, contains the ranges of the
+array that need to be sorted (initially the entire array).
+Re-using as much of this code as you can, implement a parallel version of
+this.  You should not have to change the methods \texttt{swap}, \texttt{partition},
+or \texttt{sortrange} for this.  Create \texttt{NWORKERS} ``worker threads''
+that should replace the \texttt{qsort} code.
+Each worker loops until \textit{todo}
+is empty and sorts the ranges that it finds until then.  The \texttt{main}
+thread needs to wait until all workers are done.
 \end{problems}
+
+\begin{figure}
+\begin{code}
+\harmonysource{qsort}
+\end{code}
+\caption{\harmonylink{code/qsort.hny} Iterative qsort() implementation.}
+\label{fig:qsort}
+\end{figure}
+
+\begin{figure}
+\begin{code}
+\harmonysource{qsorttest}
+\end{code}
+\caption{\harmonylink{code/qsorttest.hny} Test program for \autoref{fig:qsort}.}
+\label{fig:qsorttest}
+\end{figure}
 

### 2020-11-27 book/paper.tex
Latest

@@
 \begin{problems}
+\item \label{ex:bbmesa} Implement a solution to the bounded buffer problem using Mesa condition
+variables.
 \item Implement a ``try lock'' module using Mesa condition variables
@@
 \begin{problems}
-\item The \texttt{produce} method in \autoref{fig:boundedbuffer} cannot be used
+\item The \texttt{enqueue} method you implemented in \autoref{ex:bbmesa} cannot be used
 in interrupt handlers for two reasons: (1) it is not interrupt-safe, and (2)
@@
 say, a keyboard interrupt handler could place an event on a shared queue.
-Similar to \autoref{ex:nbproduce}, implement
-a new method \texttt{iproduce}(\textit{item}) that does not
+Implement a new method \texttt{i\_enqueue}(\textit{item}) that does not
 block.  Instead, it should return \texttt{False} if the buffer is full and \texttt{True}
@@
 \end{code}
-\caption{\harmonylink{code/barrier.hny} Barrier synchronization with semaphores.}
+\caption{\harmonylink{code/barrier.hny} Barrier synchronization with queues.}
 \label{fig:barrier}
@@
 
-Counting semaphores work well for implementing barrier synchronization.
-\autoref{fig:barrier} shows an example.  There is a 0-initialized
-semaphore for each of the \texttt{N} threads.
-Before thread $i$ enters a round, it first sends a signal to every
-other thread and then waits until it receives a signal from
-every other thread.  Thread $i$ sends a signal to thread $j$
-by incrementing \texttt{sema[$j$]} (using \texttt{V}),
-while thread $i$ waits for a signal be decrementing \texttt{sema[$i$]}
-(using \texttt{P}).
+Blocking queues work well for implementing barrier synchronization.
+\autoref{fig:barrier} shows an example.  There is a queue
+for each of the \texttt{N} threads.
+Before thread $i$ enters a round, it first sends enqueues a message to every
+other thread and then waits until it receives a message from
+every other thread.
+In this simple case, each message contains \texttt{None}, but in practice useful
+information may be exchanged between the threads.
 
@@
 \item See if you can implement barrier synchronization for \texttt{N} threads
-with just three semaphores.  Busy waiting is not allowed.  How about just two?
+with just three binary semaphores.  Busy waiting is not allowed.  How about just two?
 (As always, the Little Book of Semaphores~\cite{Downey09} is a good resource

### 2020-11-28 book/paper.tex
Renamed enqueue/dequeue into put/get, consistent with Python Queue

@@
 \item $q = \mathtt{Queue}()$: allocate a new queue;
-\item $\mathtt{enqueue}(q, v)$: add $v$ to the tail of the queue;
-\item $r = \mathtt{dequeue}(q)$: returns $r = ()$ if $q$ is empty or $r = (v,)$ (a singleton tuple)
+\item $\mathtt{put}(q, v)$: add $v$ to the tail of the queue;
+\item $r = \mathtt{get}(q)$: returns $r = ()$ if $q$ is empty or $r = (v,)$ (a singleton tuple)
 if $v$ was at the head of the queue.
@@
 and tail pointer and a lock.
-$\mathtt{queue.enqueue}(q, v)$ and $\mathtt{queue.dequeue}(q)$ both take a pointer $q$ to the
+$\mathtt{queue.put}(q, v)$ and $\mathtt{queue.get}(q)$ both take a pointer $q$ to the
 queue object because both may modify the queue.
@@
 The implementation in \autoref{fig:queueMS}~\cite{MS98} is similar but uses
-separate locks for the head and tail, allowing an enqueue and a dequeue operation
+separate locks for the head and tail, allowing a \texttt{put} and a \texttt{get} operation
 to proceed concurrently.  To avoid contention between the head and the tail,
@@
 A \emph{bounded buffer} (aka \emph{ring buffer}) is
-a queue with the usual enqueue/dequeue interface,
+a queue with the usual \texttt{put/get} interface,
 but implemented using a circular buffer
@@
 There is no \texttt{release\_one} method.  Instead, one can conclude
-that \texttt{enqueue} causes the queue to be non-empty, and
+that \texttt{put} causes the queue to be non-empty, and
 \texttt{signal} will check if there are any threads waiting for
@@
 \item \texttt{Queue()} returns a new message queue;
-\item \texttt{enqueue}($q$, \textit{item}) adds \textit{item} to the queue pointed to by $q$;
-\item \texttt{dequeue}($q$) waits for and returns an item on the queue pointed to by $q$.
+\item \texttt{put}($q$, \textit{item}) adds \textit{item} to the queue pointed to by $q$;
+\item \texttt{get}($q$) waits for and returns an item on the queue pointed to by $q$.
 \end{itemize}
@@
 Note that a \texttt{Queue} behaves much like a zero-initialized semaphore.
-\texttt{enqueue} is much like \texttt{V}, except that it is accompanied by data.
-\texttt{dequeue} is much like \texttt{P}, except that it also returns data.
+\texttt{put} is much like \texttt{V}, except that it is accompanied by data.
+\texttt{get} is much like \texttt{P}, except that it also returns data.
 Thus, synchronized queues can be considered a generalization of counting semaphores.
@@
 \begin{problems}
-\item The \texttt{enqueue} method you implemented in \autoref{ex:bbmesa} cannot be used
+\item The \texttt{put} method you implemented in \autoref{ex:bbmesa} cannot be used
 in interrupt handlers for two reasons: (1) it is not interrupt-safe, and (2)
@@
 say, a keyboard interrupt handler could place an event on a shared queue.
-Implement a new method \texttt{i\_enqueue}(\textit{item}) that does not
+Implement a new method \texttt{i\_put}(\textit{item}) that does not
 block.  Instead, it should return \texttt{False} if the buffer is full and \texttt{True}
@@
 for each of the \texttt{N} threads.
-Before thread $i$ enters a round, it first sends enqueues a message to every
+Before thread $i$ enters a round, it first sends a message to every
 other thread and then waits until it receives a message from
@@
 \hline
-\texttt{dequeue}($q$) & return next element of $q$, blocking if empty \\
-\hline
-\texttt{enqueue}($q$, \textit{item}) & add \textit{item} to $a$ \\
+\texttt{get}($q$) & return next element of $q$, blocking if empty \\
+\hline
+\texttt{put}($q$, \textit{item}) & add \textit{item} to $a$ \\
 \hline

### 2020-11-28 book/paper.tex
Latest

@@
 
+There are at least two reasons why Harmony is not object-oriented.  First,
+object-orientation often adds layers of indirection that would make it
+harder to model check and also to interpret the results.  Consider, for example,
+a lock.  In Python, a lock is an object.  A lock variable would contain a
+reference to a lock object.  In Harmony, a lock variable contains the value
+of the lock itself.  Second, many concurrency solutions in the literature are
+expressed in C or some other low-level language that does not support
+object-orientation, but instead use \texttt{malloc} and \texttt{free}.
+
 \section{Constants, Global and Local Variables}

### 2020-11-28 book/paper.tex
Improved chapter order
Latest

@@
 and \emph{message passing} as an approach to synchronization.
+\item \autoref{ch:barrier} describes \emph{barrier synchronization},
+useful in high-performance computing applications such as parallel simulations.
 \item \autoref{ch:interrupts} discusses how to handle interrupts,
 a problem closely related to synchronization among multiple threads.
-\item \autoref{ch:abp} presents a problem and a solution to the distributed
-systems problem of having two threads communicate reliably over an unreliable
-network.
 \item \autoref{ch:nonblocking} introduces \emph{non-blocking} or
@@
 steps.
-\item \autoref{ch:barrier} describes \emph{barrier synchronization},
-useful in high-performance computing applications such as parallel simulations.
+\item \autoref{ch:abp} presents a problem and a solution to the distributed
+systems problem of having two threads communicate reliably over an unreliable
+network.
 \end{itemize}
@@
 
+\chapter{Barrier Synchronization}
+\label{ch:barrier}
+\index{barrier synchronization}%
+
+\glsadd{barrier synchronization}%
+
+\begin{figure}
+\begin{center}
+\includegraphics[width=\textwidth]{figures/barrier-crop.pdf}
+\end{center}
+\caption{High-level state diagram specification of barrier synchronization
+with three threads and two rounds.  The three numbers specify how many
+rounds each thread has completed.}
+\label{fig:barrierdiagram}
+\end{figure}
+
+\begin{figure}
+\begin{code}
+\harmonysource{barrier}
+\end{code}
+\caption{\harmonylink{code/barrier.hny} Barrier synchronization with queues.}
+\label{fig:barrier}
+\end{figure}
+
+Barrier synchronization is a problem that comes up in high-performance
+parallel computing, used, among others, for scalable simulation.
+A barrier is almost the opposite of a critical section:
+the intention is to get a group of threads to run some code at the
+same time, instead of having them execute it one at a time.
+More precisely, with barrier synchronization the threads execute in rounds.
+Between each round there is a so-called \emph{barrier} where threads wait
+until all threads have completed the previous round, before they start the
+next one.
+For example, in an iterative matrix algorithm, the matrix may be
+cut up into fragments.  During a round, the threads run concurrently,
+one for each fragment.  The next round is not allowed to start
+until all threads have completed processing their fragment.
+
+\autoref{fig:barrierdiagram} shows a high-level depiction of barrier synchronization
+with three and threads.  Initially all threads are in round~0.  Then each thread
+can start and finish the round.  However, none of the threads can
+progress to the next round until all threads have finished the
+current round.
+
+Blocking queues work well for implementing barrier synchronization.
+\autoref{fig:barrier} shows an example.  There is a queue
+for each of the \texttt{N} threads.
+Before thread $i$ enters a round, it first sends a message to every
+other thread and then waits until it receives a message from
+every other thread.
+In this simple case, each message contains \texttt{None}, but in practice useful
+information may be exchanged between the threads.
+
+The \textit{round} array is kept to check the correctness of this
+approach.  Each thread increments its entry every time it enters
+a round.  If the algorithm is correct, it can never be that two threads
+are more than one round apart.
+
+\section*{Exercises}
+\begin{problems}
+\item See if you can implement barrier synchronization for \texttt{N} threads
+with just three binary semaphores.  Busy waiting is not allowed.  How about just two?
+(As always, the Little Book of Semaphores~\cite{Downey09} is a good resource
+for solving synchronization problems with semaphores.
+Look for the \emph{double turnstile}
+\index{double turnstile}%
+solution.)
+\item Implement barrier synchronization with Mesa condition variables.
+(You may want to use a double turnstile approach here as well.)
+\end{problems}
+
 \chapter{Interrupts}
@@
 The method also needs to be interrupt-safe.
+\end{problems}
+
+\chapter{Non-Blocking Synchronization}
+\label{ch:nonblocking}
+\index{non-blocking synchronization}%
+
+\glsadd{non-blocking synchronization}%
+
+\begin{figure}
+\begin{code}
+\harmonysource{hw}
+\end{code}
+\caption{\harmonylink{code/hw.hny} Non-blocking queue.}
+\label{fig:hw}
+\end{figure}
+
+So far we have concentrated on critical sections to synchronize multiple
+threads.  Certainly, preventing multiple threads from accessing
+certain code at the same time simplifies how to think about synchronization.
+However, it can lead to starvation.  Even in the absence of starvation,
+if some thread is slow for some reason while being in the critical section,
+the other threads have to wait for it to finish executing the critical section.
+Also, using synchronization primitives in interrupt handlers is tricky
+to get right (\autoref{ch:interrupts}) and might run too long.
+In this chapter, we will have a look at how one can develop concurrent
+code in which threads do not have to wait for other threads to complete their
+ongoing operations.
+
+As a first example, we will revisit the producer/consumer problem.
+The code in \autoref{fig:hw} is based on code developed by Herlihy and
+Wing~\cite{HW87}.
+The code is a ``proof of existence'' for non-blocking synchronization; it
+is not necessarily practical.
+There are two variables.  \textit{items} is an unbounded array with each
+entry initialized to \texttt{()}.  \textit{back} is an index into the
+array and points to the next slot where a new value is inserted.
+The code uses two interlock instructions:
+\begin{itemize}
+\item \texttt{inc}($p$): atomically increments \texttt{!}$p$ and returns
+the old value;
+\item \texttt{exch}($p$): sets \texttt{!}$p$ to \texttt{()} and returns
+the old value.
+\end{itemize}
+
+Method \texttt{produce}(\textit{item}) uses \texttt{inc}(?\textit{back})
+to allocate
+the next available slot in the \textit{items} array.
+It stores the item as a singleton tuple.
+Method \texttt{consume()} repeatedly scans the array, up to the
+\textit{back} index, trying to find an item to return.
+To check an entry, it uses \texttt{exch()}
+to atomically remove an item from a slot if there is one.
+This way, if two or more threads attempt to extract an item from
+the same slot, at most one will succeed.
+
+There is no critical section.  If one thread is executing instructions
+very slowly, this does not negatively impact the other threads, as it
+would with solutions based on critical sections.
+On the contrary, it helps them because it creates less contention.
+Unfortunately, the solution is not practical for the following reasons:
+\begin{itemize}
+\item The \textit{items} array must be of infinite size if an unbounded number
+of items may be produced;
+\item Each slot in the array is only used once, which is inefficient;
+\item the \texttt{inc} and \texttt{exch} interlock instructions are not
+universally available on existing processors.
+\end{itemize}
+However, in the literature there are many examples of practical
+non-blocking (aka \emph{wait-free})
+\index{wait-free synchronization}%
+synchronization algorithms.
+% TODO~\cite{}.
+
+There are also various algorithms where read-only access is wait-free
+but updates do require a lock---these are useful in situations where
+most accesses are read-only and the performance of updates is less
+critical. %~\cite{}.
+\autoref{fig:lst1} and \autoref{fig:lst2} implements an ordered linked
+list of integers without duplicates.
+There are two update operations:
+values can be added using \texttt{lst\_insert} or deleted using
+\texttt{lst\_remove}.
+There is also a read-only operation: \texttt{lst\_contains} checks if
+a particular value is in the list.
+\glsadd{linearizable}%
+
+The read-only operation \texttt{lst\_contains} is wait\_free: it scans
+through the list without having to obtain a lock.
+Nonetheless, the implementation of the list is
+\emph{linearizable}~\cite{HW90},
+\index{linearizable}%
+a strong notion of consistency that
+makes it appear as if each of the operations executes atomically at
+some point between their invocation and return.
+
+The list has two ``book-end'' nodes with values \texttt{-inf} and
+\texttt{inf} (similar to the Python \texttt{math.inf} constant).
+Each node has a lock, a value, and \textit{next}, a pointer to the next node
+(which is \texttt{None} for the final \texttt{inf} node).
+The \texttt{lst\_find}(\textit{lst}, $v$) method is a helper function for update
+methods that
+finds and locks two consecutive nodes \textit{before} and \textit{after}
+such that $\mathit{before}.\mathtt{data.value} < v \le \mathit{after}.\mathtt{data.value}$.
+It does so by performing something called \emph{hand-over-hand locking}.
+\index{hand-over-hand locking}%
+It first locks the first node, which is the \texttt{-inf} node.
+Then, iteratively, it obtains a lock on the next node and release the
+lock on the last one, and so on, similar to climbing a tree
+hand-over-hand.
+
+Having the two adjoining nodes locked, implementing \texttt{lst\_insert}
+and \texttt{lst\_remove} is straightforward.
+The \texttt{lst\_contains} method can simply scan through the list
+because the \textit{next} pointers are updated atomically in such a
+way that they always point to a legal suffix of the list.
+An invariant of the algorithm is that at any point in time the
+list is ``valid,'' starting with a \texttt{-inf} node and ending
+with a \texttt{inf} node.
+
+Determining if an implementation of a concurrent data structure
+is linearizable involves finding what are known as the
+\emph{linearization points}
+\index{linearization point}%
+of the operations in an execution.  These are the unique
+points in time at which an operation appears to execute atomically.
+The linearization points for the \texttt{lst\_insert} and
+\texttt{lst\_remove} operations coincide exactly with the update
+of the \textit{before}.\texttt{next} pointer.
+The linearization point of a \texttt{lst\_contains} method
+execution depends on whether the value is found or not.
+If found, it coincides with retrieving the pointer to the node
+that has the value.
+If not found, it coincides with retrieving the pointer to the
+\texttt{inf} node.
+
+\begin{figure}
+\begin{code}
+\harmonysource{lst1}
+\end{code}
+\caption{\harmonylink{code/lst1.hny} List with non-blocking read operation, part 1.}
+\label{fig:lst1}
+\end{figure}
+
+\begin{figure}
+\begin{code}
+\harmonysource{lst2}
+\end{code}
+\caption{\harmonylink{code/lst2.hny} List with non-blocking read operation, part 2.}
+\label{fig:lst2}
+\end{figure}
+
+\section*{Exercises}
+\begin{problems}
+\item Add read-only methods to the data structure in \autoref{fig:lst1}
+and \autoref{fig:lst2} that report the size of the list, the minimum value in the
+list, the maximum value in the list, and the sum of the values in the list.
+Are they linearizable?  If so, what are their linearization points?
+\item A \emph{seqlock}
+\index{seqlock}%
+consists of a lock and a version number.
+An update operation acquires the lock, increments the version number, makes the
+changes to the data structure, and then releases the lock.  A read-only operation
+does not use the lock.  Instead, it retrieves the version number,
+reads the data structure, and then checks if the
+version number has changed.  If so, the read-only operation is retried.
+Use a seqlock to implement a bank much like \autoref{ex:bank}, with
+one seqlock for the entire bank (i.e., no locks on individual accounts).
+Method \texttt{transfer} is an update operation; method \texttt{total} is a
+read-only operation.  Explain how a seqlock can lead to starvation.
 \end{problems}
@@
 for each client, or solve the problem using a single pair of channels.
-\end{problems}
-
-\chapter{Non-Blocking Synchronization}
-\label{ch:nonblocking}
-\index{non-blocking synchronization}%
-
-\glsadd{non-blocking synchronization}%
-
-\begin{figure}
-\begin{code}
-\harmonysource{hw}
-\end{code}
-\caption{\harmonylink{code/hw.hny} Non-blocking queue.}
-\label{fig:hw}
-\end{figure}
-
-So far we have concentrated on critical sections to synchronize multiple
-threads.  Certainly, preventing multiple threads from accessing
-certain code at the same time simplifies how to think about synchronization.
-However, it can lead to starvation.  Even in the absence of starvation,
-if some thread is slow for some reason while being in the critical section,
-the other threads have to wait for it to finish executing the critical section.
-Also, using synchronization primitives in interrupt handlers is tricky
-to get right (\autoref{ch:interrupts}) and might run too long.
-In this chapter, we will have a look at how one can develop concurrent
-code in which threads do not have to wait for other threads to complete their
-ongoing operations.
-
-As a first example, we will revisit the producer/consumer problem.
-The code in \autoref{fig:hw} is based on code developed by Herlihy and
-Wing~\cite{HW87}.
-The code is a ``proof of existence'' for non-blocking synchronization; it
-is not necessarily practical.
-There are two variables.  \textit{items} is an unbounded array with each
-entry initialized to \texttt{()}.  \textit{back} is an index into the
-array and points to the next slot where a new value is inserted.
-The code uses two interlock instructions:
-\begin{itemize}
-\item \texttt{inc}($p$): atomically increments \texttt{!}$p$ and returns
-the old value;
-\item \texttt{exch}($p$): sets \texttt{!}$p$ to \texttt{()} and returns
-the old value.
-\end{itemize}
-
-Method \texttt{produce}(\textit{item}) uses \texttt{inc}(?\textit{back})
-to allocate
-the next available slot in the \textit{items} array.
-It stores the item as a singleton tuple.
-Method \texttt{consume()} repeatedly scans the array, up to the
-\textit{back} index, trying to find an item to return.
-To check an entry, it uses \texttt{exch()}
-to atomically remove an item from a slot if there is one.
-This way, if two or more threads attempt to extract an item from
-the same slot, at most one will succeed.
-
-There is no critical section.  If one thread is executing instructions
-very slowly, this does not negatively impact the other threads, as it
-would with solutions based on critical sections.
-On the contrary, it helps them because it creates less contention.
-Unfortunately, the solution is not practical for the following reasons:
-\begin{itemize}
-\item The \textit{items} array must be of infinite size if an unbounded number
-of items may be produced;
-\item Each slot in the array is only used once, which is inefficient;
-\item the \texttt{inc} and \texttt{exch} interlock instructions are not
-universally available on existing processors.
-\end{itemize}
-However, in the literature there are many examples of practical
-non-blocking (aka \emph{wait-free})
-\index{wait-free synchronization}%
-synchronization algorithms.
-% TODO~\cite{}.
-
-There are also various algorithms where read-only access is wait-free
-but updates do require a lock---these are useful in situations where
-most accesses are read-only and the performance of updates is less
-critical. %~\cite{}.
-\autoref{fig:lst1} and \autoref{fig:lst2} implements an ordered linked
-list of integers without duplicates.
-There are two update operations:
-values can be added using \texttt{lst\_insert} or deleted using
-\texttt{lst\_remove}.
-There is also a read-only operation: \texttt{lst\_contains} checks if
-a particular value is in the list.
-\glsadd{linearizable}%
-
-The read-only operation \texttt{lst\_contains} is wait\_free: it scans
-through the list without having to obtain a lock.
-Nonetheless, the implementation of the list is
-\emph{linearizable}~\cite{HW90},
-\index{linearizable}%
-a strong notion of consistency that
-makes it appear as if each of the operations executes atomically at
-some point between their invocation and return.
-
-The list has two ``book-end'' nodes with values \texttt{-inf} and
-\texttt{inf} (similar to the Python \texttt{math.inf} constant).
-Each node has a lock, a value, and \textit{next}, a pointer to the next node
-(which is \texttt{None} for the final \texttt{inf} node).
-The \texttt{lst\_find}(\textit{lst}, $v$) method is a helper function for update
-methods that
-finds and locks two consecutive nodes \textit{before} and \textit{after}
-such that $\mathit{before}.\mathtt{data.value} < v \le \mathit{after}.\mathtt{data.value}$.
-It does so by performing something called \emph{hand-over-hand locking}.
-\index{hand-over-hand locking}%
-It first locks the first node, which is the \texttt{-inf} node.
-Then, iteratively, it obtains a lock on the next node and release the
-lock on the last one, and so on, similar to climbing a tree
-hand-over-hand.
-
-Having the two adjoining nodes locked, implementing \texttt{lst\_insert}
-and \texttt{lst\_remove} is straightforward.
-The \texttt{lst\_contains} method can simply scan through the list
-because the \textit{next} pointers are updated atomically in such a
-way that they always point to a legal suffix of the list.
-An invariant of the algorithm is that at any point in time the
-list is ``valid,'' starting with a \texttt{-inf} node and ending
-with a \texttt{inf} node.
-
-Determining if an implementation of a concurrent data structure
-is linearizable involves finding what are known as the
-\emph{linearization points}
-\index{linearization point}%
-of the operations in an execution.  These are the unique
-points in time at which an operation appears to execute atomically.
-The linearization points for the \texttt{lst\_insert} and
-\texttt{lst\_remove} operations coincide exactly with the update
-of the \textit{before}.\texttt{next} pointer.
-The linearization point of a \texttt{lst\_contains} method
-execution depends on whether the value is found or not.
-If found, it coincides with retrieving the pointer to the node
-that has the value.
-If not found, it coincides with retrieving the pointer to the
-\texttt{inf} node.
-
-\begin{figure}
-\begin{code}
-\harmonysource{lst1}
-\end{code}
-\caption{\harmonylink{code/lst1.hny} List with non-blocking read operation, part 1.}
-\label{fig:lst1}
-\end{figure}
-
-\begin{figure}
-\begin{code}
-\harmonysource{lst2}
-\end{code}
-\caption{\harmonylink{code/lst2.hny} List with non-blocking read operation, part 2.}
-\label{fig:lst2}
-\end{figure}
-
-\section*{Exercises}
-\begin{problems}
-\item Add read-only methods to the data structure in \autoref{fig:lst1}
-and \autoref{fig:lst2} that report the size of the list, the minimum value in the
-list, the maximum value in the list, and the sum of the values in the list.
-Are they linearizable?  If so, what are their linearization points?
-\item A \emph{seqlock}
-\index{seqlock}%
-consists of a lock and a version number.
-An update operation acquires the lock, increments the version number, makes the
-changes to the data structure, and then releases the lock.  A read-only operation
-does not use the lock.  Instead, it retrieves the version number,
-reads the data structure, and then checks if the
-version number has changed.  If so, the read-only operation is retried.
-Use a seqlock to implement a bank much like \autoref{ex:bank}, with
-one seqlock for the entire bank (i.e., no locks on individual accounts).
-Method \texttt{transfer} is an update operation; method \texttt{total} is a
-read-only operation.  Explain how a seqlock can lead to starvation.
-\end{problems}
-
-\chapter{Barrier Synchronization}
-\label{ch:barrier}
-\index{barrier synchronization}%
-
-\glsadd{barrier synchronization}%
-
-\begin{figure}
-\begin{center}
-\includegraphics[width=\textwidth]{figures/barrier-crop.pdf}
-\end{center}
-\caption{High-level state diagram specification of barrier synchronization
-with three threads and two rounds.  The three numbers specify how many
-rounds each thread has completed.}
-\label{fig:barrierdiagram}
-\end{figure}
-
-\begin{figure}
-\begin{code}
-\harmonysource{barrier}
-\end{code}
-\caption{\harmonylink{code/barrier.hny} Barrier synchronization with queues.}
-\label{fig:barrier}
-\end{figure}
-
-Barrier synchronization is a problem that comes up in high-performance
-parallel computing, used, among others, for scalable simulation.
-A barrier is almost the opposite of a critical section:
-the intention is to get a group of threads to run some code at the
-same time, instead of having them execute it one at a time.
-More precisely, with barrier synchronization the threads execute in rounds.
-Between each round there is a so-called \emph{barrier} where threads wait
-until all threads have completed the previous round, before they start the
-next one.
-For example, in an iterative matrix algorithm, the matrix may be
-cut up into fragments.  During a round, the threads run concurrently,
-one for each fragment.  The next round is not allowed to start
-until all threads have completed processing their fragment.
-
-\autoref{fig:barrierdiagram} shows a high-level depiction of barrier synchronization
-with three and threads.  Initially all threads are in round~0.  Then each thread
-can start and finish the round.  However, none of the threads can
-progress to the next round until all threads have finished the
-current round.
-
-Blocking queues work well for implementing barrier synchronization.
-\autoref{fig:barrier} shows an example.  There is a queue
-for each of the \texttt{N} threads.
-Before thread $i$ enters a round, it first sends a message to every
-other thread and then waits until it receives a message from
-every other thread.
-In this simple case, each message contains \texttt{None}, but in practice useful
-information may be exchanged between the threads.
-
-The \textit{round} array is kept to check the correctness of this
-approach.  Each thread increments its entry every time it enters
-a round.  If the algorithm is correct, it can never be that two threads
-are more than one round apart.
-
-\section*{Exercises}
-\begin{problems}
-\item See if you can implement barrier synchronization for \texttt{N} threads
-with just three binary semaphores.  Busy waiting is not allowed.  How about just two?
-(As always, the Little Book of Semaphores~\cite{Downey09} is a good resource
-for solving synchronization problems with semaphores.
-Look for the \emph{double turnstile}
-\index{double turnstile}%
-solution.)
-\item Implement barrier synchronization with Mesa condition variables.
-(You may want to use a double turnstile approach here as well.)
 \end{problems}
@@
 
-There are at least two reasons why Harmony is not object-oriented.  First,
+There are at least two reasons why Harmony is not object-oriented.
+First,
 object-orientation often adds layers of indirection that would make it
@@
 reference to a lock object.  In Harmony, a lock variable contains the value
-of the lock itself.  Second, many concurrency solutions in the literature are
+of the lock itself.
+Thus, the following statement means something quite different in Python and
+Harmony:
+
+\begin{code}
+\begin{verbatim}
+x = y = Lock();
+\end{verbatim}
+\end{code}
+
+In Python, this creates two variables $x$ and $y$ referring to the same lock.
+In Harmony, the two variables will be two different locks.  If you want two
+variable refering to the same lock in Harmony, you would write:
+
+\begin{code}
+\begin{verbatim}
+x = y = malloc(Lock());
+\end{verbatim}
+\end{code}
+
+The second reason for Harmony not being object-oriented is that
+many concurrency solutions in the literature are
 expressed in C or some other low-level language that does not support

### 2020-11-29 book/paper.tex
Latest
Fixed implementation of =>
Starting to add generalized barrier synchronization

@@
 \end{code}
-\caption{\harmonylink{code/barrier.hny} Barrier synchronization with queues.}
+\caption{\harmonylink{code/qbarrier.hny} Barrier synchronization with queues.}
 \label{fig:barrier}
@@
 are more than one round apart.
+
+\begin{figure}
+\begin{code}
+\harmonysource{barrier}
+\end{code}
+\caption{\harmonylink{code/barrier.hny} Generalized barrier synchronization.}
+\label{fig:genbarrier}
+\end{figure}
 
@@
 \texttt{$e$ or $e$ or ...} & two or more booleans \\
+\hline
+\texttt{$e$ => $e$} & two booleans (implication) \\
 \hline
@@
 evaluates to \texttt{True}.
+A sequence of comparison operations, such as $a < b < c$, is evaluated left to right
+but stops as soon as one of the comparisons fails.
 
@@
 For those familiar with logic gates, \textbf{not and} is the equivalent of \texttt{NAND}.
+Similarly, \textbf{not} \texttt{$=>$} is non-implication.
 

### 2020-12-01 book/paper.tex
Latest

@@
 
+More generally, barrier synchronization can be abstracted as follows.
+We want to create a \texttt{Barrier}$(n)$ object, with operations
+\texttt{enter()} and \texttt{exit()}.  It is helpful to use
+a roller coaster car with $n$ seats as a metaphor:
+\begin{itemize}
+\item the car cannot contain more than $n$ people;
+\item the car won't take off until $n$ people are in the car;
+\item no new people can enter the car until all $n$ people have left it.
+\end{itemize}
+Notice there are two different waiting conditions:
+\begin{enumerate}
+\item waiting for the car to empty out;
+\item waiting for the car to fill up.
+\end{enumerate}
+Similarly, \texttt{enter}() blocks until $n$ threads have invoked it.
+If more threads try to \texttt{enter}(), they will block until the
+previous $n$ threads have invoked \texttt{exit}().
+
+\autoref{fig:genbarrier} shows the code.
+Method~\texttt{enter}() has two \texttt{wait} loops, one for each
+waiting condition.
+This is sometimes called \emph{double turnstile}.
+Each loop uses a different condition variable.
+The first loop waits while the barrier is fully occupied, while
+the second waits for the barrier to fill up.
+The $n^\mathit{th}$ thread to arrive at the barrier notifies the
+$n-1$ threads that are waiting in the second loop.
+
+Method \texttt{exit}() counts how many threads have left the
+barrier.  When the last one leaves, it notifies all the threads
+that are waiting in the first loop.
+
 \section*{Exercises}
@@
 solution.)
-\item Implement barrier synchronization with Mesa condition variables.
-(You may want to use a double turnstile approach here as well.)
 \end{problems}

### 2020-12-06 harmony.py
Latest

@@ -489,6 +489,7 @@ def isreserved(s):
         "elif",
         "else",
         "end",
+        "except",
         "False",
         "fun",
         "for",
@@ -510,15 +511,16 @@ def isreserved(s):
         "not",
         "or",
         "pass",
-        "processes",
         "setintlevel",
         "spawn",
         "stop",
         "such",
         "that",
         "trap",
+        "try",
         "True",
         "while"
+        "with"
     }
 
 def isname(s):
@@ -527,7 +529,7 @@ def isname(s):
 
 def isunaryop(s):
     return s in { "!", "-", "~", "all", "any", "atLabel", "bagsize", "choose",
-    "min", "max", "nametag", "not", "keys", "hash", "len", "processes" }
+    "min", "max", "nametag", "not", "keys", "hash", "len" }
 
 def isxbinop(s):
     return s in {
@@ -710,6 +712,21 @@ def strValue(v):
             return ".0x%02X"%ord(v[0])
     assert False, v
 
+def jsonValue(v):
+    if isinstance(v, Value):
+        return v.jdump()
+    if isinstance(v, bool):
+        return '{ "type": "bool", "value": "%s" }'%str(v)
+    if isinstance(v, int) or isinstance(v, float):
+        return '{ "type": "int", "value": "%s" }'%str(v)
+    if isinstance(v, str):
+        if isname(v):
+            return '{ "type": "atom", "value": "%s" }'%str(v)
+        else:
+            assert len(v) == 1, v
+            return '{ "type": "char", "value": "02X" "'%ord(v[0])
+    assert False, v
+
 def strVars(v):
     assert isinstance(v, DictValue)
     result = ""
@@ -733,6 +750,9 @@ class Value:
     def __str__(self):
         return self.__repr__()
 
+    def jdump(self):
+        assert False
+
 class PcValue(Value):
     def __init__(self, pc):
         self.pc = pc
@@ -748,6 +768,9 @@ class PcValue(Value):
 
     def key(self):
         return (3, self.pc)
+
+    def jdump(self):
+        return '{ "type": "pc", "value": "%d" }'%self.pc
 
 class DictValue(Value):
     def __init__(self, d):
@@ -769,6 +792,15 @@ class DictValue(Value):
                 result += ", ";
             result += strValue(k) + ":" + strValue(self.d[k])
         return "dict{ " + result + " }"
+
+    def jdump(self):
+        result = ""
+        keys = sorted(self.d.keys(), key=keyValue)
+        for k in keys:
+            if result != "":
[…]

### 2020-12-09 harmony.py
Latest

@@ -528,8 +528,8 @@ def isname(s):
                     all(isnamechar(c) for c in s)
 
 def isunaryop(s):
-    return s in { "!", "-", "~", "all", "any", "atLabel", "bagsize", "choose",
-    "min", "max", "nametag", "not", "keys", "hash", "len" }
+    return s in { "!", "-", "~", "abs", "all", "any", "atLabel", "bagsize",
+        "choose", "min", "max", "nametag", "not", "keys", "hash", "len" }
 
 def isxbinop(s):
     return s in {
@@ -1878,6 +1878,10 @@ class NaryOp(Op):
                 if not self.checktype(state, context, sa, isinstance(e, bool)):
                     return
                 context.push(not e)
+            elif op == "abs":
+                if not self.checktype(state, context, sa, isinstance(e, int)):
+                    return
+                context.push(abs(e))
             elif op == "atLabel":
                 if not context.atomic:
                     context.failure = "not in atomic block: " + str(self.op)

### 2020-12-09 harmony.py
got rid of badsize

@@ -479,7 +479,6 @@ def isreserved(s):
         "atLabel",
         "atomic",
         "await",
-        "bagsize",
         "call",
         "choose",
         "const",
@@ -528,7 +527,7 @@ def isname(s):
                     all(isnamechar(c) for c in s)
 
 def isunaryop(s):
-    return s in { "!", "-", "~", "abs", "all", "any", "atLabel", "bagsize",
+    return s in { "!", "-", "~", "abs", "all", "any", "atLabel",
         "choose", "min", "max", "nametag", "not", "keys", "hash", "len" }
 
 def isxbinop(s):
@@ -1951,10 +1950,6 @@ class NaryOp(Op):
                 context.push(SetValue(set(e.d.keys())))
             elif op == "hash":
                 context.push((e,).__hash__())
-            elif op == "bagsize":
-                if not self.checktype(state, context, sa, isinstance(e, DictValue)):
-                    return
-                context.push(sum(e.d.values()))
             else:
                 assert False, self
         elif self.n == 2:

### 2020-12-09 ops.c
Added ~

@@ -1471,6 +1471,14 @@ uint64_t f_intersection(struct state *state, struct context *ctx, uint64_t *args
     return result;
 }
 
+uint64_t f_invert(struct state *state, struct context *ctx, uint64_t *args, int n){
+    assert(n == 1);
+    int64_t e = args[0];
+    assert((e & VALUE_MASK) == VALUE_INT);
+    e >>= VALUE_BITS;
+    return ((~e) << VALUE_BITS) | VALUE_INT;
+}
+
 uint64_t f_isEmpty(struct state *state, struct context *ctx, uint64_t *args, int n){
     assert(n == 1);
     uint64_t e = args[0];
@@ -1878,6 +1886,7 @@ struct op_info op_table[] = {
 struct f_info f_table[] = {
 	{ "+", f_plus },
 	{ "-", f_minus },
+	{ "~", f_invert },
 	{ "*", f_times },
 	{ "/", f_div },
 	{ "//", f_div },

### 2020-12-27 book/paper.tex
Latest
got rid of badsize
Added ~

@@
 
-\chapter{List of Values}
+\chapter{Value Types}
 \label{ap:values}
@@
 times it occurs in the bag.
+The syntax is \texttt{bag\{ $e$, $e$, ... \}}.
 
@@
 
+Generic operators on Harmony values include:
+
+\begin{center}
+\begin{tabular}{|l|l|}
+\hline
+\texttt{$e$ == $e$} & equivalence \\
+\hline
+\texttt{$e$ != $e$} & inequivalence \\
+\hline
+\texttt{$e$ < $e$, $e$ <= $e$, $e$ > $e$, $e$ >= $e$} & comparison\\
+\hline
+\end{tabular}
+\end{center}
+
+\section{Booleans}
+
+The Boolean type has only two possible values: \texttt{False} and
+\texttt{True}.  Unlike in Python, in Harmony booleans are distinct
+from integers, and in particular $\mathtt{False} < 0$.  In statements
+and expressions where booleans are expected, it is not possible to
+substibute values of other types.
+
+Operations on booleans include:
+\begin{center}
+\begin{tabular}{|l|l|}
+\hline
+\texttt{$e$ and $e$ and ...} & conjuction \\
+\hline
+\texttt{$e$ or $e$ or ...} & disjunction \\
+\hline
+\texttt{$e$ => $e$, $e$ not => $e$} & implication \\
+\hline
+\texttt{not $e$} & negation \\
+\hline
+\texttt{$v$ if $e$ else $v'$} & $v$ or $v'$ depending on $e$ \\
+\hline
+\texttt{any $s$, all $s$} & disjunction / conjunction for set $s$ \\
+\hline
+\end{tabular}
+\end{center}
+
+\section{Integers}
+
+The integer type supports any whole number, as well as $-\mathtt{inf}$
+and $\mathtt{inf}$.  In the Python-based model checker, integers are
+infinite precision.  In the C-based model checker, integers are implemented
+by two's complement 61-bit words, and $-\mathtt{inf}$ is represented by
+the minimal integer and $\mathtt{inf}$ is represented by the maximal integer.
+
+Operations on integers include:
+
+\begin{center}
+\begin{tabular}{|l|l|}
+\hline
+\texttt{-$e$} & negation \\
+\hline
+\texttt{abs($e$)} & absolute value \\
+\hline
+\texttt{$e$ + $e$ + ...} & sum \\
+\hline
+\texttt{$e$ - $e$} & difference \\
+\hline
+\texttt{$e$ * $e$ * $e$ ...} & product \\
+\hline
+\texttt{$e$ / $e$, $e$ // $e$} & integer division \\
+\hline
+\texttt{$e$ \% $e$, $e$ mod $e$} & integer division remainder \\
+\hline
+\texttt{$e$ ** $e$} & power \\
+\hline
+\texttt{\string~$e$} & binary inversion \\
+\hline
+\texttt{$e$ \& $e$ \& ...} & binary and \\
+\hline
+\texttt{$e$ | $e$ | ...} & binary or \\
+\hline
+\texttt{$e$ \string^ $e$ \string^ ...} & binary exclusive or \\
+\hline
+\texttt{$e << e$} & binary shift left \\
+\hline
+\texttt{$e >> e$} & binary shift right \\
+\hline
+\texttt{$\{ e..e' \}$} & set of integers from $e$ to $e'$ inclusive \\
+\hline
+\end{tabular}
+\end{center}
+
+\section{Atoms}
+
+Atoms are essentially names and consist of one or more unicode characters.
+If they do not contain special characters and do not start with a digit,
+then an atom can be represented by a ``.'' followed by the characters.
+For example, \texttt{.hello} is
+a representation of the atom ``hello''.  A special character can be
+represented by \texttt{.0xXX}, where \texttt{XX} is the hexidecimal
+unicode for the character.
+Atoms should not be confused with strings.
+There are no special operations on atoms.
+(In the future, operators may be introduced that convert between strings
+and atoms.)
+
+\section{Set}
+
+In Harmony you can create a set of any collection of Harmony values.
+Its syntax is \texttt{\{ $v_0, v_1, ...$ \}}.
+Python users: note that in Harmony the empty set is denoted as $\{\}$.
+
+The \texttt{set} module (\autoref{ap:set})
+contains various convenient routines that operate on sets.
+Native operations on sets include:
+
+\begin{center}
+\begin{tabular}{|l|l|}
+\hline
+\texttt{len $e$} & cardinality \\
+\hline
+\texttt{$e$ - $e$} & set difference \\
+\hline
+\texttt{$e$ \& $e$ \& ...} & intersection \\
+\hline
+\texttt{$e$ | $e$ | ...} & union \\
+\hline
+\texttt{$e$ \string^ $e$ \string^ ...} & inclusion/exclusion (elements in odd number of sets) \\
+\hline
+\texttt{choose $e$} & select an element (Harmony will try all) \\
+\hline
+\texttt{min $e$} & minimum element \\
+\hline
+\texttt{max $e$} & maximum element \\
+\hline
+\end{tabular}
+\end{center}
+
+\section{Dictionary}
+
+A dictionary maps a set of values (known as \emph{keys})
+to another set of values.
+The generic syntax of a dictionary is
+\texttt{dict\{ $k_0:v_0, k_1:v_1, ...$ \}}.
+If there are duplicate keys in the list, then only the one with
+the maximum value survives.
+Therefore the order of the keys in the dictionary does not matter.
+
+There are various special cases of dictionaries,
+including lists, tuples, strings,
+and bags (multisets) that are individually described below.
+
+Operations on dictionaries include the following:
+
+\begin{center}
+\begin{tabular}{|l|l|}
+\hline
+\texttt{$d~k$} & indexing \\
+\hline
+\texttt{len $d$} & the number of keys in $d$ \\
+\hline
+\texttt{keys $d$} & the set of keys in $d$ \\
+\hline
+\texttt{$v$ [not] in keys $d$} & check if $v$ is a \emph{key} in $d$ \\
+\hline
+\texttt{$v$ [not] in $d$} & check if $v$ is a \emph{value} in $d$ \\
+\hline
+\texttt{$d$ \& $d$ \& ...} & dictionary intersection \\
+\hline
+\texttt{$d$ | $d$ | ...} & dictionary union \\
+\hline
+\end{tabular}
+\end{center}
+
+Because in Harmony brackets are used for parsing purposes only,
+you can write $d[k]$ (or $d(k)$) instead of $d~k$.
+However, if $k$ is an atom, like \texttt{.id}, then you might
+prefer the notation $k\mathtt{.id}$.
+
+Python users beware: the Harmony \texttt{$v$ in $d$} operator checks
+if there is some key $k$ such that $d[k] = v$.  In Python, the same
+syntax checks if $v$ is a key in $d$.  The difference exists because
+in Harmony a list is a special case of a dictionary.
+
+Dictionary intersection and dictionary union are defined so that
+they work well with bags.  With disjoint dictionaries, intersection
+and union work as expected.  If there is a key in the intersection, then
+dictionary intersection retains the minimum value while dictionary
+union retains the maximum value.  Unlike Python, Harmony
+dictionary intersection and union are commutative and associative.
+
+\section{List or Tuple}
+
+In Harmony, there is no distinction between a list or a tuple.
+Both are a special case of dictionary.  In particular, a list of
+$n$ values is represented by a dictionary that maps the integers
+$0, ..., n-1$ to those values.  Hence, if $t$ is a list or tuple,
+then the notation $t[0]$ returns the first element of the list.
+
+You can denote a list by a sequence of values, each value terminated
+by a comma.  As per usual, you can use brackets or parentheses
+at your discretion.  For Python users, the important thing to note
+is that a singleton lists in Harmony must contain a comma.  For
+example $[1,]$ is a list containing the value $1$, while $[1]$
+is simply the value $1$.
+
+The \texttt{list} module (\autoref{ap:list}) contains various
+convenient routines that operate on lists or tuples.
+Native operations on lists or tuples include the following:
+
+\begin{center}
+\begin{tabular}{|l|l|}
+\hline
+\texttt{$t~k$} & indexing \\
+\hline
+\texttt{$t$ + $t$ + ...} & concatenation \\
+\hline
+\texttt{$t$ * $n$} & $n$ copies of $t$ concatenated \\
+\hline
+\texttt{$v$ [not] in $t$} & check if $v$ is a \emph{value} in $t$ \\
+\hline
+\texttt{len $t$} & the length of $t$ \\
+\hline
+\texttt{$t$ \& $t$ \& ...} & pairwise list minimum \\
+\hline
+\texttt{$t$ | $t$ | ...} & pairwise list maximum \\
+\hline
+\end{tabular}
+\end{center}
+
+\section{String}
+
+In Harmony, a string is a list of single-character atoms.
+The Python string notations mostly work in Harmony as well.
+For example, \texttt{'abc'} is a three-character string
+consisting of the atoms \texttt{.a}, \texttt{.b}, and
+\texttt{.c}.
+\texttt{"abc'} is the same three-character string.
+Multi-line strings can be terminated by triple quotes or
+triple double-quotes.
+
+\section{Bag or Multiset}
+
+Harmony has native support for bags (aka multisets), which are sets
+that can contain multiple copies of the same element.
+It is represented by a dictionary that maps each element to its
+multiplicity.
+So \texttt{bag\{ 10, 10, 12 \} = dict\{ 10: 2, 12: 1 \}}.
+
+The \texttt{bag} module (\autoref{ap:bag}) contains various
+convenient routines that operate on lists or tuples.
+Native operations on bags include the following:
+
+\begin{center}
+\begin{tabular}{|l|l|}
+\hline
+\texttt{$v$ [not] in keys $b$} & check if $v$ is in $b$ \\
+\hline
+\texttt{$t$ \& $t$ \& ...} & bag intersection\\
+\hline
+\texttt{$t$ | $t$ | ...} & bag union \\
+\hline
+\end{tabular}
+\end{center}
+
+\section{Program Counter}
+
+A program counter is an integer that can be used to index into Harmony
+bytecode.  When you define a method or a lambda function, you are
+creating a constant of the program counter type.
+Note, however, that labels are currently just integers.  This may
+change in the future.
+
+\section{Address}
+
+In Harmony, each shared variable has an address, which is essentially
+a list of Harmony values.
+For example, the address of variable \texttt{d.v[3]} consists of the
+list \texttt{.d}, \texttt{.v}, and 3.  The only way to construct an
+address in Harmony is using the \texttt{?} operator.
+\texttt{?d.v[3]} returns the address of variable \texttt{d.v[3]}.
+An address can be dereferenced to return the value of the variable.
+If $a$ is an address, then $!a$ returns the value.
+Like C, Harmony supports the shorthand \texttt{$a$->v} for 
+the expression \texttt{($!a$).v}.
+
+\section{Context}
+
+A context value captures the state of a thread.
+A context is itself composed over various Harmony values.
+Some of these values can be extracted.  In particular, if $c$
+is a context, then:
+
+\begin{center}
+\begin{tabular}{|l|l|}
+\hline
+\texttt{\texttt{c}.name} & returns the name of the thread's main method \\
+\hline
+\texttt{\texttt{c}.entry} & returns the program counter of the thread's main method \\
+\hline
+\texttt{\texttt{c}.arg} & returns the argument value of the thread's main method \\
+\hline
+\texttt{\texttt{c}.this} & returns the thread-local state \\
+\hline
+\texttt{\texttt{c}.mode} & returns the \emph{mode} of the context \\
+\hline
+\end{tabular}
+\end{center}
+
+The modes of a context are as follows:
+
+\begin{center}
+\begin{tabular}{|l|l|}
+\hline
+\texttt{normal} & context of a executing thread \\
+\hline
+\texttt{stopped} & context of a stopped thread \\
+\hline
+\texttt{terminated} & context of a terminated thread \\
+\hline
+\texttt{failed} & context of a failed thread \\
+\hline
+\end{tabular}
+\end{center}
+
 \chapter{List of Operators}
@@
 \hline
-\texttt{-$e$, \string~$e$} & an integer \\
+\texttt{-$e$, \string~$e$, abs($e$)} & an integer \\
 \hline
@@
 \hline
-\texttt{$e$ \& $e$ \& ...} & two or more integers or sets \\
-\hline
-\texttt{$e$ | $e$ | ...} & two or more integers or sets \\
-\hline
-\texttt{$e$ \string^ $e$ \string^ ...} & two or more integers or sets \\
+\texttt{$e$ \& $e$ \& ..., $e$ | $e$ | ..., $e$ \string^ $e$ \string^ ...} & two or more integers or sets \\
 \hline
@@
 \hline
-\texttt{bagsize $e$} & a bag \\
-\hline
 \texttt{atLabel $e$} & atom (corresponding to a label) \\
@@
 \texttt{nametag $e$} & $e$ must be \texttt{()} \\
-\hline
-\texttt{threads $e$} & $e$ must be \texttt{()} \\
 \hline
@@
 \section{The \texttt{bag} module}
+\label{ap:bag}
 
@@
 \section{The \texttt{set} module}
+\label{ap:set}
 

### 2020-12-27 book/paper.tex
Renamed process to thread

@@
 or a \emph{released} state.
-Processes trying to acquire an already acquired binary semaphore
+Threads trying to acquire an already acquired binary semaphore
 have to wait until the binary semaphore is released.
@@
 check for each waiting gate if its waiting condition hold and if there are
-processes trying to get through the gate.  If there is such a gate,
+threads trying to get through the gate.  If there is such a gate,
 it must select one and open that gate.  If there is no such gate,
@@
 
-The modes of a context are as follows:
+The possible modes of a context are as follows:
 

### 2020-12-27 book/paper.tex
Working on revision

@@
 It increments \textit{count} and sets \textit{done}[\textit{self}] to \texttt{True}.
-Method \texttt{main} waits until all \textit{done} flags are set to \texttt{True}.
+Method \texttt{main} waits until both \textit{done} flags are set to \texttt{True}.
 (\textbf{await} $c$ is semantically the same as \textbf{while not} $c$: \textbf{pass}.)
@@
 reported if the condition fails, in this case
-the actual value of \textit{count}.
+the value of \textit{count}.
 
@@
 And because Harmony tries every possible execution, it will consider
-such executions as well.
+that particular execution as well.
 
@@
 \index{interleaving}%
-in unspecified and often random ways.
+in unspecified and often unpredictable ways.
 A program is correct if it works for any interleaving of threads.
@@
 
+The exercises below have you try the same thing in Python.  As you will
+see, the bug is not easily triggered when you run a Python version of
+the program.  But in Harmony Murphy's Law applies:
+if something can go wrong, it will.  Usually that is not a good thing,
+but in Harmony it is.  It allows you to find bugs in your concurrent
+programs much more easily than with a conventional programming language.
+
 \section*{Exercises}
@@
 time---the important thing is to develop an understanding of concurrency.
+(Also, you do not get to use a \emph{lock}, yet.)
 \end{problems}
@@
 \section*{Harmony Values}
-\label{ap:harmonyvalues}
 
@@
 \end{code}
-This is unfamiliar to Python programers, but in Harmony square brackets can be used
+The meaning of $d$ $a$ $b$ $...$ is ((($d$ $a$) $b$) $...$).
+This notation is unfamiliar to Python programmers,
+but in Harmony square brackets can be used
 in the same way as parentheses, so you can express the same thing in the form
@@
 $d$\texttt{[.count]} (although both will work).
-Thus using atoms, a dictionary can be made to look much like a Python object.
-The meaning of $d$ $a$ $b$ $...$ is ((($d$ $a$) $b$) $...$).
+Thus, using atoms, a dictionary can be made to look much like a Python object.
 
@@
 As in Python, you can create singleton tuples by including a comma.
-For example, \texttt{(1,)}.
+For example, \texttt{(1,)} is a tuple consisting just of the number 1.
+Importantly, $(1) = 1 \ne (1,)$.
 
@@
 
-Appendix~\ref{ap:harmonyvalues} provides details on all the values that
+\autoref{ap:values} provides details on all the values that
 Harmony currently supports.
@@
 can spawn an unbounded number of other threads and threads may terminate.
-Each thread has an immutable (but not necessarily unique) \emph{name tag},
-\index{name tag}%
-a program counter,
+Each thread has a program counter that indexes into the code,
 \index{program counter}%
 a stack of Harmony values,
-and a single mutable general purpose \emph{register}
+and two private \emph{registers} that each contain a Harmony value.
 \index{register}%
-that contains a Harmony value.
 \end{itemize}
 
-A name tag consists of the name of the main method of the thread,
-along with an optional tag specified in the \texttt{spawn}
-\index{spawn statement}%
-\index{tag}%
-statement.
-The default tag is the argument to the method.
-A name tag is represented as a dictionary with keys \texttt{.name}
-and \texttt{.tag}, but we shall often abbreviate it using the notation
-\texttt{name/tag}.
-For example, in \autoref{fig:inc}, the created threads have name tags
-\texttt{incrementer/0}, \texttt{incrementer/1}, and \texttt{main/()}.
-
 \glsadd{context}%
-
+\index{thread-local}%
+One of the registers contains the local variables of the method that
+the thread is currently executing.  It is saved and restored by
+method invocations.  The other register is called \emph{this} and is
+used for so-called \emph{thread-local state}, which contains variables that
+can be used by all methods.
 The state of a thread is called a \emph{context} (aka \emph{continuation}):
@@
 \index{continuation}%
-it contains the values of
-its name tag, program counter, stack, and register.
+it contains the values of its program counter, stack, and registers.
 The state of a HVM
@@
 have the same context.
-The initial state of the Harmony memory is the empty dictionary, \texttt{dict\{\}}.
-The context bag has an initial context in it with name tag
-\texttt{\_\_init\_\_/()}, pc 0, register \texttt{()}, and an empty stack.
-Each machine instruction updates the state in some way.
-% \autoref{fig:incstate} shows an example of a reachable state for the program in \autoref{fig:inc}.
 
@@
 
-It may seem strange that there is only one memory location and that each
-thread has only one register.  However, this is not a limitation because
-Harmony values are unbounded trees.
-Both the memory and the register of a thread always contain
-a dictionary that maps atoms to Harmony values.  We call this a \emph{directory}.
+It may seem strange that there is only one memory location.
+However, this is not a limitation because Harmony values are unbounded trees.
+The memory contains a dictionary that maps atoms to Harmony values.
+We call this a \emph{directory}.
 \index{directory}%
@@
 \texttt{.done} is a dictionary with keys 0 and 1.
+An address is itself a Harmony value.
 
@@
 \index{stack machine}%
-but it also has a register.
+but it also has two registers.
+Like shared memory, the registers usually contain dictionaries so they
+can represent the values of multiple named variables.
+
 All instructions are atomically executed.
 Most instructions pop values from the stack or push values onto the stack.
-At first there is one thread, with name tag \texttt{\_\_init\_\_/()},
-that initializes the state.
+At first there is one thread, named \texttt{\_\_init\_\_},
+which initializes the state.
 It starts executing at instruction 0 and keeps executing until the last
@@
 
-Appendix~\ref{ap:harmonybytecode} provides a list of all HVM machine instructions,
+\autoref{ap:harmonybytecode} provides a list of all HVM machine instructions,
 in case you want to read about the details.
@@
 names of its arguments.
-The code generated from $\mathit{count} := \mathit{count} + 1$ in line~6 of
+The code generated from $\mathit{count} := \mathit{count} + 1$ in line~5 of
 \texttt{Up.hny} is as follows (see \autoref{fig:inccode}):
@@
 the shortest trace, but also the list of threads
-at that state, along with their name tags and program counters.
+at that state, along with their program counters.
 \item \texttt{Stopped States}: Similar to non-terminating states,
@@
 \begin{enumerate}
-\item The name tag of the thread;
+\item The main method and argument of the thread;
 \item The sequence of program counters of the HVM instructions that the thread executed;
@@
 If so, it uses the \texttt{head} and \texttt{tail}
-methods from the \texttt{list} module (see Appendix~\ref{ap:module})
+methods from the \texttt{list} module (see \autoref{ap:module})
 to resume the first thread that got
@@
 \item Implement a ``try lock'' module using Mesa condition variables
-(see also Exercise~\ref{ex:trylock}).  It should
+(see also \autoref{ex:trylock}).  It should
 have the following API:
@@
 Addresses are lexicographically ordered accordingly;
-\item contexts (Appendix~\ref{app:context}) are ordered first by name tag, then by program counter, then by hash.
+\item contexts (\autoref{app:context}) are ordered first by name tag, then by program counter, then by hash.
 \end{itemize}
@@
 These constants can be added to dictionaries.
-For example, in Figure~\ref{fig:petersonmethods}
+For example, in \autoref{fig:petersonmethods}
 you can add the \texttt{P\_enter} and

### 2020-12-27 book/paper.tex
Latest

@@
 \begin{verbatim}
-#states = 71 diameter = 5                                                                              ==== Safety violation ====
-P0: __init__/()   [0-4,31-43] 43 { .count: 0, .done: [False, False] }
-P1: incrementer/0 [5-8]        9 { .count: 0, .done: [False, False] }
-P2: incrementer/1 [5-15]      15 { .count: 1, .done: [False, True ] }
-P1: incrementer/0 [9-15]      15 { .count: 1, .done: [True,  True ] }
-P3: main/()       [17-27]     27 { .count: 1, .done: [True,  True ] }
+#states = 53 diameter = 5
+==== Safety violation ====
+T0: __init__() [0-4,31-46] 46 { count: 0, done[0]: False, done[1]: False }
+T1: incrementer(0) [5-8]    9 { count: 0, done[0]: False, done[1]: False }
+T2: incrementer(1) [5-15]  15 { count: 1, done[0]: False, done[1]: True  }
+T1: incrementer(0) [9-15]  15 { count: 1, done[0]: True,  done[1]: True  }
+T3: main()         [17-27] 27 { count: 1, done[0]: True,  done[1]: True  }
 >>> Harmony Assertion (file=code/Up.hny, line=10) failed: 1
-Open file:///Users/rvr/github/harmony.new/harmony.html for more information
+Open file:///Users/rvr/github/harmony/harmony.html for more information
 \end{verbatim}
@@
 The program got to the faulty execution in 5 ``steps.''
-The output has four columns:
+The output has five columns:
 \begin{enumerate}
+\item A thread identifier;
 \item The main method and argument of the thread;

### 2020-12-27 book/paper.tex
Latest

@@
 T0: __init__() [0-4,31-46] 46 { count: 0, done[0]: False, done[1]: False }
-T1: incrementer(0) [5-8]    9 { count: 0, done[0]: False, done[1]: False }
-T2: incrementer(1) [5-15]  15 { count: 1, done[0]: False, done[1]: True  }
-T1: incrementer(0) [9-15]  15 { count: 1, done[0]: True,  done[1]: True  }
+T1: incrementer(0) [ 5- 8]  9 { count: 0, done[0]: False, done[1]: False }
+T2: incrementer(1) [ 5-15] 15 { count: 1, done[0]: False, done[1]: True  }
+T1: incrementer(0) [ 9-15] 15 { count: 1, done[0]: True,  done[1]: True  }
 T3: main()         [17-27] 27 { count: 1, done[0]: True,  done[1]: True  }
@@
 \begin{figure}
-\includegraphics[width=\textwidth]{figures/Up1.pdf}
+\includegraphics[width=\textwidth]{figures/Up1.png}
 \caption{The
@@
 \begin{itemize}
-\item Thread \texttt{incrementer/0}
+\item Thread \texttt{incrementer(0)}
 executed instructions 5 through 8, loading the value of
 \textit{count} but stopping just before storing 1 into \textit{count};
-\item Thread \texttt{incrementer/1}
+\item Thread \texttt{incrementer(1)}
 executed instructions 5 through 15, storing 1 into
 into \textit{count} and storing \texttt{True} into \textit{done}[1];
-\item Thread \texttt{incrementer/0}
+\item Thread \texttt{incrementer(0)}
 continues execution, executing instructions 9 through 15
@@
 \hline
-name tag & a dictionary with atoms \texttt{.name} and \texttt{.tag} \\
+name & the name of the main method that the thread is executing \\
+\hline
+argument & the argument given to the main method \\
 \hline
@@
 \hline
-register & a dictionary mapping atoms (names of variables) to values \\
+method variables & a dictionary mapping atoms (names of method variables) to values \\
+\hline
+thread-local variables & a dictionary mapping atoms (names of thread-local variables) to values \\
 \hline
@@
 \begin{itemize}
-\item The name in a name tag is the name of the method that the thread
-is executing;
-\item The tag in a name tag the argument to the method by default.  Optionally
-\texttt{spawn} allows the tag to be specified explicitly;
 \item The frame pointer points to the current \emph{stack frame},
@@
 
-There is initially a single context with nametag
+There is initially a single context with name
 \texttt{\_\_init\_\_/()} and program counter 0.  It starts executing

### 2020-12-28 book/paper.tex
Latest

@@
 To understand the problem of concurrent computing, it
-is important to have a basic understanding of the HVM.
+is important to have a basic understanding of machine instructions,
+and in our case those of the HVM.
 
@@
 A Harmony program is translated into HVM bytecode.
-The HVM is not an ordinary virtual machine, but its architecture
+To make it amenable to efficient model checking,
+the HVM is not an ordinary virtual machine, but its architecture
 is nonetheless representative of conventional computers and
@@
 \index{thread-local}%
-One of the registers contains the local variables of the method that
-the thread is currently executing.  It is saved and restored by
+One of the registers of a thread contains the local variables of the
+method that the thread is currently executing.  It is saved and restored by
 method invocations.  The other register is called \emph{this} and is
@@
 However, this is not a limitation because Harmony values are unbounded trees.
-The memory contains a dictionary that maps atoms to Harmony values.
+The shared memory is a dictionary that maps atoms (names of shared
+variables) to other Harmony values.
 We call this a \emph{directory}.
 \index{directory}%
-A directory represents the state of a collection of variables named by the atoms.
+Thus, a directory represents the state of a collection of variables named
+by the atoms.
 %
@@
 Each node in a directory tree is then identified
-by a sequence of atoms, like a path name in the file system hierarchy.  We call
-such a sequence the \emph{address}
+by a sequence of Harmony values, like a path name in the file system
+hierarchy.  We call such a sequence an \emph{address}.
 \index{address}%
-of a Harmony value, and it is relative to the
-``root'' of the directory tree.
 For example, in \autoref{fig:inc} the memory is a dictionary with two
@@
 \texttt{.done} is a dictionary with keys 0 and 1.
+So, for example, the address of \texttt{done[0]} is the sequence
+[ \texttt{.done}, 0 ].
 An address is itself a Harmony value.
@@
 can represent the values of multiple named variables.
+As mentioned before, one register contains a dictionary with the local
+variables of the method that the thread is currently executing,
+while the other contains thread-local state.
 
@@
 \textit{count} variable onto the stack.
-\item the \texttt{Push} instruction pushes the constant 1
+\item The \texttt{Push} instruction pushes the constant 1
 onto the stack of the thread.
@@
 where $L$
-is the atom containing the name of the label (in this case, \texttt{.cs}).
-The operator returns a bag (multiset) of name tags of threads executing at that
-label.  The bag is represented by a dictionary that maps each element
+is the atom containing the name of the label (in this case, \texttt{cs}).
+The operator returns a bag (multiset) of (method, argument) pairs of threads
+executing at that label.
+In our case there are two threads: \texttt{(thread, 0)} and \texttt{(thread, 1)}.
+A bag is represented by a dictionary that maps each element
 in the bag to the number of times the element appears in the bag.
@@
 use it in normal code.
-The assertion also makes use of the \textbf{nametag}() operator
-\index{nametag operator}%
-that returns the name tag of the current thread.
 If you run the code through Harmony, the assertion should fail because
@@
 mutual exclusion for an algorithm such as Peterson's.
-For that, we have to understand a little bit more about how the Harmony
-virtual machine (HVM) works.
+It turns out that doing so is not easy.
+If you are interested in learning more about concurrent programming
+but not necessarily in how to prove concurrent programs correct, you
+may choose to skip the rest of this chapter.
+If you are still here, you have to understand a little bit more about
+how the Harmony virtual machine (HVM) works.
 In \autoref{ch:harmonymachine} we talked about the concept of \emph{state}:
@@
 \index{non-determinism}%
-in Harmony.
+in Harmony.\footnote{Actually, there is a third type called
+an \emph{interrupt}, discussed in \autoref{ch:interrupts}.}
 One is when
@@
 But we do not have such a predicate: we know how to define the set
-of reachable states, but given an arbitrary state it is not easy to
-see whether it is reachable or not.
+of reachable states inductively, but given an arbitrary state it is not
+easy to see whether it is reachable or not.
 
@@
 \index{inductive invariant}%
-An inductive invariant $\mathcal{I}$ is a predicate over states that satisfies the following:
+An inductive invariant~$\mathcal{I}$ is a predicate over states that satisfies the following:
 \begin{itemize}
@@
 by Harmony),
-but not an inductive invariant (it will turn out that $s$ is not reachable).
+but not an inductive invariant.
+It will turn out that $s$ is not reachable.
 
@@
 Although it does not, by itself, imply exclusivity, we can show that
-$\mathcal{F}(i)$ is an inductive invariant (for both threads 1 and 2).
+$\mathcal{F}(i)$ is an inductive invariant (for both threads~0 and~1).
 To wit, it holds in the initial state, because in the initial state thread $i$ does
@@
 \hline
-Assert & pop $b$ and check that it is \texttt{True} \\
-\hline
-AtomicInc & increment the atomic counter of this context \\
-\hline
-AtomicDec & decrement the atomic counter of this context \\
+Assert, Assert2 & pop $b$ and check that it is \texttt{True}.  Assert2 also pops value to print \\
+\hline
+AtomicInc/Dec & increment/decrement the atomic counter of this context \\
+\hline
+Bag & pop $n$ and $n$ values and push a bag \\
 \hline
@@
 \hline
-ReadonlyInc & increment the read-only counter of this context \\
-\hline
-ReadonlyDec & decrement the read-only counter of this context \\
+ReadonlyInc/Dec & increment/decrement the read-only counter of this context \\
 \hline

### 2020-12-28 book/paper.tex
Latest

@@
 For this, as mentioned in \autoref{ch:harmonymachine},
-each shared variable has an \emph{address}.
+each shared variable has an \emph{address}, itself a Harmony value.
 \index{address}%
@@
 Often, pointers point to dictionaries, and so if $p$ is such a pointer,
-then $!(p).\mathtt{field}$ would evaluate to the specified field in the dictionary.
+then $(!p).\mathtt{field}$ would evaluate to the specified field in the dictionary.
 Note that the parentheses in this expression
@@
 \texttt{!}(\textit{p}\texttt{.field}).
-$!(p).\mathtt{field}$ is such a common expression that, like C, Harmony supports the
+$(!p).\mathtt{field}$ is such a common expression that, like C, Harmony supports the
 shorthand $p-$$>$\texttt{field}, which greatly improves readability.
@@
 \index{module}%
-This would
-make the code re-usable by other applications.
+This would make the code usable by multiple applications.
 
@@
 algorithm gets around the problem, it is not efficient, especially
-if generalized to multiple threads.  Instead, multi-core processors provide
+if generalized to multiple threads.
+Worse, most modern processors make accessing shared memory locations
+non-atomic to allow for efficient caching and other optimizations.
+So Peterson's algorithm would not even work on such processors.
+Instead, multi-core processors provide
 so-called \emph{interlock instructions}:
@@
 
-Many CPUs have an atomic ``test-and-set'' (TAS)
+To support implementing locks,
+many CPUs have an atomic ``test-and-set'' (TAS)
 \index{test-and-set}%

### 2020-12-28 book/paper.tex
Latest

@@
 critical section at the same time.
-
+For those who read through the previous chapter:
 $\mathcal{I}_1 \land \forall i: \mathcal{I}_2(i)$ is an inductive invariant.
+
 To see that invariant $\mathcal{I}_1$ is maintained, note that
@@
 
-\begin{figure}
-\begin{code}
-\harmonysource{spinlockInv}
-\end{code}
-\caption{\harmonylink{code/spinlockInv.hny} Checking invariants.}
-\label{fig:tasinv}
-\end{figure}
-
-Harmony can check these invariants as well.  \autoref{fig:tas} already
-has the code to check $\mathcal{I}_2(i)$.  But how would one go about checking an
+Harmony can check these invariants as well.
+$\mathcal{I}_2(i)$ is checked by the \texttt{assert} statement.
+But how would one go about checking an
 invariant like $\mathcal{I}_1$?  Invariants must hold for every state.
@@
 been initialized).
-
-We can do this by adding another thread that
-checks the invariant.  \autoref{fig:tasinv} shows the code.
-Method \texttt{checkInvariant()} checks to see if the invariant holds
-in a state.  It introduces a new feature of Harmony: the ability to have
-variables local to a method.  In this case, the thread variable \textit{sum}
-is used to compute the number of shared variables that have value
-\texttt{False}.
-The method is invoked by a thread that runs alongside
-the other threads.
-In Harmony, \textbf{assert} statements are executed atomically, so the
-evaluation of the assertion is not interleaved with the execution
-of other threads.
-Because Harmony tries every possible execution, the thread is guaranteed
-to find violations of the invariant if it does not hold.
+Harmony supports checking such invariants using the \texttt{invariant}
+keyword.
+The expression counts the number of \texttt{False} values and checks that
+the result is less than or equal to 1.
+Harmony checks the expression in every reachable state.
 
@@
 write out the invariants that need to hold and check them using Harmony.
-\item People who use an ATM often first check their balance and then withdraw
-a certain amount of money not exceeding their balance.  A negative balance
-is not allowed.  \autoref{fig:atm} shows two operations on bank accounts:
-one to check the balance and one to withdraw money.
-Note that all operations on accounts are carefully protected by a lock
-(i.e., there are no data races).
-The \texttt{customer}
-method models going to a particular ATM and withdrawing money not exceeding
-the balance.  Method \texttt{checker} looks for negative balances.
-Running the code through Harmony reveals that there is a bug.
-It is a common type of concurrency bug known as \emph{Time Of Check Time Of
-Execution} (TOCTOE).
-\index{Time Of Check Time Of Execution}
-\index{TOCTOE}
-In this case, by the time the withdraw operation is performed,
-the balance can have changed.
-\begin{enumerate}
-\item Fix the code in \autoref{fig:atm}.  Note, you should leave the
-customer code the same. You are only allowed to change the
-\texttt{atm\_} methods, and you cannot use the \textbf{atomic}
-statement.
-\item Is it necessary to obtain a lock in \texttt{atm\_check\_balance()}?
-\end{enumerate}
 \end{problems}
-
-\begin{figure}
-\begin{code}
-\harmonysource{atm}
-\end{code}
-\caption{\harmonylink{code/atm.hny} Withdrawing money from an ATM.}
-\label{fig:atm}
-\end{figure}
 
@@
 Hint: you do not have to change the existing code.
+\item People who use an ATM often first check their balance and then withdraw
+a certain amount of money not exceeding their balance.  A negative balance
+is not allowed.  \autoref{fig:atm} shows two operations on bank accounts:
+one to check the balance and one to withdraw money.
+Note that all operations on accounts are carefully protected by a lock
+(i.e., there are no data races).
+The \texttt{customer}
+method models going to a particular ATM and withdrawing money not exceeding
+the balance.  Method \texttt{checker} looks for negative balances.
+Running the code through Harmony reveals that there is a bug.
+It is a common type of concurrency bug known as \emph{Time Of Check Time Of
+Execution} (TOCTOE).
+\index{Time Of Check Time Of Execution}
+\index{TOCTOE}
+In this case, by the time the withdraw operation is performed,
+the balance can have changed.
+\begin{enumerate}
+\item Fix the code in \autoref{fig:atm}.  Note, you should leave the
+customer code the same. You are only allowed to change the
+\texttt{atm\_} methods, and you cannot use the \textbf{atomic}
+statement.
+\item Is it necessary to obtain a lock in \texttt{atm\_check\_balance()}?
+\end{enumerate}
 \end{problems}
@@
 \label{fig:xy}
+\end{figure}
+
+\begin{figure}
+\begin{code}
+\harmonysource{atm}
+\end{code}
+\caption{\harmonylink{code/atm.hny} Withdrawing money from an ATM.}
+\label{fig:atm}
 \end{figure}

### 2020-12-29 book/paper.tex
Latest

@@
 In Harmony, a thread can suspend itself and save its context (state) in a
-list.  Recall that the context of a thread consists of its name tag,
-its program counter, and the contents of its register and stack.
+list.  Recall that the context of a thread contains
+its program counter, stack, and registers.
 A context is a regular (if complex) Harmony value.
@@
 the balance can have changed.
-\begin{enumerate}
-\item Fix the code in \autoref{fig:atm}.  Note, you should leave the
+Fix the code in \autoref{fig:atm}.  Note, you should leave the
 customer code the same. You are only allowed to change the
@@
 statement.
-\item Is it necessary to obtain a lock in \texttt{atm\_check\_balance()}?
-\end{enumerate}
 \end{problems}
@@
 
-\texttt{wait} adds the nametag of the thread to the bag.
+\texttt{wait} adds the context of the thread to the bag.
 This increments the number of threads in the bag with the same context.
@@
 \hline
-\texttt{nametag $e$} & $e$ must be \texttt{()} \\
+\texttt{get\_context $e$} & $e$ must be \texttt{()} \\
 \hline

### 2020-12-29 book/paper.tex
Latest

@@
 \item $\mathtt{put}(q, v)$: add $v$ to the tail of the queue;
-\item $r = \mathtt{get}(q)$: returns $r = ()$ if $q$ is empty or $r = (v,)$ (a singleton tuple)
-if $v$ was at the head of the queue.
+\item $r = \mathtt{get}(q)$: returns $r = \mathtt{None}$
+if $q$ is empty or $r = v$ if $v$ was at the head of the queue.
 \end{itemize}
@@
 points where items are dequeued.
-If the buffer is full, the enqueur must wait; if the buffer is empty, the
+If the buffer is full, the enqueuer must wait; if the buffer is empty, the
 dequeuer must wait.
@@
 
-Split Binary Semaphores (SBS) is a general technique to implement conditional
-waiting.  It was originally proposed by
+The Split Binary Semaphore (SBS) approach is a general technique for
+implementing conditional waiting.  It was originally proposed by
 Tony Hoare and popularized by Edsger Dijkstra~\cite{EWD703}.
@@
 An ordinary critical section has no waiting conditions.
-Reader/writer locks have two waiting conditions:
-\begin{enumerate}
-\item readers waiting for a writer to release the lock;
-\item writers waiting for another writer or all readers to release the lock.
-\end{enumerate}
-Similarly, a bounded buffer has two waiting conditions:
+For example, a bounded buffer has two waiting conditions:
 \begin{enumerate}
@@
 \end{enumerate}
-So each will require 3 binary semaphores if the SBS technique is applied.
+So it will require 3 binary semaphores if the SBS technique is applied.
 
@@
 In particular, when a thread leaves the critical section it should
-check for each waiting gate if its waiting condition hold and if there are
+check for each waiting gate if its waiting condition holds and if there are
 threads trying to get through the gate.  If there is such a gate,
@@
 
-Note that the following invariants holds:
+Note that the following invariants hold:
 \begin{itemize}
@@
 The waiting gates each consist of a pair: a counter that counts how many
-threads are waiting behind the gate, and a binary semaphore initialized
+threads are waiting behind the gate and a binary semaphore initialized
 in the acquired state (signifying that the gate is closed).
@@
 have to wait until all readers have left the critical section or until a writer has left.
-
+%
 The state of a reader/writer lock thus consists of the following:
@@
 this state.
-Each of these methods first has to acquire the \textit{mutex}
+First they have to acquire the \textit{mutex}
 (i.e., enter the main gate).
-\texttt{read\_acquire} and \texttt{write\_acquire} must check to see
+After acquiring the \textit{mutex},
+\texttt{read\_acquire} and \texttt{write\_acquire} each must check to see
 if it has to wait.
 If so, it increments the count associated with its respective gate,
-opens a gate (using \texttt{release\_one}, and then blocks until
+opens a gate (using method \texttt{release\_one}), and then blocks until
 its waiting gate opens up.
@@
 First the method acquires the \textit{mutex}.
-Then, in the case that the thread, say $p$,
+Then, in the case that the thread
 finds that there is a writer in the critical section
@@
 can acquire the bound lock at the same time.
-\item Implement a \emph{bounded stack}.  Method \texttt{push}(\textit{item})
-should block if the stack is full.  Method \texttt{pop()} should block
-until the stack is non-empty and then return an item.
-Also implement test programs for your solution.
+% \item Implement a \emph{bounded stack}.  Method \texttt{push}(\textit{item})
+% should block if the stack is full.  Method \texttt{pop()} should block
+% until the stack is non-empty and then return an item.
+% Also implement test programs for your solution.
 \item \label{ex:gpu} Implement a thread-safe \emph{GPU allocator} by modifying
@@
 a \texttt{downgrade} method to the code in \autoref{fig:RWsplitsema}.
+(Similarly, you may want to try to implement an \emph{upgrade} of a
+read lock to a write lock.)
 % How easy or hard would it be to do the same for the code in
@@
 \emph{safety property}---it prevents something ``bad'' from
-happening, like a reader and writer thread both entering the
-critical section.  The \emph{progress} property is an example
+happening, like a reader and writer thread both acquiring a
+reader/writer lock.
+The \emph{progress} property is an example
 of a \emph{liveness property}---guaranteeing that something good
@@
 thread never gets a turn.  This is an example of
-\texttt{starvation}.  With a spinlock, this scenario could
+\emph{starvation}.  With a spinlock, this scenario could
 even happen with two threads.  Initially both threads
@@
 from acquiring the lock yet again.
-
-It is worth noting that Peterson's Algorithm (\autoref{ch:peterson})
+%
+(It is worth noting that Peterson's Algorithm (\autoref{ch:peterson})
 does not suffer from starvation, thanks to the \texttt{turn} variable
 that alternates between 0 and 1 when two threads are contending for
-the critical section.
+the critical section.)
 
@@
 which complicates the code.  Also, it requires a deep understanding of
-the SBS method to remember to place a call \texttt{signal} in the
+the SBS method to remember to place a call to \texttt{signal} in the
 \texttt{read\_acquire} method that releases additional readers that
@@
 It is represented by a dictionary containing
-a bag of name tags of threads waiting on the condition variable.
+a bag of contexts of threads waiting on the condition variable.
 (The \texttt{synchS} library instead uses a list of contexts.)
@@
 Addresses are lexicographically ordered accordingly;
-\item contexts (\autoref{app:context}) are ordered first by name tag, then by program counter, then by hash.
+\item contexts (\autoref{app:context}) are ordered first by name, then by program counter, then by the remaining content.
 \end{itemize}
@@
 \texttt{spawn m $e$ [, $t$]} & \texttt{m} is a method,
-$e$ is an expression, $t$ is a tag (an expression) \\
+$e$ is an expression, $t$ is the initial thread-local state (an expression) \\
 \hline
@@
 \hline
-Spawn & pop tag, argument, and method and spawn a new context \\
+Spawn & pop initial thread-local state, argument, and method and spawn a new context \\
 \hline

### 2020-12-31 book/paper.tex
Fixing various things in the book

@@
 
-Mesa monitors provide another important and convenient option:
-operation \texttt{notifyAll}
-notifies \emph{all} threads that are waiting on a condition
-instead of just one.
-Having such an operation is not possible with Hoare monitors because using
+Mesa monitors also allow notifying multiple threads.
+For example, a thread can invoke \texttt{notify()} twice---if there are
+two or more threads waiting on the condition variable, two will be resumed.
+Operation \texttt{notifyAll}
+\index{notifyAll}%
+(aka \texttt{broadcast)})
+\index{broadcast}%
+notifies \emph{all} threads that are waiting
+on a condition.
+Signaling multiple threads is not possible with Hoare monitors because with
 Hoare monitors control
@@
 can only be done if there is just one such thread.
-\index{notifyAll}%
-(aka \texttt{broadcast)})
-\index{broadcast}%
-lets all threads in that bedroom into the hall.
 
@@
 adopted by all major programming languages.
-That said, few programming languages provide full support for monitors.
+That said, few programming languages provide full syntactical support
+for monitors, instead opting to support monitor semantics through library
+calls.
 In Java, each object has a hidden lock \emph{and} a hidden condition variable
@@
 for a section of code.
-In C and C++, support for locks and condition variables is through libraries.
+In C and C++, support for locks and condition variables is entirely
+through libraries.
 
@@
 
-Like C, Java, and Python, Harmony does not have built-in language
-support for monitors, but provides library support through the
-Harmony \texttt{synch} module.
+Like C, Java, and Python, Harmony does not have syntactical
+support for monitors. Harmony provides support for monitors
+through the Harmony \texttt{synch} module.
 \autoref{fig:cv} shows the implementation of
@@
 that protects the critical region.
-(Again, there is only one ``gate.'')
 There are two condition variables: readers wait on \textit{r\_cond} and
@@
 variable has not been notified.
-In particular, one should always be able to replace \texttt{wait}
+As a rule of thumb, one should always be able to replace \texttt{wait}
 by \texttt{unlock} followed by \texttt{lock}.
@@
 \texttt{release\_wlock} executes \texttt{notifyAll}(?\textit{r\_cond})
-as well as \texttt{notify}(?\textit{w\_cond)}.  Again, because we
+as well as \texttt{notify}(?\textit{w\_cond)}.  Because we
 do not keep track of the number of waiting readers or writers, we
@@
 either all the readers get to enter next or one of the writers gets
-to enter next.
+to enter next.  (If you want to prevent waking up both readers and
+a writer, then you can keep track of how many threads are waiting
+in the code.)
 
@@
 to replace each \texttt{notifyAll} with \texttt{notify} and see if
-every possible execution of the the application still terminates.
+every possible execution of the application still terminates.
 
@@
 A thread does not have to return all the GPUs it allocated at once.
+(You may want to try implementing this with Split Binary Semaphores.  It is not as easy.)
 \item The specification in the previous question makes the solution unfair.
@@
 this for ever, but for the purposes of this book, philosophers can leave
-the table when they're not eating.
+the table when they want to leave.
 
@@
 thread for each philosopher and a lock for each fork.  If you
-run it, Harmony complains that the execution may not terminate, with all five
-threads being blocked trying to acquire the lock.
+run it, Harmony complains that the execution may not be able to terminate,
+with all five threads being blocked trying to acquire the lock.
 
@@
 waits for other resources that it needs;
-\item \emph{No Preemption}: Resources cannot be forcibly taken away from threads that
+\item \emph{No Preemption}: resources cannot be forcibly taken away from threads that
 allocated them;
-\item \emph{Circular Wait}: There exists a directed circular chain of threads, each waiting
+\item \emph{Circular Wait}: there exists a directed circular chain of threads, each waiting
 to allocate a resource held by the next.
@@
 Preventing deadlock thus means preventing that one of these conditions occurs.
-However, mutual exclusion is not easily prevented (although, for some resources it is
+However, mutual exclusion is not easily prevented in general
+(although, for some resources it is
 possible, as demonstrated in \autoref{ch:nonblocking}).
@@
 it cannot cause deadlock.
-\item Implement an extra thread that checks, once, if the total of
-the balances is the same as that at the beginning (using an assertion).
-Is once enough to determine that no money gets lost or created, or should it do
-so continually?
+\item Add an invariant that checks that the total of the balances never
+changes.  Note that the invariant only holds if none of the locks are
+held.
 \end{problems}
@@
 \begin{code}
-\harmonysource{barrier}
+\harmonysource{qbarrier}
 \end{code}

### 2020-12-31 book/paper.tex
Latest

@@
 handles one message at a time, serializing all access to the data it owns.
-As message queues are FCFS (First-Come-First-Served), starvation is prevented.
+As message queues are FIFO (First-In-First-Out), starvation is prevented.
 
@@
 
-Note that a \texttt{Queue} behaves much like a zero-initialized semaphore.
+For those familiar with counting semaphores: note that a \texttt{Queue}
+behaves much like a zero-initialized counting semaphore.
 \texttt{put} is much like \texttt{V}, except that it is accompanied by data.
 \texttt{get} is much like \texttt{P}, except that it also returns data.
-Thus, synchronized queues can be considered a generalization of counting semaphores.
+Thus, synchronized queues can be considered a generalization of
+counting semaphores.
 
@@
 \begin{figure}
-\begin{center}
-\includegraphics[width=\textwidth]{figures/barrier-crop.pdf}
-\end{center}
-\caption{High-level state diagram specification of barrier synchronization
-with three threads and two rounds.  The three numbers specify how many
-rounds each thread has completed.}
-\label{fig:barrierdiagram}
-\end{figure}
-
-\begin{figure}
 \begin{code}
@@
 
+\begin{comment}
+\begin{figure}
+\begin{center}
+\includegraphics[width=\textwidth]{figures/barrier-crop.pdf}
+\end{center}
+\caption{High-level state diagram specification of barrier synchronization
+with three threads and two rounds.  The three numbers specify how many
+rounds each thread has completed.}
+\label{fig:barrierdiagram}
+\end{figure}
+
 \autoref{fig:barrierdiagram} shows a high-level depiction of barrier synchronization
@@
 current round.
+\end{comment}
 
@@
 \end{enumerate}
-Similarly, \texttt{enter}() blocks until $n$ threads have invoked it.
-If more threads try to \texttt{enter}(), they will block until the
-previous $n$ threads have invoked \texttt{exit}().
 
@@
 Each loop uses a different condition variable.
-The first loop waits while the barrier is fully occupied, while
-the second waits for the barrier to fill up.
-The $n^\mathit{th}$ thread to arrive at the barrier notifies the
-$n-1$ threads that are waiting in the second loop.
-
-Method \texttt{exit}() counts how many threads have left the
-barrier.  When the last one leaves, it notifies all the threads
-that are waiting in the first loop.
+The first loop waits until the car has emptied out, while
+the second waits for the car to fill up.
+
+Notice how variables \textit{entered} and \textit{left}
+are maintained.  They are incremented one-by-one, but they
+are not decremented one-by-one.  Instead, the last thread to
+enter sets \textit{left} to zero (and notifies the threads
+that were waiting to for the car to fill up), while the last
+thread to exit sets \textit{entered} to zero (and notifies
+the threads that were waiting for the car to empty out).
 
@@
 \begin{problems}
-\item See if you can implement barrier synchronization for \texttt{N} threads
-with just three binary semaphores.  Busy waiting is not allowed.  How about just two?
+\item Implement barrier synchronization for \texttt{N} threads
+with just three binary semaphores.  Busy waiting is not allowed.
+Can you implement barrier synchronization with two binary semaphores?
 (As always, the Little Book of Semaphores~\cite{Downey09} is a good resource
@@
 solution.)
+\item Imagine a pool hall with \texttt{N} tables.  A table is \emph{full}
+from the time there are two players until both players have left.
+When someone arrives, they can join a table that is not full, preferably
+one that has a player ready to start playing.
+Implement a simulation of such a pool hall.
 \end{problems}
@@
 \hline
-\texttt{any $s$, all $s$} & disjunction / conjunction for set $s$ \\
+\texttt{any $s$, all $s$} & disjunction / conjunction for set or list $s$ \\
 \hline
@@
 
+Harmony also supports \emph{set comprehension}.  In its simplest form,
+\texttt{\{ f($v$) for $v$ in $s$ \}} returns a set that is constructed
+by applying \texttt{f} to all elements in $s$ (where $s$ is a set or
+a list).
+This is known as \emph{mapping}.  But set comprehension is much more
+powerful and can include joining multiple sets (using nested for
+loops) and filtering (using the \texttt{such that} keywords).
+
+For example:
+\texttt{\{ $x$ + $y$ for $x$ in $s$ for $y$ in $s$ such that $(x * y) == 4$ \}}
+returns a set that is constructed by summing pairs of elements from $s$
+that, when multiplied, have the value 4.
+
 \section{Dictionary}
@@
 
+Dictionaries support comprehension.  The  basic form is:
+\texttt{dict\{ f($v$) for $v$ in $s$ \}}.  This creates a dictionary
+that maps each element $v$ in $s$ to $\mathtt{f}(v)$.
+
 There are various special cases of dictionaries,
@@
 at your discretion.  For Python users, the important thing to note
-is that a singleton lists in Harmony must contain a comma.  For
+is that a singleton list in Harmony must contain a comma.  For
 example $[1,]$ is a list containing the value $1$, while $[1]$
@@
 \end{center}
+
+Lists and tuples support comprehension as well.
+In their most basic form: \texttt{f($v$) for $v$ in $s$}.
+For example, to check if any element in a list $t$ is even, you
+can write:
+
+\texttt{any($(x \% 2) == 0$ for $x$ in $t$)}.
 
@@
 \texttt{.c}.
-\texttt{"abc'} is the same three-character string.
+\texttt{"abc"} is the same three-character string.
+Various special characters (including quotes, newlines, etc.)
+can be \emph{escaped} using a backslash.
 Multi-line strings can be terminated by triple quotes or
@@
 multiplicity.
-So \texttt{bag\{ 10, 10, 12 \} = dict\{ 10: 2, 12: 1 \}}.
+So \texttt{bag\{ 10, 10, 12 \} = dict\{ 10:2, 12:1 \}}.
 
@@
 
+Bags also support comprehension.
+
 \section{Program Counter}
@@
 A program counter is an integer that can be used to index into Harmony
-bytecode.  When you define a method or a lambda function, you are
-creating a constant of the program counter type.
-Note, however, that labels are currently just integers.  This may
-change in the future.
+bytecode.  When you define a method, a lambda function, or a label,
+you are creating a constant of the program counter type.
+
+Operations on program counters include the following:
+\begin{center}
+\begin{tabular}{|l|l|}
+\hline
+\texttt{$m$ $a$} & invoke method with program counter $m$ and argument $a$\\
+\hline
+\texttt{atLabel $l$} & return a bag of (method, argument) pairs of threads executing at label $l$ \\
+\hline
+\end{tabular}
+\end{center}
+
+You can create lambda functions similarly to Python.  For example:
+\texttt{lambda$(x,y)$:$x+y$}.
 
@@
 
-\chapter{List of Operators}
-
-Harmony currently supports the following operators:
-\vspace{1em}
-
-{\small
-\begin{tabular}{|l|l|}
-\hline
-\texttt{$e$ == $e$}, \texttt{$e$ != $e$} & two Harmony values of the same type \\
-\hline
-\texttt{$e$ < $e$, $e$ <= $e$, $e$ > $e$, $e$ >= $e$} & any two Harmony values \\
-\hline
-\texttt{$e$ and $e$ and ...} & two or more booleans \\
-\hline
-\texttt{$e$ or $e$ or ...} & two or more booleans \\
-\hline
-\texttt{$e$ => $e$} & two booleans (implication) \\
-\hline
-\texttt{not $e$} & a boolean \\
-\hline
-\texttt{-$e$, \string~$e$, abs($e$)} & an integer \\
-\hline
-\texttt{$e$ + $e$ + ...} & two or more integers, tuples, or lists \\
-\hline
-\texttt{$e$ - $e$} & two integers or sets \\
-\hline
-\texttt{$e$ \& $e$ \& ..., $e$ | $e$ | ..., $e$ \string^ $e$ \string^ ...} & two or more integers or sets \\
-\hline
-\texttt{$e$ * $e$} & two integers or an integer and a list \\
-\hline
-\texttt{$e$ / $e$, $e$ // $e$} & two integers (integer division) \\
-\hline
-\texttt{$e$ \% $e$, $e$ mod $e$} & two integers (division remainder) \\
-\hline
-\texttt{$e$ ** $e$, $e << e$, $e >> e$} & two integers \\
-\hline
-\texttt{$\{ e..e \}$} & two integers \\
-\hline
-\texttt{$e$ [not] in $e$} & first is any Harmony value, second is a set or dict \\
-\hline
-\texttt{$e$ if $e$ else $e$} & middle $e$ must be a boolean \\
-\hline
-\texttt{!$e$} & an address \\
-\hline
-\texttt{choose $e$} & a set \\
-\hline
-\texttt{min $e$, max $e$, any $e$, all $e$, len $e$} & a set or a dict \\
-\hline
-\texttt{keys $e$} & a dictionary (includes tuple and list) \\
-\hline
-\texttt{atLabel $e$} & atom (corresponding to a label) \\
-\hline
-\texttt{get\_context $e$} & $e$ must be \texttt{()} \\
-\hline
-\texttt{setintlevel} \textit{e} & \textit{e} is a boolean \\
-\hline
-\texttt{stop} \textit{lv} & \textit{lv} is an lvalue \\
-\hline
-\texttt{?}\textit{lv} & \textit{lv} is an lvalue \\
-\hline
-\texttt{lambda $a$: $e$ end} & $a$ is a bounded variable, $e$ an expression\\
-\hline
-\end{tabular}
-}
-\vspace{1em}
-
-\texttt{+} concatenates when applied to lists or tuples.
-\texttt{-} computes set difference when applied to two sets.
-\texttt{\&} computes the intersection when applied to sets.
-\texttt{|} computes the union when applied to sets.
-\texttt{\^} computes $(s \cup t) \backslash (s \cap t)$ when applied to sets $s$ and $t$.
-\texttt{$\{ x .. y \}$} computes the set consisting of the integers $x$
-through $y$ inclusive.  It returns the empty set if $x > y$.
-If $d$ is a dictionary, $e$ \texttt{in} $d$ tests if $e$ is in the set of
-\emph{values} of $d$, not the set of \emph{keys} (this is different from Python).
-
-An \emph{lvalue}
-\index{lvalue}%
-(short for left hand value of an assignment statement)
-is something that can be assigned.  This can be a
-shared variable, a thread variable, or a dereferenced pointer
-variable.  It can also be indexed.  Examples include
-\textit{var}, \textit{var}[$e$], and \texttt{!}\textit{p}.
-
-Harmony also supports the following comprehensions:
-
-\vspace{1em}
-\begin{tabular}{|l|l|}
-\hline
-Set comprehension & \texttt{\{ f($v$) for $v$ in $s$ \}} \\
-\hline
-List comprehension & \texttt{[ f($v$) for $v$ in $s$ ]} \\
-\hline
-Dict comprehension & \texttt{dict\{ f($v$) for $v$ in $s$ \}} \\
-\hline
-\end{tabular}
-\vspace{1em}
-
-In each case, $s$ must be a set or a dictionary.
-\autoref{ap:details} provides more information on comprehensions.
+A thread can call \texttt{get\_context()} to retrieve its current
+context.  Also, the function \texttt{contexts()} returns a
+bag of all contexts (including terminated ones).
+Both these functions should only be called in \texttt{atomic} mode
+and used for checking invariants.
 

### 2021-01-02 book/paper.tex
Latest

@@
 semaphores (SBS).
-\texttt{signal} checks to see if the condition's count is non-zero, and if so
+\texttt{signal} checks to see if the condition's count is non-zero, if so
 releases the condition's semaphore,
@@
 There is no \texttt{release\_one} method.  Instead, one can conclude
-that \texttt{put} causes the queue to be non-empty, and
+that \texttt{put} guarantees that the queue will be non-empty, and
 \texttt{signal} will check if there are any threads waiting for
@@
 is such a thread).  Instead, $p$ remains in the critical section
-until it leaves the monitor explicitly.  At that point, any thread that
+until it leaves the monitor explicitly
+(by calling either \texttt{release()} or \texttt{wait()}).
+At that point, any thread that
 was notified will have a chance to enter the critical section, but they compete
@@
 loops until the condition of interest becomes valid.
-A Mesa monitor therefore is more closely related to \emph{busy waiting}
+A Mesa monitor therefore is more closely related to busy waiting
 than to split binary semaphores.
@@
 The so-called
-``Mesa monitor semantics'' or ``Mesa condition variable'' semantics
-have become more popular than Hoare monitor semantics, and have been
+``Mesa monitor semantics'' or ``Mesa condition variable semantics''
+have become more popular than Hoare monitor semantics and have been
 adopted by all major programming languages.
@@
 
-Like C, Java, and Python, Harmony does not have syntactical
-support for monitors. Harmony provides support for monitors
+Harmony provides support for monitors
 through the Harmony \texttt{synch} module.
@@
 This increments the number of threads in the bag with the same context.
-\texttt{wait} then waits until that count is restored to the
+\texttt{wait} then loops until that count is restored to the
 value that it had upon entry to \texttt{wait}.
@@
 
-While Hoare moniters embody the split binary semaphore approach,
-Mesa monitors are much closer to the busy-waiting approach.
-The Mesa code then has to be careful to invoke \textbf{notify} or
-\textbf{notifyAll} in the right places.
+% While Hoare moniters embody the split binary semaphore approach,
+% Mesa monitors are much closer to the busy-waiting approach.
+When using Mesa condition variable you have to be careful to invoke
+\textbf{notify} or \textbf{notifyAll} in the right places.
 Much of the complexity of programming with Mesa condition variables
@@
 \texttt{notifyAll}.  As a rule of thumb: be conservative---it is
-better to wake up too many threads than too few.  Waking up too
+better to wake up too many threads than too few.
+So in case of doubt, use \texttt{notifyAll}.
+Waking up too
 many threads may lead to some inefficiency, but waking up too few
 may cause the application to get stuck.  Harmony can be particularly
-helpful here, as it examines every corner case.  You can safely try
-to replace each \texttt{notifyAll} with \texttt{notify} and see if
-every possible execution of the application still terminates.
+helpful here, as it examines each and every corner case.
+You can try to replace each \texttt{notifyAll} with \texttt{notify} and see
+if every possible execution of the application still terminates.
 
@@
 Imagine five philosopers sitting around a table, each with a plate of food in
-front of them and a fork between each two plates.  Each philosopher requires
+front of them and a fork between every two plates.  Each philosopher requires
 two forks to eat.  To start eating, a philosopher first picks up the fork on
@@
 this for ever, but for the purposes of this book, philosophers can leave
-the table when they want to leave.
+the table when they are not using any forks.
 
@@
 
-Threads can be \emph{interrupted}.  An interrupt is a notification of some event
-such as a keystroke, a timer expiring, the reception of a network packet, the completion
-of a disk operation, and so on.  We distinguish \emph{interrupts} and \emph{exceptions}.
-An exception is caused by the thread executing an invalid machine instruction such as
-divide-by-zero.  An interrupt is caused by some peripheral device and can be handled in
-Harmony.
-In other words: an interrupt is a notification, while an exception is an error.
-\index{exception}
+Threads can be \emph{interrupted}.  An interrupt is a notification
+of some event such as a keystroke, a timer expiring, the reception
+of a network packet, the completion of a disk operation, and so on.
+We distinguish \emph{interrupts} and \emph{exceptions}.  An exception
+is caused by the thread executing an invalid machine instruction
+such as divide-by-zero.  An interrupt is caused by some peripheral
+device and can be handled in Harmony.  In other words: an interrupt
+is a notification, while an exception is an error.  \index{exception}
 
@@
 invokes \textit{handler argument} at some later, unspecified time.
-Only one of these asynchronous events can be outstanding at a time; a new call to
-\textbf{trap} overwrites any outstanding one.
+Thus you can think of \textbf{trap} as setting a timer.
+Only one of these asynchronous events can be outstanding at a time;
+a new call to \textbf{trap} overwrites any outstanding one.
 \autoref{fig:trap} gives an example of how \textbf{trap} might be used.

### 2021-01-17 book/paper.tex
Updated list of machine instructions
Cut now takes two arguments

@@
 \hline
-Bag & pop $n$ and $n$ values and push a bag \\
-\hline
 Continue & no-op (but causes a context switch) \\
@@
 \hline
-Cut & pop set or dict and push minimum and remainder \\
+Cut & cut a set into its smallest element and the remainder \\
 \hline
@@
 DelVar [$v$] & delete thread variable $v$ \\
-\hline
-Dict & pop $n$ and $n$ key/value pairs and push a dictionary \\
 \hline
@@
 Return & pop return address, push \texttt{result}, and restore program counter \\
-\hline
-Set & pop $n$ and $n$ elements and push a set of the elements \\
 \hline

### 2021-01-17 global.h

@@ -106,6 +106,10 @@ bool dict_tryload(uint64_t dict, uint64_t key, uint64_t *result);
 uint64_t dict_remove(uint64_t dict, uint64_t key);
 uint64_t bag_add(uint64_t bag, uint64_t v);
 
+struct env_Cut {
+    uint64_t set, var;
+};
+
 struct env_DelVar {
     uint64_t name;
 };

### 2021-01-19 book/paper.tex
Latest

@@
 
+\begin{figure}[h]
+\begin{center}
+\begin{tabular}{ccc}
+\begin{tcolorbox}[width=0.4\linewidth]
+\harmonysource{prog1}
+\end{tcolorbox}
+& \rule{0.05\linewidth}{0in} &
+\begin{tcolorbox}[width=0.4\linewidth]
+\harmonysource{prog2}
+\end{tcolorbox}
+\\
+(a) \harmonylink{code/prog1.hny} Sequential && (b) \harmonylink{code/prog2.hny} Concurrent
+\end{tabular}
+\end{center}
+\caption{ A sequential and a concurrent program.}
+\label{fig:progs}
+\end{figure}
+
 \glsadd{thread}%
@@
 running in parallel while sharing variables.
-\autoref{fig:inc} presents a simple example.
+\autoref{fig:progs} shows two programs.  Program (a) is sequential.
+It sets \textit{shared} to \texttt{True}, asserts that
+\textit{shared} = \texttt{True}, and finally sets
+\textit{shared} to \texttt{False}.
+If you run the program through Harmony, it will find not problems.
+Program (b) is concurrent.  It executes methods \texttt{f()} and
+\texttt{g()} in parallel.
+If method \texttt{g()} runs and completes before \texttt{f()}, then
+the assertion in \texttt{f()} will fail when \texttt{f()} runs.
+Harmony will find this problem.
+This problem is an example of non-determinism: methods \texttt{f()}
+and \texttt{g()} can run in either order.
+
+\autoref{fig:inc} presents a more subtle example that illustrates
+non-atomicity.
 The program
@@
 
-\section{Booleans}
+\section{Boolean}
 
@@
 
-\section{Integers}
+\section{Integer}
 
@@
 
-\section{Atoms}
+\section{Atom}
 
@@
 Go & pop context and value, push value on context's stack, and add to context bag \\
+\hline
+IncVar $v$ & increment thread variable $v$ \\
 \hline

### 2021-01-19 book/paper.tex
Latest

@@
 non-atomicity.
-The program
-initializes two shared variables: an integer \textit{count} and
+The program initializes two shared variables:
+an integer \textit{count} and
 an array \textit{done} with two booleans.
+The program then spawns three threads.
+The first runs \texttt{incrementer(0)}, the second runs
+\texttt{incrementer(1)}.
+
 Method \texttt{incrementer} takes a parameter called \textit{self}.
 It increments \textit{count} and sets \textit{done}[\textit{self}] to \texttt{True}.
-Method \texttt{main} waits until both \textit{done} flags are set to \texttt{True}.
+It then waits until the other thread is done.
 (\textbf{await} $c$ is semantically the same as \textbf{while not} $c$: \textbf{pass}.)
@@
 
-The program spawns three threads.
-The first runs \texttt{incrementer(0)}, the second runs
-\texttt{incrementer(1)}, and the last runs \texttt{main()}.
 Note that although the threads are \emph{spawned} one at a time,
@@
 \begin{verbatim}
+   0 Frame __init__ ()
 code/Up.hny:1 count = 0;
-   0 Push 0
-   1 Store count
+   1 Push 0
+   2 Store count
 code/Up.hny:2 done = [ False, False ];
-   2 Push [False, False]
-   3 Store done
+   3 Push [False, False]
+   4 Store done
 code/Up.hny:4 def incrementer(self):
-   4 Jump 31
-   5 Frame incrementer self
+   5 Jump 32
+   6 Frame incrementer self
 code/Up.hny:5     count = count + 1;
-   6 Load count
-   7 Push 1
-   8 2-ary +
-   9 Store count
+   7 Load count
+   8 Push 1
+   9 2-ary +
+   10 Store count
 code/Up.hny:6     done[self] = True;
-   10 Push ?done
-   11 LoadVar self
-   12 Address
-   13 Push True
-   14 Store
-   15 Return
-code/Up.hny:8 def main():
-   16 Jump 31
-   17 Frame main ()
-code/Up.hny:9     await all(done);
-   18 Load done
-   19 1-ary all
-   20 JumpCond False 18
-code/Up.hny:10     assert count == 2, count;
-   21 ReadonlyInc
-   22 AtomicInc
-   23 Load count
-   24 Push 2
-   25 2-ary ==
-   26 Load count
-   27 Assert2
-   28 AtomicDec
-   29 ReadonlyDec
-   30 Return
-code/Up.hny:13 spawn incrementer(0);
-   31 Push 0
-   32 Dup
-   33 Push PC(5)
-   34 Spawn
+   11 Push ?done
+   12 LoadVar self
+   13 Address
+   14 Push True
+   15 Store
+code/Up.hny:7     await done[1 - self];
+   16 Push 1
+   17 LoadVar self
+   18 2-ary -
+   19 Load done
+   20 Apply
+   21 JumpCond False 16
+code/Up.hny:8     assert count == 2, count;
+   22 ReadonlyInc
+   23 AtomicInc
+   24 Load count
+   25 Push 2
+   26 2-ary ==
+   27 Load count
+   28 Assert2
+   29 AtomicDec
+   30 ReadonlyDec
+   31 Return
 \end{verbatim}

### 2021-01-19 book/paper.tex
Latest

@@
 \begin{verbatim}
-#states = 53 diameter = 5
-==== Safety violation ====
-T0: __init__() [0-4,31-46] 46 { count: 0, done[0]: False, done[1]: False }
-T1: incrementer(0) [ 5- 8]  9 { count: 0, done[0]: False, done[1]: False }
-T2: incrementer(1) [ 5-15] 15 { count: 1, done[0]: False, done[1]: True  }
-T1: incrementer(0) [ 9-15] 15 { count: 1, done[0]: True,  done[1]: True  }
-T3: main()         [17-27] 27 { count: 1, done[0]: True,  done[1]: True  }
->>> Harmony Assertion (file=code/Up.hny, line=10) failed: 1
-Open file:///Users/rvr/github/harmony/harmony.html for more information
+#states 32
+Safety Violation
+T0: __init__() [0-5,32-40] { count: 0, done: [ False, False ] }
+T1: incrementer(0) [6 - 9] { count: 0, done: [ False, False ] }
+T2: incrementer(1) [6 -18] { count: 1, done: [ False, True  ] }
+T1: incrementer(0) [10-28] { count: 1, done: [ True,  True  ] }
+Harmony assertion failed: 1
+open file:///Users/rvr/github/harmony/harmony.html for more information
 \end{verbatim}

### 2021-01-19 brief.py

@@ -1,7 +1,7 @@
 import json
 
 def json_kv(js):
-    return json_string(js["key"]) + ": " + json_string(js["value"])
+    return (json_string(js["key"]), json_string(js["value"]))
 
 def json_idx(js):
     if js["type"] == "atom":
@@ -22,7 +22,12 @@ def json_string(js):
     if type == "dict":
         if v == []:
             return "()"
-        return "dict{ " + ", ".join([ json_kv(kv) for kv in v ]) + " }" 
+        lst = [ json_kv(kv) for kv in v ]
+        keys = [ k for k,v in lst ]
+        if keys == [str(i) for i in range(len(v))]:
+            return "[ " + ", ".join([v for k,v in lst]) + " ]" 
+        else:
+            return "dict{ " + ", ".join([k + ": " + v for k,v in lst]) + " }" 
     if type == "pc":
         return "PC(%s)"%v
     if type == "address":
@@ -43,42 +48,77 @@ def print_vars(d):
         print(" %s: %s"%(k, json_string(v)), end="")
     print(" }")
 
-def print_range(mis, start, end, first):
-    if not first:
-        print(",", end="")
-    if start + 1 == end:
-        print("%s"%mis[start]["pc"], end="")
-    else:
-        print("%s-%s"%(mis[start]["pc"], mis[end-1]["pc"]), end="")
+def print_range(start, end):
+    return "%d-%d"%(start, end)
 
-def print_megastep(mas):
-    print("T%s: %s ["%(mas["tid"], mas["name"]), end="")
+tid = None
+name = None
+start = 0
+steps = ""
+interrupted = False
+lastmis = {}
+shared = {}
+failure = ""
+
+def flush():
+    global tid, name, steps, start, lastmis, shared
+
+    if tid != None:
+        print("T%s: %s ["%(tid, name), end="")
+        if steps != "":
+            steps += ","
+        steps += print_range(start, int(lastmis["pc"]))
+        print(steps + "] ", end="");
+        print_vars(shared);
+
+def print_macrostep(mas):
+    global tid, name, start, steps, lastmis, interrupted, shared, failure
+
     mis = mas["microsteps"]
-    start = 0
-    first = True
-    for i in range(1, len(mis)):
-        if "interrupt" in mis[i-1]:
-            print_range(mis, start, i-1, first)
-            first = False
-            start = i
-            print(",interrupt", end="")
+    if mas["tid"] != tid:
+        flush()
+        tid = mas["tid"]
+        name = mas["name"]
+        interrupted = False
[…]

### 2021-01-19 charm.c

@@ -183,9 +183,12 @@ void onestep(struct node *node, uint64_t ctx, uint64_t choice, bool interrupt,
             gettimeofday(&tv, NULL);
             double now = tv.tv_sec + (double) tv.tv_usec / 1000000;
             if (now - lasttime > 1) {
-                char *p = value_string(cc->name);
-                printf("%s pc=%d states=%d queue=%d\n", p, cc->pc, enqueued, enqueued - dequeued);
-                free(p);
+                if (lasttime != 0) {
+                    char *p = value_string(cc->name);
+                    printf("%s pc=%d states=%d queue=%d\n",
+                            p, cc->pc, enqueued, enqueued - dequeued);
+                    free(p);
+                }
                 lasttime = now;
             }
             timecnt = 1;

### 2021-01-19 harmony.py

@@ -5519,6 +5519,7 @@ def main():
         if r != 0:
             print("charm model checker failed")
             sys.exit(r);
+        os.system("python3 brief.py");
         r = os.system("python3 charm.py");
         if r == 0:
             print("open file://" + os.getcwd() + "/harmony.html for more information")

### 2021-01-19 ops.c

@@ -524,7 +524,6 @@ void op_Assert(const void *env, struct state *state, struct context **pctx){
         ctx_failure(*pctx, "assert can only be applied to bool values");
     }
     if (v == VALUE_FALSE) {
-        // printf("HARMONY ASSERTION FAILED\n");
         ctx_failure(*pctx, "Harmony assertion failed");
     }
     else {
@@ -540,7 +539,6 @@ void op_Assert2(const void *env, struct state *state, struct context **pctx){
     }
     if (v == VALUE_FALSE) {
         char *p = value_string(e);
-        printf("HARMONY ASSERTION FAILED: %s\n", p);
         ctx_failure(*pctx, "Harmony assertion failed: %s", p);
         free(p);
     }

### 2021-01-19 book/paper.tex
Latest

@@
 T0: __init__() [0-5,32-40] { count: 0, done: [ False, False ] }
-T1: incrementer(0) [6 - 9] { count: 0, done: [ False, False ] }
-T2: incrementer(1) [6 -18] { count: 1, done: [ False, True  ] }
+T1: incrementer(0) [ 6- 9] { count: 0, done: [ False, False ] }
+T2: incrementer(1) [ 6-18] { count: 1, done: [ False, True  ] }
 T1: incrementer(0) [10-28] { count: 1, done: [ True,  True  ] }
@@
 execution in the program.
-In this case, it executes instructions 0 through 4 first.
+In this case, it executes instructions 0 through 5 first.
 The last instruction is a \texttt{JUMP} instruction that sets the
-program counter to 31 (skipping over the code for the methods).
-At program counter 5 is the code for the \texttt{incrementer} method.
-All methods start with a \texttt{Frame} instruction and end with a \texttt{Return}
-instruction.
+program counter to 32 (skipping over the code for \texttt{incrementer} method).
+At program counter 6 is the code for the \texttt{incrementer} method.
+All methods start with a \texttt{Frame} instruction and end with a
+\texttt{Return} instruction.
 
@@
 
-\begin{enumerate} \setcounter{enumi}{5}
+\begin{enumerate} \setcounter{enumi}{6}
 \item The \texttt{Load} instruction pushes the value of the
@@
 \begin{figure}
-\includegraphics[width=\textwidth]{figures/Up1.png}
+\begin{center}
+\includegraphics[width=0.9\textwidth]{figures/Up1.png}
+\end{center}
 \caption{The
 \harmonyref{Up.html}{HTML output of running Harmony on \autoref{fig:inc}}.
-There are three sections.
+There are four sections.
 The top shows the steps to the problematic state.
-The bottom left shows the source code and bytecode
-of the program.
-The bottom right shows the state of each thread after the selected step.
-Each row in the stack trace of a thread shows the current program counter,
-the method that is being evaluated, and the line of code that is
-being evaluated.}
+Each step shows a thread identifier, the method the thread is executing,
+a ``timeline'' containing a block for each instruction executed, the
+program counter of the thread, and the values of the shared variables
+at those steps.
+The middle section shows the line of code that is being executed.
+The bottom left shows the bytecode of the program.
+The bottom right shows the state of each thread.
+Each row in the stack trace of a thread shows
+the method that is being evaluated and the values of its method variables.}
 \label{fig:inchtml1}
@@
 It also includes certain infinite loops.
-Harmony will print at trace of the shortest bad execution that it found.
+Harmony will print a trace of the shortest bad execution that it found.
 \item \texttt{Non-terminating States}: Harmony found one or more states
@@
 \item The sequence of program counters of the HVM instructions that the thread executed;
-\item The current program counter of the thread;
 \item The contents of the shared memory.
@@
 \item Thread \texttt{incrementer(0)}
-executed instructions 5 through 8, loading the value of
+executed instructions 6 through 9, loading the value of
 \textit{count} but stopping just before storing 1 into \textit{count};
 \item Thread \texttt{incrementer(1)}
-executed instructions 5 through 15, storing 1 into
+executed instructions 6 through 18, storing 1 into
 into \textit{count} and storing \texttt{True} into \textit{done}[1];
 \item Thread \texttt{incrementer(0)}
-continues execution, executing instructions 9 through 15
-storing value 1 into \textit{count}
-and storing \texttt{True} into \textit{done}[0].
+continues execution, executing instructions 10 through 28
+storing value 1 into \textit{count} (instruction 10),
+storing \texttt{True} into \textit{done}[0] (instructions 11 through 15)
+finding that \textit{done}[1] is \texttt{True} (instructions 16 through 21),
+and finally detecting that the assertion is violated
+(instructions 22 through 28).
 \end{itemize}
-Finally, thread \texttt{main()} runs and finds the problem.
-This makes precise the concurrency issue that we encountered.
 
@@
 \autoref{fig:inchtml1}.
+You can go to any point in the execution by clicking on one of the
+blocks in the ``execution timelines.''  Using the left and right arrow
+keys you can go forwards and backwards one instruction at a time.
+With the down and up arrow keys you can go through the execution one
+step at a time.
+If you hit enter, it will to go to the next line of Harmony code.
 If you hover the mouse over a machine instruction, it provides a

### 2021-01-19 charm.py

@@ -98,7 +98,7 @@ def html_megastep(glob, step, tid, name, nmicrosteps, width, f):
     print("  </td>", file=f)
 
     for i in range(width):
-      print("  <td>", file=f)
+      print("  <td align='center'>", file=f)
       print("  </td>", file=f)
     print("</tr>", file=f)
 

### 2021-01-20 book/paper.tex
Latest

@@
 
-\begin{figure}
-\begin{code}
-\begin{verbatim}
-#states 32
-Safety Violation
-T0: __init__() [0-5,32-40] { count: 0, done: [ False, False ] }
-T1: incrementer(0) [ 6- 9] { count: 0, done: [ False, False ] }
-T2: incrementer(1) [ 6-18] { count: 1, done: [ False, True  ] }
-T1: incrementer(0) [10-28] { count: 1, done: [ True,  True  ] }
-Harmony assertion failed: 1
-open file:///Users/rvr/github/harmony/harmony.html for more information
-\end{verbatim}
-\end{code}
-\caption{The text output of running Harmony on \autoref{fig:inc}.}
-\label{fig:incoutput}
-\end{figure}
-
 \glsadd{stack machine}%
@@
 
+\begin{figure}
+\begin{code}
+\begin{verbatim}
+#states 32
+Safety Violation
+T0: __init__() [0-5,32-40] { count: 0, done: [ False, False ] }
+T1: incrementer(0) [ 6- 9] { count: 0, done: [ False, False ] }
+T2: incrementer(1) [ 6-18] { count: 1, done: [ False, True  ] }
+T1: incrementer(0) [10-28] { count: 1, done: [ True,  True  ] }
+Harmony assertion failed: 1
+open file:///Users/rvr/github/harmony/harmony.html for more information
+\end{verbatim}
+\end{code}
+\caption{The text output of running Harmony on \autoref{fig:inc}.}
+\label{fig:incoutput}
+\end{figure}
+
 \autoref{ap:harmonybytecode} provides a list of all HVM machine instructions,
@@
 \texttt{Up.hny} program.
-
-\begin{figure}
-\begin{center}
-\includegraphics[width=0.9\textwidth]{figures/Up1.png}
-\end{center}
-\caption{The
-\harmonyref{Up.html}{HTML output of running Harmony on \autoref{fig:inc}}.
-There are four sections.
-The top shows the steps to the problematic state.
-Each step shows a thread identifier, the method the thread is executing,
-a ``timeline'' containing a block for each instruction executed, the
-program counter of the thread, and the values of the shared variables
-at those steps.
-The middle section shows the line of code that is being executed.
-The bottom left shows the bytecode of the program.
-The bottom right shows the state of each thread.
-Each row in the stack trace of a thread shows
-the method that is being evaluated and the values of its method variables.}
-\label{fig:inchtml1}
-\end{figure}
 
@@
 Next it reports an execution that failed this assertion.
-The program got to the faulty execution in 5 ``steps.''
-The output has five columns:
+The program got to the faulty execution in 4 ``steps.''
+The output has four columns:
 \begin{enumerate}
@@
 
-The first step is initialization.
-It sets shared variable \textit{count} to 0 and
+\begin{figure}
+\begin{center}
+\includegraphics[width=0.9\textwidth]{figures/Up1.pdf}
+\end{center}
+\caption{The
+\harmonyref{Up.html}{HTML output of running Harmony on \autoref{fig:inc}}.
+There are four sections.
+The top shows the steps to the problematic state.
+Each step shows a thread identifier, the method the thread is executing,
+a ``timeline'' containing a block for each instruction executed, the
+program counter of the thread, and the values of the shared variables
+at those steps.
+The middle section shows the line of code that is being executed.
+The bottom left shows the bytecode of the program.
+The bottom right shows the state of each thread.
+Each row in the stack trace of a thread shows
+the method that is being evaluated and the values of its method variables.}
+\label{fig:inchtml1}
+\end{figure}
+
+The four steps in the execution are as follows:
+
+\begin{enumerate}
+\item Thread \texttt{\_\_init\_\_} (with identifier T0)
+sets shared variable \textit{count} to 0 and
 shared variable \textit{done} to \texttt{[False, False]}.
-If we look at the next three steps, we see that:
-\begin{itemize}
-\item Thread \texttt{incrementer(0)}
+\item Thread \texttt{incrementer(0)} (with identifier T1)
 executed instructions 6 through 9, loading the value of
 \textit{count} but stopping just before storing 1 into \textit{count};
-\item Thread \texttt{incrementer(1)}
+\item Thread \texttt{incrementer(1)} (with identifier T2)
 executed instructions 6 through 18, storing 1 into
@@
 (instructions 22 through 28).
-\end{itemize}
+\end{enumerate}
 
@@
 \autoref{fig:inchtml1}.
-You can go to any point in the execution by clicking on one of the
-blocks in the ``execution timelines.''  Using the left and right arrow
-keys you can go forwards and backwards one instruction at a time.
-With the down and up arrow keys you can go through the execution one
-step at a time.
-If you hit enter, it will to go to the next line of Harmony code.
+
+In the top right, the HTML file contains the reported issue in red.
+Underneath it, a table shows the four steps in the execution.
+Instead of listing explicitly the program counters of the executed
+instructions, the HTML file contains a list of blocks for each
+executed instruction.  We call this the \emph{time line}.
+You can click on such a block to see the state of the Harmony
+virtual machine just after executing that instruction.
+The step that is being executed is highlighted in green.
+The table also lists the program counter of the thread at each step,
+and the values of the shared variables.
+
+Underneath the table it shows the line of Harmony code that is being
+executed in blue.
+
+The bottom left shows
+the bytecode of the program being executed.  It has alternating
+grey and white sections.  Each section corresponds to a line of
+Harmony code.  The instruction that is about to be executed,
+if any, is highlighted in red.  (In this case, the state shown
+is a failed state and no instruction will be executed next.)
 If you hover the mouse over a machine instruction, it provides a
-brief explanation.
+brief explanation of what the instruction does.
+
+The bottom right contains a table with the state of each thread.
+The thread that is executing is highlighted in green.
+Status information for a thread can include:
+\begin{itemize}
+\item[] \texttt{runnable}: the thread is runnable but not
+currently running.  In Harmony, threads are interleaved and so
+at most one thread is actually running;
+\item[] \texttt{terminated}: the thread has completed all its
+instructions;
+\item[] \texttt{failed}: the thread has encountered an error,
+such as violating an assertion or divide by zero;
+\item[] \texttt{blocked}: the thread cannot make progress until
+another thread has updated the shared state;
+\item[] \texttt{atomic}: the thread is in \emph{atomic} mode,
+not allowing other threads to be scheduled.  This is, for example,
+the case when an assertion is being checked;
+\item[] \texttt{read-only}: the thread is in \emph{read-only} mode,
+not able to modify shared state.  Assertions can execute
+arbitrary code including methods, but they are not allowed to
+modify the shared state.
+\end{itemize}
+
+Next to the status, the table shows the \emph{stack trace} of the
+thread.  A stack trace is a list of methods that are being invoked.
+In this case, the \texttt{implementer} method does not invoke any
+other methods, and so the list is of length 1.
+For each entry in the stack trace, it shows the method name and
+arguments, as well as the variables of the method.
+
+When you load the HTML file, it shows the state after executing
+the last instruction.
+As mentioned above, you can go to any point in the execution by
+clicking on one of the blocks in the time lines.
+There are also various handy keyboard shortcuts:
+\begin{itemize}
+\item \emph{Right arrow}: go to the next instruction;
+\item \emph{Left arrow}: go to the previous instruction;
+\item \emph{Down arrow}: go to the next step;
+\item \emph{Up arrow}: go to the previous step;
+\item \emph{Enter (aka Return)}: go to the next line of Harmony code;
+\item \emph{0}: go to the initial state.
+\end{itemize}
+
+If you want to see an animation of the entire execution, one instruction
+at a time, you can first hit 0 and then hold down the right arrow.
+If you want to see it one line of Harmony code at a time, hold down the
+enter (aka return) key instead.
+If you hold down the down arrow key, the movie will
+go by very quickly.
 
@@
     in broad steps, how running the program can get into a bad state.
-\item What if we moved line~6 of \autoref{fig:incenter} to after the \textbf{if}
-    statement (between lines~9 and~10)?  Do you think that would work?  Run it through
+\item What if we moved line~5 of \autoref{fig:incenter} to after the \textbf{if}
+    statement (between lines~8 and~9)?  Do you think that would work?  Run it through
     Harmony and describe either why it works or why it does not work.

### 2021-01-20 book/paper.tex
Latest

@@
 at most one thread is actually running;
+\item[] \texttt{running}: the thread is currently executing
+instructions;
 \item[] \texttt{terminated}: the thread has completed all its
@@
 \item[] \texttt{blocked}: the thread cannot make progress until
-another thread has updated the shared state;
+another thread has updated the shared state.  For example, this
+occurs when one of the implementers is waiting for the other
+to set its \textit{done} flag;
 \item[] \texttt{atomic}: the thread is in \emph{atomic} mode,
@@
 There are also various handy keyboard shortcuts:
-\begin{itemize}
-\item \emph{Right arrow}: go to the next instruction;
-\item \emph{Left arrow}: go to the previous instruction;
-\item \emph{Down arrow}: go to the next step;
-\item \emph{Up arrow}: go to the previous step;
-\item \emph{Enter (aka Return)}: go to the next line of Harmony code;
-\item \emph{0}: go to the initial state.
-\end{itemize}
+\begin{tabular}{rl}
+\emph{Right arrow}: & go to the next instruction; \\
+\emph{Left arrow}: & go to the previous instruction; \\
+\emph{Down arrow}: & go to the next step; \\
+\emph{Up arrow}: & go to the previous step; \\
+\emph{Enter (aka Return)}: & go to the next line of Harmony code; \\
+\emph{0}: & go to the initial state.
+\end{tabular}
 

### 2021-01-22 src/harmony/charm.js
Latest and greatest

@@ -50,9 +50,31 @@ function json_string_set(obj) {
 
 function json_string_dict(obj) {
   if (obj.length == 0) {
-    return "()"
+    return "( )"
   }
+
+  var islist = true;
+  for (var i = 0; i < obj.length; i++) {
+    if (obj[i].key.type != "int" || obj[i].key.value != i.toString()) {
+      islist = false;
+      break;
+    }
+  }
+
   var result = "";
+  if (islist) {
+    for (var i = 0; i < obj.length; i++) {
+      if (i != 0) {
+        result += ", ";
+      }
+      result += json_string(obj[i].value);
+    }
+    if (obj.length == 1) {
+      result += ",";
+    }
+    return "[" + result + "]";
+  }
+
   for (var i = 0; i < obj.length; i++) {
     if (result != "") {
       result += ", ";
@@ -130,8 +152,14 @@ function convert_vars(obj) {
   return result;
 }
 
-function stackTrace(table, trace, failure) {
+function stackTrace(tid, trace, failure) {
+  var table = threads[tid].tracetable;
   table.innerHTML = "";
+  if (trace.length == 0) {
+    var row = table.insertRow();
+    var mcell = row.insertCell();
+    mcell.innerHTML = threads[tid].name;
+  }
   for (var i = 0; i < trace.length; i++) {
     var row = table.insertRow();
 
@@ -377,6 +405,7 @@ function init_microstep(masidx, misidx) {
     var vals = mis.push.map(x => json_string(x));
     microsteps[t].stack = microsteps[t].stack.concat(vals);
   }
+  // microsteps[t].choose = microsteps[t].stack;
 }
 
 function init_macrostep(i) {
@@ -422,21 +451,22 @@ function run_microstep(t) {
   }
 
   if (mis.failure != null) {
-    stackTrace(threads[mis.tid].tracetable, mis.trace, mis.failure);
+    stackTrace(mis.tid, mis.trace, mis.failure);
   }
   else {
-    stackTrace(threads[mis.tid].tracetable, mis.trace, mis.choose);
+    stackTrace(mis.tid, mis.trace, mis.choose);
   }
 
   for (var ctx = 0; ctx < mis.contexts.length; ctx++) {
     var tid = parseInt(mis.contexts[ctx].tid);
[…]

### 2021-01-22 src/harmony/genhtml.py

@@ -181,24 +181,27 @@ m4_include(charm.js)
         print("</div>", file=f)
 
     def html_botright(self, f):
-        print("<table border='1' id='threadtable'>", file=f)
+        print("<table border='1'", file=f)
         print("  <thead>", file=f)
         print("    <tr>", file=f)
+        print("      <th colspan='4'>Threads</th>", file=f)
+        print("    </tr>", file=f)
+        print("    <tr>", file=f)
         print("      <th>", file=f)
-        print("        Thread", file=f)
+        print("        ID", file=f)
         print("      </th>", file=f)
         print("      <th>", file=f)
         print("        Status", file=f)
         print("      </th>", file=f)
         print("      <th>", file=f)
-        print("        Trace", file=f)
+        print("        Stack Trace", file=f)
         print("      </th>", file=f)
         print("      <th>", file=f)
-        print("        Stack", file=f)
+        print("        Stack Top", file=f)
         print("      </th>", file=f)
         print("    </tr>", file=f)
         print("  </thead>", file=f)
-        print("  <tbody>", file=f)
+        print("  <tbody id='threadtable'>", file=f)
         maxtid = 0
         for i in range(self.nthreads):
             print("    <tr id='thread%d'>"%i, file=f)
@@ -212,8 +215,7 @@ m4_include(charm.js)
             print("        <table id='threadinfo%d' border='1'>"%i, file=f)
             print("        </table>", file=f)
             print("      </td>", file=f)
-            print("      <td align='center'>", file=f)
-            print("        xxx", file=f)
+            print("      <td align='left'>", file=f)
             print("      </td>", file=f)
             print("    </tr>", file=f)
         print("  </tbody>", file=f)
@@ -226,11 +228,13 @@ m4_include(charm.js)
         self.html_top(f)
         print("    </td>", file=f)
         print("  </tr>", file=f)
+        print("  <tr><td></td></tr>", file=f)
         print("  <tr>", file=f)
         print("    <td colspan='2'>", file=f)
         print("      <h3 style='color:blue;' id='coderow'>CODE GOES HERE</h3>", file=f)
         print("    </td>", file=f)
         print("  </tr>", file=f)
+        print("  <tr><td></td></tr>", file=f)
         print("  <tr>", file=f)
         print("    <td valign='top'>", file=f)
         self.html_botleft(f)

### 2021-01-22 book/paper.tex
Latest and greatest
Latest

@@
 verifies that the value of \textit{count} equals 2.
-Like Python, Harmony supports \texttt{assert} statements with two arguments:
-the first is a Boolean condition, and the second is a value that is
-reported if the condition fails, in this case
-the value of \textit{count}.
+% Like Python, Harmony supports \texttt{assert} statements with two arguments:
+% the first is a Boolean condition, and the second is a value that is
+% reported if the condition fails, in this case
+% the value of \textit{count}.
 
@@
 Each row in the stack trace of a thread shows
-the method that is being evaluated and the values of its method variables.}
+the method that is being evaluated and
+the values of its method variables.}
 \label{fig:inchtml1}
@@
 
-Next to the status, the table shows the \emph{stack trace} of the
-thread.  A stack trace is a list of methods that are being invoked.
+The stack of each thread is subdivided into two parts: the
+\emph{stack trace} and the \emph{stack top}.
+A stack trace is a list of methods that are being invoked.
 In this case, the \texttt{implementer} method does not invoke any
@@
 arguments, as well as the variables of the method.
+The stack top shows the values on the stack beyond the stack trace.
 
@@
 \begin{figure}
-\includegraphics[width=\textwidth]{figures/naiveLock.png}
+\begin{center}
+\includegraphics[width=0.8\textwidth]{figures/naiveLock.pdf}
+\end{center}
 \caption{The \harmonyref{naiveLock.html}{HTML output of running Harmony on \autoref{fig:uplock}}.}
@@
 \begin{figure}
-\includegraphics[width=\textwidth]{figures/naiveFlags.png}
+\begin{center}
+\includegraphics[width=0.8\textwidth]{figures/naiveFlags.pdf}
+\end{center}
 \caption{The \harmonyref{naiveFlags.html}{HTML output of running Harmony on \autoref{fig:upflags}}.}

### 2021-01-23 book/paper.tex
Latest
Book updated to use where

@@
 \newcommand{\harmonylink}[1]{%
-[\href{https://www.cs.cornell.edu/home/rvr/harmony/#1}{\underline{#1}}]%
+[\href{https://harmony.cs.cornell.edu/#1}{\underline{#1}}]%
 }
@@
 \newcommand{\harmonyref}[2]{%
-\href{https://www.cs.cornell.edu/home/rvr/harmony/output/#1}{\underline{#2}}%
+\href{https://harmony.cs.cornell.edu/output/#1}{\underline{#2}}%
 }
@@
 powerful and can include joining multiple sets (using nested for
-loops) and filtering (using the \texttt{such that} keywords).
+loops) and filtering (using the \texttt{where} keyword).
 
@@
 \texttt{for} $j$ in \{ 1..10 \}
-\texttt{such that} $i < j$: ...
+\texttt{where} $i < j$: ...
 \end{code}
@@
 Set and list comprehensions are similar to Python, except that filtering uses
-the keywords \texttt{such that} instead of \texttt{if}.
+the keyword \texttt{where} instead of \texttt{if}.
 Dictionary comprehensions are a little different.

### 2021-01-24 book/paper.tex
Starting to update the book and remove semicolons

@@
 \begin{itemize}
-\item Every statement in Harmony must be terminated by a semicolon
-(even \textbf{def} statements).
-In Python semicolons are optional and rarely used.
-There is no syntactic significance to indentation in Harmony.
-Correct indentation in Harmony is encouraged but not enforced.
 \item Harmony only supports basic operator precedence or associativity.
@@
 \begin{code}
-\textbf{assert} (\textit{count} == 1) or (\textit{count} == 2);
+\textbf{assert} (\textit{count} == 1) or (\textit{count} == 2)
 \end{code}
@@
    0 Frame __init__ ()
-code/Up.hny:1 count = 0;
+code/Up.hny:1 count = 0
    1 Push 0
    2 Store count
-code/Up.hny:2 done = [ False, False ];
+code/Up.hny:2 done = [ False, False ]
    3 Push [False, False]
@@
    6 Frame incrementer self
-code/Up.hny:5     count = count + 1;
+code/Up.hny:5     count = count + 1
    7 Load count
@@
    10 Store count
-code/Up.hny:6     done[self] = True;
+code/Up.hny:6     done[self] = True
    11 Push ?done
@@
    15 Store
-code/Up.hny:7     await done[1 - self];
+code/Up.hny:7     await done[1 - self]
    16 Push 1
@@
    21 JumpCond False 16
-code/Up.hny:8     assert count == 2, count;
+code/Up.hny:8     assert count == 2
    22 ReadonlyInc
@@
 \begin{code}
-\texttt{await False;}
+\texttt{await False}
 \end{code}

### 2021-01-24 book/paper.tex
Latest

@@
 Harmony makes extensive use of dictionaries.
-A dictionary maps values, known as \emph{keys}, to values.
-The Harmony dictionary syntax and properties are a little different from Python.
+A dictionary maps keys to values.
 Unlike Python, any Harmony value can be a key, including another
 dictionary.
-Harmony dictionaries are written as
-$\mathtt{dict}\{ k_0: v_0, ~ k_1: v_1, ~ ... \}$.
+Dictionaries are written as
+$\{ k_0: v_0, ~ k_1: v_1, ~ ... \}$.
 If $d$ is a dictionary, and $k$ is a key, then the
@@
 \end{code}
-However, if $d$ = \texttt{dict\{ .count: 3 \}}, then you can write
+However, if $d = \{ \mathtt{.count}: 3 \}$, then you can write
 $d$.\texttt{count} (which has value 3) instead of having to write
@@
 \texttt{(5, False)} is the same Harmony value as
-\texttt{dict\{ 0:5, 1:False \}}.
-The empty tuple \texttt{()} is the same value as \texttt{dict\{\}}.
+\texttt{\{ 0:5, 1:False \}}.
+The empty tuple is written as \texttt{()}.
 Note that this is different from the empty set, which is \texttt{\{\}}.
@@
 is the same Harmony value as \texttt{(a, b, c)} (which looks like a Python tuple),
-which in turn is the same Harmony value as \texttt{dict\{ 0:a, 1:b, 2:c \}}.
+which in turn is the same Harmony value as \texttt{\{ 0:a, 1:b, 2:c \}}.
 So if $x$ = \texttt{[False, True]},
@@
 \hline
-Dictionary & \texttt{dict\{ .account: 12345, .valid: False \} } \\
+Dictionary & \texttt{\{ .account: 12345, .valid: False \} } \\
 \hline
@@
 The generic syntax of a dictionary is
-\texttt{dict\{ $k_0:v_0, k_1:v_1, ...$ \}}.
+$\{ k_0:v_0, k_1:v_1, ... \}$.
 If there are duplicate keys in the list, then only the one with
@@
 Dictionaries support comprehension.  The  basic form is:
-\texttt{dict\{ f($v$) for $v$ in $s$ \}}.  This creates a dictionary
-that maps each element $v$ in $s$ to $\mathtt{f}(v)$.
+\texttt{\{ f($k$):g($k$) for $k$ in $s$ \}}.
 
@@
 different than in Python.  For example, if $x$ = [.a, .b, .c], then
-\texttt{del} $x$[1] results in $x$ = \texttt{dict}\{ 0:.a, 2:.c \}, \emph{not}
+\texttt{del} $x$[1] results in $x$ = \{ 0:.a, 2:.c \}, \emph{not}
 $x$ = [.a, .c].
@@
 \index{comprehensions}%
-Set and list comprehensions are similar to Python, except that filtering uses
+Comprehensions are similar to Python, except that filtering uses
 the keyword \texttt{where} instead of \texttt{if}.
-Dictionary comprehensions are a little different.
-The dictionary comprehension \texttt{dict}\{ f($i$) \texttt{for} $i$ \texttt{in} $I$ \}
-creates a dictionary of $i$:f($i$) entries.
 

### 2021-01-24 book/paper.tex
Latest

@@
 times it occurs in the bag.
-The syntax is \texttt{bag\{ $e$, $e$, ... \}}.
 
@@
 
+Native operations on strings include the following:
+
+\begin{center}
+\begin{tabular}{|l|l|}
+\hline
+\texttt{$s~k$} & indexing \\
+\hline
+\texttt{$s$ + $s$ + ...} & concatenation \\
+\hline
+\texttt{$s$ * $n$} & $n$ copies of $s$ concatenated \\
+\hline
+\texttt{$c$ [not] in $s$} & check if $c$ (a one-character atom) is in $s$ \\
+\hline
+\texttt{len $s$} & the length of $s$ \\
+\hline
+\end{tabular}
+\end{center}
+
 \section{Bag or Multiset}
 
-Harmony has native support for bags (aka multisets), which are sets
-that can contain multiple copies of the same element.
-It is represented by a dictionary that maps each element to its
-multiplicity.
-So \texttt{bag\{ 10, 10, 12 \} = dict\{ 10:2, 12:1 \}}.
-
+A bag is represented by a dictionary that maps each element to its
+multiplicity, for example: \{ 10:2, 12:1 \}.
 The \texttt{bag} module (\autoref{ap:bag}) contains various
@@
 \end{center}
-
-Bags also support comprehension.
 

### 2021-01-24 book/paper.tex
Latest

@@
 that particular execution as well.
+What would the value of \textit{count} be at the end of that execution?
 
@@
 
+or like so:
+
+\vspace{1em}
+\begin{code}
+\harmonysource{dinersfix2}
+\end{code}
+\vspace{1em}
+
 This completes all the Havender methods.
@@
 $\{ k_0:v_0, k_1:v_1, ... \}$.
+Different from Python, the empty dictionary is either $()$ or $[]$
+(because $\{\}$ is the empty set in Harmony).
 If there are duplicate keys in the list, then only the one with
@@
 \begin{verbatim}
-dict{ .turn: 0, .flags: [ False, False ], .enter: P_enter, .exit: P_exit }
+{ .turn: 0, .flags: [ False, False ], .enter: P_enter, .exit: P_exit }
 \end{verbatim}

### 2021-01-24 book/sources/pp.py

@@ -71,6 +71,7 @@ def isreserved(s):
         "from",
         "False",
         "for",
+        "get_context",
         "go",
         "hash",
         "if",

### 2021-01-29 book/paper.tex
Now creates model checker in $HOME/.charm
Latest

@@
 Harmony is heavily influenced by Leslie Lamport's work on
-TLA+, TLC, and PlusCal~\cite{Lamport02, Lamport09}.
+TLA+, TLC, and PlusCal~\cite{Lamport02, Lamport09}, as well as
+Gerard Holzmann's work on Promela and SPIN~\cite{SPIN}.
 Harmony is designed to have a lower learning curve than those
@@
 In Harmony, a thread can suspend itself and save its context (state) in a
-list.  Recall that the context of a thread contains
+shared variable.  Recall that the context of a thread contains
 its program counter, stack, and registers.
@@
 \begin{code}
-\textbf{stop} $L$
-\end{code}
-
-Here $L$ is a shared variable containing a list.
+\textbf{stop} $s$
+\end{code}
+
+This causes the context of the thread to be saved in $s$ and
+the thread to be no longer running.
 Another thread can revive the thread using the \textbf{go}
@@
 acquires the binary semaphore if available and suspends the invoking thread if not.
+In the latter case, the context of the thread is added to the end of the list of contexts.
 Note that \textbf{stop} is called within an \textbf{atomic} statement---this is
@@
 \hline
-\texttt{signal($,$. $m$)} & signal a condition variable\\
+\texttt{signal($c$, $m$)} & signal a condition variable\\
 \hline

### 2021-01-29 src/charm/charm.c

@@ -135,7 +135,6 @@ void check_invariants(struct node *node, struct context **pctx){
             b = false;
         }
         if (!b) {
-            printf("INVARIANT FAILED\n");
             struct failure *f = new_alloc(struct failure);
             f->type = FAIL_INVARIANT;
             f->ctx = node->before;
@@ -305,8 +304,7 @@ void onestep(struct node *node, uint64_t ctx, uint64_t choice, bool interrupt,
         next->len = node->len + weight;
         graph_add(next);
 
-        // TODO.  Is this the right place?  Don't check when choice?
-        if (sc->invariants != VALUE_SET) {
+        if (sc->choosing == 0 && sc->invariants != VALUE_SET) {
             check_invariants(next, pinv_ctx);
         }
 

### 2021-01-29 book/paper.tex
Forgot any/all

@@
 \hline
-\texttt{len $e$} & cardinality \\
-\hline
-\texttt{$e$ - $e$} & set difference \\
-\hline
-\texttt{$e$ \& $e$ \& ...} & intersection \\
-\hline
-\texttt{$e$ | $e$ | ...} & union \\
-\hline
-\texttt{$e$ \string^ $e$ \string^ ...} & inclusion/exclusion (elements in odd number of sets) \\
-\hline
-\texttt{choose $e$} & select an element (Harmony will try all) \\
-\hline
-\texttt{min $e$} & minimum element \\
-\hline
-\texttt{max $e$} & maximum element \\
+\texttt{len $s$} & cardinality \\
+\hline
+\texttt{$s$ - $s$} & set difference \\
+\hline
+\texttt{$s$ \& $s$ \& ...} & intersection \\
+\hline
+\texttt{$s$ | $s$ | ...} & union \\
+\hline
+\texttt{$s$ \string^ $s$ \string^ ...} & inclusion/exclusion (elements in odd number of sets) \\
+\hline
+\texttt{choose $s$} & select an element (Harmony will try all) \\
+\hline
+\texttt{min $s$} & minimum element \\
+\hline
+\texttt{max $s$} & maximum element \\
+\hline
+\texttt{any $s$} & \texttt{True} if any value is \texttt{True} \\
+\hline
+\texttt{all $s$} & \texttt{True} if all values are \texttt{True} \\
 \hline
@@
 \hline
+\texttt{min $d$} & the minimum value in $d$ \\
+\hline
+\texttt{max $d$} & the maximum value in $d$ \\
+\hline
+\texttt{any $d$} & \texttt{True} if any value is \texttt{True} \\
+\hline
+\texttt{all $d$} & \texttt{True} if all values are \texttt{True} \\
+\hline
 \texttt{$d$ \& $d$ \& ...} & dictionary intersection \\
@@
 \texttt{len $t$} & the length of $t$ \\
+\hline
+\texttt{min $t$} & the minimum value in $t$ \\
+\hline
+\texttt{max $t$} & the maximum value in $t$ \\
+\hline
+\texttt{any $t$} & \texttt{True} if any value is \texttt{True} \\
+\hline
+\texttt{all $t$} & \texttt{True} if all values are \texttt{True} \\
 \hline

### 2021-01-31 book/paper.tex
Latest
Fixed some typos

@@
 \textit{shared} to \texttt{False}.
-If you run the program through Harmony, it will find not problems.
+If you run the program through Harmony, it will not find any problems
+because there is only one execution possible and 1) in that execution
+the assertion does not fail and 2) the execution terminates.
 Program (b) is concurrent.  It executes methods \texttt{f()} and
@@
 an array \textit{done} with two booleans.
-The program then spawns three threads.
-The first runs \texttt{incrementer(0)}, the second runs
+The program then spawns two threads.
+The first runs \texttt{incrementer(0)}; the second runs
 \texttt{incrementer(1)}.
@@
 It then waits until the other thread is done.
-(\textbf{await} $c$ is semantically the same as \textbf{while not} $c$: \textbf{pass}.)
+(\textbf{await} $c$ is shorthand for \textbf{while not} $c$: \textbf{pass}) and semantically the same.
 After that, method \texttt{incrementer}
-verifies that the value of \textit{count} equals 2.
+verifies that the value of \textit{count} equals~2.
 % Like Python, Harmony supports \texttt{assert} statements with two arguments:
@@
 values into memory.  Importantly, it does not have instructions to atomically
-increment or decrement values in memory locations.
+increment or decrement values in shared memory locations.
 So to increment a value in memory,
@@
 twice and ends up being 2.  However, the following is also
-a possible interleaving of incrementers~0 and~1:
+a possible interleaving of \texttt{incrementer(0)} and \texttt{incrementer(1)}:
 \begin{enumerate}
-\item incrementer 0 loads the value of \textit{count}, which is 0;
-\item incrementer 1 loads the value of \textit{count}, which is still 0;
-\item incrementer 1 adds one to the value that it loaded (0), and
+\item \texttt{incrementer(0)} loads the value of \textit{count}, which is 0;
+\item \texttt{incrementer(1)} loads the value of \textit{count}, which is still 0;
+\item \texttt{incrementer(1)} adds one to the value that it loaded (0), and
 stores $1$ into \textit{count};
-\item incrementer 0 adds one to the value that it loaded (0), and
+\item \texttt{incrementer(0)} adds one to the value that it loaded (0), and
 stores $1$ into \textit{count};
-\item incrementer 0 sets \textit{done}[0] to \texttt{True};
-\item incrementer 1 sets \textit{done}[1] to \texttt{True}.
+\item \texttt{incrementer(0)} sets \textit{done}[0] to \texttt{True};
+\item \texttt{incrementer(1)} sets \textit{done}[1] to \texttt{True}.
 \end{enumerate}
@@
 \end{code}
-\caption{\harmonylink{modules/syncS.hny} The binary semaphore interface in the \texttt{synchS} module uses suspension.}
+\caption{\harmonylink{modules/synchS.hny} The binary semaphore interface in the \texttt{synchS} module uses suspension.}
 \label{fig:suspension}

### 2021-01-31 book/paper.tex
More updates

@@
 However, if the model checker does not find any bugs, we do not
-know for sure that the algorithm works for lists with more than
+know for sure that the algorithm works for lists of more than
 five numbers or for lists that have values other than the numbers
@@
 \item \autoref{ch:peterson} introduces \emph{Peterson's Algorithm}, an
-elegant solution to implementating a critical section.
+elegant (although not very efficient or practical) solution to implementating a critical section.
 \item \autoref{ch:method} gives some more details on the Harmony
 language needed for the rest of the book.
-\item \autoref{ch:spinlock} introduces ``hardware'' \emph{locks}
+\item \autoref{ch:spinlock} introduces atomic \emph{locks}
 for implemented critical sections.
 \item \autoref{ch:synch} presents Harmony modules
-that supports locks and other synchronization primitives.
+that support locks and other synchronization primitives.
 \item \autoref{ch:cds} gives an introduction to building concurrent
@@
 \item \autoref{ch:interrupts} discusses how to handle interrupts,
-a problem closely related to synchronization among multiple threads.
+a problem closely related to---but not the same as---synchronizing threads.
 \item \autoref{ch:nonblocking} introduces \emph{non-blocking} or
@@
 
+Harmony is a programming language that borrows much of
+Python's syntax.
 Like Python, Harmony is an imperative,
@@
 \item Python is object-oriented, supporting classes with methods and
-inheritance; Harmony has objects but does not support classes.  Harmony supports
-pointers to objects and methods.
+inheritance; Harmony has objects but does not support classes.
+On the other hand, Harmony supports pointers to objects and methods.
 \end{itemize}
@@
 argument.  The \{ \textit{x..y} \} notation generates a set containing the numbers
-from~\textit{x} to~\textit{y} (inclusive).  The last two lines in the program are
+from~\textit{x} to~\textit{y} (inclusive).
+(Harmony does not have iterators and in particular does
+not have a \texttt{range} operator.)
+The last two lines in the program are
 the most interesting.

### 2021-01-31 book/paper.tex
Some fixes

@@
 \item \texttt{incrementer(1)} loads the value of \textit{count}, which is still 0;
-\item \texttt{incrementer(1)} adds one to the value that it loaded (0), and
+\item \texttt{incrementer(1)} adds 1 to the value that it loaded (0), and
 stores $1$ into \textit{count};
-\item \texttt{incrementer(0)} adds one to the value that it loaded (0), and
+\item \texttt{incrementer(0)} adds 1 to the value that it loaded (0), and
 stores $1$ into \textit{count};

### 2021-01-31 book/paper.tex
Pass over Chapter 4

@@
 The empty tuple is written as \texttt{()}.
-Note that this is different from the empty set, which is \texttt{\{\}}.
+% Note that this is different from the empty set, which is \texttt{\{\}}.
 As in Python, you can create singleton tuples by including a comma.
@@
 \index{multiset}%
-of contexts.  It is a multiset of contexts because two different threads can
+of contexts.  It is a multiset of contexts because two threads can
 have the same context.
@@
 which initializes the state.
-It starts executing at instruction 0 and keeps executing until the last
-execution in the program.
+It starts executing at instruction 0 and keeps executing until
+it reaches the last instruction in the program.
 In this case, it executes instructions 0 through 5 first.
-The last instruction is a \texttt{JUMP} instruction that sets the
-program counter to 32 (skipping over the code for \texttt{incrementer} method).
-At program counter 6 is the code for the \texttt{incrementer} method.
-All methods start with a \texttt{Frame} instruction and end with a
-\texttt{Return} instruction.
+The last instruction in that sequence is a \texttt{JUMP}
+instruction that sets the program counter to 32
+(skipping over the code for \texttt{incrementer} method).
+The \texttt{\_\_init\_\_} thread then executes instructions
+32 through 40 and finishes.
+Once initialization completes, any threads that were spawned
+(in this case \texttt{incrementer(0)} and \texttt{incrementer(0)})
+can run.
 
@@
 
+At program counter 6 is the code for the \texttt{incrementer} method.
+All methods start with a \texttt{Frame} instruction and end with a
+\texttt{Return} instruction.
 \autoref{ap:harmonybytecode} provides a list of all HVM machine instructions,
@@
 
-Once initialization completes, any threads that were spawned can run.
 You can think of Harmony as trying every possible interleaving of threads executing
@@
 
-Harmony can report a variety of failure types:
+Harmony can report the following failure types:
 \begin{itemize}
@@
 at that state, along with their program counters.
-\item \texttt{Stopped States}: Similar to non-terminating states,
-these are states in which some threads are left that cannot
-terminate.  \autoref{ch:synch} will explain what it means for
-a thread to be stopped.
+% \item \texttt{Stopped States}: Similar to non-terminating states,
+% these are states in which some threads are left that cannot
+% terminate.  \autoref{ch:synch} will explain what it means for
+% a thread to be stopped.
 \end{itemize}
 
-In our case, Harmony reports a safety violation, in particular it
-reports that the assertion and that
-\textit{count} has value 1 instead of 2.
+In our case, Harmony reports a safety violation.
+In particular, it reports that the assertion failed
+because \textit{count} has value 1 instead of 2.
 Next it reports an execution that failed this assertion.
-The program got to the faulty execution in 4 ``steps.''
+The program got to the failed assertion in 4 ``steps.''
 The output has four columns:
@@
 \item Thread \texttt{incrementer(0)} (with identifier T1)
-executed instructions 6 through 9, loading the value of
+executes instructions 6 through 9, loading the value of
 \textit{count} but stopping just before storing 1 into \textit{count};
 \item Thread \texttt{incrementer(1)} (with identifier T2)
-executed instructions 6 through 18, storing 1 into
-into \textit{count} and storing \texttt{True} into \textit{done}[1];
+executes instructions 6 through 18, storing 1 into
+\textit{count} and storing \texttt{True} into \textit{done}[1];
 \item Thread \texttt{incrementer(0)}
@@
 instructions, the HTML file contains a list of blocks for each
-executed instruction.  We call this the \emph{time line}.
+executed instruction.  We call this the \emph{timeline}.
 You can click on such a block to see the state of the Harmony
-virtual machine just after executing that instruction.
+virtual machine just after executing the corresponding instruction.
 The step that is being executed is highlighted in green.
@@
 As mentioned above, you can go to any point in the execution by
-clicking on one of the blocks in the time lines.
+clicking on one of the blocks in the timeline.
 There are also various handy keyboard shortcuts:

### 2021-01-31 book/paper.tex
Pass over Ch 5

@@
 at the beginning and at the end of the critical section, but in the middle
-the invariant may be temporarily broken---this is no issue as critical
+the invariant may be temporarily broken---this is not a problem as critical
 sections guarantee that no other
@@
 believes it can acquire the lock, and stores \texttt{True} in
-\texttt{lockTaken} and moves on to the critical section.
-Finally, thread~0 moves on, also stores \texttt{True} and also moves
+\textit{lockTaken} and moves on to the critical section.
+Finally, thread~0 moves on, also stores \texttt{True}
+into \textit{lockTaken} and also moves
 into the critical section.  Thread~0 is the one that detects the
@@
 does have a problem: if both try to enter the critical section at the same
-time, they may end up waiting for one another indefinitely.  Thus the
+time, they may end up waiting for one another indefinitely.
+(This is a form of \emph{deadlock}, which will be discussed in
+\autoref{ch:deadlock}.)
+Thus the
 solution violates \emph{progress}.
@@
 thread $1-i$, politely, ends up waiting indefinitely for its turn.
-Too bad, it would have been a great solution if both threads try to enter the
-critical section continually.
+Too bad, because it would have been a great solution if both
+threads try to enter the critical section ad infinitum.
 
@@
 mutual exclusion and progress---as
-we will find out, it is possible but not easy.
+we will find out, it is possible but not obvious.
 \end{problems}

### 2021-02-01 book/paper.tex
Latest

@@
   description={is the execution of a machine instruction by a thread, updating
-        its state.  Harmony distinguishes micro steps (machine instructions) and macro
-        step (a sequence of instructions with at most one effect visible by
-        other threads)}
+        its state.}
 }
@@
 Next it reports an execution that failed this assertion.
-The program got to the failed assertion in 4 ``steps.''
+The program got to the failed assertion in 4 ``turns.''
 The output has four columns:
@@
 There are four sections.
-The top shows the steps to the problematic state.
-Each step shows a thread identifier, the method the thread is executing,
+The top shows the turns to the problematic state.
+Each turn shows a thread identifier, the method the thread is executing,
 a ``timeline'' containing a block for each instruction executed, the
 program counter of the thread, and the values of the shared variables
-at those steps.
+at those turns.
 The middle section shows the line of code that is being executed.
@@
 
-The four steps in the execution are as follows:
+The four turns in the execution are as follows:
 
@@
 In the top right, the HTML file contains the reported issue in red.
-Underneath it, a table shows the four steps in the execution.
+Underneath it, a table shows the four turns in the execution.
 Instead of listing explicitly the program counters of the executed
@@
 virtual machine just after executing the corresponding instruction.
-The step that is being executed is highlighted in green.
-The table also lists the program counter of the thread at each step,
+The turn that is being executed is highlighted in green.
+The table also lists the program counter of the thread at each turn,
 and the values of the shared variables.
@@
 \emph{Left arrow}: & go to the previous instruction; \\
-\emph{Down arrow}: & go to the next step; \\
-\emph{Up arrow}: & go to the previous step; \\
+\emph{Down arrow}: & go to the next turn; \\
+\emph{Up arrow}: & go to the previous turn; \\
 \emph{Enter (aka Return)}: & go to the next line of Harmony code; \\
@@
 \index{non-determinism}%
-in Harmony.\footnote{Actually, there is a third type called
-an \emph{interrupt}, discussed in \autoref{ch:interrupts}.}
+in Harmony.
+% \footnote{Actually, there is a third type called
+% an \emph{interrupt}, discussed in \autoref{ch:interrupts}.}
 One is when
@@
 if it is either the initial state
-or it can be reached from the initial state through a trace.
+or it can be reached from the initial state through a finite trace.
 A state is final
@@
 Also, in state $s$, $\mathit{turn} = 1$.  Now let
-thread~1 make a sequence of steps.  Because $\mathit{turn} = 1$,
+thread~1 make a step.  Because $\mathit{turn} = 1$,
 thread~1 will stop waiting and also enter the critical
@@
 \item states in which $\lnot \mathtt{thread}(i)@[8 \cdots 15]$
+(because false implies anything)
 \end{enumerate}
@@
 $\mathit{turn} = 1$, and thread 1 just finished the step in line 7,
-setting $\mathit{flags}[1]$ to true.  Notice that $C(0)$ is violated.
+setting $\mathit{flags}[1]$ to true.  Notice that $C(0)$ is now violated.
 
@@
 Instead, we will use the following property: $\mathcal{G}(i) =
-\mathtt{thread}(i)@cs \Rightarrow \mathcal{C}(i) \lor \mathtt{thread}(1-i)@[8]$.
+\mathtt{thread}(i)@\mathtt{cs} \Rightarrow \mathcal{C}(i) \lor \mathtt{thread}(1-i)@\mathtt{gate}$.
 That is, if thread $i$ is at the critical section, then
@@
 We claim that $\mathcal{G}(i)$ is an inductive invariant.
-First, since neither thread exists in the initial state, it is clear that
-$\mathcal{G}(i)$ holds in the initial state.
+First, since neither thread in in the critical section in the initial state,
+it is clear that $\mathcal{G}(i)$ holds in the initial state.
 Without loss of generality, suppose $i=0$ (a benefit from the fact that the algorithm is
@@
 that, particularly in concurrent programming, such control flow directions are highly
-error-prone.  It's easy to forget to, say, release a lock when returning a value in the
+error-prone.  It's too easy to forget to, say, release a lock when returning a value in the
 middle of a method, a major source of bugs in practice.
@@
 \texttt{P\_mutex} is a method that returns a ``mutex,'' which, in this
-case, is a dictionary that contains Peterson's Algorithm's state:
+case, is a dictionary that contains Peterson's Algorithm's shared memory state:
 a turn variable and two flags.
@@
 Another invariant, $\mathcal{I}_2(i)$, is that if thread $i$ is executing between
-lines $18 \cdots 25$ (which includes the critical section),
+lines $18 \cdots 22$ (which includes the critical section),
 then \textit{private}[$i$] == \texttt{False}.

### 2021-02-01 book/paper.tex
Pass over Chapters 1-10

@@
 A thread usually does not keep explicit track of whether it has a lock
-or not because it is usually implied by the control flow of the program.
+or not because it is usually implied by the control flow of the program---a
+thread implicitly \emph{knows} that when it is executing in a critical
+section it has the lock.
 So there is no need to keep \textit{private} as a shared
@@
 include compare-and swap and fetch-and-add.  \texttt{atomic} statements
-are not intended to \emph{replace} locks or other synchonization primitives.
+are not intended to replace locks or other synchonization primitives.
 When solving synchronization problems you should not directly use
@@
 Notice that the module hides the implementation of
-binary semaphores.
+binary semaphores and locks.
 The \texttt{synch} module includes a variety of other useful
@@
 A context is a regular (if complex) Harmony value.
-The syntax of the expression is as follows:
+The syntax of the expression that a thread executes to suspend itself
+is as follows:
 
@@
 that uses suspension instead of spinlocks.
-\autoref{fig:suspension} shows the same interface implemented
-using suspension.
+\autoref{fig:suspension} shows the same interface
+as \autoref{fig:spinlocks} implemented using suspension.
 The interface is implemented as follows:
@@
 method models going to a particular ATM and withdrawing money not exceeding
-the balance.  Method \texttt{checker} looks for negative balances.
+the balance.
 Running the code through Harmony reveals that there is a bug.
@@
 and a \texttt{tail} pointer to the last element in the list.
-The \texttt{head} pointer is \textbf{None} if and only if the queue is empty.
-(\textbf{None} is a special address value that is not the address of any
+The \texttt{head} pointer is \texttt{None} if and only if the queue is empty.
+(\texttt{None} is a special address value that is not the address of any
 memory location.)
 
-\texttt{queue.Queue()} returns a new queue object consisting of a \textbf{None} head
+\texttt{queue.Queue()} returns a new queue object consisting of a \texttt{None} head
 and tail pointer and a lock.
@@
 Note that neither the \texttt{head} nor \texttt{tail} pointer are
-ever \textbf{None}.
+ever \texttt{None}.
 

### 2021-02-01 book/paper.tex
Pass over Chapter 11
Made pass over Ch 12

@@
 are simply reading the data structure.
-In many applications, reads are the majority of all accesses.
+In many applications, reads are the majority of all accesses,
+and read operations do not conflict with one another.
 Allowing reads to proceed concurrently can significantly improve performance.
@@
 proposed by Dijkstra~\cite{EWD329}.
+Multiple producers and multiple consumers may all share
+the same bounded buffer.
 
@@
 consume responses, while the server consumes requests and produces responses.
-%
-Multiple producers and multiple consumers may all share
-the same bounded buffer.
 
@@
 \item At any time, at most one gate is open;
-\item If some gate is open, then no thread is in the critical section;
-\item If some thread is in the critical section, all gates are closed;
+\item If some gate is open, then no thread is in the critical section.  Equivalently, if some thread is in the critical section, all gates are closed;
 \item At any time, at most one thread is in the critical section.
@@
 \item \textit{nwriters}: the number of writers in the critical section;
-\item \textit{mutex}: the ``main gate'' binary semaphore;
-\item \textit{r\_gate}: the ``waiting gate'' used by readers, consisting of a binary semaphore and the number of readers waiting to enter;
-\item \textit{w\_gate}: the ``waiting gate'' used by writers, similar to the readers' gate.
+\item \textit{mutex}: the main gate binary semaphore;
+\item \textit{r\_gate}: the waiting gate used by readers, consisting of a binary semaphore and the number of readers waiting to enter;
+\item \textit{w\_gate}: the waiting gate used by writers, similar to the readers' gate.
 \end{itemize}
@@
 the readers' gate, leaves the critical section (\texttt{release\_one}),
-and then tries to acquire the semaphore associated with the gate.
+and then tries to acquire the semaphore associated with the waiting gate.
 This will cause it to block until some thread opens that gate.
@@
 \textit{mutex} but the readers' gate's binary semaphore.
-\item One of the waiting reader can now re-enter the critical section.
-When it does, the reader decrements the gate's counter (from 2 to 1),
+\item One of the waiting readers can now re-enter the critical section.
+When it does, the reader decrements the gate's counter (from 2 to 1)
 and increments \textit{nreaders} (from~0 to~1).
@@
 (Similarly, you may want to try to implement an \emph{upgrade} of a
-read lock to a write lock.)
+read lock to a write lock.  Why is this problematic?)
 % How easy or hard would it be to do the same for the code in

### 2021-02-01 book/paper.tex
Pass over Ch 13
Latest

@@
 \autoref{ex:onelane}.
-If you want to use the queue method, you can change the signature of the
-methods to enter the bridge, adding an argument that contains a pointer to a
-semaphore.
+% If you want to use the queue method, you can change the signature of the
+% methods to enter the bridge, adding an argument that contains a pointer to a
+% semaphore.
 \end{problems}
@@
 a set of variables and methods that operate on those variables.
-There is a split binary semaphore associated with each such class.
-The \textit{mutex} is hidden: it is automatically acquired when invoking a
-method and released upon exit.
-The other semaphores are encapsulated in \emph{condition variables}.
-\index{condition variable}%
-Each condition variable also keeps track of the number of threads waiting
-on the condition variable.
-
+A monitor also has special variables called \emph{condition variables},
+one per waiting condition.
 There are two operations on condition variables: \texttt{wait}
@@
 \index{signal}%
-\autoref{fig:hoare} presents the semantics by giving an implementation of
-\texttt{wait} and \texttt{signal} as well as the code that is used to enter
-and exit monitor methods.
+
+Harmony does not have language support for Hoare monitors but it
+has a module called \texttt{hoare}.
+\autoref{fig:hoare} shows its implementation.
+A monitor uses a hidden split binary semaphore.
+The mutex semaphore is acquired when entering a monitor and released
+upon exit.
+Each condition variable maintains a binary semaphore and a counter for
+the number of threads waiting on the condition.
 \texttt{wait} increments the condition's counter, releases the monitor mutex,
@@
 \autoref{fig:hoaretest} presents a bounded buffer implemented using
-Hoare monitors.  It is written in much the same way you would if using
+Hoare monitors.
+It is written in much the same way you would if using
 the SBS technique (see \autoref{ex:bbsbs}).

### 2021-02-01 book/paper.tex
Pass over Ch 14

@@
 \texttt{wait} increments the condition's counter, releases the monitor mutex,
-blocks while trying to acquire the condition's semaphore, and after resuming
+blocks while trying to acquire the condition's semaphore, and upon resuming
 decrements the counter---in much the same way as we have seen for split binary
@@
 the SBS technique (see \autoref{ex:bbsbs}).
-There is no \texttt{release\_one} method.  Instead, one can conclude
+However, there is no \texttt{release\_one} method.
+Instead, one can conclude
 that \texttt{put} guarantees that the queue will be non-empty, and
@@
 solution would be to peek at the counters inside the respective condition
-variable, but that breaks the abstraction.  The alternative is for the
+variables, but that breaks the abstraction.  The alternative is for the
 reader/writer implementation to keep track of that state explicitly,
@@
 
-As in Hoare monitors, there is a hidden mutex associated with each monitor,
-and the mutex is automatically acquired upon entry to a method and released
+As in Hoare monitors, there is a hidden mutex associated with each Mesa monitor,
+and the mutex must be acquired upon entry to a method and released
 upon exit.
@@
 until it leaves the monitor explicitly
-(by calling either \texttt{release()} or \texttt{wait()}).
+(by calling either \texttt{release} or \texttt{wait}).
 At that point, any thread that
@@
 with other threads trying to enter the critical section.
-Basically, there is just one gate to enter the critical section, instead
-of a main gate and a gate per waiting condition.
 
@@
 
+Basically, there is just one gate to enter the critical section, instead
+of a main gate and a gate per waiting condition.
 This is a very important difference.  In Hoare monitors, when a thread
@@
 Mesa monitors also allow notifying multiple threads.
-For example, a thread can invoke \texttt{notify()} twice---if there are
+For example, a thread can invoke \texttt{notify} twice---if there are
 two or more threads waiting on the condition variable, two will be resumed.
@@
 
-Harmony provides support for monitors
+Harmony provides support for Mesa monitors
 through the Harmony \texttt{synch} module.
@@
 the condition that the thread is waiting for.  Many implementation
-of condition variables depend on this, and optimized implementations
+of Mesa condition variables depend on this, and optimized implementations
 of condition variables often allow so-called ``spurious wakeups,''
@@
 available, but it should grant them as soon as they are available.
+It returns a set of $n$ GPU identifiers.
 Method \texttt{gpuReleaseSet(s)} takes a set of GPU identifiers as argument.
@@
 algorithm, and \autoref{fig:qsorttest} an accompanying test program.
-The array to be sorted is stored in shared variable \textit{textqs.arr}.
-Another shared variable, \textit{testqs.todo}, contains the ranges of the
+The array to be sorted is stored in shared variable \textit{textqs}.\texttt{arr}.
+Another shared variable, \textit{testqs}.\texttt{todo}, contains the ranges of the
 array that need to be sorted (initially the entire array).

### 2021-02-01 book/paper.tex
Latest

@@
 or non-existent variable, dividing a set by an integer, and so on.
-It also includes certain infinite loops.
+% It also includes certain infinite loops.
 Harmony will print a trace of the shortest bad execution that it found.
@@
 The only way forward is if the other thread releases the spinlock.
-A thread that is in an infinite loop is also considered blocked,
-although Harmony can often detect infinite loops specifically.
+A thread that is in an infinite loop is also considered blocked.
+% although Harmony can often detect infinite loops specifically.
 
@@
 reading a non-existent key in a dictionary, accessing a non-existent
-variable, infinite loops of microsteps, as well as when an assertion fails;
+variable, as well as when an assertion fails;
 \item The execution of a thread in \emph{atomic mode} does not get interleaved

### 2021-02-01 src/charm/charm.c

@@ -151,7 +151,8 @@ void check_invariants(struct node *node, struct context **pctx){
 }
 
 void onestep(struct node *node, uint64_t ctx, uint64_t choice, bool interrupt,
-        struct dict *visited, struct queue *todo, struct context **pinv_ctx){
+        struct dict *visited, struct queue *todo, struct context **pinv_ctx,
+        bool infloop_detect){
     // Make a copy of the state
     struct state *sc = new_alloc(struct state);
     memcpy(sc, node->state, sizeof(*sc));
@@ -173,13 +174,13 @@ void onestep(struct node *node, uint64_t ctx, uint64_t choice, bool interrupt,
         interrupt_invoke(&cc);
     }
     else if (sc->choosing == 0 && cc->trap_pc != 0 && !cc->interruptlevel) {
-        onestep(node, ctx, choice, true, visited, todo, pinv_ctx);
+        onestep(node, ctx, choice, true, visited, todo, pinv_ctx, infloop_detect);
     }
 
     // Copy the choice
     uint64_t choice_copy = choice;
 
-    bool choosing = false;
+    bool choosing = false, infinite_loop = false;;
     struct dict *infloop = NULL;        // infinite loop detector
     for (int loopcnt = 0;; loopcnt++) {
         int pc = cc->pc;
@@ -213,7 +214,7 @@ void onestep(struct node *node, uint64_t ctx, uint64_t choice, bool interrupt,
             (*oi->op)(code[pc].env, sc, &cc);
         }
 
-        if (cc->phase != CTX_END && cc->failure == 0 && loopcnt > 1000) {
+        if (cc->phase != CTX_END && cc->failure == 0 && (infloop_detect || loopcnt > 1000)) {
             if (infloop == NULL) {
                 infloop = dict_new(0);
             }
@@ -224,13 +225,21 @@ void onestep(struct node *node, uint64_t ctx, uint64_t choice, bool interrupt,
             combo->state = *sc;
             memcpy(&combo->context, cc, sizeof(*cc) + stacksize);
             void **p = dict_insert(infloop, combo, combosize);
+            free(combo);
             if (*p == (void *) 0) {
                 *p = (void *) 1;
             }
-            else {
+            else if (infloop_detect) {
                 cc->failure = value_put_atom("infinite loop", 13);
+                infinite_loop = true;
             }
-            free(combo);
+            else {
+                // start over, as twostep does not have loopcnt optimization
[…]

### 2021-02-02 book/paper.tex
Latest

@@
 \hline
-ctxbag & a bag of non-stopped contexts \\
+ctxbag & a bag of runnable contexts \\
 \hline
@@
 choosing & if not \texttt{None}, indicates a context that is choosing \\
-\hline
-initializing & boolean indicating that the state is not fully initialized. \\
 \hline
@@
 \texttt{\_\_init\_\_/()} and program counter 0.  It starts executing
-in atomic mode until it finishes executing the last instruction in
-the code.  All states until then are \texttt{initializing} states.
+in atomic mode until it finishes executing the last
+\texttt{Return} instruction.
 Other threads, created through \texttt{spawn} statements, do not
@@
 
-A \emph{micro step}
-is the execution of a single HVM machine instruction
+A \emph{step} is the execution of a single HVM machine instruction
 by a context.
-Each micro step generates a new state.
+Each step generates a new state.
 When there are multiple contexts, the HVM can interleave them.
-However, trying to interleave every microstep would be needlessly expensive,
-as many micro steps involve changes to a context that are invisible to
+However, trying to interleave every step would be needlessly expensive,
+as many steps involve changes to a context that are invisible to
 other contexts.
 
-A \emph{macro step}
-\index{macro step}%
-can involve multiple micro steps.  The following
-instructions start a new macro step: \texttt{Load}, \texttt{Store},
+A \emph{stride}
+\index{stride}%
+can involve multiple steps.  The following
+instructions start a new stride: \texttt{Load}, \texttt{Store},
 \texttt{AtomicInc}, and \texttt{Continue}.  The HVM
-interleaves macro steps, not micro steps.  Like micro steps, each
-macro step involves a single context.  Unlike a micro step, a macro
-step can leave the state unchanged.
+interleaves stides, not steps.  Like steps, each
+stride involves a single context.  Unlike a step, a stride
+can leave the state unchanged (because its steps lead back
+to where the stride started).
 
 Executing a Harmony program results in a graph where the nodes are Harmony
-states and the edges are macro steps.
+states and the edges are strides.
 When a state is \texttt{choosing}, the edges from that state are
@@
 
-Consecutive macro steps by the same thread are called \emph{mega steps}.
+Consecutive strides by the same thread are called a \emph{turn}.
 Each state maintains the shortest path to it from the initial state in terms
-of mega steps.
-The diameter of the graph is the length of the longest path found.
+of turns.
+The diameter of the graph is the length of the longest path found in
+terms of turns.
 

### 2021-02-02 book/paper.tex
Latest

@@
 $ harmony triangle.hny
-#states = 13 diameter = 1
-#components: 13
-no issues found
+#states 13
+13 components, 0 bad states
+No issues
 $
@@
 $ harmony -c N=100 triangle.hny
-#states = 103 diameter = 1
-#components: 103
-no issues found
+#states 103
+103 components, 0 bad states
+No issues
 $
@@
 acquired a given number of times.  It is supported by the \texttt{synch}
-module.
-
-This concept can be generalized using something called the
+module.  The \texttt{P} or ``procure'' operation acquires a counting
+semaphore.  That is, it tries to decrement the semaphore, blocking while
+the semaphore has a value of 0.  The \texttt{V} or ``vacate'' operation
+increments the semaphore.
+
+This avoidance technique can be generalized using something called the
 Banker's Algorithm~\cite{EWD108}, but it is outside the scope of this book.

### 2021-02-02 book/sources/BBhoare.tex
Latest

\>{\tiny 1}\'\>\texttt{\textbf{import}}~hoare\\

\>{\tiny 2}\'\>\\

\>{\tiny 3}\'\>\texttt{\textbf{def}}~BB(\textit{size}):\\

\>{\tiny 4}\'\>~~~~\textit{result}~=~\{\\

\>{\tiny 5}\'\>~~~~~~~~~~~~.mon:~hoare.Monitor(),\\

\>{\tiny 6}\'\>~~~~~~~~~~~~.prod:~hoare.Condition(),~.cons:~hoare.Condition(),\\

\>{\tiny 7}\'\>~~~~~~~~~~~~.buf:~\{~\textit{x}:()~\texttt{\textbf{for}}~\textit{x}~\texttt{\textbf{in}}~\{1..\textit{size}\}~\},\\

\>{\tiny 8}\'\>~~~~~~~~~~~~.head:~1,~.tail:~1,\\

\>{\tiny 9}\'\>~~~~~~~~~~~~.count:~0,~.size:~size\\

\>{\tiny 10}\'\>~~~~~~~~\}\\

\>{\tiny 11}\'\>~~~~\\

\>{\tiny 12}\'\>\texttt{\textbf{def}}~put(\textit{bb},~\textit{item}):\\

\>{\tiny 13}\'\>~~~~hoare.enter(?\textit{bb}$\rightarrow$mon)\\

\>{\tiny 14}\'\>~~~~\texttt{\textbf{if}}~\textit{bb}$\rightarrow$count~==~\textit{bb}$\rightarrow$size:\\

\>{\tiny 15}\'\>~~~~~~~~hoare.wait(?\textit{bb}$\rightarrow$prod,~?\textit{bb}$\rightarrow$mon)\\

\>{\tiny 16}\'\>~~~~\textit{bb}$\rightarrow$buf[\textit{bb}$\rightarrow$tail]~=~\textit{item}\\

\>{\tiny 17}\'\>~~~~\textit{bb}$\rightarrow$tail~=~(\textit{bb}$\rightarrow$tail~\%~\textit{bb}$\rightarrow$size)~+~1\\

\>{\tiny 18}\'\>~~~~\textit{bb}$\rightarrow$count~+=~1\\

\>{\tiny 19}\'\>~~~~hoare.signal(?\textit{bb}$\rightarrow$cons,~?\textit{bb}$\rightarrow$mon)\\

\>{\tiny 20}\'\>~~~~hoare.exit(?\textit{bb}$\rightarrow$mon)\\

\>{\tiny 21}\'\>\\

\>{\tiny 22}\'\>\texttt{\textbf{def}}~get(\textit{bb}):\\

\>{\tiny 23}\'\>~~~~hoare.enter(?\textit{bb}$\rightarrow$mon)\\

\>{\tiny 24}\'\>~~~~\texttt{\textbf{if}}~\textit{bb}$\rightarrow$count~==~0:\\

\>{\tiny 25}\'\>~~~~~~~~hoare.wait(?\textit{bb}$\rightarrow$cons,~?\textit{bb}$\rightarrow$mon)\\

\>{\tiny 26}\'\>~~~~\textit{result}~=~\textit{bb}$\rightarrow$buf[\textit{bb}$\rightarrow$head]\\

\>{\tiny 27}\'\>~~~~\textit{bb}$\rightarrow$head~=~(\textit{bb}$\rightarrow$head~\%~\textit{bb}$\rightarrow$size)~+~1\\

\>{\tiny 28}\'\>~~~~\textit{bb}$\rightarrow$count~--=~1\\

\>{\tiny 29}\'\>~~~~hoare.signal(?\textit{bb}$\rightarrow$prod,~?\textit{bb}$\rightarrow$mon)\\

\>{\tiny 30}\'\>~~~~hoare.exit(?\textit{bb}$\rightarrow$mon)

### 2021-02-02 book/sources/PetersonInductive.tex
Latest

\>{\tiny 1}\'\>\textit{flags}~=~[~\texttt{\textbf{False}},~\texttt{\textbf{False}}~]\\

\>{\tiny 2}\'\>\textit{turn}~=~\texttt{\textbf{choose}}(\{0,~1\})\\

\>{\tiny 3}\'\>\\

\>{\tiny 4}\'\>\texttt{\textbf{def}}~thread(\textit{self}):\\

\>{\tiny 5}\'\>~~~~\texttt{\textbf{while}}~\texttt{\textbf{choose}}(\{~\texttt{\textbf{False}},~\texttt{\textbf{True}}~\}):\\

\>{\tiny 6}\'\>~~~~~~~~\emph{\# Enter critical section}\\

\>{\tiny 7}\'\>~~~~~~~~\textit{flags}[\textit{self}]~=~\texttt{\textbf{True}}\\

\>{\tiny 8}\'\>~~~~~~~~@gate:~\textit{turn}~=~1~--~\textit{self}\\

\>{\tiny 9}\'\>~~~~~~~~\texttt{\textbf{await}}~(\texttt{\textbf{not}}~\textit{flags}[1~--~\textit{self}])~\texttt{\textbf{or}}~(\textit{turn}~==~\textit{self})\\

\>{\tiny 10}\'\>\\

\>{\tiny 11}\'\>~~~~~~~~\emph{\# Critical section}\\

\>{\tiny 12}\'\>~~~~~~~~@cs:~\texttt{\textbf{assert}}~(\texttt{\textbf{not}}~\textit{flags}[1~--~\textit{self}])~\texttt{\textbf{or}}~(\textit{turn}~==~\textit{self})~\texttt{\textbf{or}}\\

\>{\tiny 13}\'\>~~~~~~~~~~~~~~~~~~~~(\texttt{\textbf{atLabel}}(gate)~==~\{(thread,~1~--~\textit{self}):~1\})\\

\>{\tiny 14}\'\>\\

\>{\tiny 15}\'\>~~~~~~~~\emph{\# Leave critical section}\\

\>{\tiny 16}\'\>~~~~~~~~\textit{flags}[\textit{self}]~=~\texttt{\textbf{False}}\\

\>{\tiny 17}\'\>\\

\>{\tiny 18}\'\>\texttt{\textbf{spawn}}~thread(0)\\

\>{\tiny 19}\'\>\texttt{\textbf{spawn}}~thread(1)

### 2021-02-02 book/sources/PetersonMethod.tex
Latest

\>{\tiny 1}\'\>\texttt{\textbf{def}}~P\_enter(\textit{pm},~\textit{pid}):\\

\>{\tiny 2}\'\>~~~~\textit{pm}$\rightarrow$flags[\textit{pid}]~=~\texttt{\textbf{True}}\\

\>{\tiny 3}\'\>~~~~\textit{pm}$\rightarrow$turn~=~1~--~\textit{pid}\\

\>{\tiny 4}\'\>~~~~\texttt{\textbf{await}}~(\texttt{\textbf{not}}~\textit{pm}$\rightarrow$flags[1~--~\textit{pid}])~\texttt{\textbf{or}}~(\textit{pm}$\rightarrow$turn~==~\textit{pid})\\

\>{\tiny 5}\'\>\\

\>{\tiny 6}\'\>\texttt{\textbf{def}}~P\_exit(\textit{pm},~\textit{pid}):\\

\>{\tiny 7}\'\>~~~~\textit{pm}$\rightarrow$flags[\textit{pid}]~=~\texttt{\textbf{False}}\\

\>{\tiny 8}\'\>\\

\>{\tiny 9}\'\>\texttt{\textbf{def}}~P\_mutex():\\

\>{\tiny 10}\'\>~~~~\textit{result}~=~\{~.turn:~\texttt{\textbf{choose}}(\{0,~1\}),~.flags:~[~\texttt{\textbf{False}},~\texttt{\textbf{False}}~]~\}\\

\>{\tiny 11}\'\>\\

\>{\tiny 12}\'\>\emph{\#\#\#\# The code above can go into its own Harmony module \#\#\#\#}\\

\>{\tiny 13}\'\>\\

\>{\tiny 14}\'\>\textit{mutex}~=~P\_mutex()\\

\>{\tiny 15}\'\>\\

\>{\tiny 16}\'\>\texttt{\textbf{def}}~thread(\textit{self}):\\

\>{\tiny 17}\'\>~~~~\texttt{\textbf{while}}~\texttt{\textbf{choose}}(\{~\texttt{\textbf{False}},~\texttt{\textbf{True}}~\}):\\

\>{\tiny 18}\'\>~~~~~~~~P\_enter(?\textit{mutex},~\textit{self})\\

\>{\tiny 19}\'\>~~~~~~~~@cs:~\texttt{\textbf{assert}}~\texttt{\textbf{atLabel}}(cs)~==~\{~(thread,~\textit{self}):~1~\}\\

\>{\tiny 20}\'\>~~~~~~~~P\_exit(?\textit{mutex},~\textit{self})\\

\>{\tiny 21}\'\>\\

\>{\tiny 22}\'\>\texttt{\textbf{spawn}}~thread(0)\\

\>{\tiny 23}\'\>\texttt{\textbf{spawn}}~thread(1)

### 2021-02-02 book/sources/RWfair.tex
Latest

\>{\tiny 1}\'\>\texttt{\textbf{from}}~synch~\texttt{\textbf{import}}~BinSema,~acquire,~release\\

\>{\tiny 2}\'\>\\

\>{\tiny 3}\'\>\texttt{\textbf{def}}~RWlock():\\

\>{\tiny 4}\'\>~~~~\textit{result}~=~\{\\

\>{\tiny 5}\'\>~~~~~~~~~~~~.nreaders:~0,~.nwriters:~0,~.mutex:~BinSema(\texttt{\textbf{False}}),\\

\>{\tiny 6}\'\>~~~~~~~~~~~~.r\_gate:~\{~.sema:~BinSema(\texttt{\textbf{True}}),~.count:~0~\},\\

\>{\tiny 7}\'\>~~~~~~~~~~~~.w\_gate:~\{~.sema:~BinSema(\texttt{\textbf{True}}),~.count:~0~\}\\

\>{\tiny 8}\'\>~~~~~~~~\}\\

\>{\tiny 9}\'\>\\

\>{\tiny 10}\'\>\texttt{\textbf{def}}~read\_acquire(\textit{rw}):\\

\>{\tiny 11}\'\>~~~~acquire(?\textit{rw}$\rightarrow$mutex)\\

\>{\tiny 12}\'\>~~~~\texttt{\textbf{if}}~(\textit{rw}$\rightarrow$nwriters~$>$~0)~\texttt{\textbf{or}}~(\textit{rw}$\rightarrow$w\_gate.count~$>$~0):\\

\>{\tiny 13}\'\>~~~~~~~~\textit{rw}$\rightarrow$r\_gate.count~+=~1;~release(?\textit{rw}$\rightarrow$mutex)\\

\>{\tiny 14}\'\>~~~~~~~~acquire(?\textit{rw}$\rightarrow$r\_gate.sema);~\textit{rw}$\rightarrow$r\_gate.count~--=~1\\

\>{\tiny 15}\'\>~~~~\textit{rw}$\rightarrow$nreaders~+=~1\\

\>{\tiny 16}\'\>~~~~\texttt{\textbf{if}}~\textit{rw}$\rightarrow$r\_gate.count~$>$~0:\\

\>{\tiny 17}\'\>~~~~~~~~release(?\textit{rw}$\rightarrow$r\_gate.sema)\\

\>{\tiny 18}\'\>~~~~\texttt{\textbf{else}}:\\

\>{\tiny 19}\'\>~~~~~~~~release(?\textit{rw}$\rightarrow$mutex)\\

\>{\tiny 20}\'\>\\

\>{\tiny 21}\'\>\texttt{\textbf{def}}~read\_release(\textit{rw}):\\

\>{\tiny 22}\'\>~~~~acquire(?\textit{rw}$\rightarrow$mutex)\\

\>{\tiny 23}\'\>~~~~\textit{rw}$\rightarrow$nreaders~--=~1\\

\>{\tiny 24}\'\>~~~~\texttt{\textbf{if}}~(\textit{rw}$\rightarrow$w\_gate.count~$>$~0)~\texttt{\textbf{and}}~(\textit{rw}$\rightarrow$nreaders~==~0):\\

\>{\tiny 25}\'\>~~~~~~~~release(?\textit{rw}$\rightarrow$w\_gate.sema)\\

\>{\tiny 26}\'\>~~~~\texttt{\textbf{else}}:\\

\>{\tiny 27}\'\>~~~~~~~~release(?\textit{rw}$\rightarrow$mutex)\\

\>{\tiny 28}\'\>~~~~\\

\>{\tiny 29}\'\>\texttt{\textbf{def}}~write\_acquire(\textit{rw}):\\

\>{\tiny 30}\'\>~~~~acquire(?\textit{rw}$\rightarrow$mutex)\\

\>{\tiny 31}\'\>~~~~\texttt{\textbf{if}}~(\textit{rw}$\rightarrow$nreaders~+~\textit{rw}$\rightarrow$nwriters)~$>$~0:\\

\>{\tiny 32}\'\>~~~~~~~~\textit{rw}$\rightarrow$w\_gate.count~+=~1;~release(?\textit{rw}$\rightarrow$mutex)\\

\>{\tiny 33}\'\>~~~~~~~~acquire(?\textit{rw}$\rightarrow$w\_gate.sema);~\textit{rw}$\rightarrow$w\_gate.count~--=~1\\

\>{\tiny 34}\'\>~~~~\textit{rw}$\rightarrow$nwriters~+=~1\\

\>{\tiny 35}\'\>~~~~release(?\textit{rw}$\rightarrow$mutex)\\

\>{\tiny 36}\'\>\\

\>{\tiny 37}\'\>\texttt{\textbf{def}}~write\_release(\textit{rw}):\\

\>{\tiny 38}\'\>~~~~acquire(?\textit{rw}$\rightarrow$mutex)\\

\>{\tiny 39}\'\>~~~~\textit{rw}$\rightarrow$nwriters~--=~1\\

\>{\tiny 40}\'\>~~~~\texttt{\textbf{if}}~\textit{rw}$\rightarrow$r\_gate.count~$>$~0:\\

\>{\tiny 41}\'\>~~~~~~~~release(?\textit{rw}$\rightarrow$r\_gate.sema)\\

\>{\tiny 42}\'\>~~~~\texttt{\textbf{elif}}~\textit{rw}$\rightarrow$w\_gate.count~$>$~0:\\

\>{\tiny 43}\'\>~~~~~~~~release(?\textit{rw}$\rightarrow$w\_gate.sema)\\

\>{\tiny 44}\'\>~~~~\texttt{\textbf{else}}:\\

\>{\tiny 45}\'\>~~~~~~~~release(?\textit{rw}$\rightarrow$mutex)

### 2021-02-02 book/sources/RWsbs.tex
Latest

\>{\tiny 1}\'\>\texttt{\textbf{from}}~synch~\texttt{\textbf{import}}~BinSema,~acquire,~release\\

\>{\tiny 2}\'\>\\

\>{\tiny 3}\'\>\texttt{\textbf{def}}~RWlock():\\

\>{\tiny 4}\'\>~~~~\textit{result}~=~\{\\

\>{\tiny 5}\'\>~~~~~~~~~~~~.nreaders:~0,~.nwriters:~0,~.mutex:~BinSema(\texttt{\textbf{False}}),\\

\>{\tiny 6}\'\>~~~~~~~~~~~~.r\_gate:~\{~.sema:~BinSema(\texttt{\textbf{True}}),~.count:~0~\},\\

\>{\tiny 7}\'\>~~~~~~~~~~~~.w\_gate:~\{~.sema:~BinSema(\texttt{\textbf{True}}),~.count:~0~\}\\

\>{\tiny 8}\'\>~~~~~~~~\}\\

\>{\tiny 9}\'\>\\

\>{\tiny 10}\'\>\texttt{\textbf{def}}~release\_one(\textit{rw}):\\

\>{\tiny 11}\'\>~~~~\texttt{\textbf{if}}~(\textit{rw}$\rightarrow$nwriters~==~0)~\texttt{\textbf{and}}~(\textit{rw}$\rightarrow$r\_gate.count~$>$~0):\\

\>{\tiny 12}\'\>~~~~~~~~release(?\textit{rw}$\rightarrow$r\_gate.sema)\\

\>{\tiny 13}\'\>~~~~\texttt{\textbf{elif}}~((\textit{rw}$\rightarrow$nreaders~+~\textit{rw}$\rightarrow$nwriters)~==~0)~\texttt{\textbf{and}}~(\textit{rw}$\rightarrow$w\_gate.count~$>$~0):\\

\>{\tiny 14}\'\>~~~~~~~~release(?\textit{rw}$\rightarrow$w\_gate.sema)\\

\>{\tiny 15}\'\>~~~~\texttt{\textbf{else}}:\\

\>{\tiny 16}\'\>~~~~~~~~release(?\textit{rw}$\rightarrow$mutex)\\

\>{\tiny 17}\'\>\\

\>{\tiny 18}\'\>\texttt{\textbf{def}}~read\_acquire(\textit{rw}):\\

\>{\tiny 19}\'\>~~~~acquire(?\textit{rw}$\rightarrow$mutex)\\

\>{\tiny 20}\'\>~~~~\texttt{\textbf{if}}~\textit{rw}$\rightarrow$nwriters~$>$~0:\\

\>{\tiny 21}\'\>~~~~~~~~\textit{rw}$\rightarrow$r\_gate.count~+=~1;~release\_one(\textit{rw})\\

\>{\tiny 22}\'\>~~~~~~~~acquire(?\textit{rw}$\rightarrow$r\_gate.sema);~\textit{rw}$\rightarrow$r\_gate.count~--=~1\\

\>{\tiny 23}\'\>~~~~\textit{rw}$\rightarrow$nreaders~+=~1\\

\>{\tiny 24}\'\>~~~~release\_one(\textit{rw})\\

\>{\tiny 25}\'\>\\

\>{\tiny 26}\'\>\texttt{\textbf{def}}~read\_release(\textit{rw}):\\

\>{\tiny 27}\'\>~~~~acquire(?\textit{rw}$\rightarrow$mutex);~\textit{rw}$\rightarrow$nreaders~--=~1;~release\_one(\textit{rw})\\

\>{\tiny 28}\'\>\\

\>{\tiny 29}\'\>\texttt{\textbf{def}}~write\_acquire(\textit{rw}):\\

\>{\tiny 30}\'\>~~~~acquire(?\textit{rw}$\rightarrow$mutex)\\

\>{\tiny 31}\'\>~~~~\texttt{\textbf{if}}~(\textit{rw}$\rightarrow$nreaders~+~\textit{rw}$\rightarrow$nwriters)~$>$~0:\\

\>{\tiny 32}\'\>~~~~~~~~\textit{rw}$\rightarrow$w\_gate.count~+=~1;~release\_one(\textit{rw})\\

\>{\tiny 33}\'\>~~~~~~~~acquire(?\textit{rw}$\rightarrow$w\_gate.sema);~\textit{rw}$\rightarrow$w\_gate.count~--=~1\\

\>{\tiny 34}\'\>~~~~\textit{rw}$\rightarrow$nwriters~+=~1\\

\>{\tiny 35}\'\>~~~~release\_one(\textit{rw})\\

\>{\tiny 36}\'\>\\

\>{\tiny 37}\'\>\texttt{\textbf{def}}~write\_release(\textit{rw}):\\

\>{\tiny 38}\'\>~~~~acquire(?\textit{rw}$\rightarrow$mutex);~\textit{rw}$\rightarrow$nwriters~--=~1;~release\_one(\textit{rw})

### 2021-02-02 book/sources/UpEnter.tex
Latest

\>{\tiny 1}\'\>\textit{count}~=~0\\

\>{\tiny 2}\'\>\textit{entered}~=~\textit{done}~=~[~\texttt{\textbf{False}},~\texttt{\textbf{False}}~]\\

\>{\tiny 3}\'\>\\

\>{\tiny 4}\'\>\texttt{\textbf{def}}~incrementer(\textit{self}):\\

\>{\tiny 5}\'\>~~~~\textit{entered}[\textit{self}]~=~\texttt{\textbf{True}}\\

\>{\tiny 6}\'\>~~~~\texttt{\textbf{if}}~\textit{entered}[1~--~\textit{self}]:~~~~~~~~~\emph{\# if the other thread has already started}\\

\>{\tiny 7}\'\>~~~~~~~~\texttt{\textbf{await}}~\textit{done}[1~--~\textit{self}]~~~~~\emph{\# wait until it is done}\\

\>{\tiny 8}\'\>~~~~\textit{count}~=~\textit{count}~+~1\\

\>{\tiny 9}\'\>~~~~\textit{done}[\textit{self}]~=~\texttt{\textbf{True}}\\

\>{\tiny 10}\'\>~~~~\texttt{\textbf{await}}~\textit{done}[1~--~\textit{self}]\\

\>{\tiny 11}\'\>~~~~\texttt{\textbf{assert}}~\textit{count}~==~2\\

\>{\tiny 12}\'\>\\

\>{\tiny 13}\'\>\texttt{\textbf{spawn}}~incrementer(0)\\

\>{\tiny 14}\'\>\texttt{\textbf{spawn}}~incrementer(1)

### 2021-02-02 book/sources/abp.tex
Latest

\>{\tiny 1}\'\>\textit{s\_chan}~=~\textit{r\_chan}~=~()\\

\>{\tiny 2}\'\>\textit{s\_seq}~=~\textit{r\_seq}~=~0\\

\>{\tiny 3}\'\>\\

\>{\tiny 4}\'\>\texttt{\textbf{def}}~net\_send(\textit{pchan},~\textit{m},~\textit{reliable}):\\

\>{\tiny 5}\'\>~~~~!\textit{pchan}~=~\textit{m}~\texttt{\textbf{if}}~(\textit{reliable}~\texttt{\textbf{or}}~\texttt{\textbf{choose}}(\{~\texttt{\textbf{False}},~\texttt{\textbf{True}}~\}))~\texttt{\textbf{else}}~()\\

\>{\tiny 6}\'\>\\

\>{\tiny 7}\'\>\texttt{\textbf{def}}~net\_recv(\textit{pchan}):\\

\>{\tiny 8}\'\>~~~~\textit{result}~=~!\textit{pchan}\\

\>{\tiny 9}\'\>\\

\>{\tiny 10}\'\>\texttt{\textbf{def}}~app\_send(\textit{payload}):\\

\>{\tiny 11}\'\>~~~~\textit{s\_seq}~=~1~--~\textit{s\_seq}\\

\>{\tiny 12}\'\>~~~~\texttt{\textbf{let}}~\textit{m},~\textit{blocked}~=~\{~.seq:~\textit{s\_seq},~.payload:~payload~\},~\texttt{\textbf{True}}:\\

\>{\tiny 13}\'\>~~~~~~~~\texttt{\textbf{while}}~\textit{blocked}:\\

\>{\tiny 14}\'\>~~~~~~~~~~~~net\_send(?\textit{s\_chan},~\textit{m},~\texttt{\textbf{False}})\\

\>{\tiny 15}\'\>~~~~~~~~~~~~\texttt{\textbf{let}}~\textit{response}~=~net\_recv(?\textit{r\_chan}):\\

\>{\tiny 16}\'\>~~~~~~~~~~~~~~~~\textit{blocked}~=~(\textit{response}~==~())~\texttt{\textbf{or}}~(\textit{response}.ack~!=~\textit{s\_seq})\\

\>{\tiny 17}\'\>~~~~~~~~~~~~\\

\>{\tiny 18}\'\>\texttt{\textbf{def}}~app\_recv(\textit{reliable}):\\

\>{\tiny 19}\'\>~~~~\textit{r\_seq}~=~1~--~\textit{r\_seq}\\

\>{\tiny 20}\'\>~~~~\texttt{\textbf{let}}~\textit{blocked}~=~\texttt{\textbf{True}}:\\

\>{\tiny 21}\'\>~~~~~~~~\texttt{\textbf{while}}~\textit{blocked}:\\

\>{\tiny 22}\'\>~~~~~~~~~~~~\texttt{\textbf{let}}~\textit{m}~=~net\_recv(?\textit{s\_chan}):\\

\>{\tiny 23}\'\>~~~~~~~~~~~~~~~~\texttt{\textbf{if}}~\textit{m}~!=~():\\

\>{\tiny 24}\'\>~~~~~~~~~~~~~~~~~~~~net\_send(?\textit{r\_chan},~\{~.ack:~\textit{m}.seq~\},~\textit{reliable})\\

\>{\tiny 25}\'\>~~~~~~~~~~~~~~~~~~~~\texttt{\textbf{if}}~\textit{m}.seq~==~\textit{r\_seq}:\\

\>{\tiny 26}\'\>~~~~~~~~~~~~~~~~~~~~~~~~\textit{result}~=~\textit{m}.payload\\

\>{\tiny 27}\'\>~~~~~~~~~~~~~~~~~~~~~~~~\textit{blocked}~=~\texttt{\textbf{False}}

### 2021-02-02 book/sources/atm.tex
Latest

\>{\tiny 1}\'\>\texttt{\textbf{from}}~synch~\texttt{\textbf{import}}~Lock,~acquire,~release\\

\>{\tiny 2}\'\>\\

\>{\tiny 3}\'\>\texttt{\textbf{const}}~N\_ACCOUNTS~=~2\\

\>{\tiny 4}\'\>\texttt{\textbf{const}}~N\_CUSTOMERS~=~2\\

\>{\tiny 5}\'\>\texttt{\textbf{const}}~N\_ATMS~=~2\\

\>{\tiny 6}\'\>\texttt{\textbf{const}}~MAX\_BALANCE~=~1\\

\>{\tiny 7}\'\>\\

\>{\tiny 8}\'\>\textit{accounts}~=~[~\{~.lock:~Lock(),~.balance:~\texttt{\textbf{choose}}(\{0..MAX\_BALANCE\})\}\\

\>{\tiny 9}\'\>~~~~~~~~~~~~~~~~~~~~~~~~~~~~\texttt{\textbf{for}}~\textit{i}~\texttt{\textbf{in}}~\{1..N\_ACCOUNTS\}~]\\

\>{\tiny 10}\'\>\\

\>{\tiny 11}\'\>\texttt{\textbf{invariant}}~\texttt{\textbf{min}}(\textit{accounts}[\textit{acct}].balance~\texttt{\textbf{for}}~\textit{acct}~\texttt{\textbf{in}}~\{0..N\_ACCOUNTS--1\})~$>=$~0\\

\>{\tiny 12}\'\>\\

\>{\tiny 13}\'\>\texttt{\textbf{def}}~atm\_check\_balance(\textit{acct}):~~~~\emph{\# return the balance on acct}\\

\>{\tiny 14}\'\>~~~~acquire(?\textit{accounts}[\textit{acct}].lock)\\

\>{\tiny 15}\'\>~~~~\textit{result}~=~\textit{accounts}[\textit{acct}].balance\\

\>{\tiny 16}\'\>~~~~release(?\textit{accounts}[\textit{acct}].lock)\\

\>{\tiny 17}\'\>\\

\>{\tiny 18}\'\>\texttt{\textbf{def}}~atm\_withdraw(\textit{acct},~\textit{amount}):~\emph{\# withdraw amount from acct}\\

\>{\tiny 19}\'\>~~~~acquire(?\textit{accounts}[\textit{acct}].lock)\\

\>{\tiny 20}\'\>~~~~\textit{accounts}[\textit{acct}].balance~--=~\textit{amount}\\

\>{\tiny 21}\'\>~~~~\textit{result}~=~\texttt{\textbf{True}}~~~~~~~~~~~~~~\emph{\# return success}\\

\>{\tiny 22}\'\>~~~~release(?\textit{accounts}[\textit{acct}].lock)\\

\>{\tiny 23}\'\>\\

\>{\tiny 24}\'\>\texttt{\textbf{def}}~customer(\textit{atm},~\textit{acct},~\textit{amount}):\\

\>{\tiny 25}\'\>~~~~\texttt{\textbf{let}}~\textit{bal}~=~atm\_check\_balance(\textit{acct}):\\

\>{\tiny 26}\'\>~~~~~~~~\texttt{\textbf{if}}~\textit{amount}~$<=$~\textit{bal}:\\

\>{\tiny 27}\'\>~~~~~~~~~~~~atm\_withdraw(\textit{acct},~\textit{amount})\\

\>{\tiny 28}\'\>~~~~~~~~\\

\>{\tiny 29}\'\>\texttt{\textbf{for}}~\textit{i}~\texttt{\textbf{in}}~\{1..N\_ATMS\}:\\

\>{\tiny 30}\'\>~~~~\texttt{\textbf{spawn}}~customer(\textit{i},~\texttt{\textbf{choose}}(\{0..N\_ACCOUNTS--1\}),\\

\>{\tiny 31}\'\>~~~~~~~~~~~~~~~~~~~~~~\texttt{\textbf{choose}}(\{0..MAX\_BALANCE\}))

### 2021-02-02 book/sources/lst1.tex
Latest

\>{\tiny 1}\'\>\texttt{\textbf{from}}~synch~\texttt{\textbf{import}}~Lock,~acquire,~release\\

\>{\tiny 2}\'\>\texttt{\textbf{from}}~alloc~\texttt{\textbf{import}}~malloc,~free\\

\>{\tiny 3}\'\>\\

\>{\tiny 4}\'\>\texttt{\textbf{def}}~lst\_node(\textit{v},~\textit{n}):~~~~~\emph{\# allocate and initialize a new list node}\\

\>{\tiny 5}\'\>~~~~\textit{result}~=~malloc(\{~.lock:~Lock(),~.value:~\textit{v},~.next:~\textit{n}~\})\\

\>{\tiny 6}\'\>\\

\>{\tiny 7}\'\>\texttt{\textbf{def}}~lst\_contains(\textit{lst},~\textit{v}):~~~\emph{\# check if v is in the list}\\

\>{\tiny 8}\'\>~~~~\texttt{\textbf{let}}~\textit{n}~=~\textit{lst}:\\

\>{\tiny 9}\'\>~~~~~~~~\texttt{\textbf{while}}~\textit{n}$\rightarrow$value~$<$~\textit{v}:\\

\>{\tiny 10}\'\>~~~~~~~~~~~~\textit{n}~=~\textit{n}$\rightarrow$next\\

\>{\tiny 11}\'\>~~~~~~~~\textit{result}~=~\textit{n}$\rightarrow$value~==~\textit{v}

### 2021-02-02 book/sources/qsort.tex
Latest

\>{\tiny 1}\'\>\texttt{\textbf{def}}~Qsort(\textit{arr}):\\

\>{\tiny 2}\'\>~~~~\textit{result}~=~\{~.arr:~arr,~.todo:~\{~(0,~\texttt{\textbf{len}}(arr)~--~1)~\}~\}\\

\>{\tiny 3}\'\>\\

\>{\tiny 4}\'\>\texttt{\textbf{def}}~swap(\textit{p},~\textit{q}):~~~~~~~~~~~~~~~\emph{\# swap !p and !q}\\

\>{\tiny 5}\'\>~~~~!\textit{p},~!\textit{q}~=~!\textit{q},~!\textit{p};~\\

\>{\tiny 6}\'\>\\

\>{\tiny 7}\'\>\texttt{\textbf{def}}~partition(\textit{qs},~\textit{lo},~\textit{hi}):\\

\>{\tiny 8}\'\>~~~~\textit{result}~=~\textit{lo}\\

\>{\tiny 9}\'\>~~~~\texttt{\textbf{for}}~\textit{i}~\texttt{\textbf{in}}~\{\textit{lo}..\textit{hi}~--~1\}:\\

\>{\tiny 10}\'\>~~~~~~~~\texttt{\textbf{if}}~\textit{qs}$\rightarrow$arr[\textit{i}]~$<=$~\textit{qs}$\rightarrow$arr[\textit{hi}]:\\

\>{\tiny 11}\'\>~~~~~~~~~~~~swap(?\textit{qs}$\rightarrow$arr[\textit{result}],~?\textit{qs}$\rightarrow$arr[\textit{i}])\\

\>{\tiny 12}\'\>~~~~~~~~~~~~\textit{result}~+=~1\\

\>{\tiny 13}\'\>~~~~swap(?\textit{qs}$\rightarrow$arr[\textit{result}],~?\textit{qs}$\rightarrow$arr[\textit{hi}]);~\\

\>{\tiny 14}\'\>\\

\>{\tiny 15}\'\>\texttt{\textbf{def}}~sortrange(\textit{qs},~\textit{range}):\\

\>{\tiny 16}\'\>~~~~\texttt{\textbf{let}}~\textit{lo},~\textit{hi}~=~\textit{range}~\texttt{\textbf{let}}~\textit{pivot}~=~partition(\textit{qs},~\textit{lo},~\textit{hi}):\\

\>{\tiny 17}\'\>~~~~~~~~\texttt{\textbf{if}}~(\textit{pivot}~--~1)~$>$~\textit{lo}:\\

\>{\tiny 18}\'\>~~~~~~~~~~~~\textit{qs}$\rightarrow$todo~$\vert$=~\{~(\textit{lo},~\textit{pivot}~--~1)~\}\\

\>{\tiny 19}\'\>~~~~~~~~\texttt{\textbf{if}}~(\textit{pivot}~+~1)~$<$~\textit{hi}:\\

\>{\tiny 20}\'\>~~~~~~~~~~~~\textit{qs}$\rightarrow$todo~$\vert$=~\{~(\textit{pivot}~+~1,~\textit{hi})~\}\\

\>{\tiny 21}\'\>\\

\>{\tiny 22}\'\>\texttt{\textbf{def}}~sort(\textit{qs}):\\

\>{\tiny 23}\'\>~~~~\texttt{\textbf{while}}~\textit{qs}$\rightarrow$todo~!=~\{\}:\\

\>{\tiny 24}\'\>~~~~~~~~\texttt{\textbf{let}}~\textit{range}~=~\texttt{\textbf{choose}}(\textit{qs}$\rightarrow$todo):\\

\>{\tiny 25}\'\>~~~~~~~~~~~~\textit{qs}$\rightarrow$todo~--=~\{~\textit{range}~\}\\

\>{\tiny 26}\'\>~~~~~~~~~~~~sortrange(\textit{qs},~\textit{range})\\

\>{\tiny 27}\'\>~~~~\textit{result}~=~\textit{qs}$\rightarrow$arr

### 2021-02-02 book/sources/qsorttest.tex
Latest

\>{\tiny 1}\'\>\texttt{\textbf{import}}~qsort,~\texttt{\textbf{bag}}\\

\>{\tiny 2}\'\>\\

\>{\tiny 3}\'\>\texttt{\textbf{const}}~NITEMS~=~4\\

\>{\tiny 4}\'\>\\

\>{\tiny 5}\'\>\textit{a}~=~[~\texttt{\textbf{choose}}(\{1..NITEMS\})~\texttt{\textbf{for}}~\textit{i}~\texttt{\textbf{in}}~\{1..\texttt{\textbf{choose}}(\{1..NITEMS\})\}~]\\

\>{\tiny 6}\'\>\textit{testqs}~=~qsort.Qsort(\textit{a})\\

\>{\tiny 7}\'\>\textit{sa}~=~qsort.sort(?\textit{testqs})\\

\>{\tiny 8}\'\>\texttt{\textbf{assert}}~\texttt{\textbf{all}}(\textit{sa}[\textit{i}~--~1]~$<=$~\textit{sa}[\textit{i}]~\texttt{\textbf{for}}~\textit{i}~\texttt{\textbf{in}}~\{1..\texttt{\textbf{len}}(\textit{sa})--1\})~\emph{\# sorted?}\\

\>{\tiny 9}\'\>\texttt{\textbf{assert}}~\texttt{\textbf{bag}}.fromList(\textit{a})~==~\texttt{\textbf{bag}}.fromList(\textit{sa});~\emph{\# is it a permutation?}

### 2021-02-02 book/sources/queueMS.tex
Latest

\>{\tiny 1}\'\>\texttt{\textbf{from}}~synch~\texttt{\textbf{import}}~Lock,~acquire,~release\\

\>{\tiny 2}\'\>\texttt{\textbf{from}}~alloc~\texttt{\textbf{import}}~malloc,~free\\

\>{\tiny 3}\'\>\\

\>{\tiny 4}\'\>\texttt{\textbf{def}}~Queue():\\

\>{\tiny 5}\'\>~~~~\texttt{\textbf{let}}~\textit{dummy}~=~malloc(\{~.value:~(),~.next:~\texttt{\textbf{None}}~\}):\\

\>{\tiny 6}\'\>~~~~~~~~\textit{result}~=~\{~.head:~\textit{dummy},~.tail:~\textit{dummy},~.hdlock:~Lock(),~.tllock:~Lock()~\}\\

\>{\tiny 7}\'\>\\

\>{\tiny 8}\'\>\texttt{\textbf{def}}~put(\textit{q},~\textit{v}):\\

\>{\tiny 9}\'\>~~~~\texttt{\textbf{let}}~\textit{node}~=~malloc(\{~.value:~\textit{v},~.next:~\texttt{\textbf{None}}~\}):\\

\>{\tiny 10}\'\>~~~~~~~~acquire(?\textit{q}$\rightarrow$tllock)\\

\>{\tiny 11}\'\>~~~~~~~~\textit{q}$\rightarrow$tail$\rightarrow$next~=~\textit{node}\\

\>{\tiny 12}\'\>~~~~~~~~\textit{q}$\rightarrow$tail~=~\textit{node}\\

\>{\tiny 13}\'\>~~~~~~~~release(?\textit{q}$\rightarrow$tllock)\\

\>{\tiny 14}\'\>\\

\>{\tiny 15}\'\>\texttt{\textbf{def}}~get(\textit{q}):\\

\>{\tiny 16}\'\>~~~~acquire(?\textit{q}$\rightarrow$hdlock)\\

\>{\tiny 17}\'\>~~~~\texttt{\textbf{let}}~\textit{dummy}~=~\textit{q}$\rightarrow$head\\

\>{\tiny 18}\'\>~~~~\texttt{\textbf{let}}~\textit{node}~=~\textit{dummy}$\rightarrow$next:\\

\>{\tiny 19}\'\>~~~~~~~~\texttt{\textbf{if}}~\textit{node}~==~\texttt{\textbf{None}}:\\

\>{\tiny 20}\'\>~~~~~~~~~~~~\textit{result}~=~\texttt{\textbf{None}}\\

\>{\tiny 21}\'\>~~~~~~~~\texttt{\textbf{else}}:\\

\>{\tiny 22}\'\>~~~~~~~~~~~~free(\textit{dummy})\\

\>{\tiny 23}\'\>~~~~~~~~~~~~\textit{result}~=~\textit{node}$\rightarrow$value\\

\>{\tiny 24}\'\>~~~~~~~~~~~~\textit{q}$\rightarrow$head~=~\textit{node}\\

\>{\tiny 25}\'\>~~~~release(?\textit{q}$\rightarrow$hdlock)

### 2021-02-02 book/sources/triangle.tex
Latest

\>{\tiny 1}\'\>\texttt{\textbf{const}}~N~=~10\\

\>{\tiny 2}\'\>\\

\>{\tiny 3}\'\>\texttt{\textbf{def}}~triangle(\textit{n}):~~~\emph{\# computes the n'th triangle number}\\

\>{\tiny 4}\'\>~~~~\textit{result}~=~0\\

\>{\tiny 5}\'\>~~~~\texttt{\textbf{for}}~\textit{i}~\texttt{\textbf{in}}~\{1..\textit{n}\}:~~~~~~\emph{\# for each integer from 1 to n inclusive}\\

\>{\tiny 6}\'\>~~~~~~~~\textit{result}~+=~\textit{i}~~~~~~\emph{\# add i to result}\\

\>{\tiny 7}\'\>\\

\>{\tiny 8}\'\>\textit{x}~=~\texttt{\textbf{choose}}(\{0..N\})~~~~~~~\emph{\# select an x between 0 and N inclusive}\\

\>{\tiny 9}\'\>\texttt{\textbf{assert}}~triangle(\textit{x})~==~((\textit{x}~*~(\textit{x}~+~1))~/~2)

### 2021-02-03 book/paper.tex
Fixed typo
Latest

@@
 Once initialization completes, any threads that were spawned
-(in this case \texttt{incrementer(0)} and \texttt{incrementer(0)})
+(in this case \texttt{incrementer(0)} and \texttt{incrementer(1)})
 can run.
@@
 Worse, most modern processors make accessing shared memory locations
-non-atomic to allow for efficient caching and other optimizations.
+non-atomic to allow for efficient caching, out-of-order execution,
+and other optimizations.
 So Peterson's algorithm would not even work on such processors.
@@
 \begin{code}
-\harmonysource{actor}
-\end{code}
-\caption{\harmonylink{code/actor.hny} A producer/consumer actor.}
+\harmonysource{counter}
+\end{code}
+\caption{\harmonylink{code/counter.hny} An illustration of the actor approach.}
 \label{fig:actor}
@@
 
-\begin{figure}
-\begin{code}
-\harmonysource{actortest}
-\end{code}
-\caption{\harmonylink{code/actortest.hny} Test code for producer/consumer actor.}
-\label{fig:actortest}
-\end{figure}
-
 Some programming languages favor a different way of implementing
-synchronization using so-called \emph{actors}~\cite{HBS73}.  Actors are
+synchronization using so-called \emph{actors}~\cite{HBS73,Agha86}.  Actors are
 threads that have only private memory and communicate through
@@
 
-\autoref{fig:actor} shows a Harmony solution to the
-``multiple producer / multiple consumer problem'' (with an unbounded buffer)
-using an actor (as illustrated in \autoref{fig:actorpic}).
-\texttt{pc\_actor} is essentially a matchmaker: it matches
-\texttt{.produce} requests with \texttt{.consume} requests.
-Sometimes it may have buffered \texttt{.consume} requests and is waiting for
-\texttt{.produce} requests, and sometimes it is vice versa.
-The variable \textit{balance} specifies what the situation is: if it is positive
-then it has buffered \texttt{.produce} requests;
-if it is negative
-then it has buffered \texttt{.consume} requests.
-Note that both \textit{balance} and \textit{requests} are variables that are
-local to the \texttt{pc\_actor} thread.
-
-Each request message is a dictionary with two fields.  One of the fields is
-\texttt{.type}, which either is \texttt{.produce} or \texttt{.consume}.
-In case of \texttt{.produce}, the second field is \texttt{.item} and holds
-the data that is produced.
-In case of \texttt{.consume}, the second field is \texttt{.queue} and holds
-a pointer to the queue where the response message must be enqueued.
-\texttt{.produce} messages do not get a response.
-
-The \texttt{pc\_actor} thread has two arguments.  $q$ is a pointer
-to the queue on which it receives messages.  \textit{nrequests} is the total
-number of requests that it will handle.  We need the latter because in Harmony
-all threads are required to terminate.  In a more realistic implementation,
-the \texttt{pc\_actor} would be in an infinite loop awaiting requests.
-
-The shared memory consists entirely of message queues.
-The \texttt{produce} method takes two arguments: a pointer to the queue
-of the \texttt{pc\_actor} and the item.  The \texttt{consume} method
-takes two queue pointer arguments.  The first is a pointer to the queue
-of the \texttt{pc\_actor}.  The second is a pointer to the queue of the
-invoker: it is the queue to which the response must be posted.
-
-\autoref{fig:actortest} shows how this code may be used.  It spawns
-\texttt{NITEMS} producer and consumer threads.  Note that in that
-case the \texttt{pc\_actor} is expected to receive \texttt{2 * NITEMS}
-requests.
+\autoref{fig:actor} illustrates the actor approach.
+There are three client threads that each want to be assigned
+a unique identifier from the set $\{ 0, 1, 2 \}$.
+Normally one would use a shared 0-initialized counter and a lock.
+Each client would acquire the lock, get the value of the counter
+and increment it, and release the lock.
+Instead, in the actor approach the counter is managed by a
+separate server thread.
+Each client sends a request to the server, consisting in this case
+of simply the queue to which the server must send the response.
+The server maintains a local, zero-initialized counter variable.
+Upon receiving a request, it returns a response with the value of
+the counter and increments the counter.  No lock is required.
+The code is tested using the \textit{done} array.
+
+This illustration is an example of the \emph{client/server} model.
+Here a single actor implements some service, and clients send request
+messages and receive response messages.  The model is particularly
+popular in distributed systems, where each actor runs on a separate
+machine and the queues are message channels.  For example, the server
+can be a web server, and its clients are web browsers.
 
@@
 \begin{problems}
-\item \label{ex:cltsvr} A popular model is the \emph{client/server} model.
-Here a single actor implements some service, like computing the square of a
-number.
-A client can send a message containing an integer to the server, and the
-server returns a response message containing the square of that integer.
-Implement this.
 \item Actors and message queues are good for building pipelines.
@@
 \begin{problems}
-\item \autoref{ex:cltsvr} explored the \emph{client/server model}.  It is popular
+\item \autoref{ch:actor} explored the \emph{client/server model}.  It is popular
 in distributed systems as well.

### 2021-02-03 book/paper.tex
Latest

@@
 An implementation of a data structure that can be safely accessed by multiple
-threads (or threads) and free of race conditions is called \emph{thread-safe}.
+threads and free of race conditions is called \emph{thread-safe}.
 \index{thread safety}%
@@
 Heather Zheng,
+Lorenzo Alvisi,
 Saleh Hassen,

### 2021-02-03 book/paper.tex
Latest

@@
 \begin{problems}
+\item The solution in \autoref{fig:dinerscv} can be simplified by, instead
+of having a condition variable per fork, having a condition variable per
+diner.  It is the same number of condition variables, but you will not
+need to have \textbf{if} statements nested inside the \textbf{while} loop
+waiting for the forks.  See if you can figure it out.
 \item \label{ex:bank} \autoref{fig:bank} shows an implementation of a bank with various

### 2021-02-03 book/paper.tex
Latest

@@
 Barrier synchronization is a problem that comes up in high-performance
-parallel computing, used, among others, for scalable simulation.
+parallel computing.
+It is used, among others, for scalable simulation.
 A barrier is almost the opposite of a critical section:
@@
 \label{fig:genbarrier}
+\end{figure}
+
+\begin{figure}
+\begin{code}
+\harmonysource{barriertest}
+\end{code}
+\caption{\harmonylink{code/barriertest.hny} Test program for \autoref{fig:genbarrier}.}
+\label{fig:barriertest}
 \end{figure}
@@
 the threads that were waiting for the car to empty out).
+
+\autoref{fig:barriertest} is a test program.  The threads check
+that all threads within the barrier are in the same round.
 
@@
 handlers must run quickly and cannot wait for other events.  So it is ok to invoke
-synchronization calls such as \texttt{V} and \texttt{notify}, but calls such as
-\texttt{P} and \texttt{wait} should only be used if it is certain that they will not
+non-blocking synchronization calls such as \texttt{notify}, but calls such as
+\texttt{acquire} and \texttt{wait} should only be used if it is certain that they will not
 block for long.  Informally, interrupt handlers can be

### 2021-02-04 book/paper.tex
Latest

@@
 Also, using synchronization primitives in interrupt handlers is tricky
-to get right (\autoref{ch:interrupts}) and might run too long.
+to get right (\autoref{ch:interrupts}) and might be too slow.
 In this chapter, we will have a look at how one can develop concurrent
-code in which threads do not have to wait for other threads to complete their
-ongoing operations.
+code in which threads do not have to wait for other threads (or interrupt
+handlers) to complete their ongoing operations.
 
@@
 There are two variables.  \textit{items} is an unbounded array with each
-entry initialized to \texttt{()}.  \textit{back} is an index into the
+entry initialized to \texttt{None}.  \textit{back} is an index into the
 array and points to the next slot where a new value is inserted.
@@
 the old value;
-\item \texttt{exch}($p$): sets \texttt{!}$p$ to \texttt{()} and returns
+\item \texttt{exch}($p$): sets \texttt{!}$p$ to \texttt{None} and returns
 the old value.
@@
 critical. %~\cite{}.
-\autoref{fig:lst1} and \autoref{fig:lst2} implements an ordered linked
+\autoref{fig:linkedlist} implements an ordered linked
 list of integers without duplicates.
-There are two update operations:
-values can be added using \texttt{lst\_insert} or deleted using
-\texttt{lst\_remove}.
-There is also a read-only operation: \texttt{lst\_contains} checks if
+(\autoref{fig:lltest} contains test code.)
+Values can be added using \texttt{insert}.
+There is also a read-only operation: \texttt{contains} checks if
 a particular value is in the list.
@@
 
-The read-only operation \texttt{lst\_contains} is wait\_free: it scans
+The read-only operation \texttt{contains} is wait\_free: it scans
 through the list without having to obtain a lock.
@@
 (which is \texttt{None} for the final \texttt{inf} node).
-The \texttt{lst\_find}(\textit{lst}, $v$) method is a helper function for update
-methods that
+The \texttt{insert(\textit{lst}, $v$)} method first
 finds and locks two consecutive nodes \textit{before} and \textit{after}
@@
 hand-over-hand.
-
-Having the two adjoining nodes locked, implementing \texttt{lst\_insert}
-and \texttt{lst\_remove} is straightforward.
-The \texttt{lst\_contains} method can simply scan through the list
+The \texttt{contains} method can simply scan through the list
 because the \textit{next} pointers are updated atomically in such a
@@
 points in time at which an operation appears to execute atomically.
-The linearization points for the \texttt{lst\_insert} and
-\texttt{lst\_remove} operations coincide exactly with the update
+The linearization point for the \texttt{insert} operation coincides exactly with the update
 of the \textit{before}.\texttt{next} pointer.
-The linearization point of a \texttt{lst\_contains} method
+The linearization point of a \texttt{contains} method
 execution depends on whether the value is found or not.
@@
 \begin{code}
-\harmonysource{lst1}
-\end{code}
-\caption{\harmonylink{code/lst1.hny} List with non-blocking read operation, part 1.}
-\label{fig:lst1}
-\end{figure}
-
-\begin{figure}
-\begin{code}
-\harmonysource{lst2}
-\end{code}
-\caption{\harmonylink{code/lst2.hny} List with non-blocking read operation, part 2.}
-\label{fig:lst2}
+\harmonysource{linkedlist}
+\end{code}
+\caption{\harmonylink{code/linkedlist.hny} Linked list with non-blocking read operation.}
+\label{fig:linkedlist}
+\end{figure}
+
+\begin{figure}
+\begin{code}
+\harmonysource{lltest}
+\end{code}
+\caption{\harmonylink{code/lltest.hny} Test code for \autoref{fig:linkedlist}.}
+\label{fig:lltest}
 \end{figure}
@@
 \begin{problems}
-\item Add read-only methods to the data structure in \autoref{fig:lst1}
-and \autoref{fig:lst2} that report the size of the list, the minimum value in the
+\item Add read-only methods to the data structure in \autoref{fig:linkedlist}
+that report the size of the list, the minimum value in the
 list, the maximum value in the list, and the sum of the values in the list.

### 2021-02-04 src/harmony/charm.js

@@ -443,7 +443,12 @@ function get_shared(shared, path) {
     return "";
   }
   if (path.length == 1) {
-    return shared[path[0]];
+    if (typeof shared[path[0]] === "string") {
+      return shared[path[0]];
+    }
+    else {
+      return JSON.stringify(shared[path[0]]);
+    }
   }
   return get_shared(shared[path[0]], path.slice(1));
 }

### 2021-02-04 book/paper.tex
Improved queue exercises

@@
 a pointer to a mutex and the thread identifier (0 or 1).
-$\mathit{pm}\rightarrow\mathtt{turn}$ is the value of the \texttt{.turn} key
+$\mathit{pm}$$\rightarrow$$\mathtt{turn}$ is the value of the \texttt{.turn} key
 in the dictionary that \textit{pm} points to.
@@
 \begin{problems}
-\item \label{ex:qcontains} Add a method $\mathtt{contains}(q, v)$ to both \autoref{fig:queue} and \autoref{fig:queueMS}
+\item \label{ex:qcontains} Add a method $\mathtt{contains}(q, v)$ to \autoref{fig:queue}
 that checks to see if $v$ is in queue $q$.
-\item Add a method $\mathtt{remove}(q, v)$ to both \autoref{fig:queue} and \autoref{fig:queueMS}
+\item Add a method $\mathtt{length}(q)$ to \autoref{fig:queue} that returns the length of the given queue.
+The complexity of the method should be $O(1)$, which is to say that you should maintain
+the length of the queue as a field member and update it in \texttt{put} and \texttt{get}.
+\item \label{ex:qlength} Write a method $\mathtt{check}(q)$
+that checks the integrity of the queue in \autoref{fig:queue}.
+In particular, it should check the following integrity properties:
+\begin{itemize}
+\item If the list is empty, $q$$\rightarrow$$\mathtt{tail}$ should be \texttt{None}.
+Otherwise, the last element in the linked list starting from $q$$\rightarrow$$\mathtt{head}$
+should equal $q$$\rightarrow$$\mathtt{head}$;
+Moreover, $q$$\rightarrow$$\texttt{tail}$$\rightarrow$$next$ should be \texttt{None}.
+\item The length field that you added in \autoref{ex:qlength} should equal the length
+of the list.
+\end{itemize}
+Method $\mathtt{check}(q)$ should not obtain a lock; instead add the following line
+just before releasing the lock in \texttt{put} and \texttt{get}:
+\begin{code}
+\textbf{assert} \texttt{check()}
+\end{code}
+\item Add a method $\mathtt{remove}(q, v)$ to \autoref{fig:queue}
 that removes all occurrences of $v$, if any, from queue $q$.

### 2021-02-04 book/paper.tex
Moved linked list to concurrent data structures

@@
 The most common use for locks is in building concurrent data structures.
-By way of example, we will demonstrate how to build a concurrent queue.
+By way of example, we will first demonstrate how to build a concurrent queue.
 The \texttt{queue} module will have the following API:
@@
 \end{figure}
+
+\begin{figure}
+\begin{code}
+\harmonysource{linkedlist}
+\end{code}
+\caption{\harmonylink{code/linkedlist.hny} Linked list with non-blocking read operation.}
+\label{fig:linkedlist}
+\end{figure}
+
+\begin{figure}
+\begin{code}
+\harmonysource{lltest}
+\end{code}
+\caption{\harmonylink{code/lltest.hny} Test code for \autoref{fig:linkedlist}.}
+\label{fig:lltest}
+\end{figure}
+
+\index{big lock}%
+\index{lock granularity}%
+\index{fine-grained lock}%
+\index{coarse-grained lock}%
+A queue has the nice property that usually only the head or the tail is accessed.
+However, in many data structures it is necessary to ``walk'' the data structure,
+an operation that can take significant time.
+In such a case, a single lock (known as a ``big lock'')
+for the entire data structure might restrict concurrency to an unacceptable level.
+To reduce the granularity of locking, each node in the data structure must be
+endowed with its own lock instead.
+
+\autoref{fig:linkedlist} implements an ordered linked
+list of integers without duplicates.
+(\autoref{fig:lltest} contains test code.)
+Values can be added using \texttt{insert} or deleted using \texttt{remove}.
+Method \texttt{contains} checks if a particular value is in the list.
+
+The list has two dummy ``book-end'' nodes with values \texttt{-inf} and
+\texttt{inf} (similar to the Python \texttt{math.inf} constant).
+An invariant of the algorithm is that at any point in time the
+list is ``valid,'' starting with a \texttt{-inf} node and ending
+with a \texttt{inf} node.
+
+\index{hand-over-hand locking}%
+Each node has a lock, a value, and \textit{next}, a pointer to the next node
+(which is \texttt{None} for the final \texttt{inf} node).
+The \texttt{\_find(\textit{lst}, $v$)} helper method first
+finds and locks two consecutive nodes \textit{before} and \textit{after}
+such that $\mathit{before}$$\rightarrow$$\mathtt{data.value} < v \le \mathit{after}$$\rightarrow$$\mathtt{data.value}$.
+It does so by performing something called \emph{hand-over-hand locking}.
+It first locks the first node, which is the \texttt{-inf} node.
+Then, iteratively, it obtains a lock on the next node and release the
+lock on the last one, and so on, similar to climbing a tree
+hand-over-hand.
+Using \texttt{\_find} the \texttt{insert}, \texttt{remove}, and \texttt{contains}
+methods are fairly straightforward.
+
+\glsadd{linearizable}%
+The implementation of the list is
+\emph{linearizable}~\cite{HW90},
+\index{linearizable}%
+a strong notion of consistency that
+makes it appear as if each of the operations executes atomically at
+some point between their invocation and return.
+Determining if an implementation of a concurrent data structure
+is linearizable involves finding what are known as the
+\emph{linearization points}
+\index{linearization point}%
+of the operations in an execution.  These are the unique
+points in time at which an operation appears to execute atomically.
+The linearization point for the \texttt{insert} operation coincides exactly with the update
+of the \textit{before}.\texttt{next} pointer.
+The linearization point of a \texttt{contains} method
+execution depends on whether the value is found or not.
+If found, it coincides with retrieving the pointer to the node
+that has the value.
+If not found, it coincides with retrieving the pointer to the
+\texttt{inf} node.
 
@@
 \texttt{bintree} with methods
+\item Add methods to the data structure in \autoref{fig:linkedlist}
+that report the size of the list, the minimum value in the
+list, the maximum value in the list, and the sum of the values in the list.
+Are they linearizable?  If so, what are their linearization points?
+\item Create a thread-safe sorted binary tree.  Implement a module \texttt{bintree}
+with methods
 $\mathtt{BinTree}()$ to create a new binary tree, $\mathtt{insert}(t, v)$ that inserts $v$ into
@@
 lock per binary tree.
-\item Bonus problem: create a binary tree that uses, instead of a single lock per tree,
+\item Create a binary tree that uses, instead of a single lock per tree,
 a lock for each node in the tree.
@@
 
+\begin{comment}
 There are also various algorithms where read-only access is wait-free
@@
 critical. %~\cite{}.
-\autoref{fig:linkedlist} implements an ordered linked
-list of integers without duplicates.
-(\autoref{fig:lltest} contains test code.)
-Values can be added using \texttt{insert}.
-There is also a read-only operation: \texttt{contains} checks if
-a particular value is in the list.
-\glsadd{linearizable}%
-
-The read-only operation \texttt{contains} is wait\_free: it scans
-through the list without having to obtain a lock.
-Nonetheless, the implementation of the list is
-\emph{linearizable}~\cite{HW90},
-\index{linearizable}%
-a strong notion of consistency that
-makes it appear as if each of the operations executes atomically at
-some point between their invocation and return.
-
-The list has two ``book-end'' nodes with values \texttt{-inf} and
-\texttt{inf} (similar to the Python \texttt{math.inf} constant).
-Each node has a lock, a value, and \textit{next}, a pointer to the next node
-(which is \texttt{None} for the final \texttt{inf} node).
-The \texttt{insert(\textit{lst}, $v$)} method first
-finds and locks two consecutive nodes \textit{before} and \textit{after}
-such that $\mathit{before}.\mathtt{data.value} < v \le \mathit{after}.\mathtt{data.value}$.
-It does so by performing something called \emph{hand-over-hand locking}.
-\index{hand-over-hand locking}%
-It first locks the first node, which is the \texttt{-inf} node.
-Then, iteratively, it obtains a lock on the next node and release the
-lock on the last one, and so on, similar to climbing a tree
-hand-over-hand.
-The \texttt{contains} method can simply scan through the list
-because the \textit{next} pointers are updated atomically in such a
-way that they always point to a legal suffix of the list.
-An invariant of the algorithm is that at any point in time the
-list is ``valid,'' starting with a \texttt{-inf} node and ending
-with a \texttt{inf} node.
-
-Determining if an implementation of a concurrent data structure
-is linearizable involves finding what are known as the
-\emph{linearization points}
-\index{linearization point}%
-of the operations in an execution.  These are the unique
-points in time at which an operation appears to execute atomically.
-The linearization point for the \texttt{insert} operation coincides exactly with the update
-of the \textit{before}.\texttt{next} pointer.
-The linearization point of a \texttt{contains} method
-execution depends on whether the value is found or not.
-If found, it coincides with retrieving the pointer to the node
-that has the value.
-If not found, it coincides with retrieving the pointer to the
-\texttt{inf} node.
-
-\begin{figure}
-\begin{code}
-\harmonysource{linkedlist}
-\end{code}
-\caption{\harmonylink{code/linkedlist.hny} Linked list with non-blocking read operation.}
-\label{fig:linkedlist}
-\end{figure}
-
-\begin{figure}
-\begin{code}
-\harmonysource{lltest}
-\end{code}
-\caption{\harmonylink{code/lltest.hny} Test code for \autoref{fig:linkedlist}.}
-\label{fig:lltest}
-\end{figure}
+\end{comment}
 
@@
 \begin{problems}
-\item Add read-only methods to the data structure in \autoref{fig:linkedlist}
-that report the size of the list, the minimum value in the
-list, the maximum value in the list, and the sum of the values in the list.
-Are they linearizable?  If so, what are their linearization points?
 \item A \emph{seqlock}

### 2021-02-05 book/paper.tex
Latest
Fixed some typos

@@
 \end{code}
-\caption{\harmonylink{code/linkedlist.hny} Linked list with non-blocking read operation.}
+\caption{\harmonylink{code/linkedlist.hny} Linked list with fine-grained locking.}
 \label{fig:linkedlist}
@@
 Otherwise, the last element in the linked list starting from $q$$\rightarrow$$\mathtt{head}$
-should equal $q$$\rightarrow$$\mathtt{head}$;
-Moreover, $q$$\rightarrow$$\texttt{tail}$$\rightarrow$$next$ should be \texttt{None}.
+should equal $q$$\rightarrow$$\mathtt{head}$. Moreover,
+$q$$\rightarrow$$\texttt{tail}$$\rightarrow$$next$ should be \texttt{None};
 \item The length field that you added in \autoref{ex:qlength} should equal the length
@@
 
-\section{The \texttt{Hoare} module}
+\section{The \texttt{hoare} module}
 \label{ap:hoare}
@@
 In Harmony, the two variables will be two different locks.  If you want two
-variable refering to the same lock in Harmony, you would write:
+variables referring to the same lock in Harmony, you would write:
 
@@
 (\textit{a b}) \texttt{+} (\textit{c d}).
-Unary operators are have the next highest precedence,
+Unary operators have the next highest precedence,
 and the remaining operators have the lowest precedence.
@@
 Because tuples and lists are dictionaries, the \texttt{del} statement is
-different than in Python.  For example, if $x$ = [.a, .b, .c], then
+different from Python.  For example, if $x$ = [.a, .b, .c], then
 \texttt{del} $x$[1] results in $x$ = \{ 0:.a, 2:.c \}, \emph{not}

### 2021-02-08 book/paper.tex
Latest

@@
 
-Even though they are not strictly necessary, we encourage to include invariants
-in your Harmony code. They can provide important insights into why the code works.
+Even though they are not strictly necessary, we encourage you
+to include invariants in your Harmony code.
+They can provide important insights into why the code works.
 
@@
 For $\mathcal{I}_2$ we only need an assertion at label \texttt{@cs} because the
-premisse is that there is a thread at that label.  However, we would
+premise is that there is a thread at that label.  However, we would
 like to check $\mathcal{I}_1$ in \emph{every state} (after the variables have
@@
 
-First and foremost I would like to think my student Haobin Ni with
+First and foremost I would like to thank my student Haobin Ni with
 whom I've had numerous discussions about the design of Harmony.
@@
 Brendon Nguyen,
+Hartek Sabharwal,
 Heather Zheng,

### 2021-02-23 book/paper.tex
Latest

@@
 }
+\newglossaryentry{sequential consistency}
+{
+  name=sequential consistency,
+  description={is a consistency model in which shared memory accesses are
+    executed in an order consistent with the program order}
+}
 \newglossaryentry{thread}
@@
 \begin{document}
-\maketitle
+\begin{titlepage}
+    \centering
+    \vfill
+    {\bfseries\huge
+        Concurrent Programming in Harmony
+    }
+    \vfill
+    \includegraphics[width=6cm]{figures/harmony-icon.png}
+    \vfill
+    \vfill
+    \hfill \Large Robbert van Renesse
+\end{titlepage}
 
@@
 \begin{itemize}
-\item \texttt{Safety violation}: this means something went wrong with
+\item \texttt{Safety violation}: This means something went wrong with
 at least one of the executions of the program that it tried.  This
@@
 Harmony will print a trace of the shortest bad execution that it found.
-\item \texttt{Non-terminating States}: Harmony found one or more states
+\item \texttt{Non-terminating State}: Harmony found one or more states
 from which there does not exist an execution such that all threads
@@
 at that state, along with their program counters.
+\item \texttt{Active Busy Waiting}:  There are states in which
+some thread cannot make progress without the help of another thread but
+does not block;
+\item \texttt{Data Race}: There are states in which two or more threads
+concurrently access a shared variable, at least one of which is a store
+operation.
 % \item \texttt{Stopped States}: Similar to non-terminating states,
@@
 
+\glsadd{sequential consistency}%
 In 1981, Gary L.~Peterson came up with a beautiful solution to the mutual exclusion
@@
 in \autoref{fig:peterson}.
+(The first line specifies that the \textit{flags} and \textit{turn}
+variables should have \emph{sequential consistency}---it prevents
+Harmony from complaining about data races involving these variables,
+explained in \autoref{ch:synch}.)
+
 A thread first indicates its interest in entering the critical section
@@
 \begin{problems}
-\item \autoref{fig:csonebit} presents another correct solution to the
+\item \autoref{fig:csonebit} presents another solution to the
 mutual exclusion problem.  It is similar to the one in
@@
 if it finds that the other thread is either trying to enter the critical
-section or already has.  Compare this algorithm with Peterson's.  What are
-the advantages and disadvantages?
+section or already has.  Compare this algorithm with Peterson's.
+Why does Harmony complain about \emph{active busy waiting}?
 \item
@@
 \hline
+Sequential & pop an address of a variable that has sequential consistency \\
+\hline
 SetIntLevel & pop $e$, set interrupt level to $e$, and push old interrupt level \\
@@
 and the assertion would fail if executed by Python.
-In Harmony, lists are values.  So when $x$ is updated in Line~3, it does
+In Harmony, lists are values.  So when $x$ is updated in Line~2, it does
 not affect the value of $y$.  The assertion succeeds.

### 2021-02-23 book/sources/pp.py

@@ -89,6 +89,7 @@ def isreserved(s):
         "not",
         "or",
         "pass",
+        "sequential",
         "setintlevel",
         "spawn",
         "stop",

### 2021-02-23 book/paper.tex
Fixed a bunch of things

@@
 
+\index{sequential consistency}%
 All instructions are atomically executed.
+The Harmony memory model is \emph{sequentially consistent}: all
+accesses are in program order.
 Most instructions pop values from the stack or push values onto the stack.
@@
 some thread cannot make progress without the help of another thread but
-does not block;
+does not block (see \autoref{ch:condwait});
 \item \texttt{Data Race}: There are states in which two or more threads
 concurrently access a shared variable, at least one of which is a store
-operation.
+operation (see \autoref{ch:synch}).
 % \item \texttt{Stopped States}: Similar to non-terminating states,
@@
 \end{itemize}
+
+Harmony checks for these types of failure conditions in the given
+order: if there are multiple failure conditions, only the first is
+reported.
 
@@
 \end{code}
-\item A colleague of the author asked if the first two assignments in
+\item \label{ex:reorder}
+A colleague of the author asked if the first two assignments in
 Peterson's algorithm (setting \textit{flags}[\textit{self}])
@@
 and other optimizations.
-So Peterson's algorithm would not even work on such processors.
-Instead, multi-core processors provide
-so-called \emph{interlock instructions}:
-\index{interlock instruction}%
+And even compilers may re-order memory accesses when optimizing code.
+%
+Peterson's algorithm relies on a sequential consistent memory model,
+and hence the \textbf{sequential} statement: without it Harmony will
+complain about data races.
+We saw in \autoref{ex:reorder} that a seemingly harmless reordering
+of instructions can break the code.
+The memory models of C and Java are not sequentially consistent, and
+they do not provide a \textbf{sequential} statement either, so one
+cannot implement Peterson's algorithm in those languages very easily.
+
+\glsadd{interlock instruction}%
+For proper synchronization, multi-core processors provide so-called
+\emph{interlock instructions}: \index{interlock instruction}%
 special machine instructions
 that can read memory and then write it in an indivisible step.
-
-\glsadd{interlock instruction}%
-
+%
 While the HVM does not have any specific built-in interlock instructions,
@@
 states of the data structure.
+
+\index{data race}
 When there is a bug in a program because some code omitted obtaining
@@
 \emph{data race}.
-\index{data race}
+In Harmony, a data race has a specific meaning: a state in which multiple
+threads are trying to access the same variable, and at least one of those
+accesses updates the variable.
+However, there are two exceptions.
+First, using the \textbf{sequential} statement, you can specify that
+concurrent access to the specified variables is intended.
+Second, if the accesses are within an \textbf{atomic} statement, then they
+are not considered part of a data race.
+However, for the same reason that you should not use \textbf{atomic}
+statements in your ordinary synchronization code, you should not use
+\textbf{sequential} statements either.  Neither is supported by regular
+programming languages.  But both can be very useful for testing.
 
@@
 chapters.
+Also notice the use of the sequentially consistent \textit{done}
+variable for testing purposes.
 
@@
 lock.
+We distinguish \emph{passive busy waiting} and \emph{active busy waiting}.
+A process that is waiting for a condition without changing the state
+(like in a spinlock) is passively busy waiting.
+A process that is waiting for a condition while changing the state
+(such as repeatedly trying to dequeue an item, which requires
+acquiring a lock) is actively busy waiting.
+
 We would like to find a solution to \emph{conditional waiting}
@@
 of the time.
-
 Before we do so, we will give two classic examples of synchronization

### 2021-02-23 book/paper.tex
Latest

@@
 Trishita Tiwari,
+William Ma,
 Yidan Wang,

### 2021-02-23 book/paper.tex
Latest

@@
 they do not provide a \textbf{sequential} statement either, so one
-cannot implement Peterson's algorithm in those languages very easily.
+cannot implement Peterson's algorithm in those languages directly.
 
@@
 When there is a bug in a program because some code omitted obtaining
-a lock before accessing the data structure, that is known as a
+a lock before accessing a shared data structure, that is known as a
 \emph{data race}.
-In Harmony, a data race has a specific meaning: a state in which multiple
+More precisely, a data race happens when there is a state in which multiple
 threads are trying to access the same variable, and at least one of those
 accesses updates the variable.
-However, there are two exceptions.
+In many environments, including C and Java programs, the behavior of
+concurrent load and store operations have tricky or even undefined
+semantics.  One should therefore avoid data races, which is why Harmony
+reports them even though Harmony has sequentially consistent memory.
+
+Harmony does not report data races in two cases.
 First, using the \textbf{sequential} statement, you can specify that
@@
 are not considered part of a data race.
-However, for the same reason that you should not use \textbf{atomic}
+For the same reason that you should not use \textbf{atomic}
 statements in your ordinary synchronization code, you should not use
 \textbf{sequential} statements either.  Neither is supported by regular
-programming languages.  But both can be very useful for testing.
+programming languages.  They exist in Harmony for the purpose of
+defining interlock instructions and for testing.
+The statements should only appear in modules that define synchronization
+primitives or in test programs.
 

### 2021-02-23 book/paper.tex
Latest

@@
 some thread cannot make progress without the help of another thread but
-does not block (see \autoref{ch:condwait});
+does not block;
 \item \texttt{Data Race}: There are states in which two or more threads
 concurrently access a shared variable, at least one of which is a store
-operation (see \autoref{ch:synch}).
+operation.
 % \item \texttt{Stopped States}: Similar to non-terminating states,
@@
 order: if there are multiple failure conditions, only the first is
-reported.
-
-In our case, Harmony reports a safety violation.
+reported.  \emph{Active busy waiting} (\autoref{ch:condwait}) is not
+technically an indication of a synchronization problem, but instead an
+indication of an inefficient solution to a synchronization problem---
+one that uses up significantly CPU cycles.
+A \emph{data race} (\autoref{ch:synch}) may not be a bug either---whether or
+not it is might depend on the semantics of the underlying memory
+operations and are therefore generally undesirable.
+
+Going back to our example, Harmony reports a safety violation.
 In particular, it reports that the assertion failed

### 2021-02-23 book/paper.tex
Latest

@@
     variables}
+}
+\newglossaryentry{data race}
+{
+  name=data race,
+  description={is when there are two or more threads concurrently accessing a
+    shared variable, at least one of which is an update to the variable}
 }
@@
 If they don't, try to understand why.
+If you get \emph{active busy waiting} or \emph{data race} reports, you probably
+found a correct solution; you'll learn later how to suppress those.
 Do not despair if you can't figure out how to develop a solution that satisfies both
-mutual exclusion and progress---as
-we will find out, it is possible but not obvious.
+mutual exclusion and progress---as we will find out, it is possible but not obvious.
 \end{problems}
@@
 
-\index{data race}
+\index{data race}%
+\glsadd{data race}%
 When there is a bug in a program because some code omitted obtaining

### 2021-02-23 src/charm/charm.c
Latest

@@ -1488,27 +1488,29 @@ int main(int argc, char **argv){
         printf("%d components, %d bad states\n", ncomponents, nbad);
     }
 
-    FILE *df = fopen("charm.dump", "w");
-	for (int i = 0; i < graph_size; i++) {
-		struct node *node = graph[i];
-		assert(node->id == i);
-		fprintf(df, "\nNode %d:\n", node->id);
-		fprintf(df, "    component: %d\n", node->component);
-		if (node->parent != NULL) {
-			fprintf(df, "    parent: %d\n", node->parent->id);
-		}
-		fprintf(df, "    vars: %s\n", value_string(node->state->vars));
-		fprintf(df, "    fwd:\n");
-		int eno = 0;
-		for (struct edge *edge = node->fwd; edge != NULL; edge = edge->next, eno++) {
-			fprintf(df, "        %d:\n", eno);
-			struct context *ctx = value_get(edge->ctx, NULL);
-			fprintf(df, "            context: %s %s %d\n", value_string(ctx->name), value_string(ctx->arg), ctx->pc);
-			fprintf(df, "            choice: %s\n", value_string(edge->choice));
-			fprintf(df, "            node: %d (%d)\n", edge->node->id, edge->node->component);
-		}
-	}
-	fclose(df);
+    if (false) {
+        FILE *df = fopen("charm.dump", "w");
+        for (int i = 0; i < graph_size; i++) {
+            struct node *node = graph[i];
+            assert(node->id == i);
+            fprintf(df, "\nNode %d:\n", node->id);
+            fprintf(df, "    component: %d\n", node->component);
+            if (node->parent != NULL) {
+                fprintf(df, "    parent: %d\n", node->parent->id);
+            }
+            fprintf(df, "    vars: %s\n", value_string(node->state->vars));
+            fprintf(df, "    fwd:\n");
+            int eno = 0;
+            for (struct edge *edge = node->fwd; edge != NULL; edge = edge->next, eno++) {
+                fprintf(df, "        %d:\n", eno);
+                struct context *ctx = value_get(edge->ctx, NULL);
+                fprintf(df, "            context: %s %s %d\n", value_string(ctx->name), value_string(ctx->arg), ctx->pc);
+                fprintf(df, "            choice: %s\n", value_string(edge->choice));
+                fprintf(df, "            node: %d (%d)\n", edge->node->id, edge->node->component);
+            }
+        }
+        fclose(df);
+    }
 
     FILE *out = fopen("charm.json", "w");
     assert(out != NULL);

### 2021-02-24 book/paper.tex
Latest

@@
 \textit{private}[$i$] for each thread $i$ is initialized to \texttt{True}.
+Method \texttt{thread} demonstrates how a local variable can be declared
+in Harmony using the \textbf{let} statement---the syntax is similar to
+\textbf{for} statements.  \autoref{sec:vars} provides further information.
+In this case $p$ contains the address of the \textit{private} entry
+for the current thread.
+
 An important invariant, $\mathcal{I}_1$, of the program is that at any time at most
@@
 
-\section{Constants, Global and Local Variables}
+\section{Constants, Global and Local Variables}\label{sec:vars}
 
@@
 William Ma,
+Xiangyu Zhang,
 Yidan Wang,

### 2021-03-01 book/paper.tex
Latest
Fixed x[...] code generation

@@
 A stack trace is a list of methods that are being invoked.
-In this case, the \texttt{implementer} method does not invoke any
+In this case, the \texttt{incrementer} method does not invoke any
 other methods, and so the list is of length 1.
@@
 
+Finally, methods can have local variables.
+Different from Python, Harmony local variables must be
+explicitly bound within a scope.
+You have already seen examples: the arguments to the method,
+the variable \textit{result}, and the bound variable (or
+variables) within a \texttt{for} statement.
+The \texttt{let} statement allows declaring additional
+local variables.  For example:
+\texttt{let} $x = 3$: $y$ $+$$=$ $x$ adds 3 to the global variable
+$y$.
+See \autoref{sec:vars} for more information.
+
 \chapter{Spinlock}
@@
 \textit{private}[$i$] for each thread $i$ is initialized to \texttt{True}.
-Method \texttt{thread} demonstrates how a local variable can be declared
-in Harmony using the \textbf{let} statement---the syntax is similar to
-\textbf{for} statements.  \autoref{sec:vars} provides further information.
-In this case $p$ contains the address of the \textit{private} entry
-for the current thread.
 

### 2021-03-01 book/paper.tex
Latest.  Removed Michael&Scott algorithm as it may be broken

@@
 
+\begin{comment}
 The implementation in \autoref{fig:queueMS}~\cite{MS98} is similar but uses
@@
 ever \texttt{None}.
+\end{comment}
 

### 2021-03-01 book/paper.tex
Latest

--- 2021-03-01 book/paper.tex
@@
 
-The implementation in \autoref{fig:queueMS}~\cite{MS98} is similar but uses
-separate locks for the head and tail, allowing a \texttt{put} and a \texttt{get} operation
+\autoref{fig:queueMS} shows another
+concurrent queue implementation~\cite{MS96}.
+It is well-known, but what is not often realized is that
+it requires sequentially consistent memory, which is not
+said explicitly in the paper.
+As a result, the algorithm is not guaranteed to work
+correctly with most modern programming languages and
+computer hardware.  But it is still useful to study it.
+The implementation uses separate
+locks for the head and the tail,
+allowing a \texttt{put} and a \texttt{get} operation
 to proceed concurrently.  To avoid contention between the head and the tail,
@@
 ever \texttt{None}.
+The problem is when the queue is empty and there are concurrent
+\texttt{get} and \texttt{put} operations.
+They obtain separate locks and then concurrently access the
+\texttt{next} field in the dummy node---a data race with
+undefined semantics in most environments.
 

### 2021-03-02 src/charm/charm.c
Improved error reporting

@@ -1160,7 +1160,10 @@ static void enum_loc(void *env, const void *key, unsigned int key_size,
 
     // TODO.  Should cache the contents of the file
     FILE *fp = fopen(cfile, "r");
-    assert(fp != NULL);
+    if (fp == NULL) {
+        fprintf(stderr, "Can't open '%s'\n", cfile);
+        exit(1);
+    }
     char buf[1024];
     while (fgets(buf, 1024, fp) != NULL) {
         if (--lineno == 0) {
@@ -1514,7 +1517,10 @@ int main(int argc, char **argv){
     }
 
     FILE *out = fopen("charm.json", "w");
-    assert(out != NULL);
+    if (out == NULL) {
+        fprintf(stderr, "charm: can't create charm.json");
+        exit(1);
+    }
     fprintf(out, "{\n");
 
     if (queue_empty(failures) && queue_empty(warnings)) {

### 2021-03-04 book/paper.tex
Improved error reporting
Latest version

@@
 \item What if we moved line~5 of \autoref{fig:incenter} to after the \textbf{if}
-    statement (between lines~8 and~9)?  Do you think that would work?  Run it through
+    statement (between lines~7 and~8)?  Do you think that would work?  Run it through
     Harmony and describe either why it works or why it does not work.
@@
 multiple operations.  Maintaining ``consistency'' is certainly much easier
-(although not necessarily impossible) if during this time no other
+if during this time no other
 thread also tries to access the binary tree.
@@
 Heather Zheng,
+Jacob Brugh,
 Lorenzo Alvisi,

### 2021-03-13 book/paper.tex
Better parse errors
Latest

@@
 \glsadd{linearizable}%
-The implementation of the list is
+Like the queue in \autoref{fig:queue}, the implementation of the list is
 \emph{linearizable}~\cite{HW90},
@@
 
-As a first example, we will revisit the producer/consumer problem.
+As an example, we will revisit the producer/consumer problem.
 The code in \autoref{fig:hw} is based on code developed by Herlihy and
@@
 \end{itemize}
-However, in the literature there are many examples of practical
+However, in the literature you can find examples of practical
 non-blocking (aka \emph{wait-free})

### 2021-03-14 book/paper.tex
Latest

@@
 \hline
-\texttt{pass} & do nothing\\
+\texttt{assert $b$ [, $e$]} & $b$ is a boolean.  Optionally report value of expression $e$\\
+\hline
+\texttt{atomic: S} & \texttt{S} a list of statements\\
+\hline
+\texttt{await $b$} & $b$ is a boolean\\
+\hline
+\texttt{const a = $e$} & \texttt{a} is a bounded variable, $e$ is a constant expression\\
+\hline
+\texttt{def m $a$: S} & \texttt{m} is an identifier, $a$ a bounded variable, \texttt{S} a list of statements\\
 \hline
 \texttt{del} \textit{lv} & delete\\
-\hline
-\texttt{assert $b$ [, $e$]} & $b$ is a boolean.  Optionally report value of expression $e$\\
-\hline
-\texttt{const a = $e$} & \texttt{a} is a bounded variable, $e$ is a constant expression\\
-\hline
-\texttt{def m $a$: S} & \texttt{m} is an identifier, $a$ a bounded variable, \texttt{S} a list of statements\\
-\hline
-\texttt{let $a$ = $e$ [let ...]: S} & $a$ is a bounded variable, $e$ is an expression, \texttt{S} a list of statements\\
-\hline
-\texttt{if $b$: S else: S} & $b$ is a boolean, \texttt{S} a list of statements\\
-\hline
-\texttt{while $b$: S} & $b$ is a boolean, \texttt{S} a list of statements\\
-\hline
-\texttt{await $b$} & $b$ is a boolean\\
 \hline
@@
 \hline
-\texttt{atomic: S} & \texttt{S} a list of statements\\
+\texttt{from m import ...} & \texttt{m} identifies a module \\
+\hline
+\texttt{go $c$ $e$} & $c$ is a context, $e$ is an expression \\
+\hline
+\texttt{if $b$: S else: S} & $b$ is a boolean, \texttt{S} a list of statements\\
+\hline
+\texttt{import m, ...} & \texttt{m} identifies a module \\
+\hline
+\texttt{let $a$ = $e$ [let ...]: S} & $a$ is a bounded variable, $e$ is an expression, \texttt{S} a list of statements\\
+\hline
+\texttt{pass} & do nothing\\
+\hline
+\texttt{sequential v, ...} & \texttt{v} has sequential consistency \\
 \hline
 \texttt{spawn m $e$ [, $t$]} & \texttt{m} is a method,
-$e$ is an expression, $t$ is the initial thread-local state (an expression) \\
+$e$ is an expression, $t$ is the initial thread-local state \\
 \hline
@@
 \hline
-\texttt{go $c$ $e$} & $c$ is a context, $e$ is an expression \\
-\hline
-\texttt{import m, ...} & \texttt{m} identifies a module \\
-\hline
-\texttt{from m import ...} & \texttt{m} identifies a module \\
+\texttt{while $b$: S} & $b$ is a boolean, \texttt{S} a list of statements\\
 \hline
@@
 \texttt{tas}(\textit{lk}) & test-and-set on \texttt{!}\textit{lk} \\
+\hline
+\texttt{cas}(\textit{ptr}, \textit{old}, \textit{new}) & compare-and-swap on \texttt{!}\textit{ptr} \\
 \hline

### 2021-03-15 book/paper.tex
Latest

@@
 verifies that the value of \textit{count} equals~2.
-% Like Python, Harmony supports \texttt{assert} statements with two arguments:
+% Like Python, Harmony supports \textbf{assert} statements with two arguments:
 % the first is a Boolean condition, and the second is a value that is
@@
 And even compilers may re-order memory accesses when optimizing code.
-%
+
 Peterson's algorithm relies on a sequential consistent memory model,
@@
 complain about data races.
-We saw in \autoref{ex:reorder} that a seemingly harmless reordering
-of instructions can break the code.
-The memory models of C and Java are not sequentially consistent, and
-they do not provide a \textbf{sequential} statement either, so one
-cannot implement Peterson's algorithm in those languages directly.
+More precisely, the \textbf{sequential} statement says that the
+program relies on memory load and store instructions operating on
+the indicated variables to be performed sequentially.
+The default memory models of C and Java are not sequentially
+consistent.
+The unfortunately named \texttt{volatile} keyword in Java has
+a similar function as Harmony's \texttt{sequential} keyword.
+Like many constructions in Java, its \texttt{volatile} keyword
+was borrowd from C and C++.
+However, in C and C++ they do \emph{not} provide sequential
+consistency, and one cannot implement Peterson's algorithm in
+C or C++ directly.
 
@@
 Harmony can check these invariants as well.
-$\mathcal{I}_2(i)$ is checked by the \texttt{assert} statement.
+$\mathcal{I}_2(i)$ is checked by the \textbf{assert} statement.
 But how would one go about checking an
@@
 The former ensures that while the
-\texttt{atomic} statement is executing no other thread can execute.
+\textbf{atomic} statement is executing no other thread can execute.
 The latter allows multiple threads to run concurrently,
@@
 
-In Harmony, \texttt{atomic} statements allow you to \emph{implement} your own
+In Harmony, \textbf{atomic} statements allow you to \emph{implement} your own
 interlock primitives such as test-and-set.  Other common examples
-include compare-and swap and fetch-and-add.  \texttt{atomic} statements
+include compare-and swap and fetch-and-add.  \textbf{atomic} statements
 are not intended to replace locks or other synchonization primitives.
 When solving synchronization problems you should not directly use
-\texttt{atomic} statements but use the synchronization primitives that are available
+\textbf{atomic} statements but use the synchronization primitives that are available
 to you.  But if you want to design a new synchronization primitive, then
@@
 You can also use \textbf{atomic} statements in your test code.
-In fact, as mentioned before, \texttt{assert} statements are executed atomically.
+In fact, as mentioned before, \textbf{assert} statements are executed atomically.
 
@@
 are not considered part of a data race.
+
+\begin{comment}
 For the same reason that you should not use \textbf{atomic}
@@
 primitives or in test programs.
+\end{comment}
 

### 2021-03-16 book/paper.tex
Latest

@@
 Like many constructions in Java, its \texttt{volatile} keyword
-was borrowd from C and C++.
+was borrowdd from C and C++.
 However, in C and C++ they do \emph{not} provide sequential
@@
 Sunwook Kim,
+Terryn Jung,
 Trishita Tiwari,

### 2021-03-21 book/paper.tex
Latest

@@
 
-In \texttt{release\_rlock}, notice that \texttt{notify}(?\textit{w\_cond})
+In \texttt{read\_release}, notice that \texttt{notify}(?\textit{w\_cond})
 is invoked when there are no readers left, \emph{without} checking
@@
 
-\texttt{release\_wlock} executes \texttt{notifyAll}(?\textit{r\_cond})
+\texttt{write\_release} executes \texttt{notifyAll}(?\textit{r\_cond})
 as well as \texttt{notify}(?\textit{w\_cond)}.  Because we
@@
 or, alternatively, up to one waiting writer can enter.  So
-\texttt{release\_wlock} wakes up all potential candidates.  There
+\texttt{write\_release} wakes up all potential candidates.  There
 are two things to note here.  First, unlike split binary semaphores
@@
 Lorenzo Alvisi,
+Maria Martucci,
 Saleh Hassen,

### 2021-03-27 book/paper.tex
Added a chapter to the book on testing

@@
 on luck, Harmony will run \emph{all possible executions} of a particular
-test program.  So even if a bug is unlikely to occur, if the test
+test program.  So, even if a bug is unlikely to occur, if the test
 \emph{can} expose the bug it \emph{will}.  Moreover, if the bug is
@@
 
-So if model checking does not prove a program correct, why is it
+If model checking does not prove a program correct, why is it
 useful?
@@
 
-So what is Harmony?
+So, what is Harmony?
 Harmony is a concurrent programming language.  It was designed to teach
@@
 problem and the bounded buffer problem.
+\item \autoref{ch:testing} discusses approaches to testing
+concurrent code in Harmony.
 \item \autoref{ch:sbs} presents \emph{Split Binary Semaphores}, a
@@
 you initialize it to 1.  (You do not need to understand the error report at this time.  They will be explained in more detail in \autoref{ch:harmonymachine}.)
-\item Write a Harmony program that computes squares by repeated adding.  So the program
+\item Write a Harmony program that computes squares by repeated adding.  So, the program
 should compute the square of $x$ by adding $x$ to an initial value of 0 $x$ times.
@@
 increment or decrement values in shared memory locations.
-So to increment a value in memory,
+So, to increment a value in memory,
 the machine must do at least three machine instructions.  Conceptually:
@@
 which in turn is the same Harmony value as \texttt{\{ 0:a, 1:b, 2:c \}}.
-So if $x$ = \texttt{[False, True]},
+So, if $x$ = \texttt{[False, True]},
 then $x$[0] = \texttt{False} and $x$[1] = \texttt{True}, just like in Python.
@@
 section, entering a state that is not exclusive.
-So exclusivity is an invariant (holds over every reachable state, as demonstrated
+So, exclusivity is an invariant (holds over every reachable state, as demonstrated
 by Harmony),
@@
 
-So we are looking for an inductive invariant that \emph{implies} exclusivity.
+We are looking for an inductive invariant that \emph{implies} exclusivity.
 In other words, the set of states where the inductive invariant holds
@@
 $\mathcal{F}(i)$ continues to hold as well.
-So $\mathcal{F}(i)$ is an inductive invariant (for both threads~0 and~1).
+So, $\mathcal{F}(i)$ is an inductive invariant (for both threads~0 and~1).
 
@@
 \texttt{tas} method is invoked in).
-So there are two cases:
+There are two cases:
 \begin{enumerate}
@@
 A thread usually does not keep explicit track of whether it has a lock
-or not because it is usually implied by the control flow of the program---a
+or not, because it is usually implied by the control flow of the program---a
 thread implicitly \emph{knows} that when it is executing in a critical
 section it has the lock.
-So there is no need to keep \textit{private} as a shared
+There is no need to keep \textit{private} as a shared
 variable---we only did so to be able to show and check the invariants.
@@
 \end{itemize}
-See \autoref{fig:queuetest} for a simple test program that uses the queue.
+
+\noindent
+See \autoref{fig:queuedemo} for a simple demonstration program that uses the queue.
 
@@
 When they are done, they release the lock.
+
+An important thing to note in \autoref{fig:queuedemo}
+is Lines~7 and~8.
+It would be incorrect to replace these by:
+
+\begin{code}
+\textbf{assert} \texttt{queue.get}(\textit{q}) in \{ \textbf{None}, 1, 2 \}
+\end{code}
+
+The reason is that \texttt{queue.get()} changes the state by
+acquiring a lock, but the expressions in \textbf{assert}
+statements (or \textbf{invariant} statements)
+are not allowed to change the state.
+In general, when calling methods in \textbf{assert} or
+\textbf{invariant} statements, one has to be convinced that
+those methods cannot change the state in any way.
 
@@
 \begin{code}
-\harmonysource{queuetest}
-\end{code}
-\caption{\harmonylink{code/queuetest.hny} Test program for a concurrent queue.}
-\label{fig:queuetest}
+\harmonysource{queuedemo}
+\end{code}
+\caption{\harmonylink{code/queuedemo.hny} Using a concurrent queue.}
+\label{fig:queuedemo}
 \end{figure}
@@
 \end{problems}
+
+\chapter{Testing}
+\label{ch:testing}
+
+\begin{figure}
+\begin{code}
+\harmonysource{queuespec}
+\end{code}
+\caption{\harmonylink{code/queuespec.hny} Queue specification.}
+\label{fig:queuespec}
+\end{figure}
+
+\begin{figure}
+\begin{code}
+\harmonysource{qtestseq}
+\end{code}
+\caption{\harmonylink{code/qtestseq.hny} Sequential queue test.}
+\label{fig:qtestseq}
+\end{figure}
+
+Testing is a way to increase confidence in the correctness
+of an implementation.
+\autoref{fig:queuedemo} demonstrates how concurrent queues
+may be used, but it is not a very thorough test program
+for an implementation such as the one in \autoref{fig:queue}
+and does little to increase our confidence in its correctness.
+In this chapter, we will look at approaches to testing
+concurrent code.
+
+The first step in testing a concurrent data structure is
+whether it works in the absence of concurrency.
+The objective is to test if the queue \emph{implementation}
+in \autoref{fig:queue} meets the sequential
+queue \emph{specification}.
+In order to do so, we should formalize the sequential queue specification.
+A specification is just code.
+\autoref{fig:queuespec} shows a sequential specification of a
+queue in Harmony (exploiting some methods from the \texttt{list}
+module described in \autoref{ap:list}).
+To check if the queue implementation meets the specification,
+we need to see if any sequence of queue operations in the
+implementation matches a corresponding sequence in the
+specification.
+We say that the implementation and the specification should
+have the same \emph{behavior}.
+
+\autoref{fig:qtestseq} presents a test program that does
+exactly this, for sequences of up to \texttt{NOPS}
+queue operations.
+It maintains two queues:
+\begin{itemize}
+\item[] \textit{implq}: the queue of the implementation;
+\item[] \textit{specq}: the queue of the specification
+\end{itemize}
+For each operation, the code first decides whether to
+perform a \texttt{get} or \texttt{put} operation.
+In the case of a \texttt{put} operation, the code also
+decides which value to append to the queue.
+All operations are performed on both the implementation
+queue and the specification queue.
+In the case of \texttt{get}, the result of the operation
+is checked against that of the specification.
+
+As with any other test program,
+\autoref{fig:qtestseq} may not trigger extant bugs, but
+it nonetheless inspires reasonable confidence that the
+queue implementation is correct, at least sequentially.
+The higher \texttt{NOPS}, the higher the confidence.
+It is possible to write similar programs in other languages
+such as Python, but the \texttt{choose} expression in Harmony
+makes it relatively easy to explore all corner cases.
+
+For example, a common programming mistake is to forget to
+update the \texttt{tail} pointer in \texttt{get()} in case
+the queue becomes empty.  Normally, it is a surprisingly
+tricky bug to find.  You can comment out those lines
+in \autoref{fig:queue} and run the test program---it should
+easily find the bug and the output explains exactly how
+the bug can happen, adding confidence that the test program
+is reasonably thorough.
+
+\begin{figure}
+\begin{code}
+\harmonysource{qtest1}
+\end{code}
+\caption{\harmonylink{code/qtest1.hny} Concurrent \texttt{put()}.}
+\label{fig:qtest1}
+\end{figure}
+
+The test program also finds some common mistakes in using
+locks, such as forgetting to release a lock when the queue
+is empty, but it is not designed to find concurrency bugs
+in general.  If you remove all \texttt{acquire()} and
+\texttt{release()} calls from \autoref{fig:queue},
+the test program will not (and should not)
+find any errors.
+The next step is to test if the queue implementation
+meets the \emph{concurrent} queue specification or not.
+But we have not yet shown what the
+concurrent queue specification is.
+
+We briefly mentioned the notion of \emph{linearizability}
+in \autoref{ch:cds}.
+Basically, we want a concurrent queue to behave
+consistently with a sequential queue in that all
+\texttt{put}
+and \texttt{get} operations should appear to happen
+in a total order.
+Moreover, we want to make sure that if some
+\texttt{put} or \texttt{get} operation $o_1$ finished
+before another operation $o_2$ started, then $o_1$
+should appear to happen before $o_2$ in the total order.
+If these two conditions are met, then we say that
+the concurrent queue implementation is linearizable.
+
+In general, if a data structure is protected by
+a single lock and every operation on that data
+structure starts with acquiring the lock and ends
+with releasing the lock, it will automatically be
+linearizable.
+The queue implementation in \autoref{fig:queue} does
+not quite match this pattern, as the \texttt{put}
+operation allocates a new node before acquiring
+the lock.
+However, in this case that is not a problem, as
+the new node has no dependencies on the queue
+when it is allocated.
+
+While it is possible to check linearizability with
+model checking, Harmony does not (yet) have explicit
+support for this.  Instead, we will present a few test
+programs that try to identify incorrect behavior of
+the queue implementation in the presence of concurrency.
+
+\autoref{fig:qtest1} shows a program that
+checks concurrent \texttt{put} operations using
+\texttt{N} threads.  Thread \texttt{putter($v$)}
+adds $v$ to the queue.
+Thread \texttt{main()} waits until all \texttt{putter}
+threads have finished, and then dequeues \texttt{N}
+items, verifying that they form the set
+$\{ 0 .. \mathtt{N} - 1 \}$.
+Note that it is perfectly fine to use \textbf{sequential}
+and \textbf{atomic} statements in test code---one should
+only avoid those in production code.
+
+\begin{figure}
+\begin{code}
+\harmonysource{qtest2}
+\end{code}
+\caption{\harmonylink{code/qtest2.hny} Concurrent \texttt{get()}.}
+\label{fig:qtest2}
+\end{figure}
+
+Similarly, \autoref{fig:qtest2} shows a program that
+checks concurrent \texttt{get} operations using
+\texttt{N} threads.
+The queue is initialized with the sequence
+$0 .. \mathtt{N} - 1$.
+Thread \texttt{getter($v$)} removes an entry from the
+queue and makes sure that no other thread got the
+same entry.
+Thread \texttt{main()} waits until all \texttt{getter}
+threads have finished and checks to make sure that
+the queue is empty.
+
+\begin{figure}
+\begin{code}
+\harmonysource{qtest3}
+\end{code}
+\caption{\harmonylink{code/qtest3.hny} Mix of concurrent
+\texttt{get()} and \texttt{put()} operations.}
+\label{fig:qtest3}
+\end{figure}
+
+\autoref{fig:qtest3} checks a mixture of
+concurrent \texttt{get} and \texttt{put} operations.
+It starts \texttt{N} \texttt{getter} threads and
+\texttt{N} \texttt{putter} threads.
+Thread \texttt{main} waits until all have finished,
+and then checks to make sure that whatever is left
+on the queue was not also consumed by some
+\texttt{getter} thread.
+Finally, it makes sure that the queue is empty
+once all the values that were added
+to it by the \texttt{putter} threads have been consumed.
+
+A common mistake that people make is to use
+a globally shared variable in the implementation of
+a concurrent data structure.
+The \texttt{get()} method of the queue implementation
+uses a local variable \textit{node}, declared using
+a \textbf{let} statement, but it would be easy to
+mistakenly use a global variable \textit{node} instead.
+If there is only one queue, this is not an issue.
+So, it is important to test not just one queue at a
+time, but also multiple queues.
+\autoref{fig:qtest4} gives an example that
+tests the \texttt{get} method for use of global
+variables.  Even if the assertion does not fail,
+Harmony might find a data race if there are global
+variables shared between different queues.
+
+While having five different test programs is fine,
+it would be nice if we can combine them all in one
+test program.
+This has to be done with some care, as running the
+concurrent test programs simultaneously would
+lead to a state space explosion.
+Instead, we can leverage the \textbf{choose}
+expression.
+\autoref{fig:multitest} shows how one might go
+about this.
+Harmony will run all the tests, but the
+tests do not interfere with one another.
+
+\begin{figure}
+\begin{code}
+\harmonysource{qtest4}
+\end{code}
+\caption{\harmonylink{code/qtest4.hny} Test with multiple queues.}
+\label{fig:qtest4}
+\end{figure}
+
+\begin{figure}
+\begin{code}
+\harmonysource{multitest}
+\end{code}
+\caption{\harmonylink{code/multitest.hny} Running a
+sequential test followed by several concurrent tests.}
+\label{fig:multitest}
+\end{figure}
 
@@
 
-So far we have looked at how to protect a single resource.
+So far, we have looked at how to protect a single resource.
 Some types of resources may have multiple units.
@@
 \end{enumerate}
-So it will require 3 binary semaphores if the SBS technique is applied.
+So, it will require 3 binary semaphores if the SBS technique is applied.
 
@@
 program allows are a subset of the traces in the property.
-So far we have pursued two properties: \emph{mutual exclusion}
+So far, we have pursued two properties: \emph{mutual exclusion}
 and \emph{progress}.  The former is an example of a
@@
 reader.  Unfortunately for the writer, this scenario can repeat itself
-indefinitely.  So even if neither reader was in the critical section
+indefinitely.  So, even if neither reader was in the critical section
 all of the time, and the second reader arrived well after the writer,
@@
 through the main gate, other threads may have entered first and falsified
-the condition.  So in Mesa, threads always have to check the condition
+the condition.  So, in Mesa, threads always have to check the condition
 again after resuming from the \texttt{wait} operation.  This is accomplished
@@
 \texttt{wait} within a \textbf{while} loop, re-checking the condition
-that they are waiting for.  So if both type of threads are waiting,
+that they are waiting for.  So, if both type of threads are waiting,
 either all the readers get to enter next or one of the writers gets
@@
 better to wake up too many threads than too few.
-So in case of doubt, use \texttt{notifyAll}.
+In case of doubt, use \texttt{notifyAll}.
 Waking up too
@@
 In the case of the Dining Philosophers, we want to avoid the situation where each
-diner picks up a fork.  So if we can prevent more than four diners from starting to
+diner picks up a fork.  If we can prevent more than four diners from starting to
 eat at the same time, then we can avoid the conditions for deadlock from ever
@@
 asynchronous events.  Because interrupts may be disabled only briefly, interrupt
-handlers must run quickly and cannot wait for other events.  So it is ok to invoke
+handlers must run quickly and cannot wait for other events.  It is ok to invoke
 non-blocking synchronization calls such as \texttt{notify}, but calls such as
@@
 
-So far we have concentrated on critical sections to synchronize multiple
+So far, we have concentrated on critical sections to synchronize multiple
 threads.  Certainly, preventing multiple threads from accessing
@@
 and the assertion would fail if executed by Python.
-In Harmony, lists are values.  So when $x$ is updated in Line~2, it does
+In Harmony, lists are values.  So, when $x$ is updated in Line~2, it does
 not affect the value of $y$.  The assertion succeeds.
@@
 We call this \emph{application}, and application is left-associative.
-So \textit{a b c} is interpreted as (\textit{a b}) $c$: $b$ is applied to $a$,
+So, \textit{a b c} is interpreted as (\textit{a b}) $c$: $b$ is applied to $a$,
 and then $c$ is applied to the result.
@@
 There are three classes of precedence.
-Application has the highest precedence.  So \texttt{!}\textit{a b} is interpreted as
+Application has the highest precedence.  So, \texttt{!}\textit{a b} is interpreted as
 \texttt{!}(\textit{a b}) and \textit{a b} \texttt{+} \textit{c d} is interpreted as
@@
 and the remaining operators have the lowest precedence.
-So $-2 + 3$ evaluates to 1, not $-5$.
+For example, $-2 + 3$ evaluates to 1, not $-5$.
 
@@
 
-In almost all expressions, subexpressions are evaluated left to right.  So $a[b] + c$
+In almost all expressions, subexpressions are evaluated left to right.
+So, $a[b] + c$
 first evaluates $a$, then $b$ (and then applies $b$ to $a$), and then $c$.  The one
@@
 one element must be followed by a comma.
-So the statement \texttt{$x$ = 1,;} assigns a singleton tuple (or list) to $x$.
+For example, the statement \texttt{$x$ = 1,;} assigns a singleton tuple (or list) to $x$.
 

### 2021-03-28 src/charm/charm.c
Working on improving data race (but broken at the moment)

@@ -17,19 +17,19 @@ struct combined {           // combination of current state and current context
 };
 
 struct component {
-    bool good;          // terminating or out-going edge
-    int size;           // #states
-    int representative; // lowest numbered state in the component
+    bool good;              // terminating or out-going edge
+    int size;               // #states
+    int representative;     // lowest numbered state in the component
 };
 
 struct edge {
-    struct edge *next;      // linked list maintenance
-    uint64_t ctx, choice;   // ctx that made the microstep, choice if any
-    bool interrupt;         // set if state change is an interrupt
-    struct node *node;      // resulting node (state)
-    uint64_t after;         // resulting context
-    int weight;             // 1 if context switch; 0 otherwise
-    struct access_info ai;  // to detect data races
+    struct edge *next;       // linked list maintenance
+    uint64_t ctx, choice;    // ctx that made the microstep, choice if any
+    bool interrupt;          // set if state change is an interrupt
+    struct node *node;       // resulting node (state)
+    uint64_t after;          // resulting context
+    int weight;              // 1 if context switch; 0 otherwise
+    struct access_info *ai;  // to detect data races
 };
 
 struct node {
@@ -57,6 +57,7 @@ struct failure {
     struct node *node;      // failed state
     uint64_t ctx;           // context that failed (before it failed)
     uint64_t choice;        // choice if any
+    uint64_t address;       // in case of data race
 };
 
 struct code *code;
@@ -158,6 +159,16 @@ void check_invariants(struct node *node, struct context **pctx){
     }
 }
 
+// For tracking data races
+struct access_info *ai_alloc(int multiplicity, int atomic, int pc){
+    struct access_info *ai = calloc(1, sizeof(*ai));
+
+    ai->multiplicity = multiplicity;
+    ai->atomic = atomic;
+    ai->pc = pc;
+    return ai;
+}
+
 void onestep(struct node *node, uint64_t ctx, uint64_t choice, bool interrupt,
         struct dict *visited, struct queue *todo, struct context **pinv_ctx,
         bool infloop_detect, int multiplicity, double timeout){
@@ -190,9 +201,7 @@ void onestep(struct node *node, uint64_t ctx, uint64_t choice, bool interrupt,
 
     bool choosing = false, infinite_loop = false;
     struct dict *infloop = NULL;        // infinite loop detector
-    struct access_info ai;
-    memset(&ai, 0, sizeof(ai));
[…]

### 2021-04-01 book/paper.tex
Working on improving data race (but broken at the moment)
Latest
Got rid of the "interlock" term.  Also provided more information on why load and store operations may not be sequentially consistent.

@@
       threads with private variables that communicate through message passing}
+}
+\newglossaryentry{atomic instruction}
+{
+  name=atomic instruction,
+  description={a machine instruction that may involve  multiple memory load and/or store
+    operations and is executed atomically}
 }
@@
     with high probability}
-}
-\newglossaryentry{interlock instruction}
-{
-  name=interlock instruction,
-  description={a machine instruction that involves multiple memory load and/or store
-    operations, executed atomically}
 }
@@
 if generalized to multiple threads.
-Worse, most modern processors make accessing shared memory locations
-non-atomic to allow for efficient caching, out-of-order execution,
-and other optimizations.
-And even compilers may re-order memory accesses when optimizing code.
-
-Peterson's algorithm relies on a sequential consistent memory model,
+Worse, Peterson relies on load and store operations to be executed
+atomically, but this may not be the case.
+There are a variety of possible reasons for this.
+
+\begin{itemize}
+\item Variables may have more bits than the processor's data bus.
+For example, variables may have 32 bits, but the data bus may only
+have 16 bits.  Thus to store or load a variable takes two 16-bit
+operations each.  Take, for example, a variable that has value
+0xFFFFFFFF, and consider a concurrent load and store operation on
+the variable.
+The store operation wants to clear the variable, but because it
+takes two store operations on the bus, the load operation may
+return either 0xFFFF0000 or 0x0000FFFF, a value that the variable
+never was supposed to have.
+This is the case even if the processor supports a 32-bit load or
+store machine instruction: on the data bus it is still two
+operations.
+\item Modern processors sometimes re-orders load and store operations
+(out-of-order execution) for improved performance.  On a sequential processor,
+the re-ordering is not a problem as the processor only re-orders
+operations on independent memory locations.  However, as \autoref{ex:reorder}
+showed, Peterson's algorithm breaks down if operations are re-ordered.
+Some memory caches can also cause non-atomic behavior of memory when shared
+among multiple cores.
+\item Even compilers, in their code generation, may make optimizations
+that can reorder operations, or even eliminate operations, on variables.
+For example, a compiler may decide that it is unnecessary to read the same
+variable more than once, because how could it possibly change if there
+are no store operations?
+\end{itemize}
+
+Peterson's algorithm relies on a \emph{sequential consistent memory model}
 and hence the \textbf{sequential} statement: without it Harmony will
@@
 program relies on memory load and store instructions operating on
-the indicated variables to be performed sequentially.
+the indicated variables to be performed sequentially, and that this
+order should be consistent with the order of operations invoked on each
+thread.
 The default memory models of C and Java are not sequentially
@@
 Like many constructions in Java, its \texttt{volatile} keyword
-was borrowdd from C and C++.
+was borrowed from C and C++.
 However, in C and C++ they do \emph{not} provide sequential
@@
 
-\glsadd{interlock instruction}%
+\glsadd{atomic instruction}%
 For proper synchronization, multi-core processors provide so-called
-\emph{interlock instructions}: \index{interlock instruction}%
+\emph{atomic instructions}: \index{atomic instruction}%
 special machine instructions
@@
 %
-While the HVM does not have any specific built-in interlock instructions,
+While the HVM does not have any specific built-in atomic instructions
+besides loading and storing variables,
 it does have support for executing multiple instructions atomically.
@@
 We can use \textbf{atomic} statement blocks to implement a wide variety of
-interlock operations.
+atomic operations.
 For example, we could fix the program in \autoref{fig:inc} by
@@
 but uses test-and-set to check and set the lock variable atomically.
+The lock is also cleared in an \textbf{atomic} block to prevent a data race.
 This approach is general for any number of threads.
@@
 In Harmony, \textbf{atomic} statements allow you to \emph{implement} your own
-interlock primitives such as test-and-set.  Other common examples
+atomic primitives such as test-and-set.  Other common examples
 include compare-and swap and fetch-and-add.  \textbf{atomic} statements
@@
 programming languages.  They exist in Harmony for the purpose of
-defining interlock instructions and for testing.
+defining atomic instructions and for testing.
 The statements should only appear in modules that define synchronization
@@
 that removes all occurrences of $v$, if any, from queue $q$.
-\item The test program in \autoref{fig:queuetest} is not very thorough.  Design
+\item The test program in \autoref{fig:queuedemo} is not very thorough.  Design
 and implement a better one.
@@
 array and points to the next slot where a new value is inserted.
-The code uses two interlock instructions:
+The code uses two atomic operations:
 \begin{itemize}
@@
 \item Each slot in the array is only used once, which is inefficient;
-\item the \texttt{inc} and \texttt{exch} interlock instructions are not
+\item the \texttt{inc} and \texttt{exch} atomic operations are not
 universally available on existing processors.
@@
 \hline
-\texttt{wait($c$, $m$)} & wait on condition variable pointed to be $c$ in monitor pointed to by $m$\\
+\texttt{wait($c$, $m$)} & wait on condition variable pointed to by $c$ in monitor pointed to by $m$\\
 \hline
@@
 Heather Zheng,
+Jack Rehmann,
 Jacob Brugh,
@@
 Maria Martucci,
+Phillip O'Reggio,
 Saleh Hassen,

### 2021-04-07 book/paper.tex
Latest

@@
 method that the thread is currently executing.  It is saved and restored by
-method invocations.  The other register is called \emph{this} and is
+method invocations.  The other register is called \textbf{this} and is
 used for so-called \emph{thread-local state}, which contains variables that
@@
 \caption{\harmonylink{code/naiveLock.hny} Na\"{\i}ve implementation of a shared lock.}
-\label{fig:uplock}
+\label{fig:naivelock}
 \end{figure}
@@
 \end{center}
-\caption{The \harmonyref{naiveLock.html}{HTML output of running Harmony on \autoref{fig:uplock}}.}
+\caption{The \harmonyref{naiveLock.html}{HTML output of running Harmony on \autoref{fig:naivelock}}.}
 \label{fig:naiveLockhtml}
@@
 \caption{\harmonylink{code/naiveFlags.hny} Na\"{\i}ve use of flags to solve mutual exclusion.}
-\label{fig:upflags}
+\label{fig:naiveflags}
 \end{figure}
@@
 \end{center}
-\caption{The \harmonyref{naiveFlags.html}{HTML output of running Harmony on \autoref{fig:upflags}}.}
+\caption{The \harmonyref{naiveFlags.html}{HTML output of running Harmony on \autoref{fig:naiveflags}}.}
 \label{fig:naiveflagshtml}
@@
 \caption{\harmonylink{code/naiveTurn.hny} Na\"{\i}ve use of turn variable to solve mutual exclusion.}
-\label{fig:upturn}
+\label{fig:naiveturn}
 \end{figure}
@@
 Using a lock is a good thought, but how does one implement one?
-\autoref{fig:uplock} presents a mutual exclusion attempt based on a
+\autoref{fig:naivelock} presents a mutual exclusion attempt based on a
 na\"{\i}ve (and, as it turns out, broken) implementation of a lock.
@@
 Preventing multiple threads from updating the same variable,
-\autoref{fig:upflags} presents a solution based on each thread having
+\autoref{fig:naiveflags} presents a solution based on each thread having
 a flag indicating that it is trying to enter the critical section.
@@
 The algorithm is an amalgam of the (incorrect) algorithms in
-\autoref{fig:upflags} and \autoref{fig:upturn}, and is presented
+\autoref{fig:naiveflags} and \autoref{fig:naiveturn}, and is presented
 in \autoref{fig:peterson}.
@@
 mutual exclusion problem.  It is similar to the one in
-\autoref{fig:upflags}, but has a thread \emph{back out and try again}
+\autoref{fig:naiveflags}, but has a thread \emph{back out and try again}
 if it finds that the other thread is either trying to enter the critical
@@
 
-\autoref{fig:uplock} showed a faulty attempt at solving mutual
+\autoref{fig:naivelock} showed a faulty attempt at solving mutual
 exclusion using a lock.  The problem with the implementation of the
@@
 \end{code}
-\caption{\harmonylink{code/csTAS.hny} Fixed version of \autoref{fig:uplock} using test-and-set.}
+\caption{\harmonylink{code/csTAS.hny} Fixed version of \autoref{fig:naivelock} using test-and-set.}
 \label{fig:tas2}
@@
 \autoref{fig:tas2} shows a more straightforward implementation of spinlock.
-The solution is similar to the na\"{i}ve solution of \autoref{fig:uplock},
+The solution is similar to the na\"{i}ve solution of \autoref{fig:naivelock},
 but uses test-and-set to check and set the lock variable atomically.
@@
 \end{code}
-\item Add a method $\mathtt{remove}(q, v)$ to \autoref{fig:queue}
+\item \label{ex:qremove} Add a method $\mathtt{remove}(q, v)$ to \autoref{fig:queue}
 that removes all occurrences of $v$, if any, from queue $q$.
-\item The test program in \autoref{fig:queuedemo} is not very thorough.  Design
-and implement a better one.
+\item The test program in \autoref{fig:queuedemo} is a not thorough test program.
+Design and implement a test program for \autoref{fig:queuedemo}.
+Make sure you \emph{test} the test program by trying it out against some buggy
+queue implementations.
 \item Create a thread-safe sorted binary tree.  Implement a module
@@
 \begin{code}
+\harmonysource{cslock}
+\end{code}
+\caption{\harmonylink{code/cslock.hny} Critical section testing without labels.}
+\label{fig:cslock}
+\end{figure}
+
+\begin{figure}
+\begin{code}
 \harmonysource{queuespec}
 \end{code}
-\caption{\harmonylink{code/queuespec.hny} Queue specification.}
+\caption{\harmonylink{code/queuespec.hny} Sequential queue specification.}
 \label{fig:queuespec}
@@
 \label{fig:qtestseq}
+\end{figure}
+
+\begin{figure}
+\begin{code}
+\harmonysource{queuelin}
+\end{code}
+\caption{\harmonylink{code/queuelin.hny} Queue implementation with linearization points.}
+\label{fig:queuelin}
+\end{figure}
+
+\begin{figure}
+\begin{code}
+\harmonysource{qtestconc}
+\end{code}
+\caption{\harmonylink{code/qtestconc.hny} Concurrent queue test.}
+\label{fig:qtestconc}
 \end{figure}
@@
 and does little to increase our confidence in its correctness.
+To wit, if \texttt{get()} always returned 1, the program
+would find no problems.
 In this chapter, we will look at approaches to testing
@@
 
-The first step in testing a concurrent data structure is
-whether it works in the absence of concurrency.
-The objective is to test if the queue \emph{implementation}
-in \autoref{fig:queue} meets the sequential
-queue \emph{specification}.
-In order to do so, we should formalize the sequential queue specification.
-A specification is just code.
+The framework in \autoref{fig:csbarebones} works well for thoroughly
+testing mutual exclusion and progress in critical sections, but depends
+on \texttt{atLabel}, an unusual operator specific to Harmony that is mostly
+useful for testing mutual exclusion.
+It turns out that we do not need labels to express this.
+\autoref{fig:cslock} tests mutual exclusion and progress for
+critical sections implemented using locks.
+The test uses a simple atomic counter that is incremented before entering
+the critical section and decremented after leaving it.
+Mutual exclusion holds if the counter never goes over 1.
+
+\emph{Test programs themselves should be tested}.
+Just because \emph{your} test program works with \emph{your} implementation
+does not mean your implementation is correct---it may be that both are
+broken.
+In the case of testing mutual exclusion with atomic counters, you will want
+to see if you can use this method to find the problems in \autoref{fig:naivelock},
+\autoref{fig:naiveflags},
+and \autoref{fig:naiveturn}.
+You also want to check if an atomic counter solution works with
+\autoref{fig:peterson} and \autoref{fig:tas}.
+
+As with critical sections, when testing a concurrent data structure
+we need a specification for its correctness.
+A good place to start is thinking about a \emph{sequential} specification
+for queues.
+A specification is simply a program, written at a high level of
+abstraction.
 \autoref{fig:queuespec} shows a sequential specification of a
@@
 module described in \autoref{ap:list}).
+
+First we can check if the queue implementation
+in \autoref{fig:queue} meets the sequential
+queue specification in \autoref{fig:queuespec}.
 To check if the queue implementation meets the specification,
@@
 have the same \emph{behavior}.
-
+%
 \autoref{fig:qtestseq} presents a test program that does
@@
 decides which value to append to the queue.
-All operations are performed on both the implementation
-queue and the specification queue.
-In the case of \texttt{get}, the result of the operation
-is checked against that of the specification.
+All operations are performed on both the queue implementation
+and the queue specification.
+In the case of \texttt{get}, the results of the operation
+on both the implementation and specification are checked against one another.
 
@@
 makes it relatively easy to explore all corner cases.
-
+%
 For example, a common programming mistake is to forget to
@@
 is reasonably thorough.
-
-\begin{figure}
-\begin{code}
-\harmonysource{qtest1}
-\end{code}
-\caption{\harmonylink{code/qtest1.hny} Concurrent \texttt{put()}.}
-\label{fig:qtest1}
-\end{figure}
 
@@
 when it is allocated.
+
+Still, it would be useful to check in Harmony that
+\autoref{fig:queue} is linearizable.
+To do this, instead of applying the operations sequentially,
+we want tthe test program to invoke the operations concurrently,
+consider all possible interleavings, and see if the result
+is consistent with an appropriate sequential execution of the
+operations.
+
+Harmony provides support, but requires that the programmer
+identify what are known as \emph{linearization points}
+in the implementation that indicate exactly \emph{which} sequential
+execution the concurrent execution must align with.
+\autoref{fig:queuelin} is a copy of \autoref{fig:queue} extended with
+linearization points.
+For each operation (\texttt{get} and \texttt{put}), the corresponding
+linearization point must occur somewhere between acquiring and releasing
+the lock.
+%
+Each linearization point execution is assigned a logical timestamp.
+Logical timestamps are numbered $0, 1, ...$
+To do so, we have added a counter (\texttt{time}) to the \texttt{Queue}.
+Method \texttt{\_linpoint} saves the current counter in
+\textbf{this}.\texttt{qtime} and increments the counter.
+The \textbf{this} dictionary maintains \emph{thread-local state} associated
+with the thread (\autoref{ch:harmonymachine})---it contains variables that
+can be accessed by any method in the thread.
+
+Given the linearization points, \autoref{fig:qtestconc} shows how
+linearizability can be tested.
+The test program is similar to the sequential test program
+(\autoref{fig:qtestseq} but starts a thread for each operation.
+The operations are executed concurrently on the concurrent queue
+implementation of \autoref{fig:queuelin}, but they are executed sequentially
+on the sequential queue specification of \autoref{fig:queuespec}.
+To that end, the test program maintains a global time variable
+\textit{qtime}, and each thread waits until the timestamp assigned to
+the last concurrent queue operation matches \textit{qtime} before invoking
+the sequential operation in the specification.
+Afterward, it atomically increments the shared \textit{qtime} variable.
+
+\section*{Exercises}
+\begin{problems}
+\item Write a sequential specification for \autoref{fig:linkedlist}.
+(Hint: it implements a set of integers.)
+Write a Harmony program that checks if \autoref{fig:linkedlist} meets
+the sequential specification.
+\item Extend \autoref{fig:linkedlist} with linearization points and
+check if the implementation is linearizable.
+\item Check if the extended queue implementation of \autoref{ex:qremove}
+is linearizable.
+\end{problems}
+
+\begin{comment}
+
+\begin{figure}
+\begin{code}
+\harmonysource{qtest1}
+\end{code}
+\caption{\harmonylink{code/qtest1.hny} Concurrent \texttt{put()}.}
+\label{fig:qtest1}
+\end{figure}
 
@@
 \end{figure}
+
+\end{comment}
 

### 2021-04-07 book/sources/pp.py

@@ -93,6 +93,7 @@ def isreserved(s):
         "setintlevel",
         "spawn",
         "stop",
+        "this",
         "trap",
         "True",
         "where",
@@ -127,6 +128,7 @@ def lexer(s, file):
     line = 1
     column = 1
     nextconst = False;
+    importline = False
     while s != "":
         if column == 1:
             if line != 1: print("\\\\\n")
@@ -144,6 +146,7 @@ def lexer(s, file):
             line += 1
             column = 1
             nextconst = False
+            importline = False
             continue
 
         # skip over line comments
@@ -161,6 +164,7 @@ def lexer(s, file):
                 s = s[1:]
             print("}", end="")
             nextconst = False
+            importline = False
             continue
 
         # skip over nested comments
@@ -185,6 +189,7 @@ def lexer(s, file):
                     s = s[1:]
                     column += 1
             nextconst = False
+            importline = False
             continue
 
         # see if it's a multi-character token.  Match with the longest one
@@ -216,9 +221,10 @@ def lexer(s, file):
             found = s[:i].replace("_", "\\_")
             if isreserved(found):
                 print("\\texttt{\\textbf{%s}}"%found, end="")
-                nextconst = found in [ "def", "const", "import", "from" ]
+                nextconst = found in [ "def", "const" ]
+                importline = found in [ "import", "from" ]
             elif isletter(found[0]):
-                if nextconst:
+                if nextconst or importline:
                     constants.add(found)
                     print("%s"%found, end="")
                     nextconst = False
@@ -266,6 +272,7 @@ def lexer(s, file):
             str += '"'
             print("''''", end="")
             nextconst = False
+            importline = False
             result += [ (str, file, line, column) ]
             s = s[i:]
             column += i

### 2021-04-07 book/paper.tex
Latest

@@
 \emph{Test programs themselves should be tested}.
-Just because \emph{your} test program works with \emph{your} implementation
-does not mean your implementation is correct---it may be that both are
-broken.
+Just because a test program works with a particular implementation
+does not mean the implementation is correct---it may be that the implementation
+is incorrect but the test program does not have enough coverage to find any
+bugs in the implementation.
 In the case of testing mutual exclusion with atomic counters, you will want
@@
 and \autoref{fig:naiveturn}.
-You also want to check if an atomic counter solution works with
-\autoref{fig:peterson} and \autoref{fig:tas}.
+Conversely, a test program may be broken in that it finds bugs that do not exist.
+So, in this case,
+you also want to check if an atomic counter solution works with known
+correct solutions such as \autoref{fig:peterson} and \autoref{fig:tas}
+if possible.
 
@@
 specification.
-We say that the implementation and the specification should
+We say that the implementation and the specification
 have the same \emph{behavior}.
@@
 in \autoref{fig:queue} and run the test program---it should
-easily find the bug and the output explains exactly how
-the bug can happen, adding confidence that the test program
+easily find the bug and explain exactly how
+the bug manifests itself, adding confidence that the test program
 is reasonably thorough.
@@
 To do this, instead of applying the operations sequentially,
-we want tthe test program to invoke the operations concurrently,
+we want the test program to invoke the operations concurrently,
 consider all possible interleavings, and see if the result
@@
 
-Harmony provides support, but requires that the programmer
-identify what are known as \emph{linearization points}
+Harmony provides support for testing linearizability,
+but requires that the programmer
+identifies what are known as \emph{linearization points}
 in the implementation that indicate exactly \emph{which} sequential
@@
 The test program is similar to the sequential test program
-(\autoref{fig:qtestseq} but starts a thread for each operation.
+(\autoref{fig:qtestseq}) but starts a thread for each operation.
 The operations are executed concurrently on the concurrent queue
@@
 Afterward, it atomically increments the shared \textit{qtime} variable.
-
-\section*{Exercises}
-\begin{problems}
-\item Write a sequential specification for \autoref{fig:linkedlist}.
-(Hint: it implements a set of integers.)
-Write a Harmony program that checks if \autoref{fig:linkedlist} meets
-the sequential specification.
-\item Extend \autoref{fig:linkedlist} with linearization points and
-check if the implementation is linearizable.
-\item Check if the extended queue implementation of \autoref{ex:qremove}
-is linearizable.
-\end{problems}
-
-\begin{comment}
+This results in the operations being executed sequentially against the
+sequential specification in the same order of the linearization points
+of the concurrent specification.
 
@@
 
-While it is possible to check linearizability with
-model checking, Harmony does not (yet) have explicit
-support for this.  Instead, we will present a few test
+If you want a purely black box test for testing a queue implementation,
+then adding linearization points is not possible.
+We will present a few test
 programs that try to identify incorrect behavior of
-the queue implementation in the presence of concurrency.
+a black box queue implementation in the presence of concurrency.
 
@@
 $\{ 0 .. \mathtt{N} - 1 \}$.
-Note that it is perfectly fine to use \textbf{sequential}
-and \textbf{atomic} statements in test code---one should
-only avoid those in production code.
+% Note that it is perfectly fine to use \textbf{sequential}
+% and \textbf{atomic} statements in test code---one should
+% only avoid those in production code.
 
@@
 
-\end{comment}
+\section*{Exercises}
+\begin{problems}
+\item Write a sequential specification for \autoref{fig:linkedlist}.
+(Hint: it implements a set of integers.)
+Write a Harmony program that checks if \autoref{fig:linkedlist} meets
+the sequential specification.
+\item Extend \autoref{fig:linkedlist} with linearization points and
+check if the implementation is linearizable.
+\item Check if the extended queue implementation of \autoref{ex:qremove}
+is linearizable.
+\end{problems}
 

### 2021-04-08 book/paper.tex
Latest

@@
 Harmony from complaining about data races involving these variables,
-explained in \autoref{ch:synch}.)
+explained in \autoref{ch:spinlock}.)
 

### 2021-04-12 book/paper.tex
Latest

@@
 data structures.
+\item \autoref{ch:testing} discusses approaches to testing
+concurrent code in Harmony.
 \item \autoref{ch:condwait} talks about threads having to wait for
@@
 problem and the bounded buffer problem.
-\item \autoref{ch:testing} discusses approaches to testing
-concurrent code in Harmony.
 \item \autoref{ch:sbs} presents \emph{Split Binary Semaphores}, a
@@
 Jacob Brugh,
+Liam Arzola,
 Lorenzo Alvisi,

### 2021-04-15 book/paper.tex
Improved barrier synch description

@@
 
-\autoref{fig:genbarrier} shows the code.
+But this poses a complication.  Suppose, for example, that there are two
+seats in the car, and there is one person in the car.  Does that mean that
+the car is not yet full, or not yet empty?  We have to distinguish
+those situations.  To this end, it is useful to think of the car as going through
+$2 \cdot n$ stages:
+\begin{itemize}
+\item[] Stage 0: the car is empty;
+\item[] Stage $1 ... n-1$: the car is filling up;
+\item[] Stage $n$: the car is full;
+\item[] Stage $n+1 ... 2n - 1$: the car is emptying out.
+\end{itemize}
+
+\autoref{fig:genbarrier} shows code that implements barrier synchronization
+using these stages.
 Method~\texttt{enter}() has two \texttt{wait} loops, one for each
@@
 the second waits for the car to fill up.
-
-Notice how variables \textit{entered} and \textit{left}
-are maintained.  They are incremented one-by-one, but they
-are not decremented one-by-one.  Instead, the last thread to
-enter sets \textit{left} to zero (and notifies the threads
-that were waiting to for the car to fill up), while the last
-thread to exit sets \textit{entered} to zero (and notifies
-the threads that were waiting for the car to empty out).
-
 \autoref{fig:barriertest} is a test program.  The threads check

### 2021-04-16 book/paper.tex
Latest

@@
 
+\begin{figure}
+\begin{code}
+\harmonysource{clock}
+\end{code}
+\caption{\harmonylink{code/clock.hny} Harmony program that finds page replacement anomalies.}
+\label{fig:clock}
+\end{figure}
+
+If you are ready to learn about how locks are implemented in practice,
+you can now skip the rest of this chapter.  But if you would like to
+see a cool example of using the concepts introduced in this chapter,
+hang on for a sequential Harmony program that finds anomalies in page
+replacement algorithms.
+In 1969, B\'{e}l\'{a}ady et al. published a paper~\cite{Belady69}
+that showed that making a
+cache larger does not necessarily lead to a higher hit ratio.  He showed
+this for a scenario using a FIFO replacement policy when the cache is full. 
+The program in \autoref{fig:clock} will find exactly the same scenario
+if you define \texttt{FIFO} to be \textbf{True}.  Moreover, if you define
+\texttt{FIFO} to be \textbf{False}, it will find a scenario for the
+CLOCK replacement policy~\cite{clock}, often used in modern operating systems.
+
+In this program, \texttt{CLOCK} maintains the state of a cache
+(in practice, typically pages in memory).
+The set \texttt{recent} maintains whether an access to the cache
+for a particular reference was recent or not.
+(It is not used if \texttt{FIFO} is \textbf{True}.)
+The integer \texttt{misses} maintains the number of cache misses.
+Method \texttt{ref}(\textit{ck}, $x$) is invoked when $x$ is
+referenced and checked against the cache \textit{ck}.
+
+The program declares two caches: one with 3 entries (\textit{clock3})
+and one with 4 entries (\textit{clock4}).
+The interesting part is in the last block of code.
+It runs every sequence of references of up to 100 entries, using
+references in the range 1 through 5.
+Note that all the constants chosen in this program (3, 4, 5, 100)
+are the result of some experimentation---you can try other ones.
+To reduce the search space, the first four references are pinned
+to 1, 2, 3, and 4.
+Further reducing the search space, the program never repeats the
+same reference twice in a row (using the local variable \texttt{last}).
+
+The two things to note here is the use of the \textbf{choose}
+expression and the \textbf{assert} statement.  Using \texttt{choose}
+we are able to express searching through all possible strings of
+references without a complicated nested iteration.
+Using \texttt{assert} we are able to express the anomaly we are
+looking for.
+
+In case you want to check if you get the right results.  For \texttt{FIFO},
+the program finds (in a few minutes) the same anomaly that
+B\'{e}l\'{a}dy et al. found: 1 2 3 4 1 2 5 1 2 3 4 5.
+For the \texttt{CLOCK} algorithm the program actually finds a shorter
+reference string: 1 2 3 4 2 1 2 5 1 2.
+
 \chapter{Spinlock}

### 2021-04-18 book/paper.tex
New chapter: debugging

@@
 concurrent code in Harmony.
+\item \autoref{ch:debugging} instead goes into how to find a bug
+in concurrent code using the Harmony output.
 \item \autoref{ch:condwait} talks about threads having to wait for
@@
 \end{problems}
+
+\chapter{Debugging}
+\label{ch:debugging}
+
+\begin{figure}
+\begin{code}
+\harmonysource{queuebug}
+\end{code}
+\caption{\harmonylink{code/queuebug.hny} A buggy queue implementation.}
+\label{fig:queuebug}
+\end{figure}
+
+\begin{figure}
+\begin{center}
+\includegraphics[width=\textwidth]{figures/queuefix1.pdf}
+\end{center}
+\caption{Harmony output.}
+\label{fig:queuefix1}
+\end{figure}
+
+So you wrote a Harmony program and Harmony reports a bug.
+Often you may just be able to figure it out by staring at the code and
+going through some easy scenarios, but what if you don't?
+The output of Harmony can be helpful in that case.
+
+\autoref{fig:queuebug} contains an attempt at a queue implementation
+where the queue is implemented by a linked list, with the first node
+being a \texttt{dummy} node to prevent data races.  Each node in the
+list contains a lock.  The \texttt{put}() method walks the list until
+it gets to the last node, each time acquiring the lock to access the
+node's fields.  When \texttt{put}() gets to the last node in the list,
+it appends a new one.  The \texttt{get}() method locks the first (dummy)
+node, removes the second from the list and frees it.  The method returns
+the value from the removed node.
+
+Let us run the code through some of the test programs in the last
+chapter.  Harmony does not detect any issues with the sequential test
+in \autoref{fig:qtestseq}.  The concurrent \texttt{put} test of
+\autoref{fig:qtest1} as well as the concurrent \texttt{get} test of
+\autoref{fig:qtest2} also find no problems.  However, when we run
+the new queue code through the test in \autoref{fig:qtest3}, Harmony
+finds a problem.
+
+\autoref{fig:queuefix1} shows the Harmony output of running
+the test in \autoref{fig:qtest3} against the queue code
+in \autoref{fig:queuebug}.
+There is quite a bit of information in the Harmony output, and it's
+important to learn to navigate through it.
+Let's start with looking at the red text.  Harmony found a safety
+violation (something bad happened during one of the possible executions),
+and in particular \texttt{putter}(1) (thread T2) was trying to dereference
+\texttt{alloc}.\textit{pool}[0].\texttt{lock} in turn 5.
+
+The \texttt{alloc} module maintains a shared array \textit{pool} that
+it uses for dynamic allocation.  Apparently \texttt{putter}(1) tried to access
+\textit{pool}[0], but it does not exist, meaning that either it was not
+yet allocated, or it had been freed since it was allocated.
+When we look at the top half of the figure, we see that in fact
+\texttt{putter}(0) (T1) allocated \textit{pool}[0] in turn 2, while
+\texttt{getter}() (T4) released it in turn 4.
+
+So how did we get there?  We can start by single stepping through the
+actions of \texttt{putter}(0) by clicking on its first block.
+By hitting return repeatedly, we can go through the lines of code that
+it is executing.  Doing so, we can see that it allocates
+\textit{pool}[0] and uses that as the node to contain value 0 and add
+this node to the queue \textit{qtest}.  At the end of its turn,
+\texttt{putter}(0) has just finished \texttt{put}(0).  Looking at the
+state in the top right, everything looks good.  The first node
+(\textit{testq}) points to the allocated node with the value 0 in it,
+and its \texttt{next} pointer is \textbf{None}.  The locks on both nodes
+are released.  So far so good.
+
+Next \texttt{putter}(1) (thread T2) takes a turn.  It gets to line 10 in
+\autoref{fig:queuebug} where it's trying to acquire
+$q$$\rightarrow$\texttt{lock}.  If we look at the bottom right,
+we see that $q$ points to \textit{pool}[0], the node that has value 0
+in it.  Again, so far so good.
+
+But before \texttt{putter}(1) obtains
+the lock on \textit{pool}[0], \texttt{getter}() (thread T4) starts
+running.  \texttt{getter}() acquires the lock on \textit{qtest}, which
+still points to \textit{pool}[0].  \texttt{getter} then extracts the
+value from \textit{pool}[0] and then releases it.
+At this point \textit{pool}[0] no longer exists, but then
+\texttt{putter}(1) starts running again from where it left off.
+It was about to acquire the lock on \textit{pool}[0], but now that
+node no longer exists.
+
+To fix the code, we have to use hand-over-hand locking (\autoref{ch:cds}).
+\autoref{fig:queuefix} shows an implementation that uses the same
+data structure as \autoref{fig:queuebug} but uses hand-over-hand locking,
+both for \texttt{put}() and for \texttt{get}().  It passes all tests.
+
+\begin{figure}
+\begin{code}
+\harmonysource{queuefix}
+\end{code}
+\caption{\harmonylink{code/queuefix.hny} Queue implementation with hand-over-hand locking.}
+\label{fig:queuefix}
+\end{figure}
 

### 2021-04-18 book/paper.tex
Latest

@@
 \begin{center}
-\includegraphics[width=\textwidth]{figures/queuefix1.pdf}
+\includegraphics[width=\textwidth]{figures/queuebug1.pdf}
 \end{center}
-\caption{Harmony output.}
-\label{fig:queuefix1}
+\caption{Harmony output of running \autoref{fig:qtest3} against
+\autoref{fig:queuebug}.}
+\label{fig:queuebug1}
 \end{figure}
@@
 
-\autoref{fig:queuefix1} shows the Harmony output of running
+\autoref{fig:queuebug1} shows the Harmony output of running
 the test in \autoref{fig:qtest3} against the queue code

### 2021-04-18 book/paper.tex
Latest

@@
 \end{center}
-\caption{Harmony output of running \autoref{fig:qtest3} against
-\autoref{fig:queuebug}.}
+\caption{\harmonyref{queuebug.html}{Harmony output of running \autoref{fig:qtest3} against
+\autoref{fig:queuebug}}.}
 \label{fig:queuebug1}
@@
 chapter.  Harmony does not detect any issues with the sequential test
-in \autoref{fig:qtestseq}.  The concurrent \texttt{put} test of
+in \autoref{fig:qtestseq}.
+(Run this using the \texttt{-m} flag like this:
+\texttt{harmony -m queue=queuebug code/qtestseq.hny})
+The concurrent \texttt{put} test of
 \autoref{fig:qtest1} as well as the concurrent \texttt{get} test of
@@
 
-To fix the code, we have to use hand-over-hand locking (\autoref{ch:cds}).
+To fix the code without changing the data structure, we can use
+hand-over-hand locking (\autoref{ch:cds}).
 \autoref{fig:queuefix} shows an implementation that uses the same

### 2021-04-18 book/paper.tex
Latest

@@
 hand-over-hand locking (\autoref{ch:cds}).
-\autoref{fig:queuefix} shows an implementation that uses the same
-data structure as \autoref{fig:queuebug} but uses hand-over-hand locking,
+\autoref{fig:queuefix} shows an implementation
+that uses hand-over-hand locking
 both for \texttt{put}() and for \texttt{get}().  It passes all tests.

### 2021-07-11 book/paper.tex
Latest

@@
 \hline
-\texttt{add}(\textit{pb}, $e$) & add one copy of $e$ to bag \texttt{!}\textit{pb}\\
-\hline
-\texttt{remove}(\textit{pb}, $e$) & remove one copy of $e$ from bag \texttt{!}\textit{pb}\\
+\texttt{add}($b$, $e$) & add one copy of $e$ to bag $b$ \\
+\hline
+\texttt{remove}($b$, $e$) & remove one copy of $e$ from bag $b$ \\
+\hline
+\texttt{combinations}($b$, $k$) & return set of all \emph{subbags} of size $k$ \\
 \hline
@@
 \texttt{issuperset($s$, $t$)} & returns whether $s$ is a superset of $t$ \\
+\hline
+\texttt{add}($s$, $e$) & returns $s \cup \{ e \}$ \\
+\hline
+\texttt{remove}($s$, $e$) & returns $s - \{ e \}$ \\
+\hline
+\texttt{combinations}($b$, $k$) & returns set of all subsets of size $k$ \\
 \hline
@@
 register (\emph{i.e.}, the values of the thread variables) onto the
-stack.  It initializes the \texttt{result} variable to the empty dictionary.
+stack.  It initializes the \texttt{result} variable to \texttt{None}.
 The \texttt{Return} instruction restores the thread register by popping
@@
 Also, each method has an implicitly declared
-\textit{result} variable.
+\textit{result} variable, which is initialized to \texttt{None}.
 Local variables are tightly scoped and cannot be shared
@@
 \texttt{where} $i < j$: ...
+\end{code}
+
+In case of dictionaries, you can access both keys and values:
+\begin{code}
+\texttt{for} $k:v$ in \{ .a: 1, .b: 2 \}: ...
 \end{code}

### 2021-07-11 book/paper.tex
Latest
Starting to add networking stuff

@@
 
+\chapter{Distributed Consensus}
+\label{ch:consensus}
+
+\begin{figure}
+\begin{code}
+\harmonysource{bosco}
+\end{code}
+\caption{\harmonylink{code/bosco.hny} A simple fault-tolerant consensus protocol.}
+\label{fig:bosco}
+\end{figure}
+
+\chapter{Paxos}
+\label{ch:paxos}
+
+\begin{figure}
+\begin{code}
+\harmonysource{paxos}
+\end{code}
+\caption{\harmonylink{code/paxos.hny} A version of the Paxos protocol.}
+\label{fig:paxos}
+\end{figure}
+
 \bibliographystyle{alpha}
@@
 \hline
-\texttt{for $a$ in $e$: S} & $a$ is a bounded variable, $e$ is a set,
+\texttt{for $a$ in $e$: S} & $a$ is a bounded variable, $e$ is a set or dictionary,
                             \texttt{S} a list of statements\\
@@
 \texttt{pass} & do nothing\\
+\hline
+\texttt{select $a$ in $e$: S} & $a$ is a bounded variable, $e$ is a set,
+                            \texttt{S} a list of statements\\
 \hline

### 2021-07-11 book/sources/pp.py

@@ -89,6 +89,7 @@ def isreserved(s):
         "not",
         "or",
         "pass",
+        "select",
         "sequential",
         "setintlevel",
         "spawn",

### 2021-07-11 book/paper.tex
Writing the consensus chapter

@@
 (You may assume that \texttt{load} and \texttt{store} instructions of the
-underlying machine architecture are atomic (indivisible)---in fact they are.)
+underlying virtual machine architecture are atomic (indivisible)---in fact
+they are.)
 \end{itemize}
@@
 \index{HVM}%
+The Harmony compiler places the bytecode for file $x$.\texttt{htm} in
+file $x$.\texttt{hvm}.
+The model checker, called \emph{Charm}, places its output in a file
+called $x$.\texttt{hco} and a stylized version of the same output
+in an HTML file called $x$.\texttt{htm}.
+
 To understand the problem of concurrent computing, it
@@
 \end{code}
-\caption{\harmonylink{code/bosco.hny} A simple fault-tolerant consensus protocol.}
+\caption{\harmonylink{code/bosco.hny} A crash-tolerant consensus protocol.}
 \label{fig:bosco}
 \end{figure}
+
+Distributed consensus is the problem of having a collection of processors agree
+on a single value over a network.  This often has to be done in adverse circumstances,
+for example in the face of processor failures.
+In Harmony, processors can be modeled by threads and there are a variety of ways in
+which one can model a network using shared variables.
+In this chapter, we will present a simple consensus algorithm that can tolerate
+fewer than $1/3^{rd}$ of processors failing by crashing.
+
+More precisely, we will
+define a constant \texttt{F} that contains the maximum number of failures, and
+assume there are $\texttt{N} = 3\texttt{F} + 1$ processors.
+Each processor \emph{proposes} a value, which we assume here is from the
+set \{ 0, 1 \}.
+By the definition of consensus, we want the following two propertiew:
+\begin{enumerate}
+\item Validity: a processor can only decide a value that has been proposed;
+\item Agreement: if two processors decide, they decide the same value.
+\end{enumerate}
+The consensus problem is surprisingly tricky to solve in the face of processor
+failures.
+
+We will model the network as a bag (aka multiset) of messages.  The bag itself is
+modeled in Harmony as a dictionary that maps unique elements in the bag to their
+multiplicity.  So, for example, $\{ .\texttt{a}:1, .\texttt{b}:2 \}$ is the bag that consists of
+one $.\texttt{a}$ and two copies of $.\texttt{b}$.
+The \texttt{bag} module (\autoref{ap:bag}) contains a variety of handy functions on bags.
+In this particular algorithm, all messages are broadcast to all processors, so they do
+not require a destination address.
+
+\autoref{fig:bosco} presents our algorithm.  Besides the network variable, we will use a
+shared list of proposals and a shared set of decisions.  The invariants we want to maintain
+on those variables is that $\textit{decisions} \subseteq \textit{proposals}$
+and $|\textit{decisions}| \leq 1$.  Since the initial value of \textit{decisions} is
+$\emptyset$, the invariants clearly hold initially.
+
+The $N$ processors after go through a sequence of \emph{rounds} in which they wait for
+$\texttt{N} - \texttt{F}$ messages, update their state based on the messages, and
+broadcast a new messages containing their new state.
+The reason that a processor waits for $\texttt{N} - \texttt{F}$ rather than \texttt{N}
+messages is because of failures: up to $\texttt{F}$ processors may never send a message
+and so it would be unwise to wait for all $\texttt{N}$.  You might be tempted to use
+a timer and time out on waiting for a particular processor.  But how would you initialize
+that timer?  While we will assume that the network is reliable, there is no guarantee
+on that messages arrive within a particular time.
+We call a set of $\texttt{N} - \texttt{F}$ processors a \emph{quorum}.
+A quorum must suffice for the algorithm to make progress.
+
+The state of a processor consists
+of its current round number and proposal.  Therefore, messages consist of a tuple containing
+a round number and a proposal. To start things, each processor first
+broadcasts its initial round number (0) and proposal.
+The number of rounds that are necessary to achieve consensus is not bounded.  But
+Harmony can only check finite models, so there is a constant \texttt{NROUNDS} that
+limits the number of rounds.  It may be that no decisions are made, but that does not
+violate either Validity or Agreement.  We only check that \emph{if} decisions are
+made, they satisfy Validity and Agreement.
+
+\index{select statement}%
+In Line~21, a processor waits for $\texttt{N} - \texttt{F}$ messages using
+the Harmony \textbf{select} statement.
+This statement takes the form \textbf{select} $v$ \textbf{in} $s$: \emph{statement block}
+where $s$ is a set and $v$ is variable that is bound to an element of $s$.
+The important properties of a \textbf{select} statement are as follows:
+\begin{itemize}
+\item it waits until $s$ is non-empty;
+\item it is executed atomically;
+\item $v$ is selected non-deterministically, like in the \texttt{choose} operator.
+\end{itemize}
+
+Since Harmony has to check all possible executions of the protocol, the
+\texttt{receive}(\textit{round}, $k$) method returns all \emph{subbags} of messages
+for the given round that have size $k = \texttt{N} - \texttt{F}$.  The method uses a
+dictionary comprehension to filter out all messages for the given \textit{round}
+and then uses the \texttt{bag.combinations} method to find all combinations of size $k$.
+The \textbf{select} statement waits until there is at least one such combination and
+then chooses an element that is bound to the \textit{quorum} variable.
+The body of the \textbf{select} statement is then executed atomically.
+This is usually how distributed algorithms are modeled, because they can only interact
+through the network.  There is no need to interleave the different processes other
+than when messages are delivered.  By executing the body atomically, a lot of
+unnecessary interleavings are avoided and this reduces the state space that must
+be explored by the model checker significantly.
+
+The body of the \textbf{select} statement contains the core of the algorithm.
+Note that $\texttt{N} - \texttt{F} = 2\texttt{F} + 1$, so that the number of
+messages is guaranteed to be odd.  Also, because there are only 0 and 1 values, there
+must exist a majority of zeroes or ones.  Variable \textit{count}[0] stores the number
+of zeroes and \texttt{count}[1] stores the number of ones received in the round.
+The rules of the algorithm are simple:
+\begin{itemize}
+\item update \textit{proposal} to be the majority value;
+\item if the quorum was unanimous, decide the value.
+\end{itemize}
+After that, proceed with the next round.
 

### 2021-07-12 book/paper.tex
Latest

@@
 \usepackage{xcolor}
-\usepackage[colorlinks=true, linkcolor=blue, urlcolor=blue]{hyperref}
+\usepackage[colorlinks=true, linkcolor=blue, urlcolor=blue, citecolor=violet]{hyperref}
 \def\itemautorefname{Exercise}%
@@
 set \{ 0, 1 \}.
-By the definition of consensus, we want the following two propertiew:
+By the definition of consensus, we want the following two properties:
 \begin{enumerate}
-\item Validity: a processor can only decide a value that has been proposed;
-\item Agreement: if two processors decide, they decide the same value.
+\item \emph{Validity}: a processor can only decide a value that has been proposed;
+\item \emph{Agreement}: if two processors decide, they decide the same value.
 \end{enumerate}
@@
 
-\autoref{fig:bosco} presents our algorithm.  Besides the network variable, we will use a
+\autoref{fig:bosco} presents our algorithm.  Besides the network variable, it uses a
 shared list of proposals and a shared set of decisions.  The invariants we want to maintain
@@
 
-The $N$ processors after go through a sequence of \emph{rounds} in which they wait for
+The $N$ processors go through a sequence of \emph{rounds} in which they wait for
 $\texttt{N} - \texttt{F}$ messages, update their state based on the messages, and
-broadcast a new messages containing their new state.
+broadcast messages containing their new state.
 The reason that a processor waits for $\texttt{N} - \texttt{F}$ rather than \texttt{N}
@@
 that timer?  While we will assume that the network is reliable, there is no guarantee
-on that messages arrive within a particular time.
+that messages arrive within a particular time.
 We call a set of $\texttt{N} - \texttt{F}$ processors a \emph{quorum}.
@@
 where $s$ is a set and $v$ is variable that is bound to an element of $s$.
-The important properties of a \textbf{select} statement are as follows:
+The important properties of this \textbf{select} statement are as follows:
 \begin{itemize}
@@
 The \textbf{select} statement waits until there is at least one such combination and
-then chooses an element that is bound to the \textit{quorum} variable.
+then chooses an element, which is bound to the \textit{quorum} variable.
 The body of the \textbf{select} statement is then executed atomically.
@@
 \item update \textit{proposal} to be the majority value;
-\item if the quorum was unanimous, decide the value.
+\item if the quorum is unanimous, decide the value.
 \end{itemize}
@@
 
+One may wonder how failures can occur in this model.  They are not explicitly modeled, but
+Harmony tries every possible execution, including executions in which some processors make
+no execution steps unless there are no other processors that can run.  The way this algorithm
+works, this includes executions in which $\texttt{N} - \texttt{F}$ processors terminate
+before the remaining $\texttt{F}$ processors start executing.  To the first set of processors,
+this is indistinguishable from executions in which the last set of processors have failed.
+
 \chapter{Paxos}
@@
 \begin{code}
-\harmonysource{paxos}
-\end{code}
-\caption{\harmonylink{code/paxos.hny} A version of the Paxos protocol.}
-\label{fig:paxos}
-\end{figure}
+\harmonysource{paxos1}
+\end{code}
+\caption{\harmonylink{code/paxos.hny} A version of the Paxos protocol, Part 1.}
+\label{fig:paxos1}
+\end{figure}
+
+\begin{figure}
+\begin{code}
+\harmonysource{paxos2}
+\end{code}
+\caption{\harmonylink{code/paxos.hny} A version of the Paxos protocol, Part 2.}
+\label{fig:paxos2}
+\end{figure}
+
+Paxos~\cite{Paxos} is the most well-known family of consensus protocols for
+environments in which few or no assumptions are made about timing.
+In this chapter we present a basic version of a Paxos protocol, one that is
+\emph{single-decree} (only tries to make a single decision).
+It uses two kinds of processors: \emph{leaders} and \emph{acceptors}.
+In order to tolerate \texttt{F} crash failures, you need at least $\texttt{F}+1$ leaders
+and $2\texttt{F} + 1$ acceptors, but leaders and acceptors can be colocated, so
+in total only $2\texttt{F} + 1$ independently failing processors are needed.
+Here we provide only a rudimentary introduction to Paxos; for more detailed
+information refer to~\cite{Paxos}.
+
+As in the consensus protocol of \autoref{ch:consensus}, Paxos uses rounds of messaging.
+However, Paxos uses two kinds of rounds: ``Phase 1'' and ``Phase 2'' rounds.
+The communication pattern is also different.  Rounds are identified by a so-called
+\emph{ballot number} combined with the phase number.
+Different leaders are in charge of different ballot numbers.
+Leaders broadcast ``Type A''
+messages to the acceptors, which respond point-to-point with ``Type B'' messages.
+
+\autoref{fig:paxos1} and \autoref{fig:paxos2} contain the code for this Paxos protocol.
+A leader first broadcasts a ``p1a'' message (Phase 1, Type A) to the acceptors
+containing only the ballot number and then waits for $\texttt{N} - \texttt{F}$ ``p1b''
+responses from the acceptors.
 

### 2021-07-12 book/paper.tex
Latest
Put in a description of the Paxos protocol

@@
 network.
+\item \autoref{ch:consensus} demonstrates a fault-tolerant distributed consensus
+algorithm (aka protocol) expressed in Harmony.
+\item \autoref{ch:paxos} shows how one can specify and check the well-known
+Paxos consensus protocol.
 \end{itemize}
@@
 
-\autoref{fig:paxos1} and \autoref{fig:paxos2} contain the code for this Paxos protocol.
-A leader first broadcasts a ``p1a'' message (Phase 1, Type A) to the acceptors
-containing only the ballot number and then waits for $\texttt{N} - \texttt{F}$ ``p1b''
-responses from the acceptors.
+\autoref{fig:paxos1} and \autoref{fig:paxos2} contain the code for
+this Paxos protocol.  Paxos is perhaps best understood starting
+with the second phase.  At the end of the first phase the leader
+broadcasts a \texttt{2.A} message (Phase 2, Type A) to the acceptors
+containing the ballot number and a proposal and then waits for
+$\texttt{N} - \texttt{F}$ matching \texttt{2.B} responses from the
+acceptors.  If each response contains the ballot number and the
+proposal, then the proposal is deemed decided.  But one or more of
+the responses can contain a higher ballot number, in which case the
+leader has to try again with an even higher ballot number.
+
+It is not possible that an acceptor responds with a smaller ballot
+number.  This is because acceptors maintain two state variables.
+One is \textit{ballot}, the highest ballot number they have seen.
+Second is a variable called \textit{last\_accepted} that, if not
+\texttt{None}, contains the last proposal the acceptor has
+\emph{accepted} and the corresponding ballot number.  The acceptor
+also contains a set \emph{received} that contains (ballot, phase)
+tuples identifiying all rounds that the ballot has already participated
+in.  An acceptor waits for a message for a round that is not in
+\emph{received}.  If its ballot number is higher than what it has
+seen before, the acceptor moves into that ballot.  If the phase is
+2, then the acceptor accepts the proposal and remembers when it did
+so by saving the (ballot, proposal) tuple in \textit{last\_accepted}.
+In all cases, the acceptor responds with the current values of
+\textit{ballot} and \textit{last\_accepted}.
+
+In its first phase, a leader of a ballot must come up with a proposal that
+cannot conflict with a proposal of an earlier ballot that may already
+have been decided.  To this end, the leader broadcasts a \texttt{2.A}
+message to the acceptors and awaits their \textit{last\_accepted}
+values.  If all acceptors respond with \texttt{None}, then the leader
+is free to choose its own proposal.  Otherwise the leader updates
+its proposal with the one corresponding to the highest ballot number.
+The leader then moves on to the second round.
 
@@
 A context is itself composed over various Harmony values.
+
+\begin{comment}
 Some of these values can be extracted.  In particular, if $c$
@@
 \end{center}
+\end{comment}
 
 A thread can call \texttt{get\_context()} to retrieve its current
-context.  Also, the function \texttt{contexts()} returns a
+context.
+
+\begin{comment}
+Also, the function \texttt{contexts()} returns a
 bag of all contexts (including terminated ones).
@@
 and used for checking invariants.
+\end{comment}
 

### 2021-07-22 book/paper.tex
Added exercises to Bosco protocol

@@
 assume there are $\texttt{N} = 3\texttt{F} + 1$ processors.
-Each processor \emph{proposes} a value, which we assume here is from the
+Each processor \emph{proposes} a value, which we assume here to be from the
 set \{ 0, 1 \}.
-By the definition of consensus, we want the following two properties:
+By the usual definition of consensus, we want the following two properties:
 \begin{enumerate}
 \item \emph{Validity}: a processor can only decide a value that has been proposed;
-\item \emph{Agreement}: if two processors decide, they decide the same value.
+\item \emph{Agreement}: if two processors decide, then they decide the same value.
 \end{enumerate}
 The consensus problem is surprisingly tricky to solve in the face of processor
-failures.
+failures and without making assumptions about how long it takes to send
+and receive a message.
 
@@
 
-The $N$ processors go through a sequence of \emph{rounds} in which they wait for
+The $\mathtt{N}$ processors go through a sequence of \emph{rounds} in which they wait for
 $\texttt{N} - \texttt{F}$ messages, update their state based on the messages, and
@@
 The state of a processor consists
-of its current round number and proposal.  Therefore, messages consist of a tuple containing
+of its current round number and proposal.  Therefore, messages contain
 a round number and a proposal. To start things, each processor first
@@
 where $s$ is a set and $v$ is variable that is bound to an element of $s$.
-The important properties of this \textbf{select} statement are as follows:
+The properties of the \textbf{select} statement are as follows:
 \begin{itemize}
@@
 
+\section*{Exercises}
+\begin{problems}
+\item While Harmony checks that the Validity and Agreement properties
+are not violated, it does not check that the algorithm ever decides.
+Indeed, in unlucky executions the algorithm may never decide.  But
+how would you check that there exist execution in which the algorithm
+does decide?
+\item The algorithm as given works in the face of crash failures.
+A more challenging class to tolerate are \emph{arbitrary failures} in
+which up to \texttt{F} processors may send arbitrary messages, including
+conflicting messages to different peers (equivocation).
+The algorithm can tolerate those failures if you use $\mathtt{N} = 5\mathtt{F} - 1$ processors instead of $\mathtt{N} = 3\mathtt{F} - 1$.  Check that.
+\end{problems}
+
 \chapter{Paxos}
@@
 the responses can contain a higher ballot number, in which case the
-leader has to try again with an even higher ballot number.
+leader has to try again with an even higher ballot number.  This is
+where the first phase comes in.
 
@@
 have been decided.  To this end, the leader broadcasts a \texttt{2.A}
-message to the acceptors and awaits their \textit{last\_accepted}
-values.  If all acceptors respond with \texttt{None}, then the leader
+message to the acceptors and awaits $\mathit{N} - \mathit{F}$ of their
+\textit{last\_accepted}
+values.  If all those acceptors responded with \texttt{None}, then the leader
 is free to choose its own proposal.  Otherwise the leader updates

### 2021-07-22 book/paper.tex
Latest

@@
 Indeed, in unlucky executions the algorithm may never decide.  But
-how would you check that there exist execution in which the algorithm
+how would you check that there exist executions in which the algorithm
 does decide?
@@
 The algorithm can tolerate those failures if you use $\mathtt{N} = 5\mathtt{F} - 1$ processors instead of $\mathtt{N} = 3\mathtt{F} - 1$.  Check that.
+\item In 1983 Michael Ben-Or presented a randomized algorithm that can
+tolerate crash failures with
+just $\mathtt{N} = 2\mathtt{F} - 1$ processors~\cite{BenOr83}.
+Implement this algorithm.
 \end{problems}

### 2021-07-22 book/paper.tex
Latest

@@
 a list of Harmony values.
-For example, the address of variable \texttt{d.v[3]} consists of the
+For example, the address of variable $d.\mathtt{v}[3]$ consists of the
 list \texttt{.d}, \texttt{.v}, and 3.  The only way to construct an
-address in Harmony is using the \texttt{?} operator.
-\texttt{?d.v[3]} returns the address of variable \texttt{d.v[3]}.
+address in Harmony is using the $?$ operator.
+$?d.\mathtt{v}[3]$ returns the address of variable $d.\mathtt{v}[3]$.
 An address can be dereferenced to return the value of the variable.

### 2021-07-22 book/paper.tex
Latest

@@
 have been decided.  To this end, the leader broadcasts a \texttt{2.A}
-message to the acceptors and awaits $\mathit{N} - \mathit{F}$ of their
+message to the acceptors and awaits $\mathtt{N} - \mathtt{F}$ of their
 \textit{last\_accepted}
@@
 The leader then moves on to the second round.
+
+In the initialization, the \texttt{proposals} variable is initialized
+to either $[\mathtt{False}, \mathtt{False}]$ or
+to $[\mathtt{True}, \mathtt{False}]$ (in case $\texttt{NBALLOTS} = 2$).
+The other combinations of \texttt{True} and \texttt{False} are symmetric,
+so there is no need to run the model checker through those scenarios as
+well, halving the running time.
+
+\section*{Exercises}
+\begin{problems}
+\item Perhaps the trickiest detail of the algorithm is that, in Line~14
+of \autoref{fig:paxos2}, the leader selects the proposal with the highest
+ballot number it receives.  Change the \texttt{max} operator in that statement
+with \texttt{choose} and see if it finds a problem.  First try with
+$\texttt{NBALLOTS} = 2$ and then with $\texttt{NBALLOTS} = 3$.
+(Warning, the latter may take a long time.)  If it finds a problem,
+analyze the output and see what went wrong.
+\end{problems}
 

### 2021-07-22 book/paper.tex
Latest

@@
 
-Python is object-oriented, but Harmony is not.  This can lead to some
-unexpected differences.  For example, consider the following code:
+Python is object-oriented, but Harmony is not.  For Python programmers,
+this can lead to some unexpected differences.
+For example, consider the following code:
 
@@
 In Harmony, the two variables will be two different locks.  If you want two
-variables referring to the same lock in Harmony, you would write:
+variables referring to the same lock in Harmony, you might write:
 
@@
 \begin{verbatim}
-x = y = malloc(Lock());
+lock = Lock()
+x = y = ?lock
 \end{verbatim}
@@
 While in theory one method can be declared within another,
-they cannot share variables either.
+they cannot share local variables either.
 All other variables are global and must be initialized
@@
 
-There are three classes of precedence.
+There are three levels of precedence.
 Application has the highest precedence.  So, \texttt{!}\textit{a b} is interpreted as
@@
 3 to $z$;
-\item \texttt{$x$, ($y$, $z$) = 1, 2;} generates an runtime error because 2 cannot
+\item \texttt{$x$, ($y$, $z$) = 1, 2}: generates an runtime error because 2 cannot
 be matched with ($y$, $z$);
-\item \texttt{$x[0]$, $x[1]$ = $x[1]$, $x[0]$;} swaps the first two elements of list $x$.
+\item \texttt{$x[0]$, $x[1]$ = $x[1]$, $x[0]$}: swaps the first two elements of list $x$.
 \end{itemize}
@@
 
-Pattern matching can also be used in \texttt{const} and \texttt{let}
-statements.
+Pattern matching can also be used in \texttt{const}, \texttt{let},
+and \texttt{select} statements.
 
@@
 
-In case of dictionaries, you can access both keys and values:
+In case of dictionaries, you can access both keys and values (not
+supported in Python):
 \begin{code}
@@
 Haobin even contributed some code to the Harmony compiler.
+
+Kevin Sun and Anthony Yang built a beautiful VSCode extension for Harmony
+called HarmonyLang and proceeded to build an animator for Harmony
+executions and two cloud-based Harmony offerings, which you can can learn
+about at \url{http://harmony.cs.cornell.edu}.  They also developed much of
+that web site.  Later they were joined by Shi Chong Zhao, who also
+made significant contributions.
 
@@
 Finally, I would like to thank my family who had to suffer as I obsessed
-over writing the code and the book during the turbulent months of
+over writing the code and the book, at home, during the turbulent months of
 May and June 2020.

### 2021-07-23 book/paper.tex
Latest
Fixed evaluation order of applicattions

@@
 of having a condition variable per fork, having a condition variable per
-diner.  It is the same number of condition variables, but you will not
+diner.  It uses the same number of condition variables, but you will not
 need to have \textbf{if} statements nested inside the \textbf{while} loop
@@
 of \autoref{fig:paxos2}, the leader selects the proposal with the highest
-ballot number it receives.  Change the \texttt{max} operator in that statement
+ballot number it receives.  Replace the \texttt{max} operator in that statement
 with \texttt{choose} and see if it finds a problem.  First try with
@@
 \texttt{*}, \texttt{/}, \texttt{//}, \texttt{\%},
-\texttt{\string\&}, \texttt{|}, \texttt{\string\^},
-\texttt{and}, and \texttt{or}\\
+\texttt{\&}, \texttt{|}, \texttt{\^},
+\texttt{and}, \texttt{or}\\
 \hline
@@
 \hline
-\texttt{def m $a$: S} & \texttt{m} is an identifier, $a$ a bounded variable, \texttt{S} a list of statements\\
+\texttt{def m $a$: S} & \texttt{m} is an identifier, $a$ a bounded variable, \texttt{S} is a list of statements\\
 \hline
@@
 \hline
-\texttt{for $a$ in $e$: S} & $a$ is a bounded variable, $e$ is a set or dictionary,
-                            \texttt{S} a list of statements\\
+\texttt{for $a$ in $e$: S} & $a$ is a bounded variable, $e$ is a set or dictionary \\
 \hline
@@
 \hline
-\texttt{if $b$: S else: S} & $b$ is a boolean, \texttt{S} a list of statements\\
+\texttt{if $b$: S else: S} & $b$ is a boolean, \texttt{S} is a list of statements\\
 \hline
@@
 \hline
-\texttt{let $a$ = $e$ [let ...]: S} & $a$ is a bounded variable, $e$ is an expression, \texttt{S} a list of statements\\
+\texttt{let $a$ = $e$ [let ...]: S} & $a$ is a bounded variable, $e$ is an expression\\
 \hline
@@
 \texttt{spawn m $e$ [, $t$]} & \texttt{m} is a method,
-$e$ is an expression, $t$ is the initial thread-local state \\
+$e$ is an expression, $t$ is the thread-local state \\
 \hline

### 2021-07-23 src/charm/ops.c

@@ -538,8 +538,8 @@ void op_Address(const void *env, struct state *state, struct context **pctx){
 }
 
 void op_Apply(const void *env, struct state *state, struct context **pctx){
-    uint64_t method = ctx_pop(pctx);
     uint64_t e = ctx_pop(pctx);
+    uint64_t method = ctx_pop(pctx);
 
     uint64_t type = method & VALUE_MASK;
     switch (type) {

### 2021-07-24 book/paper.tex
Latest
Fixed some bugs

@@
 For example:
-\texttt{\{ $x$ + $y$ for $x$ in $s$ for $y$ in $s$ such that $(x * y) == 4$ \}}
+\texttt{\{ $x$ + $y$ for $x$ in $s$ for $y$ in $s$ where $(x * y) == 4$ \}}
 returns a set that is constructed by summing pairs of elements from $s$
@@
 
-Harmony currently supports the following statements:
+Harmony currently supports the following statements (below, \texttt{S} is a list of statements):
 
@@
 \hline
-\texttt{def m $a$: S} & \texttt{m} is an identifier, $a$ a bounded variable, \texttt{S} is a list of statements\\
+\texttt{def m $a$: S} & \texttt{m} is an identifier, $a$ a bounded variable\\
 \hline
@@
 \hline
-\texttt{for $a$ in $e$: S} & $a$ is a bounded variable, $e$ is a set or dictionary \\
+\texttt{for $a[:b]$ in $e$ [where $c$]: S} & $a$ and $b$ are bounded variables, $e$ is a set or dictionary \\
 \hline
@@
 \end{tabular}
+
+\section{Assignment}
+
+The statement $x = 3$ changes the state by assigning~3 to
+variable $x$ (assuming~$x$ was not already~3).
+$x$ may be a local variable or a shared variable.
+The statement $x = y = 3$ first updates~$y$, then~$x$.
+The statement $x[\mathtt{f}()] = y[\mathtt{g}()] = 3$ first computes
+the address of $x[\mathtt{f}()]$, then computes the address of
+$y[\mathtt{g}()]$, but first assigns to $y[\mathtt{g}()]$ and then
+to $x[\mathtt{f}()]$.
+The statement $a,b = c$ assumes that $c$ is a tuple with two values.
+It first evaluates the addresses of $a$ and $b$ and first assigns to
+the latter and then the former.
+If $c$ is not a tuple with two values, then Harmony will report an error.
+
+The statement $x~+$$= 3$ loads $x$, adds 3, and then stores the
+results in $x$.  In this case, it is equivalent to $x = x + 3$.
+However, in general this is not so.  For example, $x[\mathtt{f}()]~+$$= 3$
+only evaluates $\mathtt{f}()$ once.  Unlike Python, however,
+$x~+$$= [3]$ is equivalent to $x = x + [3]$ in Harmony.
+(In Python, the following two compound statements lead to different results
+for $y$: $x = y = []; x~+$$= [1]$ and $x = y = []; x = x + [1]$.)
+
+\section{\textbf{assert}}
+
+The statement \textbf{assert $b$} evaluates $b$ and reports an error
+if $b$ is false.  It should be considered a no-op---it is part of the
+\emph{specification}, not part of the \emph{implementation} of the algorithm.
+In particular, it specifies an invariant: whenever the program counter is
+at the location where the \textbf{assert} statement is, then $b$ is
+always true.
+
+If $b$ is an expression, then it is evaluated atomically.
+Moreover, the expression is not allowed to change the state.
+If it does change the state, Harmony will report an error as well.
+
+As in Python, you can specify an additional expression:
+\textbf{assert $b$, $e$}.  The value of $e$ will be reported as part
+of the error should $b$ evaluate to false.
+
+\section{\textbf{atomic}}
+
+The statement \textbf{atomic $S_1, S_2, ...$} evaluates statements
+$S_1, S_2, ...$ atomically in that no other threads can run while
+this atomic block is executing.  Typically an atomic block runs
+to completion before any other thread can run.  The only exception
+to this is if the atomic block executes a \textbf{stop} expression.
+In this case, another thread can run.  When the original thread
+is resumed (using a \textbf{go} statement), it is once again
+atomically executing.
+
+The \textbf{atomic} statement is useful for implementing synchronization
+primitives such as test-and-set.  It is also useful for testing.
+It is not a replacement for lock/unlock, and should not generally be used
+for synchronization otherwise.  Lock/unlock does allow other
+threads to run concurrently---just not in the same critical section.
+
+\section{\textbf{await}}
+
+The statement \textbf{await $b$} is equivalent to \textbf{while !$b$: pass}.
+It is intended to improve readability of your code.
+
+\section{\textbf{const}}
+
+The expression \textbf{const} \texttt{N = 3} introduces a new constant
+\texttt{N} with the value 3.  Evaluating \texttt{N} does not lead to
+loading from a memory location.  The assignment can be overridden with
+the \texttt{-c} flag: \texttt{harmony -cN=4} executes the model checker
+with 4 assigned to \texttt{N} instead of 3.
+Harmony also supports \textbf{const} \texttt{N, M = 3, 4}, which
+assigns 3 to \texttt{N} and 4 to \texttt{M}.
 
@@
 Python iterates over the keys rather than the values;
-\item If you want to iterate over the keys of a dictionary $d$, use
-\texttt{for} $k$ \texttt{in} \texttt{keys} $d$;
+\item If you want to iterate over the keys of a dictionary $d$, use either
+\begin{itemize}
+\item \texttt{for} $k$ \texttt{in} \texttt{keys} $d$; or
+\item \texttt{for} $k:\_$ \texttt{in} $d$
+\end{itemize}
 \item The \texttt{list} module (\autoref{ap:list}) provides methods \texttt{values}(),

### 2021-07-24 book/paper.tex
Working on an explanation of every type of statement

@@
 
-\chapter{Value Types}
+\chapter{Value Types and Operators}
 \label{ap:values}
@@
 \hline
+$e$ & $e$ is an expression\\
+\hline
 \textit{lv} = [lv =]... $e$ & \textit{lv} is an lvalue and $e$ is an expression\\
@@
 \hline
+\texttt{invariant e} & \texttt{e} is an invariant\\
+\hline
 \texttt{let $a$ = $e$ [let ...]: S} & $a$ is a bounded variable, $e$ is an expression\\
@@
 \end{tabular}
+
+\section{Single expression evaluation}
+
+Any expression by itself can be used as a statement.  The most common
+form of this is a function application, for example: \texttt{f}().
+This statement evaluates \texttt{f}() but ignores its result.
+It is equivalent to the assignment statement $\_ = \mathtt{f}()$.
 
@@
 The statement $x = y = 3$ first updates~$y$, then~$x$.
-The statement $x[\mathtt{f}()] = y[\mathtt{g}()] = 3$ first computes
+The statement $x[\mathtt{f}()] = y[\mathtt{g}()] = \mathtt{h}()$ first computes
 the address of $x[\mathtt{f}()]$, then computes the address of
-$y[\mathtt{g}()]$, but first assigns to $y[\mathtt{g}()]$ and then
-to $x[\mathtt{f}()]$.
+$y[\mathtt{g}()]$, then evaluates $\mathtt{h}()$, then assigns the resulting value to $y[\mathtt{g}()]$ (using its previously computed address), and finally assigns the same value to $x[\mathtt{f}()]$ (again using its previously computed address).
 The statement $a,b = c$ assumes that $c$ is a tuple with two values.
@@
 If $c$ is not a tuple with two values, then Harmony will report an error.
+
+Assigning to $\_$ evaluates the righthand side expression but is otherwise
+a no-op.
 
@@
 
-The statement \textbf{atomic $S_1, S_2, ...$} evaluates statements
+The statement \textbf{atomic $S_1; S_2; ...$} evaluates statements
 $S_1, S_2, ...$ atomically in that no other threads can run while
@@
 assigns 3 to \texttt{N} and 4 to \texttt{M}.
+
+\section{\textbf{def}}
+
+The statement \textbf{def} \texttt{m} $a: S_1; S_2: ...$
+defines a new program counter constant \texttt{m} referring to
+a method that takes an argument $a$ and executes the statements
+$S_1, S_2, ...$.  The argument $a$ can be a tuple pattern similar
+to those used in \texttt{let} and \texttt{for} statements.
+Examples include $()$, $(x,)$, $(x, y)$, and $(x, (y, z))$.
+The given local variable names variable names are assigned
+upon application.
+It is allowed, but discouraged, to updates those local variables
+in statements $S_1, S_2, ...$.
+Each method has a predefined local variable
+\textit{result}, initialized to \texttt{None},
+that is returned by the method.  Harmony does
+not support a \texttt{return} statement that \emph{breaks out}
+of the code before executing the last statement.
+
+\section{\textbf{del}}
+
+The statement \textbf{del} $x$ removes variable $x$ from the
+state.  $x$ can be either a local or a shared variable.  For
+example, the statement \textbf{del} $x.\texttt{age}$ removes
+the \texttt{.age} field from dictionary $x$.
+Harmony automatically removes top-level local variables that
+are no longer in use
+from the state in order to attempt to reduce the number of states
+that are evaluated during model checking.
+
+Because Harmony lists are dictionaries, deleting from lists is different
+from Python:
+$x = [\mathtt{.a}, \mathtt{.b}]; \mathbf{del}~x[0]$ results in $x$ having value
+$\{1: \mathtt{.b}\}$ rather than $[\mathtt{.b}]$ (which is $\{0: \mathtt{b}\}$).
+
+\section{\textbf{for ... in [...where]}}
+
+The statement \textbf{for} $x$ \textbf{in} $y: S_1; S_2; ...$
+iterates over $y$ and executes for each element the statements
+$S_1, S_2, ...$.
+which must be a set or a dictionary.  $y$ is evaluated only once at
+the beginning of the evaluation of this statement.  In case of a set,
+the result is sorted (using Harmony's global order on all values).
+For each element, the statements $S_1, S_2, ...$ are executed with
+local variable $y$ having the value of the element.
+$x$ can be a pattern such as $(a)$ or $(a, (b, c))$.
+If the pattern cannot be matched, Harmony detects and error.
+It is allowed, but discouraged, to assign different values to $x$
+within statements $S_1, S2, ...$.
+
+The statement also supports nesting and filtering.
+Nesting is of the form
+\textbf{for} $x_1$ \textbf{in} $y_1$ \textbf{for} $x_2$ \textbf{in} $y_2$:
+$S_1; S_2; ...$,
+which is equivalent to the statement
+\textbf{for} $x_1$ \textbf{in} $y_1$: \textbf{for} $x_2$ \textbf{in} $y_2$:
+$S_1; S_2; ...$.
+Filtering is of the form
+\textbf{for} $x$ \textbf{in} $y$ \textbf{where} $z: S_1; S_2; ...$.
+For example,
+\textbf{for} $x$ \textbf{in} \{ 1 .. 10 \} \textbf{where} $(x \% 2) == 0:
+S_1; S_2; ...$ only evaluates statements $S_1, S_2, ...$ for even $x$,
+that is, 2, 4, 6, 8, and 10.
+
+\section{\textbf{from ... import}}
+
+\section{\textbf{go}}
+
+\section{\textbf{if ... elif ... else}}
+
+\section{\textbf{import}}
+
+\section{\textbf{invariant}}
+
+\section{\textbf{let}}
+
+\section{\textbf{pass}}
+
+\section{\textbf{select ... in}}
+
+\section{\textbf{sequential}}
+
+\section{\textbf{spawn}}
+
+\section{\textbf{trap}}
+
+\section{\textbf{while}}
 

### 2021-07-24 book/paper.tex
Added explanation for each of the statements

@@
 assigns 3 to \texttt{N} and 4 to \texttt{M}.
+Harmony also has limited support for \emph{constant folding}.  For example,
+\textbf{const} \texttt{N = 3 + 4} assigns value 7 to constant \texttt{N}.
 
@@
 
+Harmony does not support \textbf{break} or \textbf{continue} statements.
+
 \section{\textbf{from ... import}}
 
+The statement \textbf{from} \texttt{x} \textbf{import} \texttt{a, b, ...}
+imports module \texttt{x} and makes its constants $a, b, ...$ also constants
+in the current module.  If a module is imported more than once, its code
+is only included the first time.  The constants will typically be the
+names of methods (program counter constants) within the module.
+
+You can import all constants from a module \texttt{m}
+(including program counter constants) using the statement
+\textbf{from} \texttt{m} \textbf{import} \texttt{*}.
+This, however, excludes constants that start with the character \_: those
+are considered \emph{private} to the module.
+
 \section{\textbf{go}}
 
-\section{\textbf{if ... elif ... else}}
+The statement \textbf{go} $c e$ starts a thread with context $c$ that
+has executed a \textbf{stop} expression.  The \textbf{stop} expression
+returns value $e$.
+The same context can be started multiple times, allowing threads to
+\emph{fork}.
+
+\section{\textbf{if ... [elif ...]* [else]}}
+
+Harmony supports \textbf{if} statements.
+In its most basic form, \textbf{if} $c: S_1; S_2; ...$ evaluates $c$
+and executes statements $S_1, S_2, ...$ if and only if boolean expression
+$c$ evaluated to true.
+Harmony checks that $c$ is either \texttt{False} or \texttt{True}---if neither
+is the case, Harmony reports an error.
+The statement \textbf{if} $c: S_1, S_2, ...$ \textbf{else}:
+$T_1; T_2; ...$ is similar, but executes statements $T_1, T_2, ...$
+if and only if $c$ evaluated to false.
+You can think of \textbf{elif} $c:$ as shorthand for
+\textbf{else}: \textbf{if} $c:$.
 
@@
 
+The statement \textbf{import} $\mathtt{m}_1, \mathtt{m}_2, ...$ imports modules
+$\mathtt{m}_1, \mathtt{m}_2, ...$ in that order.
+If a module is imported more than once,
+its code is only included the first time.
+The constants (including method constants) and shared variables declared 
+in that module can subsequently be referenced by prepending ``\texttt{m}.''.
+For example, method \texttt{f()} in imported module \texttt{m} is invoked
+by calling \texttt{m.f()}.  If you would prefer to invoke it simply as
+\texttt{f()}, then you have to import using the statement
+\textbf{from} \texttt{m} \textbf{import} \texttt{f}.
+
 \section{\textbf{invariant}}
 
+The statement \textbf{invariant} $c$ declares that boolean expression $c$
+is an invariant.  $c$ is only allowed to read shared variables and
+is evaluated atomically after every state change.
+If it evaluated to \texttt{False} Harmony reports and error.  Harmony
+also reports an error if the expression evaluates to a value other
+than \texttt{False} or \texttt{True}.
+
 \section{\textbf{let}}
 
+You can introduce new local variables in a method using the
+\texttt{let} expression.
+The statement \textbf{let} $a = b: S_1; S_2, ...$ evaluates $b$, assigns
+the result to local variable $a$, and evaluates statements
+$S_1, S_2, ...$.  \textbf{let} supports pattern matching, so you can
+write \texttt{let} $x, (y, z) = b: S_1; S_2, ...$.  This will only work
+if $b$ is a tuple with two elements, the second of which also being a
+tuple with two elements---if not, Harmony will report an error.
+The variables may be updates in statements $S_1, S_2, ...$.
+
+\textbf{let} statements may be nested, such as
+\textbf{let} $a_1 = b_1$ \textbf{let} $a_2 = b_2: S_1; S_2; ...$.
+Doing so can improve readability by reducing indentation
+compared to writing them as separate statements.
+Compare the following two examples:
+
+\begin{code}
+\begin{tabbing}
+XXX\=XXX\=XXX\=XXX\kill
+\>\textbf{let} $a = y$: \\
+\>\>\textbf{let} $b = z$: \\
+\>\>\>...
+\end{tabbing}
+\end{code}
+\begin{code}
+\begin{tabbing}
+XXX\=XXX\=XXX\=XXX\kill
+\>\textbf{let} $a = y$ \\
+\>\textbf{let} $b = z$: \\
+\>\>...
+\end{tabbing}
+\end{code}
+
 \section{\textbf{pass}}
 
+The \textbf{pass} statement does nothing.
+
 \section{\textbf{select ... in}}
 
+The statement \textbf{select} $x$ \textbf{in} $y: S_1; S_2; ...$
+requires that $y$ evaluates to a set value.
+The statement does the following three things atomically:
+\begin{itemize}
+\item it waits until $y$ is non-empty;
+\item it selects one element of $y$ non-deterministically (using
+a \texttt{choose} expression);
+\item it executes statements $S_1, S_2, ...$ with the selected element
+assigned to local variable $x$.
+\end{itemize}
+$x$ may be a pattern, like in \texttt{let}, \texttt{for}, and \texttt{def}
+statements.
+Harmony reports an error if $y$ evaluates to a value that is not a set.
+If \textit{waiting} is an unused local variable,
+then \textbf{select} $x$ \textbf{in} $y: S_1; S_2; ...$ is equivalent to
+
+\begin{code}
+\begin{tabbing}
+XXX\=XXX\=XXX\=XXX\=XXX\=XXX\=XXX\=XXX\kill
+\>\textbf{let} $\mathit{waiting} = \mathtt{True}:$ \\
+\>\>\textbf{while} \textit{waiting}: \\
+\>\>\>\textbf{atomic}: \\
+\>\>\>\>\textbf{if} $y$ != \{\}:\\
+\>\>\>\>\>\textbf{let} $x = \mathbf{choose}(y)$:\\
+\>\>\>\>\>\>$S_1$ \\
+\>\>\>\>\>\>$S_2$ \\
+\>\>\>\>\>\>$...$ \\
+\>\>\>\>\>$\mathit{waiting} = \mathtt{False}$
+\end{tabbing}
+\end{code}
+
+The statement is particularly useful in programming network protocols
+when having to wait for one or more messages and executing a set
+of actions atomically after the desired messages have arrived.
+
 \section{\textbf{sequential}}
 
+In Harmony, shared variable \texttt{Load} and \texttt{Store} operations
+are atomic and have \emph{sequential consistency}.
+However, Harmony does check for \emph{data races}.  A data race occurs
+when two or more threads simultaneously access the same shared variable,
+with at least one of the accesses being a \texttt{Store} operation
+\emph{outside of an atomic block}.
+If so, Harmony will report an error.  This error can be suppressed by
+declaring the shared variable as sequential.  In particular, the statement
+\texttt{sequential} $x, y, ...$ specifies that the algorithm assumes
+that the given variables have sequential consistency.
+
+Note that few modern processors support sequentially consistent memory
+by default, as doing so would lead to high overhead.
+
 \section{\textbf{spawn}}
 
+The statement \texttt{spawn} $m$ $a$ starts a new thread that executes
+method $m$ with argument $a$. $m$ must be a program counter constant,
+and $a$ is typically a tuple containing
+zero or more parameters to be passed to the method.
+
+The default thread-local state of the thread, called \textit{self},
+is the empty dictionary by default.  It can be specified by adding
+a parameter: \texttt{spawn} $m$ $a, e$ specifies that $e$ should be
+the initial value of the thread-local state.
+
 \section{\textbf{trap}}
 
+The statement \textbf{trap} $m$ $a$ specifies that the current thread
+should execute method $m$ with argument $a$ and some future, unspecified,
+time.  It models a timer interrupt or any kind of asynchronous event to
+be handled by the thread.  Such interrupts can be disabled by setting
+the interrupt level of the thread to \texttt{True} using the
+\textbf{setintlevel} operator.
+
 \section{\textbf{while}}
+
+The statement \textbf{while} $c: S_1; S_2; ...$ executes statements
+$S_1, S_2, ...$ repeatedly as long as $c$ evaluates to \texttt{True}.
+Harmony does not support \textbf{break} or \textbf{continue} statements.
 

### 2021-07-24 book/paper.tex
Adding leader election algorithm
Added leader election to the book

@@
 network.
+\item \autoref{ch:leader} presents a protocol for electing a leader on
+a ring of processors, where each processor is uniquely identified and 
+only knows its successor on the ring.
 \item \autoref{ch:consensus} demonstrates a fault-tolerant distributed consensus
@@
 
+\chapter{Leader Election}
+\label{ch:leader}
+
+\begin{figure}
+\begin{code}
+\harmonysource{leader}
+\end{code}
+\caption{\harmonylink{code/leader.hny} A leader election protocol on a ring.}
+\label{fig:leader}
+\end{figure}
+
+Leader election is the problem of electing a unique leader in a network
+of processors.  Typically the processors have only limited information.
+In the version that we present, each processor has a unique identifier.  The
+processors are organized in a ring, but each processor only knows its
+own identifier and the identifier of its successor.  We assume that
+each processor can reliably send messages to its successor.
+
+The protocol that we present elects as leader the processor with the
+highest identifier and works in two phases: in the first
+phase, each processor sends its identifier to its successor.  When
+a processor receives an identifier that is larger than its own
+identifier, it forward that to its successor as well.  If a processor
+receives its own identifier, it knows that it is the leader.  That
+processor then starts the second phase by sending a message around the
+ring notifying the other processors of the leader's identifier.
+
+\autoref{fig:leader} describes the protocol and its test cases in Harmony.
+To test the protocol, the code first generates an identifier for each
+processor, chosen non-deterministically from a set of \texttt{NIDS}
+identifiers.  It also keeps track in the variable \textit{leader} of
+what the highest identifier is so it can later be checked.
+
+The network is modeled simply as a set of messages.
+The \texttt{send} method atomically adds a message to this set.
+Messages are tuples with three fields:
+(\textit{dst}, \textit{id}, \textit{found}).
+\textit{dst} is the identifier of the destination processor;
+\textit{id} is the identifier that is being forwarded; and
+\textit{found} is a boolean indicating the second phase of the protocol.
+The \texttt{receive}(\textit{self}) method looks for all messages
+destined for the processor with identifier \textit{self}.
+
+Method \texttt{processor}{\textit{self}, \textit{succ}) is the
+code for a processor with identifier \textit{self} and successor
+\textit{succ}.
+It starts simply by sending its own identifier to its successor.
+The processor then loops until it discovers the identifier of the
+leader in the second phase of the protocol.
+A processor waits for a message using the Harmony \textbf{select} statement.
+This statement takes the form \textbf{select} $v$ \textbf{in} $s$: \emph{statement block}
+where $s$ is a set and $v$ is variable that is bound to an element of $s$.
+The properties of the \textbf{select} statement are as follows:
+\begin{itemize}
+\item it waits until $s$ is non-empty;
+\item it is executed atomically;
+\item $v$ is selected non-deterministically, like in the \texttt{choose} operator.
+\end{itemize}
+\index{select statement}%
+
+If a processor receives its own identifier, it knows its the leader.
+The Harmony code checks this using an assertion.
+In real code the processor could not do this as it does not know
+the identifier of the leader, but assertions are only there to check
+correctness.
+The processor then sends a message to its successor that the leader
+has been found.
+If the processor receives an identifier higher than its own, the
+processor knows that it cannot be the leader.
+In that case, it simply forwards the message.
+A processor stops when it receives a message that indicates that
+the leader has been identified.
+
+Note that there is a lot of non-determinism in the specification, leading
+to a lot of executions that must be checked.
+First, every possible permutation of identifiers for the processors
+is tried.  When there are multiple messages to receive by a processor,
+every possible order is tried (including receiving the same message
+multiple times).  Fortunately, the \texttt{select} statement is executed
+atomically, otherwise the body of the statement could lead to additional
+thread interleavings.  Because in practice the different processors do
+not share memory, it is not necessary to check those interleavings.
+
 \chapter{Distributed Consensus}
@@
 
-\index{select statement}%
 In Line~21, a processor waits for $\texttt{N} - \texttt{F}$ messages using
 the Harmony \textbf{select} statement.
-This statement takes the form \textbf{select} $v$ \textbf{in} $s$: \emph{statement block}
-where $s$ is a set and $v$ is variable that is bound to an element of $s$.
-The properties of the \textbf{select} statement are as follows:
-\begin{itemize}
-\item it waits until $s$ is non-empty;
-\item it is executed atomically;
-\item $v$ is selected non-deterministically, like in the \texttt{choose} operator.
-\end{itemize}
-
 Since Harmony has to check all possible executions of the protocol, the

### 2021-07-24 book/paper.tex
Made some fixes

@@
 The protocol that we present elects as leader the processor with the
-highest identifier and works in two phases: in the first
+highest identifier~\cite{CR79} and works in two phases: in the first
 phase, each processor sends its identifier to its successor.  When
@@
 
-Method \texttt{processor}{\textit{self}, \textit{succ}) is the
+Method \texttt{processor}(\textit{self}, \textit{succ}) is the
 code for a processor with identifier \textit{self} and successor

### 2021-07-24 book/paper.tex
Latest

@@
 processors are organized in a ring, but each processor only knows its
-own identifier and the identifier of its successor.  We assume that
-each processor can reliably send messages to its successor.
+own identifier and the identifier of its successor.  Having already
+looked into the problem of how to make the network reliable,
+we assume here that each processor can reliably send messages to its
+successor.
 
@@
 Distributed consensus is the problem of having a collection of processors agree
-on a single value over a network.  This often has to be done in adverse circumstances,
+on a single value over a network.
+Without failures, this can be solved using leader election: first elect a leader, then have that leader decide a value.
+But consensus often has to be done in adverse circumstances,
 for example in the face of processor failures.

### 2021-07-25 book/paper.tex
Latest

@@
 
-Leader election is the problem of electing a unique leader in a network
-of processors.  Typically the processors have only limited information.
-In the version that we present, each processor has a unique identifier.  The
-processors are organized in a ring, but each processor only knows its
-own identifier and the identifier of its successor.  Having already
-looked into the problem of how to make the network reliable,
-we assume here that each processor can reliably send messages to its
-successor.
+Leader election is the problem of electing a unique leader in a
+network of processors.  Typically this is challenging because the
+processors have only limited information.  In the version that we
+present, each processor has a unique identifier.  The processors
+are organized in a ring, but each processor only knows its own
+identifier and the identifier of its successor on the ring.  Having
+already looked into the problem of how to make the network reliable,
+we assume here that each processor can reliably send messages to
+its successor.
 
@@
 identifier, it forward that to its successor as well.  If a processor
-receives its own identifier, it knows that it is the leader.  That
+receives its own identifier, it discovers that it is the leader.  That
 processor then starts the second phase by sending a message around the
@@
 
+\section*{Exercises}
+\begin{problems}
+\item Check if the code finds a unique leader if identifiers are not unique.
+\end{problems}
+
 \chapter{Distributed Consensus}
@@
 The algorithm can tolerate those failures if you use $\mathtt{N} = 5\mathtt{F} - 1$ processors instead of $\mathtt{N} = 3\mathtt{F} - 1$.  Check that.
-\item In 1983 Michael Ben-Or presented a randomized algorithm that can
+\item In 1983, Michael Ben-Or presented a randomized algorithm that can
 tolerate crash failures with

### 2021-07-25 book/paper.tex
Latest

@@
 Harmony is heavily influenced by Leslie Lamport's work on
-TLA+, TLC, and PlusCal~\cite{Lamport02, Lamport09}, as well as
-Gerard Holzmann's work on Promela and SPIN~\cite{SPIN}.
+TLA+, TLC, and PlusCal~\cite{Lamport02, Lamport09},
+Gerard Holzmann's work on Promela and SPIN~\cite{SPIN},
+and University of Washington's DSLabs system~\cite{MWA19}.
 Harmony is designed to have a lower learning curve than those
@@
 \item Check if the code finds a unique leader if identifiers are not unique.
+\item Messages are added atomically to the network.  Is this necessary?
+What happens if you remove the \textbf{atomic} keyword?  Explain what happens.
 \end{problems}
@@
 only evaluates $\mathtt{f}()$ once.  Unlike Python, however,
-$x~+$$= [3]$ is equivalent to $x = x + [3]$ in Harmony.
+$x~+$$= [3,]$ is equivalent to $x = x + [3,]$ in Harmony.
 (In Python, the following two compound statements lead to different results
-for $y$: $x = y = []; x~+$$= [1]$ and $x = y = []; x = x + [1]$.)
+for $y$: $x = y = []; x~+$$= [3]$ and $x = y = []; x = x + [3]$.)
 
@@
 assigns 3 to \texttt{N} and 4 to \texttt{M}.
-Harmony also has limited support for \emph{constant folding}.  For example,
+Harmony has limited support for \emph{constant folding}.  For example,
 \textbf{const} \texttt{N = 3 + 4} assigns value 7 to constant \texttt{N}.
@@
 $S_1, S_2, ...$.
-which must be a set or a dictionary.  $y$ is evaluated only once at
+$y$ must be a set or a dictionary.  $y$ is evaluated only once at
 the beginning of the evaluation of this statement.  In case of a set,
 the result is sorted (using Harmony's global order on all values).
+In case of a dictionary, the statement iterates over the values of
+the dictionary in the order of the keys.  This makes iterating over
+lists intuitive and identical to Python.
 For each element, the statements $S_1, S_2, ...$ are executed with
@@
 within statements $S_1, S2, ...$.
+
+If $y$ is a dictionary, Harmony also supports the form
+\textbf{for} $k:v$ \textbf{in} $y: S_1; S_2; ...$.  This works similar,
+except that $k$ is bound to the key and $v$ is bound to the value.
 
@@
 \textbf{from} \texttt{m} \textbf{import} \texttt{*}.
-This, however, excludes constants that start with the character \_: those
-are considered \emph{private} to the module.
+This, however, excludes constants whose names start with the character
+\texttt{\_}: those are considered \emph{private} to the module.
 
@@
 
-The statement \textbf{go} $c e$ starts a thread with context $c$ that
+The statement \textbf{go} $c$ $e$ starts a thread with context $c$ that
 has executed a \textbf{stop} expression.  The \textbf{stop} expression
@@
 is evaluated atomically after every state change.
-If it evaluated to \texttt{False} Harmony reports and error.  Harmony
+If it ever evaluates to \texttt{False} Harmony reports and error.  Harmony
 also reports an error if the expression evaluates to a value other
@@
 \begin{verbatim}
-x = y = Lock();
+x = y = Lock()
 \end{verbatim}
@@
 Each (non-reserved) identifier in a Harmony program refers to either
-a constant, a global variable, or a local variable.
+a constant, a global shared variable, a local variable, or a module.
 Constants are declared using \texttt{const} statements.
@@
 \texttt{def} statements (i.e., arguments),
-\texttt{let} statements, and in \texttt{for} loops.
+\texttt{let} statements, \texttt{for} loops, and \texttt{select} statements.
 Also, each method has an implicitly declared
 \textit{result} variable, which is initialized to \texttt{None}.
+Each thread has a variable called \textit{this} that contains
+the thread-local state.
 Local variables are tightly scoped and cannot be shared
@@
 
+\begin{comment}
 \section{For Loops and Comprehensions}
@@
 the keyword \texttt{where} instead of \texttt{if}.
+\end{comment}
 

### 2021-07-26 book/paper.tex
Added eternal option to spawn Implemented 2pc

@@
 separate server thread.
+The server never terminates, so it is spawned with the keyword
+\textbf{eternal} to suppress non-terminating state warnings.
 Each client sends a request to the server, consisting in this case
@@
 \hline
-\texttt{spawn m $e$ [, $t$]} & \texttt{m} is a method,
+\texttt{spawn [eternal] m $e$ [, $t$]} & \texttt{m} is a method,
 $e$ is an expression, $t$ is the thread-local state \\
@@
 
+Harmony normally checks that all threads eventually terminate.
+If a thread may never terminate, you should spawn it with
+\texttt{spawn eternal} $m$ $a$ to suppress those checks.
+
 \section{\textbf{trap}}
@@
 \hline
-Spawn & pop initial thread-local state, argument, and method and spawn a new context \\
+Spawn [eternal] & pop initial thread-local state, argument, and method and spawn a new context \\
 \hline

### 2021-07-26 src/charm/charm.c

@@ -1271,6 +1271,24 @@ static int fail_cmp(void *f1, void *f2){
     return node_cmp(fail1->node, fail2->node);
 }
 
+bool all_eternal(uint64_t ctxbag){
+    int size;
+    uint64_t *vals = value_get(ctxbag, &size);
+    size /= sizeof(uint64_t);
+    bool all = true;
+    for (int i = 0; i < size; i += 2) {
+        assert((vals[i] & VALUE_MASK) == VALUE_CONTEXT);
+        assert((vals[i + 1] & VALUE_MASK) == VALUE_INT);
+        struct context *ctx = value_get(vals[i], NULL);
+        assert(ctx != NULL);
+        if (!ctx->eternal) {
+            all = false;
+            break;
+        }
+    }
+    return all;
+}
+
 void usage(char *prog){
     fprintf(stderr, "Usage: %s [-c] [-t maxtime] file.json\n", prog);
     exit(1);
@@ -1526,7 +1544,8 @@ int main(int argc, char **argv){
             if (comp->good) {
                 continue;
             }
-            if (node->state->ctxbag == VALUE_DICT && node->state->stopbag == VALUE_DICT) {
+            // TODO.  In case of ctxbag, all contexts should probably be blocked
+            if (all_eternal(node->state->ctxbag) && all_eternal(node->state->stopbag)) {
                 comp->good = true;
                 continue;
             }

### 2021-07-26 src/charm/ops.c

@@ -1217,6 +1217,10 @@ uint64_t bag_add(uint64_t bag, uint64_t v){
 
 void op_Spawn(const void *env, struct state *state, struct context **pctx){
     extern int code_len;
+    const struct env_Spawn *se = env;
+
+    uint64_t thisval = ctx_pop(pctx);
+    uint64_t arg = ctx_pop(pctx);
 
     uint64_t pc = ctx_pop(pctx);
     if ((pc & VALUE_MASK) != VALUE_PC) {
@@ -1227,8 +1231,6 @@ void op_Spawn(const void *env, struct state *state, struct context **pctx){
 
     assert(pc < code_len);
     assert(strcmp(code[pc].oi->name, "Frame") == 0);
-    uint64_t arg = ctx_pop(pctx);
-    uint64_t thisval = ctx_pop(pctx);
 
     struct context *ctx = new_alloc(struct context);
 
@@ -1240,6 +1242,7 @@ void op_Spawn(const void *env, struct state *state, struct context **pctx){
     ctx->pc = pc;
     ctx->vars = VALUE_DICT;
     ctx->interruptlevel = VALUE_FALSE;
+    ctx->eternal = se->eternal;
     ctx_push(&ctx, (CALLTYPE_PROCESS << VALUE_BITS) | VALUE_INT);
     ctx_push(&ctx, arg);
     uint64_t v = value_put_context(ctx);
@@ -1503,7 +1506,6 @@ void *init_ReadonlyInc(struct dict *map){ return NULL; }
 void *init_Return(struct dict *map){ return NULL; }
 void *init_Sequential(struct dict *map){ return NULL; }
 void *init_SetIntLevel(struct dict *map){ return NULL; }
-void *init_Spawn(struct dict *map){ return NULL; }
 void *init_Trap(struct dict *map){ return NULL; }
 
 void *init_Cut(struct dict *map){
@@ -1698,6 +1700,25 @@ void *init_Push(struct dict *map){
     assert(jv->type == JV_MAP);
     struct env_Push *env = new_alloc(struct env_Push);
     env->value = value_from_json(jv->u.map);
+    return env;
+}
+
+void *init_Spawn(struct dict *map){
+    struct env_Spawn *env = new_alloc(struct env_Spawn);
+    struct json_value *eternal = dict_lookup(map, "eternal", 7);
+    if (eternal == NULL) {
+        env->eternal = false;
+    }
+    else {
+		assert(eternal->type == JV_ATOM);
[…]

### 2021-07-30 book/paper.tex
Added code for 2pc
Latest

@@
 only knows its successor on the ring.
+\item \autoref{ch:2pc} describes the two-phase commit protocol used to implement
+atomic database transactions.
 \item \autoref{ch:consensus} demonstrates a fault-tolerant distributed consensus
@@
 
+\chapter{Two Phase Commit}
+\label{ch:2pc}
+
+\begin{figure}
+\begin{code}
+\harmonysource{2pc1}
+\end{code}
+\caption{\harmonylink{code/2pc.hny} Two Phase Commit protocol: code for banks.}
+\label{fig:2pc1}
+\end{figure}
+
+\begin{figure}
+\begin{code}
+\harmonysource{2pc2}
+\end{code}
+\caption{\harmonylink{code/2pc.hny} Two Phase Commit protocol: code for transaction coordinators.}
+\label{fig:2pc2}
+\end{figure}
+
+Modern databases support multiple clients concurrently accessing the data
+stored in them.  Making matters more complicated, they store data on disk, but
+we will ignore this in this book.
+The complication we address here is that databases may be \emph{sharded},
+where different parts of the data are stored on different servers with
+different disks.  The different servers may even be under different
+authoritive domains, such as multiple banks maintaining the accounts of their
+clients.
+
+In database terminology, a \emph{transaction} is an operation on the database.
+The operation can be quite complex.  A transaction should be executed with at
+least the following two properties:
+\begin{itemize}
+\item \emph{all-or-nothing}: a transaction should either complete, or it should
+be a no-op.  It should never partially execute and then give up because of
+some kind of error or something.  Database
+people call this \emph{atomicity}, but it is not the same kind of atomicity that
+we have been discussing in this book.
+\item \emph{serialized}: any two concurrent transactions should appear to
+execute in some order.  Database people call this \emph{isolation}: one transaction
+should not be able to witness the intermediate state of another transaction
+in execution.
+\end{itemize}
+
+We will use as an example a distributed database that maintains accounts.
+For simplicity we will assume a collection of banks, each maintaining a
+single account.  There are two kinds of transactions: \emph{transfer}
+(similar to \autoref{ex:bank}) and \emph{inquiry}.  In this example, a transfer
+is a transaction that moves some balance between two accounts.
+An inquiry is a transaction over all accounts and checks that the sum of the
+balances across the accounts remains the same.
+
+Executing such transactions must be done with care.  Consider what
+would happen if transactions are not all-or-nothing or are not
+serialized.  A transfer consists of two operations: withdrawing
+funds from one account and depositing the same amount of funds in
+the other.  These two operations can be done concurrently, but if
+the withdrawal fails (for example, because there are not sufficient
+funds in the source account) then the whole transaction should fail
+and become a no-op.  Even if this is not the case, a concurrent
+inquiry transaction may accidentally witness a state in which either
+the withdrawal or the deposit happened, but not both.  And matters
+get more complicated with multiple concurrent transfers.
+
+The Two-Phase Commit protocol~\cite{Gray78} is a protocol that can be
+used to implement transactions across multiple database servers---banks
+in this case.
+Each transaction has a \emph{coordinator} that sends a \texttt{PREPARE}
+message to each of the servers involved in the transaction, asking them
+to prepare to commit to a transaction.  A server can either respond with
+\texttt{YES} if it is ready to commit and will avoid doing anything
+that might jeopardize this (like committing a conflicting transaction),
+or \texttt{NO} if it does not want to participate in the transaction.
+If all servers respond with \texttt{YES}, then the coordinator can
+decide to \emph{commit} the transaction.  Otherwise the coordinator
+must decide to \emph{abort} the transaction.  In the second phase, the
+servers that responded with \texttt{YES} (if any) must be notified in
+either case to inform them of the decision.
+
 \chapter{Distributed Consensus}

### 2021-07-30 book/paper.tex
Latest

@@
 stored in them.  Making matters more complicated, they store data on disk, but
-we will ignore this in this book.
+we will ignore that in this book.
 The complication we address here is that databases may be \emph{sharded},
-where different parts of the data are stored on different servers with
-different disks.  The different servers may even be under different
+where different parts of the data are stored on different servers.
+The different servers may even be under different
 authoritive domains, such as multiple banks maintaining the accounts of their
@@
 
-In database terminology, a \emph{transaction} is an operation on the database.
-The operation can be quite complex.  A transaction should be executed with at
-least the following two properties:
+In database terminology, a \emph{transaction} is an operation on a database.
+The operation can be quite complex, and the execution of a transaction
+should have at least the following two properties:
 \begin{itemize}
@@
 We will use as an example a distributed database that maintains accounts.
-For simplicity we will assume a collection of banks, each maintaining a
+For simplicity we will model this as a collection of banks, each maintaining a
 single account.  There are two kinds of transactions: \emph{transfer}
 (similar to \autoref{ex:bank}) and \emph{inquiry}.  In this example, a transfer
-is a transaction that moves some balance between two accounts.
+is a transaction that moves some funds between two accounts.
 An inquiry is a transaction over all accounts and checks that the sum of the
@@
 message to each of the servers involved in the transaction, asking them
-to prepare to commit to a transaction.  A server can either respond with
+to prepare to commit to their part in a particular transaction.
+A server can either respond with
 \texttt{YES} if it is ready to commit and will avoid doing anything
@@
 must decide to \emph{abort} the transaction.  In the second phase, the
-servers that responded with \texttt{YES} (if any) must be notified in
-either case to inform them of the decision.
+servers that responded with \texttt{YES} (if any) must be notified
+to inform them of the coordinator's decision.
+
+Different transactions can have different coordinators.  In our
+implementation, each bank and each coordinator is a thread.
+\autoref{fig:2pc1} shows the code for a bank.
+The state of a bank consists of the following local variables:
+\begin{itemize}
+\item \textit{self}: the bank's identifier;
+\item \textit{balance}: the current balance in the account;
+\item \textit{status}: either \texttt{.idle} or
+$(\mathtt{.busy}, \mathit{coordinator}, \mathit{oldbalance})$,
+where \textit{coordinator} is the identifier of the coordinator
+of a transaction and \textit{oldbalance} is the balance
+just before the transaction started;
+\item \textit{received}: the set of messages received and handled so far.
+\end{itemize}
+
+\noindent
+Messages sent to a bank have the following dictionary format:
+$\{ \mathtt{.dst}, \mathtt{.src}, \mathtt{.request}, \mathtt{.amount} \}$:
+\begin{itemize}
+\item \texttt{.dst}: identifier of the bank;
+\item \texttt{.src}: identifier of the coordinator that sent the message;
+\item \texttt{.request}: request type, which is either
+\texttt{.withdraw}, \texttt{.deposit}, \texttt{.commit}, or \texttt{.abort};
+\item \texttt{.amount}: amount to withdraw or deposit.
+\end{itemize}
+
+\noindent
+A bank waits for a message destined to itself that it has not yet received.
+In case of a withdrawal when the bank is idle and there are sufficient
+funds, the bank changes its status to \texttt{.busy} and keeps track of
+who to coordinator is
 

### 2021-07-30 book/paper.tex
Latest

@@
 \item \textit{balance}: the current balance in the account;
-\item \textit{status}: either \texttt{.idle} or
-$(\mathtt{.busy}, \mathit{coordinator}, \mathit{oldbalance})$,
-where \textit{coordinator} is the identifier of the coordinator
-of a transaction and \textit{oldbalance} is the balance
+\item \textit{status}: either contains () if the bank is not involved in an
+ongoing transaction or contains the balance
 just before the transaction started;

### 2021-07-30 book/paper.tex
Latest

@@
 In case of a withdrawal when the bank is idle and there are sufficient
-funds, the bank changes its status to \texttt{.busy} and keeps track of
-who to coordinator is
+funds, the bank saves the current balance in \textit{status} to indicate
+an ongoing transaction and what its original balance was.
+The bank then responds with an \texttt{.yes} message to the coordinator,
+including the new balance.
+Otherwise, the bank responds with a \texttt{.no} message.
+Deposits are similar, except that it is not necessary to check for
+sufficient funds.
+In case of a \texttt{.commit} message, the bank changes its status
+to (), indicating that there is no ongoing transaction.  In case of
+a \texttt{.abort} message, the bank restores \textit{balance} first.
+
+\autoref{fig:2pc2} contains the code for transfers and inquiries, as well
+as tests.
 

### 2021-07-31 book/paper.tex
Added more 2pc material

@@
 only knows its successor on the ring.
-\item \autoref{ch:2pc} describes the two-phase commit protocol used to implement
-atomic database transactions.
+\item \autoref{ch:2pc} describes atomic database transactions and the two-phase commit protocol used to implement them.
 \item \autoref{ch:consensus} demonstrates a fault-tolerant distributed consensus
@@
 
-\chapter{Two Phase Commit}
+\chapter{Transactions and Two Phase Commit}
 \label{ch:2pc}
@@
 
-Modern databases support multiple clients concurrently accessing the data
-stored in them.  Making matters more complicated, they store data on disk, but
-we will ignore that in this book.
+Modern databases support multiple clients concurrently accessing the data.
+They store data on disk, but we will ignore that in this book.
+(If you want to model a disk, this is probably best done as a separate
+thread.)
 The complication we address here is that databases may be \emph{sharded},
@@
 single account.  There are two kinds of transactions: \emph{transfer}
-(similar to \autoref{ex:bank}) and \emph{inquiry}.  In this example, a transfer
+(similar to \autoref{ex:bank}) and \emph{check}.  In this example, a transfer
 is a transaction that moves some funds between two accounts.
-An inquiry is a transaction over all accounts and checks that the sum of the
+A check is a transaction over all accounts and checks that the sum of the
 balances across the accounts remains the same.
@@
 and become a no-op.  Even if this is not the case, a concurrent
-inquiry transaction may accidentally witness a state in which either
+check transaction may accidentally witness a state in which either
 the withdrawal or the deposit happened, but not both.  And matters
@@
 that might jeopardize this (like committing a conflicting transaction),
-or \texttt{NO} if it does not want to participate in the transaction.
+or with \texttt{NO} if it does not want to participate in the transaction.
 If all servers respond with \texttt{YES}, then the coordinator can
@@
 an ongoing transaction and what its original balance was.
-The bank then responds with an \texttt{.yes} message to the coordinator,
+The bank then responds with a \texttt{.yes} message to the coordinator,
 including the new balance.
@@
 as tests.
+The \texttt{receive()} method is used by coordinators in a \textbf{select}
+statement to wait for a response from each bank involved in a transaction.
+Argument \texttt{self} is the
+identifier of the coordinator and \texttt{sources} is the set of banks.
+It returns the empty set if there not yet responses from all banks.
+Otherwise it returns a singleton set containing the set of responses,
+one for each source.
+
+The \texttt{transfer}() method contains the code for the transfer
+transaction.  Argument \textit{self} is the identifier of the coordinator,
+\textit{b1} is the source bank, \textit{b2} is the destination bank,
+and \textit{amt} is the amount to transfer.
+The coordinator sends a \texttt{PREPARE} message containing a
+\texttt{.withdraw} request to \textit{b1} and a \texttt{PREPARE}
+message containing a \texttt{.deposit} request to \textit{b2}.
+It then waits for responses from each.  If both responses are
+\texttt{.yes}, then it commits the transaction, otherwise it aborts
+the transaction.
+
+The \texttt{check}() method checks if the sum of the balances equals
+\textit{total}, the sum of the initial balances.
+The code is similar to \textit{transfer}, except that it always
+aborts the transaction---there is no need to ever commit it.
+As a code-saving hack: the balance inquiry is done by withdrawing \$0.
+
+As for testing, the initial balances are picked arbitrarily between
+0 and \texttt{MAX\_BALANCE} (and Harmony as always will try every
+possible set of choices).
+Each coordinator chooses whether to do a transfer or a check.  In
+case of a transfer, it also chooses the source bank and the destination
+bank.
+
+While the protocol perhaps seems simple enough, there are a lot of
+\textbf{if} statements in the code, making it hard to reason about
+correctness.
+Model checking is useful to see if there are corner
+cases where the code does not work.
+While confidence increases by
+increasing the number of banks or the number of coordinators, doing so
+quickly increases the number of possible states so that model checking
+may become infeasible.
+
+\section*{Exercises}
+\begin{problems}
+\item In \autoref{ex:bank} the code ran into a deadlock.  Can the code
+in this chapter run into a deadlock?  Explain.
+\item Transactions can fail for two reasons: a transfer transaction can
+fail because of insufficient funds, but in general transaction can fail
+if there is a conflict with another transaction.  The latter can be
+fixed by retrying the transaction until it commits.  Implement this.
+\item One way to reduce the number of conflicts between transactions
+is to distinguish read and write operations.  Two read operations
+(in our case, operations that withdraw \$0) do not conflict,
+so a bank could have multiple ongoing read operations
+for different transactions.  Implement this.
+\end{problems}
 

### 2021-08-01 book/paper.tex
Latest

@@
 \item \autoref{ch:2pc} describes atomic database transactions and the two-phase commit protocol used to implement them.
+\item \autoref{ch:abd} presents a protocol for a fault-tolerant
+replicated object that supports only read and write operations.
 \item \autoref{ch:consensus} demonstrates a fault-tolerant distributed consensus
@@
 for different transactions.  Implement this.
+\item Two-phase commit can tolerate servers failing.  If a server does not
+respond within some reasonable amount of time, the coordinator can abort
+the transaction.  Implement this.
 \end{problems}
+
+\chapter{Replicated Atomic Read/Write Register}
+\label{ch:abd}
+
+\begin{figure}
+\begin{code}
+\harmonysource{abd1}
+\end{code}
+\caption{\harmonylink{code/abd.hny} An implementation of a replicated atomic read/write register.}
+\label{fig:abd1}
+\end{figure}
+
+\begin{figure}
+\begin{code}
+\harmonysource{abd2}
+\end{code}
+\caption{\harmonylink{code/abd.hny} Testing the code in \autoref{fig:abd1}.}
+\label{fig:abd2}
+\end{figure}
 
@@
 \hline
-Set & \texttt{\{ 1, 2, 3 \}, \{ False, .id, 3 \} } \\
+Set & \texttt{\{\}, \{ 1, 2, 3 \}, \{ False, .id, 3 \} } \\
 \hline

### 2021-08-01 book/paper.tex
Latest

@@
 
+\index{logical timestamp}%
 Harmony provides support for testing linearizability,
@@
 \begin{code}
-\harmonysource{abd1}
+\harmonysource{abd}
 \end{code}
 \caption{\harmonylink{code/abd.hny} An implementation of a replicated atomic read/write register.}
-\label{fig:abd1}
-\end{figure}
-
-\begin{figure}
-\begin{code}
-\harmonysource{abd2}
-\end{code}
-\caption{\harmonylink{code/abd.hny} Testing the code in \autoref{fig:abd1}.}
-\label{fig:abd2}
-\end{figure}
+\label{fig:abd}
+\end{figure}
+
+\begin{figure}
+\begin{code}
+\harmonysource{abdtest}
+\end{code}
+\caption{\harmonylink{code/abdtest.hny} Testing the code in \autoref{fig:abd}.}
+\label{fig:abdtest}
+\end{figure}
+
+A \emph{register} is an object that you can read or write.  In a distributed
+system, it can be useful to have shared registers.  A simple shared register
+would be maintained by a server, and clients can read or write the
+shared register by exchanging messages with the server.
+We call two operations such that one does not finish before the other starts
+\emph{concurrent}.
+Since messages are delivered one at a time to the server,
+concurrent operations on the shared register appear atomic.
+In particular, we have the following three desirable properties:
+\begin{enumerate}
+\item All write operations are ordered;
+\item A read operation returns either the last value written or the value of
+a concurrent write operation.
+\item If read operation $r_1$ finishes before read operation $r_2$ starts,
+then $r_2$ cannot return a value that is older than the value returned
+by $r_1$.
+\end{enumerate}
+
+\index{single point of failure}%
+Unfortunately, however, the server is a \emph{single point of failure}:
+if it fails, all its clients suffer.
+We would therefore like to find a solution that can survive the crash
+of a server.
+In this chapter, we will \emph{replicate} the object: maintain multiple
+copies.
+One could, for example, imaging that clients write to all copies and read
+from any single one.
+While this solves the single-point-of-failure problem, we lose all the
+nice properties above.  For one, it is not guaranteed that all servers
+see all write operations in the same order.
+
+\index{logical timestamp}%
+We present a protocol here that is based on the work by
+Hagit Attiya, Amotz Bar-Noy, and Danny Dolev~\cite{ABD95}.
+In order to tolerate \texttt{N} failures,
+it uses $\mathtt{N} = 2\mathtt{F} + 1$ servers.
+Each server maintains not only the value of the object, but also a
+\emph{logical timestamp} (see also \autoref{ch:testing}).
+All write operation will be ordered by a unique timestamp.
 

### 2021-08-02 book/paper.tex
Working on atomic read/write register

@@
 A \emph{register} is an object that you can read or write.  In a distributed
-system, it can be useful to have shared registers.  A simple shared register
+system, it can be useful to have registers that are shared.
+A simple shared register
 would be maintained by a server, and clients can read or write the
@@
 \index{logical timestamp}%
-We present a protocol here that is based on the work by
-Hagit Attiya, Amotz Bar-Noy, and Danny Dolev~\cite{ABD95}.
-In order to tolerate \texttt{N} failures,
+We present a protocol preserving these properties that is based on the
+work by Hagit Attiya, Amotz Bar-Noy, and Danny Dolev~\cite{ABD95}.
+In order to tolerate \texttt{F} failures,
 it uses $\mathtt{N} = 2\mathtt{F} + 1$ servers.
-Each server maintains not only the value of the object, but also a
-\emph{logical timestamp} (see also \autoref{ch:testing}).
-All write operation will be ordered by a unique timestamp.
+In other words, the register survives as long as a strict majority of
+its copies survive.
+All write operation will be ordered by a unique \emph{logical timestamp}
+(see also \autoref{ch:testing}).
+Each server maintains not only the value of the object, but also the
+timestamp of its corresponding write operation.
+
+Each read and write operation consists of two \emph{phases}.  In a phase,
+a client broadcasts a request to all servers and waits for responses
+from a majority ($\texttt{N} - \texttt{F}$ or equivalently
+$\texttt{F} + 1$ servers).  Note that because we are assuming that no
+more than \texttt{F} servers can fail, doing so is safe, in that a client
+cannot indefinitely block as long as that assumption holds.
+
+In the first phase, an operation asks each server for its current
+timestamp and value.  After receiving $\mathtt{N} - \mathtt{F}$ responses,
+the client determines the response with the highest timestamp.
+In case of a write operation, the client then computes a new unique
+timestamp that is strictly higher than the highest it has seen.
+To make this work, timestamps are actually lexicographically ordered
+tuples consisting of an integer and the unique identifier of the
+client that writes the value.  So if $(t, c)$ is the highest timestamp
+observed, and client $c'$ needs to create a new timestamp, it can select
+$(t + 1, c')$.  After all $(t + 1, c') > (t, u)$ and no other client
+will create the same timestamp.
+
+Suppose client $c'$ is trying to write a value $v$.
+In phase 2, client $c'$ broadcasts a request containing timestamp $(t+1, c')$
+and $v$.  Each server that receives the request compares $(t+1, c')$ against
+its current timestamp.  If $(t+1, c')$ is larger than its current timestamp,
+it adopts the new timestamp and its corresponding value $v$.  In either case,
+the server responds to the client.
+Upon $c'$ receiving a response from $\mathtt{N} - \mathtt{F}$ servers,
+the write operation completes.
+In case of a read operation, client $c'$ simply \emph{writes back} the
+highest timestamp it saw along with its corresponding value.
+
+\autoref{fig:abd} contains the code for a server, as well as the code
+for read and write operations.
+For efficiency of model checking, the servers are anonymous---otherwise
+we would have to consider every permutation of states of those servers.
+Because the servers are anonymous, they may end up sending the same
+exact message, but clients are waiting for a particular number of
+messages.
+Because of this, 
+we will model the network as a bag (aka multiset) of messages.
+The bag itself is
+modeled in Harmony as a dictionary that maps unique elements in the bag to their
+multiplicity.  So, for example, $\{ .\texttt{a}:1, .\texttt{b}:2 \}$ is the bag that consists of
+one $.\texttt{a}$ and two copies of $.\texttt{b}$.
+The \texttt{bag} module (\autoref{ap:bag}) contains a variety of handy functions on bags.
+
+In case of a server, its timestamp $t$ is initialized to
+$(0, \mathtt{None})$, and its value to \texttt{None}.
+Each server also keeps track of all the requests its already received
+so it doesn't handle the same request twice.
+The rest of the code is fairly straightforward.
+
+Read and write operations are both invoked with a unique identifier
+\textit{uid}.  Both start with broadcasting a \texttt{.read}
+request to all servers and then waiting for a response from
+$\mathtt{N} - \mathtt{F}$ servers.
+The \texttt{receive()} function uses the \texttt{bag.combinations} method
+to find all combinations of subsets of responses of size
+$\mathtt{N} - \mathtt{F}$.
+The second phase of each operation is similar.
+
+\autoref{fig:abdtest} tests the third property of the properties listed
+above.  The \texttt{writer()} method writes a new value \texttt{.token}.
+The \texttt{reader()} method reads twice and makes sure that if the
+first read operation returns \texttt{.token}, then so does the second.
+This illustrates the importance of the second phase of the \texttt{read}
+operation.  You can comment out Lines~32 and~33 in \autoref{fig:abd} and
+see what goes wrong.
 
@@
 
-We will model the network as a bag (aka multiset) of messages.  The bag itself is
-modeled in Harmony as a dictionary that maps unique elements in the bag to their
-multiplicity.  So, for example, $\{ .\texttt{a}:1, .\texttt{b}:2 \}$ is the bag that consists of
-one $.\texttt{a}$ and two copies of $.\texttt{b}$.
-The \texttt{bag} module (\autoref{ap:bag}) contains a variety of handy functions on bags.
+\autoref{fig:bosco} presents our algorithm.  Besides the network variable, it uses a
+shared list of proposals and a shared set of decisions.
 In this particular algorithm, all messages are broadcast to all processors, so they do
 not require a destination address.
-
-\autoref{fig:bosco} presents our algorithm.  Besides the network variable, it uses a
-shared list of proposals and a shared set of decisions.  The invariants we want to maintain
+The invariants we want to maintain
 on those variables is that $\textit{decisions} \subseteq \textit{proposals}$

### 2021-08-02 book/paper.tex
Latest

@@
 system, it can be useful to have registers that are shared.
-A simple shared register
-would be maintained by a server, and clients can read or write the
+A simple shared register implementation
+would have its value maintained by a server, and clients can read or write the
 shared register by exchanging messages with the server.
@@
 copies.
-One could, for example, imaging that clients write to all copies and read
+One could, for example, imagine that clients write to all copies and read
 from any single one.
@@
 nice properties above.  For one, it is not guaranteed that all servers
-see all write operations in the same order.
+receive and process all write operations in the same order.
 
@@
 
-In the first phase, an operation asks each server for its current
+In the first phase, a client asks each server for its current
 timestamp and value.  After receiving $\mathtt{N} - \mathtt{F}$ responses,
@@
 client that writes the value.  So if $(t, c)$ is the highest timestamp
-observed, and client $c'$ needs to create a new timestamp, it can select
-$(t + 1, c')$.  After all $(t + 1, c') > (t, u)$ and no other client
+observed by client $c'$, and $c'$  needs to create a new timestamp,
+it can select $(t + 1, c')$.
+After all $(t + 1, c') > (t, u)$ and no other client
 will create the same timestamp.
@@
 In case of a read operation, client $c'$ simply \emph{writes back} the
-highest timestamp it saw along with its corresponding value.
+highest timestamp it saw in the first phase along with its corresponding value.
 
@@
 
-In case of a server, its timestamp $t$ is initialized to
-$(0, \mathtt{None})$, and its value to \texttt{None}.
+A server initializes its timestamp $t$ to
+$(0, \mathtt{None})$ and its value to \texttt{None}.
 Each server also keeps track of all the requests its already received
@@
 Read and write operations are both invoked with a unique identifier
-\textit{uid}.  Both start with broadcasting a \texttt{.read}
+\textit{uid}.  Both start by broadcasting a \texttt{.read}
 request to all servers and then waiting for a response from
@@
 
-More precisely, we will
-define a constant \texttt{F} that contains the maximum number of failures, and
-assume there are $\texttt{N} = 3\texttt{F} + 1$ processors.
+More precisely,
+constant \texttt{F} contains the maximum number of failures, and
+we will assume there are $\texttt{N} = 3\texttt{F} + 1$ processors.
 Each processor \emph{proposes} a value, which we assume here to be from the

### 2021-08-02 book/paper.tex
Moved some text

@@
 
+One may wonder how failures can occur in this model.  They are not
+explicitly modeled, but Harmony tries every possible execution.
+This includes executions in which the clients terminate before
+$\texttt{F}$ of the servers start executing.  To the clients, this
+is indistinguishable from executions in which those
+servers have failed.
+
 \chapter{Distributed Consensus}
@@
 After that, proceed with the next round.
-
-One may wonder how failures can occur in this model.  They are not explicitly modeled, but
-Harmony tries every possible execution, including executions in which some processors make
-no execution steps unless there are no other processors that can run.  The way this algorithm
-works, this includes executions in which $\texttt{N} - \texttt{F}$ processors terminate
-before the remaining $\texttt{F}$ processors start executing.  To the first set of processors,
-this is indistinguishable from executions in which the last set of processors have failed.
 

### 2021-08-03 book/paper.tex
Latest

@@
 
+\begin{figure}
+\begin{code}
+\harmonysource{bosco2}
+\end{code}
+\caption{\harmonylink{code/bosco2.hny} Reducing the state space.}
+\label{fig:bosco2}
+\end{figure}
+
+While one can run this code within little time for $\mathtt{F} = 1$, for
+$\mathtt{F} = 2$ the state space to explore is already quite large.
+One way to reduce the state space to explore is the following realization:
+each processor only considers messages for the round that it is in.
+If a message is for an old round, the processor will ignore it;
+if a message is for a future round, the processor will buffer it.
+So one can simplify the model and have each processor wait
+for \emph{all} \texttt{N} messages in a round
+instead of $\mathtt{N} - \mathtt{F}$.
+It would still have to choose to consider just $\mathtt{N} - \mathtt{F}$
+out of those \texttt{N} messages, but executions in which some processors
+are left behind in all rounds are no longer considered.
+It still includes executions where some subset of $\mathtt{N} - \mathtt{F}$
+processors only choose each other messages and essentially ignore the
+messages of the remaining \texttt{F} processors, so the resulting model
+is just as good.
+
+\autoref{fig:bosco2} shows the code for this model.  Running this with
+$\mathtt{F} = 2$ does not take very long and this approach is a good
+blueprint for testing other round-based protocols (of which there are
+many).
+
 \section*{Exercises}

### 2021-08-07 book/paper.tex
Working on chain replication chapter

@@
 }
+\newglossaryentry{replication}
+{
+  name=replication,
+  description={maintains multiple copies of some resource to improve availability in the face of failures}
+}
 \newglossaryentry{semaphore}
@@
       the contents of its shared memory and the set of contexts}
+}
+\newglossaryentry{state machine replication}
+{
+  name=state machine replication,
+  description={is a replication technique in which a collection of deterministic state machines process the same inputs in the same order}
 }
@@
 \item \autoref{ch:2pc} describes atomic database transactions and the two-phase commit protocol used to implement them.
+\item \autoref{ch:chain} describes \emph{state machine replication} and
+the \emph{chain replication} protocol to support replication.
 \item \autoref{ch:abd} presents a protocol for a fault-tolerant
@@
 
+\chapter{Chain Replication}
+\label{ch:chain}
+
+\begin{figure}
+\begin{code}
+\harmonysource{chain}
+\end{code}
+\caption{\harmonylink{code/chain.hny} Chain Replication.}
+\label{fig:chain}
+\end{figure}
+
+\glsadd{replication}%
+\index{replication}%
+As you have probably experienced, computers can crash.
+If you are running a web service, you may not be able to afford
+a long outage.  If you are running software that flies a plane,
+then an outage for any length of time could lead to a disaster.
+To deal with service outages caused by computers crashing, you may want
+to \emph{replicate} the service onto multiple computers.  As long as
+one of the computers survives, the service remains available.
+
+\glsadd{state machine replication}%
+\index{state machine replication}%
+Besides availability, it is usually important that the replicated
+services acts as if it were a single one.  This requires that the
+replicas of the service must coordinate their actions.
+One general approach is the
+\emph{Replicated State Machine Approach}~\cite{Lam78,S90}.
+In this approach, you model your service as a deterministic state
+machine.  The replicas each run a copy of the state machine, started
+in the same state.  As long as the replicas handle the same inputs
+in the same order, determinism guarantees that they produce the same
+outputs in the same order.
+The trick then is to ensure that all replicas handle the
+same requests in the same order and to do so in a way that continues
+to work even if some subset of replicas crash.
+
+\emph{Chain Replication}~\cite{vRS04} is a protocol to do just this.
+In Chain Replication, the replicas are organized in a linear chain.
+Each replica monitors its direct predecessor in the chain.  If it has
+no predecessors, we call it the \emph{head}.  The head may change over
+time, as a head may crash and replaced by another.
+We model the state machine as a \emph{history}: a sequence of operations.
+Only the head is allowed to introduce new updates.
+When it does so, it advertises the new update along with the position
+in the history.  When the direct successor receives the next update
+it is expecting, it appends the update to its history and likely
+advertises the new update.  When the update is applied to the history
+of the \emph{tail} replica (the replica that has no live successors),
+then the update is considered final.
+
+So, when a replica fails, its successors should find out about it.
+In practice, one server can detect the failure of another server by
+pinging it.  If a server does not receive a response to its ping within
+some maximum amount of time, then the server considers its peer crashed.
+Note that this, in general is not a safe thing to do---the network or
+the peer may be temporarily slow but the peer is not necessarily crashed
+when the timer expires.
+Nonetheless, we will assume here that failure detection does not make
+mistakes and that eventually every failure is detected.
+This is sometimes called the \emph{Fail-Stop} failure model~\cite{SS83},
+which is distinct from the often more realistic \emph{Crash} failure model
+where processes can crash but accurate detection is not available.
+
+\autoref{fig:chain} shows an implemenation of chain replication.
+
 \chapter{Replicated Atomic Read/Write Register}

### 2021-08-07 book/paper.tex
Latest

@@
 \autoref{fig:chain} shows an implemenation of chain replication.
+The code starts \texttt{NREPLICAS} \texttt{replica} threads numbered
+0 through $\mathtt{NREPLICAS} - 1$.  At least one of the replicas 
+must be immortal---the code selects one of them although none of the
+replicas is required to fail during execution.
+Each replica maintains its history and a set of its predecessors.
+It then loops until either it fails or it has applied all
+\texttt{NOPS} operations to its history.
+$\mathbf{choose}(\{ \mathit{immortal}, \texttt{True} \})$ can only
+evalute to \texttt{False} in case $\textit{immortal}$ is false.
+What it does within the loop depends on whether it is the head
+(has no alive predecessors) or not.
+Because failure detection is accurate, at most one replica can think
+it is the head at any time.  Moreover, eventually, when it has
+detected all its predecessors having failed, always some replica
+thinks it is the head.
+
+Messages have the format $(\mathit{source}, \mathit{payload})$,
+where \textit{source} is the identifier of the replica that generated
+the message and \textit{payload} has information about the state of
+the replica.  The payload can take one of two forms.  The first is
+\texttt{.crash}, indicating that the source of the message has
+crashed.  The code models a replica crashing by that replica broadcasting
+this message on the network.  The other form is a tuple
+$(\mathit{index}, \mathit{operation})$, which means that the source
+placed the given operation at the given index in its history.
 

### 2021-08-08 book/paper.tex
Finished draft of chain replication

@@
 \autoref{fig:progs} shows two programs.  Program (a) is sequential.
-It sets \textit{shared} to \texttt{True}, asserts that
-\textit{shared} = \texttt{True}, and finally sets
-\textit{shared} to \texttt{False}.
+It sets \textit{shared} to \textbf{True}, asserts that
+\textit{shared} = \textbf{True}, and finally sets
+\textit{shared} to \textbf{False}.
 If you run the program through Harmony, it will not find any problems
@@
 Method \texttt{incrementer} takes a parameter called \textit{self}.
-It increments \textit{count} and sets \textit{done}[\textit{self}] to \texttt{True}.
+It increments \textit{count} and sets \textit{done}[\textit{self}] to \textbf{True}.
 It then waits until the other thread is done.
@@
 stores $1$ into \textit{count};
-\item \texttt{incrementer(0)} sets \textit{done}[0] to \texttt{True};
-\item \texttt{incrementer(1)} sets \textit{done}[1] to \texttt{True}.
+\item \texttt{incrementer(0)} sets \textit{done}[0] to \textbf{True};
+\item \texttt{incrementer(1)} sets \textit{done}[1] to \textbf{True}.
 \end{enumerate}
@@
 Harmony values are recursively defined:
-they include booleans (\texttt{False} and \texttt{True}),
+they include booleans (\textbf{False} and \textbf{True}),
 integers (but not floating point numbers),
@@
 So, if $x$ = \texttt{[False, True]},
-then $x$[0] = \texttt{False} and $x$[1] = \texttt{True}, just like in Python.
+then $x$[0] = \textbf{False} and $x$[1] = \textbf{True}, just like in Python.
 However, when creating a singleton list, make sure you include the
 comma, as in \texttt{[False,]}.  The expression \texttt{[False]} just means
-\texttt{False}.
+\textbf{False}.
 
@@
 executes instructions 6 through 18, storing 1 into
-\textit{count} and storing \texttt{True} into \textit{done}[1];
+\textit{count} and storing \textbf{True} into \textit{done}[1];
 \item Thread \texttt{incrementer(0)}
@@
 storing value 1 into \textit{count} (instruction 10),
-storing \texttt{True} into \textit{done}[0] (instructions 11 through 15)
-finding that \textit{done}[1] is \texttt{True} (instructions 16 through 21),
+storing \textbf{True} into \textit{done}[0] (instructions 11 through 15)
+finding that \textit{done}[1] is \textbf{True} (instructions 16 through 21),
 and finally detecting that the assertion is violated
@@
 na\"{\i}ve (and, as it turns out, broken) implementation of a lock.
-Initially the lock is not owned, indicated by \textit{lockTaken} being \texttt{False}.
-To enter the critical section, a thread waits until \textit{lockTaken} is \texttt{False}
-and then sets it to \texttt{True} to indicate that the lock has been taken.
+Initially the lock is not owned, indicated by \textit{lockTaken} being \textbf{False}.
+To enter the critical section, a thread waits until \textit{lockTaken} is \textbf{False}
+and then sets it to \textbf{True} to indicate that the lock has been taken.
 The thread then executes the critical section.  Finally the thread
-releases the lock by setting \textit{lockTaken} back to \texttt{False}.
+releases the lock by setting \textit{lockTaken} back to \textbf{False}.
 
@@
 Thread~0 finds that the lock is available, but just before it stores
-\texttt{True} in \textit{lockTaken} in instruction~9, thread~1 gets
+\textbf{True} in \textit{lockTaken} in instruction~9, thread~1 gets
 to run.
@@
 to see what it does.)
-Because \textit{lockTaken} is still \texttt{False}, it too
-believes it can acquire the lock, and stores \texttt{True} in
+Because \textit{lockTaken} is still \textbf{False}, it too
+believes it can acquire the lock, and stores \textbf{True} in
 \textit{lockTaken} and moves on to the critical section.
-Finally, thread~0 moves on, also stores \texttt{True}
+Finally, thread~0 moves on, also stores \textbf{True}
 into \textit{lockTaken} and also moves
@@
 To see why, first note the following invariant: if thread $i$ is in the
-critical section, then \textit{flags}[$i$] == \texttt{True}.
+critical section, then \textit{flags}[$i$] == \textbf{True}.
 Without loss of generality,
@@
 Thread 0 can only reach the critical section if at some time $t_1$,
-$t_1 > t_0$, it finds that \textit{flags}[1] == \texttt{False}.
-Because of the invariant, \textit{flags}[1] == \texttt{False} implies that
+$t_1 > t_0$, it finds that \textit{flags}[1] == \textbf{False}.
+Because of the invariant, \textit{flags}[1] == \textbf{False} implies that
 thread~1 is not in the critical section at time $t_1$.
 Let $t_2$ be the time at which thread~0 sets \textit{flags}[0]
-to \texttt{False}.  Thread~0 is in the critical section sometime
+to \textbf{False}.  Thread~0 is in the critical section sometime
 between $t_1$ and $t_2$.
 It is easy to see that thread~1 cannot enter the critical section
-between $t_1$ and $t_2$, because \textit{flags}[1] == \texttt{False} at
+between $t_1$ and $t_2$, because \textit{flags}[1] == \textbf{False} at
 time $t_1$.  To reach the critical section between $t_1$ and $t_2$,
-it would first have to set \textit{flags}[1] to \texttt{True} and
-then wait until \textit{flags}[0] == \texttt{False}.  But that does not happen
+it would first have to set \textit{flags}[1] to \textbf{True} and
+then wait until \textit{flags}[0] == \textbf{False}.  But that does not happen
 until time $t_2$.
@@
 until either the other thread is nowhere near the critical section
-($\mathit{flag}[1 - \mathit{self}] == \mathtt{False}$) or has given way
+($\mathit{flag}[1 - \mathit{self}] == \mathbf{False}$) or has given way
 ($\mathit{turn} == \mathit{self}$).
@@
 \begin{enumerate}
-\item \textit{flags}[1] = \texttt{False};
+\item \textit{flags}[1] = \textbf{False};
 \item \textit{turn} = 0;
@@
 In the first case, if thread~1 takes a step, there are two possibilities:
-either $flags[1]$ will still be \texttt{False} (in which case the first case
-continues to hold), or $flags[1]$ will be \texttt{True}
+either $flags[1]$ will still be \textbf{False} (in which case the first case
+continues to hold), or $flags[1]$ will be \textbf{True}
 and thread~1 will be at label \texttt{@gate} (in which case the third case
@@
 Because thread~0 entered the critical section, we know that
-\textit{flags}[1] = \texttt{False} or \textit{turn} == 0 because
+\textit{flags}[1] = \textbf{False} or \textit{turn} == 0 because
 of the \texttt{await} condition.
@@
 Peterson's algorithm (setting \textit{flags}[\textit{self}])
-to \texttt{True} and \textit{turn} to $1 - \mathit{self}$) can be reversed.
+to \textbf{True} and \textit{turn} to $1 - \mathit{self}$) can be reversed.
 After all, they are different variables assigned independent values---in a
@@
 private variable (the ``test'')
-and then sets the shared variable to \texttt{True} (``set'').
+and then sets the shared variable to \textbf{True} (``set'').
 
@@
 The program uses $\mathtt{N}+1$ variables.
-Variable \textit{shared} is initialized to \texttt{False} while
-\textit{private}[$i$] for each thread $i$ is initialized to \texttt{True}.
+Variable \textit{shared} is initialized to \textbf{False} while
+\textit{private}[$i$] for each thread $i$ is initialized to \textbf{True}.
 
 An important invariant, $\mathcal{I}_1$, of the program is that at any time at most
-one of these variables is \texttt{False}.
+one of these variables is \textbf{False}.
 Another invariant, $\mathcal{I}_2(i)$, is that if thread $i$ is executing between
 lines $18 \cdots 22$ (which includes the critical section),
-then \textit{private}[$i$] == \texttt{False}.
+then \textit{private}[$i$] == \textbf{False}.
 Between the two (i.e., $\mathcal{I}_1 \land \forall i: \mathcal{I}_2(i)$),
@@
 To see that invariant $\mathcal{I}_1$ is maintained, note that
-\texttt{!}$p$ == \texttt{True} upon entry of \texttt{tas}
+\texttt{!}$p$ == \textbf{True} upon entry of \texttt{tas}
 (because of the condition on the \textbf{while} loop that the
@@
 \begin{enumerate}
-\item \texttt{!}$s$ is \texttt{False} upon entry to \texttt{tas}.
-Then upon exit \texttt{!}$p$ == \texttt{False} and \texttt{!}$s$ == \texttt{True},
+\item \texttt{!}$s$ is \textbf{False} upon entry to \texttt{tas}.
+Then upon exit \texttt{!}$p$ == \textbf{False} and \texttt{!}$s$ == \textbf{True},
 maintaining the invariant.
-\item \texttt{!}$s$ is \texttt{True} upon entry to \texttt{tas}.
+\item \texttt{!}$s$ is \textbf{True} upon entry to \texttt{tas}.
 Then upon exit nothing has changed, maintaining the invariant.
@@
 because we can assume, by the induction hypothesis, that $\textit{private}[i]$ is
-\texttt{True} just before exiting the critical section.
+\textbf{True} just before exiting the critical section.
 Invariant $\mathcal{I}_2(i)$ is obvious as (i) thread $i$ only proceeds to the critical
-section if \textit{private}[$i$] is \texttt{False}, and (ii) no other thread modifies
+section if \textit{private}[$i$] is \textbf{False}, and (ii) no other thread modifies
 \textit{private}[$i$].
@@
 keyword.
-The expression counts the number of \texttt{False} values and checks that
+The expression counts the number of \textbf{False} values and checks that
 the result is less than or equal to 1.
@@
 This interface is like \texttt{acquire}(\textit{b}) but never blocks.  It
-returns \texttt{True} if the binary semaphore was available (and now acquired) or \texttt{False}
+returns \textbf{True} if the binary semaphore was available (and now acquired) or \textbf{False}
 if the binary semaphore was already acquired.
@@
 \texttt{tryAcquire} should not wait.
-Instead it should return \texttt{True} if the lock was successfully
-acquired and \texttt{False} if the lock was not available.
+Instead it should return \textbf{True} if the lock was successfully
+acquired and \textbf{False} if the lock was not available.
 % \item Implement semaphores using Mesa condition variables.
@@
 temporarily.  In Harmony, this can be done by setting the \textit{interrupt level}
-of a thread to \texttt{True} using the \textbf{setintlevel} interface.
+of a thread to \textbf{True} using the \textbf{setintlevel} interface.
 \autoref{fig:trap4} illustrates how this is done.
 Note that it is not necessary to change the interrupt level during servicing an
-interrupt, because it is automatically set to \texttt{True} upon entry to the interrupt
-handler and restored to \texttt{False} upon exit.
+interrupt, because it is automatically set to \textbf{True} upon entry to the interrupt
+handler and restored to \textbf{False} upon exit.
 It is important that the \textit{main}() code re-enables interrupts after incrementing
@@
 Implement a new method \texttt{i\_put}(\textit{item}) that does not
-block.  Instead, it should return \texttt{False} if the buffer is full and \texttt{True}
+block.  Instead, it should return \textbf{False} if the buffer is full and \textbf{True}
 if the item was successfully enqueued.
@@
 Only the last invocation of \texttt{app\_recv}(\textit{reliable}) is invoked with
-\textit{reliable} == \texttt{True}.  It causes the last acknowledgment to be sent
+\textit{reliable} == \textbf{True}.  It causes the last acknowledgment to be sent
 reliably.  It allows the receiver (Bob) to stop, as well as the sender (Alice)
@@
 \end{code}
-\caption{\harmonylink{code/chain.hny} Chain Replication.}
+\caption{\harmonylink{code/chain.hny} Checking the correctness of Chain Replication.}
 \label{fig:chain}
@@
 services acts as if it were a single one.  This requires that the
-replicas of the service must coordinate their actions.
-One general approach is the
-\emph{Replicated State Machine Approach}~\cite{Lam78,S90}.
-In this approach, you model your service as a deterministic state
+replicas of the service coordinate their actions.
+The \emph{Replicated State Machine Approach}~\cite{Lam78,S90} is a
+general approach to do just this.
+First, you model your service as a deterministic state
 machine.  The replicas each run a copy of the state machine, started
@@
 
-\emph{Chain Replication}~\cite{vRS04} is a protocol to do just this.
+\emph{Chain Replication}~\cite{vRS04} is such a replication protocol.
 In Chain Replication, the replicas are organized in a linear chain.
@@
 Only the head is allowed to introduce new updates.
-When it does so, it advertises the new update along with the position
+When it does so, it advertises the new update along with its position
 in the history.  When the direct successor receives the next update
-it is expecting, it appends the update to its history and likely
+it is expecting, it appends the update to its history and likewise
 advertises the new update.  When the update is applied to the history
@@
 where processes can crash but accurate detection is not available.
+We will consider that failure model in the upcoming chapters.
 
@@
 replicas is required to fail during execution.
+In practice, of course, a replica does not know whether it is immortal
+or not---it should just assume so.  The immortality is only used for
+modeling the system.
+
 Each replica maintains its history and a set of its predecessors.
@@
 \texttt{NOPS} operations to its history.
-$\mathbf{choose}(\{ \mathit{immortal}, \texttt{True} \})$ can only
-evalute to \texttt{False} in case $\textit{immortal}$ is false.
+$\mathbf{choose}(\{ \mathit{immortal}, \textbf{True} \})$ can only
+evaluate to \textbf{False} in case $\textit{immortal}$ is false.
 What it does within the loop depends on whether it is the head
@@
 Because failure detection is accurate, at most one replica can think
-it is the head at any time.  Moreover, eventually, when it has
-detected all its predecessors having failed, always some replica
+it is the head at any time.  Moreover, when it has
+detected all its predecessors having failed, eventually some replica
 thinks it is the head.
@@
 crashed.  The code models a replica crashing by that replica broadcasting
-this message on the network.  The other form is a tuple
+this message on the network.  The other form of payload is a tuple
 $(\mathit{index}, \mathit{operation})$, which means that the source
 placed the given operation at the given index in its history.
+
+If a replica thinks it is the head, it adds an operation to its
+history.
+In practice this would be some operation that the head would have received
+from a client.  Clients send their operations to the head, which adds them
+to its history in the order received.
+We are not modeling clients of the service.
+Instead, in the model, we add \textit{self} to the history until the
+history have \texttt{NOPS} operations on it or until the head crashes,
+whichever comes first.  The head also broadcasts this update, which
+is intended for its successor.  Because of failures, it may be that
+more than one replica is a successor of the head at different times.
+
+Otherwise, the replica awaits a message from its direct predecessor.
+If the payload is \texttt{.crash}, then the replica removes this
+predecessor from its set of predecessors.  For example, if there
+are initially three replicas numbered 0, 1, and 2, and replica~1
+crashes, then replica~2 removes~1 from its set of predecessors, making
+replica~0 its direct (and only) predecessor.  If, instead, the payload
+is a tuple, then it must be the case that it contains the next operation
+to add to its history.  The replica adds the operation to its history
+and broadcasts the update, again intended for its successor.
+
+When the replica successfully adds all \texttt{NOPS} updates to its
+history without crashing, the code makes sure that any other such replica
+ends in the same state.  Variable \textit{final} is used to check the
+different histories against one another.
 
@@
 In the initialization, the \texttt{proposals} variable is initialized
-to either $[\mathtt{False}, \mathtt{False}]$ or
-to $[\mathtt{True}, \mathtt{False}]$ (in case $\texttt{NBALLOTS} = 2$).
-The other combinations of \texttt{True} and \texttt{False} are symmetric,
+to either $[\mathbf{False}, \mathbf{False}]$ or
+to $[\mathbf{True}, \mathbf{False}]$ (in case $\texttt{NBALLOTS} = 2$).
+The other combinations of \textbf{True} and \textbf{False} are symmetric,
 so there is no need to run the model checker through those scenarios as
@@
 \hline
-Boolean & \texttt{False}, \texttt{True} \\
+Boolean & \textbf{False}, \textbf{True} \\
 \hline
@@
 are ordered by type according to the table above.
-So, for example, \texttt{\texttt{True} < 0 < .xyz < \{ 0 \})}.
+So, for example, \texttt{\textbf{True} < 0 < .xyz < \{ 0 \})}.
 Within types, the following rules apply:
@@
 
-The Boolean type has only two possible values: \texttt{False} and
-\texttt{True}.  Unlike in Python, in Harmony booleans are distinct
-from integers, and in particular $\mathtt{False} < 0$.  In statements
+The Boolean type has only two possible values: \textbf{False} and
+\textbf{True}.  Unlike in Python, in Harmony booleans are distinct
+from integers, and in particular $\mathbf{False} < 0$.  In statements
 and expressions where booleans are expected, it is not possible to
@@
 \hline
-\texttt{any $s$} & \texttt{True} if any value is \texttt{True} \\
-\hline
-\texttt{all $s$} & \texttt{True} if all values are \texttt{True} \\
+\texttt{any $s$} & \textbf{True} if any value is \textbf{True} \\
+\hline
+\texttt{all $s$} & \textbf{True} if all values are \textbf{True} \\
 \hline
@@
 \hline
-\texttt{any $d$} & \texttt{True} if any value is \texttt{True} \\
-\hline
-\texttt{all $d$} & \texttt{True} if all values are \texttt{True} \\
+\texttt{any $d$} & \textbf{True} if any value is \textbf{True} \\
+\hline
+\texttt{all $d$} & \textbf{True} if all values are \textbf{True} \\
 \hline
@@
 \hline
-\texttt{any $t$} & \texttt{True} if any value is \texttt{True} \\
-\hline
-\texttt{all $t$} & \texttt{True} if all values are \texttt{True} \\
+\texttt{any $t$} & \textbf{True} if any value is \textbf{True} \\
+\hline
+\texttt{all $t$} & \textbf{True} if all values are \textbf{True} \\
 \hline
@@
 $c$ evaluated to true.
-Harmony checks that $c$ is either \texttt{False} or \texttt{True}---if neither
+Harmony checks that $c$ is either \textbf{False} or \textbf{True}---if neither
 is the case, Harmony reports an error.
@@
 is evaluated atomically after every state change.
-If it ever evaluates to \texttt{False} Harmony reports and error.  Harmony
+If it ever evaluates to \textbf{False} Harmony reports and error.  Harmony
 also reports an error if the expression evaluates to a value other
-than \texttt{False} or \texttt{True}.
+than \textbf{False} or \textbf{True}.
 
@@
 XXX\=XXX\=XXX\=XXX\=XXX\=XXX\=XXX\=XXX\kill
-\>\textbf{let} $\mathit{waiting} = \mathtt{True}:$ \\
+\>\textbf{let} $\mathit{waiting} = \mathbf{True}:$ \\
 \>\>\textbf{while} \textit{waiting}: \\
@@
 \>\>\>\>\>\>$...$ \\
-\>\>\>\>\>$\mathit{waiting} = \mathtt{False}$
+\>\>\>\>\>$\mathit{waiting} = \mathbf{False}$
 \end{tabbing}
@@
 be handled by the thread.  Such interrupts can be disabled by setting
-the interrupt level of the thread to \texttt{True} using the
+the interrupt level of the thread to \textbf{True} using the
 \textbf{setintlevel} operator.
@@
 The statement \textbf{while} $c: S_1; S_2; ...$ executes statements
-$S_1, S_2, ...$ repeatedly as long as $c$ evaluates to \texttt{True}.
+$S_1, S_2, ...$ repeatedly as long as $c$ evaluates to \textbf{True}.
 Harmony does not support \textbf{break} or \textbf{continue} statements.
@@
 \hline
-\texttt{Lock}() & return a binary semaphore initialized to \texttt{False} \\
+\texttt{Lock}() & return a binary semaphore initialized to \textbf{False} \\
 \hline
@@
 \hline
-Assert, Assert2 & pop $b$ and check that it is \texttt{True}.  Assert2 also pops value to print \\
+Assert, Assert2 & pop $b$ and check that it is \textbf{True}.  Assert2 also pops value to print \\
 \hline
@@
 Similarly for \textbf{or}, where evaluation stops once one of the subexpressions
-evaluates to \texttt{True}.
+evaluates to \textbf{True}.
 A sequence of comparison operations, such as $a < b < c$, is evaluated left to right

### 2021-08-08 book/paper.tex
Latest

@@
 \autoref{fig:leader} describes the protocol and its test cases in Harmony.
-To test the protocol, the code first generates an identifier for each
-processor, chosen non-deterministically from a set of \texttt{NIDS}
-identifiers.  It also keeps track in the variable \textit{leader} of
-what the highest identifier is so it can later be checked.
-
-The network is modeled simply as a set of messages.
+In Harmony, processors can be modeled by threads and there are a variety of ways in
+which one can model a network using shared variables.
+Here, we model the network as a set of messages.
 The \texttt{send} method atomically adds a message to this set.
@@
 destined for the processor with identifier \textit{self}.
+
+To test the protocol, the code first generates an identifier for each
+processor, chosen non-deterministically from a set of \texttt{NIDS}
+identifiers.  It also keeps track in the variable \textit{leader} of
+what the highest identifier is so it can later be checked.
 
@@
 of a server.
-In this chapter, we will \emph{replicate} the object: maintain multiple
-copies.
+While we could use Chain Replication to replicate the register, in
+this chapter we will use a solution that does not assume that
+crashes can be accurately detected.
+
+We will again replicate the register object: maintain multiple copies,
+but we will not use the replicated state machine approach.
 One could, for example, imagine that clients write to all copies and read
@@
 on a single value over a network.
+For example, in state machine replication, the state machines have to agree
+on which operation to apply next.
 Without failures, this can be solved using leader election: first elect a leader, then have that leader decide a value.
@@
 for example in the face of processor failures.
-In Harmony, processors can be modeled by threads and there are a variety of ways in
-which one can model a network using shared variables.
 In this chapter, we will present a simple consensus algorithm that can tolerate
@@
 As in the consensus protocol of \autoref{ch:consensus}, Paxos uses rounds of messaging.
-However, Paxos uses two kinds of rounds: ``Phase 1'' and ``Phase 2'' rounds.
-The communication pattern is also different.  Rounds are identified by a so-called
+The communication pattern, however, is different.
+Similar to the atomic read/write register protocol in \autoref{ch:abd},
+Paxos uses two kinds of rounds: ``Phase 1'' and ``Phase 2'' rounds.
+Rounds are identified by a so-called
 \emph{ballot number} combined with the phase number.
@@
 Leaders broadcast ``Type A''
-messages to the acceptors, which respond point-to-point with ``Type B'' messages.
+messages to the acceptors, which respond point-to-point with ``Type B''
+messages.
 

### 2021-08-08 book/paper.tex
Latest

@@
 \index{state machine replication}%
+\index{determinism}%
 Besides availability, it is usually important that the replicated
-services acts as if it were a single one.  This requires that the
+service acts as if it were a single one.  This requires that the
 replicas of the service coordinate their actions.
@@
 same requests in the same order and to do so in a way that continues
-to work even if some subset of replicas crash.
+to work even if some strict subset of replicas crash.
 
@@
 In Chain Replication, the replicas are organized in a linear chain.
-Each replica monitors its direct predecessor in the chain.  If it has
+Each replica monitors its direct predecessor in the chain.  If a replica has
 no predecessors, we call it the \emph{head}.  The head may change over
-time, as a head may crash and replaced by another.
+time, as a head may crash and replaced by another, but at any point in
+time there is only one head
+(and only one \emph{tail}, possibly the same replica).
 We model the state machine as a \emph{history}: a sequence of operations.
@@
 advertises the new update.  When the update is applied to the history
-of the \emph{tail} replica (the replica that has no live successors),
+of the tail replica (the replica that has no live successors),
 then the update is considered final.
@@
 some maximum amount of time, then the server considers its peer crashed.
-Note that this, in general is not a safe thing to do---the network or
+Note that this, in general, is not a safe thing to do---the network or
 the peer may be temporarily slow but the peer is not necessarily crashed
@@
 Nonetheless, we will assume here that failure detection does not make
-mistakes and that eventually every failure is detected.
-This is sometimes called the \emph{Fail-Stop} failure model~\cite{SS83},
+mistakes and that eventually every failure is eventually detected.
+This is called the \emph{Fail-Stop} failure model~\cite{SS83},
 which is distinct from the often more realistic \emph{Crash} failure model
@@
 The code starts \texttt{NREPLICAS} \texttt{replica} threads numbered
-0 through $\mathtt{NREPLICAS} - 1$.  At least one of the replicas 
+0 through $\mathtt{NREPLICAS} - 1$, with initially 0 being the head
+and $\mathtt{NREPLICAS} - 1$ being the tail.  At least one of the replicas 
 must be immortal---the code selects one of them although none of the
 replicas is required to fail during execution.
-In practice, of course, a replica does not know whether it is immortal
-or not---it should just assume so.  The immortality is only used for
-modeling the system.
+Of course, a replica does not know whether it is immortal
+or not in practice---it should just assume so.
+The immortality of one of the replicas is only used for modeling the system.
 
@@
 $\mathbf{choose}(\{ \mathit{immortal}, \textbf{True} \})$ can only
-evaluate to \textbf{False} in case $\textit{immortal}$ is false.
+evaluate to \textbf{False} in case $\textit{immortal}$ is false; otherwise
+it will always evaluate to \textbf{True}.
 What it does within the loop depends on whether it is the head
@@
 Because failure detection is accurate, at most one replica can think
-it is the head at any time.  Moreover, when it has
+it is the head at any time (and, if so, it is in fact the head).
+Moreover, when it has
 detected all its predecessors having failed, eventually some replica
@@
 crashed.  The code models a replica crashing by that replica broadcasting
-this message on the network.  The other form of payload is a tuple
+a \texttt{.crash} message on the network.
+The other form of payload is a tuple
 $(\mathit{index}, \mathit{operation})$, which means that the source
@@
 We are not modeling clients of the service.
-Instead, in the model, we add \textit{self} to the history until the
-history have \texttt{NOPS} operations on it or until the head crashes,
-whichever comes first.  The head also broadcasts this update, which
-is intended for its successor.  Because of failures, it may be that
-more than one replica is a successor of the head at different times.
-
-Otherwise, the replica awaits a message from its direct predecessor.
+Instead, in the model, the head adds its identifier \textit{self} to the
+history until the
+history has \texttt{NOPS} operations on it or until the head crashes,
+whichever comes first.  The head also broadcasts each update, which
+are intended for its successor.  Because of failures, it may be that
+at different times a replica has different successors, hence the
+broadcast.
+
+Otherwise, the replica does not believe it is the head and
+awaits a message from its direct predecessor.
 If the payload is \texttt{.crash}, then the replica removes this

### 2021-08-08 book/paper.tex
Improved the leader election tests

@@
 
-To test the protocol, the code first generates an identifier for each
+To test the protocol, the code first chooses the number of processors
+and generates an identifier for each
 processor, chosen non-deterministically from a set of \texttt{NIDS}

### 2021-08-08 book/paper.tex
Added Needham/Schroeder to the book

@@
 Paxos consensus protocol.
+\item \autoref{ch:ns} demonstrates finding a bug in the original Needham-Schroeder
+authentication protocol.
 \end{itemize}
@@
 
+\chapter{Needham-Schroeder Authentication Protocol}
+\label{ch:ns}
+
+\begin{figure}
+\begin{code}
+\harmonysource{needhamschroeder}
+\end{code}
+\caption{\harmonylink{code/needhamschroeder.hny} Needham-Schroeder protocol and an attack.}
+\label{fig:ns}
+\end{figure}
+
+The Needham-Schroeder protocol~\cite{NS78} is a security protocol in which two parties
+authenticate one another by exchanging large and recently created random numbers
+called \emph{nonces} that nobody else should be able to read.
+The version we describe here uses \emph{public key cryptography}~\cite{pkc}: with public key
+cryptography it is possible to create a message for a particular destination that
+only that destination can read.  We denote with $\langle m \rangle_p$ a message
+$m$ encrypted for $p$ so that only $p$ can decrypt the message and see that it contains $m$..
+
+The original protocol goes something like this.  Suppose Alice wants to communicate
+with Bob.  The three steps are as follows:
+
+\begin{enumerate}
+\item Alice creates a new nonce $N_A$ and sends $\langle 1, A, N_A \rangle_\mathtt{Bob}$
+to Bob;
+\item Upon receipt, Bob creates a new nonce $N_B$ and sends
+$\langle 2, N_A, N_B \rangle_\mathtt{Alice}$ to Alice;
+\item Alice sends $\langle 3, N_B \rangle_\mathtt{Bob}$ to Bob.
+\end{enumerate}
+
+When Bob receives $N_B$ in the last step, it decides that it is Alice that is
+trying to communicate at this time.  Since it created $N_B$ recently and sent it
+encrypted to Alice, Bob does not have to worry that the message was an old message
+that was replayed.  Also, if Alice is honest, only Alice can have seen the message.
+
+\autoref{fig:ns} shows the protocol implemented in Harmony.
+A message $\langle m \rangle_p$ is encoded in Harmony as a dictionary
+$\{ \mathtt{.dst}: p, \mathtt{.contents}: m \}$.
+
+Unfortunately, the protocol turns out to be incorrect, but it took 17 years before
+somebody noticed~\cite{Lowe95}.  The bug is easily found using model
+checking~\cite{Lowe96}.
+
 \bibliographystyle{alpha}

### 2021-08-08 book/paper.tex
Latest

@@
 called \emph{nonces} that nobody else should be able to read.
-The version we describe here uses \emph{public key cryptography}~\cite{pkc}: with public key
+The version we describe here uses \emph{public key cryptography}~\cite{DH76}: with public key
 cryptography it is possible to create a message for a particular destination that
 only that destination can read.  We denote with $\langle m \rangle_p$ a message
-$m$ encrypted for $p$ so that only $p$ can decrypt the message and see that it contains $m$..
-
-The original protocol goes something like this.  Suppose Alice wants to communicate
-with Bob.  The three steps are as follows:
+$m$ encrypted for $p$ so that only $p$ can decrypt the message and see that it contains $m$.
+
+Suppose Alice wants to communicate with Bob.  The three critical steps in the
+Needham-Schroeder protocol are as follows:
 
@@
 When Bob receives $N_B$ in the last step, it decides that it is Alice that is
-trying to communicate at this time.  Since it created $N_B$ recently and sent it
+trying to communicate at this time.  Since Bob created $N_B$ recently and sent it
 encrypted to Alice, Bob does not have to worry that the message was an old message
-that was replayed.  Also, if Alice is honest, only Alice can have seen the message.
+that was replayed.  Also, if Alice is honest, only Alice can have seen the message
+containing $N_B$.
 
@@
 A message $\langle m \rangle_p$ is encoded in Harmony as a dictionary
-$\{ \mathtt{.dst}: p, \mathtt{.contents}: m \}$.
+$\{ \mathtt{.dst}: p, \mathtt{.contents}: m \}$.  The code for Alice and Bob
+simply follows the steps listed above.
 
 Unfortunately, the protocol turns out to be incorrect, but it took 17 years before
-somebody noticed~\cite{Lowe95}.  The bug is easily found using model
-checking~\cite{Lowe96}.
+somebody noticed~\cite{Lowe95}.  Model checking can be used to find the bug~\cite{Lowe96}.
+To demonstate the bug, we need a third party, which we will call Carter.
+We want to make sure that Carter cannot impersonate Alice or Bob.
+However, it is possible that Alice tries to set up an authenticated connection to Carter
+using the Needham-Schroeder protocol.  That in itself is not a problem.
+
+The code in \autoref{fig:ns} has Alice try to set up a connection to either Bob
+or Carter.  Bob only accepts connections with Alice.  Carter, when receiving a message
+that it can decrypt, will try to find an attack by sending every possible message to
+every possible destination.  In particular, it keeps track of every nonce that it has
+seen and will try to construct messages with them to send to Alice and Bob.
+If, at some point, Carter finds the nonce that Bob generates for Alice, then Carter
+can impersonate Alice by sending a type~3 message containing that nonce to Bob.
+
+Running the code in \autoref{fig:ns} quickly finds a viable attack.  The attack goes
+like this:
+
+\begin{enumerate}
+\item Alice creates a new nonce $N_A$ and sends $\langle 1, A, N_A \rangle_\mathtt{Carter}$
+to Carter;
+\item Upon receipt, Carter sends $\langle 1, A, N_A \rangle_\mathtt{Bob}$ to Bob;
+\item Upon receipt, Bob creates a new nonce $N_B$ and sends
+$\langle 2, N_A, N_B \rangle_\mathtt{Alice}$ to Alice;
+\item Alice thinks the message came from Carter (because it contains $N_A$, which
+Alice sent to Carter) and
+sends $\langle 3, N_B \rangle_\mathtt{Carter}$ to Carter.
+\item Carter learns $N_B$, which it was not supposed to.
+\end{enumerate}
+
+Carter can use $N_B$ to send $\langle 3, N_B \rangle_\mathtt{Bob}$ to Bob.
+Bob now thinks it is talking to Alice, even though Alice never tried to
+communicate with Bob.
 

### 2021-08-09 book/paper.tex
Latest
Improved Bosco code

@@
 \end{code}
-\caption{\harmonylink{code/chain.hny} Checking the correctness of Chain Replication.}
+\caption{\harmonylink{code/chain.hny} Checking correctness of Chain Replication.}
 \label{fig:chain}
@@
 shared list of proposals and a shared set of decisions.
+To reduce the state space to explore, not all permutations of zeroes and ones are
+explored.  With 5 processors ($\mathtt{F} = 2)$, say, we only explore the cases where
+no processors propose 1, where exactly one processors proposes 1, and where 2 processors
+proposes 1.
 In this particular algorithm, all messages are broadcast to all processors, so they do
@@
 
-In Line~21, a processor waits for $\texttt{N} - \texttt{F}$ messages using
+In Line~22, a processor waits for $\texttt{N} - \texttt{F}$ messages using
 the Harmony \textbf{select} statement.
@@
 
-While one can run this code within little time for $\mathtt{F} = 1$, for
-$\mathtt{F} = 2$ the state space to explore is already quite large.
+While one can run this code within little time for $\mathtt{F} = 2$, for
+$\mathtt{F} = 3$ the state space to explore is already quite large.
 One way to reduce the state space to explore is the following realization:
@@
 \autoref{fig:bosco2} shows the code for this model.  Running this with
-$\mathtt{F} = 2$ does not take very long and this approach is a good
+$\mathtt{F} = 3$ does not take very long and this approach is a good
 blueprint for testing other round-based protocols (of which there are

### 2021-08-09 book/paper.tex
Renamed Carter to Corey

@@
 somebody noticed~\cite{Lowe95}.  Model checking can be used to find the bug~\cite{Lowe96}.
-To demonstate the bug, we need a third party, which we will call Carter.
-We want to make sure that Carter cannot impersonate Alice or Bob.
-However, it is possible that Alice tries to set up an authenticated connection to Carter
+To demonstate the bug, we need a third party, which we will call Corey.
+We want to make sure that Corey cannot impersonate Alice or Bob.
+However, it is possible that Alice tries to set up an authenticated connection to Corey
 using the Needham-Schroeder protocol.  That in itself is not a problem.
@@
 The code in \autoref{fig:ns} has Alice try to set up a connection to either Bob
-or Carter.  Bob only accepts connections with Alice.  Carter, when receiving a message
+or Corey.  Bob only accepts connections with Alice.  Corey, when receiving a message
 that it can decrypt, will try to find an attack by sending every possible message to
@@
 seen and will try to construct messages with them to send to Alice and Bob.
-If, at some point, Carter finds the nonce that Bob generates for Alice, then Carter
+If, at some point, Corey finds the nonce that Bob generates for Alice, then Corey
 can impersonate Alice by sending a type~3 message containing that nonce to Bob.
@@
 \begin{enumerate}
-\item Alice creates a new nonce $N_A$ and sends $\langle 1, A, N_A \rangle_\mathtt{Carter}$
-to Carter;
-\item Upon receipt, Carter sends $\langle 1, A, N_A \rangle_\mathtt{Bob}$ to Bob;
+\item Alice creates a new nonce $N_A$ and sends $\langle 1, A, N_A \rangle_\mathtt{Corey}$
+to Corey;
+\item Upon receipt, Corey sends $\langle 1, A, N_A \rangle_\mathtt{Bob}$ to Bob;
 \item Upon receipt, Bob creates a new nonce $N_B$ and sends
 $\langle 2, N_A, N_B \rangle_\mathtt{Alice}$ to Alice;
-\item Alice thinks the message came from Carter (because it contains $N_A$, which
-Alice sent to Carter) and
-sends $\langle 3, N_B \rangle_\mathtt{Carter}$ to Carter.
-\item Carter learns $N_B$, which it was not supposed to.
+\item Alice thinks the message came from Corey (because it contains $N_A$, which
+Alice sent to Corey) and
+sends $\langle 3, N_B \rangle_\mathtt{Corey}$ to Corey.
+\item Corey learns $N_B$, which it was not supposed to.
 \end{enumerate}
 
-Carter can use $N_B$ to send $\langle 3, N_B \rangle_\mathtt{Bob}$ to Bob.
+Corey can use $N_B$ to send $\langle 3, N_B \rangle_\mathtt{Bob}$ to Bob.
 Bob now thinks it is talking to Alice, even though Alice never tried to

### 2021-08-09 book/paper.tex
Latest

@@
 called \emph{nonces} that nobody else should be able to read.
+The nonces should only be used once for an instantiation of the protocol between
+honest participants.
 The version we describe here uses \emph{public key cryptography}~\cite{DH76}: with public key
@@
 
-When Bob receives $N_B$ in the last step, it decides that it is Alice that is
-trying to communicate at this time.  Since Bob created $N_B$ recently and sent it
-encrypted to Alice, Bob does not have to worry that the message was an old message
-that was replayed.  Also, if Alice is honest, only Alice can have seen the message
-containing $N_B$.
+When Bob receives $\langle 1, A, N_A \rangle_\mathtt{Bob}$, Bob does not know for sure
+that the message came from Alice, and even if it came from Alice, it does not know if
+Alice sent the message recently or if it was replayed by some adversary.
+When Alice receives $\langle 2, N_A, N_B \rangle_\mathtt{Alice}$, Alice \emph{does}
+know that, if Bob is honest, (1) Bob and only Bob could have created this message, and
+(2) Bob must have done so recently (since Alice created $N_A$).
+When Bob receives $\langle 3, N_B \rangle_\mathtt{Bob}$, Bob decides that it is Alice
+that is trying to communicate at this time.  Since Bob created $N_B$ recently and sent it
+encrypted to Alice, Bob does not have to worry that the type 3 message was an old message
+that was replayed by some adversary.
+Also, if Alice is honest, only Alice can have seen the message containing $N_B$.
+
+Thus, the intended security property of this protocol is that if Alice and Bob are both
+honest, only Alice and Bob learn $N_A$ and $N_B$ and only if they are trying to communicate
+with one another.
 
@@
 \item Upon receipt, Corey sends $\langle 1, A, N_A \rangle_\mathtt{Bob}$ to Bob;
-\item Upon receipt, Bob creates a new nonce $N_B$ and sends
+\item Upon receipt, Bob, as usual, creates a new nonce $N_B$ and sends
 $\langle 2, N_A, N_B \rangle_\mathtt{Alice}$ to Alice;
 \item Alice thinks the message came from Corey (because it contains $N_A$, which
-Alice sent to Corey) and
+Alice created for Corey and sent to Corey) and
 sends $\langle 3, N_B \rangle_\mathtt{Corey}$ to Corey.
-\item Corey learns $N_B$, which it was not supposed to.
+\item Corey learns $N_B$, which it was not supposed to as Bob is only willing to
+talk to Alice and created $N_B$ explicitly for Alice.
 \end{enumerate}
@@
 Corey can use $N_B$ to send $\langle 3, N_B \rangle_\mathtt{Bob}$ to Bob.
-Bob now thinks it is talking to Alice, even though Alice never tried to
-communicate with Bob.
+This is identical to the situation in which Alice tried to set up a connection to Bob,
+so Bob now thinks it is talking to Alice, even though Alice never tried to
+communicate with Bob.  The security property is violated because Corey discovers $N_B$
+even though Alice and Bob are both honest and are not trying to communicate with one
+another.  As a result, Carter has successfully impersonated Alice to Bob.
 

### 2021-08-11 book/sources/pp.py
Working on implementation of "possibly" language feature

@@ -90,6 +90,7 @@ def isreserved(s):
         "not",
         "or",
         "pass",
+        "possibly",
         "select",
         "sequential",
         "setintlevel",

### 2021-08-11 book/paper.tex
Working on implementation of "possibly" language feature
Added possibly to the book

@@
 
+Note the use of the \textbf{possibly} statement in this code.
+It is similar to \textbf{assert}, but where \texttt{assert} checks that
+the predicate holds in \emph{all} executions, \texttt{possibly} checks
+that the predicate holds in at least one execution.
+The purpose here is to check that the code doesn't simply always abort---there
+are executions where the code commits.
+In practice, hopefully \emph{most} executions commit, but this is not something
+the Harmony model checker can check.
+
 The \texttt{check}() method checks if the sum of the balances equals
@@
 The nonces should only be used once for an instantiation of the protocol between
-honest participants.
+honest participants (i.e., participants that follow the protocol).
 The version we describe here uses \emph{public key cryptography}~\cite{DH76}: with public key
@@
 \texttt{pass} & do nothing\\
+\hline
+\texttt{possibly $b$ [, $b$, ...]} & each $b$ is a boolean \\
 \hline
@@
 The \textbf{pass} statement does nothing.
+
+\section{\textbf{possibly}}
+
+The statement \textbf{possibly $b_1, b_2, ...$} evaluates each predicate $b_i$
+and checks if it holds in at least one execution.
+At completion, Harmony reports which of the predicates never held.
 

### 2021-08-12 book/paper.tex
improved
Latest

@@
 that was replayed by some adversary.
-Also, if Alice is honest, only Alice can have seen the message containing $N_B$.
-
-Thus, the intended security property of this protocol is that if Alice and Bob are both
-honest, only Alice and Bob learn $N_A$ and $N_B$ and only if they are trying to communicate
-with one another.
+Also, if Alice is honest, it seems only Alice can have seen the message containing $N_B$.
+
+Thus, the intended security properties of this protocol are symmetric.
+Assuming Alice and Bob are both honest:
+\begin{itemize}
+\item if Alice finishes the protocol with Bob and received $B_N$ from Bob, then
+nobody but Alice and Bob can learn $N_B$.
+\item if Bob finishes the protocol with Alice and received $A_N$ from Alice, then
+nobody but Bob and Alice can learn $N_A$.
+\end{itemize}
+
+After the protocol, Alice can include $N_A$ in messages to Bob and Bob can
+include $N_B$ in messages to Alice to authenticate the sources of those messages
+to one another.
 
@@
 somebody noticed~\cite{Lowe95}.  Model checking can be used to find the bug~\cite{Lowe96}.
-To demonstate the bug, we need a third party, which we will call Corey.
+To demonstate the bug, we need to model the environment.
+In particular, we introduce a third party, which we will call Corey.
 We want to make sure that Corey cannot impersonate Alice or Bob.
 However, it is possible that Alice tries to set up an authenticated connection to Corey
-using the Needham-Schroeder protocol.  That in itself is not a problem.
-
-The code in \autoref{fig:ns} has Alice try to set up a connection to either Bob
+using the Needham-Schroeder protocol.  That in itself should not be a problem if the
+protocol were correct.
+
+The code in \autoref{fig:ns} has Alice either not do anything, or has Alice try to set
+up a connection to either Bob
 or Corey.  Bob only accepts connections with Alice.  Corey, when receiving a message
@@
 seen and will try to construct messages with them to send to Alice and Bob.
-If, at some point, Corey finds the nonce that Bob generates for Alice, then Corey
-can impersonate Alice by sending a type~3 message containing that nonce to Bob.
+If Bob finishes the protocol, it checks to see if Alice actually tried to connect
+to Bob.  If not, the assertion fails and an attack is found.
 
@@
 \item Upon receipt, Corey sends $\langle 1, A, N_A \rangle_\mathtt{Bob}$ to Bob;
-\item Upon receipt, Bob, as usual, creates a new nonce $N_B$ and sends
+\item Upon receipt, Bob, believing it is engaging in the protocol with Alice,
+creates a new nonce $N_B$ and sends
 $\langle 2, N_A, N_B \rangle_\mathtt{Alice}$ to Alice;
@@
 sends $\langle 3, N_B \rangle_\mathtt{Corey}$ to Corey.
-\item Corey learns $N_B$, which it was not supposed to as Bob is only willing to
-talk to Alice and created $N_B$ explicitly for Alice.
+\item Corey learns $N_B$ and sends $\langle 3, N_B \rangle_\mathtt{Bob}$ to Bob.
+\item Bob receiving $\langle 3, N_B \rangle_\mathtt{Bob}$ is identical to the
+situation in which Alice tried to set up a connection to Bob,
+so Bob now thinks it is talking to Alice, even though Alice never tried to
+communicate with Bob.
 \end{enumerate}
-
-Corey can use $N_B$ to send $\langle 3, N_B \rangle_\mathtt{Bob}$ to Bob.
-This is identical to the situation in which Alice tried to set up a connection to Bob,
-so Bob now thinks it is talking to Alice, even though Alice never tried to
-communicate with Bob.  The security property is violated because Corey discovers $N_B$
-even though Alice and Bob are both honest and are not trying to communicate with one
-another.  As a result, Carter has successfully impersonated Alice to Bob.
+The security property is violated.  In particular, Bob, duped by Corey, finished
+the protocol with Alice and received $A_N$, and even though Bob and Alice
+are both honest, Corey has a copy of $A_N$.  So Corey is now able to impersonate
+Alice to Bob (but not vice versa because Alice did not try to authenticate Bob).
 
@@
 
-The statement \textbf{possibly $b_1, b_2, ...$} evaluates each predicate $b_i$
-and checks if it holds in at least one execution.
+The statement \textbf{possibly $b_1, b_2, ...$} atomically evaluates all
+predicates $b_i$.
 At completion, Harmony reports which of the predicates never held.
@@
 Pop & pop a value of the stack and discard it \\
+\hline
+Probably $i$ & pop a value to check the $i^{th}$ predicate in a \textbf{probably} statement \\
 \hline
@@
 my colleague Fred Schneider.  He suggested I write this book after
-demonstrating Harmony to him.
+demonstrating Harmony to him.  Being a foremost security expert, he also
+assisted significantly with the chapter on the Needham-Schroeder protocol.
 

### 2021-08-12 book/paper.tex
Latest

@@
 that the predicate holds in at least one execution.
-The purpose here is to check that the code doesn't simply always abort---there
-are executions where the code commits.
-In practice, hopefully \emph{most} executions commit, but this is not something
-the Harmony model checker can check.
+The purpose is to check that the code doesn't simply always abort---there
+exist executions where the code commits.
+In practice, hopefully \emph{most} executions commit, but Harmony is not
+capable of probabilistic reasoning.
 
@@
 After that, proceed with the next round.
+
+The \texttt{possibly} statement checks if there exist executions where 0 is
+decided and if there exist executions where 1 is decided.  Doing so is useful
+because the assertions would certainly not fail if no decision is ever made.
+Also, it is nice to know that both 0 and 1 can be decided.
 
@@
 \hline
-Probably $i$ & pop a value to check the $i^{th}$ predicate in a \textbf{probably} statement \\
+Possibly $i$ & pop a value to check the $i^{th}$ predicate in a \textbf{possibly} statement \\
 \hline

### 2021-08-12 book/paper.tex
Latest
Better

@@
 
-This would fix the issue, but more likely it is the program that must
-be fixed, not the specification.
+This would fix the issue,\footnote{Actually, Harmony still complains, this time
+about a \emph{data race}, about which you will learn in \autoref{ch:harmonymachine}.}
+but more likely it is the program that must be fixed, not the specification.
 
@@
 \begin{problems}
-\item While Harmony checks that the Validity and Agreement properties
-are not violated, it does not check that the algorithm ever decides.
-Indeed, in unlucky executions the algorithm may never decide.  But
-how would you check that there exist executions in which the algorithm
-does decide?
 \item The algorithm as given works in the face of crash failures.

### 2021-08-14 book/paper.tex
Latest

@@
 This book is intended to help people with understanding and
-developing concurrent code.  In particular, it uses a new tool
+developing concurrent code, which includes programs for distributed
+systems.  In particular, it uses a new tool
 called Harmony that helps with \emph{testing} concurrent code.
-The approach is based on \emph{model checking}:
+The approach is based on \emph{model checking}~\cite{CES86}:
 \index{model checking}%
@@
 Harmony is a concurrent programming language.  It was designed to teach
-the basics of concurrent programming, but it is also useful for
+the basics of concurrent and distributed programming, but it is also useful for
 testing new concurrent algorithms or even sequential and distributed
@@
 properties and does not suffer from bugs.
+Interestingly, Harmony does not have input/output statements, so there is
+no ``Hello World'' program.
 
@@
 
+The appendices contain various details about Harmony itself, including
+an appendix that explains how Harmony works (\autoref{app:howitworks}) and
+an appendix on installing and running Harmony (\autoref{app:install}).
+
 \chapter{Introduction to Harmony}
@@
 
-\chapter{The Harmony Virtual Machine}
-
-The Harmony Virtual Machine (HVM) has the following state:
+\chapter{The Harmony Virtual Machine}\label{app:hvm}
+
+The Harmony Virtual Machine (HVM, \autoref{ch:harmonymachine})
+has the following state:
 
@@
 of the form \texttt{(* comment *)}.
+
+\chapter{How Harmony Works}\label{app:howitworks}
+
+This appendix gives a very brief overview of how Harmony works.
+In a nutshell, Harmony goes through the following three phases:
+
+\begin{enumerate}
+\item The Harmony \emph{compiler} turns your Harmony program into
+bytecode.  A recursive descent parser and code generator written in
+Python (see \texttt{harmony.py}) turns an \texttt{x.hny} program
+into \texttt{x.hvm}, a JSON file containing the corresponding
+bytecode.
+\item The Harmony \emph{model checker} evaluates the state space
+that the program (now in bytecode) can generate.  The model checker
+is written in C as it needs to be highly efficient (see \texttt{charm.c}).
+The model checker starts from the initial state, and then, iteratively,
+checks for each state that it has found what next steps are possible and
+generates the next states using the Harmony virtual machine
+(\autoref{app:hvm}).  If the model is finite, eventually the model
+checker will generate a graph with all possible states.
+If there is a problematic path in this graph (see below), then it will
+report the shortest such path in the \texttt{x.hco} output file in JSON
+format.
+\item The \texttt{x.hco} output file is translated twice by
+\texttt{harmony.py}.  There is a so-called \emph{brief output} that
+is written to standard output.  The more comprehensive output is placed
+in the \texttt{x.htm} HTML output file, allowing you to navigate the
+problematic path and all the details of each of the states on the path.
+\end{enumerate}
+
+\section{Compiler}
+
+The Harmony compiler, in order to stay true to the Harmony source program,
+does not do much in the way of optimizations.  The main optimizations that
+it does are:
+
+\begin{itemize}
+\item Constant folding: (simple) expressions consisting only of constants
+are evaluated by the compiler rather than by the model checker;
+\item Jump threading: Harmony eliminates jump to jump instructions;
+\item Dead variable elimination: Harmony removes method variables that
+are no longer in use from the state in order to reduce the state space
+to be explored.
+\end{itemize}
+
+\section{Model Checker}
+
+The Harmony model checker, called \emph{Charm}, takes the output from
+the compiler and explores the entire state space in breadth-first order.
+Even Harmony does not really support input, there are three sources of
+non-determinism that make this exploration non-trivial:
+
+\begin{itemize}
+\item \textbf{choose} \emph{expressions}: Harmony's ability to let the program
+choose a value from a set;
+\item \emph{thread interleaving}: different threads run pseudo-concurrently
+with their instructions interleaved in arbitrary ways;
+\item \emph{interrupts}: Harmony programs can set interrupts that can go
+off at arbitrary times.
+\end{itemize}
+
+Charm has some tricks to significantly reduce the state space to explore.
+\begin{itemize}
+\item A thread can have local state (program counter, stack,
+method variables, and thread-local state variables).
+That state is called the \emph{context} of the thread.
+The context of a thread cannot be accessed by other
+threads, nor by \textbf{invariant} statements.
+So the model checker only interleaves threads at \textbf{Load} and
+\textbf{Store} instructions where a thread interacts with global
+variables.
+\item Threads are anonymous, and therefore two or more threads can
+have the same context.  The state of the model checker therefore maintains
+a \emph{bag} (multiset) of contexts rather are than a \emph{set} of contexts.
+Thus even if there are hundreds of threads, there may be only tens of
+possible context states.
+\end{itemize}
+
+That said, \emph{state space explosion} is still a possibility, and
+Harmony programmers should keep this in mind when writing and testing
+their programs.  Do not be too ambitious: start with small tests and
+gradually build them up as necessary.
+
+The model checker stops either when it finds a failing execution or when
+it has explored the entire state space, whichever comes first.  An execution
+can fail for a variety of reasons:
+
+\begin{itemize}
+\item An invariant failing: Harmony evaluates all invariants in all states
+that if finds---if one fails, Harmony stops further exploration;
+\item An assertion failing;
+\item A \emph{silly} error: this includes reading variables that have not
+been assigned, trying to add a set to an integer, taking the length of
+something that is not a set of a dictionary, and so on;
+\item An infinite loop: a thread goes into an infinite loop without
+accessing shared variables.
+\end{itemize}
+
+\section{Model Checker Output Analysis}
+
+The output of the model checker is a graph that is typically very large.
+If some execution failed, then Harmony will simply report the path of
+that failing execution.  But otherwise there may be the following
+outcomes:
+
+\begin{itemize}
+\item No issues: no failing executions and each program can terminate;
+\item Non-terminating states: some executions lead to some form of deadlock
+or other issue that causes some (non-eternal) threads not to be able to
+terminate;
+\item Race conditions: there are executions in which two threads access
+the same shared state variable, with at least one of those accesses being a
+\textbf{Store} operation;
+\item Busy waiting: executions in which threads are actively waiting for
+some condition, usually by releasing and reacquiring locks.
+\end{itemize}
+
+\noindent
+In order to diagnose these outcomes, Harmony must analyze the graph.
+
+The first thing that Harmony does is to locate non-terminating
+states, if any.  To do this, Harmony first determines the \emph{strongly
+connected components} of the graph using Kosaraju's algorithm.
+A component (subgraph) of a graph is strongly connected if each
+vertex (state) in the component can be reached from each other vertex.
+The components then form a Directed Acyclic Graph (DAG).
+The DAG is easier to analyze than the original graph.
+One can easily determine the sink components
+(the components with no outgoing edges).  If such a component
+has non-eternal threads in it, then each state in that component
+is a non-terminating state.
+
+To find race conditions, the model checker looks in the graph for
+states in which there are multiple threads that can make a step.
+If there is a step in which multiple threads access the same shared
+variable, at least one of those accesses is a store operation, and
+at least one of those threads is not in atomic mode, then Harmony
+reports the shortest path to such a state.
+
+To show how Harmony detects busy waiting, we will first show how
+Harmony determines if a thread is blocked or not.  A thread is considered
+blocked if it cannot terminate without the help of another thread.
+For example, a thread waiting for a lock is blocked and cannot terminate
+until another thread releases the lock.  Determining whether a thread
+is blocked in a particular state can be done within the confines of the
+connected component: the analyzer tries all possible executions of the
+thread.  If it cannot ``escape'' the connected component by doing so, it
+is considered blocked.
+A thread is considered \emph{busy waiting} if it is blocked, but it is
+also changing the shared state while doing so.  A thread that is waiting
+on a spinlock only observes the state.
+
+Finally, in the output, each thread has a unique identifier:
+\texttt{T0} is the initialization thread; \texttt{T}$n$ is the $n^{th}$
+spawned thread that executes.  This seems to contradict the fact that
+Harmony threads are anonymous.  The output analyzer assigns these
+identifiers \emph{a posteriori} to the threads in the state graph
+by keeping track, along the reported execution path, what state each
+thread is in.  So by examining the initial context of the thread that
+is running from some particular state, it can determine if that context
+corresponds to the current context of some thread that ran previously
+or if the context belongs to a new thread that has not run before.
+
+\chapter{Installing and Running Harmony}\label{app:install}
+
+There are currently three options for using Harmony:
+
+\begin{enumerate}
+\item You can download and install a command-line version of Harmony on your
+computer;
+\item You can download and install \emph{HarmonyLang}, a VSCode plug-in for Harmony;
+\item You can run Harmony programs in the cloud using \emph{Harmony IDE}.
+\end{enumerate}
+
+Below you can find more information on each of these options.
+
+\section{Command-Line}
+
+You can get the latest released developer version of Harmony by
+downloading from \url{harmony.cs.cornell.edu}.  It includes the latest
+bug fixes and features at any time.  Harmony is developed and maintained
+by the author on both MacOSX and Linux, and so these are the preferred
+platforms at this time.  A Windows version is also available, but may not
+be the latest version and is currently not as well supported.
+
+When you download the \texttt{.zip} file, you will get the following
+files:
+
+\begin{itemize}
+\item \texttt{README.txt}: installation documentation;
+\item \texttt{archive.xml}: portable compressed archive of the code base;
+\item \texttt{install.py}: Python3 program to install and update the code base.
+\end{itemize}
+
+First place this directory (folder) where you would like it---the
+\texttt{Downloads} folder is at best a good temporary place.  You can put
+the directory in your home directory, for example.
+
+Installation requires Python3 and a 64-bit C compiler.  The developer uses
+recent versions of \texttt{gcc} and \texttt{clang} for development.
+Run \texttt{python3 install.py} to install the code base.
+It extracts files from archive.xml and installs them in the current
+directory.
+It will try to compile the model checker using \texttt{gcc}.
+If that fails under Windows, it will install an executable that
+is pre-compiled on a Windows 10 machine.  If these options do not work,
+you can compile the file \texttt{charm.c} by hand using a 64-bit C
+compile and place the output in \texttt{charm.exe}.
+
+After installation, there is a \texttt{harmony} file for use under
+MacOSX and Linux, and a \texttt{harmony.bat} file for use under Windows.
+You should be able to run \texttt{harmony --help} in any of these
+environments.
+
+If you would like to run \texttt{harmony} from any directory, you have to
+add the current directory to your search path.  Under MacOSX and Linux,
+you will have to set the \texttt{PATH} environment variable.
+See \url{http://www.linux-migration.org/ch02s06.html} for more information.
+Under Windows, search "Edit environment variables" in the search bar.
+You can add the directory either to the \texttt{Path} associated with
+your account or to the system \texttt{Path}.
+If you do not install \texttt{harmony} in your search path, you may have
+to run \texttt{./harmony} in the installation directory instead of
+just \texttt{harmony}.
+
+The installation directory will have the following subdirectories:
+
+\begin{itemize}
+\item \texttt{code}: contains all the code examples from this book;
+\item \texttt{modules}: contains the Harmony modules;
+\item \texttt{python}: contains the Python examples from the book.
+\end{itemize}
+
+For example, you can try \texttt{harmony code/Diners.hny} to run the
+Dining Philosophers code.
+Harmony currently produces three output files:
+\begin{itemize}
+\item \texttt{code/Diners.hvm}: the bytecode in JSON format;
+\item \texttt{code/Diners.hco}: the output of the model checker in JSON format;
+\item \texttt{code/Diners.htm}: the model checker output converted to HTML format.
+\end{itemize}
+
+You are probably only interested to see the last one, which you should be
+able to view in any web browser of your choice, including Safari, Chrome, Edge, or Firefox.
+
+\section{HarmonyLang for VSCode}
+
+The VSCode plug-in, developed by Kevin Sun and Anthony Yang,
+is available from the VSCode market place: \url{https://marketplace.visualstudio.com/items?itemName=kevinsun-dev-cornell.harmonylang}.
+This page comes with installation documentation.  VSCode also sports a
+wonderful animator for Harmony output.
+
+HarmonyLang is regularly updated to include the latest Harmony distribution.
+
+\section{Harmony IDE for the cloud}
+
+\texttt{Harmony IDE} is an Integrated Development Environment for Harmony,
+also developed by Kevin Sun and Anthony Yang,
+that runs in the cloud.  An important advantage of using \texttt{Harmony IDE}
+is that you do not have to install anything.  You can even run in on your
+smartphone.  However, since Harmony programs can use significant compute
+resources, \texttt{Harmony IDE} must put a limit on the amount of compute
+resources you can use.  So while \texttt{Harmony IDE} is currently a great
+way to try out Harmony, if you become a serious user of Harmony you
+will probably want to install either the command-line version or
+\texttt{HarmonyLang}.
 

### 2021-08-14 book/paper.tex
Latest

@@
 
+The model checker also counts the number of times each \textbf{Possibly}
+instruction is invoked with the value \textbf{True}.  It reports those
+\textbf{Possibly} instructions that either have never been reached or
+never have been executed with the value \textbf{True}.
+
 \section{Model Checker Output Analysis}
@@
 able to view in any web browser of your choice, including Safari, Chrome, Edge, or Firefox.
+
+You can see if you are running the latest version of Harmony at any
+time by running \texttt{Python3 install.py --check}.  If you would like
+to update your installation, run \texttt{Python3 install.puy --update}.
 

### 2021-08-17 book/paper.tex
Consolidating appendices

@@
 
-\chapter{Value Types and Operators}
+\chapter{Harmony Language Details}
+
+\section{Value Types and Operators}
 \label{ap:values}
@@
 
-\section{Boolean}
+\subsection*{Boolean}
 
@@
 
-\section{Integer}
+\subsection*{Integer}
 
@@
 
-\section{Atom}
+\subsection*{Atom}
 
@@
 
-\section{Set}
+\subsection*{Set}
 
@@
 
-\section{Dictionary}
+\subsection*{Dictionary}
 
@@
 
-\section{List or Tuple}
+\subsection*{List or Tuple}
 
@@
 
-\section{String}
+\subsection*{String}
 
@@
 
-\section{Bag or Multiset}
+\subsection*{Bag or Multiset}
 
@@
 
-\section{Program Counter}
+\subsection*{Program Counter}
 
@@
 
-\section{Address}
+\subsection*{Address}
 
@@
 
-\section{Context}
+\subsection*{Context}
 
@@
 
-\chapter{List of Statements}
+\section{Statements}
 
@@
 
-\section{Single expression evaluation}
+\subsection*{Single expression evaluation}
 
@@
 
-\section{Assignment}
+\subsection*{Assignment}
 
@@
 
-\section{\textbf{assert}}
+\subsection*{\textbf{assert}}
 
@@
 
-\section{\textbf{atomic}}
+\subsection*{\textbf{atomic}}
 
@@
 
-\section{\textbf{await}}
+\subsection*{\textbf{await}}
 
@@
 
-\section{\textbf{const}}
+\subsection*{\textbf{const}}
 
@@
 
-\section{\textbf{def}}
+\subsection*{\textbf{def}}
 
@@
 
-\section{\textbf{del}}
+\subsection*{\textbf{del}}
 
@@
 
-\section{\textbf{for ... in [...where]}}
+\subsection*{\textbf{for ... in [...where]}}
 
@@
 
-\section{\textbf{from ... import}}
+\subsection*{\textbf{from ... import}}
 
@@
 
-\section{\textbf{go}}
+\subsection*{\textbf{go}}
 
@@
 
-\section{\textbf{if ... [elif ...]* [else]}}
+\subsection*{\textbf{if ... [elif ...]* [else]}}
 
@@
 
-\section{\textbf{import}}
+\subsection*{\textbf{import}}
 
@@
 
-\section{\textbf{invariant}}
+\subsection*{\textbf{invariant}}
 
@@
 
-\section{\textbf{let}}
+\subsection*{\textbf{let}}
 
@@
 
-\section{\textbf{pass}}
+\subsection*{\textbf{pass}}
 
@@
 
-\section{\textbf{possibly}}
+\subsection*{\textbf{possibly}}
 
@@
 
-\section{\textbf{select ... in}}
+\subsection*{\textbf{select ... in}}
 
@@
 
-\section{\textbf{sequential}}
+\subsection*{\textbf{sequential}}
 
@@
 
-\section{\textbf{spawn}}
+\subsection*{\textbf{spawn}}
 
@@
 
-\section{\textbf{trap}}
+\subsection*{\textbf{trap}}
 
@@
 
-\section{\textbf{while}}
+\subsection*{\textbf{while}}
 
@@
 
-\chapter{List of Modules}
+\section{Harmony is not object-oriented}
+
+Python is object-oriented, but Harmony is not.  For Python programmers,
+this can lead to some unexpected differences.
+For example, consider the following code:
+
+\begin{code}
+\harmonysource{oo}
+\end{code}
+
+In Python, lists are objects.  Thus $x$ and $y$ point to the same list,
+and the assertion would fail if executed by Python.
+In Harmony, lists are values.  So, when $x$ is updated in Line~2, it does
+not affect the value of $y$.  The assertion succeeds.
+Harmony supports references to values (\autoref{ch:method}),
+allowing programs to implement shared objects.
+
+Because Harmony does not have objects, it also does not have object methods.
+However, Harmony methods and lambdas are program counter constants.
+These constants can be added to dictionaries.
+For example, in \autoref{fig:petersonmethods}
+you can add the \texttt{P\_enter} and
+\texttt{P\_exit} methods to the \texttt{P\_mutex} dictionary
+like so:
+\begin{code}
+\begin{verbatim}
+{ .turn: 0, .flags: [ False, False ], .enter: P_enter, .exit: P_exit }
+\end{verbatim}
+\end{code}
+That would allow you to simulate object methods.
+
+There are at least two reasons why Harmony is not object-oriented.
+First,
+object-orientation often adds layers of indirection that would make it
+harder to model check and also to interpret the results.  Consider, for example,
+a lock.  In Python, a lock is an object.  A lock variable would contain a
+reference to a lock object.  In Harmony, a lock variable contains the value
+of the lock itself.
+Thus, the following statement means something quite different in Python and
+Harmony:
+
+\begin{code}
+\begin{verbatim}
+x = y = Lock()
+\end{verbatim}
+\end{code}
+
+In Python, this creates two variables $x$ and $y$ referring to the same lock.
+In Harmony, the two variables will be two different locks.  If you want two
+variables referring to the same lock in Harmony, you might write:
+
+\begin{code}
+\begin{verbatim}
+lock = Lock()
+x = y = ?lock
+\end{verbatim}
+\end{code}
+
+The second reason for Harmony not being object-oriented is that
+many concurrency solutions in the literature are
+expressed in C or some other low-level language that does not support
+object-orientation, but instead use \texttt{malloc} and \texttt{free}.
+
+\section{Constants, Global and Local Variables}\label{sec:vars}
+
+Each (non-reserved) identifier in a Harmony program refers to either
+a constant, a global shared variable, a local variable, or a module.
+Constants are declared using \texttt{const} statements.
+Those constants are computed at compile-time.
+
+Local variables all declared.  They can be declared in
+\texttt{def} statements (i.e., arguments),
+\texttt{let} statements, \texttt{for} loops, and \texttt{select} statements.
+Also, each method has an implicitly declared
+\textit{result} variable, which is initialized to \texttt{None}.
+Each thread has a variable called \textit{this} that contains
+the thread-local state.
+Local variables are tightly scoped and cannot be shared
+between threads.
+While in theory one method can be declared within another,
+they cannot share local variables either.
+All other variables are global and must be initialized
+before any threads are spawned.
+
+While arguments to a method and variables in \texttt{for} loops
+can be modified, we discourage it for improved code readability.
+
+\section{Operator Precedence}
+
+In Harmony, there is no syntactic difference between applying an argument to a function
+or an index to a dictionary.  Both use the syntax \textit{a b c ...}.
+We call this \emph{application}, and application is left-associative.
+So, \textit{a b c} is interpreted as (\textit{a b}) $c$: $b$ is applied to $a$,
+and then $c$ is applied to the result.
+For readability, it may help to write $a(b)$ for function application and
+$a[b]$ for indexing.  In case $b$ is an atom, you can also write $a.b$ for indexing.
+
+There are three levels of precedence.
+Application has the highest precedence.  So, \texttt{!}\textit{a b} is interpreted as
+\texttt{!}(\textit{a b}) and \textit{a b} \texttt{+} \textit{c d} is interpreted as
+(\textit{a b}) \texttt{+} (\textit{c d}).
+Unary operators have the next highest precedence,
+and the remaining operators have the lowest precedence.
+For example, $-2 + 3$ evaluates to 1, not $-5$.
+
+Associative operators ($+$, $*$, $|$, $\string&$, $\string^$, \textbf{and}, \textbf{or})
+are interpreted as general $n$-ary operators, and you are allowed to write
+$a + b + c$.  However, $a - b - c$ is illegal, as is any combination of operators with an
+arity larger than one, such as $a + b < c$.
+In such cases you have to add parentheses or brackets to indicate what
+the intended evaluation order is, such as $(a + b) < c$.
+
+In almost all expressions, subexpressions are evaluated left to right.
+So, $a[b] + c$
+first evaluates $a$, then $b$ (and then applies $b$ to $a$), and then $c$.  The one
+exception is the expression $a$ \textbf{if} $c$ \textbf{else} $b$, where $c$ is evaluated
+first.  In that expression, only $a$ or $b$ is evaluated depending on the value of $c$.
+In the expression $a$ \textbf{and} $b$ \textbf{and} $...$, evaluation is left
+to right but stops once one of the subexpressions evaluates to \textbf{False}.
+Similarly for \textbf{or}, where evaluation stops once one of the subexpressions
+evaluates to \textbf{True}.
+A sequence of comparison operations, such as $a < b < c$, is evaluated left to right
+but stops as soon as one of the comparisons fails.
+
+As an aside:
+the expression $a$ \textbf{not in} $b$ is equivalent to \textbf{not} ($a$ \textbf{in} $b$).
+Harmony generalizes this construct for any pair of a unary (except '$-$') and a binary operator.
+In particular, $a$ \textbf{not and} $b$ is the same as \textbf{not} ($a$ \textbf{and} $b$).  
+For those familiar with logic gates, \textbf{not and} is the equivalent of \texttt{NAND}.
+Similarly, \textbf{not} \texttt{$=>$} is non-implication.
+
+\section{Tuples, Lists, and Pattern Matching}
+
+Harmony's tuples and, equivalently, lists, are just special cases of dictionaries.
+They can be bracketed either by '(' and ')' or by '[' and ']', but
+the brackets are often optional.  Importantly, with a singleton list, the
+one element must be followed by a comma.
+For example, the statement \texttt{$x$ = 1,;} assigns a singleton tuple (or list) to $x$.
+
+Because tuples and lists are dictionaries, the \texttt{del} statement is
+different from Python.  For example, if $x$ = [.a, .b, .c], then
+\texttt{del} $x$[1] results in $x$ = \{ 0:.a, 2:.c \}, \emph{not}
+$x$ = [.a, .c].
+Harmony also does not support special slicing syntax like Python.
+To modify lists, use the \texttt{subseq} method in the \texttt{list} module
+(\autoref{ap:list}).
+
+Harmony allows pattern matching against nested tuples in various
+language constructs.
+\index{pattern matching}%
+The following are the same in Python and Harmony:
+\begin{itemize}
+\item \texttt{$x,$ = 1,}: assigns 1 to $x$;
+\item \texttt{$x$, $y$ = 1, (2, 3)}: assigns 1 to $x$ and (2, 3) to $y$;
+\item \texttt{$x$, ($y$, $z$) = 1, (2, 3)}: assigns 1 to $x$, 2 to $y$, and
+3 to $z$;
+\item \texttt{$x$, ($y$, $z$) = 1, 2}: generates an runtime error because 2 cannot
+be matched with ($y$, $z$);
+\item \texttt{$x[0]$, $x[1]$ = $x[1]$, $x[0]$}: swaps the first two elements of list $x$.
+\end{itemize}
+
+As in Python, pattern matching can also be used in \texttt{for} statements.
+For example:
+\begin{code}
+\texttt{for} key, value in [ (1, 2), (3, 4) ]:
+    ...
+\end{code}
+
+Harmony (but not Python)
+also allows pattern matching in defining and invoking methods.
+For example, you can write:
+\begin{code}
+\texttt{def} f[$a$, ($b$, $c$)]: ...
+\end{code}
+and then call \texttt{f[1, (2, 3)]}.
+Note that the more familiar: \texttt{def} g($a$) defines a method $g$ with
+a single argument $a$.  Invoking g(1, 2) would assign the tuple (1, 2) to
+$a$.  This is not consistent with Python syntax.  For single argument methods,
+you may want to declare as follows: \texttt{def} g($a,$).
+Calling g(1,) assigns 1 to $a$, while calling g(1, 2) would result in a
+runtime error as (1, 2) cannot be matched with ($a$,).
+
+Pattern matching can also be used in \texttt{const}, \texttt{let},
+and \texttt{select} statements.
+
+\begin{comment}
+\section{For Loops and Comprehensions}
+
+While Harmony does not support general iterators such as Python does,
+Harmony allows iterating over sets and dictionaries (and thus lists
+and tuples).  The details are a little different from Python:
+
+\begin{itemize}
+\item When iterating over a set, the set is always traversed in order
+(see \autoref{ap:values} for how Harmony values are ordered);
+\item In case of a dictionary, the iteration is over the \emph{values} of the
+dictionary, but in the order of the keys.  In the case of lists, this
+works much the same as in Python, but in the case of general dictionaries,
+Python iterates over the keys rather than the values;
+\item If you want to iterate over the keys of a dictionary $d$, use either
+\begin{itemize}
+\item \texttt{for} $k$ \texttt{in} \texttt{keys} $d$; or
+\item \texttt{for} $k:\_$ \texttt{in} $d$
+\end{itemize}
+\item The \texttt{list} module (\autoref{ap:list}) provides methods \texttt{values}(),
+\texttt{items}(), \texttt{enumerate}(), and \texttt{reversed}() for
+other types of iteration supported by Python.
+\end{itemize}
+
+Harmony supports nesting and filtering in \texttt{for loops}.
+For example:
+\begin{code}
+\texttt{for} $i$ in \{ 1..10 \}
+\texttt{for} $j$ in \{ 1..10 \}
+\texttt{where} $i < j$: ...
+\end{code}
+
+In case of dictionaries, you can access both keys and values (not
+supported in Python):
+\begin{code}
+\texttt{for} $k:v$ in \{ .a: 1, .b: 2 \}: ...
+\end{code}
+
+Harmony also supports set, list, and dictionary comprehensions.
+\index{comprehensions}%
+Comprehensions are similar to Python, except that filtering uses
+the keyword \texttt{where} instead of \texttt{if}.
+\end{comment}
+
+\section{Dynamic Allocation}
+
+\begin{figure}
+\begin{code}
+\harmonysource{stacktest}
+\end{code}
+\caption{\harmonylink{code/stacktest.hny} Testing a stack implementation.}
+\label{fig:stacktest}
+\end{figure}
+
+\begin{figure}
+\begin{code}
+\harmonysource{stack1}
+\end{code}
+\caption{\harmonylink{code/stack1.hny} Stack implemented using a dynamically updated list.}
+\label{fig:stack1}
+\end{figure}
+
+\begin{figure}
+\begin{code}
+\harmonysource{stack2}
+\end{code}
+\caption{\harmonylink{code/stack2.hny} Stack implemented using static lists.}
+\label{fig:stack2}
+\end{figure}
+
+\begin{figure}
+\begin{code}
+\harmonysource{stack3}
+\end{code}
+\caption{\harmonylink{code/stack3.hny} Stack implemented using a recursive tuple data structure.}
+\label{fig:stack3}
+\end{figure}
+
+\begin{figure}
+\begin{code}
+\harmonysource{stack4}
+\end{code}
+\caption{\harmonylink{code/stack4.hny} Stack implemented using a linked list.}
+\label{fig:stack4}
+\end{figure}
+
+Harmony supports various options for dynamic allocation.
+By way of example, consider a stack.
+\autoref{fig:stacktest} presents a test program for a stack.
+We present four different stack implementations to illustrate
+options for dynamic allocation:
+\begin{enumerate}
+\item[] \autoref{fig:stack1} uses a single list to represent the
+stack.  It is updated to perform \texttt{push} and \texttt{pop} operations;
+\item[] \autoref{fig:stack2} also uses a list but, instead of updating
+the list, it replaces the list with a new one for each operation;
+\item[] \autoref{fig:stack3} represents a stack as a recursively nested tuple
+$(v, f)$, where $v$ is the element on top of the stack and $r$ is a stack
+that is the remainder;
+\item[] \autoref{fig:stack4} implements a stack as a linked list with nodes
+allocated using the \texttt{alloc} module.
+\end{enumerate}
+
+While the last option is the most versatile (it allows cyclic
+data structures), Harmony does not support garbage collection
+for memory allocated this way and so allocated memory that is no
+longer in use must be explicitly released using \texttt{free}.
+
+\section{Comments}
+
+Harmony supports the same commenting conventions as Python.
+In addition, Harmony supports nested multi-line comments
+of the form \texttt{(* comment *)}.
+
+\chapter{Modules}
 \label{ap:module}
@@
 \hline
-\texttt{exit($m$)} & exit a monitor.\\
+\texttt{exit($m$)} & exit a monitor\\
 \hline
@@
 \hline
-\texttt{remove}($s$, $e$) & returns $s - \{ e \}$ \\
+\texttt{remove}($s$, $e$) & returns $s \backslash \{ e \}$ \\
 \hline
@@
 For Python programmers: note that $s <= t$ does not check if $s$ is a subset of
-$t$ when $s$ and $t$ are sets, as ``$<=$'' implements a total order on all Harmony values
-including sets.
+$t$ when $s$ and $t$ are sets, as ``$<=$'' implements a total order on all Harmony values including sets (and the subset relation is not a total order).
 
@@
 
-\chapter{List of Machine Instructions}
+\chapter{The Harmony Virtual Machine}\label{app:hvm}
+
+The Harmony Virtual Machine (HVM, \autoref{ch:harmonymachine})
+has the following state:
+
+\vspace{1em}
+\begin{tabular}{|l|l|}
+\hline
+code & a list of HVM machine instructions \\
+\hline
+labels & a dictionary of atoms to program counters \\
+\hline
+variables & a dictionary mapping atoms to values \\
+\hline
+ctxbag & a bag of runnable contexts \\
+\hline
+stopbag & a bag of stopped contexts \\
+\hline
+choosing & if not \texttt{None}, indicates a context that is choosing \\
+\hline
+\end{tabular}
+\vspace{1em}
+
+There is initially a single context with name
+\texttt{\_\_init\_\_()} and program counter 0.  It starts executing
+in atomic mode until it finishes executing the last
+\texttt{Return} instruction.
+Other threads, created through \texttt{spawn} statements, do not
+start executing until then.
+
+A \emph{step} is the execution of a single HVM machine instruction
+by a context.
+Each step generates a new state.
+When there are multiple contexts, the HVM can interleave them.
+However, trying to interleave every step would be needlessly expensive,
+as many steps involve changes to a context that are invisible to
+other contexts.
+
+A \emph{stride}
+\index{stride}%
+can involve multiple steps.  The following
+instructions start a new stride: \texttt{Load}, \texttt{Store},
+\texttt{AtomicInc}, and \texttt{Continue}.  The HVM
+interleaves stides, not steps.  Like steps, each
+stride involves a single context.  Unlike a step, a stride
+can leave the state unchanged (because its steps lead back
+to where the stride started).
+
+Executing a Harmony program results in a graph where the nodes are Harmony
+states and the edges are strides.
+When a state is \texttt{choosing}, the edges from that state are
+by a single context, one for each choice.  If not, the edges from
+the state are one per context.
+
+Consecutive strides by the same thread are called a \emph{turn}.
+Each state maintains the shortest path to it from the initial state in terms
+of turns.
+The diameter of the graph is the length of the longest path found in
+terms of turns.
+
+If some states have a problem, the state with the shortest path is reported.
+Problematic states include states that experienced exceptions.
+If there are no exceptions, Harmony computes the strongly connected components (SCCs)
+of the graph (the number of such components are printed as part of the output).
+The sink SCCs should each consist of a terminal state without any threads.
+If not, again the state with the shortest path is reported.
+
+If there are no problematic states, Harmony reports ``no issues found'' and outputs
+in the HTML file the state with the longest path.
+
+\section{Machine Instructions}
 \label{ap:harmonybytecode}
@@
 
-\chapter{Contexts and Threads}
+\section{Contexts and Threads}
 \label{app:context}
@@
 \end{itemize}
-
-\chapter{The Harmony Virtual Machine}\label{app:hvm}
-
-The Harmony Virtual Machine (HVM, \autoref{ch:harmonymachine})
-has the following state:
-
-\vspace{1em}
-\begin{tabular}{|l|l|}
-\hline
-code & a list of HVM machine instructions \\
-\hline
-labels & a dictionary of atoms to program counters \\
-\hline
-variables & a dictionary mapping atoms to values \\
-\hline
-ctxbag & a bag of runnable contexts \\
-\hline
-stopbag & a bag of stopped contexts \\
-\hline
-choosing & if not \texttt{None}, indicates a context that is choosing \\
-\hline
-\end{tabular}
-\vspace{1em}
-
-There is initially a single context with name
-\texttt{\_\_init\_\_/()} and program counter 0.  It starts executing
-in atomic mode until it finishes executing the last
-\texttt{Return} instruction.
-Other threads, created through \texttt{spawn} statements, do not
-start executing until then.
-
-A \emph{step} is the execution of a single HVM machine instruction
-by a context.
-Each step generates a new state.
-When there are multiple contexts, the HVM can interleave them.
-However, trying to interleave every step would be needlessly expensive,
-as many steps involve changes to a context that are invisible to
-other contexts.
-
-A \emph{stride}
-\index{stride}%
-can involve multiple steps.  The following
-instructions start a new stride: \texttt{Load}, \texttt{Store},
-\texttt{AtomicInc}, and \texttt{Continue}.  The HVM
-interleaves stides, not steps.  Like steps, each
-stride involves a single context.  Unlike a step, a stride
-can leave the state unchanged (because its steps lead back
-to where the stride started).
-
-Executing a Harmony program results in a graph where the nodes are Harmony
-states and the edges are strides.
-When a state is \texttt{choosing}, the edges from that state are
-by a single context, one for each choice.  If not, the edges from
-the state are one per context.
-
-Consecutive strides by the same thread are called a \emph{turn}.
-Each state maintains the shortest path to it from the initial state in terms
-of turns.
-The diameter of the graph is the length of the longest path found in
-terms of turns.
-
-If some states have a problem, the state with the shortest path is reported.
-Problematic states include states that experienced exceptions.
-If there are no exceptions, Harmony computes the strongly connected components (SCCs)
-of the graph (the number of such components are printed as part of the output).
-The sink SCCs should each consist of a terminal state without any threads.
-If not, again the state with the shortest path is reported.
-
-If there are no problematic states, Harmony reports ``no issues found'' and outputs
-in the HTML file the state with the longest path.
-
-\chapter{Harmony Language Details}
-\label{ap:details}
-
-The Harmony language borrows heavily from Python.  However, there are
-some important differences that we will describe in this chapter.
-
-\section{Harmony is not object-oriented}
-
-Python is object-oriented, but Harmony is not.  For Python programmers,
-this can lead to some unexpected differences.
-For example, consider the following code:
-
-\begin{code}
-\harmonysource{oo}
-\end{code}
-
-In Python, lists are objects.  Thus $x$ and $y$ point to the same list,
-and the assertion would fail if executed by Python.
-In Harmony, lists are values.  So, when $x$ is updated in Line~2, it does
-not affect the value of $y$.  The assertion succeeds.
-Harmony supports references to values (\autoref{ch:method}),
-allowing programs to implement shared objects.
-
-Because Harmony does not have objects, it also does not have object methods.
-However, Harmony methods and lambdas are program counter constants.
-These constants can be added to dictionaries.
-For example, in \autoref{fig:petersonmethods}
-you can add the \texttt{P\_enter} and
-\texttt{P\_exit} methods to the \texttt{P\_mutex} dictionary
-like so:
-\begin{code}
-\begin{verbatim}
-{ .turn: 0, .flags: [ False, False ], .enter: P_enter, .exit: P_exit }
-\end{verbatim}
-\end{code}
-That would allow you to simulate object methods.
-
-There are at least two reasons why Harmony is not object-oriented.
-First,
-object-orientation often adds layers of indirection that would make it
-harder to model check and also to interpret the results.  Consider, for example,
-a lock.  In Python, a lock is an object.  A lock variable would contain a
-reference to a lock object.  In Harmony, a lock variable contains the value
-of the lock itself.
-Thus, the following statement means something quite different in Python and
-Harmony:
-
-\begin{code}
-\begin{verbatim}
-x = y = Lock()
-\end{verbatim}
-\end{code}
-
-In Python, this creates two variables $x$ and $y$ referring to the same lock.
-In Harmony, the two variables will be two different locks.  If you want two
-variables referring to the same lock in Harmony, you might write:
-
-\begin{code}
-\begin{verbatim}
-lock = Lock()
-x = y = ?lock
-\end{verbatim}
-\end{code}
-
-The second reason for Harmony not being object-oriented is that
-many concurrency solutions in the literature are
-expressed in C or some other low-level language that does not support
-object-orientation, but instead use \texttt{malloc} and \texttt{free}.
-
-\section{Constants, Global and Local Variables}\label{sec:vars}
-
-Each (non-reserved) identifier in a Harmony program refers to either
-a constant, a global shared variable, a local variable, or a module.
-Constants are declared using \texttt{const} statements.
-Those constants are computed at compile-time.
-
-Local variables all declared.  They can be declared in
-\texttt{def} statements (i.e., arguments),
-\texttt{let} statements, \texttt{for} loops, and \texttt{select} statements.
-Also, each method has an implicitly declared
-\textit{result} variable, which is initialized to \texttt{None}.
-Each thread has a variable called \textit{this} that contains
-the thread-local state.
-Local variables are tightly scoped and cannot be shared
-between threads.
-While in theory one method can be declared within another,
-they cannot share local variables either.
-All other variables are global and must be initialized
-before any threads are spawned.
-
-While arguments to a method and variables in \texttt{for} loops
-can be modified, we discourage it for improved code readability.
-
-\section{Operator Precedence}
-
-In Harmony, there is no syntactic difference between applying an argument to a function
-or an index to a dictionary.  Both use the syntax \textit{a b c ...}.
-We call this \emph{application}, and application is left-associative.
-So, \textit{a b c} is interpreted as (\textit{a b}) $c$: $b$ is applied to $a$,
-and then $c$ is applied to the result.
-For readability, it may help to write $a(b)$ for function application and
-$a[b]$ for indexing.  In case $b$ is an atom, you can also write $a.b$ for indexing.
-
-There are three levels of precedence.
-Application has the highest precedence.  So, \texttt{!}\textit{a b} is interpreted as
-\texttt{!}(\textit{a b}) and \textit{a b} \texttt{+} \textit{c d} is interpreted as
-(\textit{a b}) \texttt{+} (\textit{c d}).
-Unary operators have the next highest precedence,
-and the remaining operators have the lowest precedence.
-For example, $-2 + 3$ evaluates to 1, not $-5$.
-
-Associative operators ($+$, $*$, $|$, $\string&$, $\string^$, \textbf{and}, \textbf{or})
-are interpreted as general $n$-ary operators, and you are allowed to write
-$a + b + c$.  However, $a - b - c$ is illegal, as is any combination of operators with an
-arity larger than one, such as $a + b < c$.
-In such cases you have to add parentheses or brackets to indicate what
-the intended evaluation order is, such as $(a + b) < c$.
-
-In almost all expressions, subexpressions are evaluated left to right.
-So, $a[b] + c$
-first evaluates $a$, then $b$ (and then applies $b$ to $a$), and then $c$.  The one
-exception is the expression $a$ \textbf{if} $c$ \textbf{else} $b$, where $c$ is evaluated
-first.  In that expression, only $a$ or $b$ is evaluated depending on the value of $c$.
-In the expression $a$ \textbf{and} $b$ \textbf{and} $...$, evaluation is left
-to right but stops once one of the subexpressions evaluates to \textbf{False}.
-Similarly for \textbf{or}, where evaluation stops once one of the subexpressions
-evaluates to \textbf{True}.
-A sequence of comparison operations, such as $a < b < c$, is evaluated left to right
-but stops as soon as one of the comparisons fails.
-
-As an aside:
-the expression $a$ \textbf{not in} $b$ is equivalent to \textbf{not} ($a$ \textbf{in} $b$).
-Harmony generalizes this construct for any pair of a unary (except '$-$') and a binary operator.
-In particular, $a$ \textbf{not and} $b$ is the same as \textbf{not} ($a$ \textbf{and} $b$).  
-For those familiar with logic gates, \textbf{not and} is the equivalent of \texttt{NAND}.
-Similarly, \textbf{not} \texttt{$=>$} is non-implication.
-
-\section{Tuples, Lists, and Pattern Matching}
-
-Harmony's tuples and, equivalently, lists, are just special cases of dictionaries.
-They can be bracketed either by '(' and ')' or by '[' and ']', but
-the brackets are often optional.  Importantly, with a singleton list, the
-one element must be followed by a comma.
-For example, the statement \texttt{$x$ = 1,;} assigns a singleton tuple (or list) to $x$.
-
-Because tuples and lists are dictionaries, the \texttt{del} statement is
-different from Python.  For example, if $x$ = [.a, .b, .c], then
-\texttt{del} $x$[1] results in $x$ = \{ 0:.a, 2:.c \}, \emph{not}
-$x$ = [.a, .c].
-Harmony also does not support special slicing syntax like Python.
-To modify lists, use the \texttt{subseq} method in the \texttt{list} module
-(\autoref{ap:list}).
-
-Harmony allows pattern matching against nested tuples in various
-language constructs.
-\index{pattern matching}%
-The following are the same in Python and Harmony:
-\begin{itemize}
-\item \texttt{$x,$ = 1,}: assigns 1 to $x$;
-\item \texttt{$x$, $y$ = 1, (2, 3)}: assigns 1 to $x$ and (2, 3) to $y$;
-\item \texttt{$x$, ($y$, $z$) = 1, (2, 3)}: assigns 1 to $x$, 2 to $y$, and
-3 to $z$;
-\item \texttt{$x$, ($y$, $z$) = 1, 2}: generates an runtime error because 2 cannot
-be matched with ($y$, $z$);
-\item \texttt{$x[0]$, $x[1]$ = $x[1]$, $x[0]$}: swaps the first two elements of list $x$.
-\end{itemize}
-
-As in Python, pattern matching can also be used in \texttt{for} statements.
-For example:
-\begin{code}
-\texttt{for} key, value in [ (1, 2), (3, 4) ]:
-    ...
-\end{code}
-
-Harmony (but not Python)
-also allows pattern matching in defining and invoking methods.
-For example, you can write:
-\begin{code}
-\texttt{def} f[$a$, ($b$, $c$)]: ...
-\end{code}
-and then call \texttt{f[1, (2, 3)]}.
-Note that the more familiar: \texttt{def} g($a$) defines a method $g$ with
-a single argument $a$.  Invoking g(1, 2) would assign the tuple (1, 2) to
-$a$.  This is not consistent with Python syntax.  For single argument methods,
-you may want to declare as follows: \texttt{def} g($a,$).
-Calling g(1,) assigns 1 to $a$, while calling g(1, 2) would result in a
-runtime error as (1, 2) cannot be matched with ($a$,).
-
-Pattern matching can also be used in \texttt{const}, \texttt{let},
-and \texttt{select} statements.
-
-\begin{comment}
-\section{For Loops and Comprehensions}
-
-While Harmony does not support general iterators such as Python does,
-Harmony allows iterating over sets and dictionaries (and thus lists
-and tuples).  The details are a little different from Python:
-
-\begin{itemize}
-\item When iterating over a set, the set is always traversed in order
-(see \autoref{ap:values} for how Harmony values are ordered);
-\item In case of a dictionary, the iteration is over the \emph{values} of the
-dictionary, but in the order of the keys.  In the case of lists, this
-works much the same as in Python, but in the case of general dictionaries,
-Python iterates over the keys rather than the values;
-\item If you want to iterate over the keys of a dictionary $d$, use either
-\begin{itemize}
-\item \texttt{for} $k$ \texttt{in} \texttt{keys} $d$; or
-\item \texttt{for} $k:\_$ \texttt{in} $d$
-\end{itemize}
-\item The \texttt{list} module (\autoref{ap:list}) provides methods \texttt{values}(),
-\texttt{items}(), \texttt{enumerate}(), and \texttt{reversed}() for
-other types of iteration supported by Python.
-\end{itemize}
-
-Harmony supports nesting and filtering in \texttt{for loops}.
-For example:
-\begin{code}
-\texttt{for} $i$ in \{ 1..10 \}
-\texttt{for} $j$ in \{ 1..10 \}
-\texttt{where} $i < j$: ...
-\end{code}
-
-In case of dictionaries, you can access both keys and values (not
-supported in Python):
-\begin{code}
-\texttt{for} $k:v$ in \{ .a: 1, .b: 2 \}: ...
-\end{code}
-
-Harmony also supports set, list, and dictionary comprehensions.
-\index{comprehensions}%
-Comprehensions are similar to Python, except that filtering uses
-the keyword \texttt{where} instead of \texttt{if}.
-\end{comment}
-
-\section{Dynamic Allocation}
-
-\begin{figure}
-\begin{code}
-\harmonysource{stacktest}
-\end{code}
-\caption{\harmonylink{code/stacktest.hny} Testing a stack implementation.}
-\label{fig:stacktest}
-\end{figure}
-
-\begin{figure}
-\begin{code}
-\harmonysource{stack1}
-\end{code}
-\caption{\harmonylink{code/stack1.hny} Stack implemented using a dynamically updated list.}
-\label{fig:stack1}
-\end{figure}
-
-\begin{figure}
-\begin{code}
-\harmonysource{stack2}
-\end{code}
-\caption{\harmonylink{code/stack2.hny} Stack implemented using static lists.}
-\label{fig:stack2}
-\end{figure}
-
-\begin{figure}
-\begin{code}
-\harmonysource{stack3}
-\end{code}
-\caption{\harmonylink{code/stack3.hny} Stack implemented using a recursive tuple data structure.}
-\label{fig:stack3}
-\end{figure}
-
-\begin{figure}
-\begin{code}
-\harmonysource{stack4}
-\end{code}
-\caption{\harmonylink{code/stack4.hny} Stack implemented using a linked list.}
-\label{fig:stack4}
-\end{figure}
-
-Harmony supports various options for dynamic allocation.
-By way of example, consider a stack.
-\autoref{fig:stacktest} presents a test program for a stack.
-We present four different stack implementations to illustrate
-options for dynamic allocation:
-\begin{enumerate}
-\item[] \autoref{fig:stack1} uses a single list to represent the
-stack.  It is updated to perform \texttt{push} and \texttt{pop} operations;
-\item[] \autoref{fig:stack2} also uses a list but, instead of updating
-the list, it replaces the list with a new one for each operation;
-\item[] \autoref{fig:stack3} represents a stack as a recursively nested tuple
-$(v, f)$, where $v$ is the element on top of the stack and $r$ is a stack
-that is the remainder;
-\item[] \autoref{fig:stack4} implements a stack as a linked list with nodes
-allocated using the \texttt{alloc} module.
-\end{enumerate}
-
-While the last option is the most versatile (it allows cyclic
-data structures), Harmony does not support garbage collection
-for memory allocated this way and so allocated memory that is no
-longer in use must be explicitly released using \texttt{free}.
-
-\section{Comments}
-
-Harmony supports the same commenting conventions as Python.
-In addition, Harmony supports nested multi-line comments
-of the form \texttt{(* comment *)}.
 
@@
 the compiler and explores the entire state space in breadth-first order.
-Even Harmony does not really support input, there are three sources of
+Even though Harmony does not really support input, there are three sources of
 non-determinism that make this exploration non-trivial:

### 2021-08-17 book/paper.tex
Added a few more problems

@@
 It then waits until the other thread is done.
-(\textbf{await} $c$ is shorthand for \textbf{while not} $c$: \textbf{pass}) and semantically the same.
+(\textbf{await} $c$ is shorthand for \textbf{while not} $c$: \textbf{pass}.)
 After that, method \texttt{incrementer}
@@
 analyze the output and see what went wrong.
+\item \cite{MWA19} discusses a buggy version of Paxos.  In this version,
+the responses to the second phase are matched not by ballot number but
+by the value of the proposal.  Implement this version and, using Harmony,
+find the problem this causes.
 \end{problems}
@@
 Alice to Bob (but not vice versa because Alice did not try to authenticate Bob).
+
+\section*{Exercises}
+\begin{problems}
+\item Figure out how to fix the protocol.
+\item There were two versions of the Needham-Schroeder protocol: the
+Symmetric Key protocol and the Public Key protocol.  In this chapter
+we only discussed the latter, but the former also had a problem.  See
+if you can find it using Harmony.
+\end{problems}
 

### 2021-10-31 book/paper.tex
Latest
Added "where" clause to select statement
Introducing "countLabel", which counts the number of threads at a label

@@
 \texttt{atLabel $l$} & return a bag of (method, argument) pairs of threads executing at label $l$ \\
+\texttt{countLabel $l$} & return the number of threads executing at label $l$ \\
 \hline
@@
 
-\subsection*{\textbf{for ... in [...where]}}
+\subsection*{\textbf{for ... in ... [where ...]}}
 
@@
 At completion, Harmony reports which of the predicates never held.
-
-\subsection*{\textbf{select ... in}}
+While \texttt{assert} statements check that nothing ever goes wrong,
+\texttt{possibly} statements can check that certain things sometimes
+go right.
+
+\subsection*{\textbf{select ... in ... [ where ... ]}}
 
@@
 of actions atomically after the desired messages have arrived.
+
+The optional \textbf{where} clause can be used to filter the set down to the elements
+that satisfy the condition following the \textbf{where} keyword.
 

### 2021-10-31 book/paper.tex
Added code for ticket lock Starting to update book

@@
 Harmony in fact supports this.
-It has an operator \textbf{atLabel}~$L$,
-\index{atLabel operator}%
+It has an operator \textbf{countLabel}~$L$,
+\index{countLabel operator}%
 where $L$
 is the atom containing the name of the label (in this case, \texttt{cs}).
-The operator returns a bag (multiset) of (method, argument) pairs of threads
-executing at that label.
-In our case there are two threads: \texttt{(thread, 0)} and \texttt{(thread, 1)}.
-A bag is represented by a dictionary that maps each element
-in the bag to the number of times the element appears in the bag.
-Method \textbf{atLabel} only exists for specification purposes---do not
+The operator returns the number of threads executing at that label.
+Method \textbf{countLabel} only exists for specification purposes---do not
 use it in normal code.
@@
 A thread usually does not keep explicit track of whether it has a lock
-or not, because it is usually implied by the control flow of the program---a
+or not, because it is implied by the control flow of the program---a
 thread implicitly \emph{knows} that when it is executing in a critical
@@
 testing mutual exclusion and progress in critical sections, but depends
-on \texttt{atLabel}, an unusual operator specific to Harmony that is mostly
+on \texttt{countLabel}, an unusual operator specific to Harmony that is mostly
 useful for testing mutual exclusion.
@@
 
+In Harmony, a bag is usually represented by a dictionary that maps
+the elements of the bag to their multiplicities.
+For example, the value \texttt{\{ .a: 2, .b: 3 \}} represents a bag
+with two copies of \texttt{.a} and three copies of \texttt{.b}.
+The \texttt{bag} module (\autoref{ap:bag}) contains a variety of handy
+functions on bags.
+
 \texttt{wait} adds the context of the thread to the bag.
@@
 messages.
-Because of this, 
-we will model the network as a bag (aka multiset) of messages.
-The bag itself is
-modeled in Harmony as a dictionary that maps unique elements in the bag to their
-multiplicity.  So, for example, $\{ .\texttt{a}:1, .\texttt{b}:2 \}$ is the bag that consists of
-one $.\texttt{a}$ and two copies of $.\texttt{b}$.
-The \texttt{bag} module (\autoref{ap:bag}) contains a variety of handy functions on bags.
+Because of this, we will model the network as a bag of messages.
 

### 2021-11-01 book/paper.tex
Started working on book

@@
 language needed for the rest of the book.
+\item \autoref{ch:specification} talks about how Harmony can be used as
+a specification language.  It introduces how can can specify atomic
+constructs.
 \item \autoref{ch:spinlock} introduces atomic \emph{locks}
@@
 
+\chapter{Specification}
+\label{ch:specification}
+
+\begin{figure}
+\begin{code}
+\harmonysource{lockspec}
+\end{code}
+\caption{\harmonylink{modules/synch.hny} Specification of a lock.}
+\label{fig:lockspec}
+\end{figure}
+
+\begin{figure}
+\begin{code}
+\harmonysource{cssynch}
+\end{code}
+\caption{\harmonylink{code/cssynch.hny} Using a lock to implement a
+critical section.}
+\label{fig:cssynch}
+\end{figure}
+
+So far we have used Harmony to \emph{implement} various algorithms.
+But Harmony can also be used to \emph{specify} what an algorithm is
+supposed to do.
+For example, \autoref{fig:lockspec} specifies the intended behavior
+of a lock.
+In this case, a lock is a boolean, initially \texttt{False},
+with two operations, \texttt{acquire()} and \texttt{release()}.
+The \texttt{acquire()} operation waits until the lock is \texttt{False}
+and then sets it to \texttt{True} in an atomic operation.
+The \texttt{release()} operation sets the lock back to \texttt{False}.
+The code is similar to \autoref{fig:naivelock}, except that waiting
+for the lock to become available and taking it is executed as an
+atomic operation.
+
+The code in \autoref{fig:lockspec} is taken from Harmony's \texttt{synch}
+module.  \autoref{fig:cssynch} shows how locks may be used.  Note that
+the code of \autoref{fig:lockspec} is executable in Harmony.
+However, the \texttt{atomically} keyword is not available in general
+programming languages and should not be used for implementation.
+Peterson's algorithm is an implementation of a lock, although only for
+two processes.  In the following chapters we will look at more general
+ways of implementing locks using atomic constructions that are usually
+available in the underlying hardware.
+
+In Harmony, any statement can be preceded by the \texttt{atomically}
+keyword.  It means that statement as a whole is to be executed atomically.
+The \texttt{atomically} keyword can be used to specify the behavior of
+methods such as \texttt{acquire} and \texttt{release}.  But note that
+an actual executable program such as in \autoref{fig:cssynch} should not
+use the keyword.  In fact, if we want to make the entire program
+executable, we have to also show how \texttt{acquire} and \texttt{release}
+are implemented, not just how they are specified.
+\autoref{ch:spinlock} presents such an implementation.
+
+The code in \autoref{fig:lockspec} also uses the Harmony \texttt{when}
+statement.  A \texttt{when} statement waits until a time in which
+condition holds (not necessarily the first time) and then executes the
+statement block.  The \texttt{await} \textit{condition} statement is
+the same as \texttt{when} \textit{condition}: \texttt{pass}.
+Combined with the \texttt{atomically} keyword, the entire statement
+is executed atomically at a time that the condition holds.
+
 \chapter{Spinlock}
@@
 
-\autoref{fig:naivelock} showed a faulty attempt at solving mutual
-exclusion using a lock.  The problem with the implementation of the
-lock is that checking the lock and setting it if it is available is
-not \emph{atomic}.  Thus multiple threads contending for the lock
-can all ``grab the lock'' at the same time.  While Peterson's
-algorithm gets around the problem, it is not efficient, especially
-if generalized to multiple threads.
+Peterson's algorithm implements locks, but it is not efficient,
+especially if generalized to multiple threads.
 Worse, Peterson relies on load and store operations to be executed
@@
 it does have support for executing multiple instructions atomically.
-This feature is available in the Harmony language in two ways.
-First, any Harmony statement can be made atomic by placing a label in front
-of it.  Second, a group of Harmony statements can be made atomic
-through the \textbf{atomic}
-\index{atomic statement}%
-statement.
-We can use \textbf{atomic} statement blocks to implement a wide variety of
+Any Harmony statement can be made atomic by placing either a label in front
+of it or the keyword \texttt{atomically}.
+We can use atomic statements to implement a wide variety of
 atomic operations.
@@
 \begin{code}
-\harmonysource{csTAS}
-\end{code}
-\caption{\harmonylink{code/csTAS.hny} Fixed version of \autoref{fig:naivelock} using test-and-set.}
+\harmonysource{taslock}
+\end{code}
+\caption{\harmonylink{code/taslock.hny} Implementation of a spinlock
+using test-and-set.}
 \label{fig:tas2}
@@
 but uses test-and-set to check and set the lock variable atomically.
-The lock is also cleared in an \textbf{atomic} block to prevent a data race.
+The lock is also cleared in an atomic statement to prevent a data race.
 This approach is general for any number of threads.
@@
 \emph{atomic section} (the statements executed within an
-\textbf{atomic} statement) and a \emph{critical section}
+atomic block of statements) and a \emph{critical section}
 (protected by a lock of some sort).
 The former ensures that while the
-\textbf{atomic} statement is executing no other thread can execute.
+statements are executing no other thread can execute.
 The latter allows multiple threads to run concurrently,
@@
 
-In Harmony, \textbf{atomic} statements allow you to \emph{implement} your own
+In Harmony, atomic statements allow you to \emph{implement} your own
 atomic primitives such as test-and-set.  Other common examples
-include compare-and swap and fetch-and-add.  \textbf{atomic} statements
+include compare-and swap and fetch-and-add.  Atomic statements
 are not intended to replace locks or other synchonization primitives.
 When solving synchronization problems you should not directly use
-\textbf{atomic} statements but use the synchronization primitives that are available
+atomic statements but use the synchronization primitives that are available
 to you.  But if you want to design a new synchronization primitive, then
-use \textbf{atomic} by all means.
-You can also use \textbf{atomic} statements in your test code.
+use \textbf{atomically} by all means.
+You can also use atomic statements in your test code.
 In fact, as mentioned before, \textbf{assert} statements are executed atomically.
@@
 concurrent access to the specified variables is intended.
-Second, if the accesses are within an \textbf{atomic} statement, then they
+Second, if the accesses are within an atomic statement block, then they
 are not considered part of a data race.
@@
 \begin{comment}
-For the same reason that you should not use \textbf{atomic}
+For the same reason that you should not use atomic
 statements in your ordinary synchronization code, you should not use
@@
 In the latter case, the context of the thread is added to the end of the list of contexts.
-Note that \textbf{stop} is called within an \textbf{atomic} statement---this is
-the only exception to an atomic statement running to completion.  While the
+Note that \textbf{stop} is called within an atomic statement block---this is
+the only exception to such an atomic statement block running to completion.
+While the
 thread is running no other threads can run, but when the thread suspends itself
@@
 customer code the same. You are only allowed to change the
-\texttt{atm\_} methods, and you cannot use the \textbf{atomic}
-statement.
+\texttt{atm\_} methods, and you cannot use the \textbf{atomically}
+keyword.
 \end{problems}
@@
 \item Messages are added atomically to the network.  Is this necessary?
-What happens if you remove the \textbf{atomic} keyword?  Explain what happens.
+What happens if you remove the \textbf{atomically} keyword?  Explain what happens.
 \end{problems}
@@
 
-\begin{comment}
-Also, the function \texttt{contexts()} returns a
-bag of all contexts (including terminated ones).
-Both these functions should only be called in \texttt{atomic} mode
-and used for checking invariants.
-\end{comment}
-
 \section{Statements}
@@
 
-\subsection*{\textbf{atomic}}
+\subsection*{\textbf{atomically}}
 

### 2021-11-01 book/paper.tex
Working on book

@@
 for implemented critical sections.
-\item \autoref{ch:synch} presents Harmony modules
-that support locks and other synchronization primitives.
+\item \autoref{ch:synch} looks at various ways in which the lock specification
+in \autoref{ch:specification} can be implemented.
 \item \autoref{ch:cds} gives an introduction to building concurrent
@@
 explicitly bound within a scope.
-You have already seen examples: the arguments to the method,
-the variable \textit{result}, and the bound variable (or
-variables) within a \texttt{for} statement.
-The \texttt{let} statement allows declaring additional
-local variables.  For example:
+Method variables are either mutable or immutable.
+The arguments to a method and the bound variable (or
+variables) within a \texttt{for} statement are immutable;
+the variable \textit{result} is mutable.
+Using the \texttt{var} statement, new mutable local variables can be
+declared.
+For example, \texttt{var} $x = 3$ declares a new mutable local variable
+$x$.
+The \texttt{let} statement allows declaring new immutable local variables.
+For example:
 \texttt{let} $x = 3$: $y$ $+$$=$ $x$ adds 3 to the global variable
@@
 
+\begin{figure}
+\begin{code}
+\harmonysource{UpLock}
+\end{code}
+\caption{\harmonylink{code/UpLock.hny} Program of \autoref{fig:inc} fixed with a lock.}
+\label{fig:incfixed}
+\end{figure}
+
 So far we have used Harmony to \emph{implement} various algorithms.
@@
 
-The code in \autoref{fig:lockspec} is taken from Harmony's \texttt{synch}
-module.  \autoref{fig:cssynch} shows how locks may be used.  Note that
-the code of \autoref{fig:lockspec} is executable in Harmony.
+The code in \autoref{fig:lockspec} is similar to the code in
+Harmony's \texttt{synch} module.
+(The module generalizes locks to \emph{binary semaphores}, but the
+lock interface is the same.)
+\autoref{fig:cssynch} shows how locks may be used to implement
+a critical section.
+\autoref{fig:incfixed} gives an example of how locks may be used
+to fix the program of \autoref{fig:inc}.
+
+Note that the code of \autoref{fig:lockspec} is executable in Harmony.
 However, the \texttt{atomically} keyword is not available in general
@@
 Peterson's algorithm is an implementation of a lock, although only for
-two processes.  In the following chapters we will look at more general
+two processes.  In the following chapters, we will look at more general
 ways of implementing locks using atomic constructions that are usually
@@
 an actual executable program such as in \autoref{fig:cssynch} should not
-use the keyword.  In fact, if we want to make the entire program
-executable, we have to also show how \texttt{acquire} and \texttt{release}
+use the \texttt{atomically} keyword.
+In fact, if we want to make the entire program executable on hardware,
+we have to also show how \texttt{Lock}, \texttt{acquire}, and \texttt{release}
 are implemented, not just how they are specified.
@@
 condition holds (not necessarily the first time) and then executes the
-statement block.  The \texttt{await} \textit{condition} statement is
-the same as \texttt{when} \textit{condition}: \texttt{pass}.
+statement block.  The ``\texttt{await} \textit{condition}'' statement is
+the same as ``\texttt{when} \textit{condition}: \texttt{pass}''.
 Combined with the \texttt{atomically} keyword, the entire statement
 is executed atomically at a time that the condition holds.
+
+It is important to appreciate the difference between an
+\emph{atomic section} (the statements executed within an
+atomic block of statements) and a \emph{critical section}
+(protected by a lock of some sort).
+The former ensures that while the
+statements are executing no other thread can execute.
+The latter allows multiple threads to run concurrently,
+just not within the critical section.
+The former is rarely available to a programmer (e.g., none of
+Python, C, or Java support it), while the latter
+is very common.
+
+% In Harmony, atomic statements allow you to \emph{implement} your own
+% atomic primitives such as test-and-set.  Other common examples
+% include compare-and swap and fetch-and-add.
+Atomic statements
+are not intended to replace locks or other synchonization primitives.
+When solving synchronization problems you should not directly use
+atomic statements but use the synchronization primitives that are available
+to you.  But if you want to \emph{specify} a new synchronization primitive,
+then use \textbf{atomically} by all means.
+You can also use atomic statements in your test code.
+In fact, as mentioned before, \textbf{assert} statements are included to
+test if certain conditions hold in every execution and are executed
+atomically.
 
@@
 operation.
-Method \texttt{tas} in \autoref{fig:tas} shows its specification.
+Method \texttt{test\_and\_set} in \autoref{fig:tas} shows its specification.
 Here $s$ points to a shared Boolean variable and $p$
@@
 critical section at the same time.
-For those who read through the previous chapter:
+For those who read through \autoref{ch:peterson}
 $\mathcal{I}_1 \land \forall i: \mathcal{I}_2(i)$ is an inductive invariant.
@@
 To see that invariant $\mathcal{I}_1$ is maintained, note that
-\texttt{!}$p$ == \textbf{True} upon entry of \texttt{tas}
+\texttt{!}$p$ == \textbf{True} upon entry of \texttt{test\_and\_set}
 (because of the condition on the \textbf{while} loop that the
-\texttt{tas} method is invoked in).
+\texttt{test\_and\_set} method is invoked in).
 There are two cases:
 \begin{enumerate}
-\item \texttt{!}$s$ is \textbf{False} upon entry to \texttt{tas}.
+\item \texttt{!}$s$ is \textbf{False} upon entry to \texttt{test\_and\_set}.
 Then upon exit \texttt{!}$p$ == \textbf{False} and \texttt{!}$s$ == \textbf{True},
 maintaining the invariant.
-\item \texttt{!}$s$ is \textbf{True} upon entry to \texttt{tas}.
+\item \texttt{!}$s$ is \textbf{True} upon entry to \texttt{test\_and\_set}.
 Then upon exit nothing has changed, maintaining the invariant.
@@
 
-\chapter{Blocking}
+\chapter{Lock Implementations}
 \label{ch:synch}
@@
 \end{code}
-\caption{\harmonylink{code/taslock.hny} Implementation of a spinlock
-using test-and-set.}
+\caption{\harmonylink{code/taslock.hny} Implementation of the lock
+specification in \autoref{fig:lockspec} using a spinlock based on
+test-and-set.}
 \label{fig:tas2}
@@
 
-In \autoref{fig:tas} we have shown a solution based on a shared
-variable and a private variable for each thread.   The private
-variables themselves are actually implemented as shared variables,
-but they are accessed only by their respective threads.
-A thread usually does not keep explicit track of whether it has a lock
-or not, because it is implied by the control flow of the program---a
-thread implicitly \emph{knows} that when it is executing in a critical
-section it has the lock.
-There is no need to keep \textit{private} as a shared
-variable---we only did so to be able to show and check the invariants.
-\autoref{fig:tas2} shows a more straightforward implementation of spinlock.
-The solution is similar to the na\"{i}ve solution of \autoref{fig:naivelock},
-but uses test-and-set to check and set the lock variable atomically.
-The lock is also cleared in an atomic statement to prevent a data race.
-This approach is general for any number of threads.
-
-It is important to appreciate the difference between an
-\emph{atomic section} (the statements executed within an
-atomic block of statements) and a \emph{critical section}
-(protected by a lock of some sort).
-The former ensures that while the
-statements are executing no other thread can execute.
-The latter allows multiple threads to run concurrently,
-just not within the critical section.
-The former is rarely available to a programmer (e.g., none of
-Python, C, or Java support it), while the latter
-is very common.
-
-In Harmony, atomic statements allow you to \emph{implement} your own
-atomic primitives such as test-and-set.  Other common examples
-include compare-and swap and fetch-and-add.  Atomic statements
-are not intended to replace locks or other synchonization primitives.
-When solving synchronization problems you should not directly use
-atomic statements but use the synchronization primitives that are available
-to you.  But if you want to design a new synchronization primitive, then
-use \textbf{atomically} by all means.
-You can also use atomic statements in your test code.
-In fact, as mentioned before, \textbf{assert} statements are executed atomically.
-
-\begin{figure}
-\begin{code}
-\harmonysource{lockintf}
-\end{code}
-\caption{\harmonylink{modules/synch.hny} The binary semaphore interface and implementation in the \texttt{synch} module.}
-\label{fig:spinlocks}
-\end{figure}
-
-\begin{figure}
-\begin{code}
-\harmonysource{UpLock}
-\end{code}
-\caption{\harmonylink{code/UpLock.hny} Program of \autoref{fig:inc} fixed with a lock.}
-\label{fig:incfixed}
+\begin{figure}
+\begin{code}
+\harmonysource{ticket}
+\end{code}
+\caption{\harmonylink{code/ticket.hny} Implementation of the lock
+specification in \autoref{fig:lockspec} using a ticket lock.}
+\label{fig:ticket}
 \end{figure}
@@
 
+In \autoref{fig:tas} we have shown a lock implementation based on a shared
+variable and a private variable for each thread.   The private
+variables themselves are actually implemented as shared variables,
+but they are accessed only by their respective threads.
+A thread usually does not keep explicit track of whether it has a lock
+or not, because it is implied by the control flow of the program---a
+thread implicitly \emph{knows} that when it is executing in a critical
+section it has the lock.
+There is no need to keep \textit{private} as a shared
+variable---we only did so to be able to show and check the invariants.
+\autoref{fig:tas2} shows a more straightforward implementation of a spinlock.
+% The solution is similar to the na\"{i}ve solution of \autoref{fig:naivelock},
+% but uses test-and-set to check and set the lock variable atomically.
+The lock is also cleared in an atomic statement to prevent a data race.
+This approach is general for any number of threads.
+
+You can test the spinlock with the program in \autoref{fig:cssynch}
+using the command \texttt{harmony -m synch=taslock code/cssynch.hny}.
+The \texttt{-m} flag tells Harmony to use the \texttt{taslock.hny} file
+rather than the standard \texttt{synch} module (which only contains
+a specification of the lock methods).
+
+\index{starvation}%
+The spinlock implementation suffers potentially from \emph{starvation}:
+an unlucky thread may never be able to get the lock while other threads
+successfully acquire the lock one after another.  It could even happen
+with just two threads: one thread might successfully acquire the lock
+repeatedly in a loop, while another thread is never lucky enough the
+acquire the thread in between.
+\autoref{ch:starvation} will talk more about starvation and how
+to prevent it.
+
+A \emph{ticket lock} (\autoref{fig:ticket} is an implementation of a
+lock that prevents starvation using an atomic \emph{fetch-and-increment}
+operator.  It is inspired by European bakeries.  A European bakery has a
+clearly displayed counter (usually just two digits) and a ticket dispenser.
+Tickets are numbered 0 through 99 and repeat over and over again
+(in the case of a two digit counter).
+When a customer walks into the bakery, they draw a number from the dispenser
+and wait until their number comes up.
+Every time a customer has been helped, the counter is incremented.
+(Note that this only works if there can be no more than 100 customers in
+the bakery at a time.)
+
+\autoref{fig:ticket} similarly uses two variables for a lock,
+\textit{counter} and \textit{dispenser}.  When a thread acquires the
+lock, it fetches the current dispenser value and increments it
+modulo \texttt{MAX\_THREADS}, all in one atomic operation.
+In practice, \texttt{MAX\_THREADS} would
+be a number like $2^32$ or $2^64$, but since the Harmony model checker
+checks every possible state, limiting it to a small number such as done
+here will significantly reduce the time to model check a Harmony program.
+Plus it is easier to check that it fails when you run it with more than
+\texttt{MAX\_THREADS} threads.
+You can test the implementation using the command
+\texttt{harmony -m synch=ticket code/cssynch.hny}.
+To see it fail, try
+\texttt{harmony -c NTHREADS=10 -m synch=ticket code/cssynch.hny}.
+
 \begin{comment}
@@
 
+\begin{comment}
 \index{binary semaphore}
@@
 binary semaphore.
-
-Harmony has a module called \texttt{synch} that includes support for
-binary semaphores and locks.
-\autoref{fig:spinlocks} shows how they are implemented, and
-\autoref{fig:incfixed} gives an example of how they may be used,
-in this case to fix the program of \autoref{fig:inc}.
-Notice that the module hides the implementation of
-binary semaphores and locks.
-The \texttt{synch} module includes a variety of other useful
-synchronization primitives, which will be discussed in later
-chapters.
-Also notice the use of the sequentially consistent \textit{done}
-variable for testing purposes.
+\end{comment}
 
@@
 
+We now turn to a radically different way of implementing locks, one
+that is commonly provided by operating systems to user processes.
 We call a thread \emph{blocked}
@@
 another thread changes the state first.
-A thread trying to
-acquire a test-and-set spinlock held by another thread is a good example
+A thread trying to acquire a lock held by another thread is a good example
 of a thread being blocked.
-The only way forward is if the other thread releases the spinlock.
+The only way forward is if the other thread releases the lock.
 A thread that is in an infinite loop is also considered blocked.
-% although Harmony can often detect infinite loops specifically.
 
@@
 \end{code}
-\caption{\harmonylink{modules/synchS.hny} The binary semaphore interface in the \texttt{synchS} module uses suspension.}
+\caption{\harmonylink{modules/synchS.hny} The lock implementation in the \texttt{synchS} module uses suspension.}
 \label{fig:suspension}
@@
 
-There is a second version of the \texttt{synch} module, called \texttt{synchS},
-that uses suspension instead of spinlocks.
-\autoref{fig:suspension} shows the same interface
-as \autoref{fig:spinlocks} implemented using suspension.
-The interface is implemented as follows:
+\autoref{fig:suspension} shows the lock interface using suspension.
+It is implemented as follows:
 \begin{itemize}
-\item A binary semaphore maintains both a boolean indicating whether the
-binary semaphore is currently acquired and a list of contexts of threads that want to
-acquire the binary semaphore.
+\item A lock maintains both a boolean indicating whether the
+lock is currently acquired and a list of contexts of threads that want to
+acquire the lock.
 \item
@@
 \index{acquire}%
-acquires the binary semaphore if available and suspends the invoking thread if not.
+acquires the lock if available and suspends the invoking thread if not.
 In the latter case, the context of the thread is added to the end of the list of contexts.
@@
 \index{release}%
-checks to see if any threads are waiting to acquire the binary semaphore.
+checks to see if any threads are waiting to acquire the lock.
 If so, it uses the \texttt{head} and \texttt{tail}
@@
 \texttt{choose} to pick an arbitrary one.  Selecting the first is a common
-choice in binary semaphore implementations as it prevents \emph{starvation}:
-\index{starvation}%
-every thread
-gets a chance to acquire the binary semaphore (assuming every thread eventually releases
-it).  \autoref{ch:starvation} will talk more about starvation and how
-to prevent it.
-
-Harmony allows you to select which version of the \texttt{synch} module you would
-like to use with the \texttt{-m} flag.
-\index{module}%
-For example,
-
-\begin{code}
-\begin{verbatim}
-harmony -m synch=synchS x.hny
-\end{verbatim}
-\end{code}
-
-runs the file \texttt{x.hny} using the suspension version of the \texttt{synch} module.
-You will find that using the \texttt{synchS} module often leads to
-the model checker searching a
-significantly larger state space than using the \texttt{synch} module.
-Part of the reason is that the \texttt{synchS} module keeps track of the order
-in which threads wait to acquire a binary semaphore, while the \texttt{synch} module does not.
+choice in lock implementations as it prevents starvation.
+
+You will find that using the \emph{implementation} of a lock instead
+of the \emph{specification} of a lock (in the \texttt{synch} module)
+often leads to the model checker searching a
+significantly larger state space.
+Thus it makes sense to model check larger programs in a modular fashion:
+model check one module implementation at a time, using
+specifications for the other modules.
 

### 2021-11-07 book/paper.tex
Added queue spec to chapter

@@
 
+\begin{figure}
+\begin{code}
+\harmonysource{queuespec}
+\end{code}
+\caption{\harmonylink{code/queuespec.hny} A sequential specification for a queue.}
+\label{fig:queueseq}
+\end{figure}
+
+\begin{figure}
+\begin{code}
+\harmonysource{queue}
+\end{code}
+\caption{\harmonylink{code/queue.hny} A concurrent specification for a queue.}
+\label{fig:queue}
+\end{figure}
+
+\begin{figure}
+\begin{code}
+\harmonysource{queuedemo}
+\end{code}
+\caption{\harmonylink{code/queuedemo.hny} Using a concurrent queue.}
+\label{fig:queuedemo}
+\end{figure}
+
 The most common use for locks is in building concurrent data structures.
 By way of example, we will first demonstrate how to build a concurrent queue.
-The \texttt{queue} module will have the following API:
+The \texttt{queue} module has the following API:
 \begin{itemize}
@@
 \noindent
-See \autoref{fig:queuedemo} for a simple demonstration program that uses the queue.
+\autoref{fig:queueseq} shows a sequential specification for such a
+queue in Harmony (exploiting some methods from the \texttt{list}
+module described in \autoref{ap:list}),
+but it cannot be used with threads concurrently accessing this queue.
+\autoref{fig:queue} shows the corresponding concurrent specification.
+See \autoref{fig:queuedemo} for a simple demonstration program that uses
+a concurrent queue.
 
 We will first implement the queue as a linked list.
-The implementation in \autoref{fig:queue}
+The implementation in \autoref{fig:queueconc}
 uses the \texttt{alloc} module for dynamic allocation
@@
 \begin{code}
-\harmonysource{queuedemo}
-\end{code}
-\caption{\harmonylink{code/queuedemo.hny} Using a concurrent queue.}
-\label{fig:queuedemo}
-\end{figure}
-
-\begin{figure}
-\begin{code}
-\harmonysource{queue}
-\end{code}
-\caption{\harmonylink{code/queue.hny}A basic concurrent queue data structure.}
-\label{fig:queue}
+\harmonysource{queueconc}
+\end{code}
+\caption{\harmonylink{code/queueconc.hny} An implementation of a concurrent queue data structure.}
+\label{fig:queueconc}
 \end{figure}
@@
 \begin{code}
-\harmonysource{queuespec}
-\end{code}
-\caption{\harmonylink{code/queuespec.hny} Sequential queue specification.}
-\label{fig:queuespec}
-\end{figure}
-
-\begin{figure}
-\begin{code}
-\harmonysource{qtestseq}
-\end{code}
-\caption{\harmonylink{code/qtestseq.hny} Sequential queue test.}
-\label{fig:qtestseq}
-\end{figure}
-
-\begin{figure}
-\begin{code}
 \harmonysource{queuelin}
@@
 abstraction.
-\autoref{fig:queuespec} shows a sequential specification of a
-queue in Harmony (exploiting some methods from the \texttt{list}
-module described in \autoref{ap:list}).
+\autoref{fig:queueseq} shows a sequential specification of a
+queue in Harmony.
 

### 2021-11-07 book/paper.tex
Latest

@@
 \begin{code}
+\harmonysource{setobj}
+\end{code}
+\caption{\harmonylink{code/setobj.hny} Specification of a concurrent
+set object.}
+\label{fig:setobj}
+\end{figure}
+
+\begin{figure}
+\begin{code}
 \harmonysource{linkedlist}
 \end{code}
-\caption{\harmonylink{code/linkedlist.hny} Linked list with fine-grained locking.}
+\caption{\harmonylink{code/linkedlist.hny} Implementation of a set of integers using a linked list with fine-grained locking.}
 \label{fig:linkedlist}
@@
 \begin{code}
-\harmonysource{lltest}
-\end{code}
-\caption{\harmonylink{code/lltest.hny} Test code for \autoref{fig:linkedlist}.}
-\label{fig:lltest}
+\harmonysource{intsettest}
+\end{code}
+\caption{\harmonylink{code/intsettest.hny} Test code for \autoref{fig:linkedlist}.}
+\label{fig:intsettest}
 \end{figure}
@@
 
-\autoref{fig:linkedlist} implements an ordered linked
-list of integers without duplicates.
-(\autoref{fig:lltest} contains test code.)
-Values can be added using \texttt{insert} or deleted using \texttt{remove}.
-Method \texttt{contains} checks if a particular value is in the list.
-
+\autoref{fig:setobj} gives the specification of a concurrent set object.
+\texttt{SetObject()} returns a pointer to a variable that
+contains an empty set, rather than returning an empty set.
+Values can be added using \texttt{insert()} or deleted using \texttt{remove()}.
+Method \texttt{contains()} checks if a particular value is in the list.
+
+\autoref{fig:linkedlist} implements a concurrent set of integers using
+an ordered linked list of integers without duplicates.
+(\autoref{fig:intsettest} contains test code.)
 The list has two dummy ``book-end'' nodes with values \texttt{-inf} and

### 2021-11-12 book/paper.tex
First steps towards concurrent model checker

@@
 in \autoref{fig:queue} meets the sequential
-queue specification in \autoref{fig:queuespec}.
+queue specification in \autoref{fig:queueseq}.
 To check if the queue implementation meets the specification,
@@
 implementation of \autoref{fig:queuelin}, but they are executed sequentially
-on the sequential queue specification of \autoref{fig:queuespec}.
+on the sequential queue specification of \autoref{fig:queueseq}.
 To that end, the test program maintains a global time variable

### 2021-11-15 book/paper.tex
Updated the appendices to describe "when" statements

@@
 
-Harmony currently supports the following statements (below, \texttt{S} is a list of statements):
+Harmony currently supports the following statements
+(below, \texttt{S} is a list of statements):
 
@@
 \hline
-\texttt{atomic: S} & \texttt{S} a list of statements\\
-\hline
 \texttt{await $b$} & $b$ is a boolean\\
@@
 \hline
-\texttt{let $a$ = $e$ [let ...]: S} & $a$ is a bounded variable, $e$ is an expression\\
+\texttt{let $a$ = $e$: S} & $a$ is a bounded variable, $e$ is an expression\\
 \hline
@@
 \hline
+\texttt{var $v$ = $e$} & $v$ is a new variable, $e$ is an expression\\
+\hline
+\texttt{when $b$: S} & $b$ is a boolean, \texttt{S} a list of statements\\
+\hline
+\texttt{when exists $a$ in $e$: S} & $a$ is a bounded variable, $e$ is an expression\\
+\hline
 \texttt{while $b$: S} & $b$ is a boolean, \texttt{S} a list of statements\\
@@
 \end{tabular}
+
+\begin{itemize}
+\item Bounded variables are read-only.
+\item A statement can be preceded by the \texttt{atomically} keyword to make
+the statement atomic.
+\item Multiple \texttt{for} statements can be combined into a single statement.
+\item Multiple \texttt{let} and \texttt{when} statements can be combined into
+a single statement.
+\end{itemize}
 
@@
 
-The statement \textbf{atomic $S_1; S_2; ...$} evaluates statements
-$S_1, S_2, ...$ atomically in that no other threads can run while
-this atomic block is executing.  Typically an atomic block runs
-to completion before any other thread can run.  The only exception
+A statement can be preceded by the \texttt{atomically} keyword to
+make the statement atomic.
+The statement \textbf{atomically: $S_1; S_2; ...$} evaluates statements
+$S_1, S_2, ...$ atomically.
+This means that the statement runs indivisibly---no other thread can
+interleave in the atomic statement.  The only exception
 to this is if the atomic block executes a \textbf{stop} expression.
@@
 
-The \textbf{atomic} statement is useful for implementing synchronization
-primitives such as test-and-set.  It is also useful for testing.
+\textbf{atomically} statements are useful for specification and
+implementing synchronization primitives such as test-and-set.
+It is also useful for testing.
 It is not a replacement for lock/unlock, and should not generally be used
@@
 
-The statement \textbf{await $b$} is equivalent to \textbf{while !$b$: pass}.
+The statement \textbf{await $b$} is equivalent to \textbf{when $b$: pass}.
 It is intended to improve readability of your code.
@@
 Examples include $()$, $(x,)$, $(x, y)$, and $(x, (y, z))$.
-The given local variable names variable names are assigned
-upon application.
-It is allowed, but discouraged, to updates those local variables
-in statements $S_1, S_2, ...$.
+The given local variable names are assigned upon application and
+are read-only.
 Each method has a predefined local variable
@@
 
-You can introduce new local variables in a method using the
-\texttt{let} expression.
+You can introduce new bound variables in a method using the
+\texttt{let} statement.
 The statement \textbf{let} $a = b: S_1; S_2, ...$ evaluates $b$, assigns
-the result to local variable $a$, and evaluates statements
+the result to read-only variable $a$, and evaluates statements
 $S_1, S_2, ...$.  \textbf{let} supports pattern matching, so you can
@@
 tuple with two elements---if not, Harmony will report an error.
-The variables may be updates in statements $S_1, S_2, ...$.
 
@@
 go right.
-
-\subsection*{\textbf{select ... in ... [ where ... ]}}
-
-The statement \textbf{select} $x$ \textbf{in} $y: S_1; S_2; ...$
-requires that $y$ evaluates to a set value.
-The statement does the following three things atomically:
-\begin{itemize}
-\item it waits until $y$ is non-empty;
-\item it selects one element of $y$ non-deterministically (using
-a \texttt{choose} expression);
-\item it executes statements $S_1, S_2, ...$ with the selected element
-assigned to local variable $x$.
-\end{itemize}
-$x$ may be a pattern, like in \texttt{let}, \texttt{for}, and \texttt{def}
-statements.
-Harmony reports an error if $y$ evaluates to a value that is not a set.
-If \textit{waiting} is an unused local variable,
-then \textbf{select} $x$ \textbf{in} $y: S_1; S_2; ...$ is equivalent to
-
-\begin{code}
-\begin{tabbing}
-XXX\=XXX\=XXX\=XXX\=XXX\=XXX\=XXX\=XXX\kill
-\>\textbf{let} $\mathit{waiting} = \mathbf{True}:$ \\
-\>\>\textbf{while} \textit{waiting}: \\
-\>\>\>\textbf{atomic}: \\
-\>\>\>\>\textbf{if} $y$ != \{\}:\\
-\>\>\>\>\>\textbf{let} $x = \mathbf{choose}(y)$:\\
-\>\>\>\>\>\>$S_1$ \\
-\>\>\>\>\>\>$S_2$ \\
-\>\>\>\>\>\>$...$ \\
-\>\>\>\>\>$\mathit{waiting} = \mathbf{False}$
-\end{tabbing}
-\end{code}
-
-The statement is particularly useful in programming network protocols
-when having to wait for one or more messages and executing a set
-of actions atomically after the desired messages have arrived.
-
-The optional \textbf{where} clause can be used to filter the set down to the elements
-that satisfy the condition following the \textbf{where} keyword.
 
@@
 
+\subsection*{\textbf{var}}
+
+You can introduce new local variables in a method using the
+\texttt{var} statement.
+The statement \textbf{var} $a = b$ evaluates $b$ and assigns
+the result to local variable $a$.
+\textbf{var} supports pattern matching, so you can
+write \texttt{var} $x, (y, z) = b$.  This will only work
+if $b$ is a tuple with two elements, the second of which also being a
+tuple with two elements---if not, Harmony will report an error.
+
+\subsection*{\textbf{when}}
+
+The statement \textbf{when} $c: S_1; S_2; ...$ executes statements
+$S_1, S_2, ...$ after waiting until $c$ evaluates to \textbf{True}.
+\texttt{when} statements are most useful when combined with the
+\texttt{atomically} keyword.
+If \textit{waiting} is an unused local variable,
+then \textbf{atomically when} $c: S_1; S_2; ...$ is equivalent to
+
+\begin{code}
+\begin{tabbing}
+XXX\=XXX\=XXX\=XXX\=XXX\=XXX\=XXX\=XXX\kill
+\>\textbf{var} $\mathit{waiting} = \mathbf{True}$ \\
+\>\textbf{while} \textit{waiting}: \\
+\>\>\textbf{atomically}: \\
+\>\>\>\textbf{if} $c$:\\
+\>\>\>\>$S_1$ \\
+\>\>\>\>$S_2$ \\
+\>\>\>\>$...$ \\
+\>\>\>\>$\mathit{waiting} = \mathbf{False}$
+\end{tabbing}
+\end{code}
+
+Multiple \texttt{let} and \texttt{when} statements can be combined.
+The expressions before the colon are re-evaluated repeated until all
+\texttt{when} conditions are satisfied.
+
+\subsection*{\textbf{when exists ... in ...}}
+
+The statement \textbf{when exists} $x$ \textbf{in} $y: S_1; S_2; ...$
+requires that $y$ evaluates to a set value.
+The statement does the following three things:
+\begin{itemize}
+\item it waits until $y$ is non-empty;
+\item it selects one element of $y$ non-deterministically (using
+a \texttt{choose} expression);
+\item it executes statements $S_1, S_2, ...$ with the selected element
+bound to read-only variable $x$.
+\end{itemize}
+$x$ may be a pattern, like in \texttt{let}, \texttt{for}, and \texttt{def}
+statements.
+Harmony reports an error if $y$ evaluates to a value that is not a set.
+
+\texttt{when} statements are most useful when combined with the
+\texttt{atomically} keyword.
+If \textit{waiting} is an unused local variable,
+then \textbf{atomically when exists} $x$ \textbf{in} $y: S_1; S_2; ...$ is equivalent to
+
+\begin{code}
+\begin{tabbing}
+XXX\=XXX\=XXX\=XXX\=XXX\=XXX\=XXX\=XXX\kill
+\>\textbf{var} $\mathit{waiting} = \mathbf{True}:$ \\
+\>\textbf{while} \textit{waiting}: \\
+\>\>\textbf{atomic}: \\
+\>\>\>\textbf{if} $y$ != \{\}:\\
+\>\>\>\>\textbf{let} $x = \mathbf{choose}(y)$:\\
+\>\>\>\>\>$S_1$ \\
+\>\>\>\>\>$S_2$ \\
+\>\>\>\>\>$...$ \\
+\>\>\>\>$\mathit{waiting} = \mathbf{False}$
+\end{tabbing}
+\end{code}
+
+The statement is particularly useful in programming network protocols
+when having to wait for one or more messages and executing a set
+of actions atomically after the desired messages have arrived.
+
 \subsection*{\textbf{while}}
@@
 Each (non-reserved) identifier in a Harmony program refers to either
-a constant, a global shared variable, a local variable, or a module.
+a global constant, a global shared variable, a local bound variable,
+a local mutable variable, or a module.
 Constants are declared using \texttt{const} statements.
@@
 
-Local variables all declared.  They can be declared in
+Local variables all declared.
+Bound variables can be declared in
 \texttt{def} statements (i.e., arguments),
-\texttt{let} statements, \texttt{for} loops, and \texttt{select} statements.
+\texttt{let} statements, \texttt{for} loops, and \texttt{when exists} statements.
+Mutable variables can be declared with the \texttt{var} statement.
 Also, each method has an implicitly declared
 \textit{result} variable, which is initialized to \texttt{None}.
-Each thread has a variable called \textit{this} that contains
+Each thread has a mutable variable called \textit{this} that contains
 the thread-local state.
@@
 before any threads are spawned.
-
-While arguments to a method and variables in \texttt{for} loops
-can be modified, we discourage it for improved code readability.
 

### 2021-11-15 src/charm/charm.c

@@ -1718,7 +1718,7 @@ int main(int argc, char **argv){
         fclose(df);
     }
 
-    if (false) {
+    if (true) {
         FILE *df = fopen("charm.dump", "w");
         assert(df != NULL);
         char **table = malloc(global->graph.size * sizeof(char*));
@@ -1728,34 +1728,6 @@ int main(int argc, char **argv){
             fprintf(df, "%s\n", table[i]);
         }
         fclose(df);
-        for (int i = 0; i < global->graph.size; i++) {
-            for (int j = i + 1; j < global->graph.size; j++) {
-                if (strcmp(table[i], table[j]) == 0) {
-                    uint64_t *cb1, *cb2;
-                    int n1, n2;
-                    cb1 = value_get(global->graph.nodes[i]->state->ctxbag, &n1);
-                    cb2 = value_get(global->graph.nodes[j]->state->ctxbag, &n2);
-                    printf("SAME %d %d %d %llx %llx\n", i, j,
-                        memcmp(
-                            global->graph.nodes[i]->state,
-                            global->graph.nodes[j]->state,
-                            sizeof(struct state)
-                        ),
-                        (uint64_t) cb1,
-                        (uint64_t) cb2
-                    );
-                    void dict_check(struct dict *dict, const void *key, unsigned int keyn);
-                    dict_check(global->values.dicts, cb1, n1);
-                    dict_check(global->values.dicts, cb2, n2);
-                    printf("SIZE %d %d\n", n1, n2);
-                    n1 /= sizeof(uint64_t);
-                    for (int k = 0; k < n1; k++) {
-                        printf("%llx %llx\n", *cb1++, *cb2++);
-                    }
-                    exit(0);
-                }
-            }
-        }
     }
 
     // Look for data races

### 2021-11-15 book/paper.tex
Got rid of any mention of select statements in the book

@@
 leader in the second phase of the protocol.
-A processor waits for a message using the Harmony \textbf{select} statement.
-This statement takes the form \textbf{select} $v$ \textbf{in} $s$: \emph{statement block}
+A processor waits for a message using the Harmony
+\textbf{atomically when exists} statement.
+This statement takes the form \textbf{atomically when exists} $v$ \textbf{in} $s$: \emph{statement block}
 where $s$ is a set and $v$ is variable that is bound to an element of $s$.
-The properties of the \textbf{select} statement are as follows:
+The properties of the statement are as follows:
 \begin{itemize}
@@
 every possible order is tried (including receiving the same message
-multiple times).  Fortunately, the \texttt{select} statement is executed
+multiple times).  Fortunately, the \texttt{atomically when exists} statement
+is executed
 atomically, otherwise the body of the statement could lead to additional
@@
 as tests.
-The \texttt{receive()} method is used by coordinators in a \textbf{select}
+The \texttt{receive()} method is used by coordinators in an
+\textbf{atomically when exists}
 statement to wait for a response from each bank involved in a transaction.
@@
 In Line~22, a processor waits for $\texttt{N} - \texttt{F}$ messages using
-the Harmony \textbf{select} statement.
+the Harmony \textbf{atomically when exists} statement.
 Since Harmony has to check all possible executions of the protocol, the
@@
 and then uses the \texttt{bag.combinations} method to find all combinations of size $k$.
-The \textbf{select} statement waits until there is at least one such combination and
+The \textbf{atomically when exists} statement waits until there is at least one such combination and
 then chooses an element, which is bound to the \textit{quorum} variable.
-The body of the \textbf{select} statement is then executed atomically.
+The body of the statement is then executed atomically.
 This is usually how distributed algorithms are modeled, because they can only interact
@@
 
-The body of the \textbf{select} statement contains the core of the algorithm.
+The body of the \textbf{atomically when exists} statement contains the core of the algorithm.
 Note that $\texttt{N} - \texttt{F} = 2\texttt{F} + 1$, so that the number of
@@
 Pattern matching can also be used in \texttt{const}, \texttt{let},
-and \texttt{select} statements.
+and \texttt{when exists} statements.
 

### 2021-11-15 book/paper.tex
Updated book

@@
 data structures.
+\item \autoref{ch:finegrained} gives an example of fine-grained locking
+methods that allow more concurrency than coarse-grained approaches..
 \item \autoref{ch:testing} discusses approaches to testing
@@
 \end{center}
-\caption{ A sequential and a concurrent program.}
+\caption{A sequential and a concurrent program.}
 \label{fig:progs}
@@
 
-\begin{figure}
-\begin{code}
+\begin{figure}[h]
+\begin{center}
+\begin{tabular}{cc}
+\begin{tcolorbox}[width=0.41\linewidth]
 \harmonysource{queuespec}
-\end{code}
-\caption{\harmonylink{code/queuespec.hny} A sequential specification for a queue.}
-\label{fig:queueseq}
-\end{figure}
-
-\begin{figure}
-\begin{code}
+\end{tcolorbox}
+&
+\begin{tcolorbox}[width=0.52\linewidth]
 \harmonysource{queue}
-\end{code}
-\caption{\harmonylink{code/queue.hny} A concurrent specification for a queue.}
-\label{fig:queue}
+\end{tcolorbox}
+\\
+(a) \harmonylink{code/queuespec.hny} Sequential & (b) \harmonylink{code/queue.hny} Concurrent queue
+\end{tabular}
+\end{center}
+\caption{A sequential and a concurrent specification of a queue.}
+\label{fig:queues}
 \end{figure}
@@
 \noindent
-\autoref{fig:queueseq} shows a sequential specification for such a
+\autoref{fig:queues}(a) shows a sequential specification for such a
 queue in Harmony (exploiting some methods from the \texttt{list}
@@
 but it cannot be used with threads concurrently accessing this queue.
-\autoref{fig:queue} shows the corresponding concurrent specification.
+\autoref{fig:queues}(b) shows the corresponding concurrent specification.
 See \autoref{fig:queuedemo} for a simple demonstration program that uses
@@
 
-\texttt{queue.Queue()} returns a new queue object consisting of a \texttt{None} head
+\texttt{Queue()} returns a new queue object consisting of a \texttt{None} head
 and tail pointer and a lock.
-$\mathtt{queue.put}(q, v)$ and $\mathtt{queue.get}(q)$ both take a pointer $q$ to the
+The $\mathtt{put}(q, v)$ and $\mathtt{get}(q)$ methods both take a pointer $q$ to the
 queue object because both may modify the queue.
@@
 
+\section*{Exercises}
+\begin{problems}
+\item \label{ex:qcontains} Add a method $\mathtt{contains}(q, v)$ to \autoref{fig:queues}(b)
+that checks to see if $v$ is in queue $q$.
+\item Add a method $\mathtt{length}(q)$ to \autoref{fig:queueconc} that returns the length of the given queue.
+The complexity of the method should be $O(1)$, which is to say that you should maintain
+the length of the queue as a field member and update it in \texttt{put} and \texttt{get}.
+\item \label{ex:qlength} Write a method $\mathtt{check}(q)$
+that checks the integrity of the queue in \autoref{fig:queueconc}.
+In particular, it should check the following integrity properties:
+\begin{itemize}
+\item If the list is empty, $q$$\rightarrow$$\mathtt{tail}$ should be \texttt{None}.
+Otherwise, the last element in the linked list starting from $q$$\rightarrow$$\mathtt{head}$
+should equal $q$$\rightarrow$$\mathtt{head}$. Moreover,
+$q$$\rightarrow$$\texttt{tail}$$\rightarrow$$next$ should be \texttt{None};
+\item The length field that you added in \autoref{ex:qlength} should equal the length
+of the list.
+\end{itemize}
+Method $\mathtt{check}(q)$ should not obtain a lock; instead add the following line
+just before releasing the lock in \texttt{put} and \texttt{get}:
+\begin{code}
+\textbf{assert} \texttt{check()}
+\end{code}
+\item \label{ex:qremove} Add a method $\mathtt{remove}(q, v)$ to \autoref{fig:queueconc}
+that removes all occurrences of $v$, if any, from queue $q$.
+\item The test program in \autoref{fig:queuedemo} is a not thorough test program.
+Design and implement a test program for \autoref{fig:queuedemo}.
+Make sure you \emph{test} the test program by trying it out against some buggy
+queue implementations.
+\end{problems}
+
+\chapter{Fine-Grained Locking}
+\label{ch:finegrained}
+
 \begin{figure}
@@
 \glsadd{linearizable}%
-Like the queue in \autoref{fig:queue}, the implementation of the list is
+Like the queue in \autoref{fig:queues}(b), the implementation of the list is
 \emph{linearizable}~\cite{HW90},
@@
 \begin{problems}
-\item \label{ex:qcontains} Add a method $\mathtt{contains}(q, v)$ to \autoref{fig:queue}
-that checks to see if $v$ is in queue $q$.
-\item Add a method $\mathtt{length}(q)$ to \autoref{fig:queue} that returns the length of the given queue.
-The complexity of the method should be $O(1)$, which is to say that you should maintain
-the length of the queue as a field member and update it in \texttt{put} and \texttt{get}.
-\item \label{ex:qlength} Write a method $\mathtt{check}(q)$
-that checks the integrity of the queue in \autoref{fig:queue}.
-In particular, it should check the following integrity properties:
-\begin{itemize}
-\item If the list is empty, $q$$\rightarrow$$\mathtt{tail}$ should be \texttt{None}.
-Otherwise, the last element in the linked list starting from $q$$\rightarrow$$\mathtt{head}$
-should equal $q$$\rightarrow$$\mathtt{head}$. Moreover,
-$q$$\rightarrow$$\texttt{tail}$$\rightarrow$$next$ should be \texttt{None};
-\item The length field that you added in \autoref{ex:qlength} should equal the length
-of the list.
-\end{itemize}
-Method $\mathtt{check}(q)$ should not obtain a lock; instead add the following line
-just before releasing the lock in \texttt{put} and \texttt{get}:
-\begin{code}
-\textbf{assert} \texttt{check()}
-\end{code}
-\item \label{ex:qremove} Add a method $\mathtt{remove}(q, v)$ to \autoref{fig:queue}
-that removes all occurrences of $v$, if any, from queue $q$.
-\item The test program in \autoref{fig:queuedemo} is a not thorough test program.
-Design and implement a test program for \autoref{fig:queuedemo}.
-Make sure you \emph{test} the test program by trying it out against some buggy
-queue implementations.
-\item Create a thread-safe sorted binary tree.  Implement a module
-\texttt{bintree} with methods
 \item Add methods to the data structure in \autoref{fig:linkedlist}
@@
 \label{fig:cslock}
+\end{figure}
+\end{comment}
+
+\begin{figure}
+\begin{code}
+\harmonysource{qtestseq}
+\end{code}
+\caption{\harmonylink{code/qtestseq.hny} Sequential queue test.}
+\label{fig:qtestseq}
 \end{figure}
@@
 may be used, but it is not a very thorough test program
-for an implementation such as the one in \autoref{fig:queue}
+for an implementation such as the one in \autoref{fig:queueconc}
 and does little to increase our confidence in its correctness.
@@
 
+\begin{comment}
 The framework in \autoref{fig:csbarebones} works well for thoroughly
@@
 Mutual exclusion holds if the counter never goes over 1.
-
-\emph{Test programs themselves should be tested}.
-Just because a test program works with a particular implementation
-does not mean the implementation is correct---it may be that the implementation
-is incorrect but the test program does not have enough coverage to find any
-bugs in the implementation.
-In the case of testing mutual exclusion with atomic counters, you will want
-to see if you can use this method to find the problems in \autoref{fig:naivelock},
-\autoref{fig:naiveflags},
-and \autoref{fig:naiveturn}.
-Conversely, a test program may be broken in that it finds bugs that do not exist.
-So, in this case,
-you also want to check if an atomic counter solution works with known
-correct solutions such as \autoref{fig:peterson} and \autoref{fig:tas}
-if possible.
+\end{comment}
 
@@
 we need a specification for its correctness.
-A good place to start is thinking about a \emph{sequential} specification
-for queues.
-A specification is simply a program, written at a high level of
-abstraction.
-\autoref{fig:queueseq} shows a sequential specification of a
+\autoref{fig:queues}(a) shows a sequential specification of a
 queue in Harmony.
-
 First we can check if the queue implementation
-in \autoref{fig:queue} meets the sequential
-queue specification in \autoref{fig:queueseq}.
+in \autoref{fig:queueconc} meets the sequential
+queue specification in \autoref{fig:queues}(a).
 To check if the queue implementation meets the specification,
@@
 
+\emph{Test programs themselves should be tested}.
+Just because a test program works with a particular implementation
+does not mean the implementation is correct---it may be that the implementation
+is incorrect but the test program does not have enough coverage to find any
+bugs in the implementation.
+So run a test program like this with a variety of queue implementations that
+have known bugs in them and make sure that the test program finds them.
+\begin{comment}
+In the case of testing mutual exclusion with atomic counters, you will want
+to see if you can use this method to find the problems in \autoref{fig:naivelock},
+\autoref{fig:naiveflags},
+and \autoref{fig:naiveturn}.
+\end{comment}
+Conversely, a test program may be broken in that it finds bugs that do not exist.
+\begin{comment}
+So, in this case,
+you also want to check if an atomic counter solution works with known
+correct solutions such as \autoref{fig:peterson} and \autoref{fig:tas}
+if possible.
+\end{comment}
+
 As with any other test program,
@@
 tricky bug to find.  You can comment out those lines
-in \autoref{fig:queue} and run the test program---it should
+in \autoref{fig:queueconc} and run the test program---it should
 easily find the bug and explain exactly how
@@
 in general.  If you remove all \texttt{acquire()} and
-\texttt{release()} calls from \autoref{fig:queue},
+\texttt{release()} calls from \autoref{fig:queueconc},
 the test program will not (and should not)
@@
 linearizable.
-The queue implementation in \autoref{fig:queue} does
+The queue implementation in \autoref{fig:queueconc} does
 not quite match this pattern, as the \texttt{put}
@@
 Still, it would be useful to check in Harmony that
-\autoref{fig:queue} is linearizable.
+\autoref{fig:queueconc} is linearizable.
 To do this, instead of applying the operations sequentially,
@@
 execution the concurrent execution must align with.
-\autoref{fig:queuelin} is a copy of \autoref{fig:queue} extended with
+\autoref{fig:queuelin} is a copy of \autoref{fig:queueconc} extended with
 linearization points.
@@
 implementation of \autoref{fig:queuelin}, but they are executed sequentially
-on the sequential queue specification of \autoref{fig:queueseq}.
+on the sequential queue specification of \autoref{fig:queues}(a).
 To that end, the test program maintains a global time variable
@@
 To fix the code without changing the data structure, we can use
-hand-over-hand locking (\autoref{ch:cds}).
+hand-over-hand locking (\autoref{ch:finegrained}).
 \autoref{fig:queuefix} shows an implementation

### 2021-11-15 book/paper.tex
Added spec of RW locks

@@
 \begin{code}
+\harmonysource{RW}
+\end{code}
+\caption{\harmonylink{code/RW.hny} Specification of reader/writer locks.}
+\label{fig:rwspec}
+\end{figure}
+
+\begin{figure}
+\begin{code}
 \harmonysource{RWtest}
@@
 \begin{code}
-\harmonysource{RW}
-\end{code}
-\caption{\harmonylink{code/RW.hny} Busy-Waiting Reader/Writer Lock implementation.}
+\harmonysource{RWbusy}
+\end{code}
+\caption{\harmonylink{code/RWbusy.hny} Busy-Waiting Reader/Writer Lock implementation.}
 \label{fig:rwbusy}
@@
 A reader/writer lock is an object whose abstract (and opaque)
-state contains two integer counters:
+state contains two integer counters (see \autoref{fig:rwspec}):
 \begin{enumerate}

### 2021-11-15 book/paper.tex
Latest

@@
 
+\begin{figure}
+\begin{code}
+\harmonysource{boundedbuffer}
+\end{code}
+\caption{\harmonylink{code/boundedbuffer.hny} Bounded buffer specification.}
+\label{fig:boundedbuffer}
+\end{figure}
+
 \section{Bounded Buffer}
@@
 dequeuer must wait.
-This problem is known as the ``Producer/Consumer Problem'' and was
+\autoref{fig:boundedbuffer} specifies a bounded buffer.
+Coming up with a good implementation is known as the
+``Producer/Consumer Problem'' and was
 proposed by Dijkstra~\cite{EWD329}.
@@
 consume responses, while the server consumes requests and produces responses.
+
+\index{flow control}%
+Unlike an ordinary queue, where queues can grow arbitrarily, bounded buffers
+provide \emph{flow control}: if the consumer runs faster than the producer
+(or producers), it will automatically block until there are new requests.
+Similarly, if the producers add requests at a rate that is higher than the
+consumers can deal with, the producers are blocked.  While a buffer of size 1
+already provides those properties, a larger buffer is able to deal with short
+spikes without blocking anybody.
 

### 2021-11-15 book/paper.tex
Some more minor updates to the book

@@
 
-\index{ring buffer}%
-A \emph{bounded buffer} (aka \emph{ring buffer}) is
+A \emph{bounded buffer} is
 a queue with the usual \texttt{put/get} interface,
-but implemented using a circular buffer
-\index{circular buffer}%
-of a certain length and two pointers:
-the \emph{tail} points where new items are enqueued and the \emph{head}
-points where items are dequeued.
+but implemented using a buffer
+of a certain maximum length.
 If the buffer is full, the enqueuer must wait; if the buffer is empty, the
@@
 \autoref{fig:boundedbuffer} specifies a bounded buffer.
+It is similar to the implementation in \autoref{fig:queues}(b) but adds
+checking for bounds.
 Coming up with a good implementation is known as the

### 2021-11-22 book/paper.tex
Made some minor improvements to the book

@@
 Use parentheses liberally to remove ambiguity.
-\item Harmony does not (currently) support floating point, iterators, or I/O;
-Harmony does support \textbf{for} loops and various ``comprehensions.''
+\item Harmony does not support floating point or I/O;
 \item Python is object-oriented, supporting classes with methods and
@@
 \end{itemize}
-There are also less important ones that you will discover as
+There are also less important differences that you will discover as
 you get more familiar with programming in Harmony.
@@
 the name listed in the caption of the example.)
-The example is sequential and has a method \texttt{triangle} that takes
+The example is a sequential program
+and has a method \texttt{triangle} that takes
 an integer number as argument.  Each method has a variable called
@@
 method (there is no \textbf{return} statement in Harmony).  The method
-also has a variable called \textit{n} containing the value of the
+also has a bound variable called \textit{n} containing the value of the
 argument.  The \{ \textit{x..y} \} notation generates a set containing the numbers
@@
 \begin{problems}
-\item Harmony programs can usually be easily translated into Python.  For example,
+\item Harmony programs can usually be easily translated into
+Python by hand.  For example,
 \autoref{fig:incpy} is a Python version of \autoref{fig:inc}.
@@
 For example, \texttt{(1,)} is a tuple consisting just of the number 1.
-Importantly, $(1) = 1 \ne (1,)$.
+Importantly, $(1) = 1 \ne (1,) = \{ 0:1 \}$.
 
@@
 We would like to place an assertion at the \texttt{@cs} label that
-specifies that only the current thread can be there.
+specifies that only one (the current) thread can be there.
 
@@
 Unfortunately, if we run the program through Harmony,
-we find that the assertion still fails.
+we find that the assertion fails.
 \autoref{fig:naiveLockhtml} shows the Harmony output.
@@
 To see why, first note the following invariant: if thread $i$ is in the
-critical section, then \textit{flags}[$i$] == \textbf{True}.
+critical section, then \textit{flags}[$i$] = \textbf{True}.
 Without loss of generality,
@@
 Thread 0 can only reach the critical section if at some time $t_1$,
-$t_1 > t_0$, it finds that \textit{flags}[1] == \textbf{False}.
-Because of the invariant, \textit{flags}[1] == \textbf{False} implies that
+$t_1 > t_0$, it finds that \textit{flags}[1] = \textbf{False}.
+Because of the invariant, \textit{flags}[1] = \textbf{False} implies that
 thread~1 is not in the critical section at time $t_1$.
@@
 It is easy to see that thread~1 cannot enter the critical section
-between $t_1$ and $t_2$, because \textit{flags}[1] == \textbf{False} at
+between $t_1$ and $t_2$, because \textit{flags}[1] = \textbf{False} at
 time $t_1$.  To reach the critical section between $t_1$ and $t_2$,
 it would first have to set \textit{flags}[1] to \textbf{True} and
-then wait until \textit{flags}[0] == \textbf{False}.  But that does not happen
+then wait until \textit{flags}[0] = \textbf{False}.  But that does not happen
 until time $t_2$.
@@
 An invariant of this solution is that while thread~$i$ is in the critical
-section, \textit{turn} == $i$.
+section, $\mathit{turn} = i$.
 Since \textit{turn} cannot be 0 and 1 at
@@
 until either the other thread is nowhere near the critical section
-($\mathit{flag}[1 - \mathit{self}] == \mathbf{False}$) or has given way
-($\mathit{turn} == \mathit{self}$).
+($\mathit{flag}[1 - \mathit{self}] = \mathbf{False}$) or has given way
+($\mathit{turn} = \mathit{self}$).
 Running the algorithm with Harmony shows that it satisfies both mutual
@@
 Because thread~0 entered the critical section, we know that
-\textit{flags}[1] = \textbf{False} or \textit{turn} == 0 because
+$\mathit{flags}[1] = \textbf{False}$ or $\mathit{turn} = 0$ because
 of the \texttt{await} condition.
@@
 expression \texttt{!}$p$ is the value of the shared variable.
-In particular, \texttt{!?}$x$ == $x$.
-This is similar to how C pointers work (\texttt{*\string&}$x$ == $x$).
+In particular, \texttt{!?}$x$ = $x$.
+This is similar to how C pointers work (\texttt{*\string&}$x$ = $x$).
 
@@
 Harmony's \texttt{synch} module.
-(The module generalizes locks to \emph{binary semaphores}, but the
-lock interface is the same.)
+(The module generalizes locks to \emph{binary semaphores}
+(\autoref{ch:sbs}), but the lock interface is the same.)
 \autoref{fig:cssynch} shows how locks may be used to implement
@@
 The \texttt{atomically} keyword can be used to specify the behavior of
-methods such as \texttt{acquire} and \texttt{release}.  But note that
+methods such as \texttt{acquire} and \texttt{release}.  But
 an actual executable program such as in \autoref{fig:cssynch} should not
@@
 are not intended to replace locks or other synchonization primitives.
-When solving synchronization problems you should not directly use
+When implementing synchronization solutions you should not directly use
 atomic statements but use the synchronization primitives that are available
@@
 lines $18 \cdots 22$ (which includes the critical section),
-then \textit{private}[$i$] == \textbf{False}.
+then \textit{private}[$i$] = \textbf{False}.
 Between the two (i.e., $\mathcal{I}_1 \land \forall i: \mathcal{I}_2(i)$),
@@
 To see that invariant $\mathcal{I}_1$ is maintained, note that
-\texttt{!}$p$ == \textbf{True} upon entry of \texttt{test\_and\_set}
+\texttt{!}$p$ = \textbf{True} upon entry of \texttt{test\_and\_set}
 (because of the condition on the \textbf{while} loop that the
@@
 \item \texttt{!}$s$ is \textbf{False} upon entry to \texttt{test\_and\_set}.
-Then upon exit \texttt{!}$p$ == \textbf{False} and \texttt{!}$s$ == \textbf{True},
+Then upon exit \texttt{!}$p$ = \textbf{False} and \texttt{!}$s$ = \textbf{True},
 maintaining the invariant.
@@
 acquire the thread in between.
-\autoref{ch:starvation} will talk more about starvation and how
-to prevent it.
+% \autoref{ch:starvation} will talk more about starvation and how
+% to prevent it.
 
@@
 In practice, \texttt{MAX\_THREADS} would
-be a number like $2^32$ or $2^64$, but since the Harmony model checker
-checks every possible state, limiting it to a small number such as done
+be a number like $2^{32}$ or $2^{64}$, but since the Harmony model checker
+checks every possible state, limiting \texttt{MAX\_THREADS}
+to a small number such as done
 here will significantly reduce the time to model check a Harmony program.
@@
 A thread that wants to read first waits until there are no writers:
-\textit{nwriters} == 0.  Then it increments the number of readers.
+\textit{nwriters} = 0.  Then it increments the number of readers.
 Similarly, a
@@
 A bound lock
-with \texttt{M == 1} is an ordinary lock.
+with \texttt{M = 1} is an ordinary lock.
 You should define a constant \texttt{M} and two methods:
@@
 Only the last invocation of \texttt{app\_recv}(\textit{reliable}) is invoked with
-\textit{reliable} == \textbf{True}.  It causes the last acknowledgment to be sent
+\textit{reliable} = \textbf{True}.  It causes the last acknowledgment to be sent
 reliably.  It allows the receiver (Bob) to stop, as well as the sender (Alice)
@@
 \hline
-\texttt{const a = $e$} & \texttt{a} is a bounded variable, $e$ is a constant expression\\
-\hline
-\texttt{def m $a$: S} & \texttt{m} is an identifier, $a$ a bounded variable\\
+\texttt{const a = $e$} & \texttt{a} is a bound variable, $e$ is a constant expression\\
+\hline
+\texttt{def m $a$: S} & \texttt{m} is an identifier, $a$ a bound variable\\
 \hline
@@
 \hline
-\texttt{for $a[:b]$ in $e$ [where $c$]: S} & $a$ and $b$ are bounded variables, $e$ is a set or dictionary \\
+\texttt{for $a[:b]$ in $e$ [where $c$]: S} & $a$ and $b$ are bound variables, $e$ is a set or dictionary \\
 \hline
@@
 \hline
-\texttt{let $a$ = $e$: S} & $a$ is a bounded variable, $e$ is an expression\\
+\texttt{let $a$ = $e$: S} & $a$ is a bound variable, $e$ is an expression\\
 \hline
@@
 \hline
-\texttt{select $a$ in $e$: S} & $a$ is a bounded variable, $e$ is a set,
+\texttt{select $a$ in $e$: S} & $a$ is a bound variable, $e$ is a set,
                             \texttt{S} a list of statements\\
@@
 \hline
-\texttt{when exists $a$ in $e$: S} & $a$ is a bounded variable, $e$ is an expression\\
+\texttt{when exists $a$ in $e$: S} & $a$ is a bound variable, $e$ is an expression\\
 \hline
@@
 \begin{itemize}
-\item Bounded variables are read-only.
+\item Bound variables are read-only.
 \item A statement can be preceded by the \texttt{atomically} keyword to make

### 2021-12-03 book/sources/pp.py
Starting to incorporate nice pics into the book

@@ -93,7 +93,7 @@ def isreserved(s):
         "not",
         "or",
         "pass",
-        "possibly",
+        "print",
         "select",
         "sequential",
         "setintlevel",
@@ -249,7 +249,7 @@ def lexer(s, file):
 
         # string
         if s[0] == '"':
-            print("````", end="")
+            print('\\texttt{"}', end="")
             i = 1
             str = '"'
             while i < len(s) and s[i] != '"':
@@ -278,7 +278,7 @@ def lexer(s, file):
             if i < len(s):
                 i += 1
             str += '"'
-            print("''''", end="")
+            print('\\texttt{"}', end="")
             nextconst = False
             importline = False
             result += [ (str, file, line, column) ]

### 2021-12-03 book/paper.tex
Starting to incorporate nice pics into the book
Starting to work on book again...

@@
 properties and does not suffer from bugs.
-Interestingly, Harmony does not have input/output statements, so there is
-no ``Hello World'' program.
 
@@
 
+\chapter{Hello World}
+\label{ch:helloworld}
+
+\begin{figure}[h]
+\begin{center}
+\begin{tabular}{ccc}
+\includegraphics[width=0.35\textwidth]{../code/hello1.png}
+& \rule{0.05\linewidth}{0in} &
+\includegraphics[width=0.45\textwidth]{../code/hello2.png} \\
+\begin{tcolorbox}[width=0.4\linewidth]
+\harmonysource{hello1}
+\end{tcolorbox}
+& \rule{0.05\linewidth}{0in} &
+\begin{tcolorbox}[width=0.4\linewidth]
+\harmonysource{hello2}
+\end{tcolorbox}
+\\
+(a) \harmonylink{code/hello1.hny} && (b) \harmonylink{code/hello2.hny}
+\end{tabular}
+\end{center}
+\caption{Hello World!}
+\label{fig:helloworld}
+\end{figure}
+
+The first programming book I read cover to cover was \emph{The C Programming
+Language} (first edition) by Brian W. Kernighan and Dennis M. Ritchie, which
+was around 1980.  In 1990, Dennis was my boss at AT\&T Bell Labs, while Brian
+was my colleague in the same lab.  The first program in the book printed
+the string ``hello, world''.  Since then, most programming tutorials for
+pretty much any programming language start with that example.  Harmony, too,
+has a Hello World program.
+\autoref{fig:helloworld}(a) shows the program and the corresponding output.
+The output is a \emph{Deterministic State Machine} (DFA).  The green circle
+represents the initial state and the double circle represents the final
+state.  There is one \emph{transition}, labeled with the string
+``hello world''.  The DFA describes (or \emph{recognizes}) all possible
+outputs that the program can generate.  In this case, there is only one.
+
+Now look at \autoref{fig:helloworld}{b).  It has two \texttt{print} statements.
+The output is another DFA, although this time there is an intermediate state.
+Still, there is only one possible output, and it consists of the strings
+``hello'' and ``world''.
+
+But programs can usually have more than one execution and produce multiple
+different outputs as a result.  \autoref{fig:choice} demonstrates
+\emph{nondetermistic choice} in Harmony programs.  In this case, the
+program chooses to either print ``hello'' or ``world''.  The corresponding
+DFA captures both possibilities.
+
+\begin{figure}
+\begin{center}
+\includegraphics[width=0.45\textwidth]{../code/hello3.png}
+\end{center}
+\begin{code}
+\harmonysource{hello3}
+\end{code}
+\caption{\harmonylink{code/hello3.hny} Nondeterministic choice}
+\label{fig:choice}
+\end{figure}
+
+\begin{figure}
+\begin{center}
+\includegraphics[width=0.15\textwidth]{../code/hello4.png}
+\end{center}
+\begin{code}
+\harmonysource{hello4}
+\end{code}
+\caption{\harmonylink{code/hello4.hny} Loop for some time, but don't have to
+go on for ever}
+\label{fig:hello4}
+\end{figure}
+
+\begin{figure}[h]
+\begin{center}
+\begin{tabular}{ccc}
+\includegraphics[width=0.45\textwidth]{../code/hello5.png}
+& \rule{0.05\linewidth}{0in} &
+\includegraphics[width=0.45\textwidth]{../code/hello6.png} \\
+\begin{tcolorbox}[width=0.4\linewidth]
+\harmonysource{hello5}
+\end{tcolorbox}
+& \rule{0.05\linewidth}{0in} &
+\begin{tcolorbox}[width=0.4\linewidth]
+\harmonysource{hello6}
+\end{tcolorbox}
+\\
+(a) \harmonylink{code/hello5.hny} && (b) \harmonylink{code/hello6.hny}
+\end{tabular}
+\end{center}
+\caption{Method calls and threads}
+\label{fig:methodthread}
+\end{figure}
+
+\begin{figure}
+\begin{center}
+\includegraphics[width=0.9\textwidth]{../code/hello7.png}
+\end{center}
+\begin{code}
+\harmonysource{hello7}
+\end{code}
+\caption{\harmonylink{code/hello7.hny} Interleaving threads}
+\label{fig:hello7}
+\end{figure}
+
+\begin{figure}
+\begin{center}
+\includegraphics[width=0.8\textwidth]{../code/hello8.png}
+\end{center}
+\begin{code}
+\harmonysource{hello8}
+\end{code}
+\caption{\harmonylink{code/hello8.hny} More interleaving}
+\label{fig:hello8}
+\end{figure}
+
 \chapter{Introduction to Harmony}

### 2021-12-03 book/paper.tex
Made a bunch of improvements

@@
 
-\chapter{Hello World}
-\label{ch:helloworld}
-
-\begin{figure}[h]
+\chapter{Hello World!}
+\label{ch:harmonyintro}
+
+\begin{figure}
 \begin{center}
-\begin{tabular}{ccc}
-\includegraphics[width=0.35\textwidth]{../code/hello1.png}
-& \rule{0.05\linewidth}{0in} &
-\includegraphics[width=0.45\textwidth]{../code/hello2.png} \\
-\begin{tcolorbox}[width=0.4\linewidth]
+\includegraphics[width=0.45\textwidth]{../code/hello1.png}
+\end{center}
+\begin{code}
 \harmonysource{hello1}
-\end{tcolorbox}
-& \rule{0.05\linewidth}{0in} &
-\begin{tcolorbox}[width=0.4\linewidth]
-\harmonysource{hello2}
-\end{tcolorbox}
-\\
-(a) \harmonylink{code/hello1.hny} && (b) \harmonylink{code/hello2.hny}
-\end{tabular}
-\end{center}
-\caption{Hello World!}
+\end{code}
+\caption{\harmonylink{code/hello1.hny} Hello World!}
 \label{fig:helloworld}
@@
 Language} (first edition) by Brian W. Kernighan and Dennis M. Ritchie, which
-was around 1980.  In 1990, Dennis was my boss at AT\&T Bell Labs, while Brian
-was my colleague in the same lab.  The first program in the book printed
+was around 1980.  I did not know at the time that 10 years later Dennis,
+the designer of the C programming language, would be my boss at AT\&T Bell Labs
+in Murray Hill, NJ, while Brian would be my colleague in the same lab.
+The first C program in the book printed
 the string ``hello, world''.  Since then, most programming tutorials for
-pretty much any programming language start with that example.  Harmony, too,
-has a Hello World program.
-\autoref{fig:helloworld}(a) shows the program and the corresponding output.
+pretty much any programming language start with that example.
+
+Harmony, too, has a Hello World program.
+\autoref{fig:helloworld} shows the program and the corresponding output.
+(The code for examples in this book can be found in the \texttt{code} folder under
+the name listed in the caption of the example.)
 The output is a \emph{Deterministic State Machine} (DFA).  The green circle
@@
 
-Now look at \autoref{fig:helloworld}{b).  It has two \texttt{print} statements.
-The output is another DFA, although this time there is an intermediate state.
-Still, there is only one possible output, and it consists of the strings
-``hello'' and ``world''.
-
+\index{choose operator}%
 But programs can usually have more than one execution and produce multiple
-different outputs as a result.  \autoref{fig:choice} demonstrates
+different outputs as a result.  This is usually as a result of different
+inputs, but Harmony programs do not have inputs. Instead,
+\autoref{fig:choice} demonstrates
 \emph{nondetermistic choice} in Harmony programs.  In this case, the
@@
 DFA captures both possibilities.
+You can think of the \texttt{choose} operator as enumerating all the possible
+inputs to the program.
+
+\autoref{fig:hello4} shows a program that has an infinite number of outputs
+by using a loop.  Harmony usually requires that any program must be able
+to terminate, so the loop is conditioned on a nondeterministic choice between
+\texttt{False} and \texttt{True}.  The possible outputs consists of zero or
+more copies of the string ``hello world''.  Note that this single state DFA
+(where the initial state and the final state happend to be the same)
+captures an infinite number of possible outputs.
 
@@
 \end{code}
-\caption{\harmonylink{code/hello3.hny} Nondeterministic choice}
+\caption{\harmonylink{code/hello3.hny} Two possible outputs}
 \label{fig:choice}
@@
 \end{code}
-\caption{\harmonylink{code/hello4.hny} Loop for some time, but don't have to
-go on for ever}
+\caption{\harmonylink{code/hello4.hny} An infinite number of outputs}
 \label{fig:hello4}
@@
 \end{center}
-\caption{Method calls and threads}
-\label{fig:methodthread}
-\end{figure}
+\caption{Methods and threads}
+\label{fig:threads}
+\end{figure}
+
+\autoref{fig:threads} demonstrates \emph{methods} and \emph{threads} in Harmony.
+In \autoref{fig:threads}, the code simply prints the string.
+strings ``hello'' and ``world''.  Notice that this leads to an intermediate
+state after ``hello'' is printed but before ``world'' is.  However, there
+is still only one execution possible. \autoref{fig:threads}(b) shows two
+threads, one printing ``hello'' and one printing ``world''.  Because the
+threads run concurrently, the program can either output ``hello world'' or
+``world hello''.  Printing in Harmony is atomic, so ``hweolrllod'' is not
+a possible  output.
 
@@
 \end{code}
-\caption{\harmonylink{code/hello8.hny} More interleaving}
+\caption{\harmonylink{code/hello8.hny} Less interleaving}
 \label{fig:hello8}
@@
 
-\chapter{Introduction to Harmony}
-\label{ch:harmonyintro}
+\autoref{fig:hello7} shows two threads, one printing the strings
+``hello'' and ``Robbert'', while the other prints ``hello'' and
+``Lesley''.  Now there are four possible outputs depending on
+how the two threads are interleaved, including
+``hello hello Lesley Robbert''.  This is probably not what
+the programmer wanted.  \autoref{fig:hello8} shows another
+important feature of Harmony: \emph{atomic blocks}.  The
+program is similar to \autoref{fig:hello7}, but the programmer
+specified that the two print statements in a thread should
+be executed as an atomic unit.  As a result, there are only
+two thread interleavings possible.
 
@@
 Use parentheses liberally to remove ambiguity.
-\item Harmony does not support floating point or I/O;
+\item Harmony does not support floating point;
 \item Python is object-oriented, supporting classes with methods and
@@
 \end{code}
-\caption{\harmonylink{code/triangle.hny} Computing triangle numbers.}
+\caption{\harmonylink{code/triangle.hny} Computing triangle numbers}
 \label{fig:triangle}
@@
 
-\autoref{fig:triangle} shows a simple example of a Harmony program.
-(The code for examples in this book can be found in the \texttt{code} folder under
-the name listed in the caption of the example.)
+\autoref{fig:triangle} shows another example of a Harmony program.
 The example is a sequential program
@@
 (For this to work, make sure \texttt{harmony} is in your command shell's search path.)
-Essentially, the $\texttt{choose}($S$)$
-\index{choose operator}%
-operator provides the input to the program by selecting some value from the
-set~$S$, while the $\textbf{assert}$ statement checks that the output is
-correct.  If the program is correct, the output of Harmony is the size of the
+The $\textbf{assert}$ statement checks that the output is
+correct.  If the program is correct, Harmony reports the size of the
 ``state graph'' (13 states in this case).  If not, Harmony also
@@
 
-In Harmony, constants have a default value,
+In Harmony, constants have a default specified value,
 but those can be overridden on the command
@@
 \end{center}
-\caption{A sequential and a concurrent program.}
+\caption{A sequential and a concurrent program}
 \label{fig:progs}
@@
 the assertion does not fail and 2) the execution terminates.
-Program (b) is concurrent.  It executes methods \texttt{f()} and
+Program (b) is concurrent---it executes methods \texttt{f()} and
 \texttt{g()} in parallel.
@@
 \end{code}
-\caption{\harmonylink{code/Up.hny} Incrementing twice in parallel.}
+\caption{\harmonylink{code/Up.hny} Incrementing twice in parallel}
 \label{fig:inc}
@@
 \end{code}
-\caption{\harmonylink{python/Up.py} Python implementation of \autoref{fig:inc}.}
+\caption{\harmonylink{python/Up.py} Python implementation of \autoref{fig:inc}}
 \label{fig:incpy}
@@
 \end{code}
-\caption{\harmonylink{python/UpMany.py} Incrementing \texttt{N} times.}
+\caption{\harmonylink{python/UpMany.py} Incrementing \texttt{N} times}
 \label{fig:incmany}
@@
 \end{code}
-\caption{The first part of the HVM bytecode corresponding to \autoref{fig:inc}.}
+\caption{The first part of the HVM bytecode corresponding to \autoref{fig:inc}}
 \label{fig:inccode}
@@
 \end{code}
-\caption{The text output of running Harmony on \autoref{fig:inc}.}
+\caption{The text output of running Harmony on \autoref{fig:inc}}
 \label{fig:incoutput}
@@
 \end{code}
-\caption{\harmonylink{code/UpEnter.hny} Broken attempt at fixing the code of \autoref{fig:inc}.}
+\caption{\harmonylink{code/UpEnter.hny} Broken attempt at fixing the code of \autoref{fig:inc}}
 \label{fig:incenter}
@@
 \end{code}
-\caption{\harmonylink{code/csbarebones.hny} A bare bones critical section with two threads.}
+\caption{\harmonylink{code/csbarebones.hny} A bare bones critical section with two threads}
 \label{fig:csbarebones}
@@
 \end{code}
-\caption{\harmonylink{code/cs.hny} Harmony model of a critical section.}
+\caption{\harmonylink{code/cs.hny} Harmony model of a critical section}
 \label{fig:cs}
@@
 \end{code}
-\caption{\harmonylink{code/naiveLock.hny} Na\"{\i}ve implementation of a shared lock.}
+\caption{\harmonylink{code/naiveLock.hny} Na\"{\i}ve implementation of a shared lock}
 \label{fig:naivelock}
@@
 \end{center}
-\caption{The \harmonyref{naiveLock.html}{HTML output of running Harmony on \autoref{fig:naivelock}}.}
+\caption{The \harmonyref{naiveLock.html}{HTML output of running Harmony on \autoref{fig:naivelock}}}
 \label{fig:naiveLockhtml}
@@
 \end{code}
-\caption{\harmonylink{code/naiveFlags.hny} Na\"{\i}ve use of flags to solve mutual exclusion.}
+\caption{\harmonylink{code/naiveFlags.hny} Na\"{\i}ve use of flags to solve mutual exclusion}
 \label{fig:naiveflags}
@@
 \end{center}
-\caption{The \harmonyref{naiveFlags.html}{HTML output of running Harmony on \autoref{fig:naiveflags}}.}
+\caption{The \harmonyref{naiveFlags.html}{HTML output of running Harmony on \autoref{fig:naiveflags}}}
 \label{fig:naiveflagshtml}
@@
 \end{code}
-\caption{\harmonylink{code/naiveTurn.hny} Na\"{\i}ve use of turn variable to solve mutual exclusion.}
+\caption{\harmonylink{code/naiveTurn.hny} Na\"{\i}ve use of turn variable to solve mutual exclusion}
 \label{fig:naiveturn}
@@
 \end{center}
-\caption{Venn diagram classifying all states and a trace.}
+\caption{Venn diagram classifying all states and a trace}
 \label{fig:states}
@@
 \end{code}
-\caption{\harmonylink{code/csonebit.hny} Mutual exclusion using a flag per thread.}
+\caption{\harmonylink{code/csonebit.hny} Mutual exclusion using a flag per thread}
 \label{fig:csonebit}
@@
 \end{code}
-\caption{\harmonylink{code/PetersonMethod.hny} Peterson's Algorithm accessed through methods.}
+\caption{\harmonylink{code/PetersonMethod.hny} Peterson's Algorithm accessed through methods}
 \label{fig:petersonmethods}
@@
 \end{code}
-\caption{\harmonylink{code/clock.hny} Harmony program that finds page replacement anomalies.}
+\caption{\harmonylink{code/clock.hny} Harmony program that finds page replacement anomalies}
 \label{fig:clock}
@@
 \end{code}
-\caption{\harmonylink{modules/synch.hny} Specification of a lock.}
+\caption{\harmonylink{modules/synch.hny} Specification of a lock}
 \label{fig:lockspec}
@@
 \caption{\harmonylink{code/cssynch.hny} Using a lock to implement a
-critical section.}
+critical section}
 \label{fig:cssynch}
@@
 \end{code}
-\caption{\harmonylink{code/UpLock.hny} Program of \autoref{fig:inc} fixed with a lock.}
+\caption{\harmonylink{code/UpLock.hny} Program of \autoref{fig:inc} fixed with a lock}
 \label{fig:incfixed}
@@
 \end{code}
-\caption{\harmonylink{code/spinlock.hny} Mutual Exclusion using a ``spinlock'' based on test-and-set.}
+\caption{\harmonylink{code/spinlock.hny} Mutual Exclusion using a ``spinlock'' based on test-and-set}
 \label{fig:tas}
@@
 specification in \autoref{fig:lockspec} using a spinlock based on
-test-and-set.}
+test-and-set}
 \label{fig:tas2}
@@
 \caption{\harmonylink{code/ticket.hny} Implementation of the lock
-specification in \autoref{fig:lockspec} using a ticket lock.}
+specification in \autoref{fig:lockspec} using a ticket lock}
 \label{fig:ticket}
@@
 \end{code}
-\caption{\harmonylink{modules/synchS.hny} The lock implementation in the \texttt{synchS} module uses suspension.}
+\caption{\harmonylink{modules/synchS.hny} The lock implementation in the \texttt{synchS} module uses suspension}
 \label{fig:suspension}
@@
 \end{code}
-\caption{\harmonylink{code/xy.hny} Incomplete code with desired invariant $x + y = 100$.}
+\caption{\harmonylink{code/xy.hny} Incomplete code with desired invariant $x + y = 100$}
 \label{fig:xy}
@@
 \end{code}
-\caption{\harmonylink{code/atm.hny} Withdrawing money from an ATM.}
+\caption{\harmonylink{code/atm.hny} Withdrawing money from an ATM}
 \label{fig:atm}
@@
 \end{center}
-\caption{A sequential and a concurrent specification of a queue.}
+\caption{A sequential and a concurrent specification of a queue}
 \label{fig:queues}
@@
 \end{code}
-\caption{\harmonylink{code/queuedemo.hny} Using a concurrent queue.}
+\caption{\harmonylink{code/queuedemo.hny} Using a concurrent queue}
 \label{fig:queuedemo}
@@
 \end{code}
-\caption{\harmonylink{code/queueconc.hny} An implementation of a concurrent queue data structure.}
+\caption{\harmonylink{code/queueconc.hny} An implementation of a concurrent queue data structure}
 \label{fig:queueconc}
@@
 \end{code}
-\caption{\harmonylink{code/queueMS.hny} A queue with separate locks for enqueuing and dequeuing items.}
+\caption{\harmonylink{code/queueMS.hny} A queue with separate locks for enqueuing and dequeuing items}
 \label{fig:queueMS}
@@
 \caption{\harmonylink{code/setobj.hny} Specification of a concurrent
-set object.}
+set object}
 \label{fig:setobj}
@@
 \end{code}
-\caption{\harmonylink{code/linkedlist.hny} Implementation of a set of integers using a linked list with fine-grained locking.}
+\caption{\harmonylink{code/linkedlist.hny} Implementation of a set of integers using a linked list with fine-grained locking}
 \label{fig:linkedlist}
@@
 \end{code}
-\caption{\harmonylink{code/intsettest.hny} Test code for \autoref{fig:linkedlist}.}
+\caption{\harmonylink{code/intsettest.hny} Test code for \autoref{fig:linkedlist}}
 \label{fig:intsettest}
@@
 \end{code}
-\caption{\harmonylink{code/cslock.hny} Critical section testing without labels.}
+\caption{\harmonylink{code/cslock.hny} Critical section testing without labels}
 \label{fig:cslock}
@@
 \end{code}
-\caption{\harmonylink{code/qtestseq.hny} Sequential queue test.}
+\caption{\harmonylink{code/qtestseq.hny} Sequential queue test}
 \label{fig:qtestseq}
@@
 \end{code}
-\caption{\harmonylink{code/queuelin.hny} Queue implementation with linearization points.}
+\caption{\harmonylink{code/queuelin.hny} Queue implementation with linearization points}
 \label{fig:queuelin}
@@
 \end{code}
-\caption{\harmonylink{code/qtestconc.hny} Concurrent queue test.}
+\caption{\harmonylink{code/qtestconc.hny} Concurrent queue test}
 \label{fig:qtestconc}
@@
 \end{code}
-\caption{\harmonylink{code/qtest1.hny} Concurrent \texttt{put()}.}
+\caption{\harmonylink{code/qtest1.hny} Concurrent \texttt{put()}}
 \label{fig:qtest1}
@@
 \end{code}
-\caption{\harmonylink{code/qtest2.hny} Concurrent \texttt{get()}.}
+\caption{\harmonylink{code/qtest2.hny} Concurrent \texttt{get()}}
 \label{fig:qtest2}
@@
 \caption{\harmonylink{code/qtest3.hny} Mix of concurrent
-\texttt{get()} and \texttt{put()} operations.}
+\texttt{get()} and \texttt{put()} operations}
 \label{fig:qtest3}
@@
 \end{code}
-\caption{\harmonylink{code/qtest4.hny} Test with multiple queues.}
+\caption{\harmonylink{code/qtest4.hny} Test with multiple queues}
 \label{fig:qtest4}
@@
 \caption{\harmonylink{code/multitest.hny} Running a
-sequential test followed by several concurrent tests.}
+sequential test followed by several concurrent tests}
 \label{fig:multitest}
@@
 \end{code}
-\caption{\harmonylink{code/queuebug.hny} A buggy queue implementation.}
+\caption{\harmonylink{code/queuebug.hny} A buggy queue implementation}
 \label{fig:queuebug}
@@
 \caption{\harmonyref{queuebug.html}{Harmony output of running \autoref{fig:qtest3} against
-\autoref{fig:queuebug}}.}
+\autoref{fig:queuebug}}}
 \label{fig:queuebug1}
@@
 \end{code}
-\caption{\harmonylink{code/queuefix.hny} Queue implementation with hand-over-hand locking.}
+\caption{\harmonylink{code/queuefix.hny} Queue implementation with hand-over-hand locking}
 \label{fig:queuefix}
@@
 \end{code}
-\caption{\harmonylink{code/RW.hny} Specification of reader/writer locks.}
+\caption{\harmonylink{code/RW.hny} Specification of reader/writer locks}
 \label{fig:rwspec}
@@
 \end{code}
-\caption{\harmonylink{code/RWtest.hny} Test code for reader/writer locks.}
+\caption{\harmonylink{code/RWtest.hny} Test code for reader/writer locks}
 \label{fig:rwtest}
@@
 \end{code}
-\caption{\harmonylink{code/RWbusy.hny} Busy-Waiting Reader/Writer Lock implementation.}
+\caption{\harmonylink{code/RWbusy.hny} Busy-Waiting Reader/Writer Lock implementation}
 \label{fig:rwbusy}
@@
 \end{code}
-\caption{\harmonylink{code/RWbusychk.hny} Checking for Busy Waiting.}
+\caption{\harmonylink{code/RWbusychk.hny} Checking for Busy Waiting}
 \label{fig:rwblock}
@@
 \end{code}
-\caption{\harmonylink{code/RWlock.hny} Reader/Writer with Two Locks.}
+\caption{\harmonylink{code/RWlock.hny} Reader/Writer with Two Locks}
 \label{fig:rw2lock}
@@
 \end{code}
-\caption{\harmonylink{code/RWmulti.hny} Checking that multiple readers can acquire the read lock.}
+\caption{\harmonylink{code/RWmulti.hny} Checking that multiple readers can acquire the read lock}
 \label{fig:rwmulti}
@@
 \end{code}
-\caption{\harmonylink{modules/synch.hny} The \texttt{Semaphore} interface in the \texttt{synch} module.}
+\caption{\harmonylink{modules/synch.hny} The \texttt{Semaphore} interface in the \texttt{synch} module}
 \label{fig:semaphore}
@@
 \end{code}
-\caption{\harmonylink{code/DinersSema.hny} Italian Philosophers.}
+\caption{\harmonylink{code/DinersSema.hny} Italian Philosophers}
 \label{fig:dinerssema}
@@
 \end{code}
-\caption{\harmonylink{code/boundedbuffer.hny} Bounded buffer specification.}
+\caption{\harmonylink{code/boundedbuffer.hny} Bounded buffer specification}
 \label{fig:boundedbuffer}
@@
 \end{center}
-\caption{High-level state diagram specification of producers/consumers.}
+\caption{High-level state diagram specification of producers/consumers}
 \label{fig:pc}
@@
 \end{code}
-\caption{\harmonylink{code/BBsema.hny} Bounded Buffer implementation using semaphores.}
+\caption{\harmonylink{code/BBsema.hny} Bounded Buffer implementation using semaphores}
 \label{fig:boundedbuffer}
@@
 \end{code}
-\caption{\harmonylink{code/BBsematest.hny} Test program for \autoref{fig:boundedbuffer}.}
+\caption{\harmonylink{code/BBsematest.hny} Test program for \autoref{fig:boundedbuffer}}
 \label{fig:bbtest}
@@
 \end{code}
-\caption{Script for testing producer/consumer synchronization.}
+\caption{Script for testing producer/consumer synchronization}
 \label{fig:pcscript}
@@
 \caption{\harmonylink{code/BBsemadata.hny} Test whether the bounded buffer delivers the right data in the
-right order.}
+right order}
 \label{fig:BBsemadata}
@@
 \end{code}
-\caption{\harmonylink{code/RWsbs.hny} Reader/Writer Lock using Split Binary Semaphores.}
+\caption{\harmonylink{code/RWsbs.hny} Reader/Writer Lock using Split Binary Semaphores}
 \label{fig:RWsplitsema}
@@
 \end{code}
-\caption{\harmonylink{code/gpu.hny} A thread-unsafe GPU allocator.}
+\caption{\harmonylink{code/gpu.hny} A thread-unsafe GPU allocator}
 \label{fig:gpu}
@@
 \end{code}
-\caption{\harmonylink{code/RWfair.hny} Reader/Writer Lock SBS implementation addressing fairness.}
+\caption{\harmonylink{code/RWfair.hny} Reader/Writer Lock SBS implementation addressing fairness}
 \label{fig:RWfair}
@@
 \end{code}
-\caption{\harmonylink{code/RWqueue.hny} Reader/Writer Lock SBS implementation using a queue.}
+\caption{\harmonylink{code/RWqueue.hny} Reader/Writer Lock SBS implementation using a queue}
 \label{fig:RWqueue}
@@
 \end{code}
-\caption{\harmonylink{code/RWqtest.hny} Test program for \autoref{fig:RWqueue}.}
+\caption{\harmonylink{code/RWqtest.hny} Test program for \autoref{fig:RWqueue}}
 \label{fig:RWqtest}
@@
 \end{code}
-\caption{\harmonylink{modules/hoare.hny} Implementation of Hoare monitors.}
+\caption{\harmonylink{modules/hoare.hny} Implementation of Hoare monitors}
 \label{fig:hoare}
@@
 \caption{\harmonylink{code/BBhoare.hny} Bounded Buffer implemented using
-a Hoare monitor.}
+a Hoare monitor}
 \label{fig:hoaretest}
@@
 \caption{\harmonylink{modules/synch.hny}
-Implementation of condition variables in the \texttt{synch} module.}
+Implementation of condition variables in the \texttt{synch} module}
 \label{fig:cv}
@@
 \end{code}
-\caption{\harmonylink{code/RWcv.hny} Reader/Writer Lock using Mesa-style condition variables.}
+\caption{\harmonylink{code/RWcv.hny} Reader/Writer Lock using Mesa-style condition variables}
 \label{fig:RWcv}
@@
 \end{code}
-\caption{\harmonylink{code/qsort.hny} Iterative qsort() implementation.}
+\caption{\harmonylink{code/qsort.hny} Iterative qsort() implementation}
 \label{fig:qsort}
@@
 \end{code}
-\caption{\harmonylink{code/qsorttest.hny} Test program for \autoref{fig:qsort}.}
+\caption{\harmonylink{code/qsorttest.hny} Test program for \autoref{fig:qsort}}
 \label{fig:qsorttest}
@@
 \end{code}
-\caption{\harmonylink{code/Diners.hny} Dining Philosophers.}
+\caption{\harmonylink{code/Diners.hny} Dining Philosophers}
 \label{fig:diners}
@@
 \end{code}
-\caption{\harmonylink{code/DinersCV.hny} Dining Philosophers that grab both forks at the same time.}
+\caption{\harmonylink{code/DinersCV.hny} Dining Philosophers that grab both forks at the same time}
 \label{fig:dinerscv}
@@
 \caption{\harmonylink{code/DinersAvoid.hny} Dining Philosophers that carefully avoid getting into a deadlock
-scenario.}
+scenario}
 \label{fig:dinersavoid}
@@
 \end{code}
-\caption{\harmonylink{code/bank.hny} Bank accounts.}
+\caption{\harmonylink{code/bank.hny} Bank accounts}
 \label{fig:bank}
@@
 \end{code}
-\caption{\harmonylink{code/counter.hny} An illustration of the actor approach.}
+\caption{\harmonylink{code/counter.hny} An illustration of the actor approach}
 \label{fig:actor}
@@
 \end{code}
-\caption{\harmonylink{code/qbarrier.hny} Barrier synchronization with queues.}
+\caption{\harmonylink{code/qbarrier.hny} Barrier synchronization with queues}
 \label{fig:barrier}
@@
 \end{code}
-\caption{\harmonylink{code/barrier.hny} Generalized barrier synchronization.}
+\caption{\harmonylink{code/barrier.hny} Generalized barrier synchronization}
 \label{fig:genbarrier}
@@
 \end{code}
-\caption{\harmonylink{code/barriertest.hny} Test program for \autoref{fig:genbarrier}.}
+\caption{\harmonylink{code/barriertest.hny} Test program for \autoref{fig:genbarrier}}
 \label{fig:barriertest}
@@
 \end{code}
-\caption{\harmonylink{code/trap.hny} How to use \textbf{trap}.}
+\caption{\harmonylink{code/trap.hny} How to use \textbf{trap}}
 \label{fig:trap}
@@
 \end{code}
-\caption{\harmonylink{code/trap2.hny} A race condition with interrupts.}
+\caption{\harmonylink{code/trap2.hny} A race condition with interrupts}
 \label{fig:trap2}
@@
 \end{code}
-\caption{\harmonylink{code/trap3.hny} Locks do not work with interrupts.}
+\caption{\harmonylink{code/trap3.hny} Locks do not work with interrupts}
 \label{fig:trap3}
@@
 \end{code}
-\caption{\harmonylink{code/trap4.hny} Disabling and enabling interrupts.}
+\caption{\harmonylink{code/trap4.hny} Disabling and enabling interrupts}
 \label{fig:trap4}
@@
 \end{code}
-\caption{\harmonylink{code/trap5.hny} Example of an interrupt-safe method.}
+\caption{\harmonylink{code/trap5.hny} Example of an interrupt-safe method}
 \label{fig:trap5}
@@
 \end{code}
-\caption{\harmonylink{code/trap6.hny} Code that is both interrupt-safe and thread-safe.}
+\caption{\harmonylink{code/trap6.hny} Code that is both interrupt-safe and thread-safe}
 \label{fig:trap6}
@@
 \end{code}
-\caption{\harmonylink{code/hw.hny} Non-blocking queue.}
+\caption{\harmonylink{code/hw.hny} Non-blocking queue}
 \label{fig:hw}
@@
 \end{code}
-\caption{\harmonylink{code/abp.hny} Alternating Bit Protocol.}
+\caption{\harmonylink{code/abp.hny} Alternating Bit Protocol}
 \label{fig:abp}
@@
 \end{code}
-\caption{\harmonylink{code/abptest.hny} Test code for alternating bit protocol.}
+\caption{\harmonylink{code/abptest.hny} Test code for alternating bit protocol}
 \label{fig:abptest}
@@
 \end{code}
-\caption{\harmonylink{code/leader.hny} A leader election protocol on a ring.}
+\caption{\harmonylink{code/leader.hny} A leader election protocol on a ring}
 \label{fig:leader}
@@
 \end{code}
-\caption{\harmonylink{code/2pc.hny} Two Phase Commit protocol: code for banks.}
+\caption{\harmonylink{code/2pc.hny} Two Phase Commit protocol: code for banks}
 \label{fig:2pc1}
@@
 \end{code}
-\caption{\harmonylink{code/2pc.hny} Two Phase Commit protocol: code for transaction coordinators.}
+\caption{\harmonylink{code/2pc.hny} Two Phase Commit protocol: code for transaction coordinators}
 \label{fig:2pc2}
@@
 \end{code}
-\caption{\harmonylink{code/chain.hny} Checking correctness of Chain Replication.}
+\caption{\harmonylink{code/chain.hny} Checking correctness of Chain Replication}
 \label{fig:chain}
@@
 \end{code}
-\caption{\harmonylink{code/abd.hny} An implementation of a replicated atomic read/write register.}
+\caption{\harmonylink{code/abd.hny} An implementation of a replicated atomic read/write register}
 \label{fig:abd}
@@
 \end{code}
-\caption{\harmonylink{code/abdtest.hny} Testing the code in \autoref{fig:abd}.}
+\caption{\harmonylink{code/abdtest.hny} Testing the code in \autoref{fig:abd}}
 \label{fig:abdtest}
@@
 \end{code}
-\caption{\harmonylink{code/bosco.hny} A crash-tolerant consensus protocol.}
+\caption{\harmonylink{code/bosco.hny} A crash-tolerant consensus protocol}
 \label{fig:bosco}
@@
 \end{code}
-\caption{\harmonylink{code/bosco2.hny} Reducing the state space.}
+\caption{\harmonylink{code/bosco2.hny} Reducing the state space}
 \label{fig:bosco2}
@@
 \end{code}
-\caption{\harmonylink{code/paxos.hny} A version of the Paxos protocol, Part 1.}
+\caption{\harmonylink{code/paxos.hny} A version of the Paxos protocol, Part 1}
 \label{fig:paxos1}
@@
 \end{code}
-\caption{\harmonylink{code/paxos.hny} A version of the Paxos protocol, Part 2.}
+\caption{\harmonylink{code/paxos.hny} A version of the Paxos protocol, Part 2}
 \label{fig:paxos2}
@@
 \end{code}
-\caption{\harmonylink{code/needhamschroeder.hny} Needham-Schroeder protocol and an attack.}
+\caption{\harmonylink{code/needhamschroeder.hny} Needham-Schroeder protocol and an attack}
 \label{fig:ns}

### 2021-12-03 src/charm/charm.c

@@ -280,8 +280,7 @@ static bool onestep(
 
         // See if we need to break out of this step
         struct instr_t *next_instr = &global->code.instrs[step->ctx->pc];
-        if (!step->ctx->atomicFlag && sc->ctxbag != VALUE_DICT &&
-                                (next_instr->breakable || next_instr->log)) {
+        if (!step->ctx->atomicFlag && (next_instr->breakable || next_instr->log)) {
             if (!step->ctx->atomicFlag && step->ctx->atomic > 0) {
                 step->ctx->atomicFlag = true;
             }
@@ -1866,49 +1865,26 @@ int main(int argc, char **argv){
                 else {
                     fprintf(out, ",\n");
                 }
-                if (edge->nlog == 0) {
-                    fprintf(out, "    {\n");
-                    fprintf(out, "      \"src\": %d,\n", node->id);
-                    fprintf(out, "      \"dst\": %d,\n", edge->node->id);
-                    fprintf(out, "      \"log\": \"\"\n");
-                    fprintf(out, "    }");
-                }
-                else if (edge->nlog == 1) {
-                    fprintf(out, "    {\n");
-                    fprintf(out, "      \"src\": %d,\n", node->id);
-                    fprintf(out, "      \"dst\": %d,\n", edge->node->id);
-                    char *p = json_escape_value(edge->log[0]);
-                    fprintf(out, "      \"log\": \"%s\"\n", p);
-                    free(p);
-                    fprintf(out, "    }");
-                }
-                else {
-                    fprintf(out, "    {\n");
-                    fprintf(out, "      \"src\": %d,\n", node->id);
-                    fprintf(out, "      \"dst\": \"i%d\",\n", --extra);
-                    char *p = json_escape_value(edge->log[0]);
-                    fprintf(out, "      \"log\": \"%s\"\n", p);
-                    free(p);
-                    fprintf(out, "    },\n");
-
-                    for (int j = 1; j < edge->nlog - 1; j++) {
-                        fprintf(out, "    {\n");
-                        fprintf(out, "      \"src\": \"i%d\",\n", extra);
-                        fprintf(out, "      \"dst\": \"i%d\",\n", --extra);
-                        p = json_escape_value(edge->log[j]);
-                        fprintf(out, "      \"log\": \"%s\"\n", p);
-                        free(p);
-                        fprintf(out, "    },\n");
+                fprintf(out, "    {\n");
+                fprintf(out, "      \"src\": %d,\n", node->id);
[…]

### 2021-12-03 src/harmony/behavior.py

@@ -56,18 +56,32 @@ def behavior_parse(js, minify, output):
             final_states.add(idx)
         states.add(idx)
 
-    for edge in js['edges']:
-        src = str(edge["src"])
-        dst = str(edge["dst"])
-        if edge["log"] != "":
-            input_symbols.add(edge["log"])
+    def add_edge(src, val, dst):
         assert dst != initial_state
         assert src not in final_states
-        val = edge["log"]
         if val in transitions[src]:
             transitions[src][val].add(dst)
         else:
             transitions[src][val] = {dst}
+
+    intermediate = 0
+    for edge in js['edges']:
+        src = str(edge["src"])
+        dst = str(edge["dst"])
+        if edge["log"] == []:
+            add_edge(src, "", dst)
+        else:
+            for e in edge["log"][:-1]:
+                input_symbols.add(e)
+                inter = "s%d"%intermediate
+                intermediate += 1
+                states.add(inter)
+                transitions[inter] = {}
+                add_edge(src, e, inter)
+                src = inter
+            e = edge["log"][-1]
+            add_edge(src, e, dst)
+            input_symbols.add(e)
 
     # print("states", states, file=sys.stderr)
     # print("final", final_states, file=sys.stderr)

### 2021-12-03 book/paper.tex
Fixed bug

@@
 
-The first programming book I read cover to cover was \emph{The C Programming
+The first programming book that I read cover to cover was \emph{The C Programming
 Language} (first edition) by Brian W. Kernighan and Dennis M. Ritchie, which
@@
 DFA captures both possibilities.
-You can think of the \texttt{choose} operator as enumerating all the possible
+You can think of the \texttt{choose} operator as enumerating all possible
 inputs to the program.
@@
 \autoref{fig:threads} demonstrates \emph{methods} and \emph{threads} in Harmony.
-In \autoref{fig:threads}, the code simply prints the string.
-strings ``hello'' and ``world''.  Notice that this leads to an intermediate
+In \autoref{fig:threads}(a), the code simply prints the
+strings ``hello'' and ``world'', in that order.
+Notice that this leads to an intermediate
 state after ``hello'' is printed but before ``world'' is.  However, there
@@
 (Harmony does not have iterators and in particular does
-not have a \texttt{range} operator.)
+not have a \texttt{range} operator like Python.)
 The last two lines in the program are
@@
 time---the important thing is to develop an understanding of concurrency.
-(Also, you do not get to use a \emph{lock}, yet.)
+(Also, you do not get to use the \texttt{atomically} keyword
+or a \emph{lock}, yet.)
 \end{problems}

### 2021-12-03 src/charm/charm.c

@@ -1365,6 +1365,7 @@ void process_results(
         if ((next = *p) == NULL) {
             next = *p = node;
             graph_add(&global->graph, node);   // sets node->id
+            assert(node->id != 0);
             minheap_insert(todo[0], node);
             global->enqueued++;
             must_free = false;
@@ -1382,7 +1383,7 @@ void process_results(
             minheap_insert(global->failures, f);
         }
 
-        // printf("ADD %d <=> %d\n", parent->id, node->id);
+        // printf("ADD %d => %d\n", parent->id, next->id);
 
         // Add a forward edge from parent
         struct edge *fwd = new_alloc(struct edge);
@@ -1409,8 +1410,8 @@ void process_results(
         bwd->ai = node->ai;
         bwd->log = node->log;
         bwd->nlog = node->nlog;
-        bwd->next = node->bwd;
-        node->bwd = bwd;
+        bwd->next = next->bwd;
+        next->bwd = bwd;
 
         if (must_free) {
             free(node);
@@ -1740,7 +1741,7 @@ int main(int argc, char **argv){
         printf("%d components, %d bad states\n", ncomponents, nbad);
     }
 
-    if (false) {
+    if (true) {
         FILE *df = fopen("charm.dump", "w");
         assert(df != NULL);
         for (int i = 0; i < global->graph.size; i++) {
@@ -1760,6 +1761,29 @@ int main(int argc, char **argv){
                 fprintf(df, "            context: %s %s %d\n", value_string(ctx->name), value_string(ctx->arg), ctx->pc);
                 fprintf(df, "            choice: %s\n", value_string(edge->choice));
                 fprintf(df, "            node: %d (%d)\n", edge->node->id, edge->node->component);
+                fprintf(df, "            log:");
+                for (int j = 0; j < edge->nlog; j++) {
+                    char *p = value_string(edge->log[j]);
+                    fprintf(df, " %s", p);
+                    free(p);
+                }
+                fprintf(df, "\n");
+            }
+            fprintf(df, "    bwd:\n");
+            eno = 0;
+            for (struct edge *edge = node->bwd; edge != NULL; edge = edge->next, eno++) {
+                fprintf(df, "        %d:\n", eno);
+                struct context *ctx = value_get(edge->ctx, NULL);
+                fprintf(df, "            context: %s %s %d\n", value_string(ctx->name), value_string(ctx->arg), ctx->pc);
+                fprintf(df, "            choice: %s\n", value_string(edge->choice));
[…]

### 2021-12-06 book/paper.tex
Got rid of atoms.  Now there are only strings

@@
 that map Harmony values to other Harmony values.
-%
-Another type of Harmony value is the \emph{atom}.
-\index{atom}%
-It is essentially
-just a name.  An atom is denoted using a period followed by the
-name.  For example, \texttt{.main} is an atom.
 
@@
 \index{countLabel operator}%
-where $L$
-is the atom containing the name of the label (in this case, \texttt{cs}).
+where $L$ is the name of the label (in this case, \texttt{cs}).
 The operator returns the number of threads executing at that label.
@@
 \hline
-Atom & \texttt{.example}, \texttt{.test1}, \texttt{.0x4A} \\
+String & \texttt{"example"}, \texttt{.example} \\
 \hline
@@
 
-\subsection*{Atom}
-
-Atoms are essentially names and consist of one or more unicode characters.
+\subsection*{String}
+
+A string is a sequence of zero or more unicode characters.
 If they do not contain special characters and do not start with a digit,
-then an atom can be represented by a ``.'' followed by the characters.
-For example, \texttt{.hello} is
-a representation of the atom ``hello''.  A special character can be
-represented by \texttt{.0xXX}, where \texttt{XX} is the hexidecimal
-unicode for the character.
-Atoms should not be confused with strings.
-There are no special operations on atoms.
-(In the future, operators may be introduced that convert between strings
-and atoms.)
+then a string can be represented by a ``.'' followed by the characters.
+For example, \texttt{.hello} is the same as the string \texttt{"hello"}.
 
@@
 you can write $d[k]$ (or $d(k)$) instead of $d~k$.
-However, if $k$ is an atom, like \texttt{.id}, then you might
+However, if $k$ is a string like \texttt{.id}, then you might
 prefer the notation $k\mathtt{.id}$.
@@
 \texttt{any($(x \% 2) == 0$ for $x$ in $t$)}.
-
-\subsection*{String}
-
-In Harmony, a string is a list of single-character atoms.
-The Python string notations mostly work in Harmony as well.
-For example, \texttt{'abc'} is a three-character string
-consisting of the atoms \texttt{.a}, \texttt{.b}, and
-\texttt{.c}.
-\texttt{"abc"} is the same three-character string.
-Various special characters (including quotes, newlines, etc.)
-can be \emph{escaped} using a backslash.
-Multi-line strings can be terminated by triple quotes or
-triple double-quotes.
-
-Native operations on strings include the following:
-
-\begin{center}
-\begin{tabular}{|l|l|}
-\hline
-\texttt{$s~k$} & indexing \\
-\hline
-\texttt{$s$ + $s$ + ...} & concatenation \\
-\hline
-\texttt{$s$ * $n$} & $n$ copies of $s$ concatenated \\
-\hline
-\texttt{$c$ [not] in $s$} & check if $c$ (a one-character atom) is in $s$ \\
-\hline
-\texttt{len $s$} & the length of $s$ \\
-\hline
-\end{tabular}
-\end{center}
 
@@
 For readability, it may help to write $a(b)$ for function application and
-$a[b]$ for indexing.  In case $b$ is an atom, you can also write $a.b$ for indexing.
+$a[b]$ for indexing.  In case $b$ is a simple string,
+you can also write $a.b$ for indexing.
 
@@
 which consists of the caller's frame pointer and variables, the argument to
-the method, an ``invocation type atom'' (\texttt{normal}, \texttt{interrupt}, or
+the method, an ``invocation type'' (\texttt{normal}, \texttt{interrupt}, or
 \texttt{thread}), and the return address (in case of \texttt{normal}).

### 2021-12-06 book/paper.tex
Fixed some "atom" leftovers

@@
 integers (but not floating point numbers),
-strings (enclosed by double quotes),
-sets of Harmony values,
-and dictionaries
+strings (enclosed by single or double quotes),
+sets of Harmony values, and dictionaries
 \index{dictionary}%
 that map Harmony values to other Harmony values.
+Strings that start with a letter or an underscore and only contain
+letters, digits, and underscores can be written without quotes by
+preceding it with a dot.  So \texttt{.example} is the same string
+as \texttt{"example"}.
 
@@
 $d$.\texttt{count} (which has value 3) instead of having to write
-$d$\texttt{[.count]} (although both will work).
-Thus, using atoms, a dictionary can be made to look much like a Python object.
+$d$\texttt{[.count]} or $d$\texttt{["count"]} (although any of those
+will work).
+Thus a dictionary can be made to look much like a Python object.
 
@@
 However, this is not a limitation because Harmony values are unbounded trees.
-The shared memory is a dictionary that maps atoms (names of shared
+The shared memory is a dictionary that maps strings (names of shared
 variables) to other Harmony values.
@@
 Thus, a directory represents the state of a collection of variables named
-by the atoms.
+by the strings.
 %
@@
 
-Tuples, lists, strings, and bags are all special cases of dictionaries.
+Tuples, lists, and bags are all special cases of dictionaries.
 Both tuples and lists map indexes (starting at 0) to Harmony values.
@@
 is optional.
-A string is represented as a tuple of its characters.
-Characters are one-character atoms, which can be expressed
-in hexadecimal unicode using the syntax \texttt{.0xXX}.
 A bag or multiset is a dictionary that maps a value to how many
@@
 \item integers are ordered in the natural way;
-\item atoms are lexicographically ordered;
+\item strings are lexicographically ordered;
 \item program counters are ordered by their integer values;
@@
 ordered;
-\item an address is a list of atoms.  \texttt{None} is the empty list of atoms.
+\item an address is a list of Harmony values.  \texttt{None} is the empty list.
 Addresses are lexicographically ordered accordingly;
@@
 \hline
-labels & a dictionary of atoms to program counters \\
-\hline
-variables & a dictionary mapping atoms to values \\
+variables & a dictionary mapping strings to values \\
 \hline
@@
 \hline
-method variables & a dictionary mapping atoms (names of method variables) to values \\
-\hline
-thread-local variables & a dictionary mapping atoms (names of thread-local variables) to values \\
+method variables & a dictionary mapping strings (names of method variables) to values \\
+\hline
+thread-local variables & a dictionary mapping strings (names of thread-local variables) to values \\
 \hline
@@
 \item The register of a thread always contains a dictionary, mapping
-atoms to arbitrary values.  The atoms correspond to the variable names
+strings to arbitrary values.  The strings correspond to the variable names
 in a Harmony program.

### 2021-12-07 book/paper.tex
Starting to talk about behaviors
Added buggy concurrent queue code to illustrate usefulness of behaviors

@@
 
-\chapter{Testing}
+\chapter{Testing: Checking Behaviors}
 \label{ch:testing}
@@
 As with critical sections, when testing a concurrent data structure
-we need a specification for its correctness.
+we need a specification.
 \autoref{fig:queues}(a) shows a sequential specification of a
@@
 find any errors.
+
 The next step is to test if the queue implementation
 meets the \emph{concurrent} queue specification or not.
+\autoref{fig:queues}(b) shows the concurrent version of
+\autoref{fig:queues}(a)---it simply makes all operations
+(except instantiation itself) atomic.
+
 But we have not yet shown what the

### 2021-12-11 book/paper.tex
Latest

@@
 \autoref{fig:helloworld} shows the program and the corresponding output.
-(The code for examples in this book can be found in the \texttt{code} folder under
-the name listed in the caption of the example.)
+From the command line, you can run it as follows:
+\begin{code}
+\begin{verbatim}
+$ harmony -o hello1.png code/hello1.hny
+\end{verbatim}
+\end{code}
+For this to work, make sure \texttt{harmony} is in your command shell's
+search path.
+The code for examples in this book can be found in the \texttt{code}
+folder under the name listed in the caption of the example.
+In this case, the file \texttt{code/hello1.hny} contains the
+code in \autoref{fig:helloworld}.
+The \texttt{-o hello1.png} arguments tell harmony to write the output
+in the file \texttt{hello1.png}
 The output is a \emph{Deterministic State Machine} (DFA).  The green circle
@@
 \emph{nondetermistic choice} in Harmony programs.  In this case, the
-program chooses to either print ``hello'' or ``world''.  The corresponding
+program chooses to print either ``hello'' or ``world''.  The corresponding
 DFA captures both possibilities.
@@
 to terminate, so the loop is conditioned on a nondeterministic choice between
-\texttt{False} and \texttt{True}.  The possible outputs consists of zero or
+\texttt{False} and \texttt{True}.  The possible outputs consist of zero or
 more copies of the string ``hello world''.  Note that this single state DFA
@@
 
-``Running'' this Harmony program will try all possible executions, which
+Running this Harmony program will try all possible executions, which
 includes all possible values for $x$.  Try it out (here \texttt{\$}
@@
 
-(For this to work, make sure \texttt{harmony} is in your command shell's search path.)
 The $\textbf{assert}$ statement checks that the output is
@@
 
-\begin{figure}
-\begin{code}
-\harmonysource{queuelin}
-\end{code}
-\caption{\harmonylink{code/queuelin.hny} Queue implementation with linearization points}
-\label{fig:queuelin}
-\end{figure}
-
-\begin{figure}
-\begin{code}
-\harmonysource{qtestconc}
-\end{code}
-\caption{\harmonylink{code/qtestconc.hny} Concurrent queue test}
-\label{fig:qtestconc}
-\end{figure}
-
 Testing is a way to increase confidence in the correctness
@@
 \end{itemize}
-For each operation, the code first decides whether to
+For each operation, the code first chooses whether to
 perform a \texttt{get} or \texttt{put} operation.
 In the case of a \texttt{put} operation, the code also
-decides which value to append to the queue.
+chooses which value to append to the queue.
 All operations are performed on both the queue implementation
@@
 
+\begin{figure}
+\begin{center}
+\includegraphics[width=\textwidth]{../code/qtestpar.png}
+\end{center}
+\begin{code}
+\harmonysource{qtestpar}
+\end{code}
+\caption{\harmonylink{code/qtestpar.hny} Concurrent queue test.  The
+behavior DFA is for $\mathtt{NOPS} = 2$.}
+\label{fig:qtestpar}
+\end{figure}
+
+\begin{figure}
+\begin{code}
+\harmonysource{queueseq}
+\end{code}
+\caption{\harmonylink{code/queueseq.hny} Sequential but not a concurrent
+queue implementation}
+\label{fig:queuebug}
+\end{figure}
+
 The next step is to test if the queue implementation
 meets the \emph{concurrent} queue specification or not.
-\autoref{fig:queues}(b) shows the concurrent version of
-\autoref{fig:queues}(a)---it simply makes all operations
-(except instantiation itself) atomic.
-
-But we have not yet shown what the
-concurrent queue specification is.
-
-We briefly mentioned the notion of \emph{linearizability}
-in \autoref{ch:cds}.
-Basically, we want a concurrent queue to behave
-consistently with a sequential queue in that all
-\texttt{put}
-and \texttt{get} operations should appear to happen
-in a total order.
-Moreover, we want to make sure that if some
-\texttt{put} or \texttt{get} operation $o_1$ finished
-before another operation $o_2$ started, then $o_1$
-should appear to happen before $o_2$ in the total order.
-If these two conditions are met, then we say that
-the concurrent queue implementation is linearizable.
-
-In general, if a data structure is protected by
-a single lock and every operation on that data
-structure starts with acquiring the lock and ends
-with releasing the lock, it will automatically be
-linearizable.
-The queue implementation in \autoref{fig:queueconc} does
-not quite match this pattern, as the \texttt{put}
-operation allocates a new node before acquiring
-the lock.
-However, in this case that is not a problem, as
-the new node has no dependencies on the queue
-when it is allocated.
-
-Still, it would be useful to check in Harmony that
-\autoref{fig:queueconc} is linearizable.
-To do this, instead of applying the operations sequentially,
-we want the test program to invoke the operations concurrently,
-consider all possible interleavings, and see if the result
-is consistent with an appropriate sequential execution of the
-operations.
-
-\index{logical timestamp}%
-Harmony provides support for testing linearizability,
-but requires that the programmer
-identifies what are known as \emph{linearization points}
-in the implementation that indicate exactly \emph{which} sequential
-execution the concurrent execution must align with.
-\autoref{fig:queuelin} is a copy of \autoref{fig:queueconc} extended with
-linearization points.
-For each operation (\texttt{get} and \texttt{put}), the corresponding
-linearization point must occur somewhere between acquiring and releasing
-the lock.
-%
-Each linearization point execution is assigned a logical timestamp.
-Logical timestamps are numbered $0, 1, ...$
-To do so, we have added a counter (\texttt{time}) to the \texttt{Queue}.
-Method \texttt{\_linpoint} saves the current counter in
-\textbf{this}.\texttt{qtime} and increments the counter.
-The \textbf{this} dictionary maintains \emph{thread-local state} associated
-with the thread (\autoref{ch:harmonymachine})---it contains variables that
-can be accessed by any method in the thread.
-
-Given the linearization points, \autoref{fig:qtestconc} shows how
-linearizability can be tested.
-The test program is similar to the sequential test program
-(\autoref{fig:qtestseq}) but starts a thread for each operation.
-The operations are executed concurrently on the concurrent queue
-implementation of \autoref{fig:queuelin}, but they are executed sequentially
-on the sequential queue specification of \autoref{fig:queues}(a).
-To that end, the test program maintains a global time variable
-\textit{qtime}, and each thread waits until the timestamp assigned to
-the last concurrent queue operation matches \textit{qtime} before invoking
-the sequential operation in the specification.
-Afterward, it atomically increments the shared \textit{qtime} variable.
-This results in the operations being executed sequentially against the
-sequential specification in the same order of the linearization points
-of the concurrent specification.
+\autoref{fig:queues}(b) shows the concurrent queue specification.
+It is similar to the sequential specification in \autoref{fig:queues}(a)
+but makes all operations (except instantiation itself) atomic.
+Testing the implementation of a concurrent queue specification is
+trickier than testing the implementation of a sequential one because
+there are many more scenarios to check.
+
+We would like a way that systematically compares behaviors of the
+concurrent queue implementation with behaviors of the concurrent
+queue specification.  But we cannot do this by composing the specification
+and the implementation and simply run the same test operations on both as we
+did before---concurrency make the operations non-determistic and thus
+the specification and implementation of a single execution might produce
+different results, even if both are correct.
+Instead, we will create a test program that tries various concurrent
+combinations of queue operations, and run it twice: once against the
+specification of the concurrent queue and once against the
+implementation.  We will then check if the behaviors obtained from
+running the implementation is a subset of the behaviors obtained
+from the specification.
+
+We will start with a test program that tries concurrent combinations
+of various queue operations.
+\autoref{fig:qtestpar} shows the test program.  It starts
+\texttt{NOPS} threads and each choose to do either a \texttt{put}
+or a \texttt{get} operation.  In case of a \texttt{put} operation,
+the thread enqueues its own name.  In order to output the behaviors,
+each thread prints what operation it is about to perform,
+and afterwards it prints that the operation has
+completed (including the return value if any).
+The figure shows the deterministic finite automaton when the test
+program is run against the specification in the case
+$\texttt{NOPS} = 2$---for $\texttt{NOPS} = 4$ it would be much
+too large to print here.  Since there are no cycles in the DFA, you
+can follow some paths through this DFA and see that they are valid
+interleavings of the threads.
+
+If you run the same test program against the implementation
+of \autoref{fig:queueconc}, you will get the same output.  You can
+try this for various $\texttt{NOPS}$, although it gets increasingly
+harder to check by hand that the generated DFAs are the same as
+\texttt{NOPS} increases.
+Now run the test program against \autoref{fig:queueseq}, which is
+clearly an incorrect implementation of the concurrent queue specification
+because it contains no synchronization code.
+However, Harmony does not immediately detect
+any problems.  In fact, for $\mathtt{NOPS} = 2$ it even generates
+the same set of behaviors.
+This is because the test program only outputs the behaviors---it does
+not check if they are correct.
+
+Harmony does have a way to check the behavior of one program against
+the behavior of another.  In particular, we want to check if the
+behaviors of the implementations we have match the behavior of the
+specification.  The following shows, for example, how to check the
+\texttt{queueconc.hny} implementation on the command line:
+
+\begin{code}
+\begin{verbatim}
+$ harmony -o queue4.hfa code/qtestpar.hny
+$ harmony -B queue4.hfa -m queue=queueconc code/qtestpar.hny
+\end{verbatim}
+\end{code}
+
+The first command runs the \texttt{code/qtestpar.hny} program (with the
+default 4 threads)
+and writes a representation of the output DFA in the file \texttt{queue4.hfa}.
+The second command runs the same test program, but using the queue
+implementation in the file \texttt{code/queueconc.hny}.  Moreover, it reads
+the DFA in \texttt{queue4.hfa} to check if every behavior of the second
+run of the test program is also a behavior of the first run.
+You can try the same using the \texttt{code/queueseq.hny} implementation
+and find that this implementation has behaviors that are not allowed by
+the specification.
+
+\begin{comment}
 
@@
 
+\end{comment}
+
 \section*{Exercises}
@@
 the sequential specification.
-\item Extend \autoref{fig:linkedlist} with linearization points and
-check if the implementation is linearizable.
-\item Check if the extended queue implementation of \autoref{ex:qremove}
-is linearizable.
+\item Also create a concurrent specification and a concurrent test program.
+Check if the behaviors of \autoref{fig:linkedlist} match those of the
+specification.
+% \item Extend \autoref{fig:linkedlist} with linearization points and
+% check if the implementation is linearizable.
+% \item Check if the extended queue implementation of \autoref{ex:qremove}
+% is linearizable.
 \end{problems}
@@
 \begin{code}
-\harmonysource{queuebug}
-\end{code}
-\caption{\harmonylink{code/queuebug.hny} A buggy queue implementation}
-\label{fig:queuebug}
+\harmonysource{queuebroken}
+\end{code}
+\caption{\harmonylink{code/queuebroken.hny} Another buggy queue implementation}
+\label{fig:queuebroken}
 \end{figure}
@@
 \caption{\harmonyref{queuebug.html}{Harmony output of running \autoref{fig:qtest3} against
-\autoref{fig:queuebug}}}
+\autoref{fig:queuebroken}}}
 \label{fig:queuebug1}
@@
 
-\autoref{fig:queuebug} contains an attempt at a queue implementation
+\autoref{fig:queuebroken} contains an attempt at a queue implementation
 where the queue is implemented by a linked list, with the first node
@@
 (Run this using the \texttt{-m} flag like this:
-\texttt{harmony -m queue=queuebug code/qtestseq.hny})
+\texttt{harmony -m queue=queuebroken code/qtestseq.hny})
 The concurrent \texttt{put} test of
@@
 the test in \autoref{fig:qtest3} against the queue code
-in \autoref{fig:queuebug}.
+in \autoref{fig:queuebroken}.
 There is quite a bit of information in the Harmony output, and it's
@@
 Next \texttt{putter}(1) (thread T2) takes a turn.  It gets to line 10 in
-\autoref{fig:queuebug} where it's trying to acquire
+\autoref{fig:queuebroken} where it's trying to acquire
 $q$$\rightarrow$\texttt{lock}.  If we look at the bottom right,
@@
 
+\chapter{Directly checking linearizability}
+
+\begin{figure}
+\begin{code}
+\harmonysource{queuelin}
+\end{code}
+\caption{\harmonylink{code/queuelin.hny} Queue implementation with linearization points}
+\label{fig:queuelin}
+\end{figure}
+
+\begin{figure}
+\begin{code}
+\harmonysource{qtestconc}
+\end{code}
+\caption{\harmonylink{code/qtestconc.hny} Concurrent queue test}
+\label{fig:qtestconc}
+\end{figure}
+
+We briefly mentioned the notion of \emph{linearizability}
+in \autoref{ch:cds}.
+Basically, we want a concurrent queue to behave
+consistently with a sequential queue in that all
+\texttt{put}
+and \texttt{get} operations should appear to happen
+in a total order.
+Moreover, we want to make sure that if some
+\texttt{put} or \texttt{get} operation $o_1$ finished
+before another operation $o_2$ started, then $o_1$
+should appear to happen before $o_2$ in the total order.
+If these two conditions are met, then we say that
+the concurrent queue implementation is linearizable.
+
+In general, if a data structure is protected by
+a single lock and every operation on that data
+structure starts with acquiring the lock and ends
+with releasing the lock, it will automatically be
+linearizable.
+The queue implementation in \autoref{fig:queueconc} does
+not quite match this pattern, as the \texttt{put}
+operation allocates a new node before acquiring
+the lock.
+However, in this case that is not a problem, as
+the new node has no dependencies on the queue
+when it is allocated.
+
+Still, it would be useful to check in Harmony that
+\autoref{fig:queueconc} is linearizable.
+To do this, instead of applying the operations sequentially,
+we want the test program to invoke the operations concurrently,
+consider all possible interleavings, and see if the result
+is consistent with an appropriate sequential execution of the
+operations.
+
+\index{logical timestamp}%
+Harmony provides support for testing linearizability,
+but requires that the programmer
+identifies what are known as \emph{linearization points}
+in the implementation that indicate exactly \emph{which} sequential
+execution the concurrent execution must align with.
+\autoref{fig:queuelin} is a copy of \autoref{fig:queueconc} extended with
+linearization points.
+For each operation (\texttt{get} and \texttt{put}), the corresponding
+linearization point must occur somewhere between acquiring and releasing
+the lock.
+%
+Each linearization point execution is assigned a logical timestamp.
+Logical timestamps are numbered $0, 1, ...$
+To do so, we have added a counter (\texttt{time}) to the \texttt{Queue}.
+Method \texttt{\_linpoint} saves the current counter in
+\textbf{this}.\texttt{qtime} and increments the counter.
+The \textbf{this} dictionary maintains \emph{thread-local state} associated
+with the thread (\autoref{ch:harmonymachine})---it contains variables that
+can be accessed by any method in the thread.
+
+Given the linearization points, \autoref{fig:qtestconc} shows how
+linearizability can be tested.
+The test program is similar to the sequential test program
+(\autoref{fig:qtestseq}) but starts a thread for each operation.
+The operations are executed concurrently on the concurrent queue
+implementation of \autoref{fig:queuelin}, but they are executed sequentially
+on the sequential queue specification of \autoref{fig:queues}(a).
+To that end, the test program maintains a global time variable
+\textit{qtime}, and each thread waits until the timestamp assigned to
+the last concurrent queue operation matches \textit{qtime} before invoking
+the sequential operation in the specification.
+Afterward, it atomically increments the shared \textit{qtime} variable.
+This results in the operations being executed sequentially against the
+sequential specification in the same order of the linearization points
+of the concurrent specification.
+
 \chapter{Installing and Running Harmony}\label{app:install}

### 2021-12-11 src/charm/charm.c

@@ -1483,14 +1483,14 @@ static void pr_state(struct global_t *global, FILE *fp, struct state *state, int
 }
 
 static void usage(char *prog){
-    fprintf(stderr, "Usage: %s [-c] [-t maxtime] file.json\n", prog);
+    fprintf(stderr, "Usage: %s [-c] [-t<maxtime>] [-B<dfafile>] -o<outfile> file.json\n", prog);
     exit(1);
 }
 
 int main(int argc, char **argv){
     bool cflag = false;
     int i, maxtime = 300000000 /* about 10 years */;
-    char *dfafile = NULL;
+    char *outfile = NULL, *dfafile = NULL;
     for (i = 1; i < argc; i++) {
         if (*argv[i] != '-') {
             break;
@@ -1509,6 +1509,9 @@ int main(int argc, char **argv){
         case 'B':
             dfafile = &argv[i][2];
             break;
+        case 'o':
+            outfile = &argv[i][2];
+            break;
         case 'x':
             printf("Charm model checker working\n");
             return 0;
@@ -1516,23 +1519,10 @@ int main(int argc, char **argv){
             usage(argv[0]);
         }
     }
-    if (argc - i > 1) {
+    if (argc - i != 1 || outfile == NULL) {
         usage(argv[0]);
     }
-    char *fname = i == argc ? "harmony.hvm" : argv[i];
-
-    char *outfile, *dotloc = strrchr(fname, '.');
-    int r;
-    if (dotloc == NULL) {
-        r = asprintf(&outfile, "%s.hco", fname);
-    }
-    else {
-        r = asprintf(&outfile, "%.*s.hco", (int) (dotloc - fname), fname);
-    }
-    if (r < 0) {       // mostly to suppress asprintf return value warning
-        outfile = "harmony.hco";
-    }
-
+    char *fname = argv[i];
     double timeout = gettime() + maxtime;
 
     // initialize modules

### 2021-12-11 src/harmony/brief.py

@@ -2,8 +2,6 @@ def brief_kv(js):
     return (brief_string(js["key"]), brief_string(js["value"]))
 
 def brief_idx(js):
-    if js["type"] == "atom":
-        return brief_string(js)
     return "[" + brief_string(js) + "]"
 
 def brief_string(js):
@@ -12,7 +10,7 @@ def brief_string(js):
     if type in { "bool", "int" }:
         return v
     if type == "atom":
-        return "." + v
+        return '"' + v + '"'
     if type == "set":
         if v == []:
             return "{}"
@@ -115,14 +113,12 @@ class Brief:
                 self.failure = self.lastmis["failure"]
             self.interrupted = "interrupt" in self.lastmis and self.lastmis["interrupt"] == "True"
 
-    def run(self, stem, outputflag):
-        fname = stem + ".hco"
-        with open(fname) as f:
+    def run(self, outputfiles):
+        with open(outputfiles["hco"]) as f:
             top = json.load(f)
             assert isinstance(top, dict)
             if top["issue"] == "No issues":
-                if outputflag:
-                    behavior_parse(top, True, stem);
+                behavior_parse(top, True, outputfiles);
                 return True
 
             # print("Issue:", top["issue"])

### 2021-12-11 src/harmony/charm.js

@@ -95,12 +95,7 @@ function json_string_address(obj) {
   }
   var result = "?" + obj[0].value;
   for (var i = 1; i < obj.length; i++) {
-    if (obj[i].type == "atom") {
-      result += "." + obj[i].value;
-    }
-    else {
-      result += "[" + json_string(obj[i]) + "]";
-    }
+    result += "[" + json_string(obj[i]) + "]";
   }
   return result;
 }
@@ -118,7 +113,7 @@ function json_string(obj) {
     return obj.value;
     break;
   case "atom":
-    return "." + obj.value;
+    return '"' + obj.value + '"';
   case "set":
     return json_string_set(obj.value);
   case "dict":

### 2021-12-11 src/harmony/genhtml.py

@@ -160,10 +160,10 @@ m4_include(charm.js)
         print("  </div>", file=f)
         print("</div>", file=f)
 
-    def html_botright(self, f, stem, outputflag):
+    def html_botright(self, f, outputfiles):
         if self.nthreads == 0:
-            if outputflag:
-                print("<img src='%s.png' alt='DFA image'>"%pathlib.PurePosixPath(stem).name, file=f)
+            if outputfiles["png"] != None:
+                print("<img src='%s' alt='DFA image'>"%outputfiles["png"], file=f)
             return
         print("<table border='1'", file=f)
         print("  <thead>", file=f)
@@ -205,7 +205,7 @@ m4_include(charm.js)
         print("  </tbody>", file=f)
         print("</table>", file=f)
 
-    def html_outer(self, f, stem, outputflag):
+    def html_outer(self, f, outputfiles):
         print("<table>", file=f)
         print("  <tr>", file=f)
         print("    <td colspan='2'>", file=f)
@@ -227,7 +227,7 @@ m4_include(charm.js)
         self.html_botleft(f)
         print("    </td>", file=f)
         print("    <td valign='top'>", file=f)
-        self.html_botright(f, stem, outputflag)
+        self.html_botright(f, outputfiles)
         print("    </td>", file=f)
         print("  </tr>", file=f)
         print("</table>", file=f)
@@ -242,7 +242,7 @@ m4_include(charm.js)
         print("  " + str(path), end="", file=f)
         return index + 1
 
-    def html_script(self, f, stem):
+    def html_script(self, f, outputfiles):
         print("<script>", file=f)
         print("var nthreads = %d;"%self.nthreads, file=f)
         print("var nmegasteps = %d;"%self.nmegasteps, file=f)
@@ -251,16 +251,16 @@ m4_include(charm.js)
         print(file=f)
         print("];", file=f)
         print("var state =", file=f)
-        self.file_include(stem + ".hco", f)
+        self.file_include(outputfiles["hco"], f)
         print(";", file=f)
         print(self.js, file=f)
         # file_include("charm.js", f)
         print("</script>", file=f)
 
-    def html_body(self, f, stem, outputflag):
+    def html_body(self, f, outputfiles):
         print("<body>", file=f)
-        self.html_outer(f, stem, outputflag)
-        self.html_script(f, stem)
+        self.html_outer(f, outputfiles)
+        self.html_script(f, outputfiles)
         print("</body>", file=f)
 
     def html_head(self, f):
[…]

### 2021-12-15 book/paper.tex
Latest

@@
 \begin{code}
-\textbf{assert} \texttt{queue.get}(\textit{q}) in \{ \textbf{None}, 1, 2 \}
+\textbf{assert} \texttt{queue.get}($q$) in \{ \textbf{None}, 1, 2 \}
 \end{code}
@@
 queue implementation}
-\label{fig:queuebug}
+\label{fig:queueseq}
 \end{figure}
@@
 \begin{center}
-\includegraphics[width=\textwidth]{figures/queuebug1.pdf}
+\includegraphics[width=\textwidth]{figures/queuebroken1.png}
 \end{center}
-\caption{\harmonyref{queuebug.html}{Harmony output of running \autoref{fig:qtest3} against
-\autoref{fig:queuebroken}}}
-\label{fig:queuebug1}
+\caption{\harmonyref{queuebug.html}{Harmony output of running
+\autoref{fig:qtestpar} against \autoref{fig:queuebroken}}}
+\label{fig:queuebroken1}
 \end{figure}
@@
 
-Let us run the code through some of the test programs in the last
+Let us run the code through the test programs in the last
 chapter.  Harmony does not detect any issues with the sequential test
@@
 \texttt{harmony -m queue=queuebroken code/qtestseq.hny})
-The concurrent \texttt{put} test of
-\autoref{fig:qtest1} as well as the concurrent \texttt{get} test of
-\autoref{fig:qtest2} also find no problems.  However, when we run
-the new queue code through the test in \autoref{fig:qtest3}, Harmony
-finds a problem.
-
-\autoref{fig:queuebug1} shows the Harmony output of running
-the test in \autoref{fig:qtest3} against the queue code
-in \autoref{fig:queuebroken}.
-There is quite a bit of information in the Harmony output, and it's
-important to learn to navigate through it.
+However, when we run
+the new queue code through the test in \autoref{fig:qtestpar}, Harmony
+finds a problem (even without specifying a behavior).  The command line
+for this is \texttt{harmony -m queue=queuebroken code/qtestpar.hny})
+
+Before we go look at the details of what went wrong, we want to make
+sure we generate the simplest scenario.  So first we want to explore
+what the smallest \texttt{NOPS} (number of operations or number of
+threads) that causes the bug to occur.  With some experimentation,
+we find that \texttt{NOPS} = 2 does not find a problem, but
+\texttt{NOPS} = 3 does
+(\texttt{harmony -m queue=queuebroken -c NOPS=3 code/qtestpar.hny})).
+\autoref{fig:queuebroken1} shows the Harmony output.
+
+There is quite a bit of information in the Harmony output, and while
+it may seem intimidating, we have to learn to navigate through it
+step-by-step.
 Let's start with looking at the red text.  Harmony found a safety
 violation (something bad happened during one of the possible executions),
-and in particular \texttt{putter}(1) (thread T2) was trying to dereference
-\texttt{alloc}.\textit{pool}[0].\texttt{lock} in turn 5.
+and in particular \texttt{thread}(2) (thread T2) was trying to dereference
+\texttt{?alloc["pool"][0]["lock"]}
+(aka \texttt{?alloc}.\textit{pool}[0].\texttt{lock}).
 
 The \texttt{alloc} module maintains a shared array \textit{pool} that
-it uses for dynamic allocation.  Apparently \texttt{putter}(1) tried to access
+it uses for dynamic allocation.  Apparently T2 tried to access
 \textit{pool}[0], but it does not exist, meaning that either it was not
 yet allocated, or it had been freed since it was allocated.
-When we look at the top half of the figure, we see that in fact
-\texttt{putter}(0) (T1) allocated \textit{pool}[0] in turn 2, while
-\texttt{getter}() (T4) released it in turn 4.
-
-So how did we get there?  We can start by single stepping through the
+When we look at the top half of the figure, we see that in fact thread
+T1 allocated \textit{pool}[0] in turn 2, but T3 freed it in turn 4.
+Looking back down, we see that T1 executed \texttt{thread}(1) and has
+since terminated, while T3 is executing \texttt{thread}(3).
+
+Looking further at the stack traces, we can see that
+T3 was in the process of executing
+\texttt{release}(?$q$.\texttt{lock}) within
+\texttt{get}(?$q$).  Looking in the top green row,
+we see that the value of $q$ is
+\{ \texttt{.lock}: \texttt{True}, \texttt{.next}: \texttt{None}, \texttt{.value}: \texttt{None} \}.
+T1 is currently executing
+\texttt{acquire}(\texttt{?alloc}.\textit{pool}[0].\texttt{lock})
+within \texttt{put}(?$q$, 2), but 
+\texttt{alloc}.\textit{pool}[0] does not exist.
+The corresponding line of Harmony code is
+\textbf{atomically when not} !\textit{binsema}
+in line 25 of the \texttt{sync} module.
+
+So how did we get there?  In the top we can see that the order of
+events was the following:
+\begin{enumerate}
+\item[0:] initialization completed, leaving the value of $q$ at
+\{ \texttt{.lock}: \texttt{False}, \texttt{.next}: \texttt{None}, \texttt{.value}: \texttt{None} \};
+\item[1:] thread T1 (\texttt{thread}(1)) ran and finished executing
+\texttt{put}(1) (see the output column for that clue: the thread printed that).
+We can see that $q$.\texttt{next} now points to \texttt{alloc}.\textit{pool}[0],
+which the thread must have allocated.  The contents is 
+\{ \texttt{.lock}: \texttt{False}, \texttt{.next}: \texttt{None}, \texttt{.value}: 1 \}, as expected;
+\item[2:] thread T2 (\texttt{thread}(1)) started running, calling
+\texttt{put}(?$q$, 2).  We can see it got as far as putting 2 on the queue;
+\item[3:] thread T3 (\texttt{thread}(1)) started running, calling
+\texttt{get}($?q$).  We can also see that it freed \textit{pool}[0].
+We know that it is in the process of releasing $q$.\texttt{lock};
+\item[4:] thread T2 resumes and tries to access \textit{pool}[0], which
+no longer exists because T3 just freed it.
+\end{enumerate}
+
+Clearly there was a race in which T2 was trying to lock
+?\textit{pool}[0].\texttt{lock} (which contained the node with the value 1)
+while T3 was freeing that very same node, and T2 lost the race.
+
+We can start by single stepping through the
 actions of \texttt{putter}(0) by clicking on its first block.

### 2021-12-16 book/paper.tex
Latest
Still working on debugging chapter

@@
 \texttt{release}(?$q$.\texttt{lock}) within
-\texttt{get}(?$q$).  Looking in the top green row,
-we see that the value of $q$ is
-\{ \texttt{.lock}: \texttt{True}, \texttt{.next}: \texttt{None}, \texttt{.value}: \texttt{None} \}.
+\texttt{get}(?$q$).
+% Looking in the top green row, we see that the value of $q$ is
+% \{ \texttt{.lock}: \texttt{True}, \texttt{.next}: \texttt{None}, \texttt{.value}: \texttt{None} \}.
 T1 is currently executing
@@
 \begin{enumerate}
-\item[0:] initialization completed, leaving the value of $q$ at
+\item[0:] initialization completed, with $q$ being
 \{ \texttt{.lock}: \texttt{False}, \texttt{.next}: \texttt{None}, \texttt{.value}: \texttt{None} \};
@@
 \item[2:] thread T2 (\texttt{thread}(1)) started running, calling
-\texttt{put}(?$q$, 2).  We can see it got as far as putting 2 on the queue;
+\texttt{put}(?$q$, 2).  We can see it got as far as putting 2 on the queue,
+but it is not yet done.  It is currently trying to acquire
+\texttt{alloc}.\textit{pool}[0].\texttt{lock};
 \item[3:] thread T3 (\texttt{thread}(1)) started running, calling
-\texttt{get}($?q$).  We can also see that it freed \textit{pool}[0].
-We know that it is in the process of releasing $q$.\texttt{lock};
+\texttt{get}($?q$).  We can also see that it freed \textit{pool}[0],
+and is now releasing $q$.\texttt{lock};
 \item[4:] thread T2 resumes and tries to access \textit{pool}[0], which
-no longer exists because T3 just freed it.
+no longer exists (because T3 just freed it).
 \end{enumerate}
@@
 while T3 was freeing that very same node, and T2 lost the race.
+More precisely, T2 was executing \texttt{put}(?$q$, 2), when T3
+preempted it with \texttt{get}(?$q$) and removed the node that T2
+was trying to access.  But why did the locks not prevent this?
+
+\index{reserve debugging}%
+It is time to start stepping through the code that has been executed
+before this happened.  This is sometimes known as
+\emph{reverse debugging}.  In fact, Harmony allows you to step through
+code forwards and backwards.  In this case, we first want to see what
+T2 is doing.  You can click on its first (top-left) orange box to
+time-travel to that part in the execution.  Now by hitting
+$<$return$>$ repeatedly, we can quickly skip through the code.
+T2 first calls \texttt{put}(?$q$, 1) and then allocates a new
+node initialized with a lock.  Keep stepping until it executes
+$\mathit{nq} = q$.  Hit $<$return$>$ once more and inspect the
+state of T2 in the lower-right corner.  You can see that variable
+\textit{nq} is initialized to~?$q$.  T2 then enters into the
+\textbf{while} loop and tries to acquire
+$\mathit{nq}$$\rightarrow$$\mathtt{lock}$.  This succeeds, and
+next T2 executes \textbf{let} $n = \mathit{nq}$$\rightarrow$$\mathtt{next}$.
+Now $n = ?\mathtt{alloc}.\mathit{alloc}[0]$, which is not \texttt{None}.
+It then releases $\mathit{nq}$$\rightarrow$$\mathtt{lock}$
+(\textit{nq} points to $q$).  It then sets \textit{nq} to $n$, which
+is still $?\mathtt{alloc}.\mathit{alloc}[0]$.  Finally, it calls
+$\mathtt{acquire}(?\mathit{nq}$$\rightarrow$$\mathtt{lock})$.
+But before it can complete that operation, T3 runs next.
+
+T3 chooses "get" and then goes on to invoke $\mathtt{get}(?q)$.
+This first successfull acquires $q$$\rightarrow$$\texttt{lock}$.
+It then finds out that $q$$\rightarrow$$\texttt{next}$ points to
+$\texttt{alloc}.\textit{pool}[0]$.  It sets \textit{node} to
+$\texttt{alloc}.\textit{pool}[0]$ as well and sets
+$q$$\rightarrow$$\texttt{next}$ to 
+$\mathtt{node}$$\rightarrow$$\texttt{next}$.
+
+XXXXXXXX
 

### 2021-12-16 book/paper.tex
Latest

@@
 next T2 executes \textbf{let} $n = \mathit{nq}$$\rightarrow$$\mathtt{next}$.
-Now $n = ?\mathtt{alloc}.\mathit{alloc}[0]$, which is not \texttt{None}.
+Now $n =\,?\mathtt{alloc}.\mathit{alloc}[0]$, which is not \texttt{None}.
 It then releases $\mathit{nq}$$\rightarrow$$\mathtt{lock}$
@@
 T3 chooses "get" and then goes on to invoke $\mathtt{get}(?q)$.
-This first successfull acquires $q$$\rightarrow$$\texttt{lock}$.
-It then finds out that $q$$\rightarrow$$\texttt{next}$ points to
-$\texttt{alloc}.\textit{pool}[0]$.  It sets \textit{node} to
+This first successfully acquires $q$$\rightarrow$$\texttt{lock}$.
+T3 then finds out that $q$$\rightarrow$$\texttt{next}$ points to
+$\texttt{alloc}.\textit{pool}[0]$.  T3 sets \textit{node} to
 $\texttt{alloc}.\textit{pool}[0]$ as well and sets
@@
 $\mathtt{node}$$\rightarrow$$\texttt{next}$.
-
-XXXXXXXX
-
-We can start by single stepping through the
-actions of \texttt{putter}(0) by clicking on its first block.
-By hitting return repeatedly, we can go through the lines of code that
-it is executing.  Doing so, we can see that it allocates
-\textit{pool}[0] and uses that as the node to contain value 0 and add
-this node to the queue \textit{qtest}.  At the end of its turn,
-\texttt{putter}(0) has just finished \texttt{put}(0).  Looking at the
-state in the top right, everything looks good.  The first node
-(\textit{testq}) points to the allocated node with the value 0 in it,
-and its \texttt{next} pointer is \textbf{None}.  The locks on both nodes
-are released.  So far so good.
-
-Next \texttt{putter}(1) (thread T2) takes a turn.  It gets to line 10 in
-\autoref{fig:queuebroken} where it's trying to acquire
-$q$$\rightarrow$\texttt{lock}.  If we look at the bottom right,
-we see that $q$ points to \textit{pool}[0], the node that has value 0
-in it.  Again, so far so good.
-
-But before \texttt{putter}(1) obtains
-the lock on \textit{pool}[0], \texttt{getter}() (thread T4) starts
-running.  \texttt{getter}() acquires the lock on \textit{qtest}, which
-still points to \textit{pool}[0].  \texttt{getter} then extracts the
-value from \textit{pool}[0] and then releases it.
-At this point \textit{pool}[0] no longer exists, but then
-\texttt{putter}(1) starts running again from where it left off.
-It was about to acquire the lock on \textit{pool}[0], but now that
-node no longer exists.
+T3 sets \textit{result} to 
+$\mathtt{node}$$\rightarrow$$\texttt{value}$ (which is 1)
+and then frees \textit{node}.  This is where the problem is---T2 is
+about to acquire the lock in that same node.
 

### 2021-12-16 src/charm/charm.c
Renamed "Log" to Print

@@ -293,11 +293,11 @@ static bool onestep(
         // See if we need to break out of this step.  It's complicated.
         // If the atomicFlag is set, then definitely not.
         // If the atomicFlag is not set, then it depends on whether the instruction
-        // is "breakable" (Load, Store, or Del), or a print instruction (Log).
+        // is "breakable" (Load, Store, or Del), or a print instruction (Print).
         // If neither, then no need to break---that would lead to unnecessary
         // state explosion.  Otherwise it depends on whether the atomic counter
         // is zero or not.  If not zero, then we should set the atomic flag.
-        // For a Log instruction, it also depends on whether one already
+        // For a Print instruction, it also depends on whether one already
         // happened.  If not, we don't need to break.
         struct instr_t *next_instr = &global->code.instrs[step->ctx->pc];
         if (!step->ctx->atomicFlag && (next_instr->breakable || next_instr->log)) {

### 2021-12-17 book/paper.tex
Renamed "Log" to Print
Missing files

@@
 \begin{center}
-\includegraphics[width=0.45\textwidth]{../code/hello1.png}
+\includegraphics[width=0.45\textwidth]{figures/hello1.png}
 \end{center}
@@
 \begin{center}
-\includegraphics[width=0.45\textwidth]{../code/hello3.png}
+\includegraphics[width=0.45\textwidth]{figures/hello3.png}
 \end{center}
@@
 \begin{center}
-\includegraphics[width=0.15\textwidth]{../code/hello4.png}
+\includegraphics[width=0.15\textwidth]{figures/hello4.png}
 \end{center}
@@
 \begin{tabular}{ccc}
-\includegraphics[width=0.45\textwidth]{../code/hello5.png}
+\includegraphics[width=0.45\textwidth]{figures/hello5.png}
 & \rule{0.05\linewidth}{0in} &
-\includegraphics[width=0.45\textwidth]{../code/hello6.png} \\
+\includegraphics[width=0.45\textwidth]{figures/hello6.png} \\
 \begin{tcolorbox}[width=0.4\linewidth]
@@
 \begin{center}
-\includegraphics[width=0.9\textwidth]{../code/hello7.png}
+\includegraphics[width=0.9\textwidth]{figures/hello7.png}
 \end{center}
@@
 \begin{center}
-\includegraphics[width=0.8\textwidth]{../code/hello8.png}
+\includegraphics[width=0.8\textwidth]{figures/hello8.png}
 \end{center}
@@
 % is linearizable.
+\item Rewrite \autoref{fig:qtestseq} so it only imports \texttt{queue}
+and runs \texttt{NOPS} nondeterministically chosen operations against
+it (similar in style to \autoref{fig:qtestpar} but without threads).
+Then use behaviors to check that \autoref{fig:queueconc} and
+\autoref{fig:queueseq} are correct sequential implementations of
+the queue.  Check your test program by also trying it on one or two
+buggy queue implementations.
 \end{problems}
@@
 \hline
-Possibly $i$ & pop a value to check the $i^{th}$ predicate in a \textbf{possibly} statement \\
+Print & pop a value and add to the print history \\
 \hline

### 2021-12-17 book/paper.tex
Figure missing
Added missing period

@@
 The \texttt{-o hello1.png} arguments tell harmony to write the output
-in the file \texttt{hello1.png}
+in the file \texttt{hello1.png}.
 The output is a \emph{Deterministic State Machine} (DFA).  The green circle
@@
 \begin{center}
-\includegraphics[width=\textwidth]{../code/qtestpar.png}
+\includegraphics[width=\textwidth]{figures/qtestpar.png}
 \end{center}

### 2021-12-17 book/paper.tex
Some small updates

@@
 \end{itemize}
-\index{select statement}%
 
@@
 
+\begin{comment}
 Note the use of the \textbf{possibly} statement in this code.
@@
 capable of probabilistic reasoning.
+\end{comment}
 
@@
 
+\begin{comment}
 The \texttt{possibly} statement checks if there exist executions where 0 is
@@
 Also, it is nice to know that both 0 and 1 can be decided.
+\end{comment}
 
@@
 \hline
+\texttt{@l: S} & \texttt{l} is a label \\
+\hline
 \texttt{assert $b$ [, $e$]} & $b$ is a boolean.  Optionally report value of expression $e$\\
@@
 \hline
-\texttt{possibly $b$ [, $b$, ...]} & each $b$ is a boolean \\
-\hline
-\texttt{select $a$ in $e$: S} & $a$ is a bound variable, $e$ is a set,
-                            \texttt{S} a list of statements\\
+\texttt{print $b$ [, $b$, ...]} & each $b$ is a boolean \\
 \hline
@@
 
-\subsection*{\textbf{possibly}}
-
-The statement \textbf{possibly $b_1, b_2, ...$} atomically evaluates all
-predicates $b_i$.
-At completion, Harmony reports which of the predicates never held.
-While \texttt{assert} statements check that nothing ever goes wrong,
-\texttt{possibly} statements can check that certain things sometimes
-go right.
+\subsection*{\textbf{print}}
+
+The statement \textbf{print $e$} atomically evaluates $e$ in read-only
+mode and adds the result to the print log.  The print log is used to
+create an ``external behavior DFA'' for the Harmony program.
 

### 2021-12-21 book/paper.tex
Working on a new preprocessor for the book Also removed '@' from labels Fixed a few problems here and there

@@
     to complete each round}
+}
+\newglossaryentry{behavior}
+{
+  name=behavior,
+  description={is a sequence of states.  A trace uniquely defines a behavior but not vice versa}
 }
@@
   name=property,
-  description={describes a set of execution traces or histories that are allowed by
+  description={describes a set of execution traces or behaviors or histories that are allowed by
     a program.  Safety properties are properties in which ``no bad things happen,''
@@
   name=trace,
-  description={is a sequence of steps, starting from an initial state.  An infinite
-    trace is also called a \emph{behavior}}
+  description={is a sequence of steps, starting from an initial state}
 }
@@
 
+<{:const:p}>
+<{:const:hello}>
+
 \begin{figure}
@@
 \begin{code}
-\harmonysource{hello1}
+<{:input:../code/hello1.hny}>
 \end{code}
@@
 to terminate, so the loop is conditioned on a nondeterministic choice between
-\texttt{False} and \texttt{True}.  The possible outputs consist of zero or
+<{False}> and <{True}>.  The possible outputs consist of zero or
 more copies of the string ``hello world''.  Note that this single state DFA
@@
 \begin{code}
-\harmonysource{hello3}
+<{:input:../code/hello3.hny}>
 \end{code}
@@
 \begin{code}
-\harmonysource{hello4}
+<{:input:../code/hello4.hny}>
 \end{code}
@@
 \begin{tcolorbox}[width=0.4\linewidth]
-\harmonysource{hello5}
+<{:input:../code/hello5.hny}>
 \end{tcolorbox}
@@
 \begin{tcolorbox}[width=0.4\linewidth]
-\harmonysource{hello6}
+<{:input:../code/hello6.hny}>
 \end{tcolorbox}
@@
 \begin{code}
-\harmonysource{hello7}
+<{:input:../code/hello7.hny}>
 \end{code}
@@
 \begin{code}
-\harmonysource{hello8}
+<{:input:../code/hello8.hny}>
 \end{code}
@@
 
-\begin{figure}
-\begin{code}
-\harmonysource{triangle}
+<{:const:triangle}>
+
+\begin{figure}
+\begin{code}
+<{:input:../code/triangle.hny}>
 \end{code}
@@
 an integer number as argument.  Each method has a variable called
-\textit{result} that eventually contains the result of the
+<{result}> that eventually contains the result of the
 method (there is no \textbf{return} statement in Harmony).  The method
-also has a bound variable called \textit{n} containing the value of the
-argument.  The \{ \textit{x..y} \} notation generates a set containing the numbers
-from~\textit{x} to~\textit{y} (inclusive).
+also has a bound variable called <{n}> containing the value of the
+argument.  The <{{ x..y }}> notation generates a set containing the numbers
+from~<{x}> to~<{y}> (inclusive).
 (Harmony does not have iterators and in particular does
@@
 the most interesting.
-The first assigns to \textit{x} some unspecified value in the range \texttt{0..N}
-and the second verifies that $\mathtt{triangle}(x)$ equals $x(x+1)/2$.
+The first assigns to <{x}> some unspecified value in the range \texttt{0..N}
+and the second verifies that <{triangle(x)}> equals $x(x+1)/2$.
 
@@
 \begin{problems}
-\item See what happens if, instead of initializing \textit{result} to 0,
+\item See what happens if, instead of initializing <{result}> to 0,
 you initialize it to 1.  (You do not need to understand the error report at this time.  They will be explained in more detail in \autoref{ch:harmonymachine}.)
@@
 
+<{:const:f}>
+<{:const:g}>
+
 \begin{figure}[h]
@@
 \begin{tcolorbox}[width=0.4\linewidth]
-\harmonysource{prog1}
+<{:input:../code/prog1.hny}>
 \end{tcolorbox}
@@
 \begin{tcolorbox}[width=0.4\linewidth]
-\harmonysource{prog2}
+<{:input:../code/prog2.hny}>
 \end{tcolorbox}
@@
 \autoref{fig:progs} shows two programs.  Program (a) is sequential.
-It sets \textit{shared} to \textbf{True}, asserts that
-\textit{shared} = \textbf{True}, and finally sets
-\textit{shared} to \textbf{False}.
+It sets <{shared}> to <{True}>, asserts that
+<{shared}> = <{True}>, and finally sets
+<{shared}> to <{False}>.
 If you run the program through Harmony, it will not find any problems
@@
 
+<{:const:incrementer}>
+
 \autoref{fig:inc} presents a more subtle example that illustrates
@@
 The program initializes two shared variables:
-an integer \textit{count} and
-an array \textit{done} with two booleans.
+an integer <{count}> and
+an array <{done}> with two booleans.
 The program then spawns two threads.
-The first runs \texttt{incrementer(0)}; the second runs
-\texttt{incrementer(1)}.
-
-Method \texttt{incrementer} takes a parameter called \textit{self}.
-It increments \textit{count} and sets \textit{done}[\textit{self}] to \textbf{True}.
+The first runs <{incrementer(0)}>; the second runs <{incrementer(1)}>.
+
+Method <{incrementer}> takes a parameter called <{self}>.
+It increments <{count}> and sets <{done[self]}> to <{True}>.
 It then waits until the other thread is done.
-(\textbf{await} $c$ is shorthand for \textbf{while not} $c$: \textbf{pass}.)
-After that, method \texttt{incrementer}
-verifies that the value of \textit{count} equals~2.
+(<{await c}> is shorthand for <{while not c: pass}>.)
+After that, method <{incrementer}>
+verifies that the value of <{count}> equals~2.
 % Like Python, Harmony supports \textbf{assert} statements with two arguments:
@@
 % reported if the condition fails, in this case
-% the value of \textit{count}.
+% the value of <{count}>.
 
@@
 they will execute concurrently.  It is, for example, quite possible
-that \texttt{incrementer(1)} finishes before \texttt{incrementer(0)}
+that \texttt{incrementer(1)} finishes before <{incrementer(0)}>
 even gets going.
@@
 that particular execution as well.
-What would the value of \textit{count} be at the end of that execution?
+What would the value of <{count}> be at the end of that execution?
 
@@
 \begin{code}
-\harmonysource{Up}
+<{:input:../code/Up.hny}>
 \end{code}
@@
 \item Before you run the program, what do you think will happen?  Is the
-program correct in that \textit{count} will always end up being 2?
+program correct in that <{count}> will always end up being 2?
 (You may assume that \texttt{load} and \texttt{store} instructions of the
@@
 
-If the threads run one at a time, then \textit{count} will be incremented
+If the threads run one at a time, then <{count}> will be incremented
 twice and ends up being 2.  However, the following is also
-a possible interleaving of \texttt{incrementer(0)} and \texttt{incrementer(1)}:
+a possible interleaving of <{incrementer(0)}> and \texttt{incrementer(1)}:
 \begin{enumerate}
-\item \texttt{incrementer(0)} loads the value of \textit{count}, which is 0;
-\item \texttt{incrementer(1)} loads the value of \textit{count}, which is still 0;
+\item <{incrementer(0)}> loads the value of <{count}>, which is 0;
+\item \texttt{incrementer(1)} loads the value of <{count}>, which is still 0;
 \item \texttt{incrementer(1)} adds 1 to the value that it loaded (0), and
-stores $1$ into \textit{count};
-\item \texttt{incrementer(0)} adds 1 to the value that it loaded (0), and
-stores $1$ into \textit{count};
-\item \texttt{incrementer(0)} sets \textit{done}[0] to \textbf{True};
-\item \texttt{incrementer(1)} sets \textit{done}[1] to \textbf{True}.
+stores $1$ into <{count}>;
+\item <{incrementer(0)}> adds 1 to the value that it loaded (0), and
+stores $1$ into <{count}>;
+\item <{incrementer(0)}> sets <{done[0]}> to <{True}>;
+\item \texttt{incrementer(1)} sets <{done[1]}> to <{True}>.
 \end{enumerate}
 
-The result in this particular interleaving is that \textit{count} ends up
+The result in this particular interleaving is that <{count}> ends up
 being 1.
@@
 \begin{code}
-\textbf{assert} (\textit{count} == 1) or (\textit{count} == 2)
+<{assert (count == 1) or (count == 2)}>
 \end{code}
@@
 \item \autoref{fig:incmany} is a version of \autoref{fig:incpy} that has each
-incrementer thread increment \textit{count} \texttt{N} times.  Run \autoref{fig:incmany}
+incrementer thread increment <{count}> \texttt{N} times.  Run \autoref{fig:incmany}
 10 times (using Python).
-Report how many times the assertion fails and what the value of \textit{count}
+Report how many times the assertion fails and what the value of <{count}>
 was for each of the failed runs.
@@
 Harmony values are recursively defined:
-they include booleans (\textbf{False} and \textbf{True}),
+they include booleans (<{False}> and <{True}>),
 integers (but not floating point numbers),
@@
 $\{ k_0: v_0, ~ k_1: v_1, ~ ... \}$.
-If $d$ is a dictionary, and $k$ is a key, then the
-following expression retrieves the Harmony value that $k$ maps to in $d$:
-\begin{code}
-$d$ $k$
-\end{code}
-The meaning of $d$ $a$ $b$ $...$ is ((($d$ $a$) $b$) $...$).
+If <{d}> is a dictionary, and <{k}> is a key, then the
+following expression retrieves the Harmony value that <{k}> maps to in <{d}>:
+\begin{code}
+<{d k}>
+\end{code}
+The meaning of <{d a b}> $...$ is (<{((d a) b)}> $...$).
 This notation is unfamiliar to Python programmers,
@@
 \begin{code}
-$d$[$k$]
-\end{code}
-However, if $d = \{ \mathtt{.count}: 3 \}$, then you can write
-$d$.\texttt{count} (which has value 3) instead of having to write
-$d$\texttt{[.count]} or $d$\texttt{["count"]} (although any of those
+<{d[k]}>
+\end{code}
+However, if <{d = { .count: 3 }}>, then you can write
+<{d.count}>
+(which has value 3) instead of having to write
+<{d[.count]}> or <{d["count"]}> (although any of those
 will work).
@@
 the indexes into the tuple.  For example, the tuple
-\texttt{(5, False)} is the same Harmony value as
-\texttt{\{ 0:5, 1:False \}}.
-The empty tuple is written as \texttt{()}.
-% Note that this is different from the empty set, which is \texttt{\{\}}.
+<{(5, False)}> is the same Harmony value as
+<{{ 0:5, 1:False }}>.
+The empty tuple is written as <{()}>.
 As in Python, you can create singleton tuples by including a comma.
-For example, \texttt{(1,)} is a tuple consisting just of the number 1.
+For example, <{(1,)}> is a tuple consisting just of the number 1.
 Importantly, $(1) = 1 \ne (1,) = \{ 0:1 \}$.
@@
 Again, square brackets and parentheses work the same in Harmony, so
-\texttt{[a, b, c]} (which looks like a Python list)
-is the same Harmony value as \texttt{(a, b, c)} (which looks like a Python tuple),
-which in turn is the same Harmony value as \texttt{\{ 0:a, 1:b, 2:c \}}.
-So, if $x$ = \texttt{[False, True]},
-then $x$[0] = \textbf{False} and $x$[1] = \textbf{True}, just like in Python.
+<{[a, b, c]}> (which looks like a Python list)
+is the same Harmony value as <{(a, b, c)}> (which looks like a Python tuple),
+which in turn is the same Harmony value as <{{ 0:a, 1:b, 2:c }}>.
+So, if <{x = [False, True]}>,
+then <{x[0] = False}> and <{x[1] = True}>, just like in Python.
 However, when creating a singleton list, make sure you include the
-comma, as in \texttt{[False,]}.  The expression \texttt{[False]} just means
-\textbf{False}.
+comma, as in <{[False,]}>.  The expression <{[False]}> just means
+<{False}>.
 
@@
 For example, dictionaries have two handy unary operators.
-If $d$ is a
-dictionary, then \texttt{keys}~$d$ (or equivalently \texttt{keys}($d$))
-returns the set of keys and \texttt{len}~$d$ returns the size of
+If <{d}> is a
+dictionary, then <{keys d}> (or equivalently <{keys(d)}>)
+returns the set of keys and <{len d}> returns the size of
 this set.
@@
 instruction that sets the program counter to 32
-(skipping over the code for \texttt{incrementer} method).
+(skipping over the code for <{incrementer}> method).
 The \texttt{\_\_init\_\_} thread then executes instructions
@@
 Once initialization completes, any threads that were spawned
-(in this case \texttt{incrementer(0)} and \texttt{incrementer(1)})
+(in this case <{incrementer(0)}> and \texttt{incrementer(1)})
 can run.
@@
 
-At program counter 6 is the code for the \texttt{incrementer} method.
+At program counter 6 is the code for the <{incrementer}> method.
 All methods start with a \texttt{Frame} instruction and end with a
@@
 \item The \texttt{Load} instruction pushes the value of the
-\textit{count} variable onto the stack.
+<{count}> variable onto the stack.
 \item The \texttt{Push} instruction pushes the constant 1
@@
 \item \texttt{2-ary} is a \texttt{+} operation with 2 arguments.
-It pops two values from the stack (the value of \textit{count} and 1),
+It pops two values from the stack (the value of <{count}> and 1),
 adds them, and pushes the result back onto the stack.
 \item The \texttt{Store} instruction pops
-a Harmony value (the sum of the \textit{count} variable and 1) and
-stores it in the \textit{count} variable.
+a Harmony value (the sum of the <{count}> variable and 1) and
+stores it in the <{count}> variable.
 \end{enumerate}
@@
 at least one of the executions of the program that it tried.  This
-can include a failing assertion, divide by zero, using an uninitialized
+can include a failing assertion, behavior violations, divide by zero,
+using an uninitialized
 or non-existent variable, dividing a set by an integer, and so on.
@@
 at that state, along with their program counters.
+\item \texttt{Behavior Violation}:  The program terminated in a state
+not allowed by the behavioral specification.
 \item \texttt{Active Busy Waiting}:  There are states in which
@@
 order: if there are multiple failure conditions, only the first is
-reported.  \emph{Active busy waiting} (\autoref{ch:condwait}) is not
+reported.
+You will learn about behavioral violations in \autoref{ch:testing}.
+\emph{Active busy waiting} (\autoref{ch:condwait}) is not
 technically an indication of a synchronization problem, but instead an
@@
 operations and are therefore generally undesirable.
+Harmony may also warn about behaviors, in particular if the generated
+behavior is only a subset of the provided behavior.
 
@@
 In particular, it reports that the assertion failed
-because \textit{count} has value 1 instead of 2.
+because <{count}> has value 1 instead of 2.
 Next it reports an execution that failed this assertion.
@@
 \item Thread \texttt{\_\_init\_\_} (with identifier T0)
-sets shared variable \textit{count} to 0 and
-shared variable \textit{done} to \texttt{[False, False]}.
-\item Thread \texttt{incrementer(0)} (with identifier T1)
+sets shared variable <{count}> to 0 and
+shared variable <{done}> to \texttt{[False, False]}.
+\item Thread <{incrementer(0)}> (with identifier T1)
 executes instructions 6 through 9, loading the value of
-\textit{count} but stopping just before storing 1 into \textit{count};
+<{count}> but stopping just before storing 1 into <{count}>;
 \item Thread \texttt{incrementer(1)} (with identifier T2)
 executes instructions 6 through 18, storing 1 into
-\textit{count} and storing \textbf{True} into \textit{done}[1];
-\item Thread \texttt{incrementer(0)}
+<{count}> and storing <{True}> into <{done[1]}>;
+\item Thread <{incrementer(0)}>
 continues execution, executing instructions 10 through 28
-storing value 1 into \textit{count} (instruction 10),
-storing \textbf{True} into \textit{done}[0] (instructions 11 through 15)
-finding that \textit{done}[1] is \textbf{True} (instructions 16 through 21),
+storing value 1 into <{count}> (instruction 10),
+storing <{True}> into <{done[0]}> (instructions 11 through 15)
+finding that <{done[1]}> is <{True}> (instructions 16 through 21),
 and finally detecting that the assertion is violated
@@
 occurs when one of the implementers is waiting for the other
-to set its \textit{done} flag;
+to set its <{done}> flag;
 \item[] \texttt{atomic}: the thread is in \emph{atomic} mode,
@@
 A stack trace is a list of methods that are being invoked.
-In this case, the \texttt{incrementer} method does not invoke any
+In this case, the <{incrementer}> method does not invoke any
 other methods, and so the list is of length 1.
@@
 \begin{code}
-\harmonysource{UpEnter}
+<{:input:../code/UpEnter.hny}>
 \end{code}
@@
 
-\begin{figure}
-\begin{code}
-\harmonysource{csbarebones}
+<{:const:thread}>
+<{:const:cs}>
+<{:const:gate}>
+
+\begin{figure}
+\begin{code}
+<{:input:../code/csbarebones.hny}>
 \end{code}
@@
 \begin{code}
-\harmonysource{cs}
+<{:input:../code/cs.hny}>
 \end{code}
@@
 At the heart of the problem is that we would like make sure that, when
-the \textit{count} variable is being updated, no other thread is
+the <{count}> variable is being updated, no other thread is
 trying to do the same thing.  This is called a \emph{critical section}
@@
 \autoref{fig:csbarebones} shows the Harmony code.
-Here \texttt{@cs} is a \emph{label},
+Here \texttt{cs} is a \emph{label},
 \index{label}%
@@
 \index{mutual exclusion}%
-We would like to place an assertion at the \texttt{@cs} label that
+We would like to place an assertion at the \texttt{cs} label that
 specifies that only one (the current) thread can be there.
@@
 \begin{code}
-\texttt{await False}
+<{await False}>
 \end{code}
 This code will surely prevent two or more threads from being
-at label \texttt{@cs} at the same time.
+at label \texttt{cs} at the same time.
 But it does so by preventing \emph{any} thread from reaching
@@
 \begin{code}
-\harmonysource{naiveLock}
+<{:input:../code/naiveLock.hny}>
 \end{code}
@@
 \begin{code}
-\harmonysource{naiveFlags}
+<{:input:../code/naiveFlags.hny}>
 \end{code}
@@
 \begin{code}
-\harmonysource{naiveTurn}
+<{:input:../code/naiveTurn.hny}>
 \end{code}
@@
 na\"{\i}ve (and, as it turns out, broken) implementation of a lock.
-Initially the lock is not owned, indicated by \textit{lockTaken} being \textbf{False}.
-To enter the critical section, a thread waits until \textit{lockTaken} is \textbf{False}
-and then sets it to \textbf{True} to indicate that the lock has been taken.
+Initially the lock is not owned, indicated by \textit{lockTaken} being <{False}>.
+To enter the critical section, a thread waits until \textit{lockTaken} is <{False}>
+and then sets it to <{True}> to indicate that the lock has been taken.
 The thread then executes the critical section.  Finally the thread
-releases the lock by setting \textit{lockTaken} back to \textbf{False}.
+releases the lock by setting \textit{lockTaken} back to <{False}>.
 
@@
 Thread~0 finds that the lock is available, but just before it stores
-\textbf{True} in \textit{lockTaken} in instruction~9, thread~1 gets
+<{True}> in \textit{lockTaken} in instruction~9, thread~1 gets
 to run.
@@
 to see what it does.)
-Because \textit{lockTaken} is still \textbf{False}, it too
-believes it can acquire the lock, and stores \textbf{True} in
+Because \textit{lockTaken} is still <{False}>, it too
+believes it can acquire the lock, and stores <{True}> in
 \textit{lockTaken} and moves on to the critical section.
-Finally, thread~0 moves on, also stores \textbf{True}
+Finally, thread~0 moves on, also stores <{True}>
 into \textit{lockTaken} and also moves
@@
 The \textit{lockTaken} variable
-suffers from the same sort of race condition as the \textit{count} variable
+suffers from the same sort of race condition as the <{count}> variable
 in \autoref{fig:inc}: testing and setting the lock
@@
 
-To see why, first note the following invariant: if thread $i$ is in the
-critical section, then \textit{flags}[$i$] = \textbf{True}.
+To see why, first note the following invariant: if thread <{i}> is in the
+critical section, then \textit{flags}[<{i}>] = <{True}>.
 Without loss of generality,
@@
 Thread 0 can only reach the critical section if at some time $t_1$,
-$t_1 > t_0$, it finds that \textit{flags}[1] = \textbf{False}.
-Because of the invariant, \textit{flags}[1] = \textbf{False} implies that
+$t_1 > t_0$, it finds that \textit{flags}[1] = <{False}>.
+Because of the invariant, \textit{flags}[1] = <{False}> implies that
 thread~1 is not in the critical section at time $t_1$.
 Let $t_2$ be the time at which thread~0 sets \textit{flags}[0]
-to \textbf{False}.  Thread~0 is in the critical section sometime
+to <{False}>.  Thread~0 is in the critical section sometime
 between $t_1$ and $t_2$.
 It is easy to see that thread~1 cannot enter the critical section
-between $t_1$ and $t_2$, because \textit{flags}[1] = \textbf{False} at
+between $t_1$ and $t_2$, because \textit{flags}[1] = <{False}> at
 time $t_1$.  To reach the critical section between $t_1$ and $t_2$,
-it would first have to set \textit{flags}[1] to \textbf{True} and
-then wait until \textit{flags}[0] = \textbf{False}.  But that does not happen
+it would first have to set \textit{flags}[1] to <{True}> and
+then wait until \textit{flags}[0] = <{False}>.  But that does not happen
 until time $t_2$.
@@
 Each thread politely lets the other thread have a turn first.
-When \textit{turn} = $i$, thread~$i$ can
+When \textit{turn} = <{i}>, thread~<{i}> can
 enter the critical section, while thread $1-i$ has to wait.
-An invariant of this solution is that while thread~$i$ is in the critical
+An invariant of this solution is that while thread~<{i}> is in the critical
 section, $\mathit{turn} = i$.
@@
 \emph{progress}, albeit for a different reason:
-if thread~$i$ terminates instead of entering the critical section,
+if thread~<{i}> terminates instead of entering the critical section,
 thread $1-i$, politely, ends up waiting indefinitely for its turn.
@@
 \begin{code}
-\texttt{await False}
+<{await False}>
 \end{code}
@@
 \begin{code}
-\harmonysource{Peterson}
+<{:input:../code/Peterson.hny}>
 \end{code}
@@
 Unfortunately, exclusivity is not an \emph{inductive} invariant.
-To see why, consider the following state $s$: let thread~0 be at label \texttt{@cs}
+To see why, consider the following state $s$: let thread~0 be at label \texttt{cs}
 and thread~1 be at the start of the \textbf{await} statement.
@@
 $\mathcal{F}(i) = \mathtt{thread}(i)@[8 \cdots 15] \Rightarrow \mathit{flags}[i]$,
-that is, if thread~$i$ is executing in lines 8 through 15, then $\mathit{flags}[i]$
+that is, if thread~<{i}> is executing in lines 8 through 15, then $\mathit{flags}[i]$
 is set.
@@
 $\mathcal{F}(i)$ is an inductive invariant (for both threads~0 and~1).
-To wit, it holds in the initial state, because in the initial state thread $i$ does
+To wit, it holds in the initial state, because in the initial state thread <{i}> does
 not even exist yet.
@@
 $\mathcal{F}(i)$ also holds in a next state.
-Since only thread~$i$ ever changes $\mathit{flags}[i]$, we only need to
-consider steps by thread~$i$.
+Since only thread~<{i}> ever changes $\mathit{flags}[i]$, we only need to
+consider steps by thread~<{i}>.
 Since $\mathcal{F}(i)$ holds, there are two cases to consider:
@@
 \end{enumerate}
-In each case we need to show that if thread $i$ takes a step, 
+In each case we need to show that if thread <{i}> takes a step, 
 $\mathcal{F}(i)$ still holds.
-In the first case, there is only one step that thread $i$ can take that would
+In the first case, there is only one step that thread <{i}> can take that would
 set $\mathit{flags}[i]$ to false: the step in line 15.  But executing the line
 would also take the thread out of lines $8 \cdots 15$, so $\mathcal{F}(i)$ continues to hold.
-In the second case (thread $i$ is not executing in lines $8 \cdots 15$), the only step
-that would cause thread $i$ to execute in lines $8 \cdots 15$ would be the step in line 7.
+In the second case (thread <{i}> is not executing in lines $8 \cdots 15$), the only step
+that would cause thread <{i}> to execute in lines $8 \cdots 15$ would be the step in line 7.
 But in that case $\mathit{flags}[i]$ would end up being true, so
@@
 While $\mathcal{F}(i)$ does not imply mutual exclusion, it does imply the following useful
-invariant: $\mathtt{thread}(i)@cs \Rightarrow \mathit{flags}[i]$: when thread $i$ is
+invariant: $\mathtt{thread}(i)@cs \Rightarrow \mathit{flags}[i]$: when thread <{i}> is
 at the critical section, $\mathit{flags}[i]$ is set.  This seems obvious from the code,
@@
 We need a stronger inductive invariant than $\mathcal{F}(i)$ to prove mutual exclusion.
-What else do we know when thread $i$ is in the critical section?
+What else do we know when thread <{i}> is in the critical section?
 Let $\mathcal{C}(i) = \lnot\mathit{flags}[1 - i] \lor \mathit{turn} = i$, that is,
-the condition on the \textbf{await} statement for thread~$i$.
-In a sequential program, $\mathcal{C}(i)$ would clearly hold if thread $i$ is in
+the condition on the \textbf{await} statement for thread~<{i}>.
+In a sequential program, $\mathcal{C}(i)$ would clearly hold if thread <{i}> is in
 the critical section: $\mathtt{thread}(i)@cs \Rightarrow \mathcal{C}(i)$.
@@
 \begin{code}
-\harmonysource{PetersonInductive}
+<{:input:../code/PetersonInductive.hny}>
 \end{code}
@@
 \mathtt{thread}(i)@\mathtt{cs} \Rightarrow \mathcal{C}(i) \lor \mathtt{thread}(1-i)@\mathtt{gate}$.
-That is, if thread $i$ is at the critical section, then
+That is, if thread <{i}> is at the critical section, then
 either $\mathcal{C}(i)$ holds or thread $1-i$ is about to execute line 8.
 \autoref{fig:petersonproof} formalizes $\mathcal{G}(i)$ in Harmony.
-The label \texttt{@gate} refers to the step that sets \textit{turn} to $1-i$.
+The label \texttt{gate} refers to the step that sets \textit{turn} to $1-i$.
 You can run \autoref{fig:petersonproof} to determine
@@
 \textit{flags} are true.  By $\mathcal{G}$ and the fact that neither thread
-is at label \texttt{@gate}, we know that both $C(0)$ and $C(1)$ must hold.
+is at label \texttt{gate}, we know that both $C(0)$ and $C(1)$ must hold.
 This then implies that $\mathit{turn} = 0 \land \mathit{turn} = 1$, providing
@@
 
-First consider the case that thread~0 is at label \texttt{@cs}.  If thread~0
+First consider the case that thread~0 is at label \texttt{cs}.  If thread~0
 were to take a step, then in the next state thread~0 would be no longer
@@
 \begin{enumerate}
-\item \textit{flags}[1] = \textbf{False};
+\item \textit{flags}[1] = <{False}>;
 \item \textit{turn} = 0;
-\item thread~1 is at label \texttt{@gate}.
+\item thread~1 is at label \texttt{gate}.
 \end{enumerate}
@@
 In the first case, if thread~1 takes a step, there are two possibilities:
-either $flags[1]$ will still be \textbf{False} (in which case the first case
-continues to hold), or $flags[1]$ will be \textbf{True}
-and thread~1 will be at label \texttt{@gate} (in which case the third case
+either $flags[1]$ will still be <{False}> (in which case the first case
+continues to hold), or $flags[1]$ will be <{True}>
+and thread~1 will be at label \texttt{gate} (in which case the third case
 will hold).
@@
 if the second case holds before the step, it will also hold after the step.
-Finally, if thread~1 is at label \texttt{@gate} before the step, then after
+Finally, if thread~1 is at label \texttt{gate} before the step, then after
 the step \textit{turn} will equal 0, and therefore the second case will hold
@@
 Because thread~0 entered the critical section, we know that
-$\mathit{flags}[1] = \textbf{False}$ or $\mathit{turn} = 0$ because
+$\mathit{flags}[1] = <{False}>$ or $\mathit{turn} = 0$ because
 of the \texttt{await} condition.
@@
 page.  The page claimed that the following predicate is invariant:
-if thread $i$ is in the critical section, then $\mathcal{C}(i)$ (i.e.,
-$\mathcal{G}$ without the disjunct that thread $1-i$ is at label \texttt{@gate}).
+if thread <{i}> is in the critical section, then $\mathcal{C}(i)$ (i.e.,
+$\mathcal{G}$ without the disjunct that thread $1-i$ is at label \texttt{gate}).
 To demonstrate that this predicate is not invariant, you can remove the
@@
 Peterson's algorithm (setting \textit{flags}[\textit{self}])
-to \textbf{True} and \textit{turn} to $1 - \mathit{self}$) can be reversed.
+to <{True}> and \textit{turn} to $1 - \mathit{self}$) can be reversed.
 After all, they are different variables assigned independent values---in a
@@
 \begin{code}
-\harmonysource{csonebit}
+<{:input:../code/csonebit.hny}>
 \end{code}
@@
 \begin{code}
-$r$ = m $a$;
+<{r = m a}>
 \end{code}
@@
 
-\begin{figure}
-\begin{code}
-\harmonysource{PetersonMethod}
+<{:const:P_enter}>
+<{:const:P_exit}>
+<{:const:P_mutex}>
+
+\begin{figure}
+\begin{code}
+<{:input:../code/PetersonMethod.hny}>
 \end{code}
@@
 Harmony does not have a \textbf{return} statement.  (You can assign a return value
-to a method by setting the \textit{result} variable.)  Neither does it support
+to a method by setting the <{result}> variable.)  Neither does it support
 \textbf{break} or \textbf{continue} statements in loops.  One reason for their absence is
@@
 variables) within a \texttt{for} statement are immutable;
-the variable \textit{result} is mutable.
-Using the \texttt{var} statement, new mutable local variables can be
+the variable <{result}> is mutable.
+Using the <{var}> statement, new mutable local variables can be
 declared.
-For example, \texttt{var} $x = 3$ declares a new mutable local variable
+For example, <{var}> $x = 3$ declares a new mutable local variable
 $x$.
@@
 
-\begin{figure}
-\begin{code}
-\harmonysource{clock}
+<{:const:ref}>
+
+\begin{figure}
+\begin{code}
+<{:input:../code/clock.hny}>
 \end{code}
@@
 The program in \autoref{fig:clock} will find exactly the same scenario
-if you define \texttt{FIFO} to be \textbf{True}.  Moreover, if you define
-\texttt{FIFO} to be \textbf{False}, it will find a scenario for the
+if you define \texttt{FIFO} to be <{True}>.  Moreover, if you define
+\texttt{FIFO} to be <{False}>, it will find a scenario for the
 CLOCK replacement policy~\cite{clock}, often used in modern operating systems.
@@
 for a particular reference was recent or not.
-(It is not used if \texttt{FIFO} is \textbf{True}.)
+(It is not used if \texttt{FIFO} is <{True}>.)
 The integer \texttt{misses} maintains the number of cache misses.
@@
 \begin{code}
-\harmonysource{lockspec}
+<{:input:../code/lockspec.hny}>
 \end{code}
@@
 \begin{code}
-\harmonysource{cssynch}
+<{:input:../code/cssynch.hny}>
 \end{code}
@@
 \begin{code}
-\harmonysource{UpLock}
+<{:input:../code/UpLock.hny}>
 \end{code}
@@
 of a lock.
-In this case, a lock is a boolean, initially \texttt{False},
+In this case, a lock is a boolean, initially <{False}>,
 with two operations, \texttt{acquire()} and \texttt{release()}.
-The \texttt{acquire()} operation waits until the lock is \texttt{False}
-and then sets it to \texttt{True} in an atomic operation.
-The \texttt{release()} operation sets the lock back to \texttt{False}.
+The \texttt{acquire()} operation waits until the lock is <{False}>
+and then sets it to <{True}> in an atomic operation.
+The \texttt{release()} operation sets the lock back to <{False}>.
 The code is similar to \autoref{fig:naivelock}, except that waiting
@@
 
+<{:const:test_and_set}>
+<{:const:clear}>
+<{:const:atomic_inc}>
+<{:const:fetch_and_increment}>
+
 \glsadd{spinlock}%
@@
 \begin{code}
-\harmonysource{spinlock}
+<{:input:../code/spinlock.hny}>
 \end{code}
@@
 \begin{code}
-\harmonysource{atomicinc}
+<{:input:../code/atomicinc.hny}>
 \end{code}
@@
 private variable (the ``test'')
-and then sets the shared variable to \textbf{True} (``set'').
+and then sets the shared variable to <{True}> (``set'').
 
@@
 The program uses $\mathtt{N}+1$ variables.
-Variable \textit{shared} is initialized to \textbf{False} while
-\textit{private}[$i$] for each thread $i$ is initialized to \textbf{True}.
+Variable <{shared}> is initialized to <{False}> while
+\textit{private}[<{i}>] for each thread <{i}> is initialized to <{True}>.
 
 An important invariant, $\mathcal{I}_1$, of the program is that at any time at most
-one of these variables is \textbf{False}.
-Another invariant, $\mathcal{I}_2(i)$, is that if thread $i$ is executing between
+one of these variables is <{False}>.
+Another invariant, $\mathcal{I}_2(i)$, is that if thread <{i}> is executing between
 lines $18 \cdots 22$ (which includes the critical section),
-then \textit{private}[$i$] = \textbf{False}.
+then \textit{private}[<{i}>] = <{False}>.
 Between the two (i.e., $\mathcal{I}_1 \land \forall i: \mathcal{I}_2(i)$),
@@
 To see that invariant $\mathcal{I}_1$ is maintained, note that
-\texttt{!}$p$ = \textbf{True} upon entry of \texttt{test\_and\_set}
+\texttt{!}$p$ = <{True}> upon entry of \texttt{test\_and\_set}
 (because of the condition on the \textbf{while} loop that the
@@
 \begin{enumerate}
-\item \texttt{!}$s$ is \textbf{False} upon entry to \texttt{test\_and\_set}.
-Then upon exit \texttt{!}$p$ = \textbf{False} and \texttt{!}$s$ = \textbf{True},
+\item \texttt{!}$s$ is <{False}> upon entry to \texttt{test\_and\_set}.
+Then upon exit \texttt{!}$p$ = <{False}> and \texttt{!}$s$ = <{True}>,
 maintaining the invariant.
-\item \texttt{!}$s$ is \textbf{True} upon entry to \texttt{test\_and\_set}.
+\item \texttt{!}$s$ is <{True}> upon entry to \texttt{test\_and\_set}.
 Then upon exit nothing has changed, maintaining the invariant.
@@
 because we can assume, by the induction hypothesis, that $\textit{private}[i]$ is
-\textbf{True} just before exiting the critical section.
-Invariant $\mathcal{I}_2(i)$ is obvious as (i) thread $i$ only proceeds to the critical
-section if \textit{private}[$i$] is \textbf{False}, and (ii) no other thread modifies
-\textit{private}[$i$].
+<{True}> just before exiting the critical section.
+Invariant $\mathcal{I}_2(i)$ is obvious as (i) thread <{i}> only proceeds to the critical
+section if \textit{private}[<{i}>] is <{False}>, and (ii) no other thread modifies
+\textit{private}[<{i}>].
 
@@
 invariant like $\mathcal{I}_1$?  Invariants must hold for every state.
-For $\mathcal{I}_2$ we only need an assertion at label \texttt{@cs} because the
+For $\mathcal{I}_2$ we only need an assertion at label \texttt{cs} because the
 premise is that there is a thread at that label.  However, we would
@@
 keyword.
-The expression counts the number of \textbf{False} values and checks that
+The expression counts the number of <{False}> values and checks that
 the result is less than or equal to 1.
@@
 \begin{code}
-\harmonysource{taslock}
+<{:input:../code/taslock.hny}>
 \end{code}
@@
 \begin{code}
-\harmonysource{ticket}
+<{:input:../code/ticket.hny}>
 \end{code}
@@
 \begin{code}
-\harmonysource{locksusp}
+<{:input:../code/locksusp.hny}>
 \end{code}
@@
 \begin{code}
-\textbf{stop} $s$
+<{stop s}>
 \end{code}
@@
 \begin{code}
-\textbf{go} $C$ $R$;
-\end{code}
-
-Here $C$ is a context and $R$ is a Harmony value.
-It causes a thread with context $C$ to be added to the state that has
+<{go c r}>
+\end{code}
+
+Here <{c}> is a context and <{r}> is a Harmony value.
+It causes a thread with context <{c}> to be added to the state that has
 just executed the \textbf{stop}
 \index{stop expression}%
-expression.  The \textbf{stop} expression returns the value $R$.
+expression.  The \textbf{stop} expression returns the value <{r}>.
 
@@
 This interface is like \texttt{acquire}(\textit{b}) but never blocks.  It
-returns \textbf{True} if the binary semaphore was available (and now acquired) or \textbf{False}
+returns <{True}> if the binary semaphore was available (and now acquired) or <{False}>
 if the binary semaphore was already acquired.
@@
 
-\begin{figure}
-\begin{code}
-\harmonysource{xy}
+<{:const:setX}>
+<{:const:getXY}>
+<{:const:checker}>
+
+\begin{figure}
+\begin{code}
+<{:input:../code/xy.hny}>
 \end{code}
@@
 
-\begin{figure}
-\begin{code}
-\harmonysource{atm}
+<{:const:atm_check_balance}>
+<{:const:atm_withdraw}>
+<{:const:customer}>
+
+\begin{figure}
+\begin{code}
+<{:input:../code/atm.hny}>
 \end{code}
@@
 \label{ch:cds}
+
+<{:const:sender}>
+<{:const:receiver}>
 
@@
 \begin{tabular}{cc}
-\begin{tcolorbox}[width=0.41\linewidth]
-\harmonysource{queuespec}
+\begin{tcolorbox}[width=0.42\linewidth]
+<{:input:../code/queuespec.hny}>
 \end{tcolorbox}
 &
-\begin{tcolorbox}[width=0.52\linewidth]
-\harmonysource{queue}
+\begin{tcolorbox}[width=0.53\linewidth]
+<{:input:../code/queue.hny}>
 \end{tcolorbox}
@@
 \begin{code}
-\harmonysource{queuedemo}
+<{:input:../code/queuedemo.hny}>
 \end{code}
@@
 and a \texttt{tail} pointer to the last element in the list.
-The \texttt{head} pointer is \texttt{None} if and only if the queue is empty.
-(\texttt{None} is a special address value that is not the address of any
+The \texttt{head} pointer is <{None}> if and only if the queue is empty.
+(<{None}> is a special address value that is not the address of any
 memory location.)
 
-\texttt{Queue()} returns a new queue object consisting of a \texttt{None} head
+\texttt{Queue()} returns a new queue object consisting of a <{None}> head
 and tail pointer and a lock.
@@
 \begin{code}
-\textbf{assert} \texttt{queue.get}($q$) in \{ \textbf{None}, 1, 2 \}
+<{assert queue.get(q) in { None, 1, 2 }}>
 \end{code}
@@
 Note that neither the \texttt{head} nor \texttt{tail} pointer are
-ever \texttt{None}.
+ever <{None}>.
 The problem is when the queue is empty and there are concurrent
@@
 \begin{code}
-\harmonysource{queueconc}
+<{:input:../code/queueconc.hny}>
 \end{code}
@@
 \begin{code}
-\harmonysource{queueMS}
+<{:input:../code/queueMS.hny}>
 \end{code}
@@
 \begin{itemize}
-\item If the list is empty, $q$$\rightarrow$$\mathtt{tail}$ should be \texttt{None}.
+\item If the list is empty, $q$$\rightarrow$$\mathtt{tail}$ should be <{None}>.
 Otherwise, the last element in the linked list starting from $q$$\rightarrow$$\mathtt{head}$
 should equal $q$$\rightarrow$$\mathtt{head}$. Moreover,
-$q$$\rightarrow$$\texttt{tail}$$\rightarrow$$next$ should be \texttt{None};
+$q$$\rightarrow$$\texttt{tail}$$\rightarrow$$next$ should be <{None}>;
 \item The length field that you added in \autoref{ex:qlength} should equal the length
@@
 just before releasing the lock in \texttt{put} and \texttt{get}:
-\begin{code}
-\textbf{assert} \texttt{check()}
+<{:const:check}>
+\begin{code}
+<{assert check()}>
 \end{code}
@@
 
-\begin{figure}
-\begin{code}
-\harmonysource{setobj}
+<{:module:setobj}>
+<{:const:SetObject}>
+<{:const:insert}>
+<{:const:remove}>
+<{:const:contains}>
+<{:const:_find}>
+<{:const:_node}>
+<{:const:thread1}>
+<{:const:thread2}>
+
+\begin{figure}
+\begin{code}
+<{:input:../code/setobj.hny}>
 \end{code}
@@
 \begin{code}
-\harmonysource{linkedlist}
+<{:input:../code/linkedlist.hny}>
 \end{code}
@@
 \begin{code}
-\harmonysource{intsettest}
+<{:input:../code/intsettest.hny}>
 \end{code}
@@
 Each node has a lock, a value, and \textit{next}, a pointer to the next node
-(which is \texttt{None} for the final \texttt{inf} node).
+(which is <{None}> for the final \texttt{inf} node).
 The \texttt{\_find(\textit{lst}, $v$)} helper method first
@@
 \label{ch:testing}
+<{:module:queueconc}>
 
@@
 \begin{code}
-\harmonysource{cslock}
+<{:input:../code/cslock.hny}>
 \end{code}
@@
 \begin{code}
-\harmonysource{qtestseq}
+<{:input:../code/qtestseq.hny}>
 \end{code}
@@
 \begin{code}
-\harmonysource{qtestpar}
+<{:input:../code/qtestpar.hny}>
 \end{code}
@@
 \begin{code}
-\harmonysource{queueseq}
+<{:input:../code/queueseq.hny}>
 \end{code}
@@
 \begin{code}
-\harmonysource{qtest1}
+<{:input:../code/qtest1.hny}>
 \end{code}
@@
 \begin{code}
-\harmonysource{qtest2}
+<{:input:../code/qtest2.hny}>
 \end{code}
@@
 \begin{code}
-\harmonysource{qtest3}
+<{:input:../code/qtest3.hny}>
 \end{code}
@@
 \begin{code}
-\harmonysource{qtest4}
+<{:input:../code/qtest4.hny}>
 \end{code}
@@
 \begin{code}
-\harmonysource{multitest}
+<{:input:../code/multitest.hny}>
 \end{code}
@@
 \begin{code}
-\harmonysource{queuebroken}
+<{:input:../code/queuebroken.hny}>
 \end{code}
@@
 % Looking in the top green row, we see that the value of $q$ is
-% \{ \texttt{.lock}: \texttt{True}, \texttt{.next}: \texttt{None}, \texttt{.value}: \texttt{None} \}.
+% \{ \texttt{.lock}: <{True}>, \texttt{.next}: <{None}>, \texttt{.value}: <{None}> \}.
 T1 is currently executing
@@
 \item[0:] initialization completed, with $q$ being
-\{ \texttt{.lock}: \texttt{False}, \texttt{.next}: \texttt{None}, \texttt{.value}: \texttt{None} \};
+\{ \texttt{.lock}: <{False}>, \texttt{.next}: <{None}>, \texttt{.value}: <{None}> \};
 \item[1:] thread T1 (\texttt{thread}(1)) ran and finished executing
@@
 which the thread must have allocated.  The contents is 
-\{ \texttt{.lock}: \texttt{False}, \texttt{.next}: \texttt{None}, \texttt{.value}: 1 \}, as expected;
+\{ \texttt{.lock}: <{False}>, \texttt{.next}: <{None}>, \texttt{.value}: 1 \}, as expected;
 \item[2:] thread T2 (\texttt{thread}(1)) started running, calling
@@
 next T2 executes \textbf{let} $n = \mathit{nq}$$\rightarrow$$\mathtt{next}$.
-Now $n =\,?\mathtt{alloc}.\mathit{alloc}[0]$, which is not \texttt{None}.
+Now $n =\,?\mathtt{alloc}.\mathit{alloc}[0]$, which is not <{None}>.
 It then releases $\mathit{nq}$$\rightarrow$$\mathtt{lock}$
@@
 $\mathtt{node}$$\rightarrow$$\texttt{next}$.
-T3 sets \textit{result} to 
+T3 sets <{result}> to 
 $\mathtt{node}$$\rightarrow$$\texttt{value}$ (which is 1)
@@
 \begin{code}
-\harmonysource{queuefix}
+<{:input:../code/queuefix.hny}>
 \end{code}
@@
 \label{ch:condwait}
+
+<{:module:RW}>
+<{:const:RWlock}>
+<{:const:read_acquire}>
+<{:const:read_release}>
+<{:const:write_acquire}>
+<{:const:write_release}>
+<{:const:BoundedBuffer}>
 
@@
 \begin{code}
-\harmonysource{RW}
+<{:input:../code/RW.hny}>
 \end{code}
@@
 \begin{code}
-\harmonysource{RWtest}
+<{:input:../code/RWtest.hny}>
 \end{code}
@@
 \begin{code}
-\harmonysource{RWbusy}
+<{:input:../code/RWbusy.hny}>
 \end{code}
@@
 
-\begin{comment}
-We will explore various ways of implementing reader/writer locks in
-this chapter and future ones.
-
-\autoref{fig:rwbusy} presents a solution that uses a
-single (ordinary) lock and two counters: one that maintains the number
-of readers and one that maintains the number of writers.
-\autoref{fig:rwtest} shows how the code may be used.
-We will call the lock the \emph{mutex} to distinguish it clearly from
-the reader/writer lock that we are implementing.
-The mutex is used to protect shared access to the counters.
-The program shows a thread that executes in a loop.
-Each time, it decides whether to read or write.
-The critical section is spread between two labels:
-readers access \texttt{@rcs} and writers access \texttt{@wcs}.
-The specification is that if a reader is at label \texttt{@rcs},
-no writer is allowed to be at label \texttt{@wcs}.  Vice versa, if
-a writer is at label \texttt{@wcs}, no reader is allowed to be at
-label \texttt{@rcs} \emph{and} there cannot be another writer at
-label \texttt{@wcs}.
-
-\autoref{fig:rdwr} shows a high-level
-specification for two threads.
-
-A thread that wants to read first waits until there are no writers:
-\textit{nwriters} = 0.  Then it increments the number of readers.
-Similarly, a
-thread that wants to write waits until there are no readers \emph{or} writers.
-Then it increments the number of writers.
-The important invariants in this code are:
-\begin{itemize}
-\item $n$ readers at \texttt{@rcs} $\Rightarrow \mathit{nreaders} \ge n$,
-\item $n$ writers at \texttt{@wcs} $\Rightarrow \mathit{nwriters} \ge n$,
-\item $(\mathit{nreaders} \ge 0 \land \mathit{nwriters} = 0) \lor
-    (\mathit{nreaders} = 0 \land 0 \le \mathit{nwriters} \le 1)$.
-\end{itemize}
-It is easy to see that the invariants hold and imply the reader/writer
-specification.
-The solution also supports progress: if no thread is in the critical
-section then any thread can enter.  Better still: if any reader is in the
-critical section, any other reader is also able to enter.
-
-\begin{figure}
-\begin{code}
-\harmonysource{RWbusychk}
-\end{code}
-\caption{\harmonylink{code/RWbusychk.hny} Checking for Busy Waiting}
-\label{fig:rwblock}
-\end{figure}
-
-\glsadd{busy waiting}%
-
-While correct, it is not considered a good solution.
-The solution is an example of what is called \emph{busy-waiting}
-\index{busy waiting}%
-(aka \emph{spin-waiting}):
-\index{spin waiting}%
-threads spin in a loop until some desirable application-level condition is met,
-wasting CPU cycles.
-\autoref{fig:inc} also has an example of busy-waiting: the \texttt{main}
-thread waits for the other two threads to finish by checking their \texttt{done}
-flags.
-
-The astute reader might wonder if obtaining the mutex itself is an
-example of busy-waiting.  After all, the Harmony \texttt{synch} implementation
-of \texttt{lock()} spins in a loop until the mutex is available
-(see \autoref{fig:spinlocks}).
-However, there is a big difference.  As we pointed out, the threads that
-are waiting for a spinlock are \emph{blocked}.
-In most operating systems and programming language runtimes,
-when a thread acquires a mutex, the thread is placed on a scheduling
-queue and stops using CPU cycles until the mutex becomes available.
-Harmony supports this with the \texttt{synchS} module.
-Busy waiting disables all this: even when using the \texttt{synchS}
-module, readers and writers only get suspended temporarily in case
-there is contention for the mutex.
-A thread trying to obtain a read or write lock in
-\autoref{fig:rwbusy} tries to obtain the mutex repeatedly.
-
-Thus, it is considered ok to have threads be blocked while waiting
-for application-specific conditions, but not for them to
-busy-wait for application-specific conditions.
-
-Harmony does not complain about \autoref{fig:rwbusy}, but, using
-the right test, Harmony can be used to check for busy-waiting.
-\autoref{fig:rwblock} presents such a test.
-Note that the initialization code obtains a write lock, preventing
-all readers and writers from entering the critical section.
-In that case, those threads should ideally be blocked and Harmony
-can test for that by running it with the \texttt{-b} flag.
-This flag tells Harmony to check
-that all states are non-terminating and lead to a state in which
-all threads are blocked.
-
-\section*{Exercises}
-\begin{problems}
-\item Draw additional states and steps in \autoref{fig:rdwr}
-for three threads.
-\end{problems}
-
-\chapter{Reader/Writer Locks with Blocking}
-\label{ch:rdwr}
-
-\begin{figure}
-\begin{code}
-\harmonysource{RWlock}
-\end{code}
-\caption{\harmonylink{code/RWlock.hny} Reader/Writer with Two Locks}
-\label{fig:rw2lock}
-\end{figure}
-
-\begin{figure}
-\begin{code}
-\harmonysource{RWmulti}
-\end{code}
-\caption{\harmonylink{code/RWmulti.hny} Checking that multiple readers can acquire the read lock}
-\label{fig:rwmulti}
-\end{figure}
-
-\autoref{fig:rw2lock} presents a reader/writer lock implementation
-that does not busy-wait.
-You can test the code by running \texttt{harmony -m RW=RWlock RWtest.hny}.
-It uses two ordinary locks: \textit{rwlock} is used by both readers and writers,
-while \textit{rlock} is only used by readers.
-\textit{rwlock} is held either when readers are at label \texttt{@rcs}
-or when a writer is at label \texttt{@wcs}.
-\textit{rlock} is used to protect the \textit{nreaders} variable that
-counts the number of readers in the critical section.
-
-The invariants that imply the reader/writer specification
-(which are again easy to verify) are as follows:
-
-\begin{itemize}
-\item $n$ readers at \texttt{@rcs} $\Rightarrow \mathit{nreaders} \ge n$
-\item $\exists$ writer at \texttt{@wcs} $\Rightarrow \mathit{nreaders} = 0$ and writer holds \textit{rwlock}
-\item at most one writer can hold \textit{rwlock}
-\end{itemize}
-
-A writer just acquires \textit{rwlock} to enter the critical section
-and releases it to exit.  The \emph{first} reader to enter the critical
-section acquires \textit{rwlock} and the \emph{last} reader to exit
-the critical section releases \textit{rwlock}.
-The implementation satisfies progress: if no thread is in the critical
-section then any thread can enter.
-
-It is instructive to see what happens when a writer is in the critical
-section and two readers try to enter.  The first reader successfully
-acquires \textit{rlock} but blocks when trying to acquire
-\textit{rwlock}, which is held by the writer.  The second reader blocks
-trying acquire \textit{rlock} because it is held by the first reader.
-When the writer leaves the critical section, the first reader acquires
-\textit{rwlock}, sets \textit{nreaders} to 1, and releases \textit{rlock}.
-\textit{rwlock} is still held.
-Then the second reader acquires \textit{rlock} and, assuming the first
-reader is still in the critical section, increments \textit{nreaders} to 2
-and enters the critical section \emph{without} acquiring \textit{rwlock}.
-\textit{rwlock} is essentially jointly held by both readers.
-It does not matter in which order they leave: the second will release
-\textit{rwlock}.
-
-While Harmony does not detect any issues, we must remember what it is checking
-for: it checks to make sure that there are never a reader and a writer
-in the critical section, and there are never two writers in the critical
-section.  Moreover, it checks progress: all executions eventually terminate.
-What it does \emph{not} check is if the code allows more than one reader
-in the critical section.  Indeed, we could have implemented the reader/writer
-lock using an ordinary lock, never allowing
-more than one thread in the critical section, and Harmony would not have
-complained.
-
-\autoref{fig:rwmulti} contains a new test that ensures that indeed more
-than one reader can enter the critical section at a time.  It does so by
-having threads sometimes wait in the critical section until all other readers
-are there.  If other readers cannot enter, then that thread enters in an
-infinite loop that Harmony will readily detect.  Try it out!
-
-An important lesson here is that one should not celebrate too early if Harmony
-does not find any issues.  While a test may be successful, a test may not
-explore all desirable executions.  While Harmony can help explore all cornercases
-in a test, it cannot find problems that the test is not looking for.
-
-\section*{Exercises}
-\begin{problems}
-\item Why do you suppose the code in \autoref{fig:rwmulti} uses a flag
-per thread rather than a simple counter?
-Can you write a version of this test that uses a counter instead
-of a flag per thread?
-% \item Write a ``library implementation'' of reader/writer locks so multiple
-% reader/writer locks can be instantiated.
-\end{problems}
-
-\chapter{Semaphores}
-\label{ch:semaphore}
-\index{semaphore}%
-
-\glsadd{semaphore}%
-
-\begin{figure}
-\begin{code}
-\harmonysource{semaphore}
-\end{code}
-\caption{\harmonylink{modules/synch.hny} The \texttt{Semaphore} interface in the \texttt{synch} module}
-\label{fig:semaphore}
-\end{figure}
-
-\begin{figure}
-\begin{code}
-\harmonysource{DinersSema}
-\end{code}
-\caption{\harmonylink{code/DinersSema.hny} Italian Philosophers}
-\label{fig:dinerssema}
-\end{figure}
-
-So far, we have looked at how to protect a single resource.
-Some types of resources may have multiple units.
-If there are only $n$ units, no more than $n$ units can be used at a time;
-if all are in use and a thread comes along that needs one of the units,
-it has to wait until another thread releases one of the units.
-Note that we cannot solve this problem simply using a lock per unit---allocating
-a unit requires access to the entire collection.
-
-Consider the following synchronization problem that we call
-\emph{Italian Philosophers} (a simplification of
-the famous Dining Philosophers problem described in \autoref{ch:deadlock}).
-There are five philosophers sitting around a table, each with a plate of
-spaghetti in front of them.
-There is a glass sitting in the middle of the table with only
-three forks in it.  A philosopher needs one of those forks to eat, but clearly,
-no more than three philosophers can eat at a time.
-Unfortunately, a lock can only count to one.
-
-Introduced by Dijkstra,
-a \emph{semaphore} is a synchronization primitive that
-can solve this kind of problem~\cite{EWD35}.
-A semaphore is essentially
-a counter that can be incremented and decremented but is not allowed to go
-below zero.  The semaphore counter is typically initialized to the number of
-units of a resource initially available.
-When allocating a resource, a thread decrements the
-counter using the \texttt{P}
-\index{P}%
-\index{procure}%
-operation.  You can think of \texttt{P} standing
-for \emph{Procure}, as in procuring the resource associated with the semaphore.
-The \texttt{P} operation blocks the invoking thread if the counter is zero.
-To release the resource, a thread increments the resource using the
-\texttt{V}
-\index{V}%
-\index{vacate}%
-operation.  You can think of \texttt{V} standing for \emph{vacate},
-as in vacating the resource.
-One can think of a semaphore as a generalization of a lock.  In fact, a
-lock can be implemented by a semaphore initialized to 1.  Then \texttt{P}
-procures the lock, and \texttt{V} vacates the lock.
-
-\autoref{fig:semaphore} shows the \texttt{synch} module implementation of
-semaphores.
-It represents a semaphore as an integer.
-\autoref{fig:dinerssema} shows a solution to the Italian Philosphers problem
-using a semaphore.
-
-You may find the free ``The Little Book of Semaphores'' by
-Allen Downey a great resource for working with semaphores~\cite{Downey09}.
-
-\section*{Exercises}
-\begin{problems}
-\item How might you check with Harmony that at most three
-Italian philosophers can eat at the same time?
-\item Implement a test program that checks to see that
-the solution allows more than one Italian philosopher to eat at the same time?
-\end{problems}
-\end{comment}
-
-\begin{figure}
-\begin{code}
-\harmonysource{boundedbuffer}
+\begin{figure}
+\begin{code}
+<{:input:../code/boundedbuffer.hny}>
 \end{code}
@@
 \begin{code}
-\harmonysource{BBsema}
+<{:input:../code/BBsema.hny}>
 \end{code}
@@
 \begin{code}
-\harmonysource{BBsematest}
+<{:input:../code/BBsematest.hny}>
 \end{code}
@@
 \begin{code}
-\harmonysource{BBsemadata}
+<{:input:../code/BBsemadata.hny}>
 \end{code}
@@
 the correct order.  There are the same number of producers and consumers.
-Each producer $i$ produces two tuples: $(i, 0)$ and $(i, 1)$ in that order.
+Each producer <{i}> produces two tuples: $(i, 0)$ and $(i, 1)$ in that order.
 Each consumer consumes two tuples and checks that the tuples are different
@@
 
-\begin{figure}
-\begin{code}
-\harmonysource{RWsbs}
+<{:const:release_one}>
+
+\begin{figure}
+\begin{code}
+<{:input:../code/RWsbs.hny}>
 \end{code}
@@
 \end{enumerate}
+
+<{:const:gpuAlloc}>
+<{:const:gpuRelease}>
 
@@
 \begin{code}
-\harmonysource{gpu}
+<{:input:../code/gpu.hny}>
 \end{code}
@@
 {\small
-\harmonysource{RWfair}
+<{:input:../code/RWfair.hny}>
 }
@@
 {\small
-\harmonysource{RWqueue}
+<{:input:../code/RWqueue.hny}>
 }
@@
 \begin{code}
-\harmonysource{RWqtest}
+<{:input:../code/RWqtest.hny}>
 \end{code}
@@
 
+<{:module:hoare}>
+<{:const:Monitor}>
+<{:const:Condition}>
+<{:const:enter}>
+<{:const:exit}>
+
 \glsadd{monitor}%
@@
 \begin{code}
-\harmonysource{hoare}
+<{:input:../code/hoare.hny}>
 \end{code}
@@
 \begin{code}
-\harmonysource{BBhoare}
+<{:input:../code/BBhoare.hny}>
 \end{code}
@@
 \begin{code}
-\harmonysource{mesa}
+<{:input:../code/mesa.hny}>
 \end{code}
@@
 \begin{code}
-\harmonysource{RWcv}
+<{:input:../code/RWcv.hny}>
 \end{code}
@@
 
+<{:module:qsort}>
+<{:const:Qsort}>
+<{:const:swap}>
+<{:const:partition}>
+<{:const:sortrange}>
+<{:const:sort}>
+
 \section*{Exercises}
@@
 \texttt{tryAcquire} should not wait.
-Instead it should return \textbf{True} if the lock was successfully
-acquired and \textbf{False} if the lock was not available.
+Instead it should return <{True}> if the lock was successfully
+acquired and <{False}> if the lock was not available.
 % \item Implement semaphores using Mesa condition variables.
@@
 \begin{code}
-\harmonysource{qsort}
+<{:input:../code/qsort.hny}>
 \end{code}
@@
 \begin{code}
-\harmonysource{qsorttest}
+<{:input:../code/qsorttest.hny}>
 \end{code}
@@
 
+<{:const:diner}>
+
 \glsadd{deadlock}%
@@
 \begin{code}
-\harmonysource{Diners}
+<{:input:../code/Diners.hny}>
 \end{code}
@@
 \begin{code}
-\harmonysource{DinersCV}
+<{:input:../code/DinersCV.hny}>
 \end{code}
@@
 \begin{code}
-\harmonysource{baddblwait}
+<{:input:../code/baddblwait.hny}>
 \end{code}
@@
 \begin{code}
-\harmonysource{dinersfix}
+<{:input:../code/dinersfix.hny}>
 \end{code}
@@
 \begin{code}
-\harmonysource{dinersfix2}
+<{:input:../code/dinersfix2.hny}>
 \end{code}
@@
 \begin{code}
-\harmonysource{DinersAvoid}
+<{:input:../code/DinersAvoid.hny}>
 \end{code}
@@
 \end{figure}
+
+<{:const:transfer}>
 
@@
 \begin{code}
-\harmonysource{bank}
+<{:input:../code/bank.hny}>
 \end{code}
@@
 \index{message passing}%
+
+<{:const:server}>
+<{:const:client}>
 
@@
 \begin{code}
-\harmonysource{counter}
+<{:input:../code/counter.hny}>
 \end{code}
@@
 the counter and increments the counter.  No lock is required.
-The code is tested using the \textit{done} array.
+The code is tested using the <{done}> array.
 
@@
 
+<{:const:Barrier}>
+
 \glsadd{barrier synchronization}%
@@
 \begin{code}
-\harmonysource{qbarrier}
+<{:input:../code/qbarrier.hny}>
 \end{code}
@@
 for each of the \texttt{N} threads.
-Before thread $i$ enters a round, it first sends a message to every
+Before thread <{i}> enters a round, it first sends a message to every
 other thread and then waits until it receives a message from
 every other thread.
-In this simple case, each message contains \texttt{None}, but in practice useful
+In this simple case, each message contains <{None}>, but in practice useful
 information may be exchanged between the threads.
@@
 \begin{code}
-\harmonysource{barrier}
+<{:input:../code/barrier.hny}>
 \end{code}
@@
 \begin{code}
-\harmonysource{barriertest}
+<{:input:../code/barriertest.hny}>
 \end{code}
@@
 
-\begin{figure}
-\begin{code}
-\harmonysource{trap}
+<{:const:handler}>
+<{:const:main}>
+<{:const:increment}>
+
+\begin{figure}
+\begin{code}
+<{:input:../code/trap.hny}>
 \end{code}
@@
 \begin{code}
-\harmonysource{trap2}
+<{:input:../code/trap2.hny}>
 \end{code}
@@
 \begin{code}
-\harmonysource{trap3}
+<{:input:../code/trap3.hny}>
 \end{code}
@@
 \begin{code}
-\harmonysource{trap4}
+<{:input:../code/trap4.hny}>
 \end{code}
@@
 \begin{code}
-\textbf{trap} \textit{handler} \textit{argument}
+<{trap handler argument}>
 \end{code}
@@
 Here, the \textit{main}() thread loops until the interrupt has occurred and
-the \textit{done} flag has been set.
+the <{done}> flag has been set.
 
 But now consider \autoref{fig:trap2}.  The difference with \autoref{fig:trap} is
-that both the \textit{main}() and \textit{handler}() methods increment \textit{count}.
+that both the \textit{main}() and \textit{handler}() methods increment <{count}>.
 This is not unlike the example we gave in \autoref{fig:inc}, except that only a single
@@
 it through Harmony to see for yourself.  If the interrupt occurs after \textit{main}()
-reads \textit{count} (and thus still has value 0) but before \textit{main()} writes the
+reads <{count}> (and thus still has value 0) but before \textit{main()} writes the
 updated value~1, then the interrupt handler will also read value~0 and write value~1.
@@
 temporarily.  In Harmony, this can be done by setting the \textit{interrupt level}
-of a thread to \textbf{True} using the \textbf{setintlevel} interface.
+of a thread to <{True}> using the \textbf{setintlevel} interface.
 \autoref{fig:trap4} illustrates how this is done.
 Note that it is not necessary to change the interrupt level during servicing an
-interrupt, because it is automatically set to \textbf{True} upon entry to the interrupt
-handler and restored to \textbf{False} upon exit.
+interrupt, because it is automatically set to <{True}> upon entry to the interrupt
+handler and restored to <{False}> upon exit.
 It is important that the \textit{main}() code re-enables interrupts after incrementing
-\textit{count}.  What would happen if \textit{main}() left interrupts disabled?
+<{count}>.  What would happen if \textit{main}() left interrupts disabled?
 
@@
 \begin{code}
-\harmonysource{trap5}
+<{:input:../code/trap5.hny}>
 \end{code}
@@
 \begin{code}
-\harmonysource{trap6}
+<{:input:../code/trap6.hny}>
 \end{code}
@@
 interrupt levels.
-To increment \textit{count}, the interrupt level must be \textit{True} and
+To increment <{count}>, the interrupt level must be <{True}> and
 \textit{countlock} must be held.
@@
 Implement a new method \texttt{i\_put}(\textit{item}) that does not
-block.  Instead, it should return \textbf{False} if the buffer is full and \textbf{True}
+block.  Instead, it should return <{False}> if the buffer is full and <{True}>
 if the item was successfully enqueued.
@@
 
+<{:const:inc}>
+<{:const:exch}>
+<{:const:produce}>
+<{:const:consume}>
+
 \glsadd{non-blocking synchronization}%
@@
 \begin{code}
-\harmonysource{hw}
+<{:input:../code/hw.hny}>
 \end{code}
@@
 There are two variables.  \textit{items} is an unbounded array with each
-entry initialized to \texttt{None}.  \textit{back} is an index into the
+entry initialized to <{None}>.  \textit{back} is an index into the
 array and points to the next slot where a new value is inserted.
@@
 the old value;
-\item \texttt{exch}($p$): sets \texttt{!}$p$ to \texttt{None} and returns
+\item \texttt{exch}($p$): sets \texttt{!}$p$ to <{None}> and returns
 the old value.
@@
 
-\begin{figure}
-\begin{code}
-\harmonysource{abp}
+<{:const:net_send}>
+<{:const:net_recv}>
+<{:const:app_send}>
+<{:const:app_recv}>
+<{:const:send}>
+<{:const:receive}>
+<{:const:processor}>
+<{:const:broadcast}>
+
+\begin{figure}
+\begin{code}
+<{:input:../code/abp.hny}>
 \end{code}
@@
 \begin{code}
-\harmonysource{abptest}
+<{:input:../code/abptest.hny}>
 \end{code}
@@
 application message
-$m$, then $m = m_i$ for some $i$ and Bob will already have received
+$m$, then $m = m_i$ for some <{i}> and Bob will already have received
 application messages $m_1, ..., m_i$ (safety).
@@
 Only the last invocation of \texttt{app\_recv}(\textit{reliable}) is invoked with
-\textit{reliable} = \textbf{True}.  It causes the last acknowledgment to be sent
+\textit{reliable} = <{True}>.  It causes the last acknowledgment to be sent
 reliably.  It allows the receiver (Bob) to stop, as well as the sender (Alice)
@@
 \begin{code}
-\harmonysource{leader}
+<{:input:../code/leader.hny}>
 \end{code}
@@
 
-\begin{figure}
-\begin{code}
-\harmonysource{2pc1}
+<{:const:bank}>
+
+\begin{figure}
+\begin{code}
+<{:input:../code/2pc1.hny}>
 \end{code}
@@
 \begin{code}
-\harmonysource{2pc2}
+<{:input:../code/2pc2.hny}>
 \end{code}
@@
 
-\begin{figure}
-\begin{code}
-\harmonysource{chain}
+<{:const:replica}>
+
+\begin{figure}
+\begin{code}
+<{:input:../code/chain.hny}>
 \end{code}
@@
 \texttt{NOPS} operations to its history.
-$\mathbf{choose}(\{ \mathit{immortal}, \textbf{True} \})$ can only
-evaluate to \textbf{False} in case $\textit{immortal}$ is false; otherwise
-it will always evaluate to \textbf{True}.
+$\mathbf{choose}(\{ \mathit{immortal}, <{True}> \})$ can only
+evaluate to <{False}> in case $\textit{immortal}$ is false; otherwise
+it will always evaluate to <{True}>.
 What it does within the loop depends on whether it is the head
@@
 
-\begin{figure}
-\begin{code}
-\harmonysource{abd}
+<{:module:abd}>
+<{:const:read}>
+<{:const:write}>
+<{:const:reader}>
+<{:const:writer}>
+
+\begin{figure}
+\begin{code}
+<{:input:../code/abd.hny}>
 \end{code}
@@
 \begin{code}
-\harmonysource{abdtest}
+<{:input:../code/abdtest.hny}>
 \end{code}
@@
 A server initializes its timestamp $t$ to
-$(0, \mathtt{None})$ and its value to \texttt{None}.
+$(0, \mathtt{None})$ and its value to <{None}>.
 Each server also keeps track of all the requests its already received
@@
 \begin{code}
-\harmonysource{bosco}
+<{:input:../code/bosco.hny}>
 \end{code}
@@
 Since Harmony has to check all possible executions of the protocol, the
-\texttt{receive}(\textit{round}, $k$) method returns all \emph{subbags} of messages
+\texttt{receive}(\textit{round}, <{k}>) method returns all \emph{subbags} of messages
 for the given round that have size $k = \texttt{N} - \texttt{F}$.  The method uses a
 dictionary comprehension to filter out all messages for the given \textit{round}
-and then uses the \texttt{bag.combinations} method to find all combinations of size $k$.
+and then uses the \texttt{bag.combinations} method to find all combinations of size <{k}>.
 The \textbf{atomically when exists} statement waits until there is at least one such combination and
@@
 messages is guaranteed to be odd.  Also, because there are only 0 and 1 values, there
-must exist a majority of zeroes or ones.  Variable \textit{count}[0] stores the number
-of zeroes and \texttt{count}[1] stores the number of ones received in the round.
+must exist a majority of zeroes or ones.  Variable <{count[0]}> stores the number
+of zeroes and <{count[1]}> stores the number of ones received in the round.
 The rules of the algorithm are simple:
@@
 \begin{code}
-\harmonysource{bosco2}
+<{:input:../code/bosco2.hny}>
 \end{code}
@@
 
-\begin{figure}
-\begin{code}
-\harmonysource{paxos1}
+<{:const:leader}>
+<{:const:acceptor}>
+
+\begin{figure}
+\begin{code}
+<{:input:../code/paxos1.hny}>
 \end{code}
@@
 \begin{code}
-\harmonysource{paxos2}
+<{:input:../code/paxos2.hny}>
 \end{code}
@@
 Second is a variable called \textit{last\_accepted} that, if not
-\texttt{None}, contains the last proposal the acceptor has
+<{None}>, contains the last proposal the acceptor has
 \emph{accepted} and the corresponding ballot number.  The acceptor
@@
 \textit{last\_accepted}
-values.  If all those acceptors responded with \texttt{None}, then the leader
+values.  If all those acceptors responded with <{None}>, then the leader
 is free to choose its own proposal.  Otherwise the leader updates
@@
 to $[\mathbf{True}, \mathbf{False}]$ (in case $\texttt{NBALLOTS} = 2$).
-The other combinations of \textbf{True} and \textbf{False} are symmetric,
+The other combinations of <{True}> and <{False}> are symmetric,
 so there is no need to run the model checker through those scenarios as
@@
 
-\begin{figure}
-\begin{code}
-\harmonysource{needhamschroeder}
+<{:const:alice}>
+<{:const:bob}>
+<{:const:corey}>
+
+\begin{figure}
+\begin{code}
+<{:input:../code/needhamschroeder.hny}>
 \end{code}
@@
 \hline
-Boolean & \textbf{False}, \textbf{True} \\
+Boolean & <{False}>, <{True}> \\
 \hline
@@
 are ordered by type according to the table above.
-So, for example, \texttt{\textbf{True} < 0 < .xyz < \{ 0 \})}.
+So, for example, \texttt{<{True}> < 0 < .xyz < \{ 0 \})}.
 Within types, the following rules apply:
@@
 ordered;
-\item an address is a list of Harmony values.  \texttt{None} is the empty list.
+\item an address is a list of Harmony values.  <{None}> is the empty list.
 Addresses are lexicographically ordered accordingly;
@@
 
-The Boolean type has only two possible values: \textbf{False} and
-\textbf{True}.  Unlike in Python, in Harmony booleans are distinct
+The Boolean type has only two possible values: <{False}> and
+<{True}>.  Unlike in Python, in Harmony booleans are distinct
 from integers, and in particular $\mathbf{False} < 0$.  In statements
@@
 \hline
-\texttt{any $s$} & \textbf{True} if any value is \textbf{True} \\
-\hline
-\texttt{all $s$} & \textbf{True} if all values are \textbf{True} \\
+\texttt{any $s$} & <{True}> if any value is <{True}> \\
+\hline
+\texttt{all $s$} & <{True}> if all values are <{True}> \\
 \hline
@@
 Dictionaries support comprehension.  The basic form is:
-\texttt{\{ f($k$):g($k$) for $k$ in $s$ \}}.
+\texttt{\{ f(<{k}>):g(<{k}>) for <{k}> in $s$ \}}.
 
@@
 \hline
-\texttt{len $d$} & the number of keys in $d$ \\
-\hline
-\texttt{keys $d$} & the set of keys in $d$ \\
-\hline
-\texttt{$v$ [not] in keys $d$} & check if $v$ is a \emph{key} in $d$ \\
-\hline
-\texttt{$v$ [not] in $d$} & check if $v$ is a \emph{value} in $d$ \\
-\hline
-\texttt{min $d$} & the minimum value in $d$ \\
-\hline
-\texttt{max $d$} & the maximum value in $d$ \\
-\hline
-\texttt{any $d$} & \textbf{True} if any value is \textbf{True} \\
-\hline
-\texttt{all $d$} & \textbf{True} if all values are \textbf{True} \\
-\hline
-\texttt{$d$ \& $d$ \& ...} & dictionary intersection \\
-\hline
-\texttt{$d$ | $d$ | ...} & dictionary union \\
+\texttt{len <{d}>} & the number of keys in <{d}> \\
+\hline
+\texttt{keys <{d}>} & the set of keys in <{d}> \\
+\hline
+\texttt{$v$ [not] in keys <{d}>} & check if $v$ is a \emph{key} in <{d}> \\
+\hline
+\texttt{$v$ [not] in <{d}>} & check if $v$ is a \emph{value} in <{d}> \\
+\hline
+\texttt{min <{d}>} & the minimum value in <{d}> \\
+\hline
+\texttt{max <{d}>} & the maximum value in <{d}> \\
+\hline
+\texttt{any <{d}>} & <{True}> if any value is <{True}> \\
+\hline
+\texttt{all <{d}>} & <{True}> if all values are <{True}> \\
+\hline
+\texttt{<{d}> \& <{d}> \& ...} & dictionary intersection \\
+\hline
+\texttt{<{d}> | <{d}> | ...} & dictionary union \\
 \hline
@@
 you can write $d[k]$ (or $d(k)$) instead of $d~k$.
-However, if $k$ is a string like \texttt{.id}, then you might
+However, if <{k}> is a string like \texttt{.id}, then you might
 prefer the notation $k\mathtt{.id}$.
 
-Python users beware: the Harmony \texttt{$v$ in $d$} operator checks
-if there is some key $k$ such that $d[k] = v$.  In Python, the same
-syntax checks if $v$ is a key in $d$.  The difference exists because
+Python users beware: the Harmony \texttt{$v$ in <{d}>} operator checks
+if there is some key <{k}> such that $d[k] = v$.  In Python, the same
+syntax checks if $v$ is a key in <{d}>.  The difference exists because
 in Harmony a list is a special case of a dictionary.
@@
 \hline
-\texttt{any $t$} & \textbf{True} if any value is \textbf{True} \\
-\hline
-\texttt{all $t$} & \textbf{True} if all values are \textbf{True} \\
+\texttt{any $t$} & <{True}> if any value is <{True}> \\
+\hline
+\texttt{all $t$} & <{True}> if all values are <{True}> \\
 \hline
@@
 \hline
-\texttt{@l: S} & \texttt{l} is a label \\
+\texttt{l: S} & \texttt{l} is a label \\
 \hline
@@
 Each method has a predefined local variable
-\textit{result}, initialized to \texttt{None},
+<{result}>, initialized to <{None}>,
 that is returned by the method.  Harmony does
@@
 \textbf{for} $k:v$ \textbf{in} $y: S_1; S_2; ...$.  This works similar,
-except that $k$ is bound to the key and $v$ is bound to the value.
+except that <{k}> is bound to the key and $v$ is bound to the value.
 
@@
 $c$ evaluated to true.
-Harmony checks that $c$ is either \textbf{False} or \textbf{True}---if neither
+Harmony checks that $c$ is either <{False}> or <{True}>---if neither
 is the case, Harmony reports an error.
@@
 is evaluated atomically after every state change.
-If it ever evaluates to \textbf{False} Harmony reports and error.  Harmony
+If it ever evaluates to <{False}> Harmony reports and error.  Harmony
 also reports an error if the expression evaluates to a value other
-than \textbf{False} or \textbf{True}.
+than <{False}> or <{True}>.
 
@@
 be handled by the thread.  Such interrupts can be disabled by setting
-the interrupt level of the thread to \textbf{True} using the
+the interrupt level of the thread to <{True}> using the
 \textbf{setintlevel} operator.
@@
 You can introduce new local variables in a method using the
-\texttt{var} statement.
+<{var}> statement.
 The statement \textbf{var} $a = b$ evaluates $b$ and assigns
@@
 \textbf{var} supports pattern matching, so you can
-write \texttt{var} $x, (y, z) = b$.  This will only work
+write <{var}> $x, (y, z) = b$.  This will only work
 if $b$ is a tuple with two elements, the second of which also being a
@@
 The statement \textbf{when} $c: S_1; S_2; ...$ executes statements
-$S_1, S_2, ...$ after waiting until $c$ evaluates to \textbf{True}.
+$S_1, S_2, ...$ after waiting until $c$ evaluates to <{True}>.
 \texttt{when} statements are most useful when combined with the
@@
 The statement \textbf{while} $c: S_1; S_2; ...$ executes statements
-$S_1, S_2, ...$ repeatedly as long as $c$ evaluates to \textbf{True}.
+$S_1, S_2, ...$ repeatedly as long as $c$ evaluates to <{True}>.
 Harmony does not support \textbf{break} or \textbf{continue} statements.
@@
 \begin{code}
-\harmonysource{oo}
+<{:input:../code/oo.hny}>
 \end{code}
@@
 \texttt{let} statements, \texttt{for} loops, and \texttt{when exists} statements.
-Mutable variables can be declared with the \texttt{var} statement.
+Mutable variables can be declared with the <{var}> statement.
 Also, each method has an implicitly declared
-\textit{result} variable, which is initialized to \texttt{None}.
+<{result}> variable, which is initialized to <{None}>.
 Each thread has a mutable variable called \textit{this} that contains
@@
 In the expression $a$ \textbf{and} $b$ \textbf{and} $...$, evaluation is left
-to right but stops once one of the subexpressions evaluates to \textbf{False}.
+to right but stops once one of the subexpressions evaluates to <{False}>.
 Similarly for \textbf{or}, where evaluation stops once one of the subexpressions
-evaluates to \textbf{True}.
+evaluates to <{True}>.
 A sequence of comparison operations, such as $a < b < c$, is evaluated left to right
@@
 \begin{code}
-\texttt{for} key, value in [ (1, 2), (3, 4) ]:
+<{for key, value in [ (1, 2), (3, 4) ]:}>
     ...
@@
 \begin{code}
-\texttt{def} f[$a$, ($b$, $c$)]: ...
+<{def f[a, (b, c)]}>: ...
 \end{code}
@@
 Python iterates over the keys rather than the values;
-\item If you want to iterate over the keys of a dictionary $d$, use either
+\item If you want to iterate over the keys of a dictionary <{d}>, use either
 \begin{itemize}
-\item \texttt{for} $k$ \texttt{in} \texttt{keys} $d$; or
-\item \texttt{for} $k:\_$ \texttt{in} $d$
+\item \texttt{for} <{k}> \texttt{in} \texttt{keys} <{d}>; or
+\item \texttt{for} $k:\_$ \texttt{in} <{d}>
 \end{itemize}
@@
 \begin{code}
-\texttt{for} $i$ in \{ 1..10 \}
+\texttt{for} <{i}> in \{ 1..10 \}
 \texttt{for} $j$ in \{ 1..10 \}
@@
 
-\begin{figure}
-\begin{code}
-\harmonysource{stacktest}
+<{:module:stack}>
+<{:const:Stack}>
+<{:const:stacktest}>
+<{:const:push}>
+<{:const:pop}>
+
+\begin{figure}
+\begin{code}
+<{:input:../code/stacktest.hny}>
 \end{code}
@@
 \begin{code}
-\harmonysource{stack1}
+<{:input:../code/stack1.hny}>
 \end{code}
@@
 \begin{code}
-\harmonysource{stack2}
+<{:input:../code/stack2.hny}>
 \end{code}
@@
 \begin{code}
-\harmonysource{stack3}
+<{:input:../code/stack3.hny}>
 \end{code}
@@
 \begin{code}
-\harmonysource{stack4}
+<{:input:../code/stack4.hny}>
 \end{code}
@@
 The usage is similar to \texttt{malloc} and \texttt{free} in C.
-\texttt{malloc}() is specified to return \texttt{None} when running out of
+\texttt{malloc}() is specified to return <{None}> when running out of
 memory, although this is an impossible outcome in the current
@@
 \hline
-\texttt{combinations}($b$, $k$) & return set of all \emph{subbags} of size $k$ \\
+\texttt{combinations}($b$, <{k}>) & return set of all \emph{subbags} of size <{k}> \\
 \hline
@@
 \hline
-\texttt{combinations}($b$, $k$) & returns set of all subsets of size $k$ \\
+\texttt{combinations}($b$, <{k}>) & returns set of all subsets of size <{k}> \\
 \hline
@@
 \hline
-\texttt{Lock}() & return a binary semaphore initialized to \textbf{False} \\
+\texttt{Lock}() & return a binary semaphore initialized to <{False}> \\
 \hline
@@
 \hline
-choosing & if not \texttt{None}, indicates a context that is choosing \\
+choosing & if not <{None}>, indicates a context that is choosing \\
 \hline
@@
 \hline
-Apply & pop $m$ and $i$ and apply $i$ to $m$, pushing a value \\
-\hline
-Assert, Assert2 & pop $b$ and check that it is \textbf{True}.  Assert2 also pops value to print \\
+Apply & pop $m$ and <{i}> and apply <{i}> to $m$, pushing a value \\
+\hline
+Assert, Assert2 & pop $b$ and check that it is <{True}>.  Assert2 also pops value to print \\
 \hline
@@
 \hline
-Move $i$ & move stack element at offset $i$ to top of the stack \\
+Move <{i}> & move stack element at offset <{i}> to top of the stack \\
 \hline
@@
 \item The effect of the \texttt{Apply} instructions depends much on $m$.
-If $m$ is a dictionary, then \texttt{Apply} finds $i$ in the dictionary
+If $m$ is a dictionary, then \texttt{Apply} finds <{i}> in the dictionary
 and pushes the value.
@@
 register (\emph{i.e.}, the values of the thread variables) onto the
-stack.  It initializes the \texttt{result} variable to \texttt{None}.
+stack.  It initializes the \texttt{result} variable to <{None}>.
 The \texttt{Return} instruction restores the thread register by popping
@@
 \texttt{Continue} instruction.
+\item There are two versions of \texttt{AtomicInc}: \emph{lazy} or \emph{eager}.
+When eager, an atomic section immediately causes a \emph{switch point}
+(switch between threads).
+When lazy, the state change does not happen until the first \texttt{Load}
+\texttt{Store}, or \texttt{Print} instruction.  If there are no such
+instructions, the atomic section may not even cause a switch point.
 \end{itemize}
@@
 \texttt{harmony.py}.  There is a so-called \emph{brief output} that
-is written to standard output.  The more comprehensive output is placed
-in the \texttt{x.htm} HTML output file, allowing you to navigate the
-problematic path and all the details of each of the states on the path.
+is written to standard output.  The rest depends on whether there was
+a problem with the execution or not.  If there was a problem, the more
+comprehensive output is placed in the \texttt{x.htm} HTML output file,
+allowing you to navigate the problematic path and all the details of
+each of the states on the path.  If not, a DFA of the print behavior
+is generated and compared with a provided DFA if specified with the
+\texttt{-B} flag.
 \end{enumerate}
@@
 
+A thread can be in \emph{atomic} mode or not.
+In atomic mode, the execution of the thread is not interleaved with
+other threads.
+A thread can also be in \emph{read-only} mode or not.
+In read-only mode, the thread cannot write or deleted shared variables.
+
 Charm has some tricks to significantly reduce the state space to explore.
@@
 threads, nor by \textbf{invariant} statements.
-So the model checker only interleaves threads at \textbf{Load} and
-\textbf{Store} instructions where a thread interacts with global
+So the model checker only interleaves threads at \textbf{Load},
+\textbf{Store}, and \textbf{Del} instructions where a thread
+interacts with global
 variables.
@@
 \item An assertion failing;
+\item A behavior violation: this is when the sequence of printed values
+are not recognized by the provided DFA (using the \texttt{-B} flag);
 \item A \emph{silly} error: this includes reading variables that have not
@@
 
-The model checker also counts the number of times each \textbf{Possibly}
-instruction is invoked with the value \textbf{True}.  It reports those
-\textbf{Possibly} instructions that either have never been reached or
-never have been executed with the value \textbf{True}.
-
 \section{Model Checker Output Analysis}
 
-The output of the model checker is a graph that is typically very large.
+\index{Kripke structure}%
+The output of the model checker is a graph (a so-called \emph{Kripke
+structure}) that is typically very large.
 If some execution failed, then Harmony will simply report the path of
@@
 
-Finally, in the output, each thread has a unique identifier:
+In the output, each thread has a unique identifier:
 \texttt{T0} is the initialization thread; \texttt{T}$n$ is the $n^{th}$
@@
 
+If there are no issues, Harmony also generates a DFA of the print
+behavior.  Starting with the original state graph or Kripke structure,
+the edges are inspected.  If there are multiple print operations on
+an edge, additional states are inserted so that there are either 0
+or 1 print operations on an edge.  This graph of nodes (states) and edges
+(transitions)
+forms a Non-deterministic Finite Automaton (NFA) with $\epsilon$-transitions
+(transitions without print operations).  Harmony turns the NFA into a DFA
+and by default also minimizes the DFA (although not strictly
+necxessary).  The DFA can be fed into another run of the model checker
+to check that its print operations are consistent with the provided DFA.
+
 \chapter{Directly checking linearizability}
 
-\begin{figure}
-\begin{code}
-\harmonysource{queuelin}
+<{:module:queuelin}>
+<{:module:queuespec}>
+<{:const:_linpoint}>
+
+\begin{figure}
+\begin{code}
+<{:input:../code/queuelin.hny}>
 \end{code}
@@
 \begin{code}
-\harmonysource{qtestconc}
+<{:input:../code/qtestconc.hny}>
 \end{code}

### 2021-12-21 book/sources/preproc.py

@@ -1,0 +1,357 @@
+import sys
+import os
+import getopt
+import traceback
+import collections
+import time
+import math
+
+nametypes = {}
+
+def islower(c):
+    return c in "abcdefghijklmnopqrstuvwxyz"
+
+def isupper(c):
+    return c in "ABCDEFGHIJKLMNOPQRSTUVWXYZ"
+
+def isletter(c):
+    return islower(c) or isupper(c)
+
+def isnumeral(c):
+    return c in "0123456789"
+
+def isalnum(c):
+    return isletter(c) or isnumeral(c)
+
+def isnamechar(c):
+    return isalnum(c) or c == "_"
+
+def isprint(c):
+    return isinstance(c, str) and len(c) == 1 and (
+        isalnum(c) or c in " ~`!@#$%^&*()-_=+[{]}\\|;:'\",<.>/?")
+
+def isnumber(s):
+    return all(isnumeral(c) for c in s)
+
+def allcaps(s):
+    return all(isupper(c) or c == '_' for c in s)
+
+reserved = {
+    "all",
+    "and",
+    "any",
+    "as",
+    "assert",
+    "atLabel",
+    "atomic",
+    "atomically",
+    "await",
+    "bag",
+    "choose",
+    "const",
+    "countLabel",
+    "def",
+    "del",
+    "dict",
+    "elif",
+    "else",
+    "end",
+    "eternal",
+    "exists",
+    "from",
+    "False",
+    "for",
+    "get_context",
+    "go",
+    "hash",
+    "if",
+    "import",
+    "in",
+    "inf",
+    "invariant",
+    "keys",
+    "lambda",
+    "len",
+    "let",
+    "max",
+    "min",
+    "None",
+    "not",
+    "or",
+    "pass",
+    "print",
+    "select",
+    "sequential",
+    "setintlevel",
+    "spawn",
+    "stop",
+    "this",
+    "trap",
+    "True",
+    "var",
+    "when",
+    "where",
+    "while"
+}
+
+constants = {
+    "False",
+    "True",
+    "None",
+    "inf",
+    "synch",
+    "queue",
+    "list",
+    "bag",
+    "alloc",
+    "Lock",
+    "acquire",
+    "release",
+    "wait",
+    "notify",
+    "notifyAll",
+    "signal",
+    "P",
+    "V",
+    "malloc",
+    "free",
+    "Queue",
+    "put",
+    "get"
+}
+
+def isname(s):
[…]

### 2021-12-21 src/charm/charm.c

@@ -159,7 +159,7 @@ static bool onestep(
     assert(step->ctx->failure == 0);
 
     struct global_t *global = w->global;
-    int dfa_state = node->dfa_state;
+    step->dfa_state = node->dfa_state;
 
     // See if we should also try an interrupt.
     if (interrupt) {
@@ -199,17 +199,6 @@ static bool onestep(
 
         struct instr_t *instrs = global->code.instrs;
         struct op_info *oi = instrs[pc].oi;
-        if (global->dfa != NULL && instrs[pc].log) {
-            assert(step->ctx->sp > 0);
-            int nstate = dfa_step(global->dfa, dfa_state, step->ctx->stack[step->ctx->sp - 1]);
-            if (nstate < 0) {
-                char *p = value_string(step->ctx->stack[step->ctx->sp - 1]);
-                value_ctx_failure(step->ctx, &global->values, "Behavior failure on %s", p);
-                free(p);
-                break;
-            }
-            dfa_state = nstate;
-        }
         if (instrs[pc].choose) {
             assert(step->ctx->sp > 0);
             step->ctx->stack[step->ctx->sp - 1] = choice;
@@ -348,12 +337,6 @@ static bool onestep(
         sc->ctxbag = value_bag_add(&global->values, sc->ctxbag, after, 1);
     }
 
-    if (sc->ctxbag == VALUE_DICT && global->dfa != NULL &&
-                    step->ctx->failure == 0 &&
-                    !dfa_is_final(global->dfa, dfa_state)) {
-        value_ctx_failure(step->ctx, &global->values, "Behavior failure: not a final state");
-    }
-
     // Weight of this step
     int weight = ctx == node->after ? 0 : 1;
 
@@ -367,7 +350,7 @@ static bool onestep(
     next->len = node->len + weight;
     next->steps = node->steps + loopcnt;
     next->weight = weight;
-    next->dfa_state = dfa_state;
+    next->dfa_state = step->dfa_state;
 
     next->ai = step->ai;
     step->ai = NULL;
@@ -381,6 +364,17 @@ static bool onestep(
     }
     else if (sc->choosing == 0 && sc->invariants != VALUE_SET) {
         check_invariants(w, next, &w->inv_step);
+    }
+
+    if (sc->ctxbag == VALUE_DICT && global->dfa != NULL &&
+                    step->ctx->failure == 0 &&
+                    !dfa_is_final(global->dfa, step->dfa_state)) {
[…]

### 2021-12-21 src/charm/ops.h

@@ -15,6 +15,7 @@ struct op_info *ops_get(char *opname, int size);
 struct step {
     struct context *ctx;
     struct access_info *ai;
+    int dfa_state;          // behavior DFA
     uint64_t *log;
     int nlog;
 };

### 2021-12-21 book/paper.tex
Updating book to use preproc Also renamed bag.count to bag.multiplicity to avoid clashes with variables called count

@@
 DFA captures both possibilities.
-You can think of the \texttt{choose} operator as enumerating all possible
+You can think of the <{choose}> operator as enumerating all possible
 inputs to the program.
@@
 The example is a sequential program
-and has a method \texttt{triangle} that takes
+and has a method <{triangle}> that takes
 an integer number as argument.  Each method has a variable called
@@
 Running this Harmony program will try all possible executions, which
-includes all possible values for $x$.  Try it out (here \texttt{\$}
+includes all possible values for <{x}>.  Try it out (here \texttt{\$}
 represents a shell prompt):
@@
 \index{constant}%
-For example, if you want to test the code for \texttt{N = 100}, run:
+For example, if you want to test the code for <{N = 100}>, run:
 \begin{code}
@@
 \item Write a Harmony program that computes squares by repeated adding.  So, the program
-should compute the square of $x$ by adding $x$ to an initial value of 0 $x$ times.
+should compute the square of <{x}> by adding <{x}> to an initial value of 0 <{x}> times.
 \end{problems}
@@
 the assertion does not fail and 2) the execution terminates.
-Program (b) is concurrent---it executes methods \texttt{f()} and
-\texttt{g()} in parallel.
-If method \texttt{g()} runs and completes before \texttt{f()}, then
-the assertion in \texttt{f()} will fail when \texttt{f()} runs.
+Program (b) is concurrent---it executes methods <{f()}> and
+<{g()}> in parallel.
+If method <{g()}> runs and completes before <{f()}>, then
+the assertion in <{f()}> will fail when <{f()}> runs.
 Harmony will find this problem.
-This problem is an example of non-determinism: methods \texttt{f()}
-and \texttt{g()} can run in either order.
+This problem is an example of non-determinism: methods <{f()}>
+and <{g()}> can run in either order.
 
@@
 time---the important thing is to develop an understanding of concurrency.
-(Also, you do not get to use the \texttt{atomically} keyword
+(Also, you do not get to use the <{atomically}> keyword
 or a \emph{lock}, yet.)
@@
 \index{HVM}%
-The Harmony compiler places the bytecode for file $x$.\texttt{htm} in
-file $x$.\texttt{hvm}.
+The Harmony compiler places the bytecode for file <{x}>.\texttt{htm} in
+file <{x}>.\texttt{hvm}.
 The model checker, called \emph{Charm}, places its output in a file
-called $x$.\texttt{hco} and a stylized version of the same output
-in an HTML file called $x$.\texttt{htm}.
+called <{x}>.\texttt{hco} and a stylized version of the same output
+in an HTML file called <{x}>.\texttt{htm}.
 
@@
 na\"{\i}ve (and, as it turns out, broken) implementation of a lock.
-Initially the lock is not owned, indicated by \textit{lockTaken} being <{False}>.
-To enter the critical section, a thread waits until \textit{lockTaken} is <{False}>
+Initially the lock is not owned, indicated by <{lockTaken}> being <{False}>.
+To enter the critical section, a thread waits until <{lockTaken}> is <{False}>
 and then sets it to <{True}> to indicate that the lock has been taken.
 The thread then executes the critical section.  Finally the thread
-releases the lock by setting \textit{lockTaken} back to <{False}>.
+releases the lock by setting <{lockTaken}> back to <{False}>.
 
@@
 Thread~0 finds that the lock is available, but just before it stores
-<{True}> in \textit{lockTaken} in instruction~9, thread~1 gets
+<{True}> in <{lockTaken}> in instruction~9, thread~1 gets
 to run.
@@
 to see what it does.)
-Because \textit{lockTaken} is still <{False}>, it too
+Because <{lockTaken}> is still <{False}>, it too
 believes it can acquire the lock, and stores <{True}> in
-\textit{lockTaken} and moves on to the critical section.
+<{lockTaken}> and moves on to the critical section.
 Finally, thread~0 moves on, also stores <{True}>
-into \textit{lockTaken} and also moves
+into <{lockTaken}> and also moves
 into the critical section.  Thread~0 is the one that detects the
 problem.
-The \textit{lockTaken} variable
+The <{lockTaken}> variable
 suffers from the same sort of race condition as the <{count}> variable
@@
 To see why, first note the following invariant: if thread <{i}> is in the
-critical section, then \textit{flags}[<{i}>] = <{True}>.
+critical section, then <{flags[i] = True}>.
 Without loss of generality,
-suppose that thread~0 sets \textit{flags}[0] at time $t_0$.
+suppose that thread~0 sets <{flags[0]}> at time $t_0$.
 Thread 0 can only reach the critical section if at some time $t_1$,
-$t_1 > t_0$, it finds that \textit{flags}[1] = <{False}>.
-Because of the invariant, \textit{flags}[1] = <{False}> implies that
+$t_1 > t_0$, it finds that <{flags[1] = False}>.
+Because of the invariant, <{flags[1] = False}> implies that
 thread~1 is not in the critical section at time $t_1$.
-Let $t_2$ be the time at which thread~0 sets \textit{flags}[0]
+Let $t_2$ be the time at which thread~0 sets <{flags[0]}>
 to <{False}>.  Thread~0 is in the critical section sometime
@@
 It is easy to see that thread~1 cannot enter the critical section
-between $t_1$ and $t_2$, because \textit{flags}[1] = <{False}> at
+between $t_1$ and $t_2$, because <{flags[1] = False}> at
 time $t_1$.  To reach the critical section between $t_1$ and $t_2$,
-it would first have to set \textit{flags}[1] to <{True}> and
-then wait until \textit{flags}[0] = <{False}>.  But that does not happen
+it would first have to set <{flags[1]}> to <{True}> and
+then wait until <{flags[0] = False}>.  But that does not happen
 until time $t_2$.
@@
 The final na\"{\i}ve solution that we propose
-is based on a variable called \textit{turn}.
+is based on a variable called <{turn}>.
 Each thread politely lets the other thread have a turn first.
-When \textit{turn} = <{i}>, thread~<{i}> can
+When <{turn}> = <{i}>, thread~<{i}> can
 enter the critical section, while thread $1-i$ has to wait.
@@
 section, $\mathit{turn} = i$.
-Since \textit{turn} cannot be 0 and 1 at
+Since <{turn}> cannot be 0 and 1 at
 the same time, mutual exclusion is satisfied.
@@
 in \autoref{fig:peterson}.
-(The first line specifies that the \textit{flags} and \textit{turn}
+(The first line specifies that the \textit{flags} and <{turn}>
 variables should have \emph{sequential consistency}---it prevents
@@
 there is more than one thread that can make a step.  The other is
-when a thread executes a \texttt{choose} operation and there is
+when a thread executes a <{choose}> operation and there is
 more than one choice.
@@
 One way to classify states is to define a predicate over states.
-All states in which $x$ = 1, or all states where
+All states in which <{x}> = 1, or all states where
 there are two or more threads executing, are examples of such predicates.
@@
 Unfortunately, exclusivity is not an \emph{inductive} invariant.
-To see why, consider the following state $s$: let thread~0 be at label \texttt{cs}
+To see why, consider the following state <{s}>: let thread~0 be at label \texttt{cs}
 and thread~1 be at the start of the \textbf{await} statement.
-Also, in state $s$, $\mathit{turn} = 1$.  Now let
+Also, in state <{s}>, $\mathit{turn} = 1$.  Now let
 thread~1 make a step.  Because $\mathit{turn} = 1$,
@@
 but not an inductive invariant.
-It will turn out that $s$ is not reachable.
+It will turn out that <{s}> is not reachable.
 
@@
 \autoref{fig:petersonproof} formalizes $\mathcal{G}(i)$ in Harmony.
-The label \texttt{gate} refers to the step that sets \textit{turn} to $1-i$.
+The label \texttt{gate} refers to the step that sets <{turn}> to $1-i$.
 You can run \autoref{fig:petersonproof} to determine
@@
 \item \textit{flags}[1] = <{False}>;
-\item \textit{turn} = 0;
+\item <{turn}> = 0;
 \item thread~1 is at label \texttt{gate}.
@@
 will hold).
-We know that thread~1 never sets \textit{turn} to 1, so
+We know that thread~1 never sets <{turn}> to 1, so
 if the second case holds before the step, it will also hold after the step.
 Finally, if thread~1 is at label \texttt{gate} before the step, then after
-the step \textit{turn} will equal 0, and therefore the second case will hold
+the step <{turn}> will equal 0, and therefore the second case will hold
 after the step.
@@
 A colleague of the author asked if the first two assignments in
-Peterson's algorithm (setting \textit{flags}[\textit{self}])
-to <{True}> and \textit{turn} to $1 - \mathit{self}$) can be reversed.
+Peterson's algorithm (setting <{flags[self]}>
+to <{True}> and <{turn}> to <{1 - self}> can be reversed.
 After all, they are different variables assigned independent values---in a
@@
 
-A method \texttt{m} with argument $a$ is invoked in its
+A method <{m}> with argument <{a}> is invoked in its
 most basic form as follows (assigning the result to~\texttt{r}).
@@
 That's right, no parentheses are required.  In fact, if you invoke
-\texttt{m}($a$), the argument is ($a$), which is the same
-as $a$.
-If you invoke \texttt{m()}, the argument is \texttt{()},
+<{m(a)}>, the argument is (<{a}>), which is the same
+as <{a}>.
+If you invoke <{m()}>, the argument is <{()}>,
 which is the empty tuple.
-If you invoke \texttt{m}($a$, $b$), the argument is ($a$, $b$),
-the tuple consisting of values $a$ and $b$.
+If you invoke <{m(a, b)}>, the argument is <{(a, b)}>,
+the tuple consisting of values <{a}> and <{b}>.
 
@@
 If \texttt{f} is either a method or a
-dictionary, and $x$ is an arbitrary Harmony value, then
-\texttt{f} $x$, \texttt{f}($x$), and \texttt{f}[$x$] are all
+dictionary, and <{x}> is an arbitrary Harmony value, then
+<{f x}>, <{f(x)}>, and <{f[x]}> are all
 the same expression in Harmony.
@@
 <{:const:P_mutex}>
+<{:var:p}>
 
@@
 \index{address}%
-If $x$ is a shared variable, then the expression ?$x$ is the address of $x$.
+If <{x}> is a shared variable, then the expression <{?x}> is the address of <{x}>.
 If a variable contains an address, we call that variable a \emph{pointer}.
 \index{pointer}%
-If $p$ is a pointer to a shared variable, then the
-expression \texttt{!}$p$ is the value of the shared variable.
-In particular, \texttt{!?}$x$ = $x$.
-This is similar to how C pointers work (\texttt{*\string&}$x$ = $x$).
-
-Often, pointers point to dictionaries, and so if $p$ is such a pointer,
-then $(!p).\mathtt{field}$ would evaluate to the specified field in the dictionary.
+If <{p}> is a pointer to a shared variable, then the
+expression \texttt{!}<{p}> is the value of the shared variable.
+In particular, <{!?x = x}>.
+This is similar to how C pointers work (\texttt{*\string&}<{x}> = <{x}>).
+
+Often, pointers point to dictionaries, and so if <{p}> is such a pointer,
+then <{(!p).field}> would evaluate to the specified field in the dictionary.
 Note that the parentheses in this expression
-are needed, as \texttt{!}\textit{p}\texttt{.field} would wrongly evaluate
-\texttt{!}(\textit{p}\texttt{.field}).
-$(!p).\mathtt{field}$ is such a common expression that, like C, Harmony supports the
-shorthand $p-$$>$\texttt{field}, which greatly improves readability.
+are needed, as <{!p.field}> would wrongly evaluate
+<{!(p.field)}>.
+<{(!p).field}> is such a common expression that, like C, Harmony supports the
+shorthand <{p->field}>, which greatly improves readability.
 
@@
 a turn variable and two flags.
-Both methods \texttt{P\_enter} and \texttt{P\_exit} take two arguments:
+Both methods <{P_enter}> and <{P_exit}> take two arguments:
 a pointer to a mutex and the thread identifier (0 or 1).
-$\mathit{pm}$$\rightarrow$$\mathtt{turn}$ is the value of the \texttt{.turn} key
-in the dictionary that \textit{pm} points to.
+<{pm->turn}> is the value of the <{.turn}> key
+in the dictionary that <{pm}> points to.
 
 You can put the first three methods in its own Harmony source file
-and include it using the Harmony \texttt{import} statement.
+and include it using the Harmony <{import}> statement.
 \index{import statement}%
@@
 The arguments to a method and the bound variable (or
-variables) within a \texttt{for} statement are immutable;
+variables) within a <{for}> statement are immutable;
 the variable <{result}> is mutable.
@@
 For example, <{var}> $x = 3$ declares a new mutable local variable
-$x$.
-The \texttt{let} statement allows declaring new immutable local variables.
+<{x}>.
+The <{let}> statement allows declaring new immutable local variables.
 For example:
-\texttt{let} $x = 3$: $y$ $+$$=$ $x$ adds 3 to the global variable
-$y$.
+<{let}> $x = 3$: <{y}> $+$$=$ <{x}> adds 3 to the global variable
+<{y}>.
 See \autoref{sec:vars} for more information.
@@
 The integer \texttt{misses} maintains the number of cache misses.
-Method \texttt{ref}(\textit{ck}, $x$) is invoked when $x$ is
+Method \texttt{ref}(\textit{ck}, <{x}>) is invoked when <{x}> is
 referenced and checked against the cache \textit{ck}.
@@
 The two things to note here is the use of the \textbf{choose}
-expression and the \textbf{assert} statement.  Using \texttt{choose}
+expression and the \textbf{assert} statement.  Using <{choose}>
 we are able to express searching through all possible strings of
@@
 Note that the code of \autoref{fig:lockspec} is executable in Harmony.
-However, the \texttt{atomically} keyword is not available in general
+However, the <{atomically}> keyword is not available in general
 programming languages and should not be used for implementation.
@@
 
-In Harmony, any statement can be preceded by the \texttt{atomically}
+In Harmony, any statement can be preceded by the <{atomically}>
 keyword.  It means that statement as a whole is to be executed atomically.
-The \texttt{atomically} keyword can be used to specify the behavior of
+The <{atomically}> keyword can be used to specify the behavior of
 methods such as \texttt{acquire} and \texttt{release}.  But
 an actual executable program such as in \autoref{fig:cssynch} should not
-use the \texttt{atomically} keyword.
+use the <{atomically}> keyword.
 In fact, if we want to make the entire program executable on hardware,
@@
 
-The code in \autoref{fig:lockspec} also uses the Harmony \texttt{when}
-statement.  A \texttt{when} statement waits until a time in which
+The code in \autoref{fig:lockspec} also uses the Harmony <{when}>
+statement.  A <{when}> statement waits until a time in which
 condition holds (not necessarily the first time) and then executes the
 statement block.  The ``\texttt{await} \textit{condition}'' statement is
-the same as ``\texttt{when} \textit{condition}: \texttt{pass}''.
-Combined with the \texttt{atomically} keyword, the entire statement
+the same as ``<{when}> \textit{condition}: <{pass}>''.
+Combined with the <{atomically}> keyword, the entire statement
 is executed atomically at a time that the condition holds.
@@
 Any Harmony statement can be made atomic by placing either a label in front
-of it or the keyword \texttt{atomically}.
+of it or the keyword <{atomically}>.
 We can use atomic statements to implement a wide variety of
@@
 Method \texttt{test\_and\_set} in \autoref{fig:tas} shows its specification.
-Here $s$ points to a shared Boolean variable and $p$
+Here <{s}> points to a shared Boolean variable and <{p}>
 to a private Boolean variable, belonging to some thread.
@@
 To see that invariant $\mathcal{I}_1$ is maintained, note that
-\texttt{!}$p$ = <{True}> upon entry of \texttt{test\_and\_set}
+\texttt{!}<{p}> = <{True}> upon entry of \texttt{test\_and\_set}
 (because of the condition on the \textbf{while} loop that the
@@
 \begin{enumerate}
-\item \texttt{!}$s$ is <{False}> upon entry to \texttt{test\_and\_set}.
-Then upon exit \texttt{!}$p$ = <{False}> and \texttt{!}$s$ = <{True}>,
+\item \texttt{!}<{s}> is <{False}> upon entry to \texttt{test\_and\_set}.
+Then upon exit \texttt{!}<{p}> = <{False}> and \texttt{!}<{s}> = <{True}>,
 maintaining the invariant.
-\item \texttt{!}$s$ is <{True}> upon entry to \texttt{test\_and\_set}.
+\item \texttt{!}<{s}> is <{True}> upon entry to \texttt{test\_and\_set}.
 Then upon exit nothing has changed, maintaining the invariant.
@@
 
-This causes the context of the thread to be saved in $s$ and
+This causes the context of the thread to be saved in <{s}> and
 the thread to be no longer running.
@@
 have picked the last one, and yet another implementation could have used
-\texttt{choose} to pick an arbitrary one.  Selecting the first is a common
+<{choose}> to pick an arbitrary one.  Selecting the first is a common
 choice in lock implementations as it prevents starvation.
@@
 Report how many states were explored by Harmony for each module.
-\item \label{ex:xy} \autoref{fig:xy} shows a Harmony program with two variables $x$
-(initially 0) and $y$ (initially 100) that can be accessed through methods
+\item \label{ex:xy} \autoref{fig:xy} shows a Harmony program with two variables <{x}>
+(initially 0) and <{y}> (initially 100) that can be accessed through methods
 \texttt{setX} and \texttt{getXY}.  An application invariant is that \texttt{getXY}
@@
 \item $q = \mathtt{Queue}()$: allocate a new queue;
-\item $\mathtt{put}(q, v)$: add $v$ to the tail of the queue;
+\item $\mathtt{put}(q, v)$: add <{v}> to the tail of the queue;
 \item $r = \mathtt{get}(q)$: returns $r = \mathtt{None}$
-if $q$ is empty or $r = v$ if $v$ was at the head of the queue.
+if <{q}> is empty or $r = v$ if <{v}> was at the head of the queue.
 \end{itemize}
@@
 of nodes in the list using \texttt{malloc}() and \texttt{free}().
-\texttt{malloc}($v$) returns a new memory location initialized to $v$,
+\texttt{malloc}(<{v}>) returns a new memory location initialized to <{v}>,
 which should be released with \texttt{free}() when it is no longer in use.
@@
 and tail pointer and a lock.
-The $\mathtt{put}(q, v)$ and $\mathtt{get}(q)$ methods both take a pointer $q$ to the
+The $\mathtt{put}(q, v)$ and $\mathtt{get}(q)$ methods both take a pointer <{q}> to the
 queue object because both may modify the queue.
@@
 locks for the head and the tail,
-allowing a \texttt{put} and a \texttt{get} operation
+allowing a <{put}> and a <{get}> operation
 to proceed concurrently.  To avoid contention between the head and the tail,
@@
 The problem is when the queue is empty and there are concurrent
-\texttt{get} and \texttt{put} operations.
+<{get}> and <{put}> operations.
 They obtain separate locks and then concurrently access the
@@
 \item \label{ex:qcontains} Add a method $\mathtt{contains}(q, v)$ to \autoref{fig:queues}(b)
-that checks to see if $v$ is in queue $q$.
+that checks to see if <{v}> is in queue <{q}>.
 \item Add a method $\mathtt{length}(q)$ to \autoref{fig:queueconc} that returns the length of the given queue.
 The complexity of the method should be $O(1)$, which is to say that you should maintain
-the length of the queue as a field member and update it in \texttt{put} and \texttt{get}.
+the length of the queue as a field member and update it in <{put}> and <{get}>.
 \item \label{ex:qlength} Write a method $\mathtt{check}(q)$
@@
 \begin{itemize}
-\item If the list is empty, $q$$\rightarrow$$\mathtt{tail}$ should be <{None}>.
-Otherwise, the last element in the linked list starting from $q$$\rightarrow$$\mathtt{head}$
-should equal $q$$\rightarrow$$\mathtt{head}$. Moreover,
-$q$$\rightarrow$$\texttt{tail}$$\rightarrow$$next$ should be <{None}>;
+\item If the list is empty, <{q->tail}> should be <{None}>.
+Otherwise, the last element in the linked list starting from <{q->head}>
+should equal <{q->head}>. Moreover,
+<{q->tail->next}> should be <{None}>;
 \item The length field that you added in \autoref{ex:qlength} should equal the length
@@
 \end{itemize}
-Method $\mathtt{check}(q)$ should not obtain a lock; instead add the following line
-just before releasing the lock in \texttt{put} and \texttt{get}:
+Method <{check(q)>} should not obtain a lock; instead add the following line
+just before releasing the lock in <{put}> and <{get}>:
 <{:const:check}>
@@
 \item \label{ex:qremove} Add a method $\mathtt{remove}(q, v)$ to \autoref{fig:queueconc}
-that removes all occurrences of $v$, if any, from queue $q$.
+that removes all occurrences of <{v}>, if any, from queue <{q}>.
 \item The test program in \autoref{fig:queuedemo} is a not thorough test program.
@@
 (which is <{None}> for the final \texttt{inf} node).
-The \texttt{\_find(\textit{lst}, $v$)} helper method first
-finds and locks two consecutive nodes \textit{before} and \textit{after}
-such that $\mathit{before}$$\rightarrow$$\mathtt{data.value} < v \le \mathit{after}$$\rightarrow$$\mathtt{data.value}$.
+The \texttt{\_find(\textit{lst}, <{v}>)} helper method first
+finds and locks two consecutive nodes <{before}> and \textit{after}
+such that <{before->data.value < v <= after->data.value}>.
 It does so by performing something called \emph{hand-over-hand locking}.
@@
 hand-over-hand.
-Using \texttt{\_find} the \texttt{insert}, \texttt{remove}, and \texttt{contains}
+Using \texttt{\_find} the <{insert}>, \texttt{remove}, and \texttt{contains}
 methods are fairly straightforward.
@@
 points in time at which an operation appears to execute atomically.
-The linearization point for the \texttt{insert} operation coincides exactly with the update
-of the \textit{before}.\texttt{next} pointer.
+The linearization point for the <{insert}> operation coincides exactly with the update
+of the <{before.next}> pointer.
 The linearization point of a \texttt{contains} method
@@
 with methods
-$\mathtt{BinTree}()$ to create a new binary tree, $\mathtt{insert}(t, v)$ that inserts $v$ into
-tree $t$, and $\mathtt{contains}(t, v)$ that checks if $v$ is in tree $t$.  Use a single
+$\mathtt{BinTree}()$ to create a new binary tree, $\mathtt{insert}(t, v)$ that inserts <{v}> into
+tree <{t}>, and $\mathtt{contains}(t, v)$ that checks if <{v}> is in tree <{t}>.  Use a single
 lock per binary tree.
@@
 For each operation, the code first chooses whether to
-perform a \texttt{get} or \texttt{put} operation.
-In the case of a \texttt{put} operation, the code also
+perform a <{get}> or <{put}> operation.
+In the case of a <{put}> operation, the code also
 chooses which value to append to the queue.
@@
 and the queue specification.
-In the case of \texttt{get}, the results of the operation
+In the case of <{get}>, the results of the operation
 on both the implementation and specification are checked against one another.
@@
 It is possible to write similar programs in other languages
-such as Python, but the \texttt{choose} expression in Harmony
+such as Python, but the <{choose}> expression in Harmony
 makes it relatively easy to explore all corner cases.
@@
 \autoref{fig:qtestpar} shows the test program.  It starts
-\texttt{NOPS} threads and each choose to do either a \texttt{put}
-or a \texttt{get} operation.  In case of a \texttt{put} operation,
+\texttt{NOPS} threads and each choose to do either a <{put}>
+or a <{get}> operation.  In case of a <{put}> operation,
 the thread enqueues its own name.  In order to output the behaviors,
@@
 \autoref{fig:qtest1} shows a program that
-checks concurrent \texttt{put} operations using
-\texttt{N} threads.  Thread \texttt{putter($v$)}
-adds $v$ to the queue.
+checks concurrent <{put}> operations using
+\texttt{N} threads.  Thread \texttt{putter(<{v}>)}
+adds <{v}> to the queue.
 Thread \texttt{main()} waits until all \texttt{putter}
@@
 Similarly, \autoref{fig:qtest2} shows a program that
-checks concurrent \texttt{get} operations using
+checks concurrent <{get}> operations using
 \texttt{N} threads.
@@
 $0 .. \mathtt{N} - 1$.
-Thread \texttt{getter($v$)} removes an entry from the
+Thread \texttt{getter(<{v}>)} removes an entry from the
 queue and makes sure that no other thread got the
@@
 \autoref{fig:qtest3} checks a mixture of
-concurrent \texttt{get} and \texttt{put} operations.
+concurrent <{get}> and <{put}> operations.
 It starts \texttt{N} \texttt{getter} threads and
@@
 The \texttt{get()} method of the queue implementation
-uses a local variable \textit{node}, declared using
+uses a local variable <{node}>, declared using
 a \textbf{let} statement, but it would be easy to
-mistakenly use a global variable \textit{node} instead.
+mistakenly use a global variable <{node}> instead.
 If there is only one queue, this is not an issue.
@@
 \autoref{fig:qtest4} gives an example that
-tests the \texttt{get} method for use of global
+tests the <{get}> method for use of global
 variables.  Even if the assertion does not fail,
@@
 being a \texttt{dummy} node to prevent data races.  Each node in the
-list contains a lock.  The \texttt{put}() method walks the list until
+list contains a lock.  The <{put}>() method walks the list until
 it gets to the last node, each time acquiring the lock to access the
-node's fields.  When \texttt{put}() gets to the last node in the list,
-it appends a new one.  The \texttt{get}() method locks the first (dummy)
+node's fields.  When <{put}>() gets to the last node in the list,
+it appends a new one.  The <{get}>() method locks the first (dummy)
 node, removes the second from the list and frees it.  The method returns
@@
 T3 was in the process of executing
-\texttt{release}(?$q$.\texttt{lock}) within
-\texttt{get}(?$q$).
-% Looking in the top green row, we see that the value of $q$ is
+\texttt{release}(?<{q}>.\texttt{lock}) within
+<{get}>(?<{q}>).
+% Looking in the top green row, we see that the value of <{q}> is
 % \{ \texttt{.lock}: <{True}>, \texttt{.next}: <{None}>, \texttt{.value}: <{None}> \}.
@@
 \texttt{acquire}(\texttt{?alloc}.\textit{pool}[0].\texttt{lock})
-within \texttt{put}(?$q$, 2), but 
+within <{put}>(?<{q}>, 2), but 
 \texttt{alloc}.\textit{pool}[0] does not exist.
@@
 \begin{enumerate}
-\item[0:] initialization completed, with $q$ being
+\item[0:] initialization completed, with <{q}> being
 \{ \texttt{.lock}: <{False}>, \texttt{.next}: <{None}>, \texttt{.value}: <{None}> \};
 \item[1:] thread T1 (\texttt{thread}(1)) ran and finished executing
-\texttt{put}(1) (see the output column for that clue: the thread printed that).
-We can see that $q$.\texttt{next} now points to \texttt{alloc}.\textit{pool}[0],
+<{put}>(1) (see the output column for that clue: the thread printed that).
+We can see that <{q}>.\texttt{next} now points to \texttt{alloc}.\textit{pool}[0],
 which the thread must have allocated.  The contents is 
@@
 \item[2:] thread T2 (\texttt{thread}(1)) started running, calling
-\texttt{put}(?$q$, 2).  We can see it got as far as putting 2 on the queue,
+<{put}>(?<{q}>, 2).  We can see it got as far as putting 2 on the queue,
 but it is not yet done.  It is currently trying to acquire
@@
 \item[3:] thread T3 (\texttt{thread}(1)) started running, calling
-\texttt{get}($?q$).  We can also see that it freed \textit{pool}[0],
-and is now releasing $q$.\texttt{lock};
+<{get}>($?q$).  We can also see that it freed \textit{pool}[0],
+and is now releasing <{q}>.\texttt{lock};
 \item[4:] thread T2 resumes and tries to access \textit{pool}[0], which
@@
 while T3 was freeing that very same node, and T2 lost the race.
-More precisely, T2 was executing \texttt{put}(?$q$, 2), when T3
-preempted it with \texttt{get}(?$q$) and removed the node that T2
+More precisely, T2 was executing <{put}>(?<{q}>, 2), when T3
+preempted it with <{get}>(?<{q}>) and removed the node that T2
 was trying to access.  But why did the locks not prevent this?
@@
 $<$return$>$ repeatedly, we can quickly skip through the code.
-T2 first calls \texttt{put}(?$q$, 1) and then allocates a new
+T2 first calls <{put}>(?<{q}>, 1) and then allocates a new
 node initialized with a lock.  Keep stepping until it executes
@@
 state of T2 in the lower-right corner.  You can see that variable
-\textit{nq} is initialized to~?$q$.  T2 then enters into the
+<{nq}> is initialized to~?<{q}>.  T2 then enters into the
 \textbf{while} loop and tries to acquire
-$\mathit{nq}$$\rightarrow$$\mathtt{lock}$.  This succeeds, and
-next T2 executes \textbf{let} $n = \mathit{nq}$$\rightarrow$$\mathtt{next}$.
+<{nq->lock}>.  This succeeds, and
+next T2 executes \textbf{let} <{n = nq->next}>.
 Now $n =\,?\mathtt{alloc}.\mathit{alloc}[0]$, which is not <{None}>.
-It then releases $\mathit{nq}$$\rightarrow$$\mathtt{lock}$
-(\textit{nq} points to $q$).  It then sets \textit{nq} to $n$, which
+It then releases <{nq->lock}>
+(<{nq}> points to <{q}>).  It then sets <{nq}> to $n$, which
 is still $?\mathtt{alloc}.\mathit{alloc}[0]$.  Finally, it calls
-$\mathtt{acquire}(?\mathit{nq}$$\rightarrow$$\mathtt{lock})$.
+<{acquire(?nq->lock)}>.
 But before it can complete that operation, T3 runs next.
@@
 T3 chooses "get" and then goes on to invoke $\mathtt{get}(?q)$.
-This first successfully acquires $q$$\rightarrow$$\texttt{lock}$.
-T3 then finds out that $q$$\rightarrow$$\texttt{next}$ points to
-$\texttt{alloc}.\textit{pool}[0]$.  T3 sets \textit{node} to
-$\texttt{alloc}.\textit{pool}[0]$ as well and sets
-$q$$\rightarrow$$\texttt{next}$ to 
-$\mathtt{node}$$\rightarrow$$\texttt{next}$.
-T3 sets <{result}> to 
-$\mathtt{node}$$\rightarrow$$\texttt{value}$ (which is 1)
-and then frees \textit{node}.  This is where the problem is---T2 is
+This first successfully acquires <{q}->lock}>.
+T3 then finds out that <{q->next}> points to <{alloc.pool[0]}>.
+T3 sets <{node}> to <{alloc.pool[0]}> as well and sets
+<{q->{next}> to <{node->next}>.
+T3 sets <{result}> to <{node->value}> (which is 1)
+and then frees <{node}>.  This is where the problem is---T2 is
 about to acquire the lock in that same node.
@@
 that uses hand-over-hand locking
-both for \texttt{put}() and for \texttt{get}().  It passes all tests.
+both for <{put}>() and for <{get}>().  It passes all tests.
 
@@
 Instead, one can conclude
-that \texttt{put} guarantees that the queue will be non-empty, and
+that <{put}> guarantees that the queue will be non-empty, and
 \texttt{signal} will check if there are any threads waiting for
@@
 \index{notify}%
-When a thread $p$ invokes \texttt{notify}, it does not immediately pass 
+When a thread <{p}> invokes \texttt{notify}, it does not immediately pass 
 control to a thread that is waiting on the corresponding condition (if there
-is such a thread).  Instead, $p$ remains in the critical section
+is such a thread).  Instead, <{p}> remains in the critical section
 until it leaves the monitor explicitly
@@
 On the left is a Hoare monitor.
-When some thread $p$ is in the bathroom, it can invoke \textbf{signal}
+When some thread <{p}> is in the bathroom, it can invoke \textbf{signal}
 on some non-empty bedroom associated with a condition that holds.
-This select a thread $q$ in that bedroom.
-As a result of invoking \textbf{signal}, $p$ goes back into the hall
-and $q$ enters the bathroom.\footnote{Per Brinch Hansen' monitors have
+This select a thread <{q}> in that bedroom.
+As a result of invoking \textbf{signal}, <{p}> goes back into the hall
+and <{q}> enters the bathroom.\footnote{Per Brinch Hansen' monitors have
 a \textbf{continue} operation instead of \textbf{signal}.  The difference
@@
 than returning to the hall.}
-Importantly, when $q$ enters the bathroom, the condition that $q$ was
+Importantly, when <{q}> enters the bathroom, the condition that <{q}> was
 waiting for is guaranteed to hold.
@@
 On the right is a Mesa monitor.  The only way into the bathroom is through
-the hall.  When thread $p$ invokes \textbf{notify} on a bedroom, it
-selects a thread $q$ in that bedroom, if any.
-Thread $q$ goes into the hall,
+the hall.  When thread <{p}> invokes \textbf{notify} on a bedroom, it
+selects a thread <{q}> in that bedroom, if any.
+Thread <{q}> goes into the hall,
 joining any other threads that are also waiting to
@@
 If there are no threads in that bedroom, \textbf{notify} is a no-op.
-Importantly, in either case $p$ remains in the bathroom.
-Finally, when $p$ leaves the bathroom, one of the
-threads in the hall can enter the bathroom.  But that may not be $q$.
+Importantly, in either case <{p}> remains in the bathroom.
+Finally, when <{p}> leaves the bathroom, one of the
+threads in the hall can enter the bathroom.  But that may not be <{q}>.
 Assuming every thread eventually leaves the bathroom and the system is
-fair, eventually $q$ will enter the bathroom.  However, it is no longer
+fair, eventually <{q}> will enter the bathroom.  However, it is no longer
 guaranteed that the condition it was waiting for still holds.
-Therefore, the first thing $q$ must
+Therefore, the first thing <{q}> must
 do is check, again, to see if the condition holds.  If not, it should go back
@@
 \item \texttt{Queue()} returns a new message queue;
-\item \texttt{put}($q$, \textit{item}) adds \textit{item} to the queue pointed to by $q$;
-\item \texttt{get}($q$) waits for and returns an item on the queue pointed to by $q$.
+\item <{put}>(<{q}>, \textit{item}) adds \textit{item} to the queue pointed to by <{q}>;
+\item <{get}>(<{q}>) waits for and returns an item on the queue pointed to by <{q}>.
 \end{itemize}
@@
 behaves much like a zero-initialized counting semaphore.
-\texttt{put} is much like \texttt{V}, except that it is accompanied by data.
-\texttt{get} is much like \texttt{P}, except that it also returns data.
+<{put}> is much like \texttt{V}, except that it is accompanied by data.
+<{get}> is much like \texttt{P}, except that it also returns data.
 Thus, synchronized queues can be considered a generalization of
@@
 \begin{problems}
-\item The \texttt{put} method you implemented in \autoref{ex:bbmesa} cannot be used
+\item The <{put}> method you implemented in \autoref{ex:bbmesa} cannot be used
 in interrupt handlers for two reasons: (1) it is not interrupt-safe, and (2)
@@
 \begin{itemize}
-\item \texttt{inc}($p$): atomically increments \texttt{!}$p$ and returns
+\item \texttt{inc}(<{p}>): atomically increments \texttt{!}<{p}> and returns
 the old value;
-\item \texttt{exch}($p$): sets \texttt{!}$p$ to <{None}> and returns
+\item \texttt{exch}(<{p}>): sets \texttt{!}<{p}> to <{None}> and returns
 the old value.
@@
 Also, the network does not create messages out of nothing: if
-message $m$ is received, then message $m$ was sent.
+message <{m}> is received, then message <{m}> was sent.
 
@@
 application message
-$m$, then $m = m_i$ for some <{i}> and Bob will already have received
+<{m}>, then $m = m_i$ for some <{i}> and Bob will already have received
 application messages $m_1, ..., m_i$ (safety).
 Also, if the network is fair and Alice sends application message
-$m$, then eventually Bob should deliver $m$ (liveness).
+<{m}>, then eventually Bob should deliver <{m}> (liveness).
 
@@
 channel as a variable that either contains a network message or nothing
-(we use \texttt{()} in the model).  Let \textit{s\_chan} be the channel
+(we use <{()}> in the model).  Let \textit{s\_chan} be the channel
 from Alice to Bob and \textit{r\_chan} the channel from Bob to Alice.
-\texttt{net\_send}(\textit{pchan}, $m$, \textit{reliable}) models sending
-a message $m$ to
+\texttt{net\_send}(\textit{pchan}, <{m}>, \textit{reliable}) models sending
+a message <{m}> to
 \texttt{!}\textit{pchan}, where \textit{pchan} is either ?\textit{s\_chan}
 or ?\textit{r\_chan}.
-The method places either $m$ (to model a successful send)
-or \texttt{()} (to model loss) in \texttt{!}\textit{pchan}.
+The method places either <{m}> (to model a successful send)
+or <{()}> (to model loss) in \texttt{!}\textit{pchan}.
 The use of the \textit{reliable} flag will be explained later.
@@
 
-Method \texttt{app\_send}($m$) retransmits $m$ until
+Method \texttt{app\_send}(<{m}>) retransmits <{m}> until
 an acknowledgment is received.
@@
 \textbf{atomically when exists} statement.
-This statement takes the form \textbf{atomically when exists} $v$ \textbf{in} $s$: \emph{statement block}
-where $s$ is a set and $v$ is variable that is bound to an element of $s$.
+This statement takes the form <{atomically when exists v in s}>: \emph{statement block}
+where <{s}> is a set and <{v}> is variable that is bound to an element of <{s}>.
 The properties of the statement are as follows:
 \begin{itemize}
-\item it waits until $s$ is non-empty;
+\item it waits until <{s}> is non-empty;
 \item it is executed atomically;
-\item $v$ is selected non-deterministically, like in the \texttt{choose} operator.
+\item <{v}> is selected non-deterministically, like in the <{choose}> operator.
 \end{itemize}
@@
 
-Suppose client $c'$ is trying to write a value $v$.
+Suppose client $c'$ is trying to write a value <{v}>.
 In phase 2, client $c'$ broadcasts a request containing timestamp $(t+1, c')$
-and $v$.  Each server that receives the request compares $(t+1, c')$ against
+and <{v}>.  Each server that receives the request compares $(t+1, c')$ against
 its current timestamp.  If $(t+1, c')$ is larger than its current timestamp,
-it adopts the new timestamp and its corresponding value $v$.  In either case,
+it adopts the new timestamp and its corresponding value <{v}>.  In either case,
 the server responds to the client.
@@
 
-A server initializes its timestamp $t$ to
+A server initializes its timestamp <{t}> to
 $(0, \mathtt{None})$ and its value to <{None}>.
@@
 \label{ch:consensus}
+
+<{:const:multiplicity}>
+<{:const:size}>
+<{:const:empty}>
+<{:const:add}>
+<{:const:combinations}>
 
@@
 ballot number it receives.  Replace the \texttt{max} operator in that statement
-with \texttt{choose} and see if it finds a problem.  First try with
+with <{choose}> and see if it finds a problem.  First try with
 $\texttt{NBALLOTS} = 2$ and then with $\texttt{NBALLOTS} = 3$.
@@
 only that destination can read.  We denote with $\langle m \rangle_p$ a message
-$m$ encrypted for $p$ so that only $p$ can decrypt the message and see that it contains $m$.
+<{m}> encrypted for <{p}> so that only <{p}> can decrypt the message and see that it contains <{m}>.
 
@@
 \hline
-Integer & \texttt{-inf, ..., -2, -1, 0, 1, 2, ..., inf} \\
-\hline
-String & \texttt{"example"}, \texttt{.example} \\
+Integer & <{-inf, ..., -2, -1, 0, 1, 2, ..., inf}> \\
+\hline
+String & <{"example"}>, <{.example}> \\
 \hline
@@
 \hline
-Dictionary & \texttt{\{ .account: 12345, .valid: False \} } \\
-\hline
-Set & \texttt{\{\}, \{ 1, 2, 3 \}, \{ False, .id, 3 \} } \\
-\hline
-Address & \texttt{?lock, ?flags[2], None} \\
-\hline
-Context & (generated by \textbf{stop} expression) \\
+Dictionary & <{{ .account: 12345, .valid: False }}> \\
+\hline
+Set & <{{}, { 1, 2, 3 }, { False, .id, 3 }}> \\
+\hline
+Address & <{?lock, ?flags[2], None}> \\
+\hline
+Context & (generated by <{stop}> expression) \\
 \hline
@@
 Both tuples and lists map indexes (starting at 0) to Harmony values.
-Their format is either \texttt{($e$, $e$, ..., $e$,)} or
-\texttt{[$e$, $e$, ..., $e$,]}.
+Their format is either <{(e, e, ..., e,)}> or <{[e, e, ..., e,]}>.
 If the tuple or list has two or more elements, then the final comma
@@
 are ordered by type according to the table above.
-So, for example, \texttt{<{True}> < 0 < .xyz < \{ 0 \})}.
+So, for example, <{True < 0 < .xyz < { 0 }}>.
 Within types, the following rules apply:
@@
 \begin{itemize}
-\item \texttt{False < True};
+\item <{False < True}>;
 \item integers are ordered in the natural way;
@@
 \hline
-\texttt{$e$ == $e$} & equivalence \\
-\hline
-\texttt{$e$ != $e$} & inequivalence \\
-\hline
-\texttt{$e$ < $e$, $e$ <= $e$, $e$ > $e$, $e$ >= $e$} & comparison\\
+\texttt{<{e}> == <{e}>} & equivalence \\
+\hline
+\texttt{<{e}> != <{e}>} & inequivalence \\
+\hline
+\texttt{<{e}> < <{e}>, <{e}> <= <{e}>, <{e}> > <{e}>, <{e}> >= <{e}>} & comparison\\
 \hline
@@
 \hline
-\texttt{$e$ and $e$ and ...} & conjuction \\
-\hline
-\texttt{$e$ or $e$ or ...} & disjunction \\
-\hline
-\texttt{$e$ => $e$, $e$ not => $e$} & implication \\
-\hline
-\texttt{not $e$} & negation \\
-\hline
-\texttt{$v$ if $e$ else $v'$} & $v$ or $v'$ depending on $e$ \\
-\hline
-\texttt{any $s$, all $s$} & disjunction / conjunction for set or list $s$ \\
+\texttt{<{e}> and <{e}> and ...} & conjuction \\
+\hline
+\texttt{<{e}> or <{e}> or ...} & disjunction \\
+\hline
+\texttt{<{e}> => <{e}>, <{e}> not => <{e}>} & implication \\
+\hline
+\texttt{not <{e}>} & negation \\
+\hline
+\texttt{<{v}> if <{e}> else $v'$} & <{v}> or $v'$ depending on <{e}> \\
+\hline
+\texttt{any <{s}>, all <{s}>} & disjunction / conjunction for set or list <{s}> \\
 \hline
@@
 \hline
-\texttt{-$e$} & negation \\
-\hline
-\texttt{abs($e$)} & absolute value \\
-\hline
-\texttt{$e$ + $e$ + ...} & sum \\
-\hline
-\texttt{$e$ - $e$} & difference \\
-\hline
-\texttt{$e$ * $e$ * $e$ ...} & product \\
-\hline
-\texttt{$e$ / $e$, $e$ // $e$} & integer division \\
-\hline
-\texttt{$e$ \% $e$, $e$ mod $e$} & integer division remainder \\
-\hline
-\texttt{$e$ ** $e$} & power \\
-\hline
-\texttt{\string~$e$} & binary inversion \\
-\hline
-\texttt{$e$ \& $e$ \& ...} & binary and \\
-\hline
-\texttt{$e$ | $e$ | ...} & binary or \\
-\hline
-\texttt{$e$ \string^ $e$ \string^ ...} & binary exclusive or \\
+\texttt{-<{e}>} & negation \\
+\hline
+\texttt{abs(<{e}>)} & absolute value \\
+\hline
+\texttt{<{e}> + <{e}> + ...} & sum \\
+\hline
+\texttt{<{e}> - <{e}>} & difference \\
+\hline
+\texttt{<{e}> * <{e}> * <{e}> ...} & product \\
+\hline
+\texttt{<{e}> / <{e}>, <{e}> // <{e}>} & integer division \\
+\hline
+\texttt{<{e}> \% <{e}>, <{e}> mod <{e}>} & integer division remainder \\
+\hline
+\texttt{<{e}> ** <{e}>} & power \\
+\hline
+\texttt{\string~<{e}>} & binary inversion \\
+\hline
+\texttt{<{e}> \& <{e}> \& ...} & binary and \\
+\hline
+\texttt{<{e}> | <{e}> | ...} & binary or \\
+\hline
+\texttt{<{e}> \string^ <{e}> \string^ ...} & binary exclusive or \\
 \hline
@@
 \hline
-\texttt{$\{ e..e' \}$} & set of integers from $e$ to $e'$ inclusive \\
+\texttt{$\{ e..e' \}$} & set of integers from <{e}> to $e'$ inclusive \\
 \hline
@@
 then a string can be represented by a ``.'' followed by the characters.
-For example, \texttt{.hello} is the same as the string \texttt{"hello"}.
+For example, <{.example}> is the same as the string <{"example"}>.
 
@@
 \hline
-\texttt{len $s$} & cardinality \\
-\hline
-\texttt{$s$ - $s$} & set difference \\
-\hline
-\texttt{$s$ \& $s$ \& ...} & intersection \\
-\hline
-\texttt{$s$ | $s$ | ...} & union \\
-\hline
-\texttt{$s$ \string^ $s$ \string^ ...} & inclusion/exclusion (elements in odd number of sets) \\
-\hline
-\texttt{choose $s$} & select an element (Harmony will try all) \\
-\hline
-\texttt{min $s$} & minimum element \\
-\hline
-\texttt{max $s$} & maximum element \\
-\hline
-\texttt{any $s$} & <{True}> if any value is <{True}> \\
-\hline
-\texttt{all $s$} & <{True}> if all values are <{True}> \\
+\texttt{len <{s}>} & cardinality \\
+\hline
+\texttt{<{s}> - <{s}>} & set difference \\
+\hline
+\texttt{<{s}> \& <{s}> \& ...} & intersection \\
+\hline
+\texttt{<{s}> | <{s}> | ...} & union \\
+\hline
+\texttt{<{s}> \string^ <{s}> \string^ ...} & inclusion/exclusion (elements in odd number of sets) \\
+\hline
+\texttt{choose <{s}>} & select an element (Harmony will try all) \\
+\hline
+\texttt{min <{s}>} & minimum element \\
+\hline
+\texttt{max <{s}>} & maximum element \\
+\hline
+\texttt{any <{s}>} & <{True}> if any value is <{True}> \\
+\hline
+\texttt{all <{s}>} & <{True}> if all values are <{True}> \\
 \hline
@@
 Harmony also supports \emph{set comprehension}.  In its simplest form,
-\texttt{\{ f($v$) for $v$ in $s$ \}} returns a set that is constructed
-by applying \texttt{f} to all elements in $s$ (where $s$ is a set or
+\texttt{\{ f(<{v}>) for <{v}> in <{s}> \}} returns a set that is constructed
+by applying \texttt{f} to all elements in <{s}> (where <{s}> is a set or
 a list).
@@
 For example:
-\texttt{\{ $x$ + $y$ for $x$ in $s$ for $y$ in $s$ where $(x * y) == 4$ \}}
-returns a set that is constructed by summing pairs of elements from $s$
+\texttt{\{ <{x}> + <{y}> for <{x}> in <{s}> for <{y}> in <{s}> where $(x * y) == 4$ \}}
+returns a set that is constructed by summing pairs of elements from <{s}>
 that, when multiplied, have the value 4.
@@
 $\{ k_0:v_0, k_1:v_1, ... \}$.
-Different from Python, the empty dictionary is either $()$ or $[]$
+Different from Python, the empty dictionary is either <{()}> or $[]$
 (because $\{\}$ is the empty set in Harmony).
@@
 Dictionaries support comprehension.  The basic form is:
-\texttt{\{ f(<{k}>):g(<{k}>) for <{k}> in $s$ \}}.
+\texttt{\{ f(<{k}>):g(<{k}>) for <{k}> in <{s}> \}}.
 
@@
 \hline
-\texttt{$v$ [not] in keys <{d}>} & check if $v$ is a \emph{key} in <{d}> \\
-\hline
-\texttt{$v$ [not] in <{d}>} & check if $v$ is a \emph{value} in <{d}> \\
+\texttt{<{v}> [not] in keys <{d}>} & check if <{v}> is a \emph{key} in <{d}> \\
+\hline
+\texttt{<{v}> [not] in <{d}>} & check if <{v}> is a \emph{value} in <{d}> \\
 \hline
@@
 
-Python users beware: the Harmony \texttt{$v$ in <{d}>} operator checks
+Python users beware: the Harmony \texttt{<{v}> in <{d}>} operator checks
 if there is some key <{k}> such that $d[k] = v$.  In Python, the same
-syntax checks if $v$ is a key in <{d}>.  The difference exists because
+syntax checks if <{v}> is a key in <{d}>.  The difference exists because
 in Harmony a list is a special case of a dictionary.
@@
 $n$ values is represented by a dictionary that maps the integers
-$0, ..., n-1$ to those values.  Hence, if $t$ is a list or tuple,
+$0, ..., n-1$ to those values.  Hence, if <{t}> is a list or tuple,
 then the notation $t[0]$ returns the first element of the list.
@@
 \hline
-\texttt{$t$ + $t$ + ...} & concatenation \\
-\hline
-\texttt{$t$ * $n$} & $n$ copies of $t$ concatenated \\
-\hline
-\texttt{$v$ [not] in $t$} & check if $v$ is a \emph{value} in $t$ \\
-\hline
-\texttt{len $t$} & the length of $t$ \\
-\hline
-\texttt{min $t$} & the minimum value in $t$ \\
-\hline
-\texttt{max $t$} & the maximum value in $t$ \\
-\hline
-\texttt{any $t$} & <{True}> if any value is <{True}> \\
-\hline
-\texttt{all $t$} & <{True}> if all values are <{True}> \\
-\hline
-\texttt{$t$ \& $t$ \& ...} & pairwise list minimum \\
-\hline
-\texttt{$t$ | $t$ | ...} & pairwise list maximum \\
+\texttt{<{t}> + <{t}> + ...} & concatenation \\
+\hline
+\texttt{<{t}> * $n$} & $n$ copies of <{t}> concatenated \\
+\hline
+\texttt{<{v}> [not] in <{t}>} & check if <{v}> is a \emph{value} in <{t}> \\
+\hline
+\texttt{len <{t}>} & the length of <{t}> \\
+\hline
+\texttt{min <{t}>} & the minimum value in <{t}> \\
+\hline
+\texttt{max <{t}>} & the maximum value in <{t}> \\
+\hline
+\texttt{any <{t}>} & <{True}> if any value is <{True}> \\
+\hline
+\texttt{all <{t}>} & <{True}> if all values are <{True}> \\
+\hline
+\texttt{<{t}> \& <{t}> \& ...} & pairwise list minimum \\
+\hline
+\texttt{<{t}> | <{t}> | ...} & pairwise list maximum \\
 \hline
@@
 Lists and tuples support comprehension as well.
-In their most basic form: \texttt{f($v$) for $v$ in $s$}.
-For example, to check if any element in a list $t$ is even, you
+In their most basic form: \texttt{f(<{v}>) for <{v}> in <{s}>}.
+For example, to check if any element in a list <{t}> is even, you
 can write:
 
-\texttt{any($(x \% 2) == 0$ for $x$ in $t$)}.
+\texttt{any($(x \% 2) == 0$ for <{x}> in <{t}>)}.
 
@@
 \hline
-\texttt{$v$ [not] in keys $b$} & check if $v$ is in $b$ \\
-\hline
-\texttt{$t$ \& $t$ \& ...} & bag intersection\\
-\hline
-\texttt{$t$ | $t$ | ...} & bag union \\
+\texttt{<{v}> [not] in keys <{b}>} & check if <{v}> is in <{b}> \\
+\hline
+\texttt{<{t}> \& <{t}> \& ...} & bag intersection\\
+\hline
+\texttt{<{t}> | <{t}> | ...} & bag union \\
 \hline
@@
 \hline
-\texttt{$m$ $a$} & invoke method with program counter $m$ and argument $a$\\
-\hline
-\texttt{atLabel $l$} & return a bag of (method, argument) pairs of threads executing at label $l$ \\
-\texttt{countLabel $l$} & return the number of threads executing at label $l$ \\
+<{m a}> & invoke method with program counter <{m}> and argument <{a}>\\
+\hline
+<{atLabel l}> & return a bag of (method, argument) pairs of threads executing at label <{l}> \\ \hline
+<{countLabel l}> & return the number of threads executing at label <{l}> \\
 \hline
@@
 An address can be dereferenced to return the value of the variable.
-If $a$ is an address, then $!a$ returns the value.
-Like C, Harmony supports the shorthand \texttt{$a$->v} for 
+If <{a}> is an address, then $!a$ returns the value.
+Like C, Harmony supports the shorthand \texttt{<{a}>->v} for 
 the expression \texttt{($!a$).v}.
@@
 \begin{comment}
-Some of these values can be extracted.  In particular, if $c$
+Some of these values can be extracted.  In particular, if <{c}>
 is a context, then:
@@
 Harmony currently supports the following statements
-(below, \texttt{S} is a list of statements):
+(below, <{S}> is a list of statements):
 
@@
 \hline
-$e$ & $e$ is an expression\\
-\hline
-\textit{lv} = [lv =]... $e$ & \textit{lv} is an lvalue and $e$ is an expression\\
-\hline
-\textit{lv} \texttt{[op]=} $e$ & \texttt{op} is one of \texttt{+}, \texttt{-},
+<{e}> & <{e}> is an expression\\
+\hline
+<{lv = [lv =]... e}> & <{lv}> is an lvalue and <{e}> is an expression\\
+\hline
+<{lv} [op]= e}> & <{op}> is one of \texttt{+}, \texttt{-},
 \texttt{*}, \texttt{/}, \texttt{//}, \texttt{\%},
-\texttt{\&}, \texttt{|}, \texttt{\^},
-\texttt{and}, \texttt{or}\\
-\hline
-\texttt{l: S} & \texttt{l} is a label \\
-\hline
-\texttt{assert $b$ [, $e$]} & $b$ is a boolean.  Optionally report value of expression $e$\\
-\hline
-\texttt{await $b$} & $b$ is a boolean\\
-\hline
-\texttt{const a = $e$} & \texttt{a} is a bound variable, $e$ is a constant expression\\
-\hline
-\texttt{def m $a$: S} & \texttt{m} is an identifier, $a$ a bound variable\\
-\hline
-\texttt{del} \textit{lv} & delete\\
-\hline
-\texttt{for $a[:b]$ in $e$ [where $c$]: S} & $a$ and $b$ are bound variables, $e$ is a set or dictionary \\
-\hline
-\texttt{from m import ...} & \texttt{m} identifies a module \\
-\hline
-\texttt{go $c$ $e$} & $c$ is a context, $e$ is an expression \\
-\hline
-\texttt{if $b$: S else: S} & $b$ is a boolean, \texttt{S} is a list of statements\\
-\hline
-\texttt{import m, ...} & \texttt{m} identifies a module \\
-\hline
-\texttt{invariant e} & \texttt{e} is an invariant\\
-\hline
-\texttt{let $a$ = $e$: S} & $a$ is a bound variable, $e$ is an expression\\
-\hline
-\texttt{pass} & do nothing\\
-\hline
-\texttt{print $b$ [, $b$, ...]} & each $b$ is a boolean \\
-\hline
-\texttt{sequential v, ...} & \texttt{v} has sequential consistency \\
-\hline
-\texttt{spawn [eternal] m $e$ [, $t$]} & \texttt{m} is a method,
-$e$ is an expression, $t$ is the thread-local state \\
-\hline
-\texttt{trap m $e$} & \texttt{m} is a method and $e$ is an expression \\
-\hline
-\texttt{var $v$ = $e$} & $v$ is a new variable, $e$ is an expression\\
-\hline
-\texttt{when $b$: S} & $b$ is a boolean, \texttt{S} a list of statements\\
-\hline
-\texttt{when exists $a$ in $e$: S} & $a$ is a bound variable, $e$ is an expression\\
-\hline
-\texttt{while $b$: S} & $b$ is a boolean, \texttt{S} a list of statements\\
+\texttt{\&}, \texttt{|}, \texttt{\^}, <{and}>, <{or}>\\
+\hline
+<{l: S}> & <{l}> is a label \\
+\hline
+<{assert b [, e]}> & <{b}> is a boolean.  Optionally report value of expression <{e}>\\
+\hline
+<{await b}> & <{b}> is a boolean\\
+\hline
+<{const a = e}> & <{a}> is a bound variable, <{e}> is a constant expression\\
+\hline
+<{def m a: S}> & <{m}> is an identifier, <{a}> a bound variable\\
+\hline
+<{del lv}> & delete <{lv}>\\
+\hline
+<{for a[:b] in e [where c]: S}> & <{a}> and <{b}> are bound variables, <{e}> is a set or dictionary \\
+\hline
+<{from m import ...}> & <{m}> identifies a module \\
+\hline
+<{go c e}> & <{c}> is a context, <{e}> is an expression \\
+\hline
+<{if b: S else: S}> & <{b}> is a boolean, <{S}> is a list of statements\\
+\hline
+<{import m, ...}> & <{m}> identifies a module \\
+\hline
+<{invariant e}> & <{e}> is an invariant\\
+\hline
+<{let a = e: S} & <{a}> is a bound variable, <{e}> is an expression\\
+\hline
+<{pass}> & do nothing\\
+\hline
+<{print e>}> & <{e}> is an expression \\
+\hline
+<{sequential v, ...}> & <{v}> has sequential consistency \\
+\hline
+<{spawn [eternal] m e [, t]}> & <{m}> is a method,
+<{e}> is an expression, <{t}> is the thread-local state \\
+\hline
+<{trap m e}> & <{m}> is a method and <{e}> is an expression \\
+\hline
+<{var v = e}> & <{v}> is a new variable, <{e}> is an expression\\
+\hline
+<{when b: S}> & <{b}> is a boolean, <{S}> a list of statements\\
+\hline
+<{when exists a in e: S}> & <{a}> is a bound variable, <{e}> is an expression\\
+\hline
+<{while b: S}> & <{b}> is a boolean, <{S}> a list of statements\\
 \hline
@@
 \item Bound variables are read-only.
-\item A statement can be preceded by the \texttt{atomically} keyword to make
+\item A statement can be preceded by the <{atomically}> keyword to make
 the statement atomic.
-\item Multiple \texttt{for} statements can be combined into a single statement.
-\item Multiple \texttt{let} and \texttt{when} statements can be combined into
+\item Multiple <{for}> statements can be combined into a single statement.
+\item Multiple <{let}> and <{when}> statements can be combined into
 a single statement.
@@
 
-The statement $x = 3$ changes the state by assigning~3 to
-variable $x$ (assuming~$x$ was not already~3).
-$x$ may be a local variable or a shared variable.
-The statement $x = y = 3$ first updates~$y$, then~$x$.
+The statement <{x = 3}> changes the state by assigning~3 to
+variable <{x}> (assuming~<{x}> was not already~3).
+<{x}> may be a local variable or a shared variable.
+The statement $x = y = 3$ first updates~<{y}>, then~<{x}>.
 The statement $x[\mathtt{f}()] = y[\mathtt{g}()] = \mathtt{h}()$ first computes
@@
 $y[\mathtt{g}()]$, then evaluates $\mathtt{h}()$, then assigns the resulting value to $y[\mathtt{g}()]$ (using its previously computed address), and finally assigns the same value to $x[\mathtt{f}()]$ (again using its previously computed address).
-The statement $a,b = c$ assumes that $c$ is a tuple with two values.
-It first evaluates the addresses of $a$ and $b$ and first assigns to
+The statement $a,b = c$ assumes that <{c}> is a tuple with two values.
+It first evaluates the addresses of <{a}> and <{b}> and first assigns to
 the latter and then the former.
-If $c$ is not a tuple with two values, then Harmony will report an error.
+If <{c}> is not a tuple with two values, then Harmony will report an error.
 
@@
 
-The statement $x~+$$= 3$ loads $x$, adds 3, and then stores the
-results in $x$.  In this case, it is equivalent to $x = x + 3$.
+The statement $x~+$$= 3$ loads <{x}>, adds 3, and then stores the
+results in <{x}>.  In this case, it is equivalent to $x = x + 3$.
 However, in general this is not so.  For example, $x[\mathtt{f}()]~+$$= 3$
@@
 (In Python, the following two compound statements lead to different results
-for $y$: $x = y = []; x~+$$= [3]$ and $x = y = []; x = x + [3]$.)
+for <{y}>: $x = y = []; x~+$$= [3]$ and $x = y = []; x = x + [3]$.)
 
@@
 
-The statement \textbf{assert $b$} evaluates $b$ and reports an error
-if $b$ is false.  It should be considered a no-op---it is part of the
+The statement \textbf{assert <{b}>} evaluates <{b}> and reports an error
+if <{b}> is false.  It should be considered a no-op---it is part of the
 \emph{specification}, not part of the \emph{implementation} of the algorithm.
 In particular, it specifies an invariant: whenever the program counter is
-at the location where the \textbf{assert} statement is, then $b$ is
+at the location where the \textbf{assert} statement is, then <{b}> is
 always true.
 
-If $b$ is an expression, then it is evaluated atomically.
+If <{b}> is an expression, then it is evaluated atomically.
 Moreover, the expression is not allowed to change the state.
@@
 As in Python, you can specify an additional expression:
-\textbf{assert $b$, $e$}.  The value of $e$ will be reported as part
-of the error should $b$ evaluate to false.
+\textbf{assert <{b}>, <{e}>}.  The value of <{e}> will be reported as part
+of the error should <{b}> evaluate to false.
 
@@
 
-A statement can be preceded by the \texttt{atomically} keyword to
+A statement can be preceded by the <{atomically}> keyword to
 make the statement atomic.
-The statement \textbf{atomically: $S_1; S_2; ...$} evaluates statements
-$S_1, S_2, ...$ atomically.
+The statement <{atomically: S1; S2; ...}> evaluates statements
+<{S1, S2, ...}> atomically.
 This means that the statement runs indivisibly---no other thread can
@@
 
-The statement \textbf{await $b$} is equivalent to \textbf{when $b$: pass}.
+The statement \textbf{await <{b}>} is equivalent to \textbf{when <{b}>: pass}.
 It is intended to improve readability of your code.
@@
 
-The statement \textbf{def} \texttt{m} $a: S_1; S_2: ...$
-defines a new program counter constant \texttt{m} referring to
-a method that takes an argument $a$ and executes the statements
-$S_1, S_2, ...$.  The argument $a$ can be a tuple pattern similar
-to those used in \texttt{let} and \texttt{for} statements.
-Examples include $()$, $(x,)$, $(x, y)$, and $(x, (y, z))$.
+The statement <{def m a: S1; S2: ...}>
+defines a new program counter constant <{m}> referring to
+a method that takes an argument <{a}> and executes the statements
+<{S1, S2, ...}>.  The argument <{a}> can be a tuple pattern similar
+to those used in <{let}> and <{for}> statements.
+Examples include <{()}>, $(x,)$, $(x, y)$, and $(x, (y, z))$.
 The given local variable names are assigned upon application and
@@
 
-The statement \textbf{del} $x$ removes variable $x$ from the
-state.  $x$ can be either a local or a shared variable.  For
+The statement \textbf{del} <{x}> removes variable <{x}> from the
+state.  <{x}> can be either a local or a shared variable.  For
 example, the statement \textbf{del} $x.\texttt{age}$ removes
-the \texttt{.age} field from dictionary $x$.
+the \texttt{.age} field from dictionary <{x}>.
 Harmony automatically removes top-level local variables that
@@
 from Python:
-$x = [\mathtt{.a}, \mathtt{.b}]; \mathbf{del}~x[0]$ results in $x$ having value
+$x = [\mathtt{.a}, \mathtt{.b}]; \mathbf{del}~x[0]$ results in <{x}> having value
 $\{1: \mathtt{.b}\}$ rather than $[\mathtt{.b}]$ (which is $\{0: \mathtt{b}\}$).
@@
 
-The statement \textbf{for} $x$ \textbf{in} $y: S_1; S_2; ...$
-iterates over $y$ and executes for each element the statements
-$S_1, S_2, ...$.
-$y$ must be a set or a dictionary.  $y$ is evaluated only once at
+The statement <{for x in y: S1; S2; ...}>
+iterates over <{y}> and executes for each element the statements
+<{S1, S2, ...}>.
+<{y}> must be a set or a dictionary.  <{y}> is evaluated only once at
 the beginning of the evaluation of this statement.  In case of a set,
@@
 lists intuitive and identical to Python.
-For each element, the statements $S_1, S_2, ...$ are executed with
-local variable $y$ having the value of the element.
-$x$ can be a pattern such as $(a)$ or $(a, (b, c))$.
+For each element, the statements <{S1, S2, ...}> are executed with
+local variable <{y}> having the value of the element.
+<{x}> can be a pattern such as $(a)$ or $(a, (b, c))$.
 If the pattern cannot be matched, Harmony detects and error.
-It is allowed, but discouraged, to assign different values to $x$
-within statements $S_1, S2, ...$.
-
-If $y$ is a dictionary, Harmony also supports the form
-\textbf{for} $k:v$ \textbf{in} $y: S_1; S_2; ...$.  This works similar,
-except that <{k}> is bound to the key and $v$ is bound to the value.
+It is allowed, but discouraged, to assign different values to <{x}>
+within statements <{S1, S2, ...}>.
+
+If <{y}> is a dictionary, Harmony also supports the form
+<{for k:v in y: S1; S2; ...}>.  This works similar,
+except that <{k}> is bound to the key and <{v}> is bound to the value.
 
@@
 Nesting is of the form
-\textbf{for} $x_1$ \textbf{in} $y_1$ \textbf{for} $x_2$ \textbf{in} $y_2$:
-$S_1; S_2; ...$,
+<{for x1 in y1 for x2 in y2: S1; S2; ...}>,
 which is equivalent to the statement
-\textbf{for} $x_1$ \textbf{in} $y_1$: \textbf{for} $x_2$ \textbf{in} $y_2$:
-$S_1; S_2; ...$.
+<{for x1 in y1: for x2 in y2: S1; S2; ...}>.
 Filtering is of the form
-\textbf{for} $x$ \textbf{in} $y$ \textbf{where} $z: S_1; S_2; ...$.
+<{for x in y where z: S1; S2; ...}>.
 For example,
-\textbf{for} $x$ \textbf{in} \{ 1 .. 10 \} \textbf{where} $(x \% 2) == 0:
-S_1; S_2; ...$ only evaluates statements $S_1, S_2, ...$ for even $x$,
+<{for x in 1 .. 10 where (x % 2) == 0: S1; S2; ...}>
+only evaluates statements <{S1, S2, ...}> for even <{x}>,
 that is, 2, 4, 6, 8, and 10.
@@
 
-You can import all constants from a module \texttt{m}
+You can import all constants from a module <{m}>
 (including program counter constants) using the statement
-\textbf{from} \texttt{m} \textbf{import} \texttt{*}.
+\textbf{from} <{m}> \textbf{import} \texttt{*}.
 This, however, excludes constants whose names start with the character
@@
 
-The statement \textbf{go} $c$ $e$ starts a thread with context $c$ that
+The statement \textbf{go} <{c}> <{e}> starts a thread with context <{c}> that
 has executed a \textbf{stop} expression.  The \textbf{stop} expression
-returns value $e$.
+returns value <{e}>.
 The same context can be started multiple times, allowing threads to
@@
 Harmony supports \textbf{if} statements.
-In its most basic form, \textbf{if} $c: S_1; S_2; ...$ evaluates $c$
+In its most basic form, \textbf{if} $c: S_1; S_2; ...$ evaluates <{c}>
 and executes statements $S_1, S_2, ...$ if and only if boolean expression
-$c$ evaluated to true.
-Harmony checks that $c$ is either <{False}> or <{True}>---if neither
+<{c}> evaluated to true.
+Harmony checks that <{c}> is either <{False}> or <{True}>---if neither
 is the case, Harmony reports an error.
@@
 $T_1; T_2; ...$ is similar, but executes statements $T_1, T_2, ...$
-if and only if $c$ evaluated to false.
+if and only if <{c}> evaluated to false.
 You can think of \textbf{elif} $c:$ as shorthand for
@@
 The constants (including method constants) and shared variables declared 
-in that module can subsequently be referenced by prepending ``\texttt{m}.''.
-For example, method \texttt{f()} in imported module \texttt{m} is invoked
+in that module can subsequently be referenced by prepending ``<{m}>.''.
+For example, method <{f()}> in imported module <{m}> is invoked
 by calling \texttt{m.f()}.  If you would prefer to invoke it simply as
-\texttt{f()}, then you have to import using the statement
-\textbf{from} \texttt{m} \textbf{import} \texttt{f}.
+<{f()}>, then you have to import using the statement
+\textbf{from} <{m}> \textbf{import} \texttt{f}.
 
@@
 
-The statement \textbf{invariant} $c$ declares that boolean expression $c$
-is an invariant.  $c$ is only allowed to read shared variables and
+The statement \textbf{invariant} <{c}> declares that boolean expression <{c}>
+is an invariant.  <{c}> is only allowed to read shared variables and
 is evaluated atomically after every state change.
@@
 You can introduce new bound variables in a method using the
-\texttt{let} statement.
-The statement \textbf{let} $a = b: S_1; S_2, ...$ evaluates $b$, assigns
-the result to read-only variable $a$, and evaluates statements
+<{let}> statement.
+The statement \textbf{let} $a = b: S_1; S_2, ...$ evaluates <{b}>, assigns
+the result to read-only variable <{a}>, and evaluates statements
 $S_1, S_2, ...$.  \textbf{let} supports pattern matching, so you can
-write \texttt{let} $x, (y, z) = b: S_1; S_2, ...$.  This will only work
-if $b$ is a tuple with two elements, the second of which also being a
+write <{let}> $x, (y, z) = b: S_1; S_2, ...$.  This will only work
+if <{b}> is a tuple with two elements, the second of which also being a
 tuple with two elements---if not, Harmony will report an error.
@@
 \begin{code}
-\begin{tabbing}
-XXX\=XXX\=XXX\=XXX\kill
-\>\textbf{let} $a = y$: \\
-\>\>\textbf{let} $b = z$: \\
-\>\>\>...
-\end{tabbing}
-\end{code}
-\begin{code}
-\begin{tabbing}
-XXX\=XXX\=XXX\=XXX\kill
-\>\textbf{let} $a = y$ \\
-\>\textbf{let} $b = z$: \\
-\>\>...
-\end{tabbing}
+<{:inline:}>
+let a = y:
+    let b = z:
+        ...
+<{:end:}>
+\end{code}
+
+\begin{code}
+<{:inline:}>
+let a = y
+let b = z:
+<{:end:}>
 \end{code}
@@
 
-The statement \textbf{print $e$} atomically evaluates $e$ in read-only
+The statement \textbf{print <{e}>} atomically evaluates <{e}> in read-only
 mode and adds the result to the print log.  The print log is used to
@@
 
-The statement \texttt{spawn} $m$ $a$ starts a new thread that executes
-method $m$ with argument $a$. $m$ must be a program counter constant,
-and $a$ is typically a tuple containing
+The statement \texttt{spawn} <{m}> <{a}> starts a new thread that executes
+method <{m}> with argument <{a}>. <{m}> must be a program counter constant,
+and <{a}> is typically a tuple containing
 zero or more parameters to be passed to the method.
@@
 is the empty dictionary by default.  It can be specified by adding
-a parameter: \texttt{spawn} $m$ $a, e$ specifies that $e$ should be
+a parameter: \texttt{spawn} <{m}> $a, e$ specifies that <{e}> should be
 the initial value of the thread-local state.
@@
 If a thread may never terminate, you should spawn it with
-\texttt{spawn eternal} $m$ $a$ to suppress those checks.
+\texttt{spawn eternal} <{m}> <{a}> to suppress those checks.
 
@@
 
-The statement \textbf{trap} $m$ $a$ specifies that the current thread
-should execute method $m$ with argument $a$ and some future, unspecified,
+The statement \textbf{trap} <{m}> <{a}> specifies that the current thread
+should execute method <{m}> with argument <{a}> and some future, unspecified,
 time.  It models a timer interrupt or any kind of asynchronous event to
@@
 <{var}> statement.
-The statement \textbf{var} $a = b$ evaluates $b$ and assigns
-the result to local variable $a$.
+The statement \textbf{var} $a = b$ evaluates <{b}> and assigns
+the result to local variable <{a}>.
 \textbf{var} supports pattern matching, so you can
 write <{var}> $x, (y, z) = b$.  This will only work
-if $b$ is a tuple with two elements, the second of which also being a
+if <{b}> is a tuple with two elements, the second of which also being a
 tuple with two elements---if not, Harmony will report an error.
@@
 
-The statement \textbf{when} $c: S_1; S_2; ...$ executes statements
-$S_1, S_2, ...$ after waiting until $c$ evaluates to <{True}>.
-\texttt{when} statements are most useful when combined with the
-\texttt{atomically} keyword.
-If \textit{waiting} is an unused local variable,
-then \textbf{atomically when} $c: S_1; S_2; ...$ is equivalent to
-
-\begin{code}
-\begin{tabbing}
-XXX\=XXX\=XXX\=XXX\=XXX\=XXX\=XXX\=XXX\kill
-\>\textbf{var} $\mathit{waiting} = \mathbf{True}$ \\
-\>\textbf{while} \textit{waiting}: \\
-\>\>\textbf{atomically}: \\
-\>\>\>\textbf{if} $c$:\\
-\>\>\>\>$S_1$ \\
-\>\>\>\>$S_2$ \\
-\>\>\>\>$...$ \\
-\>\>\>\>$\mathit{waiting} = \mathbf{False}$
-\end{tabbing}
-\end{code}
-
-Multiple \texttt{let} and \texttt{when} statements can be combined.
+The statement <{when c: S1; S2; ...}> executes statements
+<{S1, S2, ...}> after waiting until <{c}> evaluates to <{True}>.
+<{when}> statements are most useful when combined with the
+<{atomically}> keyword.
+If <{waiting}> is an unused local variable,
+then <{atomically when c: S1; S2; ...}> is equivalent to
+
+\begin{code}
+<{:inline:}>
+var waiting = True
+while waiting:
+    atomically:
+        if c:
+            S1
+            S2
+            ...
+            waiting = False
+<{:end:}>
+\end{code}
+
+Multiple <{let}> and <{when}> statements can be combined.
 The expressions before the colon are re-evaluated repeated until all
-\texttt{when} conditions are satisfied.
-
-\subsection*{\textbf{when exists ... in ...}}
-
-The statement \textbf{when exists} $x$ \textbf{in} $y: S_1; S_2; ...$
-requires that $y$ evaluates to a set value.
+<{when}> conditions are satisfied.
+
+\subsection*{<{when exists ... in ...}>}
+
+The statement <{when exists} x in y: S1; S2; ...}>
+requires that <{y}> evaluates to a set value.
 The statement does the following three things:
 \begin{itemize}
-\item it waits until $y$ is non-empty;
-\item it selects one element of $y$ non-deterministically (using
-a \texttt{choose} expression);
-\item it executes statements $S_1, S_2, ...$ with the selected element
-bound to read-only variable $x$.
+\item it waits until <{y}> is non-empty;
+\item it selects one element of <{y}> non-deterministically (using
+a <{choose}> expression);
+\item it executes statements <{S1, S2, ...}> with the selected element
+bound to read-only variable <{x}>.
 \end{itemize}
-$x$ may be a pattern, like in \texttt{let}, \texttt{for}, and \texttt{def}
+<{x}> may be a pattern, like in <{let}>, <{for}>, and <{def}>
 statements.
-Harmony reports an error if $y$ evaluates to a value that is not a set.
-
-\texttt{when} statements are most useful when combined with the
-\texttt{atomically} keyword.
-If \textit{waiting} is an unused local variable,
-then \textbf{atomically when exists} $x$ \textbf{in} $y: S_1; S_2; ...$ is equivalent to
-
-\begin{code}
-\begin{tabbing}
-XXX\=XXX\=XXX\=XXX\=XXX\=XXX\=XXX\=XXX\kill
-\>\textbf{var} $\mathit{waiting} = \mathbf{True}:$ \\
-\>\textbf{while} \textit{waiting}: \\
-\>\>\textbf{atomic}: \\
-\>\>\>\textbf{if} $y$ != \{\}:\\
-\>\>\>\>\textbf{let} $x = \mathbf{choose}(y)$:\\
-\>\>\>\>\>$S_1$ \\
-\>\>\>\>\>$S_2$ \\
-\>\>\>\>\>$...$ \\
-\>\>\>\>$\mathit{waiting} = \mathbf{False}$
-\end{tabbing}
+Harmony reports an error if <{y}> evaluates to a value that is not a set.
+
+<{when}> statements are most useful when combined with the
+<{atomically}> keyword.
+If <{waiting}> is an unused local variable,
+then <{atomically when exists x in y: S1; S2; ...}> is equivalent to
+
+\begin{code}
+<{:inline:}>
+var waiting = True:
+    while waiting:
+        atomically:
+            if y != {}:
+                let x = choose(y):
+                    S1
+                    S2
+                    ...
+                waiting = False
+<{:end:}>
 \end{code}
@@
 
-The statement \textbf{while} $c: S_1; S_2; ...$ executes statements
-$S_1, S_2, ...$ repeatedly as long as $c$ evaluates to <{True}>.
+The statement <{while c: S1; S2; ...}> executes statements
+<{S1, S2, ...}> repeatedly as long as <{c}> evaluates to <{True}>.
 Harmony does not support \textbf{break} or \textbf{continue} statements.
@@
 \begin{code}
-<{:input:../code/oo.hny}>
-\end{code}
-
-In Python, lists are objects.  Thus $x$ and $y$ point to the same list,
+<{:inline:}>
+x = y = [ 1, 2 ]
+x[0] = 3
+assert y[0] == 1
+<{:end:}>
+\end{code}
+
+In Python, lists are objects.  Thus <{x}> and <{y}> point to the same list,
 and the assertion would fail if executed by Python.
-In Harmony, lists are values.  So, when $x$ is updated in Line~2, it does
-not affect the value of $y$.  The assertion succeeds.
+In Harmony, lists are values.  So, when <{x}> is updated in Line~2, it does
+not affect the value of <{y}>.  The assertion succeeds.
 Harmony supports references to values (\autoref{ch:method}),
@@
 \begin{code}
-\begin{verbatim}
+<{:var:enter}>
+<{:var:exit}>
+<{:inline:}>
 { .turn: 0, .flags: [ False, False ], .enter: P_enter, .exit: P_exit }
-\end{verbatim}
+<{:end:}>
 \end{code}
@@
 \begin{code}
-\begin{verbatim}
+<{:inline:}>
 x = y = Lock()
-\end{verbatim}
-\end{code}
-
-In Python, this creates two variables $x$ and $y$ referring to the same lock.
+<{:end:}>
+\end{code}
+
+In Python, this creates two variables <{x}> and <{y}> referring to the same lock.
 In Harmony, the two variables will be two different locks.  If you want two
@@
 \begin{code}
-\begin{verbatim}
+<{:inline:}>
 lock = Lock()
 x = y = ?lock
-\end{verbatim}
+<{:end:}>
 \end{code}
@@
 a local mutable variable, or a module.
-Constants are declared using \texttt{const} statements.
+Constants are declared using <{const}> statements.
 Those constants are computed at compile-time.
@@
 Bound variables can be declared in
-\texttt{def} statements (i.e., arguments),
-\texttt{let} statements, \texttt{for} loops, and \texttt{when exists} statements.
+<{def}> statements (i.e., arguments),
+<{let}> statements, <{for}> loops, and \texttt{when exists} statements.
 Mutable variables can be declared with the <{var}> statement.
@@
 We call this \emph{application}, and application is left-associative.
-So, \textit{a b c} is interpreted as (\textit{a b}) $c$: $b$ is applied to $a$,
-and then $c$ is applied to the result.
+So, \textit{a b c} is interpreted as (\textit{a b}) <{c}>: <{b}> is applied to <{a}>,
+and then <{c}> is applied to the result.
 For readability, it may help to write $a(b)$ for function application and
-$a[b]$ for indexing.  In case $b$ is a simple string,
+$a[b]$ for indexing.  In case <{b}> is a simple string,
 you can also write $a.b$ for indexing.
@@
 
-Associative operators ($+$, $*$, $|$, $\string&$, $\string^$, \textbf{and}, \textbf{or})
+Associative operators ($+$, $*$, $|$, $\string&$, $\string^$, <{and}>, \textbf{or})
 are interpreted as general $n$-ary operators, and you are allowed to write
@@
 So, $a[b] + c$
-first evaluates $a$, then $b$ (and then applies $b$ to $a$), and then $c$.  The one
-exception is the expression $a$ \textbf{if} $c$ \textbf{else} $b$, where $c$ is evaluated
-first.  In that expression, only $a$ or $b$ is evaluated depending on the value of $c$.
-In the expression $a$ \textbf{and} $b$ \textbf{and} $...$, evaluation is left
+first evaluates <{a}>, then <{b}> (and then applies <{b}> to <{a}>), and then <{c}>.  The one
+exception is the expression <{a}> \textbf{if} <{c}> \textbf{else} <{b}>, where <{c}> is evaluated
+first.  In that expression, only <{a}> or <{b}> is evaluated depending on the value of <{c}>.
+In the expression <{a}> <{and}> <{b}> <{and}> $...$, evaluation is left
 to right but stops once one of the subexpressions evaluates to <{False}>.
@@
 As an aside:
-the expression $a$ \textbf{not in} $b$ is equivalent to \textbf{not} ($a$ \textbf{in} $b$).
+the expression <{a}> \textbf{not in} <{b}> is equivalent to <{not}> (<{a}> \textbf{in} <{b}>).
 Harmony generalizes this construct for any pair of a unary (except '$-$') and a binary operator.
-In particular, $a$ \textbf{not and} $b$ is the same as \textbf{not} ($a$ \textbf{and} $b$).  
-For those familiar with logic gates, \textbf{not and} is the equivalent of \texttt{NAND}.
-Similarly, \textbf{not} \texttt{$=>$} is non-implication.
+In particular, <{a not and b}> is the same as <{not (a and b)}>.  
+For those familiar with logic gates, <{not and}> is the equivalent of \texttt{NAND}.
+Similarly, <{not =>}> is non-implication.
 
@@
 one element must be followed by a comma.
-For example, the statement \texttt{$x$ = 1,;} assigns a singleton tuple (or list) to $x$.
+For example, the statement \texttt{<{x}> = 1,;} assigns a singleton tuple (or list) to <{x}>.
 
 Because tuples and lists are dictionaries, the \texttt{del} statement is
-different from Python.  For example, if $x$ = [.a, .b, .c], then
-\texttt{del} $x$[1] results in $x$ = \{ 0:.a, 2:.c \}, \emph{not}
-$x$ = [.a, .c].
+different from Python.  For example, if <{x}> = [.a, .b, .c], then
+\texttt{del} <{x}>[1] results in <{x}> = \{ 0:.a, 2:.c \}, \emph{not}
+<{x}> = [.a, .c].
 Harmony also does not support special slicing syntax like Python.
@@
 \begin{itemize}
-\item \texttt{$x,$ = 1,}: assigns 1 to $x$;
-\item \texttt{$x$, $y$ = 1, (2, 3)}: assigns 1 to $x$ and (2, 3) to $y$;
-\item \texttt{$x$, ($y$, $z$) = 1, (2, 3)}: assigns 1 to $x$, 2 to $y$, and
+\item <{x, = 1,}>: assigns 1 to <{x}>;
+\item <{x, y = 1, (2, 3)}>: assigns 1 to <{x}> and (2, 3) to <{y}>;
+\item <{x, (y, z) = 1, (2, 3)}>: assigns 1 to <{x}>, 2 to <{y}>, and
 3 to $z$;
-\item \texttt{$x$, ($y$, $z$) = 1, 2}: generates an runtime error because 2 cannot
-be matched with ($y$, $z$);
-\item \texttt{$x[0]$, $x[1]$ = $x[1]$, $x[0]$}: swaps the first two elements of list $x$.
+\item <{x, (y, z) = 1, 2}>: generates an runtime error because 2 cannot
+be matched with <{(y, z)}>;
+\item <{x[0], x[1] = x[1], x[0]}>: swaps the first two elements of list <{x}>.
 \end{itemize}
 
-As in Python, pattern matching can also be used in \texttt{for} statements.
+As in Python, pattern matching can also be used in <{for}> statements.
 For example:
@@
 \end{code}
-and then call \texttt{f[1, (2, 3)]}.
-Note that the more familiar: \texttt{def} g($a$) defines a method $g$ with
-a single argument $a$.  Invoking g(1, 2) would assign the tuple (1, 2) to
-$a$.  This is not consistent with Python syntax.  For single argument methods,
-you may want to declare as follows: \texttt{def} g($a,$).
-Calling g(1,) assigns 1 to $a$, while calling g(1, 2) would result in a
-runtime error as (1, 2) cannot be matched with ($a$,).
-
-Pattern matching can also be used in \texttt{const}, \texttt{let},
-and \texttt{when exists} statements.
+and then call <{f[1, (2, 3)]}>.
+Note that the more familiar: <{def g(a)}> defines a method <{g}> with
+a single argument <{a}>.  Invoking <{g(1, 2)}> would assign the tuple (1, 2) to
+<{a}>.  This is not consistent with Python syntax.  For single argument methods,
+you may want to declare as follows: <{def}> g($a,$).
+Calling g(1,) assigns 1 to <{a}>, while calling g(1, 2) would result in a
+runtime error as (1, 2) cannot be matched with (<{a}>,).
+
+Pattern matching can also be used in <{const}>, <{let}>,
+and <{when exists}> statements.
 
@@
 \begin{itemize}
-\item \texttt{for} <{k}> \texttt{in} \texttt{keys} <{d}>; or
-\item \texttt{for} $k:\_$ \texttt{in} <{d}>
+\item <{for}> <{k}> \texttt{in} \texttt{keys} <{d}>; or
+\item <{for}> $k:\_$ \texttt{in} <{d}>
 \end{itemize}
@@
 \begin{code}
-\texttt{for} <{i}> in \{ 1..10 \}
-\texttt{for} $j$ in \{ 1..10 \}
-\texttt{where} $i < j$: ...
+<{for i in { 1..10 } for j in { 1..10 } where i < j: ...}>
 \end{code}
@@
 \begin{code}
-\texttt{for} $k:v$ in \{ .a: 1, .b: 2 \}: ...
+<{for k:v in .a: 1, .b: 2: ...}>
 \end{code}
@@
 \item[] \autoref{fig:stack3} represents a stack as a recursively nested tuple
-$(v, f)$, where $v$ is the element on top of the stack and $r$ is a stack
+$(v, f)$, where <{v}> is the element on top of the stack and $r$ is a stack
 that is the remainder;
@@
 \hline
-\texttt{malloc($v$)} & return a pointer to a memory location initialized to $v$ \\
-\hline
-\texttt{free($p$)} & free an allocated memory location $p$ \\
+\texttt{malloc(<{v}>)} & return a pointer to a memory location initialized to <{v}> \\
+\hline
+\texttt{free(<{p}>)} & free an allocated memory location <{p}> \\
 \hline
@@
 \hline
-\texttt{fromSet($s$)} & create a bag from set $s$\\
-\hline
-\texttt{fromList($t$)} & convert list $t$ into a bag \\
-\hline
-\texttt{count($b$, $e$)} & count how many times $e$ occurs in bag $b$\\
-\hline
-\texttt{bchoose($b$)} & like \texttt{choose(s)}, but applied to a bag\\
-\hline
-\texttt{add}($b$, $e$) & add one copy of $e$ to bag $b$ \\
-\hline
-\texttt{remove}($b$, $e$) & remove one copy of $e$ from bag $b$ \\
-\hline
-\texttt{combinations}($b$, <{k}>) & return set of all \emph{subbags} of size <{k}> \\
+\texttt{fromSet(<{s}>)} & create a bag from set <{s}>\\
+\hline
+\texttt{fromList(<{t}>)} & convert list <{t}> into a bag \\
+\hline
+\texttt{multiplicity(<{b}>, <{e}>)} & count how many times <{e}> occurs in bag <{b}>\\
+\hline
+\texttt{bchoose(<{b}>)} & like \texttt{choose(s)}, but applied to a bag\\
+\hline
+\texttt{add}(<{b}>, <{e}>) & add one copy of <{e}> to bag <{b}> \\
+\hline
+\texttt{remove}(<{b}>, <{e}>) & remove one copy of <{e}> from bag <{b}> \\
+\hline
+\texttt{combinations}(<{b}>, <{k}>) & return set of all \emph{subbags} of size <{k}> \\
 \hline
@@
 \hline
-\texttt{enter($m$)} & enter a monitor.  $m$ points to a monitor mutex\\
-\hline
-\texttt{exit($m$)} & exit a monitor\\
+\texttt{enter(<{m}>)} & enter a monitor.  <{m}> points to a monitor mutex\\
+\hline
+\texttt{exit(<{m}>)} & exit a monitor\\
 \hline
@@
 \hline
-\texttt{wait($c$, $m$)} & wait on condition variable pointed to by $c$ in monitor pointed to by $m$\\
-\hline
-\texttt{signal($c$, $m$)} & signal a condition variable\\
+\texttt{wait(<{c}>, <{m}>)} & wait on condition variable pointed to by <{c}> in monitor pointed to by <{m}>\\
+\hline
+\texttt{signal(<{c}>, <{m}>)} & signal a condition variable\\
 \hline
@@
 \hline
-\texttt{subseq($t$, $b$, $f$)} & return a \emph{slice} of list $t$ starting
-at index $b$ and ending just before $f$\\
-\hline
-\texttt{append($t$, $e$)} & append $e$ to list $t$\\
-\hline
-\texttt{head($t$)} & return the first element of list $t$\\
-\hline
-\texttt{tail($t$)} & return all but the first element of list $t$\\
-\hline
-\texttt{reversed($t$)} & reverse a list \\
-\hline
-\texttt{sorted($t$)} & sorted set or list \\
-\hline
-\texttt{set($t$)} & convert values of a dict or list into a set \\
-\hline
-\texttt{list($t$)} & convert set into a list \\
-\hline
-\texttt{values($t$)} & convert values of a dict into a list sorted by key \\
-\hline
-\texttt{items($t$)} & convert dict into (key, value) list sorted by key \\
-\hline
-\texttt{enumerate($t$)} & like Python enumerate \\
-\hline
-\texttt{sum($t$)} & return the sum of all elements in $t$\\
-\hline
-\texttt{qsort($t$)} & sort list $t$ using quicksort\\
+\texttt{subseq(<{t}>, <{b}>, $f$)} & return a \emph{slice} of list <{t}> starting
+at index <{b}> and ending just before $f$\\
+\hline
+\texttt{append(<{t}>, <{e}>)} & append <{e}> to list <{t}>\\
+\hline
+\texttt{head(<{t}>)} & return the first element of list <{t}>\\
+\hline
+\texttt{tail(<{t}>)} & return all but the first element of list <{t}>\\
+\hline
+\texttt{reversed(<{t}>)} & reverse a list \\
+\hline
+\texttt{sorted(<{t}>)} & sorted set or list \\
+\hline
+\texttt{set(<{t}>)} & convert values of a dict or list into a set \\
+\hline
+\texttt{list(<{t}>)} & convert set into a list \\
+\hline
+\texttt{values(<{t}>)} & convert values of a dict into a list sorted by key \\
+\hline
+\texttt{items(<{t}>)} & convert dict into (key, value) list sorted by key \\
+\hline
+\texttt{enumerate(<{t}>)} & like Python enumerate \\
+\hline
+\texttt{sum(<{t}>)} & return the sum of all elements in <{t}>\\
+\hline
+\texttt{qsort(<{t}>)} & sort list <{t}> using quicksort\\
 \hline
@@
 \hline
-\texttt{issubset($s$, $t$)} & returns whether $s$ is a subset of $t$ \\
-\hline
-\texttt{issuperset($s$, $t$)} & returns whether $s$ is a superset of $t$ \\
-\hline
-\texttt{add}($s$, $e$) & returns $s \cup \{ e \}$ \\
-\hline
-\texttt{remove}($s$, $e$) & returns $s \backslash \{ e \}$ \\
-\hline
-\texttt{combinations}($b$, <{k}>) & returns set of all subsets of size <{k}> \\
+\texttt{issubset(<{s}>, <{t}>)} & returns whether <{s}> is a subset of <{t}> \\
+\hline
+\texttt{issuperset(<{s}>, <{t}>)} & returns whether <{s}> is a superset of <{t}> \\
+\hline
+\texttt{add}(<{s}>, <{e}>) & returns $s \cup \{ e \}$ \\
+\hline
+\texttt{remove}(<{s}>, <{e}>) & returns $s \backslash \{ e \}$ \\
+\hline
+\texttt{combinations}(<{b}>, <{k}>) & returns set of all subsets of size <{k}> \\
 \hline
@@
 \vspace{1em}
-For Python programmers: note that $s <= t$ does not check if $s$ is a subset of
-$t$ when $s$ and $t$ are sets, as ``$<=$'' implements a total order on all Harmony values including sets (and the subset relation is not a total order).
+For Python programmers: note that <{s <= t}> does not check if <{s}> is a subset of
+<{t}> when <{s}> and <{t}> are sets, as ``<{<=}>'' implements a total order on all Harmony values including sets (and the subset relation is not a total order).
 
@@
 \hline
-\texttt{BinSem}($v$) & return a binary semaphore initialized to $v$ \\
+\texttt{BinSem}(<{v}>) & return a binary semaphore initialized to <{v}> \\
 \hline
@@
 \hline
-\texttt{wait}($c$, \textit{lk}) & wait on condition variable \texttt{!}$c$
+\texttt{wait}(<{c}>, \textit{lk}) & wait on condition variable \texttt{!}<{c}>
 and lock \textit{lk} \\ \hline
-\texttt{notify}($c$) & notify a thread waiting on condition variable \texttt{!}$c$ \\
-\hline
-\texttt{notifyAll}($c$) & notify all threads waiting on condition variable \texttt{!}$c$ \\
+\texttt{notify}(<{c}>) & notify a thread waiting on condition variable \texttt{!}<{c}> \\
+\hline
+\texttt{notifyAll}(<{c}>) & notify all threads waiting on condition variable \texttt{!}<{c}> \\
 \hline
@@
 \hline
-\texttt{get}($q$) & return next element of $q$, blocking if empty \\
-\hline
-\texttt{put}($q$, \textit{item}) & add \textit{item} to $a$ \\
+<{get}>(<{q}>) & return next element of <{q}>, blocking if empty \\
+\hline
+<{put}>(<{q}>, \textit{item}) & add \textit{item} to <{a}> \\
 \hline
@@
 \hline
-Apply & pop $m$ and <{i}> and apply <{i}> to $m$, pushing a value \\
-\hline
-Assert, Assert2 & pop $b$ and check that it is <{True}>.  Assert2 also pops value to print \\
+Apply & pop <{m}> and <{i}> and apply <{i}> to <{m}>, pushing a value \\
+\hline
+Assert, Assert2 & pop <{b}> and check that it is <{True}>.  Assert2 also pops value to print \\
 \hline
@@
 \hline
-Del [$v$] & delete shared variable $v$ \\
-\hline
-DelVar [$v$] & delete thread variable $v$ \\
+Del [<{v}>] & delete shared variable <{v}> \\
+\hline
+DelVar [<{v}>] & delete thread variable <{v}> \\
 \hline
@@
 \hline
-Frame \texttt{m} $a$ & start method \texttt{m} with arguments $a$,
+Frame <{m}> <{a}> & start method <{m}> with arguments <{a}>,
 initializing variables.  \\
@@
 \hline
-IncVar $v$ & increment thread variable $v$ \\
-\hline
-Jump $p$ & set program counter to $p$ \\
-\hline
-JumpCond $e$ $p$ & pop expression and, if equal to $e$, set program counter to $p$ \\
-\hline
-Load [$v$] & push the value of a shared variable onto the stack \\
-\hline
-LoadVar [$v$] & push the value of a thread variable onto the stack \\
+IncVar <{v}> & increment thread variable <{v}> \\
+\hline
+Jump <{p}> & set program counter to <{p}> \\
+\hline
+JumpCond <{e}> <{p}> & pop expression and, if equal to <{e}>, set program counter to <{p}> \\
+\hline
+Load [<{v}>] & push the value of a shared variable onto the stack \\
+\hline
+LoadVar [<{v}>] & push the value of a thread variable onto the stack \\
 \hline
@@
 \hline
-$n$-ary \textit{op} & apply $n$-ary operator \textit{op} to the top $n$ elements on the stack \\
+$n$-ary <{op}> & apply $n$-ary operator <{op}> to the top $n$ elements on the stack \\
 \hline
@@
 \hline
-Push $c$ & push constant $c$ onto the stack \\
+Push <{c}> & push constant <{c}> onto the stack \\
 \hline
@@
 \hline
-SetIntLevel & pop $e$, set interrupt level to $e$, and push old interrupt level \\
+SetIntLevel & pop <{e}>, set interrupt level to <{e}>, and push old interrupt level \\
 \hline
@@
 \hline
-Stop [$v$] & save context into shared variable $v$ and remove from context bag \\
-\hline
-Store [$v$] & pop a value from the stack and store it in a shared variable \\
-\hline
-StoreVar [$v$] & pop a value from the stack and store it in a thread variable \\
+Stop [<{v}>] & save context into shared variable <{v}> and remove from context bag \\
+\hline
+Store [<{v}>] & pop a value from the stack and store it in a shared variable \\
+\hline
+StoreVar [<{v}>] & pop a value from the stack and store it in a thread variable \\
 \hline
@@
 value on the stack.
-\item The effect of the \texttt{Apply} instructions depends much on $m$.
-If $m$ is a dictionary, then \texttt{Apply} finds <{i}> in the dictionary
+\item The effect of the \texttt{Apply} instructions depends much on <{m}>.
+If <{m}> is a dictionary, then \texttt{Apply} finds <{i}> in the dictionary
 and pushes the value.
-If $m$ is a program counter, then \texttt{Apply} invokes method $m$ by
+If <{m}> is a program counter, then \texttt{Apply} invokes method <{m}> by
 pushing the current program counter and setting the program counter to
-$m$.  $m$ is supposed to leave the result on the stack.
+<{m}>.  <{m}> is supposed to leave the result on the stack.
 \item The \texttt{Frame} instruction pushes the value of the thread
@@
 \begin{itemize}
-\item \texttt{m()} invokes method \texttt{m} with the empty dictionary \texttt{()} as argument;
-\item \texttt{m($a$)} invokes method \texttt{m} with argument $a$;
-\item \texttt{m($a$, $b$, $c$)} invokes method \texttt{m} with tuple \texttt{($a$, $b$, $c$)} as argument.
+\item <{m()}> invokes method <{m}> with the empty dictionary <{()}> as argument;
+\item <{m(a)}> invokes method <{m}> with argument <{a}>;
+\item <{m(a, b, c)}> invokes method <{m}> with tuple <{(a, b, c)}> as argument.
 \end{itemize}
@@
 consistently with a sequential queue in that all
-\texttt{put}
-and \texttt{get} operations should appear to happen
+<{put}>
+and <{get}> operations should appear to happen
 in a total order.
 Moreover, we want to make sure that if some
-\texttt{put} or \texttt{get} operation $o_1$ finished
+<{put}> or <{get}> operation $o_1$ finished
 before another operation $o_2$ started, then $o_1$
@@
 The queue implementation in \autoref{fig:queueconc} does
-not quite match this pattern, as the \texttt{put}
+not quite match this pattern, as the <{put}>
 operation allocates a new node before acquiring
@@
 linearization points.
-For each operation (\texttt{get} and \texttt{put}), the corresponding
+For each operation (<{get}> and <{put}>), the corresponding
 linearization point must occur somewhere between acquiring and releasing

### 2021-12-21 book/sources/preproc.py

@@ -34,7 +34,7 @@ def isnumber(s):
     return all(isnumeral(c) for c in s)
 
 def allcaps(s):
-    return all(isupper(c) or c == '_' for c in s)
+    return all(isupper(c) or isnumeral(c) or c == '_' for c in s)
 
 reserved = {
     "all",
@@ -308,12 +308,14 @@ def loadAll(filename):
             all += line
     return all
 
-def doInput(filename):
-    all = loadAll(filename)
+def doInline(s, filename):
     print("\\begin{tabbing}")
     print("X\\=XX\\=XXX\\kill")
-    lexer(all, filename)
+    lexer(s, filename)
     print("\\end{tabbing}")
+
+def doInput(filename):
+    doInline(loadAll(filename), filename)
 
 def parse(s, filename):
     if s == "":
@@ -342,8 +344,16 @@ def doProcess(filename):
         print(all[:i], end="")
         all = all[i+2:]
         j = all.index("}>")
-        parse(all[:j], filename)
+        cmd = all[:j]
         all = all[j+2:]
+        if cmd == ":inline:":
+            if all[0] == "\n":
+                all = all[1:]
+            end = all.index("<{:end:}>")
+            doInline(all[:end], filename)
+            all = all[end+9:]
+        else:
+            parse(cmd, filename)
 
 def main():
     for k in reserved:

### 2021-12-22 book/paper.tex
Fixed some json escape issues

@@
 \end{itemize}
-Method <{check(q)>} should not obtain a lock; instead add the following line
+Method <{check(q)}> should not obtain a lock; instead add the following line
 just before releasing the lock in <{put}> and <{get}>:

### 2021-12-22 src/charm/charm.c

@@ -440,33 +440,6 @@ void print_vars(FILE *file, uint64_t v){
     fprintf(file, " }");
 }
 
-static char *json_escape(const char *s, unsigned int len){
-	struct strbuf sb;
-
-	strbuf_init(&sb);
-	while (len > 0) {
-		switch (*s) {		// TODO.  More cases
-		case '"':
-			strbuf_append(&sb, "\\\"", 2);
-			break;
-		case '\\':
-			strbuf_append(&sb, "\\\\", 2);
-			break;
-        case '\n':
-			strbuf_append(&sb, "\\n", 2);
-			break;
-        case '\r':
-			strbuf_append(&sb, "\\r", 2);
-			break;
-		default:
-			strbuf_append(&sb, s, 1);
-		}
-		s++;
-		len--;
-	}
-	return strbuf_getstr(&sb);
-}
-
 char *json_escape_value(uint64_t v){
     char *s = value_string(v);
     char *r = json_escape(s, strlen(s));

### 2021-12-22 src/charm/json.c

@@ -366,3 +366,30 @@ struct json_value *json_lookup_value(struct dict *map, char *key){
 	}
 	return jv;
 }
+
+char *json_escape(const char *s, unsigned int len){
+	struct strbuf sb;
+
+	strbuf_init(&sb);
+	while (len > 0) {
+		switch (*s) {		// TODO.  More cases
+		case '"':
+			strbuf_append(&sb, "\\\"", 2);
+			break;
+		case '\\':
+			strbuf_append(&sb, "\\\\", 2);
+			break;
+        case '\n':
+			strbuf_append(&sb, "\\n", 2);
+			break;
+        case '\r':
+			strbuf_append(&sb, "\\r", 2);
+			break;
+		default:
+			strbuf_append(&sb, s, 1);
+		}
+		s++;
+		len--;
+	}
+	return strbuf_getstr(&sb);
+}

### 2021-12-22 src/charm/value.c

@@ -309,12 +309,18 @@ static char *value_string_atom(uint64_t v) {
 	strbuf_init(&sb);
     strbuf_append(&sb, "\"", 1);
 	while (size > 0) {
-		switch (*s) {		// TODO.  More cases
+		switch (*s) {
 		case '"':
 			strbuf_append(&sb, "\\\"", 2);
 			break;
 		case '\\':
 			strbuf_append(&sb, "\\\\", 2);
+			break;
+		case '\n':
+			strbuf_append(&sb, "\\n", 2);
+			break;
+		case '\r':
+			strbuf_append(&sb, "\\r", 2);
 			break;
 		default:
 			strbuf_append(&sb, s, 1);
@@ -332,7 +338,9 @@ static char *value_json_atom(uint64_t v) {
     char *s = dict_retrieve(p, &size), *r;
     assert(size > 0);
     if (size > 1 || (isprint(s[0]) && s[0] != '"' && s[0] != '\\')) {
-        alloc_printf(&r, "{ \"type\": \"atom\", \"value\": \"%.*s\" }", size, s);
+        char *esc = json_escape(s, size);
+        alloc_printf(&r, "{ \"type\": \"atom\", \"value\": \"%s\" }", esc);
+        free(esc);
     }
     else {
         alloc_printf(&r, "{ \"type\": \"char\", \"value\": \"%02x\" }", s[0]);

### 2021-12-23 book/paper.tex
print is no longer atomic or read-only
Latest version of book

@@
 developing concurrent code, which includes programs for distributed
-systems.  In particular, it uses a new tool
+systems.  In particular, it uses a tool
 called Harmony that helps with \emph{testing} concurrent code.
@@
 Because model checking tries every possible execution, the test
-program needs to be simple. Otherwise it may take longer than we
+program needs to be simple---otherwise it may take longer than we
 care to wait for or run out of memory.
@@
 useful?
-To answer that question, let us consider a sorting algorithm.
+To answer that question, consider a sorting algorithm.
 Suppose we create a test program, a model, that tries sorting
@@
 \item \autoref{ch:specification} talks about how Harmony can be used as
-a specification language.  It introduces how can can specify atomic
+a specification language.  It introduces how to specify atomic
 constructs.
@@
 Paxos consensus protocol.
-\item \autoref{ch:ns} demonstrates finding a bug in the original Needham-Schroeder
-authentication protocol.
+\item \autoref{ch:ns} demonstrates using Harmony to find a (known) bug in the original Needham-Schroeder authentication protocol.
 \end{itemize}
@@
 more copies of the string ``hello world''.  Note that this single state DFA
-(where the initial state and the final state happend to be the same)
+(where the initial state and the final state happen to be the same)
 captures an infinite number of possible outputs.
@@
 \begin{problems}
+\item Add the line <{print(x, triangle(x))}> to the end of the
+program and create an output png file.  Before you look at it, what
+do you think it should look like?
 \item See what happens if, instead of initializing <{result}> to 0,
@@
 It sets <{shared}> to <{True}>, asserts that
-<{shared}> = <{True}>, and finally sets
-<{shared}> to <{False}>.
+<{shared}> = <{True}> and finally sets <{shared}> to <{False}>.
 If you run the program through Harmony, it will not find any problems
@@
 the assertion in <{f()}> will fail when <{f()}> runs.
-Harmony will find this problem.
 This problem is an example of non-determinism: methods <{f()}>
 and <{g()}> can run in either order.
+In one order, the assertion fails, while in the other it does not.
+But since Harmony will find all possible executions, it will find
+the problematic one.
 
@@
 
-The statement \textbf{print <{e}>} atomically evaluates <{e}> in read-only
-mode and adds the result to the print log.  The print log is used to
+The statement \textbf{print <{e}>} evaluates <{e}>
+and adds the result to the print log.  The print log is used to
 create an ``external behavior DFA'' for the Harmony program.
@@
 called HarmonyLang and proceeded to build an animator for Harmony
-executions and two cloud-based Harmony offerings, which you can can learn
+executions and two cloud-based Harmony offerings, which you can learn
 about at \url{http://harmony.cs.cornell.edu}.  They also developed much of

### 2021-12-23 book/paper.tex
Updated bosco

@@
 
-The state of a processor consists
-of its current round number and proposal.  Therefore, messages contain
-a round number and a proposal. To start things, each processor first
-broadcasts its initial round number (0) and proposal.
+The state of a processor consists of its current round number (initially 0)
+and an estimate (initially the proposal).  Therefore, messages contain
+a round number and an estimate. To start things, each processor first
+broadcasts its initial round number and initial estimate.
 The number of rounds that are necessary to achieve consensus is not bounded.  But
@@
 \begin{itemize}
-\item update \textit{proposal} to be the majority value;
+\item update <{estimate}> to be the majority value;
 \item if the quorum is unanimous, decide the value.
@@
 After that, proceed with the next round.
-
-\begin{comment}
-The \texttt{possibly} statement checks if there exist executions where 0 is
-decided and if there exist executions where 1 is decided.  Doing so is useful
-because the assertions would certainly not fail if no decision is ever made.
-Also, it is nice to know that both 0 and 1 can be decided.
-\end{comment}
 

### 2021-12-23 book/paper.tex
Latest

@@
 
-The exercises below have you try the same thing in Python.  As you will
+The exercises below have you try the same thing (having threads
+concurrently increment an integer variable) in Python.  As you will
 see, the bug is not easily triggered when you run a Python version of
@@
 
-The following exercises are intended to show you that while it is just
-as easy to write concurrent programs in Python, it is much easier to
-find concurrency bugs using Harmony.
+% The following exercises are intended to show you that while it is just
+% as easy to write concurrent programs in Python, it is much easier to
+% find concurrency bugs using Harmony.
 

### 2021-12-23 book/paper.tex
Added my own NFA->DFA conversion code in the hopes of getting rid of automata library some day

@@
 
-\begin{comment}
-\begin{figure}
-\begin{code}
-<{:input:../code/RWbusy.hny}>
-\end{code}
-\caption{\harmonylink{code/RWbusy.hny} Busy-Waiting Reader/Writer Lock implementation}
-\label{fig:rwbusy}
-\end{figure}
-
-\begin{figure}
-\begin{center}
-\includegraphics[width=2.3in]{figures/rdwr.png}
-\end{center}
-\caption{High-level state diagram specification of reader/writer locks with
-up to two threads.
-The first number in a state gives the number of threads; the second number is the
-number of threads reading in the critical section; the third is the number of
-threads writing in the critical section.}
-\label{fig:rdwr}
-\end{figure}
-\end{comment}
-
 Locks are useful when accessing a shared data structure.  By preventing

### 2021-12-23 src/charm/graph.h

@@ -32,8 +32,8 @@ struct edge {
     uint64_t after;          // resulting context
     int weight;              // 1 if context switch; 0 otherwise
     struct access_info *ai;  // to detect data races
-    uint64_t *log;           // history
-    int nlog;                // size of history
+    uint64_t *log;           // print history
+    int nlog;                // size of print history
 };
 
 enum fail_type {

### 2021-12-23 src/harmony/behavior.py

@@ -1,3 +1,5 @@
+homebrewDFA = False
+
 def read_hfa(file, dfa):
     with open(file) as fd:
         js = json.load(fd)
@@ -20,7 +22,7 @@ def read_hfa(file, dfa):
             transitions[e["src"]][symbol] = e["dst"]
 
     hfa = DFA(
-        states=set(states),
+        states=states,
         input_symbols=symbols,
         transitions=transitions,
         initial_state=initial,
@@ -83,6 +85,22 @@ def behavior_show_diagram(dfa, path=None):
         graph.write_png(path)
     return graph
 
+error_state = frozenset({}) if homebrewDFA else "{}"
+
+def eps_closure_rec(states, transitions, current, output):
+    if current in output:
+        return
+    output.add(current)
+    t = transitions[current]
+    if '' in t:
+        for s in t['']:
+            eps_closure_rec(states, transitions, s, output)
+
+def eps_closure(states, transitions, current):
+    x = set()
+    eps_closure_rec(states, transitions, current, x)
+    return frozenset(x)
+
 def behavior_parse(js, minify, outputfiles, behavior):
     if outputfiles["hfa"] == None and outputfiles["png"] == None and outputfiles["gv"] == None and behavior == None:
         return
@@ -137,20 +155,58 @@ def behavior_parse(js, minify, outputfiles, behavior):
             labels[symbol] = e
 
     # print("states", states, file=sys.stderr)
+    # print("initial", initial_state, file=sys.stderr)
     # print("final", final_states, file=sys.stderr)
     # print("symbols", input_symbols, file=sys.stderr)
     # print("transitions", transitions, file=sys.stderr)
 
     nfa = NFA(
-        states=set(states),
+        states=states,
         input_symbols=input_symbols,
         transitions=transitions,
         initial_state=initial_state,
         final_states=final_states
     )
-
     print("NFA -> DFA", file=sys.stderr)
-    intermediate = DFA.from_nfa(nfa)  # returns an equivalent DFA
+
+    if homebrewDFA:
+        # Compute the epsilon closure for each state
+        eps_closures = { s:eps_closure(states, transitions, s) for s in states }
+
+        # Convert the NFA into a DFA
+        dfa_transitions = {}
+        dfa_initial = eps_closures[initial_state]
+        q = [dfa_initial]       # queue of states to handle
+        while q != []:
+            current = q.pop()
+            if current in dfa_transitions:
+                continue
+            dfa_transitions[current] = {}
[…]

### 2021-12-24 book/paper.tex
Fixes to book

@@
 file <{x}>.\texttt{hvm}.
-The model checker, called \emph{Charm}, places its output in a file
+The model checker, called \emph{Charm}, executes the code in
+<{x}>.\texttt{htm} and places its output in a file
 called <{x}>.\texttt{hco} and a stylized version of the same output
@@
 letters, digits, and underscores can be written without quotes by
-preceding it with a dot.  So \texttt{.example} is the same string
-as \texttt{"example"}.
+preceding it with a dot.  So <{.example}> is the same string
+as <{"example"}>.
 
@@
 A dictionary maps keys to values.
-Unlike Python, any Harmony value can be a key, including another
-dictionary.
+Unlike Python, which requires that keys must be hashable,
+any Harmony value can be a key, including another dictionary.
 Dictionaries are written as
@@
 
-\autoref{ap:values} provides details on all the values that
+\autoref{ap:values} provides details on all the types of values that
 Harmony currently supports.
@@
 
+\index{register}%
+\glsadd{context}%
+\index{thread-local}%
 Instead of bits and bytes, a HVM manipulates Harmony values.
@@
 a stack of Harmony values,
-and two private \emph{registers} that each contain a Harmony value.
-\index{register}%
+and a private \emph{register} that contains a
+Harmony value.\footnote{Currently, another thread register
+contains thread-local data.  We do not use it (yet) in this book.}
 \end{itemize}
 
-\glsadd{context}%
-\index{thread-local}%
-One of the registers of a thread contains the local variables of the
+The register of a thread contains the local variables of the
 method that the thread is currently executing.  It is saved and restored by
-method invocations.  The other register is called \textbf{this} and is
-used for so-called \emph{thread-local state}, which contains variables that
-can be used by all methods.
+method invocations.
 The state of a thread is called a \emph{context} (aka \emph{continuation}):
@@
 it contains the values of its program counter, stack, and registers.
-The state of a HVM
+The HVM state
 consists of the value of its memory and the multiset (or \emph{bag})
@@
 of contexts.  It is a multiset of contexts because two threads can
-have the same context.
+have the same context at the same time.
 
@@
 For example, in \autoref{fig:inc} the memory is a dictionary with two
-entries: \texttt{.count} and \texttt{.done}.  And the value of entry
-\texttt{.done} is a dictionary with keys 0 and 1.
-So, for example, the address of \texttt{done[0]} is the sequence
-[ \texttt{.done}, 0 ].
+entries: <{.count}> and <{done}>.  And the value of entry
+<{done}> is a dictionary with keys 0 and 1.
+So, for example, the address of <{done[0]}> is the sequence <{[done, 0]}>.
 An address is itself a Harmony value.

### 2021-12-24 book/paper.tex
Updated chapter 4

@@
 code/Up.hny:4 def incrementer(self):
-   5 Jump 32
+   5 Jump 35
    6 Frame incrementer self
@@
    10 Store count
-code/Up.hny:6     done[self] = True
-   11 Push ?done
-   12 LoadVar self
-   13 Address
-   14 Push True
-   15 Store
-code/Up.hny:7     await done[1 - self]
-   16 Push 1
-   17 LoadVar self
-   18 2-ary -
-   19 Load done
-   20 Apply
-   21 JumpCond False 16
-code/Up.hny:8     assert count == 2
-   22 ReadonlyInc
-   23 AtomicInc
-   24 Load count
-   25 Push 2
-   26 2-ary ==
-   27 Load count
-   28 Assert2
-   29 AtomicDec
-   30 ReadonlyDec
-   31 Return
 \end{verbatim}
@@
 \index{stack machine}%
-but it also has two registers.
-Like shared memory, the registers usually contain dictionaries so they
+but it also a register.
+Like shared memory, the register contains a dictionary so it
 can represent the values of multiple named variables.
-As mentioned before, one register contains a dictionary with the local
-variables of the method that the thread is currently executing,
-while the other contains thread-local state.
-
+%
 \index{sequential consistency}%
@@
 The last instruction in that sequence is a \texttt{JUMP}
-instruction that sets the program counter to 32
+instruction that sets the program counter to 35
 (skipping over the code for <{incrementer}> method).
-The \texttt{\_\_init\_\_} thread then executes instructions
-32 through 40 and finishes.
+The \texttt{\_\_init\_\_} thread then executes the
+remaining instructions and finishes.
 Once initialization completes, any threads that were spawned
@@
 \begin{verbatim}
-#states 32
+#states 44 (time 0.001+0.001)
 Safety Violation
-T0: __init__() [0-5,32-40] { count: 0, done: [ False, False ] }
-T1: incrementer(0) [ 6- 9] { count: 0, done: [ False, False ] }
-T2: incrementer(1) [ 6-18] { count: 1, done: [ False, True  ] }
-T1: incrementer(0) [10-28] { count: 1, done: [ True,  True  ] }
-Harmony assertion failed: 1
-open file:///Users/rvr/github/harmony/harmony.html for more information
+T0: __init__()     [0-5,35-43]   { count: 0, done: [ False, False ] }
+T2: incrementer(1) [6-9]         { count: 0, done: [ False, False ] }
+T1: incrementer(0) [6-20]        { count: 1, done: [ True,  False ] }
+T2: incrementer(1) [10-24,26-31] { count: 1, done: [ True,  True  ] }
+Harmony assertion failed
+open code/Up.htm for more information
 \end{verbatim}
@@
 at that state, along with their program counters.
-\item \texttt{Behavior Violation}:  The program terminated in a state
-not allowed by the behavioral specification.
+\item \texttt{Behavior Violation}:  The program can terminate in a state
+not allowed by the behavioral specification (\autoref{ch:testing}).
 \item \texttt{Active Busy Waiting}:  There are states in which
-some thread cannot make progress without the help of another thread but
-does not block;
+some thread cannot make progress without the help of another thread,
+but does not block (\autoref{ch:condwait}).
 \item \texttt{Data Race}: There are states in which two or more threads
 concurrently access a shared variable, at least one of which is a store
-operation.
+operation (\autoref{ch:synch}).
 % \item \texttt{Stopped States}: Similar to non-terminating states,
@@
 reported.
-You will learn about behavioral violations in \autoref{ch:testing}.
 \emph{Active busy waiting} (\autoref{ch:condwait}) is not
@@
 indication of an inefficient solution to a synchronization problem---
-one that uses up significantly CPU cycles.
-A \emph{data race} (\autoref{ch:synch}) may not be a bug either---whether or
+one that uses up CPU cycles unnecessarily.
+A \emph{data race} may not be a bug either---whether or
 not it is might depend on the semantics of the underlying memory
@@
 Going back to our example, Harmony reports a safety violation.
-In particular, it reports that the assertion failed
-because <{count}> has value 1 instead of 2.
-Next it reports an execution that failed this assertion.
-The program got to the failed assertion in 4 ``turns.''
+In particular, it reports that an assertion failed.
+The program got to the failed assertion in four ``turns.''
 The output has four columns:
@@
 \begin{center}
-\includegraphics[width=0.9\textwidth]{figures/Up1.pdf}
+\includegraphics[width=0.9\textwidth]{figures/Up1.png}
 \end{center}
@@
 a ``timeline'' containing a block for each instruction executed, the
-program counter of the thread, and the values of the shared variables
-at those turns.
+program counter of the thread, the values of the shared variables
+at those turns, and any values the program may have printed.
 The middle section shows the line of code that is being executed.
@@
 \item Thread \texttt{\_\_init\_\_} (with identifier T0)
-sets shared variable <{count}> to 0 and
-shared variable <{done}> to \texttt{[False, False]}.
-\item Thread <{incrementer(0)}> (with identifier T1)
+executes instructions 0 through 5 and 35 through 43, setting
+shared variable <{count}> to 0 and
+shared variable <{done}> to <{[False, False]}>.
+\item Thread <{incrementer(0)}> (with identifier T2)
 executes instructions 6 through 9, loading the value of
 <{count}> but stopping just before storing 1 into <{count}>;
-\item Thread \texttt{incrementer(1)} (with identifier T2)
-executes instructions 6 through 18, storing 1 into
+\item Thread \texttt{incrementer(1)} (with identifier T1)
+executes instructions 6 through 20, storing 1 into
 <{count}> and storing <{True}> into <{done[1]}>;
-\item Thread <{incrementer(0)}>
-continues execution, executing instructions 10 through 28
+\item Thread <{incrementer(0)}> (T2)
+continues execution, executing instructions 10 through 24
 storing value 1 into <{count}> (instruction 10),
-storing <{True}> into <{done[0]}> (instructions 11 through 15)
-finding that <{done[1]}> is <{True}> (instructions 16 through 21),
-and finally detecting that the assertion is violated
-(instructions 22 through 28).
+storing <{True}> into <{done[0]}>,
+finding that <{done[1]}> is <{True}>,
+and finally detecting that the assertion is violated.
 \end{enumerate}
@@
 The table also lists the program counter of the thread at each turn,
-and the values of the shared variables.
-
+the values of the shared variables, and any values the thread
+may have printed (none in this case).
+%
 Underneath the table it shows the line of Harmony code that is being

### 2021-12-24 book/paper.tex
Made a pass over Chapter 5

@@
 machine instructions.  A counter is a very simple example of
-a data structure, but as we have seen it too requires multiple instructions.
+a data structure (it is an array of bits), but---as we have
+seen---incrementing it requires multiple instructions.
 A more involved one would be accessing a binary tree.
@@
 An implementation of a data structure that can be safely accessed by multiple
-threads and free of race conditions is called \emph{thread-safe}.
+threads and is free of race conditions is called \emph{thread-safe}.
 \index{thread safety}%
@@
 \autoref{fig:csbarebones} shows the Harmony code.
-Here \texttt{cs} is a \emph{label},
+Here <{cs}> is a \emph{label},
 \index{label}%
@@
 \index{mutual exclusion}%
-We would like to place an assertion at the \texttt{cs} label that
+We would like to place an assertion at the <{cs}> label that
 specifies that only one (the current) thread can be there.
@@
 \index{countLabel operator}%
-where $L$ is the name of the label (in this case, \texttt{cs}).
+where $L$ is the name of the label (in this case, <{cs}>).
 The operator returns the number of threads executing at that label.
@@
 This code will surely prevent two or more threads from being
-at label \texttt{cs} at the same time.
+at label <{cs}> at the same time.
 But it does so by preventing \emph{any} thread from reaching
@@
 both choices.)
-
-\begin{comment}
-\begin{figure}
-\begin{center}
-\includegraphics[width=3.5in]{figures/mutex-crop.pdf}
-\end{center}
-\caption{High-level state diagram specification of mutual exclusion with up to two threads.
-The first number in a state gives the number of threads; the second number is the
-number of threads in the critical section.}
-\label{fig:mutex}
-\end{figure}
-
-\autoref{fig:mutex} shows a high-level state diagram
-\index{state diagram}%
-of mutual exclusion and progress.
-The circles represent states, while the arrows represent possible state
-transitions (high-level steps).  The label in each circle summarizes the state;
-the label on an arrow describes the transition.  In this case,
-the label contains two numbers: the total number of threads and the number
-of threads in the critical section.
-\end{comment}
-
-A thread that is in the critical
-section cannot terminate until after leaving the critical section.
+As it turns out, there is an advantage to doing it this way: we can also
+test if a thread can enter when there is no other thread trying to enter
+the critical section.  As we will see below, this is not always obvious.
+
 We will now consider various approaches toward implementing this
@@
 \end{code}
-\caption{\harmonylink{code/naiveLock.hny} Na\"{\i}ve implementation of a shared lock}
+\begin{center}
+\includegraphics[width=0.8\textwidth]{figures/naiveLock.png}
+\end{center}
+\caption{\harmonylink{code/naiveLock.hny} Na\"{\i}ve implementation of a shared lock and the \harmonyref{naiveLock.html}{HTML output} of running Harmony on \autoref{fig:naivelock}}
 \label{fig:naivelock}
@@
 \begin{figure}
+\begin{code}
+<{:input:../code/naiveFlags.hny}>
+\end{code}
 \begin{center}
-\includegraphics[width=0.8\textwidth]{figures/naiveLock.pdf}
+\includegraphics[width=0.8\textwidth]{figures/naiveFlags.png}
 \end{center}
-\caption{The \harmonyref{naiveLock.html}{HTML output of running Harmony on \autoref{fig:naivelock}}}
-\label{fig:naiveLockhtml}
-\end{figure}
-
-\begin{figure}
-\begin{code}
-<{:input:../code/naiveFlags.hny}>
-\end{code}
-\caption{\harmonylink{code/naiveFlags.hny} Na\"{\i}ve use of flags to solve mutual exclusion}
+\caption{\harmonylink{code/naiveFlags.hny} Na\"{\i}ve use of flags to solve mutual exclusion and the \harmonyref{naiveFlags.html}{HTML output} of running Harmony on \autoref{fig:naiveflags}}
 \label{fig:naiveflags}
-\end{figure}
-
-\begin{figure}
-\begin{center}
-\includegraphics[width=0.8\textwidth]{figures/naiveFlags.pdf}
-\end{center}
-\caption{The \harmonyref{naiveFlags.html}{HTML output of running Harmony on \autoref{fig:naiveflags}}}
-\label{fig:naiveflagshtml}
 \end{figure}
@@
 Using a lock is a good thought, but how does one implement one?
-\autoref{fig:naivelock} presents a mutual exclusion attempt based on a
-na\"{\i}ve (and, as it turns out, broken) implementation of a lock.
+\autoref{fig:naivelock} presents an attempt at mutual exclusion based on a
+na\"{\i}ve (and, as it turns out, incorrect) implementation of a lock.
 Initially the lock is not owned, indicated by <{lockTaken}> being <{False}>.
@@
 and then sets it to <{True}> to indicate that the lock has been taken.
-The thread then executes the critical section.  Finally the thread
+The thread then executes the critical section.  Finally, the thread
 releases the lock by setting <{lockTaken}> back to <{False}>.
 
-Unfortunately, if we run the program through Harmony,
-we find that the assertion fails.
-\autoref{fig:naiveLockhtml} shows the Harmony output.
-Thread~0 finds that the lock is available, but just before it stores
-<{True}> in <{lockTaken}> in instruction~9, thread~1 gets
-to run.
-(Recall that you can hover your mouse over a machine instruction in order
-to see what it does.)
-Because <{lockTaken}> is still <{False}>, it too
-believes it can acquire the lock, and stores <{True}> in
-<{lockTaken}> and moves on to the critical section.
-Finally, thread~0 moves on, also stores <{True}>
-into <{lockTaken}> and also moves
-into the critical section.  Thread~0 is the one that detects the
-problem.
-The <{lockTaken}> variable
-suffers from the same sort of race condition as the <{count}> variable
-in \autoref{fig:inc}: testing and setting the lock
-consists of several instructions.  It is thus possible
-for both threads to believe the lock is available and to obtain the lock
-at the same time.
+Unfortunately, if we run the program through Harmony, we find that
+the assertion fails.  \autoref{fig:naivelock} also shows the Harmony
+output.  <{thread(1)}> finds that the lock is available, but just
+before it stores <{True}> in <{lockTaken}>, <{thread(0)}> gets to
+run.  (Recall that you can hover your mouse over a machine instruction
+in order to see what it does.) Because <{lockTaken}> is still
+<{False}>, it too believes it can acquire the lock, and stores
+<{True}> in <{lockTaken}> and moves on to the critical section.
+Finally, <{thread(1)}> moves on, also stores <{True}> into <{lockTaken}>
+and also moves into the critical section.  <{thread(1)}> is the one
+that detects the problem.  The <{lockTaken}> variable suffers from
+the same sort of race condition as the <{count}> variable in
+\autoref{fig:inc}: testing and setting the lock consists of several
+instructions.  It is thus possible for both threads to believe the
+lock is available and to obtain the lock at the same time.
 
@@
 
-However, if you run the program through Harmony (\autoref{fig:naiveflagshtml}),
+However, if you run the program through Harmony,
 it turns out the solution
@@
 Each thread politely lets the other thread have a turn first.
-When <{turn}> = <{i}>, thread~<{i}> can
+When <{turn = i}>, thread~<{i}> can
 enter the critical section, while thread $1-i$ has to wait.
 An invariant of this solution is that while thread~<{i}> is in the critical
-section, $\mathit{turn} = i$.
+section, <{turn = i}>.
 Since <{turn}> cannot be 0 and 1 at
@@
 Unfortunately, exclusivity is not an \emph{inductive} invariant.
-To see why, consider the following state <{s}>: let thread~0 be at label \texttt{cs}
+To see why, consider the following state <{s}>: let thread~0 be at label <{cs}>
 and thread~1 be at the start of the \textbf{await} statement.
@@
 
-First consider the case that thread~0 is at label \texttt{cs}.  If thread~0
+First consider the case that thread~0 is at label <{cs}>.  If thread~0
 were to take a step, then in the next state thread~0 would be no longer
@@
 invariant like $\mathcal{I}_1$?  Invariants must hold for every state.
-For $\mathcal{I}_2$ we only need an assertion at label \texttt{cs} because the
+For $\mathcal{I}_2$ we only need an assertion at label <{cs}> because the
 premise is that there is a thread at that label.  However, we would
@@
 
-\texttt{release\_one()} is the function that a thread uses when leaving
+<{release_one()}> is the function that a thread uses when leaving
 the critical section.  It must check to see if there is a waiting gate
@@
 If so, it selects one and opens that gate.  In the given code,
-\texttt{release\_one()} first checks the readers' gate and then the
+<{release_one()}> first checks the readers' gate and then the
 writers' gate, but the other way around would have worked as well.
-If neither waiting gate qualifies, then \texttt{release\_one()} 
+If neither waiting gate qualifies, then <{release_one()}> 
 has to open the main gate (i.e., release \textit{mutex}).
@@
 \textit{nwriters}, which must be 1 at this time, and thus becomes 0.
-\item It then calls \texttt{release\_one()}.
-\texttt{release\_one()} finds that there are no writers in the critical section
+\item It then calls <{release_one()}>.
+<{release_one()}> finds that there are no writers in the critical section
 and there are two readers waiting.  It therefore releases not
@@
 and increments \textit{nreaders} (from~0 to~1).
-The reader finally calls \texttt{release\_one()}.
-\item Again, \texttt{release\_one()} finds that there are no writers and
+The reader finally calls <{release_one()}>.
+\item Again, <{release_one()}> finds that there are no writers and
 that there are readers waiting, so again it releases the readers' semaphore.
@@
 from~1 to~2.
-\item Finally, the second reader  calls \texttt{release\_one()}.
-This time \texttt{release\_one()} does not find any threads waiting,
+\item Finally, the second reader  calls <{release_one()}>.
+This time <{release_one()}> does not find any threads waiting,
 and so it releases \textit{mutex}.
@@
 % Implement counting semaphores using the split binary semaphores technique.
-\item Several of the instantiations of \texttt{release\_one()} in
+\item Several of the instantiations of <{release_one()}> in
 \autoref{fig:RWsplitsema} can be replaced by simply releasing \textit{mutex}.

### 2021-12-24 book/paper.tex
Started working on Peterson chapter

@@
 
+<{:const:__init__}>
 <{:const:p}>
@@
 (The first line specifies that the \textit{flags} and <{turn}>
-variables should have \emph{sequential consistency}---it prevents
+variables are assumed to satisfy \emph{sequential consistency}---it prevents
 Harmony from complaining about data races involving these variables,
@@
 until either the other thread is nowhere near the critical section
-($\mathit{flag}[1 - \mathit{self}] = \mathbf{False}$) or has given way
-($\mathit{turn} = \mathit{self}$).
+(<{flag[1 - self] = False}>) or has given way (<{turn = self}>).
 Running the algorithm with Harmony shows that it satisfies both mutual
@@
 The HVM starts in an initial state in which there is only
-one thread and its program counter is~0.  A \emph{trace}
+one thread (<{__init__}>()) and its program counter is~0.
+A \emph{trace}
 \index{trace}%

### 2021-12-24 book/paper.tex
Fixed Peterson's chapter

@@
 \index{trace}%
-is a sequence of steps starting from the initial state.
+is a sequence of steps starting from the initial state,
+resulting in a sequence of states.
 When making a step, there are two sources of non-determinism
@@
 But we do not have such a predicate: we know how to define the set
-of reachable states inductively, but given an arbitrary state it is not
+of reachable states inductively, but---given an arbitrary state---it is not
 easy to see whether it is reachable or not.
@@
 Let us begin with considering the following important property:
-$\mathcal{F}(i) = \mathtt{thread}(i)@[8 \cdots 15] \Rightarrow \mathit{flags}[i]$,
-that is, if thread~<{i}> is executing in lines 8 through 15, then $\mathit{flags}[i]$
+$\mathcal{F}(i) = \mathtt{thread}(i)@[10 \cdots 17] \Rightarrow \mathit{flags}[i]$,
+that is, if thread~<{i}> is executing in lines 10 through 17, then $\mathit{flags}[i]$
 is set.
@@
 \item states in which $\mathit{flags}[i] = \texttt{true}$
-\item states in which $\lnot \mathtt{thread}(i)@[8 \cdots 15]$
+\item states in which $\lnot \mathtt{thread}(i)@[10 \cdots 17]$
 (because false implies anything)
 \end{enumerate}
-In each case we need to show that if thread <{i}> takes a step, 
+In each case, we need to show that if thread <{i}> takes a step, then
 $\mathcal{F}(i)$ still holds.
 In the first case, there is only one step that thread <{i}> can take that would
-set $\mathit{flags}[i]$ to false: the step in line 15.  But executing the line
-would also take the thread out of lines $8 \cdots 15$, so $\mathcal{F}(i)$ continues to hold.
-In the second case (thread <{i}> is not executing in lines $8 \cdots 15$), the only step
-that would cause thread <{i}> to execute in lines $8 \cdots 15$ would be the step in line 7.
+set $\mathit{flags}[i]$ to false: the step from line 17 to line 18.  But executing the line
+would also take the thread out of lines $10 \cdots 17$, so $\mathcal{F}(i)$ continues to hold.
+In the second case (thread~<{i}> is not executing in lines $10 \cdots 17$), the only step
+that would cause thread~<{i}> to execute in lines $10 \cdots 17$ would be the step in line 9.
 But in that case $\mathit{flags}[i]$ would end up being true, so
@@
 but now you know how to prove it.
+We will use a similar technique to prove the exclusivity is invariant.
 
@@
 However, because thread $1-i$ is executing concurrently, this property does not
-hold.  In particular, suppose thread 0 is at the critical section, $\mathit{flags}[0]$ = true,
-$\mathit{turn} = 1$, and thread 1 just finished the step in line 7,
-setting $\mathit{flags}[1]$ to true.  Notice that $C(0)$ is now violated.
+hold.
+You can use Harmony to verify this.  Just place the following command
+in the critical section of the program:
+
+\begin{code}
+<{assert (not flags[1 - self]) or (turn == self)}>
+\end{code}
 
@@
 
-Instead, we will use the following property: $\mathcal{G}(i) =
-\mathtt{thread}(i)@\mathtt{cs} \Rightarrow \mathcal{C}(i) \lor \mathtt{thread}(1-i)@\mathtt{gate}$.
+When running Harmony, this assertion will fail.
+You can check the HTML output to see what happened.
+Suppose thread 0 is at the critical section, $\mathit{flags}[0]$ = true,
+$\mathit{turn} = 1$, and thread 1 just finished the step in line 7,
+setting $\mathit{flags}[1]$ to true.  Then $C(0)$ is violated.
+But it suggests a new property:
+$\mathcal{G}(i) =
+\mathtt{thread}(i)@\mathtt{cs} \Rightarrow \mathcal{C}(i) \lor \mathtt{thread}(1-i)@10$.
 That is, if thread <{i}> is at the critical section, then
-either $\mathcal{C}(i)$ holds or thread $1-i$ is about to execute line 8.
+either $\mathcal{C}(i)$ holds or thread $1-i$ is about to execute line 10.
 \autoref{fig:petersonproof} formalizes $\mathcal{G}(i)$ in Harmony.
-The label \texttt{gate} refers to the step that sets <{turn}> to $1-i$.
+The label \texttt{gate} refers to line 10, that is,
+the step that sets <{turn}> to $1-i$.
 You can run \autoref{fig:petersonproof} to determine
@@
 $\mathcal{G}$ without the disjunct that thread $1-i$ is at label \texttt{gate}).
-To demonstrate that this predicate is not invariant, you can remove the
-disjunct from \autoref{fig:petersonproof} and run it to get a
-counterexample.
+We already saw that this is not an invariant.
+(The author fixed the Wikipedia page with the help of Fred B.~Schneider.)
 
@@
 predicate is not invariant, and therefore not an inductive invariant either.
-(The author fixed the Wikipedia page with the help of Fred B.~Schneider.)
+Moreover, analyzing the counterexample provided by Harmony may well
+suggest how to fix the predicate.
 

### 2021-12-24 book/paper.tex
Proofread through chapter 7

@@
 A method <{m}> with argument <{a}> is invoked in its
-most basic form as follows (assigning the result to~\texttt{r}).
+most basic form as follows (assigning the result to~<{r}>).
 \begin{code}
@@
 error-prone.  It's too easy to forget to, say, release a lock when returning a value in the
-middle of a method, a major source of bugs in practice.
-
-Harmony is not an object-oriented language like Python is.  In Python
+middle of a method---a major source of bugs in practice.
+
+Harmony is not an object-oriented language like Python is.  In Python,
 you can pass a reference to an object to a method, and that method
@@
 explicitly bound within a scope.
-Method variables are either mutable or immutable.
+Method variables are either mutable (writable) or immutable (read-only).
 The arguments to a method and the bound variable (or
@@
 replacement algorithms.
-In 1969, B\'{e}l\'{a}ady et al. published a paper~\cite{Belady69}
+In 1969, B\'{e}l\'{a}dy et al.~published a paper~\cite{Belady69}
 that showed that making a
@@
 The program in \autoref{fig:clock} will find exactly the same scenario
-if you define \texttt{FIFO} to be <{True}>.  Moreover, if you define
-\texttt{FIFO} to be <{False}>, it will find a scenario for the
+if you define <{FIFO}> to be <{True}>.  Moreover, if you define
+<{FIFO}> to be <{False}>, it will find a scenario for the
 CLOCK replacement policy~\cite{clock}, often used in modern operating systems.
 
-In this program, \texttt{CLOCK} maintains the state of a cache
+In this program, <{CLOCK}> maintains the state of a cache
 (in practice, typically pages in memory).
-The set \texttt{recent} maintains whether an access to the cache
+The set <{recent}> maintains whether an access to the cache
 for a particular reference was recent or not.
-(It is not used if \texttt{FIFO} is <{True}>.)
-The integer \texttt{misses} maintains the number of cache misses.
-Method \texttt{ref}(\textit{ck}, <{x}>) is invoked when <{x}> is
-referenced and checked against the cache \textit{ck}.
-
-The program declares two caches: one with 3 entries (\textit{clock3})
-and one with 4 entries (\textit{clock4}).
+(It is not used if <{FIFO}> is <{True}>.)
+The integer <{misses}> maintains the number of cache misses.
+Method <{ref(ck, x)}> is invoked when <{x}> is
+referenced and checked against the cache <{ck}>.
+
+The program declares two caches: one with 3 entries (<{clock3}>)
+and one with 4 entries (<{clock4}>).
 The interesting part is in the last block of code.
@@
 Further reducing the search space, the program never repeats the
-same reference twice in a row (using the local variable \texttt{last}).
+same reference twice in a row (using the local variable <{last}>).
 
 The two things to note here is the use of the \textbf{choose}
-expression and the \textbf{assert} statement.  Using <{choose}>
+expression and the \textbf{assert} statement.  Using <{choose}>,
 we are able to express searching through all possible strings of
 references without a complicated nested iteration.
-Using \texttt{assert} we are able to express the anomaly we are
+Using \texttt{assert}, we are able to express the anomaly we are
 looking for.
 
-In case you want to check if you get the right results.  For \texttt{FIFO},
-the program finds (in a few minutes) the same anomaly that
-B\'{e}l\'{a}dy et al. found: 1 2 3 4 1 2 5 1 2 3 4 5.
-For the \texttt{CLOCK} algorithm the program actually finds a shorter
+In case you want to check if you get the right results.  For <{FIFO}>,
+the program finds the same anomaly that
+B\'{e}l\'{a}dy et al.~found: 1 2 3 4 1 2 5 1 2 3 4 5.
+For the <{CLOCK}> algorithm the program actually finds a shorter
 reference string: 1 2 3 4 2 1 2 5 1 2.

### 2021-12-24 book/paper.tex
Proofread through Chapter 8

@@
 
-So far we have used Harmony to \emph{implement} various algorithms.
+So far, we have used Harmony to \emph{implement} various algorithms.
 But Harmony can also be used to \emph{specify} what an algorithm is
@@
 methods such as \texttt{acquire} and \texttt{release}.  But
-an actual executable program such as in \autoref{fig:cssynch} should not
-use the <{atomically}> keyword.
-In fact, if we want to make the entire program executable on hardware,
-we have to also show how \texttt{Lock}, \texttt{acquire}, and \texttt{release}
+an actual executable program---such as the one in
+\autoref{fig:cssynch}---should not use the <{atomically}> keyword
+because---on a normal machine---it cannot be directly compiled into
+machine code.
+If we want to make the program executable on hardware,
+we have to show how \texttt{Lock}, \texttt{acquire}, and \texttt{release}
 are implemented, not just how they are specified.
@@
 atomic statements but use the synchronization primitives that are available
-to you.  But if you want to \emph{specify} a new synchronization primitive,
+to you.  But if you want to \emph{specify} a synchronization primitive,
 then use \textbf{atomically} by all means.

### 2021-12-24 book/paper.tex
Latest

@@
 operations on independent memory locations.  However, as \autoref{ex:reorder}
-showed, Peterson's algorithm breaks down if operations are re-ordered.
+showed, Peterson's algorithm breaks down if such seemingly independent
+operations are re-ordered.
 Some memory caches can also cause non-atomic behavior of memory when shared
@@
 variable more than once, because how could it possibly change if there
-are no store operations?
+are no store operations in between?
 \end{itemize}
@@
 was borrowed from C and C++.
-However, in C and C++ they do \emph{not} provide sequential
+However, in C and C++, they do \emph{not} provide sequential
 consistency, and one cannot implement Peterson's algorithm in

### 2021-12-24 book/paper.tex
Pass over Chapter 9 finished

@@
 verifies that the value of <{count}> equals~2.
-% Like Python, Harmony supports \textbf{assert} statements with two arguments:
-% the first is a Boolean condition, and the second is a value that is
-% reported if the condition fails, in this case
-% the value of <{count}>.
 
@@
 Method \texttt{test\_and\_set} in \autoref{fig:tas} shows its specification.
-Here <{s}> points to a shared Boolean variable and <{p}>
-to a private Boolean variable, belonging to some thread.
+Here <{s}> points to a shared boolean variable and <{p}>
+to a private boolean variable, belonging to some thread.
 The operation copies the value of the shared variable to the
@@
 it can acquire the lock.
-The program uses $\mathtt{N}+1$ variables.
+The program uses $\mathtt{N}+1$ boolean variables.
 Variable <{shared}> is initialized to <{False}> while
@@
 one of these variables is <{False}>.
-Another invariant, $\mathcal{I}_2(i)$, is that if thread <{i}> is executing between
-lines $18 \cdots 22$ (which includes the critical section),
+Another invariant, $\mathcal{I}_2(i)$, is that if thread <{i}> is
+in the critical section,
 then \textit{private}[<{i}>] = <{False}>.
@@
 critical section at the same time.
-For those who read through \autoref{ch:peterson}
-$\mathcal{I}_1 \land \forall i: \mathcal{I}_2(i)$ is an inductive invariant.
+% For those who read through \autoref{ch:peterson}
+% $\mathcal{I}_1 \land \forall i: \mathcal{I}_2(i)$ is an inductive invariant.
 
@@
 
-The Boolean type has only two possible values: <{False}> and
+The boolean type has only two possible values: <{False}> and
 <{True}>.  Unlike in Python, in Harmony booleans are distinct

### 2021-12-24 book/paper.tex
Finished pass over Chapter 10

@@
 The code in \autoref{fig:lockspec} is similar to the code in
-Harmony's \texttt{synch} module.
+Harmony's <{synch}> module.
 (The module generalizes locks to \emph{binary semaphores}
@@
 
-In \autoref{fig:tas} we have shown a lock implementation based on a shared
+\autoref{fig:tas} shows a lock implementation based on a shared
 variable and a private variable for each thread.   The private
@@
 The \texttt{-m} flag tells Harmony to use the \texttt{taslock.hny} file
-rather than the standard \texttt{synch} module (which only contains
-a specification of the lock methods).
+for the <{synch}> module rather than the standard <{synch}> module
+(which contains only a specification of the lock methods).
 
@@
 with just two threads: one thread might successfully acquire the lock
-repeatedly in a loop, while another thread is never lucky enough the
-acquire the thread in between.
+repeatedly in a loop, while another thread is never lucky enough to
+acquire the lock in between.
 % \autoref{ch:starvation} will talk more about starvation and how
@@
 lock that prevents starvation using an atomic \emph{fetch-and-increment}
-operator.  It is inspired by European bakeries.  A European bakery has a
+operator.  It is inspired by European bakeries.  A European bakery often has a
 clearly displayed counter (usually just two digits) and a ticket dispenser.
@@
 checks every possible state, limiting \texttt{MAX\_THREADS}
-to a small number such as done
-here will significantly reduce the time to model check a Harmony program.
+to a small number
+significantly reduces the time to model check a Harmony program.
 Plus it is easier to check that it fails when you run it with more than
@@
 shared variable.  Recall that the context of a thread contains
-its program counter, stack, and registers.
+its program counter, stack, and register (containing the current
+method's variables).
 A context is a regular (if complex) Harmony value.
@@
 You will find that using the \emph{implementation} of a lock instead
-of the \emph{specification} of a lock (in the \texttt{synch} module)
+of the \emph{specification} of a lock (in the <{synch}> module)
 often leads to the model checker searching a
@@
 \item
-Run \autoref{fig:incfixed} using (i) \texttt{synch} and (ii) \texttt{synchS}.
+Run \autoref{fig:incfixed} using (i) <{synch}> and (ii) \texttt{synchS}.
 Report how many states were explored by Harmony for each module.
@@
 \item \label{ex:trylock} Implement \texttt{tryAcquire}(\textit{b}) as an additional
-interface for both the \texttt{synch} and \texttt{synchS} modules.
+interface for both the <{synch}> and \texttt{synchS} modules.
 This interface is like \texttt{acquire}(\textit{b}) but never blocks.  It
-returns <{True}> if the binary semaphore was available (and now acquired) or <{False}>
-if the binary semaphore was already acquired.
+returns <{True}> if the lock was available (and now acquired) or <{False}>
+if the lock was already acquired.
 Hint: you do not have to change the existing code.
@@
 \caption{\harmonylink{modules/synch.hny}
-Implementation of condition variables in the \texttt{synch} module}
+Implementation of condition variables in the <{synch}> module}
 \label{fig:cv}
@@
 Harmony provides support for Mesa monitors
-through the Harmony \texttt{synch} module.
+through the Harmony <{synch}> module.
 \autoref{fig:cv} shows the implementation of
-condition variables in the \texttt{synch} module.
+condition variables in the <{synch}> module.
 \texttt{Condition}() creates a new condition variable.
@@
 four.  A counting semaphore is like a binary semaphore, but can be
-acquired a given number of times.  It is supported by the \texttt{synch}
+acquired a given number of times.  It is supported by the <{synch}>
 module.  The \texttt{P} or ``procure'' operation acquires a counting
@@
 
-The \texttt{synch} library supports a synchronized message queue,
+The <{synch}> library supports a synchronized message queue,
 similar to the \texttt{Queue} object in Python.
@@
 It is useful to create an abstract network that reliably sends messages
-between threads, much like the FIFO queue in the \texttt{synch} module.
+between threads, much like the FIFO queue in the <{synch}> module.
 For this, we need a network protocol that Alice and Bob can run.
@@
 
-\section{The \texttt{synch} module}
+\section{The <{synch}> module}
 
 \index{synch module}%
-The \texttt{synch} module provides the following methods:
+The <{synch}> module provides the following methods:
 

### 2021-12-24 book/paper.tex
Made a pass over Chapter 11

@@
 \item \autoref{fig:incmany} is a version of \autoref{fig:incpy} that has each
-incrementer thread increment <{count}> \texttt{N} times.  Run \autoref{fig:incmany}
+incrementer thread increment <{count}> <{N}> times.  Run \autoref{fig:incmany}
 10 times (using Python).
@@
 was for each of the failed runs.
-Also experiment with lower values of \texttt{N}.
-How large does \texttt{N} need to be for assertions to fail?
-(Try powers of 10 for \texttt{N}.)
+Also experiment with lower values of <{N}>.
+How large does <{N}> need to be for assertions to fail?
+(Try powers of 10 for <{N}>.)
 \item Can you think of a fix to \autoref{fig:inc}?  Try one or two different fixes
@@
 \end{code}
-\caption{\harmonylink{python/UpMany.py} Incrementing \texttt{N} times}
+\caption{\harmonylink{python/UpMany.py} Incrementing <{N}> times}
 \label{fig:incmany}
@@
 \autoref{fig:tas} goes on to implement mutual exclusion for
-a set of \texttt{N} threads.
+a set of <{N}> threads.
 The approach is called \emph{spinlock},
@@
 checks to see if any threads are waiting to acquire the lock.
-If so, it uses the \texttt{head} and \texttt{tail}
+If so, it uses the <{head}> and <{tail}>
 methods from the \texttt{list} module (see \autoref{ap:module})
@@
 \\
-(a) \harmonylink{code/queuespec.hny} Sequential & (b) \harmonylink{code/queue.hny} Concurrent queue
+(a) \harmonylink{code/queuespec.hny} Sequential & (b) \harmonylink{code/queue.hny} Concurrent
 \end{tabular}
@@
 By way of example, we will first demonstrate how to build a concurrent queue.
-The \texttt{queue} module has the following API:
+The \texttt{queue} module can be used as follows:
 \begin{itemize}
-\item $q = \mathtt{Queue}()$: allocate a new queue;
-\item $\mathtt{put}(q, v)$: add <{v}> to the tail of the queue;
-\item $r = \mathtt{get}(q)$: returns $r = \mathtt{None}$
-if <{q}> is empty or $r = v$ if <{v}> was at the head of the queue.
+\item <{x = Queue()}>: initialize a new queue <{x}>;
+\item <{put(?x, v)}>: add <{v}> to the tail of <{x}>;
+\item <{r = get(?x)}>: returns <{r = None}>
+if <{x}> is empty or <{r = v}> if <{v}> was at the head of <{x}>.
 \end{itemize}
@@
 which should be released with \texttt{free}() when it is no longer in use.
-The queue maintains a \texttt{head} pointer to the first element in the list
-and a \texttt{tail} pointer to the last element in the list.
-The \texttt{head} pointer is <{None}> if and only if the queue is empty.
+The queue maintains a <{head}> pointer to the first element in the list
+and a <{tail}> pointer to the last element in the list.
+The <{head}> pointer is <{None}> if and only if the queue is empty.
 (<{None}> is a special address value that is not the address of any
@@
 
-\texttt{Queue()} returns a new queue object consisting of a <{None}> head
-and tail pointer and a lock.
-The $\mathtt{put}(q, v)$ and $\mathtt{get}(q)$ methods both take a pointer <{q}> to the
+\texttt{Queue()} returns the initial value for a queue object
+consisting of a <{None}> head and tail pointer and a lock.
+The <{put(q, v)}> and <{get(q)}> methods both take a pointer <{q}> to the
 queue object because both may modify the queue.
@@
 are not allowed to change the state.
-In general, when calling methods in \textbf{assert} or
-\textbf{invariant} statements, one has to be convinced that
-those methods cannot change the state in any way.
 
@@
 said explicitly in the paper.
-As a result, the algorithm is not guaranteed to work
-correctly with most modern programming languages and
-computer hardware.  But it is still useful to study it.
+As a result, the algorithm is must be coded very carefully to work
+correctly with modern programming languages and
+computer hardware.
 The implementation uses separate
@@
 Except initially, the dummy node is the last node that was dequeued.
-Note that neither the \texttt{head} nor \texttt{tail} pointer are
+Note that neither the <{head}> nor <{tail}> pointer are
 ever <{None}>.
@@
 They obtain separate locks and then concurrently access the
-\texttt{next} field in the dummy node---a data race with
+<{next}> field in the dummy node---a data race with
 undefined semantics in most environments.
@@
 
+<{:const:contains}>
+<{:const:check}>
+<{:const:length}>
+
 \section*{Exercises}
 \begin{problems}
-\item \label{ex:qcontains} Add a method $\mathtt{contains}(q, v)$ to \autoref{fig:queues}(b)
-that checks to see if <{v}> is in queue <{q}>.
-\item Add a method $\mathtt{length}(q)$ to \autoref{fig:queueconc} that returns the length of the given queue.
+\item \label{ex:qcontains} Add a method <{contains(q, v)}> to
+\autoref{fig:queues}(b) that checks to see if <{v}> is in queue <{q}>.
+\item Add a method <{length(q)}> to \autoref{fig:queueconc} that returns the length of the given queue.
 The complexity of the method should be $O(1)$, which is to say that you should maintain
 the length of the queue as a field member and update it in <{put}> and <{get}>.
-\item \label{ex:qlength} Write a method $\mathtt{check}(q)$
+\item \label{ex:qlength} Write a method <{check(q)}>
 that checks the integrity of the queue in \autoref{fig:queueconc}.
@@
 just before releasing the lock in <{put}> and <{get}>:
-<{:const:check}>
 \begin{code}
@@
 that removes all occurrences of <{v}>, if any, from queue <{q}>.
-\item The test program in \autoref{fig:queuedemo} is a not thorough test program.
+\item The test program in \autoref{fig:queuedemo} is not a thorough test program.
 Design and implement a test program for \autoref{fig:queuedemo}.
 Make sure you \emph{test} the test program by trying it out against some buggy
-queue implementations.
+queue implementations.  (You will learn more about testing concurrent programs in \autoref{ch:testing}.)
 \end{problems}
@@
 and does little to increase our confidence in its correctness.
-To wit, if \texttt{get()} always returned 1, the program
+To wit, if <{get()}> always returned 1, the program
 would find no problems.
@@
 For example, a common programming mistake is to forget to
-update the \texttt{tail} pointer in \texttt{get()} in case
+update the <{tail}> pointer in <{get()}> in case
 the queue becomes empty.  Normally, it is a surprisingly
@@
 checks concurrent <{put}> operations using
-\texttt{N} threads.  Thread \texttt{putter(<{v}>)}
+<{N}> threads.  Thread \texttt{putter(<{v}>)}
 adds <{v}> to the queue.
 Thread \texttt{main()} waits until all \texttt{putter}
-threads have finished, and then dequeues \texttt{N}
+threads have finished, and then dequeues <{N}>
 items, verifying that they form the set
@@
 \end{code}
-\caption{\harmonylink{code/qtest2.hny} Concurrent \texttt{get()}}
+\caption{\harmonylink{code/qtest2.hny} Concurrent <{get()}>}
 \label{fig:qtest2}
@@
 checks concurrent <{get}> operations using
-\texttt{N} threads.
+<{N}> threads.
 The queue is initialized with the sequence
@@
 same entry.
-Thread \texttt{main()} waits until all \texttt{getter}
+Thread \texttt{main()} waits until all <{getter}>
 threads have finished and checks to make sure that
@@
 \caption{\harmonylink{code/qtest3.hny} Mix of concurrent
-\texttt{get()} and \texttt{put()} operations}
+<{get()}> and \texttt{put()} operations}
 \label{fig:qtest3}
@@
 concurrent <{get}> and <{put}> operations.
-It starts \texttt{N} \texttt{getter} threads and
-\texttt{N} \texttt{putter} threads.
-Thread \texttt{main} waits until all have finished,
+It starts <{N}> <{getter}> threads and
+<{N}> \texttt{putter} threads.
+Thread <{main}> waits until all have finished,
 and then checks to make sure that whatever is left
 on the queue was not also consumed by some
-\texttt{getter} thread.
+<{getter}> thread.
 Finally, it makes sure that the queue is empty
@@
 a concurrent data structure.
-The \texttt{get()} method of the queue implementation
+The <{get()}> method of the queue implementation
 uses a local variable <{node}>, declared using
@@
 <{put}>(1) (see the output column for that clue: the thread printed that).
-We can see that <{q}>.\texttt{next} now points to \texttt{alloc}.\textit{pool}[0],
+We can see that <{q}>.<{next}> now points to \texttt{alloc}.\textit{pool}[0],
 which the thread must have allocated.  The contents is 
@@
 \item \label{ex:boundlock} Using busy waiting, implement a ``bound lock'' that allows
-up to \texttt{M} threads to acquire it at the same time.\footnote{A bound lock is a restricted version of a \emph{counting} semaphore.}
+up to <{M}> threads to acquire it at the same time.\footnote{A bound lock is a restricted version of a \emph{counting} semaphore.}
 A bound lock
 with \texttt{M = 1} is an ordinary lock.
-You should define a constant \texttt{M} and two methods:
+You should define a constant <{M}> and two methods:
 \texttt{acquire\_bound\_lock()}
@@
 \item Write a test program for your bound lock
-that checks that no more than \texttt{M} threads can acquire the
+that checks that no more than <{M}> threads can acquire the
 bound lock.
 \item Write a test program for bound locks
-that checks that up to \texttt{M} threads
+that checks that up to <{M}> threads
 can acquire the bound lock at the same time.
@@
 \autoref{fig:gpu}.
-There are \texttt{N} GPUs identified by the numbers
-1 through \texttt{N}.  Method \texttt{gpuAlloc()} returns the identifier
+There are <{N}> GPUs identified by the numbers
+1 through <{N}>.  Method \texttt{gpuAlloc()} returns the identifier
 of an available GPU, blocking if there is currently no available GPU.
@@
 <{:const:sort}>
+<{:const:main}>
 
@@
 algorithm, and \autoref{fig:qsorttest} an accompanying test program.
-The array to be sorted is stored in shared variable \textit{textqs}.\texttt{arr}.
-Another shared variable, \textit{testqs}.\texttt{todo}, contains the ranges of the
+The array to be sorted is stored in shared variable <{textqs.arr}>.
+Another shared variable, <{testqs.todo}>, contains the ranges of the
 array that need to be sorted (initially the entire array).
 Re-using as much of this code as you can, implement a parallel version of
-this.  You should not have to change the methods \texttt{swap}, \texttt{partition},
-or \texttt{sortrange} for this.  Create \texttt{NWORKERS} ``worker threads''
-that should replace the \texttt{qsort} code.
-Each worker loops until \textit{todo}
-is empty and sorts the ranges that it finds until then.  The \texttt{main}
+this.  You should not have to change the methods <{swap}>, <{partition}>,
+or <{sortrange}> for this.  Create <{NWORKERS}> ``worker threads''
+that should replace the <{qsort}> code.
+Each worker loops until <{todo}>
+is empty and sorts the ranges that it finds until then.  The <{main}>
 thread needs to wait until all workers are done.
@@
 \item Do you see what the problem is?
-\item Does it depend on \texttt{N}, the number of philosophers?
+\item Does it depend on <{N}>, the number of philosophers?
 \item Does it matter in what order the philosophers lay down their forks?
@@
 <{:const:transfer}>
+<{:const:total}>
 
@@
 threads.  Can you fix the problem?
-\item Add a method \texttt{total()} to the solution of the previous question
+\item Add a method <{total()}> to the solution of the previous question
 that computes the total over all balances.
@@
 \begin{enumerate}
-\item an actor that generates a sequence of integers 1 through \texttt{N};
+\item an actor that generates a sequence of integers 1 through <{N}>;
 \item an actor that receives integers and forwards only those that are prime;
@@
 \autoref{fig:barrier} shows an example.  There is a queue
-for each of the \texttt{N} threads.
+for each of the <{N}> threads.
 Before thread <{i}> enters a round, it first sends a message to every
@@
 \begin{problems}
-\item Implement barrier synchronization for \texttt{N} threads
+\item Implement barrier synchronization for <{N}> threads
 with just three binary semaphores.  Busy waiting is not allowed.
@@
 solution.)
-\item Imagine a pool hall with \texttt{N} tables.  A table is \emph{full}
+\item Imagine a pool hall with <{N}> tables.  A table is \emph{full}
 from the time there are two players until both players have left.
@@
 a client broadcasts a request to all servers and waits for responses
-from a majority ($\texttt{N} - \texttt{F}$ or equivalently
+from a majority ($<{N}> - \texttt{F}$ or equivalently
 $\texttt{F} + 1$ servers).  Note that because we are assuming that no
@@
 constant \texttt{F} contains the maximum number of failures, and
-we will assume there are $\texttt{N} = 3\texttt{F} + 1$ processors.
+we will assume there are $<{N}> = 3\texttt{F} + 1$ processors.
 Each processor \emph{proposes} a value, which we assume here to be from the
@@
 The $\mathtt{N}$ processors go through a sequence of \emph{rounds} in which they wait for
-$\texttt{N} - \texttt{F}$ messages, update their state based on the messages, and
+$<{N}> - \texttt{F}$ messages, update their state based on the messages, and
 broadcast messages containing their new state.
-The reason that a processor waits for $\texttt{N} - \texttt{F}$ rather than \texttt{N}
+The reason that a processor waits for $<{N}> - \texttt{F}$ rather than <{N}>
 messages is because of failures: up to $\texttt{F}$ processors may never send a message
-and so it would be unwise to wait for all $\texttt{N}$.  You might be tempted to use
+and so it would be unwise to wait for all $<{N}>$.  You might be tempted to use
 a timer and time out on waiting for a particular processor.  But how would you initialize
@@
 that messages arrive within a particular time.
-We call a set of $\texttt{N} - \texttt{F}$ processors a \emph{quorum}.
+We call a set of $<{N}> - \texttt{F}$ processors a \emph{quorum}.
 A quorum must suffice for the algorithm to make progress.
@@
 
-In Line~22, a processor waits for $\texttt{N} - \texttt{F}$ messages using
+In Line~22, a processor waits for $<{N}> - \texttt{F}$ messages using
 the Harmony \textbf{atomically when exists} statement.
@@
 \texttt{receive}(\textit{round}, <{k}>) method returns all \emph{subbags} of messages
-for the given round that have size $k = \texttt{N} - \texttt{F}$.  The method uses a
+for the given round that have size $k = <{N}> - \texttt{F}$.  The method uses a
 dictionary comprehension to filter out all messages for the given \textit{round}
@@
 The body of the \textbf{atomically when exists} statement contains the core of the algorithm.
-Note that $\texttt{N} - \texttt{F} = 2\texttt{F} + 1$, so that the number of
+Note that $<{N}> - \texttt{F} = 2\texttt{F} + 1$, so that the number of
 messages is guaranteed to be odd.  Also, because there are only 0 and 1 values, there
@@
 So one can simplify the model and have each processor wait
-for \emph{all} \texttt{N} messages in a round
+for \emph{all} <{N}> messages in a round
 instead of $\mathtt{N} - \mathtt{F}$.
 It would still have to choose to consider just $\mathtt{N} - \mathtt{F}$
-out of those \texttt{N} messages, but executions in which some processors
+out of those <{N}> messages, but executions in which some processors
 are left behind in all rounds are no longer considered.
@@
 containing the ballot number and a proposal and then waits for
-$\texttt{N} - \texttt{F}$ matching \texttt{2.B} responses from the
+$<{N}> - \texttt{F}$ matching \texttt{2.B} responses from the
 acceptors.  If each response contains the ballot number and the
@@
 
-The expression \textbf{const} \texttt{N = 3} introduces a new constant
-\texttt{N} with the value 3.  Evaluating \texttt{N} does not lead to
+The expression <{const N = 3}> introduces a new constant
+<{N}> with the value 3.  Evaluating <{N}> does not lead to
 loading from a memory location.  The assignment can be overridden with
 the \texttt{-c} flag: \texttt{harmony -cN=4} executes the model checker
-with 4 assigned to \texttt{N} instead of 3.
-Harmony also supports \textbf{const} \texttt{N, M = 3, 4}, which
-assigns 3 to \texttt{N} and 4 to \texttt{M}.
+with 4 assigned to <{N}> instead of 3.
+Harmony also supports <{const N, M = 3, 4}>, which
+assigns 3 to <{N}> and 4 to <{M}>.
 Harmony has limited support for \emph{constant folding}.  For example,
-\textbf{const} \texttt{N = 3 + 4} assigns value 7 to constant \texttt{N}.
+<{const N = 3 + 4}> assigns value 7 to constant <{N}>.
 
@@
 to those used in <{let}> and <{for}> statements.
-Examples include <{()}>, $(x,)$, $(x, y)$, and $(x, (y, z))$.
+Examples include <{()}>, <{(x,)}>, <{(x, y)}>, and <{(x, (y, z))}>.
 The given local variable names are assigned upon application and

### 2021-12-24 book/paper.tex
Latest
Pass over chapter 12

@@
 said explicitly in the paper.
-As a result, the algorithm is must be coded very carefully to work
+As a result, the algorithm must be coded very carefully to work
 correctly with modern programming languages and
@@
 \begin{code}
+<{:input:../code/intsettest.hny}>
+\end{code}
+\caption{\harmonylink{code/intsettest.hny} Test code for set objects}
+\label{fig:intsettest}
+\end{figure}
+
+\begin{figure}
+\begin{code}
 <{:input:../code/linkedlist.hny}>
@@
 \label{fig:linkedlist}
-\end{figure}
-
-\begin{figure}
-\begin{code}
-<{:input:../code/intsettest.hny}>
-\end{code}
-\caption{\harmonylink{code/intsettest.hny} Test code for \autoref{fig:linkedlist}}
-\label{fig:intsettest}
 \end{figure}
@@
 \texttt{SetObject()} returns a pointer to a variable that
-contains an empty set, rather than returning an empty set.
-Values can be added using \texttt{insert()} or deleted using \texttt{remove()}.
-Method \texttt{contains()} checks if a particular value is in the list.
-
-\autoref{fig:linkedlist} implements a concurrent set of integers using
+contains an empty set, rather than returning an empty set \emph{value}.
+As such, it is more like an object in an object-oriented language than
+like a value in its own right.
+Values can be added to the set object using
+<{insert()}> or deleted using <{remove()}>.
+Method <{contains()}> checks if a particular value is in the list.
+\autoref{fig:intsettest} contains a simple (although not very thorough)
+test program to demonstrate the use of set objects.
+
+\autoref{fig:linkedlist} implements a concurrent set object using
 an ordered linked list of integers without duplicates.
-(\autoref{fig:intsettest} contains test code.)
-The list has two dummy ``book-end'' nodes with values \texttt{-inf} and
-\texttt{inf} (similar to the Python \texttt{math.inf} constant).
+The list has two dummy ``book-end'' nodes with values <{-inf}> and
+<{inf}> (similar to the Python \texttt{math.inf} constant).
 An invariant of the algorithm is that at any point in time the
-list is ``valid,'' starting with a \texttt{-inf} node and ending
-with a \texttt{inf} node.
+list is ``valid,'' starting with a <{-inf}> node and ending
+with an <{inf}> node.
 
 \index{hand-over-hand locking}%
-Each node has a lock, a value, and \textit{next}, a pointer to the next node
-(which is <{None}> for the final \texttt{inf} node).
-The \texttt{\_find(\textit{lst}, <{v}>)} helper method first
-finds and locks two consecutive nodes <{before}> and \textit{after}
+Each node has a lock, a value, and <{next}>, a pointer to the next node
+(which is <{None}> for the <{inf}> node).
+The <{_find(lst, v)}> helper method first
+finds and locks two consecutive nodes <{before}> and <{after}>
 such that <{before->data.value < v <= after->data.value}>.
 It does so by performing something called \emph{hand-over-hand locking}.
-It first locks the first node, which is the \texttt{-inf} node.
+It first locks the first node, which is the <{-inf}> node.
 Then, iteratively, it obtains a lock on the next node and release the
-lock on the last one, and so on, similar to climbing a tree
+lock on the last one, and so on, similar to climbing a rope
 hand-over-hand.
-Using \texttt{\_find} the <{insert}>, \texttt{remove}, and \texttt{contains}
-methods are fairly straightforward.
-
+Using \texttt{\_find}, the <{insert}>, \texttt{remove}, and \texttt{contains}
+methods are fairly straightforward to implement.
+
+\begin{comment}
 \glsadd{linearizable}%
@@
 If not found, it coincides with retrieving the pointer to the
-\texttt{inf} node.
+<{inf}> node.
+\end{comment}
 
@@
 list, the maximum value in the list, and the sum of the values in the list.
-Are they linearizable?  If so, what are their linearization points?
+(All these should ignore the two end nodes.)
+% Are they linearizable?  If so, what are their linearization points?
 \item Create a thread-safe sorted binary tree.  Implement a module \texttt{bintree}

### 2021-12-24 book/paper.tex
Pass over Chapter 13

@@
 letters, digits, and underscores can be written without quotes by
-preceding it with a dot.  So <{.example}> is the same string
+preceding it with a dot.  So, <{.example}> is the same string
 as <{"example"}>.
@@
 would find no problems.
+Similarly, \autoref{fig:intsettest} is not a good test program
+for something as complicated as \autoref{fig:linkedlist}.
 In this chapter, we will look at approaches to testing
@@
 
-As with critical sections, when testing a concurrent data structure
-we need a specification.
-\autoref{fig:queues}(a) shows a sequential specification of a
+As with critical sections---when testing a concurrent data structure---we
+need a specification.
+For example, \autoref{fig:queues}(a) shows a sequential specification of a
 queue in Harmony.
-First we can check if the queue implementation
+First, we can check if the queue implementation
 in \autoref{fig:queueconc} meets the sequential
@@
 \autoref{fig:qtestseq} presents a test program that does
-exactly this, for sequences of up to \texttt{NOPS}
+exactly this, for sequences of up to <{NOPS}>
 queue operations.
@@
 \begin{itemize}
-\item[] \textit{implq}: the queue of the implementation;
-\item[] \textit{specq}: the queue of the specification
+\item[] \textit{specq}: the queue of the specification;
+\item[] \textit{implq}: the queue of the implementation.
 \end{itemize}
@@
 bugs in the implementation.
-So run a test program like this with a variety of queue implementations that
+So, run a test program like this with a variety of queue implementations that
 have known bugs in them and make sure that the test program finds them.
@@
 Conversely, a test program may be broken in that it finds bugs that do not exist.
+In my experience, it is often harder to implement the test program than
+the algorithm that the test program tests.
+
 \begin{comment}
@@
 queue implementation is correct, at least sequentially.
-The higher \texttt{NOPS}, the higher the confidence.
+The higher <{NOPS}>, the higher the confidence.
 It is possible to write similar programs in other languages
@@
 the test program will not (and should not)
-find any errors.
+find any errors, but it would be an incorrect implementation
+of a concurrent queue.
 
@@
 
-We would like a way that systematically compares behaviors of the
+We would like a way that---similar to the sequential test---systematically
+compares behaviors of the
 concurrent queue implementation with behaviors of the concurrent
@@
 implementation.  We will then check if the behaviors obtained from
-running the implementation is a subset of the behaviors obtained
+running the implementation are also behaviors obtained
 from the specification.
@@
 \autoref{fig:qtestpar} shows the test program.  It starts
-\texttt{NOPS} threads and each choose to do either a <{put}>
+<{NOPS}> threads, and each choose to do either a <{put}>
 or a <{get}> operation.  In case of a <{put}> operation,
-the thread enqueues its own name.  In order to output the behaviors,
+the thread enqueues its own name (which is provided as an argument
+to the thread).  In order to capture the behaviors,
 each thread prints what operation it is about to perform,
@@
 completed (including the return value if any).
-The figure shows the deterministic finite automaton when the test
+This is probably much like you would do if you were trying to find a
+bug in a program.
+
+\autoref{fig:qtestpar} also shows the deterministic finite automaton
+that describes the possible outputs when the test
 program is run against the specification in the case
-$\texttt{NOPS} = 2$---for $\texttt{NOPS} = 4$ it would be much
+$<{NOPS}> = 2$---for $<{NOPS}> = 4$ it would be much
 too large to print here.  Since there are no cycles in the DFA, you
@@
 interleavings of the threads.
+You can obtain this output yourself by running
+
+\begin{code}
+\begin{verbatim}
+$ harmony -c NOPS=2 -o spec.png code/qtestpar.hny
+\end{verbatim}
+\end{code}
 
 If you run the same test program against the implementation
-of \autoref{fig:queueconc}, you will get the same output.  You can
-try this for various $\texttt{NOPS}$, although it gets increasingly
+of \autoref{fig:queueconc}, you will get the same output:
+
+\begin{code}
+\begin{verbatim}
+$ harmony -c NOPS=2 -o impl.png -m queue=queueconc code/qtestpar.hny
+\end{verbatim}
+\end{code}
+
+You can
+try this for various $<{NOPS}>$, although it gets increasingly
 harder to check by hand that the generated DFAs are the same as
-\texttt{NOPS} increases.
+<{NOPS}> increases.
 Now run the test program against \autoref{fig:queueseq}, which is
@@
 
-Harmony does have a way to check the behavior of one program against
-the behavior of another.  In particular, we want to check if the
-behaviors of the implementations we have match the behavior of the
+Harmony does have a way to check the behaviors of one program against
+the behaviors of another.  In particular, we want to check if the
+behaviors of the implementations we have match behaviors of the
 specification.  The following shows, for example, how to check the
@@
 \begin{problems}
-\item Write a sequential specification for \autoref{fig:linkedlist}.
-(Hint: it implements a set of integers.)
-Write a Harmony program that checks if \autoref{fig:linkedlist} meets
-the sequential specification.
-\item Also create a concurrent specification and a concurrent test program.
-Check if the behaviors of \autoref{fig:linkedlist} match those of the
-specification.
+\item Write a Harmony program that checks if
+\autoref{fig:linkedlist} satisfies the specification
+of \autoref{fig:setobj} \emph{sequentially}.
+\item Write a Harmony program that checks if
+\autoref{fig:linkedlist} satisfies the specification
+of \autoref{fig:setobj} \emph{concurrently}.
 % \item Extend \autoref{fig:linkedlist} with linearization points and
@@
 \item Rewrite \autoref{fig:qtestseq} so it only imports \texttt{queue}
-and runs \texttt{NOPS} nondeterministically chosen operations against
+and runs <{NOPS}> nondeterministically chosen operations against
 it (similar in style to \autoref{fig:qtestpar} but without threads).
@@
 
-So you wrote a Harmony program and Harmony reports a bug.
+So, you wrote a Harmony program and Harmony reports a bug.
 Often you may just be able to figure it out by staring at the code and
@@
 Before we go look at the details of what went wrong, we want to make
-sure we generate the simplest scenario.  So first we want to explore
-what the smallest \texttt{NOPS} (number of operations or number of
+sure we generate the simplest scenario.  So, first we want to explore
+what the smallest <{NOPS}> (number of operations or number of
 threads) that causes the bug to occur.  With some experimentation,
-we find that \texttt{NOPS} = 2 does not find a problem, but
-\texttt{NOPS} = 3 does
+we find that <{NOPS}> = 2 does not find a problem, but
+<{NOPS}> = 3 does
 (\texttt{harmony -m queue=queuebroken -c NOPS=3 code/qtestpar.hny})).
@@
 
-So how did we get there?  In the top we can see that the order of
+So, how did we get there?  In the top we can see that the order of
 events was the following:
@@
 It then loops until either it fails or it has applied all
-\texttt{NOPS} operations to its history.
+<{NOPS}> operations to its history.
 $\mathbf{choose}(\{ \mathit{immortal}, <{True}> \})$ can only
@@
 history until the
-history has \texttt{NOPS} operations on it or until the head crashes,
+history has <{NOPS}> operations on it or until the head crashes,
 whichever comes first.  The head also broadcasts each update, which
@@
 
-When the replica successfully adds all \texttt{NOPS} updates to its
+When the replica successfully adds all <{NOPS}> updates to its
 history without crashing, the code makes sure that any other such replica
@@
 tuples consisting of an integer and the unique identifier of the
-client that writes the value.  So if $(t, c)$ is the highest timestamp
+client that writes the value.  So, if $(t, c)$ is the highest timestamp
 observed by client $c'$, and $c'$  needs to create a new timestamp,
@@
 if a message is for a future round, the processor will buffer it.
-So one can simplify the model and have each processor wait
+So, one can simplify the model and have each processor wait
 for \emph{all} <{N}> messages in a round
@@
 the protocol with Alice and received $A_N$, and even though Bob and Alice
-are both honest, Corey has a copy of $A_N$.  So Corey is now able to impersonate
+are both honest, Corey has a copy of $A_N$.  So, Corey is now able to impersonate
 Alice to Bob (but not vice versa because Alice did not try to authenticate Bob).
@@
 threads, nor by \textbf{invariant} statements.
-So the model checker only interleaves threads at \textbf{Load},
+So, the model checker only interleaves threads at \textbf{Load},
 \textbf{Store}, and \textbf{Del} instructions where a thread
@@
 by keeping track, along the reported execution path, what state each
-thread is in.  So by examining the initial context of the thread that
+thread is in.  So, by examining the initial context of the thread that
 is running from some particular state, it can determine if that context
@@
 resources, \texttt{Harmony IDE} must put a limit on the amount of compute
-resources you can use.  So while \texttt{Harmony IDE} is currently a great
+resources you can use.  So, while \texttt{Harmony IDE} is currently a great
 way to try out Harmony, if you become a serious user of Harmony you

### 2021-12-24 book/paper.tex
Made a pass over Ch 14

@@
 \item \autoref{fig:incmany} is a version of \autoref{fig:incpy} that has each
-incrementer thread increment <{count}> <{N}> times.  Run \autoref{fig:incmany}
+incrementer thread increment <{count N}> times.  Run \autoref{fig:incmany}
 10 times (using Python).
@@
 The <{atomically}> keyword can be used to specify the behavior of
-methods such as \texttt{acquire} and \texttt{release}.  But
+methods such as <{acquire}> and <{release}>.  But
 an actual executable program---such as the one in
@@
 If we want to make the program executable on hardware,
-we have to show how \texttt{Lock}, \texttt{acquire}, and \texttt{release}
+we have to show how \texttt{Lock}, <{acquire}>, and <{release}>
 are implemented, not just how they are specified.
@@
 \texttt{!}<{p}> = <{True}> upon entry of \texttt{test\_and\_set}
-(because of the condition on the \textbf{while} loop that the
+(because of the condition on the <{while}> loop that the
 \texttt{test\_and\_set} method is invoked in).
@@
 \item
-\texttt{acquire}()
+<{acquire}>()
 \index{acquire}%
@@
 \item
-\texttt{release}()
+<{release}>()
 \index{release}%
@@
 interface for both the <{synch}> and \texttt{synchS} modules.
-This interface is like \texttt{acquire}(\textit{b}) but never blocks.  It
+This interface is like <{acquire}>(\textit{b}) but never blocks.  It
 returns <{True}> if the lock was available (and now acquired) or <{False}>
@@
 The implementation in \autoref{fig:queueconc}
-uses the \texttt{alloc} module for dynamic allocation
+uses the <{alloc}> module for dynamic allocation
 \index{dynamic allocation}%
@@
 concurrent <{get}> and <{put}> operations.
-It starts <{N}> <{getter}> threads and
+It starts <{N getter}> threads and
 <{N}> \texttt{putter} threads.
@@
 uses a local variable <{node}>, declared using
-a \textbf{let} statement, but it would be easy to
+a <{let}> statement, but it would be easy to
 mistakenly use a global variable <{node}> instead.
@@
 
-So, you wrote a Harmony program and Harmony reports a bug.
+So, you wrote a Harmony program and Harmony reports a problem.
 Often you may just be able to figure it out by staring at the code and
@@
 the new queue code through the test in \autoref{fig:qtestpar}, Harmony
-finds a problem (even without specifying a behavior).  The command line
-for this is \texttt{harmony -m queue=queuebroken code/qtestpar.hny})
+reports a safety violation (even without specifying a behavior).
+The command line to reproduce this is:
+
+\begin{code}
+\begin{verbatim}
+harmony -m queue=queuebroken code/qtestpar.hny
+\end{verbatim}
+\end{code}
 
 Before we go look at the details of what went wrong, we want to make
-sure we generate the simplest scenario.  So, first we want to explore
+sure that we generate the simplest scenario.  So, first we want to explore
 what the smallest <{NOPS}> (number of operations or number of
 threads) that causes the bug to occur.  With some experimentation,
-we find that <{NOPS}> = 2 does not find a problem, but
-<{NOPS}> = 3 does
+we find that <{NOPS = 2}> does not find a problem, but
+<{NOPS = 3}> does
 (\texttt{harmony -m queue=queuebroken -c NOPS=3 code/qtestpar.hny})).
@@
 violation (something bad happened during one of the possible executions),
-and in particular \texttt{thread}(2) (thread T2) was trying to dereference
-\texttt{?alloc["pool"][0]["lock"]}
-(aka \texttt{?alloc}.\textit{pool}[0].\texttt{lock}).
-
-The \texttt{alloc} module maintains a shared array \textit{pool} that
+and in particular <{thread}>(2) (thread T2) was trying to dereference
+the address <{?alloc["pool"][0]["lock"]}> (aka <{?alloc.pool[0].lock}>).
+
+The <{alloc}> module maintains a shared array <{pool}> that
 it uses for dynamic allocation.  Apparently T2 tried to access
-\textit{pool}[0], but it does not exist, meaning that either it was not
+<{pool}>[0], but it does not exist, meaning that either it was not
 yet allocated, or it had been freed since it was allocated.
 When we look at the top half of the figure, we see that in fact thread
-T1 allocated \textit{pool}[0] in turn 2, but T3 freed it in turn 4.
-Looking back down, we see that T1 executed \texttt{thread}(1) and has
-since terminated, while T3 is executing \texttt{thread}(3).
+T1 allocated <{pool}>[0] in turn 2, but T3 freed it in turn 4.
+Looking back down, we see that T1 executed <{thread(1)}> and has
+since terminated, while T3 is executing <{thread(3)}>.
 
@@
 T3 was in the process of executing
-\texttt{release}(?<{q}>.\texttt{lock}) within
-<{get}>(?<{q}>).
-% Looking in the top green row, we see that the value of <{q}> is
-% \{ \texttt{.lock}: <{True}>, \texttt{.next}: <{None}>, \texttt{.value}: <{None}> \}.
-T1 is currently executing
-\texttt{acquire}(\texttt{?alloc}.\textit{pool}[0].\texttt{lock})
-within <{put}>(?<{q}>, 2), but 
-\texttt{alloc}.\textit{pool}[0] does not exist.
+<{release(?q.lock)}> within <{get(?q}>).
+T1 is currently executing <{acquire(?alloc.pool[0].lock)}>
+within <{put(?q, 2)}>, but 
+<{alloc.pool}>[0] does not exist.
 The corresponding line of Harmony code is
@@
 \item[0:] initialization completed, with <{q}> being
-\{ \texttt{.lock}: <{False}>, \texttt{.next}: <{None}>, \texttt{.value}: <{None}> \};
-\item[1:] thread T1 (\texttt{thread}(1)) ran and finished executing
-<{put}>(1) (see the output column for that clue: the thread printed that).
-We can see that <{q}>.<{next}> now points to \texttt{alloc}.\textit{pool}[0],
+<{{ .lock: False, .next: None, .value: None }}>;
+\item[1:] thread T1 (<{thread(1)}>) ran and finished executing
+<{put(1)}> (see the output column for that clue: the thread printed that).
+We can see that <{q.next}> now points to <{alloc.pool[0]}>,
 which the thread must have allocated.  The contents is 
-\{ \texttt{.lock}: <{False}>, \texttt{.next}: <{None}>, \texttt{.value}: 1 \}, as expected;
-\item[2:] thread T2 (\texttt{thread}(1)) started running, calling
-<{put}>(?<{q}>, 2).  We can see it got as far as putting 2 on the queue,
+<{{ .lock: False, .next: None, .value: 1 }}>, as expected;
+\item[2:] thread T2 (<{thread(1)}>) started running, calling
+<{put(?q, 2)}>.  We can see it got as far as putting 2 on the queue,
 but it is not yet done.  It is currently trying to acquire
-\texttt{alloc}.\textit{pool}[0].\texttt{lock};
-\item[3:] thread T3 (\texttt{thread}(1)) started running, calling
-<{get}>($?q$).  We can also see that it freed \textit{pool}[0],
-and is now releasing <{q}>.\texttt{lock};
-\item[4:] thread T2 resumes and tries to access \textit{pool}[0], which
+<{alloc.pool[0].lock}>;
+\item[3:] thread T3 (<{thread(1)}>) started running, calling
+<{get(?q)}>.  We can also see that it freed <{pool[0]}>,
+and is now releasing <{q.lock}>;
+\item[4:] thread T2 resumes and tries to access <{pool[0]}>, which
 no longer exists (because T3 just freed it).
@@
 Clearly there was a race in which T2 was trying to lock
-?\textit{pool}[0].\texttt{lock} (which contained the node with the value 1)
+<{pool[0].lock}> (which contained the node with the value 1)
 while T3 was freeing that very same node, and T2 lost the race.
-More precisely, T2 was executing <{put}>(?<{q}>, 2), when T3
-preempted it with <{get}>(?<{q}>) and removed the node that T2
+More precisely, T2 was executing <{put(?q, 2)}>, when T3
+preempted it with <{get(?q)}> and removed the node that T2
 was trying to access.  But why did the locks not prevent this?
@@
 \emph{reverse debugging}.  In fact, Harmony allows you to step through
-code forwards and backwards.  In this case, we first want to see what
+an execution forwards and backwards.
+In this case, we first want to see what
 T2 is doing.  You can click on its first (top-left) orange box to
-time-travel to that part in the execution.  Now by hitting
-$<$return$>$ repeatedly, we can quickly skip through the code.
-T2 first calls <{put}>(?<{q}>, 1) and then allocates a new
-node initialized with a lock.  Keep stepping until it executes
-$\mathit{nq} = q$.  Hit $<$return$>$ once more and inspect the
+time-travel to that part in the execution.
+Now by hitting $\langle\mathtt{return}\rangle$ repeatedly,
+we can quickly skip through the code.
+T2 first calls <{put(?q, 1)}> and then allocates a new
+node initialized with a lock.  Keep stepping until it executes <{nq = q}>.
+Hit $\langle\mathtt{return}\rangle$ once more and inspect the
 state of T2 in the lower-right corner.  You can see that variable
-<{nq}> is initialized to~?<{q}>.  T2 then enters into the
-\textbf{while} loop and tries to acquire
-<{nq->lock}>.  This succeeds, and
-next T2 executes \textbf{let} <{n = nq->next}>.
-Now $n =\,?\mathtt{alloc}.\mathit{alloc}[0]$, which is not <{None}>.
+<{nq}> is initialized to~<{?q}>.  T2 then enters into the
+<{while}> loop and tries to acquire <{nq->lock}>.
+This succeeds, and next T2 executes <{let n = nq->next}>.
+Now <{n = ?alloc.pool[0]}>, which is not <{None}>.
 It then releases <{nq->lock}>
 (<{nq}> points to <{q}>).  It then sets <{nq}> to $n$, which
-is still $?\mathtt{alloc}.\mathit{alloc}[0]$.  Finally, it calls
-<{acquire(?nq->lock)}>.
+is still <{alloc.pool[0]}>.  Finally, it calls <{acquire(?nq->lock)}>.
 But before it can complete that operation, T3 runs next.
 
-T3 chooses "get" and then goes on to invoke $\mathtt{get}(?q)$.
+T3 chooses <{"get"}> and then goes on to invoke <{get(?q)}>.
 This first successfully acquires <{q}->lock}>.
@@
 T3 sets <{node}> to <{alloc.pool[0]}> as well and sets
-<{q->{next}> to <{node->next}>.
+<{q->next}> to <{node->next}>.
 T3 sets <{result}> to <{node->value}> (which is 1)
@@
 that uses hand-over-hand locking
-both for <{put}>() and for <{get}>().  It passes all tests.
+both for <{put()}> and for <{get()}>.  It passes all tests.
 
@@
 until it leaves the monitor explicitly
-(by calling either \texttt{release} or \texttt{wait}).
+(by calling either <{release}> or \texttt{wait}).
 At that point, any thread that
@@
 again after resuming from the \texttt{wait} operation.  This is accomplished
-by wrapping each \texttt{wait} operation in a \textbf{while} statement that
+by wrapping each \texttt{wait} operation in a <{while}> statement that
 loops until the condition of interest becomes valid.
@@
 
-Note that \texttt{wait} is always invoked within a \textbf{while}
+Note that \texttt{wait} is always invoked within a <{while}>
 loop that checks for the condition that the thread is waiting for.
-It is \emph{imperative} that there is always a \textbf{while} loop
+It is \emph{imperative} that there is always a <{while}> loop
 around any invocation of \texttt{wait} containing the negation of
@@
 As a rule of thumb, one should always be able to replace \texttt{wait}
-by \texttt{unlock} followed by \texttt{lock}.
+by <{unlock}> followed by <{lock}>.
 This turns the solution into a busy-waiting one, inefficient but still
@@
 readers and writers are awakened---this is ok because both execute
-\texttt{wait} within a \textbf{while} loop, re-checking the condition
+\texttt{wait} within a <{while}> loop, re-checking the condition
 that they are waiting for.  So, if both type of threads are waiting,
@@
 \item \textit{tl} = \texttt{TryLock}() \# \emph{create a try lock}
-\item \texttt{acquire}(?\textit{tl}) \# \emph{acquire a try lock}
+\item <{acquire}>(?\textit{tl}) \# \emph{acquire a try lock}
 \item \texttt{tryAcquire}(?\textit{tl}) \# \emph{attempt to acquire a try lock}
-\item \texttt{release}(?\textit{tl}) \# \emph{release a try lock}
+\item <{release}>(?\textit{tl}) \# \emph{release a try lock}
 \end{enumerate}
@@
 \index{multiple conditions, waiting on}%
-The condition in the \textbf{while} statement is the negation of the
+The condition in the <{while}> statement is the negation of the
 condition that the diner is waiting and consists of two disjuncts.
-Within the \textbf{while} statement,
+Within the <{while}> statement,
 there is an \textbf{if} statement for each disjunct.
 The code waits for either or both forks if necessary.  After that, it goes
-back to the top of the \textbf{while} loop.
+back to the top of the <{while}> loop.
 
@@
 diner.  It uses the same number of condition variables, but you will not
-need to have \textbf{if} statements nested inside the \textbf{while} loop
+need to have \textbf{if} statements nested inside the <{while}> loop
 waiting for the forks.  See if you can figure it out.
@@
 
-Try moving \texttt{lock}() to the beginning of the \texttt{increment} method
-and \texttt{unlock}() to the end of \texttt{increment} and see what happens.
+Try moving <{lock()}> to the beginning of the <{increment}> method
+and <{unlock()}> to the end of <{increment}> and see what happens.
 While Harmony will only report one faulty run, this incorrect code can lead to
@@
 non-blocking synchronization calls such as \texttt{notify}, but calls such as
-\texttt{acquire} and \texttt{wait} should only be used if it is certain that they will not
+<{acquire}> and \texttt{wait} should only be used if it is certain that they will not
 block for long.  Informally, interrupt handlers can be
@@
 
-The statement \textbf{go} <{c}> <{e}> starts a thread with context <{c}> that
+The statement \textbf{go} <{c e}> starts a thread with context <{c}> that
 has executed a \textbf{stop} expression.  The \textbf{stop} expression
@@
 
-\subsection*{\textbf{let}}
+\subsection*{<{let}>}
 
@@
 <{let}> statement.
-The statement \textbf{let} $a = b: S_1; S_2, ...$ evaluates <{b}>, assigns
+The statement <{let}> $a = b: S_1; S_2, ...$ evaluates <{b}>, assigns
 the result to read-only variable <{a}>, and evaluates statements
-$S_1, S_2, ...$.  \textbf{let} supports pattern matching, so you can
+$S_1, S_2, ...$.  <{let}> supports pattern matching, so you can
 write <{let}> $x, (y, z) = b: S_1; S_2, ...$.  This will only work
@@
 
-\textbf{let} statements may be nested, such as
-\textbf{let} $a_1 = b_1$ \textbf{let} $a_2 = b_2: S_1; S_2; ...$.
+<{let}> statements may be nested, such as
+<{let}> $a_1 = b_1$ <{let}> $a_2 = b_2: S_1; S_2; ...$.
 Doing so can improve readability by reducing indentation
@@
 
-The statement \texttt{spawn} <{m}> <{a}> starts a new thread that executes
+The statement \texttt{spawn} <{m a}> starts a new thread that executes
 method <{m}> with argument <{a}>. <{m}> must be a program counter constant,
@@
 If a thread may never terminate, you should spawn it with
-\texttt{spawn eternal} <{m}> <{a}> to suppress those checks.
+\texttt{spawn eternal} <{m a}> to suppress those checks.
 
@@
 
-The statement \textbf{trap} <{m}> <{a}> specifies that the current thread
+The statement \textbf{trap} <{m a}> specifies that the current thread
 should execute method <{m}> with argument <{a}> and some future, unspecified,
@@
 
-\subsection*{\textbf{while}}
+\subsection*{<{while}>}
 
@@
 first.  In that expression, only <{a}> or <{b}> is evaluated depending on the value of <{c}>.
-In the expression <{a}> <{and}> <{b}> <{and}> $...$, evaluation is left
+In the expression <{a and b and}> $...$, evaluation is left
 to right but stops once one of the subexpressions evaluates to <{False}>.
@@
 \begin{itemize}
-\item <{for}> <{k}> \texttt{in} \texttt{keys} <{d}>; or
+\item <{for k}> \texttt{in} \texttt{keys} <{d}>; or
 \item <{for}> $k:\_$ \texttt{in} <{d}>
@@
 \item[] \autoref{fig:stack4} implements a stack as a linked list with nodes
-allocated using the \texttt{alloc} module.
+allocated using the <{alloc}> module.
 \end{enumerate}
@@
 
-\section{The \texttt{alloc} module}
-
-The \texttt{alloc} module
+\section{The <{alloc}> module}
+
+The <{alloc}> module
 \index{alloc module}%
@@
 \hline
-\texttt{acquire}(\textit{bs}) & acquire binary semaphore \texttt{!}\textit{bs} \\
-\hline
-\texttt{release}(\textit{bs}) & release binary semaphore \texttt{!}\textit{bs} \\
+<{acquire}>(\textit{bs}) & acquire binary semaphore \texttt{!}\textit{bs} \\
+\hline
+<{release}>(\textit{bs}) & release binary semaphore \texttt{!}\textit{bs} \\
 \hline
@@
 \hline
-Frame <{m}> <{a}> & start method <{m}> with arguments <{a}>,
+Frame <{m a}> & start method <{m}> with arguments <{a}>,
 initializing variables.  \\
@@
 \hline
-JumpCond <{e}> <{p}> & pop expression and, if equal to <{e}>, set program counter to <{p}> \\
+JumpCond <{e p}> & pop expression and, if equal to <{e}>, set program counter to <{p}> \\
 \hline
@@
 the method, an ``invocation type'' (\texttt{normal}, \texttt{interrupt}, or
-\texttt{thread}), and the return address (in case of \texttt{normal}).
+<{thread}>), and the return address (in case of \texttt{normal}).
 \item A thread terminates when it reaches the \texttt{Return} instruction
-of the top-level method (when the stack frame is of type \texttt{thread})
+of the top-level method (when the stack frame is of type <{thread}>)
 or when it hits an exception.  Exceptions include divide by zero,

### 2021-12-24 book/paper.tex
Pass over Ch 15

@@
 section.
-The name \textit{mutex} is often used to denote a variable or value
+The name <{mutex}> is often used to denote a variable or value
 that is used for mutual exclusion.
@@
 If we want to make the program executable on hardware,
-we have to show how \texttt{Lock}, <{acquire}>, and <{release}>
+we have to show how <{Lock}>, <{acquire}>, and <{release}>
 are implemented, not just how they are specified.
@@
 make progress until another thread enqueues an item.
-However, the thread is not considered \emph{blocked} because it is
+However, the thread is not considered blocked because it is
 changing the shared state by repeatedly acquiring and releasing the
 lock.
-We distinguish \emph{passive busy waiting} and \emph{active busy waiting}.
 A process that is waiting for a condition without changing the state
-(like in a spinlock) is passively busy waiting.
+(like in a spinlock) is \emph{blocked}.
 A process that is waiting for a condition while changing the state
 (such as repeatedly trying to dequeue an item, which requires
-acquiring a lock) is actively busy waiting.
+acquiring a lock) is \emph{actively busy waiting}.
 
@@
 \begin{enumerate}
-\item \textit{nreaders}: the number of readers
-\item \textit{nwriters}: the number of writers
+\item <{nreaders}>: the number of readers
+\item <{nwriters}>: the number of writers
 \end{enumerate}
@@
 
-There are four operations on a reader/writer lock \textit{rw}:
+There are four operations on a reader/writer lock <{rw}>:
 \begin{itemize}
-\item \texttt{read\_acquire}(\textit{rw}): waits until $\mathit{nwriters} = 0$
-and then increments \textit{nreaders};
-\item \texttt{read\_release}(\textit{rw}): decrements $\mathit{nreaders}$;
-\item \texttt{write\_acquire}(\textit{rw}): waits until
-$\mathit{nreaders} = \mathit{nwriters} = 0$
-and then sets \textit{nwriters} to 1;
-\item \texttt{write\_release}(\textit{rw}): sets $\mathit{nwriters}$ to 0.
+\item <{read_acquire(rw)}>: waits until <{nwriters = 0}>
+and then increments <{nreaders}>;
+\item <{read_release(rw)}>: decrements <{nreaders}>;
+\item <{write_acquire(rw)}>: waits until <{nreaders = nwriters = 0}>
+and then sets <{nwriters}> to 1;
+\item <{write_release(rw)}>: sets <{nwriters}> to 0.
 \end{itemize}
@@
 In particular, a thread can only release a reader/writer lock for reading
-if it acquired it for reading and the same for writing.
-Moreover, a thread is only allowed the acquire a reader/writer lock once.
+if it acquired the lock for reading and the same for writing.
 
@@
 of a certain maximum length.
-If the buffer is full, the enqueuer must wait; if the buffer is empty, the
-dequeuer must wait.
+If the buffer is full, an enqueuer must wait; if the buffer is empty,
+a dequeuer must wait.
 \autoref{fig:boundedbuffer} specifies a bounded buffer.
@@
 thread first has to wait until there is room in the bounded buffer.
-To this end, it invokes \texttt{P}(\textit{n\_empty}), which waits until
+To this end, it invokes <{P}>(\textit{n\_empty}), which waits until
 $\mathit{n\_empty} > 0$ and atomically decrements \textit{n\_empty} once this
@@
 \begin{itemize}
-\item \textit{nreaders}: the number of readers in the critical section;
-\item \textit{nwriters}: the number of writers in the critical section;
-\item \textit{mutex}: the main gate binary semaphore;
+\item <{nreaders}>: the number of readers in the critical section;
+\item <{nwriters}>: the number of writers in the critical section;
+\item <{mutex}>: the main gate binary semaphore;
 \item \textit{r\_gate}: the waiting gate used by readers, consisting of a binary semaphore and the number of readers waiting to enter;
@@
 this state.
-First they have to acquire the \textit{mutex}
+First they have to acquire the <{mutex}>
 (i.e., enter the main gate).
-After acquiring the \textit{mutex},
+After acquiring the <{mutex}>,
 \texttt{read\_acquire} and \texttt{write\_acquire} each must check to see
@@
 If neither waiting gate qualifies, then <{release_one()}> 
-has to open the main gate (i.e., release \textit{mutex}).
+has to open the main gate (i.e., release <{mutex}>).
 
 Let us examine \texttt{read\_acquire} more carefully.
-First the method acquires the \textit{mutex}.
+First the method acquires the <{mutex}>.
 Then, in the case that the thread
@@
 \begin{enumerate}
-\item After acquiring \textit{mutex}, the writer decrements
-\textit{nwriters}, which must be 1 at this time, and thus becomes 0.
+\item After acquiring <{mutex}>, the writer decrements
+<{nwriters}>, which must be 1 at this time, and thus becomes 0.
 \item It then calls <{release_one()}>.
@@
 and there are two readers waiting.  It therefore releases not
-\textit{mutex} but the readers' gate's binary semaphore.
+<{mutex}> but the readers' gate's binary semaphore.
 \item One of the waiting readers can now re-enter the critical section.
 When it does, the reader decrements the gate's counter (from 2 to 1)
-and increments \textit{nreaders} (from~0 to~1).
+and increments <{nreaders}> (from~0 to~1).
 The reader finally calls <{release_one()}>.
@@
 \item The second reader can now enter the critical section.
-It decrements the gate's count from 1 to 0 and increments \textit{nreaders}
+It decrements the gate's count from 1 to 0 and increments <{nreaders}>
 from~1 to~2.
@@
 This time <{release_one()}> does not find any threads waiting,
-and so it releases \textit{mutex}.
+and so it releases <{mutex}>.
 There are now two threads that are holding the reader/writer lock.
@@
 \item Several of the instantiations of <{release_one()}> in
-\autoref{fig:RWsplitsema} can be replaced by simply releasing \textit{mutex}.
+\autoref{fig:RWsplitsema} can be replaced by simply releasing <{mutex}>.
 Which ones?
@@
 one per waiting condition.
-There are two operations on condition variables: \texttt{wait}
+There are two operations on condition variables: <{wait}>
 \index{wait}%
@@
 the number of threads waiting on the condition.
-\texttt{wait} increments the condition's counter, releases the monitor mutex,
+<{wait}> increments the condition's counter, releases the monitor mutex,
 blocks while trying to acquire the condition's semaphore, and upon resuming
@@
 and, unlike \texttt{release\_one}, re-enter the critical
-section afterwards by acquiring the \textit{mutex}.
+section afterwards by acquiring the <{mutex}>.
 
@@
 Mesa monitors also have condition variables that a thread can wait on.
-Like in Hoare monitors, the \texttt{wait} operation releases the mutex.
+Like in Hoare monitors, the <{wait}> operation releases the mutex.
 The most important difference is in what \texttt{signal} does.
 To make the distinction more clear, we shall call the corresponding Mesa
-operation \texttt{notify} rather than \texttt{signal}.
+operation <{notify}> rather than \texttt{signal}.
 \index{notify}%
-When a thread <{p}> invokes \texttt{notify}, it does not immediately pass 
+When a thread <{p}> invokes <{notify}>, it does not immediately pass 
 control to a thread that is waiting on the corresponding condition (if there
@@
 until it leaves the monitor explicitly
-(by calling either <{release}> or \texttt{wait}).
+(by calling either <{release}> or <{wait}>).
 At that point, any thread that
@@
 the condition.  So, in Mesa, threads always have to check the condition
-again after resuming from the \texttt{wait} operation.  This is accomplished
-by wrapping each \texttt{wait} operation in a <{while}> statement that
+again after resuming from the <{wait}> operation.  This is accomplished
+by wrapping each <{wait}> operation in a <{while}> statement that
 loops until the condition of interest becomes valid.
@@
 Mesa monitors also allow notifying multiple threads.
-For example, a thread can invoke \texttt{notify} twice---if there are
+For example, a thread can invoke <{notify}> twice---if there are
 two or more threads waiting on the condition variable, two will be resumed.
-Operation \texttt{notifyAll}
+Operation <{notifyAll}>
 \index{notifyAll}%
@@
 Methods declared with the \texttt{synchronized} keyword automatically
-obtain the lock.  Java objects also support \texttt{wait}, \texttt{notify},
-and \texttt{notifyAll}.
+obtain the lock.  Java objects also support <{wait}>, <{notify}>,
+and <{notifyAll}>.
 In addition, Java supports explicit allocations of locks
@@
 condition variables in the <{synch}> module.
-\texttt{Condition}() creates a new condition variable.
+<{Condition}>() creates a new condition variable.
 It is represented by a dictionary containing
@@
 
-\texttt{wait} adds the context of the thread to the bag.
+<{wait}> adds the context of the thread to the bag.
 This increments the number of threads in the bag with the same context.
-\texttt{wait} then loops until that count is restored to the
-value that it had upon entry to \texttt{wait}.
-\texttt{notify} removes an arbitrary context from
+<{wait}> then loops until that count is restored to the
+value that it had upon entry to <{wait}>.
+<{notify}> removes an arbitrary context from
 the bag, allowing one of the threads with that context to
 resume and re-acquire the lock associated with the monitor.
-\texttt{notifyAll} empties out the entire bag, allowing all threads
+<{notifyAll}> empties out the entire bag, allowing all threads
 in the bag to resume.
@@
 To illustrate how Mesa condition variables are used in practice, we demonstrate using an implementation of reader/writer locks.
-\autoref{fig:RWcv} shows the code.  \textit{mutex} is the shared lock
+\autoref{fig:RWcv} shows the code.  <{mutex}> is the shared lock
 that protects the critical region.
-There are two condition variables: readers wait on \textit{r\_cond} and
-writers wait on \textit{w\_cond}.
+There are two condition variables: readers wait on <{r_cond}> and
+writers wait on <{w_cond}>.
 The implementation also keeps track of the number of
@@
 
-Note that \texttt{wait} is always invoked within a <{while}>
+Note that <{wait}> is always invoked within a <{while}>
 loop that checks for the condition that the thread is waiting for.
 It is \emph{imperative} that there is always a <{while}> loop
-around any invocation of \texttt{wait} containing the negation of
+around any invocation of <{wait}> containing the negation of
 the condition that the thread is waiting for.  Many implementation
@@
 of condition variables often allow so-called ``spurious wakeups,''
-where \texttt{wait} may sometimes return even if the conditon
+where <{wait}> may sometimes return even if the conditon
 variable has not been notified.
-As a rule of thumb, one should always be able to replace \texttt{wait}
+As a rule of thumb, one should always be able to replace <{wait}>
 by <{unlock}> followed by <{lock}>.
@@
 
-In \texttt{read\_release}, notice that \texttt{notify}(?\textit{w\_cond})
+In \texttt{read\_release}, notice that <{notify}>(?\textit{w\_cond})
 is invoked when there are no readers left, \emph{without} checking
 if there are writers waiting to enter.  This is
-ok, because calling \texttt{notify} is a no-op if no thread is
+ok, because calling <{notify}> is a no-op if no thread is
 waiting.
 
-\texttt{write\_release} executes \texttt{notifyAll}(?\textit{r\_cond})
-as well as \texttt{notify}(?\textit{w\_cond)}.  Because we
+\texttt{write\_release} executes <{notifyAll}>(?\textit{r\_cond})
+as well as <{notify}>(?\textit{w\_cond)}.  Because we
 do not keep track of the number of waiting readers or writers, we
@@
 \autoref{fig:RWsplitsema}), with Mesa monitors all readers are
-awakened in one fell swoop using \texttt{notifyAll}.  Second, both
+awakened in one fell swoop using <{notifyAll}>.  Second, both
 readers and writers are awakened---this is ok because both execute
-\texttt{wait} within a <{while}> loop, re-checking the condition
+<{wait}> within a <{while}> loop, re-checking the condition
 that they are waiting for.  So, if both type of threads are waiting,
@@
 Much of the complexity of programming with Mesa condition variables
-is in figuring out when to invoke \texttt{notify} and when to invoke
-\texttt{notifyAll}.  As a rule of thumb: be conservative---it is
+is in figuring out when to invoke <{notify}> and when to invoke
+<{notifyAll}>.  As a rule of thumb: be conservative---it is
 better to wake up too many threads than too few.
-In case of doubt, use \texttt{notifyAll}.
+In case of doubt, use <{notifyAll}>.
 Waking up too
@@
 helpful here, as it examines each and every corner case.
-You can try to replace each \texttt{notifyAll} with \texttt{notify} and see
+You can try to replace each <{notifyAll}> with <{notify}> and see
 if every possible execution of the application still terminates.
@@
 acquired a given number of times.  It is supported by the <{synch}>
-module.  The \texttt{P} or ``procure'' operation acquires a counting
+module.  The <{P}> or ``procure'' operation acquires a counting
 semaphore.  That is, it tries to decrement the semaphore, blocking while
-the semaphore has a value of 0.  The \texttt{V} or ``vacate'' operation
+the semaphore has a value of 0.  The <{V}> or ``vacate'' operation
 increments the semaphore.
@@
 \item \texttt{Queue()} returns a new message queue;
-\item <{put}>(<{q}>, \textit{item}) adds \textit{item} to the queue pointed to by <{q}>;
-\item <{get}>(<{q}>) waits for and returns an item on the queue pointed to by <{q}>.
+\item <{put(q}>, \textit{item}) adds \textit{item} to the queue pointed to by <{q}>;
+\item <{get(q}>) waits for and returns an item on the queue pointed to by <{q}>.
 \end{itemize}
@@
 behaves much like a zero-initialized counting semaphore.
-<{put}> is much like \texttt{V}, except that it is accompanied by data.
-<{get}> is much like \texttt{P}, except that it also returns data.
+<{put}> is much like <{V}>, except that it is accompanied by data.
+<{get}> is much like <{P}>, except that it also returns data.
 Thus, synchronized queues can be considered a generalization of
@@
 using these stages.
-Method~\texttt{enter}() has two \texttt{wait} loops, one for each
+Method~\texttt{enter}() has two <{wait}> loops, one for each
 waiting condition.
@@
 handlers must run quickly and cannot wait for other events.  It is ok to invoke
-non-blocking synchronization calls such as \texttt{notify}, but calls such as
-<{acquire}> and \texttt{wait} should only be used if it is certain that they will not
+non-blocking synchronization calls such as <{notify}>, but calls such as
+<{acquire}> and <{wait}> should only be used if it is certain that they will not
 block for long.  Informally, interrupt handlers can be
@@
 \hline
-\texttt{BinSem}(<{v}>) & return a binary semaphore initialized to <{v}> \\
-\hline
-\texttt{Lock}() & return a binary semaphore initialized to <{False}> \\
+<{BinSem(v}>) & return a binary semaphore initialized to <{v}> \\
+\hline
+<{Lock}>() & return a binary semaphore initialized to <{False}> \\
 \hline
@@
 \hline
-\texttt{Condition}() & return a condition variable \\
-\hline
-\texttt{wait}(<{c}>, \textit{lk}) & wait on condition variable \texttt{!}<{c}>
+<{Condition}>() & return a condition variable \\
+\hline
+<{wait(c}>, \textit{lk}) & wait on condition variable \texttt{!}<{c}>
 and lock \textit{lk} \\ \hline
-\texttt{notify}(<{c}>) & notify a thread waiting on condition variable \texttt{!}<{c}> \\
-\hline
-\texttt{notifyAll}(<{c}>) & notify all threads waiting on condition variable \texttt{!}<{c}> \\
-\hline
-\texttt{Semaphore}(\textit{cnt}) & return a counting semaphore initialized to \textit{cnt} \\
-\hline
-\texttt{P}(\textit{sema}) & procure \texttt{!}\textit{sema}  \\
-\hline
-\texttt{V}(\textit{sema}) & vacate \texttt{!}\textit{sema}  \\
+<{notify(c}>) & notify a thread waiting on condition variable \texttt{!}<{c}> \\
+\hline
+<{notifyAll(c}>) & notify all threads waiting on condition variable \texttt{!}<{c}> \\
+\hline
+<{Semaphore(cnt}>) & return a counting semaphore initialized to <{cnt}> \\
+\hline
+<{P(sema}>) & procure \texttt{!}<{sema}>  \\
+\hline
+<{V(sema}>) & vacate \texttt{!}<{sema}>  \\
 \hline
@@
 \hline
-<{get}>(<{q}>) & return next element of <{q}>, blocking if empty \\
-\hline
-<{put}>(<{q}>, \textit{item}) & add \textit{item} to <{a}> \\
+<{get(q}>) & return next element of <{q}>, blocking if empty \\
+\hline
+<{put(q}>, \textit{item}) & add \textit{item} to <{a}> \\
 \hline

### 2021-12-24 book/paper.tex
Pass over Chapter 16

@@
 Tony Hoare and popularized by Edsger Dijkstra~\cite{EWD703}.
-SBS is an extension of a critical section that is protected by a single
-binary semaphore.
+A \emph{binary semaphore} is a generalization of a lock.  While a lock
+is always initialized in the released state, a binary semaphore---if so
+desired---can be initialized in the acquired state.
+SBS is an extension of a critical section that is protected by a lock.
 If there are $n$ \emph{waiting conditions},
 then SBS uses $n+1$ binary semaphores to protect the critical section.
-An ordinary critical section has no waiting conditions.
-For example, a bounded buffer has two waiting conditions:
+An ordinary critical section has no waiting conditions and therefore
+uses just one binary semaphore (because $n = 0$).
+But, for example, a bounded buffer has two waiting conditions:
 \begin{enumerate}
@@
 
-Think of each of these semaphores as a gate that a thread must go
+Think of each of these binary semaphores as a gate that a thread must go
 through in order to enter the critical section.  A gate is either open
@@
 the critical section.
-In particular, when a thread leaves the critical section it should
+In particular, when a thread leaves the critical section, it should
 check for each waiting gate if its waiting condition holds and if there are
-threads trying to get through the gate.  If there is such a gate,
-it must select one and open that gate.  If there is no such gate,
+threads trying to get through the gate.  If there is such a gate, then
+it must select one and open that gate.  If there is no such gate, then
 it must open the main gate.
@@
 Finally, if a thread is executing in the critical section and needs to
-wait for a particular condition, it leaves the critical section and waits
-for the gate associated with that condition to open.
-
-Note that the following invariants hold:
+wait for a particular condition, then it leaves the critical section and
+waits for the gate associated with that condition to open.
+
+The following invariants hold:
 \begin{itemize}
 \item At any time, at most one gate is open;
-\item If some gate is open, then no thread is in the critical section.  Equivalently, if some thread is in the critical section, all gates are closed;
+\item If some gate is open, then no thread is in the critical section.  Equivalently, if some thread is in the critical section, then all gates are closed;
 \item At any time, at most one thread is in the critical section.
@@
 \item <{nreaders}>: the number of readers in the critical section;
-\item <{nwriters}>: the number of writers in the critical section;
+\item <{nwriters}>: the number of writers in the critical section (0 or 1);
 \item <{mutex}>: the main gate binary semaphore;
@@
 \texttt{read\_acquire} and \texttt{write\_acquire} each must check to see
-if it has to wait.
+if the thread has to wait.
 If so, it increments the count associated with its respective gate,
@@
 <{release_one()}> first checks the readers' gate and then the
-writers' gate, but the other way around would have worked as well.
+writers' gate, but the other way around works as well.
 If neither waiting gate qualifies, then <{release_one()}> 
@@
 Let us examine \texttt{read\_acquire} more carefully.
-First the method acquires the <{mutex}>.
+First, the method acquires <{mutex}>.
 Then, in the case that the thread
@@
 ($\mathit{nwriters > 0}$), it increments the counter associated with
-the readers' gate, leaves the critical section (\texttt{release\_one}),
-and then tries to acquire the semaphore associated with the waiting gate.
-This will cause it to block until some thread opens that gate.
+the readers' gate, leaves the critical section (\texttt{release\_one}), and
+then tries to acquire the binary semaphore associated with the waiting gate.
+This causes the thread to block until some other thread opens that gate.
 
@@
 and so it releases <{mutex}>.
-There are now two threads that are holding the reader/writer lock.
+There are now two reader threads that are holding the reader/writer lock.
 \end{enumerate}
@@
 % Implement counting semaphores using the split binary semaphores technique.
-\item Several of the instantiations of <{release_one()}> in
+\item Several of the calls to <{release_one()}> in
 \autoref{fig:RWsplitsema} can be replaced by simply releasing <{mutex}>.
@@
 1 through <{N}>.  Method \texttt{gpuAlloc()} returns the identifier
-of an available GPU, blocking if there is currently no available GPU.
+of an available GPU, blocking if there is currently no GPU available.
 Method \texttt{gpuRelease}(\textit{gpu}) releases the given GPU.  It never needs

### 2021-12-24 book/paper.tex
Pass over Ch 17

@@
 from acquiring the lock yet again.
-%
-(It is worth noting that Peterson's Algorithm (\autoref{ch:peterson})
+
+Peterson's Algorithm (\autoref{fig:peterson})
 does not suffer from starvation, thanks to the \texttt{turn} variable
 that alternates between 0 and 1 when two threads are contending for
-the critical section.)
+the critical section.
+Ticket locks (\autoref{fig:ticket}) are also free from starvation.
 

### 2021-12-24 book/paper.tex
Pass over Ch 18

@@
 
-Tony Hoare, who came up with the concept of split binary semaphores, devised
-an abstraction of the concept in a programming language paradigm called
-\emph{monitors}~\cite{Hoare74}.
+Tony Hoare, who came up with the concept of split binary semaphores (SBS),
+devised an abstraction of the concept in a programming language paradigm
+called \emph{monitors}~\cite{Hoare74}.
 \index{monitor}%
@@
 and
-\texttt{signal}.
+<{signal}>.
 \index{signal}%
 
-Harmony does not have language support for Hoare monitors but it
+Harmony does not have language support for monitors, but it
 has a module called \texttt{hoare}.
 \autoref{fig:hoare} shows its implementation.
-A monitor uses a hidden split binary semaphore.
+A Hoare monitor uses a hidden split binary semaphore.
 The mutex semaphore is acquired when entering a monitor and released
@@
 the number of threads waiting on the condition.
-<{wait}> increments the condition's counter, releases the monitor mutex,
+Method <{wait}> increments the condition's counter, releases the monitor mutex,
 blocks while trying to acquire the condition's semaphore, and upon resuming
-decrements the counter---in much the same way as we have seen for split binary
-semaphores (SBS).
-\texttt{signal} checks to see if the condition's count is non-zero, if so
+decrements the counter---in much the same way as we have seen for SBS.
+Method <{signal}> checks to see if the condition's count is non-zero, if so
 releases the condition's semaphore,
@@
 that <{put}> guarantees that the queue will be non-empty, and
-\texttt{signal} will check if there are any threads waiting for
-this event.  If so, \texttt{signal} will pass control to one such thread
+<{signal}> will check if there are any threads waiting for
+this event.  If so, <{signal}> will pass control to one such thread
 and, unlike \texttt{release\_one}, re-enter the critical
@@
 which complicates the code.  Also, it requires a deep understanding of
-the SBS method to remember to place a call to \texttt{signal} in the
+the SBS method to remember to place a call to <{signal}> in the
 \texttt{read\_acquire} method that releases additional readers that
@@
 Like in Hoare monitors, the <{wait}> operation releases the mutex.
-The most important difference is in what \texttt{signal} does.
+The most important difference is in what <{signal}> does.
 To make the distinction more clear, we shall call the corresponding Mesa
-operation <{notify}> rather than \texttt{signal}.
+operation <{notify}> rather than <{signal}>.
 \index{notify}%
-When a thread <{p}> invokes <{notify}>, it does not immediately pass 
+Unlike <{signal}>, when a thread <{p}> invokes <{notify}>
+it does not immediately pass 
 control to a thread that is waiting on the corresponding condition (if there
-is such a thread).  Instead, <{p}> remains in the critical section
-until it leaves the monitor explicitly
-(by calling either <{release}> or <{wait}>).
-At that point, any thread that
-was notified will have a chance to enter the critical section, but they compete
+is such a thread).  Instead, <{p}> continues executing in the critical section
+until it leaves the monitor (by calling <{release}>) or
+releases the monitor (by calling <{wait}>).
+Either way, any thread that was notified will now have a chance to enter
+the critical section, but they compete
 with other threads trying to enter the critical section.
@@
 On the right is a Mesa monitor.  The only way into the bathroom is through
-the hall.  When thread <{p}> invokes \textbf{notify} on a bedroom, it
+the hall.  When thread <{p}> invokes <{notify}> on a bedroom, it
 selects a thread <{q}> in that bedroom, if any.
@@
 enter the bathroom.
-If there are no threads in that bedroom, \textbf{notify} is a no-op.
+If there are no threads in that bedroom, <{notify}> is a no-op.
 Importantly, in either case <{p}> remains in the bathroom.
@@
 the elements of the bag to their multiplicities.
-For example, the value \texttt{\{ .a: 2, .b: 3 \}} represents a bag
-with two copies of \texttt{.a} and three copies of \texttt{.b}.
-The \texttt{bag} module (\autoref{ap:bag}) contains a variety of handy
+For example, the value <{{ .a: 2, .b: 3 }}> represents a bag
+with two copies of <{.a}> and three copies of <{.b}>.
+The <{bag}> module (\autoref{ap:bag}) contains a variety of handy
 functions on bags.
 
-<{wait}> adds the context of the thread to the bag.
+Method <{wait}> adds the context of the thread to the bag.
 This increments the number of threads in the bag with the same context.
@@
 value that it had upon entry to <{wait}>.
-<{notify}> removes an arbitrary context from
+Method <{notify}> removes an arbitrary context from
 the bag, allowing one of the threads with that context to
@@
 As a rule of thumb, one should always be able to replace <{wait}>
-by <{unlock}> followed by <{lock}>.
+by <{release}> followed by <{acquirire}>.
 This turns the solution into a busy-waiting one, inefficient but still
@@
 % Mesa monitors are much closer to the busy-waiting approach.
-When using Mesa condition variable you have to be careful to invoke
-\textbf{notify} or \textbf{notifyAll} in the right places.
+When using Mesa condition variables, you have to be careful to invoke
+<{notify}> or <{notifyAll}> in the right places.
 Much of the complexity of programming with Mesa condition variables

### 2021-12-24 book/paper.tex
Started working on deadlock chapter

@@
 forks.  Each philosopher repeats this procedure.  Dijkstra had them repeating
-this for ever, but for the purposes of this book, philosophers can leave
-the table when they are not using any forks.
+this for ever, but for the purposes of this book,
+philosophers can---if they wish---leave the table when they are
+not using any forks.
 
@@
 \begin{code}
-<{:input:../code/dinersfix.hny}>
+<{:inline:}>
+if left < right:
+    synch.acquire(?forks[left])
+    synch.acquire(?forks[right])
+else:
+    synch.acquire(?forks[right])
+    synch.acquire(?forks[left])
+<{:end:}>
 \end{code}
@@
 \begin{code}
-<{:input:../code/dinersfix2.hny}>
+<{:inline:}>
+synch.acquire(?forks[min(left, right)])
+synch.acquire(?forks[max(left, right)])
+<{:end:}>
 \end{code}

### 2021-12-25 book/paper.tex
Working on updating paper.tex to use new preprocessor

@@
 that sorts all lists of up to five numbers correctly but fails to
-do so for a list of 6 numbers.  (Hint: simply use an \textbf{if}
+do so for a list of 6 numbers.  (Hint: simply use an <{if}>
 statement.)
@@
 
-The $\textbf{assert}$ statement checks that the output is
+The <{assert}> statement checks that the output is
 correct.  If the program is correct, Harmony reports the size of the
@@
 It sets <{shared}> to <{True}>, asserts that
-<{shared}> = <{True}> and finally sets <{shared}> to <{False}>.
+<{shared = True}> and finally sets <{shared}> to <{False}>.
 If you run the program through Harmony, it will not find any problems
@@
     in broad steps, how running the program can get into a bad state.
-\item What if we moved line~5 of \autoref{fig:incenter} to after the \textbf{if}
+\item What if we moved line~5 of \autoref{fig:incenter} to after the <{if}>
     statement (between lines~7 and~8)?  Do you think that would work?  Run it through
@@
 in \autoref{fig:peterson}.
-(The first line specifies that the \textit{flags} and <{turn}>
+(The first line specifies that the <{flags}> and <{turn}>
 variables are assumed to satisfy \emph{sequential consistency}---it prevents
@@
 The HVM starts in an initial state in which there is only
-one thread (<{__init__}>()) and its program counter is~0.
+one thread (<{__init__()}>) and its program counter is~0.
 A \emph{trace}
@@
 \autoref{fig:petersonproof} formalizes $\mathcal{G}(i)$ in Harmony.
-The label \texttt{gate} refers to line 10, that is,
+The label <{gate}> refers to line 10, that is,
 the step that sets <{turn}> to $1-i$.
@@
 the critical section.  By $\mathcal{F}$ it must be the case that both
-\textit{flags} are true.  By $\mathcal{G}$ and the fact that neither thread
-is at label \texttt{gate}, we know that both $C(0)$ and $C(1)$ must hold.
+<{flags}> are true.  By $\mathcal{G}$ and the fact that neither thread
+is at label <{gate}>, we know that both $C(0)$ and $C(1)$ must hold.
 This then implies that $\mathit{turn} = 0 \land \mathit{turn} = 1$, providing
@@
 \begin{enumerate}
-\item \textit{flags}[1] = <{False}>;
-\item <{turn}> = 0;
-\item thread~1 is at label \texttt{gate}.
+\item <{flags[1] = False}>;
+\item <{turn = 0}>;
+\item thread~1 is at label <{gate}>.
 \end{enumerate}
@@
 continues to hold), or $flags[1]$ will be <{True}>
-and thread~1 will be at label \texttt{gate} (in which case the third case
+and thread~1 will be at label <{gate}> (in which case the third case
 will hold).
@@
 if the second case holds before the step, it will also hold after the step.
-Finally, if thread~1 is at label \texttt{gate} before the step, then after
+Finally, if thread~1 is at label <{gate}> before the step, then after
 the step <{turn}> will equal 0, and therefore the second case will hold
@@
 if thread <{i}> is in the critical section, then $\mathcal{C}(i)$ (i.e.,
-$\mathcal{G}$ without the disjunct that thread $1-i$ is at label \texttt{gate}).
+$\mathcal{G}$ without the disjunct that thread $1-i$ is at label <{gate}>).
 We already saw that this is not an invariant.
@@
 In particular, <{!?x = x}>.
-This is similar to how C pointers work (\texttt{*\string&}<{x}> = <{x}>).
+This is similar to how C pointers work (\texttt{*\string&}<{x = x}>).
 
@@
 condition holds (not necessarily the first time) and then executes the
-statement block.  The ``\texttt{await} \textit{condition}'' statement is
-the same as ``<{when}> \textit{condition}: <{pass}>''.
+statement block.  The ``<{await condition}>'' statement is
+the same as ``<{when condition: pass}>''.
 Combined with the <{atomically}> keyword, the entire statement
@@
 operation.
-Method \texttt{test\_and\_set} in \autoref{fig:tas} shows its specification.
+Method <{test_and_set}> in \autoref{fig:tas} shows its specification.
 Here <{s}> points to a shared boolean variable and <{p}>
@@
 Variable <{shared}> is initialized to <{False}> while
-\textit{private}[<{i}>] for each thread <{i}> is initialized to <{True}>.
+<{private}[i]}> for each thread <{i}> is initialized to <{True}>.
 
@@
 Another invariant, $\mathcal{I}_2(i)$, is that if thread <{i}> is
-in the critical section,
-then \textit{private}[<{i}>] = <{False}>.
+in the critical section, then <{private}[i] = False}>.
 Between the two (i.e., $\mathcal{I}_1 \land \forall i: \mathcal{I}_2(i)$),
@@
 To see that invariant $\mathcal{I}_1$ is maintained, note that
-\texttt{!}<{p}> = <{True}> upon entry of \texttt{test\_and\_set}
+<{!p = True}> upon entry of <{test_and_set}>
 (because of the condition on the <{while}> loop that the
-\texttt{test\_and\_set} method is invoked in).
+<{test_and_set}> method is invoked in).
 There are two cases:
 \begin{enumerate}
-\item \texttt{!}<{s}> is <{False}> upon entry to \texttt{test\_and\_set}.
-Then upon exit \texttt{!}<{p}> = <{False}> and \texttt{!}<{s}> = <{True}>,
+\item <{!s}> is <{False}> upon entry to <{test_and_set}>.
+Then upon exit <{!p = False}> and <{!s = True}>,
 maintaining the invariant.
-\item \texttt{!}<{s}> is <{True}> upon entry to \texttt{test\_and\_set}.
+\item <{!s}> is <{True}> upon entry to <{test_and_set}>.
 Then upon exit nothing has changed, maintaining the invariant.
@@
 Invariant $\mathcal{I}_1$ is also easy to verify for exiting the critical section
-because we can assume, by the induction hypothesis, that $\textit{private}[i]$ is
+because we can assume, by the induction hypothesis, that $<{private}>[i]$ is
 <{True}> just before exiting the critical section.
 Invariant $\mathcal{I}_2(i)$ is obvious as (i) thread <{i}> only proceeds to the critical
-section if \textit{private}[<{i}>] is <{False}>, and (ii) no other thread modifies
-\textit{private}[<{i}>].
+section if <{private}>[<{i}>] is <{False}>, and (ii) no other thread modifies
+<{private}>[<{i}>].
 
@@
 section it has the lock.
-There is no need to keep \textit{private} as a shared
+There is no need to keep <{private}> as a shared
 variable---we only did so to be able to show and check the invariants.
@@
 \autoref{fig:ticket} similarly uses two variables for a lock,
-\textit{counter} and \textit{dispenser}.  When a thread acquires the
+<{counter}> and <{dispenser}>.  When a thread acquires the
 lock, it fetches the current dispenser value and increments it
-modulo \texttt{MAX\_THREADS}, all in one atomic operation.
-In practice, \texttt{MAX\_THREADS} would
+modulo <{MAX_THREADS}>, all in one atomic operation.
+In practice, <{MAX_THREADS}> would
 be a number like $2^{32}$ or $2^{64}$, but since the Harmony model checker
-checks every possible state, limiting \texttt{MAX\_THREADS}
+checks every possible state, limiting <{MAX_THREADS}>
 to a small number
@@
 Plus it is easier to check that it fails when you run it with more than
-\texttt{MAX\_THREADS} threads.
+<{MAX_THREADS}> threads.
 You can test the implementation using the command
@@
 \end{code}
-\caption{\harmonylink{modules/synchS.hny} The lock implementation in the \texttt{synchS} module uses suspension}
+\caption{\harmonylink{modules/synchS.hny} The lock implementation in the <{synchS}> module uses suspension}
 \label{fig:suspension}
@@
 \item
-<{acquire}>()
+<{acquire()}>
 \index{acquire}%
@@
 \item
-<{release}>()
+<{release()}>
 \index{release}%
@@
 \item
-Run \autoref{fig:incfixed} using (i) <{synch}> and (ii) \texttt{synchS}.
+Run \autoref{fig:incfixed} using (i) <{synch}> and (ii) <{synchS}>.
 Report how many states were explored by Harmony for each module.
@@
 should return a pair that sums to 100.  Add the necessary synchronization code.
-\item \label{ex:trylock} Implement \texttt{tryAcquire}(\textit{b}) as an additional
-interface for both the <{synch}> and \texttt{synchS} modules.
-This interface is like <{acquire}>(\textit{b}) but never blocks.  It
+\item \label{ex:trylock} Implement <{tryAcquire(b)}> as an additional
+interface for both the <{synch}> and <{synchS}> modules.
+This interface is like <{acquire(b)}> but never blocks.  It
 returns <{True}> if the lock was available (and now acquired) or <{False}>
@@
 \index{dynamic allocation}%
-of nodes in the list using \texttt{malloc}() and \texttt{free}().
-\texttt{malloc}(<{v}>) returns a new memory location initialized to <{v}>,
-which should be released with \texttt{free}() when it is no longer in use.
+of nodes in the list using <{malloc()}> and <{free()}>.
+<{malloc(v}>) returns a new memory location initialized to <{v}>,
+which should be released with <{free()}> when it is no longer in use.
 The queue maintains a <{head}> pointer to the first element in the list
@@
 hand-over-hand.
-Using \texttt{\_find}, the <{insert}>, \texttt{remove}, and \texttt{contains}
+Using \texttt{\_find}, the <{insert}>, <{remove}>, and \texttt{contains}
 methods are fairly straightforward to implement.
@@
 \begin{itemize}
-\item[] \textit{specq}: the queue of the specification;
-\item[] \textit{implq}: the queue of the implementation.
+\item[] <{specq}>: the queue of the specification;
+\item[] <{implq}>: the queue of the implementation.
 \end{itemize}
@@
 being a \texttt{dummy} node to prevent data races.  Each node in the
-list contains a lock.  The <{put}>() method walks the list until
+list contains a lock.  The <{put()}> method walks the list until
 it gets to the last node, each time acquiring the lock to access the
-node's fields.  When <{put}>() gets to the last node in the list,
-it appends a new one.  The <{get}>() method locks the first (dummy)
+node's fields.  When <{put()}> gets to the last node in the list,
+it appends a new one.  The <{get()}> method locks the first (dummy)
 node, removes the second from the list and frees it.  The method returns
@@
 it uses for dynamic allocation.  Apparently T2 tried to access
-<{pool}>[0], but it does not exist, meaning that either it was not
+<{pool[0]}>, but it does not exist, meaning that either it was not
 yet allocated, or it had been freed since it was allocated.
 When we look at the top half of the figure, we see that in fact thread
-T1 allocated <{pool}>[0] in turn 2, but T3 freed it in turn 4.
+T1 allocated <{pool[0]}> in turn 2, but T3 freed it in turn 4.
 Looking back down, we see that T1 executed <{thread(1)}> and has
@@
 within <{put(?q, 2)}>, but 
-<{alloc.pool}>[0] does not exist.
+<{alloc.pool[0]}> does not exist.
 The corresponding line of Harmony code is
-\textbf{atomically when not} !\textit{binsema}
+<{atomically when not !binsema}>
 in line 25 of the \texttt{sync} module.
@@
 
-\begin{comment}
-\autoref{fig:pc} gives a high-level description.  The
-two numbers in a state specify
-the number of items that the producers still can produce and
-the number of items that the consumers have consumed.  In this
-description there are three items to produce initially.
-
-\begin{figure}
-\begin{code}
-<{:input:../code/BBsemadata.hny}>
-\end{code}
-\caption{\harmonylink{code/BBsemadata.hny} Test whether the bounded buffer delivers the right data in the
-right order}
-\label{fig:BBsemadata}
-\end{figure}
-
-\autoref{fig:boundedbuffer} presents the implementation of a bounded
-buffer using semaphores.  It features a circular bounded buffer \textit{buf} with
-slots numbered 1 through \texttt{NSLOTS}.  There are two indexes into
-the buffer: \textit{b\_in} specifies where the next produced item is inserted,
-while \textit{b\_out} specifies from where the next can be consumed.
-As there may be multiple producers and multiple
-consumers, updates to the indexes, which are shared variables, must be protected.
-To this end, we use two semaphores as locks: \textit{l\_in} for \textit{b\_in}
-and \textit{l\_out} for \textit{b\_out}.
-
-Without additional synchronization, the indexes may overflow and point to invalid
-entries in the circular buffer.
-For this, there is a semaphore \textit{n\_full} that
-keeps track of the number of filled entries in the buffer and a semaphore
-\textit{n\_empty} that keeps track of the number of empty entries in the buffer.
-
-To add an item to the bounded buffer (\texttt{produce}(\textit{item})), the producing
-thread first has to wait until there is room in the bounded buffer.
-To this end, it invokes <{P}>(\textit{n\_empty}), which waits until
-$\mathit{n\_empty} > 0$ and atomically decrements \textit{n\_empty} once this
-is the case.  In other words, the producer procures an empty slot.
-Next, it adds the item to the next position in the buffer.
-Since there may be more than one empty slot, multiple producers may attempt
-to do the same thing concurrently.  Hence the use of the \textit{l\_in}
-semaphore used as a lock.  Finally, once the item is added, the thread
-increments the \textit{n\_full} semaphore to signal to consumers that
-an item is available.
-The consumer code is symmetric.
-
-Note the quite different usage of the \textit{l\_} semaphores and \textit{n\_}
-semaphores.
-People often call semaphores that are used as locks \emph{binary semaphores},
-\index{binary semaphore}%
-as they only take on the values 0 and 1.
-They protect critical sections.
-The second type of semaphores are called \emph{counting semaphores}.
-\index{counting semaphore}%
-They can be used to send signals between threads.
-However, binary and counting semaphores are implemented the same way.
-
-\autoref{fig:bbtest} invokes the code of \autoref{fig:boundedbuffer}.
-We can run this through Harmony, but what would it test?  There are
-no assert statements in the code.  We can split the functionality
-of this code into two and test each separately.  One is the
-synchronization part: producers should never run more than
-\texttt{NSLOTS} ahead of the consumers, but they should be able to
-produce up to \texttt{NSLOTS} items even if there are no consumers.
-The other part is the data: we want to make sure that the items are
-sent through the buffer in order and without loss.  We will first
-focus on the synchronization.
-
-The first part can be tested with the code given in \autoref{fig:pc}, by
-experimenting with different non-negative values for the three constants.
-In particular, any run with a positive \texttt{NSLOTS} and
-$\mathtt{NCONSS} \le \mathtt{NPRODS} \le \mathtt{NCONSS} + \mathtt{NSLOTS}$
-should terminate, while other values should lead to threads blocking.
-To this this, one can write a script like the one in \autoref{fig:pcscript}.
-
-\autoref{fig:BBsemadata} tests whether the correct data is sent in
-the correct order.  There are the same number of producers and consumers.
-Each producer <{i}> produces two tuples: $(i, 0)$ and $(i, 1)$ in that order.
-Each consumer consumes two tuples and checks that the tuples are different
-and that, if the two tuples come from the same producer, then they have to
-be in the order sent.
-Moreover, consumers add the tuples they received to the set
-\textit{received} (atomically---a lock could have been used here as well).
-The \texttt{main()} thread waits for all consumers to have received
-exactly the set of tuples that the producers produce.
-
-\section*{Exercises}
-\begin{problems}
-\end{problems}
-\end{comment}
-
 \chapter{Split Binary Semaphores}
@@
 \item <{mutex}>: the main gate binary semaphore;
-\item \textit{r\_gate}: the waiting gate used by readers, consisting of a binary semaphore and the number of readers waiting to enter;
-\item \textit{w\_gate}: the waiting gate used by writers, similar to the readers' gate.
+\item <{r_gate}>: the waiting gate used by readers, consisting of a binary semaphore and the number of readers waiting to enter;
+\item <{w_gate}>: the waiting gate used by writers, similar to the readers' gate.
 \end{itemize}
@@
 Each of the
-\texttt{read\_acquire}, \texttt{read\_release},
-\texttt{write\_acquire}, and \texttt{write\_release} methods must maintain
+<{read_acquire}>, <{read_release}>,
+<{write_acquire}>, and <{write_release}> methods must maintain
 this state.
@@
 After acquiring the <{mutex}>,
-\texttt{read\_acquire} and \texttt{write\_acquire} each must check to see
+<{read_acquire}> and <{write_acquire}> each must check to see
 if the thread has to wait.
@@
 
-Let us examine \texttt{read\_acquire} more carefully.
+Let us examine <{read_acquire}> more carefully.
 First, the method acquires <{mutex}>.
@@
 section and there are two readers waiting.  Let us see what happens when
-the writer calls \texttt{write\_release}:
+the writer calls <{write_release}>:
 \begin{enumerate}
@@
 can acquire the bound lock at the same time.
-% \item Implement a \emph{bounded stack}.  Method \texttt{push}(\textit{item})
+% \item Implement a \emph{bounded stack}.  Method <{push(item}>)
 % should block if the stack is full.  Method \texttt{pop()} should block
@@
 There are <{N}> GPUs identified by the numbers
-1 through <{N}>.  Method \texttt{gpuAlloc()} returns the identifier
+1 through <{N}>.  Method <{gpuAlloc()}> returns the identifier
 of an available GPU, blocking if there is currently no GPU available.
-Method \texttt{gpuRelease}(\textit{gpu}) releases the given GPU.  It never needs
+Method <{gpuRelease(gpu}>) releases the given GPU.  It never needs
 to block.
@@
 to a read lock when its done writing but not done reading.  Add
-a \texttt{downgrade} method to the code in \autoref{fig:RWsplitsema}.
+a <{downgrade}> method to the code in \autoref{fig:RWsplitsema}.
 (Similarly, you may want to try to implement an \emph{upgrade} of a
@@
 method has a custom way of selecting which gate to open.  In particular,
-\texttt{read\_release} prefers the write gate, while \texttt{write\_release}
+<{read_release}> prefers the write gate, while <{write_release}>
 prefers the read gate.
@@
 the SBS method to remember to place a call to <{signal}> in the
-\texttt{read\_acquire} method that releases additional readers that
+<{read_acquire}> method that releases additional readers that
 may be waiting to acquire the lock.
@@
 condition variables in the <{synch}> module.
-<{Condition}>() creates a new condition variable.
+<{Condition()}> creates a new condition variable.
 It is represented by a dictionary containing
 a bag of contexts of threads waiting on the condition variable.
-(The \texttt{synchS} library instead uses a list of contexts.)
+(The <{synchS}> library instead uses a list of contexts.)
 
@@
 
-In \texttt{read\_release}, notice that <{notify}>(?\textit{w\_cond})
+In <{read_release}>, notice that <{notify(?w_cond)}>
 is invoked when there are no readers left, \emph{without} checking
@@
 
-\texttt{write\_release} executes <{notifyAll}>(?\textit{r\_cond})
-as well as <{notify}>(?\textit{w\_cond)}.  Because we
+<{write_release}> executes <{notifyAll(?r_cond)}>
+as well as <{notify(?w_cond)}>.  Because we
 do not keep track of the number of waiting readers or writers, we
@@
 or, alternatively, up to one waiting writer can enter.  So
-\texttt{write\_release} wakes up all potential candidates.  There
+<{write_release}> wakes up all potential candidates.  There
 are two things to note here.  First, unlike split binary semaphores
@@
 \begin{enumerate}
-\item \textit{tl} = \texttt{TryLock}() \# \emph{create a try lock}
-\item <{acquire}>(?\textit{tl}) \# \emph{acquire a try lock}
-\item \texttt{tryAcquire}(?\textit{tl}) \# \emph{attempt to acquire a try lock}
-\item <{release}>(?\textit{tl}) \# \emph{release a try lock}
+\item <{tl = TryLock()}> \# \emph{create a try lock}
+\item <{acquire(?tl}>) \# \emph{acquire a try lock}
+\item <{tryAcquire(?tl}>) \# \emph{attempt to acquire a try lock}
+\item <{release(?tl}>) \# \emph{release a try lock}
 \end{enumerate}
 \noindent
-\texttt{tryAcquire} should not wait.
+<{tryAcquire}> should not wait.
 Instead it should return <{True}> if the lock was successfully
@@
 Within the <{while}> statement,
-there is an \textbf{if} statement for each disjunct.
+there is an <{if}> statement for each disjunct.
 The code waits for either or both forks if necessary.  After that, it goes
@@
 diner.  It uses the same number of condition variables, but you will not
-need to have \textbf{if} statements nested inside the <{while}> loop
+need to have <{if}> statements nested inside the <{while}> loop
 waiting for the forks.  See if you can figure it out.
@@
 \item \texttt{Queue()} returns a new message queue;
-\item <{put(q}>, \textit{item}) adds \textit{item} to the queue pointed to by <{q}>;
+\item <{put(q}>, <{item}>) adds <{item}> to the queue pointed to by <{q}>;
 \item <{get(q}>) waits for and returns an item on the queue pointed to by <{q}>.
@@
 
-The \textit{round} array is kept to check the correctness of this
+The <{round}> array is kept to check the correctness of this
 approach.  Each thread increments its entry every time it enters
@@
 \end{code}
-\caption{\harmonylink{code/trap.hny} How to use \textbf{trap}}
+\caption{\harmonylink{code/trap.hny} How to use <{trap}>}
 \label{fig:trap}
@@
 \end{code}
-invokes \textit{handler argument} at some later, unspecified time.
-Thus you can think of \textbf{trap} as setting a timer.
+invokes <{handler argument}> at some later, unspecified time.
+Thus you can think of <{trap}> as setting a timer.
 Only one of these asynchronous events can be outstanding at a time;
-a new call to \textbf{trap} overwrites any outstanding one.
-\autoref{fig:trap} gives an example of how \textbf{trap} might be used.
-Here, the \textit{main}() thread loops until the interrupt has occurred and
+a new call to <{trap}> overwrites any outstanding one.
+\autoref{fig:trap} gives an example of how <{trap}> might be used.
+Here, the <{main()}> thread loops until the interrupt has occurred and
 the <{done}> flag has been set.
@@
 But now consider \autoref{fig:trap2}.  The difference with \autoref{fig:trap} is
-that both the \textit{main}() and \textit{handler}() methods increment <{count}>.
+that both the <{main()}> and <{handler()}> methods increment <{count}>.
 This is not unlike the example we gave in \autoref{fig:inc}, except that only a single
 thread is involved now.  And, indeed, it suffers from a similar race condition; run
-it through Harmony to see for yourself.  If the interrupt occurs after \textit{main}()
-reads <{count}> (and thus still has value 0) but before \textit{main()} writes the
+it through Harmony to see for yourself.
+If the interrupt occurs after <{main()}>
+reads <{count}> (and thus still has value 0) but before <{main()}> writes the
 updated value~1, then the interrupt handler will also read value~0 and write value~1.
-We say that the code in \autoref{fig:trap2} is not \textit{interrupt-safe} (as opposed
-to not being \textit{thread-safe}).
+We say that the code in \autoref{fig:trap2} is not \emph{interrupt-safe} (as opposed
+to not being \emph{thread-safe}).
 \index{interrupt-safety}
@@
 The issue is that a thread can only acquire a lock once.  If the interrupt happens
-after \textit{main}() acquires the lock but before \textit{main}() releases it, the
-\textit{handler}() method will block trying to acquire the lock, even though
+after <{main()}> acquires the lock but before <{main()}> releases it, the
+<{handler}>() method will block trying to acquire the lock, even though
 it is being acquired by the same thread that already holds the lock.
@@
 Instead, the way one fixes interrupt-safety issues is through disabling interrupts
-temporarily.  In Harmony, this can be done by setting the \textit{interrupt level}
-of a thread to <{True}> using the \textbf{setintlevel} interface.
+temporarily.  In Harmony, this can be done by setting the \emph{interrupt level}
+of a thread to <{True}> using the <{setintlevel}> interface.
 \autoref{fig:trap4} illustrates how this is done.
@@
 handler and restored to <{False}> upon exit.
-It is important that the \textit{main}() code re-enables interrupts after incrementing
-<{count}>.  What would happen if \textit{main}() left interrupts disabled?
-
-\textbf{setintlevel}(\textit{il}) sets the interrupt level to \textit{il} and returns
+It is important that the <{main()}> code re-enables interrupts after incrementing
+<{count}>.  What would happen if <{main()}> left interrupts disabled?
+
+<{setintlevel(il}>) sets the interrupt level to <{il}> and returns
 the prior interrupt level.  Returning the old level is handy when writing interrupt-safe
@@
 To increment <{count}>, the interrupt level must be <{True}> and
-\textit{countlock} must be held.
+<{countlock}> must be held.
 \autoref{fig:trap6} gives an example of how this might be done.
@@
 say, a keyboard interrupt handler could place an event on a shared queue.
-Implement a new method \texttt{i\_put}(\textit{item}) that does not
+Implement a new method \texttt{i\_put}(<{item}>) that does not
 block.  Instead, it should return <{False}> if the buffer is full and <{True}>
@@
 is not necessarily practical.
-There are two variables.  \textit{items} is an unbounded array with each
-entry initialized to <{None}>.  \textit{back} is an index into the
+There are two variables.  <{items}> is an unbounded array with each
+entry initialized to <{None}>.  <{back}> is an index into the
 array and points to the next slot where a new value is inserted.
@@
 \begin{itemize}
-\item \texttt{inc}(<{p}>): atomically increments \texttt{!}<{p}> and returns
+\item <{inc(p}>): atomically increments \texttt{!}<{p}> and returns
 the old value;
-\item \texttt{exch}(<{p}>): sets \texttt{!}<{p}> to <{None}> and returns
+\item <{exch(p}>): sets \texttt{!}<{p}> to <{None}> and returns
 the old value.
@@
 
-Method \texttt{produce}(\textit{item}) uses \texttt{inc}(?\textit{back})
+Method <{produce(item}>) uses <{inc}>(?<{back}>)
 to allocate
-the next available slot in the \textit{items} array.
+the next available slot in the <{items}> array.
 It stores the item as a singleton tuple.
-Method \texttt{consume()} repeatedly scans the array, up to the
-\textit{back} index, trying to find an item to return.
-To check an entry, it uses \texttt{exch()}
+Method <{consume()}> repeatedly scans the array, up to the
+<{back}> index, trying to find an item to return.
+To check an entry, it uses <{exch()}>
 to atomically remove an item from a slot if there is one.
@@
 \begin{itemize}
-\item The \textit{items} array must be of infinite size if an unbounded number
+\item The <{items}> array must be of infinite size if an unbounded number
 of items may be produced;
 \item Each slot in the array is only used once, which is inefficient;
-\item the \texttt{inc} and \texttt{exch} atomic operations are not
+\item the <{inc}> and <{exch}> atomic operations are not
 universally available on existing processors.
@@
 \index{sequence number}%
-\textit{s\_seq} and \textit{r\_seq} resp.
+<{s_seq}> and <{r_seq}> resp.
 Alice sends a network message to Bob containing an
@@
 channel as a variable that either contains a network message or nothing
-(we use <{()}> in the model).  Let \textit{s\_chan} be the channel
-from Alice to Bob and \textit{r\_chan} the channel from Bob to Alice.
-\texttt{net\_send}(\textit{pchan}, <{m}>, \textit{reliable}) models sending
-a message <{m}> to
-\texttt{!}\textit{pchan}, where \textit{pchan} is either ?\textit{s\_chan}
-or ?\textit{r\_chan}.
+(we use <{()}> in the model).  Let <{s_chan}> be the channel
+from Alice to Bob and <{r_chan}> the channel from Bob to Alice.
+<{net_send(pchan, m, reliable)}> models sending a message <{m}> to
+<{!pchan}>, where <{pchan}> is either <{?s_chan}> or <{?r_chan}>.
 The method places either <{m}> (to model a successful send)
-or <{()}> (to model loss) in \texttt{!}\textit{pchan}.
-The use of the \textit{reliable} flag will be explained later.
-\texttt{net\_recv}(\textit{pchan}) models checking \texttt{!pchan} for
+or <{()}> (to model loss) in <{!pchan}>.
+The use of the <{reliable}> flag will be explained later.
+\texttt{net\_recv}(<{pchan}>) models checking <{!pchan}> for
 the next message.
 
-Method \texttt{app\_send}(<{m}>) retransmits <{m}> until
+Method <{app_send(m}>) retransmits <{m}> until
 an acknowledgment is received.
-Method \texttt{app\_recv}(\textit{reliable}) returns the next successfully received
+Method <{app_recv(reliable}>) returns the next successfully received
 message.
@@
 
-Only the last invocation of \texttt{app\_recv}(\textit{reliable}) is invoked with
-\textit{reliable} = <{True}>.  It causes the last acknowledgment to be sent
+Only the last invocation of <{app_recv(reliable)}> is invoked with
+<{reliable = True}>.  It causes the last acknowledgment to be sent
 reliably.  It allows the receiver (Bob) to stop, as well as the sender (Alice)
@@
 Messages are tuples with three fields:
-(\textit{dst}, \textit{id}, \textit{found}).
-\textit{dst} is the identifier of the destination processor;
-\textit{id} is the identifier that is being forwarded; and
-\textit{found} is a boolean indicating the second phase of the protocol.
-The \texttt{receive}(\textit{self}) method looks for all messages
-destined for the processor with identifier \textit{self}.
+(<{dst}>, <{id}>, <{found}>).
+<{dst}> is the identifier of the destination processor;
+<{id}> is the identifier that is being forwarded; and
+<{found}> is a boolean indicating the second phase of the protocol.
+The \texttt{receive}(<{self}>) method looks for all messages
+destined for the processor with identifier <{self}>.
 
@@
 processor, chosen non-deterministically from a set of \texttt{NIDS}
-identifiers.  It also keeps track in the variable \textit{leader} of
+identifiers.  It also keeps track in the variable <{leader}> of
 what the highest identifier is so it can later be checked.
 
-Method \texttt{processor}(\textit{self}, \textit{succ}) is the
-code for a processor with identifier \textit{self} and successor
-\textit{succ}.
+Method \texttt{processor}(<{self}>, <{succ}>) is the
+code for a processor with identifier <{self}> and successor
+<{succ}>.
 It starts simply by sending its own identifier to its successor.
@@
 \begin{itemize}
-\item \textit{self}: the bank's identifier;
-\item \textit{balance}: the current balance in the account;
-\item \textit{status}: either contains () if the bank is not involved in an
+\item <{self}>: the bank's identifier;
+\item <{balance}>: the current balance in the account;
+\item <{status}>: either contains () if the bank is not involved in an
 ongoing transaction or contains the balance
 just before the transaction started;
-\item \textit{received}: the set of messages received and handled so far.
+\item <{received}>: the set of messages received and handled so far.
 \end{itemize}
@@
 In case of a withdrawal when the bank is idle and there are sufficient
-funds, the bank saves the current balance in \textit{status} to indicate
+funds, the bank saves the current balance in <{status}> to indicate
 an ongoing transaction and what its original balance was.
@@
 to (), indicating that there is no ongoing transaction.  In case of
-a \texttt{.abort} message, the bank restores \textit{balance} first.
+a \texttt{.abort} message, the bank restores <{balance}> first.
 
@@
 The \texttt{transfer}() method contains the code for the transfer
-transaction.  Argument \textit{self} is the identifier of the coordinator,
-\textit{b1} is the source bank, \textit{b2} is the destination bank,
-and \textit{amt} is the amount to transfer.
+transaction.  Argument <{self}> is the identifier of the coordinator,
+<{b1}> is the source bank, <{b2}> is the destination bank,
+and <{amt}> is the amount to transfer.
 The coordinator sends a \texttt{PREPARE} message containing a
-\texttt{.withdraw} request to \textit{b1} and a \texttt{PREPARE}
-message containing a \texttt{.deposit} request to \textit{b2}.
+\texttt{.withdraw} request to <{b1}> and a \texttt{PREPARE}
+message containing a \texttt{.deposit} request to <{b2}>.
 It then waits for responses from each.  If both responses are
@@
 The \texttt{check}() method checks if the sum of the balances equals
-\textit{total}, the sum of the initial balances.
-The code is similar to \textit{transfer}, except that it always
+<{total}>, the sum of the initial balances.
+The code is similar to <{transfer}>, except that it always
 aborts the transaction---there is no need to ever commit it.
@@
 While the protocol perhaps seems simple enough, there are a lot of
-\textbf{if} statements in the code, making it hard to reason about
+<{if}> statements in the code, making it hard to reason about
 correctness.
@@
 $\mathbf{choose}(\{ \mathit{immortal}, <{True}> \})$ can only
-evaluate to <{False}> in case $\textit{immortal}$ is false; otherwise
+evaluate to <{False}> in case $<{immortal}>$ is false; otherwise
 it will always evaluate to <{True}>.
@@
 Messages have the format $(\mathit{source}, \mathit{payload})$,
-where \textit{source} is the identifier of the replica that generated
-the message and \textit{payload} has information about the state of
+where <{source}> is the identifier of the replica that generated
+the message and <{payload}> has information about the state of
 the replica.  The payload can take one of two forms.  The first is
@@
 We are not modeling clients of the service.
-Instead, in the model, the head adds its identifier \textit{self} to the
+Instead, in the model, the head adds its identifier <{self}> to the
 history until the
@@
 history without crashing, the code makes sure that any other such replica
-ends in the same state.  Variable \textit{final} is used to check the
+ends in the same state.  Variable <{final}> is used to check the
 different histories against one another.
@@
 work by Hagit Attiya, Amotz Bar-Noy, and Danny Dolev~\cite{ABD95}.
-In order to tolerate \texttt{F} failures,
+In order to tolerate <{F}> failures,
 it uses $\mathtt{N} = 2\mathtt{F} + 1$ servers.
@@
 a client broadcasts a request to all servers and waits for responses
-from a majority ($<{N}> - \texttt{F}$ or equivalently
-$\texttt{F} + 1$ servers).  Note that because we are assuming that no
-more than \texttt{F} servers can fail, doing so is safe, in that a client
+from a majority ($<{N}> - <{F}>$ or equivalently
+$<{F}> + 1$ servers).  Note that because we are assuming that no
+more than <{F}> servers can fail, doing so is safe, in that a client
 cannot indefinitely block as long as that assumption holds.
@@
 Read and write operations are both invoked with a unique identifier
-\textit{uid}.  Both start by broadcasting a \texttt{.read}
+<{uid}>.  Both start by broadcasting a \texttt{.read}
 request to all servers and then waiting for a response from
@@
 This includes executions in which the clients terminate before
-$\texttt{F}$ of the servers start executing.  To the clients, this
+$<{F}>$ of the servers start executing.  To the clients, this
 is indistinguishable from executions in which those
@@
 More precisely,
-constant \texttt{F} contains the maximum number of failures, and
-we will assume there are $<{N}> = 3\texttt{F} + 1$ processors.
+constant <{F}> contains the maximum number of failures, and
+we will assume there are $<{N}> = 3<{F}> + 1$ processors.
 Each processor \emph{proposes} a value, which we assume here to be from the
@@
 The invariants we want to maintain
-on those variables is that $\textit{decisions} \subseteq \textit{proposals}$
-and $|\textit{decisions}| \leq 1$.  Since the initial value of \textit{decisions} is
+on those variables is that $<{decisions}> \subseteq <{proposals}>$
+and $|<{decisions}>| \leq 1$.  Since the initial value of <{decisions}> is
 $\emptyset$, the invariants clearly hold initially.
@@
 The $\mathtt{N}$ processors go through a sequence of \emph{rounds} in which they wait for
-$<{N}> - \texttt{F}$ messages, update their state based on the messages, and
+$<{N}> - <{F}>$ messages, update their state based on the messages, and
 broadcast messages containing their new state.
-The reason that a processor waits for $<{N}> - \texttt{F}$ rather than <{N}>
-messages is because of failures: up to $\texttt{F}$ processors may never send a message
+The reason that a processor waits for $<{N}> - <{F}>$ rather than <{N}>
+messages is because of failures: up to $<{F}>$ processors may never send a message
 and so it would be unwise to wait for all $<{N}>$.  You might be tempted to use
@@
 that messages arrive within a particular time.
-We call a set of $<{N}> - \texttt{F}$ processors a \emph{quorum}.
+We call a set of $<{N}> - <{F}>$ processors a \emph{quorum}.
 A quorum must suffice for the algorithm to make progress.
@@
 
-In Line~22, a processor waits for $<{N}> - \texttt{F}$ messages using
+In Line~22, a processor waits for $<{N}> - <{F}>$ messages using
 the Harmony \textbf{atomically when exists} statement.
 Since Harmony has to check all possible executions of the protocol, the
-\texttt{receive}(\textit{round}, <{k}>) method returns all \emph{subbags} of messages
-for the given round that have size $k = <{N}> - \texttt{F}$.  The method uses a
-dictionary comprehension to filter out all messages for the given \textit{round}
+\texttt{receive}(<{round}>, <{k}>) method returns all \emph{subbags} of messages
+for the given round that have size $k = <{N}> - <{F}>$.  The method uses a
+dictionary comprehension to filter out all messages for the given <{round}>
 and then uses the \texttt{bag.combinations} method to find all combinations of size <{k}>.
 The \textbf{atomically when exists} statement waits until there is at least one such combination and
-then chooses an element, which is bound to the \textit{quorum} variable.
+then chooses an element, which is bound to the <{quorum}> variable.
 The body of the statement is then executed atomically.
@@
 The body of the \textbf{atomically when exists} statement contains the core of the algorithm.
-Note that $<{N}> - \texttt{F} = 2\texttt{F} + 1$, so that the number of
+Note that $<{N}> - <{F}> = 2<{F}> + 1$, so that the number of
 messages is guaranteed to be odd.  Also, because there are only 0 and 1 values, there
@@
 processors only choose each other messages and essentially ignore the
-messages of the remaining \texttt{F} processors, so the resulting model
+messages of the remaining <{F}> processors, so the resulting model
 is just as good.
@@
 A more challenging class to tolerate are \emph{arbitrary failures} in
-which up to \texttt{F} processors may send arbitrary messages, including
+which up to <{F}> processors may send arbitrary messages, including
 conflicting messages to different peers (equivocation).
@@
 It uses two kinds of processors: \emph{leaders} and \emph{acceptors}.
-In order to tolerate \texttt{F} crash failures, you need at least $\texttt{F}+1$ leaders
-and $2\texttt{F} + 1$ acceptors, but leaders and acceptors can be colocated, so
-in total only $2\texttt{F} + 1$ independently failing processors are needed.
+In order to tolerate <{F}> crash failures, you need at least $<{F}>+1$ leaders
+and $2<{F}> + 1$ acceptors, but leaders and acceptors can be colocated, so
+in total only $2<{F}> + 1$ independently failing processors are needed.
 Here we provide only a rudimentary introduction to Paxos; for more detailed
@@
 containing the ballot number and a proposal and then waits for
-$<{N}> - \texttt{F}$ matching \texttt{2.B} responses from the
+<{N - F}> matching \texttt{2.B} responses from the
 acceptors.  If each response contains the ballot number and the
@@
 number.  This is because acceptors maintain two state variables.
-One is \textit{ballot}, the highest ballot number they have seen.
-Second is a variable called \textit{last\_accepted} that, if not
+One is <{ballot}>, the highest ballot number they have seen.
+Second is a variable called <{last_accepted}> that, if not
 <{None}>, contains the last proposal the acceptor has
@@
 2, then the acceptor accepts the proposal and remembers when it did
-so by saving the (ballot, proposal) tuple in \textit{last\_accepted}.
+so by saving the (ballot, proposal) tuple in <{last_accepted}>.
 In all cases, the acceptor responds with the current values of
-\textit{ballot} and \textit{last\_accepted}.
+<{ballot}> and <{last_accepted}>.
 
@@
 have been decided.  To this end, the leader broadcasts a \texttt{2.A}
-message to the acceptors and awaits $\mathtt{N} - \mathtt{F}$ of their
-\textit{last\_accepted}
+message to the acceptors and awaits <{N - F}> of their <{last_accepted}>
 values.  If all those acceptors responded with <{None}>, then the leader
@@
 
-Harmony supports \textbf{if} statements.
-In its most basic form, \textbf{if} $c: S_1; S_2; ...$ evaluates <{c}>
+Harmony supports <{if}> statements.
+In its most basic form, <{if}> $c: S_1; S_2; ...$ evaluates <{c}>
 and executes statements $S_1, S_2, ...$ if and only if boolean expression
@@
 is the case, Harmony reports an error.
-The statement \textbf{if} $c: S_1, S_2, ...$ \textbf{else}:
+The statement <{if}> $c: S_1, S_2, ...$ <{else}>:
 $T_1; T_2; ...$ is similar, but executes statements $T_1, T_2, ...$
@@
 You can think of \textbf{elif} $c:$ as shorthand for
-\textbf{else}: \textbf{if} $c:$.
+<{else}>: <{if}> $c:$.
 
@@
 
-The statement \texttt{spawn} <{m a}> starts a new thread that executes
+The statement <{spawn}> <{m a}> starts a new thread that executes
 method <{m}> with argument <{a}>. <{m}> must be a program counter constant,
@@
 
-The default thread-local state of the thread, called \textit{self},
+The default thread-local state of the thread, called <{self}>,
 is the empty dictionary by default.  It can be specified by adding
-a parameter: \texttt{spawn} <{m}> $a, e$ specifies that <{e}> should be
+a parameter: <{spawn}> <{m}> $a, e$ specifies that <{e}> should be
 the initial value of the thread-local state.
@@
 
-\subsection*{\textbf{trap}}
-
-The statement \textbf{trap} <{m a}> specifies that the current thread
+\subsection*{<{trap}>}
+
+The statement <{trap m a}> specifies that the current thread
 should execute method <{m}> with argument <{a}> and some future, unspecified,
@@
 the interrupt level of the thread to <{True}> using the
-\textbf{setintlevel} operator.
-
-\subsection*{\textbf{var}}
+<{setintlevel}> operator.
+
+\subsection*{<{var}>}
 
@@
 <{var}> statement.
-The statement \textbf{var} $a = b$ evaluates <{b}> and assigns
+The statement <{var a = b}> evaluates <{b}> and assigns
 the result to local variable <{a}>.
-\textbf{var} supports pattern matching, so you can
-write <{var}> $x, (y, z) = b$.  This will only work
+<{var}> supports pattern matching, so you can
+write <{var x, (y, z) = b}>.  This will only work
 if <{b}> is a tuple with two elements, the second of which also being a
@@
 expressed in C or some other low-level language that does not support
-object-orientation, but instead use \texttt{malloc} and \texttt{free}.
+object-orientation, but instead use <{malloc}> and <{free}>.
 
@@
 <{result}> variable, which is initialized to <{None}>.
-Each thread has a mutable variable called \textit{this} that contains
+Each thread has a mutable variable called <{this}> that contains
 the thread-local state.
@@
 In Harmony, there is no syntactic difference between applying an argument to a function
-or an index to a dictionary.  Both use the syntax \textit{a b c ...}.
+or an index to a dictionary.  Both use the syntax <{a b c ...}>.
 We call this \emph{application}, and application is left-associative.
-So, \textit{a b c} is interpreted as (\textit{a b}) <{c}>: <{b}> is applied to <{a}>,
+So, <{a b c}> is interpreted as (<{a b}>) <{c}>: <{b}> is applied to <{a}>,
 and then <{c}> is applied to the result.
@@
 There are three levels of precedence.
-Application has the highest precedence.  So, \texttt{!}\textit{a b} is interpreted as
-\texttt{!}(\textit{a b}) and \textit{a b} \texttt{+} \textit{c d} is interpreted as
-(\textit{a b}) \texttt{+} (\textit{c d}).
+Application has the highest precedence.  So, \texttt{!}<{a b}> is interpreted as
+\texttt{!}(<{a b}>) and <{a b}> \texttt{+} <{c d}> is interpreted as
+(<{a b}>) \texttt{+} (<{c d}>).
 Unary operators have the next highest precedence,
@@
 In almost all expressions, subexpressions are evaluated left to right.
-So, $a[b] + c$
+So, <{a[b] + c}>
 first evaluates <{a}>, then <{b}> (and then applies <{b}> to <{a}>), and then <{c}>.  The one
-exception is the expression <{a}> \textbf{if} <{c}> \textbf{else} <{b}>, where <{c}> is evaluated
+exception is the expression <{a if c else b}>, where <{c}> is evaluated
 first.  In that expression, only <{a}> or <{b}> is evaluated depending on the value of <{c}>.
@@
 \item[] \autoref{fig:stack1} uses a single list to represent the
-stack.  It is updated to perform \texttt{push} and \texttt{pop} operations;
+stack.  It is updated to perform <{push}> and <{pop}> operations;
 \item[] \autoref{fig:stack2} also uses a list but, instead of updating
@@
 \item[] \autoref{fig:stack3} represents a stack as a recursively nested tuple
-$(v, f)$, where <{v}> is the element on top of the stack and $r$ is a stack
+$(v, f)$, where <{v}> is the element on top of the stack and <{r}> is a stack
 that is the remainder;
@@
 for memory allocated this way and so allocated memory that is no
-longer in use must be explicitly released using \texttt{free}.
+longer in use must be explicitly released using <{free}>.
 
@@
 \hline
-\texttt{malloc(<{v}>)} & return a pointer to a memory location initialized to <{v}> \\
-\hline
-\texttt{free(<{p}>)} & free an allocated memory location <{p}> \\
+<{malloc(v)}> & return a pointer to a memory location initialized to <{v}> \\
+\hline
+<{free(p)}> & free an allocated memory location <{p}> \\
 \hline
@@
 
-The usage is similar to \texttt{malloc} and \texttt{free} in C.
-\texttt{malloc}() is specified to return <{None}> when running out of
+The usage is similar to <{malloc}> and <{free}> in C.
+<{malloc()}> is specified to return <{None}> when running out of
 memory, although this is an impossible outcome in the current
@@
 \hline
-\texttt{empty()} & returns an empty bag\\
+<{empty()}> & returns an empty bag\\
 \hline
@@
 \hline
-\texttt{add}(<{b}>, <{e}>) & add one copy of <{e}> to bag <{b}> \\
-\hline
-\texttt{remove}(<{b}>, <{e}>) & remove one copy of <{e}> from bag <{b}> \\
-\hline
-\texttt{combinations}(<{b}>, <{k}>) & return set of all \emph{subbags} of size <{k}> \\
+<{add(b, e)}> & add one copy of <{e}> to bag <{b}> \\
+\hline
+<{remove(b, e)}> & remove one copy of <{e}> from bag <{b}> \\
+\hline
+<{combinations(b, k)}> & return set of all \emph{subbags} of size <{k}> \\
 \hline
@@
 \hline
-\texttt{issubset(<{s}>, <{t}>)} & returns whether <{s}> is a subset of <{t}> \\
-\hline
-\texttt{issuperset(<{s}>, <{t}>)} & returns whether <{s}> is a superset of <{t}> \\
-\hline
-\texttt{add}(<{s}>, <{e}>) & returns $s \cup \{ e \}$ \\
-\hline
-\texttt{remove}(<{s}>, <{e}>) & returns $s \backslash \{ e \}$ \\
-\hline
-\texttt{combinations}(<{b}>, <{k}>) & returns set of all subsets of size <{k}> \\
+<{issubset(s, t)}> & returns whether <{s}> is a subset of <{t}> \\
+\hline
+<{issuperset(s, t)}> & returns whether <{s}> is a superset of <{t}> \\
+\hline
+<{add(s, e)}> & returns $s \cup \{ e \}$ \\
+\hline
+<{remove(s, e)}> & returns $s \backslash \{ e \}$ \\
+\hline
+<{combinations(b, k)}> & returns set of all subsets of size <{k}> \\
 \hline
@@
 \hline
-\texttt{tas}(\textit{lk}) & test-and-set on \texttt{!}\textit{lk} \\
-\hline
-\texttt{cas}(\textit{ptr}, \textit{old}, \textit{new}) & compare-and-swap on \texttt{!}\textit{ptr} \\
-\hline
-<{BinSem(v}>) & return a binary semaphore initialized to <{v}> \\
-\hline
-<{Lock}>() & return a binary semaphore initialized to <{False}> \\
-\hline
-<{acquire}>(\textit{bs}) & acquire binary semaphore \texttt{!}\textit{bs} \\
-\hline
-<{release}>(\textit{bs}) & release binary semaphore \texttt{!}\textit{bs} \\
-\hline
-<{Condition}>() & return a condition variable \\
-\hline
-<{wait(c}>, \textit{lk}) & wait on condition variable \texttt{!}<{c}>
-and lock \textit{lk} \\ \hline
-<{notify(c}>) & notify a thread waiting on condition variable \texttt{!}<{c}> \\
-\hline
-<{notifyAll(c}>) & notify all threads waiting on condition variable \texttt{!}<{c}> \\
-\hline
-<{Semaphore(cnt}>) & return a counting semaphore initialized to <{cnt}> \\
-\hline
-<{P(sema}>) & procure \texttt{!}<{sema}>  \\
-\hline
-<{V(sema}>) & vacate \texttt{!}<{sema}>  \\
+<{tas(lk)}> & test-and-set on <{!lk}> \\
+\hline
+<{cas}(ptr, old, new)}> & compare-and-swap on <{!ptr}> \\
+\hline
+<{BinSem(v)}> & return a binary semaphore initialized to <{v}> \\
+\hline
+<{Lock()}> & return a binary semaphore initialized to <{False}> \\
+\hline
+<{acquire(bs)}> & acquire binary semaphore \texttt{!}<{bs}> \\
+\hline
+<{release(bs)}> & release binary semaphore \texttt{!}<{bs}> \\
+\hline
+<{Condition()}> & return a condition variable \\
+\hline
+<{wait(c}>, <{lk}>) & wait on condition variable \texttt{!}<{c}>
+and lock <{lk}> \\ \hline
+<{notify(c)}> & notify a thread waiting on condition variable \texttt{!}<{c}> \\
+\hline
+<{notifyAll(c)}> & notify all threads waiting on condition variable \texttt{!}<{c}> \\
+\hline
+<{Semaphore(cnt)}> & return a counting semaphore initialized to <{cnt}> \\
+\hline
+<{P(sema)}> & procure \texttt{!}<{sema}>  \\
+\hline
+<{V(sema)}> & vacate \texttt{!}<{sema}>  \\
 \hline
@@
 \hline
-<{put(q}>, \textit{item}) & add \textit{item} to <{a}> \\
+<{put(q, item)}> & add <{item}> to <{a}> \\
 \hline
@@
 There is initially a single context with name
-\texttt{\_\_init\_\_()} and program counter 0.  It starts executing
+<{__init__()}> and program counter 0.  It starts executing
 in atomic mode until it finishes executing the last
 \texttt{Return} instruction.
-Other threads, created through \texttt{spawn} statements, do not
+Other threads, created through <{spawn}> statements, do not
 start executing until then.
@@
 To that end, the test program maintains a global time variable
-\textit{qtime}, and each thread waits until the timestamp assigned to
-the last concurrent queue operation matches \textit{qtime} before invoking
+<{qtime}>, and each thread waits until the timestamp assigned to
+the last concurrent queue operation matches <{qtime}> before invoking
 the sequential operation in the specification.
-Afterward, it atomically increments the shared \textit{qtime} variable.
+Afterward, it atomically increments the shared <{qtime}> variable.
 This results in the operations being executed sequentially against the

### 2021-12-25 book/paper.tex
Fixed some stuff

@@
 
+<{:const:tryAcquire}>
+
 \section*{Exercises}
@@
 <{:const:gpuRelease}>
+<{:const:acquire_bound_lock}>
+<{:const:release_bound_lock}>
+<{:const:nb_enter}>
+<{:const:nb_leave}>
+<{:const:sb_enter}>
+<{:const:sb_leave}>
 
@@
 You should define a constant <{M}> and two methods:
-\texttt{acquire\_bound\_lock()}
-and \texttt{release\_bound\_lock()}.
+<{acquire_bound_lock()}> and <{release_bound_lock()}>.
 (Bound locks are useful for situations where too many threads working
@@
 Write methods \texttt{OLBlock()} to create a new ``one lane bridge'' lock,
-\texttt{nb\_enter()} that a car must invoke before going northbound on
-the bridge and \texttt{nb\_leave()} that the car must invoke after leaving
-the bridge.  Similarly write \texttt{sb\_enter()} and \texttt{sb\_leave()}
+<{nb_enter()}> that a car must invoke before going northbound on
+the bridge and <{nb_leave()}> that the car must invoke after leaving
+the bridge.  Similarly write <{sb_enter()}> and <{sb_leave()}>
 for southbound cars.
@@
 As a rule of thumb, one should always be able to replace <{wait}>
-by <{release}> followed by <{acquirire}>.
+by <{release}> followed by <{acquire}>.
 This turns the solution into a busy-waiting one, inefficient but still
@@
 
+<{:const:TryLock}>
 <{:module:qsort}>
@@
 algorithm, and \autoref{fig:qsorttest} an accompanying test program.
-The array to be sorted is stored in shared variable <{textqs.arr}>.
+The array to be sorted is stored in shared variable <{testqs.arr}>.
 Another shared variable, <{testqs.todo}>, contains the ranges of the

### 2021-12-25 book/paper.tex
Fixed some stuff

@@
 of nodes in the list using <{malloc()}> and <{free()}>.
-<{malloc(v}>) returns a new memory location initialized to <{v}>,
+<{malloc(v)}> returns a new memory location initialized to <{v}>,
 which should be released with <{free()}> when it is no longer in use.
@@
 can acquire the bound lock at the same time.
-% \item Implement a \emph{bounded stack}.  Method <{push(item}>)
+% \item Implement a \emph{bounded stack}.  Method <{push(item)}>
 % should block if the stack is full.  Method \texttt{pop()} should block
@@
 of an available GPU, blocking if there is currently no GPU available.
-Method <{gpuRelease(gpu}>) releases the given GPU.  It never needs
+Method <{gpuRelease(gpu)}> releases the given GPU.  It never needs
 to block.
@@
 \item \texttt{Queue()} returns a new message queue;
-\item <{put(q}>, <{item}>) adds <{item}> to the queue pointed to by <{q}>;
-\item <{get(q}>) waits for and returns an item on the queue pointed to by <{q}>.
+\item <{put(q, item)}> adds <{item}> to the queue pointed to by <{q}>;
+\item <{get(q)}> waits for and returns an item on the queue pointed to by <{q}>.
 \end{itemize}
@@
 
-<{setintlevel(il}>) sets the interrupt level to <{il}> and returns
+<{setintlevel(il)}> sets the interrupt level to <{il}> and returns
 the prior interrupt level.  Returning the old level is handy when writing interrupt-safe
@@
 \begin{itemize}
-\item <{inc(p}>): atomically increments \texttt{!}<{p}> and returns
+\item <{inc(p)}>: atomically increments \texttt{!}<{p}> and returns
 the old value;
-\item <{exch(p}>): sets \texttt{!}<{p}> to <{None}> and returns
+\item <{exch(p)}>: sets \texttt{!}<{p}> to <{None}> and returns
 the old value.
@@
 
-Method <{produce(item}>) uses <{inc}>(?<{back}>)
-to allocate
+Method <{produce(item)}> uses <{inc(?back)}> to allocate
 the next available slot in the <{items}> array.
@@
 
-Method <{app_send(m}>) retransmits <{m}> until
+Method <{app_send(m)}> retransmits <{m}> until
 an acknowledgment is received.
-Method <{app_recv(reliable}>) returns the next successfully received
+Method <{app_recv(reliable)}> returns the next successfully received
 message.
@@
 \hline
-<{wait(c}>, <{lk}>) & wait on condition variable \texttt{!}<{c}>
+<{wait(c, lk)}> & wait on condition variable \texttt{!}<{c}>
 and lock <{lk}> \\ \hline
@@
 \hline
-<{get(q}>) & return next element of <{q}>, blocking if empty \\
+<{get(q)}> & return next element of <{q}>, blocking if empty \\
 \hline

### 2021-12-25 book/paper.tex
Pass over Ch 19

@@
 The condition in the <{while}> statement is the negation of the
-condition that the diner is waiting and consists of two disjuncts.
+condition that the diner is waiting for and consists of two disjuncts.
 Within the <{while}> statement,

### 2021-12-25 book/paper.tex
Pass over Ch 20

@@
 
-\texttt{Queue()} returns the initial value for a queue object
+<{Queue()}> returns the initial value for a queue object
 consisting of a <{None}> head and tail pointer and a lock.
@@
 The <{synch}> library supports a synchronized message queue,
-similar to the \texttt{Queue} object in Python.
+similar to the <{Queue}> object in Python.
 Its interface is as follows:
 \begin{itemize}
-\item \texttt{Queue()} returns a new message queue;
+\item <{Queue()}> returns an empty queue;
 \item <{put(q, item)}> adds <{item}> to the queue pointed to by <{q}>;
@@
 
-For those familiar with counting semaphores: note that a \texttt{Queue}
+For those familiar with counting semaphores: note that a <{Queue}>
 behaves much like a zero-initialized counting semaphore.
@@
 the counter and increments the counter.  No lock is required.
-The code is tested using the <{done}> array.
 
@@
 \hline
-\texttt{Queue}() & return a synchronized queue object \\
+<{Queue}>() & return a synchronized queue object \\
 \hline
@@
 Logical timestamps are numbered $0, 1, ...$
-To do so, we have added a counter (\texttt{time}) to the \texttt{Queue}.
+To do so, we have added a counter (\texttt{time}) to the <{Queue}>.
 Method \texttt{\_linpoint} saves the current counter in

### 2021-12-25 book/paper.tex
More formatting stuff

@@
 If <{p}> is a pointer to a shared variable, then the
-expression \texttt{!}<{p}> is the value of the shared variable.
+expression <{!p}> is the value of the shared variable.
 In particular, <{!?x = x}>.
@@
 
+<{:const:i_put}>
+
 \section*{Exercises}
@@
 say, a keyboard interrupt handler could place an event on a shared queue.
-Implement a new method \texttt{i\_put}(<{item}>) that does not
+Implement a new method <{i_put(item)}> that does not
 block.  Instead, it should return <{False}> if the buffer is full and <{True}>
@@
 \begin{itemize}
-\item <{inc(p)}>: atomically increments \texttt{!}<{p}> and returns
+\item <{inc(p)}>: atomically increments <{!p}> and returns
 the old value;
-\item <{exch(p)}>: sets \texttt{!}<{p}> to <{None}> and returns
-the old value.
+\item <{exch(p)}>: sets <{!p}> to <{None}> and returns the old value.
 \end{itemize}
@@
 There are three levels of precedence.
-Application has the highest precedence.  So, \texttt{!}<{a b}> is interpreted as
-\texttt{!}(<{a b}>) and <{a b}> \texttt{+} <{c d}> is interpreted as
-(<{a b}>) \texttt{+} (<{c d}>).
+Application has the highest precedence.  So, <{!a b}> is interpreted as
+<{!(a b)}> and <{a b + c d}> is interpreted as <{(a b) + (c d)}>.
 Unary operators have the next highest precedence,
@@
 \hline
-<{acquire(bs)}> & acquire binary semaphore \texttt{!}<{bs}> \\
-\hline
-<{release(bs)}> & release binary semaphore \texttt{!}<{bs}> \\
+<{acquire(bs)}> & acquire binary semaphore <{!bs}> \\
+\hline
+<{release(bs)}> & release binary semaphore <{!bs}> \\
 \hline
@@
 \hline
-<{wait(c, lk)}> & wait on condition variable \texttt{!}<{c}>
+<{wait(c, lk)}> & wait on condition variable <{!c}>
 and lock <{lk}> \\ \hline
-<{notify(c)}> & notify a thread waiting on condition variable \texttt{!}<{c}> \\
-\hline
-<{notifyAll(c)}> & notify all threads waiting on condition variable \texttt{!}<{c}> \\
+<{notify(c)}> & notify a thread waiting on condition variable <{!c}> \\
+\hline
+<{notifyAll(c)}> & notify all threads waiting on condition variable <{!c}> \\
 \hline
@@
 \hline
-<{P(sema)}> & procure \texttt{!}<{sema}>  \\
-\hline
-<{V(sema)}> & vacate \texttt{!}<{sema}>  \\
+<{P(sema)}> & procure <{!sema}>  \\
+\hline
+<{V(sema)}> & vacate <{!sema}>  \\
 \hline

### 2021-12-25 book/paper.tex
More preproc stuff

@@
 To see why, consider the following state <{s}>: let thread~0 be at label <{cs}>
-and thread~1 be at the start of the \textbf{await} statement.
+and thread~1 be at the start of the <{await}> statement.
 Also, in state <{s}>, $\mathit{turn} = 1$.  Now let
@@
 Let $\mathcal{C}(i) = \lnot\mathit{flags}[1 - i] \lor \mathit{turn} = i$, that is,
-the condition on the \textbf{await} statement for thread~<{i}>.
+the condition on the <{await}> statement for thread~<{i}>.
 In a sequential program, $\mathcal{C}(i)$ would clearly hold if thread <{i}> is in
@@
 The two things to note here is the use of the \textbf{choose}
-expression and the \textbf{assert} statement.  Using <{choose}>,
+expression and the <{assert}> statement.  Using <{choose}>,
 we are able to express searching through all possible strings of
@@
 to you.  But if you want to \emph{specify} a synchronization primitive,
-then use \textbf{atomically} by all means.
+then use <{atomically}> by all means.
 You can also use atomic statements in your test code.
-In fact, as mentioned before, \textbf{assert} statements are included to
+In fact, as mentioned before, <{assert}> statements are included to
 test if certain conditions hold in every execution and are executed
@@
 Harmony can check these invariants as well.
-$\mathcal{I}_2(i)$ is checked by the \textbf{assert} statement.
+$\mathcal{I}_2(i)$ is checked by the <{assert}> statement.
 But how would one go about checking an
@@
 customer code the same. You are only allowed to change the
-\texttt{atm\_} methods, and you cannot use the \textbf{atomically}
+\texttt{atm\_} methods, and you cannot use the <{atomically}>
 keyword.
@@
 The reason is that \texttt{queue.get()} changes the state by
-acquiring a lock, but the expressions in \textbf{assert}
+acquiring a lock, but the expressions in <{assert}>
 statements (or \textbf{invariant} statements)
@@
 Each entry in the queue is a pair consisting of a type of access
-(\texttt{.read} or \texttt{.write}) and a pointer to the semaphore of the
+(<{.read}> or <{.write}>) and a pointer to the semaphore of the
 thread that is waiting.  Method \texttt{release\_one} considers semaphores
@@
 \item Messages are added atomically to the network.  Is this necessary?
-What happens if you remove the \textbf{atomically} keyword?  Explain what happens.
+What happens if you remove the <{atomically}> keyword?  Explain what happens.
 \end{problems}
@@
 in this case.
-Each transaction has a \emph{coordinator} that sends a \texttt{PREPARE}
+Each transaction has a \emph{coordinator} that sends a <{PREPARE}>
 message to each of the servers involved in the transaction, asking them
@@
 \noindent
-Messages sent to a bank have the following dictionary format:
-$\{ \mathtt{.dst}, \mathtt{.src}, \mathtt{.request}, \mathtt{.amount} \}$:
+Messages sent to a bank are dictionaries with the following fields:
 \begin{itemize}
-\item \texttt{.dst}: identifier of the bank;
-\item \texttt{.src}: identifier of the coordinator that sent the message;
-\item \texttt{.request}: request type, which is either
-\texttt{.withdraw}, \texttt{.deposit}, \texttt{.commit}, or \texttt{.abort};
-\item \texttt{.amount}: amount to withdraw or deposit.
+\item <{.dst}>: identifier of the bank;
+\item <{.src}>: identifier of the coordinator that sent the message;
+\item <{.request}>: request type, which is either
+<{.withdraw}>, <{.deposit}>, <{.commit}>, or <{.abort}>;
+\item <{.amount}>: amount to withdraw or deposit.
 \end{itemize}
@@
 an ongoing transaction and what its original balance was.
-The bank then responds with a \texttt{.yes} message to the coordinator,
+The bank then responds with a <{.yes}> message to the coordinator,
 including the new balance.
-Otherwise, the bank responds with a \texttt{.no} message.
+Otherwise, the bank responds with a <{.no}> message.
 Deposits are similar, except that it is not necessary to check for
 sufficient funds.
-In case of a \texttt{.commit} message, the bank changes its status
+In case of a <{.commit}> message, the bank changes its status
 to (), indicating that there is no ongoing transaction.  In case of
-a \texttt{.abort} message, the bank restores <{balance}> first.
+a <{.abort}> message, the bank restores <{balance}> first.
 
@@
 as tests.
-The \texttt{receive()} method is used by coordinators in an
+The <{receive()}> method is used by coordinators in an
 \textbf{atomically when exists}
@@
 and <{amt}> is the amount to transfer.
-The coordinator sends a \texttt{PREPARE} message containing a
-\texttt{.withdraw} request to <{b1}> and a \texttt{PREPARE}
-message containing a \texttt{.deposit} request to <{b2}>.
+The coordinator sends a <{PREPARE}> message containing a
+<{.withdraw}> request to <{b1}> and a <{PREPARE}>
+message containing a <{.deposit}> request to <{b2}>.
 It then waits for responses from each.  If both responses are
-\texttt{.yes}, then it commits the transaction, otherwise it aborts
+<{.yes}>, then it commits the transaction, otherwise it aborts
 the transaction.
@@
 Note the use of the \textbf{possibly} statement in this code.
-It is similar to \textbf{assert}, but where \texttt{assert} checks that
+It is similar to <{assert}>, but where \texttt{assert} checks that
 the predicate holds in \emph{all} executions, \texttt{possibly} checks
@@
 the replica.  The payload can take one of two forms.  The first is
-\texttt{.crash}, indicating that the source of the message has
+<{.crash}>, indicating that the source of the message has
 crashed.  The code models a replica crashing by that replica broadcasting
-a \texttt{.crash} message on the network.
+a <{.crash}> message on the network.
 The other form of payload is a tuple
@@
 awaits a message from its direct predecessor.
-If the payload is \texttt{.crash}, then the replica removes this
+If the payload is <{.crash}>, then the replica removes this
 predecessor from its set of predecessors.  For example, if there
@@
 Read and write operations are both invoked with a unique identifier
-<{uid}>.  Both start by broadcasting a \texttt{.read}
+<{uid}>.  Both start by broadcasting a <{.read}>
 request to all servers and then waiting for a response from
-$\mathtt{N} - \mathtt{F}$ servers.
-The \texttt{receive()} function uses the \texttt{bag.combinations} method
+<{N - F}> servers.
+The <{receive()}> function uses the <{bag.combinations}> method
 to find all combinations of subsets of responses of size
@@
 \autoref{fig:abdtest} tests the third property of the properties listed
-above.  The \texttt{writer()} method writes a new value \texttt{.token}.
+above.  The <{writer()}> method writes a new value <{.token}>.
 The \texttt{reader()} method reads twice and makes sure that if the
-first read operation returns \texttt{.token}, then so does the second.
+first read operation returns <{.token}>, then so does the second.
 This illustrates the importance of the second phase of the \texttt{read}
@@
 \hline
-\texttt{<{e}> == <{e}>} & equivalence \\
-\hline
-\texttt{<{e}> != <{e}>} & inequivalence \\
-\hline
-\texttt{<{e}> < <{e}>, <{e}> <= <{e}>, <{e}> > <{e}>, <{e}> >= <{e}>} & comparison\\
+<{e == e}> & equivalence \\
+\hline
+<{e != e}> & inequivalence \\
+\hline
+<{e < e, e <= e, e > e, e >= e}> & comparison\\
 \hline
@@
 \hline
-\texttt{<{e}> and <{e}> and ...} & conjuction \\
-\hline
-\texttt{<{e}> or <{e}> or ...} & disjunction \\
-\hline
-\texttt{<{e}> => <{e}>, <{e}> not => <{e}>} & implication \\
-\hline
-\texttt{not <{e}>} & negation \\
-\hline
-\texttt{<{v}> if <{e}> else $v'$} & <{v}> or $v'$ depending on <{e}> \\
-\hline
-\texttt{any <{s}>, all <{s}>} & disjunction / conjunction for set or list <{s}> \\
+<{e and e and ...}> & conjuction \\
+\hline
+<{e or e or ...}> & disjunction \\
+\hline
+<{e => e, e not => e}> & implication \\
+\hline
+<{not e}> & negation \\
+\hline
+<{v if e else v'}> & <{v}> or <{v'}> depending on <{e}> \\
+\hline
+<{any s, all s}> & disjunction / conjunction for set or list <{s}> \\
 \hline
@@
 \hline
-\texttt{-<{e}>} & negation \\
-\hline
-\texttt{abs(<{e}>)} & absolute value \\
-\hline
-\texttt{<{e}> + <{e}> + ...} & sum \\
-\hline
-\texttt{<{e}> - <{e}>} & difference \\
-\hline
-\texttt{<{e}> * <{e}> * <{e}> ...} & product \\
-\hline
-\texttt{<{e}> / <{e}>, <{e}> // <{e}>} & integer division \\
-\hline
-\texttt{<{e}> \% <{e}>, <{e}> mod <{e}>} & integer division remainder \\
-\hline
-\texttt{<{e}> ** <{e}>} & power \\
-\hline
-\texttt{\string~<{e}>} & binary inversion \\
-\hline
-\texttt{<{e}> \& <{e}> \& ...} & binary and \\
-\hline
-\texttt{<{e}> | <{e}> | ...} & binary or \\
-\hline
-\texttt{<{e}> \string^ <{e}> \string^ ...} & binary exclusive or \\
-\hline
-\texttt{$e << e$} & binary shift left \\
-\hline
-\texttt{$e >> e$} & binary shift right \\
-\hline
-\texttt{$\{ e..e' \}$} & set of integers from <{e}> to $e'$ inclusive \\
+<{-e}> & negation \\
+\hline
+<{abs(e)}> & absolute value \\
+\hline
+<{e + e + ...}> & sum \\
+\hline
+<{e - e}> & difference \\
+\hline
+<{e * e * e ...}> & product \\
+\hline
+<{e / e, e // e}> & integer division \\
+\hline
+<{e % e, e mod e}> & integer division remainder \\
+\hline
+<{e ** e}> & power \\
+\hline
+<{~e}> & binary inversion \\
+\hline
+<{e & e & ...}> & binary and \\
+\hline
+<{e | e | ...}> & binary or \\
+\hline
+<{e ^ e ^ ...}> & binary exclusive or \\
+\hline
+<{e << e}> & binary shift left \\
+\hline
+<{e >> e}> & binary shift right \\
+\hline
+<{{ e..e' }}> & set of integers from <{e}> to $e'$ inclusive \\
 \hline
@@
 \hline
-\texttt{len <{s}>} & cardinality \\
-\hline
-\texttt{<{s}> - <{s}>} & set difference \\
-\hline
-\texttt{<{s}> \& <{s}> \& ...} & intersection \\
-\hline
-\texttt{<{s}> | <{s}> | ...} & union \\
-\hline
-\texttt{<{s}> \string^ <{s}> \string^ ...} & inclusion/exclusion (elements in odd number of sets) \\
-\hline
-\texttt{choose <{s}>} & select an element (Harmony will try all) \\
-\hline
-\texttt{min <{s}>} & minimum element \\
-\hline
-\texttt{max <{s}>} & maximum element \\
-\hline
-\texttt{any <{s}>} & <{True}> if any value is <{True}> \\
-\hline
-\texttt{all <{s}>} & <{True}> if all values are <{True}> \\
+<{len s}> & cardinality \\
+\hline
+<{s - s}> & set difference \\
+\hline
+<{s & s & ...}> & intersection \\
+\hline
+<{s | s | ...}> & union \\
+\hline
+<{s ^ s ^ ...}> & inclusion/exclusion (elements in odd number of sets) \\
+\hline
+<{choose s}> & select an element (Harmony will try all) \\
+\hline
+<{min s}> & minimum element \\
+\hline
+<{max s}> & maximum element \\
+\hline
+<{any s}> & <{True}> if any value is <{True}> \\
+\hline
+<{all s}> & <{True}> if all values are <{True}> \\
 \hline
@@
 \hline
-\texttt{$d~k$} & indexing \\
-\hline
-\texttt{len <{d}>} & the number of keys in <{d}> \\
-\hline
-\texttt{keys <{d}>} & the set of keys in <{d}> \\
-\hline
-\texttt{<{v}> [not] in keys <{d}>} & check if <{v}> is a \emph{key} in <{d}> \\
-\hline
-\texttt{<{v}> [not] in <{d}>} & check if <{v}> is a \emph{value} in <{d}> \\
-\hline
-\texttt{min <{d}>} & the minimum value in <{d}> \\
-\hline
-\texttt{max <{d}>} & the maximum value in <{d}> \\
-\hline
-\texttt{any <{d}>} & <{True}> if any value is <{True}> \\
-\hline
-\texttt{all <{d}>} & <{True}> if all values are <{True}> \\
-\hline
-\texttt{<{d}> \& <{d}> \& ...} & dictionary intersection \\
-\hline
-\texttt{<{d}> | <{d}> | ...} & dictionary union \\
+<{d k}> & indexing \\
+\hline
+<{len d}> & the number of keys in <{d}> \\
+\hline
+<{keys d}> & the set of keys in <{d}> \\
+\hline
+<{<{v [not] in keys d}> & check if <{v}> is [not] a key in <{d}> \\
+\hline
+<{v [not] in d}> & check if <{v}> is [not] a value in <{d}> \\
+\hline
+<{min d}> & the minimum value in <{d}> \\
+\hline
+<{max d}> & the maximum value in <{d}> \\
+\hline
+<{any d}> & <{True}> if any value is <{True}> \\
+\hline
+<{all d}> & <{True}> if all values are <{True}> \\
+\hline
+<{d & d & ...}> & dictionary intersection \\
+\hline
+<{d | d | ...}> & dictionary union \\
 \hline
@@
 you can write $d[k]$ (or $d(k)$) instead of $d~k$.
-However, if <{k}> is a string like \texttt{.id}, then you might
+However, if <{k}> is a string like <{.id}>, then you might
 prefer the notation $k\mathtt{.id}$.
@@
 \hline
-\texttt{$t~k$} & indexing \\
-\hline
-\texttt{<{t}> + <{t}> + ...} & concatenation \\
-\hline
-\texttt{<{t}> * $n$} & $n$ copies of <{t}> concatenated \\
-\hline
-\texttt{<{v}> [not] in <{t}>} & check if <{v}> is a \emph{value} in <{t}> \\
-\hline
-\texttt{len <{t}>} & the length of <{t}> \\
-\hline
-\texttt{min <{t}>} & the minimum value in <{t}> \\
-\hline
-\texttt{max <{t}>} & the maximum value in <{t}> \\
-\hline
-\texttt{any <{t}>} & <{True}> if any value is <{True}> \\
-\hline
-\texttt{all <{t}>} & <{True}> if all values are <{True}> \\
-\hline
-\texttt{<{t}> \& <{t}> \& ...} & pairwise list minimum \\
-\hline
-\texttt{<{t}> | <{t}> | ...} & pairwise list maximum \\
+<{t k}> & indexing \\
+\hline
+<{t + t + ...}> & concatenation \\
+\hline
+<{t * n}> & <{n}> copies of <{t}> concatenated \\
+\hline
+<{v [not] in t}> & check if <{v}> is [not] a value in <{t}> \\
+\hline
+<{len t}> & the length of <{t}> \\
+\hline
+<{min t}> & the minimum value in <{t}> \\
+\hline
+<{max t}> & the maximum value in <{t}> \\
+\hline
+<{any t}> & <{True}> if any value is <{True}> \\
+\hline
+<{all t}> & <{True}> if all values are <{True}> \\
+\hline
+<{t & t & ...}> & pairwise list minimum \\
+\hline
+<{t | t | ...}> & pairwise list maximum \\
 \hline
@@
 \hline
-\texttt{<{v}> [not] in keys <{b}>} & check if <{v}> is in <{b}> \\
-\hline
-\texttt{<{t}> \& <{t}> \& ...} & bag intersection\\
-\hline
-\texttt{<{t}> | <{t}> | ...} & bag union \\
+<{v [not] in keys b}> & check if <{v}> is [not] in <{b}> \\
+\hline
+<{t & t & ...}> & bag intersection\\
+\hline
+<{t | t | ...}> & bag union \\
 \hline
@@
 a list of Harmony values.
-For example, the address of variable $d.\mathtt{v}[3]$ consists of the
-list \texttt{.d}, \texttt{.v}, and 3.  The only way to construct an
-address in Harmony is using the $?$ operator.
-$?d.\mathtt{v}[3]$ returns the address of variable $d.\mathtt{v}[3]$.
+For example, the address of variable <{d.v[3]}> consists of the
+list <{.d}>, <{.v}>, and 3.  The only way to construct an
+address in Harmony is using the <{?}> operator.
+<{?d.v[3]}> returns the address of variable <{d.v[3]}>.
 An address can be dereferenced to return the value of the variable.
-If <{a}> is an address, then $!a$ returns the value.
-Like C, Harmony supports the shorthand \texttt{<{a}>->v} for 
-the expression \texttt{($!a$).v}.
+If <{a}> is an address, then <{!a}> returns the value.
+Like C, Harmony supports the shorthand <{a->v}> for the expression <{(!a).v}>.
 
@@
 
-\subsection*{\textbf{assert}}
-
-The statement \textbf{assert <{b}>} evaluates <{b}> and reports an error
+\subsection*{<{assert}>}
+
+The statement <{assert b}> evaluates <{b}> and reports an error
 if <{b}> is false.  It should be considered a no-op---it is part of the
@@
 In particular, it specifies an invariant: whenever the program counter is
-at the location where the \textbf{assert} statement is, then <{b}> is
+at the location where the <{assert}> statement is, then <{b}> is
 always true.
@@
 As in Python, you can specify an additional expression:
-\textbf{assert <{b}>, <{e}>}.  The value of <{e}> will be reported as part
+<{assert b, e}>.  The value of <{e}> will be reported as part
 of the error should <{b}> evaluate to false.
 
-\subsection*{\textbf{atomically}}
+\subsection*{<{atomically}>}
 
@@
 
-\textbf{atomically} statements are useful for specification and
+<{atomically}> statements are useful for specification and
 implementing synchronization primitives such as test-and-set.
@@
 
-\subsection*{\textbf{await}}
-
-The statement \textbf{await <{b}>} is equivalent to \textbf{when <{b}>: pass}.
+\subsection*{<{await}>}
+
+The statement <{await b}> is equivalent to \textbf{when <{b}>: pass}.
 It is intended to improve readability of your code.
 
-\subsection*{\textbf{const}}
+\subsection*{<{const}>}
 
@@
 
-\subsection*{\textbf{def}}
+\subsection*{<{def}>}
 
@@
 
-\subsection*{\textbf{del}}
-
-The statement \textbf{del} <{x}> removes variable <{x}> from the
+\subsection*{<{del}>}
+
+The statement <{del x}> removes variable <{x}> from the
 state.  <{x}> can be either a local or a shared variable.  For
-example, the statement \textbf{del} $x.\texttt{age}$ removes
-the \texttt{.age} field from dictionary <{x}>.
+example, the statement <{del x.age}> removes
+the <{.age}> field from dictionary <{x}>.
 Harmony automatically removes top-level local variables that
@@
 from Python:
-$x = [\mathtt{.a}, \mathtt{.b}]; \mathbf{del}~x[0]$ results in <{x}> having value
-$\{1: \mathtt{.b}\}$ rather than $[\mathtt{.b}]$ (which is $\{0: \mathtt{b}\}$).
+<{x = [.a, .b]; del x[0]}> results in <{x}> having value
+<{{1: .b}}> rather than <{[.b]}> (which is <{{0: .b}}>).
 

### 2021-12-25 book/paper.tex
Latest

@@
 Harmony in fact supports this.
-It has an operator \textbf{countLabel}~$L$,
+It has an operator <{countLabel l}>,
 \index{countLabel operator}%
-where $L$ is the name of the label (in this case, <{cs}>).
+where <{l}> is the name of the label (in this case, <{cs}>).
 The operator returns the number of threads executing at that label.
-Method \textbf{countLabel} only exists for specification purposes---do not
+Method <{countLabel}> only exists for specification purposes---do not
 use it in normal code.
@@
 
-The two things to note here is the use of the \textbf{choose}
+The two things to note here is the use of the <{choose}>
 expression and the <{assert}> statement.  Using <{choose}>,
@@
 Peterson's algorithm relies on a \emph{sequential consistent memory model}
-and hence the \textbf{sequential} statement: without it Harmony will
+and hence the <{sequential}> statement: without it Harmony will
 complain about data races.
-More precisely, the \textbf{sequential} statement says that the
+More precisely, the <{sequential}> statement says that the
 program relies on memory load and store instructions operating on
@@
 consistent.
-The unfortunately named \texttt{volatile} keyword in Java has
-a similar function as Harmony's \texttt{sequential} keyword.
+The unfortunately named \textbf{volatile} keyword in Java has
+a similar function as Harmony's <{sequential}> keyword.
 Like many constructions in Java, its \texttt{volatile} keyword
@@
 Harmony does not report data races in two cases.
-First, using the \textbf{sequential} statement, you can specify that
+First, using the <{sequential}> statement, you can specify that
 concurrent access to the specified variables is intended.
@@
 statements in your ordinary synchronization code, you should not use
-\textbf{sequential} statements either.  Neither is supported by regular
+<{sequential}> statements either.  Neither is supported by regular
 programming languages.  They exist in Harmony for the purpose of
@@
 the thread to be no longer running.
-Another thread can revive the thread using the \textbf{go}
+Another thread can revive the thread using the <{go}>
 \index{go statement}%
@@
 It causes a thread with context <{c}> to be added to the state that has
-just executed the \textbf{stop}
+just executed the <{stop}>
 \index{stop expression}%
-expression.  The \textbf{stop} expression returns the value <{r}>.
+expression.  The <{stop}> expression returns the value <{r}>.
 
@@
 In the latter case, the context of the thread is added to the end of the list of contexts.
-Note that \textbf{stop} is called within an atomic statement block---this is
+Note that <{stop}> is called within an atomic statement block---this is
 the only exception to such an atomic statement block running to completion.
@@
 acquiring a lock, but the expressions in <{assert}>
-statements (or \textbf{invariant} statements)
+statements (or <{invariant}> statements)
 are not allowed to change the state.
@@
 $\{ 0 .. \mathtt{N} - 1 \}$.
-% Note that it is perfectly fine to use \textbf{sequential}
-% and \textbf{atomic} statements in test code---one should
-% only avoid those in production code.
 
@@
 lead to a state space explosion.
-Instead, we can leverage the \textbf{choose}
+Instead, we can leverage the <{choose}>
 expression.
@@
 On the left is a Hoare monitor.
-When some thread <{p}> is in the bathroom, it can invoke \textbf{signal}
+When some thread <{p}> is in the bathroom, it can invoke <{signal}>
 on some non-empty bedroom associated with a condition that holds.
 This select a thread <{q}> in that bedroom.
-As a result of invoking \textbf{signal}, <{p}> goes back into the hall
+As a result of invoking <{signal}>, <{p}> goes back into the hall
 and <{q}> enters the bathroom.\footnote{Per Brinch Hansen' monitors have
-a \textbf{continue} operation instead of \textbf{signal}.  The difference
+a \textbf{continue} operation instead of <{signal}>.  The difference
 is that threads that invoke \textbf{continue} leave the monitor rather
@@
 The server never terminates, so it is spawned with the keyword
-\textbf{eternal} to suppress non-terminating state warnings.
+<{eternal}> to suppress non-terminating state warnings.
 Each client sends a request to the server, consisting in this case
@@
 <{found}> is a boolean indicating the second phase of the protocol.
-The \texttt{receive}(<{self}>) method looks for all messages
+The <{receive}>(<{self}>) method looks for all messages
 destined for the processor with identifier <{self}>.
@@
 A processor waits for a message using the Harmony
-\textbf{atomically when exists} statement.
+<{atomically when exists}> statement.
 This statement takes the form <{atomically when exists v in s}>: \emph{statement block}
@@
 The <{receive()}> method is used by coordinators in an
-\textbf{atomically when exists}
+<{atomically when exists}>
 statement to wait for a response from each bank involved in a transaction.
@@
 the transaction.
-
-\begin{comment}
-Note the use of the \textbf{possibly} statement in this code.
-It is similar to <{assert}>, but where \texttt{assert} checks that
-the predicate holds in \emph{all} executions, \texttt{possibly} checks
-that the predicate holds in at least one execution.
-The purpose is to check that the code doesn't simply always abort---there
-exist executions where the code commits.
-In practice, hopefully \emph{most} executions commit, but Harmony is not
-capable of probabilistic reasoning.
-\end{comment}
 
@@
 
-In Line~22, a processor waits for $<{N}> - <{F}>$ messages using
-the Harmony \textbf{atomically when exists} statement.
+In Line~22, a processor waits for <{N - F}> messages using
+the Harmony <{atomically when exists}> statement.
 Since Harmony has to check all possible executions of the protocol, the
-\texttt{receive}(<{round}>, <{k}>) method returns all \emph{subbags} of messages
-for the given round that have size $k = <{N}> - <{F}>$.  The method uses a
+<{receive(round, k)}> method returns all \emph{subbags} of messages
+for the given round that have size <{k = N - F}>.  The method uses a
 dictionary comprehension to filter out all messages for the given <{round}>
-and then uses the \texttt{bag.combinations} method to find all combinations of size <{k}>.
-The \textbf{atomically when exists} statement waits until there is at least one such combination and
+and then uses the <{bag.combinations}> method to find all combinations of size <{k}>.
+The <{atomically when exists}> statement waits until there is at least one such combination and
 then chooses an element, which is bound to the <{quorum}> variable.
@@
 
-The body of the \textbf{atomically when exists} statement contains the core of the algorithm.
-Note that $<{N}> - <{F}> = 2<{F}> + 1$, so that the number of
+The body of the <{atomically when exists}> statement contains the core of the algorithm.
+Note that <{N - F = 2F + 1}>, so that the number of
 messages is guaranteed to be odd.  Also, because there are only 0 and 1 values, there
@@
 interleave in the atomic statement.  The only exception
-to this is if the atomic block executes a \textbf{stop} expression.
+to this is if the atomic block executes a <{stop}> expression.
 In this case, another thread can run.  When the original thread
-is resumed (using a \textbf{go} statement), it is once again
+is resumed (using a <{go}> statement), it is once again
 atomically executing.
@@
 
-The statement <{await b}> is equivalent to \textbf{when <{b}>: pass}.
+The statement <{await b}> is equivalent to <{when b: pass}>.
 It is intended to improve readability of your code.
@@
 
-\subsection*{\textbf{for ... in ... [where ...]}}
+\subsection*{<{for ... in ... [where ...]}>}
 
@@
 
-\subsection*{\textbf{from ... import}}
-
-The statement \textbf{from} \texttt{x} \textbf{import} \texttt{a, b, ...}
-imports module \texttt{x} and makes its constants $a, b, ...$ also constants
+\subsection*{<{from ... import}>}
+
+The statement <{from x import a, b, ...}>
+imports module <{x}> and makes its constants <{a, b, ...}> also constants
 in the current module.  If a module is imported more than once, its code
@@
 (including program counter constants) using the statement
-\textbf{from} <{m}> \textbf{import} \texttt{*}.
+<{from m import *}>.
 This, however, excludes constants whose names start with the character
@@
 
-\subsection*{\textbf{go}}
-
-The statement \textbf{go} <{c e}> starts a thread with context <{c}> that
-has executed a \textbf{stop} expression.  The \textbf{stop} expression
+\subsection*{<{go}>}
+
+The statement <{go c e}> starts a thread with context <{c}> that
+has executed a <{stop}> expression.  The <{stop}> expression
 returns value <{e}>.
@@
 
-\subsection*{\textbf{if ... [elif ...]* [else]}}
+\subsection*{<{if ... [elif ...]* [else]}>}
 
 Harmony supports <{if}> statements.
-In its most basic form, <{if}> $c: S_1; S_2; ...$ evaluates <{c}>
-and executes statements $S_1, S_2, ...$ if and only if boolean expression
+In its most basic form, <{if c: S1; S2; ...}> evaluates <{c}>
+and executes statements <{S1, S2, ...}> if and only if boolean expression
 <{c}> evaluated to true.
@@
 is the case, Harmony reports an error.
-The statement <{if}> $c: S_1, S_2, ...$ <{else}>:
-$T_1; T_2; ...$ is similar, but executes statements $T_1, T_2, ...$
-if and only if <{c}> evaluated to false.
-You can think of \textbf{elif} $c:$ as shorthand for
-<{else}>: <{if}> $c:$.
-
-\subsection*{\textbf{import}}
-
-The statement \textbf{import} $\mathtt{m}_1, \mathtt{m}_2, ...$ imports modules
-$\mathtt{m}_1, \mathtt{m}_2, ...$ in that order.
+The statement <{if c: S1, S2, ... else: T1; T2; ...}>
+is similar, but executes statements <{T1, T2, ...}>
+if and only if <{c}> evaluated to <{False}>.
+You can think of <{elif}> $c:$ as shorthand for <{else: if c:}>.
+
+\subsection*{<{import}>}
+
+The statement <{import m1, m2, ...}> imports modules
+<{m1, m2, ...}> in that order.
 If a module is imported more than once,
@@
 For example, method <{f()}> in imported module <{m}> is invoked
-by calling \texttt{m.f()}.  If you would prefer to invoke it simply as
+by calling <{m.f()}>.  If you would prefer to invoke it simply as
 <{f()}>, then you have to import using the statement
-\textbf{from} <{m}> \textbf{import} \texttt{f}.
-
-\subsection*{\textbf{invariant}}
-
-The statement \textbf{invariant} <{c}> declares that boolean expression <{c}>
+<{from m import f}>.
+
+\subsection*{<{invariant}>}
+
+The statement <{invariant c}> declares that boolean expression <{c}>
 is an invariant.  <{c}> is only allowed to read shared variables and
@@
 <{let}> statement.
-The statement <{let}> $a = b: S_1; S_2, ...$ evaluates <{b}>, assigns
+The statement <{let a = b: S1; S2, ...}> evaluates <{b}>, assigns
 the result to read-only variable <{a}>, and evaluates statements
-$S_1, S_2, ...$.  <{let}> supports pattern matching, so you can
-write <{let}> $x, (y, z) = b: S_1; S_2, ...$.  This will only work
+<{S1, S2, ...}>.  <{let}> supports pattern matching, so you can
+write <{let x, (y, z) = b: S1; S2, ...}>.  This will only work
 if <{b}> is a tuple with two elements, the second of which also being a
@@
 <{let}> statements may be nested, such as
-<{let}> $a_1 = b_1$ <{let}> $a_2 = b_2: S_1; S_2; ...$.
+<{let a1 = b1 let a2 = b2: S1; S2; ...}>.
 Doing so can improve readability by reducing indentation
@@
 
-\subsection*{\textbf{pass}}
-
-The \textbf{pass} statement does nothing.
-
-\subsection*{\textbf{print}}
-
-The statement \textbf{print <{e}>} evaluates <{e}>
+\subsection*{<{pass}>}
+
+The <{pass}> statement does nothing.
+
+\subsection*{<{print}>}
+
+The statement <{print e}> evaluates <{e}>
 and adds the result to the print log.  The print log is used to
@@
 
-\subsection*{\textbf{sequential}}
+\subsection*{<{sequential}>}
 
@@
 declaring the shared variable as sequential.  In particular, the statement
-\texttt{sequential} $x, y, ...$ specifies that the algorithm assumes
+<{sequential x, y, ...}> specifies that the algorithm assumes
 that the given variables have sequential consistency.
@@
 
-\subsection*{\textbf{spawn}}
-
-The statement <{spawn}> <{m a}> starts a new thread that executes
+\subsection*{<{spawn}>}
+
+The statement <{spawn m a}> starts a new thread that executes
 method <{m}> with argument <{a}>. <{m}> must be a program counter constant,
@@
 is the empty dictionary by default.  It can be specified by adding
-a parameter: <{spawn}> <{m}> $a, e$ specifies that <{e}> should be
+a parameter: <{spawn m a, e}> specifies that <{e}> should be
 the initial value of the thread-local state.
@@
 If a thread may never terminate, you should spawn it with
-\texttt{spawn eternal} <{m a}> to suppress those checks.
+<{spawn eternal m a}> to suppress those checks.
 
@@
 
-\subsection*{\textbf{when}}
+\subsection*{<{when}>}
 
@@
 
-Associative operators ($+$, $*$, $|$, $\string&$, $\string^$, <{and}>, \textbf{or})
+Associative operators ($+$, $*$, $|$, $\string&$, $\string^$, <{and}>, <{or}>)
 are interpreted as general $n$-ary operators, and you are allowed to write
@@
 to right but stops once one of the subexpressions evaluates to <{False}>.
-Similarly for \textbf{or}, where evaluation stops once one of the subexpressions
+Similarly for <{or}>, where evaluation stops once one of the subexpressions
 evaluates to <{True}>.
@@
 As an aside:
-the expression <{a}> \textbf{not in} <{b}> is equivalent to <{not}> (<{a}> \textbf{in} <{b}>).
+the expression <{a not in b}> is equivalent to <{not (a in b)}>.
 Harmony generalizes this construct for any pair of a unary (except '$-$') and a binary operator.
@@
 \begin{itemize}
-\item \textbf{choose} \emph{expressions}: Harmony's ability to let the program
+\item <{choose}> \emph{expressions}: Harmony's ability to let the program
 choose a value from a set;
@@
 The context of a thread cannot be accessed by other
-threads, nor by \textbf{invariant} statements.
+threads, nor by <{invariant}> statements.
 So, the model checker only interleaves threads at \textbf{Load},
@@
 To do so, we have added a counter (\texttt{time}) to the <{Queue}>.
-Method \texttt{\_linpoint} saves the current counter in
-\textbf{this}.\texttt{qtime} and increments the counter.
-The \textbf{this} dictionary maintains \emph{thread-local state} associated
+Method <{_linpoint}> saves the current counter in
+<{this.qtime}> and increments the counter.
+The <{this}> dictionary maintains \emph{thread-local state} associated
 with the thread (\autoref{ch:harmonymachine})---it contains variables that

### 2021-12-25 book/paper.tex
Working on new barrier synchronization chapter

@@
 \begin{center}
-\includegraphics[width=0.8\textwidth]{figures/hello8.png}
+\includegraphics[width=0.9\textwidth]{figures/hello8.png}
 \end{center}
@@
 <{:const:Barrier}>
+<{:const:bwait}>
+
+\begin{figure}
+\begin{code}
+<{:input:../code/barriertest.hny}>
+\end{code}
+\caption{\harmonylink{code/barriertest.hny} Test program for \autoref{fig:genbarrier}}
+\label{fig:barriertest}
+\end{figure}
+
+\begin{figure}
+\begin{code}
+<{:input:../code/barrier.hny}>
+\end{code}
+\caption{\harmonylink{code/barrier.hny} Barrier implementation}
+\label{fig:genbarrier}
+\end{figure}
+
+\begin{figure}
+\begin{code}
+<{:input:../code/barriertest2.hny}>
+\end{code}
+\caption{\harmonylink{code/barriertest2.hny} Demonstrating the double-barrier pattern}
+\label{fig:barriertest2}
+\end{figure}
 
 \glsadd{barrier synchronization}%
-
-\begin{figure}
-\begin{code}
-<{:input:../code/qbarrier.hny}>
-\end{code}
-\caption{\harmonylink{code/qbarrier.hny} Barrier synchronization with queues}
-\label{fig:barrier}
-\end{figure}
 
@@
 parallel computing.
-It is used, among others, for scalable simulation.
+The Harmony model checker uses it.
 A barrier is almost the opposite of a critical section:
@@
 same time, instead of having them execute it one at a time.
-More precisely, with barrier synchronization the threads execute in rounds.
-Between each round there is a so-called \emph{barrier} where threads wait
-until all threads have completed the previous round, before they start the
-next one.
+More precisely, with barrier synchronization, the threads execute in rounds.
+Between each round, there is a so-called \emph{barrier} where threads wait
+until all threads have completed the previous round and reached the
+barrier---before they start the next round.
 For example, in an iterative matrix algorithm, the matrix may be
@@
 
-\begin{comment}
-\begin{figure}
-\begin{center}
-\includegraphics[width=\textwidth]{figures/barrier-crop.pdf}
-\end{center}
-\caption{High-level state diagram specification of barrier synchronization
-with three threads and two rounds.  The three numbers specify how many
-rounds each thread has completed.}
-\label{fig:barrierdiagram}
-\end{figure}
-
-\autoref{fig:barrierdiagram} shows a high-level depiction of barrier synchronization
-with three and threads.  Initially all threads are in round~0.  Then each thread
-can start and finish the round.  However, none of the threads can
-progress to the next round until all threads have finished the
-current round.
-\end{comment}
-
-Blocking queues work well for implementing barrier synchronization.
-\autoref{fig:barrier} shows an example.  There is a queue
-for each of the <{N}> threads.
-Before thread <{i}> enters a round, it first sends a message to every
-other thread and then waits until it receives a message from
-every other thread.
-In this simple case, each message contains <{None}>, but in practice useful
-information may be exchanged between the threads.
-
-The <{round}> array is kept to check the correctness of this
-approach.  Each thread increments its entry every time it enters
-a round.  If the algorithm is correct, it can never be that two threads
-are more than one round apart.
-
-\begin{figure}
-\begin{code}
-<{:input:../code/barrier.hny}>
-\end{code}
-\caption{\harmonylink{code/barrier.hny} Generalized barrier synchronization}
-\label{fig:genbarrier}
-\end{figure}
-
-\begin{figure}
-\begin{code}
-<{:input:../code/barriertest.hny}>
-\end{code}
-\caption{\harmonylink{code/barriertest.hny} Test program for \autoref{fig:genbarrier}}
-\label{fig:barriertest}
-\end{figure}
-
-More generally, barrier synchronization can be abstracted as follows.
-We want to create a \texttt{Barrier}$(n)$ object, with operations
-\texttt{enter()} and \texttt{exit()}.  It is helpful to use
-a roller coaster car with $n$ seats as a metaphor:
+A barrier is used as follows:
 \begin{itemize}
-\item the car cannot contain more than $n$ people;
-\item the car won't take off until $n$ people are in the car;
-\item no new people can enter the car until all $n$ people have left it.
+\item <{b = Barrier(n)}>: initialize a barrier <{b}> for a collection of
+<{n}> threads;
+\item <{bwait(?b)}>: wait until all threads have reached the barrier
 \end{itemize}
-Notice there are two different waiting conditions:
-\begin{enumerate}
-\item waiting for the car to empty out;
-\item waiting for the car to fill up.
-\end{enumerate}
-
-But this poses a complication.  Suppose, for example, that there are two
-seats in the car, and there is one person in the car.  Does that mean that
-the car is not yet full, or not yet empty?  We have to distinguish
-those situations.  To this end, it is useful to think of the car as going through
-$2 \cdot n$ stages:
-\begin{itemize}
-\item[] Stage 0: the car is empty;
-\item[] Stage $1 ... n-1$: the car is filling up;
-\item[] Stage $n$: the car is full;
-\item[] Stage $n+1 ... 2n - 1$: the car is emptying out.
-\end{itemize}
-
-\autoref{fig:genbarrier} shows code that implements barrier synchronization
-using these stages.
-Method~\texttt{enter}() has two <{wait}> loops, one for each
-waiting condition.
-This is sometimes called \emph{double turnstile}.
-Each loop uses a different condition variable.
-The first loop waits until the car has emptied out, while
-the second waits for the car to fill up.
-\autoref{fig:barriertest} is a test program.  The threads check
-that all threads within the barrier are in the same round.
+
+\autoref{fig:barriertest} is a test program for barriers.
+It uses an integer array <{round}> with one entry per thread.
+Each thread, in a loop, waits for all threads to get to the barrier
+before incrementing its round number.  If the barrier works as
+advertised, two threads should never be more than one round apart.
+
+When implementing a barrier, a complication to worry about is that
+a barrier can be used over and over again.  If this were not the
+case, then a solution based on a lock, a condition variable,
+and a counter initialized to the number of threads could be used.
+The threads would decrement the counter and wait on the condition
+variable until the counter reaches 0.
+
+\autoref{fig:genbarrier} shows how one might implement a reusable barrier.
+Besides a counter <{.left}> that counts how many threads still have to
+reach the barrier, it uses a counter <{.cycle}> that is incremented
+after each use of the barrier---to deal with the complication above.
+The last thread that reaches the barrier restores
+<{.left}> to the number of threads
+(<{.required}>) and increments the cycle counter.
+The other threads are waiting for the cycle counter to be incremented.
+The cycle counter is allowed to wrap around---in fact, a single bit
+suffices for the counter.
+
+A common design pattern with barriers in parallel programs,
+demonstrated in \autoref{fig:barriertest2}, is to use the barrier
+twice in each round.  Before a round starts, one of the threads---let's
+call it the coordinator---sets up the work that needs to be done
+while the other threads wait.  Then all threads do the work and go
+on until they reach a second barrier.  The second barrier is used
+so the coordinator can wait for all threads to be done before setting
+up the work for the next round.
 

### 2021-12-25 book/paper.tex
Pass over Chapter 22

@@
 
-You would be excused if you wanted to solve the problem using locks, similar to
-\autoref{fig:incfixed}.  \autoref{fig:trap3} shows how one might go about this.
-But locks are intended to solve synchronization issues between multiple threads.
-If you run the code through Harmony, you will find that the code may not terminate.
-The issue is that a thread can only acquire a lock once.  If the interrupt happens
-after <{main()}> acquires the lock but before <{main()}> releases it, the
-<{handler}>() method will block trying to acquire the lock, even though
-it is being acquired by the same thread that already holds the lock.
+You would be excused if you wanted to solve the problem using locks,
+similar to \autoref{fig:incfixed}.  \autoref{fig:trap3} shows how
+one might go about this.  But locks are intended to solve synchronization
+issues between multiple threads.  But an interrupt handler is not
+run by another thread---it is run by the same thread that experienced
+the interrupt.  If you run the code through Harmony, you will find
+that the code may not terminate.  The issue is that a thread can
+only acquire a lock once.  If the interrupt happens after <{main()}>
+acquires the lock but before <{main()}> releases it, the <{handler}>()
+method will block trying to acquire the lock, even though it is
+being acquired by the same thread that already holds the lock.
 
@@
 asynchronous events.  Because interrupts may be disabled only briefly, interrupt
-handlers must run quickly and cannot wait for other events.  It is ok to invoke
+handlers must run quickly and cannot wait for other events.
+\begin{comment}
+It is ok to invoke
 non-blocking synchronization calls such as <{notify}>, but calls such as
@@
 \emph{producers} but not \emph{consumers} of synchronization events.
+\end{comment}
 

### 2021-12-25 book/paper.tex
Pass over Ch 23 and 24

@@
 
+<{:module:abp}>
 <{:const:net_send}>
@@
 application message
-<{m}>, then $m = m_i$ for some <{i}> and Bob will already have received
-application messages $m_1, ..., m_i$ (safety).
+<{m}>, then $m = m_i$ for some <{i}> and, unless $m$ is the very first
+message, Bob will already have received
+application messages $m_1, ..., m_{i-1}$ (safety).
 Also, if the network is fair and Alice sends application message

### 2021-12-26 book/paper.tex
Pass over Ch 25

@@
 
-The protocol that we present elects as leader the processor with the
-highest identifier~\cite{CR79} and works in two phases: in the first
-phase, each processor sends its identifier to its successor.  When
-a processor receives an identifier that is larger than its own
-identifier, it forward that to its successor as well.  If a processor
-receives its own identifier, it discovers that it is the leader.  That
-processor then starts the second phase by sending a message around the
-ring notifying the other processors of the leader's identifier.
-
-\autoref{fig:leader} describes the protocol and its test cases in Harmony.
-In Harmony, processors can be modeled by threads and there are a variety of ways in
-which one can model a network using shared variables.
-Here, we model the network as a set of messages.
-The \texttt{send} method atomically adds a message to this set.
-Messages are tuples with three fields:
-(<{dst}>, <{id}>, <{found}>).
-<{dst}> is the identifier of the destination processor;
-<{id}> is the identifier that is being forwarded; and
-<{found}> is a boolean indicating the second phase of the protocol.
-The <{receive}>(<{self}>) method looks for all messages
-destined for the processor with identifier <{self}>.
+The protocol that we present elects as leader the processor with
+the highest identifier~\cite{CR79} and works in two phases: in the
+first phase, each processor sends its identifier to its successor.
+When a processor receives an identifier that is larger than its own
+identifier, it forwards that identifier to its successor as well.
+If a processor receives its own identifier, it discovers that it
+is the leader.  That processor then starts the second phase by
+sending a message around the ring notifying the other processors
+of the leader's identifier.
+
+\autoref{fig:leader} describes the protocol and its test cases in
+Harmony.  In Harmony, processors can be modeled by threads and there
+are a variety of ways in which one can model a network using shared
+variables.  Here, we model the network as a set of messages.  The
+\texttt{send} method atomically adds a message to this set.  Messages
+are tuples with three fields: (<{dst}>, <{id}>, <{found}>).  <{dst}>
+is the identifier of the destination processor; <{id}> is the
+identifier that is being forwarded; and <{found}> is a boolean
+indicating the second phase of the protocol.  The <{receive}>(<{self}>)
+method looks for all messages destined for the processor with
+identifier <{self}>.
 
@@
 identifiers.  It also keeps track in the variable <{leader}> of
-what the highest identifier is so it can later be checked.
-
-Method \texttt{processor}(<{self}>, <{succ}>) is the
-code for a processor with identifier <{self}> and successor
-<{succ}>.
+what the highest identifier is, so it can later be checked.
+
+Method <{processor(self, succ)}> is the code for a processor with
+identifier <{self}> and successor <{succ}>.
 It starts simply by sending its own identifier to its successor.
@@
 <{atomically when exists}> statement.
-This statement takes the form <{atomically when exists v in s}>: \emph{statement block}
+This statement takes the form
+\begin{code}
+<{atomically when exists v in s}>: \emph{statement block}
+\end{code}
 where <{s}> is a set and <{v}> is variable that is bound to an element of <{s}>.

### 2021-12-26 book/paper.tex
Working on formatting

@@
 <{:const:g}>
+<{:const:h}>
 
@@
 (If you want to model a disk, this is probably best done as a separate
-thread.)
+thread that maintains the contents of the disk.)
 The complication we address here is that databases may be \emph{sharded},
@@
 The operation can be quite complex, and the execution of a transaction
-should have at least the following two properties:
+should have the following two properties:
 \begin{itemize}
@@
 
-We will use as an example a distributed database that maintains accounts.
-For simplicity we will model this as a collection of banks, each maintaining a
-single account.  There are two kinds of transactions: \emph{transfer}
+We will use as an example a distributed database that maintains bank accounts.
+For simplicity, we will model this as a collection of banks,
+each maintaining a single account.
+There are two kinds of transactions: \emph{transfer}
 (similar to \autoref{ex:bank}) and \emph{check}.  In this example, a transfer
@@
 A server can either respond with
-\texttt{YES} if it is ready to commit and will avoid doing anything
+<{YES}> if it is ready to commit and will avoid doing anything
 that might jeopardize this (like committing a conflicting transaction),
-or with \texttt{NO} if it does not want to participate in the transaction.
-If all servers respond with \texttt{YES}, then the coordinator can
+or with <{NO}> if it does not want to participate in the transaction.
+If all servers respond with <{YES}>, then the coordinator can
 decide to \emph{commit} the transaction.  Otherwise the coordinator
 must decide to \emph{abort} the transaction.  In the second phase, the
-servers that responded with \texttt{YES} (if any) must be notified
+servers that responded with <{YES}> (if any) must be notified
 to inform them of the coordinator's decision.
@@
 Addresses are lexicographically ordered accordingly;
-\item contexts (\autoref{app:context}) are ordered first by name, then by program counter, then by the remaining content.
+\item contexts (\autoref{app:context}) are ordered first by name, then by program counter, then by the remaining content in an unspecified way.
 \end{itemize}
@@
 and expressions where booleans are expected, it is not possible to
-substibute values of other types.
+substitute values of other types.
 
@@
 A string is a sequence of zero or more unicode characters.
-If they do not contain special characters and do not start with a digit,
+If it consists entirely of alphanumerical characters or underscore
+characters and does not start with a digit,
 then a string can be represented by a ``.'' followed by the characters.
@@
 $\{ k_0:v_0, k_1:v_1, ... \}$.
-Different from Python, the empty dictionary is either <{()}> or $[]$
-(because $\{\}$ is the empty set in Harmony).
+Different from Python, the empty dictionary is either <{()}> or <{[]}>
+(because <{{}}> is the empty set in Harmony).
 If there are duplicate keys in the list, then only the one with
@@
 Dictionaries support comprehension.  The basic form is:
-\texttt{\{ f(<{k}>):g(<{k}>) for <{k}> in <{s}> \}}.
+<{{ f(k):g(k) for k in s }}>.
 
@@
 \hline
-<{<{v [not] in keys d}> & check if <{v}> is [not] a key in <{d}> \\
+<{v [not] in keys d}> & check if <{v}> is [not] a key in <{d}> \\
 \hline
@@
 Because in Harmony brackets are used for parsing purposes only,
-you can write $d[k]$ (or $d(k)$) instead of $d~k$.
+you can write <{d[k]}> (or <{d(k)}>) instead of <{d k}>.
 However, if <{k}> is a string like <{.id}>, then you might
-prefer the notation $k\mathtt{.id}$.
-
-Python users beware: the Harmony \texttt{<{v}> in <{d}>} operator checks
-if there is some key <{k}> such that $d[k] = v$.  In Python, the same
+prefer the notation <{k.id}>.
+
+Python users beware: the Harmony <{v in d}> operator checks
+if there is some key <{k}> such that <{d[k] = v}>.  In Python, the same
 syntax checks if <{v}> is a key in <{d}>.  The difference exists because
@@
 Lists and tuples support comprehension as well.
-In their most basic form: \texttt{f(<{v}>) for <{v}> in <{s}>}.
+In their most basic form: <{(f(v) for v in s)}>.
 For example, to check if any element in a list <{t}> is even, you
-can write:
-
-\texttt{any($(x \% 2) == 0$ for <{x}> in <{t}>)}.
+can write: <{any((x % 2) == 0 for x in t)}>.
 
@@
 
-You can create lambda functions similarly to Python.  For example:
-\texttt{lambda$(x,y)$:$x+y$}.
+You can create lambda functions similarly to Python, except that the
+expression has to end on the keyword <{end}>.  For example:
+<{lambda(x,y): x+y end}>.
 
@@
 
-A thread can call \texttt{get\_context()} to retrieve its current
+A thread can call <{get_context()}> to retrieve its current
 context.
@@
 \hline
-<{let a = e: S} & <{a}> is a bound variable, <{e}> is an expression\\
+<{let a = e: S}> & <{a}> is a bound variable, <{e}> is an expression\\
 \hline
@@
 <{x}> may be a local variable or a shared variable.
-The statement $x = y = 3$ first updates~<{y}>, then~<{x}>.
-The statement $x[\mathtt{f}()] = y[\mathtt{g}()] = \mathtt{h}()$ first computes
-the address of $x[\mathtt{f}()]$, then computes the address of
-$y[\mathtt{g}()]$, then evaluates $\mathtt{h}()$, then assigns the resulting value to $y[\mathtt{g}()]$ (using its previously computed address), and finally assigns the same value to $x[\mathtt{f}()]$ (again using its previously computed address).
-The statement $a,b = c$ assumes that <{c}> is a tuple with two values.
+The statement <{x = y = 3}> first updates~<{y}>, then~<{x}>.
+The statement <{x[f()] = y[g()] = h()}> first computes
+the address of <{x[f()]}>, then computes the address of
+<{y[g()]}>, then evaluates <{h()}>,
+then assigns the resulting value to <{y[g()]}>
+(using its previously computed address),
+and finally assigns the same value to <{x[f()]}>
+(again using its previously computed address).
+The statement <{a,b = c}> assumes that <{c}> is a tuple with two values.
 It first evaluates the addresses of <{a}> and <{b}> and first assigns to
@@
 
-Assigning to $\_$ evaluates the righthand side expression but is otherwise
+Assigning to <{_}> evaluates the righthand side expression but is otherwise
 a no-op.
@@
 results in <{x}>.  In this case, it is equivalent to $x = x + 3$.
-However, in general this is not so.  For example, $x[\mathtt{f}()]~+$$= 3$
-only evaluates $\mathtt{f}()$ once.  Unlike Python, however,
-$x~+$$= [3,]$ is equivalent to $x = x + [3,]$ in Harmony.
+However, in general this is not so.  For example, <{x[f()] += 3}>
+only evaluates <{f()}> once.  Unlike Python, however,
+<{x += [3,]}> is equivalent to <{x = x + [3,]}> in Harmony.
 (In Python, the following two compound statements lead to different results
-for <{y}>: $x = y = []; x~+$$= [3]$ and $x = y = []; x = x + [3]$.)
+for <{y}>: <{x = y = []; x += [3]}> and <{x = y = []; x = x + [3]}>.)
 

### 2021-12-26 book/paper.tex
Fixed some more formatting
Latest

@@
 
+<{:const:tas}>
+<{:const:cas}>
+<{:const:BinSem}>
+<{:const:Semaphore}>
+
 \chapter{Hello World!}
@@
 outputs in the same order.
-The trick then is to ensure that all replicas handle the
+Therefore, the trick is to ensure that all replicas handle the
 same requests in the same order and to do so in a way that continues
@@
 (and only one \emph{tail}, possibly the same replica).
+
 We model the state machine as a \emph{history}: a sequence of operations.
@@
 Of course, a replica does not know whether it is immortal
-or not in practice---it should just assume so.
-The immortality of one of the replicas is only used for modeling the system.
+or not in practice---it should just assume that it is.
+The immortality of one of the replicas is only used for modeling
+the assumptions we make about the system.
 
@@
 <{NOPS}> operations to its history.
-$\mathbf{choose}(\{ \mathit{immortal}, <{True}> \})$ can only
-evaluate to <{False}> in case $<{immortal}>$ is false; otherwise
+<{choose({ immortal, True })}> can only
+evaluate to <{False}> in case <{immortal = False}>; otherwise
 it will always evaluate to <{True}>.
@@
 \hline
-<{cas}(ptr, old, new)}> & compare-and-swap on <{!ptr}> \\
+<{cas(ptr, old, new)}> & compare-and-swap on <{!ptr}> \\
 \hline

### 2021-12-26 book/paper.tex
Rewrote the Chain Replication chapter

@@
 \begin{code}
+<{:input:../code/rsmspec.hny}>
+\end{code}
+\caption{\harmonylink{code/rsmspec.hny} Replicated State Machine}
+\label{fig:rsmspec}
+\end{figure}
+
+\begin{figure}
+\begin{code}
 <{:input:../code/chain.hny}>
 \end{code}
-\caption{\harmonylink{code/chain.hny} Checking correctness of Chain Replication}
+\caption{\harmonylink{code/chain.hny} Chain Replication}
 \label{fig:chain}
@@
 outputs in the same order.
-Therefore, the trick is to ensure that all replicas handle the
+
+\autoref{fig:rsmspec} presents a Harmony specification of state machine
+replication.
+We model the state machine as a \emph{history}: a sequence of operations.
+Each replica has a local copy of the history.
+The network is modeled as an ordered queue of messages.
+<{NOPS}> clients each place an operation on the network.
+All but one of the replicas is allowed to crash.
+The model chooses one of the replicas that is not allowed to crash.
+Of course, a replica does not know whether it is immortal
+or not in practice---it should just assume that it is.
+The immortality of one of the replicas is only used for modeling
+the assumptions we make about the system.
+A replica loops until either it fails or it has received all operations.
+<{choose({ immortal, True })}> can only
+evaluate to <{False}> in case <{immortal = False}>; otherwise
+it will always evaluate to <{True}>.
+After a replica has received all <{NOPS}> operations, the replica
+finishes and prints its history.
+All replicas will print the same history (which should be the same
+as the network).
+
+But in reality the network is not an ordered queue and better modeled
+as a set of messages.
+The trick now is to ensure that all replicas handle the
 same requests in the same order and to do so in a way that continues
 to work even if some strict subset of replicas crash.
-
 \emph{Chain Replication}~\cite{vRS04} is such a replication protocol.
@@
 
-We model the state machine as a \emph{history}: a sequence of operations.
-Only the head is allowed to introduce new updates.
-When it does so, it advertises the new update along with its position
-in the history.  When the direct successor receives the next update
-it is expecting, it appends the update to its history and likewise
-advertises the new update.  When the update is applied to the history
-of the tail replica (the replica that has no live successors),
-then the update is considered final.
+Only the head is allowed to accept new updates from clients.
+When it does so, it adds the update to the end of its history
+and sends the history to its successor on the chain.
+When the direct successor receives such a history, it makes sure
+that the history is an extension of its own and, if so, replaces its own
+history with the one received.
+It then sends the history on to its successor, if any.
 
@@
 This is called the \emph{Fail-Stop} failure model~\cite{SS83},
-which is distinct from the often more realistic \emph{Crash} failure model
+which is distinct from the often more realistic \emph{Crash Failure} model
 where processes can crash but accurate detection is not available.
-We will consider that failure model in the upcoming chapters.
+We will consider that more realistic failure model in the upcoming chapters.
 
 \autoref{fig:chain} shows an implemenation of chain replication.
-The code starts \texttt{NREPLICAS} \texttt{replica} threads numbered
-0 through $\mathtt{NREPLICAS} - 1$, with initially 0 being the head
-and $\mathtt{NREPLICAS} - 1$ being the tail.  At least one of the replicas 
-must be immortal---the code selects one of them although none of the
-replicas is required to fail during execution.
-Of course, a replica does not know whether it is immortal
-or not in practice---it should just assume that it is.
-The immortality of one of the replicas is only used for modeling
-the assumptions we make about the system.
-
 Each replica maintains its history and a set of its predecessors.
-It then loops until either it fails or it has applied all
-<{NOPS}> operations to its history.
-<{choose({ immortal, True })}> can only
-evaluate to <{False}> in case <{immortal = False}>; otherwise
-it will always evaluate to <{True}>.
 What it does within the loop depends on whether it is the head
@@
 a <{.crash}> message on the network.
-The other form of payload is a tuple
-$(\mathit{index}, \mathit{operation})$, which means that the source
-placed the given operation at the given index in its history.
+The other form is a history of operations.
 
 If a replica thinks it is the head, it adds an operation to its
-history.
-In practice this would be some operation that the head would have received
-from a client.  Clients send their operations to the head, which adds them
-to its history in the order received.
-We are not modeling clients of the service.
-Instead, in the model, the head adds its identifier <{self}> to the
-history until the
-history has <{NOPS}> operations on it or until the head crashes,
-whichever comes first.  The head also broadcasts each update, which
-are intended for its successor.  Because of failures, it may be that
+history.  The head then broadcasts its history, which
+is intended for its successor.  Because of failures, it may be that
 at different times a replica has different successors, hence the
@@
 crashes, then replica~2 removes~1 from its set of predecessors, making
-replica~0 its direct (and only) predecessor.  If, instead, the payload
-is a tuple, then it must be the case that it contains the next operation
-to add to its history.  The replica adds the operation to its history
-and broadcasts the update, again intended for its successor.
-
-When the replica successfully adds all <{NOPS}> updates to its
-history without crashing, the code makes sure that any other such replica
-ends in the same state.  Variable <{final}> is used to check the
-different histories against one another.
+replica~0 its direct (and only) predecessor.
+Otherwise, the payload must be a history.
+The replica only accepts the history if it is an extension of its own
+history.
+When the replica successfully received all <{NOPS}> updates
+without crashing, it prints its history.
+
+The question is whether \autoref{fig:chain} has the same behavior as
+the replicated state machine specification of \autoref{fig:rsmspec}.
+This can be checked using the following two Harmony commands:
+\begin{code}
+\begin{verbatim}
+$ harmony -o rsm.hfa code/rsmspec.hny
+$ harmony -B rsm.hfa code/chain.hny
+\end{verbatim}
+\end{code}
+
+The first command outputs the behavior DFA of \texttt{code/rsmspec.hny}
+in the file \texttt{rsm.hfa}.
+The second command checks that the behavior of \texttt{code/chain.hny}
+satisfies the DFA in \texttt{rsm.hfa}.
+The DFA can also be visualized using the command
+\texttt{harmony -o rsm.png code/rsmspec.hny}
+(\autoref{fig:rsmdfa}).
+It shows all permutations of the operations are possible outputs.
+Moreover, each is output at least once, and at most
+<{NREPLICAS}> times.
+
+\begin{figure}
+\begin{center}
+\includegraphics[width=0.8\textwidth]{figures/rsmspec.png}
+\end{center}
+\caption{The DFA generated by both
+\autoref{fig:rsmspec} and \autoref{fig:chain}}
+\label{fig:rsmdfa}
+\end{figure}
 

### 2021-12-26 book/paper.tex
Added a practical note

@@
 
+The model has each replica send its entire history each time it extends
+its history.  This is fine for modeling, but in practice that would not
+scale.  In practice, a predecessor would set up a TCP connection to
+its successor and only send updates to its history along the TCP
+connection.  Because TCP connections guarantee FIFO order, this would
+be identical to the predecessor sending a series of histories, but
+much more efficient.
+
 \chapter{Replicated Atomic Read/Write Register}

### 2021-12-26 book/paper.tex
Rewrote parts of the Bosco chapter

@@
 it can acquire the lock.
-The program uses $\mathtt{N}+1$ boolean variables.
+The program uses $<{N}>+1$ boolean variables.
 Variable <{shared}> is initialized to <{False}> while
@@
 items, verifying that they form the set
-$\{ 0 .. \mathtt{N} - 1 \}$.
+$\{ 0 .. <{N}> - 1 \}$.
 
@@
 The queue is initialized with the sequence
-$0 .. \mathtt{N} - 1$.
+$0 .. <{N}> - 1$.
 Thread \texttt{getter(<{v}>)} removes an entry from the
@@
 In order to tolerate <{F}> failures,
-it uses $\mathtt{N} = 2\mathtt{F} + 1$ servers.
+it uses $<{N}> = 2<{F}> + 1$ servers.
 In other words, the register survives as long as a strict majority of
@@
 In the first phase, a client asks each server for its current
-timestamp and value.  After receiving $\mathtt{N} - \mathtt{F}$ responses,
+timestamp and value.  After receiving $<{N}> - <{F}>$ responses,
 the client determines the response with the highest timestamp.
@@
 the server responds to the client.
-Upon $c'$ receiving a response from $\mathtt{N} - \mathtt{F}$ servers,
+Upon $c'$ receiving a response from $<{N}> - <{F}>$ servers,
 the write operation completes.
@@
 to find all combinations of subsets of responses of size
-$\mathtt{N} - \mathtt{F}$.
+$<{N}> - <{F}>$.
 The second phase of each operation is similar.
@@
 \begin{figure}
-\begin{code}
-<{:input:../code/bosco.hny}>
-\end{code}
-\caption{\harmonylink{code/bosco.hny} A crash-tolerant consensus protocol}
-\label{fig:bosco}
+\begin{center}
+\includegraphics[width=0.9\textwidth]{figures/consensus.png}
+\end{center}
+\begin{code}
+<{:input:../code/consensus.hny}>
+\end{code}
+\caption{\harmonylink{code/consensus.hny} Distributed consensus code and DFA}
+\label{fig:consensus}
 \end{figure}
@@
 for example in the face of processor failures.
-In this chapter, we will present a simple consensus algorithm that can tolerate
+
+Each processor \emph{proposes} a value, which we assume here to be from the
+set \{ 0, 1 \}.
+By the usual definition of consensus, we want the following three properties:
+\begin{enumerate}
+\item \emph{Validity}: a processor can only decide a value that has been proposed;
+\item \emph{Agreement}: if two processors decide, then they decide the same value.
+\item \emph{Termination}: each processor eventually decides.
+\end{enumerate}
+The consensus problem is impossible to solve in the face of processor
+failures and without making assumptions about how long it takes to send
+and receive a message~\cite{FLP85}.
+Here we will not worry about \emph{Termination}.
+
+\autoref{fig:consensus} presents a specification for binary
+consensus---the proposals are from the set <{{0, 2}}>
+In this case there are four processors.  The proposal of processor <{i}>
+is in <{proposals[i]}>.
+The <{decision}> is chosen from the set of proposals.
+Each processor may or may not print the decision---capturing the
+absence of the \emph{Termination} property.
+It may be that no decisions are made, but that does not
+violate either Validity or Agreement.
+Thus the behavior of the program is to first print the array of
+proposals, followed by some subset of processors printing their decision.
+Notice the following properties:
+\begin{itemize}
+\item there are $16 = 2^4$ possible proposal configurations;
+\item all processors that decide decide the same value;
+\item if all processors propose 0, then all processors that decide decide 0;
+\item if all processors propose 1, then all processors that decide decide 1.
+\end{itemize}
+
+This is just the specification---in practice we do not have a shared
+variable in which we can store the decision a priori.
+We will present a simple consensus algorithm that can tolerate
 fewer than $1/3^{rd}$ of processors failing by crashing.
-
 More precisely,
@@
 we will assume there are $<{N}> = 3<{F}> + 1$ processors.
-Each processor \emph{proposes} a value, which we assume here to be from the
-set \{ 0, 1 \}.
-By the usual definition of consensus, we want the following two properties:
-\begin{enumerate}
-\item \emph{Validity}: a processor can only decide a value that has been proposed;
-\item \emph{Agreement}: if two processors decide, then they decide the same value.
-\end{enumerate}
-The consensus problem is surprisingly tricky to solve in the face of processor
-failures and without making assumptions about how long it takes to send
-and receive a message.
-
-\autoref{fig:bosco} presents our algorithm.  Besides the network variable, it uses a
-shared list of proposals and a shared set of decisions.
-To reduce the state space to explore, not all permutations of zeroes and ones are
-explored.  With 5 processors ($\mathtt{F} = 2)$, say, we only explore the cases where
-no processors propose 1, where exactly one processors proposes 1, and where 2 processors
-proposes 1.
+
+\begin{figure}
+\begin{code}
+<{:input:../code/bosco.hny}>
+\end{code}
+\caption{\harmonylink{code/bosco.hny} A crash-tolerant consensus protocol}
+\label{fig:bosco}
+\end{figure}
+
+\begin{figure}
+\begin{center}
+\includegraphics[width=0.95\textwidth]{figures/bosco.png}
+\end{center}
+\caption{The DFA for \autoref{fig:bosco}}
+\label{fig:boscodfa}
+\end{figure}
+
+\autoref{fig:bosco} presents our algorithm.  Besides the <{network}>
+variable,
+it uses a shared list of proposals and a shared set of decisions.
 In this particular algorithm, all messages are broadcast to all processors, so they do
 not require a destination address.
-The invariants we want to maintain
-on those variables is that $<{decisions}> \subseteq <{proposals}>$
-and $|<{decisions}>| \leq 1$.  Since the initial value of <{decisions}> is
-$\emptyset$, the invariants clearly hold initially.
-
-The $\mathtt{N}$ processors go through a sequence of \emph{rounds} in which they wait for
-$<{N}> - <{F}>$ messages, update their state based on the messages, and
+The $<{N}>$ processors go through a sequence of \emph{rounds} in which they wait for
+<{N - F}> messages, update their state based on the messages, and
 broadcast messages containing their new state.
-The reason that a processor waits for $<{N}> - <{F}>$ rather than <{N}>
-messages is because of failures: up to $<{F}>$ processors may never send a message
-and so it would be unwise to wait for all $<{N}>$.  You might be tempted to use
+The reason that a processor waits for <{N - F}> rather than <{N}>
+messages is because of failures: up to <{F}> processors may never send a message
+and so it would be unwise to wait for all <{N}>.  You might be tempted to use
 a timer and time out on waiting for a particular processor.  But how would you initialize
@@
 that messages arrive within a particular time.
-We call a set of $<{N}> - <{F}>$ processors a \emph{quorum}.
+We call a set of <{N - F}> processors a \emph{quorum}.
 A quorum must suffice for the algorithm to make progress.
@@
 Harmony can only check finite models, so there is a constant \texttt{NROUNDS} that
-limits the number of rounds.  It may be that no decisions are made, but that does not
-violate either Validity or Agreement.  We only check that \emph{if} decisions are
-made, they satisfy Validity and Agreement.
-
-In Line~22, a processor waits for <{N - F}> messages using
+limits the number of rounds.
+
+In Line~21, a processor waits for <{N - F}> messages using
 the Harmony <{atomically when exists}> statement.
@@
 The body of the <{atomically when exists}> statement contains the core of the algorithm.
-Note that <{N - F = 2F + 1}>, so that the number of
+Note that $<{N}> - <{F}> = 2<{F}> + 1$, so that the number of
 messages is guaranteed to be odd.  Also, because there are only 0 and 1 values, there
@@
 
+To check for correct behavior, run the following two commands:
+\begin{code}
+\begin{verbatim}
+$ harmony -o consensus.hfa code/consensus.hny
+$ harmony -B consensus.hfa code/bosco.hny
+\end{verbatim}
+\end{code}
+
+You will note that the second command prints a warning:
+``\texttt{behavior warning: strict subset of specified behavior}.''
+Thus, the set of behaviors that our algorithm generates is a subset
+of the behavior that the specification allows.
+\autoref{fig:boscodfa} shows the behavior, and indeed it is not the
+same as the behavior of \autoref{fig:consensus}.
+This is because in our algorithm the outcome is decided a priori if more than
+twothirds of the processors have the same proposal, whereas in the
+consensus specification the outcome is only decided a priori if the
+processors are initially unanimous.
+
 \begin{figure}
@@
 
-While one can run this code within little time for $\mathtt{F} = 2$, for
-$\mathtt{F} = 3$ the state space to explore is already quite large.
+While one can run this code within little time for <{F = 1}>, for
+<{F = 2}> the state space to explore is already quite large.
 One way to reduce the state space to explore is the following realization:
@@
 for \emph{all} <{N}> messages in a round
-instead of $\mathtt{N} - \mathtt{F}$.
-It would still have to choose to consider just $\mathtt{N} - \mathtt{F}$
+instead of <{N - F}>.
+It would still have to choose to consider just <{N - F}>
 out of those <{N}> messages, but executions in which some processors
 are left behind in all rounds are no longer considered.
-It still includes executions where some subset of $\mathtt{N} - \mathtt{F}$
+It still includes executions where some subset of <{N - F}>
 processors only choose each other messages and essentially ignore the
@@
 
-\autoref{fig:bosco2} shows the code for this model.  Running this with
-$\mathtt{F} = 3$ does not take very long and this approach is a good
-blueprint for testing other round-based protocols (of which there are
-many).
+Another way to reduce the state space to explore is to leverage
+symmetry.  First of all, it does not matter who proposes a particular
+value.  Also, the values 0 and 1 are not important to how the
+protocol operates.  So, with 5 processors (<{F = 2}>), say, we only
+need to explore the cases where no processors propose 1, where
+exactly one processors proposes 1, and where 2 processors proposes 1.
+
+\autoref{fig:bosco2} shows the code for this optimized model.
+Running this with <{F = 2}> does not take very long and this approach
+is a good blueprint for testing other round-based protocols (of
+which there are many).
 
@@
 conflicting messages to different peers (equivocation).
-The algorithm can tolerate those failures if you use $\mathtt{N} = 5\mathtt{F} - 1$ processors instead of $\mathtt{N} = 3\mathtt{F} - 1$.  Check that.
+The algorithm can tolerate those failures if you use $<{N}> = 5<{F}> - 1$ processors instead of $<{N}> = 3<{F}> - 1$.  Check that.
 \item In 1983, Michael Ben-Or presented a randomized algorithm that can
 tolerate crash failures with
-just $\mathtt{N} = 2\mathtt{F} - 1$ processors~\cite{BenOr83}.
+just $<{N}> = 2<{F}> - 1$ processors~\cite{BenOr83}.
 Implement this algorithm.
@@
 It uses two kinds of processors: \emph{leaders} and \emph{acceptors}.
-In order to tolerate <{F}> crash failures, you need at least $<{F}>+1$ leaders
+In order to tolerate <{F}> crash failures, you need at least $<{F}> + 1$ leaders
 and $2<{F}> + 1$ acceptors, but leaders and acceptors can be colocated, so

### 2021-12-30 book/paper.tex
Latest
Added nice picture for ABD

@@
 \begin{code}
+<{:input:../code/register.hny}>
+\end{code}
+\caption{\harmonylink{code/register.hny} An atomic read/write register}
+\label{fig:register}
+\end{figure}
+
+\begin{figure}
+\begin{center}
+\includegraphics[width=\textwidth]{figures/abdtest.png}
+\end{center}
+\begin{code}
+<{:input:../code/abdtest.hny}>
+\end{code}
+\caption{\harmonylink{code/abdtest.hny} Testing the code in \autoref{fig:abd}}
+\label{fig:abdtest}
+\end{figure}
+
+\begin{figure}
+\begin{code}
 <{:input:../code/abd.hny}>
@@
 \label{fig:abd}
-\end{figure}
-
-\begin{figure}
-\begin{code}
-<{:input:../code/abdtest.hny}>
-\end{code}
-\caption{\harmonylink{code/abdtest.hny} Testing the code in \autoref{fig:abd}}
-\label{fig:abdtest}
 \end{figure}
@@
 
-You will note that the second command prints a warning:
+Note that the second command prints a warning:
 ``\texttt{behavior warning: strict subset of specified behavior}.''
@@
 consensus specification the outcome is only decided a priori if the
-processors are initially unanimous.
+processors are initially unanimous.  Another difference is that if the
+outcome is decided a priori, all processors are guaranteed to decide.
 

### 2021-12-30 book/paper.tex
Latest

@@
 \begin{code}
+<{:input:../code/RWcheat.hny}>
+\end{code}
+\caption{\harmonylink{code/RWcheat.hny} "Cheating" reader/writer lock}
+\label{fig:rwcheat}
+\end{figure}
+
+\begin{figure}
+\begin{code}
+<{:input:../code/RWbtest.hny}>
+\end{code}
+\caption{\harmonylink{code/RWbtest.hny} A behavioral test of reader/writer locks}
+\label{fig:rwbtest}
+\end{figure}
+
+A problem with this test is that it does not find a problem
+with an implementation like the one in \autoref{fig:rwcheat}.
+This implementation implements a reader/writer lock as an ordinary
+lock, and thus lets only one thread in the critical section at a time.
+In some sense, the implementation is correct because it satisfies
+the requirements, but it is clearly not a desirable implementation.
+For a case like this one, it is better to compare behaviors between
+the specification and the implementation.
+
+\autoref{fig:rwbtest} is the same test as \autoref{fig:rwtest} but
+prints identifying information before and every lock operation.
+Now we can compare behaviors as follows:
+
+\begin{code}
+\begin{verbatim}
+$ harmony -o rw.hfa -o rwspec.png -cNOPS=2 code/RWbtest.hny
+$ harmony -B rw.hfa -o rwimpl.png -cNOPS=2 -m RW=RWcheat code/RWbtest.hny
+\end{verbatim}
+\end{code}
+
+The second command will print a warning that there are behaviors in the
+specification that are not achieved by the implementation.
+We set <{NOPS}> to 2 to make the behaviors relatively small and easier
+to compare.
+It is now possible to compare the outputs in \texttt{rwspec.png}
+and \texttt{rwimpl.png} and see what behaviors are allowed by the
+specification that are not allowed by the implementation.
+(Harmony does not yet have a mechanism to do this automatically.)
+
+\begin{figure}
+\begin{code}
 <{:input:../code/boundedbuffer.hny}>

### 2021-12-30 book/paper.tex
Latest

@@
 \setlist[problems,2]{label=(\alph*), wide =0.5em, topsep=2pt, itemsep=2pt}
+
+% \renewcommand{\dblfloatpagefraction}{0.99}
+% \renewcommand{\floatpagefraction}{.99}
+% \renewcommand{\dblfloatpagefraction}{0.99}
+% \renewcommand{\floatpagefraction}{.99}
 
@@
 it can acquire the lock.
-The program uses $<{N}>+1$ boolean variables.
+The program uses <{N + 1}> boolean variables.
 Variable <{shared}> is initialized to <{False}> while
@@
 Invariant $\mathcal{I}_1$ is also easy to verify for exiting the critical section
-because we can assume, by the induction hypothesis, that $<{private}>[i]$ is
+because we can assume, by the induction hypothesis, that <{private[i]}> is
 <{True}> just before exiting the critical section.
@@
 program is run against the specification in the case
-$<{NOPS}> = 2$---for $<{NOPS}> = 4$ it would be much
+<{NOPS = 2}>---for <{NOPS = 4}> it would be much
 too large to print here.  Since there are no cycles in the DFA, you
@@
 You can
-try this for various $<{NOPS}>$, although it gets increasingly
+try this for various <{NOPS}>, although it gets increasingly
 harder to check by hand that the generated DFAs are the same as
@@
 \end{code}
-\caption{\harmonylink{code/abdtest.hny} Testing the code in \autoref{fig:abd}}
+\caption{\harmonylink{code/abdtest.hny} Behavioral test for atomic read/write registers and the output for the case that <{NREADERS = NWRITERS = 1}>}
 \label{fig:abdtest}
@@
 
-A \emph{register} is an object that you can read or write.  In a distributed
-system, it can be useful to have registers that are shared.
+A \emph{register} is an object that you can read or write.
+In a distributed system, examples include a shared file system
+(each file is a register) or a key/value store (each key corresponds
+to a register).
 A simple shared register implementation
@@
 
+It is instructive to look at the test program and its output in
+\autoref{fig:abdtest}.  This is for the case when there is only
+a single reader thread (identified as ``1'') and a single writer
+thread (identified as ``$-1$''), but already
+there are many cases to consider.  Each thread prints information
+just before and just after doing their single operation.  The output
+shows all possible interleavings in the form of a DFA.  Note
+that if the read operation starts after the write operation
+has completed, then the read operaion must return the new value.
+However, if the two operations interleave in some way, then the
+read operation can return either the old or the new value.
+
 \index{single point of failure}%
-Unfortunately, however, the server is a \emph{single point of failure}:
+Unfortunately, a server is a \emph{single point of failure}:
 if it fails, all its clients suffer.
@@
 In order to tolerate <{F}> failures,
-it uses $<{N}> = 2<{F}> + 1$ servers.
+it uses <{N}> = 2<{F}> + 1 servers.
 In other words, the register survives as long as a strict majority of
@@
 a client broadcasts a request to all servers and waits for responses
-from a majority ($<{N}> - <{F}>$ or equivalently
-$<{F}> + 1$ servers).  Note that because we are assuming that no
+from a majority (<{N - F}> or equivalently
+<{F + 1}> servers).  Note that because we are assuming that no
 more than <{F}> servers can fail, doing so is safe, in that a client
@@
 In the first phase, a client asks each server for its current
-timestamp and value.  After receiving $<{N}> - <{F}>$ responses,
+timestamp and value.  After receiving <{N - F}> responses,
 the client determines the response with the highest timestamp.
@@
 the server responds to the client.
-Upon $c'$ receiving a response from $<{N}> - <{F}>$ servers,
+Upon $c'$ receiving a response from <{N - F}> servers,
 the write operation completes.
@@
 The <{receive()}> function uses the <{bag.combinations}> method
-to find all combinations of subsets of responses of size
-$<{N}> - <{F}>$.
+to find all combinations of subsets of responses of size <{N - F}>.
 The second phase of each operation is similar.
 
-\autoref{fig:abdtest} tests the third property of the properties listed
-above.  The <{writer()}> method writes a new value <{.token}>.
-The \texttt{reader()} method reads twice and makes sure that if the
-first read operation returns <{.token}>, then so does the second.
-This illustrates the importance of the second phase of the \texttt{read}
-operation.  You can comment out Lines~32 and~33 in \autoref{fig:abd} and
-see what goes wrong.
+\autoref{fig:abdtest} can be used to test this protocol, although
+you will notice that the model checker cannot deal with cases involving
+more than three client threads.
+Three is just enough to check the third property listed above
+(using one writer and two readers).
+Doing so illustrates the importance of the second phase of the \texttt{read}
+operation.  You can comment out Lines~34, 36, and~37 in \autoref{fig:abd}
+and to elide the second phase and see what goes wrong.
 
@@
 This includes executions in which the clients terminate before
-$<{F}>$ of the servers start executing.  To the clients, this
+<{F}> of the servers start executing.  To the clients, this
 is indistinguishable from executions in which those
@@
 constant <{F}> contains the maximum number of failures, and
-we will assume there are $<{N}> = 3<{F}> + 1$ processors.
+we will assume there are <{N}> = 3<{F}> + 1 processors.
 
@@
 not require a destination address.
-The $<{N}>$ processors go through a sequence of \emph{rounds} in which they wait for
+The <{N}> processors go through a sequence of \emph{rounds} in which they wait for
 <{N - F}> messages, update their state based on the messages, and
@@
 The body of the <{atomically when exists}> statement contains the core of the algorithm.
-Note that $<{N}> - <{F}> = 2<{F}> + 1$, so that the number of
+Note that <{N - F}> = 2<{F}> + 1, so that the number of
 messages is guaranteed to be odd.  Also, because there are only 0 and 1 values, there
@@
 It uses two kinds of processors: \emph{leaders} and \emph{acceptors}.
-In order to tolerate <{F}> crash failures, you need at least $<{F}> + 1$ leaders
+In order to tolerate <{F}> crash failures, you need at least <{F}> + 1 leaders
 and $2<{F}> + 1$ acceptors, but leaders and acceptors can be colocated, so

### 2021-12-31 book/paper.tex

@@
 problem and you may already have prototyped some solutions.
-In this chapter we will go through a few reasonable but broken attempts.
+In this chapter, we will go through a few reasonable but broken attempts.
 At the heart of the problem is that we would like make sure that, when
@@
 environments in which few or no assumptions are made about timing.
-In this chapter we present a basic version of a Paxos protocol, one that is
+In this chapter, we present a basic version of a Paxos protocol, one that is
 \emph{single-decree} (only tries to make a single decision).
@@
 this Paxos protocol.  Paxos is perhaps best understood starting
-with the second phase.  At the end of the first phase the leader
+with the second phase.  At the end of the first phase, the leader
 broadcasts a \texttt{2.A} message (Phase 2, Type A) to the acceptors
@@
 \emph{received}.  If its ballot number is higher than what it has
-seen before, the acceptor moves into that ballot.  If the phase is
-2, then the acceptor accepts the proposal and remembers when it did
+seen before, the acceptor moves into that ballot.  If the phase is~2,
+then the acceptor accepts the proposal and remembers when it did
 so by saving the (ballot, proposal) tuple in <{last_accepted}>.
@@
 
-In the initialization, the \texttt{proposals} variable is initialized
-to either $[\mathbf{False}, \mathbf{False}]$ or
-to $[\mathbf{True}, \mathbf{False}]$ (in case $\texttt{NBALLOTS} = 2$).
-The other combinations of <{True}> and <{False}> are symmetric,
-so there is no need to run the model checker through those scenarios as
-well, halving the running time.
+To run and check the Paxos code, do the following (leveraging the
+consensus specification of \autoref{fig:consensus}):
+\begin{code}
+\begin{verbatim}
+$ harmony -o consensus.hfa -cN=2 code/consensus.hny
+$ harmony -B consensus.hfa code/paxos.hny
+\end{verbatim}
+\end{code}
+
+You should get a warning that our implementation of Paxos does not
+generate all possible behaviors.
+This is because we only run the protocol for a finite number of ballots,
+and therefore at least one of the ballots will be successful.
+With an unlimited number of ballots, Paxos may never decide unless you
+make some liveness assumptions.
 
@@
 \begin{problems}
-\item Perhaps the trickiest detail of the algorithm is that, in Line~14
+\item Perhaps the trickiest detail of the algorithm is that, in Line~8
 of \autoref{fig:paxos2}, the leader selects the proposal with the highest

### 2022-01-04 book/paper.tex
More on behaviors

@@
 
-\begin{comment}
-The framework in \autoref{fig:csbarebones} works well for thoroughly
-testing mutual exclusion and progress in critical sections, but depends
-on \texttt{countLabel}, an unusual operator specific to Harmony that is mostly
-useful for testing mutual exclusion.
-It turns out that we do not need labels to express this.
-\autoref{fig:cslock} tests mutual exclusion and progress for
-critical sections implemented using locks.
-The test uses a simple atomic counter that is incremented before entering
-the critical section and decremented after leaving it.
-Mutual exclusion holds if the counter never goes over 1.
-\end{comment}
-
 As with critical sections---when testing a concurrent data structure---we
@@
 We say that the implementation and the specification
-have the same \emph{behavior}.
-%
+have the same \emph{behaviors} or are
+\emph{behaviorally equivalent}.
+
+Behaviors say something about how we got to a state.
+The same state can be reaching by multiple behaviors, and the behaviors
+are often an integral part of whether a program is correct or not.
+Just because a state satisfies some invariant---however important---does
+not mean that the state is valid given the sequence of operations.
+For example, a state in which the queue is empty is certainly a valid state in
+its own right, but if the last operation to get there was an enqueue operation,
+there must be a bug in the program.
+It can therefore be important to capture the behaviors.
+We could store behaviors in the state itself by adding what is
+known as a \emph{history variable} that keeps track of all the operations.
+While this can be useful for correctness proofs, for model checking this approach
+presents a problem: introducing this
+additional state can lead to state explosion or even turn a finite model
+(a model with a finite number of states) into an infinite one.
+We therefore use a different approach: composing an implementation with
+its specification to ensure that accept the same behaviors.
+
 \autoref{fig:qtestseq} presents a test program that does
@@
 have known bugs in them and make sure that the test program finds them.
-\begin{comment}
-In the case of testing mutual exclusion with atomic counters, you will want
-to see if you can use this method to find the problems in \autoref{fig:naivelock},
-\autoref{fig:naiveflags},
-and \autoref{fig:naiveturn}.
-\end{comment}
 Conversely, a test program may be broken in that it finds bugs that do not exist.
@@
 the algorithm that the test program tests.
-
-\begin{comment}
-So, in this case,
-you also want to check if an atomic counter solution works with known
-correct solutions such as \autoref{fig:peterson} and \autoref{fig:tas}
-if possible.
-\end{comment}
 
@@
 \begin{problems}
+\item \autoref{fig:lockspec} shows a specification of a lock.  Write a program
+that checks the behaviors of lock implementations such as \autoref{fig:tas2}
+and \autoref{fig:ticket}.  That is, it should not rely on assertions such as
+in \autoref{fig:cs}.
 \item Write a Harmony program that checks if

### 2022-01-06 book/paper.tex
Added a selection of operations on strings

@@
 \begin{problems}
+\item Write a Harmony program that uses <{choose}> instead of <{spawn}> to create the same
+output DFA as \autoref{fig:threads}(b).
 \item Add the line <{print(x, triangle(x))}> to the end of the
@@
 
+Native operations on strings include the following:
+
+\begin{center}
+\begin{tabular}{|l|l|}
+\hline
+<{s k}> & indexing \\
+\hline
+<{s + s + ...}> & concatenation \\
+\hline
+<{s * n}> & <{n}> copies of <{s}> concatenated \\
+\hline
+<{v [not] in s}> & check if <{v}> is [not] a substring in <{s}> \\
+\hline
+<{len s}> & the length of <{s}> \\
+\hline
+<{str e}> & string representation of any value <{e}> \\
+\hline
+\end{tabular}
+\end{center}
+
 \subsection*{Set}
@@
 \hline
-<{for a[:b] in e [where c]: S}> & <{a}> and <{b}> are bound variables, <{e}> is a set or dictionary \\
+<{for a[:b] in e [where c]: S}> & <{a}> and <{b}> are bound variables, <{e}> is a set, dictionary, or string \\
 \hline
@@
 <{S1, S2, ...}>.
-<{y}> must be a set or a dictionary.  <{y}> is evaluated only once at
+<{y}> must be a set, dictionary, or string.  <{y}> is evaluated only once at
 the beginning of the evaluation of this statement.  In case of a set,

### 2022-01-06 book/preproc.py

@@ -85,6 +85,7 @@ reserved = {
     "setintlevel",
     "spawn",
     "stop",
+    "str",
     "this",
     "trap",
     "True",

### 2022-01-06 src/charm/ops.c

@@ -346,6 +346,24 @@ void op_Apply(const void *env, struct state *state, struct step *step, struct gl
             step->ctx->pc++;
         }
         return;
+    case VALUE_ATOM:
+        {
+            if ((e & VALUE_MASK) != VALUE_INT) {
+                value_ctx_failure(step->ctx, &global->values, "Bad index type for string");
+                return;
+            }
+            e >>= VALUE_BITS;
+            int size;
+            char *chars = value_get(method, &size);
+            if (e >= size) {
+                value_ctx_failure(step->ctx, &global->values, "Index out of range");
+                return;
+            }
+            hvalue_t v = value_put_atom(&global->values, &chars[e], 1);
+            value_ctx_push(&step->ctx, v);
+            step->ctx->pc++;
+        }
+        return;
     case VALUE_PC:
         value_ctx_push(&step->ctx, ((step->ctx->pc + 1) << VALUE_BITS) | VALUE_PC);
         value_ctx_push(&step->ctx, (CALLTYPE_NORMAL << VALUE_BITS) | VALUE_INT);
@@ -355,8 +373,7 @@ void op_Apply(const void *env, struct state *state, struct step *step, struct gl
         return;
     default:
         {
-            extern char *json_escape_value(hvalue_t v);
-            char *x = json_escape_value(method);
+            char *x = value_string(method);
             value_ctx_failure(step->ctx, &global->values, "Can only apply to methods or dictionaries, not to: %s", x);
             free(x);
         }
@@ -477,7 +494,19 @@ void op_Cut(const void *env, struct state *state, struct step *step, struct glob
         step->ctx->pc++;
         return;
     }
-    panic("op_Cut: not a set or dict");
+    if ((v & VALUE_MASK) == VALUE_ATOM) {
+        assert(v != VALUE_ATOM);
+        int size;
+        char *chars = value_get(v, &size);
+        assert(size > 0);
+
+        ctx->vars = value_dict_store(&global->values, ctx->vars, ec->set, value_put_atom(&global->values, &chars[1], size - 1));
+        hvalue_t e = value_put_atom(&global->values, chars, 1);
+        var_match(step->ctx, ec->value, &global->values, e);
+        step->ctx->pc++;
+        return;
+    }
[…]

### 2022-01-06 src/charm/value.c

@@ -38,7 +38,9 @@ void *value_copy(hvalue_t v, int *psize){
 }
 
 hvalue_t value_put_atom(struct values_t *values, const void *p, int size){
-    assert(size > 0);
+    if (size == 0) {
+        return VALUE_ATOM;
+    }
     void *q = dict_find(values->atoms, p, size);
     return (hvalue_t) q | VALUE_ATOM;
 }
@@ -83,10 +85,9 @@ int value_cmp_int(hvalue_t v1, hvalue_t v2){
 }
 
 int value_cmp_atom(hvalue_t v1, hvalue_t v2){
-    void *p1 = (void *) v1, *p2 = (void *) v2;
     int size1, size2;
-    char *s1 = dict_retrieve(p1, &size1);
-    char *s2 = dict_retrieve(p2, &size2);
+    char *s1 = value_get(v1, &size1);
+    char *s2 = value_get(v2, &size2);
     int size = size1 < size2 ? size1 : size2;
     int cmp = strncmp(s1, s2, size);
     if (cmp != 0) {
@@ -301,9 +302,8 @@ static char *value_json_int(hvalue_t v) {
 }
 
 static char *value_string_atom(hvalue_t v) {
-    void *p = (void *) v;
     int size;
-    char *s = dict_retrieve(p, &size), *r;
+    char *s = value_get(v, &size), *r;
 	struct strbuf sb;
 
 	strbuf_init(&sb);
@@ -333,18 +333,11 @@ static char *value_string_atom(hvalue_t v) {
 }
 
 static char *value_json_atom(hvalue_t v) {
-    void *p = (void *) v;
     int size;
-    char *s = dict_retrieve(p, &size), *r;
-    assert(size > 0);
-    if (size > 1 || (isprint(s[0]) && s[0] != '"' && s[0] != '\\')) {
-        char *esc = json_escape(s, size);
-        alloc_printf(&r, "{ \"type\": \"atom\", \"value\": \"%s\" }", esc);
-        free(esc);
-    }
-    else {
-        alloc_printf(&r, "{ \"type\": \"char\", \"value\": \"%02x\" }", s[0]);
-    }
+    char *s = value_get(v, &size), *r;
+    char *esc = json_escape(s, size);
+    alloc_printf(&r, "{ \"type\": \"atom\", \"value\": \"%s\" }", esc);
+    free(esc);
     return r;
 }
 
[…]

### 2022-01-06 book/paper.tex
Added Towers of Hanoi as an example

@@
 
+<{:const:head}>
+<{:const:tail}>
+\begin{figure}
+\begin{code}
+<{:input:../code/hanoi.hny}>
+\end{code}
+\caption{\harmonylink{code/hanoi.hny} Towers of Hanoi}
+\label{fig:hanoi}
+\end{figure}
+
+As an example of using modules,
+\autoref{fig:hanoi} uses (or abuses?) Harmony to solve the
+\emph{Towers of Hanoi} problem.
+There are three towers with disks of varying sizes.
+In the initial configuration, the first tower has three disks (of
+sizes 1, 2, and 3), with the largest disk at the bottom, while the
+other two towers are empty.  You are allowed to move a top disk
+from one tower to another, but you are not allowed to stack a larger
+disk on a smaller one.  The objective is to move the disks from the
+first tower to the third one.
+
+The program uses the <{list}> module documented in \autoref{ap:list}.
+It has methods to extract the \emph{head} (first element) and
+the \emph{tail} (remaining elements) of a list.
+(The code is simple and available in \texttt{modules/list.hny}.)
+The program tries valid moves at random until it finds a solution.
+Curiously, the program then asserts <{False}>.  This is to cause
+the model checker to stop and output the trace.  If you look in the
+output column of the trace, you will find the minimal number of moves
+necessary to solve the problem.
+
 <{:const:ref}>
-
 \begin{figure}
@@
 
-\index{list module}%
+\index{hoare module}%
 The \texttt{hoare} module implements support for Hoare-style monitors

### 2022-01-06 book/paper.tex
Latest

@@
 
-As an example of using modules,
-\autoref{fig:hanoi} uses (or abuses?) Harmony to solve the
-\emph{Towers of Hanoi} problem.
-There are three towers with disks of varying sizes.
+As an example of using <{import}> and <{let}>,
+\autoref{fig:hanoi} solves the \emph{Towers of Hanoi} problem.
+If you are not familiar with this problem:
+there are three towers with disks of varying sizes.
 In the initial configuration, the first tower has three disks (of

### 2022-01-06 book/paper.tex
Latest

@@
 
+It is even cooler to remove that assertion and let Harmony generate
+all possible solutions to the problem like so:
+
+\begin{code}
+\begin{verbatim}
+$ harmony -o hanoi.png code/hanoi.hny
+\end{verbatim}
+\end{code}
+
+The resulting \texttt{hanoi.png} file contains a DFA describing
+the possible solutions.  It is a little too big to include here,
+but well worth looking at.
+
 <{:const:ref}>

### 2022-01-06 book/paper.tex
Latest

@@
 and University of Washington's DSLabs system~\cite{MWA19}.
+Some of the examples in this book are derived from those sources.
 Harmony is designed to have a lower learning curve than those
@@
 
+If you already know about concurrent and distributed programming and
+are just interested in a ``speed course'' on Harmony, I would recommend
+reading \autoref{ch:harmonyintro}, \autoref{ch:harmonymachine},
+\autoref{ch:method}, \autoref{ch:specification}, and \autoref{ch:cds}.
 The appendices contain various details about Harmony itself, including
-an appendix that explains how Harmony works (\autoref{app:howitworks}) and
-an appendix on installing and running Harmony (\autoref{app:install}).
+an appendix on convenient Harmony modules (\autoref{ap:module}),
+an appendix that explains how Harmony works (\autoref{ap:howitworks}),
+and
+an appendix on installing and running Harmony (\autoref{ap:install}).
 
@@
 Addresses are lexicographically ordered accordingly;
-\item contexts (\autoref{app:context}) are ordered first by name, then by program counter, then by the remaining content in an unspecified way.
+\item contexts (\autoref{ap:context}) are ordered first by name, then by program counter, then by the remaining content in an unspecified way.
 \end{itemize}
@@
 
-\chapter{The Harmony Virtual Machine}\label{app:hvm}
+\chapter{The Harmony Virtual Machine}\label{ap:hvm}
 
@@
 \section{Contexts and Threads}
-\label{app:context}
+\label{ap:context}
 
@@
 
-\chapter{How Harmony Works}\label{app:howitworks}
+\chapter{How Harmony Works}\label{ap:howitworks}
 
@@
 generates the next states using the Harmony virtual machine
-(\autoref{app:hvm}).  If the model is finite, eventually the model
+(\autoref{ap:hvm}).  If the model is finite, eventually the model
 checker will generate a graph with all possible states.
@@
 
-\chapter{Installing and Running Harmony}\label{app:install}
+\chapter{Installing and Running Harmony}\label{ap:install}
 

### 2022-01-06 book/paper.tex
Added an exercise

@@
 
+\section*{Exercises}
+\begin{problems}
+\item (This is just for fun or exercise as it is not a concurrent or
+distributed problem.)
+Implement a Harmony program that finds solutions to the
+``cabbage, goat, and wolf'' problem.  In this problem, a person
+accompanied by these three items has to cross a stream in a small
+boat, but can only take one item at a time. So, the person has to
+cross back and forth several times, leaving two items on one or
+the other shore by themselves.
+Unfortunately, if left to themselves, the goat would
+eat the cabbage and the wolf would eat the goat.  What crossings
+does the person need to make in order not to lose any items?
+\end{problems}
+
 \chapter{Specification}

### 2022-01-09 book/paper.tex

@@
 
+\section{Formal Specification}
+\label{ap:formal}
+
+A formal specification of the Harmony Virtual Machine is underway but
+not yet completed.
+Given a Harmony program, you can output the TLA+ specification for
+the program using the following command:
+
+\begin{code}
+\begin{verbatim}
+$ harmony -o program.tla program.hny
+\end{verbatim}
+\end{code}
+
+For many Harmony programs, such as Peterson's algorithm and the Dining
+Philosophers in this book, the result is complete enough to run through
+the TLC model checker.
+
 \chapter{How Harmony Works}\label{ap:howitworks}

### 2022-01-09 book/paper.tex
str implemented

--- 2022-01-06 book/paper.tex
@@
 \hline
+Invariant \textit{end} & code for invariant follows.  Skip to \textit{end} + 1\\
+\hline
 Jump <{p}> & set program counter to <{p}> \\
@@
 
+\section{Formal Specification}
+\label{ap:formal}
+
+A formal specification of the Harmony Virtual Machine is well underway but
+not yet completed.  At this time, there is no support yet for bitwise integer
+operations.
+Also, strings are limited to the printable characters minus
+double quotes, back quotes, or backslashes.
+Given a Harmony program, you can output the TLA+ specification for
+the program using the following command:
+
+\begin{code}
+\begin{verbatim}
+$ harmony -o program.tla program.hny
+\end{verbatim}
+\end{code}
+
+For most Harmony programs, including Peterson's algorithm and the Dining
+Philosophers in this book, the result is complete enough to run through
+the TLC model checker.
+
 \chapter{How Harmony Works}\label{ap:howitworks}

### 2022-01-10 book/paper.tex
Implemented setintlevel in TLA+
All except trap...
Latest

@@
 A formal specification of the Harmony Virtual Machine is well underway but
-not yet completed.  At this time, there is no support yet for bitwise integer
-operations.
+not yet completed.  In particular, <{trap}> is not yet specified.
 Also, strings are limited to the printable characters minus
 double quotes, back quotes, or backslashes.
+But everything else is specified.
 Given a Harmony program, you can output the TLA+ specification for
@@
 First and foremost I would like to thank my student Haobin Ni with
-whom I've had numerous discussions about the design of Harmony.
+whom I've had numerous discussions about the initial design of Harmony.
 Haobin even contributed some code to the Harmony compiler.
+Many thanks are also due to William Ma who refactored the Harmony code
+to make it easier to maintain.
+He also wrote the first version of the behavior automaton generator
+and created the first graphs using the graphviz tool.
+I have had lots of discussions with him about a wide range of
+improvements to the Harmony language, many of which came to fruition.
+I also want to thank Ariel Kellison with whom I discussed approaches
+to formally specify the Harmony virtual machine in TLA+.
 
@@
 about at \url{http://harmony.cs.cornell.edu}.  They also developed much of
-that web site.  Later they were joined by Shi Chong Zhao, who also
-made significant contributions.
+that web site and made valuable suggestions for improvements to the
+Harmony language.  Later they were joined by Shi Chong Zhao, who also
+made significant contributions.  Kevin and Anthony continue to make
+great contributions to the Harmony distribution.
 
@@
 of a concurrent system.  My experimentation with using TLC on Peterson's
-Algorithm became an aha moment for me.
+Algorithm became an aha moment for me.  I have learned so much from his
+papers.
 
 I first demonstrated Harmony to the students in my CS6480 class on systems
-and formal verification and
-received valuable feedback from them.
-
+and formal verification and received valuable feedback from them.
 The following people contributed by making comments on or finding bugs in
@@
 Trishita Tiwari,
-William Ma,
 Xiangyu Zhang,

### 2022-01-10 book/paper.tex
Trying to add a list of figures to book
Added list of figures to the book

@@
 \tableofcontents
+\listoffigures
 
@@
 \harmonyref{Up.html}{HTML output of running Harmony on \autoref{fig:inc}}.
-There are four sections.
-The top shows the turns to the problematic state.
-Each turn shows a thread identifier, the method the thread is executing,
-a ``timeline'' containing a block for each instruction executed, the
-program counter of the thread, the values of the shared variables
-at those turns, and any values the program may have printed.
-The middle section shows the line of code that is being executed.
-The bottom left shows the bytecode of the program.
-The bottom right shows the state of each thread.
-Each row in the stack trace of a thread shows
-the method that is being evaluated and
-the values of its method variables.}
+% There are four sections.
+% The top shows the turns to the problematic state.
+% Each turn shows a thread identifier, the method the thread is executing,
+% a ``timeline'' containing a block for each instruction executed, the
+% program counter of the thread, the values of the shared variables
+% at those turns, and any values the program may have printed.
+% The middle section shows the line of code that is being executed.
+% The bottom left shows the bytecode of the program.
+% The bottom right shows the state of each thread.
+% Each row in the stack trace of a thread shows
+% the method that is being evaluated and
+% the values of its method variables.
+}
 \label{fig:inchtml1}
@@
 
+\addtocontents{toc}{\protect\setcounter{tocdepth}{2}}
+
 \end{document}

### 2022-01-10 book/paper.tex
Improved figure captions a bit

@@
 \end{code}
-\caption{\harmonylink{code/hello3.hny} Two possible outputs}
+\caption{\harmonylink{code/hello3.hny} Harmony program with two possible outputs}
 \label{fig:choice}
@@
 \end{code}
-\caption{\harmonylink{code/hello4.hny} An infinite number of outputs}
+\caption{\harmonylink{code/hello4.hny} Harmony program with an infinite number of outputs}
 \label{fig:hello4}
@@
 \end{center}
-\caption{Methods and threads}
+\caption{Demonstrating Harmony methods and threads}
 \label{fig:threads}
@@
 \end{code}
-\caption{\harmonylink{code/hello7.hny} Interleaving threads}
+\caption{\harmonylink{code/hello7.hny} Various interleavings of threads}
 \label{fig:hello7}
@@
 \end{code}
-\caption{\harmonylink{code/hello8.hny} Less interleaving}
+\caption{\harmonylink{code/hello8.hny} Making groups of operations atomic reduces interleaving}
 \label{fig:hello8}
@@
 \end{code}
-\caption{\harmonylink{code/Up.hny} Incrementing twice in parallel}
+\caption{\harmonylink{code/Up.hny} Incrementing the same variable twice in parallel}
 \label{fig:inc}
@@
 \end{code}
-\caption{\harmonylink{python/UpMany.py} Incrementing <{N}> times}
+\caption{\harmonylink{python/UpMany.py} Using Python to increment <{N}> times}
 \label{fig:incmany}
@@
 \end{code}
-\caption{\harmonylink{code/UpEnter.hny} Broken attempt at fixing the code of \autoref{fig:inc}}
+\caption{\harmonylink{code/UpEnter.hny} Incorrect attempt at fixing the code of \autoref{fig:inc}}
 \label{fig:incenter}
@@
 \end{code}
-\caption{\harmonylink{code/UpLock.hny} Program of \autoref{fig:inc} fixed with a lock}
+\caption{\harmonylink{code/UpLock.hny} \autoref{fig:inc} fixed with a lock}
 \label{fig:incfixed}
@@
 \end{code}
-\caption{\harmonylink{modules/synchS.hny} The lock implementation in the <{synchS}> module uses suspension}
+\caption{\harmonylink{modules/synchS.hny} Lock implementation using suspension}
 \label{fig:suspension}
@@
 \end{code}
-\caption{\harmonylink{code/xy.hny} Incomplete code with desired invariant $x + y = 100$}
+\caption{\harmonylink{code/xy.hny} Incomplete code for \autoref{ex:xy} with desired invariant $x + y = 100$}
 \label{fig:xy}
@@
 \end{code}
-\caption{\harmonylink{code/consensus.hny} Distributed consensus code and DFA}
+\caption{\harmonylink{code/consensus.hny} Distributed consensus code and behavior DFA}
 \label{fig:consensus}
@@
 \end{center}
-\caption{The DFA for \autoref{fig:bosco}}
+\caption{The behavior DFA for \autoref{fig:bosco}}
 \label{fig:boscodfa}

### 2022-01-11 book/paper.tex
Latest

@@
 \end{code}
-\caption{\harmonylink{code/linkedlist.hny} Implementation of a set of integers using a linked list with fine-grained locking}
+\caption{\harmonylink{code/linkedlist.hny} Implementation of a set of values using a linked list with fine-grained locking}
 \label{fig:linkedlist}
@@
 \autoref{fig:linkedlist} implements a concurrent set object using
-an ordered linked list of integers without duplicates.
-The list has two dummy ``book-end'' nodes with values <{-inf}> and
-<{inf}> (similar to the Python \texttt{math.inf} constant).
+an ordered linked list without duplicates.
+The list has two dummy ``book-end'' nodes with values <{(-1, None)}> and
+<{(1, None)}>.  A value <{v}> is stored as <{(0, v)}>---note that for
+any value <{v}>, <{(-1, None) < (0, v) < (1, None)}>.
 An invariant of the algorithm is that at any point in time the
-list is ``valid,'' starting with a <{-inf}> node and ending
-with an <{inf}> node.
+list is ``valid,'' starting with a <{(-1, None)}> node and ending
+with an <{(1, None)}> node.
 
@@
 Each node has a lock, a value, and <{next}>, a pointer to the next node
-(which is <{None}> for the <{inf}> node).
+(which is <{None}> for the <{(1, None)}> node to mark the end of the list).
 The <{_find(lst, v)}> helper method first
 finds and locks two consecutive nodes <{before}> and <{after}>
-such that <{before->data.value < v <= after->data.value}>.
+such that <{before->data.value < (0, v) <= after->data.value}>.
 It does so by performing something called \emph{hand-over-hand locking}.
-It first locks the first node, which is the <{-inf}> node.
+It first locks the first node, which is the <{(-1, None)}> node.
 Then, iteratively, it obtains a lock on the next node and release the
@@
 If not found, it coincides with retrieving the pointer to the
-<{inf}> node.
+<{(1, None)}> node.
 \end{comment}
@@
 \hline
-Integer & <{-inf, ..., -2, -1, 0, 1, 2, ..., inf}> \\
+Integer & <{..., -2, -1, 0, 1, 2, ...}> \\
 \hline
@@
 
-The integer type supports any whole number, as well as $-\mathtt{inf}$
-and $\mathtt{inf}$.  In the Python-based model checker, integers are
-infinite precision.  In the C-based model checker, integers are implemented
-by two's complement 61-bit words, and $-\mathtt{inf}$ is represented by
-the minimal integer and $\mathtt{inf}$ is represented by the maximal integer.
+The integer type supports any whole number.
+In the C-based model checker, integers are currently implemented
+by two's complement 61-bit words.  The model checker checks for
+overflow on various operations.
 

### 2022-01-11 paper/paper.tex
Latest

\documentclass[twocolumn]{article}
\usepackage{fullpage}
\usepackage{graphicx}
\usepackage{framed}
\usepackage{tcolorbox}
\usepackage{verbatim}
\usepackage{fancyvrb}
\usepackage{comment}
\usepackage{amsmath}
\usepackage{geometry}
\usepackage{enumitem}
\usepackage{etoolbox}
% \usepackage{bold-extra}
\usepackage{xcolor}
\usepackage[colorlinks=true, linkcolor=blue, urlcolor=blue, citecolor=violet]{hyperref}

% \renewcommand{\topfraction}{.99}
% \renewcommand{\bottomfraction}{.99}
% \renewcommand{\textfraction}{.01}
% \renewcommand{\floatpagefraction}{.9}
% \renewcommand{\dbltopfraction}{.99}
% \renewcommand{\dblfloatpagefraction}{.9}

\widowpenalty 10000
\clubpenalty 10000
\tolerance=9999
\newcommand{\tm}{\raisebox{.9ex}{\tiny tm}}

\newcommand{\harmonysource}[1]{
\begin{tabbing}
XX\=XXX\=XXX\kill
    \input{sources/#1.tex}
\end{tabbing}
}

\newcommand{\harmonylink}[1]{%
[\href{https://harmony.cs.cornell.edu/#1}{\underline{#1}}]%
}

\newcommand{\harmonyref}[2]{%
\href{https://harmony.cs.cornell.edu/output/#1}{\underline{#2}}%
}

\newenvironment{code}{
\tcolorbox
}{
\endtcolorbox
}

\title{Concurrent Programming in Harmony}
\author{Robbert van Renesse \and William Ma \and Kevin Sun \and Anthony Yang}

\begin{document}
\maketitle

<{:const:tas}>
<{:const:cas}>
<{:const:BinSem}>
<{:const:Semaphore}>
<{:const:__init__}>
<{:const:p}>
<{:const:hello}>

\begin{figure*}
\begin{center}
\includegraphics[width=.9\textwidth]{hello.pdf}
\end{center}
\caption{A selection of Harmony programs and their outputs}
\label{fig:helloworld}
\end{figure*}

\section{Introduction}

\emph{Harmony} is a Python-like programming language designed to teach
concurrent and distributed programming to students who do not have a
background in formal methods.  Harmony programs are model checked.
Without any special annotation, Harmony can diagnose a large variety
of problems:
\begin{itemize}
\item ``Silly errors'': accessing non-existent variables,
divide-by-zero, out-of-bounds array access, trying to multiply two lists
with one another, and many more.
\item Safety violations due to assertions failing, such as an assertion to ensure mutual exclusion.
\item Liveness violations (deadlock).
\item Faulty behaviors: implementation has behaviors not seen with specification, such as violations of linearizability;
\item Incomplete behaviors: specification has behaviors not seen with implementation, such as multiple readers not able to get a reader/writer lock simultaneously;
\item Data races: variables accessed without a lock.
\item Busy Waiting: waiting for a condition without blocking.
\end{itemize}

Harmony programs have three sources of non-determinism:
<{choose}> operations that select a value from a set,
\emph{interleaving} due to multi-threading, and \emph{interrupts}.
\autoref{fig:helloworld} demonstrates the <{choose}> and the <{spawn}>
statements.  Since Harmony programs are model-checked and multiple
executions are explored across a finite set of states, the output of a
Harmony program is a finite automaton that describes all possible
outputs of the program.  Groups of statements can be made atomic
using the <{atomically}> keyword.  Finally, Harmony supports a thread
waiting until some condition is reached with the <{when}> statement.

Harmony programs compile to a virtual machine.  The Harmony virtual
machine is formally specified using TLA+, allowing Harmony programs
to run using TLC, the model checker that comes with the TLA+
toolbox.  Harmony has its own custom concurrent model checker that
scales significantly better for Harmony programs.

The Harmony language can be used both to \emph{specify} desired
behaviors and to \emph{implement} them.  To check that the implementation
satisfies the specification, the programmer writes a test program that
systematically checks available operations and prints the results.
The test program is first run against the specification, generating
a \emph{behavior automaton}.  The same test program is then run
against the implementation, along with the behavior automaton.
The Harmony model checker generates the product of the test program
and the behavior automaton to find discrepancies in the implementation
with respect to its specification.

Harmony has been used to verify the correctness of various lock
implementations, monitors and condition variables (both Hoare and Mesa),
various implementations of concurrent data structures including lock-free
ones, and barrier synchronization implementations.  Harmony has also been
used to verify a variety of distributed algorithms such as two-phase
commit, chain replication, various consensus algorithms such as Paxos,
and the Needham-Schroeder authentication protocols.

\begin{figure}
\begin{tabular}{|l|l|}
\hline
Boolean & <{False}>, <{True}> \\
\hline
Integer & <{..., -2, -1, 0, 1, 2, ...}> \\
\hline
String & <{"example"}>, <{.example}> \\
\hline
PC & (program counters) \\
\hline
Dictionary & <{{ .acct: 12345, .valid: False }}> \\
\hline
Set & <{{}, { 1, 2, 3 }, { False, .id, 3 }}> \\
\hline
Address & <{?lock, ?flags[2], None}> \\
\hline
Context & (generated by <{stop}> expression) \\
\hline
\end{tabular}
\caption{Harmony values}
\label{fig:values}
\end{figure}

\section{Harmony Virtual Machine}

The Harmony virtual machine is fully specified in approximately
1000 lines of TLA+.
Below we summarize its most important aspects.

\subsection{Harmony Values}

Harmony has 8 types of values listed in \autoref{fig:values}.
There are no exceptions: a set can contain dictionaries, a dictionary
can map dictionaries to dictionaries, and so on.  Unlike many
object-oriented programming languages, the programmer never
has to specify what it means for two values to be the same of what its
hash value is.

Any two values can be compared using the standard comparison operators.
If two values have different types, then their order is defined as in
\autoref{fig:values} (e.g., <{False < 0}>).
For two values of the same type, Harmony defines a deterministic order.
For example, for two sets, the values in the sets are first ordered and
then lexicographically compared, so that <{{ 1, 2 } < { 1, 3 }}>.
(Note that, unlike Python, the $<$ operator is a total order and
$<$ on sets does not represent the subset relationship.)
Dictionaries are lexicographically ordered by their (key, value) pairs.

The strings <{"example"}> and <{.example}> are the same---the latter
form is convenient for use in dictionaries.
An \emph{address} represents a path in a hierarchy of dictionaries, with
<{None}> being the empty path.
Examples of \emph{program counters} include method constants and program labels.
A \emph{context} represents the state of a thread.  It includes a program
counter, a map of local variable names to values, and a stack of values.
Harmony makes no difference between lists and tuples and implements
both as dictionaries from integers to values.
So <{[7, 4] = (7, 4) = { 0: 7, 1: 4 }}>.
Note that <{[7, 4][0] = 7}>, as expected.
The empty dictionary is represented as either <{()}> or <{[]}>
(<{{}}> is the empty set).

Except for <{choose}>, any operation on values is deterministic and
defined in the formal specification of the Harmony virtual machine.
For example, like Python, Harmony supports merging of dictionaries
using the union operator (<{|}>), but, unlike Python, Harmony defines
what happens when two dictionaries have the same key but different values.
In that case, Harmony will use the maximum value as defined above.
For dictionary  intersection, Harmony uses the minimum value instead.
Doing so has make dictionary union and intersection deterministic and
commutative.  Moreover, Harmony represents multisets as dictionaries
of values to multiplicities.  Using Harmony's rules on dictionary union
and intersection, union and intersection on multisets are correctly
defined.

\cite{HW90}

\bibliographystyle{alpha}
\bibliography{paper}

\end{document}

### 2022-01-11 paper/paper.tex
Latest version of paper

@@
 \usepackage{etoolbox}
+\usepackage{paralist}
 % \usepackage{bold-extra}
@@
 of problems:
-\begin{itemize}
+\begin{compactitem}
 \item ``Silly errors'': accessing non-existent variables,
@@
 \item Liveness violations (deadlock).
-\item Faulty behaviors: implementation has behaviors not seen with specification, such as violations of linearizability;
-\item Incomplete behaviors: specification has behaviors not seen with implementation, such as multiple readers not able to get a reader/writer lock simultaneously;
+\item Faulty behaviors: implementation has behaviors not seen with specification, such as violations of linearizability~\cite{HW90}.
+\item Incomplete behaviors: specification has behaviors not seen with implementation, such as multiple readers not able to get a reader/writer lock simultaneously.
 \item Data races: variables accessed without a lock.
 \item Busy Waiting: waiting for a condition without blocking.
-\end{itemize}
+\end{compactitem}
 
@@
 
-The Harmony virtual machine is fully specified in approximately
+The Harmony Virtual Machine (HVM) is fully specified in approximately
 1000 lines of TLA+.
@@
 
-\subsection{Harmony Values}
+\subsection{Values}
 
@@
 
-\cite{HW90}
+\subsection{State}
+
+The state of the HVM consists of three variables:
+\begin{compactitem}
+\item a dictionary of shared variable names to values;
+\item a bag (multiset) of contexts, one for each thread;
+\item a set of \emph{active} contexts.
+\end{compactitem}
+
+Harmony uses a multibag to keep track of the states of the threads
+because threads can be in the same state.
+Threads do not have unique identifiers, reducing state explosion in
+the model checker.
+
+A context is active if it can make a transition.
+Contexts can be atomic or not---the HVM prevents multiple atomic
+contexts from being atomic simultaneously.  Therefore, the active set
+either contains a single atomic context or it contains all contexts
+in the domain of the context bag.
+
+The initial state consists of an empty dictionary of shared variables and
+a single atomic context with program
+counter 0 in both the context bag and the active set.
+
+Harmony also allows the user to specify a determistic finite automaton
+(DFA) to check the legality of the output of the program to be
+checked.  In that case, the state of the automaton is also made
+part of the HVM state, so that the HVM maintains the product of the
+program state and the DFA state.
+
+\subsection{Transitions}
+
+The HVM makes a transition by executing an HVM instruction using
+one of the contexts in the active set.  There are about 30 basic
+transitions, which, for this paper, we classify as follows:
+
+\begin{compactitem}
+\item \emph{basic operations}: do a simple operation to a context
+such as pushing a value onto its stack, and increase its program
+counter by one.
+\item \emph{jumps}: to implement loops.
+\item \texttt{Choose}: nondetermistic selection from a set.
+\item \texttt{Apply}: to call a method \emph{or} to index into a dictionary.
+\item \texttt{Frame/Return}: method start/stop.
+\item \texttt{Spawn}: create a new thread.
+\item \texttt{AtomicInc/AtomicDec}: modify the atomic counter of the context.
+\item \texttt{SetIntLevel}: enable/disable interrupts.
+\item \texttt{Stop/Go}: manipulating continuations.
+\item \texttt{Assert}: check a value.
+\item \texttt{Print}: add a value to the print log.
+\end{compactitem}
+
+Each transition modifies at least the context bag and set of active
+contexts.
+A few operations (\texttt{Store}, \texttt{Stop}, and \texttt{Del})
+also modify one of the shared variables.
+
+\paragraph{Choose.} \texttt{Choose} pops a set value of the stack
+and pushes a selected element back onto the stack.  The model checker
+tries each selection.
+
+\paragraph{Apply.}  \texttt{Apply} pops two values $f$ and $x$ of the stack,
+\emph{applies} $x$ to $f$, and pushes the resulting value back onto the
+stack.  What \emph{applies} means depends on the type of $f$.  If $f$ is
+a PC value, then it means invoking method $f$: it pushes the return address
+and $x$ onto the stack and set the program counter to $f$.  If $f$ is a
+dictionary, then it means pushing the value corresponding to key $x$ onto
+the stack.  This way, a dictionary acts much like a method that maps keys
+to values.  If $f$ is a string, then the operation returns the $x^{th}$
+character of the string.
+
+\paragraph{Frame/Return.}  \texttt{Frame} is the first operation of
+a method and \texttt{Return} its last.  \texttt{Frame} pops a value of
+the stack and assigns it to the argument (or argument list) of the
+method.  It also saves the local variables onto the stack and set the
+value of the <{result}> variable to <{None}>.  The operation of
+\texttt{Return} depends on whether the stack is empty of not.  If empty
+\texttt{Return} terminates the thread by removing the context from the
+context bag replacing the active set with the domain of the context bag.
+If the stack is not empty, then the saved local variables are popped
+and restored, the return address is popped and assigned to the
+program counter, and the value of <{result}> is pushed onto the
+stack.
+
+\paragraph{Spawn.}  Spawn pops a program counter and argument from the
+stack and uses them to create a new context.  The new context is added
+to the context bag.  Also, if the current context is not atomic, then
+the new context is also added to the active set.
+
+\paragraph{AtomicInc/AtomicDec.}  A context contains a counter that,
+if non-zero, puts the context in atomic mode.
+<{atomically}> statements can be nested: at the beginning, the
+counter is incremented; at the end, it is decremented.
+When incremented from 0 to 1, the operation replaces the active
+set of context with just the current context.  When decremented
+from 1 to 0, the active set is replaced with the union of the
+new context and the domain of the context bag.
+
+\paragraph{SetIntLevel.}  A context contains a boolean that determines
+if interrupts are enabled or not.  \texttt{SetIntLevel} pops the
+new interrupt level and pushes the old.
+
+\paragraph{Stop/Go.}  \texttt{Stop} saves the current context (with
+its program counter incremented) in a specified shared location (usually the
+tail of a queue) and removes the context from the context bag and active set.
+If the context was atomic (which is usually the case), the active set is
+replaced with the domain of the context bag.  Even though its atomic counter
+is not zero, the stopped context is not considered atomic at this time.
+\texttt{Go} pops a context value and a return value of the stack and
+pushes the return value onto the stack of the popped context, and adds
+the context to the context bag.  If the current context is not atomic,
+then the restored context is also added to the active set.
+
+\paragraph{Assert.}  \texttt{Assert} pops a boolean value and increments
+the program counter.  If the value is <{False}>, the HVM aborts.
+
+\paragraph{Print.}  \texttt{Print} pops a value of the stack and increments
+the program counter.  In case a behavior automaton has been specified, the
+value is applied to the DFA state and checked for validity.
+To reduce state explosion, the HVM does not keep track of the value itself
+in the state (like keeping the value in a shared list as part of the state).
+
+\subsection{Model Checker}
+
+\emph{Charm}, the Harmony model checker is an explicit state model
+checker that computes a \emph{Kripke structure} of states reachable
+from the initial state in a breadth-first manner, stopping only if
+it encounters an error (such as divide-by-zero or a failing assertion)
+or until there are no more new states.
+
+There are three sources of nondetermism in Harmony: \texttt{Choose}
+operations, thread interleaving, and interrupts.
+When it comes to interleaving, an important observation is that
+local operations on the state of a context (updates to its program
+counter, local variables, or stack) are not visible to other threads.
+An important optimization that significantly
+reduces the size of the set of generated states is to treat a
+sequence of operations as atomic if only its first operation is an
+operation that accesses the shared state (\texttt{Load}, \texttt{Store},
+\texttt{Del}, or \texttt{Stop}) or a \texttt{Choose} or \texttt{Print}
+operation.  We call such operations \emph{breakable}.
+Note that this is essentially a form of Partial Order Reduction.
+Reducing the state space this way is highly effective for Harmony.
+For example, when model checking Peterson's algorithm expressed in
+Harmony, the number of states is reduced from 7927 to 104 states.
+In the case of the Knuth-Yao algorithm (which is not concurrent but
+involves choices), the number of states is reduced from 258 states
+to 14.
+
+Given a state and a context, Charm first applies the first instruction
+to the state.  Then it continues to apply instructions as long as
+the active set is the singleton set (such as when the thread is in
+atomic mode) or, if not, as long as the next instruction is not
+breakable.  Only the resulting state is added to the Kripke structure,
+and the edge indicates the starting context.  During this loop, Charm
+does check if the same state recurs---if so it aborts
+and reports an infinite loop.
+
+In the case the first instruction is a \texttt{Choose} instruction,
+Charm will try all possible choices for the set value
+on top of the stack.  If not, Charm will also check to
+see if a trap is outstanding and interrupts are enabled.  If so,
+Charm will try an interrupt transition as well.
+
+\subsection{Data Race Detection}
+
+A data race occurs when two or more threads simultaneously try to
+access the same shared data location, at least one of which is a
+\texttt{Store}, \texttt{Stop}, or \texttt{Del} instruction.  Harmony
+only considers this a problem it at least one of those threads is
+not in atomic mode.  After all, we do not want to consider accesses
+to a spinlock as a problem.  Also, the Harmony language allows
+shared variables to be tagged as \emph{sequentially consistent},
+effectively turning off data race detection.  Charm maintains on
+the edges of the Kripke a list of accesses that were made during
+the transition.  After all transitions from a state have been
+computed, Charm checks for data races.
+
+\section{Analysis}
+
+After running the Charm model checker, there are two cases to
+consider.  If Charm aborted due to some problematic transition,
+then Charm reports the shortest path as a so-called \emph{counter-example}.
+If not, the resulting graph is analyzed for various other problems,
+including deadlock, busy waiting, and executions
+that ended in a state that is not considered final according to
+behavior DFA if specified.  We detail below how these problems are
+found.
+
+\paragraph{Deadlock and Faulty Behavior.}
+
+Harmony transforms the Kripke structure into a DAG of its strongly
+connected components.  The states in the sink components of this
+DAG can be considered the terminal states of the Harmony program.
+States in which the context bag is empty represent states in which
+all threads have terminated.
+In the case the context bag is not empty, we check if there are any
+non-\emph{eternal} contexts.  If so, Harmony reports a \emph{deadlock}.
+Harmony programmers can specify that certain threads are eternal,
+such as server threads that operate in an infinite loop waiting for
+requests.
+If there is no deadlock and a behavior automaton was specified,
+Harmony also checks Harmony checks if the automaton is in a
+terminal state.  If not, a \emph{behavior violation} is reported.
+
+\paragraph{Active Busy Waiting.}
+
+A thread is considered \emph{blocked} if it can only make progress
+(modify the state) if another thread makes a transition first.
+This is not always a problem: a thread waiting on a lock in a spinloop
+is a common way of synchronizing threads, and Harmony does not report
+such issues.
+
+A thread is considered \emph{actively busy waiting} if it cannot
+escape it strongly connected component without another thread
+updating the shared state first.  A common example of this is a
+synchronization algorithm that checks for a condition in a loop,
+acquiring and releasing the lock each time.  Harmony checks each
+strongly connected component for the existence of such situations.
+
+\paragraph{Incomplete Behavior.}
+
+The Kripke structure can be considered as a non-deterministic
+automaton, with on each edge zero or more \emph{symbols} (values
+that were printed.  If a behavior automaton was specified,
+then there are now two automata: the specified automaton
+${\mathcal Spec}$ that was presented to Charm as an input as
+part of the \emph{specification}, and the generated automaton
+${\mathcal Impl}$ that was generated by the given Harmony
+\emph{implementation}.  Harmony already checks on-the-fly whether
+${\mathcal Impl} \subseteq {\mathcal Spec}$, that is, that every
+behavior (sequence of symbols) of the implementation is also a
+behavior of the specification.  Harmony also checks if
+${\mathcal Spec} \subset {\mathcal Impl}$ and, if so,
+produces a warning.
+Although not current done, it is possible to compute
+${\mathcal Spec}~\backslash~{\mathcal Impl}$
+to generate an automaton for all behaviors allowed by the
+specification that are not in the implementation.
+
+Having such behaviors is not always a problem, but it can be.  For
+example, a reader/writer lock implemented as a plain lock meets the
+specification of a reader/writer lock, but it does not allow multiple
+readers in the critical section.  Unfortunately, checking ${\mathcal
+Spec} \subset {\mathcal Impl}$ does not scale well and can only be
+done for relatively small Kripke structures.
 

### 2022-01-11 paper/paper.tex
Latest

@@
 
-\title{Concurrent Programming in Harmony}
+\title{Concurrent and Distributed Programming in Harmony}
 \author{Robbert van Renesse \and William Ma \and Kevin Sun \and Anthony Yang}
@@
 <{:const:hello}>
+<{:const:enqueue}>
+<{:const:dequeue}>
 
@@
 \end{compactitem}
+
+\begin{figure*}
+\begin{center}
+\includegraphics[width=.9\textwidth]{MS.pdf}
+\end{center}
+\caption{A reproduction of the 1996 Michael\&Scott two-lock concurrent queue
+implementation and its equivalent in Harmony (within the box).}
+\label{fig:ms}
+\end{figure*}
 
@@
 \end{figure}
+
+\begin{figure*}[h]
+\begin{center}
+\begin{tabular}{cc}
+{\footnotesize
+\begin{tcolorbox}[width=0.45\linewidth]
+<{:const:head}>
+<{:const:tail}>
+<{:const:append}>
+<{:const:test_and_set}>
+<{:inline:}>
+def Lock():
+    result = False
+
+def acquire(lk):
+    atomically when not !lk:
+        !lk = True
+
+def release(lk):
+    atomically !lk = False
+<{:end:}>
+\end{tcolorbox}
+}
+&
+{\footnotesize
+\begin{tcolorbox}[width=0.45\linewidth]
+<{:inline:}>
+def test_and_set(s):
+    atomically:
+        result = !s
+        !s = True
+
+def acquire(lk):
+    while test_and_set(lk):
+        pass
+
+<{:end:}>
+\end{tcolorbox}
+}
+\\
+(a) & (b)
+\end{tabular}
+\end{center}
+\caption{Lock specification and a spinlock implementation based on
+test-and-set.}
+\label{fig:synch}
+\end{figure*}
+
+\begin{figure*}[h]
+\begin{center}
+\begin{tabular}{cc}
+{\footnotesize
+\begin{tcolorbox}[width=0.45\linewidth]
+<{:inline:}>
+import list
+
+def Queue():
+    result = []
+
+def enqueue(q, v):
+    atomically !q = list.append(!q, v)
+
+def dequeue(q):
+    atomically:
+        if !q == []:
+            result = None
+        else:
+            result = list.head(!q)
+            !q = list.tail(!q)
+
+<{:end:}>
+\end{tcolorbox}
+}
+&
+{\footnotesize
+\begin{tcolorbox}[width=0.48\linewidth]
+<{:inline:}>
+import queue
+
+const NOPS = 4
+q = queue.Queue()
+
+def thread(self):
+    let op = choose({ "deq", "enq" }):
+        if op == "enq":
+            print("call enqueue", self)
+            queue.enqueue(?q, self)
+            print("done enqueue", self)
+        else:
+            print("call dequeue", self)
+            let v = queue.dequeue(?q):
+                print("done dequeue", self, v)
+
+for i in {1..NOPS}: spawn thread(i)
+<{:end:}>
+\end{tcolorbox}
+}
+\\
+(a) concurrent <{queue}> module & (b) test program
+\end{tabular}
+\end{center}
+\caption{The (a) specification and (b) test program for a concurrent queue.}
+\label{fig:queue}
+\end{figure*}
+
+\section{Harmony Language}
+
+\subsection{Basics}
+
+Harmony is heavily based on Python to reduce the learning curve
+for users.  Many of Python's powerful constructs and operators are
+supported are supported, including comprehensions.  Harmony does
+not support classes, and instead supports memory allocation and
+pointers in a way that is familiar to C programmers.  While this
+seems like a step back in language evolution, many synchronization
+algorithms in the literature explicitly consider memory accesses,
+and perhaps surprisingly this has been a bonus for analyzing
+such algorithms with Harmony.  Moreover, the Harmony model checker 
+can be used to check for memory leaks or access to dangling pointers.
+
+For example, \autoref{fig:ms} shows a reproduction from the well-known
+Michael\&Scott two-lock concurrent queue~\cite{MS96} along with a
+Harmony version of the same algorithm.
+Harmony uses the `<{?}>' instead of the `<{&}>' symbol to take the address
+of a memory location, but inherits the same <{q->field}> notation
+from C to access the given field in the dictionary pointed to by
+pointer <{q}>.  <{q->field}> is shorthand for <{(!q)["field"]}> in
+Harmony, where `<{!}>' is the Harmony operator to dereference a pointer.
+Harmony also complaints, correctly, about a data race when
+there is a concurrent \texttt{enqueue} and \texttt{dequeue} operation
+on an empty queue---the implementer should ensure that accesses to the
+\texttt{next} field are sequentially consistent.
+
+The Harmony example in \autoref{fig:ms} shows some other language
+features in Harmony that are not in Python.
+The <{let x = e: S}> statement introduces a new bound variable <{x}>
+with value <{e}> that is only valid within the statement block <{S}>.
+Also, Harmony does not have \texttt{return}, \texttt{break}, or
+\texttt{continue} statements to force the programmer into a purely
+nested control flow.  Each method has a variable <{result}> in which
+the return value of the method is stored.
+
+\subsection{\texttt{atomically}}
+
+Statements in Harmony, including groups of statements can be tagged with
+the \texttt{atomically} keyword to make them atomic.  What this means
+is that the statement is evaluated atomically by the Harmony virtual
+machine without interleaving of instructions of other threads.  It is
+a \emph{stronger} condition than holding a lock---a lock still allows
+concurrent execution of multiple threads, but it prevents multiple
+threads from executing in the same critical section.
+The <{atomically}> keyword is intended for <{specification}> more than
+for <{implementation}> of concurrent algorithms.
+
+For example, \autoref{fig:synch}(a) shows the specification for locks
+from the Harmony <{synch}> module that was used in \autoref{fig:ms},
+featuring the Harmony <{when}> statement.
+\autoref{fig:synch}(b) shows an implementation of the <{acquire}>
+method based on a test-and-set primitive, which in turn is
+\emph{specified}.
+Inprecisely, methods that use the <{atomically}> keyword are specifications
+while methods that do not are probably implementation.
+Whether something is a specification or an implementation is in the
+eye of the beholder.
+Importantly, however, is the notion of modularity that is implied here.
+Harmony can model check the implementation of \autoref{fig:ms} using
+the specification of a lock in \autoref{fig:synch}(a).
+Separately, Harmony can check the correctness of implementation of
+\autoref{fig:synch}(b) against the specification of
+\autoref{fig:synch}(a).
+
+The Harmony distribution comes with a collection of specifications,
+each often with multiple implementations.  The lock specification
+comes with the spinlock implementation of \autoref{fig:synch}(a), but
+also a ticket lock implementation and one based on continuations and
+a ready queue.  There are various concurrent queue implementations,
+including the lock-free implementation of~\cite{MS96}.
+
+\subsection{Checking an implementation}
+
+The Harmony language allows objects such as locks and concurrent queues
+to be specified and implemented in separate programs.  But the programs
+are not \emph{active} as they are just a collection of method definitions.
+A model checker can only check behaviors, but these programs have empty
+behaviors.
+Suppose we want to check the correctness of the concurrent queue
+implementation of \autoref{fig:ms}.  First, we would need a specification
+of a concurrent queue.  Second, we would need a test program that tries
+various concurrent interleavings of <{enqueue}> and <{dequeue}> operations.
+
+\autoref{fig:queue}(a) shows the specification of a <{queue}> module.
+A queue is represented as a list.
+The <{enqueue}> method add a value onto the end of the list,
+while the <{dequeue}> method extracts the head, both as atomic actions.
+
+To test a queue, we want to perform a set of concurrent <{enqueue}>
+and <{dequeue}> operations concurrently.
+\autoref{fig:queue}(b) shows how this can be specified in Harmony.
+The test program spawns 4 threads that each select an operation
+using the nondetermistic <{choose}> statement.  In case <{enqueue}>
+is selected, the thread enqueues its identifier <{self}>.  The program
+includes a print statement at the beginning and end of every queue
+operation.
+
+The test program is similar to what a programmer might have written
+as a unit test.  The programmer would run the program a bunch of
+times and look at the output in order to see if it looks reasonable.
+Doing so is, however, difficult and error prone, as the output tends
+to be complex and in the mess of many lines of output it is easy
+to overlook problems such as two threads dequeuing the same item
+or items being dequeued in the wrong order.  The output of concurrent
+print statement can be mangled together.  Not only that, but the
+programmer would need to run this test program many times and check
+each the output each time.  It even is hard to write a program that
+checks the output.
+
+This process is automated in Harmony.  The approach is to
+run the model checker twice.  The first time, the test program is
+run against the specification of the concurrent queue.  Harmony
+outputs a DFA such as shown in \autoref{fig:hello}.  The second
+time, the same test program is run against the implementation of the
+concurrent queue, but with the inclusion of the DFA.  The model
+checker ends up verifying the product of the implementation
+and the DFA and ensures that every behavior of the implementation
+program (in terms of its print statements) is also a behavior of
+the specification program.
 
@@
 then the restored context is also added to the active set.
+A stopped context can be re-instantiated multiple times, allowing
+programmers to implement thread forking.
 
@@
 \emph{Charm}, the Harmony model checker is an explicit state model
-checker that computes a \emph{Kripke structure} of states reachable
+checker that computes a \emph{Kripke structure} (essentially, a graph)
+of states reachable
 from the initial state in a breadth-first manner, stopping only if
@@
 Charm will try an interrupt transition as well.
+
+Part of the model checking is trivially parallelizable, allowing for
+significant speedups on multi-core machines.
 
@@
 
+\begin{figure}
+\begin{center}
+{\footnotesize
+\begin{code}
+<{:inline:}>
+from synch import Lock, acquire, release
+
+const N = 5
+forks = [Lock(),] * N
+
+def diner(which):
+    let left, right = (which, (which + 1) % N):
+        while choose({ False, True }):
+            acquire(?forks[left])
+            acquire(?forks[right])
+            # dine
+            release(?forks[left])
+            release(?forks[right])
+
+for i in {0..N-1}: spawn diner(i)
+<{:end:}>
+\end{code}
+}
+\end{center}
+\caption{The Dining Philosophers Program in Harmony}
+\label{fig:diners}
+\end{figure}
+
+\begin{figure*}
+\begin{center}
+\includegraphics[width=\textwidth]{diners.png}
+\end{center}
+\caption{The output of running the Dining Philosophers program}
+\label{fig:dinersHTML}
+\end{figure*}
+
 \paragraph{Incomplete Behavior.}
@@
 
+\paragraph{Constructing a Counter-Example.}
+
+When Charm finds a problematic state or transition, it outputs
+a detailed trace that is called a \emph{counter-example}.
+It always outputs the shortest such trace.
+However, in order to save memory, Charm does not actually keep
+the required information during the construction of the Kripke
+structure.  In particular, its edges only contain the context
+and the choice if any.  Thus, in order to reconstruct the
+counter-example, Charm runs the HVM one more time, this time
+re-evaluating the edges along the path and outputting the
+detailed information about each transition.
+
+\section{Debugging}
+
+If Charm determines a problem, it outputs the shortest counter-example that
+it could find.  However, this counter-example is at the level of the HVM,
+not the Harmony program that the programmer wrote.  It can be hard to
+dig through a list of HVM instructions.
+Harmony includes tools to quickly focus on the root cause of the
+problem.  In particular, in case of a problem, Harmony outputs a compact
+but detailed and interactive overview of the problem in HTML format that
+can be accessed from any web browser.
+
+For example, \autoref{fig:diners} shows the implementation
+in Harmony of the famous Dining Philosophers problem, where 5 philosphers
+share 5 forks (represented by 5 locks).
+The philosophers end up being deadlocked
+if they all pick up their left fork before any philosopher picks up
+their right fork.
+Harmony quickly finds the problem and produces the output
+shown in \autoref{fig:dinersHTML}.
+
+The little orange boxes represent individual HVM instructions of the
+problematic trace.  You can click on any such box to see the state
+at that time, or you can navigate in various ways (forwards one step,
+backwards one step, jump to the next line of Harmony code, or jumps
+forwards or backwards to a context switch between threads).  The actual
+HVM code is in the lowerleft of the output.
+
+However, users first want to focus on the rows in the top table.
+In this particular example
+
 \bibliographystyle{alpha}

### 2022-01-11 paper/paper.tex
Latest version of paper

@@
 <{:const:dequeue}>
+<{:const:send}>
+<{:const:receive}>
 
@@
 and the Needham-Schroeder authentication protocols.
+
+This paper provides an overview of the Harmony language and Charm,
+its model checker, and how they may be used by programmers to
+develop reliable concurrent or distributed code.
 
@@
 the specification program.
+
+\subsection{Distributed Programming}
+
+Harmony can be used to model a distributed system in which the
+only shared variable is a set of messages representing the network.
+A message is sent by atomically adding a message to the network:
+<{def send(m): atomically network |= {m}}>.
+To receive a message, a program can use the Harmony <{when}>
+statement to wait for a message.  Typically, however, the program
+does not want to receive just any message, but a message that satisfies
+certain conditions such as: wait for the next message that is addressed
+to me and has a sequence number larger than 3.
+For this, Harmony supports the <{exists clause}> and allows
+<{when}> clauses to be nested:
+<{atomically when exists m in network when (m.dst == self) and (m.seq > 3): ...}>
+With these constructs, Harmony can be effectively used as a
+model checker for distributed algorithms.
 
@@
 \end{center}
-\caption{The Dining Philosophers Program in Harmony}
+\caption{Dining Philosophers in Harmony}
 \label{fig:diners}
@@
 \end{center}
-\caption{The output of running the Dining Philosophers program}
+\caption{The HTML output of running the Dining Philosophers program}
 \label{fig:dinersHTML}
@@
 However, users first want to focus on the rows in the top table.
-In this particular example
+In this particular example, there are 6 threads, the thread that
+first runs and the five threads that are spawned by it.  They take
+``turns'' by context switching.  In this particular trace, <{diner(2)}>
+runs first and acquires <{forks[2]}>, its left fork.  In the table
+in the lower-right, you can see that the thread is currently blocked
+and in atomic as well as read-only mode (conditions in a <{when}>
+clause are not allowed to modify state).  You can also see its stack
+trace: it trying to acquire <{forks[3]}>, its right fork (executing
+the code in \autoref{fig:synch}(a)).  Next <{diner(0)}>, <{diner(3)}>,
+<{diner(4)}>, and <{diner(1)}> take similar turns.  All threads are
+blocked waiting for another thread to release a lock---a classic
+deadlock situation.
+
+Should the programmer need more detailed information, then they can
+run through the counter-example step-by-step, forwards and backwards,
+as explained above.
+
+\section{Limitations}
+
+Combinatorial state explosion.
+Production code.
+Lack of formal definition of Harmony itself.
 

### 2022-01-11 paper/paper.tex
Latest

@@
 eye of the beholder.
-Importantly, however, is the notion of modularity that is implied here.
+Important, however, is the notion of modularity that is implied here.
 Harmony can model check the implementation of \autoref{fig:ms} using
@@
 \autoref{fig:synch}(b) against the specification of
-\autoref{fig:synch}(a).
+\autoref{fig:synch}(a).  This modularity allows larger systems to
+be model checked that would not be tractable otherwise.
 
@@
 if interrupts are enabled or not.  \texttt{SetIntLevel} pops the
-new interrupt level and pushes the old.
+new interrupt level and pushes the old.  Supporting interrupts is
+important.  First, writing multi-threaded programs that also need
+to handle interrupts is difficult.  Second, distributed algorithms
+often require timer interrupts to trigger actions such as retransmissions.
 
@@
 clause are not allowed to modify state).  You can also see its stack
-trace: it trying to acquire <{forks[3]}>, its right fork (executing
+trace: it is trying to acquire <{forks[3]}>, its right fork (executing
 the code in \autoref{fig:synch}(a)).  Next <{diner(0)}>, <{diner(3)}>,
@@
 
-Combinatorial state explosion.
-Production code.
-Lack of formal definition of Harmony itself.
+While Harmony has proven itself to be effective and even fun to use
+for a large variety of synchronization problems and distributed
+algorithms, it is not a panacea.  At the core of Harmony is a model
+checker and with all its optimizations it can still easily suffer
+from combinatorial state explosion in programs that have too many
+threads or too many non-deterministic choices.  Nonetheless, because
+of its systematic state exploration, many problems can often be
+uncovered in small scenarios with few threads and few choices.
+
+The Harmony language, at least currently, is not usable for production
+code.  One reason for this is the <{atomically}> clause.  On a multi-core
+processor it would require an instruction to stop all other cores.
+Harmony was never designed to be a language for production code.  Because
+it is close to languages such as Python and C, it tends to be fairly
+easy to hand-translate code developed in Harmony to a desired target
+programming language.  However, such hand-translation could easily
+introduce a bug that did not exist in the Harmony code.
+
+Finally, the Harmony language currently does not have a formal definition
+itself, and there is no guarantee that the compiler itself is correct.
+Harmony compiles to a virtual machine code that is fully formally
+specified in TLA+, and this is the level of abstraction where model
+checking occurs.
 

### 2022-01-11 paper/paper.tex
Latest

@@
 
-Harmony is heavily based on Python to reduce the learning curve
-for users.  Many of Python's powerful constructs and operators are
+Harmony is heavily based on Python to reduce the learning curve for
+users.  Many of Python's powerful constructs and operators are
 supported are supported, including comprehensions.  Harmony does
@@
 pointers in a way that is familiar to C programmers.  While this
-seems like a step back in language evolution, many synchronization
-algorithms in the literature explicitly consider memory accesses,
-and perhaps surprisingly this has been a bonus for analyzing
-such algorithms with Harmony.  Moreover, the Harmony model checker 
-can be used to check for memory leaks or access to dangling pointers.
+perhaps seems like a step back in language evolution, many
+synchronization algorithms in the literature explicitly consider
+memory accesses, and perhaps surprisingly this has been a bonus for
+analyzing such algorithms with Harmony.  Moreover, the Harmony model
+checker can be used to check for memory leaks or access to dangling
+pointers.
 
@@
 algorithms, it is not a panacea.  At the core of Harmony is a model
-checker and with all its optimizations it can still easily suffer
+checker and---even with all its optimizations---it can still suffer
 from combinatorial state explosion in programs that have too many
@@
 uncovered in small scenarios with few threads and few choices.
+Harmony's support for anonymous threads is useful for scaling when
+there are many similar threads that do not need to know their own
+identifier.  Also, Harmony's support for modular model checking
+allows larger systems to be built from smaller ones.
 
@@
 
+\section{Related Work}
+
+Harmony does not offer much innovation on the front of model checking
+
+PlusCal and TLA+, Promela and Spin, DSLABS (UW), Java Pathfinder, Zing,
+CBMC/JBMC.
+
 \bibliographystyle{alpha}

### 2022-01-12 paper/paper.tex
Latest

@@
 \author{Robbert van Renesse \and William Ma \and Kevin Sun \and Anthony Yang}
+\date{Department of Computer Science, Cornell University}
 
@@
 background in formal methods.  Harmony programs are model checked.
-Without any special annotation, Harmony can diagnose a large variety
+Without any special formalism, Harmony can diagnose a large variety
 of problems:
@@
 \caption{A reproduction of the 1996 Michael\&Scott two-lock concurrent queue
-implementation and its equivalent in Harmony (within the box).}
+implementation and its equivalent in Harmony (within the box)}
 \label{fig:ms}
@@
 outputs of the program.  Groups of statements can be made atomic
-using the <{atomically}> keyword.  Finally, Harmony supports a thread
+using the <{atomically}> keyword.  Finally, Harmony supports
 waiting until some condition is reached with the <{when}> statement.
@@
 against the implementation, along with the behavior automaton.
-The Harmony model checker generates the product of the test program
+The Harmony model checker generates the Cartesian product of the test program
 and the behavior automaton to find discrepancies in the implementation
@@
 its model checker, and how they may be used by programmers to
-develop reliable concurrent or distributed code.
+develop code for reliable concurrent or distributed applications.
 
@@
 \caption{Lock specification and a spinlock implementation based on
-test-and-set.}
+test-and-set}
 \label{fig:synch}
@@
 \end{center}
-\caption{The (a) specification and (b) test program for a concurrent queue.}
+\caption{The (a) specification and (b) test program for a concurrent queue}
 \label{fig:queue}
@@
 users.  Many of Python's powerful constructs and operators are
-supported are supported, including comprehensions.  Harmony does
-not support classes, and instead supports memory allocation and
+supported, including comprehensions.  Harmony does
+not support classes and instead supports memory allocation and
 pointers in a way that is familiar to C programmers.  While this
@@
 
-Statements in Harmony, including groups of statements can be tagged with
+Statements in Harmony, including groups of statements, can be tagged with
 the \texttt{atomically} keyword to make them atomic.  What this means
@@
 \emph{specified}.
-Inprecisely, methods that use the <{atomically}> keyword are specifications
-while methods that do not are probably implementation.
+Imprecisely, methods that use the <{atomically}> keyword are specifications.
 Whether something is a specification or an implementation is in the
@@
 the specification of a lock in \autoref{fig:synch}(a).
-Separately, Harmony can check the correctness of implementation of
-\autoref{fig:synch}(b) against the specification of
+Separately, Harmony can check the correctness of the implementation in
+\autoref{fig:synch}(b) against the specification in
 \autoref{fig:synch}(a).  This modularity allows larger systems to
-be model checked that would not be tractable otherwise.
+be model checked in a way that would not be tractable otherwise.
 
@@
 
-\autoref{fig:queue}(a) shows the specification of a <{queue}> module.
-A queue is represented as a list.
+\autoref{fig:queue}(a) shows the specification of a concurrent
+<{queue}> module.  A queue is represented as a list.
 The <{enqueue}> method add a value onto the end of the list,
@@
 each the output each time.  It even is hard to write a program that
-checks the output.
+checks the output automatically.
 
@@
 run against the specification of the concurrent queue.  Harmony
-outputs a DFA such as shown in \autoref{fig:hello}.  The second
+outputs a DFA such as shown in \autoref{fig:helloworld}.  The second
 time, the same test program is run against the implementation of the
 concurrent queue, but with the inclusion of the DFA.  The model
-checker ends up verifying the product of the implementation
+checker ends up verifying the Cartesian product of the implementation
 and the DFA and ensures that every behavior of the implementation
@@
 A message is sent by atomically adding a message to the network:
-<{def send(m): atomically network |= {m}}>.
+<{def send(m): atomically network |= {m}}> (`<{|}>' is the union
+operator in Python).
 To receive a message, a program can use the Harmony <{when}>
@@
 object-oriented programming languages, the programmer never
-has to specify what it means for two values to be the same of what its
-hash value is.
+has to specify what it means for two values to be the same or what the
+hash of a value is.
 
 Any two values can be compared using the standard comparison operators.
-If two values have different types, then their order is defined as in
-\autoref{fig:values} (e.g., <{False < 0}>).
-For two values of the same type, Harmony defines a deterministic order.
+If two values have different types, then their order is defined as by
+the order in \autoref{fig:values} (e.g., <{False < 0}>).
+For any two values of the same type, Harmony defines a deterministic order.
 For example, for two sets, the values in the sets are first ordered and
@@
 For example, like Python, Harmony supports merging of dictionaries
-using the union operator (<{|}>), but, unlike Python, Harmony defines
+using the union operator, but, unlike Python, Harmony defines
 what happens when two dictionaries have the same key but different values.
@@
 For dictionary  intersection, Harmony uses the minimum value instead.
-Doing so has make dictionary union and intersection deterministic and
+Doing so makes dictionary union and intersection deterministic and
 commutative.  Moreover, Harmony represents multisets as dictionaries
@@
 because threads can be in the same state.
-Threads do not have unique identifiers, reducing state explosion in
-the model checker.
+Threads do not have unique identifiers, reducing combinatorial
+state explosion in the model checker.
 
@@
 a PC value, then it means invoking method $f$: it pushes the return address
-and $x$ onto the stack and set the program counter to $f$.  If $f$ is a
+and $x$ onto the stack and sets the program counter to $f$.  If $f$ is a
 dictionary, then it means pushing the value corresponding to key $x$ onto
@@
 the stack and assigns it to the argument (or argument list) of the
-method.  It also saves the local variables onto the stack and set the
+method.  It also saves the local variables onto the stack and sets the
 value of the <{result}> variable to <{None}>.  The operation of
-\texttt{Return} depends on whether the stack is empty of not.  If empty
+\texttt{Return} depends on whether the stack is empty of not.  If empty,
 \texttt{Return} terminates the thread by removing the context from the
-context bag replacing the active set with the domain of the context bag.
-If the stack is not empty, then the saved local variables are popped
-and restored, the return address is popped and assigned to the
-program counter, and the value of <{result}> is pushed onto the
+context bag and replacing the active set with the domain of the context bag.
+If the stack is not empty, then (1) the saved local variables are popped
+and restored, (2) the return address is popped and assigned to the
+program counter, and (3) the value of <{result}> is pushed onto the
 stack.
@@
 replaced with the domain of the context bag.  Even though its atomic counter
-is not zero, the stopped context is not considered atomic at this time.
-\texttt{Go} pops a context value and a return value of the stack and
+is not zero, the stopped context is not considered atomic at this time, but
+will resume being atomic once it is scheduled to run.
+
+\texttt{Go} pops a context value and a return value of the stack,
 pushes the return value onto the stack of the popped context, and adds
@@
 
-\emph{Charm}, the Harmony model checker is an explicit state model
+\emph{Charm}, the Harmony model checker, is an explicit state model
 checker that computes a \emph{Kripke structure} (essentially, a graph)
@@
 operation.  We call such operations \emph{breakable}.
-Note that this is essentially a form of Partial Order Reduction.
+Note that this is essentially a form of Partial Order Reduction~\cite{}.
+
 Reducing the state space this way is highly effective for Harmony.
@@
 
-Given a state and a context, Charm first applies the first instruction
-to the state.  Then it continues to apply instructions as long as
+Given a state and a context, Charm starts with applying the instruction
+pointed to by the context's program counter to the state.
+Then it continues to apply instructions as long as
 the active set is the singleton set (such as when the thread is in
@@
 Charm will try all possible choices for the set value
-on top of the stack.  If not, Charm will also check to
+on top of the stack.  If not, Charm will check to
 see if a trap is outstanding and interrupts are enabled.  If so,
@@
 
-Part of the model checking is trivially parallelizable, allowing for
-significant speedups on multi-core machines.
+About half of the work of the model checking is trivially parallelizable,
+allowing for significant speedups on multi-core machines.
 
@@
 effectively turning off data race detection.  Charm maintains on
-the edges of the Kripke a list of accesses that were made during
-the transition.  After all transitions from a state have been
+each edge of the Kripke structure a list of accesses that were made
+during the transition.  After all transitions from a state have been
 computed, Charm checks for data races.
@@
 States in which the context bag is empty represent states in which
-all threads have terminated.
-In the case the context bag is not empty, we check if there are any
-non-\emph{eternal} contexts.  If so, Harmony reports a \emph{deadlock}.
-Harmony programmers can specify that certain threads are eternal,
-such as server threads that operate in an infinite loop waiting for
-requests.
-If there is no deadlock and a behavior automaton was specified,
-Harmony also checks Harmony checks if the automaton is in a
-terminal state.  If not, a \emph{behavior violation} is reported.
+all threads have terminated.  In the case the context bag is not
+empty, we check if there are any non-\emph{eternal} contexts.  If
+so, Harmony reports a \emph{deadlock}.  (Harmony programmers can
+specify that certain threads are eternal, such as server threads
+that operate in an infinite loop waiting for requests.) If there
+is no deadlock and a behavior automaton was specified, Harmony also
+checks if the automaton is in a terminal state.  If not, a
+\emph{behavior violation} is reported.
 
@@
 (modify the state) if another thread makes a transition first.
-This is not always a problem: a thread waiting on a lock in a spinloop
+This is not usually a problem: a thread waiting on a lock in a spinloop
 is a common way of synchronizing threads, and Harmony does not report
-such issues.
+such issues.  We call this \emph{passive busy waiting} as the thread
+does not modify the state.
 
 A thread is considered \emph{actively busy waiting} if it cannot
-escape it strongly connected component without another thread
+escape its strongly connected component without another thread
 updating the shared state first.  A common example of this is a
@@
 acquiring and releasing the lock each time.  Harmony checks each
-strongly connected component for the existence of such situations.
+strongly connected component for the existence of such situations
+and reports them.
 
@@
 automaton, with on each edge zero or more \emph{symbols} (values
-that were printed.  If a behavior automaton was specified,
+that were printed).  If a behavior automaton was specified,
 then there are now two automata: the specified automaton
@@
 produces a warning.
-Although not current done, it is possible to compute
+Although not currently performed automatically, it is possible to compute
 ${\mathcal Spec}~\backslash~{\mathcal Impl}$
@@
 a detailed trace that is called a \emph{counter-example}.
-It always outputs the shortest such trace.
+Harmony always outputs the shortest such trace.
 However, in order to save memory, Charm does not actually keep
@@
 Harmony includes tools to quickly focus on the root cause of the
-problem.  In particular, in case of a problem, Harmony outputs a compact
+problem without having to dig through the HVM instructions.
+In particular, in case of a problem, Harmony outputs a compact
 but detailed and interactive overview of the problem in HTML format that
@@
 forwards or backwards to a context switch between threads).  The actual
-HVM code is in the lowerleft of the output.
+HVM code is in the lower left of the output.
 
@@
 in the lower-right, you can see that the thread is currently blocked
-and in atomic as well as read-only mode (conditions in a <{when}>
+and in atomic as well as read-only mode (because conditions in a <{when}>
 clause are not allowed to modify state).  You can also see its stack
@@
 the code in \autoref{fig:synch}(a)).  Next <{diner(0)}>, <{diner(3)}>,
-<{diner(4)}>, and <{diner(1)}> take similar turns.  All threads are
+<{diner(4)}>, and <{diner(1)}> take similar turns.  All threads end up
 blocked waiting for another thread to release a lock---a classic
@@
 
-Harmony does not offer much innovation on the front of model checking
-
 PlusCal and TLA+, Promela and Spin, DSLABS (UW), Java Pathfinder, Zing,

### 2022-01-13 paper/paper.tex
Latest

@@
 
+\section{Harmony Compiler}
+
+The compiler translates the Harmony language to a sequence of HVM
+instructions.  The compiler is mostly straightforward.  This section
+describes some of its unusual features.
+
+To help the programmer reason about the behavior of the generated code,
+the compiler preserves the left-to-right evaluation order of
+expresssions.  In other words, if expression $e_1$ appears before
+$e_2$ in the source language, then $e_1$ is evaluated before $e_2$
+in the HVM translation.  There is one exception, for the Harmony/Python
+$e_1~<{if}>~c~<{else}>~e_2$ expression, $c$ is evaluated first, and
+depending on the outcome, either $e_1$ or $e_2$ is evaluated.
+
+The compiler performs some standard optimizations that do not re-order
+expression evaluations, such as constant folding and jump threading.
+A more interesting optimization is \emph{dead variable elimination}:
+local variables that are no longer in use are automatically removed
+from the state using the HVM \texttt{DelVar} instruction.  The reason
+for this is to reduce state explosion by removing state that is no
+longer in use.
+
 \section{Analysis}

### 2022-01-13 paper/paper.tex
Latest paper version

@@
 concurrent and distributed programming to students who do not have a
-background in formal methods.  Harmony programs are model checked.
+background in formal methods.  Harmony programs are model checked~\cite{CES86}.
 Without any special formalism, Harmony can diagnose a large variety
@@
 Harmony programs compile to a virtual machine.  The Harmony virtual
-machine is formally specified using TLA+, allowing Harmony programs
-to run using TLC, the model checker that comes with the TLA+
-toolbox.  Harmony has its own custom concurrent model checker that
-scales significantly better for Harmony programs.
+machine is formally specified using TLA+~\cite{Lamport02}, allowing
+Harmony programs to run using TLC, the model checker that comes
+with the TLA+ toolbox.  Harmony has its own custom concurrent model
+checker that scales significantly better for Harmony programs.
 
@@
 and the Needham-Schroeder authentication protocols.
+
+Harmony's unique combination of features are a familiar and easy
+to learn high-level programming language, the ability to concisely
+specify an interface, the ability to compare behaviors of specification
+and implementation, the ability to modularly model check programs,
+and support for continuations and interrupts.  Harmony finds various common
+problems automatically without a complicated property specification
+language, such as Harmony's ability to find instances of undesirable
+busy waiting.
 
@@
 
-\paragraph{Stop/Go.}  \texttt{Stop} saves the current context (with
+\paragraph{Stop/Go.}  Harmony supports continuations as a first-class
+construct.  To enable this, \texttt{Stop} saves the current context (with
 its program counter incremented) in a specified shared location (usually the
@@
 
-PlusCal and TLA+, Promela and Spin, DSLABS (UW), Java Pathfinder, Zing,
-CBMC/JBMC.
+Harmony is preceded by many other model checkers and accompanying
+languages.  This section mentions some of Harmony's strongest influences.
+
+PlusCal~\cite{Lamport09} is an imperative programming language that,
+after translation to TLA+~\cite{Lamport02} (a formal verification
+language), can be run through the TLC model checker.  Harmony also
+supports compilation to TLA+.  While similar in spirit to Harmony,
+Harmony first compiles to a virtual machine that is close to CPUs
+like x86, ARM, and RISC-V, using a stack and common control flow
+techniques.  The resulting HVM has a TLA+ specification.  PlusCal,
+instead, \emph{trans\-piles} to TLA+.  Informally, the gap between
+PlusCal programs and their TLA+ representations is much smaller
+than the gap between Harmony programs and their TLA+ representations,
+making PlusCal more amenable to formal verification.  On the other
+hand, working effectively with PlusCal does require significant
+knowledge of TLA+, increasing the learning curve.
+
+Promela~\cite{SPIN} is a programming language that serves as input
+to the SPIN model checker.  Promela has many of the same features
+as Harmony and also includes explicit message channels.  Promela
+does not have a call stack and does not support procedure calls,
+making it hard to do modular development and verification.
+
+Zing~\cite{Zing} is a modeling language that, like Harmony, compiles to an
+intermediate language that resembles a conventional processor,
+including a call stack and dynamically allocated memory.  Like
+Harmony it supports procedure calls and concurrency.  Zing goes
+further in that it also support object-orientation.  Zing is
+mostly used as an intermediate language, transpiled from sources
+written in languages such as C\#.
+
+Java Pathfinder~\cite{Pathfinder} and JBMC~\cite{JBMC} are model
+checkers for Java bytecode, and,
+like Charm, can diagnose data races and deadlock in concurrent
+programs.  Because they operate at the level of Java bytecode, they
+have no support for atomic code blocks, which allows Harmony to be
+used both as a specification and an implementation language.
+Particularly the Harmony <{atomically when}> statement is very
+useful as a specification construct, for example to specify new
+synchronization primitives.
+
+DSLabs~\cite{MWA19} is an educational framework for model checking distributed
+systems.  Distributed algorithms are specified in Java in terms of
+nodes, messages, and timers in an event-driven style.  DSLabs does not
+support model checking sequential or concurrent programs.  Java
+also has the disadvantage that it is a verbose language.  Just the
+boilerplate for the Paxos project in DSLabs is well over 100 lines
+of Java code.  The entire Paxos implementation in Harmony is less
+than 50 lines.
 

### 2022-01-13 paper/paper.tex
Minor improvement
Added missing citation

@@
 also a ticket lock implementation and one based on continuations and
-a ready queue.  There are various concurrent queue implementations,
-including the lock-free implementation of~\cite{MS96}.
+a ready queue.  The distribution also includes various concurrent
+queue implementations, including the lock-free implementation
+of~\cite{MS96}.
 
@@
 operation.  We call such operations \emph{breakable}.
-Note that this is essentially a form of Partial Order Reduction~\cite{}.
+Note that this is essentially a form of Partial Order Reduction~\cite{Val91}.
 

### 2022-01-13 paper/paper.tex
Improved code a bit

@@
 def thread(self):
-    let op = choose({ "deq", "enq" }):
-        if op == "enq":
+    if choose({ "deq", "enq" }) == "enq":
         print("call enqueue", self)

### 2022-01-13 book/paper.tex
Improved naming of a file

@@
 queue in Harmony (exploiting some methods from the \texttt{list}
-module described in \autoref{ap:list}),
+module described in \autoref{ap:list}).
+It is a credible queue implementation,
 but it cannot be used with threads concurrently accessing this queue.
 \autoref{fig:queues}(b) shows the corresponding concurrent specification.
+It cannot be used as an implementation for a queue, as processors generally
+do not have atomic operations on lists, but it will work well as a
+specification.
 See \autoref{fig:queuedemo} for a simple demonstration program that uses
@@
 \begin{code}
-<{:input:../code/intsettest.hny}>
-\end{code}
-\caption{\harmonylink{code/intsettest.hny} Test code for set objects}
-\label{fig:intsettest}
+<{:input:../code/setobjtest.hny}>
+\end{code}
+\caption{\harmonylink{code/setobjtest.hny} Test code for set objects}
+\label{fig:setobjtest}
 \end{figure}
@@
 Method <{contains()}> checks if a particular value is in the list.
-\autoref{fig:intsettest} contains a simple (although not very thorough)
+\autoref{fig:setobjtest} contains a simple (although not very thorough)
 test program to demonstrate the use of set objects.
@@
 would find no problems.
-Similarly, \autoref{fig:intsettest} is not a good test program
+Similarly, \autoref{fig:setobjtest} is not a good test program
 for something as complicated as \autoref{fig:linkedlist}.
@@
 the specification.
-
-\begin{comment}
-
-\begin{figure}
-\begin{code}
-<{:input:../code/qtest1.hny}>
-\end{code}
-\caption{\harmonylink{code/qtest1.hny} Concurrent \texttt{put()}}
-\label{fig:qtest1}
-\end{figure}
-
-If you want a purely black box test for testing a queue implementation,
-then adding linearization points is not possible.
-We will present a few test
-programs that try to identify incorrect behavior of
-a black box queue implementation in the presence of concurrency.
-
-\autoref{fig:qtest1} shows a program that
-checks concurrent <{put}> operations using
-<{N}> threads.  Thread \texttt{putter(<{v}>)}
-adds <{v}> to the queue.
-Thread \texttt{main()} waits until all \texttt{putter}
-threads have finished, and then dequeues <{N}>
-items, verifying that they form the set
-$\{ 0 .. <{N}> - 1 \}$.
-
-\begin{figure}
-\begin{code}
-<{:input:../code/qtest2.hny}>
-\end{code}
-\caption{\harmonylink{code/qtest2.hny} Concurrent <{get()}>}
-\label{fig:qtest2}
-\end{figure}
-
-Similarly, \autoref{fig:qtest2} shows a program that
-checks concurrent <{get}> operations using
-<{N}> threads.
-The queue is initialized with the sequence
-$0 .. <{N}> - 1$.
-Thread \texttt{getter(<{v}>)} removes an entry from the
-queue and makes sure that no other thread got the
-same entry.
-Thread \texttt{main()} waits until all <{getter}>
-threads have finished and checks to make sure that
-the queue is empty.
-
-\begin{figure}
-\begin{code}
-<{:input:../code/qtest3.hny}>
-\end{code}
-\caption{\harmonylink{code/qtest3.hny} Mix of concurrent
-<{get()}> and \texttt{put()} operations}
-\label{fig:qtest3}
-\end{figure}
-
-\autoref{fig:qtest3} checks a mixture of
-concurrent <{get}> and <{put}> operations.
-It starts <{N getter}> threads and
-<{N}> \texttt{putter} threads.
-Thread <{main}> waits until all have finished,
-and then checks to make sure that whatever is left
-on the queue was not also consumed by some
-<{getter}> thread.
-Finally, it makes sure that the queue is empty
-once all the values that were added
-to it by the \texttt{putter} threads have been consumed.
-
-A common mistake that people make is to use
-a globally shared variable in the implementation of
-a concurrent data structure.
-The <{get()}> method of the queue implementation
-uses a local variable <{node}>, declared using
-a <{let}> statement, but it would be easy to
-mistakenly use a global variable <{node}> instead.
-If there is only one queue, this is not an issue.
-So, it is important to test not just one queue at a
-time, but also multiple queues.
-\autoref{fig:qtest4} gives an example that
-tests the <{get}> method for use of global
-variables.  Even if the assertion does not fail,
-Harmony might find a data race if there are global
-variables shared between different queues.
-
-While having five different test programs is fine,
-it would be nice if we can combine them all in one
-test program.
-This has to be done with some care, as running the
-concurrent test programs simultaneously would
-lead to a state space explosion.
-Instead, we can leverage the <{choose}>
-expression.
-\autoref{fig:multitest} shows how one might go
-about this.
-Harmony will run all the tests, but the
-tests do not interfere with one another.
-
-\begin{figure}
-\begin{code}
-<{:input:../code/qtest4.hny}>
-\end{code}
-\caption{\harmonylink{code/qtest4.hny} Test with multiple queues}
-\label{fig:qtest4}
-\end{figure}
-
-\begin{figure}
-\begin{code}
-<{:input:../code/multitest.hny}>
-\end{code}
-\caption{\harmonylink{code/multitest.hny} Running a
-sequential test followed by several concurrent tests}
-\label{fig:multitest}
-\end{figure}
-
-\end{comment}
 
@@
 \begin{code}
-<{:input:../code/baddblwait.hny}>
+<{:inline:}>
+while forks[left]:
+    synch.wait(?conds[left], ?mutex)
+while forks[right]:
+    synch.wait(?conds[right], ?mutex)
+<{:end:}>
 \end{code}

### 2022-01-14 paper/paper.tex
Changing the spin

@@
 
-\title{Concurrent and Distributed Programming in Harmony}
+\title{Harmony: A Verifiable Systems Programming Language}
 \author{Robbert van Renesse \and William Ma \and Kevin Sun \and Anthony Yang}
@@
 \emph{Harmony} is a Python-like programming language designed to teach
-concurrent and distributed programming to students who do not have a
-background in formal methods.  Harmony programs are model checked~\cite{CES86}.
+systems programming and verification techniques to students who do not have a
+background in formal methods.  Harmony programs, which can be multi-threaded
+or distributed, are model checked~\cite{CES86}.
 Without any special formalism, Harmony can diagnose a large variety
@@
 \item Busy Waiting: waiting for a condition without blocking.
+\item Violations of interrupt safety: a race condition caused by the execution
+of an interrupt handler.
 \end{compactitem}
@@
 outputs of the program.  Groups of statements can be made atomic
-using the <{atomically}> keyword.  Finally, Harmony supports
-waiting until some condition is reached with the <{when}> statement.
+using the <{atomically}> keyword.  Harmony supports
+waiting until some condition is reached with its <{when}> statement.
 
@@
 
+Harmony supports continuations as first-class objects, allowing the
+state of a thread can be saved in a variable and later restored.
+Continuations enable the implementation of specialized thread
+schedulers as well as the implementation of thread forking by
+reinstantiating a saved thread state multiple times.
+
 Harmony has been used to verify the correctness of various lock
@@
 and implementation, the ability to modularly model check programs,
-and support for continuations and interrupts.  Harmony finds various common
-problems automatically without a complicated property specification
-language, such as Harmony's ability to find instances of undesirable
-busy waiting.
+and support for continuations and interrupts.  In addition, Harmony
+finds various common problems automatically without a complicated
+property specification language, such as Harmony's ability to find
+instances of undesirable busy waiting.
 
@@
 its model checker, and how they may be used by programmers to
-develop code for reliable concurrent or distributed applications.
+develop specify and implement reliable concurrent and/or
+distributed systems.
 
@@
 \end{center}
-\caption{Lock specification and a spinlock implementation based on
+\caption{(a) Lock specification and (b) a spinlock implementation based on
 test-and-set}
@@
 
+Dist\-Algo~\cite{DistAlgo} is a programming language based on Python to
+simplify the specification and verification of distributed algorithms.
+Dist\-Algo includes a framework for runtime checking of safety and liveness
+properties~\cite{LS20}.
+
 \bibliographystyle{alpha}

### 2022-01-14 paper/paper.tex
More focus on systems programming

@@
 \emph{Harmony} is a Python-like programming language designed to teach
-systems programming and verification techniques to students who do not have a
-background in formal methods.  Harmony programs, which can be multi-threaded
+systems specification, implementation, and verification techniques to
+students who do not have a background in formal methods.
+Harmony programs, which can be multi-threaded
 or distributed, are model checked~\cite{CES86}.
@@
 \item ``Silly errors'': accessing non-existent variables,
-divide-by-zero, out-of-bounds array access, trying to multiply two lists
-with one another, and many more.
+divide-by-zero, out-of-bounds array access, dereferencing dangling
+pointers, trying to multiply two lists with one another,
+and many more.
 \item Safety violations due to assertions failing, such as an assertion to ensure mutual exclusion.
@@
 of an interrupt handler.
+\item Memory leaks.
 \end{compactitem}
@@
 and implementation, the ability to modularly model check programs,
-and support for continuations and interrupts.  In addition, Harmony
-finds various common problems automatically without a complicated
-property specification language, such as Harmony's ability to find
-instances of undesirable busy waiting.
+and support for dynamic memory allocation, continuations, and
+interrupts.  In addition, Harmony finds various common problems
+automatically without a complicated property specification language,
+such as Harmony's ability to find instances of undesirable busy
+waiting.
 
@@
 also a ticket lock implementation and one based on continuations and
-a ready queue.  The distribution also includes various concurrent
+a blocked queue.  The distribution also includes various concurrent
 queue implementations, including the lock-free implementation
@@
 the specification program.
+
+\begin{figure}
+\begin{center}
+{\footnotesize
+\begin{code}
+<{:inline:}>
+import list
+
+def Lock(held):
+    result = { .held: False, .blocked: [] }
+
+def acquire(lk):
+    atomically:
+        if lk->held:
+            stop lk->blocked[len lk->blocked]
+        else:
+            lk->held = True
+
+def release(lk):
+    atomically:
+        if lk->blocked == []:
+            lk->held = False
+        else:
+            go (list.head(lk->blocked)) ()
+            lk->blocked = list.tail(lk->blocked)
+<{:end:}>
+\end{code}
+}
+\end{center}
+\caption{FIFO lock using continuation}
+\label{fig:continuation}
+\end{figure}
+
+\subsection{Continuations}
+
+The Harmony <{stop x}> expression stops the current thread and
+saves its state in the variable <{x}>.  Another thread can
+resume the thread using the statement <{go x r}>, which causes
+the original thread to resume, with the <{stop x}> expression
+evaluating to the value <{r}>.
+\autoref{fig:continuation} shows another implementation of
+\autoref{fig:synch}(a) using continuations, where blocked
+threads are placed on a FIFO queue and released in that order.
+Besides synchronization and scheduling, continuations also enable
+the implementation of a primitive such as Posix \texttt{fork}.
+
+\subsection{Interrupts}
+
+Correct interrupt handling is another difficult aspect of
+systems programming.  Interrupt handlers typically have various
+stringent limitations in what they are allowed to do in order
+to ensure the absence of race conditions due to re-entrant code.
+The Harmony <{trap f(x)}> statement sets an interrupt handler
+<{f(x}}> to be called at an unspecified later time.  You can think
+of it as setting an unspecified timer, but it can model any type
+of interrupt.  In addition, the Harmony <{setintlevel L}> statement
+enables or disables interrupts and returns the old interrupt level.
 
@@
 also modify one of the shared variables.
+Below we describe in more detail some of the more unusual ones.
 
@@
 
+\begin{comment}
 \paragraph{Frame/Return.}  \texttt{Frame} is the first operation of
@@
 stack.
+\end{comment}
 
@@
 
+\begin{comment}
 \paragraph{SetIntLevel.}  A context contains a boolean that determines
@@
 often require timer interrupts to trigger actions such as retransmissions.
+\end{comment}
 
@@
 \begin{center}
-\includegraphics[width=\textwidth]{diners.png}
+\includegraphics[width=.8\textwidth]{diners.png}
 \end{center}

### 2022-01-14 paper/paper.tex
Latest

@@
 \end{center}
-\caption{A selection of Harmony programs and their outputs}
+\caption{A demonstration of various Harmony language constructs.  The DFAs
+are the actual outputs of the given programs.}
 \label{fig:helloworld}
@@
 is selected, the thread enqueues its identifier <{self}>.  The program
-includes a print statement at the beginning and end of every queue
-operation.
+includes a <{print}> statement at the beginning and end of every queue
+operation.  Think of the <{print}> statement as \emph{annotations}
+that tell Harmony what the interesting events are with respect to
+queue operations.
 
@@
 enables or disables interrupts and returns the old interrupt level.
+The model checker tries calling the interrupt handler at every
+possible point in the execution of the Harmony program.
 

### 2022-01-14 paper/paper.tex
Pass made

@@
 \item Liveness violations (deadlock).
-\item Faulty behaviors: implementation has behaviors not seen with specification, such as violations of linearizability~\cite{HW90}.
+\item Faulty behaviors: implementation has behaviors not seen in the specification, such as violations of linearizability~\cite{HW90}.
 \item Incomplete behaviors: specification has behaviors not seen with implementation, such as multiple readers not able to get a reader/writer lock simultaneously.
-\item Data races: variables accessed without a lock.
+\item Data races: shared variables accessed without a lock.
 \item Busy Waiting: waiting for a condition without blocking.
@@
 \emph{interleaving} due to multi-threading, and \emph{interrupts}.
-\autoref{fig:helloworld} demonstrates the <{choose}> and the <{spawn}>
-statements.  Since Harmony programs are model-checked and multiple
+\autoref{fig:helloworld} demonstrates various non-deterministic
+Harmony programs.
+Since Harmony programs are model-checked and all possible
 executions are explored across a finite set of states, the output of a
@@
 Harmony supports continuations as first-class objects, allowing the
-state of a thread can be saved in a variable and later restored.
+state of a thread to be saved in a variable and later restored.
 Continuations enable the implementation of specialized thread
@@
 reinstantiating a saved thread state multiple times.
+Other uses include exception handlers, coroutines, and generators.
 
@@
 its model checker, and how they may be used by programmers to
-develop specify and implement reliable concurrent and/or
-distributed systems.
+specify and implement reliable concurrent and/or distributed systems.
+
+\begin{figure*}[h]
+\begin{center}
+\begin{tabular}{cc}
+{\footnotesize
+\begin{tcolorbox}[width=0.45\linewidth]
+<{:const:head}>
+<{:const:tail}>
+<{:const:append}>
+<{:const:test_and_set}>
+<{:inline:}>
+def Lock():
+    result = False
+
+def acquire(lk):
+    atomically when not !lk:
+        !lk = True
+
+def release(lk):
+    atomically !lk = False
+<{:end:}>
+\end{tcolorbox}
+}
+&
+{\footnotesize
+\begin{tcolorbox}[width=0.45\linewidth]
+<{:inline:}>
+def test_and_set(s):
+    atomically:
+        result = !s
+        !s = True
+
+def acquire(lk):
+    while test_and_set(lk):
+        pass
+
+<{:end:}>
+\end{tcolorbox}
+}
+\\
+(a) & (b)
+\end{tabular}
+\end{center}
+\caption{(a) Lock specification and (b) a spinlock implementation based on
+test-and-set}
+\label{fig:synch}
+\end{figure*}
+
+\begin{figure*}[h]
+\begin{center}
+\begin{tabular}{cc}
+{\footnotesize
+\begin{tcolorbox}[width=0.45\linewidth]
+<{:inline:}>
+import list
+
+def Queue():
+    result = []
+
+def enqueue(q, v):
+    atomically !q = list.append(!q, v)
+
+def dequeue(q):
+    atomically:
+        if !q == []:
+            result = None
+        else:
+            result = list.head(!q)
+            !q = list.tail(!q)
+
+<{:end:}>
+\end{tcolorbox}
+}
+&
+{\footnotesize
+\begin{tcolorbox}[width=0.48\linewidth]
+<{:inline:}>
+import queue
+
+const NOPS = 4
+q = queue.Queue()
+
+def thread(self):
+    if choose({ "deq", "enq" }) == "enq":
+        print("call enqueue", self)
+        queue.enqueue(?q, self)
+        print("done enqueue", self)
+    else:
+        print("call dequeue", self)
+        let v = queue.dequeue(?q):
+            print("done dequeue", self, v)
+
+for i in {1..NOPS}: spawn thread(i)
+<{:end:}>
+\end{tcolorbox}
+}
+\\
+(a) concurrent <{queue}> module & (b) test program
+\end{tabular}
+\end{center}
+\caption{The (a) specification and (b) test program for a concurrent queue}
+\label{fig:queue}
+\end{figure*}
+
+\section{Harmony Language}
+
+\subsection{Basics}
+
+Harmony is heavily based on Python to reduce the learning curve for
+users.  Many of Python's powerful constructs and operators are
+supported, including comprehensions.  Harmony does
+not support classes and instead supports memory allocation and
+pointers in a way that is familiar to C programmers.  While this
+perhaps seems like a step back in language evolution, many
+synchronization algorithms in the literature explicitly consider
+memory accesses, and this has been a bonus for
+analyzing such algorithms with Harmony.  Moreover, the Harmony model
+checker can be used to check for memory leaks or access to dangling
+pointers.
+
+For example, \autoref{fig:ms} shows a reproduction of the well-known
+Michael\&Scott two-lock concurrent queue~\cite{MS96} along with a
+Harmony version of the same algorithm.
+Harmony uses the `<{?}>' instead of the `<{&}>' symbol to take the address
+of a memory location, but inherits the same <{q->field}> notation
+from C to access the given field in the dictionary pointed to by
+pointer <{q}>.  <{q->field}> is shorthand for <{(!q)["field"]}> in
+Harmony, where `<{!}>' is the Harmony operator to dereference a pointer.
+Harmony also complains, correctly, about a data race when
+there is a concurrent \texttt{enqueue} and \texttt{dequeue} operation
+on an empty queue---the implementer should ensure that accesses to the
+\texttt{next} field are sequentially consistent.
+
+The Harmony example in \autoref{fig:ms} shows some other language
+features in Harmony that are not in Python.
+The <{let x = e: S}> statement introduces a new bound variable <{x}>
+with value <{e}> that is only valid within the statement block <{S}>.
+Harmony does not have \texttt{return}, \texttt{break}, or
+\texttt{continue} statements to force the programmer into a purely
+nested control flow.  Each method has a variable <{result}> in which
+the return value of the method is stored.
+
+\begin{figure*}
+\begin{center}
+\begin{tabular}{l|c|c|c|c}
+           & <{NOPS=1}> & <{NOPS=2}> & <{NOPS=3}> & <{NOPS=4}> \\ \hline\hline
+queue specification    & 11/0.3 & 110/0.3 &  1,286/0.4 &  16,332/0.9 \\ \hline
+queue impl. w/o check  & 20/0.5 & 383/0.6 &  8,874/0.8 & 261,878/12.7 \\ \hline
+queue impl. with check & 20/0.5 & 425/0.6 & 10,977/1.0 & 437,226/23.6 \\
+\end{tabular}
+\end{center}
+\caption{The number of states generated by the model checker and the
+execution time (in seconds) for running \autoref{fig:queue}(b).
+This is done for three configurations, varying the value of <{NOPS}>:
+the specification (\autoref{fig:queue}(a)), a simple linked list
+implementation using locks, and the same implementation combined with
+the behavior automaton generated by the first configuration.}
+\label{fig:qcmp}
+\end{figure*}
+
+\subsection{\texttt{atomically}}
+
+Statements in Harmony, including groups of statements, can be tagged with
+the \texttt{atomically} keyword to make them atomic.  What this means
+is that the statement is evaluated atomically by the Harmony virtual
+machine without interleaving of instructions of other threads.  It is
+a \emph{stronger} condition than holding a lock---a lock still allows
+concurrent execution of multiple threads, but it prevents multiple
+threads from executing in the same critical section.
+The <{atomically}> keyword is intended for <{specification}> more than
+for <{implementation}> of concurrent algorithms.
+
+For example, \autoref{fig:synch}(a) shows the specification for locks
+from the Harmony <{synch}> module that was used in \autoref{fig:ms},
+featuring the Harmony <{when}> statement.
+\autoref{fig:synch}(b) shows an implementation of the <{acquire}>
+method based on a test-and-set primitive, which in turn is
+\emph{specified}.
+Imprecisely, methods that use the <{atomically}> keyword are specifications.
+Whether something is a specification or an implementation is in the
+eye of the beholder.
+Important, however, is the notion of modularity that is implied here.
+Harmony can model check the implementation of \autoref{fig:ms} using
+the specification of a lock in \autoref{fig:synch}(a).
+Separately, Harmony can check the correctness of the implementation in
+\autoref{fig:synch}(b) against the specification in
+\autoref{fig:synch}(a).
+
+This modularity allows larger systems to be model checked in a way
+that would not be tractable otherwise.
+\autoref{fig:qcmp} shows some measurements taken from running
+Harmony against the test program in \autoref{fig:queue}(b).
+The Harmony compiler allows both constants and modules to be
+specified on the command line, so the test program can easily be
+run with different versions of the <{queue}> module and different
+values for <{NOPS}>, the number of threads spawned to choose and
+perform a queue operation.  The table has a row for three different
+configurations and a column for four different values for <{NOPS}>.
+It is clear that running the test program against the specification
+involves significantly fewer states, and therefore less time, than
+running the test program against an implementation, particularly
+if the behavior of that implementation must be checked as well.
+
+With Harmony, one can check the implementation of a lock separately,
+but use the specification of the lock in the implementation of the
+queue.  Similarly, one can check the implementation of a queue,
+but then use the specification of the queue in a larger application.
+Doing so is essentially a form of \emph{Partial Order Reduction}~\cite{Val91},
+replacing the <{enqueue}> and <{dequeue}> implementations
+by atomic actions.
+
+The Harmony distribution comes with a collection of specifications,
+each often with multiple implementations.  The lock specification
+comes with the spinlock implementation of \autoref{fig:synch}(a), but
+also a ticket lock implementation and one based on continuations and
+a queue of blocked threads.  The distribution also includes various concurrent
+queue implementations, including the lock-free implementation
+of~\cite{MS96}.
+
+\subsection{Checking an implementation}
+
+The Harmony language allows objects such as locks and concurrent queues
+to be specified and implemented in separate programs.  But the programs
+are not \emph{active} as they are just a collection of method definitions.
+A model checker can only check behaviors, but these programs have empty
+behaviors.
+Suppose we want to check the correctness of the concurrent queue
+implementation of \autoref{fig:ms}.  First, we would need a specification
+of a concurrent queue.  Second, we would need a test program that tries
+various concurrent interleavings of <{enqueue}> and <{dequeue}> operations.
+
+\autoref{fig:queue}(a) shows the specification of a concurrent
+<{queue}> module.  A queue is represented as a list.
+The <{enqueue}> method appends a value onto the end of the list,
+while the <{dequeue}> method extracts the head, both as atomic actions.
+
+To test a queue, we want to perform a set of concurrent <{enqueue}>
+and <{dequeue}> operations concurrently.
+\autoref{fig:queue}(b) shows how this can be specified in Harmony.
+The test program spawns 4 threads that each select an operation
+using the nondetermistic <{choose}> statement.  In case <{enqueue}>
+is selected, the thread enqueues its identifier <{self}>.  The program
+includes a <{print}> statement at the beginning and end of every queue
+operation.  Think of the <{print}> statements as \emph{annotations}
+that tell Harmony what the interesting events are with respect to
+queue operations.
+
+The test program is similar to what a programmer might have written
+as a unit test.  The programmer would run the program a bunch of
+times and look at the output in order to see if it looks reasonable.
+Doing so is difficult and error prone, as the output tends
+to be complex and in the mess of many lines of output it is easy
+to overlook problems such as two threads dequeuing the same item
+or items being dequeued in the wrong order.  The output of concurrent
+print statement can be mangled together.  Not only that, but the
+programmer would need to run this test program many times and check
+the output each time.  It even is hard to write a program that
+checks the output automatically.
+
+This process is automated in Harmony.  The approach is to
+run the model checker twice.  The first time, the test program is
+run against the specification of the concurrent queue.  Harmony
+outputs a DFA such as shown in \autoref{fig:helloworld}.  The second
+time, the same test program is run against the implementation of the
+concurrent queue, but with the inclusion of the DFA.  The model
+checker ends up verifying the Cartesian product of the implementation
+and the DFA and ensures that every behavior of the implementation
+program (in terms of its print statements) is also a behavior of
+the specification program.  In case the implementation has a behavior
+that is not allowed by the specificatino, the model checker can
+provide the shortest counter-example.
+
+\begin{figure}
+\begin{center}
+{\footnotesize
+\begin{code}
+<{:inline:}>
+import list
+
+def Lock(held):
+    result = { .held: False, .blocked: [] }
+
+def acquire(lk):
+    atomically:
+        if lk->held:
+            stop lk->blocked[len lk->blocked]
+        else:
+            lk->held = True
+
+def release(lk):
+    atomically:
+        if lk->blocked == []:
+            lk->held = False
+        else:
+            go (list.head(lk->blocked)) ()
+            lk->blocked = list.tail(lk->blocked)
+<{:end:}>
+\end{code}
+}
+\end{center}
+\caption{FIFO lock using continuation}
+\label{fig:continuation}
+\end{figure}
+
+\subsection{Continuations}
+
+The Harmony <{stop x}> expression stops the current thread and saves
+its state in the variable <{x}>.  Another thread can resume the
+thread using the statement <{go x r}>, which causes the original
+thread to resume, with the <{stop x}> expression evaluating to the
+value <{r}>.  \autoref{fig:continuation} shows another implementation
+of the lock specification of \autoref{fig:synch}(a) using continuations,
+where blocked threads are placed on a FIFO queue and released in
+that order.  Besides synchronization and scheduling, continuations
+also enable the implementation of a primitive such as Posix
+\texttt{fork}.
+
+\subsection{Interrupts}
+
+Correct interrupt handling is another difficult aspect of
+systems programming.  Interrupt handlers typically have various
+stringent limitations in what they are allowed to do in order
+to ensure the absence of race conditions due to re-entrant code.
+The Harmony <{trap f(x)}> statement sets an interrupt handler
+<{f(x)}> to be called at an unspecified later time.  You can think
+of it as setting an unspecified timer, but it can model any type
+of interrupt.  In addition, the Harmony <{setintlevel L}> statement
+enables or disables interrupts and returns the old interrupt level.
+The model checker tries calling the interrupt handler at every
+possible point in the execution of the Harmony program.
 
@@
 
-\begin{figure*}[h]
-\begin{center}
-\begin{tabular}{cc}
-{\footnotesize
-\begin{tcolorbox}[width=0.45\linewidth]
-<{:const:head}>
-<{:const:tail}>
-<{:const:append}>
-<{:const:test_and_set}>
-<{:inline:}>
-def Lock():
-    result = False
-
-def acquire(lk):
-    atomically when not !lk:
-        !lk = True
-
-def release(lk):
-    atomically !lk = False
-<{:end:}>
-\end{tcolorbox}
-}
-&
-{\footnotesize
-\begin{tcolorbox}[width=0.45\linewidth]
-<{:inline:}>
-def test_and_set(s):
-    atomically:
-        result = !s
-        !s = True
-
-def acquire(lk):
-    while test_and_set(lk):
-        pass
-
-<{:end:}>
-\end{tcolorbox}
-}
-\\
-(a) & (b)
-\end{tabular}
-\end{center}
-\caption{(a) Lock specification and (b) a spinlock implementation based on
-test-and-set}
-\label{fig:synch}
-\end{figure*}
-
-\begin{figure*}[h]
-\begin{center}
-\begin{tabular}{cc}
-{\footnotesize
-\begin{tcolorbox}[width=0.45\linewidth]
-<{:inline:}>
-import list
-
-def Queue():
-    result = []
-
-def enqueue(q, v):
-    atomically !q = list.append(!q, v)
-
-def dequeue(q):
-    atomically:
-        if !q == []:
-            result = None
-        else:
-            result = list.head(!q)
-            !q = list.tail(!q)
-
-<{:end:}>
-\end{tcolorbox}
-}
-&
-{\footnotesize
-\begin{tcolorbox}[width=0.48\linewidth]
-<{:inline:}>
-import queue
-
-const NOPS = 4
-q = queue.Queue()
-
-def thread(self):
-    if choose({ "deq", "enq" }) == "enq":
-        print("call enqueue", self)
-        queue.enqueue(?q, self)
-        print("done enqueue", self)
-    else:
-        print("call dequeue", self)
-        let v = queue.dequeue(?q):
-            print("done dequeue", self, v)
-
-for i in {1..NOPS}: spawn thread(i)
-<{:end:}>
-\end{tcolorbox}
-}
-\\
-(a) concurrent <{queue}> module & (b) test program
-\end{tabular}
-\end{center}
-\caption{The (a) specification and (b) test program for a concurrent queue}
-\label{fig:queue}
-\end{figure*}
-
-\section{Harmony Language}
-
-\subsection{Basics}
-
-Harmony is heavily based on Python to reduce the learning curve for
-users.  Many of Python's powerful constructs and operators are
-supported, including comprehensions.  Harmony does
-not support classes and instead supports memory allocation and
-pointers in a way that is familiar to C programmers.  While this
-perhaps seems like a step back in language evolution, many
-synchronization algorithms in the literature explicitly consider
-memory accesses, and perhaps surprisingly this has been a bonus for
-analyzing such algorithms with Harmony.  Moreover, the Harmony model
-checker can be used to check for memory leaks or access to dangling
-pointers.
-
-For example, \autoref{fig:ms} shows a reproduction from the well-known
-Michael\&Scott two-lock concurrent queue~\cite{MS96} along with a
-Harmony version of the same algorithm.
-Harmony uses the `<{?}>' instead of the `<{&}>' symbol to take the address
-of a memory location, but inherits the same <{q->field}> notation
-from C to access the given field in the dictionary pointed to by
-pointer <{q}>.  <{q->field}> is shorthand for <{(!q)["field"]}> in
-Harmony, where `<{!}>' is the Harmony operator to dereference a pointer.
-Harmony also complaints, correctly, about a data race when
-there is a concurrent \texttt{enqueue} and \texttt{dequeue} operation
-on an empty queue---the implementer should ensure that accesses to the
-\texttt{next} field are sequentially consistent.
-
-The Harmony example in \autoref{fig:ms} shows some other language
-features in Harmony that are not in Python.
-The <{let x = e: S}> statement introduces a new bound variable <{x}>
-with value <{e}> that is only valid within the statement block <{S}>.
-Also, Harmony does not have \texttt{return}, \texttt{break}, or
-\texttt{continue} statements to force the programmer into a purely
-nested control flow.  Each method has a variable <{result}> in which
-the return value of the method is stored.
-
-\subsection{\texttt{atomically}}
-
-Statements in Harmony, including groups of statements, can be tagged with
-the \texttt{atomically} keyword to make them atomic.  What this means
-is that the statement is evaluated atomically by the Harmony virtual
-machine without interleaving of instructions of other threads.  It is
-a \emph{stronger} condition than holding a lock---a lock still allows
-concurrent execution of multiple threads, but it prevents multiple
-threads from executing in the same critical section.
-The <{atomically}> keyword is intended for <{specification}> more than
-for <{implementation}> of concurrent algorithms.
-
-For example, \autoref{fig:synch}(a) shows the specification for locks
-from the Harmony <{synch}> module that was used in \autoref{fig:ms},
-featuring the Harmony <{when}> statement.
-\autoref{fig:synch}(b) shows an implementation of the <{acquire}>
-method based on a test-and-set primitive, which in turn is
-\emph{specified}.
-Imprecisely, methods that use the <{atomically}> keyword are specifications.
-Whether something is a specification or an implementation is in the
-eye of the beholder.
-Important, however, is the notion of modularity that is implied here.
-Harmony can model check the implementation of \autoref{fig:ms} using
-the specification of a lock in \autoref{fig:synch}(a).
-Separately, Harmony can check the correctness of the implementation in
-\autoref{fig:synch}(b) against the specification in
-\autoref{fig:synch}(a).  This modularity allows larger systems to
-be model checked in a way that would not be tractable otherwise.
-
-The Harmony distribution comes with a collection of specifications,
-each often with multiple implementations.  The lock specification
-comes with the spinlock implementation of \autoref{fig:synch}(a), but
-also a ticket lock implementation and one based on continuations and
-a blocked queue.  The distribution also includes various concurrent
-queue implementations, including the lock-free implementation
-of~\cite{MS96}.
-
-\subsection{Checking an implementation}
-
-The Harmony language allows objects such as locks and concurrent queues
-to be specified and implemented in separate programs.  But the programs
-are not \emph{active} as they are just a collection of method definitions.
-A model checker can only check behaviors, but these programs have empty
-behaviors.
-Suppose we want to check the correctness of the concurrent queue
-implementation of \autoref{fig:ms}.  First, we would need a specification
-of a concurrent queue.  Second, we would need a test program that tries
-various concurrent interleavings of <{enqueue}> and <{dequeue}> operations.
-
-\autoref{fig:queue}(a) shows the specification of a concurrent
-<{queue}> module.  A queue is represented as a list.
-The <{enqueue}> method add a value onto the end of the list,
-while the <{dequeue}> method extracts the head, both as atomic actions.
-
-To test a queue, we want to perform a set of concurrent <{enqueue}>
-and <{dequeue}> operations concurrently.
-\autoref{fig:queue}(b) shows how this can be specified in Harmony.
-The test program spawns 4 threads that each select an operation
-using the nondetermistic <{choose}> statement.  In case <{enqueue}>
-is selected, the thread enqueues its identifier <{self}>.  The program
-includes a <{print}> statement at the beginning and end of every queue
-operation.  Think of the <{print}> statement as \emph{annotations}
-that tell Harmony what the interesting events are with respect to
-queue operations.
-
-The test program is similar to what a programmer might have written
-as a unit test.  The programmer would run the program a bunch of
-times and look at the output in order to see if it looks reasonable.
-Doing so is, however, difficult and error prone, as the output tends
-to be complex and in the mess of many lines of output it is easy
-to overlook problems such as two threads dequeuing the same item
-or items being dequeued in the wrong order.  The output of concurrent
-print statement can be mangled together.  Not only that, but the
-programmer would need to run this test program many times and check
-each the output each time.  It even is hard to write a program that
-checks the output automatically.
-
-This process is automated in Harmony.  The approach is to
-run the model checker twice.  The first time, the test program is
-run against the specification of the concurrent queue.  Harmony
-outputs a DFA such as shown in \autoref{fig:helloworld}.  The second
-time, the same test program is run against the implementation of the
-concurrent queue, but with the inclusion of the DFA.  The model
-checker ends up verifying the Cartesian product of the implementation
-and the DFA and ensures that every behavior of the implementation
-program (in terms of its print statements) is also a behavior of
-the specification program.
-
-\begin{figure}
-\begin{center}
-{\footnotesize
-\begin{code}
-<{:inline:}>
-import list
-
-def Lock(held):
-    result = { .held: False, .blocked: [] }
-
-def acquire(lk):
-    atomically:
-        if lk->held:
-            stop lk->blocked[len lk->blocked]
-        else:
-            lk->held = True
-
-def release(lk):
-    atomically:
-        if lk->blocked == []:
-            lk->held = False
-        else:
-            go (list.head(lk->blocked)) ()
-            lk->blocked = list.tail(lk->blocked)
-<{:end:}>
-\end{code}
-}
-\end{center}
-\caption{FIFO lock using continuation}
-\label{fig:continuation}
-\end{figure}
-
-\subsection{Continuations}
-
-The Harmony <{stop x}> expression stops the current thread and
-saves its state in the variable <{x}>.  Another thread can
-resume the thread using the statement <{go x r}>, which causes
-the original thread to resume, with the <{stop x}> expression
-evaluating to the value <{r}>.
-\autoref{fig:continuation} shows another implementation of
-\autoref{fig:synch}(a) using continuations, where blocked
-threads are placed on a FIFO queue and released in that order.
-Besides synchronization and scheduling, continuations also enable
-the implementation of a primitive such as Posix \texttt{fork}.
-
-\subsection{Interrupts}
-
-Correct interrupt handling is another difficult aspect of
-systems programming.  Interrupt handlers typically have various
-stringent limitations in what they are allowed to do in order
-to ensure the absence of race conditions due to re-entrant code.
-The Harmony <{trap f(x)}> statement sets an interrupt handler
-<{f(x}}> to be called at an unspecified later time.  You can think
-of it as setting an unspecified timer, but it can model any type
-of interrupt.  In addition, the Harmony <{setintlevel L}> statement
-enables or disables interrupts and returns the old interrupt level.
-The model checker tries calling the interrupt handler at every
-possible point in the execution of the Harmony program.
-
 \subsection{Distributed Programming}
@@
 does not want to receive just any message, but a message that satisfies
-certain conditions such as: wait for the next message that is addressed
-to me and has a sequence number larger than 3.
-For this, Harmony supports the <{exists clause}> and allows
+certain conditions such as: \emph{wait for the next message that is addressed
+to me and has a sequence number larger than 3}.
+For this, Harmony supports the <{exists}> clause and allows
 <{when}> clauses to be nested:
@@
 can map dictionaries to dictionaries, and so on.  Unlike many
-object-oriented programming languages, the programmer never
+object-oriented programming languages including Python, the programmer never
 has to specify what it means for two values to be the same or what the
@@
 counter, a map of local variable names to values, and a stack of values.
-Harmony makes no difference between lists and tuples and implements
-both as dictionaries from integers to values.
-So <{[7, 4] = (7, 4) = { 0: 7, 1: 4 }}>.
-Note that <{[7, 4][0] = 7}>, as expected.
-The empty dictionary is represented as either <{()}> or <{[]}>
-(<{{}}> is the empty set).
+% Harmony makes no difference between lists and tuples and implements
+% both as dictionaries from integers to values.
+% So <{[7, 4] = (7, 4) = { 0: 7, 1: 4 }}>.
+% Note that <{[7, 4][0] = 7}>, as expected.
+% The empty dictionary is represented as either <{()}> or <{[]}>
+% (<{{}}> is the empty set).
 
@@
 what happens when two dictionaries have the same key but different values.
-In that case, Harmony will use the maximum value as defined above.
+In that case, Harmony will use the maximum value according to the order
+defined above.
 For dictionary  intersection, Harmony uses the minimum value instead.
@@
 \paragraph{Stop/Go.}  Harmony supports continuations as a first-class
-construct.  To enable this, \texttt{Stop} saves the current context (with
-its program counter incremented) in a specified shared location (usually the
-tail of a queue) and removes the context from the context bag and active set.
-If the context was atomic (which is usually the case), the active set is
-replaced with the domain of the context bag.  Even though its atomic counter
-is not zero, the stopped context is not considered atomic at this time, but
-will resume being atomic once it is scheduled to run.
+construct.  To enable this, \texttt{Stop} saves the current context
+(with its program counter incremented) in a specified shared location
+(such as the tail of a scheduling queue) and removes the context
+from the context bag and active set.  If the context was atomic
+(which is usually the case), the active set is replaced with the
+domain of the context bag.  Even though its atomic counter is not
+zero, the stopped context is not considered atomic at this time,
+but will resume being atomic once it is scheduled to run.
 
@@
 in the state (like keeping the value in a shared list as part of the state).
+Instead, Harmony keeps printed values on the edges between the states
+in the Kripke structure.
 
@@
 
-\emph{Charm}, the Harmony model checker, is an explicit state model
+\emph{Charm}, Harmony's concurrent model checker, is an explicit state model
 checker that computes a \emph{Kripke structure} (essentially, a graph)
@@
 operation.  We call such operations \emph{breakable}.
-Note that this is essentially a form of Partial Order Reduction~\cite{Val91}.
+Note that this optimization is essentially Partial Order
+Reduction~\cite{Val91}.
 
@@
 and the edge indicates the starting context.  During this loop, Charm
-does check if the same state recurs---if so it aborts
+does check if the same state recurs---if so, it aborts
 and reports an infinite loop.
@@
 
-About half of the work of the model checking is trivially parallelizable,
-allowing for significant speedups on multi-core machines.
+In practice, about half of the work of the model checking is trivially
+parallelizable, allowing for significant speedups on multi-core machines.
 
@@
 expression evaluations, such as constant folding and jump threading.
-A more interesting optimization is \emph{dead variable elimination}:
+An unusual optimization is \emph{dead variable elimination}:
 local variables that are no longer in use are automatically removed
@@
 including deadlock, busy waiting, and executions
-that ended in a state that is not considered final according to
+that ended in a state that is not considered final according to the
 behavior DFA if specified.  We detail below how these problems are
@@
 all threads have terminated.  In the case the context bag is not
-empty, we check if there are any non-\emph{eternal} contexts.  If
-so, Harmony reports a \emph{deadlock}.  (Harmony programmers can
+empty, we check if there are any non-\emph{eternal} contexts.
+(Harmony programmers can
 specify that certain threads are eternal, such as server threads
-that operate in an infinite loop waiting for requests.) If there
+that operate in an infinite loop waiting for requests.) If
+so, Harmony reports a \emph{deadlock}. If there
 is no deadlock and a behavior automaton was specified, Harmony also
@@
 strongly connected component for the existence of such situations
-and reports them.
+and reports the shortest example.
 
@@
 \begin{center}
-\includegraphics[width=.8\textwidth]{diners.png}
+\includegraphics[width=.9\textwidth]{diners.png}
 \end{center}
@@
 readers in the critical section.  Unfortunately, checking ${\mathcal
-Spec} \subset {\mathcal Impl}$ does not scale well and can only be
-done for relatively small Kripke structures.
+Spec} \subset {\mathcal Impl}$ does not scale well in general and can
+only be done for relatively small Kripke structures.
 
@@
 Harmony always outputs the shortest such trace.
+For ease of debugging, traces are ordered first by the number of
+context switches they make, and then by the number of states on
+the path.
 However, in order to save memory, Charm does not actually keep
@@
 Harmony includes tools to quickly focus on the root cause of the
-problem without having to dig through the HVM instructions.
+problem without having to look at the HVM instructions themselves.
 In particular, in case of a problem, Harmony outputs a compact
@@
 problematic trace.  You can click on any such box to see the state
-at that time, or you can navigate in various ways (forwards one step,
+at that time and the HVM instruction it is executing,
+or you can navigate in various ways (forwards one step,
 backwards one step, jump to the next line of Harmony code, or jumps
@@
 processor it would require an instruction to stop all other cores.
-Harmony was never designed to be a language for production code.  Because
+Harmony was not designed to be a language for production code.  Because
 it is close to languages such as Python and C, it tends to be fairly
@@
 to the SPIN model checker.  Promela has many of the same features
-as Harmony and also includes explicit message channels.  Promela
+as Harmony and also includes behavior checking.  Promela
 does not have a call stack and does not support procedure calls,

### 2022-01-15 book/paper.tex
Improved concurrent queue test program

@@
 \autoref{fig:qtestpar} shows the test program.  It starts
-<{NOPS}> threads, and each choose to do either a <{put}>
-or a <{get}> operation.  In case of a <{put}> operation,
+<{NOPS}> threads doing either a <{put}>
+or a <{get}> operation.  It selects the fraction of <{put}>
+over <{get}> operations nondetermistically, although avoiding the
+uninteresting case in which there are none of one of them.
+In case of a <{put}> operation,
 the thread enqueues its own name (which is provided as an argument
@@
 
+\begin{figure}
+\begin{code}
+<{:input:../code/RWbusy.hny}>
+\end{code}
+\caption{\harmonylink{code/RWbusy.hny} Busy waiting reader/writer lock}
+\label{fig:rwbusy}
+\end{figure}
+
 A problem with this test is that it does not find a problem
@@
 (Harmony does not yet have a mechanism to do this automatically.)
+
+\autoref{fig:rwbusy} illustrates an implementation of a reader/writer lock
+that uses active busy waiting.  This is an undesirable solution, as it wastes
+CPU cycles.  Harmony complains about this solution.
 

### 2022-01-15 paper/paper.tex

@@
 <{:const:dequeue}>
+<{:const:enq_test}>
+<{:const:deq_test}>
 <{:const:send}>
@@
 
-def thread(self):
-    if choose({ "deq", "enq" }) == "enq":
+def enq_test(self):
     print("call enqueue", self)
@@
     print("done enqueue", self)
-    else:
+
+def deq_test(self):
     print("call dequeue", self)
@@
 
-for i in {1..NOPS}: spawn thread(i)
+nenq = choose {1..NOPS-1}
+for i in {1..nenq}:      spawn enq_test(i)
+for i in {1..NOPS-nenq}: spawn deq_test(i)
 <{:end:}>
@@
 \begin{tabular}{l|c|c|c|c}
-           & <{NOPS=1}> & <{NOPS=2}> & <{NOPS=3}> & <{NOPS=4}> \\ \hline\hline
-queue specification    & 11/0.3 & 110/0.3 &  1,286/0.4 &  16,332/0.9 \\ \hline
-queue impl. w/o check  & 20/0.5 & 383/0.6 &  8,874/0.8 & 261,878/12.7 \\ \hline
-queue impl. with check & 20/0.5 & 425/0.6 & 10,977/1.0 & 437,226/23.6 \\
+           & <{NOPS=2}> & <{NOPS=3}> & <{NOPS=4}> & <{NOPS=5}> \\ \hline\hline
+queue specification    &  30/0.0 &  350/0.0 &  3,385/0.0 &    32,010/0.1 \\ \hline
+queue impl. w/o check  &  90/0.0 & 1905/0.0 & 39,519/0.1 & 1,089,222/5.6 \\ \hline
+queue impl. with check & 101/0.0 & 2566/0.0 & 75,974/0.3 & 3,400,555/22.8 \\
 \end{tabular}
@@
 implementation using locks, and the same implementation combined with
-the behavior automaton generated by the first configuration.}
+the behavior automaton generated by the first configuration.  The
+hardware consisted of a 2.8 GHz Quad-Core Intel Core i7
+(2 hyperthreads per core) with 2,133 MHz LPDDR3 memory.}
 \label{fig:qcmp}

### 2022-01-15 paper/paper.tex
Fixed some interrupt issues

@@
 
-% \renewcommand{\topfraction}{.99}
-% \renewcommand{\bottomfraction}{.99}
-% \renewcommand{\textfraction}{.01}
-% \renewcommand{\floatpagefraction}{.9}
-% \renewcommand{\dbltopfraction}{.99}
-% \renewcommand{\dblfloatpagefraction}{.9}
+\renewcommand{\topfraction}{.99}
+\renewcommand{\bottomfraction}{.99}
+\renewcommand{\textfraction}{.01}
+\renewcommand{\floatpagefraction}{.9}
+\renewcommand{\dbltopfraction}{.99}
+\renewcommand{\dblfloatpagefraction}{.9}
 
@@
 Without any special formalism, Harmony can diagnose a large variety
-of problems:
+of problems in programs:
 \begin{compactitem}
@@
 \caption{The number of states generated by the model checker and the
-execution time (in seconds) for running \autoref{fig:queue}(b).
+model checking time (in seconds) for running \autoref{fig:queue}(b).
 This is done for three configurations, varying the value of <{NOPS}>:
@@
 
-This modularity allows larger systems to be model checked in a way
-that would not be tractable otherwise.
-\autoref{fig:qcmp} shows some measurements taken from running
-Harmony against the test program in \autoref{fig:queue}(b).
-The Harmony compiler allows both constants and modules to be
-specified on the command line, so the test program can easily be
-run with different versions of the <{queue}> module and different
-values for <{NOPS}>, the number of threads spawned to choose and
-perform a queue operation.  The table has a row for three different
-configurations and a column for four different values for <{NOPS}>.
-It is clear that running the test program against the specification
-involves significantly fewer states, and therefore less time, than
-running the test program against an implementation, particularly
-if the behavior of that implementation must be checked as well.
-
-With Harmony, one can check the implementation of a lock separately,
-but use the specification of the lock in the implementation of the
-queue.  Similarly, one can check the implementation of a queue,
-but then use the specification of the queue in a larger application.
-Doing so is essentially a form of \emph{Partial Order Reduction}~\cite{Val91},
-replacing the <{enqueue}> and <{dequeue}> implementations
-by atomic actions.
-
 The Harmony distribution comes with a collection of specifications,
@@
 \autoref{fig:queue}(b) shows how this can be specified in Harmony.
-The test program spawns 4 threads that each select an operation
-using the nondetermistic <{choose}> statement.  In case <{enqueue}>
-is selected, the thread enqueues its identifier <{self}>.  The program
+The test program spawns <{NOPS = 4}> threads, at least one of which
+does an <{enqueue}> operation and at least one of which does a
+<{dequeue}> operation.  A thread that does an <{enqueue}> enqueues
+its identifier <{self}>.  The program
 includes a <{print}> statement at the beginning and end of every queue
@@
 provide the shortest counter-example.
+
+\autoref{fig:qcmp} shows some measurements taken from running
+Harmony against the test program in \autoref{fig:queue}(b).
+The Harmony compiler allows both constants and modules to be
+specified on the command line, so the test program can easily be
+run with different versions of the <{queue}> module and different
+values for <{NOPS}>, the number of threads spawned to
+perform a queue operation.  The table has a row for three different
+configurations and a column for four different values for <{NOPS}>.
+It is clear that running the test program against the specification
+involves significantly fewer states, and therefore less time, than
+running the test program against an implementation, particularly
+if the behavior of that implementation must be checked as well.
+
+In practice, <{NOPS = 4}> is probably sufficient to check the
+correctness of a concurrent queue implementation.  Note, however,
+that if there are more than two operations to check, the model
+checking time will also increase.  In such cases, it usually suffices
+to run all possible pairs of read-only and read-write operations.
+
+Modularity allows larger systems to be model checked in a way
+that would not be tractable otherwise.
+With Harmony, one can check the implementation of a lock separately,
+but use the specification of the lock in the implementation of the
+queue.  Similarly, one can check the implementation of a queue,
+but then use the specification of the queue in a larger application.
+Doing so is essentially a form of \emph{Partial Order Reduction}~\cite{Val91},
+replacing the <{enqueue}> and <{dequeue}> implementations
+by atomic actions.
 

### 2022-01-15 src/charm/charm.c

@@ -199,8 +199,6 @@ static bool onestep(
     // Copy the choice
     hvalue_t choice_copy = choice;
 
-    // bool print_occurred = global->code.instrs[step->ctx->pc].choose;
-    bool print_occurred = true;
     bool choosing = false, infinite_loop = false;
     struct dict *infloop = NULL;        // infinite loop detector
     int loopcnt = 0;
@@ -323,14 +321,8 @@ static bool onestep(
         if (!step->ctx->atomicFlag && (next_instr->breakable || next_instr->log)) {
             if (step->ctx->atomic > 0) {
                 step->ctx->atomicFlag = true;
-                break;
             }
-            if (next_instr->log && !print_occurred) {
-                print_occurred = true;
-            }
-            else {
-                break;
-            }
+            break;
         }
     }
 
@@ -394,6 +386,8 @@ static bool onestep(
         check_invariants(w, next, &w->inv_step);
     }
 
+    int nctxs;
+    hvalue_t *ctxs = value_get(sc->ctxbag, &nctxs);
     minheap_insert(w->results[weight], next);
     return true;
 }
@@ -418,10 +412,12 @@ static void make_step(
     // See if we need to interrupt
     if (sc->choosing == 0 && step.ctx->trap_pc != 0 && !step.ctx->interruptlevel) {
         bool succ = onestep(w, node, sc, ctx, &step, choice, true, false, multiplicity);
-        if (!succ) {
+        if (!succ) {        // ran into an infinite loop
             (void) onestep(w, node, sc, ctx, &step, choice, true, true, multiplicity);
         }
-        // sc = new_alloc(struct state);
+
+        // Allocate another state
+        sc = new_alloc(struct state);
         memcpy(sc, node->state, sizeof(*sc));
         free(step.ctx);
         step.ctx = value_copy(ctx, NULL);
@@ -429,7 +425,7 @@ static void make_step(
 
     sc->choosing = 0;
     bool succ = onestep(w, node, sc, ctx, &step, choice, false, false, multiplicity);
-    if (!succ) {
+    if (!succ) {        // ran into an infinite loop
         (void) onestep(w, node, sc, ctx, &step, choice, false, true, multiplicity);
     }
 
@@ -923,23 +919,23 @@ hvalue_t twostep(
         diff_dump(global, file, oldstate, sc, oldctx, step.ctx, true, false, 0, NULL);
     }
 
-    bool print_occurred = true;         // TODO
[…]

### 2022-01-15 src/charm/graph.h

@@ -81,7 +81,9 @@ struct node {
 struct failure {
     enum fail_type type;
     struct node *node;      // failed state
+    struct node *parent;    // if NULL, use node->parent
     hvalue_t choice;        // choice if any
+    bool interrupt;         // interrupt transition
     hvalue_t address;       // in case of data race
 };
 

### 2022-01-16 paper/paper.tex

@@
 \caption{(a) Lock specification and (b) a spinlock implementation based on
-test-and-set}
+a specification of test-and-set}
 \label{fig:synch}
 \end{figure*}
+
+\section{Harmony Language}
+
+Harmony is heavily based on Python to reduce the learning curve for
+users.  Many of Python's powerful constructs and operators are
+supported, including comprehensions.  Harmony does
+not support classes and instead supports memory allocation and
+pointers in a way that is familiar to C programmers.  While this
+perhaps seems like a step back in language evolution, many
+synchronization algorithms in the literature explicitly consider
+memory accesses, and this has been a bonus for
+analyzing such algorithms with Harmony.  Moreover, the Harmony model
+checker can be used to check for memory leaks or access to dangling
+pointers.
+
+For example, \autoref{fig:ms} shows a reproduction of the well-known
+Michael\&Scott two-lock concurrent queue~\cite{MS96} along with a
+Harmony version of the same algorithm.
+Harmony uses the `<{?}>' instead of the `<{&}>' symbol to take the address
+of a memory location, but inherits the same <{q->field}> notation
+from C to access the given field in the dictionary pointed to by
+pointer <{q}>.  <{q->field}> is shorthand for <{(!q)["field"]}> in
+Harmony, where `<{!}>' is the Harmony operator to dereference a pointer.
+Harmony complains, correctly, about a data race when
+there is a concurrent \texttt{enqueue} and \texttt{dequeue} operation
+on an empty queue---the implementer should ensure that accesses to the
+\texttt{next} field are sequentially consistent.
+
+The Harmony example in \autoref{fig:ms} shows some other language
+features in Harmony that are not in Python.
+The <{let x = e: S}> statement introduces a new bound variable <{x}>
+with value <{e}> that is only valid within the statement block <{S}>.
+Harmony does not have \texttt{return}, \texttt{break}, or
+\texttt{continue} statements to force the programmer into a purely
+nested control flow.  Each method has a variable <{result}> in which
+the return value of the method is stored.
 
@@
 
-\section{Harmony Language}
-
-\subsection{Basics}
-
-Harmony is heavily based on Python to reduce the learning curve for
-users.  Many of Python's powerful constructs and operators are
-supported, including comprehensions.  Harmony does
-not support classes and instead supports memory allocation and
-pointers in a way that is familiar to C programmers.  While this
-perhaps seems like a step back in language evolution, many
-synchronization algorithms in the literature explicitly consider
-memory accesses, and this has been a bonus for
-analyzing such algorithms with Harmony.  Moreover, the Harmony model
-checker can be used to check for memory leaks or access to dangling
-pointers.
-
-For example, \autoref{fig:ms} shows a reproduction of the well-known
-Michael\&Scott two-lock concurrent queue~\cite{MS96} along with a
-Harmony version of the same algorithm.
-Harmony uses the `<{?}>' instead of the `<{&}>' symbol to take the address
-of a memory location, but inherits the same <{q->field}> notation
-from C to access the given field in the dictionary pointed to by
-pointer <{q}>.  <{q->field}> is shorthand for <{(!q)["field"]}> in
-Harmony, where `<{!}>' is the Harmony operator to dereference a pointer.
-Harmony also complains, correctly, about a data race when
-there is a concurrent \texttt{enqueue} and \texttt{dequeue} operation
-on an empty queue---the implementer should ensure that accesses to the
-\texttt{next} field are sequentially consistent.
-
-The Harmony example in \autoref{fig:ms} shows some other language
-features in Harmony that are not in Python.
-The <{let x = e: S}> statement introduces a new bound variable <{x}>
-with value <{e}> that is only valid within the statement block <{S}>.
-Harmony does not have \texttt{return}, \texttt{break}, or
-\texttt{continue} statements to force the programmer into a purely
-nested control flow.  Each method has a variable <{result}> in which
-the return value of the method is stored.
+\subsection{\texttt{atomically}}
+
+Statements in Harmony, including groups of statements, can be tagged with
+the \texttt{atomically} keyword to make them atomic.  What this means
+is that the statement is evaluated atomically by the Harmony virtual
+machine without interleaving of instructions from other threads or interrupt
+handlers.  It is
+a \emph{stronger} condition than holding a lock---a lock still allows
+concurrent execution of multiple threads, but it prevents multiple
+threads from executing in the same critical section.
+The <{atomically}> keyword is intended for <{specification}> more than
+for <{implementation}> of concurrent algorithms.
+
+For example, \autoref{fig:synch}(a) shows the specification for locks
+from the Harmony <{synch}> module that was used in \autoref{fig:ms},
+featuring the Harmony <{when}> statement.
+\autoref{fig:synch}(b) shows an implementation of the <{acquire}>
+method based on a test-and-set primitive, which in turn is
+\emph{specified}.
+Imprecisely, methods that use the <{atomically}> keyword are specifications.
+Important is the notion of modularity that is implied here.
+Harmony can model check the implementation of \autoref{fig:ms} using
+the specification of a lock in \autoref{fig:synch}(a).
+Separately, Harmony can check the correctness of the implementation in
+\autoref{fig:synch}(b) against the specification in
+\autoref{fig:synch}(a).
+
+The Harmony distribution comes with a collection of specifications,
+each often with multiple implementations.  The lock specification
+comes with the spinlock implementation of \autoref{fig:synch}(a), but
+also a ticket lock implementation and one based on continuations and
+a queue of blocked threads.  The distribution also includes various concurrent
+queue implementations, including the lock-free implementation
+of~\cite{MS96}.
+
+\subsection{Checking an implementation}
+
+The Harmony language allows objects such as locks and concurrent queues
+to be specified and implemented in separate programs.  But the programs
+are not \emph{active} as they are just a collection of method definitions.
+A model checker can only check behaviors, but these programs have empty
+behaviors.
+Suppose we want to check the correctness of the concurrent queue
+implementation of \autoref{fig:ms}.  First, we would need a specification
+of a concurrent queue.  Second, we would need a test program that tries
+various concurrent interleavings of <{enqueue}> and <{dequeue}> operations.
+
+\autoref{fig:queue}(a) shows the specification of a concurrent
+<{queue}> module.  A queue is represented as a list.
+The <{enqueue}> method appends a value onto the end of the list,
+while the <{dequeue}> method extracts the head, both as atomic actions.
+
+To test a queue, we want to perform a set of concurrent <{enqueue}>
+and <{dequeue}> operations concurrently.
+\autoref{fig:queue}(b) shows how this can be specified in Harmony.
+The test program spawns <{NOPS = 4}> threads, at least one of which
+does an <{enqueue}> operation and at least one of which does a
+<{dequeue}> operation.  A thread that does an <{enqueue}> enqueues
+its identifier <{self}>.  The program
+includes a <{print}> statement at the beginning and end of every queue
+operation.  Think of the <{print}> statements as \emph{annotations}
+that tell Harmony what the interesting events are with respect to
+queue operations.
 
@@
 
-\subsection{\texttt{atomically}}
-
-Statements in Harmony, including groups of statements, can be tagged with
-the \texttt{atomically} keyword to make them atomic.  What this means
-is that the statement is evaluated atomically by the Harmony virtual
-machine without interleaving of instructions of other threads.  It is
-a \emph{stronger} condition than holding a lock---a lock still allows
-concurrent execution of multiple threads, but it prevents multiple
-threads from executing in the same critical section.
-The <{atomically}> keyword is intended for <{specification}> more than
-for <{implementation}> of concurrent algorithms.
-
-For example, \autoref{fig:synch}(a) shows the specification for locks
-from the Harmony <{synch}> module that was used in \autoref{fig:ms},
-featuring the Harmony <{when}> statement.
-\autoref{fig:synch}(b) shows an implementation of the <{acquire}>
-method based on a test-and-set primitive, which in turn is
-\emph{specified}.
-Imprecisely, methods that use the <{atomically}> keyword are specifications.
-Whether something is a specification or an implementation is in the
-eye of the beholder.
-Important, however, is the notion of modularity that is implied here.
-Harmony can model check the implementation of \autoref{fig:ms} using
-the specification of a lock in \autoref{fig:synch}(a).
-Separately, Harmony can check the correctness of the implementation in
-\autoref{fig:synch}(b) against the specification in
-\autoref{fig:synch}(a).
-
-The Harmony distribution comes with a collection of specifications,
-each often with multiple implementations.  The lock specification
-comes with the spinlock implementation of \autoref{fig:synch}(a), but
-also a ticket lock implementation and one based on continuations and
-a queue of blocked threads.  The distribution also includes various concurrent
-queue implementations, including the lock-free implementation
-of~\cite{MS96}.
-
-\subsection{Checking an implementation}
-
-The Harmony language allows objects such as locks and concurrent queues
-to be specified and implemented in separate programs.  But the programs
-are not \emph{active} as they are just a collection of method definitions.
-A model checker can only check behaviors, but these programs have empty
-behaviors.
-Suppose we want to check the correctness of the concurrent queue
-implementation of \autoref{fig:ms}.  First, we would need a specification
-of a concurrent queue.  Second, we would need a test program that tries
-various concurrent interleavings of <{enqueue}> and <{dequeue}> operations.
-
-\autoref{fig:queue}(a) shows the specification of a concurrent
-<{queue}> module.  A queue is represented as a list.
-The <{enqueue}> method appends a value onto the end of the list,
-while the <{dequeue}> method extracts the head, both as atomic actions.
-
-To test a queue, we want to perform a set of concurrent <{enqueue}>
-and <{dequeue}> operations concurrently.
-\autoref{fig:queue}(b) shows how this can be specified in Harmony.
-The test program spawns <{NOPS = 4}> threads, at least one of which
-does an <{enqueue}> operation and at least one of which does a
-<{dequeue}> operation.  A thread that does an <{enqueue}> enqueues
-its identifier <{self}>.  The program
-includes a <{print}> statement at the beginning and end of every queue
-operation.  Think of the <{print}> statements as \emph{annotations}
-that tell Harmony what the interesting events are with respect to
-queue operations.
-
 The test program is similar to what a programmer might have written
@@
 the specification program.  In case the implementation has a behavior
-that is not allowed by the specificatino, the model checker can
+that is not allowed by the specification, the model checker can
 provide the shortest counter-example.
 
-\autoref{fig:qcmp} shows some measurements taken from running
+\autoref{fig:qcmp} shows measurements taken from running
 Harmony against the test program in \autoref{fig:queue}(b).
@@
 checking time will also increase.  In such cases, it usually suffices
-to run all possible pairs of read-only and read-write operations.
+to run all pairs of read-only and read-write operations.
 
@@
 only shared variable is a set of messages representing the network.
+From a model checking perspective, it is convenient that message and
+timer handling can be performed atomically, as the only interaction
+between threads is through the network variable.
 A message is sent by atomically adding a message to the network:
@@
 
-\paragraph{Stop/Go.}  Harmony supports continuations as a first-class
-construct.  To enable this, \texttt{Stop} saves the current context
+\paragraph{Stop/Go.} \texttt{Stop} saves the current context
 (with its program counter incremented) in a specified shared location
@@
 in the state (like keeping the value in a shared list as part of the state).
-Instead, Harmony keeps printed values on the edges between the states
-in the Kripke structure.
+Instead, Harmony keeps printed values on the edges between the states.
 
@@
 and reports the shortest example.
-
-\begin{figure}
-\begin{center}
-{\footnotesize
-\begin{code}
-<{:inline:}>
-from synch import Lock, acquire, release
-
-const N = 5
-forks = [Lock(),] * N
-
-def diner(which):
-    let left, right = (which, (which + 1) % N):
-        while choose({ False, True }):
-            acquire(?forks[left])
-            acquire(?forks[right])
-            # dine
-            release(?forks[left])
-            release(?forks[right])
-
-for i in {0..N-1}: spawn diner(i)
-<{:end:}>
-\end{code}
-}
-\end{center}
-\caption{Dining Philosophers in Harmony}
-\label{fig:diners}
-\end{figure}
-
-\begin{figure*}
-\begin{center}
-\includegraphics[width=.9\textwidth]{diners.png}
-\end{center}
-\caption{The HTML output of running the Dining Philosophers program}
-\label{fig:dinersHTML}
-\end{figure*}
 
@@
 the required information during the construction of the Kripke
-structure.  In particular, its edges only contain the context
-and the choice if any.  Thus, in order to reconstruct the
+structure.  In particular, its edges only contain the context,
+whether an interrupt occurred, and the choice if any.
+Thus, in order to reconstruct the
 counter-example, Charm runs the HVM one more time, this time
@@
 
+\begin{figure}
+\begin{center}
+{\footnotesize
+\begin{code}
+<{:inline:}>
+from synch import Lock, acquire, release
+
+const N = 5
+forks = [Lock(),] * N
+
+def diner(which):
+    let left, right = (which, (which + 1) % N):
+        while choose({ False, True }):
+            acquire(?forks[left])
+            acquire(?forks[right])
+            # dine
+            release(?forks[left])
+            release(?forks[right])
+
+for i in {0..N-1}: spawn diner(i)
+<{:end:}>
+\end{code}
+}
+\end{center}
+\caption{Dining Philosophers in Harmony}
+\label{fig:diners}
+\end{figure}
+
+\begin{figure*}
+\begin{center}
+\includegraphics[width=\textwidth]{diners.png}
+\end{center}
+\caption{The HTML output of running the Dining Philosophers program}
+\label{fig:dinersHTML}
+\end{figure*}
+
 \section{Debugging}
@@
 it could find.  However, this counter-example is at the level of the HVM,
-not the Harmony program that the programmer wrote.  It can be hard to
-dig through a list of HVM instructions.
+not the Harmony program that the programmer wrote.  Without help it can be
+difficult to dig through a list of HVM instructions.
 Harmony includes tools to quickly focus on the root cause of the
@@
 identifier.  Also, Harmony's support for modular model checking
-allows larger systems to be built from smaller ones.
-
-The Harmony language, at least currently, is not usable for production
-code.  One reason for this is the <{atomically}> clause.  On a multi-core
-processor it would require an instruction to stop all other cores.
-Harmony was not designed to be a language for production code.  Because
-it is close to languages such as Python and C, it tends to be fairly
+allows implementing larger systems using the specifications of smaller ones.
+
+Currently there is no compiler that generates production code from
+a Harmony program.
+Because it is close to languages such as Python and C, it tends to be fairly
 easy to hand-translate code developed in Harmony to a desired target
@@
 introduce a bug that did not exist in the Harmony code.
+A Harmony compiler would have to be able to recognize constructs such as
+the test-and-set specification in \autoref{fig:synch}(b) and generate
+a hardware test-and-set instruction.
 
@@
 than the gap between Harmony programs and their TLA+ representations,
-making PlusCal more amenable to formal verification.  On the other
+making PlusCal programs more amenable to formal verification.  On the other
 hand, working effectively with PlusCal does require significant

### 2022-01-16 paper/paper.tex
Fixed interrupt handling

@@
 \emph{Charm}, Harmony's concurrent model checker, is an explicit state model
-checker that computes a \emph{Kripke structure} (essentially, a graph)
-of states reachable
+checker that computes a \emph{Kripke structure}~\cite{Kripke1963}
+(essentially, a graph) of states reachable
 from the initial state in a breadth-first manner, stopping only if
@@
 access the same shared data location, at least one of which is a
-\texttt{Store}, \texttt{Stop}, or \texttt{Del} instruction.  Harmony
+\texttt{Store}, \texttt{Stop}, or \texttt{Del} instruction.  Data races
+can also occur in a single thread that experiences an interrupt.  Harmony
 only considers this a problem it at least one of those threads is

### 2022-01-16 src/charm/code.h

@@ -9,7 +9,7 @@
 struct instr_t {
     struct op_info *oi;
     const void *env;
-    bool choose, load, store, del, log, breakable;
+    bool choose, load, store, del, retop, print, breakable;
 };
 
 struct code_t {

### 2022-01-16 paper/paper.tex
Latest

@@
 the return value of the method is stored.
+
+In order to make deadlock detection possible, Harmony programs are
+required to eventually terminate under fair selection of its
+nondetermistic choices.  In other words, from each state there
+should always exist an extension of the execution in which the
+Harmony program terminates.  Harmony supports \emph{eternal}
+threads---threads that never terminate---for example to model
+servers.  A Harmony program is considered terminated if it is in a
+state in which all its remaining threads are eternal and \emph{blocked}.
+A thread is considered blocked if it can only update the shared
+state after another thread has updated the shared state (for example,
+a thread waiting on a spinlock).
 
@@
 all threads have terminated.  In the case the context bag is not
-empty, we check if there are any non-\emph{eternal} contexts.
-(Harmony programmers can
-specify that certain threads are eternal, such as server threads
-that operate in an infinite loop waiting for requests.) If
-so, Harmony reports a \emph{deadlock}. If there
+empty, we check if there are any non-\emph{eternal} contexts or
+contexts that are not blocked.
+If so, Harmony reports a \emph{deadlock}. If there
 is no deadlock and a behavior automaton was specified, Harmony also
@@
 
-A thread is considered \emph{blocked} if it can only make progress
-(modify the state) if another thread makes a transition first.
-This is not usually a problem: a thread waiting on a lock in a spinloop
-is a common way of synchronizing threads, and Harmony does not report
-such issues.  We call this \emph{passive busy waiting} as the thread
-does not modify the state.
+As mentioned above, a thread is considered blocked if it can only
+modify the shared state if another thread updates the shared state
+first.  This is not usually a problem: a thread waiting on a lock
+in a spinloop is a common way of synchronizing threads, and Harmony
+does not report such issues.  We call this \emph{passive busy
+waiting} as the thread does not modify the state.
 

### 2022-01-16 paper/paper.tex

@@
 \documentclass[twocolumn]{article}
-\usepackage{fullpage}
+\usepackage{usenix2019}
 \usepackage{graphicx}
@@
 \usepackage{xcolor}
-\usepackage[colorlinks=true, linkcolor=blue, urlcolor=blue, citecolor=violet]{hyperref}
+% \usepackage[colorlinks=true, linkcolor=blue, urlcolor=blue, citecolor=violet]{hyperref}
+\usepackage{hyperref}
 
@@
 Harmony's unique combination of features are a familiar and easy
-to learn high-level programming language, the ability to concisely
-specify an interface, the ability to compare behaviors of specification
-and implementation, the ability to modularly model check programs,
-and support for dynamic memory allocation, continuations, and
-interrupts.  In addition, Harmony finds various common problems
-automatically without a complicated property specification language,
-such as Harmony's ability to find instances of undesirable busy
-waiting.
+to learn high-level and highly expressive programming language, the
+ability to concisely specify an interface, the ability to compare
+behaviors of specification and implementation, the ability to
+modularly model check programs, and support for dynamic memory
+allocation, continuations, and interrupts.  In addition, Harmony
+finds various common problems automatically without a complicated
+property specification language, such as Harmony's ability to find
+instances of undesirable busy waiting.
 
@@
 users.  Many of Python's powerful constructs and operators are
-supported, including comprehensions.  Harmony does
-not support classes and instead supports memory allocation and
-pointers in a way that is familiar to C programmers.  While this
-perhaps seems like a step back in language evolution, many
-synchronization algorithms in the literature explicitly consider
-memory accesses, and this has been a bonus for
-analyzing such algorithms with Harmony.  Moreover, the Harmony model
-checker can be used to check for memory leaks or access to dangling
-pointers.
+supported, including sets, lists, tuples, dictionaries, and
+comprehensions.  In certain cases Harmony goes further than Python,
+for example by allowing sets of dictionaries (Harmony does not have
+so-called ``unhashable types'').
+
+Harmony does not
+support classes and instead supports memory allocation and pointers
+in a way that is familiar to C programmers.  While this perhaps
+seems like a step back in language evolution, many synchronization
+algorithms in the literature explicitly consider memory accesses,
+and this has been a bonus for analyzing such algorithms with Harmony.
+Moreover, the Harmony model checker can be used to check for memory
+leaks or access to dangling pointers.
 

### 2022-01-16 paper/paper.tex
More improvements to the paper

@@
 
-\title{Harmony: A Verifiable Systems Programming Language}
+\title{Harmony: A Systems Programming Language and Model Checker}
 \author{Robbert van Renesse \and William Ma \and Kevin Sun \and Anthony Yang}
@@
 \item Safety violations due to assertions failing, such as an assertion to ensure mutual exclusion.
-\item Liveness violations (deadlock).
+\item Liveness violations such as caused by deadlock or infinite loops.
 \item Faulty behaviors: implementation has behaviors not seen in the specification, such as violations of linearizability~\cite{HW90}.
@@
 
-Harmony programs compile to a virtual machine.  The Harmony virtual
+Harmony programs compile to a virtual machine.  The virtual machine
+instructions are the units of atomicity.  The Harmony virtual
 machine is formally specified using TLA+~\cite{Lamport02}, allowing
@@
 
-Harmony has been used to verify the correctness of various lock
+Harmony has been used to model check the correctness of various lock
 implementations, monitors and condition variables (both Hoare and Mesa),
@@
 ones, and barrier synchronization implementations.  Harmony has also been
-used to verify a variety of distributed algorithms such as two-phase
+used to model check a variety of distributed algorithms such as two-phase
 commit, chain replication, various consensus algorithms such as Paxos,
@@
 
+A \emph{blocked} thread is waiting on a condition
+(for example, a lock becoming available) using the <{when}> statement.
+This is implemented in the Harmony virtual machine
+by executing in a loop until the condition holds, without changing the
+shared state, and thus the thread can only be unblocked by another thread
+changing the state.
+
+\emph{Deadlock} is the situation in which threads are waiting for
+one another.
 In order to make deadlock detection possible, Harmony programs are
@@
 servers.  A Harmony program is considered terminated if it is in a
-state in which all its remaining threads are eternal and \emph{blocked}.
-A thread is considered blocked if it can only update the shared
-state after another thread has updated the shared state (for example,
-a thread waiting on a spinlock).
+state in which all its remaining threads are eternal and blocked.
 
@@
 
-Statements in Harmony, including groups of statements, can be tagged with
-the \texttt{atomically} keyword to make them atomic.  What this means
-is that the statement is evaluated atomically by the Harmony virtual
-machine without interleaving of instructions from other threads or interrupt
-handlers.  It is
+A Harmony program compiles to instructions for a virtual machine.
+Those instructions are normally the units of atomicity, but threads
+can be temporarily placed in \emph{atomic mode} to prevent instructions
+interleaving with those of other threads or interrupt handlers.
+At the Harmony language level, even simple statements should not be
+assumed to be atomic.
+However, statements---including groups of statements---can be tagged with
+the \texttt{atomically} keyword to make them execute as if they were
+a single atomic instruction.
+It is
 a \emph{stronger} condition than holding a lock---a lock still allows
@@
 concurrent queue, but with the inclusion of the DFA.  The model
-checker ends up verifying the Cartesian product of the implementation
+checker ends up executing the Cartesian product of the implementation
 and the DFA and ensures that every behavior of the implementation
@@
 For any two values of the same type, Harmony defines a deterministic order.
-For example, for two sets, the values in the sets are first ordered and
-then lexicographically compared, so that <{{ 1, 2 } < { 1, 3 }}>.
+For example, for two sets, the sets are first converted into ordered lists
+and then lexicographically compared, so that <{{ 1, 2 } < { 1, 3 }}>.
 (Note that, unlike Python, the $<$ operator is a total order and
@@
 
-Except for <{choose}>, any operation on values is deterministic and
-defined in the formal specification of the Harmony virtual machine.
+Except for <{choose}>, any operation on values is a deterministic
+function and unequivocally defined in the formal specification of the
+Harmony virtual machine.
 For example, like Python, Harmony supports merging of dictionaries
@@
 then there are now two automata: the specified automaton
-${\mathcal Spec}$ that was presented to Charm as an input as
+${\mathcal S}$ that was presented to Charm as an input as
 part of the \emph{specification}, and the generated automaton
-${\mathcal Impl}$ that was generated by the given Harmony
+${\mathcal I}$ that was generated by the given Harmony
 \emph{implementation}.  Harmony already checks on-the-fly whether
-${\mathcal Impl} \subseteq {\mathcal Spec}$, that is, that every
+${\mathcal I} \subseteq {\mathcal S}$, that is, that every
 behavior (sequence of symbols) of the implementation is also a
 behavior of the specification.  Harmony also checks if
-${\mathcal Spec} \subset {\mathcal Impl}$ and, if so,
+${\mathcal S} \subset {\mathcal I}$ and, if so,
 produces a warning.
 Although not currently performed automatically, it is possible to compute
-${\mathcal Spec}~\backslash~{\mathcal Impl}$
+${\mathcal S}~\backslash~{\mathcal I}$
 to generate an automaton for all behaviors allowed by the
@@
 readers in the critical section.  Unfortunately, checking ${\mathcal
-Spec} \subset {\mathcal Impl}$ does not scale well in general and can
+S} \subset {\mathcal I}$ does not scale well in general and can
 only be done for relatively small Kripke structures.

### 2022-01-16 paper/paper.tex
More improvements to paper

@@
 but then use the specification of the queue in a larger application.
-Doing so is essentially a form of \emph{Partial Order Reduction}~\cite{Val91},
+Doing so is essentially a form of
+\emph{Partial Order Reduction}~\cite{Owicki75,Lipton75,Val91},
 replacing the <{enqueue}> and <{dequeue}> implementations

### 2022-01-16 paper/paper.tex
Latest

@@
 A message is sent by atomically adding a message to the network:
-<{def send(m): atomically network |= {m}}> (`<{|}>' is the union
-operator in Python).
+<{def send(m): atomically network |= {m}}> (`<{|}>' is the symbol for
+the union operator in Python).
 To receive a message, a program can use the Harmony <{when}>

### 2022-01-16 paper/paper.tex
Latest
Fixed active busy wait

@@
 \end{center}
-\caption{FIFO lock using continuation}
+\caption{FIFO lock using continuations}
 \label{fig:continuation}
@@
 
-A thread is considered \emph{actively busy waiting} if it cannot
+A thread is considered \emph{actively busy waiting} if it is in a
+loop updating the shared state, but cannot
 escape its strongly connected component without another thread
-updating the shared state first.  A common example of this is a
+also updating the shared state.  A common example of this is a
 synchronization algorithm that checks for a condition in a loop,

### 2022-01-18 paper/paper.tex

@@
 and many more.
-\item Safety violations due to assertions failing, such as an assertion to ensure mutual exclusion.
-\item Liveness violations such as caused by deadlock or infinite loops.
+\item Safety violations: Harmony supports both assertions that must hold in a particular location and general invariants.
+\item Liveness violations: not reaching a final state, often due to deadlock or an infinite loop.
 \item Faulty behaviors: implementation has behaviors not seen in the specification, such as violations of linearizability~\cite{HW90}.
@@
 of an interrupt handler.
-\item Memory leaks.
+\item Memory leaks: memory unintentionally allocated in a final state.
 \end{compactitem}
@@
 
-Harmony's unique combination of features are a familiar and easy
-to learn high-level and highly expressive programming language, the
-ability to concisely specify an interface, the ability to compare
-behaviors of specification and implementation, the ability to
-modularly model check programs, and support for dynamic memory
-allocation, continuations, and interrupts.  In addition, Harmony
-finds various common problems automatically without a complicated
-property specification language, such as Harmony's ability to find
-instances of undesirable busy waiting.
+Harmony's unique combination of features are
+\begin{compactitem}
+\item a familiar and easy-to-learn high-level programming language.
+\item the ability to concisely specify an interface.
+\item the ability to compare behaviors of specification and implementation.
+\item the ability to modularly model check programs.
+\item support for dynamic memory allocation.
+\item support for continuations.
+\item support for interrupts.
+\item automatic diagnosis of various common problems such as
+undesirable busy waiting.
+\end{compactitem}
 
@@
 support classes and instead supports memory allocation and pointers
-in a way that is familiar to C programmers.  While this perhaps
+in a way that is familiar to C programmers.  While perhaps this
 seems like a step back in language evolution, many synchronization
@@
 one another.
-In order to make deadlock detection possible, Harmony programs are
+In order to make deadlock detection possible, a Harmony program is
 required to eventually terminate under fair selection of its
@@
 
-To test a queue, we want to perform a set of concurrent <{enqueue}>
-and <{dequeue}> operations concurrently.
-\autoref{fig:queue}(b) shows how this can be specified in Harmony.
-The test program spawns <{NOPS = 4}> threads, at least one of which
-does an <{enqueue}> operation and at least one of which does a
-<{dequeue}> operation.  A thread that does an <{enqueue}> enqueues
-its identifier <{self}>.  The program
-includes a <{print}> statement at the beginning and end of every queue
-operation.  Think of the <{print}> statements as \emph{annotations}
-that tell Harmony what the interesting events are with respect to
-queue operations.
+To test a concurrent queue, we need to perform a set of <{enqueue}>
+and <{dequeue}> operations concurrently.  \autoref{fig:queue}(b)
+shows how this can be specified in Harmony.  The test program spawns
+<{NOPS = 4}> threads, at least one of which does an <{enqueue}>
+operation and at least one of which does a <{dequeue}> operation.
+A thread that does an <{enqueue}> enqueues its identifier <{self}>.
+The program includes a <{print}> statement at the beginning and end
+of every queue operation.  Think of the <{print}> statements as
+\emph{annotations} that tell Harmony what the interesting events
+are with respect to queue operations.  We will use the output of
+those <{print}> statements to compare the behavior of an implementation
+with that of its specification.
 
@@
 times and look at the output in order to see if it looks reasonable.
-Doing so is difficult and error prone, as the output tends
-to be complex and in the mess of many lines of output it is easy
-to overlook problems such as two threads dequeuing the same item
-or items being dequeued in the wrong order.  The output of concurrent
-print statement can be mangled together.  Not only that, but the
+Doing so is difficult and error prone, as the output tends to be
+complex and---in the mess of many lines of output---it is easy to
+overlook problems such as two threads dequeuing the same item or
+items being dequeued in the wrong order.  Not only that, but the
 programmer would need to run this test program many times and check
-the output each time.  It even is hard to write a program that
-checks the output automatically.
+the output each time.  It is hard to write a program that checks
+the output automatically as it is difficult to characterize what
+correct outputs are.
 
@@
 \begin{compactitem}
-\item \emph{basic operations}: do a simple operation to a context
-such as pushing a value onto its stack, and increase its program
-counter by one.
-\item \emph{jumps}: to implement loops.
+\item \emph{basic operations}: a deterministic operation that only
+involves a single context and no shared state, such as pushing a value
+onto the stack or increment the program counter.
+\item \texttt{Frame/Return}: method start/end.
 \item \texttt{Choose}: nondetermistic selection from a set.
-\item \texttt{Apply}: to call a method \emph{or} to index into a dictionary.
-\item \texttt{Frame/Return}: method start/stop.
-\item \texttt{Spawn}: create a new thread.
-\item \texttt{AtomicInc/AtomicDec}: modify the atomic counter of the context.
-\item \texttt{SetIntLevel}: enable/disable interrupts.
-\item \texttt{Stop/Go}: manipulating continuations.
-\item \texttt{Assert}: check a value.
-\item \texttt{Print}: add a value to the print log.
+\item \texttt{Apply}: method call \emph{or} index into a dictionary.
+\item \texttt{Spawn}: thread creation.
+\item \texttt{AtomicInc/AtomicDec}: atomic section enter/leave.
+\item \texttt{SetIntLevel}: interrupt enable/disable.
+\item \texttt{Stop/Go}: continuation save/restart.
+\item \texttt{Assert}: value check.
+\item \texttt{Print}: output.
 \end{compactitem}
@@
 
-The little orange boxes represent individual HVM instructions of the
-problematic trace.  You can click on any such box to see the state
-at that time and the HVM instruction it is executing,
-or you can navigate in various ways (forwards one step,
-backwards one step, jump to the next line of Harmony code, or jumps
-forwards or backwards to a context switch between threads).  The actual
-HVM code is in the lower left of the output.
+The little orange boxes represent individual HVM instructions of
+the problematic trace.  The programmer can click on any such box
+to see the state at that time and the HVM instruction it is executing,
+or can navigate in various ways (forwards one step, backwards
+one step, jump to the next line of Harmony code, or jumps forwards
+or backwards to a context switch between threads).  The actual HVM
+code is in the lower left of the output.
 
@@
 techniques.  The resulting HVM has a TLA+ specification.  PlusCal,
-instead, \emph{trans\-piles} to TLA+.  Informally, the gap between
-PlusCal programs and their TLA+ representations is much smaller
-than the gap between Harmony programs and their TLA+ representations,
-making PlusCal programs more amenable to formal verification.  On the other
-hand, working effectively with PlusCal does require significant
-knowledge of TLA+, increasing the learning curve.
+instead, \emph{trans\-piles} to TLA+ (source-to-source translation).
+Informally, the gap between PlusCal programs and their TLA+
+representations is much smaller than the gap between Harmony programs
+and their TLA+ representations, making PlusCal programs more amenable
+to formal verification.  On the other hand, working effectively
+with PlusCal does require significant knowledge of TLA+, increasing
+the learning curve.
 

### 2022-01-19 paper/paper.tex
Improved description of "breakable" instructions

@@
 \begin{compactitem}
-\item \emph{basic operations}: a deterministic operation that only
+\item \emph{local operations}: a deterministic operation that only
 involves a single context and no shared state, such as pushing a value
 onto the stack or increment the program counter.
+\item \texttt{Load/Store/Del}: an operation that modifies the shared state.
+\item \texttt{Print}: output.
+\item \texttt{Choose}: nondetermistic selection from a set.
 \item \texttt{Frame/Return}: method start/end.
-\item \texttt{Choose}: nondetermistic selection from a set.
 \item \texttt{Apply}: method call \emph{or} index into a dictionary.
@@
 \item \texttt{Assert}: value check.
-\item \texttt{Print}: output.
 \end{compactitem}
@@
 operations, thread interleaving, and interrupts.
+Each of these cause branching in the state graph and contribute
+to the combinatorial explosion of states.
 When it comes to interleaving, an important observation is that
 local operations on the state of a context (updates to its program
-counter, local variables, or stack) are not visible to other threads.
+counter, local variables, or stack) are not visible to other threads
+and therefore two such operations executed by different threads
+commute.
+Operations that do not commute that way are Load, Store, and Del
+instructions, as well as the Print instruction.  We call those
+instructions \emph{breakable}.
 An important optimization that significantly
 reduces the size of the set of generated states is to treat a
-sequence of operations as atomic if only its first operation is an
-operation that accesses the shared state (\texttt{Load}, \texttt{Store},
-\texttt{Del}, or \texttt{Stop}) or a \texttt{Choose} or \texttt{Print}
-operation.  We call such operations \emph{breakable}.
-Note that this optimization is essentially Partial Order
-Reduction~\cite{Val91}.
+sequence of operations as atomic if only its first instruction
+is breakable.  Given two such sequences $\sigma_1$ and $\sigma_2$,
+the execution of any interleaving would either be equivalent to
+$\sigma_1\sigma_2$ or $\sigma_2\sigma_1$.
+This is another example of Partial Order Reduction~\cite{Val91}.
 
@@
 breakable.  Only the resulting state is added to the Kripke structure,
-and the edge indicates the starting context.  During this loop, Charm
+and the resulting edge is labeled with the starting context.
+During this loop, Charm
 does check if the same state recurs---if so, it aborts
@@
 
-In practice, about half of the work of the model checking is trivially
-parallelizable, allowing for significant speedups on multi-core machines.
+% In practice, about half of the work of the model checking is trivially
+% parallelizable, allowing for significant speedups on multi-core machines.
 

### 2022-01-21 book/paper.tex
update install section of textbook

@@
 folder under the name listed in the caption of the example.
+If you need to, you can download the sources separately from
+\url{https://harmony.cs.cornell.edu/sources.zip}.
 In this case, the file \texttt{code/hello1.hny} contains the
@@
 
-There are currently three options for using Harmony:
+There are two supported options for using Harmony:
 
@@
 \item You can download and install \emph{HarmonyLang}, a VSCode plug-in for Harmony;
-\item You can run Harmony programs in the cloud using \emph{Harmony IDE}.
 \end{enumerate}
 
-Below you can find more information on each of these options.
+In the following sections, you can find more information on each of these options.
+
+For each option, you will need to have Python version 3.6 or higher and a C compiler such as
+\texttt{gcc}, \texttt{cc}, or \texttt{clang}.
 
@@
 
-You can get the latest released developer version of Harmony by
-downloading from \url{harmony.cs.cornell.edu}.  It includes the latest
-bug fixes and features at any time.  Harmony is developed and maintained
-by the author on both MacOSX and Linux, and so these are the preferred
-platforms at this time.  A Windows version is also available, but may not
-be the latest version and is currently not as well supported.
-
-When you download the \texttt{.zip} file, you will get the following
-files:
-
-\begin{itemize}
-\item \texttt{README.txt}: installation documentation;
-\item \texttt{archive.xml}: portable compressed archive of the code base;
-\item \texttt{install.py}: Python3 program to install and update the code base.
-\end{itemize}
-
-First place this directory (folder) where you would like it---the
-\texttt{Downloads} folder is at best a good temporary place.  You can put
-the directory in your home directory, for example.
-
-Installation requires Python3 and a 64-bit C compiler.  The developer uses
-recent versions of \texttt{gcc} and \texttt{clang} for development.
-Run \texttt{python3 install.py} to install the code base.
-It extracts files from archive.xml and installs them in the current
-directory.
-It will try to compile the model checker using \texttt{gcc}.
-If that fails under Windows, it will install an executable that
-is pre-compiled on a Windows 10 machine.  If these options do not work,
-you can compile the file \texttt{charm.c} by hand using a 64-bit C
-compile and place the output in \texttt{charm.exe}.
-
-After installation, there is a \texttt{harmony} file for use under
-MacOSX and Linux, and a \texttt{harmony.bat} file for use under Windows.
-You should be able to run \texttt{harmony --help} in any of these
-environments.
-
-If you would like to run \texttt{harmony} from any directory, you have to
-add the current directory to your search path.  Under MacOSX and Linux,
-you will have to set the \texttt{PATH} environment variable.
-See \url{http://www.linux-migration.org/ch02s06.html} for more information.
-Under Windows, search "Edit environment variables" in the search bar.
-You can add the directory either to the \texttt{Path} associated with
-your account or to the system \texttt{Path}.
-If you do not install \texttt{harmony} in your search path, you may have
-to run \texttt{./harmony} in the installation directory instead of
-just \texttt{harmony}.
-
-The installation directory will have the following subdirectories:
-
-\begin{itemize}
-\item \texttt{code}: contains all the code examples from this book;
-\item \texttt{modules}: contains the Harmony modules;
-\item \texttt{python}: contains the Python examples from the book.
-\end{itemize}
-
-For example, you can try \texttt{harmony code/Diners.hny} to run the
-Dining Philosophers code.
-Harmony currently produces three output files:
-\begin{itemize}
-\item \texttt{code/Diners.hvm}: the bytecode in JSON format;
-\item \texttt{code/Diners.hco}: the output of the model checker in JSON format;
-\item \texttt{code/Diners.htm}: the model checker output converted to HTML format.
-\end{itemize}
-
-You are probably only interested to see the last one, which you should be
-able to view in any web browser of your choice, including Safari, Chrome, Edge, or Firefox.
-
-You can see if you are running the latest version of Harmony at any
-time by running \texttt{Python3 install.py --check}.  If you would like
-to update your installation, run \texttt{Python3 install.puy --update}.
+You can get the latest released version of Harmony using the
+\texttt{pip} package manager, which is bundled with distributions
+of Python3.X.
+
+On a command line/terminal, run \texttt{pip install harmony-model-checker}.
+If you are on a Windows system, then you may encounter an error message
+like \texttt{error: Microsoft Visual C++ 14.0 or greater is required...} along with a link.
+If so, follow the link to install "Microsoft C++ Build Tools".
+
+After installing Harmony, run the command \texttt{harmony} on the command line to confirm
+that it is installed.
+
+If the \texttt{harmony} command cannot be found, it is possible that the Python
+scripts directory is not included in your environment PATH. This issue may also
+have appeared as a warning message when installing \texttt{harmony-model-checker}.
+After adding the directory to the PATH and restarting the command line, you should
+be able to run \texttt{harmony}.
+
+To update Harmony, run \texttt{pip install --upgrade harmony-model-checker}.
 
@@
 
-The VSCode plug-in, developed by Kevin Sun and Anthony Yang,
-is available from the VSCode market place: \url{https://marketplace.visualstudio.com/items?itemName=kevinsun-dev-cornell.harmonylang}.
-This page comes with installation documentation.  VSCode also sports a
-wonderful animator for Harmony output.
-
-HarmonyLang is regularly updated to include the latest Harmony distribution.
-
-\section{Harmony IDE for the cloud}
-
-\texttt{Harmony IDE} is an Integrated Development Environment for Harmony,
-also developed by Kevin Sun and Anthony Yang,
-that runs in the cloud.  An important advantage of using \texttt{Harmony IDE}
-is that you do not have to install anything.  You can even run in on your
-smartphone.  However, since Harmony programs can use significant compute
-resources, \texttt{Harmony IDE} must put a limit on the amount of compute
-resources you can use.  So, while \texttt{Harmony IDE} is currently a great
-way to try out Harmony, if you become a serious user of Harmony you
-will probably want to install either the command-line version or
-\texttt{HarmonyLang}.
+The VSCode plugin HarmonyLang, developed by Kevin Sun and Anthony Yang, is available
+from the VSCode marketplace: \url{https://marketplace.visualstudio.com/items?itemName=kevinsun-dev-cornell.harmonylang}.
+This page comes with installation documentation. VSCode also sports a wonderful animator for Harmony output.
+
+The VSCode extension uses Python and pip internally to install and build Harmony automatically. However,
+if some requirements are missing, such as Python or a C compiler, then a error message pop-up will appear
+describing the issue.
+
+The VSCode automatically updates to the latest version of Harmony when the extension is activated.
+However, the update can also be directly prompted by using the Command Palette and running "Install Harmony".
 

### 2022-01-22 book/paper.tex
update installation documentation

@@
 like \texttt{error: Microsoft Visual C++ 14.0 or greater is required...} along with a link.
-If so, follow the link to install "Microsoft C++ Build Tools".
+If so, follow the link to install "Microsoft C++ Build Tools" and the "Desktop development with C++" workload.
 
 After installing Harmony, run the command \texttt{harmony} on the command line to confirm
-that it is installed.
+that it is successfully installed.
 

### 2022-01-23 book/paper.tex
Removed installation instructions from book

--- 2022-01-15 book/paper.tex
@@
 an appendix on convenient Harmony modules (\autoref{ap:module}),
-an appendix that explains how Harmony works (\autoref{ap:howitworks}),
-and
-an appendix on installing and running Harmony (\autoref{ap:install}).
+and an appendix that explains how Harmony works (\autoref{ap:howitworks}),
 
@@
 \autoref{fig:helloworld} shows the program and the corresponding output.
-From the command line, you can run it as follows:
+After installation (see \url{https://harmony.cs.cornell.edu}), you
+can run it as follows from the command line:
 \begin{code}
@@
 
+\begin{comment}
 \chapter{Installing and Running Harmony}\label{ap:install}
@@
 \texttt{HarmonyLang}.
+\end{comment}
 

### 2022-01-23 book/paper.tex

--- 2022-01-22 book/paper.tex
@@
 
-The VSCode automatically updates to the latest version of Harmony when the extension is activated.
-However, the update can also be directly prompted by using the Command Palette and running "Install Harmony".
+The VSCode plug-in, developed by Kevin Sun and Anthony Yang,
+is available from the VSCode market place: \url{https://marketplace.visualstudio.com/items?itemName=kevinsun-dev-cornell.harmonylang}.
+This page comes with installation documentation.  VSCode also sports a
+wonderful animator for Harmony output.
+
+HarmonyLang is regularly updated to include the latest Harmony distribution.
+
+\section{Harmony IDE for the cloud}
+
+\texttt{Harmony IDE} is an Integrated Development Environment for Harmony,
+also developed by Kevin Sun and Anthony Yang,
+that runs in the cloud.  An important advantage of using \texttt{Harmony IDE}
+is that you do not have to install anything.  You can even run in on your
+smartphone.  However, since Harmony programs can use significant compute
+resources, \texttt{Harmony IDE} must put a limit on the amount of compute
+resources you can use.  So, while \texttt{Harmony IDE} is currently a great
+way to try out Harmony, if you become a serious user of Harmony you
+will probably want to install either the command-line version or
+\texttt{HarmonyLang}.
+\end{comment}
 

### 2022-01-28 book/paper.tex
describing "save" expression in book

@@
 
-This causes the context of the thread to be saved in <{s}> and
+This causes the context of the thread to be saved in <{!s}> and
 the thread to be no longer running.
@@
 \begin{code}
-<{go c r}>
-\end{code}
-
-Here <{c}> is a context and <{r}> is a Harmony value.
-It causes a thread with context <{c}> to be added to the state that has
-just executed the <{stop}>
+<{go s r}>
+\end{code}
+
+Here <{s}> contains the context and <{r}> is a Harmony value.
+It causes a thread with the context contained in <{s}> to be added to
+the state that has just executed the <{stop s}>
 \index{stop expression}%
@@
 \hline
-Context & (generated by <{stop}> expression) \\
+Context & (generated by <{stop}> or <{save}> expression) \\
 \hline
@@
 
-A context value captures the state of a thread.
+A context value (aka continuation) captures the state of a thread.
 A context is itself composed over various Harmony values.
+
+The following operations generate contexts:
+\begin{center}
+\begin{tabular}{|l|l|}
+\hline
+<{save e}> & returns a Harmony value (see below) \\
+\hline
+<{stop p}> & saves context in <{!p}> and stops the thread (see below) \\
+\hline
+\end{tabular}
+\end{center}
+
+The <{save e}> expression, when invoked, returns a tuple <{(e, c)}> where
+<{c}> is the context value of the thread right after the <{save}> instruction
+finished.
+Using <{go c r}> the context can be turned into a new thread that, instead,
+returns <{r}> from the <{save e}> expression.  See \autoref{fig:fork}
+for an example of how this could be used to \emph{fork} a thread into
+two threads.
+
+The <{stop p}> expression stores the context of the thread right after
+the expression in <{!p}> (i.e., <{p}> must be an address value) and
+terminates the thread.  The thread can later be reinstantiated with
+<{go !p r}>, in which case the <{stop}> expression returns <{r}>.
+A thread can be for ever suspended using <{stop None}>.
 
@@
 \end{comment}
-
-A thread can call <{get_context()}> to retrieve its current
-context.
 
@@
 
+\begin{figure}
+\begin{code}
+<{:inline:}>
+def fork():
+    atomically:
+        let (r, ctx) = save True:
+            result = r
+            if r:
+                go ctx (False, None)
+
+def main():
+    if fork():
+        print "parent"
+    else:
+        print "child"
+
+spawn eternal main()
+<{:end:}>
+\end{code}
+\caption{Using <{save}> and <{go}> to implement \texttt{fork()}}
+\label{fig:fork}
+\end{figure}
+
 The statement <{go c e}> starts a thread with context <{c}> that
-has executed a <{stop}> expression.  The <{stop}> expression
-returns value <{e}>.
+has executed a <{stop}> or <{save}> expression.
+The <{stop}> or <{save}> expression returns value <{e}>.
 The same context can be started multiple times, allowing threads to
-\emph{fork}.
+\emph{fork}.  See \autoref{fig:fork} for an example.
 

### 2022-01-28 book/preproc.py

@@ -80,6 +80,7 @@ reserved = {
     "or",
     "pass",
     "print",
+    "save",
     "select",
     "sequential",
     "setintlevel",

### 2022-01-28 src/charm/ops.c

@@ -1160,34 +1160,29 @@ void op_Stop(const void *env, struct state *state, struct step *step, struct glo
             free(p);
             return;
         }
+        step->ctx->stopped = true;
+        step->ctx->pc++;
+
         if (av == VALUE_ADDRESS) {
-            value_ctx_failure(step->ctx, &global->values, "Stop: address is None");
             return;
         }
 
+        hvalue_t v = value_put_context(&global->values, step->ctx);
         int size;
         hvalue_t *indices = value_get(av, &size);
         size /= sizeof(hvalue_t);
-
-        step->ctx->stopped = true;
-        step->ctx->pc++;
-        hvalue_t v = value_put_context(&global->values, step->ctx);
-
         if (!ind_trystore(state->vars, indices, size, v, &global->values, &state->vars)) {
             char *x = indices_string(indices, size);
             value_ctx_failure(step->ctx, &global->values, "Stop: bad address: %s", x);
             free(x);
-            return;
         }
     }
     else {
         step->ctx->stopped = true;
         step->ctx->pc++;
         hvalue_t v = value_put_context(&global->values, step->ctx);
-
         if (!ind_trystore(state->vars, es->indices, es->n, v, &global->values, &state->vars)) {
             value_ctx_failure(step->ctx, &global->values, "Store: bad variable");
-            return;
         }
     }
 }

### 2022-01-28 paper/paper.tex
More on continuations in the paper

@@
         if lk->held:
-            stop lk->blocked[len lk->blocked]
+            stop ?lk->blocked[len lk->blocked]
         else:
@@
 
-The Harmony <{stop x}> expression stops the current thread and saves
-its state in the variable <{x}>.  Another thread can resume the
-thread using the statement <{go x r}>, which causes the original
-thread to resume, with the <{stop x}> expression evaluating to the
-value <{r}>.  \autoref{fig:continuation} shows another implementation
+<{:var:p}>
+The Harmony <{save(e)}> expression returns a tuple containing <{e}>
+and the current context.
+Another thread can instantiate a context <{c}>
+using the statement <{go c r}>, with the context
+returning <{r}> from the <{save(e)}> expression.
+This is similar to the C \texttt{setjmp/longjmp} construct.
+Similar to <{save}>, the <{stop p}> expression saves the current
+context in <{!p}>, where <{p}> is the address value of a shared variable.
+Unlike <{save}>, the context is removed from the current set of
+threads.
+\autoref{fig:continuation} shows another implementation
 of the lock specification of \autoref{fig:synch}(a) using continuations,
 where blocked threads are placed on a FIFO queue and released in
-that order.  Besides synchronization and scheduling, continuations
-also enable the implementation of a primitive such as Posix
-\texttt{fork}.
+that order.  Besides synchronization, scheduling, and exception handling,
+continuations also enable the implementation of a primitive such as Posix
+\texttt{fork} or the implementation of co-routines.
 

### 2022-01-28 book/paper.tex
format
Updated book accordingly

@@
 
-Method <{wait}> adds the context of the thread to the bag.
-This increments the number of threads in the bag with the same context.
+Method <{wait}> adds the context of the thread---used as a unique
+identifier for the thread---to the bag,
+incrementing the number of threads in the bag with the same context.
+The Harmony <{save}> expression (\autoref{ap:context}) returns a tuple
+containing a value (in this case `<{()}>' and the context of the thread.
 <{wait}> then loops until that count is restored to the
@@
 \begin{code}
+<{:const:fork}>
 <{:inline:}>

### 2022-01-28 book/paper.tex
typo in book
Added OpSave to TLA+ spec

@@
 The Harmony <{save}> expression (\autoref{ap:context}) returns a tuple
-containing a value (in this case `<{()}>' and the context of the thread.
+containing a value (in this case `<{()}>') and the context of the thread.
 <{wait}> then loops until that count is restored to the
@@
 <{go !p r}>, in which case the <{stop}> expression returns <{r}>.
-A thread can be for ever suspended using <{stop None}>.
+A thread can be for ever suspended using <{stop None}> or just <{stop()}>.
 
